### 2022-04-01 README.Rmd
intial 2

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Madrid zonas de bajas emisiones (low emission zones)

The repository is home to the analysis of low emissions zones in Madrid using spatial availability measure.

Currently, the following raw data from Madrid is contained here:

-  "parque_vehiculos_2021"  includes vehicle fleet characteristics of Madrid City. Of interest: the district and the environmental label associated to each vehicle.

- several excel documents from the Home Mobility Survey of the Community of Madrid. These documents include information like citizens' socioeconomic characteristics, origin and destination of the trip, transport mode, distance, time required, activities carried out at origin and destination, etc.

- A figure of Madrid spatial context.

<!-- badges: start -->
<!-- badges: end -->

### 2022-04-02 README.Rmd
Added all raw data and began a data preparation .rmd.

@@
 
-Currently, the following raw data from Madrid is contained here:
+Currently, the data contained in this repository is from the [Madrid City Council](https://datos.madrid.es/portal/site/egob/menuitem.c05c1f754a33a9fbe4b2e4b284f1a5a0/?vgnextoid=128b5dda42c84410VgnVCM2000000c205a0aRCRD&vgnextchannel=374512b9ace9f310VgnVCM100000171f5a0aRCRD&vgnextfmt=default) and the [Community of Madrid](https://datos.comunidad.madrid/catalogo/dataset/resultados-edm2018) databases retrieved March 31st 2022.
+It includes the following:
 
--  "parque_vehiculos_2021"  includes vehicle fleet characteristics of Madrid City. Of interest: the district and the environmental label associated to each vehicle.
+-  "parque_vehiculos_2021", "2020", and "2019",  includes vehicle fleet characteristics of Madrid City. Of interest: the district and the environmental label associated to each vehicle.
 

### 2022-04-02 data-raw/00-OD-preparation.Rmd
Added all raw data and began a data preparation .rmd.

---
title: "00-OD-preparation"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

Here we load the libraries we will be using in this .Rmd file:
```{r}
library(tidyverse) 
library(readxl) #imports excel files
library(geojsonio) #converts geoJSON files to R classes and vice versa
library(sf) # sf= simple features! These are spatial objects in R, read here: https://r-spatial.github.io/sf/index.html
library(lubridate) #for time formating
```

## Import data

```{r}
OD_v <- read_excel("EDM2018VIAJES.xlsx") #here we import the trips file, by default this function imports the first sheet in an excel file. This .Rmd file is saved in the same place as the .xlsx file so we just need to write the name of the file we want to read. run ' help(read_excel) ` to see the documentation of this function.

summary(OD_v) #descriptive stats of each row
```

You can also view a few rows of the 'OD' table like this:
```{r}
head(OD_v, 12) #the first 12 rows
```

Or view it in RStudio:
```{r}
view(OD_v)
```

I'm interested in making the table spatial! It currently does not have coordinates but we can add from the shapefile which can be downloaded from the [Open Data Portal Madrid Regional Transport Consortium](https://datos.crtm.es/datasets/zonificacionzt1259/explore?location=40.420959%2C-3.695073%2C12.93):
```{r}
# here we download the madrid zona shapefile using the API GeoJSON resource! neat
zn <- geojson_read("https://opendata.arcgis.com/datasets/97f83bab03664d4e9853acf0e431d893_0.geojson",  what = "sp")
zn<- st_as_sf(zn) %>% st_transform(crs = 4326) #transform it into a 'sf' class object and transform coordinates to WGS84

#let's save it to data-raw just in case this link/resource doesn't work in the future
st_write(zn, "zn.shp")
```

## Filtering OD trips, home to work

```{r}
OD_v_Oc_Dt <- OD_v %>% filter(VORI == "1" & VDES == "2") #only filter in origin casa trips and destination trabajo

count(OD_v_Oc_Dt) # from the 222744 trips we now have 27,796 trips
```

What can we say about these work trips?

Well, the majority of trip frequency is "A diario, de lunes a viernes" (Daily, from Monday to Friday - 1)
```{r}
hist(OD_v_Oc_Dt$VFRECUENCIA)
```
The average destination arrival time (Destino. Hora llegada) is:
```{r}

timeD <- as.POSIXct(OD_v_Oc_Dt$VDESHORAFIN, format="%H%M") #formating the column
timeD <- format(timeD, "%H:%M") # removing date so we just how time

seconds_to_period(mean(period_to_seconds(hm(timeD)), na.rm=T)) # finding mean

```
9:48am arrival.. not very helpful.

Let's see the distribution:
```{r}
hist(OD_v_Oc_Dt$VDESHORAFIN %>% as.numeric(), breaks = 24)
```
So we see that 8AM, and 9AM arrival times are most frequent.

We also know that the majority do make this trip by personal car "¿Dispone de vehículo particular para hacer este viaje?" (1):
```{r}
hist(OD_v_Oc_Dt$VVEHICULO)
```
Try exploring the rest of the variables!

## Making OD home to work trips spatial

Here we add # of origin trips to the zona sf object, so we see where trips are coming from.

First, we count the number of trips coming from each origin:
```{r}
Otrips_zona <- OD_v_Oc_Dt %>% group_by(VORIZT1259)  %>% summarise(Otrips = (n())) # we group by the origin ZT1259 ID and then we summarize the number of rows that have the same ZT using the summarize() function and the n() function to create a new 'Otrips' column.

head(Otrips_zona,12)
```

Now, we add this new column to the zonas sf object:
```{r}
zn <- zn %>% merge(Otrips_zona, by.x = "ZT1259", by.y="VORIZT1259")

head(zn %>% st_drop_geometry(), 12) # notice the new 'Otrips' column
```

a simple plot:
```{r}
ggplot() +
  geom_sf(data = zn,
          aes(fill = Otrips),
          color  = NA) +
  scale_fill_distiller(palette = "Spectral")
```

### 2022-04-06 data-raw/00-OD-preparation.Rmd
meeting with javier

@@
 ```{r}
-Otrips_zona <- OD_v_Oc_Dt %>% group_by(VORIZT1259)  %>% summarise(Otrips = (n())) # we group by the origin ZT1259 ID and then we summarize the number of rows that have the same ZT using the summarize() function and the n() function to create a new 'Otrips' column.
+Dtrips_zona <- OD_v_Oc_Dt %>% group_by(VDESZT1259)  %>% summarise(Dtrips = (n())) # we group by the origin ZT1259 ID and then we summarize the number of rows that have the same ZT using the summarize() function and the n() function to create a new 'Otrips' column.
 
-head(Otrips_zona,12)
+head(Dtrips_zona,12)
 ```
@@
 ```{r}
-zn <- zn %>% merge(Otrips_zona, by.x = "ZT1259", by.y="VORIZT1259")
+zn <- zn %>% merge(Dtrips_zona, by.x = "ZT1259", by.y="VDESZT1259")
 
@@
   geom_sf(data = zn,
-          aes(fill = Otrips),
+          aes(fill = Dtrips),
           color  = NA) +

### 2022-04-07 00-OD-preparation.Rmd
edits to OD-preparation file and finished set-up for r5r car travel time file.

--- 2022-04-06 data-raw/00-OD-preparation.Rmd
@@
 ```{r}
-OD_v <- read_excel("EDM2018VIAJES.xlsx") #here we import the trips file, by default this function imports the first sheet in an excel file.
+OD_v <- read_excel("data-raw/EDM2018VIAJES.xlsx") #here we import the trips file, by default this function imports the first sheet in an excel file.
 This .Rmd file is saved in the same place as the .xlsx file so we just need to write the name of the file we want to read.
@@
 ```{r}
-view(OD_v)
+#view(OD_v)
 ```
@@
 ```{r}
-# here we download the madrid zona shapefile using the API GeoJSON resource! neat
-zn <- geojson_read("https://opendata.arcgis.com/datasets/97f83bab03664d4e9853acf0e431d893_0.geojson",  what = "sp")
-zn<- st_as_sf(zn) %>% st_transform(crs = 4326) #transform it into a 'sf' class object and transform coordinates to WGS84
+# # here we download the madrid zona shapefile using the API GeoJSON resource! neat
+# zn <- geojson_read("https://opendata.arcgis.com/datasets/97f83bab03664d4e9853acf0e431d893_0.geojson",  what = "sp")
+# zn<- st_as_sf(zn) %>% st_transform(crs = 4326) #transform it into a 'sf' class object and transform coordinates to WGS84
+# 
+# #let's save it to data-raw just in case this link/resource doesn't work in the future
+# st_write(zn, "data-raw/zn.shp")
 
-#let's save it to data-raw just in case this link/resource doesn't work in the future
-st_write(zn, "zn.shp")
+zn <- st_read("data-raw/zn.shp")
 ```
@@
 hist(OD_v_Oc_Dt$VVEHICULO)
+
+OD_v_Oc_Dt$VVEHICULO %>% typeof()
 ```
-Try exploring the rest of the variables!
+
+Let's see how long trips are (by distance):
+```{r}
+t <- OD_v_Oc_Dt %>% filter(DISTANCIA_VIAJE <= 60) %>% select(DISTANCIA_VIAJE) %>% unlist() %>% as.numeric() 
+hist(t)
+```
 
@@
 
-First, we count the number of trips coming from each origin:
+First, we count the number of trips coming to each destination:
 ```{r}
-Dtrips_zona <- OD_v_Oc_Dt %>% group_by(VDESZT1259)  %>% summarise(Dtrips = (n())) # we group by the origin ZT1259 ID and then we summarize the number of rows that have the same ZT using the summarize() function and the n() function to create a new 'Otrips' column.
+Dtrips_zona <- OD_v_Oc_Dt %>% group_by(VDESZT1259)  %>% summarise(Dtrips = (n())) # we group by the Destination ZT1259 ID and then we summarize the number of rows that have the same ZT using the summarize() function and the n() function to create a new 'Dtrips' column.
 
@@
 ```{r}
-zn <- zn %>% merge(Dtrips_zona, by.x = "ZT1259", by.y="VDESZT1259")
+zn_t <- zn %>% merge(Dtrips_zona, by.x = "ZT1259", by.y="VDESZT1259")
 
-head(zn %>% st_drop_geometry(), 12) # notice the new 'Otrips' column
+head(zn_t %>% st_drop_geometry(), 12) # notice the new 'Otrips' column
 ```
 
-a simple plot:
+a simple plot of the number of DESTINATION home-based trips:
 ```{r}
 ggplot() +
-  geom_sf(data = zn,
+  geom_sf(data = zn_t,
           aes(fill = Dtrips),
@@
 ```
+For comparision... lets plot the number of ORIGIN home-based trips:
+```{r}
+Otrips_zona <- OD_v_Oc_Dt %>% group_by(VORIZT1259)  %>% summarise(Otrips = (n())) # we group by the origin ZT1259 ID and then we summarize the number of rows that have the same ZT using the summarize() function and the n() function to create a new 'Otrips' column.
+
+zn_t <- zn_t %>% merge(Otrips_zona, by.x = "ZT1259", by.y="VORIZT1259", all.x = TRUE)
+
+ggplot() +
+  geom_sf(data = zn_t,
+          aes(fill = Otrips),
+          color  = NA) +
+  scale_fill_distiller(palette = "Spectral")
+```
+
+Let's compare these two patterns if we only include trips in which the destination is within the LEZ.
+
+<!-- to do once Javier adds the LEZ file to data-raw or replicate how the 'zn' object was imported-->
+
+## Saving data
+```{r}
+save(zn, file = "data/zn.RData")
+```

### 2022-04-07 01-Travel-time-calculation.Rmd
edits to OD-preparation file and finished set-up for r5r car travel time file.

---
title: "01-Travel-time-calculation"
output: html_document
---
In this notebook, we take the centroids of all zones in Madrid and calculate travel time using car, transit, walk, and cycle.

Travel modes:
1) Car
2) Walk
3) Bike
4) Transit bus

# --- Libraries and input data
```{r setup, include=FALSE}
library(disk.frame)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(httr)
library(knitr)
#install.packages("kableExtra", dependencies = TRUE) #installing kableExtra and dependencies as I was recieving errors
library(kableExtra)
#install.packages("devtools")
#install.packages("Rtools")
#devtools::install_github("https://github.com/hrbrmstr/lodes.git", dependencies = TRUE) # if you need to download
library(lodes)
library(progress)
library(purrr)
library(r5r)
library(sf)
library(stplanr)
library(tidycensus)
library(tidyr)
library(tmap)
library(zoo) # for rollapplyr

# setup for disk.frame
setup_disk.frame()
options(scipen = 999)
options(java.parameters = "-Xmx6G")
options(future.globals.maxSize = Inf)
tmap_mode("view")
```

```{r setup, include=FALSE}
load(file = "data/zn.RData")
```

# --- Formating the data
```{r}

```

Format the origin and destination points (ID, Long, Lat). In this case, all origins and all destinations are the same (i.e., 1259 zones so 1259 origins and 1259 destinations, so 1,585,081 OD travel times from each origin to destination):
```{r}
# # I was having an error running st_centroid, this line fixes it.
# sf::sf_use_s2(FALSE)

#convert zn polygons into centroids
OD_all <- zn %>% st_transform(crs = 2062) %>% st_centroid()

#plot to double check that centroids are on zones
ggplot() +
  geom_sf(data = zn) +
  geom_sf(data = OD_all)
```
```{r}
#transfer coordinates located in geometry into X and Y columns, transform coordinate into crs 4326i.e. WSG 1984 (the input for r5r)
OD_all <- cbind(as.data.frame(OD_all), 
                st_coordinates(st_transform(OD_all, crs = 4326))) %>%
  rename(lon = "X", lat = "Y", id = "FID") %>%
  select(id, lon, lat)
```

#--- Download Spain OSM Network and Madrid GTFS Data

```{r set up r5 path, include=FALSE}
# the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html
dir.create("data-raw/intermed/r5_graph")
r5_path <- file.path("data-raw/intermed/r5_graph")
```

```{r download data, include=FALSE, eval=FALSE}
# downloading spain osm in the correct format
download.file(url = paste0("http://download.geofabrik.de/europe/spain/madrid-latest.osm.pbf"),
              destfile = file.path(r5_path, "osm.pbf"),
              mode = "wb")

```

```{r}
#download and import GTFS (Madrid transit data)
# download.file(url = "https://transitfeeds.com/p/emt-madrid/212/latest/download", destfile = file.path(r5_path, "MAD_transit.zip"), mode = "wb")

```

# Set Up R5 Routing

```{r build graph, include = FALSE}
r5_MAD <- setup_r5(data_path = r5_path, verbose = FALSE)
```
```{r}
# stop_r5(r5_HAM)
# rJava::.jgc(R.gc = TRUE)
```

# Calculate OD Matrices

## Car travel time
```{r}
# set up batching according to how many origin rows to process at one time
chunksize = 50 
num_chunks = ceiling(nrow(OD_all)/chunksize)

# create origin-destination pairs
origins_chunks <- as.disk.frame(OD_all,
                          outdir = "data-raw/intermed/df/Orig",
                          nchunks = num_chunks,
                          overwrite = TRUE)

start.time <- Sys.time()
pb <- txtProgressBar(0, num_chunks, style = 3)

for (i in 1:num_chunks){
  Orig_chunk <- get_chunk(origins_chunks, i)
  ttm_chunk <- travel_time_matrix(r5r_core = r5_MAD,
                          origins = Orig_chunk,
                          destinations = OD_all,
                          mode = c("CAR"), 
                          departure_datetime = as.POSIXct(strptime("2022-03-30 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "UTC+2")),
                          max_trip_duration = 120)

  # export output as disk.frame
  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
                                            nchunks = 1,
                                            outdir = "data-raw/intermed/df/output_ttm",
                                            compress = 50,
                                            overwrite = TRUE),
         add_chunk(output_df, ttm_chunk, chunk_id = i))
  setTxtProgressBar(pb, i)
}
end.time <- Sys.time()
print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))

output_OD_car <- as.data.frame(output_df)

save("output_OD_car", file = "data-raw/intermed/ttresults/output_OD_car.Rdata")
#I only made the cut off 60 minutes, so if some parcels do not have a travel time, rerun this for a 120 min cut off. The following are 120mins cutoff
```

### 2022-04-08 00-OD-preparation.Rmd
Calculated travel times for car, walk, bike, bus for all OD combinations in Madrid. Completed some preliminary comparisons to survey travel times.

@@
 ```
+Importing the spatial object which corresponds to the low-emission-zones
+```{r}
+#XXX
+```
 
-## Filtering OD trips, home to work
+## Organizing the trip level data
+
+```{r}
+#adding travel time... we don't even need r5r! we just need to decide how to remove A polluting cars
+OD_v <- OD_v %>% mutate(
+  TIEMPO_VIAJES = 
+                  as.numeric((as.POSIXct(OD_v$VDESHORAFIN, format="%H%M") -
+                  as.POSIXct(OD_v$VORIHORAINI, format="%H%M"))/60))
+
+```
+
+## Exploring the data - filtering OD trips, home to work
 
@@
 ```{r}
-
-timeD <- as.POSIXct(OD_v_Oc_Dt$VDESHORAFIN, format="%H%M") #formating the column
-timeD <- format(timeD, "%H:%M") # removing date so we just how time
-
-seconds_to_period(mean(period_to_seconds(hm(timeD)), na.rm=T)) # finding mean
-
+Dtrips <- as.POSIXct(OD_v$VDESHORAFIN, format="%H%M") %>% format("%H:%M") # formatting and removing date so we just how time
+seconds_to_period(mean(period_to_seconds(hm(Dtrips)), na.rm=T)) # finding mean
 ```
@@
 ```{r}
-hist(OD_v_Oc_Dt$VDESHORAFIN %>% as.numeric(), breaks = 24)
+hist(OD_v_Oc_Dt$VDESHORAFIN%>% as.numeric(), breaks = 24)
 ```
@@
 
-Let's see how long trips are (by distance):
+Let's see how long are short-trips are (by distance, assuming in km units):
 ```{r}
@@
 hist(t)
+```
+
+As you can see though there are some long distance trip (greater than 60km).
+There are 101 trips which are between 60km to 600km and 1 trip greater than 600km.
+For perspective though, these trips make up onl 0.3% of total trips taken ( ~100 trips out of ~28,000 trips).
+```{r}
+t <- OD_v_Oc_Dt %>% filter(DISTANCIA_VIAJE >= 60 & DISTANCIA_VIAJE <= 600) %>% select(DISTANCIA_VIAJE) %>% unlist() %>% as.numeric() 
+hist(t, breaks = 50)
+
+t <- OD_v_Oc_Dt %>% filter(DISTANCIA_VIAJE >= 600) %>% select(DISTANCIA_VIAJE) %>% unlist() %>% as.numeric() 
+hist(t, breaks = 50)
+```
+
+Let's check the primary mode: 
+```{r}
+hist(OD_v_Oc_Dt$MODO_PRIORITARIO)
 ```
@@
 save(zn, file = "data/zn.RData")
+save(OD_v, file = "data/OD_v.RData")
 ```

### 2022-04-08 01-Travel-time-calculation.Rmd
Calculated travel times for car, walk, bike, bus for all OD combinations in Madrid. Completed some preliminary comparisons to survey travel times.

@@
 # --- Formating the data
-```{r}
-
-```
-
 Format the origin and destination points (ID, Long, Lat).
@@
 #download and import GTFS (Madrid transit data)
-# download.file(url = "https://transitfeeds.com/p/emt-madrid/212/latest/download", destfile = file.path(r5_path, "MAD_transit.zip"), mode = "wb")
+download.file(url = "https://transitfeeds.com/p/emt-madrid/212/latest/download", destfile = file.path(r5_path, "MAD_transit.zip"), mode = "wb")
 
@@
                           mode = c("CAR"), 
-                          departure_datetime = as.POSIXct(strptime("2022-03-30 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "UTC+2")),
+                          departure_datetime = as.POSIXct(strptime("2022-03-30 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "CEST")),
                           max_trip_duration = 120)
@@
 ```
+
+## Transit travel time
+```{r}
+# set up batching according to how many origin rows to process at one time
+chunksize = 50 
+num_chunks = ceiling(nrow(OD_all)/chunksize)
+
+# create origin-destination pairs
+origins_chunks <- as.disk.frame(OD_all,
+                          outdir = "data-raw/intermed/df/Orig",
+                          nchunks = num_chunks,
+                          overwrite = TRUE)
+
+start.time <- Sys.time()
+pb <- txtProgressBar(0, num_chunks, style = 3)
+
+for (i in 1:num_chunks){
+  Orig_chunk <- get_chunk(origins_chunks, i)
+  ttm_chunk <- travel_time_matrix(r5r_core = r5_MAD,
+                          origins = Orig_chunk,
+                          destinations = OD_all,
+                          mode = c("TRANSIT", "WALK"),
+                          departure_datetime = as.POSIXct(strptime("2022-03-30 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "CEST")),
+                          max_trip_duration = 120)
+
+  # export output as disk.frame
+  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
+                                            nchunks = 1,
+                                            outdir = "data-raw/intermed/df/output_ttm",
+                                            compress = 50,
+                                            overwrite = TRUE),
+         add_chunk(output_df, ttm_chunk, chunk_id = i))
+  setTxtProgressBar(pb, i)
+}
+end.time <- Sys.time()
+print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))
+
+output_OD_transit <- as.data.frame(output_df)
+
+save("output_OD_transit", file = "data-raw/intermed/ttresults/output_OD_transit.Rdata")
+#I only made the cut off 60 minutes, so if some parcels do not have a travel time, rerun this for a 120 min cut off. The following are 120mins cutoff
+```
+
+## Walk travel time
+```{r}
+# set up batching according to how many origin rows to process at one time
+chunksize = 50 
+num_chunks = ceiling(nrow(OD_all)/chunksize)
+
+# create origin-destination pairs
+origins_chunks <- as.disk.frame(OD_all,
+                          outdir = "data-raw/intermed/df/Orig",
+                          nchunks = num_chunks,
+                          overwrite = TRUE)
+
+start.time <- Sys.time()
+pb <- txtProgressBar(0, num_chunks, style = 3)
+
+for (i in 1:num_chunks){
+  Orig_chunk <- get_chunk(origins_chunks, i)
+  ttm_chunk <- travel_time_matrix(r5r_core = r5_MAD,
+                          origins = Orig_chunk,
+                          destinations = OD_all,
+                          mode = c("WALK"),
+                          departure_datetime = as.POSIXct(strptime("2022-03-30 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "CEST")),
+                          max_trip_duration = 60)
+
+  # export output as disk.frame
+  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
+                                            nchunks = 1,
+                                            outdir = "data-raw/intermed/df/output_ttm",
+                                            compress = 50,
+                                            overwrite = TRUE),
+         add_chunk(output_df, ttm_chunk, chunk_id = i))
+  setTxtProgressBar(pb, i)
+}
+end.time <- Sys.time()
+print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))
+
+output_OD_walk <- as.data.frame(output_df)
+
+save("output_OD_walk", file = "data-raw/intermed/ttresults/output_OD_walk.Rdata")
+```
+
+## Bike travel time
+```{r}
+# set up batching according to how many origin rows to process at one time
+chunksize = 50 
+num_chunks = ceiling(nrow(OD_all)/chunksize)
+
+# create origin-destination pairs
+origins_chunks <- as.disk.frame(OD_all,
+                          outdir = "data-raw/intermed/df/Orig",
+                          nchunks = num_chunks,
+                          overwrite = TRUE)
+
+start.time <- Sys.time()
+pb <- txtProgressBar(0, num_chunks, style = 3)
+
+for (i in 1:num_chunks){
+  Orig_chunk <- get_chunk(origins_chunks, i)
+  ttm_chunk <- travel_time_matrix(r5r_core = r5_MAD,
+                          origins = Orig_chunk,
+                          destinations = OD_all,
+                          mode = c("BICYCLE"),
+                          departure_datetime = as.POSIXct(strptime("2022-03-30 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "CEST")),
+                          max_trip_duration = 60)
+
+  # export output as disk.frame
+  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
+                                            nchunks = 1,
+                                            outdir = "data-raw/intermed/df/output_ttm",
+                                            compress = 50,
+                                            overwrite = TRUE),
+         add_chunk(output_df, ttm_chunk, chunk_id = i))
+  setTxtProgressBar(pb, i)
+}
+end.time <- Sys.time()
+print(paste0("OD matrix calculation took ", round(difftime(end.time, start.time, units = "mins"), digits = 2), " minutes..."))
+
+output_OD_bike <- as.data.frame(output_df)
+
+save("output_OD_bike", file = "data-raw/intermed/ttresults/output_OD_bike.Rdata")
+```

### 2022-04-08 02-Data-package-preparation.Rmd
Calculated travel times for car, walk, bike, bus for all OD combinations in Madrid. Completed some preliminary comparisons to survey travel times.

---
title: "02-Data-package-preparation"
output: html_document
---

load the libraries we will be using in this .Rmd file:
```{r}
library(tidyverse) 
library(sf) # sf= simple features! These are spatial objects in R, read here: https://r-spatial.github.io/sf/index.html
library(ggplot2)
```

load car travel time data, the trip level data, and the zones
```{r}
load(file = "data-raw/intermed/ttresults/output_OD_car.Rdata")
load(file = "data/OD_v.RData")
load(file = "data/zn.RData")
```

let's compare *car* travel times to the travel times we actually have in the survey. To do this we first need to add the origin and dest ZN ids to the travel time object. Then we will join them to the CAR trips.

So first, add ZN ids:
```{r}
zn_t <- zn %>% st_drop_geometry() %>% transmute(FID, 
                                                OZT1259 = ZT1259)

OD_car <- output_OD_car %>% merge(zn_t, by.x=c("fromId"), by.y = c("FID"))

zn_t <- zn %>% st_drop_geometry() %>% transmute(FID,
                                                DZT1259 = ZT1259)

OD_car <- OD_car %>% merge(zn_t, by.x=c("fromId"), by.y = c("FID"))

OD_car <- OD_car %>% select(OZT1259, DZT1259, travel_time)
```

Now merge the calculated travel time to the trip data - ONLY car trips though (so codes 11, 12, 13, 14, 15,16) and ONLY home to work trips.
```{r}
#filter the data... now we only have ~2400 trips left out of the full data set
OD_v_car_hometowork <- OD_v %>% filter(MODO_PRIORITARIO == (10| 11 | 12 | 13 | 14 | 15 | 16)) %>% filter(VORI == "1" & VDES == "2")

OD_v_car_hometowork <- OD_v_car_hometowork  %>% merge(OD_car, by.x=c("VORIZT1259", "VDESZT1259"), by.y = c("OZT1259", "DZT1259"), all.x=TRUE)

```

Let's plot the distribution of r5r car travel times and the survey travel times
```{r}
# make data long
times1 <- OD_v_car_hometowork %>% transmute(tt = TIEMPO_VIAJES, 
                                            L = "survey")
times2 <- OD_v_car_hometowork %>% transmute(tt = travel_time, 
                                            L = "calculated")

ctimes <- rbind(times1, times2)

ggplot(data=ctimes, aes(x=tt, group=L, col=L)) +
    geom_density(adjust=1.5, alpha=.4) 
```
So.. interesting, you can see a few things

1) there are some longer than 120 car trips (I only calculated up to 120 mins). 
2) the r5r estimation (red line) have many more trips ~30 min while the survey has more at the ~10 min and ~45 min lengths.

### 2022-04-14 00-OD-preparation.Rmd
added column to identify which zone is a LEZ in 00-data-prep

@@
 ```{r}
-#XXX
+zn_LEZ <- st_read("data-raw/LEZ.shp") %>% transmute(ZT1259, 
+                                                                       LEZ = 1)
+```
+
+Add a column which indicates if a zn is within the center of madrid:
+```{r}
+zn <- zn %>% merge(zn_city %>% st_drop_geometry(), by = c("ZT1259"), all.x=TRUE)
+```
+
+plot and color in madrid city:
+```{r}
+#plot to double check that centroids are on zones
+ggplot() +
+  geom_sf(data = zn, aes(fill = as.factor(city)), col = NA) +
+  scale_fill_manual(values = c("red", "grey"), name= "Cluster Group")
+```
+```{r}
+#plot to double check that centroids are on zones
+ggplot() +
+  geom_sf(data = zn_city,) 
 ```

### 2022-04-18 00-OD-preparation.Rmd
javier s first commit - changes to OD-prep and adding silly rmd

--- 2022-04-08 00-OD-preparation.Rmd
@@
 ---
+
+fsdsjhdskjds
+```{r}
+
+```
 
@@
 ```{r}
-#XXX
+zn_LEZ <- st_read("data-raw/LEZ.shp") %>% transmute(ZT1259, 
+                                                                       LEZ = 1)
+```
+
+Add a column which indicates if a zn is within the center of madrid:
+```{r}
+zn <- zn %>% merge(zn_LEZ %>% st_drop_geometry(), by = c("ZT1259"), all.x=TRUE)
+```
+
+plot and color in madrid city:
+```{r}
+#plot to double check that centroids are on zones
+ggplot() +
+  geom_sf(data = zn, aes(fill = as.factor(LEZ)), col = NA) +
+  scale_fill_manual(values = c("red", "grey"), name= "Cluster Group")
+```
+```{r}
+#plot to double check that centroids are on zones
+ggplot() +
+  geom_sf(data = zn_LEZ) 
 ```
@@
                   as.POSIXct(OD_v$VORIHORAINI, format="%H%M"))/60))
+
+summary(OD_v$TIEMPO_VIAJES)
 

### 2022-04-18 todelete.Rmd
javier s first commit - changes to OD-prep and adding silly rmd

---
title: "test_todelete"
output: html_document
date: '2022-04-18'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2022-04-22 00-OD-preparation.Rmd
Added visuals for Madrid Region and LEZ analysis in 00-OD-Prep.

--- 2022-04-08 00-OD-preparation.Rmd
@@
 library(lubridate) #for time formating
+library(ggspatial)
 ```
@@
 I'm interested in making the table spatial!
-It currently does not have coordinates but we can add from the shapefile which can be downloaded from the [Open Data Portal Madrid Regional Transport Consortium](https://datos.crtm.es/datasets/zonificacionzt1259/explore?location=40.420959%2C-3.695073%2C12.93):
+It currently does not have coordinates but we can add from the Madrid Region shapefile which can be downloaded from the [Open Data Portal Madrid Regional Transport Consortium](https://datos.crtm.es/datasets/zonificacionzt1259/explore?location=40.420959%2C-3.695073%2C12.93):
 ```{r}
@@
 ```
-Importing the spatial object which corresponds to the low-emission-zones
+
+Importing a list of all the municipalities within the Madrid region and adding them to the zone shape file:
 ```{r}
-#XXX
+t <- read_excel("data-raw/ZonificacionZT1259 complete.xlsx") %>% select(ZT1259, Municipality)
+
+zn <- zn %>% merge(t, by = c("ZT1259"), all.x=TRUE)
+```
+
+Next, we will add a column which indicates if a zone is in a LEZ or is located within Madrid City Centre (this area contains the LEZ and is contained within the Municipality of Madrid):
+```{r}
+#IDS that correspond to CC - city center and LEZ - low emission zone
+CC_IDS <- c("079-01", "079-02", "079-03", "079-04", "079-05", "079-06", "079-07")
+LEZ_IDS <- c("079-01-001", "079-01-002", "079-01-003", "079-01-004", "079-01-005", "079-01-006", "079-01-007", "079-01-008", "079-01-009", "079-01-010", "079-01-012", "079-01-013", "079-01-014", "079-01-015", "079-01-016", "079-01-017", "079-01-018", "079-01-019","079-01-020", "079-01-021", "079-01-022", "079-01-023")
+
+#manually added column to zn which identifies if an ID is in the city center (1) or not (0). This searches for any instance of "079-01-XXX", "079-02-XXX" etc.
+zn$CC <- ifelse(grepl(paste(CC_IDS, collapse = "|"), zn$ZT1259), 1,0)
+
+#more simple to use the mutate function here since the string of IDs is complete for LEZ
+zn <- zn %>% mutate(LEZ = ifelse(ZT1259 %in% LEZ_IDS, 1, 0))
+
+```
+
+create a bounding box for the plot:
+```{r}
+#create the bounding box for the Toronto region + a little extra
+bbox_new1 <- st_bbox(zn %>% filter(CC == 1)) # current bounding box
+
+xrange <- bbox_new1$xmax - bbox_new1$xmin # range of x values
+yrange <- bbox_new1$ymax - bbox_new1$ymin # range of y values
+
+bbox_new1[1] <- bbox_new1[1] - (0.2 * xrange) # xmin - left
+bbox_new1[3] <- bbox_new1[3] + (0.2 * xrange) # xmax - right
+bbox_new1[2] <- bbox_new1[2] - (0.05 * yrange) # ymin - bottom
+bbox_new1[4] <- bbox_new1[4] + (0.05 * yrange) # ymax - top
+
+bbox_CityCentre <- bbox_new1 %>%  # take the bounding box ...
+  st_as_sfc()
+```
+
+plot:
+```{r}
+plot_rmad <- ggplot() +
+  geom_sf(data = zn, fill = "light grey", col = "dark grey", size=0.5) + #all of madrid region
+  geom_sf(data = zn %>% filter(Municipality == "Madrid"), aes(fill = "Municipality of Madrid"), col = "yellow3", size=0.5, show.legend = "polygon") + # madrid municiality
+  geom_sf(data = zn %>% filter(CC == 1), aes(fill =  "City centre"), col = NA, show.legend = "polygon") + # madrid city centre
+  geom_sf(data = zn %>%  filter(LEZ == 1), aes(fill = "Low emission zone"), col = NA, show.legend = "polygon") + #madrid low emission zone
+  geom_sf(data = bbox_CityCentre, fill = NA, col = "black", size = 0.5)+
+  scale_fill_manual(labels = c("City centre", "Low emission zone", "Municipality of Madrid"),
+                      values = c("green", "red", "yellow")) + 
+  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
+                         height = unit(0.8, "cm"), 
+                         width = unit(0.8, "cm"),
+                         style = north_arrow_orienteering(line_width = 0.25,
+                                                          line_col = "dimgrey", 
+                                                          fill = c("white","dimgrey"))) +
+  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
+                   height = unit(0.15, "cm")) +
+  scale_y_continuous(breaks = c(40, 40.4, 40.8, 41.2)) + 
+  scale_x_continuous(breaks = c(-3.0, -3.8, -4.6)) +
+  guides(fill=guide_legend(title=NULL)) +
+  theme_minimal() + theme(legend.position = c(0.25, .16))
+
+#just the city center and LEZ
+plot_ccmad <- ggplot() +
+  geom_sf(data = zn %>% filter(CC == 1), aes(fill = "City centre"), col = "green3") + # madrid city centre
+  geom_sf(data = zn %>%  filter(LEZ == 1), aes(fill = "Low emission zone"), col = "red4") + #madrid low emission zone
+  geom_sf(data = bbox_CityCentre, fill = NA, col = "black", size = 0.5)+
+  scale_fill_manual(labels = c("City centre", "Low emission zone"),
+                    values = c("green", "red")) + 
+  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
+                   height = unit(0.15, "cm"),
+                   pad_y = unit(0.05, "cm")) +
+  guides(fill=guide_legend(title=NULL)) +
+  theme_void() + theme(legend.position = "none")
+
+cowplot::plot_grid(plot_rmad, plot_ccmad, ncol = 2, rel_widths = c(2.5,1)) 
 ```
@@
 
+MODO_PRIORITARIO -> 11, 12, 13, 14, 15, 16 correspond to:
+11 '11. Private car - driver
+12 '12. Company car - driver
+13 '13. Rental car without driver
+14 '14. Private car - passenger
+15 '15. Company car- passenger
+16 '16. Car rental with driver - passenger
+
 ## Making OD home to work trips spatial
@@
 ```
-For comparision... lets plot the number of ORIGIN home-based trips:
+
+For comparison... lets plot the number of ORIGIN home-based trips:
 ```{r}
@@
 ```
+How many trips come into the LEZ from outside of the LEZ and how many of those are car trips? Just curious.
 
-Let's compare these two patterns if we only include trips in which the destination is within the LEZ.
+First, let's add a column which correspond to if the trip is made by car or not my car.
+```{r}
+OD_v_Oc_Dt <- OD_v_Oc_Dt %>% mutate(bycar = ifelse(MODO_PRIORITARIO %in% c(11, 12, 13, 14, 15, 16),
+                                                                         1, 0),
+                                    nocar = ifelse(!(MODO_PRIORITARIO %in% c(11, 12, 13, 14, 15, 16)),
+                                                                         1, 0))
+```
 
-<!-- to do once Javier adds the LEZ file to data-raw or replicate how the 'zn' object was imported-->
+Next, let's split the data - one subset of all trips that don't go into the LEZ, and another subset of trips that DO go into the LEZ
+```{r}
+OD_v_Oc_Dt_noDesLEZ <- OD_v_Oc_Dt %>% filter(!(VDESZT1259 %in% LEZ_IDS) | (VORIZT1259 %in% LEZ_IDS))
+OD_v_Oc_Dt_DesLEZ <- OD_v_Oc_Dt %>% filter((VDESZT1259 %in% LEZ_IDS) & !(VORIZT1259 %in% LEZ_IDS))
+
+count(OD_v_Oc_Dt_DesLEZ)/count(OD_v_Oc_Dt)*100
+count(OD_v_Oc_Dt_noDesLEZ) + count(OD_v_Oc_Dt_DesLEZ) ==  count(OD_v_Oc_Dt)
+```
+we can see that trips into LEZ are a small proportion - only 5.84% of total home-to-work trips in the madrid region have a destination in the LEZ (and do not originate in the LEZ).
+Now...
+how many of these are car trips?
+
+```{r}
+sum(OD_v_Oc_Dt_DesLEZ$bycar)/count(OD_v_Oc_Dt_DesLEZ)
+```
+only 19% of these LEZ trips are made by car! This may not have a huge GHG emission savings... but perhaps it has a big impact on equity?
+
+Let's plot where these car LEZ trips are coming from:
+```{r}
+D2trips_zona <- OD_v_Oc_Dt_DesLEZ %>% group_by(VORIZT1259)  %>% summarise(Dtrips = (n()),
+                                                                              bycar = sum(bycar),
+                                                                              nocar = sum(nocar))
+
+zn_t <- zn %>% merge(D2trips_zona, by.x = "ZT1259", by.y="VORIZT1259")
+```
+
+plots:
+```{r}
+# % of trips made by car
+ggplot() +
+  geom_sf(data = zn_t,
+          aes(fill = (bycar/Dtrips * 100) ),
+          color  = NA) +
+  scale_fill_distiller(palette = "Spectral") +
+  labs(title = "% work trips by origin (excluding those that originate in LEZ) made by car ")
+
+# magnitude of trips made by car 
+ggplot() +
+  geom_sf(data = zn_t,
+          aes(fill = (bycar) ),
+          color  = NA) +
+  scale_fill_distiller(palette = "Spectral") +
+  labs(title = "number of work trips by origin (excluding those that originate in LEZ) made by car ")
+```
+
+% of car trips which go into the LEZ 
+```{r}
+# Mag
+ggplot() +
+  geom_sf(data = zn_t,
+          aes(fill = (bycar/Dtrips * 100) ),
+          color  = NA) +
+  scale_fill_distiller(palette = "Spectral") +
+  labs(title = "% work trips (excluding those that originate in LEZ) made by car ")
+```
+
+```{r}
+ggplot() +
+  geom_sf(data = zn %>% filter(CC == 1), aes(fill = "City centre"), col = "green3") + # madrid city centre
+  geom_sf(data = zn %>%  filter(LEZ == 1), aes(fill = "Low emission zone"), col = "red4") + #madrid low emission zone
+  geom_sf(data = bbox_CityCentre, fill = NA, col = "black", size = 0.5)+
+  scale_fill_manual(labels = c("City centre", "Low emission zone"),
+                    values = c("green", "red")) + 
+  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
+                   height = unit(0.15, "cm"),
+                   pad_y = unit(0.05, "cm")) +
+  guides(fill=guide_legend(title=NULL)) +
+  theme_void() + theme(legend.position = "none"
+```
 

### 2022-05-02 00-OD-preparation.Rmd
Completed impedance functions in 'draft-manuscript'. This .Rmd will eventually be moved to a new repository dedicated to manuscript writing when this MADRID-LEZ data package is complete.

@@
 zn <- st_read("data-raw/zn.shp")
+ALLLEZ <- st_read("data-raw/Zona_Bajas_Emisiones_MADRID.shp")
 ```
@@
 cowplot::plot_grid(plot_rmad, plot_ccmad, ncol = 2, rel_widths = c(2.5,1)) 
+```
+
+```{r}
+
+ggplot() +
+  geom_sf(data = zn) +
+  geom_sf(data = ALLLEZ, fill="red") 
 ```
@@
 
-```{r}
-ggplot() +
-  geom_sf(data = zn %>% filter(CC == 1), aes(fill = "City centre"), col = "green3") + # madrid city centre
-  geom_sf(data = zn %>%  filter(LEZ == 1), aes(fill = "Low emission zone"), col = "red4") + #madrid low emission zone
-  geom_sf(data = bbox_CityCentre, fill = NA, col = "black", size = 0.5)+
-  scale_fill_manual(labels = c("City centre", "Low emission zone"),
-                    values = c("green", "red")) + 
-  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
-                   height = unit(0.15, "cm"),
-                   pad_y = unit(0.05, "cm")) +
-  guides(fill=guide_legend(title=NULL)) +
-  theme_void() + theme(legend.position = "none"
-```
-
 ## Saving data

### 2022-05-02 Draft-manuscript-tobemoved.Rmd
Completed impedance functions in 'draft-manuscript'. This .Rmd will eventually be moved to a new repository dedicated to manuscript writing when this MADRID-LEZ data package is complete.

---
title: "Draft-manuscript-tobemoved"
output: html_document
---

```{r knitr-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  cache = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = '', 
  out.width = "1\\linewidth")
```

```{r install-data-package, eval = FALSE, include=FALSE}
# if (!require("MadridZBE", character.only = TRUE)) {
#       remotes::install_github("soukhova/MadridZBE",
#                         build_vignettes = TRUE)
#   }
```

```{r load-packages, include=FALSE, cache=FALSE}
library(TTS2016R)
library(dplyr)
library(fitdistrplus)
library(ggplot2)
library(kableExtra)
library(patchwork)
library(sf)
library(scales)
library(ggpmisc)
library(ggrepel)
library(cowplot)
library(ggspatial)
library(spdep)
library(RColorBrewer)
library(tmap)
library(shadowtext)
library(grid)
library(readxl) #imports excel files
# library(extrafont)
# font_import()
# loadfonts(device = "win")

options(scipen = 999)
```

```{r sp_avail-function,include=FALSE}
#defining the spatial availability function
sp_avail <- function(x, o_id, d_id, pop, opp, r, f, alpha = 1){

  o_id <- rlang::enquo(o_id)
  d_id <- rlang::enquo(d_id)
  pop <- rlang::enquo(pop)
  opp <- rlang::enquo(opp)
  r <- rlang::enquo(r)
  f <- rlang::enquo(f)

  sum_pop <- x %>%
    dplyr::distinct(!!o_id,
                    .keep_all = TRUE) %>%
    dplyr::mutate(sum_pop = !!r*(!!pop)^alpha) %>%
    dplyr::pull(sum_pop) %>%
    sum()

  f_p <- dplyr::pull(x, !!r) * dplyr::pull(x, !!pop)^alpha / sum_pop

  sum_impedance <- x %>%
    dplyr::group_by(!!d_id) %>%
    dplyr::summarize(sum_impedance = sum(!!f))

  x <- x %>%
    dplyr::left_join(sum_impedance,
                     by = rlang::as_name(d_id))

  f_c <- dplyr::pull(x, !!f) / x$sum_impedance

  x$f_c <- f_c
  x$f_p <- f_p

  sum_pa <- x %>%
    dplyr::group_by(!!d_id) %>%
    dplyr::summarize(sum_pa= sum(f_p * f_c))

  x <- x %>%
    dplyr::left_join(sum_pa,
                     by = rlang::as_name(d_id))
  f_t <- (f_p * f_c) / dplyr::pull(x, sum_pa)

  dplyr::pull(x, !!opp) * f_t
}
```

\newpage

# Introduction

- Some information about LEZ generally

- Information about accessibility issues regarding LEZ; the tradition of using accessibility to evaluate these differences. Mention of spatial availability.

- What we will be doing, the aim.

\newpage

# Using accessibility to evaluate inequity background

- Information about inequity and LEZ may exasperate these issues

- lit has found that emissions reduciton should not result in inequalities..

\newpage

# Methods

## Data 
- -plots of madrid zone, description of the LEZ scheme, how effective it is now and will be in the future - change in modal shares and emissions but what about equity

## Estimated data
- prediction in change in travel mode
- prediction in private vehicle fleet label A/B/C proportions

## Spatial availability
- spatial availability explanation 
  - impedance function explanation

\newpage

# Results

## Fleet characteristics today vs. 2025
<!-- to do -->

## Impedance function calibration

```{r data-import-filter}
OD_v <- read_excel("data-raw/EDM2018VIAJES.xlsx")

OD_v_HW <- OD_v %>% filter(VORI == "1" & VDES == "2") #only filter in origin casa trips and destination trabajo

OD_v_HW <- OD_v_HW %>% 
  mutate(TIEMPO_VIAJES = 
                  as.numeric((as.POSIXct(OD_v_HW$VDESHORAFIN, format="%H%M") -
                  as.POSIXct(OD_v_HW$VORIHORAINI, format="%H%M"))/60)) %>%
  mutate(TIEMPO_VIAJES = ifelse(VDESHORAFIN == "2410", 40, 
                                         ifelse(VDESHORAFIN == "2440", 55, 
                                                ifelse(VDESHORAFIN == "2800", 420, TIEMPO_VIAJES)))) %>% # some destination times are greater than 2400 and so yield an NA different, correct these 3 trips
  mutate(TIEMPO_VIAJES = ifelse(TIEMPO_VIAJES == 0, 0.1, TIEMPO_VIAJES)) # set all 0min travel times to 0.1 min

summary(OD_v_HW$TIEMPO_VIAJES)
```

```{r data-for-impedance}
#here we separate all the "car" trips, "transit" trips, "bike" trips, and "walk" trips. We then will create 4 different imepedance function....
bus_HW_alltt <- OD_v_HW  %>% filter(MODO_PRIORITARIO >= 9) %>%
  dplyr::select(TIEMPO_VIAJES)

car_motor_HW_alltt <- OD_v_HW  %>% filter(MODO_PRIORITARIO <= 19 & MODO_PRIORITARIO >= 10) %>%
  dplyr::select(TIEMPO_VIAJES) #includes both motorcycles and cars

bike_HW_alltt <- OD_v_HW  %>% filter(MODO_PRIORITARIO <= 23 & MODO_PRIORITARIO >= 20) %>%
  dplyr::select(TIEMPO_VIAJES)

walk_HW_alltt <- OD_v_HW  %>% filter(MODO_PRIORITARIO == 24) %>%
  dplyr::select(TIEMPO_VIAJES)

```

plotting histograms:
```{r}
hist(bus_HW_alltt%>% unlist() %>% as.numeric(), breaks=100)
hist(car_motor_HW_alltt%>% unlist() %>% as.numeric(), breaks=100)
hist(bike_HW_alltt%>% unlist() %>% as.numeric(), breaks=100)
hist(walk_HW_alltt%>% unlist() %>% as.numeric(), breaks=100)
```
```{r}
#cullen grey graph - BUS
descdist(bus_HW_alltt%>% unlist() %>% as.numeric(), discrete=FALSE, boot=500)
```

```{r bus-itting-impedance-function}
# using fitdist function to fit a distribution using the default maximum likelihood estimation method and Nelder-Mead method for direct optimization
gamma_ <- fitdistrplus::fitdist(data=bus_HW_alltt%>% unlist() %>% as.numeric(), "gamma", method="mle", optim.method="Nelder-Mead") 
lnorm_ <- fitdistrplus::fitdist(data=bus_HW_alltt%>% unlist() %>% as.numeric(), "lnorm", method="mle", optim.method="Nelder-Mead")
norm_ <-fitdistrplus::fitdist(data=bus_HW_alltt%>% unlist() %>% as.numeric(), "norm", method="mle", optim.method="Nelder-Mead")
exp_ <- fitdistrplus::fitdist(data=bus_HW_alltt%>% unlist() %>% as.numeric(), "exp", method="mle", optim.method="Nelder-Mead")
# pois_ <- fitdistrplus::fitdist(data=bus_HW_alltt%>% unlist() %>% as.numeric(), "pois", method="mle", optim.method="Nelder-Mead")
# nbinom_ <- fitdistrplus::fitdist(data=bus_HW_alltt%>% unlist() %>% as.numeric(), "nbinom", method="mle", optim.method="Nelder-Mead")
# geom_ <- fitdistrplus::fitdist(data=bus_HW_alltt%>% unlist() %>% as.numeric(), "geom", method="mle", optim.method="Nelder-Mead")
# beta_ <- fitdistrplus::fitdist(data=bus_HW_alltt%>% unlist() %>% as.numeric(), "beta", method="mle", optim.method="Nelder-Mead")
logis_ <- fitdistrplus::fitdist(data=bus_HW_alltt%>% unlist() %>% as.numeric(), "logis", method="mle", optim.method="Nelder-Mead")

plot(gamma_)
plot(lnorm_)
plot(norm_)
plot(exp_)
# plot(pois_)
# plot(nbinom_)
# plot(geom_)
# plot(beta_)
plot(logis_)
```
now let's compare these models AIC and BIC to see which fits the bus data best:
```{r}
broom::glance(MASS::fitdistr(bus_HW_alltt%>% unlist() %>% as.numeric(),"gamma"))
broom::glance(MASS::fitdistr(bus_HW_alltt%>% unlist() %>% as.numeric(),"lognormal"))
broom::glance(MASS::fitdistr(bus_HW_alltt%>% unlist() %>% as.numeric(),"normal"))
broom::glance(MASS::fitdistr(bus_HW_alltt%>% unlist() %>% as.numeric(),"exponential"))
# plot(pois_)
# plot(nbinom_)
# plot(geom_)
# plot(beta_)
broom::glance(MASS::fitdistr(bus_HW_alltt%>% unlist() %>% as.numeric(),"logistic"))
```

gamma has largest logLik and the smallest AIC and BIC. we will pick this gamma function for bus trips.
```{r}
HW_bus_gamma <- gamma_
```

Let's do car/motorcycles now:
```{r car-motor-itting-impedance-function}
# using fitdist function to fit a distribution using the default maximum likelihood estimation method and Nelder-Mead method for direct optimization
gamma_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt%>% unlist() %>% as.numeric(), "gamma", method="mle", optim.method="Nelder-Mead") 
lnorm_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt%>% unlist() %>% as.numeric(), "lnorm", method="mle", optim.method="Nelder-Mead")
norm_ <-fitdistrplus::fitdist(data=car_motor_HW_alltt%>% unlist() %>% as.numeric(), "norm", method="mle", optim.method="Nelder-Mead")
exp_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt%>% unlist() %>% as.numeric(), "exp", method="mle", optim.method="Nelder-Mead")
# pois_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt%>% unlist() %>% as.numeric(), "pois", method="mle", optim.method="Nelder-Mead")
# nbinom_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt%>% unlist() %>% as.numeric(), "nbinom", method="mle", optim.method="Nelder-Mead")
# geom_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt%>% unlist() %>% as.numeric(), "geom", method="mle", optim.method="Nelder-Mead")
# beta_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt%>% unlist() %>% as.numeric(), "beta", method="mle", optim.method="Nelder-Mead")
logis_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt%>% unlist() %>% as.numeric(), "logis", method="mle", optim.method="Nelder-Mead")

plot(gamma_)
plot(lnorm_)
plot(norm_)
plot(exp_)
# plot(pois_)
# plot(nbinom_)
# plot(geom_)
# plot(beta_)
plot(logis_)
```
now let's compare these models AIC and BIC to see which fits the car data best:
```{r}
broom::glance(MASS::fitdistr(car_motor_HW_alltt%>% unlist() %>% as.numeric(),"gamma"))
broom::glance(MASS::fitdistr(car_motor_HW_alltt%>% unlist() %>% as.numeric(),"lognormal"))
broom::glance(MASS::fitdistr(car_motor_HW_alltt%>% unlist() %>% as.numeric(),"normal"))
broom::glance(MASS::fitdistr(car_motor_HW_alltt%>% unlist() %>% as.numeric(),"exponential"))
# plot(pois_)
# plot(nbinom_)
# plot(geom_)
# plot(beta_)
broom::glance(MASS::fitdistr(car_motor_HW_alltt%>% unlist() %>% as.numeric(),"logistic"))
```
gamma has largest logLik and the smallest AIC and BIC. we will also pick this gamma function for car trips.
```{r}
HW_car_motor_gamma <- gamma_
```

Let's do bike now:
```{r bike-itting-impedance-function}
# using fitdist function to fit a distribution using the default maximum likelihood estimation method and Nelder-Mead method for direct optimization
gamma_ <- fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "gamma", method="mle", optim.method="Nelder-Mead") 
lnorm_ <- fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "lnorm", method="mle", optim.method="Nelder-Mead")
norm_ <-fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "norm", method="mle", optim.method="Nelder-Mead")
exp_ <- fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "exp", method="mle", optim.method="Nelder-Mead")
# pois_ <- fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "pois", method="mle", optim.method="Nelder-Mead")
# nbinom_ <- fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "nbinom", method="mle", optim.method="Nelder-Mead")
# geom_ <- fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "geom", method="mle", optim.method="Nelder-Mead")
# beta_ <- fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "beta", method="mle", optim.method="Nelder-Mead")
logis_ <- fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "logis", method="mle", optim.method="Nelder-Mead")

plot(gamma_)
plot(lnorm_)
plot(norm_)
plot(exp_)
plot(logis_)
```
now let's compare these models AIC and BIC to see which fits the bike data best:
```{r}
broom::glance(MASS::fitdistr(bike_HW_alltt%>% unlist() %>% as.numeric(),"gamma"))
broom::glance(MASS::fitdistr(bike_HW_alltt%>% unlist() %>% as.numeric(),"lognormal"))
broom::glance(MASS::fitdistr(bike_HW_alltt%>% unlist() %>% as.numeric(),"normal"))
broom::glance(MASS::fitdistr(bike_HW_alltt%>% unlist() %>% as.numeric(),"exponential"))
broom::glance(MASS::fitdistr(bike_HW_alltt%>% unlist() %>% as.numeric(),"logistic"))
```
lognormal has largest logLik and the smallest AIC and BIC. we will pick lognormal function for bike trips! Interesting how this is different than the motorized modes.

```{r}
HW_bike_lnorm <- lnorm_
```

Let's do walk now:
```{r walk-itting-impedance-function}
# using fitdist function to fit a distribution using the default maximum likelihood estimation method and Nelder-Mead method for direct optimization
gamma_ <- fitdistrplus::fitdist(data=walk_HW_alltt%>% unlist() %>% as.numeric(), "gamma", method="mle", optim.method="Nelder-Mead") 
lnorm_ <- fitdistrplus::fitdist(data=walk_HW_alltt%>% unlist() %>% as.numeric(), "lnorm", method="mle", optim.method="Nelder-Mead")
norm_ <-fitdistrplus::fitdist(data=walk_HW_alltt%>% unlist() %>% as.numeric(), "norm", method="mle", optim.method="Nelder-Mead")
exp_ <- fitdistrplus::fitdist(data=walk_HW_alltt%>% unlist() %>% as.numeric(), "exp", method="mle", optim.method="Nelder-Mead")
# pois_ <- fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "pois", method="mle", optim.method="Nelder-Mead")
# nbinom_ <- fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "nbinom", method="mle", optim.method="Nelder-Mead")
# geom_ <- fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "geom", method="mle", optim.method="Nelder-Mead")
# beta_ <- fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "beta", method="mle", optim.method="Nelder-Mead")
logis_ <- fitdistrplus::fitdist(data=walk_HW_alltt%>% unlist() %>% as.numeric(), "logis", method="mle", optim.method="Nelder-Mead")

plot(gamma_)
plot(lnorm_)
plot(norm_)
plot(exp_)
plot(logis_)
```
now let's compare these models AIC and BIC to see which fits the bike data best:
```{r}
broom::glance(MASS::fitdistr(walk_HW_alltt%>% unlist() %>% as.numeric(),"gamma"))
broom::glance(MASS::fitdistr(walk_HW_alltt%>% unlist() %>% as.numeric(),"lognormal"))
broom::glance(MASS::fitdistr(walk_HW_alltt%>% unlist() %>% as.numeric(),"normal"))
broom::glance(MASS::fitdistr(walk_HW_alltt%>% unlist() %>% as.numeric(),"exponential"))
broom::glance(MASS::fitdistr(walk_HW_alltt%>% unlist() %>% as.numeric(),"logistic"))
```
lognormal has largest logLik and the smallest AIC and BIC. we will pick lognormal function for walk trips! Interesting how this is different than the motorized modes and takes the same form as the bike trips.

```{r}
HW_walk_lnorm <- lnorm_
```

## Spatial availability comparision

## Inequality analysis
<!-- to do -->

# Discussion

# Conclusions

### 2022-05-08 Draft-manuscript-tobemoved.Rmd
finalized 2 spatial availability scenario calculations (baseline no LEZ and optimistic LEZ)

@@
 ```{r load-packages, include=FALSE, cache=FALSE}
-library(TTS2016R)
-library(dplyr)
+library(tidyverse)
+library(sf)
 library(fitdistrplus)
 library(ggplot2)
-library(kableExtra)
+library(readxl) #imports excel files
 library(patchwork)
-library(sf)
-library(scales)
-library(ggpmisc)
-library(ggrepel)
 library(cowplot)
-library(ggspatial)
-library(spdep)
-library(RColorBrewer)
-library(tmap)
-library(shadowtext)
-library(grid)
-library(readxl) #imports excel files
-# library(extrafont)
-# font_import()
-# loadfonts(device = "win")
-
-options(scipen = 999)
 ```
@@
 #here we separate all the "car" trips, "transit" trips, "bike" trips, and "walk" trips. We then will create 4 different imepedance function....
-bus_HW_alltt <- OD_v_HW  %>% filter(MODO_PRIORITARIO >= 9) %>%
+transit_HW_alltt <- OD_v_HW  %>% filter(MODO_PRIORITARIO <= 9) %>%
   dplyr::select(TIEMPO_VIAJES)
@@
 ```{r}
-hist(bus_HW_alltt%>% unlist() %>% as.numeric(), breaks=100)
+hist(transit_HW_alltt%>% unlist() %>% as.numeric(), breaks=100)
 hist(car_motor_HW_alltt%>% unlist() %>% as.numeric(), breaks=100)
@@
 #cullen grey graph - BUS
-descdist(bus_HW_alltt%>% unlist() %>% as.numeric(), discrete=FALSE, boot=500)
+descdist(transit_HW_alltt%>% unlist() %>% as.numeric(), discrete=FALSE, boot=500)
 ```
@@
 # using fitdist function to fit a distribution using the default maximum likelihood estimation method and Nelder-Mead method for direct optimization
-gamma_ <- fitdistrplus::fitdist(data=bus_HW_alltt%>% unlist() %>% as.numeric(), "gamma", method="mle", optim.method="Nelder-Mead") 
-lnorm_ <- fitdistrplus::fitdist(data=bus_HW_alltt%>% unlist() %>% as.numeric(), "lnorm", method="mle", optim.method="Nelder-Mead")
-norm_ <-fitdistrplus::fitdist(data=bus_HW_alltt%>% unlist() %>% as.numeric(), "norm", method="mle", optim.method="Nelder-Mead")
-exp_ <- fitdistrplus::fitdist(data=bus_HW_alltt%>% unlist() %>% as.numeric(), "exp", method="mle", optim.method="Nelder-Mead")
-# pois_ <- fitdistrplus::fitdist(data=bus_HW_alltt%>% unlist() %>% as.numeric(), "pois", method="mle", optim.method="Nelder-Mead")
-# nbinom_ <- fitdistrplus::fitdist(data=bus_HW_alltt%>% unlist() %>% as.numeric(), "nbinom", method="mle", optim.method="Nelder-Mead")
-# geom_ <- fitdistrplus::fitdist(data=bus_HW_alltt%>% unlist() %>% as.numeric(), "geom", method="mle", optim.method="Nelder-Mead")
-# beta_ <- fitdistrplus::fitdist(data=bus_HW_alltt%>% unlist() %>% as.numeric(), "beta", method="mle", optim.method="Nelder-Mead")
-logis_ <- fitdistrplus::fitdist(data=bus_HW_alltt%>% unlist() %>% as.numeric(), "logis", method="mle", optim.method="Nelder-Mead")
+gamma_ <- fitdistrplus::fitdist(data=transit_HW_alltt%>% unlist() %>% as.numeric(), "gamma", method="mle", optim.method="Nelder-Mead") 
+lnorm_ <- fitdistrplus::fitdist(data=transit_HW_alltt%>% unlist() %>% as.numeric(), "lnorm", method="mle", optim.method="Nelder-Mead")
+norm_ <-fitdistrplus::fitdist(data=transit_HW_alltt%>% unlist() %>% as.numeric(), "norm", method="mle", optim.method="Nelder-Mead")
+exp_ <- fitdistrplus::fitdist(data=transit_HW_alltt%>% unlist() %>% as.numeric(), "exp", method="mle", optim.method="Nelder-Mead")
+# pois_ <- fitdistrplus::fitdist(data=transit_HW_alltt%>% unlist() %>% as.numeric(), "pois", method="mle", optim.method="Nelder-Mead")
+# nbinom_ <- fitdistrplus::fitdist(data=transit_HW_alltt%>% unlist() %>% as.numeric(), "nbinom", method="mle", optim.method="Nelder-Mead")
+# geom_ <- fitdistrplus::fitdist(data=transit_HW_alltt%>% unlist() %>% as.numeric(), "geom", method="mle", optim.method="Nelder-Mead")
+# beta_ <- fitdistrplus::fitdist(data=transit_HW_alltt%>% unlist() %>% as.numeric(), "beta", method="mle", optim.method="Nelder-Mead")
+logis_ <- fitdistrplus::fitdist(data=transit_HW_alltt%>% unlist() %>% as.numeric(), "logis", method="mle", optim.method="Nelder-Mead")
 
@@
 ```{r}
-broom::glance(MASS::fitdistr(bus_HW_alltt%>% unlist() %>% as.numeric(),"gamma"))
-broom::glance(MASS::fitdistr(bus_HW_alltt%>% unlist() %>% as.numeric(),"lognormal"))
-broom::glance(MASS::fitdistr(bus_HW_alltt%>% unlist() %>% as.numeric(),"normal"))
-broom::glance(MASS::fitdistr(bus_HW_alltt%>% unlist() %>% as.numeric(),"exponential"))
+broom::glance(MASS::fitdistr(transit_HW_alltt%>% unlist() %>% as.numeric(),"gamma"))
+broom::glance(MASS::fitdistr(transit_HW_alltt%>% unlist() %>% as.numeric(),"lognormal"))
+broom::glance(MASS::fitdistr(transit_HW_alltt%>% unlist() %>% as.numeric(),"normal"))
+broom::glance(MASS::fitdistr(transit_HW_alltt%>% unlist() %>% as.numeric(),"exponential"))
 # plot(pois_)
@@
 # plot(beta_)
-broom::glance(MASS::fitdistr(bus_HW_alltt%>% unlist() %>% as.numeric(),"logistic"))
+broom::glance(MASS::fitdistr(transit_HW_alltt%>% unlist() %>% as.numeric(),"logistic"))
 ```
 
-gamma has largest logLik and the smallest AIC and BIC. we will pick this gamma function for bus trips.
+lognormal has largest logLik and the smallest AIC and BIC. we will pick this lognormal function for transit trips.
 ```{r}
-HW_bus_gamma <- gamma_
+HW_transit_lnorm <- lnorm_
 ```
@@
 ```
-gamma has largest logLik and the smallest AIC and BIC. we will also pick this gamma function for car trips.
+gamma has largest logLik and the smallest AIC and BIC. we will pick this gamma function for car trips.
 ```{r}
@@
 ```
-lognormal has largest logLik and the smallest AIC and BIC. we will pick lognormal function for bike trips! Interesting how this is different than the motorized modes.
+lognormal has largest logLik and the smallest AIC and BIC. we will pick lognormal function for bike trips! Interesting how car mode is different from all other modes.
 
@@
 ```
-now let's compare these models AIC and BIC to see which fits the bike data best:
+now let's compare these models AIC and BIC to see which fits the walk data best:
 ```{r}
@@
 ```
-lognormal has largest logLik and the smallest AIC and BIC.
-we will pick lognormal function for walk trips!
-Interesting how this is different than the motorized modes and takes the same form as the bike trips.
+lognormal has largest logLik and the smallest AIC and BIC. we will pick lognormal function for walk trips!
 
@@
 
+Add the number of trips (by mode) and create objects which contain the number of workers (origin) and jobs (destinations) - by mode:
+```{r}
+OD_v_HW <- OD_v_HW  %>% mutate(trips_transit = ifelse(MODO_PRIORITARIO <= 9, 1,0),
+                    trips_car_motor = ifelse(MODO_PRIORITARIO <= 19 & MODO_PRIORITARIO >= 10, 1, 0),
+                    trips_bike = ifelse(MODO_PRIORITARIO <= 23 & MODO_PRIORITARIO >= 20, 1, 0),
+                    trips_walk = ifelse(MODO_PRIORITARIO == 24, 1, 0))
+
+workers <- OD_v_HW %>%
+  group_by(VORIZT1259) %>%
+  summarize(workers_transit = sum(trips_transit),
+            workers_car = sum(trips_car_motor),
+            workers_bike = sum(trips_bike),
+            workers_walk = sum(trips_walk),
+            .groups = "drop") %>%
+  transmute(ZT1259 = VORIZT1259,
+            workers_transit,
+            workers_car,
+            workers_bike,
+            workers_walk)
+
+jobs <- OD_v_HW %>%
+  group_by(VDESZT1259) %>%
+  summarize(jobs_transit = sum(trips_transit),
+            jobs_car = sum(trips_car_motor),
+            jobs_bike = sum(trips_bike),
+            jobs_walk = sum(trips_walk),
+            .groups = "drop") %>%
+  transmute(ZT1259 = VDESZT1259,
+            jobs_transit,
+            jobs_car,
+            jobs_bike,
+            jobs_walk)
+```
+
+Join workers and jobs to the ZN simple features object:
+```{r}
+load("data/zn.RData")
+
+zn <- zn %>%
+  # Join workers
+  left_join(workers,
+            by = "ZT1259") %>%
+  # Join jobs
+  left_join(jobs,
+            by = "ZT1259") %>%
+  # Replace NAs with zeros
+  mutate(workers_transit = tidyr::replace_na(workers_transit, 0),
+         workers_car = tidyr::replace_na(workers_car, 0),
+         workers_bike = tidyr::replace_na(workers_bike, 0),
+         workers_walk = tidyr::replace_na(workers_walk, 0),
+         jobs_transit = tidyr::replace_na(jobs_transit, 0),
+         jobs_car = tidyr::replace_na(jobs_car, 0),
+         jobs_bike = tidyr::replace_na(jobs_bike, 0),
+         jobs_walk = tidyr::replace_na(jobs_walk, 0))
+
+# sum(zn$workers_transit) + sum(zn$workers_bike) + sum(zn$workers_car) + sum(zn$workers_walk)
+# sum(zn$jobs_transit) + sum(zn$jobs_bike) + sum(zn$jobs_car) + sum(zn$jobs_walk)
+```
+
+add the number of workers and jobs from each zona to each trip:
+```{r}
+#add the number of jobs and workers to the od_ft_tt matrix
+OD <- OD_v_HW %>% merge(zn %>% dplyr::select(ZT1259, workers_transit,workers_car,workers_bike,workers_walk) %>% st_drop_geometry(),
+                   by.x = "VORIZT1259", by.y="ZT1259", all.x = TRUE)
+
+OD <- OD %>% merge(zn %>% dplyr::select(ZT1259, jobs_transit,jobs_car,jobs_bike,jobs_walk) %>% st_drop_geometry(),
+                   by.x = "VDESZT1259", by.y="ZT1259", all.x = TRUE)
+
+OD <- OD %>% 
+  mutate(workers_transit = ifelse(is.na(workers_transit), 0, workers_transit),
+         workers_car = ifelse(is.na(workers_car), 0, workers_car),
+         workers_bike = ifelse(is.na(workers_bike), 0, workers_bike),
+         workers_walk = ifelse(is.na(workers_walk), 0, workers_walk),
+         jobs_car = ifelse(is.na(jobs_car), 0, jobs_car),
+         jobs_transit = ifelse(is.na(jobs_transit), 0, jobs_transit),
+         jobs_bike = ifelse(is.na(jobs_bike), 0, jobs_bike),
+         jobs_walk = ifelse(is.na(jobs_walk), 0, jobs_walk))
+```
+
+now reshape the data, each row will represent an any origin to to 1 destination (i.e., destination row is unique).
+We will average the travel times, by mode, and then use this average travel time to find the 'f' (impedance value for travel for each mode):
+```{r}
+#consolidate trips so each trip is unique. workers = jobs.
+OD_test <- OD %>%
+  group_by(VORIZT1259,VDESZT1259) %>%
+  summarise(trips_transit = sum(trips_transit),
+            trips_car_motor = sum(trips_car_motor),
+            trips_bike = sum(trips_bike),
+            trips_walk = sum(trips_walk),
+            workers_transit = mean(workers_transit),
+            # workers_transit = ifelse(is.na(workers_transit), 0, workers_transit),
+            workers_car = mean(workers_car),
+            # workers_car = ifelse(is.na(workers_car), 0, workers_car),
+            workers_bike = mean(workers_bike),
+            # workers_bike = ifelse(is.na(workers_bike), 0, workers_bike),
+            workers_walk = mean(workers_walk),
+            # workers_walk = ifelse(is.na(workers_walk), 0, workers_walk),
+            jobs_transit = mean(jobs_transit),
+            jobs_car = mean(jobs_car),
+            jobs_bike = mean(jobs_bike),
+            jobs_walk = mean(jobs_walk),
+            # jobs_car = ifelse(is.na(jobs_car), 0, jobs_car),
+            # jobs_transit = ifelse(is.na(jobs_transit), 0, jobs_transit),
+            # jobs_bike = ifelse(is.na(jobs_bike), 0, jobs_bike),
+            # jobs_walk = ifelse(is.na(jobs_walk), 0, jobs_walk),
+            tt_transit = ifelse(trips_transit != 0, mean(TIEMPO_VIAJES, na.rm=T), NA),
+            tt_car = ifelse(trips_car_motor != 0, mean(TIEMPO_VIAJES, na.rm=T), NA),
+            tt_bike = ifelse(trips_bike != 0, mean(TIEMPO_VIAJES, na.rm=T), NA),
+            tt_walk = ifelse(trips_walk != 0, mean(TIEMPO_VIAJES, na.rm=T), NA)) %>% ungroup()
+
+# OD_test <- OD %>%
+#   mutate(workers_ = (workers_car+ workers_transit+ workers_bike+ workers_walk),
+#          workers_ = ifelse(is.na(workers_), 0, workers_),
+#          workers_car = ifelse(is.na(workers_car), 0, workers_car),
+#          workers_transit = ifelse(is.na(workers_transit), 0, workers_transit),
+#          workers_bike = ifelse(is.na(workers_bike), 0, workers_bike),
+#          workers_walk = ifelse(is.na(workers_walk), 0, workers_walk),
+#            jobs_ = (jobs_car+ jobs_transit+ jobs_bike+ jobs_walk),
+#          jobs_ = ifelse(is.na(jobs_), 0, jobs_),
+#          jobs_car = ifelse(is.na(jobs_car), 0, jobs_car),
+#          jobs_transit = ifelse(is.na(jobs_transit), 0, jobs_transit),
+#          jobs_bike = ifelse(is.na(jobs_bike), 0, jobs_bike),
+#          jobs_walk = ifelse(is.na(jobs_walk), 0, jobs_walk),
+#          tt_transit = ifelse(trips_transit != 0, TIEMPO_VIAJES, NA),
+#             tt_car = ifelse(trips_car_motor != 0, TIEMPO_VIAJES, NA),
+#             tt_bike = ifelse(trips_bike != 0, TIEMPO_VIAJES, NA),
+#             tt_walk = ifelse(trips_walk != 0, TIEMPO_VIAJES, NA)) %>% ungroup()
+
+#check, should add to the origin number of trips 27795 - perfect!
+# sum(OD$trips_transit) + sum(OD$trips_car_motor) + sum(OD$trips_bike) + sum(OD$trips_walk)
+```
+
+now estimate f value using these averaged travel times:
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+#dgamma is the density plot for gamma function and dlnorm is the desntiy plot for the lognormal function; adding values as "f" our impedance function
+OD_test <- OD_test %>%
+  mutate(f_car = dgamma(tt_car, HW_car_motor_gamma$estimate["shape"], HW_car_motor_gamma$estimate["rate"]),
+         f_car = ifelse(is.na(f_car), 0, f_car),
+         f_transit = dlnorm(tt_transit, HW_transit_lnorm$estimate["meanlog"], HW_transit_lnorm$estimate["sdlog"]),
+         f_transit = ifelse(is.na(f_transit), 0, f_transit),
+         f_bike = dlnorm(tt_bike, HW_bike_lnorm$estimate["meanlog"], HW_bike_lnorm$estimate["sdlog"]),
+         f_bike = ifelse(is.na(f_bike), 0, f_bike),
+         f_walk = dlnorm(tt_walk, HW_walk_lnorm$estimate["meanlog"], HW_walk_lnorm$estimate["sdlog"]),
+         f_walk = ifelse(is.na(f_walk), 0, f_walk))
+```
+
+```{=tex}
+\begin{equation}
+\label{gamma-dist}
+\begin{array}{l}\ 
+f(x, \alpha, \beta) = \frac {x^{\alpha-1}e^{-\frac{x}{\beta}}}{ \beta^{\alpha}\Gamma(\alpha)} \quad \text{for } 0 \leq x \leq \infty\\
+
+\Gamma(\alpha) =  \int_{0}^{\infty} x^{\alpha-1}e^{-x} \,dx\\
+\end{array}
+\end{equation}
+```
+
 ## Spatial availability comparision
 
-## Inequality analysis
-<!-- to do -->
+```{r}
+#calculate spatial availability
+OD_SA <- OD_test %>%
+  mutate(catch = 1) %>%
+  mutate(SA_ij_car = sp_avail(., 
+                         o_id = VORIZT1259,
+                         d_id = VDESZT1259,
+                         pop = workers_car,
+                         opp = jobs_car,
+                         r = catch,
+                         f = f_car),
+         SA_ij_car = ifelse(is.nan(SA_ij_car), 0, SA_ij_car),
+         SA_ij_transit = sp_avail(., 
+                         o_id = VORIZT1259,
+                         d_id = VDESZT1259,
+                         pop = workers_transit,
+                         opp = jobs_transit,
+                         r = catch,
+                         f = f_transit),
+         SA_ij_transit = ifelse(is.nan(SA_ij_transit), 0, SA_ij_transit),
+         SA_ij_bike = sp_avail(., 
+                         o_id = VORIZT1259,
+                         d_id = VDESZT1259,
+                         pop = workers_bike,
+                         opp = jobs_bike,
+                         r = catch,
+                         f = f_bike),
+         SA_ij_bike = ifelse(is.nan(SA_ij_bike), 0, SA_ij_bike),
+         SA_ij_walk = sp_avail(., 
+                         o_id = VORIZT1259,
+                         d_id = VDESZT1259,
+                         pop = workers_walk,
+                         opp = jobs_walk,
+                         r = catch,
+                         f = f_walk),
+         SA_ij_walk = ifelse(is.nan(SA_ij_walk), 0, SA_ij_walk))
 
-# Discussion
+#verify that the sum of all jobs is consistent with the number of jobs
+sum(OD_SA$SA_ij_car) + sum(OD_SA$SA_ij_transit) + sum(OD_SA$SA_ij_bike) + sum(OD_SA$SA_ij_walk)
 
-# Conclusions
+#verify that the sum of all jobs is consistent with the number of jobs - from another perspective
+sum_jobs <- OD_SA %>% 
+  group_by(VDESZT1259) %>% 
+  summarise(jobs = mean(jobs_transit) + mean(jobs_car)  + mean(jobs_walk) + mean(jobs_bike))
+sum(sum_jobs$jobs, na.rm = T)
+
+# same number excellent!
+there are 27434 jobs total and there are 27795 workers (and subsequently 27795 trips total).
+The mismatch in numbers arises because we only survey people who's work trip originates in the Madrid region (from home).
+This means that workers who travel outside of the madrid region still count as a worker (and a trip) + increase 'competition' for the zone....
+maybe we don't want this?
+
+```
+
+Let's try calculating SA from a different approch:
+```{r}
+# Also another question - this configuration assumes competition is contained within modes - i.e., transit workers compete for transit jobs, etc.
+This isn't the most realistic I think.
+So let's sum up all the mode specific workers and jobs and impedance values BEFORE calculating SA so we only have 1 SA
+
+#add new summed columns
+OD_SA <- OD_SA %>%
+  mutate(workers_ = (workers_car+ workers_transit+ workers_bike+ workers_walk),
+         workers_ = ifelse(is.na(workers_), 0, workers_),
+         jobs_ = (jobs_car + jobs_transit + jobs_bike + jobs_walk),
+         jobs_ = ifelse(is.na(jobs_), 0, jobs_),
+         f_ = f_car + f_transit + f_bike + f_walk)
+
+#calculate spatial availability
+OD_SA2 <- OD_SA %>%
+  mutate(catch = 1) %>%
+  mutate(SA_ij = sp_avail(., 
+                         o_id = VORIZT1259,
+                         d_id = VDESZT1259,
+                         pop = workers_,
+                         opp = jobs_,
+                         r = catch,
+                         f = f_),
+         SA_ij = ifelse(is.nan(SA_ij), 0, SA_ij))
+
+#verify that the sum of all jobs is consistent with the number of jobs
+sum(OD_SA2$SA_ij)
+
+#verify that the sum of all jobs is consistent with the number of jobs - from another perspective
+sum_jobs <- OD_SA2 %>% 
+  group_by(VDESZT1259) %>% 
+  summarise(jobs = mean(jobs_))
+sum(sum_jobs$jobs, na.rm = T)
+#Perfect it works! 
+```
+
+Let's aggregate the spatial availability by origin zone for these 2 methods (calculating SA separately... and calculating 1 SA that combines all modes)
+```{r}
+#aggregating spatial availability  
+MAD_availability <- OD_SA2 %>%
+  group_by(VORIZT1259) %>%
+  summarize(SA_i2 = sum(SA_ij),
+            SA_i1 = sum(SA_ij_car) + sum(SA_ij_bike) + sum(SA_ij_transit) + sum(SA_ij_walk),
+            workers_ = mean(workers_car, na.rm=T) + mean(workers_transit, na.rm=T) + mean(workers_bike, na.rm=T) + mean(workers_walk, na.rm=T),
+            avgtt_i = mean(mean(tt_transit) + mean(tt_walk) + mean(tt_car) + mean(tt_bike)),
+            avg_f_i = mean(mean(f_transit) + mean(f_walk) + mean(f_car) + mean(f_bike)))
+
+#Merge TO availability calculation to the TAZ sf object created for accessibility above:
+zn_test <- zn %>% merge(MAD_availability, by.x=c("ZT1259"), by.y=c("VORIZT1259"), all.x=T) 
+```
+
+```{r create-bounding-boxes}
+#create the bounding box for the Toronto region + a little extra
+bbox_new1 <- st_bbox(zn_test%>%filter(Municipality == "Madrid")) # current bounding box
+
+xrange <- bbox_new1$xmax - bbox_new1$xmin # range of x values
+yrange <- bbox_new1$ymax - bbox_new1$ymin # range of y values
+
+bbox_new1[1] <- bbox_new1[1] - (0.0 * xrange) # xmin - left
+bbox_new1[3] <- bbox_new1[3] + (0.0 * xrange) # xmax - right
+bbox_new1[2] <- bbox_new1[2] - (0.0 * yrange) # ymin - bottom
+bbox_new1[4] <- bbox_new1[4] + (0.0 * yrange) # ymax - top
+
+bbox_Madrid_muni <- bbox_new1 %>%  # take the bounding box ...
+  st_as_sfc()
+
+Madrid_muni_border <- zn_test%>%filter(Municipality == "Madrid")%>%st_union()
+```
+
+plot 1 (SA for each mode calculated seperated and then added together) and plot 2 (SA for all modes calculated together):
+```{r}
+#plot 1 main plot
+plot_MAD_avail1 <- ggplot() +
+  geom_sf(data = zn_test, aes(fill= SA_i1), color = NA) + #data
+    scale_fill_distiller(palette = "RdYlGn", #legend scale bar
+                         name = expression(V["i"]),
+                         na.value = "grey90",
+                         limits = c( round(min(zn_test$SA_i1, na.rm=T)), round(max(zn_test$SA_i1, na.rm=T)))) +
+  geom_sf(data = Madrid_muni_border, # border for Madrid muni
+          colour=alpha("dimgrey",1), 
+          size = 0.5, fill=NA, 
+          show.legend = "polygon") +
+  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
+                         height = unit(0.8, "cm"), 
+                         width = unit(0.8, "cm"),
+                         style = north_arrow_orienteering(line_width = 0.25,
+                                                          line_col = "dimgrey", 
+                                                          fill = c("white","dimgrey"))) +
+  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
+                   height = unit(0.15, "cm")) +
+  theme_void() +
+  theme(legend.position = c(1.10, 0.25), 
+        plot.margin = margin(t = 0, r = 0, b = 0, l = -80)) # positing legend and clipping out white space
+
+# plot 1 + inset map
+plot_MAD_avail1 <- ggdraw(plot_MAD_avail1) +
+  draw_plot({plot_MAD_avail1 + coord_sf(xlim = st_coordinates(bbox_Madrid_muni)[c(1,2),1],
+                                         ylim = st_coordinates(bbox_Madrid_muni)[c(2,3),2],
+                                         expand = FALSE) +
+      theme(legend.position = "none")},
+    x = 0.65, y = 0.50,  # The distance along a (0,1) x-axis (and y axis) to draw the left edge (the bottom edge) of the plot
+    width = 0.46, height = 0.46) # The width and height of the plot expressed as proportion of the entire ggdraw object
+
+#plot 2
+plot_MAD_avail2 <- ggplot() +
+  geom_sf(data = zn_test, aes(fill= SA_i2), color = NA) + #data
+    scale_fill_distiller(palette = "RdYlGn", #legend scale bar
+                         name = expression(V["i"]),
+                         na.value = "grey90",
+                         limits = c( round(min(zn_test$SA_i2, na.rm=T)), round(max(zn_test$SA_i2, na.rm=T)))) +
+  geom_sf(data = Madrid_muni_border, # border for Madrid muni
+          colour=alpha("dimgrey",1), 
+          size = 0.5, fill=NA, 
+          show.legend = "polygon") +
+  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
+                         height = unit(0.8, "cm"), 
+                         width = unit(0.8, "cm"),
+                         style = north_arrow_orienteering(line_width = 0.25,
+                                                          line_col = "dimgrey", 
+                                                          fill = c("white","dimgrey"))) +
+  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
+                   height = unit(0.15, "cm")) +
+  theme_void() +
+  theme(legend.position = c(1.10, 0.25), 
+        plot.margin = margin(t = 0, r = 0, b = 0, l = -80)) # positing legend and clipping out white space
+
+# plot 1 + inset map
+plot_MAD_avail2 <- ggdraw(plot_MAD_avail2) +
+  draw_plot({plot_MAD_avail2 + coord_sf(xlim = st_coordinates(bbox_Madrid_muni)[c(1,2),1],
+                                         ylim = st_coordinates(bbox_Madrid_muni)[c(2,3),2],
+                                         expand = FALSE) +
+      theme(legend.position = "none")},
+    x = 0.65, y = 0.50,  # The distance along a (0,1) x-axis (and y axis) to draw the left edge (the bottom edge) of the plot
+    width = 0.46, height = 0.46) # The width and height of the plot expressed as proportion of the entire ggdraw object
+
+plot_MAD_avail1
+plot_MAD_avail2
+```
+We can normalize it by the number of workers so the numeric value corresponds to available jobs per worker. First plot 1, then plot 2.
+```{r}
+#plot 1 main plot
+plot_MAD_avail_perworker_1 <- ggplot() +
+  geom_sf(data = zn_test, aes(fill= SA_i1/workers_), color = NA) + #data
+    scale_fill_distiller(palette = "RdYlGn", #legend scale bar
+                         name = expression(v["i"]),
+                         na.value = "grey90",
+                         limits = c( round(min(zn_test$SA_i1/zn_test$workers_, na.rm=T)), 
+                                     round(max(zn_test$SA_i1/zn_test$workers_, na.rm=T))))+
+  geom_sf(data = Madrid_muni_border, # border for Madrid muni
+          colour=alpha("dimgrey",1), 
+          size = 0.5, fill=NA, 
+          show.legend = "polygon") +
+  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
+                         height = unit(0.8, "cm"), 
+                         width = unit(0.8, "cm"),
+                         style = north_arrow_orienteering(line_width = 0.25,
+                                                          line_col = "dimgrey", 
+                                                          fill = c("white","dimgrey"))) +
+  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
+                   height = unit(0.15, "cm")) +
+  theme_void() +
+  theme(legend.position = c(1.10, 0.25), 
+        plot.margin = margin(t = 0, r = 0, b = 0, l = -80)) # positing legend and clipping out white space
+
+# plot 1 + inset map
+plot_MAD_avail_perworker_1 <- ggdraw(plot_MAD_avail_perworker_1) +
+  draw_plot({plot_MAD_avail_perworker_1 + coord_sf(xlim = st_coordinates(bbox_Madrid_muni)[c(1,2),1],
+                                         ylim = st_coordinates(bbox_Madrid_muni)[c(2,3),2],
+                                         expand = FALSE) +
+      theme(legend.position = "none")},
+    x = 0.65, y = 0.50,  # The distance along a (0,1) x-axis (and y axis) to draw the left edge (the bottom edge) of the plot
+    width = 0.46, height = 0.46) # The width and height of the plot expressed as proportion of the entire ggdraw object
+
+#plot 2
+plot_MAD_avail_perworker_2 <- ggplot() +
+  geom_sf(data = zn_test, aes(fill= SA_i2/workers_), color = NA) + #data
+    scale_fill_distiller(palette = "RdYlGn", #legend scale bar
+                         name = expression(v["i"]),
+                         na.value = "grey90",
+                         limits = c( round(min(zn_test$SA_i2/zn_test$workers_, na.rm=T)), 
+                                     round(max(zn_test$SA_i2/zn_test$workers_, na.rm=T))))+
+  geom_sf(data = Madrid_muni_border, # border for Madrid muni
+          colour=alpha("dimgrey",1), 
+          size = 0.5, fill=NA, 
+          show.legend = "polygon") +
+  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
+                         height = unit(0.8, "cm"), 
+                         width = unit(0.8, "cm"),
+                         style = north_arrow_orienteering(line_width = 0.25,
+                                                          line_col = "dimgrey", 
+                                                          fill = c("white","dimgrey"))) +
+  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
+                   height = unit(0.15, "cm")) +
+  theme_void() +
+  theme(legend.position = c(1.10, 0.25), 
+        plot.margin = margin(t = 0, r = 0, b = 0, l = -80)) # positing legend and clipping out white space
+
+# plot 1 + inset map
+plot_MAD_avail_perworker_2 <- ggdraw(plot_MAD_avail_perworker_2) +
+  draw_plot({plot_MAD_avail_perworker_2 + coord_sf(xlim = st_coordinates(bbox_Madrid_muni)[c(1,2),1],
+                                         ylim = st_coordinates(bbox_Madrid_muni)[c(2,3),2],
+                                         expand = FALSE) +
+      theme(legend.position = "none")},
+    x = 0.65, y = 0.50,  # The distance along a (0,1) x-axis (and y axis) to draw the left edge (the bottom edge) of the plot
+    width = 0.46, height = 0.46) # The width and height of the plot expressed as proportion of the entire ggdraw object
+
+plot_MAD_avail_perworker_1
+plot_MAD_avail_perworker_2
+```
+
+## An optimistic scenario - no cars going into madrid municipality
+
+Let's test out a scenario - all car trips going into madrid municipality are now transit trips. How will this impact the SA profile? We will using the 2nd method (calculating SA within all modes).
+
+First, let's re-assign to-madrid muni car trips to transit trips:
+```{r}
+Madrid_muni <- zn %>% filter(Municipality == "Madrid") %>% 
+  dplyr::transmute(ZT1259,
+                   MM = 1)
+
+OD_v_HW_optimistic <- OD_v_HW  %>% merge(Madrid_muni, by.x=("VDESZT1259"), by.y=("ZT1259"), all.x=T)
+
+#reassigning car trips to transit
+OD_v_HW_optimistic1 <- OD_v_HW_optimistic  %>% filter(MM == 1) %>% rowwise() %>%
+  mutate(trips_transit = sum(trips_transit,trips_car_motor, na.rm = T),
+         trips_car_motor = 0)
+
+OD_v_HW_optimistic2 <- OD_v_HW_optimistic  %>% filter(is.na(MM))
+
+OD_v_HW_optimistic <- rbind(OD_v_HW_optimistic1,OD_v_HW_optimistic2)
+
+workers <- OD_v_HW_optimistic %>%
+  group_by(VORIZT1259) %>%
+  summarize(workers_transit = sum(trips_transit),
+            workers_car = sum(trips_car_motor),
+            workers_bike = sum(trips_bike),
+            workers_walk = sum(trips_walk),
+            .groups = "drop") %>%
+  transmute(ZT1259 = VORIZT1259,
+            workers_transit,
+            workers_car,
+            workers_bike,
+            workers_walk)
+
+jobs <- OD_v_HW_optimistic %>%
+  group_by(VDESZT1259) %>%
+  summarize(jobs_transit = sum(trips_transit),
+            jobs_car = sum(trips_car_motor),
+            jobs_bike = sum(trips_bike),
+            jobs_walk = sum(trips_walk),
+            .groups = "drop") %>%
+  transmute(ZT1259 = VDESZT1259,
+            jobs_transit,
+            jobs_car,
+            jobs_bike,
+            jobs_walk)
+```
+
+Now again, join the workers and jobs to the ZN simple features object:
+```{r}
+load("data/zn.RData")
+
+zn <- zn %>%
+  # Join workers
+  left_join(workers,
+            by = "ZT1259") %>%
+  # Join jobs
+  left_join(jobs,
+            by = "ZT1259") %>%
+  # Replace NAs with zeros
+  mutate(workers_transit = tidyr::replace_na(workers_transit, 0),
+         workers_car = tidyr::replace_na(workers_car, 0),
+         workers_bike = tidyr::replace_na(workers_bike, 0),
+         workers_walk = tidyr::replace_na(workers_walk, 0),
+         jobs_transit = tidyr::replace_na(jobs_transit, 0),
+         jobs_car = tidyr::replace_na(jobs_car, 0),
+         jobs_bike = tidyr::replace_na(jobs_bike, 0),
+         jobs_walk = tidyr::replace_na(jobs_walk, 0))
+```
+
+add the number of workers and jobs from each zona to each trip:
+```{r}
+#add the number of jobs and workers to the od_ft_tt matrix
+OD_optimistic <- OD_v_HW_optimistic %>% merge(zn %>% dplyr::select(ZT1259, workers_transit,workers_car,workers_bike,workers_walk) %>% st_drop_geometry(),
+                   by.x = "VORIZT1259", by.y="ZT1259", all.x = TRUE)
+
+OD_optimistic <- OD_optimistic %>% merge(zn %>% dplyr::select(ZT1259, jobs_transit,jobs_car,jobs_bike,jobs_walk) %>% st_drop_geometry(),
+                   by.x = "VDESZT1259", by.y="ZT1259", all.x = TRUE)
+
+OD_optimistic <- OD_optimistic %>% 
+  mutate(workers_transit = ifelse(is.na(workers_transit), 0, workers_transit),
+         workers_car = ifelse(is.na(workers_car), 0, workers_car),
+         workers_bike = ifelse(is.na(workers_bike), 0, workers_bike),
+         workers_walk = ifelse(is.na(workers_walk), 0, workers_walk),
+         jobs_car = ifelse(is.na(jobs_car), 0, jobs_car),
+         jobs_transit = ifelse(is.na(jobs_transit), 0, jobs_transit),
+         jobs_bike = ifelse(is.na(jobs_bike), 0, jobs_bike),
+         jobs_walk = ifelse(is.na(jobs_walk), 0, jobs_walk))
+```
+
+now reshape the data, each row will represent an any origin to to 1 destination (i.e., destination row is unique).
+We will average the travel times, by mode, and then use this average travel time to find the 'f' (impedance value for travel for each mode):
+```{r}
+#consolidate trips so each trip is unique. workers = jobs.
+OD_optimistic <- OD_optimistic %>%
+  group_by(VORIZT1259,VDESZT1259) %>%
+  summarise(trips_transit = sum(trips_transit),
+            trips_car_motor = sum(trips_car_motor),
+            trips_bike = sum(trips_bike),
+            trips_walk = sum(trips_walk),
+            workers_transit = mean(workers_transit),
+            # workers_transit = ifelse(is.na(workers_transit), 0, workers_transit),
+            workers_car = mean(workers_car),
+            # workers_car = ifelse(is.na(workers_car), 0, workers_car),
+            workers_bike = mean(workers_bike),
+            # workers_bike = ifelse(is.na(workers_bike), 0, workers_bike),
+            workers_walk = mean(workers_walk),
+            # workers_walk = ifelse(is.na(workers_walk), 0, workers_walk),
+            jobs_transit = mean(jobs_transit),
+            jobs_car = mean(jobs_car),
+            jobs_bike = mean(jobs_bike),
+            jobs_walk = mean(jobs_walk),
+            # jobs_car = ifelse(is.na(jobs_car), 0, jobs_car),
+            # jobs_transit = ifelse(is.na(jobs_transit), 0, jobs_transit),
+            # jobs_bike = ifelse(is.na(jobs_bike), 0, jobs_bike),
+            # jobs_walk = ifelse(is.na(jobs_walk), 0, jobs_walk),
+            tt_transit = ifelse(trips_transit != 0, mean(TIEMPO_VIAJES, na.rm=T), NA),
+            tt_car = ifelse(trips_car_motor != 0, mean(TIEMPO_VIAJES, na.rm=T), NA),
+            tt_bike = ifelse(trips_bike != 0, mean(TIEMPO_VIAJES, na.rm=T), NA),
+            tt_walk = ifelse(trips_walk != 0, mean(TIEMPO_VIAJES, na.rm=T), NA)) %>% ungroup()
+
+```
+
+now estimate f value using these averaged travel times:
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+#dgamma is the density plot for gamma function and dlnorm is the desntiy plot for the lognormal function; adding values as "f" our impedance function
+OD_optimistic <- OD_optimistic %>%
+  mutate(f_car = dgamma(tt_car, HW_car_motor_gamma$estimate["shape"], HW_car_motor_gamma$estimate["rate"]),
+         f_car = ifelse(is.na(f_car), 0, f_car),
+         f_transit = dlnorm(tt_transit, HW_transit_lnorm$estimate["meanlog"], HW_transit_lnorm$estimate["sdlog"]),
+         f_transit = ifelse(is.na(f_transit), 0, f_transit),
+         f_bike = dlnorm(tt_bike, HW_bike_lnorm$estimate["meanlog"], HW_bike_lnorm$estimate["sdlog"]),
+         f_bike = ifelse(is.na(f_bike), 0, f_bike),
+         f_walk = dlnorm(tt_walk, HW_walk_lnorm$estimate["meanlog"], HW_walk_lnorm$estimate["sdlog"]),
+         f_walk = ifelse(is.na(f_walk), 0, f_walk))
+```
+
+Calculating SA:
+```{r}
+#add new summed columns
+OD_optimistic <- OD_optimistic %>%
+  mutate(workers_ = (workers_car+ workers_transit+ workers_bike+ workers_walk),
+         workers_ = ifelse(is.na(workers_), 0, workers_),
+         jobs_ = (jobs_car + jobs_transit + jobs_bike + jobs_walk),
+         jobs_ = ifelse(is.na(jobs_), 0, jobs_),
+         f_ = f_car + f_transit + f_bike + f_walk)
+
+#calculate spatial availability
+OD_optimistic <- OD_optimistic %>%
+  mutate(catch = 1) %>%
+  mutate(SA_ij = sp_avail(., 
+                         o_id = VORIZT1259,
+                         d_id = VDESZT1259,
+                         pop = workers_,
+                         opp = jobs_,
+                         r = catch,
+                         f = f_),
+         SA_ij = ifelse(is.nan(SA_ij), 0, SA_ij))
+
+#verify that the sum of all jobs is consistent with the number of jobs
+sum(OD_optimistic$SA_ij)
+
+#verify that the sum of all jobs is consistent with the number of jobs - from another perspective
+sum_jobs <- OD_optimistic %>% 
+  group_by(VDESZT1259) %>% 
+  summarise(jobs = mean(jobs_))
+sum(sum_jobs$jobs, na.rm = T)
+#Perfect it works! 
+```
+
+Let's aggregate the spatial availability by origin zone:
+```{r}
+#aggregating spatial availability  
+MAD_availability <- OD_optimistic %>%
+  group_by(VORIZT1259) %>%
+  summarize(SA_i3 = sum(SA_ij),
+            workers_ = mean(workers_car, na.rm=T) + mean(workers_transit, na.rm=T) + mean(workers_bike, na.rm=T) + mean(workers_walk, na.rm=T),
+            avgtt_i = mean(mean(tt_transit) + mean(tt_walk) + mean(tt_car) + mean(tt_bike)),
+            avg_f_i = mean(mean(f_transit) + mean(f_walk) + mean(f_car) + mean(f_bike)))
+
+#Merge TO availability calculation to the TAZ sf object created for accessibility above:
+zn_test3 <- zn %>% merge(MAD_availability, by.x=c("ZT1259"), by.y=c("VORIZT1259"), all.x=T) 
+```
+
+Here we plot plot 2 (the original scenario) and plot 3 ( this optimistic scenario)
+```{r}
+#plot 2
+plot_MAD_avail_perworker_3 <- ggplot() +
+  geom_sf(data = zn_test3, aes(fill= SA_i3/workers_), color = NA) + #data
+    scale_fill_distiller(palette = "RdYlGn", #legend scale bar
+                         name = expression(v["i"]),
+                         na.value = "grey90",
+                         limits = c( round(min(zn_test3$SA_i3/zn_test3$workers_, na.rm=T)), 
+                                     round(max(zn_test3$SA_i3/zn_test3$workers_, na.rm=T))))+
+  geom_sf(data = Madrid_muni_border, # border for Madrid muni
+          colour=alpha("dimgrey",1), 
+          size = 0.5, fill=NA, 
+          show.legend = "polygon") +
+  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
+                         height = unit(0.8, "cm"), 
+                         width = unit(0.8, "cm"),
+                         style = north_arrow_orienteering(line_width = 0.25,
+                                                          line_col = "dimgrey", 
+                                                          fill = c("white","dimgrey"))) +
+  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
+                   height = unit(0.15, "cm")) +
+  theme_void() +
+  theme(legend.position = c(1.10, 0.25), 
+        plot.margin = margin(t = 0, r = 0, b = 0, l = -80)) # positing legend and clipping out white space
+
+# plot 1 + inset map
+plot_MAD_avail_perworker_3 <- ggdraw(plot_MAD_avail_perworker_3) +
+  draw_plot({plot_MAD_avail_perworker_3 + coord_sf(xlim = st_coordinates(bbox_Madrid_muni)[c(1,2),1],
+                                         ylim = st_coordinates(bbox_Madrid_muni)[c(2,3),2],
+                                         expand = FALSE) +
+      theme(legend.position = "none")},
+    x = 0.65, y = 0.50,  # The distance along a (0,1) x-axis (and y axis) to draw the left edge (the bottom edge) of the plot
+    width = 0.46, height = 0.46) # The width and height of the plot expressed as proportion of the entire ggdraw object
+
+plot_MAD_avail_perworker_2
+plot_MAD_avail_perworker_3
+```
+
+As can be seen, the change (visability) is not that dramatic.

### 2022-05-23 00-OD-preparation.Rmd
Created an additional .Rmd exploring 1) trip chained trips, 2) spatial availability calcs, 3) mode shifting due to LEZ

@@
 ```{r}
-t <- read_excel("data-raw/ZonificacionZT1259 complete.xlsx") %>% select(ZT1259, Municipality)
+t <- read_excel("data-raw/ZonificacionZT1259 complete.xlsx") %>% dplyr::select(ZT1259, Municipality)
 
@@
 ```{r}
-t <- OD_v_Oc_Dt %>% filter(DISTANCIA_VIAJE <= 60) %>% select(DISTANCIA_VIAJE) %>% unlist() %>% as.numeric() 
+t <- OD_v_Oc_Dt %>% filter(DISTANCIA_VIAJE <= 60) %>% dplyr::select(DISTANCIA_VIAJE) %>% unlist() %>% as.numeric() 
 hist(t)
@@
 ```{r}
-t <- OD_v_Oc_Dt %>% filter(DISTANCIA_VIAJE >= 60 & DISTANCIA_VIAJE <= 600) %>% select(DISTANCIA_VIAJE) %>% unlist() %>% as.numeric() 
+t <- OD_v_Oc_Dt %>% filter(DISTANCIA_VIAJE >= 60 & DISTANCIA_VIAJE <= 600) %>% dplyr::select(DISTANCIA_VIAJE) %>% unlist() %>% as.numeric() 
 hist(t, breaks = 50)
 
-t <- OD_v_Oc_Dt %>% filter(DISTANCIA_VIAJE >= 600) %>% select(DISTANCIA_VIAJE) %>% unlist() %>% as.numeric() 
+t <- OD_v_Oc_Dt %>% filter(DISTANCIA_VIAJE >= 600) %>% dplyr::select(DISTANCIA_VIAJE) %>% unlist() %>% as.numeric() 
 hist(t, breaks = 50)

### 2022-05-23 Simplified-spatial-availability-example.Rmd
Created an additional .Rmd exploring 1) trip chained trips, 2) spatial availability calcs, 3) mode shifting due to LEZ

--- 2022-05-08 Draft-manuscript-tobemoved.Rmd
@@
 ---
-title: "Draft-manuscript-tobemoved"
+title: "Simplified-spatial-availability-example"
 output: html_document
 ---
-
-```{r knitr-setup, include=FALSE}
-knitr::opts_chunk$set(
-  echo = FALSE,
-  cache = TRUE,
-  warning = FALSE,
-  message = FALSE,
-  comment = '', 
-  out.width = "1\\linewidth")
-```
-
-```{r install-data-package, eval = FALSE, include=FALSE}
-# if (!require("MadridZBE", character.only = TRUE)) {
-#       remotes::install_github("soukhova/MadridZBE",
-#                         build_vignettes = TRUE)
-#   }
-```
 
@@
 library(cowplot)
+library(ggspatial)
 ```
@@
 
-\newpage
-
-# Introduction
-
-- Some information about LEZ generally
-
-- Information about accessibility issues regarding LEZ; the tradition of using accessibility to evaluate these differences. Mention of spatial availability.
-
-- What we will be doing, the aim.
-
-\newpage
-
-# Using accessibility to evaluate inequity background
-
-- Information about inequity and LEZ may exasperate these issues
-
-- lit has found that emissions reduciton should not result in inequalities..
-
-\newpage
-
-# Methods
-
-## Data 
-- -plots of madrid zone, description of the LEZ scheme, how effective it is now and will be in the future - change in modal shares and emissions but what about equity
-
-## Estimated data
-- prediction in change in travel mode
-- prediction in private vehicle fleet label A/B/C proportions
-
-## Spatial availability
-- spatial availability explanation 
-  - impedance function explanation
-
-\newpage
-
-# Results
-
-## Fleet characteristics today vs. 2025
-<!-- to do -->
-
-## Impedance function calibration
-
+Import the trip data:
 ```{r data-import-filter}
 OD_v <- read_excel("data-raw/EDM2018VIAJES.xlsx")
-
-OD_v_HW <- OD_v %>% filter(VORI == "1" & VDES == "2") #only filter in origin casa trips and destination trabajo
-
-OD_v_HW <- OD_v_HW %>% 
-  mutate(TIEMPO_VIAJES = 
-                  as.numeric((as.POSIXct(OD_v_HW$VDESHORAFIN, format="%H%M") -
-                  as.POSIXct(OD_v_HW$VORIHORAINI, format="%H%M"))/60)) %>%
-  mutate(TIEMPO_VIAJES = ifelse(VDESHORAFIN == "2410", 40, 
-                                         ifelse(VDESHORAFIN == "2440", 55, 
-                                                ifelse(VDESHORAFIN == "2800", 420, TIEMPO_VIAJES)))) %>% # some destination times are greater than 2400 and so yield an NA different, correct these 3 trips
-  mutate(TIEMPO_VIAJES = ifelse(TIEMPO_VIAJES == 0, 0.1, TIEMPO_VIAJES)) # set all 0min travel times to 0.1 min
-
-summary(OD_v_HW$TIEMPO_VIAJES)
 ```
 
+# A) Exploring chained-trips... should we include them?
+
+Calculate the duration of each trip
+```{r}
+OD_v <- OD_v %>%
+  mutate(TIEMPO_VIAJES =
+           (as.numeric(OD_v$VDESHORAFIN) - 
+                    as.numeric(OD_v$VORIHORAINI))) %>%
+  mutate(TIEMPO_VIAJES = ifelse(TIEMPO_VIAJES == 0, 0.1, TIEMPO_VIAJES)) # set all 0min travel times to 0.1 min
+summary(OD_v$TIEMPO_VIAJES)
+```
+Checking the trip data - it appears each OD is a row, i.e., chained trips are multiple rows.
+We want to extract every home origin (VORI = 1) to every work destination (VDEST =2) trip as a single row...
+so let's do this in multiple steps.
+
+First, the direct home to work trips:
+```{r}
+# not chained home to work trips - direct
+OD_v_HW <- OD_v %>% filter(VORI == "1" & VDES == "2") #only filter in origin casa trips and destination trabajo
+```
+
+Next, let's identify the chained trips and then consolidate them. We only care about trip O/D, travel mode, travel time, and 
+```{r}
+t <- OD_v %>% group_by(ID_HOGAR,ID_IND) %>% 
+  filter((VORI == "1" & VDES != "2" & VDES != "1") | (VDES == "2" & VORI != "1"& VORI != "2")) %>% #here the origin NEEDS to be from home, the destination can be anywhere OR the origin can be from anywhere but the destination NEEDS to be work.
+  mutate(chained = lag(VDES) - VORI)
+
+chained2 <- (ifelse(t$chained[-1]==0 |t$chained==0, 0, NA)) %>% as.data.frame()
+
+t <- cbind(t,chained2) %>% dplyr::rename("chained2" = ".")
+
+t<-t %>% filter(chained2==0) # these are all the CHAINED trips - only considering 1 location before work.  
+t
+```
+
+Now... let's only include chained trips with the same mode! 
+```{r}
+t <- t %>% group_by(ID_HOGAR, ID_IND) %>% mutate(trip = ifelse(((ID_VIAJE - lag(ID_VIAJE)) == 1)
+                                                          & ((MODO_PRIORITARIO - lag(MODO_PRIORITARIO))== 0), 
+                                                          MODO_PRIORITARIO, 0))
+
+t %>% filter(trip == 0) %>% nrow() / t %>% filter(chained == 0) %>% nrow()
+t %>% filter(trip != 0 & !is.na(trip)) %>% nrow() / t %>% filter(chained == 0) %>% nrow()
+```
+so ~27% of chained trips are not made using the same mode.
+THe rest of them are (`r t %>% filter(trip != 0 & !is.na(trip)) %>% nrow() ` trips are chained (1 intermediate destination) and use the same mode).
+
+So my question is...
+should we include these 1,371 extra chained trips (this is if we consolidate these 2 mini trips into 1 home to work trip (with an intermediate stop)).
+This only represent ~5% additional trips to the direct home to work 27K trips.
+If we include them...
+how do we count travel time?
+just sum both steps of the trip together..?
+Is that a good idea..?
+I don't think so.
+
+Let's see if this sample increases the number of ZTs spatially:
+```{r}
+#bind the t object (chained) to the workers object. 
+t <- t %>% filter(is.na(chained)) # only including origins
+t <- t[,1:20] 
+OD_v_HW_chained <- rbind(OD_v_HW, t)
+```
+
+```{r}
+#importing zones
+zn <- st_read("data-raw/zn.shp")
+ALLLEZ <- st_read("data-raw/Zona_Bajas_Emisiones_MADRID.shp")
+
+temp <- read_excel("data-raw/ZonificacionZT1259 complete.xlsx") %>% dplyr::select(ZT1259, Municipality)
+zn_t <- zn %>% merge(temp, by = c("ZT1259"), all.x=TRUE)
+```
+
+```{r}
+Otrips_zona_c <- OD_v_HW_chained %>% group_by(VORIZT1259)  %>% summarise(Otrips = (n())) # we group by the origin ZT1259 ID and then we summarize the number of rows that have the same ZT using the summarize() function and the n() function to create a new 'Otrips' column.
+
+zn_c <- zn %>% merge(Otrips_zona_c, by.x = "ZT1259", by.y="VORIZT1259", all.x = TRUE)
+
+ggplot() +
+  geom_sf(data = zn_c,
+          aes(fill = Otrips),
+          color  = NA) +
+  scale_fill_distiller(palette = "Spectral")
+
+Otrips_zona <- OD_v_HW %>% group_by(VORIZT1259)  %>% summarise(Otrips = (n())) # we group by the origin ZT1259 ID and then we summarize the number of rows that have the same ZT using the summarize() function and the n() function to create a new 'Otrips' column.
+
+zn_noc <- zn %>% merge(Otrips_zona, by.x = "ZT1259", by.y="VORIZT1259", all.x = TRUE)
+
+ggplot() +
+  geom_sf(data = zn_noc,
+          aes(fill = Otrips),
+          color  = NA) +
+  scale_fill_distiller(palette = "Spectral")
+```
+
+```{r}
+summary(zn_noc$Otrips)
+summary(zn_c$Otrips)
+```
+
+so as we can see... the number of origin ZNs with NO origin trips only goes down by 2.. (from 344 to 342). This is not a lot.
+
+Next step... let's try to see what is happening in these NA zones. What sort of origin trips are being generated? 
+```{r}
+missing_workers <- zn_noc %>% filter(is.na(Otrips)) %>% dplyr::select(ZT1259) #344 zonas are missing home to work trips.
+
+missing_workers <- missing_workers %>% merge(OD_v, by.x = "ZT1259", by.y="VORIZT1259", all.x = TRUE)
+
+temp <- missing_workers %>% st_drop_geometry() %>% group_by(ZT1259) %>% summarise(trip_count = n(),
+                                                                          Owork_perc = sum(VORI == 2)/n(),
+                                                                          Ostudy_perc = sum(VORI == 4)/n(),
+                                                                          Osportwalk_perc = sum(VORI == 10)/n(),
+                                                                          Oshop_perc = sum(VORI == 5)/n(),
+                                                                          Ohome_perc = sum(VORI == 1)/n())
+temp
+```
+```{r}
+summary(temp)
+```
+```{r}
+# write.csv(missing_workers %>% st_drop_geometry(),"ZT_Trips_missingworkers.csv", row.names = FALSE)
+```
+
+‘trip count’ is how many trips there are in each ZT.
+‘Owork_perc’ is an origin work purpose (of that trip) in each ZT, i.e.
+39.47% of origin trips are work related averaged across all ZTs.
+‘Ostudy_perc’ is for study origin trips..
+etc.
+You can see that the majority of origin trip purposes are work.
+This means that these are predominately ‘work’ related areas and none of the households surveyed live in those ZTs.
+
+And looking in excel.
+even if we consider all origin zones from work themed chained trips (i.e., home to somewhere to work) as 'residential locations' (this is incorrect) we STILL have ~170 zones with no work-to-home trips.
+
+# B) Calculating spatial availability - an exploration of 3 approaches
+
+For simplicity.
+Let's assume we are only considering home to work direct trips.
+Let's now manually calculate spatial availability $V_{ij}$.
+(where i is each origin (home) and j is each destination (work))
+
+$$
+V_{ij} = O_j\frac{F^p_{ij} \cdot F^c_{ij}}{\sum_{i=1}^K F^p_{ij} \cdot F^c_{ij}}
+$$
+The population allocation factor is based on _worker population_ for each ij trip $F^p_{ij}$. No function is needed here, we are just going to work with the proportions.
+
+Also simple - $O_{j}$ - represents the number of jobs at a destination $j$.
+
+The complexity we have for this case study is the travel cost allocation factor $F^c_{ij}$ due to the 4 modes.
+
+Think of $F^c_{ij}$ as the travel cost for each ij trip.
+This travel cost is based on the fitted impedance function which is calculated for each mode (for our sample, log-normal is selected for all modes other then car - car is gamma).
+These impedance functions are predicted using each travel time for each mode (i.e., 4 modes, so we have 4 impedance functions).
+The impedance functions will be used to predict the _mode relative_ cost of travel, specifically, the higher the value, the stronger the travel-time decay effect on spatial availability will be.
+
+Ultimately, for each origin to destination, the product of $F^c_{ij}$ and $F^p_{ij}$ is divided by the SUM of all $F^c_{ij}$ and $F^p_{ij}$ in that region $K$ (all the origins $i$ that travel to a destination $j$).
+
+We must manipulate the data before we can fit the impedance function:
 ```{r data-for-impedance}
@@
   dplyr::select(TIEMPO_VIAJES)
-
 ```
 
-plotting histograms:
-```{r}
-hist(transit_HW_alltt%>% unlist() %>% as.numeric(), breaks=100)
-hist(car_motor_HW_alltt%>% unlist() %>% as.numeric(), breaks=100)
-hist(bike_HW_alltt%>% unlist() %>% as.numeric(), breaks=100)
-hist(walk_HW_alltt%>% unlist() %>% as.numeric(), breaks=100)
-```
-```{r}
-#cullen grey graph - BUS
-descdist(transit_HW_alltt%>% unlist() %>% as.numeric(), discrete=FALSE, boot=500)
-```
+Then create the impedance functions:
+```{r fitting-impedance-function}
+# using fitdist function to fit a distribution using the default maximum likelihood estimation method and Nelder-Mead method for direct optimization
+lnorm_ <- fitdistrplus::fitdist(data=transit_HW_alltt%>% unlist() %>% as.numeric(), "lnorm", method="mle", optim.method="Nelder-Mead")
+HW_transit_lnorm <- lnorm_
 
-```{r bus-itting-impedance-function}
-# using fitdist function to fit a distribution using the default maximum likelihood estimation method and Nelder-Mead method for direct optimization
-gamma_ <- fitdistrplus::fitdist(data=transit_HW_alltt%>% unlist() %>% as.numeric(), "gamma", method="mle", optim.method="Nelder-Mead") 
-lnorm_ <- fitdistrplus::fitdist(data=transit_HW_alltt%>% unlist() %>% as.numeric(), "lnorm", method="mle", optim.method="Nelder-Mead")
-norm_ <-fitdistrplus::fitdist(data=transit_HW_alltt%>% unlist() %>% as.numeric(), "norm", method="mle", optim.method="Nelder-Mead")
-exp_ <- fitdistrplus::fitdist(data=transit_HW_alltt%>% unlist() %>% as.numeric(), "exp", method="mle", optim.method="Nelder-Mead")
-# pois_ <- fitdistrplus::fitdist(data=transit_HW_alltt%>% unlist() %>% as.numeric(), "pois", method="mle", optim.method="Nelder-Mead")
-# nbinom_ <- fitdistrplus::fitdist(data=transit_HW_alltt%>% unlist() %>% as.numeric(), "nbinom", method="mle", optim.method="Nelder-Mead")
-# geom_ <- fitdistrplus::fitdist(data=transit_HW_alltt%>% unlist() %>% as.numeric(), "geom", method="mle", optim.method="Nelder-Mead")
-# beta_ <- fitdistrplus::fitdist(data=transit_HW_alltt%>% unlist() %>% as.numeric(), "beta", method="mle", optim.method="Nelder-Mead")
-logis_ <- fitdistrplus::fitdist(data=transit_HW_alltt%>% unlist() %>% as.numeric(), "logis", method="mle", optim.method="Nelder-Mead")
+gamma_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt%>% unlist() %>% as.numeric(), "gamma", method="mle", optim.method="Nelder-Mead")
+HW_car_motor_gamma <- gamma_
 
-plot(gamma_)
-plot(lnorm_)
-plot(norm_)
-plot(exp_)
-# plot(pois_)
-# plot(nbinom_)
-# plot(geom_)
-# plot(beta_)
-plot(logis_)
-```
-now let's compare these models AIC and BIC to see which fits the bus data best:
-```{r}
-broom::glance(MASS::fitdistr(transit_HW_alltt%>% unlist() %>% as.numeric(),"gamma"))
-broom::glance(MASS::fitdistr(transit_HW_alltt%>% unlist() %>% as.numeric(),"lognormal"))
-broom::glance(MASS::fitdistr(transit_HW_alltt%>% unlist() %>% as.numeric(),"normal"))
-broom::glance(MASS::fitdistr(transit_HW_alltt%>% unlist() %>% as.numeric(),"exponential"))
-# plot(pois_)
-# plot(nbinom_)
-# plot(geom_)
-# plot(beta_)
-broom::glance(MASS::fitdistr(transit_HW_alltt%>% unlist() %>% as.numeric(),"logistic"))
-```
+lnorm_ <- fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "lnorm", method="mle", optim.method="Nelder-Mead")
+HW_bike_lnorm <- lnorm_
 
-lognormal has largest logLik and the smallest AIC and BIC. we will pick this lognormal function for transit trips.
-```{r}
-HW_transit_lnorm <- lnorm_
-```
-
-Let's do car/motorcycles now:
-```{r car-motor-itting-impedance-function}
-# using fitdist function to fit a distribution using the default maximum likelihood estimation method and Nelder-Mead method for direct optimization
-gamma_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt%>% unlist() %>% as.numeric(), "gamma", method="mle", optim.method="Nelder-Mead") 
-lnorm_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt%>% unlist() %>% as.numeric(), "lnorm", method="mle", optim.method="Nelder-Mead")
-norm_ <-fitdistrplus::fitdist(data=car_motor_HW_alltt%>% unlist() %>% as.numeric(), "norm", method="mle", optim.method="Nelder-Mead")
-exp_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt%>% unlist() %>% as.numeric(), "exp", method="mle", optim.method="Nelder-Mead")
-# pois_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt%>% unlist() %>% as.numeric(), "pois", method="mle", optim.method="Nelder-Mead")
-# nbinom_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt%>% unlist() %>% as.numeric(), "nbinom", method="mle", optim.method="Nelder-Mead")
-# geom_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt%>% unlist() %>% as.numeric(), "geom", method="mle", optim.method="Nelder-Mead")
-# beta_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt%>% unlist() %>% as.numeric(), "beta", method="mle", optim.method="Nelder-Mead")
-logis_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt%>% unlist() %>% as.numeric(), "logis", method="mle", optim.method="Nelder-Mead")
-
-plot(gamma_)
-plot(lnorm_)
-plot(norm_)
-plot(exp_)
-# plot(pois_)
-# plot(nbinom_)
-# plot(geom_)
-# plot(beta_)
-plot(logis_)
-```
-now let's compare these models AIC and BIC to see which fits the car data best:
-```{r}
-broom::glance(MASS::fitdistr(car_motor_HW_alltt%>% unlist() %>% as.numeric(),"gamma"))
-broom::glance(MASS::fitdistr(car_motor_HW_alltt%>% unlist() %>% as.numeric(),"lognormal"))
-broom::glance(MASS::fitdistr(car_motor_HW_alltt%>% unlist() %>% as.numeric(),"normal"))
-broom::glance(MASS::fitdistr(car_motor_HW_alltt%>% unlist() %>% as.numeric(),"exponential"))
-# plot(pois_)
-# plot(nbinom_)
-# plot(geom_)
-# plot(beta_)
-broom::glance(MASS::fitdistr(car_motor_HW_alltt%>% unlist() %>% as.numeric(),"logistic"))
-```
-gamma has largest logLik and the smallest AIC and BIC. we will pick this gamma function for car trips.
-```{r}
-HW_car_motor_gamma <- gamma_
-```
-
-Let's do bike now:
-```{r bike-itting-impedance-function}
-# using fitdist function to fit a distribution using the default maximum likelihood estimation method and Nelder-Mead method for direct optimization
-gamma_ <- fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "gamma", method="mle", optim.method="Nelder-Mead") 
-lnorm_ <- fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "lnorm", method="mle", optim.method="Nelder-Mead")
-norm_ <-fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "norm", method="mle", optim.method="Nelder-Mead")
-exp_ <- fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "exp", method="mle", optim.method="Nelder-Mead")
-# pois_ <- fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "pois", method="mle", optim.method="Nelder-Mead")
-# nbinom_ <- fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "nbinom", method="mle", optim.method="Nelder-Mead")
-# geom_ <- fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "geom", method="mle", optim.method="Nelder-Mead")
-# beta_ <- fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "beta", method="mle", optim.method="Nelder-Mead")
-logis_ <- fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "logis", method="mle", optim.method="Nelder-Mead")
-
-plot(gamma_)
-plot(lnorm_)
-plot(norm_)
-plot(exp_)
-plot(logis_)
-```
-now let's compare these models AIC and BIC to see which fits the bike data best:
-```{r}
-broom::glance(MASS::fitdistr(bike_HW_alltt%>% unlist() %>% as.numeric(),"gamma"))
-broom::glance(MASS::fitdistr(bike_HW_alltt%>% unlist() %>% as.numeric(),"lognormal"))
-broom::glance(MASS::fitdistr(bike_HW_alltt%>% unlist() %>% as.numeric(),"normal"))
-broom::glance(MASS::fitdistr(bike_HW_alltt%>% unlist() %>% as.numeric(),"exponential"))
-broom::glance(MASS::fitdistr(bike_HW_alltt%>% unlist() %>% as.numeric(),"logistic"))
-```
-lognormal has largest logLik and the smallest AIC and BIC. we will pick lognormal function for bike trips! Interesting how car mode is different from all other modes.
-
-```{r}
-HW_bike_lnorm <- lnorm_
-```
-
-Let's do walk now:
-```{r walk-itting-impedance-function}
-# using fitdist function to fit a distribution using the default maximum likelihood estimation method and Nelder-Mead method for direct optimization
-gamma_ <- fitdistrplus::fitdist(data=walk_HW_alltt%>% unlist() %>% as.numeric(), "gamma", method="mle", optim.method="Nelder-Mead") 
 lnorm_ <- fitdistrplus::fitdist(data=walk_HW_alltt%>% unlist() %>% as.numeric(), "lnorm", method="mle", optim.method="Nelder-Mead")
-norm_ <-fitdistrplus::fitdist(data=walk_HW_alltt%>% unlist() %>% as.numeric(), "norm", method="mle", optim.method="Nelder-Mead")
-exp_ <- fitdistrplus::fitdist(data=walk_HW_alltt%>% unlist() %>% as.numeric(), "exp", method="mle", optim.method="Nelder-Mead")
-# pois_ <- fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "pois", method="mle", optim.method="Nelder-Mead")
-# nbinom_ <- fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "nbinom", method="mle", optim.method="Nelder-Mead")
-# geom_ <- fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "geom", method="mle", optim.method="Nelder-Mead")
-# beta_ <- fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "beta", method="mle", optim.method="Nelder-Mead")
-logis_ <- fitdistrplus::fitdist(data=walk_HW_alltt%>% unlist() %>% as.numeric(), "logis", method="mle", optim.method="Nelder-Mead")
-
-plot(gamma_)
-plot(lnorm_)
-plot(norm_)
-plot(exp_)
-plot(logis_)
-```
-now let's compare these models AIC and BIC to see which fits the walk data best:
-```{r}
-broom::glance(MASS::fitdistr(walk_HW_alltt%>% unlist() %>% as.numeric(),"gamma"))
-broom::glance(MASS::fitdistr(walk_HW_alltt%>% unlist() %>% as.numeric(),"lognormal"))
-broom::glance(MASS::fitdistr(walk_HW_alltt%>% unlist() %>% as.numeric(),"normal"))
-broom::glance(MASS::fitdistr(walk_HW_alltt%>% unlist() %>% as.numeric(),"exponential"))
-broom::glance(MASS::fitdistr(walk_HW_alltt%>% unlist() %>% as.numeric(),"logistic"))
-```
-lognormal has largest logLik and the smallest AIC and BIC. we will pick lognormal function for walk trips!
-
-```{r}
 HW_walk_lnorm <- lnorm_
@@
 
-Add the number of trips (by mode) and create objects which contain the number of workers (origin) and jobs (destinations) - by mode:
+Then we add the number of trips (by mode) and create objects which contain the number of workers (origin) and jobs (destinations) - by mode:
 ```{r}
@@
 
-Join workers and jobs to the ZN simple features object:
+Next join workers and jobs to the ZN simple features object:
 ```{r}
@@
 
-add the number of workers and jobs from each zona to each trip:
+Then add the number of workers and jobs from each ZN to each trip:
 ```{r}
@@
 
-now reshape the data, each row will represent an any origin to to 1 destination (i.e., destination row is unique).
-We will average the travel times, by mode, and then use this average travel time to find the 'f' (impedance value for travel for each mode):
+Now reshape the data, each row will represent an any origin to to 1 destination (i.e., destination row is unique).
+We will average the travel times, by mode, and then use this average travel time to find the 'f' (impedance value for travel for each mode).
+Hint: this is why we needed to find the impedance function in the first place!
+The functions represent the cost of travel for each mode based on the travel time for a trip, since we are averaging together the travel time for each mode from
 ```{r}
@@
             trips_walk = sum(trips_walk),
-            workers_transit = mean(workers_transit),
-            # workers_transit = ifelse(is.na(workers_transit), 0, workers_transit),
+            workers_transit = mean(workers_transit), #workers_ and jobs_ values are retained from the ZN object - so they are the same value for each OD pair thus we use mean() here. 
             workers_car = mean(workers_car),
-            # workers_car = ifelse(is.na(workers_car), 0, workers_car),
             workers_bike = mean(workers_bike),
-            # workers_bike = ifelse(is.na(workers_bike), 0, workers_bike),
             workers_walk = mean(workers_walk),
-            # workers_walk = ifelse(is.na(workers_walk), 0, workers_walk),
             jobs_transit = mean(jobs_transit),
@@
             jobs_walk = mean(jobs_walk),
-            # jobs_car = ifelse(is.na(jobs_car), 0, jobs_car),
-            # jobs_transit = ifelse(is.na(jobs_transit), 0, jobs_transit),
-            # jobs_bike = ifelse(is.na(jobs_bike), 0, jobs_bike),
-            # jobs_walk = ifelse(is.na(jobs_walk), 0, jobs_walk),
-            tt_transit = ifelse(trips_transit != 0, mean(TIEMPO_VIAJES, na.rm=T), NA),
+            tt_transit = ifelse(trips_transit != 0, mean(TIEMPO_VIAJES, na.rm=T), NA), #time is averaged for trips from the same origin to destination + same mode - there likely is variation between trips as these come from the survey (e.g., trips taken at different time of day so they are of different length)
             tt_car = ifelse(trips_car_motor != 0, mean(TIEMPO_VIAJES, na.rm=T), NA),
@@
 
-# OD_test <- OD %>%
-#   mutate(workers_ = (workers_car+ workers_transit+ workers_bike+ workers_walk),
-#          workers_ = ifelse(is.na(workers_), 0, workers_),
-#          workers_car = ifelse(is.na(workers_car), 0, workers_car),
-#          workers_transit = ifelse(is.na(workers_transit), 0, workers_transit),
-#          workers_bike = ifelse(is.na(workers_bike), 0, workers_bike),
-#          workers_walk = ifelse(is.na(workers_walk), 0, workers_walk),
-#            jobs_ = (jobs_car+ jobs_transit+ jobs_bike+ jobs_walk),
-#          jobs_ = ifelse(is.na(jobs_), 0, jobs_),
-#          jobs_car = ifelse(is.na(jobs_car), 0, jobs_car),
-#          jobs_transit = ifelse(is.na(jobs_transit), 0, jobs_transit),
-#          jobs_bike = ifelse(is.na(jobs_bike), 0, jobs_bike),
-#          jobs_walk = ifelse(is.na(jobs_walk), 0, jobs_walk),
-#          tt_transit = ifelse(trips_transit != 0, TIEMPO_VIAJES, NA),
-#             tt_car = ifelse(trips_car_motor != 0, TIEMPO_VIAJES, NA),
-#             tt_bike = ifelse(trips_bike != 0, TIEMPO_VIAJES, NA),
-#             tt_walk = ifelse(trips_walk != 0, TIEMPO_VIAJES, NA)) %>% ungroup()
-
-#check, should add to the origin number of trips 27795 - perfect!
-# sum(OD$trips_transit) + sum(OD$trips_car_motor) + sum(OD$trips_bike) + sum(OD$trips_walk)
+#check, should add up to the origin number of trips 27795 - perfect!
+sum(OD$trips_transit) + sum(OD$trips_car_motor) + sum(OD$trips_bike) + sum(OD$trips_walk)
+sum(OD_test$trips_transit) + sum(OD_test$trips_car_motor) + sum(OD_test$trips_bike) + sum(OD_test$trips_walk)
 ```
@@
 
-```{=tex}
-\begin{equation}
-\label{gamma-dist}
-\begin{array}{l}\ 
-f(x, \alpha, \beta) = \frac {x^{\alpha-1}e^{-\frac{x}{\beta}}}{ \beta^{\alpha}\Gamma(\alpha)} \quad \text{for } 0 \leq x \leq \infty\\
-
-\Gamma(\alpha) =  \int_{0}^{\infty} x^{\alpha-1}e^{-x} \,dx\\
-\end{array}
-\end{equation}
-```
-
-## Spatial availability comparision
+## Comparing 2 spatial availability calculations
 
 ```{r}
-#calculate spatial availability
+#calculate spatial availability using the 'sp_avail' function which is defined at the top of the .Rmd
+#this is the FIRST way of calculating SA.
+This calculates SA for each mode seperately (e.g., the competition of car trips only impacts workers who travel by car, etc.) - I believe this method is less realistic.
 OD_SA <- OD_test %>%
@@
 
-Let's try calculating SA from a different approch:
+Let's try calculating SA from a different approach:
 ```{r}
-# Also another question - this configuration assumes competition is contained within modes - i.e., transit workers compete for transit jobs, etc.
-This isn't the most realistic I think.
-So let's sum up all the mode specific workers and jobs and impedance values BEFORE calculating SA so we only have 1 SA
+# Here's SA calculation 2 - instead of separating each mode, I calculate the SA for all modes together so the competition is felt for all workers (not mode specific).
 
-#add new summed columns
-OD_SA <- OD_SA %>%
+# here I sum up all the mode specific workers and jobs and impedance values BEFORE calculating SA so we only have 1 resulting SA value for the OD pairs and not 4 (one for each mode) like the other approach.
+OD_SA1 <- OD_SA %>%
   mutate(workers_ = (workers_car+ workers_transit+ workers_bike+ workers_walk),
@@
 
+# as you can see, sometimes an OD row has multiple modes taking that trip - the f_ value sums all modes in that row. 
+head(OD_SA1 %>% dplyr::select(f_, f_car, f_transit, f_bike, f_walk), 25)
+#*note: summing these all together certainly distorts the travel cost proportional allocation but the correct number of workers (by mode) are multiplied in...so the SA adds up to the number of workers which is a good check.
+Still not sure what the impact of this is...
+```
+
+Now calculate SA using this second approach (all workers (not mode specific) competite with eachother) and preform some checks:
+```{r}
 #calculate spatial availability
-OD_SA2 <- OD_SA %>%
+OD_SA2 <- OD_SA1 %>%
   mutate(catch = 1) %>%
@@
 
-Let's aggregate the spatial availability by origin zone for these 2 methods (calculating SA separately... and calculating 1 SA that combines all modes)
+Now let's try a third approach. A mix of both examples. imagine:
+if 1 origin destination pair has 2 transit trips and 2 car trips, I want to calculate the SA for each trip considering ALL 4 trips (not like in the first attempt when I considered only 2 trips transit, 2 trips car, and then added them together).
+So it would look like:
+
+t1 / t1+t2+c1+c2
+t2 / t1+t2+c1+c2
+c1 / t1+t2+c1+c2
+c2 / t1+t2+c1+c2
+
+```{r}
+o_id3 <- rep(OD_SA2$VORIZT1259,4)  %>% as.data.frame()
+d_id3 <-rep(OD_SA2$VDESZT1259,4)   %>% as.data.frame()
+
+# catch3 <- rep(1, nrow(d_id3)) %>% data.frame() 
+catch3 <- stack(OD_SA2[,3:6]) %>% dplyr::select(values) #trips (transit, car, bike, walk)
+
+workers_3 <-  rep(OD_SA2$workers_,4)  %>% as.data.frame()
+jobs_3 <- rep(OD_SA2$jobs_,4)  %>% as.data.frame()
+# workers_3 <- stack(OD_SA2[,7:10]) %>% dplyr::select(values)
+# jobs_3 <- stack(OD_SA2[,11:14]) %>% dplyr::select(values)
+#f_3 <- rep(OD_SA2$f_,4)  %>% as.data.frame()
+f_3 <- stack(OD_SA2[,c(20,19,21:22)]) %>% dplyr::select(values)
+
+OD_SA3 <- cbind(o_id3, d_id3, catch3, workers_3, jobs_3, f_3) %>% data.frame()
+names(OD_SA3)[1:6] <- c("o_id3", "d_id3", "catch3","workers_3", "jobs_3", "f_3")
+
+OD_SA3 <- OD_SA3 %>% mutate(catch3 = ifelse(catch3 != 0, 1, 0)) 
+summary(OD_SA3)
+```
+
+```{r}
+#calculate spatial availability
+OD_SA3 <- OD_SA3 %>%
+  mutate(catch = 1) %>%
+  mutate(SA_ij_3 = sp_avail(.,
+                         o_id = o_id3,
+                         d_id = d_id3,
+                         pop = workers_3, 
+                         opp = jobs_3,
+                         r = catch,
+                         f = f_3))
+
+#verify that the sum of all jobs is consistent with the number of jobs
+sum(OD_SA3$SA_ij_3)
+
+```
+What's interesting is this third approach yields the same as the second approach.
+Meaning, when we use to total sum of workers and jobs (across all modes) for each OD pair.
+repeat the calculation for each mode specific 'f' (i.e., 4 times for each OD pair), and sum up the SA_ij_3 for each OD (each OD has 4 values due to for modes) the resulting value is the same as if the summation of 'f' was done BEFORE the SA calculation.
+
+These 2 approaches are definitely different than the first approach which only considers the workers and jobs which are taken BY the mode.
+Effectively, this approach calculates the SA_ij occurring for each mode - independent of the workers / jobs / travel costs of other modes.
+This 4 SA_ij calculations can also be all summed together and equal back to 27434 (the number of total jobs)
+```{r}
+OD_SA3 %>% group_by(o_id3, d_id3) %>% summarize(f_3 = sum(f_3),
+                                                SA_ij_3 = sum(SA_ij_3))
+OD_SA2 %>% group_by(VORIZT1259, VDESZT1259) %>% summarize(f_2 = sum(f_),
+                                                          SA_ij2 = sum(SA_ij))
+OD_SA2 %>% group_by(VORIZT1259, VDESZT1259) %>% summarize(f_1 = sum(f_car, f_bike, f_walk, f_transit), 
+                                                          SA_ij1 = sum(SA_ij_car, SA_ij_bike, SA_ij_transit, SA_ij_walk))
+
+```
+
+# Visually comparing the 2 approaches
+Let's aggregate the spatial availability by origin zone for the first 2 methods (calculating SA separately... and calculating 1 SA that combines all modes) and visually compare 
 ```{r}
@@
 
-plot_MAD_avail1
-plot_MAD_avail2
+plot_MAD_avail1 # the mode specific SA calculation and then sum them together - approach 1
+plot_MAD_avail2 # the f_, worker_, job_, summed together BEFORE SA calculation - approach 2
 ```
@@
 
-plot_MAD_avail_perworker_1
-plot_MAD_avail_perworker_2
+plot_MAD_avail_perworker_1 # the mode specific SA calculation and then sum them together - approach 1
+plot_MAD_avail_perworker_2 # the f_, worker_, job_, summed together BEFORE SA calculation - approach 2
 ```
 
-## An optimistic scenario - no cars going into madrid municipality
+# C) Exploring mode shifting, first an optimistic scenario - no cars (all transit) going into madrid municipality
 
@@
   mutate(trips_transit = sum(trips_transit,trips_car_motor, na.rm = T),
-         trips_car_motor = 0)
+         trips_car_motor = 0) #16627 ODs are making trips INTO muni, this is 59% of all OD pairs
 
@@
 
-Now again, join the workers and jobs to the ZN simple features object:
+Let's quickly compare how many ODs which have car trips changed to transit had EXISTING transit trips. This may inform what travel time we use.
+```{r}
+madrid_trips_reassigned <- OD_v_HW_optimistic1 %>% 
+  group_by(VORIZT1259, VDESZT1259) %>%
+  summarise(NEWtrips_transit = sum(trips_transit),
+         NEWtrips_car_motor = sum(trips_car_motor))
+
+madrid_trips_original <-OD_v_HW %>%
+  group_by(VORIZT1259, VDESZT1259) %>%
+  summarise(trips_transit = sum(trips_transit),
+         trips_car_motor = sum(trips_car_motor))
+
+madrid_trips_reassigned <- madrid_trips_reassigned %>% merge(madrid_trips_original, by =c("VORIZT1259","VDESZT1259" ))
+#all the trip_.y, are the original trips as specified in the travel survey. the trips.x are the assigned trips (moved to transit).
+
+# percentage of transit trips that existed in the original OD pair
+
+countt <- madrid_trips_reassigned %>% mutate(transit_count = ifelse(trips_transit != 0, 1, 0))
+
+sum(countt$transit_count)/nrow(madrid_trips_reassigned)
+```
+
+We can see that only ~50% of the OD pairs had a transit trip originally - so we can only borrow transit trip times from thest ODs...
+we'd have to figure out how to subsitute the remaining 50%.
+Alternatively - we can figure out any other solution to apply to all mode-shifted trips.
+
+For simplicity, let's continue and  just take the travel time (for car) for these trips and put it on the transit impedance curve.
+
+To continue this process, join the workers and jobs to the ZN simple features object:
 ```{r}
@@
 now reshape the data, each row will represent an any origin to to 1 destination (i.e., destination row is unique).
-We will average the travel times, by mode, and then use this average travel time to find the 'f' (impedance value for travel for each mode):
+We will average the travel times, by mode, and then use this average travel time to find the 'f' (impedance value for travel for each mode).
+Note: this is a *KEY* assumption - I'm giving the new transit trips the SAME travel time as they had when traveling by car but I am predicting the travel cost using the impedance curve of TRANSIT instead of CAR - that's the only difference.
+This is certainly a strange assumption.
+A solution could be to use the r5r calculated transit travel times but as we saw they don't fit the original data travel times well...
 ```{r}

### 2022-05-26 Simplified-spatial-availability-example.Rmd
Minor fixes to travel time averages

@@
 # sum(zn$jobs_transit) + sum(zn$jobs_bike) + sum(zn$jobs_car) + sum(zn$jobs_walk)
+
+# in 'zn' now we have how many people go to work from each zone (workers) and the number of people that arrive to that zone (jobs). By mode. 
 ```
@@
 ```{r}
-#consolidate trips so each trip is unique. workers = jobs.
-OD_test <- OD %>%
+# #consolidate trips so each trip is unique. workers = jobs.
+# OD_test <- OD %>%
+#   group_by(VORIZT1259,VDESZT1259) %>%
+#   summarise(trips_transit = sum(trips_transit),
+#             trips_car_motor = sum(trips_car_motor),
+#             trips_bike = sum(trips_bike),
+#             trips_walk = sum(trips_walk),
+#             workers_transit = mean(workers_transit), #workers_ and jobs_ values are retained from the ZN object - so they are the same value for each OD pair thus we use mean() here. 
+#             workers_car = mean(workers_car),
+#             workers_bike = mean(workers_bike),
+#             workers_walk = mean(workers_walk),
+#             jobs_transit = mean(jobs_transit),
+#             jobs_car = mean(jobs_car),
+#             jobs_bike = mean(jobs_bike),
+#             jobs_walk = mean(jobs_walk)) %>% ungroup()
+# 
+# OD_test <- OD_test %>% mutate(
+#             tt_transit = ifelse(trips_transit != 0, mean(TIEMPO_VIAJES, na.rm=T), NA), #time is averaged for trips from the same origin to destination + same mode - there likely is variation between trips as these come from the survey (e.g., trips taken at different time of day so they are of different length)
+#             tt_car = ifelse(trips_car_motor != 0, mean(TIEMPO_VIAJES, na.rm=T), NA), #NOTE: CHECK!!
+#             tt_bike = ifelse(trips_bike != 0, mean(TIEMPO_VIAJES, na.rm=T), NA),
+#             tt_walk = ifelse(trips_walk != 0, mean(TIEMPO_VIAJES, na.rm=T), NA))
+
+```
+
+```{r}
+OD_test1 <- OD %>%
   group_by(VORIZT1259,VDESZT1259) %>%
@@
             jobs_bike = mean(jobs_bike),
-            jobs_walk = mean(jobs_walk),
-            tt_transit = ifelse(trips_transit != 0, mean(TIEMPO_VIAJES, na.rm=T), NA), #time is averaged for trips from the same origin to destination + same mode - there likely is variation between trips as these come from the survey (e.g., trips taken at different time of day so they are of different length)
-            tt_car = ifelse(trips_car_motor != 0, mean(TIEMPO_VIAJES, na.rm=T), NA),
-            tt_bike = ifelse(trips_bike != 0, mean(TIEMPO_VIAJES, na.rm=T), NA),
-            tt_walk = ifelse(trips_walk != 0, mean(TIEMPO_VIAJES, na.rm=T), NA)) %>% ungroup()
+            jobs_walk = mean(jobs_walk)) %>% ungroup()
+
+OD_test_transit <- OD %>% filter(trips_transit == 1) %>%
+  group_by(VORIZT1259,VDESZT1259) %>%
+  summarise(tt_transit = mean(TIEMPO_VIAJES, na.rm=T)) %>% ungroup()
+
+OD_test_car <- OD %>% filter(trips_car_motor == 1) %>%
+  group_by(VORIZT1259,VDESZT1259) %>%
+  summarise(tt_car = mean(TIEMPO_VIAJES, na.rm=T)) %>% ungroup()
+
+OD_test_bike <- OD %>% filter(trips_bike == 1) %>%
+  group_by(VORIZT1259,VDESZT1259) %>%
+  summarise(tt_bike = mean(TIEMPO_VIAJES)) %>% ungroup()
+
+OD_test_walk <- OD %>% filter(trips_walk == 1) %>%
+  group_by(VORIZT1259,VDESZT1259) %>%
+  summarise(tt_walk = mean(TIEMPO_VIAJES)) %>% ungroup()
+
+OD_test <- OD_test1 %>% merge(OD_test_transit, by=c("VORIZT1259","VDESZT1259"), all=T)
+OD_test <- OD_test %>% merge(OD_test_car, by=c("VORIZT1259","VDESZT1259"), all=T)
+OD_test <- OD_test %>% merge(OD_test_bike, by=c("VORIZT1259","VDESZT1259"), all=T)
+OD_test <- OD_test %>% merge(OD_test_walk, by=c("VORIZT1259","VDESZT1259"), all=T)
 
@@
 ```
+
+Check a sample OD pair:
+```{r}
+OD_test %>% filter(VORIZT1259 == "004-001" & VDESZT1259 == "004-001")
+OD %>% filter(VORIZT1259 == "004-001" & VDESZT1259 == "004-001" & trips_car_motor == "1" )
+```
+Good the averages make sense.
 
@@
 maybe we don't want this?
-
 ```
@@
          jobs_ = ifelse(is.na(jobs_), 0, jobs_),
-         f_ = f_car + f_transit + f_bike + f_walk)
+         f_ = f_car + f_transit + f_bike + f_walk) #CHECK: is this still proportional
 
@@
 Still not sure what the impact of this is...
+head(OD_SA1 %>% dplyr::select(workers_, workers_car, workers_transit, workers_bike, workers_walk), 25)
 ```
 
-Now calculate SA using this second approach (all workers (not mode specific) competite with eachother) and preform some checks:
+Now calculate SA using this second approach (all workers (not mode specific) compete with each other) and preform some checks:
 ```{r}

### 2022-05-26 Simplified-spatial-availability-example.Rmd
AP started reviewing the code; important change: we need to use the survey expansion factors!

@@
 #here we separate all the "car" trips, "transit" trips, "bike" trips, and "walk" trips. We then will create 4 different imepedance function....
-transit_HW_alltt <- OD_v_HW  %>% filter(MODO_PRIORITARIO <= 9) %>%
-  dplyr::select(TIEMPO_VIAJES)
-
-car_motor_HW_alltt <- OD_v_HW  %>% filter(MODO_PRIORITARIO <= 19 & MODO_PRIORITARIO >= 10) %>%
-  dplyr::select(TIEMPO_VIAJES) #includes both motorcycles and cars
-
-bike_HW_alltt <- OD_v_HW  %>% filter(MODO_PRIORITARIO <= 23 & MODO_PRIORITARIO >= 20) %>%
-  dplyr::select(TIEMPO_VIAJES)
+transit_HW_alltt <- OD_v_HW  %>% 
+  filter(MODO_PRIORITARIO <= 9) %>%
+  #dplyr::select(TIEMPO_VIAJES, ELE_G_POND_ESC2)
+  transmute(TIEMPO_VIAJES, 
+            ELE_G_POND_ESC2,
+            mode = "transit")
+
+car_motor_HW_alltt <- OD_v_HW  %>% 
+  filter(MODO_PRIORITARIO <= 19 & MODO_PRIORITARIO >= 10) %>%
+  #dplyr::select(TIEMPO_VIAJES, ELE_G_POND_ESC2) #includes both motorcycles and cars
+  transmute(TIEMPO_VIAJES, 
+            ELE_G_POND_ESC2,
+            mode = "car/motor")
+
+bike_HW_alltt <- OD_v_HW  %>% 
+  filter(MODO_PRIORITARIO <= 23 & MODO_PRIORITARIO >= 20) %>%
+  #dplyr::select(TIEMPO_VIAJES, ELE_G_POND_ESC2)
+  transmute(TIEMPO_VIAJES, 
+            ELE_G_POND_ESC2,
+            mode = "bike")
 
 walk_HW_alltt <- OD_v_HW  %>% filter(MODO_PRIORITARIO == 24) %>%
-  dplyr::select(TIEMPO_VIAJES)
+  #dplyr::select(TIEMPO_VIAJES, ELE_G_POND_ESC2)
+  transmute(TIEMPO_VIAJES, 
+            ELE_G_POND_ESC2,
+            mode = "walk")
 ```
@@
 # using fitdist function to fit a distribution using the default maximum likelihood estimation method and Nelder-Mead method for direct optimization
-lnorm_ <- fitdistrplus::fitdist(data=transit_HW_alltt%>% unlist() %>% as.numeric(), "lnorm", method="mle", optim.method="Nelder-Mead")
+lnorm_ <- fitdistrplus::fitdist(data=transit_HW_alltt$TIEMPO_VIAJES,#%>% unlist() %>% as.numeric(),
+                                "lnorm", 
+                                method="mle",
+                                #Notice the use of the weights,
+                                weights = round(transit_HW_alltt$ELE_G_POND_ESC2),
+                                optim.method="Nelder-Mead")
 HW_transit_lnorm <- lnorm_
 
-gamma_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt%>% unlist() %>% as.numeric(), "gamma", method="mle", optim.method="Nelder-Mead")
+gamma_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt$TIEMPO_VIAJES, #%>% unlist() %>% as.numeric(),
+                                "gamma", 
+                                method="mle",
+                                #Notice the use of the weights,
+                                weights = round(car_motor_HW_alltt$ELE_G_POND_ESC2), 
+                                optim.method="Nelder-Mead")
 HW_car_motor_gamma <- gamma_
 
-lnorm_ <- fitdistrplus::fitdist(data=bike_HW_alltt%>% unlist() %>% as.numeric(), "lnorm", method="mle", optim.method="Nelder-Mead")
+lnorm_ <- fitdistrplus::fitdist(data=bike_HW_alltt$TIEMPO_VIAJES, #%>% unlist() %>% as.numeric(),
+                                "lnorm", 
+                                method="mle",
+                                #Notice the use of the weights,
+                                weights = round(bike_HW_alltt$ELE_G_POND_ESC2),
+                                optim.method="Nelder-Mead")
 HW_bike_lnorm <- lnorm_
 
-lnorm_ <- fitdistrplus::fitdist(data=walk_HW_alltt%>% unlist() %>% as.numeric(), "lnorm", method="mle", optim.method="Nelder-Mead")
+lnorm_ <- fitdistrplus::fitdist(data=walk_HW_alltt$TIEMPO_VIAJES, #%>% unlist() %>% as.numeric(),
+                                "lnorm", 
+                                method="mle",
+                                #Notice the use of the weights,
+                                weights = round(walk_HW_alltt$ELE_G_POND_ESC2),
+                                optim.method="Nelder-Mead")
 HW_walk_lnorm <- lnorm_
@@
 
+To visualize the impedance functions we create a data frame and calculate the density of the distribution given the estimated parameters:
+```{r}
+x <- data.frame(t = seq(1, 300, 1))
+
+x_transit <- x %>%
+  mutate(f = dlnorm(t,
+                    HW_transit_lnorm$estimate[1],
+                    HW_transit_lnorm$estimate[2]),
+         mode = "transit")
+
+x_car_motor <- x %>%
+  mutate(f = dgamma(t,
+                    shape = HW_car_motor_gamma$estimate[1],
+                    rate = HW_car_motor_gamma$estimate[2]),
+         mode = "car/motor")
+
+x_bike <- x %>%
+  mutate(f = dlnorm(t,
+                    HW_bike_lnorm$estimate[1],
+                    HW_bike_lnorm$estimate[2]),
+         mode = "bike")
+
+x_walk <- x %>%
+  mutate(f = dlnorm(t,
+                    HW_walk_lnorm$estimate[1],
+                    HW_walk_lnorm$estimate[2]),
+         mode = "walk")
+```
+
+Plot:
+```{r}
+rbind(x_transit,
+      x_car_motor,
+      x_bike,
+      x_walk) %>%
+  mutate(mode = factor(mode,
+                       levels = c("bike", "walk", "car/motor", "transit"))) %>%
+  ggplot(aes(x = t, y = f)) + 
+  geom_line() + 
+  facet_wrap(~ mode, nrow = 2)
+```
+
+What do the impedance curves tell us about the travel behavior by different modes? Check the histograms:
+```{r}
+rbind(transit_HW_alltt,
+      car_motor_HW_alltt,
+      bike_HW_alltt,
+      walk_HW_alltt) %>%
+  mutate(mode = factor(mode,
+                       levels = c("bike", "walk", "car/motor", "transit"))) %>%
+  ggplot(aes(x = TIEMPO_VIAJES, weights = ELE_G_POND_ESC2, after_stat(density))) + 
+  geom_histogram(bins = 35) + 
+  facet_wrap(~ mode, nrow = 2)
+```
+
 Then we add the number of trips (by mode) and create objects which contain the number of workers (origin) and jobs (destinations) - by mode:
 ```{r}
-OD_v_HW <- OD_v_HW  %>% mutate(trips_transit = ifelse(MODO_PRIORITARIO <= 9, 1,0),
-                    trips_car_motor = ifelse(MODO_PRIORITARIO <= 19 & MODO_PRIORITARIO >= 10, 1, 0),
-                    trips_bike = ifelse(MODO_PRIORITARIO <= 23 & MODO_PRIORITARIO >= 20, 1, 0),
-                    trips_walk = ifelse(MODO_PRIORITARIO == 24, 1, 0))
-
+# OD_v_HW <- OD_v_HW  %>% 
+#   mutate(trips_transit = ifelse(MODO_PRIORITARIO <= 9, 1,0),
+#          trips_car_motor = ifelse(MODO_PRIORITARIO <= 19 & MODO_PRIORITARIO >= 10, 1, 0),
+#          trips_bike = ifelse(MODO_PRIORITARIO <= 23 & MODO_PRIORITARIO >= 20, 1, 0),
+#          trips_walk = ifelse(MODO_PRIORITARIO == 24, 1, 0))
+
+# Instead of creating various columns like above, create a single column with the mode
+OD_v_HW <- OD_v_HW  %>% 
+  mutate(mode = case_when(MODO_PRIORITARIO <= 9 ~ "transit",
+                          MODO_PRIORITARIO <= 19 & MODO_PRIORITARIO >= 10 ~ "car_motor",
+                          MODO_PRIORITARIO <= 23 & MODO_PRIORITARIO >= 20 ~ "bike",
+                          MODO_PRIORITARIO == 24 ~ "walk"),
+         mode = factor(mode))
+
+# workers <- OD_v_HW %>%
+#   group_by(VORIZT1259) %>%
+#   summarize(workers_transit = sum(trips_transit),
+#             workers_car = sum(trips_car_motor),
+#             workers_bike = sum(trips_bike),
+#             workers_walk = sum(trips_walk),
+#             .groups = "drop") %>%
+#   transmute(ZT1259 = VORIZT1259,
+#             workers_transit,
+#             workers_car,
+#             workers_bike,
+#             workers_walk)
+
+# The workers are classified by mode
 workers <- OD_v_HW %>%
-  group_by(VORIZT1259) %>%
-  summarize(workers_transit = sum(trips_transit),
-            workers_car = sum(trips_car_motor),
-            workers_bike = sum(trips_bike),
-            workers_walk = sum(trips_walk),
+  group_by(VORIZT1259,
+           mode) %>%
+  # The sum is of the weighted workers
+  summarize(workers = sum(ELE_G_POND_ESC2),
             .groups = "drop") %>%
-  transmute(ZT1259 = VORIZT1259,
-            workers_transit,
-            workers_car,
-            workers_bike,
-            workers_walk)
-
+  rename(ZT1259 = VORIZT1259)
+
+# jobs <- OD_v_HW %>%
+#   group_by(VDESZT1259) %>%
+#   summarize(jobs_transit = sum(trips_transit),
+#             jobs_car = sum(trips_car_motor),
+#             jobs_bike = sum(trips_bike),
+#             jobs_walk = sum(trips_walk),
+#             .groups = "drop") %>%
+#   transmute(ZT1259 = VDESZT1259,
+#             jobs_transit,
+#             jobs_car,
+#             jobs_bike,
+#             jobs_walk)
+
+# Do we need to classify the jobs by mode?
 jobs <- OD_v_HW %>%
   group_by(VDESZT1259) %>%
-  summarize(jobs_transit = sum(trips_transit),
-            jobs_car = sum(trips_car_motor),
-            jobs_bike = sum(trips_bike),
-            jobs_walk = sum(trips_walk),
+  summarize(jobs = sum(ELE_G_POND_ESC2),
             .groups = "drop") %>%
-  transmute(ZT1259 = VDESZT1259,
-            jobs_transit,
-            jobs_car,
-            jobs_bike,
-            jobs_walk)
-```
-
+  rename(ZT1259 = VDESZT1259)
+```
+
+NOW WE NEED: the travel times from the routing algorithm between origin-destination pairs.
+Anastasia did not use the routing, it may not be needed; instead she used the average travel time by mode for all trips
+
+<!--
 Next join workers and jobs to the ZN simple features object:
@@
 
-zn <- zn %>%
+# zn <- zn %>%
+#   # Join workers
+#   left_join(workers,
+#             by = "ZT1259") %>%
+#   # Join jobs
+#   left_join(jobs,
+#             by = "ZT1259") %>%
+#   # Replace NAs with zeros
+#   mutate(workers_transit = tidyr::replace_na(workers_transit, 0),
+#          workers_car = tidyr::replace_na(workers_car, 0),
+#          workers_bike = tidyr::replace_na(workers_bike, 0),
+#          workers_walk = tidyr::replace_na(workers_walk, 0),
+#          jobs_transit = tidyr::replace_na(jobs_transit, 0),
+#          jobs_car = tidyr::replace_na(jobs_car, 0),
+#          jobs_bike = tidyr::replace_na(jobs_bike, 0),
+#          jobs_walk = tidyr::replace_na(jobs_walk, 0))
+
+workers_sf <- workers %>%
   # Join workers
-  left_join(workers,
+  left_join(zn,
             by = "ZT1259") %>%
+  st_as_sf()
+
+jobs_sf <- jobs %>%
   # Join jobs
-  left_join(jobs,
+  left_join(zn,
             by = "ZT1259") %>%
-  # Replace NAs with zeros
-  mutate(workers_transit = tidyr::replace_na(workers_transit, 0),
-         workers_car = tidyr::replace_na(workers_car, 0),
-         workers_bike = tidyr::replace_na(workers_bike, 0),
-         workers_walk = tidyr::replace_na(workers_walk, 0),
-         jobs_transit = tidyr::replace_na(jobs_transit, 0),
-         jobs_car = tidyr::replace_na(jobs_car, 0),
-         jobs_bike = tidyr::replace_na(jobs_bike, 0),
-         jobs_walk = tidyr::replace_na(jobs_walk, 0))
+  st_as_sf()
 

### 2022-05-27 Simplified-spatial-availability-example.Rmd
Final audit of the procedures by AS and AP

@@
 
-Plot:
-```{r}
-rbind(x_transit,
+Bind the empirical and theoretical trip length distributions:
+```{r}
+tld_empirical <- rbind(transit_HW_alltt,
+      car_motor_HW_alltt,
+      bike_HW_alltt,
+      walk_HW_alltt) %>%
+  mutate(mode = factor(mode,
+                       levels = c("bike", "walk", "car/motor", "transit")),
+         distribution = "empirical")
+
+tld_theoretical <- rbind(x_transit,
       x_car_motor,
@@
   mutate(mode = factor(mode,
-                       levels = c("bike", "walk", "car/motor", "transit"))) %>%
-  ggplot(aes(x = t, y = f)) + 
-  geom_line() + 
+                       levels = c("bike", "walk", "car/motor", "transit")))
+```
+
+Plot:
+```{r}
+ggplot() + 
+  geom_histogram(data = tld_empirical,
+                 aes(x = TIEMPO_VIAJES, 
+                     #weights = ELE_G_POND_ESC2, 
+                     after_stat(density)),
+                 bins = 45) +
+  geom_line(data = tld_theoretical,
+            aes(x = t, y = f)) + 
   facet_wrap(~ mode, nrow = 2)
@@
 
-What do the impedance curves tell us about the travel behavior by different modes? Check the histograms:
+Double check the empirical trip length distributions:
 ```{r}
@@
   ggplot(aes(x = TIEMPO_VIAJES, weights = ELE_G_POND_ESC2, after_stat(density))) + 
-  geom_histogram(bins = 35) + 
+  geom_histogram(bins = 45) + 
   facet_wrap(~ mode, nrow = 2)
@@
   mutate(mode = case_when(MODO_PRIORITARIO <= 9 ~ "transit",
-                          MODO_PRIORITARIO <= 19 & MODO_PRIORITARIO >= 10 ~ "car_motor",
+                          MODO_PRIORITARIO <= 19 & MODO_PRIORITARIO >= 10 ~ "car/motor",
                           MODO_PRIORITARIO <= 23 & MODO_PRIORITARIO >= 20 ~ "bike",
                           MODO_PRIORITARIO == 24 ~ "walk"),
-         mode = factor(mode))
+         mode = factor(mode,
+                       levels = c("bike", "walk", "car/motor", "transit")))
 
@@
 
-NOW WE NEED: the travel times from the routing algorithm between origin-destination pairs.
-Anastasia did not use the routing, it may not be needed; instead she used the average travel time by mode for all trips
-
-<!--
 Next join workers and jobs to the ZN simple features object:
@@
 #add the number of jobs and workers to the od_ft_tt matrix
-OD <- OD_v_HW %>% merge(zn %>% dplyr::select(ZT1259, workers_transit,workers_car,workers_bike,workers_walk) %>% st_drop_geometry(),
-                        by.x = "VORIZT1259", by.y="ZT1259", all.x = TRUE)
-
-OD <- OD %>% merge(zn %>% dplyr::select(ZT1259, jobs_transit,jobs_car,jobs_bike,jobs_walk) %>% st_drop_geometry(),
-                   by.x = "VDESZT1259", by.y="ZT1259", all.x = TRUE)
-
-OD <- OD %>% 
-  mutate(workers_transit = ifelse(is.na(workers_transit), 0, workers_transit),
-         workers_car = ifelse(is.na(workers_car), 0, workers_car),
-         workers_bike = ifelse(is.na(workers_bike), 0, workers_bike),
-         workers_walk = ifelse(is.na(workers_walk), 0, workers_walk),
-         jobs_car = ifelse(is.na(jobs_car), 0, jobs_car),
-         jobs_transit = ifelse(is.na(jobs_transit), 0, jobs_transit),
-         jobs_bike = ifelse(is.na(jobs_bike), 0, jobs_bike),
-         jobs_walk = ifelse(is.na(jobs_walk), 0, jobs_walk))
+# OD <- OD_v_HW %>% merge(zn %>% dplyr::select(ZT1259, workers_transit,workers_car,workers_bike,workers_walk) %>% st_drop_geometry(),
+#                         by.x = "VORIZT1259", by.y="ZT1259", all.x = TRUE)
+# 
+# OD <- OD %>% merge(zn %>% dplyr::select(ZT1259, jobs_transit,jobs_car,jobs_bike,jobs_walk) %>% st_drop_geometry(),
+#                    by.x = "VDESZT1259", by.y="ZT1259", all.x = TRUE)
+# 
+# OD <- OD %>% 
+#   mutate(workers_transit = ifelse(is.na(workers_transit), 0, workers_transit),
+#          workers_car = ifelse(is.na(workers_car), 0, workers_car),
+#          workers_bike = ifelse(is.na(workers_bike), 0, workers_bike),
+#          workers_walk = ifelse(is.na(workers_walk), 0, workers_walk),
+#          jobs_car = ifelse(is.na(jobs_car), 0, jobs_car),
+#          jobs_transit = ifelse(is.na(jobs_transit), 0, jobs_transit),
+#          jobs_bike = ifelse(is.na(jobs_bike), 0, jobs_bike),
+#          jobs_walk = ifelse(is.na(jobs_walk), 0, jobs_walk))
 ```
@@
 ```{r}
-OD_test1 <- OD %>%
-  group_by(VORIZT1259,VDESZT1259) %>%
-  summarise(trips_transit = sum(trips_transit),
-            trips_car_motor = sum(trips_car_motor),
-            trips_bike = sum(trips_bike),
-            trips_walk = sum(trips_walk),
-            workers_transit = mean(workers_transit), #workers_ and jobs_ values are retained from the ZN object - so they are the same value for each OD pair thus we use mean() here.
-            workers_car = mean(workers_car),
-            workers_bike = mean(workers_bike),
-            workers_walk = mean(workers_walk),
-            jobs_transit = mean(jobs_transit),
-            jobs_car = mean(jobs_car),
-            jobs_bike = mean(jobs_bike),
-            jobs_walk = mean(jobs_walk)) %>% ungroup()
-
-OD_test_transit <- OD %>% filter(trips_transit == 1) %>%
-  group_by(VORIZT1259,VDESZT1259) %>%
-  summarise(tt_transit = mean(TIEMPO_VIAJES, na.rm=T)) %>% ungroup()
-
-OD_test_car <- OD %>% filter(trips_car_motor == 1) %>%
-  group_by(VORIZT1259,VDESZT1259) %>%
-  summarise(tt_car = mean(TIEMPO_VIAJES, na.rm=T)) %>% ungroup()
-
-OD_test_bike <- OD %>% filter(trips_bike == 1) %>%
-  group_by(VORIZT1259,VDESZT1259) %>%
-  summarise(tt_bike = mean(TIEMPO_VIAJES)) %>% ungroup()
-
-OD_test_walk <- OD %>% filter(trips_walk == 1) %>%
-  group_by(VORIZT1259,VDESZT1259) %>%
-  summarise(tt_walk = mean(TIEMPO_VIAJES)) %>% ungroup()
-
-OD_test <- OD_test1 %>% merge(OD_test_transit, by=c("VORIZT1259","VDESZT1259"), all=T)
-OD_test <- OD_test %>% merge(OD_test_car, by=c("VORIZT1259","VDESZT1259"), all=T)
-OD_test <- OD_test %>% merge(OD_test_bike, by=c("VORIZT1259","VDESZT1259"), all=T)
-OD_test <- OD_test %>% merge(OD_test_walk, by=c("VORIZT1259","VDESZT1259"), all=T)
-
-#check, should add up to the origin number of trips 27795 - perfect!
-sum(OD$trips_transit) + sum(OD$trips_car_motor) + sum(OD$trips_bike) + sum(OD$trips_walk)
-sum(OD_test$trips_transit) + sum(OD_test$trips_car_motor) + sum(OD_test$trips_bike) + sum(OD_test$trips_walk)
-```
+# OD_test1 <- OD %>%
+#   group_by(VORIZT1259,VDESZT1259) %>%
+#   summarise(trips_transit = sum(trips_transit),
+#             trips_car_motor = sum(trips_car_motor),
+#             trips_bike = sum(trips_bike),
+#             trips_walk = sum(trips_walk),
+#             workers_transit = mean(workers_transit), #workers_ and jobs_ values are retained from the ZN object - so they are the same value for each OD pair thus we use mean() here.
+#             workers_car = mean(workers_car),
+#             workers_bike = mean(workers_bike),
+#             workers_walk = mean(workers_walk),
+#             jobs_transit = mean(jobs_transit),
+#             jobs_car = mean(jobs_car),
+#             jobs_bike = mean(jobs_bike),
+#             jobs_walk = mean(jobs_walk)) %>% ungroup()
+
+OD <- OD_v_HW %>%
+  group_by(VORIZT1259, 
+           VDESZT1259,
+           mode) %>%
+  summarize(tt = mean(TIEMPO_VIAJES),
+            .groups = "drop")
+
+# Check the mean and median travel times by mode
+OD %>% 
+  group_by(mode) %>% 
+  summarize(mtt = mean(tt))
+OD %>% 
+  group_by(mode) %>% 
+  summarize(mtt = median(tt))
+
+# OD_test_transit <- OD %>% filter(trips_transit == 1) %>%
+#   group_by(VORIZT1259,VDESZT1259) %>%
+#   summarise(tt_transit = mean(TIEMPO_VIAJES, na.rm=T)) %>% ungroup()
+# 
+# OD_test_car <- OD %>% filter(trips_car_motor == 1) %>%
+#   group_by(VORIZT1259,VDESZT1259) %>%
+#   summarise(tt_car = mean(TIEMPO_VIAJES, na.rm=T)) %>% ungroup()
+# 
+# OD_test_bike <- OD %>% filter(trips_bike == 1) %>%
+#   group_by(VORIZT1259,VDESZT1259) %>%
+#   summarise(tt_bike = mean(TIEMPO_VIAJES)) %>% ungroup()
+# 
+# OD_test_walk <- OD %>% filter(trips_walk == 1) %>%
+#   group_by(VORIZT1259,VDESZT1259) %>%
+#   summarise(tt_walk = mean(TIEMPO_VIAJES)) %>% ungroup()
+# 
+# OD_test <- OD_test1 %>% merge(OD_test_transit, by=c("VORIZT1259","VDESZT1259"), all=T)
+# OD_test <- OD_test %>% merge(OD_test_car, by=c("VORIZT1259","VDESZT1259"), all=T)
+# OD_test <- OD_test %>% merge(OD_test_bike, by=c("VORIZT1259","VDESZT1259"), all=T)
+# OD_test <- OD_test %>% merge(OD_test_walk, by=c("VORIZT1259","VDESZT1259"), all=T)
+# 
+# #check, should add up to the origin number of trips 27795 - perfect!
+# sum(OD$trips_transit) + sum(OD$trips_car_motor) + sum(OD$trips_bike) + sum(OD$trips_walk)
+# sum(OD_test$trips_transit) + sum(OD_test$trips_car_motor) + sum(OD_test$trips_bike) + sum(OD_test$trips_walk)
+```
+
+<!--
 
@@
 Good the averages make sense.
-
-now estimate f value using these averaged travel times:
+-->
+
+Join workers and jobs to OD table:
+```{r}
+OD_2 <- OD %>%
+  left_join(workers,
+            by = c("VORIZT1259" = "ZT1259", "mode"))
+
+OD_2 <- OD_2 %>%
+  left_join(jobs,
+            by = c("VDESZT1259" = "ZT1259"))
+```
+
+Now estimate f value using these averaged travel times:
 ```{r, echo=FALSE, warning=FALSE, message=FALSE}
 #dgamma is the density plot for gamma function and dlnorm is the desntiy plot for the lognormal function; adding values as "f" our impedance function
-OD_test <- OD_test %>%
-  mutate(f_car = dgamma(tt_car, HW_car_motor_gamma$estimate["shape"], HW_car_motor_gamma$estimate["rate"]),
-         f_car = ifelse(is.na(f_car), 0, f_car),
-         f_transit = dlnorm(tt_transit, HW_transit_lnorm$estimate["meanlog"], HW_transit_lnorm$estimate["sdlog"]),
-         f_transit = ifelse(is.na(f_transit), 0, f_transit),
-         f_bike = dlnorm(tt_bike, HW_bike_lnorm$estimate["meanlog"], HW_bike_lnorm$estimate["sdlog"]),
-         f_bike = ifelse(is.na(f_bike), 0, f_bike),
-         f_walk = dlnorm(tt_walk, HW_walk_lnorm$estimate["meanlog"], HW_walk_lnorm$estimate["sdlog"]),
-         f_walk = ifelse(is.na(f_walk), 0, f_walk))
-```
+# OD_test <- OD_test %>%
+#   mutate(f_car = dgamma(tt_car, HW_car_motor_gamma$estimate["shape"], HW_car_motor_gamma$estimate["rate"]),
+#          f_car = ifelse(is.na(f_car), 0, f_car),
+#          f_transit = dlnorm(tt_transit, HW_transit_lnorm$estimate["meanlog"], HW_transit_lnorm$estimate["sdlog"]),
+#          f_transit = ifelse(is.na(f_transit), 0, f_transit),
+#          f_bike = dlnorm(tt_bike, HW_bike_lnorm$estimate["meanlog"], HW_bike_lnorm$estimate["sdlog"]),
+#          f_bike = ifelse(is.na(f_bike), 0, f_bike),
+#          f_walk = dlnorm(tt_walk, HW_walk_lnorm$estimate["meanlog"], HW_walk_lnorm$estimate["sdlog"]),
+#          f_walk = ifelse(is.na(f_walk), 0, f_walk))
+
+OD_2 <- OD_2 %>%
+  mutate(f = case_when(mode == "transit" ~ dlnorm(tt,
+                                                  HW_transit_lnorm$estimate["meanlog"], 
+                                                  HW_transit_lnorm$estimate["sdlog"]),
+                       mode == "car/motor" ~ dgamma(tt,
+                                                    HW_car_motor_gamma$estimate["shape"],
+                                                    HW_car_motor_gamma$estimate["rate"]),
+                       mode == "bike" ~ dlnorm(tt,
+                                               HW_bike_lnorm$estimate["meanlog"],
+                                               HW_bike_lnorm$estimate["sdlog"]),
+                       mode == "walk" ~ dlnorm(tt, 
+                                               HW_walk_lnorm$estimate["meanlog"], 
+                                               HW_walk_lnorm$estimate["sdlog"])))
+```
+
+## Spatial availability
+
+Calculate spatial availabilty:
+```{r}
+OD_SA <- OD_2 %>%
+  mutate(catch = 1) %>%
+  mutate(SA_ij = sp_avail(., 
+                          o_id = VORIZT1259,
+                          d_id = VDESZT1259,
+                          pop = workers,
+                          opp = jobs,
+                          r = catch,
+                          f = f))
+```
+
+Check that the totals are maintained:
+```{r}
+sum(OD_SA$SA_ij)
+sum(jobs$jobs)
+```
+
+Spatial availability by mode:
+```{r}
+OD_SA %>%
+  group_by(mode) %>%
+  summarize(V_m = sum(SA_ij))
+```
+
+Spatial availability by origin (with check):
+```{r}
+SA_i <- OD_SA %>%
+  group_by(VORIZT1259) %>%
+  summarize(V_i = sum(SA_ij))
+
+summary(SA_i)
+sum(SA_i$V_i)
+```
+
+Spatial availability by origin/mode (with check):
+```{r}
+SA_im <- OD_SA %>%
+  group_by(VORIZT1259, 
+           mode) %>%
+  summarize(V_im = sum(SA_ij),
+            .groups = "drop")
+
+summary(SA_im)
+sum(SA_im$V_im)
+```
+
+Calculate worker population by zone of origin (with check):
+```{r}
+population <- workers %>%
+  group_by(ZT1259) %>%
+  summarize(population = sum(workers))
+
+sum(population$population)
+```
+
+Join zonal worker population (with checks):
+```{r}
+SA_i <- SA_i %>%
+  left_join(population,
+            by = c("VORIZT1259" = "ZT1259"))
+sum(SA_i$V_i)
+
+SA_im <- SA_im %>%
+  left_join(population,
+            by = c("VORIZT1259" = "ZT1259"))
+sum(SA_im$V_im)
+```
+
+Calculate availability per capita:
+```{r}
+SA_i <- SA_i %>%
+  mutate(v_i = V_i/population)
+
+summary(SA_i)
+
+SA_im <- SA_im %>%
+  mutate(v_im = V_im/population)
+
+summary(SA_im)
+```
+
+Join geometry of zones to SA:
+```{r}
+SA_i <- SA_i %>%
+  left_join(zn %>%
+              dplyr::select(ZT1259),
+            by = c("VORIZT1259" = "ZT1259")) %>%
+  st_as_sf()
+
+SA_im <- SA_im %>%
+  left_join(zn %>%
+              dplyr::select(ZT1259),
+            by = c("VORIZT1259" = "ZT1259")) %>%
+  st_as_sf()
+```
+
+Preliminaries to plotting:
+```{r create-bounding-boxes}
+#create the bounding box for the Toronto region + a little extra
+bbox_new1 <- st_bbox(zn%>%filter(Municipality == "Madrid")) # current bounding box
+
+xrange <- bbox_new1$xmax - bbox_new1$xmin # range of x values
+yrange <- bbox_new1$ymax - bbox_new1$ymin # range of y values
+
+bbox_new1[1] <- bbox_new1[1] - (0.0 * xrange) # xmin - left
+bbox_new1[3] <- bbox_new1[3] + (0.0 * xrange) # xmax - right
+bbox_new1[2] <- bbox_new1[2] - (0.0 * yrange) # ymin - bottom
+bbox_new1[4] <- bbox_new1[4] + (0.0 * yrange) # ymax - top
+
+bbox_Madrid_muni <- bbox_new1 %>%  # take the bounding box ...
+  st_as_sfc()
+
+Madrid_muni_border <- zn %>%
+  filter(Municipality == "Madrid") %>%
+  st_union()
+```
+
+Plot V (spatial availability) by origin:
+```{r}
+ggplot() +
+  geom_sf(data = SA_i,
+          aes(fill = V_i),
+          color = NA) +
+  scale_fill_distiller(palette = "YlGn", #legend scale bar
+                       name = expression(v["i"]),
+                       na.value = "grey90",
+                       direction = 1,
+                       trans = "log",
+                       breaks=c(0, 10, 100, 1000, 10000))+
+  geom_sf(data = Madrid_muni_border, # border for Madrid muni
+          colour=alpha("black",1), 
+          size = 0.5, 
+          fill=NA, 
+          show.legend = "polygon") +
+  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
+                         height = unit(0.8, "cm"), 
+                         width = unit(0.8, "cm"),
+                         style = north_arrow_orienteering(line_width = 0.25,
+                                                          line_col = "dimgrey", 
+                                                          fill = c("white","dimgrey"))) +
+  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
+                   height = unit(0.15, "cm")) +
+  theme_void() +
+  theme(legend.position = c(1.10, 0.25), 
+        plot.margin = margin(t = 0, r = 0, b = 0, l = -80)) # positing legend and clipping out white space
+```
+
+Plot v (spatial availability per capita) by origin:
+```{r}
+ggplot() +
+  geom_sf(data = SA_i,
+          aes(fill = v_i),
+          color = NA) +
+  scale_fill_gradient2(name = expression(v["i"]),
+                       midpoint = 1,
+                       na.value = "grey90")+
+  geom_sf(data = Madrid_muni_border, # border for Madrid muni
+          colour=alpha("black",1), 
+          size = 0.5, 
+          fill=NA, 
+          show.legend = "polygon") +
+  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
+                         height = unit(0.8, "cm"), 
+                         width = unit(0.8, "cm"),
+                         style = north_arrow_orienteering(line_width = 0.25,
+                                                          line_col = "dimgrey", 
+                                                          fill = c("white","dimgrey"))) +
+  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
+                   height = unit(0.15, "cm")) +
+  theme_void() +
+  theme(legend.position = c(1.10, 0.25), 
+        plot.margin = margin(t = 0, r = 0, b = 0, l = -80)) # positing legend and clipping out white space
+```
+
+Plot V (spatial availability) by origin-mode:
+```{r}
+ggplot() +
+  geom_sf(data = SA_im,
+          aes(fill = V_im),
+          color = NA) +
+  scale_fill_distiller(palette = "YlGn", #legend scale bar
+                       name = expression(v["i"]),
+                       na.value = "grey90",
+                       direction = 1,
+                       trans = "log",
+                       breaks=c(0, 10, 100, 1000, 10000))+
+  geom_sf(data = Madrid_muni_border, # border for Madrid muni
+          colour=alpha("black",1), 
+          size = 0.5, 
+          fill=NA, 
+          show.legend = "polygon") +
+  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
+                         height = unit(0.8, "cm"), 
+                         width = unit(0.8, "cm"),
+                         style = north_arrow_orienteering(line_width = 0.25,
+                                                          line_col = "dimgrey", 
+                                                          fill = c("white","dimgrey"))) +
+  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
+                   height = unit(0.15, "cm")) +
+  theme_void() +
+  theme(legend.position = c(1.10, 0.25), 
+        plot.margin = margin(t = 0, r = 0, b = 0, l = -80)) + # positing legend and clipping out white space
+  facet_wrap(~mode, nrow = 2)
+```
+
+Plot v (spatial availability per capita) by origin:
+```{r}
+ggplot() +
+  geom_sf(data = SA_im,
+          aes(fill = v_im),
+          color = NA) +
+  scale_fill_gradient2(name = expression(v["i"]),
+                       midpoint = 1,
+                       na.value = "grey90")+
+  geom_sf(data = Madrid_muni_border, # border for Madrid muni
+          colour=alpha("black",1), 
+          size = 0.5, 
+          fill=NA, 
+          show.legend = "polygon") +
+  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
+                         height = unit(0.8, "cm"), 
+                         width = unit(0.8, "cm"),
+                         style = north_arrow_orienteering(line_width = 0.25,
+                                                          line_col = "dimgrey", 
+                                                          fill = c("white","dimgrey"))) +
+  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
+                   height = unit(0.15, "cm")) +
+  theme_void() +
+  theme(legend.position = c(1.10, 0.25), 
+        plot.margin = margin(t = 0, r = 0, b = 0, l = -80)) + # positing legend and clipping out white space
+  facet_wrap(~mode, nrow = 2)
+```
+<!--
 

### 2022-05-31 Simplified-spatial-availability-example.Rmd
Finished converting to ZT208 resolutions.
Am working on how to shift car trips to transit (in a way that's flexible and will allow us to easily toggle the modal shift scenarios (i.e., instead of 100% transit, maybe 80%/10/10 transit/bike/walk))

@@
 ```{r data-import-filter}
-OD_v <- read_excel("data-raw/EDM2018VIAJES.xlsx")
+load(file = "data/OD_v.RData")
+```
+
+Keep only the direct home to work trips:
+```{r}
+# not chained home to work trips - direct
+OD_v_HW <- OD_v %>% filter(VORI == "1" & VDES == "2") #only filter in origin casa trips and destination trabajo
+```
+
+Import zones:
+```{r}
+#importing zones
+load("data/zn1259.RData")
+load("data/zn208.RData")
+load("data/zn_u.RData") # a key to join the zt1259 to the zn208 + contains municipality names
 ```
@@
 
-Calculate the duration of each trip
-```{r}
-OD_v <- OD_v %>%
-  mutate(TIEMPO_VIAJES =
-           (as.numeric(OD_v$VDESHORAFIN) - 
-              as.numeric(OD_v$VORIHORAINI))) %>%
-  mutate(TIEMPO_VIAJES = ifelse(TIEMPO_VIAJES == 0, 0.1, TIEMPO_VIAJES)) # set all 0min travel times to 0.1 min
-summary(OD_v$TIEMPO_VIAJES)
-```
-Checking the trip data - it appears each OD is a row, i.e., chained trips are multiple rows.
+This needs to be re-thought for the ZT208 resolution.
+Since we don't have an issue with missing data from ZT208 zones, we will skip trying to include chained trips (including also complicates the impedance zone):
+<!-- Checking the trip data - it appears each OD is a row, i.e., chained trips are multiple rows.
 We want to extract every home origin (VORI = 1) to every work destination (VDEST =2) trip as a single row...
@@
 
-First, the direct home to work trips:
-```{r}
-# not chained home to work trips - direct
-OD_v_HW <- OD_v %>% filter(VORI == "1" & VDES == "2") #only filter in origin casa trips and destination trabajo
-```
-
 Next, let's identify the chained trips and then consolidate them. We only care about trip O/D, travel mode, travel time, and 
 ```{r}
-t <- OD_v %>% group_by(ID_HOGAR,ID_IND) %>% 
+temp <- OD_v %>% group_by(ID_HOGAR,ID_IND) %>% 
   filter((VORI == "1" & VDES != "2" & VDES != "1") | (VDES == "2" & VORI != "1"& VORI != "2")) %>% #here the origin NEEDS to be from home, the destination can be anywhere OR the origin can be from anywhere but the destination NEEDS to be work.
@@
 
-chained2 <- (ifelse(t$chained[-1]==0 |t$chained==0, 0, NA)) %>% as.data.frame()
-
-t <- cbind(t,chained2) %>% dplyr::rename("chained2" = ".")
-
-t<-t %>% filter(chained2==0) # these are all the CHAINED trips - only considering 1 location before work.  
-t
+chained2 <- (ifelse(temp$chained[-1]==0 | temp$chained==0, 0, NA)) %>% as.data.frame() #identifying chained trips using '0's
+
+temp <- cbind(temp,chained2) %>% dplyr::rename("chained2" = ".")
+
+temp <- temp %>% filter(chained2==0) # filtering only CHAINED trips - reminder these consider only 1 location before work. 1,629 trips 
 ```
@@
 ```{r}
-t <- t %>% group_by(ID_HOGAR, ID_IND) %>% mutate(trip = ifelse(((ID_VIAJE - lag(ID_VIAJE)) == 1)
+t <- temp %>% group_by(ID_HOGAR, ID_IND) %>% mutate(trip = ifelse(((ID_VIAJE - lag(ID_VIAJE)) == 1)
                                                                & ((MODO_PRIORITARIO - lag(MODO_PRIORITARIO))== 0), 
-                                                               MODO_PRIORITARIO, 0))
-
-t %>% filter(trip == 0) %>% nrow() / t %>% filter(chained == 0) %>% nrow()
-t %>% filter(trip != 0 & !is.na(trip)) %>% nrow() / t %>% filter(chained == 0) %>% nrow()
-```
-so ~27% of chained trips are not made using the same mode.
-THe rest of them are (`r t %>% filter(trip != 0 & !is.na(trip)) %>% nrow() ` trips are chained (1 intermediate destination) and use the same mode).
+                                                               MODO_PRIORITARIO, 0)) #if trip is chained the mode (number) is printed, if the trip is not chained a '0' is printed
+
+t %>% filter(trip == 0) %>% nrow() / t  %>% filter(chained == 0) %>% nrow()    #of chained trips / number of non-chained trips
+t  %>% filter(trip != 0 & !is.na(trip))%>% nrow() / t %>% filter(chained == 0) %>% nrow() #of chained trips / number of non-chained trips
+```
+so ~4% of trips are chained, and ~27% of chained trips are not made using the same mode.
+THe rest of them are (`r temp %>% filter(trip != 0 & !is.na(trip)) %>% nrow() ` trips are chained (1 intermediate destination) and use the same mode).
 
@@
 t <- t %>% filter(is.na(chained)) # only including origins
-t <- t[,1:20] 
+t <- t[,1:24] 
 OD_v_HW_chained <- rbind(OD_v_HW, t)
@@
 ```{r}
-#importing zones
-zn <- st_read("data-raw/zn.shp")
-ALLLEZ <- st_read("data-raw/Zona_Bajas_Emisiones_MADRID.shp")
-
-temp <- read_excel("data-raw/ZonificacionZT1259 complete.xlsx") %>% dplyr::select(ZT1259, Municipality)
-zn_t <- zn %>% merge(temp, by = c("ZT1259"), all.x=TRUE)
-```
-
-```{r}
-Otrips_zona_c <- OD_v_HW_chained %>% group_by(VORIZT1259)  %>% summarise(Otrips = (n())) # we group by the origin ZT1259 ID and then we summarize the number of rows that have the same ZT using the summarize() function and the n() function to create a new 'Otrips' column.
-
-zn_c <- zn %>% merge(Otrips_zona_c, by.x = "ZT1259", by.y="VORIZT1259", all.x = TRUE)
+Otrips_zona_c <- OD_v_HW_chained %>% group_by(VORI_CD_ZT208)  %>% summarise(Otrips = (n())) # we group by the origin ZT208 ID and then we summarize the number of rows that have the same ZT using the summarize() function and the n() function to create a new 'Otrips' column.
+
+zn_c <- zn208 %>% merge(Otrips_zona_c, by.x = "CD_ZT208", by.y="VORI_CD_ZT208", all.x = TRUE)
 
@@
 
-Otrips_zona <- OD_v_HW %>% group_by(VORIZT1259)  %>% summarise(Otrips = (n())) # we group by the origin ZT1259 ID and then we summarize the number of rows that have the same ZT using the summarize() function and the n() function to create a new 'Otrips' column.
-
-zn_noc <- zn %>% merge(Otrips_zona, by.x = "ZT1259", by.y="VORIZT1259", all.x = TRUE)
+Otrips_zona <- OD_v_HW %>% group_by(VORI_CD_ZT208)  %>% summarise(Otrips = (n())) # we group by the origin ZT208 ID and then we summarize the number of rows that have the same ZT using the summarize() function and the n() function to create a new 'Otrips' column.
+
+zn_noc <- zn208 %>% merge(Otrips_zona, by.x = "CD_ZT208", by.y="VORI_CD_ZT208", all.x = TRUE)
 
@@
 
-so as we can see... the number of origin ZNs with NO origin trips only goes down by 2.. (from 344 to 342). This is not a lot.
-
-Next step... let's try to see what is happening in these NA zones. What sort of origin trips are being generated? 
-```{r}
-missing_workers <- zn_noc %>% filter(is.na(Otrips)) %>% dplyr::select(ZT1259) #344 zonas are missing home to work trips.
-
-missing_workers <- missing_workers %>% merge(OD_v, by.x = "ZT1259", by.y="VORIZT1259", all.x = TRUE)
-
-temp <- missing_workers %>% st_drop_geometry() %>% group_by(ZT1259) %>% summarise(trip_count = n(),
-                                                                                  Owork_perc = sum(VORI == 2)/n(),
-                                                                                  Ostudy_perc = sum(VORI == 4)/n(),
-                                                                                  Osportwalk_perc = sum(VORI == 10)/n(),
-                                                                                  Oshop_perc = sum(VORI == 5)/n(),
-                                                                                  Ohome_perc = sum(VORI == 1)/n())
-temp
-```
-```{r}
-summary(temp)
-```
-```{r}
-# write.csv(missing_workers %>% st_drop_geometry(),"ZT_Trips_missingworkers.csv", row.names = FALSE)
-```
-
-‘trip count’ is how many trips there are in each ZT.
+so as we can see... the number of origin ZNs with NO origin trips doesn't go down - we only have 1 missing ZT using the 208 zonig system and 1 is still missing after including the chained trips.
+
+Next step... let's try to see what is happening in these NA zones. What sort of origin trips are being generated? (analysis for the 1259 zoning system)-->
+
+<!-- ```{r} -->
+<!-- missing_workers <- zn_noc %>% filter(is.na(Otrips)) %>% dplyr::select(ZT1259) #344 zonas are missing home to work trips. -->
+
+<!-- missing_workers <- missing_workers %>% merge(OD_v, by.x = "ZT1259", by.y="VORIZT1259", all.x = TRUE) -->
+
+<!-- temp <- missing_workers %>% st_drop_geometry() %>% group_by(ZT1259) %>% summarise(trip_count = n(), -->
+<!--                                                                                   Owork_perc = sum(VORI == 2)/n(), -->
+<!--                                                                                   Ostudy_perc = sum(VORI == 4)/n(), -->
+<!--                                                                                   Osportwalk_perc = sum(VORI == 10)/n(), -->
+<!--                                                                                   Oshop_perc = sum(VORI == 5)/n(), -->
+<!--                                                                                   Ohome_perc = sum(VORI == 1)/n()) -->
+<!-- temp -->
+<!-- ``` -->
+<!-- ```{r} -->
+<!-- summary(temp) -->
+<!-- ``` -->
+<!-- ```{r} -->
+<!-- # write.csv(missing_workers %>% st_drop_geometry(),"ZT_Trips_missingworkers.csv", row.names = FALSE) -->
+<!-- ``` -->
+
+<!-- ‘trip count’ is how many trips there are in each ZT.
 ‘Owork_perc’ is an origin work purpose (of that trip) in each ZT, i.e.
@@
 This means that these are predominately ‘work’ related areas and none of the households surveyed live in those ZTs.
-
-And looking in excel.
+-->
+
+<!-- And looking in excel.
 even if we consider all origin zones from work themed chained trips (i.e., home to somewhere to work) as 'residential locations' (this is incorrect) we STILL have ~170 zones with no work-to-home trips.
+-->
 
@@
 lnorm_ <- fitdistrplus::fitdist(data=transit_HW_alltt$TIEMPO_VIAJES,#%>% unlist() %>% as.numeric(),
-                                "lnorm", 
+                                "gamma", 
                                 method="mle",
@@
 
-lnorm_ <- fitdistrplus::fitdist(data=walk_HW_alltt$TIEMPO_VIAJES, #%>% unlist() %>% as.numeric(),
-                                "lnorm", 
+logis_ <- fitdistrplus::fitdist(data=walk_HW_alltt$TIEMPO_VIAJES, #%>% unlist() %>% as.numeric(),
+                                "logis", 
                                 method="mle",
@@
                                 optim.method="Nelder-Mead")
-HW_walk_lnorm <- lnorm_
+HW_walk_lnorm <- logis_
 ```
@@
 x_transit <- x %>%
-  mutate(f = dlnorm(t,
+  mutate(f = dgamma(t,
                     HW_transit_lnorm$estimate[1],
@@
 x_walk <- x %>%
-  mutate(f = dlnorm(t,
+  mutate(f = dlogis(t,
                     HW_walk_lnorm$estimate[1],
@@
 workers <- OD_v_HW %>%
-  group_by(VORIZT1259,
+  group_by(VORI_CD_ZT208,
            mode) %>%
@@
             .groups = "drop") %>%
-  rename(ZT1259 = VORIZT1259)
+  rename(CD_ZT208 = VORI_CD_ZT208)
 
@@
 jobs <- OD_v_HW %>%
-  group_by(VDESZT1259) %>%
+  group_by(VDES_CD_ZT208) %>%
   summarize(jobs = sum(ELE_G_POND_ESC2),
             .groups = "drop") %>%
-  rename(ZT1259 = VDESZT1259)
+  rename(CD_ZT208 = VDES_CD_ZT208)
 ```
@@
 ```{r}
-load("data/zn.RData")
-
 # zn <- zn %>%
@@
   # Join workers
-  left_join(zn,
-            by = "ZT1259") %>%
+  left_join(zn208,
+            by = "CD_ZT208") %>%
   st_as_sf()
@@
   # Join jobs
-  left_join(zn,
-            by = "ZT1259") %>%
+  left_join(zn208,
+            by = "CD_ZT208") %>%
   st_as_sf()
@@
 OD <- OD_v_HW %>%
-  group_by(VORIZT1259, 
-           VDESZT1259,
+  group_by(VORI_CD_ZT208,
+           VDES_CD_ZT208,
            mode) %>%
@@
   left_join(workers,
-            by = c("VORIZT1259" = "ZT1259", "mode"))
+            by = c("VORI_CD_ZT208" = "CD_ZT208", "mode"))
 
@@
   left_join(jobs,
-            by = c("VDESZT1259" = "ZT1259"))
+            by = c("VDES_CD_ZT208" = "CD_ZT208"))
 ```
@@
 OD_2 <- OD_2 %>%
-  mutate(f = case_when(mode == "transit" ~ dlnorm(tt,
-                                                  HW_transit_lnorm$estimate["meanlog"], 
-                                                  HW_transit_lnorm$estimate["sdlog"]),
+  mutate(f = case_when(mode == "transit" ~ dgamma(tt,
+                                                  HW_transit_lnorm$estimate["shape"], 
+                                                  HW_transit_lnorm$estimate["rate"]),
                        mode == "car/motor" ~ dgamma(tt,
@@
                                                HW_bike_lnorm$estimate["sdlog"]),
-                       mode == "walk" ~ dlnorm(tt, 
-                                               HW_walk_lnorm$estimate["meanlog"], 
-                                               HW_walk_lnorm$estimate["sdlog"])))
+                       mode == "walk" ~ dlogis(tt, 
+                                               HW_walk_lnorm$estimate["location"], 
+                                               HW_walk_lnorm$estimate["scale"])))
 ```
@@
   mutate(SA_ij = sp_avail(., 
-                          o_id = VORIZT1259,
-                          d_id = VDESZT1259,
+                          o_id = VORI_CD_ZT208,
+                          d_id = VDES_CD_ZT208,
                           pop = workers,
@@
 ```{r}
-sum(OD_SA$SA_ij)
+sum(OD_SA$SA_ij, na.rm = T)
 sum(jobs$jobs)
@@
 SA_i <- OD_SA %>%
-  group_by(VORIZT1259) %>%
+  group_by(VORI_CD_ZT208) %>%
   summarize(V_i = sum(SA_ij))
@@
 SA_im <- OD_SA %>%
-  group_by(VORIZT1259, 
+  group_by(VORI_CD_ZT208, 
            mode) %>%
@@
 population <- workers %>%
-  group_by(ZT1259) %>%
+  group_by(CD_ZT208) %>%
   summarize(population = sum(workers))
@@
   left_join(population,
-            by = c("VORIZT1259" = "ZT1259"))
+            by = c("VORI_CD_ZT208" = "CD_ZT208"))
 sum(SA_i$V_i)
@@
   left_join(population,
-            by = c("VORIZT1259" = "ZT1259"))
+            by = c("VORI_CD_ZT208" = "CD_ZT208"))
 sum(SA_im$V_im)
@@
 SA_i <- SA_i %>%
-  left_join(zn %>%
-              dplyr::select(ZT1259),
-            by = c("VORIZT1259" = "ZT1259")) %>%
+  left_join(zn208 %>%
+              dplyr::select(CD_ZT208),
+            by =c("VORI_CD_ZT208" = "CD_ZT208")) %>%
   st_as_sf()
@@
 SA_im <- SA_im %>%
-  left_join(zn %>%
-              dplyr::select(ZT1259),
-            by = c("VORIZT1259" = "ZT1259")) %>%
+  left_join(zn208 %>%
+              dplyr::select(CD_ZT208),
+            by = c("VORI_CD_ZT208" = "CD_ZT208")) %>%
   st_as_sf()
@@
 Preliminaries to plotting:
+
+Figure out which CD_ZT208 correspond to Madrid - we see that CD_ZT208 1 through 59 do so.
+
+```{r}
+t <- zn_u %>% dplyr::select(CD_ZT208, Municipality) %>% group_by(CD_ZT208,Municipality) %>% summarize() %>% filter(Municipality == "Madrid")
+t
+```
+
 ```{r create-bounding-boxes}
+zn208 <- zn208 %>% mutate(MM = ifelse(str_starts(ZT208,"Madrid"), 1, 0)) %>% filter(MM == 1)
+
 #create the bounding box for the Toronto region + a little extra
-bbox_new1 <- st_bbox(zn%>%filter(Municipality == "Madrid")) # current bounding box
+bbox_new1 <- st_bbox(zn208 %>%filter(MM == 1)) # current bounding box
 
@@
 
-Madrid_muni_border <- zn %>%
-  filter(Municipality == "Madrid") %>%
+sf::sf_use_s2(FALSE) #this terns off the spherical geometry component in the st_union() function - this assumes the sf is just planar. Thats' fine for our purposes
+Madrid_muni_border <- zn208 %>%
+  filter(MM == 1) %>%
   st_union()
@@
 ```
+
+The previous attempts at calculating spatial availability is commented out and included in the end of the document.
+There were three different approaches, and the approach reflective above is the 3rd approach tested.
+The first approach can be used in the case of time windows, but the 2nd and 3rd are similar just the 3rd approach is more conceptually easy to understand and code.
+
+# C) Exploring mode shifting, first an optimistic scenario - no cars (all transit) going into madrid municipality
+
+Let's test out a scenario - all car trips going into madrid municipality are now transit trips. How will this impact the SA profile? We will using the 2nd method (calculating SA within all modes).
+
+<!-- AS finished here May 31, am stuck on how to shift car trips to transit (in a way that's flexible and will allow us to easily toggle the modal shift scenarios (i.e., instead of 100% transit, maybe 80%/10/10 transit/bike/walk)) -->
+
+First, let's re-assign to-madrid muni car trips to transit trips:
+```{r}
+Madrid_muni <- zn208 %>% filter(MM == 1)
+
+# OD <- OD_v_HW %>%
+#   group_by(VORI_CD_ZT208,
+#            VDES_CD_ZT208,
+#            mode) %>%
+#   summarize(tt = mean(TIEMPO_VIAJES),
+#             .groups = "drop")
+
+OD_v_HW_VDES_Mad <- OD_v_HW %>% filter(VDES_Municipality == "Madrid") %>%
+  group_by(VORI_CD_ZT208, VDES_CD_ZT208, mode) %>% 
+  summarize(TIEMPO_VIAJES = mean(TIEMPO_VIAJES))
+
+OD_v_HW_VDES_Mad %>% mutate(tt = ifelse(mode == "car", ))
+
+OD_v_HW_optimistic1 %>% mutate(TIEMPO_VIAJES_opt = ifelse(mode="car" & MM == 1, NA, ))
+
+OD_v_HW_optimistic1 %>% group_by(mode) %>% summarize(car = mean())
+
+  mutate(TIEMPO_VIAJES_opt = ifelse(mode = "car", ))
+```
+
+```{r}
+#reassigning car trips to transit trips (i.e., a 100% shift to transit from car)
+OD_v_HW_optimistic1 <- OD_v_HW_optimistic  %>% filter(MM == 1) %>% #only CAR trips to Madrid
+  mutate(trips_transit = sum(trips_transit,trips_car_motor, na.rm = T), #only trips 
+         trips_car_motor = 0) #16627 ODs are making trips INTO muni, this is 59% of all OD pairs
+
+OD_v_HW_optimistic2 <- OD_v_HW_optimistic  %>% filter(is.na(MM))
+
+OD_v_HW_optimistic <- rbind(OD_v_HW_optimistic1,OD_v_HW_optimistic2,OD_v_HW_optimistic2)
+
+workers <- OD_v_HW_optimistic %>%
+  group_by(VORIZT1259) %>%
+  summarize(workers_transit = sum(trips_transit),
+            workers_car = sum(trips_car_motor),
+            workers_bike = sum(trips_bike),
+            workers_walk = sum(trips_walk),
+            .groups = "drop") %>%
+  transmute(ZT1259 = VORIZT1259,
+            workers_transit,
+            workers_car,
+            workers_bike,
+            workers_walk)
+
+jobs <- OD_v_HW_optimistic %>%
+  group_by(VDESZT1259) %>%
+  summarize(jobs_transit = sum(trips_transit),
+            jobs_car = sum(trips_car_motor),
+            jobs_bike = sum(trips_bike),
+            jobs_walk = sum(trips_walk),
+            .groups = "drop") %>%
+  transmute(ZT1259 = VDESZT1259,
+            jobs_transit,
+            jobs_car,
+            jobs_bike,
+            jobs_walk)
+```
+
+Let's quickly compare how many ODs which have car trips changed to transit had EXISTING transit trips. This may inform what travel time we use.
+```{r}
+madrid_trips_reassigned <- OD_v_HW_optimistic1 %>%
+  group_by(VORIZT1259, VDESZT1259) %>%
+  summarise(NEWtrips_transit = sum(trips_transit),
+            NEWtrips_car_motor = sum(trips_car_motor))
+
+madrid_trips_original <-OD_v_HW %>%
+  group_by(VORIZT1259, VDESZT1259) %>%
+  summarise(trips_transit = sum(trips_transit),
+            trips_car_motor = sum(trips_car_motor))
+
+madrid_trips_reassigned <- madrid_trips_reassigned %>% merge(madrid_trips_original, by =c("VORIZT1259","VDESZT1259" ))
+#all the trip_.y, are the original trips as specified in the travel survey. the trips.x are the assigned trips (moved to transit).
+
+# percentage of transit trips that existed in the original OD pair
+
+countt <- madrid_trips_reassigned %>% mutate(transit_count = ifelse(trips_transit != 0, 1, 0))
+
+sum(countt$transit_count)/nrow(madrid_trips_reassigned)
+```
+
+We can see that only ~50% of the OD pairs had a transit trip originally - so we can only borrow transit trip times from thest ODs...
+we'd have to figure out how to subsitute the remaining 50%.
+Alternatively - we can figure out any other solution to apply to all mode-shifted trips.
+
+For simplicity, let's continue and  just take the travel time (for car) for these trips and put it on the transit impedance curve.
+
+To continue this process, join the workers and jobs to the ZN simple features object:
+```{r}
+load("data/zn.RData")
+
+zn <- zn %>%
+  # Join workers
+  left_join(workers,
+            by = "ZT1259") %>%
+  # Join jobs
+  left_join(jobs,
+            by = "ZT1259") %>%
+  # Replace NAs with zeros
+  mutate(workers_transit = tidyr::replace_na(workers_transit, 0),
+         workers_car = tidyr::replace_na(workers_car, 0),
+         workers_bike = tidyr::replace_na(workers_bike, 0),
+         workers_walk = tidyr::replace_na(workers_walk, 0),
+         jobs_transit = tidyr::replace_na(jobs_transit, 0),
+         jobs_car = tidyr::replace_na(jobs_car, 0),
+         jobs_bike = tidyr::replace_na(jobs_bike, 0),
+         jobs_walk = tidyr::replace_na(jobs_walk, 0))
+```
+
+add the number of workers and jobs from each zona to each trip:
+```{r}
+#add the number of jobs and workers to the od_ft_tt matrix
+OD_optimistic <- OD_v_HW_optimistic %>% merge(zn %>% dplyr::select(ZT1259, workers_transit,workers_car,workers_bike,workers_walk) %>% st_drop_geometry(),
+                                              by.x = "VORIZT1259", by.y="ZT1259", all.x = TRUE)
+
+OD_optimistic <- OD_optimistic %>% merge(zn %>% dplyr::select(ZT1259, jobs_transit,jobs_car,jobs_bike,jobs_walk) %>% st_drop_geometry(),
+                                         by.x = "VDESZT1259", by.y="ZT1259", all.x = TRUE)
+
+OD_optimistic <- OD_optimistic %>%
+  mutate(workers_transit = ifelse(is.na(workers_transit), 0, workers_transit),
+         workers_car = ifelse(is.na(workers_car), 0, workers_car),
+         workers_bike = ifelse(is.na(workers_bike), 0, workers_bike),
+         workers_walk = ifelse(is.na(workers_walk), 0, workers_walk),
+         jobs_car = ifelse(is.na(jobs_car), 0, jobs_car),
+         jobs_transit = ifelse(is.na(jobs_transit), 0, jobs_transit),
+         jobs_bike = ifelse(is.na(jobs_bike), 0, jobs_bike),
+         jobs_walk = ifelse(is.na(jobs_walk), 0, jobs_walk))
+```
+
+now reshape the data, each row will represent an any origin to to 1 destination (i.e., destination row is unique).
+We will average the travel times, by mode, and then use this average travel time to find the 'f' (impedance value for travel for each mode).
+Note: this is a *KEY* assumption - I'm giving the new transit trips the SAME travel time as they had when traveling by car but I am predicting the travel cost using the impedance curve of TRANSIT instead of CAR - that's the only difference.
+This is certainly a strange assumption.
+A solution could be to use the r5r calculated transit travel times but as we saw they don't fit the original data travel times well...
+```{r}
+#consolidate trips so each trip is unique. workers = jobs.
+OD_optimistic <- OD_optimistic %>%
+  group_by(VORIZT1259,VDESZT1259) %>%
+  summarise(trips_transit = sum(trips_transit),
+            trips_car_motor = sum(trips_car_motor),
+            trips_bike = sum(trips_bike),
+            trips_walk = sum(trips_walk),
+            workers_transit = mean(workers_transit),
+            # workers_transit = ifelse(is.na(workers_transit), 0, workers_transit),
+            workers_car = mean(workers_car),
+            # workers_car = ifelse(is.na(workers_car), 0, workers_car),
+            workers_bike = mean(workers_bike),
+            # workers_bike = ifelse(is.na(workers_bike), 0, workers_bike),
+            workers_walk = mean(workers_walk),
+            # workers_walk = ifelse(is.na(workers_walk), 0, workers_walk),
+            jobs_transit = mean(jobs_transit),
+            jobs_car = mean(jobs_car),
+            jobs_bike = mean(jobs_bike),
+            jobs_walk = mean(jobs_walk),
+            # jobs_car = ifelse(is.na(jobs_car), 0, jobs_car),
+            # jobs_transit = ifelse(is.na(jobs_transit), 0, jobs_transit),
+            # jobs_bike = ifelse(is.na(jobs_bike), 0, jobs_bike),
+            # jobs_walk = ifelse(is.na(jobs_walk), 0, jobs_walk),
+            tt_transit = ifelse(trips_transit != 0, mean(TIEMPO_VIAJES, na.rm=T), NA),
+            tt_car = ifelse(trips_car_motor != 0, mean(TIEMPO_VIAJES, na.rm=T), NA),
+            tt_bike = ifelse(trips_bike != 0, mean(TIEMPO_VIAJES, na.rm=T), NA),
+            tt_walk = ifelse(trips_walk != 0, mean(TIEMPO_VIAJES, na.rm=T), NA)) %>% ungroup()
+
+```
+
+now estimate f value using these averaged travel times:
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+#dgamma is the density plot for gamma function and dlnorm is the desntiy plot for the lognormal function; adding values as "f" our impedance function
+OD_optimistic <- OD_optimistic %>%
+  mutate(f_car = dgamma(tt_car, HW_car_motor_gamma$estimate["shape"], HW_car_motor_gamma$estimate["rate"]),
+         f_car = ifelse(is.na(f_car), 0, f_car),
+         f_transit = dlnorm(tt_transit, HW_transit_lnorm$estimate["meanlog"], HW_transit_lnorm$estimate["sdlog"]),
+         f_transit = ifelse(is.na(f_transit), 0, f_transit),
+         f_bike = dlnorm(tt_bike, HW_bike_lnorm$estimate["meanlog"], HW_bike_lnorm$estimate["sdlog"]),
+         f_bike = ifelse(is.na(f_bike), 0, f_bike),
+         f_walk = dlnorm(tt_walk, HW_walk_lnorm$estimate["meanlog"], HW_walk_lnorm$estimate["sdlog"]),
+         f_walk = ifelse(is.na(f_walk), 0, f_walk))
+```
+
+Calculating SA:
+```{r}
+#add new summed columns
+OD_optimistic <- OD_optimistic %>%
+  mutate(workers_ = (workers_car+ workers_transit+ workers_bike+ workers_walk),
+         workers_ = ifelse(is.na(workers_), 0, workers_),
+         jobs_ = (jobs_car + jobs_transit + jobs_bike + jobs_walk),
+         jobs_ = ifelse(is.na(jobs_), 0, jobs_),
+         f_ = f_car + f_transit + f_bike + f_walk)
+
+#calculate spatial availability
+OD_optimistic <- OD_optimistic %>%
+  mutate(catch = 1) %>%
+  mutate(SA_ij = sp_avail(.,
+                          o_id = VORIZT1259,
+                          d_id = VDESZT1259,
+                          pop = workers_,
+                          opp = jobs_,
+                          r = catch,
+                          f = f_),
+         SA_ij = ifelse(is.nan(SA_ij), 0, SA_ij))
+
+#verify that the sum of all jobs is consistent with the number of jobs
+sum(OD_optimistic$SA_ij)
+
+#verify that the sum of all jobs is consistent with the number of jobs - from another perspective
+sum_jobs <- OD_optimistic %>%
+  group_by(VDESZT1259) %>%
+  summarise(jobs = mean(jobs_))
+sum(sum_jobs$jobs, na.rm = T)
+#Perfect it works!
+```
+
+Let's aggregate the spatial availability by origin zone:
+```{r}
+#aggregating spatial availability
+MAD_availability <- OD_optimistic %>%
+  group_by(VORIZT1259) %>%
+  summarize(SA_i3 = sum(SA_ij),
+            workers_ = mean(workers_car, na.rm=T) + mean(workers_transit, na.rm=T) + mean(workers_bike, na.rm=T) + mean(workers_walk, na.rm=T),
+            avgtt_i = mean(mean(tt_transit) + mean(tt_walk) + mean(tt_car) + mean(tt_bike)),
+            avg_f_i = mean(mean(f_transit) + mean(f_walk) + mean(f_car) + mean(f_bike)))
+
+#Merge TO availability calculation to the TAZ sf object created for accessibility above:
+zn_test3 <- zn %>% merge(MAD_availability, by.x=c("ZT1259"), by.y=c("VORIZT1259"), all.x=T)
+```
+
+Here we plot plot 2 (the original scenario) and plot 3 ( this optimistic scenario)
+```{r}
+#plot 2
+plot_MAD_avail_perworker_3 <- ggplot() +
+  geom_sf(data = zn_test3, aes(fill= SA_i3/workers_), color = NA) + #data
+  scale_fill_distiller(palette = "RdYlGn", #legend scale bar
+                       name = expression(v["i"]),
+                       na.value = "grey90",
+                       limits = c( round(min(zn_test3$SA_i3/zn_test3$workers_, na.rm=T)),
+                                   round(max(zn_test3$SA_i3/zn_test3$workers_, na.rm=T))))+
+  geom_sf(data = Madrid_muni_border, # border for Madrid muni
+          colour=alpha("dimgrey",1),
+          size = 0.5, fill=NA,
+          show.legend = "polygon") +
+  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
+                         height = unit(0.8, "cm"),
+                         width = unit(0.8, "cm"),
+                         style = north_arrow_orienteering(line_width = 0.25,
+                                                          line_col = "dimgrey",
+                                                          fill = c("white","dimgrey"))) +
+  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
+                   height = unit(0.15, "cm")) +
+  theme_void() +
+  theme(legend.position = c(1.10, 0.25),
+        plot.margin = margin(t = 0, r = 0, b = 0, l = -80)) # positing legend and clipping out white space
+
+# plot 1 + inset map
+plot_MAD_avail_perworker_3 <- ggdraw(plot_MAD_avail_perworker_3) +
+  draw_plot({plot_MAD_avail_perworker_3 + coord_sf(xlim = st_coordinates(bbox_Madrid_muni)[c(1,2),1],
+                                                   ylim = st_coordinates(bbox_Madrid_muni)[c(2,3),2],
+                                                   expand = FALSE) +
+      theme(legend.position = "none")},
+      x = 0.65, y = 0.50,  # The distance along a (0,1) x-axis (and y axis) to draw the left edge (the bottom edge) of the plot
+      width = 0.46, height = 0.46) # The width and height of the plot expressed as proportion of the entire ggdraw object
+
+plot_MAD_avail_perworker_2
+plot_MAD_avail_perworker_3
+```
+
+As can be seen, the change (visability) is not that dramatic.
+
+The commented out different spatial availability calculation approaches (also done at the ZT1259 resolution instead of ZT208)
 <!--
@@
 ```
-
-# C) Exploring mode shifting, first an optimistic scenario - no cars (all transit) going into madrid municipality
-
-Let's test out a scenario - all car trips going into madrid municipality are now transit trips. How will this impact the SA profile? We will using the 2nd method (calculating SA within all modes).
-
-First, let's re-assign to-madrid muni car trips to transit trips:
-```{r}
-Madrid_muni <- zn %>% filter(Municipality == "Madrid") %>% 
-  dplyr::transmute(ZT1259,
-                   MM = 1)
-
-OD_v_HW_optimistic <- OD_v_HW  %>% merge(Madrid_muni, by.x=("VDESZT1259"), by.y=("ZT1259"), all.x=T)
-
-#reassigning car trips to transit
-OD_v_HW_optimistic1 <- OD_v_HW_optimistic  %>% filter(MM == 1) %>% rowwise() %>%
-  mutate(trips_transit = sum(trips_transit,trips_car_motor, na.rm = T),
-         trips_car_motor = 0) #16627 ODs are making trips INTO muni, this is 59% of all OD pairs
-
-OD_v_HW_optimistic2 <- OD_v_HW_optimistic  %>% filter(is.na(MM))
-
-OD_v_HW_optimistic <- rbind(OD_v_HW_optimistic1,OD_v_HW_optimistic2)
-
-workers <- OD_v_HW_optimistic %>%
-  group_by(VORIZT1259) %>%
-  summarize(workers_transit = sum(trips_transit),
-            workers_car = sum(trips_car_motor),
-            workers_bike = sum(trips_bike),
-            workers_walk = sum(trips_walk),
-            .groups = "drop") %>%
-  transmute(ZT1259 = VORIZT1259,
-            workers_transit,
-            workers_car,
-            workers_bike,
-            workers_walk)
-
-jobs <- OD_v_HW_optimistic %>%
-  group_by(VDESZT1259) %>%
-  summarize(jobs_transit = sum(trips_transit),
-            jobs_car = sum(trips_car_motor),
-            jobs_bike = sum(trips_bike),
-            jobs_walk = sum(trips_walk),
-            .groups = "drop") %>%
-  transmute(ZT1259 = VDESZT1259,
-            jobs_transit,
-            jobs_car,
-            jobs_bike,
-            jobs_walk)
-```
-
-Let's quickly compare how many ODs which have car trips changed to transit had EXISTING transit trips. This may inform what travel time we use.
-```{r}
-madrid_trips_reassigned <- OD_v_HW_optimistic1 %>% 
-  group_by(VORIZT1259, VDESZT1259) %>%
-  summarise(NEWtrips_transit = sum(trips_transit),
-            NEWtrips_car_motor = sum(trips_car_motor))
-
-madrid_trips_original <-OD_v_HW %>%
-  group_by(VORIZT1259, VDESZT1259) %>%
-  summarise(trips_transit = sum(trips_transit),
-            trips_car_motor = sum(trips_car_motor))
-
-madrid_trips_reassigned <- madrid_trips_reassigned %>% merge(madrid_trips_original, by =c("VORIZT1259","VDESZT1259" ))
-#all the trip_.y, are the original trips as specified in the travel survey. the trips.x are the assigned trips (moved to transit).
-
-# percentage of transit trips that existed in the original OD pair
-
-countt <- madrid_trips_reassigned %>% mutate(transit_count = ifelse(trips_transit != 0, 1, 0))
-
-sum(countt$transit_count)/nrow(madrid_trips_reassigned)
-```
-
-We can see that only ~50% of the OD pairs had a transit trip originally - so we can only borrow transit trip times from thest ODs...
-we'd have to figure out how to subsitute the remaining 50%.
-Alternatively - we can figure out any other solution to apply to all mode-shifted trips.
-
-For simplicity, let's continue and  just take the travel time (for car) for these trips and put it on the transit impedance curve.
-
-To continue this process, join the workers and jobs to the ZN simple features object:
-```{r}
-load("data/zn.RData")
-
-zn <- zn %>%
-  # Join workers
-  left_join(workers,
-            by = "ZT1259") %>%
-  # Join jobs
-  left_join(jobs,
-            by = "ZT1259") %>%
-  # Replace NAs with zeros
-  mutate(workers_transit = tidyr::replace_na(workers_transit, 0),
-         workers_car = tidyr::replace_na(workers_car, 0),
-         workers_bike = tidyr::replace_na(workers_bike, 0),
-         workers_walk = tidyr::replace_na(workers_walk, 0),
-         jobs_transit = tidyr::replace_na(jobs_transit, 0),
-         jobs_car = tidyr::replace_na(jobs_car, 0),
-         jobs_bike = tidyr::replace_na(jobs_bike, 0),
-         jobs_walk = tidyr::replace_na(jobs_walk, 0))
-```
-
-add the number of workers and jobs from each zona to each trip:
-```{r}
-#add the number of jobs and workers to the od_ft_tt matrix
-OD_optimistic <- OD_v_HW_optimistic %>% merge(zn %>% dplyr::select(ZT1259, workers_transit,workers_car,workers_bike,workers_walk) %>% st_drop_geometry(),
-                                              by.x = "VORIZT1259", by.y="ZT1259", all.x = TRUE)
-
-OD_optimistic <- OD_optimistic %>% merge(zn %>% dplyr::select(ZT1259, jobs_transit,jobs_car,jobs_bike,jobs_walk) %>% st_drop_geometry(),
-                                         by.x = "VDESZT1259", by.y="ZT1259", all.x = TRUE)
-
-OD_optimistic <- OD_optimistic %>% 
-  mutate(workers_transit = ifelse(is.na(workers_transit), 0, workers_transit),
-         workers_car = ifelse(is.na(workers_car), 0, workers_car),
-         workers_bike = ifelse(is.na(workers_bike), 0, workers_bike),
-         workers_walk = ifelse(is.na(workers_walk), 0, workers_walk),
-         jobs_car = ifelse(is.na(jobs_car), 0, jobs_car),
-         jobs_transit = ifelse(is.na(jobs_transit), 0, jobs_transit),
-         jobs_bike = ifelse(is.na(jobs_bike), 0, jobs_bike),
-         jobs_walk = ifelse(is.na(jobs_walk), 0, jobs_walk))
-```
-
-now reshape the data, each row will represent an any origin to to 1 destination (i.e., destination row is unique).
-We will average the travel times, by mode, and then use this average travel time to find the 'f' (impedance value for travel for each mode).
-Note: this is a *KEY* assumption - I'm giving the new transit trips the SAME travel time as they had when traveling by car but I am predicting the travel cost using the impedance curve of TRANSIT instead of CAR - that's the only difference.
-This is certainly a strange assumption.
-A solution could be to use the r5r calculated transit travel times but as we saw they don't fit the original data travel times well...
-```{r}
-#consolidate trips so each trip is unique. workers = jobs.
-OD_optimistic <- OD_optimistic %>%
-  group_by(VORIZT1259,VDESZT1259) %>%
-  summarise(trips_transit = sum(trips_transit),
-            trips_car_motor = sum(trips_car_motor),
-            trips_bike = sum(trips_bike),
-            trips_walk = sum(trips_walk),
-            workers_transit = mean(workers_transit),
-            # workers_transit = ifelse(is.na(workers_transit), 0, workers_transit),
-            workers_car = mean(workers_car),
-            # workers_car = ifelse(is.na(workers_car), 0, workers_car),
-            workers_bike = mean(workers_bike),
-            # workers_bike = ifelse(is.na(workers_bike), 0, workers_bike),
-            workers_walk = mean(workers_walk),
-            # workers_walk = ifelse(is.na(workers_walk), 0, workers_walk),
-            jobs_transit = mean(jobs_transit),
-            jobs_car = mean(jobs_car),
-            jobs_bike = mean(jobs_bike),
-            jobs_walk = mean(jobs_walk),
-            # jobs_car = ifelse(is.na(jobs_car), 0, jobs_car),
-            # jobs_transit = ifelse(is.na(jobs_transit), 0, jobs_transit),
-            # jobs_bike = ifelse(is.na(jobs_bike), 0, jobs_bike),
-            # jobs_walk = ifelse(is.na(jobs_walk), 0, jobs_walk),
-            tt_transit = ifelse(trips_transit != 0, mean(TIEMPO_VIAJES, na.rm=T), NA),
-            tt_car = ifelse(trips_car_motor != 0, mean(TIEMPO_VIAJES, na.rm=T), NA),
-            tt_bike = ifelse(trips_bike != 0, mean(TIEMPO_VIAJES, na.rm=T), NA),
-            tt_walk = ifelse(trips_walk != 0, mean(TIEMPO_VIAJES, na.rm=T), NA)) %>% ungroup()
-
-```
-
-now estimate f value using these averaged travel times:
-```{r, echo=FALSE, warning=FALSE, message=FALSE}
-#dgamma is the density plot for gamma function and dlnorm is the desntiy plot for the lognormal function; adding values as "f" our impedance function
-OD_optimistic <- OD_optimistic %>%
-  mutate(f_car = dgamma(tt_car, HW_car_motor_gamma$estimate["shape"], HW_car_motor_gamma$estimate["rate"]),
-         f_car = ifelse(is.na(f_car), 0, f_car),
-         f_transit = dlnorm(tt_transit, HW_transit_lnorm$estimate["meanlog"], HW_transit_lnorm$estimate["sdlog"]),
-         f_transit = ifelse(is.na(f_transit), 0, f_transit),
-         f_bike = dlnorm(tt_bike, HW_bike_lnorm$estimate["meanlog"], HW_bike_lnorm$estimate["sdlog"]),
-         f_bike = ifelse(is.na(f_bike), 0, f_bike),
-         f_walk = dlnorm(tt_walk, HW_walk_lnorm$estimate["meanlog"], HW_walk_lnorm$estimate["sdlog"]),
-         f_walk = ifelse(is.na(f_walk), 0, f_walk))
-```
-
-Calculating SA:
-```{r}
-#add new summed columns
-OD_optimistic <- OD_optimistic %>%
-  mutate(workers_ = (workers_car+ workers_transit+ workers_bike+ workers_walk),
-         workers_ = ifelse(is.na(workers_), 0, workers_),
-         jobs_ = (jobs_car + jobs_transit + jobs_bike + jobs_walk),
-         jobs_ = ifelse(is.na(jobs_), 0, jobs_),
-         f_ = f_car + f_transit + f_bike + f_walk)
-
-#calculate spatial availability
-OD_optimistic <- OD_optimistic %>%
-  mutate(catch = 1) %>%
-  mutate(SA_ij = sp_avail(., 
-                          o_id = VORIZT1259,
-                          d_id = VDESZT1259,
-                          pop = workers_,
-                          opp = jobs_,
-                          r = catch,
-                          f = f_),
-         SA_ij = ifelse(is.nan(SA_ij), 0, SA_ij))
-
-#verify that the sum of all jobs is consistent with the number of jobs
-sum(OD_optimistic$SA_ij)
-
-#verify that the sum of all jobs is consistent with the number of jobs - from another perspective
-sum_jobs <- OD_optimistic %>% 
-  group_by(VDESZT1259) %>% 
-  summarise(jobs = mean(jobs_))
-sum(sum_jobs$jobs, na.rm = T)
-#Perfect it works! 
-```
-
-Let's aggregate the spatial availability by origin zone:
-```{r}
-#aggregating spatial availability  
-MAD_availability <- OD_optimistic %>%
-  group_by(VORIZT1259) %>%
-  summarize(SA_i3 = sum(SA_ij),
-            workers_ = mean(workers_car, na.rm=T) + mean(workers_transit, na.rm=T) + mean(workers_bike, na.rm=T) + mean(workers_walk, na.rm=T),
-            avgtt_i = mean(mean(tt_transit) + mean(tt_walk) + mean(tt_car) + mean(tt_bike)),
-            avg_f_i = mean(mean(f_transit) + mean(f_walk) + mean(f_car) + mean(f_bike)))
-
-#Merge TO availability calculation to the TAZ sf object created for accessibility above:
-zn_test3 <- zn %>% merge(MAD_availability, by.x=c("ZT1259"), by.y=c("VORIZT1259"), all.x=T) 
-```
-
-Here we plot plot 2 (the original scenario) and plot 3 ( this optimistic scenario)
-```{r}
-#plot 2
-plot_MAD_avail_perworker_3 <- ggplot() +
-  geom_sf(data = zn_test3, aes(fill= SA_i3/workers_), color = NA) + #data
-  scale_fill_distiller(palette = "RdYlGn", #legend scale bar
-                       name = expression(v["i"]),
-                       na.value = "grey90",
-                       limits = c( round(min(zn_test3$SA_i3/zn_test3$workers_, na.rm=T)), 
-                                   round(max(zn_test3$SA_i3/zn_test3$workers_, na.rm=T))))+
-  geom_sf(data = Madrid_muni_border, # border for Madrid muni
-          colour=alpha("dimgrey",1), 
-          size = 0.5, fill=NA, 
-          show.legend = "polygon") +
-  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
-                         height = unit(0.8, "cm"), 
-                         width = unit(0.8, "cm"),
-                         style = north_arrow_orienteering(line_width = 0.25,
-                                                          line_col = "dimgrey", 
-                                                          fill = c("white","dimgrey"))) +
-  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
-                   height = unit(0.15, "cm")) +
-  theme_void() +
-  theme(legend.position = c(1.10, 0.25), 
-        plot.margin = margin(t = 0, r = 0, b = 0, l = -80)) # positing legend and clipping out white space
-
-# plot 1 + inset map
-plot_MAD_avail_perworker_3 <- ggdraw(plot_MAD_avail_perworker_3) +
-  draw_plot({plot_MAD_avail_perworker_3 + coord_sf(xlim = st_coordinates(bbox_Madrid_muni)[c(1,2),1],
-                                                   ylim = st_coordinates(bbox_Madrid_muni)[c(2,3),2],
-                                                   expand = FALSE) +
-      theme(legend.position = "none")},
-      x = 0.65, y = 0.50,  # The distance along a (0,1) x-axis (and y axis) to draw the left edge (the bottom edge) of the plot
-      width = 0.46, height = 0.46) # The width and height of the plot expressed as proportion of the entire ggdraw object
-
-plot_MAD_avail_perworker_2
-plot_MAD_avail_perworker_3
-```
-
-As can be seen, the change (visability) is not that dramatic.

### 2023-05-18 Draft-manuscript-tobemoved.Rmd
Added 'multi-mode-acc-methods', a manuscript for multi-model accessibility draft.

@@
 
-- lit has found that emissions reduciton should not result in inequalities..
+- lit has found that emissions reduction should not result in inequalities..
 

### 2023-05-18 Multi-mode-acc-methods.qmd
Added 'multi-mode-acc-methods', a manuscript for multi-model accessibility draft.

---
title: "How does my active transport access change when car access is restricted? Measuring multi-modal accessibility with spatial availability"
format: docx
bibliography: bibliography.bib
---

# Abstract

An increasing number of studies within the domain of transportation planning are concerned with the assessment of access to opportunity inequities. Who (the characteristics of people) and how (what modes are used) people can access opportunities of interest are questions central to planning for equity and towards justice. These questions can be addressed, through the imagination of policy intervention scenarios, but methods within the current multi-modal accessibility literature are lacking. In this spirit, we present _spatial availability_ within the context of multi-modal accessibility. We first illustrate its features. We then demonstrate its use on the case study of Low Emission Zones in Madrid (Spain). In using spatial availability, a singly-constrained accessibility measure, imagining policy scenarios that move regions towards a more equitable access-to-opportunity landscape is possible.

# Introduction

<!-- outlines the use of place-based accessibility measures in the literature: equity, planning -- esp. as it relates to modal competition (brief) -->
Location-based access (or accessibility) measures are ubiquitous in the literature and are present in a variety of configurations. They aim to capture the "potential interaction" that the population at a location has with opportunities based on transportation systems and land-use. Their implementation varies widely depending on the the research question and goal.

<!-- different types of location-based accessibility measures: constrained vs. unconstrained. Introduce examples of unconstrained measures. -->
From the perspective of representing a score of _opportunities_ that can be accessed by the population at a location, location-based accessibility measures can be categorized as 1) cumulative opportunity measures, 2) weighted cumulative opportunity measures, and 3) competitive accessibility measures as defined in [@levinsonTransportAccessManual2020]. All these types of measures use a distance decay function (otherwise known as an impedance function) to discount the opportunities that can be potentially accessed based on the travel cost for the population. These three types of accessibility measures can be unified into two categories. Cumulative and weighted cumulative opportunity measures are related: in fact, they just have differing impedance functions. The cumulative opportunity measure has a binary impedance function and the weighted cumulative opportunity measure has an impedance function that can take any functional form (e.g, exponential, log-normal, or even any sort of probability density function as discussed in the Trip Length Distribution literature). From another perspective, these measure are also unconstrained. The impedance function that discounted the opportunities is not constrained by any mass effect. In this way, cumulative and weighted cumulative measures can be considered _unconstrained_ accessibility measures.

<!-- examples of constrained (competitive) measures, the introduction of spatial availability -->

The third type of accessibility measure, competitive measures, discount the number of opportunities allocated based on the population that is demanding those opportunities in addition to discounting by travel impedance (like unconstrained accessibility measures). Competitive measures hence consider both the supply of the opportunities and the demand for those opportunities by the population in addition to the travel impedance. They have been criticized for being difficult to compute (relative to unconstrained measures) and difficult to interpret as they are a ratio of supply to demand [@levinsonTransportAccessManual2020]. In @soukhovIntroducingSpatialAvailability2023, we introduce spatial availability, a type of competitive accessibility measure that is _constrained_ beside a popular _unconstrained_ accessibility measure (Hansen-type gravity model) and a competitive accessibility measure (two-step floating catchment area (2SFCA)). We title spatial availability a _constrained_ accessibility measure since it proportionally allocates all the opportunities (supply) to each origin-location based on competition from the population (demand) and travel impedance in a single step with an interpretable output. The output is the number of opportunities available in the region, per location, such that if all the values for the locations in the region were summed they would equal the total opportunities in the region.

<!-- policy applications of accessibility broadly, but what's missing is multi-modal accessibility methods -->
The consideration of both supply of opportunities and demand from the population is important for policy applications. Spatial availability can be used to create supply per capita benchmarks for locations, and the delta of supply from this benchmark to existing conditions can be used to motivate policy. Currently, conventional _unconstrained_ accessibility measures are difficult for planners to operationalize for a variety of reasons. Furthermore, exisiting competitive measures are difficult to interpret.

However, this is a gap in the literature. Researchers and transport planners needs accessibility tools that interpretable consider competition. This is the aim of this paper. An example of a case where conventional accessibility measures (unconstrained and competitive) fall short is in considering multi-modal accessibility. Take for instance, measuring the accessibility to opportunities for a region for car and transit modes. The literature suggests that these modes are in direct competition. A user, if they have access to both modes, will consider the utility (travel impedance) of using one or the other. Thus, the potential for the population to interact with opportunities is directly dependant on ... the opportunities..

<!-- introduction of multi-modal accessibility. when modes compete with one another. at the center of this is that modes SHOULD compete. competition considerations is required.-->

## An introduction to the limitations of multi-modal methods

<!-- no competition: multi-modal accessibility examples -->
@tahmasbiMultimodalAccessibilitybasedEquity2019 computes accessibility across modes. No competition between modes is considered. Methodologically...

<!-- with competition but NOT proportional allocation: multimodal accessibility examples-->
@carpentieriMultimodalAccessibilityPrimary2020 "Multimodal Accessibility to Primary Health Services for the Elderly: A Case Study of Naples, Italy"

@taoInvestigatingImpactsPublic2020a "Investigating the impacts of public transport on job accessibility in Shenzhen, China: a multi-modal approach"

@maoMeasuringSpatialAccessibility2013 "Measuring spatial accessibility to healthcare for populations with multiple transportation modes"

A clustering method to modify the cumulative accessibility measure @mohriClusteringMethodMeasuring2021
<!-- summary of the paper-->
With the inconsistencies and limitations of existing measures in mind, we propose the multimodal application of spatial availability. To illustrate this application, in Section 2, we demonstrate the formulation of spatial availability in a simple example of two modes and three population and opportunity centers in a synthetic example. In Section 3, we then demonstrate an empirical example of the low emission zone in Madrid Spain. Specifically, how an expansion in the restriction of private vehicles changes the access to opportunities for transit, cycling and walking. In Section 4, we provide concluding remarks on the strengths of the use of spatial availability as a multimodal accessibility measure, potential future uses in policy planning scenarios, and limitations of accessibility measures.

# Illustrative example: using spatial availability for multiple modes

In brief, we define the _spatial availability_ at $i$ ($V_{i}$) as the proportion of all opportunities $O$ that are allocated to $i$ from all destinations $j$:
$$
V_i = \sum_{j=1}^NO_jF^t_{ij}
$$

\noindent where:

-   $F^t_{ij}$ is a balancing factor that depends on the population and cost of movement in the system.
-   $O_j$ is the number of opportunities at $j$.
-   $V_i$ is the number of spatially available opportunities from the perspective of $i$.

The general form of spatial availability is also as a sum, and the fundamental difference between Hansen- and Shen-type accessibility is that opportunities are allocated proportionally. Balancing factor $F^t_{ij}$ consists of two components: a population-based balancing factor $F^p_{i}$ and an impedance-based balancing factor $F^c_{ij}$ that, respectively, allocate opportunities to $i$ in proportion to the size of the population of the different competing centers (the mass effect of the gravity model) and the cost of reaching opportunities (the impedance effect).

We adapt the illustrative example used in @shen1998 assuming two modes and demonstrate spatial availability's within this multimodal application. <!-- TO DO: demonstrate an illustrate of the example-->

## Proportional allocation by population

According to the gravity modelling framework, the potential for interaction depends on the mass (i.e., the population) and the friction of distance (i.e., the impedance function). We begin by describing the proposed proportional allocation mechanism based on demand by the population. Recall, the total population in the example is 210,000. The proportion of the population by population center, by the two modes ($x$, and $z$) is:
$$
\begin{array}{l}
F^p_A = \frac{50,000}{210,000} & \text{where } F^{px}_A = \frac{40,000}{210,000} \text{ & } F^{pz}_A = \frac{10,000}{210,000}\\ 
\\
F^p_B = \frac{150,000}{210,000} & \text{where } F^{px}_B = \frac{100,000}{210,000} \text{ & } F^{pz}_B = \frac{50,000}{210,000}\\
\\
F^p_C = \frac{10,000}{210,000} & \text{where } F^{px}_C = \frac{7,000}{210,000} \text{ & } F^{pz}_C = \frac{3,000}{210,000}\\
\end{array}
$$

Competition between the subsets of the populations that use the jobs matters. After all, there is a finite number of jobs (210,000 in our example). Jobs are allocated proportionally from each employment center to each population center depending on their population sizes as per the balancing factors $F^p_i$. In this way, employment center 1 allocates $100,000\cdot \frac{50,000}{210,000}= 23,809.52$ jobs to $A$. This  $100,000\cdot \frac{150,000}{210,000}= 71,428.57$ jobs to $B$; and $100,000\cdot \frac{10,000}{210,000}= 7,142.857$ jobs to $C$. Notice how this mechanism ensures that the total number of jobs at employment center 1 is preserved at 100,000.

We can verify that the number of jobs allocated is consistent with the total number of jobs in the system:
$$
\begin{array}{l}
\text{Employment center 1 to population centers A, B, and C: }\\
100,000 \cdot \frac{50,000}{210,000} + 100,000 \cdot \frac{150,000}{210,000} + 100,000 \cdot \frac{10,000}{210,000} = 100,000\\
\\
\text{Employment center 2 to population centers A, B, and C: }\\
100,000 \cdot \frac{50,000}{210,000} + 100,000 \cdot \frac{150,000}{210,000} + 100,000 \cdot \frac{10,000}{210,000} = 100,000\\
\\
\text{Employment center 3 to population centers A, B, and C: }\\
10,000 \cdot \frac{50,000}{210,000} + 10,000 \cdot \frac{150,000}{210,000} + 10,000 \cdot \frac{10,000}{210,000} = 10,000\\
\end{array}
$$

In the general case where there are $N$ population centers in the region, we define the following population-based balancing factors in Equation (\ref{eq:population-balancing-factor}):
```{=tex}
\begin{equation}
\label{eq:population-balancing-factor}
F^{pM}_{i} = \frac{P_{i}^m}{\sum_{i=1}^N P_{i}^m}
\end{equation}
```

Balancing factor $F^p_{i}$ corresponds to the proportion of the population in origin $i$ relative to the population in the region. On the right hand side of the equation, the numerator $P_{i}^m$ is the population at origin $i$. <!-- The summation in the denominator is over $i=1,\cdots,N$, and adds up to the total population of the region. Notice that we incorporate an empirical parameter $\alpha$. The role of $\alpha$ is to modulate the effect of demand by population. When $\alpha <1$, opportunities are allocated more rapidly to smaller centers relative to larger ones; $\alpha>1$ achieves the opposite effect.-->

The discussion so far concerns only the mass effect (i.e., population size) of the gravity model. In addition, the potential for interaction is thought to decrease with increasing cost, so next we define similar balancing factors but based on the impedance.

## Proportional allocation by cost

Clearly, using only balancing factor $F^p_{i}$ to calculate spatial availability $V^p_i$ does not account for the cost of reaching employment centers. Consider instead a set of balancing factors $F^c_{ij}$ that account for the friction of distance for our example. Recall that the impedance function $f(c_{ij})$ equals $\exp(-\beta\cdot c_{ij})$ where  $\beta = 0.1$ and travel time $c_{ij}$ is either 15, 30 or 60 minutes. For instance, the impedance-based balancing factors $F^c_{ij}$ would be the following for employment center 1 (employment center 2 and 3 have their own balancing factor values for each origin $i$ as will be discussed later):
$$
\begin{array}{l}
F^c_{A1} = \frac{0.223130}{0.223130 + 0.049787 + 0.000045} = 0.8174398\\
F^c_{B1} = \frac{0.049787}{0.223130 + 0.049787 + 0.000045} = 0.1823954\\
F^c_{C1} = \frac{0.000045}{0.223130 + 0.049787 + 0.000045} = 0.0001648581\\
\end{array}
$$

Balancing factors $F^c_{ij}$ use the impedance function to proportionally allocate more jobs to closer population centers, that is, to those with populations _more willing to reach the jobs_. Indeed, the factors $F^c_{ij}$ can be thought of as the proportion of the population at $i$ willing to travel to destination $j$, conditional on the travel behavior as given by the impedance function. For instance, ${81.74398}\%$ of jobs from employment center 1 are allocated to population center $A$ based on impedance.

So as follows from our example, of the 100,000 jobs at employment center 1 the number of jobs allocated to population center $A$ is $100,000\times 0.8174398 = 81,743.98$ jobs; the number allocated to population center $B$ is $100,000\times 0.1823954 = 18,239.54$ jobs; and the number allocated to population center $C$ is $100,000\times 0.0001648581 = 16.48581$ jobs. We see once more that the total number of jobs at the employment center is preserved at 100,000. In this example, the proportional allocation mechanism assigns the largest share of jobs to population center $A$, which is the closest to employment center 1 and the smallest to the more distant population center $C$.

In the general case where there are $N$ population centers and $J$ employment centers in the region, we define the following impedance-based balancing factors:
```{=tex}
\begin{equation}
\label{eq:impedance-balancing-factor}
F^c_{ij} = \frac{f(c_{ij})}{\sum_{i=1}^N f(c_{ij})}
\end{equation}
```

The total number of jobs available to $i$ from $j$ according to impedance is defined as follows:
$$
V^c_{ij} = O_j\frac{F^c_{ij}}{\sum_{i=1}^N F^c_{ij}}
$$

The total number of jobs available to $i$ from all destinations is:
$$
V^c_{i} = \sum_{j=1}^J O_j\frac{F^c_{ij}}{\sum_{i=1}^N F^c_{ij}}
$$

Like the population-based allocation factors, $F^c_{i}$ summed over $i=1,\cdots,N$ always equals to 1 (i.e., $\sum_{i=1}^{N} F^c_{ij} = 1$). As before, the sum of all spatially available jobs equals $O$, the total number of opportunities in the region:
$$
\begin{array}{l}
\sum_{i=1}^N V^c_i =\sum_{i=1}^N\sum_{j=1}^JO_j\frac{F^c_{ij}}{\sum_{i=1}^N F^c_{ij}}\\
=\sum_{i=1}^N \frac{F^c_{ij}}{\sum_{i=1}^N F^c_{ij}}\cdot\sum_{j=1}^JO_j\\
=\sum_{j=1}^J O_j = O
\end{array}
$$

We are now ready to more formally define spatial availability with due consideration to both population and travel cost effects.

## Assembling mass and impedance effects

Population and the cost of travel are both part of the gravity modelling framework. Since the balancing factors defined in the preceding sections are proportions, they can be combined multiplicatively to obtain their joint effect. This multiplicative relationship can alternatively be understood as the joint probability of allocating opportunities and is captured by Equation (\ref{eq:balancing-factors}), where $F^p_{i}$ is the population-based balancing factor that grants a larger share of the existing opportunities to larger centers and $F^c_{ij}$ is the impedance-based balancing factor that grants a larger share of the existing opportunities to closer centers. This is in line with the tradition of gravity modeling.
```{=tex}
\begin{equation}
\label{eq:balancing-factors}
F^t_{ij} = \frac{F^p_{i} \cdot F^c_{ij}}{\sum_{i=1}^N F^p_{i} \cdot F^c_{ij}}
\end{equation}
```

\noindent with $F^p_{i}$ and $F^c_{ij}$ as defined in Equations (\ref{eq:population-balancing-factor}) and (\ref{eq:impedance-balancing-factor}) respectively. The combined balancing factor $F^t_{ij}$ is used to proportionally allocate jobs from $j$ to $i$. Hence, spatial availability is given by Equation (\ref{eq:spatial-availability}).

```{=tex}
\begin{equation}
\label{eq:spatial-availability}
V_{i} = \sum_{j=1}^J O_j\ F^t_{ij}
\end{equation}
```

The terms in Equation (\ref{eq:spatial-availability}):

-   $F^t_{ij}$ is a balancing factor as defined in Equation (\ref{eq:balancing-factors}).
-   $i$ is a set of origin locations in the region $i = 1,\cdots, N$.
-   $j$ is a set of destination locations in the region $j = 1,\cdots,J$.
-   $O_j$ is the number of opportunities at location $j$.
-   $V_{i}$ is the spatial availability at $i$.

Notice that, unlike $S_i$ in Hansen-type accessibility (Equation (\ref{eq:conventional-accessibility})), the population enters the calculation of $V_{i}$ through $F^p_i$. Returning to the example in Figure \ref{fig:Fig1}, Table \ref{tab:synthetic-example} also contains the information needed to calculate $V_i$, with $\beta$ set again to 0.1. Column __V_ij__ are the jobs available to each origin from each employment center. In this column $V_{A1}=$ 59,901 is the number of jobs available at $A$ from employment center 1. Column __V_i__ (i.e., $\sum_{j=1}^JV_{ij}$) gives the total number of jobs available to origin $i$. We can verify that the total number of jobs available is consistent with the total number of jobs in the region (with some small rounding error):
$$
\sum_{i=1}^N V_i = 66,833 + 133,203 + 9,963 \approx 210,000 
$$

Compared the calculated values of $V_i$ to column __S_i__ (Hansen-type accessibility) in Table \ref{tab:synthetic-example}. The spatial availability values are more intuitive. Recall that population centers $A$ and $B$ had identical Hansen-type accessibility to employment opportunities. According to $V_i$, population center $A$ has greater job availability due to: 1) its close proximity to employment center 1; combined with 2) less competition (i.e., a majority of the population have to travel longer distances to reach employment center 1). Job availability is lower for population center $B$ due to much higher competition (150,000 people can reach 100,000 jobs at equal cost). And center $C$ has almost as many jobs available as it has population.

As discussed above, Hansen-type accessibility is not designed to preserve the number of jobs in the region. Shen-type accessibility ends up preserving the number of jobs in the region but the definitions of variables are internally obscured; the only way it preserves the number of jobs is if the effect of the impedance function is ignored when expanding the values of jobs per capita to obtain the total number of opportunities. The proportional allocation procedure described above, in contrast, consistently returns a number of jobs available that matches the total number of jobs in the region.

Since the jobs spatially available are consistent with the jobs in the region, it is possible to define a measure of spatial availability per capita as presented in Equation (\ref{eq:SA-per-capita}):
```{=tex}
\begin{equation}
\label{eq:SA-per-capita}
v_i = \frac{V_i}{P_i}
\end{equation}
```

And, since the jobs are preserved, it is possible to use the regional jobs per capita ($\frac{\sum_{j=1}^J O_j}{\sum_{i=1}^N P_i}$) as a benchmark to compare the spatial availability of jobs per capita at each origin.

In the example, since the population is equal to the number of jobs, the regional value of jobs per capita is $1.0$. To complete the illustrative example, the spatial availability of jobs per capita by origin is:
```{r SA-6, eval=TRUE, include=FALSE}
V_i$V_i[1]/50000
V_i$V_i[2]/150000
V_i$V_i[3]/10000
```

```{=tex}
\begin{equation}
\label{eq:SA-per-capita-2populations}
\begin{array}{l}
v_{1} = \frac{V_1}{P_1} =  \frac{66,833.47}{50,000} = 1.337\\
v_{2} =  \frac{V_{2}}{P_2} =  \frac{133,203.4}{150,000} = 0.888\\
v_{3} =  \frac{V_{3}}{P_3} =  \frac{9,963.171}{10,000} = 0.996\\
\end{array}
\end{equation}
```

We can see that population center $A$ has fewer jobs per capita than the regional benchmark, center $B$ has more, and center $C$ is at parity. Remarkably, the spatial availability per capita matches the values of $a_i$ in Table \ref{tab:synthetic-example}. Appendix A has a proof of the mathematical equivalence between the two measures. It is interesting to notice how Weibull (1976) @weibull_axiomatic_1976 and Shen (1998) @shen1998, as well as this paper, all reach identical expressions starting from different assumptions; this effect is known as _equifinality_ (see Ortúzar & Willumsen (2011) @ortuzar_2011_modelling, p. 333 and Williams (1981) @williams_hall_1981). This result means that Shen-type accessibility and 2SFCA can be re-conceptualized as singly-constrained accessibility measures along with the proposed spatial availability measure.

# The case of Madrid's low emission zones: multimodal accessibility landscapes

<!-- a summary about low emission zones, they are a geographic accessibility restriction. they change the accessibility landscape-->

<!-- -->

## Data

Two scenarios:
- 2018: the small low emission zone
- Applying an optimistic restriction: what if we increased the restrictions?

## Multi-modal impedance functions

## Results of

#

### 2023-05-18 Simplified-spatial-availability-example.Rmd
Added 'multi-mode-acc-methods', a manuscript for multi-model accessibility draft.

@@
 
-Calculate spatial availabilty:
+Calculate spatial availability:
 ```{r}
@@
 
-<!-- AS finished here May 31, am stuck on how to shift car trips to transit (in a way that's flexible and will allow us to easily toggle the modal shift scenarios (i.e., instead of 100% transit, maybe 80%/10/10 transit/bike/walk)) -->
+<!-- AS finished here May 31, am stuck on how to shift car trips to transit (in a way that's flexible and will allow us to easily toggle the modal shift scenarios (i.e., instead of 100% transit, maybe 80%/10/10 transit/bike/walk))
+
+- 1) add akk trips (for all modes) to the dataset as 0
+- 2) allocation factors for trips going INTO madrid LEZ, a factor of 1 (car trip) proportional to other modes by distance-->
 

### 2023-05-30 Draft-manuscript-tobemoved.Rmd
draft of intro and illustrative example - for AP review.

@@
 ```{r data-import-filter}
-OD_v <- read_excel("data-raw/EDM2018VIAJES.xlsx")
+OD_v <- read_excel("data-prep/data-raw/EDM2018VIAJES.xlsx")
 
@@
 ```{r}
-load("data/zn.RData")
-
+load("data-prep/data/zn.RData")
+zn %>%
+  # Join workers
+  left_join(workers,
+            by = "ZT1259")
 zn <- zn %>%

### 2023-05-30 Multi-mode-acc-methods.qmd
draft of intro and illustrative example - for AP review.

@@
 ---
+```{r load-packages, include=FALSE, cache=FALSE}
+# library(fitdistrplus) # Help to Fit of a Parametric Distribution to Non-Censored or Censored Data
+# library(gridExtra) # Miscellaneous Functions for "Grid" Graphics 
+# library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax
+
+# library(shadowtext) # Shadow Text Grob and Layer
+# library(skimr) # Compact and Flexible Summaries of Data
+# library(spdep) # Spatial Dependence: Weighting Schemes, Statistics
+# library(tidyr) # Tidy Messy Data
+# options(scipen = 999)
+
+library(patchwork) # The Composer of Plots
+library(sf) # Simple Features for R
+library(dplyr) # A Grammar of Data Manipulation
+library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
+library(ggforce) # Accelerating 'ggplot2'
+library(geomtextpath) # Curved Text in 'ggplot2'
+library(ggpmisc) # Miscellaneous Extensions to 'ggplot2'
+library(ggrepel) # Automatically Position Non-Overlapping Text Labels with 'ggplot2'
+library(ggspatial) # Spatial Data Framework for ggplot2
+library(huxtable) # Easily Create and Style Tables for LaTeX, HTML and Other Formats
+library(scales) # Scale Functions for Visualization
+```
+
+```{r sp_avail-function,include=FALSE}
+#defining the spatial availability function
+sp_avail <- function(x, o_id, d_id, pop, opp, r, f, alpha = 1){
+
+  o_id <- rlang::enquo(o_id)
+  d_id <- rlang::enquo(d_id)
+  pop <- rlang::enquo(pop)
+  opp <- rlang::enquo(opp)
+  r <- rlang::enquo(r)
+  f <- rlang::enquo(f)
+
+  # Sum of population in system
+  sum_pop <- x %>%
+    dplyr::distinct(!!o_id,
+                    .keep_all = TRUE) %>%
+    dplyr::mutate(sum_pop = !!r*(!!pop)^alpha) %>%
+    dplyr::pull(sum_pop) %>%
+    sum()
+
+  # Balancing factor size
+  f_p <- dplyr::pull(x, !!r) * dplyr::pull(x, !!pop)^alpha / sum_pop
+
+  # Sum of impedance
+  sum_impedance <- x %>%
+    dplyr::group_by(!!d_id) %>%
+    dplyr::summarize(sum_impedance = sum(!!f))
+
+  # Join sum of impedance to table
+  x <- x %>%
+    dplyr::left_join(sum_impedance,
+                     by = rlang::as_name(d_id))
+
+  # Balancing factor impedance
+  f_c <- dplyr::pull(x, !!f) / x$sum_impedance
+
+  # Add balancing factors to table
+  x$f_c <- f_c
+  x$f_p <- f_p
+
+  # Sum of balancing factors for mass and impedance
+  sum_pa <- x %>%
+    dplyr::group_by(!!d_id) %>%
+    dplyr::summarize(sum_pa= sum(f_p * f_c))
+
+  x <- x %>%
+    dplyr::left_join(sum_pa,
+                     by = rlang::as_name(d_id))
+
+  # Joint balancing factor
+  f_t <- (f_p * f_c) / dplyr::pull(x, sum_pa)
+
+  dplyr::pull(x, !!opp) * f_t
+}
+
+#detailed
+sp_avail_detailed <- function(x, o_id, d_id, pop, opp, r, f, alpha = 1){
+
+  o_id <- rlang::enquo(o_id)
+  d_id <- rlang::enquo(d_id)
+  pop <- rlang::enquo(pop)
+  opp <- rlang::enquo(opp)
+  r <- rlang::enquo(r)
+  f <- rlang::enquo(f)
+
+  sum_pop <- x %>%
+    dplyr::distinct(!!o_id,
+                    .keep_all = TRUE) %>%
+    dplyr::mutate(sum_pop = !!r*(!!pop)^alpha) %>%
+    dplyr::pull(sum_pop) %>%
+    sum()
+
+  f_p <- dplyr::pull(x, !!r) * dplyr::pull(x, !!pop)^alpha / sum_pop
+
+  sum_impedance <- x %>%
+    dplyr::group_by(!!d_id) %>%
+    dplyr::summarize(sum_impedance = sum(!!f))
+
+  x <- x %>%
+    dplyr::left_join(sum_impedance,
+                     by = rlang::as_name(d_id))
+
+  f_c <- dplyr::pull(x, !!f) / x$sum_impedance
+
+  x$f_c <- f_c
+  x$f_p <- f_p
+
+  sum_pa <- x %>%
+    dplyr::group_by(!!d_id) %>%
+    dplyr::summarize(sum_pa= sum(f_p * f_c))
+
+  x <- x %>%
+    dplyr::left_join(sum_pa,
+                     by = rlang::as_name(d_id))
+  x$f_t <- (f_p * f_c) / dplyr::pull(x, sum_pa)
+
+  x %>%
+    dplyr::mutate(V_ij = !!opp * f_t)
+}
+```
+
+```{r synthetic-data, include=FALSE}
+od_tt <- data.frame(i = c("A", "A", "A", "B", "B", "B", "C", "C", "C"), # Three origins
+                    j = c("1", "2", "3", "1", "2", "3", "1", "2", "3"), # Three destinations
+                    tt_z = c(15, 30, 100, 30, 15, 100, 100, 100, 15), # Travel time, z-mode (original)
+                    tt_x = c(10, 25, 80, 25, 10, 80, 80, 80, 10), # Travel time, x-mode (the faster mode - like car)
+                    pop = c(50000, 50000, 50000, 150000, 150000, 150000, 10000, 10000, 10000), # Population
+                    opp = c(100000, 100000, 10000, 100000, 100000, 10000, 100000, 100000, 10000)) # Jobs
+
+#write the dataframe long.
+od_tt_m <- data.frame(i = c("A", "A", "A", "A", "A", "A",
+                            "B", "B", "B", "B", "B", "B",
+                            "C", "C", "C", "C", "C", "C"), # Three origins
+                    j = c("1", "2", "3", "1", "2", "3", 
+                          "1", "2", "3", "1", "2", "3", 
+                          "1", "2", "3", "1", "2", "3"), # Three destinations
+                    m = c("z", "z", "z", "x", "x", "x",
+                          "z", "z", "z", "x", "x", "x",
+                          "z", "z", "z", "x", "x", "x"),
+                    tt = c(15, 30, 100, 10, 25, 80, #travel time, z-mode (original) AND THEN x-mode (the faster mode)
+                           30, 15, 100, 25,  10,  80,
+                           100, 100, 15, 80, 80, 10), 
+                    pop = c(10000, 10000, 10000, 40000, 40000, 40000,
+                            50000, 50000,50000, 100000, 100000, 100000,
+                            3000, 3000, 3000, 7000,7000, 7000),
+                    opp = c(100000, 100000, 10000, 100000, 100000, 10000,
+                            100000, 100000, 10000, 100000, 100000, 10000,
+                            100000, 100000, 10000, 100000, 100000, 10000)) 
+```
+
+```{r data-figure-with-toy-example, include=FALSE}
+od <- data.frame(id = c("A", "B", "C", "1", "2", "3"),
+                 type = c("Population", "Population", "Population", "Jobs", "Jobs", "Jobs"),
+                 size = c(50000, 150000, 10000, 100000, 100000, 10000),
+                 x = c(2.5, 2.5, 6.5, 0.5, 0.5, 6.5),
+                 y = c(8.0, 2.5, 4.5, 7.5, 2.5, 2.5))
+
+centers <- data.frame(id = c("Urban center", "Suburb", "Satellite town"),
+                 radius = rep(1.3, 3),
+                 x = c(1.5, 1.5, 6.5),
+                 y = c(2.5, 7.5, 3.5))
+
+# od_lines coded as segments
+od_lines <- data.frame(x = c(2.5, 2.5, 2.5, 2.5, 2.5, 2.5, 6.5, 6.5, 6.5),
+                       y = c(8.0, 8.0, 8.0, 2.5, 2.5, 2.5, 4.5, 4.5, 4.5),
+                       xend = c(0.5, 0.5, 6.5, 0.5, 0.5, 6.5, 0.5, 0.5, 6.5),
+                       yend = c(7.5, 2.5, 2.5, 7.5, 2.5, 2.5, 7.5, 2.5, 2.5),
+                       group = c("A", "A", "A", "B", "B", "B", "C", "C", "C"),
+                       time = c("15 min (z)\n10 min (x)", "30 min (z)\n25 min (x)", "100 min (z)\n80 min (x)",
+                                "30 min (z)\n25 min (x)", "15 min (z)\n10 min (x)", "100 min (z)\n80 min (x)",
+                                "100 min (z)\n80 min (x)", "100 min (z)\n80 min (x)", "15 min (z)\n10 min (x)"))
+
+od_table <- data.frame(Center = c("A", "B", "C", "1", "2", "3"),
+  Size = c("50,000 pop", "150,000 pop", "10,000 pop", "100,000 jobs", "100,000 jobs", "10,000 jobs"))
+```
+
+```{r, creating-toy-example-plot, eval=FALSE}
+toy_example_plot <- ggplot() + 
+  # Plot centers
+  geom_circle(data = centers,
+             aes(x0 = x, 
+                 y0 = y,
+                 r = radius),
+             color = "grey40") +
+  annotate(geom = "label", 
+           x=c(0.5, 0.6, 7),
+           y=c(9, 1, 5.2), 
+           label = c("Suburban", "Urban", "Satellite"), 
+           size=4, 
+           label.size=NA) +
+  # Connect origins and destinations with curves
+  geom_curve(data = od_lines,
+               aes(x = x,
+                   y = y,
+                   xend = xend,
+                   yend = yend,
+                   linetype = group,
+                   group = group),
+             curvature = 0.20,
+             linewidth = 1,
+             color = "black") +
+  # Text on curves
+  geom_textcurve(data = od_lines,
+               aes(x = x,
+                   y = y,
+                   xend = xend,
+                   yend = yend,
+                   linetype = group,
+                   group = group,
+                   label = time),
+             curvature = 0.20,
+             color = "black") +
+  # Plot origins and destinations
+  geom_point(data = od,
+            aes(size = size,
+                x = x, 
+                y = y, 
+                color = type,
+                shape = type)) +
+  scale_size(range = c(4, 10)) +
+  # Label origins and destinations
+  geom_text(data = od, 
+            aes(x, 
+                y, 
+                label=id), 
+            size=3) +
+  # Create table
+  annotate(geom = "table",
+           x = 8, 
+           y = 9, 
+           label = list(od_table),
+           fill = "white",
+           size=3) +
+  # Theme the plot
+  coord_fixed() + 
+  theme_void() + guides(shape = "none", size = "none", color = "none", linetype ="none")
+
+toy_example_plot
+ggsave("images/Fig1.png",dpi=600)
+```
 
@@
 <!-- examples of constrained (competitive) measures, the introduction of spatial availability -->
-
 The third type of accessibility measure, competitive measures, discount the number of opportunities allocated based on the population that is demanding those opportunities in addition to discounting by travel impedance (like unconstrained accessibility measures).
@@
 The output is the number of opportunities available in the region, per location, such that if all the values for the locations in the region were summed they would equal the total opportunities in the region.
+
+ANother example of proportional allocation, BFCA
 
@@
 This is the aim of this paper.
-An example of a case where conventional accessibility measures (unconstrained and competitive) fall short is in considering multi-modal accessibility.
+We plan to expland to this measure to multi-modal systems.
+opportunities are avialble to users of different modes.
+The cost of reaching them is not uniform.
+
+<!-- An example of a case where conventional accessibility measures (unconstrained and competitive) fall short is in considering multi-modal accessibility.
 Take for instance, measuring the accessibility to opportunities for a region for car and transit modes.
@@
 the opportunities..
+-->
 
@@
 A clustering method to modify the cumulative accessibility measure @mohriClusteringMethodMeasuring2021
+
 <!-- summary of the paper-->
@@
 $$
-V_i = \sum_{j=1}^NO_jF^t_{ij}
+V_i = \sum_{j=1}^J O_jF^t_{ij}
 $$
@@
 
--   $F^t_{ij}$ is a balancing factor that depends on the population and cost of movement in the system.
+-   $V_i$ is the number of spatially available opportunities at $i$.
 -   $O_j$ is the number of opportunities at $j$.
--   $V_i$ is the number of spatially available opportunities from the perspective of $i$.
+-   $F^t_{ij}$ is a balancing factor that depends on the demand for opportunities and cost of movement in the system.
 
-The general form of spatial availability is also as a sum, and the fundamental difference between Hansen- and Shen-type accessibility is that opportunities are allocated proportionally.
-Balancing factor $F^t_{ij}$ consists of two components: a population-based balancing factor $F^p_{i}$ and an impedance-based balancing factor $F^c_{ij}$ that, respectively, allocate opportunities to $i$ in proportion to the size of the population of the different competing centers (the mass effect of the gravity model) and the cost of reaching opportunities (the impedance effect).
+The general form of spatial availability is a sum: this is similar to Hansen-type accessibility (popular _unconstrained_ measure) and Shen-type accessibility (popular _competitive_ and _unconstrained_ measure).
+What makes spatial availability a _constrained_ measure are the balancing factors that allocate opportunities proportionally.
+The balancing factor $F^t_{ij}$ consists of two components: a population-based balancing factor $F^p_{i}$ and an impedance-based balancing factor $F^c_{ij}$ that, respectively, allocate opportunities to $i$ in proportion to the size of the population of the different competing centers (the mass effect of the gravity model) and the cost of reaching opportunities (the impedance effect).
 
-We adapt the illustrative example used in @shen1998 assuming two modes and demonstrate spatial availability's within this multimodal application.
-<!-- TO DO: demonstrate an illustrate of the example-->
+```{r, synthetic-example-plot, fig.cap="\\label{fig:Fig1} Modified synthetic example from Shen (1998) with locations of employment centers (in orange), population centers (in blue), number of jobs and population, and travel times for two modes (x mode and z mode).", fig.show='hold', fig.align='center'}
+knitr::include_graphics("images/Fig1.png")
+```
+
+```{r access-calcs-table, include=FALSE}
+#table with all the accessibility calculations
+# Impedance parameter
+b <- 0.1
+
+# Calculate impedance function
+od_tt_m <- od_tt_m %>%
+  mutate(f = exp(-b * tt))
+```
+
+```{r calculating-beta-0.1-Vi-table-creation}
+od_tt_m <- od_tt_m %>%
+  mutate(catch = 1) %>%
+  sp_avail_detailed(o_id = i, #spatial availability
+                    d_id = j, 
+                    pop = pop,
+                    opp = opp,
+                    r = catch,
+                    f = f, 
+                    alpha = 1)
+
+V_i <- od_tt_m %>%
+  group_by(i) %>%
+  summarize(V_i = sum(V_ij))
+
+od_tt_m <- od_tt_m %>%
+  left_join(V_i, by = "i") %>%
+  transmute(i,
+            j,
+            m,
+            pop,
+            opp,
+            tt,
+            V_ij = ifelse(V_ij < 0.001, "< 0.001", round(V_ij, 3)),
+            V_i = ifelse(V_i < 0.001, "< 0.001", round(V_i, 3)),
+            f = ifelse(f < 0.001, "< 0.001", round(f, 3)),
+            f_p = ifelse(f_p < 0.001, "< 0.001", round(f_p, 3)),
+            f_c = ifelse(f_c < 0.001, "< 0.001", round(f_c, 3)),
+            f_t = ifelse(f_t < 0.001, "< 0.001", round(f_t, 3)))
+```
+
+```{r merging-data-for-table}
+colnames(od_tt_m) <- c("i", "j","Mode", "Pop.", "Opp.", "TT", "V", "V_i", 
+                       "f(TT)", "F^p", "F^c", "F^t")
+
+od_tt_m_hux <- od_tt_m %>%
+  as_hux()
+```
+
+```{r toy-example-table-all-values, cache=FALSE}
+table_toy  <- od_tt_m_hux %>%
+  set_background_color(everywhere, 7, "grey90") %>%
+  set_background_color(everywhere, 8, "lightyellow") %>%
+  set_bold(1, everywhere) %>%
+  set_bold(everywhere, 1) %>%
+  # set_bottom_border(c(1,2,5,10,15,20), everywhere) %>%
+  # set_right_border(everywhere, c(1,4,7,10), brdr(1, "solid", "grey")) %>%
+  # set_align(1:20, 1, "left") %>%
+  # set_align(1:20,  2:10, "center") %>%
+  # set_number_format(c(6,10,11:15,16:18), everywhere, 3) %>% #the smaller numbers, have 3 decimals
+  # set_number_format(c(7:9,19,20), everywhere, 1) %>% # the larger numbers, have 1 decimal
+  merge_cells(2:7, 1) %>% merge_cells(8:13, 1) %>% merge_cells(14:19, 1) %>%
+  merge_cells(14:15, 2) %>% merge_cells(16:17, 2) %>% merge_cells(18:19, 2) %>% 
+  merge_cells(2:4, 3) %>% merge_cells(5:7, 3) %>% merge_cells(8:10, 3) %>% 
+  merge_cells(11:13, 3) %>% merge_cells(14:16, 3) %>% merge_cells(17:19, 3) %>% 
+  merge_cells(2:4, 4) %>% merge_cells(5:7, 4) %>% merge_cells(8:10, 4) %>% 
+  merge_cells(11:13, 4) %>% merge_cells(14:16, 4) %>% merge_cells(17:19, 4) %>% 
+  merge_cells(2:7, 5) %>% merge_cells(8:13, 5) %>% merge_cells(14:19, 5) %>%
+  merge_cells(2:7, 8) %>% merge_cells(8:13, 8) %>% merge_cells(14:19, 8) %>%
+  set_caption("Summary description of the synthetic example: Spatial availability at each origin per mode $V_{im}$ (grey) and at each origin $V_i$ (yellow)") %>%
+  set_label("tab:synthetic-example") %>% 
+  set_latex_float("h!") %>% 
+  set_width(1.1)
+
+font_size(table_toy) <- 7
+
+table_toy[1:19,]
+```
+
+We adapt the illustrative example used in @shen1998 assuming two modes and demonstrate spatial availability's within this multi-modal application.
+See the Figure \ref{fig:Fig1} and the Table \ref{tab:synthetic-example} with calculated spatial availability values (yellow - per origin and mode, and grey per origin).
+We use a negative exponential impedance function with $\beta=0.1$ as also used in @shen1998: we assume this impedance function for both modes.
+$$
+f(c_{ij}) = \exp(-\beta\cdot c_{ij})
+$$
 
@@
 \begin{array}{l}
-F^p_A = \frac{50,000}{210,000} & \text{where } F^{px}_A = \frac{40,000}{210,000} \text{ & } F^{pz}_A = \frac{10,000}{210,000}\\ 
+F^{pm}_{i} = \frac{P_{i}^m}{\sum_{m=1}^M\sum_{i=1}^N P_{i}^m}\\
+
+F^{pm}_A = \frac{50,000}{210,000} & \text{where } F^{px}_A = \frac{40,000}{210,000} \text{ & } F^{pz}_A = \frac{10,000}{210,000}\\ 
 \\
-F^p_B = \frac{150,000}{210,000} & \text{where } F^{px}_B = \frac{100,000}{210,000} \text{ & } F^{pz}_B = \frac{50,000}{210,000}\\
+F^{pm}_B = \frac{150,000}{210,000} & \text{where } F^{px}_B = \frac{100,000}{210,000} \text{ & } F^{pz}_B = \frac{50,000}{210,000}\\
 \\
-F^p_C = \frac{10,000}{210,000} & \text{where } F^{px}_C = \frac{7,000}{210,000} \text{ & } F^{pz}_C = \frac{3,000}{210,000}\\
+F^{pm}_C = \frac{10,000}{210,000} & \text{where } F^{px}_C = \frac{7,000}{210,000} \text{ & } F^{pz}_C = \frac{3,000}{210,000}\\
 \end{array}
@@
 
-Competition between the subsets of the populations that use the jobs matters.
-After all, there is a finite number of jobs (210,000 in our example).
 Jobs are allocated proportionally from each employment center to each population center depending on their population sizes as per the balancing factors $F^p_i$.
-In this way, employment center 1 allocates $100,000\cdot \frac{50,000}{210,000}= 23,809.52$ jobs to $A$.
-This  $100,000\cdot \frac{150,000}{210,000}= 71,428.57$ jobs to $B$; and $100,000\cdot \frac{10,000}{210,000}= 7,142.857$ jobs to $C$.
+In this way, employment center 1 allocates $100,000\cdot \frac{50,000}{210,000}= 23,809.52$ jobs to $A$ (the suburb).
+Similarly, employment centers 2 and 3 allocate $100,000\cdot \frac{150,000}{210,000}= 71,428.57$ jobs to $B$ (the urban city) and $100,000\cdot \frac{10,000}{210,000}= 7,142.857$ jobs to $C$ (the satellite city).
 Notice how this mechanism ensures that the total number of jobs at employment center 1 is preserved at 100,000.
+The jobs at the employment centers are allocated proportional to the population size of the population center relative to all the population center sizes.
+Larger population centers are more competitive (i.e., they demand more) than smaller population centers, so they get allocated more job opportunities.
+
+However, how do the populations access these jobs?
+In this synthetic example, the population either uses the regular mode $z$ or uses the faster mode $x$.
+Their choice in mode is already assigned: the mode selected can be seen as a characteristic of the population, or a way to subset the total population.
+In this way, the introduction of this population-characteristic of mode selection adds another layer of competition to the mass effect.
+We demonstrated that employment center 1 allocates $23,809.52$ jobs to $A$ but 20% of these jobs are allocated to population using $z$ ($10,000$ of the $50,000$ population in $A$) and 80% is allocated to the population using the faster mode $x$ ($40,000$ of the $50,000$ population in $A$).
+Similarly, employment centers 2 and 3 allocate $71,428.57$ jobs to $B$ (the urban city) of which 66% are allocated to the population using $z$ and 33% to the population using $x$ mode and the $7,142.857$ jobs allocated to $C$ (the satellite city) is split between 30% of the population using mode $z$ and 70% of the population using mode $x$.
+
+Considering the subsets of the population that use the two different modes, the total number of jobs that are assigned to each population center and then to each sub-set of population, is still preserved.
+The total sum of all the jobs allocated in the region is still 210,000 in our example.
 
@@
 \text{Employment center 1 to population centers A, B, and C: }\\
-100,000 \cdot \frac{50,000}{210,000} + 100,000 \cdot \frac{150,000}{210,000} + 100,000 \cdot \frac{10,000}{210,000} = 100,000\\
+100,000 \cdot \frac{40,000}{210,000} + 100,000 \cdot \frac{10,000}{210,000} + 100,000 \cdot \frac{100,000}{210,000} + 100,000 \cdot \frac{50,000}{210,000} + 100,000 \cdot \frac{7,000}{210,000} + 100,000 \cdot \frac{3,000}{210,000} = 100,000\\
 \\
 \text{Employment center 2 to population centers A, B, and C: }\\
-100,000 \cdot \frac{50,000}{210,000} + 100,000 \cdot \frac{150,000}{210,000} + 100,000 \cdot \frac{10,000}{210,000} = 100,000\\
+100,000 \cdot \frac{40,000}{210,000} + 100,000 \cdot \frac{10,000}{210,000} + 100,000 \cdot \frac{100,000}{210,000} + 100,000 \cdot \frac{50,000}{210,000} + 100,000 \cdot \frac{7,000}{210,000} + 100,000 \cdot \frac{3,000}{210,000} = 100,000\\
 \\
 \text{Employment center 3 to population centers A, B, and C: }\\
-10,000 \cdot \frac{50,000}{210,000} + 10,000 \cdot \frac{150,000}{210,000} + 10,000 \cdot \frac{10,000}{210,000} = 10,000\\
+10,000 \cdot \frac{40,000}{210,000} + 10,000 \cdot \frac{10,000}{210,000} + 10,000 \cdot \frac{100,000}{210,000} + 10,000 \cdot \frac{50,000}{210,000} + 10,000 \cdot \frac{7,000}{210,000} + 10,000 \cdot \frac{3,000}{210,000} = 10,000\\
 \end{array}
@@
 \label{eq:population-balancing-factor}
-F^{pM}_{i} = \frac{P_{i}^m}{\sum_{i=1}^N P_{i}^m}
+F^{pm}_{i} = \frac{P_{i}^m}{\sum_{m=1}^M\sum_{i=1}^N P_{i}^m}
 \end{equation}
@@
 Balancing factor $F^p_{i}$ corresponds to the proportion of the population in origin $i$ relative to the population in the region.
-On the right hand side of the equation, the numerator $P_{i}^m$ is the population at origin $i$.
-<!-- The summation in the denominator is over $i=1,\cdots,N$, and adds up to the total population of the region.
-Notice that we incorporate an empirical parameter $\alpha$.
-The role of $\alpha$ is to modulate the effect of demand by population.
-When $\alpha <1$, opportunities are allocated more rapidly to smaller centers relative to larger ones; $\alpha>1$ achieves the opposite effect.-->
+On the right hand side of the equation, the numerator $P_{i}^m$ is the population at origin $i$ that use each type of mode $m$ for the full set of modes $M$.
 
 The discussion so far concerns only the mass effect (i.e., population size) of the gravity model.
-In addition, the potential for interaction is thought to decrease with increasing cost, so next we define similar balancing factors but based on the impedance.
+In addition, the potential for interaction with opportunities is thought to increase with decreasing cost, so next we define similar balancing factors but based on the impedance.
 
@@
 
-Clearly, using only balancing factor $F^p_{i}$ to calculate spatial availability $V^p_i$ does not account for the cost of reaching employment centers.
-Consider instead a set of balancing factors $F^c_{ij}$ that account for the friction of distance for our example.
-Recall that the impedance function $f(c_{ij})$ equals $\exp(-\beta\cdot c_{ij})$ where  $\beta = 0.1$ and travel time $c_{ij}$ is either 15, 30 or 60 minutes.
-For instance, the impedance-based balancing factors $F^c_{ij}$ would be the following for employment center 1 (employment center 2 and 3 have their own balancing factor values for each origin $i$ as will be discussed later):
+Clearly, exclusively using the population balancing factor $F^{pm}_{i}$ to calculate spatial availability does not account for the cost of reaching employment centers.
+Consider instead a set of balancing factors $F^{cm}_{ij}$ that accounts for the friction of distance.
+For this example, we define an impedance function $f(c_{ij}^m)$ that equals $\exp(-\beta\cdot c_{ij}^m)$ where $\beta = 0.1$ and the travel time $c_{ij}^m$ is either 15, 30 or 100 minutes for $m=z$ or 10, 20, or 80 minutes for $m=x$.
+
+To illustrate, the impedance-based balancing factors $F^{cm}_{ij}$ would be the following for employment center 1 (employment center 2 and 3 have their own balancing factor values for each origin $i$ as is discussed later):
 $$
 \begin{array}{l}
-F^c_{A1} = \frac{0.223130}{0.223130 + 0.049787 + 0.000045} = 0.8174398\\
-F^c_{B1} = \frac{0.049787}{0.223130 + 0.049787 + 0.000045} = 0.1823954\\
-F^c_{C1} = \frac{0.000045}{0.223130 + 0.049787 + 0.000045} = 0.0001648581\\
+F_{ij}^{cm} = \frac{f(c_{ij}^m)}{\sum_{m=1}^M \sum_{i=1}^N f(c_{ij}^m)}\\
+
+F^{cx}_{A1} = \frac{0.223130}{0.223130 + 0.049787 + 0.000045 + 0.3678794 + 0.1353353 + 0.0003354626} = 0.287349 \\
+F^{cx}_{B1} = \frac{0.049787}{0.223130 + 0.049787 + 0.000045 + 0.3678794 + 0.1353353 + 0.0003354626} = 0.06411619\\
+F^{cx}_{C1} = \frac{0.000045}{0.223130 + 0.049787 + 0.000045 + 0.3678794 + 0.1353353 + 0.0003354626} = 0.00005795144\\
+
+F^{cz}_{A1} = \frac{0.3678794}{0.223130 + 0.049787 + 0.000045 + 0.3678794 + 0.1353353 + 0.0003354626} = 0.4737587\\
+F^{cz}_{B1} = \frac{0.1353353}{0.223130 + 0.049787 + 0.000045 + 0.3678794 + 0.1353353 + 0.0003354626} = 0.1742861\\
+F^{cz}_{C1} = \frac{0.0003354626}{0.223130 + 0.049787 + 0.000045 + 0.3678794 + 0.1353353 + 0.0003354626} = 0.000432012\\
 \end{array}
@@
 
-Balancing factors $F^c_{ij}$ use the impedance function to proportionally allocate more jobs to closer population centers, that is, to those with populations _more willing to reach the jobs_.
-Indeed, the factors $F^c_{ij}$ can be thought of as the proportion of the population at $i$ willing to travel to destination $j$, conditional on the travel behavior as given by the impedance function.
-For instance, ${81.74398}\%$ of jobs from employment center 1 are allocated to population center $A$ based on impedance.
+Balancing factors $F^{cm}_{ij}$ use the impedance function to proportionally allocate more jobs to closer population centers, that is, to those with populations _more willing to reach the jobs_.
+Indeed, the factors $F^{cm}_{ij}$ can be thought of as the proportion of the population at $i$ willing to travel to destination $j$, conditional on the travel behavior as given by the impedance function.
+For instance, ${28.7349}\%$ of jobs from employment center 1 are allocated to population $A$ using the slower mode $x$ and ${47.37587}\%$ of the jobs are allocated to the population using the faster mode $z$ based on impedance.
+Population center $A$ is assigned $47.37587\%+{28.7349}\% = 76.11077\%$ of the jobs in the region based on the impedance, but it can be seen that the population using the faster mode is assigned a larger proportion of jobs.
+The population that uses the faster mode $z$ is more competitive than the population that uses the slower mode $x$.
 
-So as follows from our example, of the 100,000 jobs at employment center 1 the number of jobs allocated to population center $A$ is $100,000\times 0.8174398 = 81,743.98$ jobs; the number allocated to population center $B$ is $100,000\times 0.1823954 = 18,239.54$ jobs; and the number allocated to population center $C$ is $100,000\times 0.0001648581 = 16.48581$ jobs.
-We see once more that the total number of jobs at the employment center is preserved at 100,000.
-In this example, the proportional allocation mechanism assigns the largest share of jobs to population center $A$, which is the closest to employment center 1 and the smallest to the more distant population center $C$.
+So as follows from our example, of the 100,000 jobs at employment center 1 the number of jobs allocated to population center $A$ is $100,000\times 0.7611077 = 76,110.77$ jobs - $28,734.9$ to population using mode $x$ and $47,375.87$ to population using mode $z$.
+The number allocated to population center $B$ is $100,000\times 0.2384023= 23,840.23$ jobs - $6,411.619$ to population using mode $x$ and $17,428.61$ to population using mode $z$; and the number allocated to population center $C$ is $100,000\times 0.0004899634 = 48.99634$ jobs - $5.795144$ to population using mode $x$ and $43.2012$ to population using mode $z$.
+We see once more that the total number of jobs at the employment center is preserved at $100,000$.
+The proportional allocation mechanism assigns the largest share of jobs to population center $A$, which is the closest to employment center 1 and the smallest to the more distant population center $C$.
+Additionally, the sum of the population using the faster mode $z$ across all three population centers gets assigned the majority of jobs from employment center $1$ ($64.81028\%$).
+<!-- NOTE: what's interesting is that when considering multiple modes employment center (1) allocates less jobs to urban (A), more jobs to suburban (B) and more jobs to remote (c).
+Why is this interesting?
+The suburban and sat.
+centers have f(c_ij) from 20, 30, 80, 100 min -- higher average values than just considering 30 and 100 min (as in the single mode example) hence relatively favouring the impedance of the inter-city commute than the intercity commute (10 and 15 min).
+The addition of a faster mode at the inter-city commutes reweighs the modal options and makes this form of travel more competitive, overall.
+-->
 
-In the general case where there are $N$ population centers and $J$ employment centers in the region, we define the following impedance-based balancing factors:
+In the general case where there are $N$ population centers in the region, we define the following impedance-based balancing factors for each $i$ $j$ flow for each $m$:
 ```{=tex}
@@
 \label{eq:impedance-balancing-factor}
-F^c_{ij} = \frac{f(c_{ij})}{\sum_{i=1}^N f(c_{ij})}
+F_{ij}^{cm} = \frac{f(c_{ij}^m)}{\sum_{m=1}^M \sum_{i=1}^N f(c_{ij}^m)}\\
 \end{equation}
 ```
-
-The total number of jobs available to $i$ from $j$ according to impedance is defined as follows:
-$$
-V^c_{ij} = O_j\frac{F^c_{ij}}{\sum_{i=1}^N F^c_{ij}}
-$$
-
-The total number of jobs available to $i$ from all destinations is:
-$$
-V^c_{i} = \sum_{j=1}^J O_j\frac{F^c_{ij}}{\sum_{i=1}^N F^c_{ij}}
-$$
-
-Like the population-based allocation factors, $F^c_{i}$ summed over $i=1,\cdots,N$ always equals to 1 (i.e., $\sum_{i=1}^{N} F^c_{ij} = 1$).
-As before, the sum of all spatially available jobs equals $O$, the total number of opportunities in the region:
-$$
-\begin{array}{l}
-\sum_{i=1}^N V^c_i =\sum_{i=1}^N\sum_{j=1}^JO_j\frac{F^c_{ij}}{\sum_{i=1}^N F^c_{ij}}\\
-=\sum_{i=1}^N \frac{F^c_{ij}}{\sum_{i=1}^N F^c_{ij}}\cdot\sum_{j=1}^JO_j\\
-=\sum_{j=1}^J O_j = O
-\end{array}
-$$
-
-We are now ready to more formally define spatial availability with due consideration to both population and travel cost effects.
 
@@
 Since the balancing factors defined in the preceding sections are proportions, they can be combined multiplicatively to obtain their joint effect.
-This multiplicative relationship can alternatively be understood as the joint probability of allocating opportunities and is captured by Equation (\ref{eq:balancing-factors}), where $F^p_{i}$ is the population-based balancing factor that grants a larger share of the existing opportunities to larger centers and $F^c_{ij}$ is the impedance-based balancing factor that grants a larger share of the existing opportunities to closer centers.
-This is in line with the tradition of gravity modeling.
+This multiplicative relationship can alternatively be understood as the joint probability of allocating opportunities and is captured by Equation (\ref{eq:balancing-factors}), where $F^{pm}_{i}$ is the population-based balancing factor that grants a larger share of the existing opportunities to the subsets of populations using mode $m$ located at population centers $i$.
+$F^{cm}_{ij}$ is the impedance-based balancing factor that grants a larger share of the existing opportunities to subsets of populations using $m$ mode in population centers $i$ with the least costly travel impedance from $i$ to $j$.
+Considering both the mass effect from population and weighting travel flows based on impedance is in line with the tradition of gravity modeling.
+Equation (\ref{eq:population-balancing-factor}) defines this combined balancing factor.
+As we demonstrate in this section, within spatial availability, we can consider modes $m$ as a subset of the population in the region.
+In doing so, makes population centers $i$ with larger subsets of populations more attractive (have a larger mass effect) and makes the subset of populations using least costly modes $m$ more attractive with respect to proportional allocation of opportunities in the region.
+
 ```{=tex}
@@
 \label{eq:balancing-factors}
-F^t_{ij} = \frac{F^p_{i} \cdot F^c_{ij}}{\sum_{i=1}^N F^p_{i} \cdot F^c_{ij}}
+F^{tm}_{ij} = \frac{F^{pm}_{i} \cdot F^{cm}_{ij}}{\sum_{m=1}^M \sum_{i=1}^N F^{pm}_{i} \cdot F^{cm}_{ij}}
 \end{equation}
@@
 
-\noindent with $F^p_{i}$ and $F^c_{ij}$ as defined in Equations (\ref{eq:population-balancing-factor}) and (\ref{eq:impedance-balancing-factor}) respectively.
-The combined balancing factor $F^t_{ij}$ is used to proportionally allocate jobs from $j$ to $i$.
-Hence, spatial availability is given by Equation (\ref{eq:spatial-availability}).
+\noindent with $F^{pm}_{i}$ and $F^{cm}_{ij}$ as defined in Equations (\ref{eq:population-balancing-factor}) and (\ref{eq:impedance-balancing-factor}) respectively.
+The combined balancing factor $F^{tm}_{ij}$ is used to proportionally allocate jobs from $j$ to $i$ based mass-effect and impedance-effect of populations using multiple modes $m$.
+Hence, spatial availability with consideration for multiple modes $m$ is given by Equation (\ref{eq:spatial-availability}).
 
@@
 \label{eq:spatial-availability}
-V_{i} = \sum_{j=1}^J O_j\ F^t_{ij}
+V^m_{i} = \sum_{j=1}^J O_j\ F^{tm}_{ij}\\
+
+V_{i} = \sum_{m=1}^M V^m_{i}\\
 \end{equation}
@@
 
--   $F^t_{ij}$ is a balancing factor as defined in Equation (\ref{eq:balancing-factors}).
+-   $F^{tm}_{ij}$ is a balancing factor as defined in Equation (\ref{eq:balancing-factors}).
+-   $m$ is a set of modes used by populations in the region $m = 1, \cdots, M$.
 -   $i$ is a set of origin locations in the region $i = 1,\cdots, N$.
@@
 -   $O_j$ is the number of opportunities at location $j$.
+-   $V^m_{i}$ is the spatial availability at $i$ for mode $m$.
 -   $V_{i}$ is the spatial availability at $i$.
 
-Notice that, unlike $S_i$ in Hansen-type accessibility (Equation (\ref{eq:conventional-accessibility})), the population enters the calculation of $V_{i}$ through $F^p_i$.
+<!-- NOTE: do we want to display this by mode,origin or origin or both? -->
+
+We calculate $V^m_{i}$ for each population center $i$ and mode $m$ and then present the spatial availability $V_{i}$ per population center in Table \ref{tab:synthetic-example}.
+
+Notice, the population enters the calculation of $V^m_{i}$ through $F^{pm}_i$.
 Returning to the example in Figure \ref{fig:Fig1}, Table \ref{tab:synthetic-example} also contains the information needed to calculate $V_i$, with $\beta$ set again to 0.1.
 Column __V_ij__ are the jobs available to each origin from each employment center.
-In this column $V_{A1}=$ 59,901 is the number of jobs available at $A$ from employment center 1.
-Column __V_i__ (i.e., $\sum_{j=1}^JV_{ij}$) gives the total number of jobs available to origin $i$.
+In this column $V_{A1}=$ $68,648.305$ is the number of jobs available at $A$ from employment center 1.
+Column __V_i__ (i.e., $\sum_{m=1}^M \sum_{j=1}^J V_{ijm}$) gives the total number of jobs available to origin $i$ from each subset of the population using modes $m$.
 We can verify that the total number of jobs available is consistent with the total number of jobs in the region (with some small rounding error):
 $$
-\sum_{i=1}^N V_i = 66,833 + 133,203 + 9,963 \approx 210,000 
+\sum_{i=A} V_{ijm} = 8070.736 + 962.474 + 1.378 + 53225.576 + 6347.407 + 40.733 = 68648.305\\
+\sum_{i=B} V_{ijm} =9004.123 + 21567.552 + 6.891 + 29690.578 + 71117.764 + 101.833 = 131488.741\\
+\sum_{i=C} V_{ijm} =0.493 + 0.263 + 2032.006 + 8.494 + 4.540 + 7817.159 = 9862.955\\
+
+\sum_{i=1}^N V_{ijm} = (68648.305) + (131488.741) + (9862.955) = 210000 
 $$
-
-Compared the calculated values of $V_i$ to column __S_i__ (Hansen-type accessibility) in Table \ref{tab:synthetic-example}.
-The spatial availability values are more intuitive.
-Recall that population centers $A$ and $B$ had identical Hansen-type accessibility to employment opportunities.
-According to $V_i$, population center $A$ has greater job availability due to: 1) its close proximity to employment center 1; combined with 2) less competition (i.e., a majority of the population have to travel longer distances to reach employment center 1).
+Table \ref{tab:synthetic-example} and the equations above display the intuition in the spatial availability calculation for the synthetic example.
+Population center $A$ has greater job availability due to: 1) its close proximity to employment center 1; combined with 2) less competition (i.e., a majority of the population have to travel longer distances to reach employment center 1).
 Job availability is lower for population center $B$ due to much higher competition (150,000 people can reach 100,000 jobs at equal cost).
@@
 
-As discussed above, Hansen-type accessibility is not designed to preserve the number of jobs in the region.
-Shen-type accessibility ends up preserving the number of jobs in the region but the definitions of variables are internally obscured; the only way it preserves the number of jobs is if the effect of the impedance function is ignored when expanding the values of jobs per capita to obtain the total number of opportunities.
-The proportional allocation procedure described above, in contrast, consistently returns a number of jobs available that matches the total number of jobs in the region.
+Further, we can also examine the competition between modes.
+The potential population using the slower mode $z$ is `r od_tt_m %>% group_by(i,Mode) %>% summarize(pop = first(Pop.), .groups = "drop") %>% filter(Mode == "z") %>% dplyr::select(pop) %>% sum()` (`r ((od_tt_m %>% group_by(i,Mode) %>% summarize(pop = first(Pop.), .groups = "drop") %>% filter(Mode == "z") %>% dplyr::select(pop) %>% sum()) / (od_tt_m %>% group_by(i,Mode) %>% summarize(pop = first(Pop.), .groups = "drop") %>% dplyr::select(pop) %>% sum())) %>% percent()` of the potential population in the region).
+Despite this, this slower mode $z$ accounts for only `r ((od_tt_m %>% group_by(i,Mode) %>% summarize(V = sum(V), .groups = "drop") %>% filter(Mode == "z") %>% dplyr::select(V) %>% sum()) / (od_tt_m %>% group_by(i,V) %>% summarize(V = sum(V), .groups = "drop") %>% dplyr::select(V) %>% sum())) %>% percent()` of the region's spatial availability in the region - the rest is allocated to the population using faster mode $x$ (of which represents `r ((od_tt_m %>% group_by(i,Mode) %>% summarize(pop = first(Pop.), .groups = "drop") %>% filter(Mode == "x") %>% dplyr::select(pop) %>% sum()) / (od_tt_m %>% group_by(i,Mode) %>% summarize(pop = first(Pop.), .groups = "drop") %>% dplyr::select(pop) %>% sum())) %>% percent()` of the population).
+This comparison between the slower mode $z$ and the other mode $x$ is detailed as follows:
 
-Since the jobs spatially available are consistent with the jobs in the region, it is possible to define a measure of spatial availability per capita as presented in Equation (\ref{eq:SA-per-capita}):
+Potential population $P$ using mode $z$ and $x$ to go to $j$:
+$$
+\sum_{m}^M P_{im} = 210000 \text{ total population}\\
+
+\sum_{m=z} P_{im} = 10000 + 50000 + 3000 = 63000\\
+\sum_{m=x} P_{im} =  40000 + 100000 + 7000 = 147000\\
+$$
+
+Spatial availability $V$ for mode $z$ and $x$:
+$$
+\sum_{m}^M V_{ijm} = 210000 \text{ total jobs}\\
+
+\sum_{m=z} V_{ijm} = 8070.736 + 962.474 + 1.378 + 9004.123 + 21567.552 + 6.891 + 0.493 + 0.263 + 2032.006  = 41645.92\\
+\sum_{m=x} V_{ijm} =  53225.576 + 6347.407 + 40.733 + 29690.578 + 71117.764 + 101.833 + 8.494 + 4.540 + 7817.159 = 168354.1\\
+$$
+
+Noteably, the population that uses the slower mode $x$ is allocated 10% less spatial availability: this means that the population using the faster mode $z$ captures 10% _more_ spatially availability opportunities than the proportion of it's population metered by distance decay demands.
+This exposes an issue of inequality.
+
+This can be explored through travel times.
+The population using the slower mode $z$ only captures 20% of the spatial availability, but accounts for `r ((od_tt_m %>% group_by(i,Mode) %>% summarize(V_tt = sum(V*TT), .groups = "drop") %>% filter(Mode == "z") %>% dplyr::select(V_tt) %>% sum()) / (od_tt_m %>% group_by(i,Mode) %>% summarize(V_tt = sum(V*TT), .groups = "drop") %>% dplyr::select(V_tt) %>% sum())) %>% percent()` of the potential travel time traveled in the region: 6% more travel time than the spatial available opportunities.
+
+Alternatively, this can be explored through proportional benchmarks.
+As the number of opportunities are preserved through the region and within the sub-populations, a spatial availability per capita as presented in Equation (\ref{eq:SA-per-capita}) is defined at each population center, mode, and/or regionally:
 ```{=tex}
@@
 \label{eq:SA-per-capita}
-v_i = \frac{V_i}{P_i}
+v_{im} = \frac{V_{im}}{P_{im}}
 \end{equation}
@@
 
-And, since the jobs are preserved, it is possible to use the regional jobs per capita ($\frac{\sum_{j=1}^J O_j}{\sum_{i=1}^N P_i}$) as a benchmark to compare the spatial availability of jobs per capita at each origin.
+The spatial availability per capita values for $A$, $B$, and $C$ are 1.37, 0.877 and 0.986, respectively.
+For the faster mode $x$, the benchmarks are evidently higher than the averages at: 1.49, 1.01, and 1.12 respectively.
+Starkly lower, especially for $B$ and $C$ are the slower mode $z$ with values of: 0.903, 0.612, and 0.678.
+See as follows:
 
-In the example, since the population is equal to the number of jobs, the regional value of jobs per capita is $1.0$.
-To complete the illustrative example, the spatial availability of jobs per capita by origin is:
-```{r SA-6, eval=TRUE, include=FALSE}
-V_i$V_i[1]/50000
-V_i$V_i[2]/150000
-V_i$V_i[3]/10000
+Potential travel time associated with $V_{ijm}$:
+$$
+\sum_{m}^M V_{ijm} tt_{ijm} = 3,009,957 \text{ minutes}\\
+
+\sum_{m=z} V_{ijm} tt_{ijm} = 774,954 \text{ minutes}\\
+\sum_{m=x}  V_{ijm} tt_{ijm} =  2,235,003 \text { minutes}\\
+$$
+
+Per capita spatial availability $v_{ijm}$, notice how the slower mode $z$ has much lower per capita values than the faster mode $x$ and is lower than the region total benchmarks:
+```{r}
+#use the following to calculate first, the regional averages, and then the origin mode specific (z and then x).
+#od_tt_m %>% group_by(i) %>% summarize(v = first(V_i)/sum(unique(Pop.)), .groups = "drop")
+#od_tt_m %>% group_by(i,Mode) %>% summarize(v = sum(V/Pop.), .groups = "drop") %>% filter(Mode == "z")
+#od_tt_m %>% group_by(i,Mode) %>% summarize(v = sum(V/Pop.), .groups = "drop") %>% filter(Mode == "x")
 ```
 
-```{=tex}
-\begin{equation}
-\label{eq:SA-per-capita-2populations}
-\begin{array}{l}
-v_{1} = \frac{V_1}{P_1} =  \frac{66,833.47}{50,000} = 1.337\\
-v_{2} =  \frac{V_{2}}{P_2} =  \frac{133,203.4}{150,000} = 0.888\\
-v_{3} =  \frac{V_{3}}{P_3} =  \frac{9,963.171}{10,000} = 0.996\\
-\end{array}
-\end{equation}
-```
+$$
+\sum_{m}^M \sum_{i=A} \frac{V_{ijm}}{P_{im}}  = 1.37 \text{ sa jobs per capita}\\
+\sum_{m}^M \sum_{i=B} \frac{V_{ijm}}{P_{im}}  = 0.877 \text{ sa jobs per capita}\\
+\sum_{m}^M \sum_{i=C} \frac{V_{ijm}}{P_{im}}  = 0.986 \text{ sa jobs per capita}\\
 
-We can see that population center $A$ has fewer jobs per capita than the regional benchmark, center $B$ has more, and center $C$ is at parity.
-Remarkably, the spatial availability per capita matches the values of $a_i$ in Table \ref{tab:synthetic-example}.
-Appendix A has a proof of the mathematical equivalence between the two measures.
-It is interesting to notice how Weibull (1976) @weibull_axiomatic_1976 and Shen (1998) @shen1998, as well as this paper, all reach identical expressions starting from different assumptions; this effect is known as _equifinality_ (see Ortúzar & Willumsen (2011) @ortuzar_2011_modelling, p.
-333 and Williams (1981) @williams_hall_1981).
-This result means that Shen-type accessibility and 2SFCA can be re-conceptualized as singly-constrained accessibility measures along with the proposed spatial availability measure.
+\sum_{m=z} \sum_{i=A} \frac{V_{ijm}}{P_{im}}  = 0.903 \text{ sa jobs per capita}\\
+\sum_{m=z} \sum_{i=B} \frac{V_{ijm}}{P_{im}}  = 0.612 \text{ sa jobs per capita}\\
+\sum_{m=z} \sum_{i=C} \frac{V_{ijm}}{P_{im}}  = 0.678 \text{ sa jobs per capita}\\
+
+\sum_{m=x} \sum_{i=A} \frac{V_{ijm}}{P_{im}}  = 1.49 \text{ sa jobs per capita}\\
+\sum_{m=x} \sum_{i=B} \frac{V_{ijm}}{P_{im}}  = 1.01 \text{ sa jobs per capita}\\
+\sum_{m=x} \sum_{i=C} \frac{V_{ijm}}{P_{im}}  = 1.12 \text{ sa jobs per capita}\\
+$$
+As a third way of interpreting spatial availability, spatial availability per capita can be related to the spatial mismatch theory.
+Opportunities should be equal to demand.
+In this way, values below 1 are particularly troubling from the perspective of equity.
 
@@
 
+## Case study background
 <!-- a summary about low emission zones, they are a geographic accessibility restriction. they change the accessibility landscape-->
 
+## Data 
 <!-- -->
+```{r}
+#import travel times
+load("./data-prep/data-raw/intermed/ttresults/output_OD_bike.Rdata")
+load("./data-prep/data-raw/intermed/ttresults/output_OD_car.Rdata")
+load("./data-prep/data-raw/intermed/ttresults/output_OD_transit.Rdata")
+load("./data-prep/data-raw/intermed/ttresults/output_OD_walk.Rdata")
+```
+
+<!-- 
+zn1259 centroids. calculate travel times from orign to destinations -> r5r. Use impedance functions per mode.
+
+with and without madrid centro LEZ.
+
+step 1: zn1259 travel times from centroids-->
 

### 2023-06-05 Multi-mode-acc-methods.qmd
Draft of empirical example - Madrid LEZ

@@
 ---
-title: "How does my active transport access change when car access is restricted? Measuring multi-modal accessibility with spatial availability"
+title: "Extending competitive accessibility for multi-modal analysis"
+author: Anastasia Soukhov, Javier Tarriño Ortiz, Julio A. Soria-Lara, Antonio Páez
 format: docx
@@
 
-An increasing number of studies within the domain of transportation planning are concerned with the assessment of access to opportunity inequities.
-Who (the characteristics of people) and how (what modes are used) people can access opportunities of interest are questions central to planning for equity and towards justice.
-These questions can be addressed, through the imagination of policy intervention scenarios, but methods within the current multi-modal accessibility literature are lacking.
-In this spirit, we present _spatial availability_ within the context of multi-modal accessibility.
-We first illustrate its features.
-We then demonstrate its use on the case study of Low Emission Zones in Madrid (Spain).
-In using spatial availability, a singly-constrained accessibility measure, imagining policy scenarios that move regions towards a more equitable access-to-opportunity landscape is possible.
+An increasing number of studies within the domain of transportation planning are concerned with the inequities in accessibility to opportunities.
+A dimension of these inequities arise from differences in access by mode type (e.g., commuting using a car as opposed to transit).
+However, methods implemented in current multi-modal accessibility literature are lacking.
+This paper presents an extension of _spatial availability_, a singly-constrained competitive accessibility measure, for the context of multi-modal accessibility analysis.
+We first illustrate the features of spatial availability that lend itself to multi-modal analysis.
+We then demonstrate its use on the case study of Low Emission Zones in Madrid (Spain) and highlight how this policy intervention changes the accessibility of populations using different modes.
+In summary, spatial availability can be used to create and interpret multi-modal policy intervention scenarios unlike previous methods: this creation and interpretation can help regions envision a more sustainable and equitable access-to-opportunity landscape.
 
@@
 ## Data 
-<!-- -->
-```{r}
+```{r load-data}
+load("data-prep/data/zn1259.RData") 
+zn1259 <- zn %>% dplyr::select(c("FID", "CD_ZT12", "ZT1259"))
+remove(zn)
+load("data-prep/data/zn208.RData")
+load("data-prep/data/zn_u.RData") # a key to join the zn1259 to the zn208 + contains municipality names
+
+zn1259 <-  merge(zn1259, zn_u, by="ZT1259")
+
+LEZ_centro <- read_sf("data-prep/data-raw/LEZ.shp") %>% st_transform(4326) %>% mutate(merge_all = 1) %>%
+    group_by(merge_all) %>%
+    summarise(geometry = sf::st_union(geometry)) %>%
+    ungroup()
+#add M30 for visualizing
+LEZ_M30 <- read_sf("data-prep/data-raw/Zona_Bajas_Emisiones_MADRID.shp") %>% st_transform(4326)
+
+#creating a list of zones that correspond to the zn208, which indicate if they are in a LEZ.
+LEZ_centro_list <- st_intersection(zn208, LEZ_centro) %>% 
+  st_drop_geometry() %>% 
+  dplyr::select(CD_ZT208) %>% 
+  unlist()
+
+LEZ_M30_list <- st_intersection(zn208, LEZ_M30) %>% 
+  st_drop_geometry() %>% 
+  dplyr::select(CD_ZT208) %>% 
+  unlist()
+
 #import travel times
@@
 
-<!-- 
-zn1259 centroids. calculate travel times from orign to destinations -> r5r. Use impedance functions per mode.
+```{r travel-data-from-survey}
+#import travel survey 2018
+load(file = "data-prep/data/OD_v.RData")
 
-with and without madrid centro LEZ.
+#Keep only the direct home to work trips (no chained home-to-x-to-work trips, they only make up a small portion of home-to-work trips anyways, see the exploration of this in the 'data-exploration.rmd' file)
+OD_v_HW <- OD_v %>% filter(VORI == "1" & VDES == "2" &
+                             VORI_Municipality == "Madrid" & VDES_Municipality == "Madrid") #VORG = 1 = home, VDES = 2
 
-step 1: zn1259 travel times from centroids-->
+#here we separate all the "car" trips, "transit" trips, "bike" trips, and "walk" trips. We then will create 4 different impedance functions
+transit_HW_alltt <- OD_v_HW  %>% 
+  filter(MODO_PRIORITARIO <= 9) %>%
+  transmute(TIEMPO_VIAJES, 
+            mode = "transit")
 
-## Data
+car_motor_HW_alltt <- OD_v_HW  %>% 
+  filter(MODO_PRIORITARIO <= 19 & MODO_PRIORITARIO >= 10) %>% #includes both motorcycles and cars
+  transmute(TIEMPO_VIAJES, 
+            mode = "car/motor")
 
-Two scenarios:
-- 2018: the small low emission zone
-- Applying an optimistic restriction: what if we increased the restrictions?
+bike_HW_alltt <- OD_v_HW  %>% 
+  filter(MODO_PRIORITARIO <= 23 & MODO_PRIORITARIO >= 20) %>%
+  transmute(TIEMPO_VIAJES, 
+            mode = "bike")
 
-## Multi-modal impedance functions
+walk_HW_alltt <- OD_v_HW  %>% filter(MODO_PRIORITARIO == 24) %>%
+  transmute(TIEMPO_VIAJES, 
+            mode = "walk")
+```
 
-## Results of
+Then create the impedance functions:
+```{r fitting-impedance-function}
+# using fitdist function to fit a distribution using the default maximum likelihood estimation method and Nelder-Mead method for direct optimization
+lnorm_ <- fitdistrplus::fitdist(data=transit_HW_alltt$TIEMPO_VIAJES,#%>% unlist() %>% as.numeric(),
+                                "gamma", 
+                                method="mle",
+                                #Notice the use of the weights,
+                                #weights = round(transit_HW_alltt$ELE_G_POND_ESC2),
+                                optim.method="Nelder-Mead")
+HW_transit_lnorm <- lnorm_
 
-#
+gamma_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt$TIEMPO_VIAJES, #%>% unlist() %>% as.numeric(),
+                                "gamma", 
+                                method="mle",
+                                #Notice the use of the weights,
+                                #weights = round(car_motor_HW_alltt$ELE_G_POND_ESC2), 
+                                optim.method="Nelder-Mead")
+HW_car_motor_gamma <- gamma_
+
+lnorm_ <- fitdistrplus::fitdist(data=bike_HW_alltt$TIEMPO_VIAJES, #%>% unlist() %>% as.numeric(),
+                                "lnorm", 
+                                method="mle",
+                                #Notice the use of the weights,
+                                #weights = round(bike_HW_alltt$ELE_G_POND_ESC2),
+                                optim.method="Nelder-Mead")
+HW_bike_lnorm <- lnorm_
+
+logis_ <- fitdistrplus::fitdist(data=walk_HW_alltt$TIEMPO_VIAJES, #%>% unlist() %>% as.numeric(),
+                                "logis", 
+                                method="mle",
+                                #Notice the use of the weights,
+                                #weights = round(walk_HW_alltt$ELE_G_POND_ESC2),
+                                optim.method="Nelder-Mead")
+HW_walk_lnorm <- logis_
+```
+To visualize the impedance functions we create a data frame and calculate the density of the distribution given the estimated parameters:
+```{r}
+x <- data.frame(t = seq(1, 300, 1))
+
+x_transit <- x %>%
+  mutate(f = dgamma(t,
+                    HW_transit_lnorm$estimate[1],
+                    HW_transit_lnorm$estimate[2]),
+         mode = "transit")
+
+x_car_motor <- x %>%
+  mutate(f = dgamma(t,
+                    shape = HW_car_motor_gamma$estimate[1],
+                    rate = HW_car_motor_gamma$estimate[2]),
+         mode = "car/motor")
+
+x_bike <- x %>%
+  mutate(f = dlnorm(t,
+                    HW_bike_lnorm$estimate[1],
+                    HW_bike_lnorm$estimate[2]),
+         mode = "bike")
+
+x_walk <- x %>%
+  mutate(f = dlogis(t,
+                    HW_walk_lnorm$estimate[1],
+                    HW_walk_lnorm$estimate[2]),
+         mode = "walk")
+```
+
+Bind the empirical and theoretical trip length distributions:
+```{r}
+tld_empirical <- rbind(transit_HW_alltt,
+      car_motor_HW_alltt,
+      bike_HW_alltt,
+      walk_HW_alltt) %>%
+  mutate(mode = factor(mode,
+                       levels = c("bike", "walk", "car/motor", "transit")),
+         distribution = "empirical")
+
+tld_theoretical <- rbind(x_transit,
+      x_car_motor,
+      x_bike,
+      x_walk) %>%
+  mutate(mode = factor(mode,
+                       levels = c("bike", "walk", "car/motor", "transit")))
+```
+
+Plot:
+```{r}
+ggplot() + 
+  geom_histogram(data = tld_empirical,
+                 aes(x = TIEMPO_VIAJES, 
+                     #weights = ELE_G_POND_ESC2, 
+                     after_stat(density)),
+                 bins = 45) +
+  geom_line(data = tld_theoretical,
+            aes(x = t, y = f)) + 
+  facet_wrap(~ mode, nrow = 2)
+
+#here, let's double check the empirical trip length distributions, as we can see from the two generated plots, our input TLD is correct!
+rbind(transit_HW_alltt,
+      car_motor_HW_alltt,
+      bike_HW_alltt,
+      walk_HW_alltt) %>%
+  mutate(mode = factor(mode,
+                       levels = c("bike", "walk", "car/motor", "transit"))) %>%
+  ggplot(aes(x = TIEMPO_VIAJES, after_stat(density),
+             #weights = ELE_G_POND_ESC2
+             )) + 
+  geom_histogram(bins = 45) + 
+  facet_wrap(~ mode, nrow = 2)
+```
+
+Next, we define $P_i$ (for the population balancing factor) for each mode and $O_j$ associated with each zone.
+To do this, we add the number of trips (by mode) and create objects which contain the number of workers (origin) and jobs (destinations) - by mode:
+```{r}
+# Instead of creating various columns like above, create a single column with the mode
+OD_v_HW <- OD_v_HW  %>% 
+  mutate(mode = case_when(MODO_PRIORITARIO <= 9 ~ "transit",
+                          MODO_PRIORITARIO <= 19 & MODO_PRIORITARIO >= 10 ~ "car/motor",
+                          MODO_PRIORITARIO <= 23 & MODO_PRIORITARIO >= 20 ~ "bike",
+                          MODO_PRIORITARIO == 24 ~ "walk"),
+         mode = factor(mode,
+                       levels = c("bike", "walk", "car/motor", "transit")))
+
+# The workers are classified by mode and origin. Ex. from Origin 1 five people use bike mode to get to destinations. This is P_i for each m
+workers <- OD_v_HW %>%
+  group_by(VORI_CD_ZT208,
+           mode) %>%
+  summarize(workers = n(), #we should be summing the number of workers by the weight but we don't have a weight... so for now, 1 row = 1 trip. 
+            .groups = "drop") %>%
+  rename(CD_ZT208 = VORI_CD_ZT208)
+
+# Calculate the number of jobs per j ($O_j$)
+jobs <- OD_v_HW %>%
+  group_by(VDES_CD_ZT208) %>%
+  summarize(jobs = n(),
+            .groups = "drop") %>%
+  rename(CD_ZT208 = VDES_CD_ZT208)
+```
+
+Next join the number of workers summed at each origin to the ZN simple features object:
+```{r}
+workers_sf <- workers %>%
+  # Join workers
+  left_join(zn208,
+            by = "CD_ZT208") %>%
+  st_as_sf()
+
+jobs_sf <- jobs %>%
+  # Join jobs
+  left_join(zn208,
+            by = "CD_ZT208") %>%
+  st_as_sf()
+```
+
+Now we reshape the data.
+Each row will represent an any origin to to 1 destination (i.e., destination row is unique).
+We will average the travel times, by mode, and then use this average travel time to find the 'f' (impedance value for travel for each mode).
+Hint: this is why we needed to find the impedance function in the first place!
+The functions represent the cost of travel for each mode based on the travel time for a trip, since we are averaging together the travel time for each mode.
+```{r}
+OD <- OD_v_HW %>%
+  group_by(VORI_CD_ZT208,
+           VDES_CD_ZT208,
+           mode) %>%
+  summarize(tt = mean(TIEMPO_VIAJES),
+            .groups = "drop")
+
+# Check the mean and median travel times by mode
+OD %>% 
+  group_by(mode) %>% 
+  summarize(mtt = mean(tt))
+OD %>% 
+  group_by(mode) %>% 
+  summarize(mtt = median(tt))
+```
+
+Join workers and jobs to OD table:
+```{r}
+OD_2 <- OD %>%
+  left_join(workers,
+            by = c("VORI_CD_ZT208" = "CD_ZT208", "mode"))
+
+OD_2 <- OD_2 %>%
+  left_join(jobs,
+            by = c("VDES_CD_ZT208" = "CD_ZT208"))
+```
+
+Preview the table for one origin and one destination, as an example:
+```{r}
+OD_2 %>% filter(VORI_CD_ZT208 == 1 & VDES_CD_ZT208 == 2) #explanation: 'workers' corresponds to the people who leave ORI 1 to go to any destination by mode (P_im) (e.g., 5 workers leave ORI 1 by bike to go to a job somewhere in the region).
+'jobs' corresponds to the number of workers that come to DES 2 as a job destination in the region (sum_m O_j).
+'tt' corresponds to the average travel time for i to j per mode (avg.
+c_ijm) (e.g., only 1 trip is made by bike from ORI 1 to DES 2 and it is 5 mins, so the average travel time is 5 mins) -- this c_ijm will be used to calculate f(c_ijm).
+```
+
+Estimate f value using these averaged travel times:
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+OD_2 <- OD_2 %>%
+  mutate(f = case_when(mode == "transit" ~ dgamma(tt,
+                                                  HW_transit_lnorm$estimate["shape"], 
+                                                  HW_transit_lnorm$estimate["rate"]),
+                       mode == "car/motor" ~ dgamma(tt,
+                                                    HW_car_motor_gamma$estimate["shape"],
+                                                    HW_car_motor_gamma$estimate["rate"]),
+                       mode == "bike" ~ dlnorm(tt,
+                                               HW_bike_lnorm$estimate["meanlog"],
+                                               HW_bike_lnorm$estimate["sdlog"]),
+                       mode == "walk" ~ dlogis(tt, 
+                                               HW_walk_lnorm$estimate["location"], 
+                                               HW_walk_lnorm$estimate["scale"])))
+```
+
+## Calculating spatial availability 
+Calculate spatial availability:
+```{r}
+OD_SA <- OD_2 %>%
+  mutate(catch = 1) %>%
+  mutate(SA_ij = sp_avail(., 
+                          o_id = VORI_CD_ZT208,
+                          d_id = VDES_CD_ZT208,
+                          pop = workers,
+                          opp = jobs,
+                          r = catch,
+                          f = f))
+```
+
+Check that the totals are maintained, it is!:
+```{r}
+sum(OD_SA$SA_ij, na.rm = T)
+sum(jobs$jobs)
+#sum(OD_SA$tt)
+```
+Spatial availability by mode, we can see that car mode captures 71% of the spatial availability but only accounts for XXXX of the tt_ijm !!:
+```{r}
+OD_SA %>%
+  group_by(mode) %>%
+  summarize(V_m = sum(SA_ij),
+            V_m_perc = sum(SA_ij)/27795,
+            tt_m = sum(tt)/780315.1)
+```
+
+Okay, let's preform a few more checks and then aggregate spatial availability by origin (zn208) and by origin,mode.
+
+Spatial availability by origin (with check).
+Here we represent spatial availability per origin across all modes for each i.
+We know that car mode represents the majority of spatial availability per i overall:
+```{r}
+SA_i <- OD_SA %>%
+  group_by(VORI_CD_ZT208) %>%
+  summarize(V_i = sum(SA_ij))
+
+summary(SA_i) 
+sum(SA_i$V_i)
+```
+
+But we can suspect that the spatial distribution of SA is uneven depending on the mode. So here we represent Spatial availability by origin, mode (with check):
+```{r}
+SA_im <- OD_SA %>%
+  group_by(VORI_CD_ZT208, 
+           mode) %>%
+  summarize(V_im = sum(SA_ij),
+            .groups = "drop")
+
+summary(SA_im)
+sum(SA_im$V_im)
+```
+
+Let's preform some more checks. Calculate worker population by zone of origin:
+```{r}
+population <- workers %>%
+  group_by(CD_ZT208) %>%
+  summarize(population = sum(workers))
+
+sum(population$population)
+```
+
+Finally, let's join the zonal worker population to the SA_i and SA_im spatial objects (with checks):
+```{r}
+SA_i <- SA_i %>%
+  left_join(population,
+            by = c("VORI_CD_ZT208" = "CD_ZT208"))
+sum(SA_i$V_i)
+
+SA_im <- SA_im %>%
+  left_join(population,
+            by = c("VORI_CD_ZT208" = "CD_ZT208"))
+sum(SA_im$V_im)
+```
+
+To facilitate the interpretation of spatial availability, here we calculate spatial availability per capita:
+```{r}
+SA_i <- SA_i %>%
+  mutate(v_i = V_i/population)
+
+summary(SA_i)
+
+SA_im <- SA_im %>%
+  mutate(v_im = V_im/population)
+
+summary(SA_im)
+```
+
+Lastly, join geometry of zones to SA so we can visualize:
+```{r}
+SA_i <- SA_i %>%
+  left_join(zn208 %>%
+              dplyr::select(CD_ZT208),
+            by =c("VORI_CD_ZT208" = "CD_ZT208")) %>%
+  st_as_sf()
+
+SA_im <- SA_im %>%
+  left_join(zn208 %>%
+              dplyr::select(CD_ZT208),
+            by = c("VORI_CD_ZT208" = "CD_ZT208")) %>%
+  st_as_sf()
+```
+
+## Plotting:
+
+Figure out which CD_ZT208 correspond to Madrid - we see that CD_ZT208 1 through 59 do so.
+```{r}
+t <- zn_u %>% dplyr::select(CD_ZT208, Municipality) %>% group_by(CD_ZT208,Municipality) %>% summarize() %>% filter(Municipality == "Madrid")
+t
+```
+
+```{r create-bounding-boxes}
+zn208 <- zn208 %>% mutate(MM = ifelse(str_starts(ZT208,"Madrid"), 1, 0)) %>% filter(MM == 1)
+
+#create the bounding box for the Toronto region + a little extra
+bbox_new1 <- st_bbox(zn208 %>%filter(MM == 1)) # current bounding box
+
+xrange <- bbox_new1$xmax - bbox_new1$xmin # range of x values
+yrange <- bbox_new1$ymax - bbox_new1$ymin # range of y values
+
+bbox_new1[1] <- bbox_new1[1] - (0.0 * xrange) # xmin - left
+bbox_new1[3] <- bbox_new1[3] + (0.0 * xrange) # xmax - right
+bbox_new1[2] <- bbox_new1[2] - (0.0 * yrange) # ymin - bottom
+bbox_new1[4] <- bbox_new1[4] + (0.0 * yrange) # ymax - top
+
+bbox_Madrid_muni <- bbox_new1 %>%  # take the bounding box ...
+  st_as_sfc()
+
+sf::sf_use_s2(FALSE) #this terns off the spherical geometry component in the st_union() function - this assumes the sf is just planar. Thats' fine for our purposes
+Madrid_muni_border <- zn208 %>%
+  filter(MM == 1) %>%
+  st_union()
+```
+
+Plot V (spatial availability) by origin:
+```{r}
+ggplot() +
+  geom_sf(data = SA_i,
+          aes(fill = V_i),
+          color = NA) +
+  scale_fill_distiller(palette = "YlGn", #legend scale bar
+                       name = expression(v["i"]),
+                       na.value = "grey90",
+                       direction = 1,
+                       #trans = "log",
+                       #breaks=c(0, 10, 100, 1000, 10000)
+                       )+
+  geom_sf(data = Madrid_muni_border, # border for Madrid muni
+          colour=alpha("black",1), 
+          size = 0.5, 
+          fill=NA, 
+          show.legend = "polygon") +
+  geom_sf(data = LEZ_centro, 
+          colour=alpha("red",1), 
+          size = 0.5, 
+          fill = NA) + 
+  geom_sf(data = LEZ_M30, 
+          colour=alpha("cyan",1), 
+          size = 0.5, 
+          fill = NA) +
+  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
+                         height = unit(0.8, "cm"), 
+                         width = unit(0.8, "cm"),
+                         style = north_arrow_orienteering(line_width = 0.25,
+                                                          line_col = "dimgrey", 
+                                                          fill = c("white","dimgrey"))) +
+  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
+                   height = unit(0.15, "cm")) +
+  theme_void() +
+  theme(legend.position = c(1.10, 0.25), 
+        plot.margin = margin(t = 0, r = 0, b = 0, l = -80)) # positing legend and clipping out white space
+```
+
+<!-- Two scenarios: -->
+<!-- - 2018: the small low emission zone -->
+<!-- - Applying an optimistic restriction: what if we increased the restrictions?  -->

### 2023-06-05 Multi-mode-acc-methods.qmd
Revised the impedance function calc. for the empirical example: now only for within Madrid home-to-work trips

@@
 
-#Keep only the direct home to work trips (no chained home-to-x-to-work trips, they only make up a small portion of home-to-work trips anyways, see the exploration of this in the 'data-exploration.rmd' file)
+#Keep only the direct home to work trips for the spatial availability calculation (no chained home-to-x-to-work trips, they only make up a small portion of home-to-work trips anyways, see the exploration of this in the 'data-exploration.rmd' file)
 OD_v_HW <- OD_v %>% filter(VORI == "1" & VDES == "2" &
-                             VORI_Municipality == "Madrid" & VDES_Municipality == "Madrid") #VORG = 1 = home, VDES = 2
+                             VORI_Municipality == "Madrid" & VDES_Municipality == "Madrid" &
+                             TIEMPO_VIAJES <= 120) #VORG = 1 = casa, VDES = 2 = trabajo
 
@@
 transit_HW_alltt <- OD_v_HW  %>% 
-  filter(MODO_PRIORITARIO <= 9) %>%
+  filter(MODO_PRIORITARIO <= 9) %>% #(1.
+Renfe Cercanías ', 2.
+Autobus interurbano, 3.
+Autobus urbano otro municipio, 4.
+Metro, 5.
+Metro ligero/tranvía, 6.
+Autobus urbano Madrid EMT, 7.
+Resto renfe, 8.
+Autobus discrecional, 9.
+Autobus de largo recorrido)
   transmute(TIEMPO_VIAJES, 
+            ELE_G_POND_ESC2,
             mode = "transit")
@@
 car_motor_HW_alltt <- OD_v_HW  %>% 
-  filter(MODO_PRIORITARIO <= 19 & MODO_PRIORITARIO >= 10) %>% #includes both motorcycles and cars
+  filter(MODO_PRIORITARIO <= 19 & MODO_PRIORITARIO >= 10) %>% #includes both motorcycles and cars (specifically: 10.
+Taxi, 11.
+Coche conductor particular, 12.
+Coche conductor empresa, 13.
+Coche conductor alquiler sin conductor, 14.
+Coche pasajero particular, 15.
+Coche pasajero empresa, 16.
+Coche pasajero alquiler con conductor, 17.
+Moto/ciclomotor particular, 18.
+Moto/ciclomotor publica, 19.
+Moto/ciclomotor empresa)
   transmute(TIEMPO_VIAJES, 
+            ELE_G_POND_ESC2,
             mode = "car/motor")
@@
 bike_HW_alltt <- OD_v_HW  %>% 
-  filter(MODO_PRIORITARIO <= 23 & MODO_PRIORITARIO >= 20) %>%
+  filter(MODO_PRIORITARIO <= 23 & MODO_PRIORITARIO >= 20) %>% #includes cycling and 'otros', specificially: 20. Bicicleta particular, 21. Bicicleta publica, 22. Bicicleta empresa, 23. Otros)
   transmute(TIEMPO_VIAJES, 
+            ELE_G_POND_ESC2,
             mode = "bike")
 
-walk_HW_alltt <- OD_v_HW  %>% filter(MODO_PRIORITARIO == 24) %>%
+walk_HW_alltt <- OD_v_HW  %>% filter(MODO_PRIORITARIO == 24) %>% #"24. Andando/pie"
   transmute(TIEMPO_VIAJES,
+            ELE_G_POND_ESC2,
             mode = "walk")
@@
 
-Then create the impedance functions:
-```{r fitting-impedance-function}
+```{r fitting-impedance-function1}
+#see which curve could fit the best using the descdist() function
+
+#transit
+transit_HW_alltt_long <- transit_HW_alltt[rep(seq_len(dim(transit_HW_alltt %>% dplyr::select(-mode))[1]), transit_HW_alltt$ELE_G_POND_ESC2), 1]
+fitdistrplus::descdist(data=transit_HW_alltt_long) #try beta, uniform, and normal
+
+#car
+car_motor_HW_alltt_long <- car_motor_HW_alltt[rep(seq_len(dim(car_motor_HW_alltt %>% dplyr::select(-mode))[1]), car_motor_HW_alltt$ELE_G_POND_ESC2), 1]
+fitdistrplus::descdist(data=car_motor_HW_alltt_long) #try beta, gamma
+
+#bike
+bike_HW_alltt_long <- bike_HW_alltt[rep(seq_len(dim(bike_HW_alltt %>% dplyr::select(-mode))[1]), bike_HW_alltt$ELE_G_POND_ESC2), 1]
+fitdistrplus::descdist(data=bike_HW_alltt_long) #try beta, gamma
+
+#walk
+walk_HW_alltt_long <- walk_HW_alltt[rep(seq_len(dim(walk_HW_alltt %>% dplyr::select(-mode))[1]), walk_HW_alltt$ELE_G_POND_ESC2), 1]
+fitdistrplus::descdist(data=walk_HW_alltt_long) #try beta, gamma, exp
+
+```
+
+```{r fitting-impedance-function2-transit}
 # using fitdist function to fit a distribution using the default maximum likelihood estimation method and Nelder-Mead method for direct optimization
-lnorm_ <- fitdistrplus::fitdist(data=transit_HW_alltt$TIEMPO_VIAJES,#%>% unlist() %>% as.numeric(),
+
+#for transit, try: uniform, normal, gamma, lognormal
+gamma_ <- fitdistrplus::fitdist(data=transit_HW_alltt$TIEMPO_VIAJES,
                                 "gamma", 
                                 method="mle",
-                                #Notice the use of the weights,
-                                #weights = round(transit_HW_alltt$ELE_G_POND_ESC2),
+                                weights = round(transit_HW_alltt$ELE_G_POND_ESC2),
                                 optim.method="Nelder-Mead")
-HW_transit_lnorm <- lnorm_
-
-gamma_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt$TIEMPO_VIAJES, #%>% unlist() %>% as.numeric(),
+unif_ <- fitdistrplus::fitdist(data=transit_HW_alltt$TIEMPO_VIAJES,
+                                "unif", 
+                                method="mle",
+                                weights = round(transit_HW_alltt$ELE_G_POND_ESC2),
+                               optim.method="Nelder-Mead")
+norm_ <- fitdistrplus::fitdist(data=transit_HW_alltt$TIEMPO_VIAJES,
+                                "norm", 
+                                method="mle",
+                                weights = round(transit_HW_alltt$ELE_G_POND_ESC2),
+                                optim.method="Nelder-Mead")
+lnorm_ <- fitdistrplus::fitdist(data=transit_HW_alltt$TIEMPO_VIAJES,
+                                "lnorm", 
+                                method="mle",
+                                weights = round(transit_HW_alltt$ELE_G_POND_ESC2),
+                                optim.method="Nelder-Mead")
+
+gamma_$aic #lowest aic and bic, so we pick this distribution.
+unif_$aic 
+norm_$aic
+lnorm_$aic
+
+gamma_$bic #lowest aic and bic, so we pick this distribution.
+unif_$bic 
+norm_$bic
+lnorm_$bic
+
+HW_transit_gamma <- gamma_
+```
+```{r fitting-impedance-function2-car}
+#for car, try: uniform, lnorm
+gamma_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt$TIEMPO_VIAJES,
                                 "gamma", 
                                 method="mle",
-                                #Notice the use of the weights,
-                                #weights = round(car_motor_HW_alltt$ELE_G_POND_ESC2), 
+                                weights = round(car_motor_HW_alltt$ELE_G_POND_ESC2),
                                 optim.method="Nelder-Mead")
-HW_car_motor_gamma <- gamma_
-
-lnorm_ <- fitdistrplus::fitdist(data=bike_HW_alltt$TIEMPO_VIAJES, #%>% unlist() %>% as.numeric(),
+
+lnorm_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt$TIEMPO_VIAJES,
                                 "lnorm", 
                                 method="mle",
-                                #Notice the use of the weights,
-                                #weights = round(bike_HW_alltt$ELE_G_POND_ESC2),
+                                weights = round(car_motor_HW_alltt$ELE_G_POND_ESC2),
                                 optim.method="Nelder-Mead")
+
+gamma_$aic #lowest aic and bic, so we pick this distribution.
+lnorm_$aic
+
+gamma_$bic #lowest aic and bic, so we pick this distribution.
+lnorm_$bic
+
+HW_car_motor_gamma <- gamma_
+```
+```{r fitting-impedance-function2-bike}
+#for bike, try: uniform, lnorm
+gamma_ <- fitdistrplus::fitdist(data=bike_HW_alltt$TIEMPO_VIAJES,
+                                "gamma", 
+                                method="mle",
+                                weights = round(bike_HW_alltt$ELE_G_POND_ESC2),
+                                optim.method="Nelder-Mead")
+
+lnorm_ <- fitdistrplus::fitdist(data=bike_HW_alltt$TIEMPO_VIAJES,
+                                "lnorm", 
+                                method="mle",
+                                weights = round(bike_HW_alltt$ELE_G_POND_ESC2),
+                                optim.method="Nelder-Mead")
+
+gamma_$aic 
+lnorm_$aic #lowest aic and bic, so we pick this distribution.
+
+gamma_$bic 
+lnorm_$bic #lowest aic and bic, so we pick this distribution.
+
 HW_bike_lnorm <- lnorm_
-
-logis_ <- fitdistrplus::fitdist(data=walk_HW_alltt$TIEMPO_VIAJES, #%>% unlist() %>% as.numeric(),
-                                "logis", 
+```
+
+```{r fitting-impedance-function2-walk}
+#for walk, try: uniform, lnorm, normal
+gamma_ <- fitdistrplus::fitdist(data=walk_HW_alltt$TIEMPO_VIAJES,
+                                "gamma", 
                                 method="mle",
-                                #Notice the use of the weights,
-                                #weights = round(walk_HW_alltt$ELE_G_POND_ESC2),
+                                weights = round(walk_HW_alltt$ELE_G_POND_ESC2),
                                 optim.method="Nelder-Mead")
-HW_walk_lnorm <- logis_
-```
+
+lnorm_ <- fitdistrplus::fitdist(data=walk_HW_alltt$TIEMPO_VIAJES,
+                                "lnorm", 
+                                method="mme",
+                                weights = round(walk_HW_alltt$ELE_G_POND_ESC2),
+                                optim.method="Nelder-Mead")
+norm_ <- fitdistrplus::fitdist(data=walk_HW_alltt$TIEMPO_VIAJES,
+                                "norm", 
+                                method="mle",
+                                weights = round(walk_HW_alltt$ELE_G_POND_ESC2),
+                                optim.method="Nelder-Mead")
+
+gamma_$aic 
+lnorm_$aic #lowest aic and bic, so we pick this distribution.
+norm_$aic
+
+gamma_$bic 
+lnorm_$bic #lowest aic and bic, so we pick this distribution.
+norm_$bic
+
+HW_walk_lnorm <- lnorm_
+
+```
+
 To visualize the impedance functions we create a data frame and calculate the density of the distribution given the estimated parameters:
 ```{r}
-x <- data.frame(t = seq(1, 300, 1))
+x <- data.frame(t = seq(1, 120, 1))
 
@@
   mutate(f = dgamma(t,
-                    HW_transit_lnorm$estimate[1],
-                    HW_transit_lnorm$estimate[2]),
+                    HW_transit_gamma$estimate[1],
+                    HW_transit_gamma$estimate[2]),
          mode = "transit")
@@
 x_walk <- x %>%
-  mutate(f = dlogis(t,
+  mutate(f = dlnorm(t,
                     HW_walk_lnorm$estimate[1],
@@
                  aes(x = TIEMPO_VIAJES, 
-                     #weights = ELE_G_POND_ESC2, 
+                     weights = ELE_G_POND_ESC2, 
                      after_stat(density)),
-                 bins = 45) +
+                 bins = 20) +
   geom_line(data = tld_theoretical,
@@
   ggplot(aes(x = TIEMPO_VIAJES, after_stat(density),
-             #weights = ELE_G_POND_ESC2
+             weights = ELE_G_POND_ESC2
              )) + 
-  geom_histogram(bins = 45) + 
+  geom_histogram(bins = 20) + 
   facet_wrap(~ mode, nrow = 2)
@@
 
-Next, we define $P_i$ (for the population balancing factor) for each mode and $O_j$ associated with each zone.
-To do this, we add the number of trips (by mode) and create objects which contain the number of workers (origin) and jobs (destinations) - by mode:
-```{r}
+```{r define-population-balancing-factor}
+# We define $P_i$ (for the population balancing factor) for each mode and $O_j$ associated with each zone.
+To do this, we add the number of trips (by mode) and create objects that contain the number of workers (origin) and jobs (destinations) - by mode:
+
 # Instead of creating various columns like above, create a single column with the mode
@@
 workers <- OD_v_HW %>%
-  group_by(VORI_CD_ZT208,
+  group_by(VORIZT1259,
            mode) %>%
-  summarize(workers = n(), #we should be summing the number of workers by the weight but we don't have a weight... so for now, 1 row = 1 trip. 
-            .groups = "drop") %>%
-  rename(CD_ZT208 = VORI_CD_ZT208)
+  summarize(workers = n()*ELE_G_POND_ESC2, #we should be summing the number of workers by the weight but we don't have a weight... so for now, 1 row = 1 trip. 
+            .groups = "drop")%>%
+  rename(ZT1259 = VORIZT1259)
 
@@
 jobs <- OD_v_HW %>%
-  group_by(VDES_CD_ZT208) %>%
-  summarize(jobs = n(),
+  group_by(VDESZT1259) %>%
+  summarize(jobs = n()*ELE_G_POND_ESC2,
             .groups = "drop") %>%
-  rename(CD_ZT208 = VDES_CD_ZT208)
+  rename(ZT1259 = VDESZT1259)
 ```
@@
   # Join workers
-  left_join(zn208,
-            by = "CD_ZT208") %>%
+  left_join(zn1259,
+            by = "ZT1259") %>%
   st_as_sf()
@@
   # Join jobs
-  left_join(zn208,
-            by = "CD_ZT208") %>%
+  left_join(zn1259,
+            by = "ZT1259") %>%
   st_as_sf()
@@
 OD <- OD_v_HW %>%
-  group_by(VORI_CD_ZT208,
-           VDES_CD_ZT208,
+  group_by(VORIZT1259,
+           VDESZT1259,
            mode) %>%
@@
   group_by(mode) %>% 
-  summarize(mtt = mean(tt))
-OD %>% 
-  group_by(mode) %>% 
-  summarize(mtt = median(tt))
+  summarize(mean_tt = mean(tt),
+            median_tt = median(tt),
+            min_tt = min(tt),
+            max_tt = max(tt))
 ```
@@
   left_join(workers,
-            by = c("VORI_CD_ZT208" = "CD_ZT208", "mode"))
+            by = c("VORIZT1259" = "ZT1259", "mode"))
 
@@
   left_join(jobs,
-            by = c("VDES_CD_ZT208" = "CD_ZT208"))
+            by = c("VDESZT1259" = "ZT1259"))
 ```
@@
 ```{r}
-OD_2 %>% filter(VORI_CD_ZT208 == 1 & VDES_CD_ZT208 == 2) #explanation: 'workers' corresponds to the people who leave ORI 1 to go to any destination by mode (P_im) (e.g., 5 workers leave ORI 1 by bike to go to a job somewhere in the region).
+OD_2 %>% filter(VORIZT1259 == 1 & VDESZT1259 == 2) #explanation: 'workers' corresponds to the people who leave ORI 1 to go to any destination by mode (P_im) (e.g., 5 workers leave ORI 1 by bike to go to a job somewhere in the region).
 'jobs' corresponds to the number of workers that come to DES 2 as a job destination in the region (sum_m O_j).

### 2023-06-06 Multi-mode-acc-methods.qmd
Small updates

@@
            mode) %>%
-  summarize(workers = n()*ELE_G_POND_ESC2, #we should be summing the number of workers by the weight but we don't have a weight... so for now, 1 row = 1 trip. 
+  summarize(workers = n(), #we should be summing the number of workers by the weight but we don't have a weight... so for now, 1 row = 1 trip. 
             .groups = "drop")%>%
@@
   group_by(VDESZT1259) %>%
-  summarize(jobs = n()*ELE_G_POND_ESC2,
+  summarize(jobs = n(),
             .groups = "drop") %>%
@@
             by = c("VDESZT1259" = "ZT1259"))
+
+#THIS IS TOO MANY!
 ```
@@
 ```{r}
-OD_2 %>% filter(VORIZT1259 == 1 & VDESZT1259 == 2) #explanation: 'workers' corresponds to the people who leave ORI 1 to go to any destination by mode (P_im) (e.g., 5 workers leave ORI 1 by bike to go to a job somewhere in the region).
+# OD_2 %>% filter(VORIZT1259 == "079-09-225" & VDESZT1259 == 2) #explanation: 'workers' corresponds to the people who leave ORI 1 to go to any destination by mode (P_im) (e.g., 5 workers leave ORI 1 by bike to go to a job somewhere in the region).
 'jobs' corresponds to the number of workers that come to DES 2 as a job destination in the region (sum_m O_j).
@@
   mutate(f = case_when(mode == "transit" ~ dgamma(tt,
-                                                  HW_transit_lnorm$estimate["shape"], 
-                                                  HW_transit_lnorm$estimate["rate"]),
+                                                  HW_transit_gamma$estimate["shape"], 
+                                                  HW_transit_gamma$estimate["rate"]),
                        mode == "car/motor" ~ dgamma(tt,
@@
                                                HW_bike_lnorm$estimate["sdlog"]),
-                       mode == "walk" ~ dlogis(tt, 
+                       mode == "walk" ~ dlnorm(tt, 
                                                HW_walk_lnorm$estimate["location"], 
@@
   mutate(SA_ij = sp_avail(., 
-                          o_id = VORI_CD_ZT208,
-                          d_id = VDES_CD_ZT208,
+                          o_id = VORIZT1259,
+                          d_id = VDESZT1259,
                           pop = workers,
@@
 SA_i <- OD_SA %>%
-  group_by(VORI_CD_ZT208) %>%
+  group_by(VORIZT1259) %>%
   summarize(V_i = sum(SA_ij))

### 2023-06-10 CIT-Presentation.Rmd
CIT-Presentation draft!

---
title: "Multi-modal spatial availability: a research proposal"
subtitle: "CIT - Tenerife 2023"
author: "Anastasia Soukhov, J. Tarriño Ortiz, J. A. Soria-Lara, A. Páez"
date: "2022-11-10"
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current% / 12" 
      highlightStyle: github
      highlightLines: true
      ratio: '16:9'
      countIncrementalSlides: false
      beforeInit: "macros.js"
---
```{r xaringan-themer, include=FALSE, warning=FALSE}
#this themer package allows us to change the .css themes through this chunk. Easier than changing the .css directly.
library(xaringanthemer)
style_mono_accent(
  base_color = "#7A003C", #maroon
  link_color = "#5E6A71", #grey
  text_bold_color = "#FDBF57", #gold
  text_slide_number_color = "#FDBF57", #gold
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i"),
  code_font_google   = google_font("Fira Mono")
)
```

```{r load-packages, include=FALSE, cache=FALSE}
library(dplyr)
library(sf)
library(ggplot2)
library(tidyverse)
library(units)
library(patchwork)
library(ggspatial)
library(ggforce)
library(geomtextpath)
library(knitr)
library(kableExtra)
options(scipen = 999)
```

```{r sp_avail-function,include=FALSE}
#defining the spatial availability function, this prints all the intermediates
sp_avail_detailed <- function(x, o_id, d_id, pop, opp, r, f, alpha = 1){

  o_id <- rlang::enquo(o_id)
  d_id <- rlang::enquo(d_id)
  pop <- rlang::enquo(pop)
  opp <- rlang::enquo(opp)
  r <- rlang::enquo(r)
  f <- rlang::enquo(f)

  sum_pop <- x %>%
    dplyr::distinct(!!o_id,
                    .keep_all = TRUE) %>%
    dplyr::mutate(sum_pop = !!r*(!!pop)^alpha) %>%
    dplyr::pull(sum_pop) %>%
    sum()

  f_p <- dplyr::pull(x, !!r) * dplyr::pull(x, !!pop)^alpha / sum_pop

  sum_impedance <- x %>%
    dplyr::group_by(!!d_id) %>%
    dplyr::summarize(sum_impedance = sum(!!f))

  x <- x %>%
    dplyr::left_join(sum_impedance,
                     by = rlang::as_name(d_id))

  f_c <- dplyr::pull(x, !!f) / x$sum_impedance

  x$f_c <- f_c
  x$f_p <- f_p

  sum_pa <- x %>%
    dplyr::group_by(!!d_id) %>%
    dplyr::summarize(sum_pa= sum(f_p * f_c))

  x <- x %>%
    dplyr::left_join(sum_pa,
                     by = rlang::as_name(d_id))
  f_t <- (f_p * f_c) / dplyr::pull(x, sum_pa)

  x %>%
    dplyr::mutate(V_ij = !!opp * !!f_t)
}
```

```{r synthetic-data, include=FALSE}
od_tt <- data.frame(i = c("A", "A", "A", "B", "B", "B", "C", "C", "C"), # Three origins
                    j = c("1", "2", "3", "1", "2", "3", "1", "2", "3"), # Three destinations
                    tt_z = c(15, 30, 100, 30, 15, 100, 100, 100, 15), # Travel time, z-mode (original)
                    tt_x = c(10, 25, 80, 25, 10, 80, 80, 80, 10), # Travel time, x-mode (the faster mode - like car)
                    pop = c(50000, 50000, 50000, 150000, 150000, 150000, 10000, 10000, 10000), # Population
                    opp = c(100000, 100000, 10000, 100000, 100000, 10000, 100000, 100000, 10000)) # Jobs

#write the dataframe long.
od_tt_m <- data.frame(i = c("A", "A", "A", "A", "A", "A",
                            "B", "B", "B", "B", "B", "B",
                            "C", "C", "C", "C", "C", "C"), # Three origins
                    j = c("1", "2", "3", "1", "2", "3", 
                          "1", "2", "3", "1", "2", "3", 
                          "1", "2", "3", "1", "2", "3"), # Three destinations
                    m = c("pt", "pt", "pt", "car", "car", "car",
                          "pt", "pt", "pt", "car", "car", "car",
                          "pt", "pt", "pt", "car", "car", "car"),
                    tt = c(15, 30, 100, 10, 25, 80, #travel time, z-mode (original) AND THEN x-mode (the faster mode)
                           30, 15, 100, 25,  10,  80,
                           100, 100, 15, 80, 80, 10), 
                    pop = c(20000, 20000, 20000, 30000, 30000, 30000,
                            50000, 50000,50000, 100000, 100000, 100000,
                            3000, 3000, 3000, 7000,7000, 7000),
                    opp = c(100000, 100000, 10000, 100000, 100000, 10000,
                            100000, 100000, 10000, 100000, 100000, 10000,
                            100000, 100000, 10000, 100000, 100000, 10000)) 
```

```{r data-figure-with-toy-example, include=FALSE}
od <- data.frame(id = c("A", "B", "C", "1", "2", "3"),
                 type = c("Population", "Population", "Population", "Jobs", "Jobs", "Jobs"),
                 size = c(50000, 150000, 10000, 100000, 100000, 10000),
                 x = c(2.5, 2.5, 6.5, 0.5, 0.5, 6.5),
                 y = c(8.0, 2.5, 4.5, 7.5, 2.5, 2.5))

centers <- data.frame(id = c("Urban center", "Suburb", "Satellite town"),
                 radius = rep(1.3, 3),
                 x = c(1.5, 1.5, 6.5),
                 y = c(2.5, 7.5, 3.5))

# od_lines coded as segments
od_lines <- data.frame(x = c(2.5, 2.5, 2.5, 2.5, 2.5, 2.5, 6.5, 6.5, 6.5),
                       y = c(8.0, 8.0, 8.0, 2.5, 2.5, 2.5, 4.5, 4.5, 4.5),
                       xend = c(0.5, 0.5, 6.5, 0.5, 0.5, 6.5, 0.5, 0.5, 6.5),
                       yend = c(7.5, 2.5, 2.5, 7.5, 2.5, 2.5, 7.5, 2.5, 2.5),
                       group = c("A", "A", "A", "B", "B", "B", "C", "C", "C"),
                       time = c("15 min (pt)\n10 min (car)", "30 min (pt)\n25 min (car)", "100 min (pt)\n80 min (car)",
                                "30 min (pt)\n25 min (car)", "15 min (pt)\n10 min (car)", "100 min (pt)\n80 min (car)",
                                "100 min (pt)\n80 min (car)", "100 min (pt)\n80 min (car)", "15 min (pt)\n10 min (car)"))

od_table <- data.frame(Center = c("A", "B", "C", "1", "2", "3"),
  Size = c("50,000 pop (40% pt)", "150,000 pop (33% pt)", "10,000 pop (30% pt)", 
           "100,000 jobs", "100,000 jobs", "10,000 jobs"))

od_table2 <- data.frame(Path = c("A to 1", "A to 2", "A to 3", 
                                "B to 1", "B to 2", "B to 3", 
                                "C to 1", "C to 2", "C to 3"),
  Value = c("15 min (pt)\n10 min (car)", "30 min (pt)\n25 min (car)", "100 min (pt)\n80 min (car)",
            "30 min (pt)\n25 min (car)", "15 min (pt)\n10 min (car)", "100 min (pt)\n80 min (car)",
            "100 min (pt)\n80 min (car)", "100 min (pt)\n80 min (car)", "15 min (pt)\n10 min (car)"))
```

```{r, create-figure-with-toy-example, echo=FALSE, eval=FALSE}
toy_example_plot <- ggplot() + 
  # Plot centers
  geom_circle(data = centers,
             aes(x0 = x, 
                 y0 = y,
                 r = radius),
             color = "grey40") +
  annotate(geom = "label", 
           x=c(0.5, 0.6, 7),
           y=c(9, 1, 5.2), 
           label = c("Suburban", "Urban", "Satellite"), 
           size=4, 
           label.size=NA) +
  # Connect origins and destinations with curves
  geom_curve(data = od_lines,
               aes(x = x,
                   y = y,
                   xend = xend,
                   yend = yend,
                   linetype = group,
                   group = group),
             curvature = 0.20,
             linewidth = 1,
             color = "black") +
  # Text on curves
  geom_textcurve(data = od_lines,
               aes(x = x,
                   y = y,
                   xend = xend,
                   yend = yend,
                   linetype = group,
                   group = group,
                   label = time),
               size = 3,
             curvature = 0.20,
             color = "black") +
  # Plot origins and destinations
  geom_point(data = od,
            aes(size = size,
                x = x, 
                y = y, 
                color = type,
                shape = type)) +
  scale_size(range = c(4, 10)) +
  # Label origins and destinations
  geom_text(data = od, 
            aes(x, 
                y, 
                label=id), 
            size=3) +
  # Create table
  annotate(geom = "table",
           x = 9, 
           y = 9, 
           label = list(od_table),
           fill = "white",
           size=3) +
  # Theme the plot
  coord_fixed() + 
  theme_void() + guides(shape = "none", size = "none", color = "none", linetype ="none")

toy_example_plot
ggsave("Presentations/toy-example.png",dpi=600)
```

## Overview

**1.** Accessibility measures and multi-modal applications

**2.** Explaining **spatial availability** within multi-modal analysis

**3.** Policy scenario planning application: low emission zones in Madrid

**4.** Implications: the changes in spatial availability of opportunities, active mobility, GHG emissions

???
XXX
---
## Multi-modal accessibility measures

.pull-left[

#### Gravity-based accessibility (Hansen-type) - no competition

$$A_i^m = \sum_{j}O_jf(c_{ij})^m$$
-   $i$ is a set of origin locations
-   $j$ is a set of destination locations
-   $O_j$ is the number of opportunities at location $j$
-   $f(\cdot)$ is an impedance function of $c_{ij}$ (cost of moving between $i$ and $j$) for each mode $m$

**E.g.**, used in _(Tahmasbi et al., 2019)_.

Suffers from the same interpretability issues _(Handy and Niemeier, 1997; Miller et al., 2018)_.

Does not include competition (demand:supply or the mass effect) _(Shen, 1998; Merlin and Hu, 2018)_.

]

???

Interpretability issues are not resolved when considering multiple modes.

**e.g.**, used in _(Tahmasbi et al., 2019)_ for quantify access $A_i$ for $m$ walking, public transit, and private car modes to retail store destinations (different $f^m(c_{ij}^m)$ for each $m$)
- Suffers from the same interpretability issues _(Handy and Niemeier, 1997; Miller et al., 2018)_
- Does not include competition (demand:supply or the mass effect) _(Shen, 1998; Merlin and Hu, 2018)_

--

.pull-right[
]

???
Accessibility:

- A way to relatively quantify the access that a spatial unit: a product of land-use and the population's means of transportation

- In other words, it measures the amount of _potential interaction_ a population has with opportunities _(Hansen, 1959)_

- It connects land-use and the population's means of transportation: the population are the people that live within the region and the opportunities are the destinations of interest to the population or of the transportation network itself.

-**Hansen**-type accessibility - the sum of opportunities multiplied by a function which measures of the cost of moving between i and j.

- this measure suffers from interpretability issues - what is travel-cost adjusted opportunities?

- it also doesn't consider competition, which has been shown to reflect empirical accessibility to competitive-opportunities worse than competitive accessibility measures.

---
count: false
## Multi-modal accessibility measures

.pull-left[
#### Competitive accessibility (Shen-type) - no proportional allocation
Shen (1998) modified Hansen-type accessibility; popularized as the **two-step floating catchment approach (2SFCA)** by _(Luo and Wang, 2003)_

$$S_i^m = \sum_j \frac{O_jf(c_{ij})^m}{\sum_m D_j^m}$$

with:
$$D_j^m = \sum_i P_{i}^m f(c_{ij})^m$$
- $S_{i}^m$ is Shen-type accessibility for each mode $m$
- $D_j^m$ is the demand at $j$ from $P_{i}$ using each mode $m$

]

.pull-right[

It considers multi-model competition following the generalized:
$$S_i = \sum_m \biggl(\frac{P_i^m}{P_i}\biggr) \sum_j \frac{O_jf(c_{ij})^m}{\sum_m \sum_i P_{i}^m f(c_{ij})^m}$$

E.g., used by _Tao et al., 2020_.

However, the resulting value is an accessibility _score_ so interpretation is obfuscated _(Soukhov et al., 2023)_

]

???
- e.g., used by _Tao et al., 2020_ to calculate the $S_i^m$ for $m$ modes public transit and car to employment as well as the generalized accessibility $S_i$

- **Shen**-type accessibility - the sum of travel-cost adjusted opportunities (supply) divided by travel-cost adjusted demand! It considers travel-cost from both the opportunities perspective and the population side. It is done in 2 steps for the 2SFCA: "The first one computes the ratio between supply and demand of each health facility; the second step computes accessibility as the weighted sum of these ratios using a distance-decay function"

It should be mentioned that Weibull (1976) was the first to arrive to this same formula from axiomatic approach. Joseph and Bantock (1982) then applied it to health care access.

---
### Impedance functions - quantifying how subsets of people travel

.left-column[

$f(c_{ij})^m$ is a function of the cost of moving between $i$ and $j$ for modes $m$.

]

.right-column[]

---
count: false
### Impedance functions - quantifying how subsets of people travel

.left-column[

$f(c_{ij})^m$ is a function of the cost of moving between $i$ and $j$ for modes $m$.

$f(c_{ij})^{t.behav.1} = \exp(-0.1\cdot c_{ij})$

]

.right-column[
```{r comparison-impedance-functions-synthetic-example1, echo=FALSE, warning=FALSE, fig.width=9, fig.height=6}
data.frame(tt = seq(0, 100, 1)) %>%
  mutate(f1 = exp(-0.1 * tt)) %>%
  pivot_longer(cols = -tt,
               names_to = "Function",
               values_to = "f") %>%
  mutate(Function = case_when(Function == "f1" ~ "beta = 0.1")) %>%
  ggplot() +
  xlab(expression(c[ij]~"(travel time in minutes)")) +
  ylab(expression('f(c'[ij]*')')) +
  geom_line(aes(x = tt, 
                y = f,
                color = Function), size=1) +
  scale_color_manual(values = c("dark green"), label = c(expression("exp(-0.1 c"[ij]*")")))+
  scale_x_continuous(breaks = seq(0, 80, 20))+
  theme_minimal()+
  theme(axis.title = element_text(size=18),
        axis.text = element_text(size=16),
        legend.position = c(0.85, 0.9),
        legend.title = element_blank(),
        legend.text = element_text(size=18))
```
]

---
count: false
### Impedance functions - quantifying how subsets of people travel

.left-column[

$f(c_{ij})^m$ is a function of the cost of moving between $i$ and $j$ for modes $m$.

$f(c_{ij})^{\text{t.behav.1}} = \exp(-0.1\cdot c_{ij})$

$f(c_{ij})^{\text{t.behav.2}} = \exp(-0.6\cdot c_{ij})$

]

.right-column[
```{r comparison-impedance-functions-synthetic-example2, warning=FALSE, echo=FALSE, fig.width=9, fig.height=6}
data.frame(tt = seq(0, 100, 1)) %>%
  mutate(f1 = exp(-0.1 * tt),
         f2 = exp(-0.6 * tt)) %>%
  pivot_longer(cols = -tt,
               names_to = "Function",
               values_to = "f") %>%
  mutate(Function = case_when(Function == "f1" ~ "beta = 0.1",
                              Function == "f2" ~ "beta = 0.6")) %>%
  ggplot() +
  xlab(expression(c[ij]~"(travel time in minutes)")) +
  ylab(expression('f(c'[ij]*')')) +
  geom_line(aes(x = tt, 
                y = f,
                color = Function), size=1) +
  scale_color_manual(values = c("dark green", "#9999CC"), label = c(expression("exp(-0.1 c"[ij]*")"), expression("exp(-0.6 c"[ij]*")")))+
  scale_x_continuous(breaks = seq(0, 80, 20))+
  theme_minimal()+
  theme(axis.title = element_text(size=18),
        axis.text = element_text(size=16),
        legend.position = c(0.85, 0.9),
        legend.title = element_blank(),
        legend.text = element_text(size=18))
```
]

---
### Simulated example - Hansen-type accessibility (no competition or proportional allocation consideration)

.pull-left[

```{r, toy-example1, echo=FALSE}
knitr::include_graphics("Presentations/toy-example.png")
```

]

--

.pull-right[

$A_i^m = \sum_jO_jf(c_{ij})^m$

Impedance function: $f(c_{ij})^{\text{car}} = f(c_{ij})^{\text{pt}} =exp(-0.1 \cdot c_{ij})$;

```{r, echo=FALSE, message=FALSE}
#set b and calculate impedance
b_x <- 0.1 #the faster mode, car
b_z <- 0.1 #the original slower mode, transit

od_tt_m <- od_tt_m |>
  mutate(f = case_when(m=="car" ~ exp(-b_x * tt),
                       m=="pt" ~ exp(-b_z * tt)))

A <- od_tt_m |>
  group_by(i,m) |>
  summarize(A = sum(opp * f),
            .groups = "drop")

A %>% kable(format = "html",
        booktabs = TRUE,
        digits = 2)
```
]

???
- explain the map; blue and orange dots are origins and employment centers respectively. Travel time within each center is 15 mins, travel time between sub and urban is 30 mins and travel time to the satellite town from sub / urban is 100 mins.

- then explain the accessibility formula; the impedance function is a negative exponential (Wilson 1971), beta 0.1, like in the example. Same for both modes.

- Accessibility can be interpreted as a score, such as, population center at A and B has the same _potential_ access to 27,292 jobs using transit and 44,999.80 jobs using car. It is difficult to understand the meaning of the value other than A and B have a higher score than C.

- We know that the size of population matters. They are different in both A and B -- the population that uses transit is also different in both centers -- how does these two levels of differing demand size impact accessibility to opportunities? We can't use hansen-type measure without further processing.
---

### Synthetic example - Shen-type accessibility (no proportional allocation consideration)

.pull-left[
```{r, echo=FALSE}
knitr::include_graphics("Presentations/toy-example.png")
```
]

--

.pull-right[
$A_i^m$ and $S_i^m = \sum_j \frac{O_j f(c_{ij})^m}{\sum_m \sum_i P_{i}^m f(c_{ij})^m}$

```{r, echo=FALSE}
#set b and calculate impedance
LOS_j <- od_tt_m |>
  group_by(j,m) |>
  summarize(opp = mean(opp),
            D_j = sum(pop * f),
            .groups = "drop") |>
  group_by(j) |>
  summarize(LOS_j = first(opp)/sum(D_j),
            opp = mean(opp),
            D_j = mean(D_j))

S_m <- od_tt_m %>%
  left_join(LOS_j |> 
              dplyr::select(j, LOS_j, opp, D_j),
            by = c("j")) %>%
  group_by(i,m) |>
  summarize(S_i_m = sum(LOS_j * f),
            pop = mean(pop),
            .groups = "drop") |>
  rename("S" = "S_i_m")

S_m <- A %>% 
  left_join(S_m, by=c("i","m"))

```
```{r, echo=FALSE, message=FALSE}
knitr::kable(S_m %>% dplyr::select(-pop),
             format = "html",
             booktabs = TRUE,
             digits = 2)
```
]

---
count: false
### Synthetic example - Shen-type accessibility (no proportional allocation consideration)

.pull-left[
```{r, echo=FALSE}
knitr::include_graphics("Presentations/toy-example.png")
```
]

--

.pull-right[
$S_i = \sum_m \biggl(\frac{P_i^m}{P_i}\biggr) \sum_j \frac{O_j f(c_{ij})^m}{\sum_m \sum_i P_{i}^m f(c_{ij})^m}$:
```{r, echo=FALSE, message=FALSE}
S <- S_m %>%
  group_by(i) |>
  summarize(S = sum(pop*S)/sum(pop),
            .groups = "drop")

knitr::kable(S,
             format = "html",
             booktabs = TRUE,
             digits = 2)
```

]

???

Now let's look at Shen's accessibility, same impedance function.

First and foremost, we see that Shen's accessibility can still be interpreted as as score, like Hansen-style accessibility. But it can also be interpreted as travel-cost adjusted jobs per travel-cost adjusted capita (potential supply over potential demand) per mode.

It also accounts for competition between modes by weighted-sum of mode-specific population over the total population at that origin $i$.

We can do better, let's use proportional allocation (NOTE: WHY??????????)

---
### Spatial availability - competitive _and_ proportional allocation

.pull-left[

- Spatial availability per origin and mode:

$V^m_{i} = \sum_{j=1}^J O_j\ F^{t,m}_{ij}$

- The proportional allocation factor combines two aspects of the problem, population and cost:

$F^{t,m}_{ij} = \frac{F^{pm}_{i} \cdot F^{cm}_{ij}}{\sum_{m=1}^M \sum_{i=1}^N F^{pm}_{i} \cdot F^{cm}_{ij}}$

- Where $F^{p,m}_{i}$ is population allocation factor:

$F^{p,m}_{i} = \frac{P^m_{i}}{\sum_{i} P^m_{i}}$

- And the $F^{c,m}_{ij}$ is a cost allocation factor:

$F^{c,m}_{ij} = \frac{f(c_{ij})^m}{\sum_{i} f(c_{ij})^m}$
]

--

.pull-right[

- Spatial availability per origin:

$V_{i} = \sum_{m=1}^M V^m_{i}$

- Considers **competition** and **proportionally allocates opportunities** (singly-constrained: $O = \sum_{j=1}^J O_j$)
]
???

- So let me introduce spatial availability, it consists of proportional allocation factors and the opportunities. the population allocation factor is the allocation of opportunities, proportional to the population size. It is always a proportion between 0 to 1.

- The cost factor allocates opportunities proportionally based on the impedance function. The impedance function can be anything - gravity-based like Hansen, cumulative, etc. It is also always a proportion between 0 to 1.

- You multiple both factors together and by opportunities --> the number of opportunities are preserved! This is why you can just add up all the spatial availabilies per mode. It all sums up.

  -supply of population and demand for opportunities are both travel-cost and population adjusted
  - why singly-constrained? the sum of all accessibility values for i equals all the opportunities for i to j (like in a singly constrained gravity model - it is attraction constrained)

---
### Spatial availability - competitive _and_ proportional allocation

.pull-left[
```{r, echo=FALSE}
knitr::include_graphics("Presentations/toy-example.png")
```
]

--

.pull-right[

$V^m_{i} = \sum_{j=1}^J O_j\ F^{t,m}_{ij}$ and $\frac{V^m_{i}}{P_{i}^m}$

```{r, echo=FALSE, message=FALSE}
od_tt_m <- od_tt_m %>%
  mutate(catch = 1) %>%
  sp_avail_detailed(o_id = i, #spatial availability
                    d_id = j, 
                    pop = pop,
                    opp = opp,
                    r = catch,
                    f = f, 
                    alpha = 1)

V_m <- od_tt_m %>%
  group_by(i,m) %>%
  summarize(V = sum(V_ij),
            v = V/first(pop))
```
```{r, echo=FALSE, message=FALSE}
knitr::kable(V_m,
             format = "html",
             booktabs = TRUE,
             digits = 2)
```

]

---
count: false
### Spatial availability - competitive _and_ proportional allocation

.pull-left[
```{r, echo=FALSE}
knitr::include_graphics("Presentations/toy-example.png")
```
]

--

.pull-right[
$S_i$ (eqv. to $\frac{V_i}{P_i}$) and $V_{i} = \sum_{m=1}^M V^m_{i}$

```{r, echo=FALSE, message=FALSE}
V <- od_tt_m %>%
  group_by(i) %>%
  summarize(V = sum(V_ij))

V <- V %>% left_join(S, by=c("i")) |> transmute(S,V)

knitr::kable(V,
             format = "html",
             booktabs = TRUE,
             digits = 2)
```

]

--

- And finally, if you divide spatial availability by the population, it can be simply normalized to be opportunities available per capita. This is equivalent to Shen-style accessibility. Both competitive and singly-constrained

???
XXX

---
### Madrid's low emission zones - a policy of modal discrimination

.pull-left[
```{r, echo=FALSE, out.width='80%', fig.align='center'}
knitr::include_graphics("Presentations/madrid-region.png")
```
]

.pull-right[
```{r, echo=FALSE, out.width='50%', fig.align='center'}
knitr::include_graphics("Presentations/city-centre.png")
```

- November 2018 - LEZ Centro
- 2023 - LEZ within M-30
- 2024 - LEZ within Madrid Municipality 
]

???
A NEEED for multi-modal analysis methods that are interpretive and comparative -- proportional allocation achieves that!
---
### Multi-modal home-to-work accessibility in Madrid: 2018 travel survey

.pull-left[
```{r, echo=FALSE, out.width='80%',fig.align='center'}
knitr::include_graphics("Presentations/SA_i_V_zn1259_plot.png")
```
Spatial Availability $V_i$ at resolution Zona 1259
]

.pull-right[
```{r, echo=FALSE, out.width='80%', fig.align='center'}
knitr::include_graphics("Presentations/SA_i_V_zn208_plot.png")
```
Spatial Availability $V_i$ at resolution Zona 208
]

???
- a need for multi-modal analysis to assess impacts of this geographic. The issue of MUAP.
---
class: center
### Multi-modal home-to-work accessibility in Madrid: 2018 travel survey
```{r, echo=FALSE, out.width='45%',fig.align='center'}
knitr::include_graphics("Presentations/SA_im_V_zn1259_plot.png")
```
Spatial Availability $V_i^m$ at resolution Zona 1259

???
- With spatial availability you can compare between modes -- however still the issue of MUAP.

---
class: center
### Multi-modal home-to-work accessibility in Madrid: 2018 travel survey
```{r, echo=FALSE, out.width='70%',fig.align='center'}
knitr::include_graphics("Presentations/SA_im_vv_zn1259_plot.png")
```
Spatial Availability by $m$ population ( $v_i^m$ ) at resolution Zona 1259

???
- Can also benchmark between modes: the spatial avail per mode-using population per $i$ -- however still the issue of MUAP
---
### Spatial availability: a way forward in multimodal transport and land-use planning

- Data-driven transport accessibility policy scenario creation

- Modifiable unit areal problem (MUAP) - need for logically consistent accessibility measurement method
  - Calculated travel time tessellation

- Questions of equity: which neighbourhoods and people _within_ neighbourhoods are not meeting the mark? 
  - Who will be most impacted?

???
- policy scenario creation
- MUAP exists, seen in Madrid transport data, the need for simulated calculation and consistent aggregation of accessibility (through proportional allocation)
- Quantitatively, this method can be used to more interpetble answer these questions of equity. Socio-economic impact, GHG emissions, across sub-sets of population (multi-modal users included).

---
count: false
class: center, middle, inverse

## Acknowledgments

```{r, echo=FALSE, out.width="20%"}
knitr::include_graphics("Presentations/MJ_logo.png")
```

```{r, echo=FALSE, out.width="30%"}
knitr::include_graphics("Presentations/mcm-sci-sees_left-col.png")
```

### Any questions? Remarks?

#### Anastasia Soukhov, **soukhoa@mcmaster.ca**

#### Spatial availability paper:
```{r, echo=FALSE, out.width="10%"}
knitr::include_graphics("Presentations/spatial-availability.png")
```

### 2023-06-10 Multi-mode-acc-methods.qmd
CIT-Presentation draft!

@@
 HW_walk_lnorm <- lnorm_
-
 ```
@@
            mode) %>%
-  summarize(workers = n(), #we should be summing the number of workers by the weight but we don't have a weight... so for now, 1 row = 1 trip. 
+  summarize(workers = sum(ELE_G_POND_ESC2), #we should be summing the number of workers by the weight but we don't have a weight... so for now, 1 row = 1 trip. 
             .groups = "drop")%>%
@@
   group_by(VDESZT1259) %>%
-  summarize(jobs = n(),
+  summarize(jobs = sum(ELE_G_POND_ESC2),
             .groups = "drop") %>%
@@
                        mode == "walk" ~ dlnorm(tt, 
-                                               HW_walk_lnorm$estimate["location"], 
-                                               HW_walk_lnorm$estimate["scale"])))
+                                               HW_walk_lnorm$estimate["meanlog"], 
+                                               HW_walk_lnorm$estimate["sdlog"])))
 ```
@@
 sum(jobs$jobs)
-#sum(OD_SA$tt)
-```
-Spatial availability by mode, we can see that car mode captures 71% of the spatial availability but only accounts for XXXX of the tt_ijm !!:
+sum(OD_SA$tt)
+```
+Spatial availability by mode, we can see that car mode captures 43% of the spatial availability but only accounts for 30% of the tt_ijm !!:
 ```{r}
@@
   summarize(V_m = sum(SA_ij),
-            V_m_perc = sum(SA_ij)/27795,
-            tt_m = sum(tt)/780315.1)
+            V_m_perc = sum(SA_ij)/847574,
+            tt_m = sum(tt)/435061.9)
 ```
@@
 SA_im <- OD_SA %>%
-  group_by(VORI_CD_ZT208, 
+  group_by(VORIZT1259, 
            mode) %>%
@@
 population <- workers %>%
-  group_by(CD_ZT208) %>%
+  group_by(ZT1259) %>%
   summarize(population = sum(workers))
@@
   left_join(population,
-            by = c("VORI_CD_ZT208" = "CD_ZT208"))
+            by = c("VORIZT1259" = "ZT1259"))
 sum(SA_i$V_i)
@@
   left_join(population,
-            by = c("VORI_CD_ZT208" = "CD_ZT208"))
+            by = c("VORIZT1259" = "ZT1259"))
 sum(SA_im$V_im)
@@
 SA_i <- SA_i %>%
-  left_join(zn208 %>%
-              dplyr::select(CD_ZT208),
-            by =c("VORI_CD_ZT208" = "CD_ZT208")) %>%
+  left_join(zn1259 %>%
+              dplyr::select(ZT1259),
+            by =c("VORIZT1259" = "ZT1259")) %>%
   st_as_sf()
@@
 SA_im <- SA_im %>%
-  left_join(zn208 %>%
-              dplyr::select(CD_ZT208),
-            by = c("VORI_CD_ZT208" = "CD_ZT208")) %>%
+  left_join(zn1259 %>%
+              dplyr::select(ZT1259),
+            by = c("VORIZT1259" = "ZT1259")) %>%
   st_as_sf()
@@
 
-Figure out which CD_ZT208 correspond to Madrid - we see that CD_ZT208 1 through 59 do so.
-```{r}
-t <- zn_u %>% dplyr::select(CD_ZT208, Municipality) %>% group_by(CD_ZT208,Municipality) %>% summarize() %>% filter(Municipality == "Madrid")
-t
-```
-
 ```{r create-bounding-boxes}
-zn208 <- zn208 %>% mutate(MM = ifelse(str_starts(ZT208,"Madrid"), 1, 0)) %>% filter(MM == 1)
-
-#create the bounding box for the Toronto region + a little extra
-bbox_new1 <- st_bbox(zn208 %>%filter(MM == 1)) # current bounding box
+zn1259_mad <- zn1259 %>% filter(Municipality == "Madrid")
+
+#create the bounding box for the Madrid city + a little extra
+bbox_new1 <- st_bbox(zn1259_mad) # current bounding box
 
@@
 sf::sf_use_s2(FALSE) #this terns off the spherical geometry component in the st_union() function - this assumes the sf is just planar. Thats' fine for our purposes
-Madrid_muni_border <- zn208 %>%
-  filter(MM == 1) %>%
+Madrid_muni_border <- zn1259_mad %>%
   st_union()
@@
   scale_fill_distiller(palette = "YlGn", #legend scale bar
-                       name = expression(v["i"]),
+                       name = expression(V["i"]),
                        na.value = "grey90",

### 2023-06-12 CIT-Presentation.Rmd
Draft CIT-Presentation 2 (resolved some figure formating issues)

@@
     nature:
-      slideNumberFormat: "%current% / 12" 
+      slideNumberFormat: "%current% / 14" 
       highlightStyle: github
@@
                    label = time),
-               size = 3,
+               size = 3.5,
              curvature = 0.20,
@@
                 label=id), 
-            size=3) +
+            size=4) +
   # Create table
@@
 -   $O_j$ is the number of opportunities at location $j$
--   $f(\cdot)$ is an impedance function of $c_{ij}$ (cost of moving between $i$ and $j$) for each mode $m$
+-   $f(\cdot)$ is an impedance function of $c_{ij}$ (cost of moving between $i$ and $j$) for each mode $m$; if binary, the measure if _cumulative opportunities_
 
-**E.g.**, used in _(Tahmasbi et al., 2019)_.
+]
+
+-- 
+.pull-right[
+
+**E.g.**, used in _(Tahmasbi et al., 2019)_ to calculate $A_i^m$ (per mode). Variations on how it is combined across modes.
 
@@
 
-Does not include competition (demand:supply or the mass effect) _(Shen, 1998; Merlin and Hu, 2018)_.
+Does not include competition (demand:supply, mass effect) _(Shen, 1998; Merlin and Hu, 2018)_.
 
-]
-
-???
-
-Interpretability issues are not resolved when considering multiple modes.
-
-**e.g.**, used in _(Tahmasbi et al., 2019)_ for quantify access $A_i$ for $m$ walking, public transit, and private car modes to retail store destinations (different $f^m(c_{ij}^m)$ for each $m$)
-- Suffers from the same interpretability issues _(Handy and Niemeier, 1997; Miller et al., 2018)_
-- Does not include competition (demand:supply or the mass effect) _(Shen, 1998; Merlin and Hu, 2018)_
-
---
-
-.pull-right[
 ]
@@
 
+Interpretability issues are not resolved when considering multiple modes.
+
+- Used in _(Tahmasbi et al., 2019)_ to quantify access $A_i$ for $m$ walking, public transit, and private car modes to retail store destinations (different $f^m(c_{ij}^m)$ for each $m$).
+Variations on how it is combined across modes (if at all) for general accessibility.
+It varies so widely because interpretation is inconsistent.
+- Suffers from the same interpretability issues _(Handy and Niemeier, 1997; Miller et al., 2018)_
+- Does not include competition (demand:supply or the mass effect) _(Shen, 1998; Merlin and Hu, 2018)_
+
 ---
-count: false
 ## Multi-modal accessibility measures
@@
 #### Competitive accessibility (Shen-type) - no proportional allocation
-Shen (1998) modified Hansen-type accessibility; popularized as the **two-step floating catchment approach (2SFCA)** by _(Luo and Wang, 2003)_
+Shen (1998) modified Hansen-type accessibility; popularized as the **two-step floating catchment area (2SFCA) method** by _(Luo and Wang, 2003)_
 
@@
 
-It considers multi-model competition following the generalized:
+Generalized multi-model competition:
 $$S_i = \sum_m \biggl(\frac{P_i^m}{P_i}\biggr) \sum_j \frac{O_jf(c_{ij})^m}{\sum_m \sum_i P_{i}^m f(c_{ij})^m}$$
@@
 
-However, the resulting value is an accessibility _score_ so interpretation is obfuscated _(Soukhov et al., 2023)_
+However, the resulting value is an accessibility _score_ so interpretation is obfuscated _(Soukhov et al., 2023)_.
 
@@
 .right-column[
-```{r comparison-impedance-functions-synthetic-example1, echo=FALSE, warning=FALSE, fig.width=9, fig.height=6}
+```{r comparison-impedance-functions-synthetic-example1, echo=FALSE, warning=FALSE, fig.width=8, fig.height=6}
 data.frame(tt = seq(0, 100, 1)) %>%
@@
 .right-column[
-```{r comparison-impedance-functions-synthetic-example2, warning=FALSE, echo=FALSE, fig.width=9, fig.height=6}
+```{r comparison-impedance-functions-synthetic-example2, warning=FALSE, echo=FALSE, fig.width=8, fig.height=6}
 data.frame(tt = seq(0, 100, 1)) %>%
@@
 
+---
+count: false
+### Impedance functions - quantifying how subsets of people travel
+
+.left-column[
+
+$f(c_{ij})^m$ is a function of the cost of moving between $i$ and $j$ for modes $m$.
+
+$f(c_{ij})^{\text{t.behav.1}} = \exp(-0.1\cdot c_{ij})$
+
+$f(c_{ij})^{\text{t.behav.2}} = \exp(-0.6\cdot c_{ij})$
+
+$f(c_{ij}) = 1$ if $c_{ij} \leq 30$ else 
+  $f(c_{ij}) = 0$
+
+]
+
+.right-column[
+```{r comparison-impedance-functions-synthetic-example3, warning=FALSE, echo=FALSE, fig.width=8, fig.height=6}
+data.frame(tt = seq(0, 100, 1)) %>%
+  mutate(f1 = exp(-0.1 * tt),
+         f2 = exp(-0.6 * tt),
+         f3 = ifelse(tt < 30, 1, 0)) %>%
+  pivot_longer(cols = -tt,
+               names_to = "Function",
+               values_to = "f") %>%
+  mutate(Function = case_when(Function == "f1" ~ "beta = 0.1",
+                              Function == "f2" ~ "beta = 0.6",
+                              Function == "f3" ~ "tt <= 30")) %>%
+  ggplot() +
+  xlab(expression(c[ij]~"(travel time in minutes)")) +
+  ylab(expression('f(c'[ij]*')')) +
+  geom_line(aes(x = tt, 
+                y = f,
+                color = Function), size=1) +
+  scale_color_manual(values = c("dark green", "#9999CC", "#CC6666"), label = c(expression("exp(-0.1 c"[ij]*")"), expression("exp(-0.6 c"[ij]*")"), expression(c[ij]~"< 30 min")))+
+  scale_x_continuous(breaks = seq(0, 80, 20))+
+  theme_minimal()+
+  theme(axis.title = element_text(size=18),
+        axis.text = element_text(size=16),
+        legend.position = c(0.85, 0.9),
+        legend.title = element_blank(),
+        legend.text = element_text(size=18))
+```
+]
 ---
@@
 ```
+
 ]
@@
 .pull-right[
-$S_i = \sum_m \biggl(\frac{P_i^m}{P_i}\biggr) \sum_j \frac{O_j f(c_{ij})^m}{\sum_m \sum_i P_{i}^m f(c_{ij})^m}$:
+$S_i = \sum_m \biggl(\frac{P_i^m}{P_i}\biggr) S_i^m$
 ```{r, echo=FALSE, message=FALSE}
@@
 
-- Spatial availability per origin and mode:
+- Spatial availability considering multi-modes:
 
-$V^m_{i} = \sum_{j=1}^J O_j\ F^{t,m}_{ij}$
+$V^m_{i} = \sum_{j}^J O_j\ F^{t,m}_{ij}$
 
@@
 
-- Spatial availability per origin:
+- Total spatial availability: $V_{i} = \sum_{m}^M V^m_{i}$
+  - Considers **competition** and **proportionally allocates opportunities** (singly-constrained: $O = \sum_{j} O_j$)
 
-$V_{i} = \sum_{m=1}^M V^m_{i}$
+- Total spatial availability per capita: $v_{i} = \sum_{m} \frac{V_i^m}{P_{i^m}}$
+  - Equivalent to Shen-style accessibility
 
-- Considers **competition** and **proportionally allocates opportunities** (singly-constrained: $O = \sum_{j=1}^J O_j$)
 ]
+
 ???
-
 - So let me introduce spatial availability, it consists of proportional allocation factors and the opportunities.
@@
 
-$V^m_{i} = \sum_{j=1}^J O_j\ F^{t,m}_{ij}$ and $\frac{V^m_{i}}{P_{i}^m}$
+$V^m_{i} = \sum_{j} O_j\ F^{t,m}_{ij}$ and $v^m_{i} = \frac{V^m_{i}}{P_{i}^m}$
 
@@
 
+???
+
+- And finally, if you divide spatial availability by the population, it can be simply normalized to be opportunities available per capita.
+This is equivalent to Shen-style accessibility.
+Both competitive and singly-constrained
+
 ---
@@
 .pull-right[
-$S_i$ (eqv. to $\frac{V_i}{P_i}$) and $V_{i} = \sum_{m=1}^M V^m_{i}$
+$S_i$ (eqv. to $v_i$) and $V_{i} = \sum_{m=1}^M V^m_{i}$
 
@@
 ]
-
---
-
-- And finally, if you divide spatial availability by the population, it can be simply normalized to be opportunities available per capita.
-This is equivalent to Shen-style accessibility.
-Both competitive and singly-constrained
 
@@
 .pull-left[
-```{r, echo=FALSE, out.width='80%',fig.align='center'}
+```{r, echo=FALSE, out.width='120%',fig.align='center'}
 knitr::include_graphics("Presentations/SA_i_V_zn1259_plot.png")
@@
 .pull-right[
-```{r, echo=FALSE, out.width='80%', fig.align='center'}
+```{r, echo=FALSE, out.width='120%', fig.align='center'}
 knitr::include_graphics("Presentations/SA_i_V_zn208_plot.png")
@@
 ### Multi-modal home-to-work accessibility in Madrid: 2018 travel survey
-```{r, echo=FALSE, out.width='70%',fig.align='center'}
+```{r, echo=FALSE, out.width='45%',fig.align='center'}
 knitr::include_graphics("Presentations/SA_im_vv_zn1259_plot.png")
@@
 ```
+
+---
+class: center, middle, inverse
+
+#Extra slides
+
+---
+### Proof: Spatial availability cancels out into Shen-style accessibility
+
+Population allocation factor:
+$F^p_{ij} = \frac{P_{i\in r}^\alpha}{\sum_{i}^K P_{i\in r}^\alpha}$
+
+$F^p_{A} = \frac{P_{A}^\alpha}{P_{A}^\alpha + P_{B}^\alpha + P_{C}^\alpha}$
+
+Cost allocation factor:
+$F^c_{ij} = \frac{f(c_{ij})}{\sum_{i=A}^K f(c_{ij})}$
+
+$F^c_{A1} = \frac{f(c_{A1})}{f(c_{A1})+f(c_{B1})+f(c_{C1})}$
+$F^c_{B1} = \frac{f(c_{A2})}{f(c_{A2})+f(c_{B2})+f(c_{C2})}$
+$F^c_{C1} = \frac{f(c_{A3})}{f(c_{A3})+f(c_{B3})+f(c_{C3})}$
+
+Now let's put it together with P, and see how the denominators end up cancelling out:
+$v_{i} = \sum_{j}\frac{O_j}{P_{i\in r}^\alpha}\frac{\frac{P_{i\in r}^\alpha}{\sum_{i}^K P_{i\in r}^\alpha} \cdot \frac{f(c_{ij})}{\sum_{i}^K f(c_{ij})}}{\sum_{i}^K \frac{P_{i\in r}^\alpha}{\sum_{i}^K P_{i\in r}^\alpha} \cdot \frac{f(c_{ij})}{\sum_{i}^K f(c_{ij})}}$
+
+???
+
+- we presented this formula to a few accessibility academics and because the way we explained it - from the perspective of proportional allocation - made it appear to be different than shen's.
+It was a surprise to see the denominators cancel out to equal shen's!
+
+---
+
+$v_{A} = \frac{O_1}{P_{A}^\alpha}(\frac{\frac{P_{A}^\alpha}{P_{A}^\alpha+P_{B}^\alpha+P_{C}^\alpha} \cdot \frac{f(c_{A1})}{f(c_{A1})+f(c_{B1})+f(c_{C1})}}{\frac{P_{A}^\alpha}{P_{A}^\alpha+P_{B}^\alpha+P_{C}^\alpha} \cdot \frac{f(c_{A1})}{f(c_{A1})+f(c_{B1})+f(c_{C1})} + \frac{P_{A}^\alpha}{P_{A}^\alpha+P_{B}^\alpha+P_{C}^\alpha} \cdot \frac{f(c_{B1})}{f(c_{A1})+f(c_{B1})+f(c_{C1})}+ \frac{P_{A}^\alpha}{P_{A}^\alpha+P_{B}^\alpha+P_{C}^\alpha} \cdot \frac{f(c_{C1})}{f(c_{A1})+f(c_{B1})+f(c_{C1})}}) +$
+
+$\frac{O_2}{P_{A}^\alpha}(\frac{\frac{P_{A}^\alpha}{P_{A}^\alpha+P_{B}^\alpha+P_{C}^\alpha} \cdot \frac{f(c_{A2})}{f(c_{A2})+f(c_{B2})+f(c_{C2})}}{\frac{P_{A}^\alpha}{P_{A}^\alpha+P_{B}^\alpha+P_{C}^\alpha} \cdot \frac{f(c_{A2})}{f(c_{A2})+f(c_{B2})+f(c_{C2})} + \frac{P_{A}^\alpha}{P_{A}^\alpha+P_{B}^\alpha+P_{C}^\alpha} \cdot \frac{f(c_{B2})}{f(c_{A2})+f(c_{B2})+f(c_{C2})}+\frac{P_{A}^\alpha}{P_{A}^\alpha+P_{B}^\alpha+P_{C}^\alpha} \cdot \frac{f(c_{C2})}{f(c_{A2})+f(c_{B2})+f(c_{C2})}} )+$
+
+$\frac{O_3}{P_{A}^\alpha}(\frac{\frac{P_{A}^\alpha}{P_{A}^\alpha+P_{B}^\alpha+P_{C}^\alpha} \cdot \frac{f(c_{A3})}{f(c_{A3})+f(c_{B3})+f(c_{C3})}}{\frac{P_{A}^\alpha}{P_{A}^\alpha+P_{B}^\alpha+P_{C}^\alpha} \cdot \frac{f(c_{A3})}{f(c_{A3})+f(c_{B3})+f(c_{C3})} + \frac{P_{A}^\alpha}{P_{A}^\alpha+P_{B}^\alpha+P_{C}^\alpha} \cdot \frac{f(c_{B3})}{f(c_{A3})+f(c_{B3})+f(c_{C3})}+\frac{P_{A}^\alpha}{P_{A}^\alpha+P_{B}^\alpha+P_{C}^\alpha} \cdot \frac{f(c_{C3})}{f(c_{A3})+f(c_{B3})+f(c_{C3})}} )$
+
+First, notice how the denominator on the denominator is the same across the summation? Let's simplify it:
+
+$v_{A} = \frac{O_1}{P_{A}^\alpha}(\frac{\frac{P_{A}^\alpha}{P_{A}^\alpha+P_{B}^\alpha+P_{C}^\alpha} \cdot \frac{f(c_{A1})}{f(c_{A1})+f(c_{B1})+f(c_{C1})}}{\frac{P_{A}^\alpha \cdot f(c_{A1}) + P_{A}^\alpha \cdot f(c_{B1}) + P_{A}^\alpha \cdot f(c_{C1})}{(P_{A}^\alpha+P_{B}^\alpha+P_{C}^\alpha) \cdot (f(c_{A1})+f(c_{B1})+f(c_{C1}))}}) +$
+$\frac{O_2}{P_{A}^\alpha}(\frac{\frac{P_{A}^\alpha}{P_{A}^\alpha+P_{B}^\alpha+P_{C}^\alpha} \cdot \frac{f(c_{A2})}{f(c_{A2})+f(c_{B2})+f(c_{C2})}}{\frac{P_{A}^\alpha \cdot f(c_{A2}) + P_{A}^\alpha \cdot f(c_{B2}) + P_{A}^\alpha \cdot f(c_{C2})}{(P_{A}^\alpha+P_{B}^\alpha+P_{C}^\alpha) \cdot (f(c_{A2})+f(c_{B2})+f(c_{C2}))}}) +$
+$\frac{O_3}{P_{A}^\alpha}(\frac{\frac{P_{A}^\alpha}{P_{A}^\alpha+P_{B}^\alpha+P_{C}^\alpha} \cdot \frac{f(c_{A3})}{f(c_{A3})+f(c_{B3})+f(c_{C3})}}{\frac{P_{A}^\alpha \cdot f(c_{A3}) + P_{A}^\alpha \cdot f(c_{B3}) + P_{A}^\alpha \cdot f(c_{C3})}{(P_{A}^\alpha+P_{B}^\alpha+P_{C}^\alpha) \cdot (f(c_{A3})+f(c_{B3})+f(c_{C3}))}} )$
+
+See how the denominator of the denominator is the same as the denominator of the numerator's denominator for each J (J=1, J=2, and J=3)? Let's cancel those out and simplify:
+
+$v_{A} = \frac{O_1}{P_{A}^\alpha}(\frac{P_{A}^\alpha \cdot f(c_{A1})}{P_{A}^\alpha \cdot f(c_{A1}) + P_{A}^\alpha \cdot f(c_{B1}) + P_{A}^\alpha \cdot f(c_{C1})} +$
+$\frac{O_2}{P_{A}^\alpha}\frac{P_{A}^\alpha \cdot f(c_{A2})}{P_{A}^\alpha \cdot f(c_{A2}) + P_{A}^\alpha \cdot f(c_{B2}) + P_{A}^\alpha \cdot f(c_{C2})} +$
+$\frac{O_3}{P_{A}^\alpha}\frac{P_{A}^\alpha \cdot f(c_{A3})}{P_{A}^\alpha \cdot f(c_{A3}) + P_{A}^\alpha \cdot f(c_{B3}) + P_{A}^\alpha \cdot f(c_{C3})} )$
+
+---
+
+Next, see how we can cancel out the $P_{A}^\alpha$? Let's do that.
+
+$v_{A} = O_1(\frac{f(c_{A1})}{P_{A}^\alpha \cdot f(c_{A1}) + P_{B}^\alpha \cdot f(c_{B1}) + P_{C}^\alpha \cdot f(c_{C1})} + O_2\frac{f(c_{A2})}{P_{A}^\alpha \cdot f(c_{A2}) + P_{B}^\alpha \cdot f(c_{B2}) + P_{C}^\alpha \cdot f(c_{C2})} + O_3\frac{f(c_{A3})}{P_{A}^\alpha \cdot f(c_{A3}) + P_{B}^\alpha \cdot f(c_{B3}) + P_{C}^\alpha \cdot f(c_{C3})} )$
+
+```{r, eval=FALSE, include=FALSE}
+#Let's check to see if it manually adds up to what shen's is
+b<-0
+r1 <- 100000*exp(-b*15)/(50000*exp(-b*15) + 150000*exp(-b*30) + 10000*exp(-b*100))
+r2 <-100000*exp(-b*30)/(50000*exp(-b*30) + 150000*exp(-b*15) + 10000*exp(-b*100))
+r3 <- 10000*exp(-b*100)/(50000*exp(-b*100) + 150000*exp(-b*100) + 10000*exp(-b*15))
+r1+r2+r3
+
+#this is shen's, the same!!!
+j1<-(100000*exp(-0.1*15))/(50000 * exp(-b * 15) + 150000 * exp(-b * 30) + 10000 * exp(-b * 100))
+j2<-(100000*exp(-0.1*30))/(50000 * exp(-b * 30) + 150000 * exp(-b * 15) + 10000 * exp(-b * 100))
+j3<-(10000*exp(-0.1*100))/(50000 * exp(-b * 100) + 150000 * exp(-b * 100) + 10000 * exp(-b * 15))
+j1 + j2 + j3
+```
+```{r, eval=FALSE, include=FALSE}
+(exp(-b*15)/((50000*exp(-b*15)) + (50000*exp(-b*30)) + (50000*exp(-b*100)))+ exp(-b*30)/((50000*exp(-b*15)) + (50000*exp(-b*30)) + (50000*exp(-b*100)))+ exp(-b*100)/((50000*exp(-b*15)) + (50000*exp(-b*30)) + (50000*exp(-b*100))))*100000
+```
+
+Shen's accessibility:
+
+$S_i = \sum_j \frac{O_jf(c_{ij})}{\sum_kP_kf(c_{jk})}$
+
+---

### 2023-06-12 Multi-mode-acc-methods.qmd
Draft CIT-Presentation 2 (resolved some figure formating issues)

@@
 #creating a list of zones that correspond to the zn208, which indicate if they are in a LEZ.
-LEZ_centro_list <- st_intersection(zn208, LEZ_centro) %>% 
-  st_drop_geometry() %>% 
-  dplyr::select(CD_ZT208) %>% 
-  unlist()
-
-LEZ_M30_list <- st_intersection(zn208, LEZ_M30) %>% 
-  st_drop_geometry() %>% 
-  dplyr::select(CD_ZT208) %>% 
-  unlist()
+# LEZ_centro_list <- st_intersection(zn208, LEZ_centro) %>% 
+#   st_drop_geometry() %>% 
+#   dplyr::select(CD_ZT208) %>% 
+#   unlist()
+
+# LEZ_M30_list <- st_intersection(zn208, LEZ_M30) %>% 
+#   st_drop_geometry() %>% 
+#   dplyr::select(CD_ZT208) %>% 
+#   unlist()
 
 #import travel times
-load("./data-prep/data-raw/intermed/ttresults/output_OD_bike.Rdata")
-load("./data-prep/data-raw/intermed/ttresults/output_OD_car.Rdata")
-load("./data-prep/data-raw/intermed/ttresults/output_OD_transit.Rdata")
-load("./data-prep/data-raw/intermed/ttresults/output_OD_walk.Rdata")
+load("./data-prep/data-raw/intermed/ttresults/calc_bikett.Rdata")
+load("./data-prep/data-raw/intermed/ttresults/calc_walktt.Rdata")
+load("./data-prep/data-raw/intermed/ttresults/calc_transittt.Rdata")
+load("./data-prep/data-raw/intermed/ttresults/calc_cartt.Rdata")
+```
+
+```{r travel-data-calculated-filtered}
+zn1259_mad <- zn1259 %>% filter(Municipality == "Madrid")
+zn1259_mad_list <- zn1259_mad %>% st_drop_geometry() %>% dplyr::select(ZT1259)
+
+calc_walktt <- calc_walktt %>% filter((VORIZT1259 %in% zn1259_mad_list$ZT1259) & (VDESZT1259 %in% zn1259_mad_list$ZT1259))
+calc_cartt <- calc_cartt %>% filter((VORIZT1259 %in% zn1259_mad_list$ZT1259) & (VDESZT1259 %in% zn1259_mad_list$ZT1259))
+calc_transittt <- calc_transittt %>% filter((VORIZT1259 %in% zn1259_mad_list$ZT1259) & (VDESZT1259 %in% zn1259_mad_list$ZT1259))
+calc_bikett <- calc_bikett %>% filter((VORIZT1259 %in% zn1259_mad_list$ZT1259) & (VDESZT1259 %in% zn1259_mad_list$ZT1259))
 ```
@@
             ELE_G_POND_ESC2,
-            mode = "transit")
+            mode = "transit",
+            VORIZT1259,
+            VDESZT1259)
 
@@
             ELE_G_POND_ESC2,
-            mode = "car/motor")
+            mode = "car/motor",
+            VORIZT1259,
+            VDESZT1259)
 
@@
             ELE_G_POND_ESC2,
-            mode = "bike")
+            mode = "bike",
+            VORIZT1259,
+            VDESZT1259)
 
@@
             ELE_G_POND_ESC2,
-            mode = "walk")
+            mode = "walk",
+            VORIZT1259,
+            VDESZT1259)
+```
+
+```{r seeing-how-similar-calc-tt-is-to-survey}
+#Let's see how similar the calc tt is to the survey (only Madrid city travel times).
+We can see that transit is the most worrisome -- calculating travel times from some arbitrary point in the center of the TAZ is not very accurate.
+Also travel times for all active modes should also be calculated for trips between 60-120mins (not only 0-60mins).
+For now let's only work with the survey data.
+
+#walk
+compare_walktts <- calc_walktt %>% merge(walk_HW_alltt, by=c("VORIZT1259", "VDESZT1259")) %>% dplyr::select(-c("ELE_G_POND_ESC2", "mode"))
+compare_walktts %>% summary()
+
+#bike
+compare_biketts <- calc_bikett %>% merge(bike_HW_alltt, by=c("VORIZT1259", "VDESZT1259")) %>% dplyr::select(-c("ELE_G_POND_ESC2", "mode"))
+compare_biketts %>% summary()
+
+#car
+compare_cartts <- calc_cartt %>% merge(car_motor_HW_alltt, by=c("VORIZT1259", "VDESZT1259")) %>% dplyr::select(-c("ELE_G_POND_ESC2", "mode"))
+compare_cartts %>% summary()
+
+#transit 
+compare_transittts <- calc_transittt %>% merge(transit_HW_alltt, by=c("VORIZT1259", "VDESZT1259")) %>% dplyr::select(-c("ELE_G_POND_ESC2", "mode"))
+compare_transittts %>% summary()
 ```
@@
 ```{r fitting-impedance-function1}
-#see which curve could fit the best using the descdist() function
+#see which curve will fit the best using the descdist() function. This is to calibrate an impedance function for each mode based on the travel survey
+
+#NOTE: maybe we can try a distance based distribution???? Not really convinced why would could - perhaps to include all modes?
 
@@
            mode) %>%
-  summarize(workers = sum(ELE_G_POND_ESC2), #we should be summing the number of workers by the weight but we don't have a weight... so for now, 1 row = 1 trip. 
+  summarize(workers = sum(ELE_G_POND_ESC2), 
+            total_ttworkers = sum(workers*mean(TIEMPO_VIAJES)),
             .groups = "drop")%>%
@@
   rename(ZT1259 = VDESZT1259)
+
+#let's also do this for zn208
+workers_zn208 <- OD_v_HW %>%
+  group_by(VORI_CD_ZT208,
+           mode) %>%
+  summarize(workers = sum(ELE_G_POND_ESC2),
+            total_ttworkers = sum(workers*mean(TIEMPO_VIAJES)),
+            .groups = "drop") %>%
+  rename(CD_ZT208 = VORI_CD_ZT208)
+
+jobs_zn208 <- OD_v_HW %>%
+  group_by(VDES_CD_ZT208) %>%
+  summarize(jobs = sum(ELE_G_POND_ESC2),
+            .groups = "drop") %>%
+  rename(CD_ZT208 = VDES_CD_ZT208)
 ```
@@
             by = "ZT1259") %>%
+  st_as_sf()
+
+#for zn208
+workers_sf_zn208 <- workers_zn208 %>%
+  left_join(zn208,
+            by = "CD_ZT208") %>%
+  st_as_sf()
+
+jobs_sf_zn208 <- jobs_zn208 %>%
+  left_join(zn208,
+            by = "CD_ZT208") %>%
   st_as_sf()
@@
 
-# Check the mean and median travel times by mode
+OD_zn208 <- OD_v_HW %>%
+  group_by(VORI_CD_ZT208,
+           VDES_CD_ZT208,
+           mode) %>%
+  summarize(tt = mean(TIEMPO_VIAJES),
+            .groups = "drop")
+
+# Check the mean and median travel times by mode, the first is at the level of zn1259 and the second is at zn208
 OD %>% 
@@
             max_tt = max(tt))
+
+OD_zn208 %>% 
+  group_by(mode) %>% 
+  summarize(mean_tt = mean(tt),
+            median_tt = median(tt),
+            min_tt = min(tt),
+            max_tt = max(tt))
 ```
@@
 
-#THIS IS TOO MANY!
-```
-
-Preview the table for one origin and one destination, as an example:
-```{r}
-# OD_2 %>% filter(VORIZT1259 == "079-09-225" & VDESZT1259 == 2) #explanation: 'workers' corresponds to the people who leave ORI 1 to go to any destination by mode (P_im) (e.g., 5 workers leave ORI 1 by bike to go to a job somewhere in the region).
-'jobs' corresponds to the number of workers that come to DES 2 as a job destination in the region (sum_m O_j).
-'tt' corresponds to the average travel time for i to j per mode (avg.
-c_ijm) (e.g., only 1 trip is made by bike from ORI 1 to DES 2 and it is 5 mins, so the average travel time is 5 mins) -- this c_ijm will be used to calculate f(c_ijm).
+#let's also do this at the level of zn208
+OD_3 <- OD_zn208 %>%
+  left_join(workers_zn208,
+            by = c("VORI_CD_ZT208" = "CD_ZT208", "mode"))
+
+OD_3 <- OD_3 %>%
+  left_join(jobs_zn208,
+            by = c("VDES_CD_ZT208" = "CD_ZT208"))
+
 ```
@@
                                                HW_walk_lnorm$estimate["sdlog"])))
+
+#also for zn208
+OD_3 <- OD_3 %>%
+  mutate(f = case_when(mode == "transit" ~ dgamma(tt,
+                                                  HW_transit_gamma$estimate["shape"], 
+                                                  HW_transit_gamma$estimate["rate"]),
+                       mode == "car/motor" ~ dgamma(tt,
+                                                    HW_car_motor_gamma$estimate["shape"],
+                                                    HW_car_motor_gamma$estimate["rate"]),
+                       mode == "bike" ~ dlnorm(tt,
+                                               HW_bike_lnorm$estimate["meanlog"],
+                                               HW_bike_lnorm$estimate["sdlog"]),
+                       mode == "walk" ~ dlnorm(tt, 
+                                               HW_walk_lnorm$estimate["meanlog"], 
+                                               HW_walk_lnorm$estimate["sdlog"])))
 ```
@@
                           f = f))
+
+#and for zn208
+OD_SA_zn208 <- OD_3 %>%
+  mutate(catch = 1) %>%
+  mutate(SA_ij = sp_avail(., 
+                          o_id = VORI_CD_ZT208,
+                          d_id = VDES_CD_ZT208,
+                          pop = workers,
+                          opp = jobs,
+                          r = catch,
+                          f = f))
 ```
@@
 sum(OD_SA$SA_ij, na.rm = T)
+sum(OD_SA_zn208$SA_ij, na.rm = T)
 sum(jobs$jobs)
+
 sum(OD_SA$tt)
-```
-Spatial availability by mode, we can see that car mode captures 43% of the spatial availability but only accounts for 30% of the tt_ijm !!:
+sum(OD_SA_zn208$tt)
+
+sum(OD_SA$total_ttworkers)
+sum(OD_SA_zn208$total_ttworkers)
+```
+Spatial availability by mode, we can see that car mode captures ~43% of the spatial availability but only accounts for ~30% of the tt_ijm !!:
 ```{r}
@@
             V_m_perc = sum(SA_ij)/847574,
-            tt_m = sum(tt)/435061.9)
-```
-
-Okay, let's preform a few more checks and then aggregate spatial availability by origin (zn208) and by origin,mode.
+            tt_m = sum(tt)/435061.9,
+            ttworkers = sum(total_ttworkers)/440728388)
+
+#for zn208
+OD_SA_zn208 %>%
+  group_by(mode) %>%
+  summarize(V_m = sum(SA_ij),
+            V_m_perc = sum(SA_ij)/847574,
+            tt_m = sum(tt)/185686.1,
+            ttworkers = sum(total_ttworkers)/1211991264)
+```
+
+Okay, let's preform a few more checks and then aggregate spatial availability by origin (for both zn1259 and zn208(less 'NAs')) and by origin,mode.
 
@@
 sum(SA_i$V_i)
+
+#zn208
+SA_i_zn208 <- OD_SA_zn208 %>%
+  group_by(VORI_CD_ZT208) %>%
+  summarize(V_i = sum(SA_ij))
+
+summary(SA_i_zn208) 
+sum(SA_i_zn208$V_i)
 ```
@@
 sum(SA_im$V_im)
+
+#zn208
+SA_im_zn208 <- OD_SA_zn208 %>%
+  group_by(VORI_CD_ZT208, 
+           mode) %>%
+  summarize(V_im = sum(SA_ij),
+            .groups = "drop")
+
+summary(SA_im_zn208)
+sum(SA_im_zn208$V_im)
 ```
@@
 sum(population$population)
+
+#zn208
+population_zn208 <- workers_zn208 %>%
+  group_by(CD_ZT208) %>%
+  summarize(population = sum(workers))
+
+sum(population_zn208$population)
 ```
@@
 sum(SA_im$V_im)
+
+#zn208
+SA_i_zn208 <- SA_i_zn208 %>%
+  left_join(population_zn208,
+            by = c("VORI_CD_ZT208" = "CD_ZT208"))
+sum(SA_i_zn208$V_i)
+
+SA_im_zn208 <- SA_im_zn208 %>%
+  left_join(population_zn208,
+            by = c("VORI_CD_ZT208" = "CD_ZT208"))
+sum(SA_im_zn208$V_im)
 ```
@@
 summary(SA_im)
+
+#zn208
+SA_i_zn208 <- SA_i_zn208 %>%
+  mutate(v_i = V_i/population)
+
+summary(SA_i_zn208)
+
+SA_im_zn208 <- SA_im_zn208 %>%
+  mutate(v_im = V_im/population)
+
+summary(SA_im_zn208)
 ```
@@
   st_as_sf()
+
+#zn208
+SA_i_zn208 <- SA_i_zn208 %>%
+  left_join(zn208 %>%
+              dplyr::select(CD_ZT208),
+            by =c("VORI_CD_ZT208" = "CD_ZT208")) %>%
+  st_as_sf()
+
+SA_im_zn208 <- SA_im_zn208 %>%
+  left_join(zn208 %>%
+              dplyr::select(CD_ZT208),
+            by = c("VORI_CD_ZT208" = "CD_ZT208")) %>%
+  st_as_sf()
 ```
@@
 ```{r create-bounding-boxes}
-zn1259_mad <- zn1259 %>% filter(Municipality == "Madrid")
-
 #create the bounding box for the Madrid city + a little extra
@@
 
-Plot V (spatial availability) by origin:
-```{r}
-ggplot() +
+Plot V (spatial availability) by origin (first zn1259 then zn208):
+```{r}
+SA_i_V_zn208_plot <- ggplot() +
+  geom_sf(data = SA_i_zn208,
+          aes(fill = V_i),
+          color = NA) +
+  scale_fill_distiller(palette = "YlGn", #legend scale bar
+                       name = expression(V["i"]),
+                       na.value = "grey90",
+                       direction = 1,
+                       label = c("14.7 (min.)", "6248 (1st qu.)", " ", "17768 (3rd qu.)\n14613 (mean)", "51655 (max.)"), 
+                       breaks = c(0, quantile(SA_i_zn208$V_i)[2], mean(SA_i_zn208$V_i), quantile(SA_i_zn208$V_i)[4], max(SA_i_zn208$V_i)),
+                       limits = c(0, max(SA_i_zn208$V_i)))+
+  geom_sf(data = Madrid_muni_border, # border for Madrid muni
+          colour=alpha("black",1), 
+          size = 0.5, 
+          fill=NA, 
+          show.legend = "polygon") +
+  geom_sf(data = LEZ_centro, 
+          colour=alpha("red",1), 
+          size = 0.5, 
+          fill = NA) + 
+  geom_sf(data = LEZ_M30, 
+          colour=alpha("cyan",1), 
+          size = 0.5, 
+          fill = NA) +
+annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
+                         height = unit(0.7, "cm"), 
+                         width = unit(0.7, "cm"),
+                         style = north_arrow_orienteering(line_width = 0.25,
+                                                          line_col = "dimgrey", 
+                                                          fill = c("white","dimgrey"))) +
+  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
+                   height = unit(0.15, "cm")) +
+  theme_void() +
+  theme(legend.key.width = unit(0.2, 'cm')) # positing legend and clipping out white space
+
+SA_i_V_plot <- ggplot() +
   geom_sf(data = SA_i,
@@
                        direction = 1,
+                       label = c("0.9 (min.)", "697 (1st qu.)", "1842 (mean)", "2520 (3rd qu.)", "7463 (max.)"), 
+                       breaks = c(0, quantile(SA_i$V_i)[2], mean(SA_i$V_i), quantile(SA_i$V_i)[4], max(SA_i$V_i)),
+                       limits = c(0, max(SA_i$V_i)))+
+  geom_sf(data = Madrid_muni_border, # border for Madrid muni
+          colour=alpha("black",1), 
+          size = 0.5, 
+          fill=NA, 
+          show.legend = "polygon") +
+  geom_sf(data = LEZ_centro, 
+          colour=alpha("red",1), 
+          size = 0.5, 
+          fill = NA) + 
+  geom_sf(data = LEZ_M30, 
+          colour=alpha("cyan",1), 
+          size = 0.5, 
+          fill = NA) +
+  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
+                         height = unit(0.7, "cm"), 
+                         width = unit(0.7, "cm"),
+                         style = north_arrow_orienteering(line_col = "dimgrey", 
+                                                          fill = c("white","dimgrey"))) +
+  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
+                   height = unit(0.15, "cm")) +
+  theme_void() +
+  theme(legend.key.width = unit(0.2, 'cm')) # positing legend and clipping out white space
+```
+
+```{r}
+SA_i_V_zn208_plot
+ggsave("Presentations/SA_i_V_zn208_plot.png")
+```
+
+```{r}
+SA_i_V_plot
+ggsave("Presentations/SA_i_V_zn1259_plot.png")
+```
+
+Plot v (spatial availability per capita) by origin (first zn1259 then zn208):
+```{r}
+SA_i_v_plot <- ggplot() +
+  geom_sf(data = SA_i,
+          aes(fill = v_i),
+          color = NA) +
+  scale_fill_gradient2(name = expression(v["i"]),
+                       midpoint = 1,
+                       na.value = "grey90")+
+  geom_sf(data = Madrid_muni_border, # border for Madrid muni
+          colour=alpha("black",1), 
+          size = 0.5, 
+          fill=NA, 
+          show.legend = "polygon") +
+  geom_sf(data = LEZ_centro, 
+          colour=alpha("red",1), 
+          size = 0.5, 
+          fill = NA) + 
+  geom_sf(data = LEZ_M30, 
+          colour=alpha("cyan",1), 
+          size = 0.5, 
+          fill = NA) +
+  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
+                         height = unit(0.7, "cm"), 
+                         width = unit(0.7, "cm"),
+                         style = north_arrow_orienteering(line_col = "dimgrey", 
+                                                          fill = c("white","dimgrey"))) +
+  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
+                   height = unit(0.15, "cm")) +
+  theme_void() +
+  theme(legend.key.width = unit(0.2, 'cm')) # positing legend and clipping out white space
+
+SA_i_v_zn208_plot <- ggplot() +
+  geom_sf(data = SA_i_zn208,
+          aes(fill = v_i),
+          color = NA) +
+  scale_fill_gradient2(name = expression(v["i"]),
+                       midpoint = 1,
+                       na.value = "grey90")+
+  geom_sf(data = Madrid_muni_border, # border for Madrid muni
+          colour=alpha("black",1), 
+          size = 0.5, 
+          fill=NA, 
+          show.legend = "polygon") +
+  geom_sf(data = LEZ_centro, 
+          colour=alpha("red",1), 
+          size = 0.5, 
+          fill = NA) + 
+  geom_sf(data = LEZ_M30, 
+          colour=alpha("cyan",1), 
+          size = 0.5, 
+          fill = NA) +
+  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
+                         height = unit(0.7, "cm"), 
+                         width = unit(0.7, "cm"),
+                         style = north_arrow_orienteering(line_col = "dimgrey", 
+                                                          fill = c("white","dimgrey"))) +
+  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
+                   height = unit(0.15, "cm")) +
+  theme_void() +
+  theme(legend.key.width = unit(0.2, 'cm')) # positing legend and clipping out white space
+```
+```{r}
+SA_i_v_zn208_plot
+ggsave("Presentations/SA_i_vv_zn208_plot.png")
+```
+```{r}
+SA_i_v_plot
+ggsave("Presentations/SA_i_vv_zn1259_plot.png")
+```
+
+Plot V (spatial availability) by origin-mode (first zn1259 then zn208):
+```{r}
+SA_im_V_plot <- ggplot() +
+  geom_sf(data = SA_im,
+          aes(fill = V_im),
+          color = NA) +
+  scale_fill_distiller(palette = "YlGn", #legend scale bar
+                       name = expression(V["im"]),
+                       na.value = "grey90",
+                       direction = 1)+
+  geom_sf(data = Madrid_muni_border, # border for Madrid munim
+          colour=alpha("black",1), 
+          size = 0.5, 
+          fill=NA, 
+          show.legend = "polygon") +
+  geom_sf(data = LEZ_centro, 
+          colour=alpha("red",1), 
+          size = 0.5, 
+          fill = NA) + 
+  geom_sf(data = LEZ_M30, 
+          colour=alpha("cyan",1), 
+          size = 0.5, 
+          fill = NA) +
+  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
+                   height = unit(0.15, "cm")) +
+  theme_void() +
+  theme(legend.key.width = unit(0.2, 'cm')) +
+  facet_wrap(~mode, nrow = 2)
+
+#zn208
+SA_im_V_zn208_plot <- ggplot() +
+  geom_sf(data = SA_im_zn208,
+          aes(fill = V_im),
+          color = NA) +
+  scale_fill_distiller(palette = "YlGn", #legend scale bar
+                       name = expression(V["im"]),
+                       na.value = "grey90",
+                       direction = 1,
                        #trans = "log",
@@
           fill = NA) +
-  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
-                         height = unit(0.8, "cm"), 
-                         width = unit(0.8, "cm"),
-                         style = north_arrow_orienteering(line_width = 0.25,
-                                                          line_col = "dimgrey", 
-                                                          fill = c("white","dimgrey"))) +
   annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
@@
   theme_void() +
-  theme(legend.position = c(1.10, 0.25), 
-        plot.margin = margin(t = 0, r = 0, b = 0, l = -80)) # positing legend and clipping out white space
+  theme(legend.key.width = unit(0.2, 'cm')) +
+  facet_wrap(~mode, nrow = 2)
+```
+```{r}
+SA_im_V_plot
+ggsave("Presentations/SA_im_V_zn1259_plot.png")
+```
+```{r}
+SA_im_V_zn208_plot
+ggsave("Presentations/SA_im_V_zn208_plot.png")
+```
+
+Plot v (spatial availability per capita) by origin (first zn1259 then zn208):
+```{r}
+SA_im_v_plot <- ggplot() +
+  geom_sf(data = SA_im,
+          aes(fill = v_im),
+          color = NA) +
+  scale_fill_gradient2(name = expression(v["i"]),
+                       midpoint = 1,
+                       mid = "grey80")+
+  geom_sf(data = Madrid_muni_border, # border for Madrid muni
+          colour=alpha("black",1), 
+          size = 0.5, 
+          fill=NA, 
+          show.legend = "polygon") +
+  geom_sf(data = LEZ_centro, 
+          colour=alpha("red",1), 
+          size = 0.5, 
+          fill = NA) + 
+  geom_sf(data = LEZ_M30, 
+          colour=alpha("cyan",1), 
+          size = 0.5, 
+          fill = NA) +
+  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
+                   height = unit(0.15, "cm")) +
+  theme_void() +
+  theme(legend.key.width = unit(0.2, 'cm')) +
+  facet_wrap(~mode, nrow = 2)
+
+SA_im_v_zn208_plot <- ggplot() +
+  geom_sf(data = SA_im_zn208,
+          aes(fill = v_im),
+          color = NA) +
+  scale_fill_gradient2(name = expression(v["i"]),
+                       midpoint = 1,
+                       mid = "grey80")+
+  geom_sf(data = Madrid_muni_border, # border for Madrid muni
+          colour=alpha("black",1), 
+          size = 0.5, 
+          fill=NA, 
+          show.legend = "polygon") +
+  geom_sf(data = LEZ_centro, 
+          colour=alpha("red",1), 
+          size = 0.5, 
+          fill = NA) + 
+  geom_sf(data = LEZ_M30, 
+          colour=alpha("cyan",1), 
+          size = 0.5, 
+          fill = NA) +
+  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
+                   height = unit(0.15, "cm")) +
+  theme_void() +
+  theme(legend.key.width = unit(0.2, 'cm')) +
+  facet_wrap(~mode, nrow = 2)
+```
+```{r}
+SA_im_v_plot
+ggsave("Presentations/SA_im_vv_zn1259_plot.png")
+```
+```{r}
+SA_im_v_zn208_plot
+ggsave("Presentations/SA_im_vv_zn208_plot.png")
 ```

### 2023-07-30 NECTAR-Coimbra-Presentation.Rmd
Presentation file commits pt.1.

--- 2023-06-12 CIT-Presentation.Rmd
@@
 ---
-title: "Multi-modal spatial availability: a research proposal"
-subtitle: "CIT - Tenerife 2023"
+title: "Quantifying multi-modal competition using spatial availability"
+subtitle: "NECTAR - Sustainable Neighbourhoods"
 author: "Anastasia Soukhov, J. Tarriño Ortiz, J. A. Soria-Lara, A. Páez"
-date: "2022-11-10"
+date: "21-06-2023"
 output:
@@
     nature:
-      slideNumberFormat: "%current% / 14" 
+      slideNumberFormat: "%current% / 17" 
       highlightStyle: github
@@
 library(kableExtra)
+library("ggpmisc")
 options(scipen = 999)
@@
                            100, 100, 15, 80, 80, 10), 
-                    pop = c(20000, 20000, 20000, 30000, 30000, 30000,
-                            50000, 50000,50000, 100000, 100000, 100000,
+                    pop = c(16666.66666, 16666.66666, 16666.66666, 33333.33333, 33333.33333, 33333.33333,
+                            60000, 60000,60000, 90000, 90000, 90000,
                             3000, 3000, 3000, 7000,7000, 7000),
@@
 od_table <- data.frame(Center = c("A", "B", "C", "1", "2", "3"),
-  Size = c("50,000 pop (40% pt)", "150,000 pop (33% pt)", "10,000 pop (30% pt)", 
+  Size = c("50,000 pop.", "150,000 pop.", "10,000 pop.", 
            "100,000 jobs", "100,000 jobs", "10,000 jobs"))
@@
   annotate(geom = "label", 
-           x=c(0.5, 0.6, 7),
-           y=c(9, 1, 5.2), 
-           label = c("Suburban", "Urban", "Satellite"), 
+           x=c(1.6, 1.7, 7.5),
+           y=c(9.1, 0.9, 5.2), 
            size=4, 
@@
              color = "black") +
-  # Text on curves
-  geom_textcurve(data = od_lines,
-               aes(x = x,
-                   y = y,
-                   xend = xend,
-                   yend = yend,
-                   linetype = group,
-                   group = group,
-                   label = time),
-               size = 3.5,
-             curvature = 0.20,
-             color = "black") +
   # Plot origins and destinations
@@
                 shape = type)) +
-  scale_size(range = c(4, 10)) +
+  scale_size(range = c(6, 16)) +
   # Label origins and destinations
@@
   # Create table
-  annotate(geom = "table",
+  annotate(geom = 'table',
            x = 9,
@@
            label = list(od_table),
-           fill = "white",
-           size=3) +
+           size=4) +
   # Theme the plot
@@
 toy_example_plot
-ggsave("Presentations/toy-example.png",dpi=600)
+ggsave("Presentations/toy-example1.png",dpi=600)
 ```
 
-## Overview
+## Accessibility competition between modes
 
-**1.** Accessibility measures and multi-modal applications
+**Motivation**: interpreting accessibility measures (location-based _potential interaction_) for policy interventions
 
-**2.** Explaining **spatial availability** within multi-modal analysis
+```{r, toy-example0, out.width='40%', fig.align='center', echo=FALSE}
+knitr::include_graphics("Presentations/access-demo.png")
+```
 
-**3.** Policy scenario planning application: low emission zones in Madrid
+---
+## Accessibility competition between modes
 
-**4.** Implications: the changes in spatial availability of opportunities, active mobility, GHG emissions
+**Motivation**: interpreting accessibility measures (location-based _potential interaction_) for policy interventions
+
+```{r, toy-example00, out.width='40%', fig.align='center', echo=FALSE}
+knitr::include_graphics("Presentations/access-demo.png")
+```
+
+### Overview
+
+**1.** Types of current accessibility measures and their limitations
+
+**2.** Multi-modal **spatial availability**
+
+**3.** Planning accessibility policy interventions considering multi-modal competition
+
+---
+class: center, middle
+## Classification of multi-modal accessibility measures
+
+##### Hansen-type accessibility _(Hansen, 1959)_
+_No competition_ and _unconstrained_: $S_i^m = \sum_j O_j f^m(c_{ij}^m)$
+
+--
+
+##### Shen-type accessibility _(Shen, 1998)_ and _(Luo and Wang, 2003)_
+**Competitive** and _unconstrained_: $a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}$
+
+--
+
+##### Spatial availability _(Soukhov et al., 2023)_
+**Competitive** and **constrained**: $V_i^m = \sum_j O_j F^{t,m}_{ij}$
 
 ???
-XXX
+
+<!-- for this slide-->
+
+Accessibility: a way to quantify the access that a spatial unit has, a product of land-use and the population's means of transportation.
+It measures the amount of _potential interaction_ a population has with opportunities _(Hansen, 1959)_
+
+What do all these types of measures have in common? How they are a weighted sum of opportunities that can be interacted with. What differs between them is how they weight the opportunities.
+
 ---
-## Multi-modal accessibility measures
+
+### Simulated example: uncompetitive and unconstrained
 
@@
 
-#### Gravity-based accessibility (Hansen-type) - no competition
-
-$$A_i^m = \sum_{j}O_jf(c_{ij})^m$$
--   $i$ is a set of origin locations
--   $j$ is a set of destination locations
--   $O_j$ is the number of opportunities at location $j$
--   $f(\cdot)$ is an impedance function of $c_{ij}$ (cost of moving between $i$ and $j$) for each mode $m$; if binary, the measure if _cumulative opportunities_
-
-]
-
--- 
-.pull-right[
-
-**E.g.**, used in _(Tahmasbi et al., 2019)_ to calculate $A_i^m$ (per mode). Variations on how it is combined across modes.
-
-Suffers from the same interpretability issues _(Handy and Niemeier, 1997; Miller et al., 2018)_.
-
-Does not include competition (demand:supply, mass effect) _(Shen, 1998; Merlin and Hu, 2018)_.
-
-]
-
-???
-Accessibility:
-
-- A way to relatively quantify the access that a spatial unit: a product of land-use and the population's means of transportation
-
-- In other words, it measures the amount of _potential interaction_ a population has with opportunities _(Hansen, 1959)_
-
-- It connects land-use and the population's means of transportation: the population are the people that live within the region and the opportunities are the destinations of interest to the population or of the transportation network itself.
-
--**Hansen**-type accessibility - the sum of opportunities multiplied by a function which measures of the cost of moving between i and j.
-
-- this measure suffers from interpretability issues - what is travel-cost adjusted opportunities?
-
-- it also doesn't consider competition, which has been shown to reflect empirical accessibility to competitive-opportunities worse than competitive accessibility measures.
-
-Interpretability issues are not resolved when considering multiple modes.
-
-- Used in _(Tahmasbi et al., 2019)_ to quantify access $A_i$ for $m$ walking, public transit, and private car modes to retail store destinations (different $f^m(c_{ij}^m)$ for each $m$).
-Variations on how it is combined across modes (if at all) for general accessibility.
-It varies so widely because interpretation is inconsistent.
-- Suffers from the same interpretability issues _(Handy and Niemeier, 1997; Miller et al., 2018)_
-- Does not include competition (demand:supply or the mass effect) _(Shen, 1998; Merlin and Hu, 2018)_
-
----
-## Multi-modal accessibility measures
-
-.pull-left[
-#### Competitive accessibility (Shen-type) - no proportional allocation
-Shen (1998) modified Hansen-type accessibility; popularized as the **two-step floating catchment area (2SFCA) method** by _(Luo and Wang, 2003)_
-
-$$S_i^m = \sum_j \frac{O_jf(c_{ij})^m}{\sum_m D_j^m}$$
-
-with:
-$$D_j^m = \sum_i P_{i}^m f(c_{ij})^m$$
-- $S_{i}^m$ is Shen-type accessibility for each mode $m$
-- $D_j^m$ is the demand at $j$ from $P_{i}$ using each mode $m$
-
-]
-
-.pull-right[
-
-Generalized multi-model competition:
-$$S_i = \sum_m \biggl(\frac{P_i^m}{P_i}\biggr) \sum_j \frac{O_jf(c_{ij})^m}{\sum_m \sum_i P_{i}^m f(c_{ij})^m}$$
-
-E.g., used by _Tao et al., 2020_.
-
-However, the resulting value is an accessibility _score_ so interpretation is obfuscated _(Soukhov et al., 2023)_.
-
-]
-
-???
-- e.g., used by _Tao et al., 2020_ to calculate the $S_i^m$ for $m$ modes public transit and car to employment as well as the generalized accessibility $S_i$
-
-- **Shen**-type accessibility - the sum of travel-cost adjusted opportunities (supply) divided by travel-cost adjusted demand!
-It considers travel-cost from both the opportunities perspective and the population side.
-It is done in 2 steps for the 2SFCA: "The first one computes the ratio between supply and demand of each health facility; the second step computes accessibility as the weighted sum of these ratios using a distance-decay function"
-
-It should be mentioned that Weibull (1976) was the first to arrive to this same formula from axiomatic approach. Joseph and Bantock (1982) then applied it to health care access.
-
----
-### Impedance functions - quantifying how subsets of people travel
-
-.left-column[
-
-$f(c_{ij})^m$ is a function of the cost of moving between $i$ and $j$ for modes $m$.
-
-]
-
-.right-column[]
-
----
-count: false
-### Impedance functions - quantifying how subsets of people travel
-
-.left-column[
-
-$f(c_{ij})^m$ is a function of the cost of moving between $i$ and $j$ for modes $m$.
-
-$f(c_{ij})^{t.behav.1} = \exp(-0.1\cdot c_{ij})$
-
-]
-
-.right-column[
-```{r comparison-impedance-functions-synthetic-example1, echo=FALSE, warning=FALSE, fig.width=8, fig.height=6}
-data.frame(tt = seq(0, 100, 1)) %>%
-  mutate(f1 = exp(-0.1 * tt)) %>%
-  pivot_longer(cols = -tt,
-               names_to = "Function",
-               values_to = "f") %>%
-  mutate(Function = case_when(Function == "f1" ~ "beta = 0.1")) %>%
-  ggplot() +
-  xlab(expression(c[ij]~"(travel time in minutes)")) +
-  ylab(expression('f(c'[ij]*')')) +
-  geom_line(aes(x = tt, 
-                y = f,
-                color = Function), size=1) +
-  scale_color_manual(values = c("dark green"), label = c(expression("exp(-0.1 c"[ij]*")")))+
-  scale_x_continuous(breaks = seq(0, 80, 20))+
-  theme_minimal()+
-  theme(axis.title = element_text(size=18),
-        axis.text = element_text(size=16),
-        legend.position = c(0.85, 0.9),
-        legend.title = element_blank(),
-        legend.text = element_text(size=18))
-```
-]
-
----
-count: false
-### Impedance functions - quantifying how subsets of people travel
-
-.left-column[
-
-$f(c_{ij})^m$ is a function of the cost of moving between $i$ and $j$ for modes $m$.
-
-$f(c_{ij})^{\text{t.behav.1}} = \exp(-0.1\cdot c_{ij})$
-
-$f(c_{ij})^{\text{t.behav.2}} = \exp(-0.6\cdot c_{ij})$
-
-]
-
-.right-column[
-```{r comparison-impedance-functions-synthetic-example2, warning=FALSE, echo=FALSE, fig.width=8, fig.height=6}
-data.frame(tt = seq(0, 100, 1)) %>%
-  mutate(f1 = exp(-0.1 * tt),
-         f2 = exp(-0.6 * tt)) %>%
-  pivot_longer(cols = -tt,
-               names_to = "Function",
-               values_to = "f") %>%
-  mutate(Function = case_when(Function == "f1" ~ "beta = 0.1",
-                              Function == "f2" ~ "beta = 0.6")) %>%
-  ggplot() +
-  xlab(expression(c[ij]~"(travel time in minutes)")) +
-  ylab(expression('f(c'[ij]*')')) +
-  geom_line(aes(x = tt, 
-                y = f,
-                color = Function), size=1) +
-  scale_color_manual(values = c("dark green", "#9999CC"), label = c(expression("exp(-0.1 c"[ij]*")"), expression("exp(-0.6 c"[ij]*")")))+
-  scale_x_continuous(breaks = seq(0, 80, 20))+
-  theme_minimal()+
-  theme(axis.title = element_text(size=18),
-        axis.text = element_text(size=16),
-        legend.position = c(0.85, 0.9),
-        legend.title = element_blank(),
-        legend.text = element_text(size=18))
-```
-]
-
----
-count: false
-### Impedance functions - quantifying how subsets of people travel
-
-.left-column[
-
-$f(c_{ij})^m$ is a function of the cost of moving between $i$ and $j$ for modes $m$.
-
-$f(c_{ij})^{\text{t.behav.1}} = \exp(-0.1\cdot c_{ij})$
-
-$f(c_{ij})^{\text{t.behav.2}} = \exp(-0.6\cdot c_{ij})$
-
-$f(c_{ij}) = 1$ if $c_{ij} \leq 30$ else 
-  $f(c_{ij}) = 0$
-
-]
-
-.right-column[
-```{r comparison-impedance-functions-synthetic-example3, warning=FALSE, echo=FALSE, fig.width=8, fig.height=6}
-data.frame(tt = seq(0, 100, 1)) %>%
-  mutate(f1 = exp(-0.1 * tt),
-         f2 = exp(-0.6 * tt),
-         f3 = ifelse(tt < 30, 1, 0)) %>%
-  pivot_longer(cols = -tt,
-               names_to = "Function",
-               values_to = "f") %>%
-  mutate(Function = case_when(Function == "f1" ~ "beta = 0.1",
-                              Function == "f2" ~ "beta = 0.6",
-                              Function == "f3" ~ "tt <= 30")) %>%
-  ggplot() +
-  xlab(expression(c[ij]~"(travel time in minutes)")) +
-  ylab(expression('f(c'[ij]*')')) +
-  geom_line(aes(x = tt, 
-                y = f,
-                color = Function), size=1) +
-  scale_color_manual(values = c("dark green", "#9999CC", "#CC6666"), label = c(expression("exp(-0.1 c"[ij]*")"), expression("exp(-0.6 c"[ij]*")"), expression(c[ij]~"< 30 min")))+
-  scale_x_continuous(breaks = seq(0, 80, 20))+
-  theme_minimal()+
-  theme(axis.title = element_text(size=18),
-        axis.text = element_text(size=16),
-        legend.position = c(0.85, 0.9),
-        legend.title = element_blank(),
-        legend.text = element_text(size=18))
-```
-]
----
-### Simulated example - Hansen-type accessibility (no competition or proportional allocation consideration)
-
-.pull-left[
-
-```{r, toy-example1, echo=FALSE}
-knitr::include_graphics("Presentations/toy-example.png")
+```{r, toy-example1, out.width='120%', echo=FALSE}
+knitr::include_graphics("Presentations/toy-example1.png")
 ```
@@
 
-$A_i^m = \sum_jO_jf(c_{ij})^m$
+##### Hansen-type accessibility
+$S_i^m = \sum_jO_jf^m(c_{ij}^m)$
 
-Impedance function: $f(c_{ij})^{\text{car}} = f(c_{ij})^{\text{pt}} =exp(-0.1 \cdot c_{ij})$;
+Impedance function: $f(c_{ij})^{\text{car}} = f(c_{ij})^{\text{pt}} =exp(-0.1 \cdot c_{ij}^m)$;
 
@@
 
-A <- od_tt_m |>
+S <- od_tt_m |>
   group_by(i,m) |>
-  summarize(A = sum(opp * f),
+  summarize(S = sum(opp * f),
             .groups = "drop")
 
-A %>% kable(format = "html",
+add_data <- data.frame(i = "TOTAL", m = "" , S = S$S%>%sum())
+S <- bind_rows(S, add_data)
+
+S %>% kable(format = "html",
         booktabs = TRUE,
-        digits = 2)
+        digits = 1)
 ```
@@
 ???
+- NOTE: the sum of all that is what it means to be unconstrained. Also, A and B are equal, that is what it means to be uncompetitive.
+
 - explain the map; blue and orange dots are origins and employment centers respectively.
@@
 
-### Synthetic example - Shen-type accessibility (no proportional allocation consideration)
+### Simulated example: competitive and unconstrained
 
 .pull-left[
-```{r, echo=FALSE}
-knitr::include_graphics("Presentations/toy-example.png")
+```{r, out.width='120%',echo=FALSE}
+knitr::include_graphics("Presentations/toy-example1.png")
 ```
@@
 .pull-right[
-$A_i^m$ and $S_i^m = \sum_j \frac{O_j f(c_{ij})^m}{\sum_m \sum_i P_{i}^m f(c_{ij})^m}$
+
+##### Shen-type accessibility
+$a_i^m = \sum_j \frac{O_j f^m(c_{ij}^m)}{\sum_m \sum_i P_{i}^m f^m(c_{ij}^m)}$
 
@@
 
-S_m <- od_tt_m %>%
+a_m <- od_tt_m %>%
   left_join(LOS_j |> 
@@
   group_by(i,m) |>
-  summarize(S_i_m = sum(LOS_j * f),
+  summarize(a_i_m = sum(LOS_j * f),
             pop = mean(pop),
             .groups = "drop") |>
-  rename("S" = "S_i_m")
-
-S_m <- A %>% 
-  left_join(S_m, by=c("i","m"))
+  rename("a" = "a_i_m")
 
@@
 ```{r, echo=FALSE, message=FALSE}
-knitr::kable(S_m %>% dplyr::select(-pop),
+knitr::kable(a_m %>% dplyr::select(-c("pop")),
              format = "html",
@@
 
+???
+-  ## NOTE: this is competitive for S.
+you have high access with less demand, A and B are different.
+It is still unconstrained when you try to retrieve the number of opportunities -- it is the totla population and not the job-seeking population
 ---
-count: false
-### Synthetic example - Shen-type accessibility (no proportional allocation consideration)
+### Simulated example: competitive and unconstrained
 
 .pull-left[
-```{r, echo=FALSE}
-knitr::include_graphics("Presentations/toy-example.png")
+```{r, out.width='120%',echo=FALSE}
+knitr::include_graphics("Presentations/toy-example1.png")
 ```
@@
 .pull-right[
-$S_i = \sum_m \biggl(\frac{P_i^m}{P_i}\biggr) S_i^m$
+
+##### Shen-type accessibility
+$a_i = \sum_m \biggl(\frac{P_i^m}{P_i}\biggr) a_i^m$
 ```{r, echo=FALSE, message=FALSE}
-S <- S_m %>%
+a <- a_m %>%
   group_by(i) |>
-  summarize(S = sum(pop*S)/sum(pop),
+  summarize(a = sum(pop*a)/sum(pop),
             .groups = "drop")
 
-knitr::kable(S,
+knitr::kable(a,
              format = "html",
@@
              digits = 2)
+
+# weighted average-- is based just on the population overall not the opportunity-skeeing populationg (i.e., not weighted by travel behaviour impedance function)
 ```
@@
 
-We can do better, let's use proportional allocation (NOTE: WHY??????????)
-
 ---
-### Spatial availability - competitive _and_ proportional allocation
+### Spatial availability: competitive _and_ constrained
 
@@
 
-- The proportional allocation factor combines two aspects of the problem, population and cost:
+- The **proportional allocation factor** combines two aspects of the problem, population and cost:
 
-$F^{t,m}_{ij} = \frac{F^{pm}_{i} \cdot F^{cm}_{ij}}{\sum_{m=1}^M \sum_{i=1}^N F^{pm}_{i} \cdot F^{cm}_{ij}}$
+$F^{t,m}_{ij} = \frac{F^{pm}_{i} \cdot F^{cm}_{ij}}{\sum_{m} \sum_{i} F^{pm}_{i} \cdot F^{cm}_{ij}}$
 
-- Where $F^{p,m}_{i}$ is population allocation factor:
+- Where $F^{p,m}_{i}$ is **population allocation** factor:
 
-$F^{p,m}_{i} = \frac{P^m_{i}}{\sum_{i} P^m_{i}}$
+$F^{p,m}_{i} = \frac{P^m_{i}}{\sum_{m}\sum_{i} P^m_{i}}$
 
-- And the $F^{c,m}_{ij}$ is a cost allocation factor:
+- And the $F^{c,m}_{ij}$ is a **cost allocation** factor:
 
-$F^{c,m}_{ij} = \frac{f(c_{ij})^m}{\sum_{i} f(c_{ij})^m}$
+$F^{c,m}_{ij} = \frac{f^m(c_{ij}^m)}{\sum_{m} \sum_{i} f^m(c_{ij}^m)}$
 ]
@@
 
-- Total spatial availability: $V_{i} = \sum_{m}^M V^m_{i}$
-  - Considers **competition** and **proportionally allocates opportunities** (singly-constrained: $O = \sum_{j} O_j$)
+##### Proportional allocation mechanism
 
-- Total spatial availability per capita: $v_{i} = \sum_{m} \frac{V_i^m}{P_{i^m}}$
-  - Equivalent to Shen-style accessibility
+- $V^m_{i}$ is always singly-constrained: $\sum_{i} V_{i} = \sum_{m}\sum_{i} V_{i}^m = \sum_{j} O_j$
+
+- So, one can calculate spatial availability....
+  - **per $i$**: ( $V_{i} = \sum_{m} V^m_{i}$ )
+  - **per mode**: ( $v^{m} = \sum_{m} \sum_{i} V^m_{i}$ )
+  - **per capita in $i$**: ( $v_{i} = \sum_{m} \frac{V_i^m}{P_{i^m}}$ )
+  - Etc. etc.,
 
@@
 -supply of population and demand for opportunities are both travel-cost and population adjusted
+
 - why singly-constrained? the sum of all accessibility values for i equals all the opportunities for i to j (like in a singly constrained gravity model - it is attraction constrained)
 
+- Can calculate spatial availability per $i$, or per capita in $i$, or per $m$.
+Per subset of population -- anything!
+This is also something that you couldn't just do with the Shen-type measure because it is unconstrained.
+
 ---
-### Spatial availability - competitive _and_ proportional allocation
+### Simulated example: competitive _and_ constrained
 
 .pull-left[
-```{r, echo=FALSE}
-knitr::include_graphics("Presentations/toy-example.png")
+
+```{r, out.width='120%',echo=FALSE}
+knitr::include_graphics("Presentations/toy-example1.png")
 ```
@@
 
-$V^m_{i} = \sum_{j} O_j\ F^{t,m}_{ij}$ and $v^m_{i} = \frac{V^m_{i}}{P_{i}^m}$
+##### Spatial availability per origin and mode
+$V^m_{i} = \sum_{j} O_j\ F^{t,m}_{ij}$
 
@@
             v = V/first(pop))
+
+add_data <- data.frame(i = "TOTAL", m = "", V = V_m$V%>%sum(), v = NA)
+V_m <- bind_rows(V_m, add_data)
+
 ```
 ```{r, echo=FALSE, message=FALSE}
-knitr::kable(V_m,
+knitr::kable(V_m %>% dplyr::select(-v),
              format = "html",
@@
 ---
-count: false
-### Spatial availability - competitive _and_ proportional allocation
+
+### Simulated example: competitive _and_ constrained
 
 .pull-left[
-```{r, echo=FALSE}
-knitr::include_graphics("Presentations/toy-example.png")
+```{r,out.width='120%', echo=FALSE}
+knitr::include_graphics("Presentations/toy-example1.png")
 ```
@@
 .pull-right[
-$S_i$ (eqv. to $v_i$) and $V_{i} = \sum_{m=1}^M V^m_{i}$
+
+##### Spatial availability per origin, per mode
+$V_{i} = \sum_{m} V^m_{i}$
 
@@
 
-V <- V %>% left_join(S, by=c("i")) |> transmute(S,V)
+V <- V %>% left_join(a, by=c("i")) |> rename("v" = "a") |> transmute(i, V, v)
 
+add_data <- data.frame(i = "TOTAL", V = V$V%>%sum(), v = NA)
+V <- bind_rows(V, add_data)
+```
+
+```{r, echo=FALSE, message=FALSE}
 knitr::kable(V,
+             format = "html",
+             digits = 2)
+```
+
+$V^m = \sum_iV_i^m$
+
+```{r, echo=FALSE, message=FALSE}
+V_m <- od_tt_m %>%
+  group_by(m) %>%
+  summarize(V = sum(V_ij)) |> 
+  transmute(m, V, 
+            v = c((30+100+7)/(50+150+10), (20+50+3)/(50+150+10)))
+
+add_data <- data.frame(m = "TOTAL", V = V_m$V%>%sum())
+V_m <- bind_rows(V_m, add_data)
+
+knitr::kable(V_m,
              format = "html",
@@
 ```
-
 - November 2018 - LEZ Centro
@@
 ---
-### Multi-modal home-to-work accessibility in Madrid: 2018 travel survey
-
-.pull-left[
-```{r, echo=FALSE, out.width='120%',fig.align='center'}
-knitr::include_graphics("Presentations/SA_i_V_zn1259_plot.png")
-```
-Spatial Availability $V_i$ at resolution Zona 1259
-]
-
-.pull-right[
-```{r, echo=FALSE, out.width='120%', fig.align='center'}
-knitr::include_graphics("Presentations/SA_i_V_zn208_plot.png")
-```
-Spatial Availability $V_i$ at resolution Zona 208
-]
-
-???
-- a need for multi-modal analysis to assess impacts of this geographic. The issue of MUAP.
----
 class: center
 ### Multi-modal home-to-work accessibility in Madrid: 2018 travel survey
-```{r, echo=FALSE, out.width='45%',fig.align='center'}
-knitr::include_graphics("Presentations/SA_im_V_zn1259_plot.png")
+```{r, echo=FALSE, out.width='70%',fig.align='center'}
+knitr::include_graphics("Presentations/im_populations_zn208_plot.png")
 ```
-Spatial Availability $V_i^m$ at resolution Zona 1259
-
-???
-- With spatial availability you can compare between modes -- however still the issue of MUAP.
+Worker population, by mode, at resolution Zona 208
 
@@
 ### Multi-modal home-to-work accessibility in Madrid: 2018 travel survey
-```{r, echo=FALSE, out.width='45%',fig.align='center'}
-knitr::include_graphics("Presentations/SA_im_vv_zn1259_plot.png")
+```{r, echo=FALSE, out.width='70%',fig.align='center'}
+knitr::include_graphics("Presentations/Hansen_im_V_zn208_plot.png")
 ```
-Spatial Availability by $m$ population ( $v_i^m$ ) at resolution Zona 1259
+Uncompetitive and unconstrained accessibility $S_i^m$ at resolution Zona 208
 
 ???
-- Can also benchmark between modes: the spatial avail per mode-using population per $i$ -- however still the issue of MUAP
+1) totals don't add up - not constrained
+2) uncompetive between modes
+
 ---
-### Spatial availability: a way forward in multimodal transport and land-use planning
+class: center
+### Multi-modal home-to-work accessibility in Madrid: 2018 travel survey
+```{r, echo=FALSE, out.width='70%',fig.align='center'}
+knitr::include_graphics("Presentations/SA_im_V_zn208_plot.png")
+```
+Spatial Availability $V_i^m$ at resolution Zona 208
 
-- Data-driven transport accessibility policy scenario creation
+???
+- SA is better for transit/walk for centro. Car mode is limited in the centro. 
+- those who live in the periphery have high accessibility, those are the ones who have cars and use transit.
 
-- Modifiable unit areal problem (MUAP) - need for logically consistent accessibility measurement method
-  - Calculated travel time tessellation
+---
+class: center
+### Multi-modal home-to-work accessibility in Madrid: 2018 travel survey
+```{r, echo=FALSE, out.width='70%', fig.align='center'}
+knitr::include_graphics("Presentations/SA_i_V_zn208_plot.png")
+```
+Spatial Availability $V_i$ at resolution Zona 208
 
-- Questions of equity: which neighbourhoods and people _within_ neighbourhoods are not meeting the mark? 
-  - Who will be most impacted?
+???
+- center -- high competition, for all modes and those that travel from close/far of the city.
+
+---
+class: middle
+### Spatial availability: planning multi-modal transportation policy
+
+- **Scenario evaluation**: accessibility when...
+  - LEZ boundaries change
+  - Vehicle characteristic restrictions change
+  - Travel behaviour changes
+  - Spatial unit of analysis changes
+
+--
+
+- **Questions of equity**: which neighbourhoods, what people, what modes?
+  - GHG emission reductions
+  - Advantages to non-car users
+  - Who are impacted disproportionally
 
@@
 - policy scenario creation
-- MUAP exists, seen in Madrid transport data, the need for simulated calculation and consistent aggregation of accessibility (through proportional allocation)
+
 - Quantitatively, this method can be used to more interpetble answer these questions of equity. Socio-economic impact, GHG emissions, across sub-sets of population (multi-modal users included).
 
+<!-- - MUAP exists, seen in Madrid transport data, the need for simulated calculation and consistent aggregation of accessibility (through proportional allocation) -->
+
 ---
-count: false
 class: center, middle, inverse
@@
 #Extra slides
+
+---
+---
+## Multi-modal accessibility measures
+
+.pull-left[
+
+#### No competition and unconstrained
+##### Hansen-type accessibility
+
+$$S_i^m = \sum_{j}O_jf^m(c_{ij}^m)$$
+-   $m$ is a set of modes
+-   $i$ is a set of origin locations
+-   $j$ is a set of destination locations
+-   $O_j$ is the number of opportunities at location $j$
+-   $f(\cdot)$ is an impedance function of $c_{ij}$ for each mode $m$ (a function of cost of moving between $i$ and $j$ for each mode $m$)
+
+]
+
+--
+.pull-right[
+
+**E.g.** used in _(Tahmasbi et al., 2019)_.
+
+**Unconstrained** - suffers from the same interpretability issues _(Handy and Niemeier, 1997; Miller et al., 2018)_.
+
+Does not include **competition** - no mass effect consideration _(Shen, 1998; Merlin and Hu, 2018)_.
+
+]
+
+???
+1. Accessibility: a sum of opportunities.
+
+2. E.g., it is used in Tahmasbi et al., 2019 to calculate accessibility per spatial unit per mode -- and compared across modes. Other literature has varied how the measure has been aggregated.
+
+It is unconstrained -- interpretability issues.
+The impedance function weights the opportunities, but what is a 'travel-cost adjusted potential opportunity' anyways?
+This difficulty in interpretation is a result of no global 'maximum' accessibility.
+difficult to make comparisons.
+
+It is not competitive -- it does not consider demand for the opportunities in how the opportunities are summed. This does not include the 'mass effect'
+
+---
+## Multi-modal accessibility measures
+
+.pull-left[
+#### Competition and unconstrained
+##### Shen-type accessibility
+
+Popularized as the **two-step floating catchment area (2SFCA) method** by _(Luo and Wang, 2003)_ (introduced by _Shen 1998_)
+
+$$a_i^m = \sum_j \frac{O_jf(c_{ij})^m}{\sum_m D_j^m}$$
+
+with:
+$$D_j^m = \sum_i P_{i}^m f(c_{ij})^m$$
+- $a_{i}^m$ is Shen-type accessibility for each mode $m$
+- $D_j^m$ is the demand at $j$ from $P_{i}$ using each mode $m$
+
+]
+
+.pull-right[
+
+Generalized multi-model competition:
+$$a_i = \sum_m \biggl(\frac{P_i^m}{P_i}\biggr) \sum_j \frac{O_jf(c_{ij})^m}{\sum_m \sum_i P_{i}^m f(c_{ij})^m}$$
+
+**E.g.**, used in _Tao et al., 2020_.
+
+Though **competitive**, it is still **unconstrained** making interpretation difficult _(Soukhov et al., 2023)_
+
+]
+
+???
+**Shen**-type accessibility - the sum of travel-cost adjusted opportunities (supply) divided by travel-cost adjusted demand!
+It considers travel-cost from both the opportunities perspective and the population side.
+It can be seen as an access score of the supply to demand.
+
+To calculate the generalized version, access is calculated for each mode and then multiplied by the population weighted average.
+
+It is used by some literature, such as _Tao et al.
+2020_ and is competitive as it does consider demand.
+However it is still not constrained.
+The totals do not sum up to anything.
+The numerator and denominator's definitions are confusing (trust me).
+And now in the multi-modal case, the score is multiplied by the pop weighted average -- further difficult to understand.
 

### 2023-07-30 LEZ-scenarios.qmd
Presentation file commit pt.2. and draft manuscript documents.

---
title: "LEZ-scenarios"
format: html
---
# Data prep
```{r load-packages, include=FALSE, cache=FALSE}
library(tidyverse)
library(sf)
library(fitdistrplus)
library(ggplot2)
library(readxl) #imports excel files
library(patchwork)
library(cowplot)
library(ggspatial)
```

```{r sp_avail-function,include=FALSE}
#defining the spatial availability function
sp_avail <- function(x, o_id, d_id, pop, opp, r, f, alpha = 1){

  o_id <- rlang::enquo(o_id)
  d_id <- rlang::enquo(d_id)
  pop <- rlang::enquo(pop)
  opp <- rlang::enquo(opp)
  r <- rlang::enquo(r)
  f <- rlang::enquo(f)

  sum_pop <- x %>%
    dplyr::distinct(!!o_id,
                    .keep_all = TRUE) %>%
    dplyr::mutate(sum_pop = !!r*(!!pop)^alpha) %>%
    dplyr::pull(sum_pop) %>%
    sum()

  f_p <- dplyr::pull(x, !!r) * dplyr::pull(x, !!pop)^alpha / sum_pop

  sum_impedance <- x %>%
    dplyr::group_by(!!d_id) %>%
    dplyr::summarize(sum_impedance = sum(!!f))

  x <- x %>%
    dplyr::left_join(sum_impedance,
                     by = rlang::as_name(d_id))

  f_c <- dplyr::pull(x, !!f) / x$sum_impedance

  x$f_c <- f_c
  x$f_p <- f_p

  sum_pa <- x %>%
    dplyr::group_by(!!d_id) %>%
    dplyr::summarize(sum_pa= sum(f_p * f_c))

  x <- x %>%
    dplyr::left_join(sum_pa,
                     by = rlang::as_name(d_id))
  f_t <- (f_p * f_c) / dplyr::pull(x, sum_pa)

  dplyr::pull(x, !!opp) * f_t
}
```

Import the trip data:
```{r data-import-filter}
load(file = "data-prep/data/OD_v.RData")
#Keep only the direct home to work trips (no chained home-to-x-to-work trips, they only make up a small portion of home-to-work trips anyways, see the exploration of this in the 'data-exploration.rmd' file)
OD_v_HW <- OD_v %>% filter(VORI == "1" & VDES == "2") #VORG = 1 = home, VDES = 2= work
```

Import zones:
```{r, warning=FALSE}
#importing zones
load("data-prep/data/zn1259.RData")
load("data-prep/data/zn208.RData")
load("data-prep/data/zn_u.RData") # a key to join the zn1259 to the zn208 + contains municipality names
LEZ_centro <- read_sf("data-prep/data-raw/LEZ.shp") %>% st_transform(4326) %>% mutate(merge_all = 1) %>%
    group_by(merge_all) %>%
    summarise(geometry = sf::st_union(geometry)) %>%
    ungroup()
LEZ_M30 <- read_sf("data-prep/data-raw/Zona_Bajas_Emisiones_MADRID.shp") %>% st_transform(4326)

#creating a list of zones that correspond to the zn208, which indicate if they are in a LEZ.
LEZ_centro_list <- st_intersection(zn208, LEZ_centro) %>% 
  st_drop_geometry() %>% 
  dplyr::select(CD_ZT208) %>% 
  unlist()

LEZ_M30_list <- st_intersection(zn208, LEZ_M30) %>% 
  st_drop_geometry() %>% 
  dplyr::select(CD_ZT208) %>% 
  unlist()
```

Let's describe **spatial availability** ($V_{ij}$). The formulation:
$$
V_{ij} = \sum_{j=1}^J O_j \frac{\sum_{m=1}^M F^{pm}_{ij} \cdot F^{cm}_{ij}}{\sum_{m=1}^M\sum_{i=1}^K F^{pm}_{ij} \cdot F^{cm}_{ij}}
$$

The population allocation factor ($F_{ij}^pm$) is based on _worker population_ $p$ for each ij (origin to destination) trip for each mode $m$. $F_{ij}^pm$ comes from the home-to-work flows (the sum of workers that leave an origin and the sum of workers that go to a destination): this is calculated for each mode traveling from i to j. This component is the 'competitive demand for opportunities'

The $O_{j}$ is simple - it represents the number of jobs at a destination $j$ (i.e., the sum of workers (not mode-specific, just all workers) that go to each destination j). This component is the supply of opportunities.

The complexity we have for this case study is the travel cost allocation factor $F^{cm}_{ij}$ due to the 4 modes (transit, car, bike, walk). Think of $F^{cm}_{ij}$ as the travel cost for each ij trip per mode. This travel cost is based on _the fitted impedance function_ $f(c^{m}_{ij})$ which is calculated for each mode (for our sample, log-normal is selected for all modes other then car - car is gamma). These impedance functions are predicted using each travel time for each mode (i.e., 4 modes, so we have 4 impedance functions). The impedance functions will be used to predict the _mode relative_ cost of travel, specifically, the larger the value, the stronger the travel-time decay effect on spatial availability will be. In other words, the larger the impedance value for an i the more $O_j$ are allocated to that origin i (a least costly travel cost). This component is the 'competitive travel cost to opportunities'.

Ultimately, for each origin to destination, the product of $F^{cm}_{ij}$ and $F^{pm}_{ij}$ for all modes $M$ (here, we have 4 modes $car$, $transit$, $walk$, $bic$) is divided by the SUM of all $F^{cm}_{ij}$ and $F^{pm}_{ij}$ in that region $K$ (all the origins $i$ that travel to a destination $j$) for all modes $M$.

## impedance functions

We must manipulate the data before we can fit the impedance function:
```{r data-for-impedance}
#here we separate all the "car" trips, "transit" trips, "bike" trips, and "walk" trips. We then will create 4 different impedance functions
transit_HW_alltt <- OD_v_HW  %>% 
  filter(MODO_PRIORITARIO <= 9) %>%
  transmute(TIEMPO_VIAJES, 
            mode = "transit")

car_motor_HW_alltt <- OD_v_HW  %>% 
  filter(MODO_PRIORITARIO <= 19 & MODO_PRIORITARIO >= 10) %>% #includes both motorcycles and cars
  transmute(TIEMPO_VIAJES, 
            mode = "car/motor")

bike_HW_alltt <- OD_v_HW  %>% 
  filter(MODO_PRIORITARIO <= 23 & MODO_PRIORITARIO >= 20) %>%
  transmute(TIEMPO_VIAJES, 
            mode = "bike")

walk_HW_alltt <- OD_v_HW  %>% filter(MODO_PRIORITARIO == 24) %>%
  transmute(TIEMPO_VIAJES, 
            mode = "walk")
```

Then create the impedance functions:
```{r fitting-impedance-function}
# using fitdist function to fit a distribution using the default maximum likelihood estimation method and Nelder-Mead method for direct optimization
lnorm_ <- fitdistrplus::fitdist(data=transit_HW_alltt$TIEMPO_VIAJES,#%>% unlist() %>% as.numeric(),
                                "gamma", 
                                method="mle",
                                #Notice the use of the weights,
                                #weights = round(transit_HW_alltt$ELE_G_POND_ESC2),
                                optim.method="Nelder-Mead")
HW_transit_lnorm <- lnorm_

gamma_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt$TIEMPO_VIAJES, #%>% unlist() %>% as.numeric(),
                                "gamma", 
                                method="mle",
                                #Notice the use of the weights,
                                #weights = round(car_motor_HW_alltt$ELE_G_POND_ESC2), 
                                optim.method="Nelder-Mead")
HW_car_motor_gamma <- gamma_

lnorm_ <- fitdistrplus::fitdist(data=bike_HW_alltt$TIEMPO_VIAJES, #%>% unlist() %>% as.numeric(),
                                "lnorm", 
                                method="mle",
                                #Notice the use of the weights,
                                #weights = round(bike_HW_alltt$ELE_G_POND_ESC2),
                                optim.method="Nelder-Mead")
HW_bike_lnorm <- lnorm_

logis_ <- fitdistrplus::fitdist(data=walk_HW_alltt$TIEMPO_VIAJES, #%>% unlist() %>% as.numeric(),
                                "logis", 
                                method="mle",
                                #Notice the use of the weights,
                                #weights = round(walk_HW_alltt$ELE_G_POND_ESC2),
                                optim.method="Nelder-Mead")
HW_walk_lnorm <- logis_
```

To visualize the impedance functions we create a data frame and calculate the density of the distribution given the estimated parameters:
```{r}
x <- data.frame(t = seq(1, 300, 1))

x_transit <- x %>%
  mutate(f = dgamma(t,
                    HW_transit_lnorm$estimate[1],
                    HW_transit_lnorm$estimate[2]),
         mode = "transit")

x_car_motor <- x %>%
  mutate(f = dgamma(t,
                    shape = HW_car_motor_gamma$estimate[1],
                    rate = HW_car_motor_gamma$estimate[2]),
         mode = "car/motor")

x_bike <- x %>%
  mutate(f = dlnorm(t,
                    HW_bike_lnorm$estimate[1],
                    HW_bike_lnorm$estimate[2]),
         mode = "bike")

x_walk <- x %>%
  mutate(f = dlogis(t,
                    HW_walk_lnorm$estimate[1],
                    HW_walk_lnorm$estimate[2]),
         mode = "walk")
```

Bind the empirical and theoretical trip length distributions:
```{r}
tld_empirical <- rbind(transit_HW_alltt,
      car_motor_HW_alltt,
      bike_HW_alltt,
      walk_HW_alltt) %>%
  mutate(mode = factor(mode,
                       levels = c("bike", "walk", "car/motor", "transit")),
         distribution = "empirical")

tld_theoretical <- rbind(x_transit,
      x_car_motor,
      x_bike,
      x_walk) %>%
  mutate(mode = factor(mode,
                       levels = c("bike", "walk", "car/motor", "transit")))
```

Plot:
```{r}
ggplot() + 
  geom_histogram(data = tld_empirical,
                 aes(x = TIEMPO_VIAJES, 
                     #weights = ELE_G_POND_ESC2, 
                     after_stat(density)),
                 bins = 45) +
  geom_line(data = tld_theoretical,
            aes(x = t, y = f)) + 
  facet_wrap(~ mode, nrow = 2)

#here, let's double check the empirical trip length distributions, as we can see from the two generated plots, our input TLD is correct!
rbind(transit_HW_alltt,
      car_motor_HW_alltt,
      bike_HW_alltt,
      walk_HW_alltt) %>%
  mutate(mode = factor(mode,
                       levels = c("bike", "walk", "car/motor", "transit"))) %>%
  ggplot(aes(x = TIEMPO_VIAJES, after_stat(density),
             #weights = ELE_G_POND_ESC2
             )) + 
  geom_histogram(bins = 45) + 
  facet_wrap(~ mode, nrow = 2)
```

## Defining P_i, O_j, and f(c_ijm)
Next, we define $P_i$ (for the population balancing factor) for each mode and $O_j$ associated with each zone. To do this, we add the number of trips (by mode) and create objects which contain the number of workers (origin) and jobs (destinations) - by mode:
```{r}
# Instead of creating various columns like above, create a single column with the mode
OD_v_HW <- OD_v_HW  %>% 
  mutate(mode = case_when(MODO_PRIORITARIO <= 9 ~ "transit",
                          MODO_PRIORITARIO <= 19 & MODO_PRIORITARIO >= 10 ~ "car/motor",
                          MODO_PRIORITARIO <= 23 & MODO_PRIORITARIO >= 20 ~ "bike",
                          MODO_PRIORITARIO == 24 ~ "walk"),
         mode = factor(mode,
                       levels = c("bike", "walk", "car/motor", "transit")))

# The workers are classified by mode and origin. Ex. from Origin 1 five people use bike mode to get to destinations. This is P_i for each m
workers <- OD_v_HW %>%
  group_by(VORI_CD_ZT208,
           mode) %>%
  summarize(workers = n(), #we should be summing the number of workers by the weight but we don't have a weight... so for now, 1 row = 1 trip. 
            .groups = "drop") %>%
  rename(CD_ZT208 = VORI_CD_ZT208)

# Calculate the number of jobs per j ($O_j$)
jobs <- OD_v_HW %>%
  group_by(VDES_CD_ZT208) %>%
  summarize(jobs = n(),
            .groups = "drop") %>%
  rename(CD_ZT208 = VDES_CD_ZT208)
```

Next join the number of workers summed at each origin to the ZN simple features object:
```{r}
workers_sf <- workers %>%
  # Join workers
  left_join(zn208,
            by = "CD_ZT208") %>%
  st_as_sf()

jobs_sf <- jobs %>%
  # Join jobs
  left_join(zn208,
            by = "CD_ZT208") %>%
  st_as_sf()
```

Now we reshape the data. Each row will represent an any origin to to 1 destination (i.e., destination row is unique). We will average the travel times, by mode, and then use this average travel time to find the 'f' (impedance value for travel for each mode). Hint: this is why we needed to find the impedance function in the first place! The functions represent the cost of travel for each mode based on the travel time for a trip, since we are averaging together the travel time for each mode.
```{r}
OD <- OD_v_HW %>%
  group_by(VORI_CD_ZT208,
           VDES_CD_ZT208,
           mode) %>%
  summarize(tt = mean(TIEMPO_VIAJES),
            .groups = "drop")

# Check the mean and median travel times by mode
OD %>% 
  group_by(mode) %>% 
  summarize(mtt = mean(tt))
OD %>% 
  group_by(mode) %>% 
  summarize(mtt = median(tt))
```

Join workers and jobs to OD table:
```{r}
OD_2 <- OD %>%
  left_join(workers,
            by = c("VORI_CD_ZT208" = "CD_ZT208", "mode"))

OD_2 <- OD_2 %>%
  left_join(jobs,
            by = c("VDES_CD_ZT208" = "CD_ZT208"))
```

Preview the table for one origin and one destination, as an example:
```{r}
OD_2 %>% filter(VORI_CD_ZT208 == 1 & VDES_CD_ZT208 == 2) #explanation: 'workers' corresponds to the people who leave ORI 1 to go to any destination by mode (P_im) (e.g., 5 workers leave ORI 1 by bike to go to a job somewhere in the region). 'jobs' corresponds to the number of workers that come to DES 2 as a job destination in the region (sum_m O_j). 'tt' corresponds to the average travel time for i to j per mode (avg. c_ijm) (e.g., only 1 trip is made by bike from ORI 1 to DES 2 and it is 5 mins, so the average travel time is 5 mins) -- this c_ijm will be used to calculate f(c_ijm). 
```

Estimate f value using these averaged travel times:
```{r, echo=FALSE, warning=FALSE, message=FALSE}
OD_2 <- OD_2 %>%
  mutate(f = case_when(mode == "transit" ~ dgamma(tt,
                                                  HW_transit_lnorm$estimate["shape"], 
                                                  HW_transit_lnorm$estimate["rate"]),
                       mode == "car/motor" ~ dgamma(tt,
                                                    HW_car_motor_gamma$estimate["shape"],
                                                    HW_car_motor_gamma$estimate["rate"]),
                       mode == "bike" ~ dlnorm(tt,
                                               HW_bike_lnorm$estimate["meanlog"],
                                               HW_bike_lnorm$estimate["sdlog"]),
                       mode == "walk" ~ dlogis(tt, 
                                               HW_walk_lnorm$estimate["location"], 
                                               HW_walk_lnorm$estimate["scale"])))
```

## Calculating spatial availability 
Calculate spatial availability:
```{r}
OD_SA <- OD_2 %>%
  mutate(catch = 1) %>%
  mutate(SA_ij = sp_avail(., 
                          o_id = VORI_CD_ZT208,
                          d_id = VDES_CD_ZT208,
                          pop = workers,
                          opp = jobs,
                          r = catch,
                          f = f))
```

Check that the totals are maintained, it is!:
```{r}
sum(OD_SA$SA_ij, na.rm = T)
sum(jobs$jobs)
#sum(OD_SA$tt)
```

Spatial availability by mode, we can see that car mode captures 71% of the spatial availability but only accounts for 47% of the tt_ijm !!:
```{r}
OD_SA %>%
  group_by(mode) %>%
  summarize(V_m = sum(SA_ij),
            V_m_perc = sum(SA_ij)/27795,
            tt_m = sum(tt)/780315.1)
```

Okay, let's preform a few more checks and then aggregate spatial availability by origin (zn208) and by origin,mode.

Spatial availability by origin (with check). Here we represent spatial availability per origin across all modes for each i. We know that car mode represents the majority of spatial availability per i overall:
```{r}
SA_i <- OD_SA %>%
  group_by(VORI_CD_ZT208) %>%
  summarize(V_i = sum(SA_ij))

summary(SA_i) 
sum(SA_i$V_i)
```

But we can suspect that the spatial distribution of SA is uneven depending on the mode. So here we represent Spatial availability by origin, mode (with check):
```{r}
SA_im <- OD_SA %>%
  group_by(VORI_CD_ZT208, 
           mode) %>%
  summarize(V_im = sum(SA_ij),
            .groups = "drop")

summary(SA_im)
sum(SA_im$V_im)
```

Let's preform some more checks. Calculate worker population by zone of origin:
```{r}
population <- workers %>%
  group_by(CD_ZT208) %>%
  summarize(population = sum(workers))

sum(population$population)
```

Finally, let's join the zonal worker population to the SA_i and SA_im spatial objects (with checks):
```{r}
SA_i <- SA_i %>%
  left_join(population,
            by = c("VORI_CD_ZT208" = "CD_ZT208"))
sum(SA_i$V_i)

SA_im <- SA_im %>%
  left_join(population,
            by = c("VORI_CD_ZT208" = "CD_ZT208"))
sum(SA_im$V_im)
```

To facilitate the interpretation of spatial availability, here we calculate spatial availability per capita:
```{r}
SA_i <- SA_i %>%
  mutate(v_i = V_i/population)

summary(SA_i)

SA_im <- SA_im %>%
  mutate(v_im = V_im/population)

summary(SA_im)
```

Lastly, join geometry of zones to SA so we can visualize:
```{r}
SA_i <- SA_i %>%
  left_join(zn208 %>%
              dplyr::select(CD_ZT208),
            by =c("VORI_CD_ZT208" = "CD_ZT208")) %>%
  st_as_sf()

SA_im <- SA_im %>%
  left_join(zn208 %>%
              dplyr::select(CD_ZT208),
            by = c("VORI_CD_ZT208" = "CD_ZT208")) %>%
  st_as_sf()
```

## Plotting:

Figure out which CD_ZT208 correspond to Madrid - we see that CD_ZT208 1 through 59 do so.
```{r}
t <- zn_u %>% dplyr::select(CD_ZT208, Municipality) %>% group_by(CD_ZT208,Municipality) %>% summarize() %>% filter(Municipality == "Madrid")
t
```

```{r create-bounding-boxes}
zn208 <- zn208 %>% mutate(MM = ifelse(str_starts(ZT208,"Madrid"), 1, 0)) %>% filter(MM == 1)

#create the bounding box for the Toronto region + a little extra
bbox_new1 <- st_bbox(zn208 %>%filter(MM == 1)) # current bounding box

xrange <- bbox_new1$xmax - bbox_new1$xmin # range of x values
yrange <- bbox_new1$ymax - bbox_new1$ymin # range of y values

bbox_new1[1] <- bbox_new1[1] - (0.0 * xrange) # xmin - left
bbox_new1[3] <- bbox_new1[3] + (0.0 * xrange) # xmax - right
bbox_new1[2] <- bbox_new1[2] - (0.0 * yrange) # ymin - bottom
bbox_new1[4] <- bbox_new1[4] + (0.0 * yrange) # ymax - top

bbox_Madrid_muni <- bbox_new1 %>%  # take the bounding box ...
  st_as_sfc()

sf::sf_use_s2(FALSE) #this terns off the spherical geometry component in the st_union() function - this assumes the sf is just planar. Thats' fine for our purposes
Madrid_muni_border <- zn208 %>%
  filter(MM == 1) %>%
  st_union()
```

Plot V (spatial availability) by origin:
```{r}
ggplot() +
  geom_sf(data = SA_i,
          aes(fill = V_i),
          color = NA) +
  scale_fill_distiller(palette = "YlGn", #legend scale bar
                       name = expression(v["i"]),
                       na.value = "grey90",
                       direction = 1,
                       #trans = "log",
                       #breaks=c(0, 10, 100, 1000, 10000)
                       )+
  geom_sf(data = Madrid_muni_border, # border for Madrid muni
          colour=alpha("black",1), 
          size = 0.5, 
          fill=NA, 
          show.legend = "polygon") +
  geom_sf(data = LEZ_centro, 
          colour=alpha("red",1), 
          size = 0.5, 
          fill = NA) + 
  geom_sf(data = LEZ_M30, 
          colour=alpha("cyan",1), 
          size = 0.5, 
          fill = NA) +
  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
                         height = unit(0.8, "cm"), 
                         width = unit(0.8, "cm"),
                         style = north_arrow_orienteering(line_width = 0.25,
                                                          line_col = "dimgrey", 
                                                          fill = c("white","dimgrey"))) +
  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
                   height = unit(0.15, "cm")) +
  theme_void() +
  theme(legend.position = c(1.10, 0.25), 
        plot.margin = margin(t = 0, r = 0, b = 0, l = -80)) # positing legend and clipping out white space
```

Plot v (spatial availability per capita) by origin:
```{r}
ggplot() +
  geom_sf(data = SA_i,
          aes(fill = v_i),
          color = NA) +
  scale_fill_gradient2(name = expression(v["i"]),
                       midpoint = 1,
                       na.value = "grey90")+
  geom_sf(data = Madrid_muni_border, # border for Madrid muni
          colour=alpha("black",1), 
          size = 0.5, 
          fill=NA, 
          show.legend = "polygon") +
  geom_sf(data = LEZ_centro, 
          colour=alpha("red",1), 
          size = 0.5, 
          fill = NA) + 
  geom_sf(data = LEZ_M30, 
          colour=alpha("cyan",1), 
          size = 0.5, 
          fill = NA) +
  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
                         height = unit(0.8, "cm"), 
                         width = unit(0.8, "cm"),
                         style = north_arrow_orienteering(line_width = 0.25,
                                                          line_col = "dimgrey", 
                                                          fill = c("white","dimgrey"))) +
  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
                   height = unit(0.15, "cm")) +
  theme_void() +
  theme(legend.position = c(1.10, 0.25), 
        plot.margin = margin(t = 0, r = 0, b = 0, l = -80)) # positing legend and clipping out white space
```

Plot V (spatial availability) by origin-mode:
```{r}
ggplot() +
  geom_sf(data = SA_im,
          aes(fill = V_im),
          color = NA) +
  scale_fill_distiller(palette = "YlGn", #legend scale bar
                       name = expression(v["i"]),
                       na.value = "grey90",
                       direction = 1,
                       #trans = "log",
                       #breaks=c(0, 10, 100, 1000, 10000)
                       )+
  geom_sf(data = Madrid_muni_border, # border for Madrid muni
          colour=alpha("black",1), 
          size = 0.5, 
          fill=NA, 
          show.legend = "polygon") +
  geom_sf(data = LEZ_centro, 
          colour=alpha("red",1), 
          size = 0.5, 
          fill = NA) + 
  geom_sf(data = LEZ_M30, 
          colour=alpha("cyan",1), 
          size = 0.5, 
          fill = NA) +
  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
                         height = unit(0.8, "cm"), 
                         width = unit(0.8, "cm"),
                         style = north_arrow_orienteering(line_width = 0.25,
                                                          line_col = "dimgrey", 
                                                          fill = c("white","dimgrey"))) +
  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
                   height = unit(0.15, "cm")) +
  theme_void() +
  theme(legend.position = c(1.10, 0.25), 
        plot.margin = margin(t = 0, r = 0, b = 0, l = -80)) + # positing legend and clipping out white space
  facet_wrap(~mode, nrow = 2)
```

Plot v (spatial availability per capita) by origin:
```{r}
ggplot() +
  geom_sf(data = SA_im,
          aes(fill = v_im),
          color = NA) +
  scale_fill_gradient2(name = expression(v["i"]),
                       midpoint = 1,
                       na.value = "grey90")+
  geom_sf(data = Madrid_muni_border, # border for Madrid muni
          colour=alpha("black",1), 
          size = 0.5, 
          fill=NA, 
          show.legend = "polygon") +
  geom_sf(data = LEZ_centro, 
          colour=alpha("red",1), 
          size = 0.5, 
          fill = NA) + 
  geom_sf(data = LEZ_M30, 
          colour=alpha("cyan",1), 
          size = 0.5, 
          fill = NA) +
  annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset
                         height = unit(0.8, "cm"), 
                         width = unit(0.8, "cm"),
                         style = north_arrow_orienteering(line_width = 0.25,
                                                          line_col = "dimgrey", 
                                                          fill = c("white","dimgrey"))) +
  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
                   height = unit(0.15, "cm")) +
  theme_void() +
  theme(legend.position = c(1.10, 0.25), 
        plot.margin = margin(t = 0, r = 0, b = 0, l = -80)) + # positing legend and clipping out white space
  facet_wrap(~mode, nrow = 2)
```

# Scenarios

## Overview and data

All scenarios use the 2018 travel survey (zn208) - home-to-work work trips, by mode. We extract the travel time and number of trips (by mode) from each origin zone and destination (by mode) to each destination zone.

The scenarios:

- S1) During LEZ-centro (2018 travel survey, no assumptions) -- **done, in the chunks above (with 1 limitation -- we are missing weights for the travel survey!)**

- S0) Before LEZ-centro (2018 travel survey, w/ Tarri thesis survey assumptions) -- **attempted below**
- S2) During LEZ-M30 (2018 travel survey w/ E-MORES modal shift assumptions) -- **attempted below**

- S3) Future Scenario in 2022: LEZ-M30 w/ no Labels B,C ? 
- S4) Future Scenario in 2025: LEZ-M50 ?
- S5) Future Scenarios in 2025: LEZ-M50 w/ no Labels B,C ?

## S0) Before LEZ-centro (2018 travel survey, w/ Tarri thesis survey assumptions)

From the survey conducted during Tarri's thesis, 50% of trips made by car to the centro were shifted to some other mode. So, let's assume same number of trips, same travel times, but we will just re-shift them modes of the trips that go into the centro.

```{r}
# tagging the trips that go into centro LEZ and that go into the M30 LEZ
OD_2 <- OD_2 %>% mutate(LEZCent = ifelse(VDES_CD_ZT208 %in% LEZ_centro_list, "Yes", NA),
                              LEZM30 = ifelse(VDES_CD_ZT208 %in% LEZ_M30_list, "Yes", NA))

#count of car trips going into the LEZ Centroid
OD_v_HW %>% filter(LEZCent == "Yes" & mode == "car/motor") %>% count()
```

1524 trips -- represents ~5% of the (un-weighted) trips. So we can assume that this portion of trips is doubled, and those pre-LEZ trips are made on non-car surveys but are shifted to car modes. 
```{r}
prop_table <- OD_2 %>% group_by(VORI_CD_ZT208, VDES_CD_ZT208) %>%
  mutate(S0 = ifelse(mode=="car/motor" & LEZCent == "Yes", workers, NA),
         S0 = sum(S0, na.rm=T),
         prop_mode = workers/sum(workers)) %>%
  filter(mode != "car/motor") %>%
  mutate(prop_mode_wo_car = ifelse(mode != "car/motor", sum(prop_mode), NA)) %>%
  dplyr::select(VORI_CD_ZT208, VDES_CD_ZT208, mode, prop_mode, prop_mode_wo_car, S0)

test <- prop_table %>% merge(OD_2 %>% group_by(VORI_CD_ZT208, VDES_CD_ZT208) %>% mutate(S0_car = ifelse(mode=="car/motor" & LEZCent == "Yes", workers, NA))
                             , by = c("VORI_CD_ZT208","VDES_CD_ZT208", "mode"), all.y =TRUE)

test <- test %>% mutate(S0_1 = ifelse(LEZCent == "Yes",
                              S0*(prop_mode/prop_mode_wo_car)),
                workers_S0 = workers - S0_1,
                workers_S0_1 = workers + S0_car,
                workers_S0_2 = coalesce(workers_S0,workers_S0_1),
                workers_S0_3 = coalesce(workers_S0_2, workers))

test
```

Calculate spatial availability:
```{r}
OD_SA_2 <- test %>%
  mutate(catch = 1) %>%
  mutate(SA_ij = sp_avail(., 
                          o_id = VORI_CD_ZT208,
                          d_id = VDES_CD_ZT208,
                          pop = workers_S0_3,
                          opp = jobs,
                          r = catch,
                          f = f))

sum(OD_SA_2$SA_ij, na.rm = T)
sum(jobs$jobs)
```

```{r}
OD_SA %>%
  group_by(mode) %>%
  summarize(V_m = sum(SA_ij),
            V_m_perc = sum(SA_ij)/27795)

OD_SA_2 %>%
  group_by(mode) %>%
  summarize(V_m = sum(SA_ij),
            V_m_perc = sum(SA_ij)/27795)
```

## S2) During LEZ-M30 (2018 travel survey w/ E-MORES modal shift assumptions)

During LEZ-M30 (w/assumptions: same number of trips, just car trips are shifted to transit )

# S3), S4), and S5)
- S3) Future Scenario: LEZ-M30 w/ no Labels B,C 
- S4) Future Scenario: LEZ-M50
- S5) Future Scenarios: LEZ-M50 w/ no Labels B,C

## Issues/questions
- We need weights for travel survey!!
- Population assumptions are needed: trip # don't stay the same in 2017, 2018, 2022, 2025. 
- Travel time assumptions are needed? Unsure how to address this.
- Are assumptions for SO and S2 good enough?
- No assumptions are made about through-traffic, is this okay? 
- Decide on future scenario assumptions.
- What census variables do we need for equity analysis?
- Emissions?

### 2023-07-30 Multi-mode-acc-methods.qmd
Presentation file commit pt.2. and draft manuscript documents.

@@
 format: docx
-bibliography: bibliography.bib
+bibliography: mybibfile.bib
 ---
@@
 
-ANother example of proportional allocation, BFCA
+Another example of proportional allocation, BFCA
 
@@
 Currently, conventional _unconstrained_ accessibility measures are difficult for planners to operationalize for a variety of reasons.
-Furthermore, exisiting competitive measures are difficult to interpret.
+Furthermore, existing competitive measures are difficult to interpret.
 
@@
 This is the aim of this paper.
-We plan to expland to this measure to multi-modal systems.
-opportunities are avialble to users of different modes.
+We plan to expand to this measure to multi-modal systems.
+opportunities are available to users of different modes.
 The cost of reaching them is not uniform.
-
-<!-- An example of a case where conventional accessibility measures (unconstrained and competitive) fall short is in considering multi-modal accessibility.
-Take for instance, measuring the accessibility to opportunities for a region for car and transit modes.
-The literature suggests that these modes are in direct competition.
-A user, if they have access to both modes, will consider the utility (travel impedance) of using one or the other.
-Thus, the potential for the population to interact with opportunities is directly dependant on ...
-the opportunities..
--->
-
-<!-- introduction of multi-modal accessibility. when modes compete with one another. at the center of this is that modes SHOULD compete. competition considerations is required.-->
 
@@
 
-# Illustrative example: using spatial availability for multiple modes
+# Illustrative example: operationalizing spatial availability for multiple modes
 
@@
 The general form of spatial availability is a sum: this is similar to Hansen-type accessibility (popular _unconstrained_ measure) and Shen-type accessibility (popular _competitive_ and _unconstrained_ measure).
-What makes spatial availability a _constrained_ measure are the balancing factors that allocate opportunities proportionally.
+What makes spatial availability both competitive and constrained are the balancing factors that allocate opportunities proportionally.
 The balancing factor $F^t_{ij}$ consists of two components: a population-based balancing factor $F^p_{i}$ and an impedance-based balancing factor $F^c_{ij}$ that, respectively, allocate opportunities to $i$ in proportion to the size of the population of the different competing centers (the mass effect of the gravity model) and the cost of reaching opportunities (the impedance effect).
+Spatial availability considers demand for opportunities based on the magnitude of demand for opportunities and the travel-cost to reach those opportunities (i.e., competition) as well as preserving the sum of opportunities (i.e., constrained) such that scores are a proportion of the total number of opportunities in the region.
+
+<!-- An example of a case where conventional accessibility measures (unconstrained and competitive) fall short is in considering multi-modal accessibility.
+Take for instance, measuring the accessibility to opportunities for a region for car and transit modes.
+The literature suggests that these modes are in direct competition.
+A user, if they have access to both modes, will consider the utility (travel impedance) of using one or the other.
+Thus, the potential for the population to interact with opportunities is directly dependant on the opportunities t  -->
+
+<!-- introduction of multi-modal accessibility. when modes compete with one another. at the center of this is that modes SHOULD compete. competition considerations is required.-->
+
+See @soukhovIntroducingSpatialAvailability2023 for details, but in short, the proportional allocation mechanism of $F^t_{ij}$ ensures that each $V_i$ value assigned to each $i$ sums to the total number of opportunities.
+This is important, because opportunities in a region are finite.
+If one person interacts with the opportunity at a given time, it is taken.
+In this way, if one person has the ability to reach more opportunities through a faster mode, for instance, than that person has more opportunities they can potentially interact with than other people.
+In a region with finite opportunities, one person being able to access more opportunities means there are fewer opportunities available for people using slower modes.
+
+can be formulated to consider  characteristics of different population groups.
+For instance, consider two sub-populations at population center $A$.
+One sub-population uses mode $x$ and the other uses mode $z$ to access jobs at employment center $1$.
+Mode $x$ is faster than mode $z$, so the sub-population that use $x$ can potentially access more jobs at employment center $1$ than the other sub-population as a result of the lower travel cost.
+Furthermore, the sub-population using the faster mode $x$ can potentially access more jobs at further employment centers (say $2$ and $3$) than the sub-population using the slower mode.
+This puts the sub-population using the faster (lower travel cost) mode $x$ at an accessibility advantage.
+If that sub-population is also within a population center that is low-population (relative to the region) and central located to employment centers (relative to the region), then this sub-population using the faster mode $x$ is the most advantaged of all sub-populations using any mode or located at any population center.
+
+The illustrative example described is represented in Figure \ref{fig:Fig1} and was inspired by the single-mode example used in @shen1998 and reconfigured in @soukhovIntroducingSpatialAvailability2023.
+Population center $A$ can be seen as a Suburban population center; it is closest to its own relatively large employment center at $A$, close to the Urban area's equally large employment center $B$, and has a population that is smaller than the Urban center $2$ and larger than the rural center $3$.
+Consider the number of jobs to be finite in the region.
+From the perspective of accessibility, the sub-population that is in a center that is the least populated, is the most centrally located to largest amount of jobs, is located away from densely populated centers, and uses the lowest travel cost mode $x$ can potentially access the most job opportunities.
+From the opposite perspective, sub-populations using the slower mode $z$ and especially those located in population centers that are not centrally located to opportunities and/or close to highly dense population centers, are at an access disadvantage.
+The number of job opportunities are finite, and the sub-populations using faster modes and/or centrally located take a larger proportion of opportunities, leaving fewer for other populations.
 
@@
 However, how do the populations access these jobs?
-In this synthetic example, the population either uses the regular mode $z$ or uses the faster mode $x$.
+In this example, the population either uses the regular mode $z$ or uses the faster mode $x$.
 Their choice in mode is already assigned: the mode selected can be seen as a characteristic of the population, or a way to subset the total population.
@@
 #see which curve will fit the best using the descdist() function. This is to calibrate an impedance function for each mode based on the travel survey
-
-#NOTE: maybe we can try a distance based distribution???? Not really convinced why would could - perhaps to include all modes?
 
@@
                                                HW_walk_lnorm$estimate["meanlog"], 
-                                               HW_walk_lnorm$estimate["sdlog"])))
+                                               HW_walk_lnorm$estimate["sdlog"])),
+         jobs_f = jobs*f) #for Hansen-type accessibility (unconstrained)
 
@@
                                                HW_walk_lnorm$estimate["meanlog"], 
-                                               HW_walk_lnorm$estimate["sdlog"])))
+                                               HW_walk_lnorm$estimate["sdlog"])),
+         jobs_f = jobs*f) #for Hansen-type accessibility (unconstrained)
+```
+
+## Calculating hansen-type accessibility (for comparision)
+```{r, echo=FALSE, warning=FALSE, message=FALSE}
+OD_Si <- OD_2 %>%
+  group_by(VORIZT1259, mode) %>%
+  summarize(S_i = sum(jobs_f)) # Hansen-type accessibility (unconstrained)
+
+#also for zn208
+OD_Si_zn208 <- OD_3 %>%
+  group_by(VORI_CD_ZT208, mode) %>%
+  summarize(S_i = sum(jobs_f)) # Hansen-type accessibility (unconstrained)
 ```
@@
   summarize(V_im = sum(SA_ij),
+            population_m = mean(workers),
             .groups = "drop")
@@
 SA_im <- SA_im %>%
+  left_join(zn1259 %>%
+              dplyr::select(ZT1259),
+            by = c("VORIZT1259" = "ZT1259")) %>%
+  st_as_sf()
+
+OD_Si <- OD_Si %>%
   left_join(zn1259 %>%
@@
 SA_im_zn208 <- SA_im_zn208 %>%
+  left_join(zn208 %>%
+              dplyr::select(CD_ZT208),
+            by = c("VORI_CD_ZT208" = "CD_ZT208")) %>%
+  st_as_sf()
+
+OD_Si_zn208 <- OD_Si_zn208 %>%
   left_join(zn208 %>%
@@
 ```{r}
+#mapping population by mode!
+ggplot() +
+  geom_sf(data = SA_im_zn208,
+          aes(fill = population_m),
+          color = NA) +
+  scale_fill_distiller(palette = "Reds", #legend scale bar
+                       na.value = "grey90",
+                       direction = 1)+
+  geom_sf(data = Madrid_muni_border, # border for Madrid muni
+          colour=alpha("black",1), 
+          size = 0.5, 
+          fill=NA, 
+          show.legend = "polygon") +
+  geom_sf(data = LEZ_centro, 
+          colour=alpha("red",1), 
+          size = 0.5, 
+          fill = NA) + 
+  geom_sf(data = LEZ_M30, 
+          colour=alpha("cyan",1), 
+          size = 0.5, 
+          fill = NA) +
+  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
+                   height = unit(0.15, "cm")) +
+  theme_void() +
+  theme(legend.key.width = unit(0.2, 'cm')) +
+  facet_wrap(~mode, nrow = 2)
+ggsave("Presentations/im_populations_zn208_plot.png")
+```
+
+```{r}
 SA_i_V_zn208_plot
@@
 
-Plot V (spatial availability) by origin-mode (first zn1259 then zn208):
+Plot V (spatial availability) by origin-mode (first zn1259 then zn208). And for comparision, plot hansen-type accessibility by origin-mode (first zn1259 then zn208):
 ```{r}
@@
 #zn208
+#the accessibility plot (Hansen-type)
+S_im_plot_zn208_plot <- ggplot() +
+  geom_sf(data = OD_Si_zn208,
+          aes(fill = S_i),
+          color = NA) +
+  scale_fill_distiller(palette = "YlGn", #legend scale bar
+                       name = expression(S["im"]),
+                       na.value = "grey90",
+                       direction = 1)+
+  geom_sf(data = Madrid_muni_border, # border for Madrid munim
+          colour=alpha("black",1), 
+          size = 0.5, 
+          fill=NA, 
+          show.legend = "polygon") +
+  geom_sf(data = LEZ_centro, 
+          colour=alpha("red",1), 
+          size = 0.5, 
+          fill = NA) + 
+  geom_sf(data = LEZ_M30, 
+          colour=alpha("cyan",1), 
+          size = 0.5, 
+          fill = NA) +
+  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
+                   height = unit(0.15, "cm")) +
+  theme_void() +
+  theme(legend.key.width = unit(0.2, 'cm')) +
+  facet_wrap(~mode, nrow = 2)
+
 SA_im_V_zn208_plot <- ggplot() +
@@
 ```{r}
+S_im_plot_zn208_plot
+ggsave("Presentations/Hansen_im_V_zn208_plot.png")
+```
+
+```{r}
 SA_im_V_zn208_plot

### 2023-07-30 TRB-manuscript/TRB-manuscript.Rmd
TRB draft - ready for AP edits

---
title: Multimodal spatial availability - the case of low emission zones in Madrid
runningheader: "Multimodal spatial availability - the case of low emission zones in Madrid"
author:
  - name: Anastasia Soukhov
    note: Corresponding Author
    position: PhD Candidate
    email: soukhoa@mcmaster.ca
    affiliation: Department of Earth, Environment and Society, McMaster University, Canada
  - name: Javier Tarriño Ortiz
    position: PhD Graduate
    email: bob@example.com
    affiliation: Universidad Politécnica de Madrid, Spain
  - name: Julio A. Soria-Lara
    position: Associate Professor
    email: derek@example.com
    affiliation: Urban and Regional Planning, Universidad Politécnica de Madrid, Spain
  - name: Antonio Páez
    position: Associate Professor
    email: paezha@mcmaster.ca
    affiliation: Department of Earth, Environment and Society, McMaster University, Canada
abstract: |
  An increasing number of studies within the domain of transportation planning are concerned with the inequities in accessibility to opportunities. A dimension of these inequities arises from differences in access by mode type (e.g., commuting using a car as opposed to transit). However, methods implemented in current accessibility literature are lacking within the context of multi-modal analysis. This paper presents an extension of spatial availability, a singly-constrained competitive accessibility measure, for the context of multi-modal accessibility analysis. We first illustrate the features of spatial availability that lend itself to multi-modal analysis. We then demonstrate its use on the case study of Low Emission Zones in Madrid (Spain) and highlight how this policy intervention changes the accessibility of populations using different modes. In summary, spatial availability can be used to create and interpret multi-modal policy intervention scenarios unlike previous methods: this creation and interpretation can help regions envision a more sustainable and equitable access-to-opportunity landscape.

keywords: ["Multimodal", "Accessibility", "Equity", "Policy scenarios", "Low emission zones"]
bibliography: mybibfile.bib
biblio-style: unsrtnat
classoption:
  - numbered #  When numbered option is activated, lines are numbered.
output: rticles::trb_article
wordcount: 7500
---
```{r knitr-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  cache = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = '', 
  out.width = "1\\linewidth")

options(scipen=999) #eliminate scientific notation in print outs.
```

```{r load-packages, include=FALSE, cache=FALSE}
# library(fitdistrplus) # Help to Fit of a Parametric Distribution to Non-Censored or Censored Data
# library(gridExtra) # Miscellaneous Functions for "Grid" Graphics 
# library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax

# library(shadowtext) # Shadow Text Grob and Layer
# library(skimr) # Compact and Flexible Summaries of Data
# library(spdep) # Spatial Dependence: Weighting Schemes, Statistics
# library(tidyr) # Tidy Messy Data
# options(scipen = 999)

library(patchwork) # The Composer of Plots
library(sf) # Simple Features for R
library(dplyr) # A Grammar of Data Manipulation
library(ggplot2) # Create Elegant Data Visualisations Using the Grammar of Graphics
library(ggforce) # Accelerating 'ggplot2'
library(geomtextpath) # Curved Text in 'ggplot2'
library(ggpmisc) # Miscellaneous Extensions to 'ggplot2'
library(ggrepel) # Automatically Position Non-Overlapping Text Labels with 'ggplot2'
library(ggspatial) # Spatial Data Framework for ggplot2
library(flextable) # Functions for Tabular Reporting
library(scales) # Scale Functions for Visualization
library(TAM)
library(officer)
library(ggnewscale)
```

```{r sp_avail-function,include=FALSE}
#defining the spatial availability function
sp_avail <- function(x, o_id, d_id, pop, opp, r, f, alpha = 1){

  o_id <- rlang::enquo(o_id)
  d_id <- rlang::enquo(d_id)
  pop <- rlang::enquo(pop)
  opp <- rlang::enquo(opp)
  r <- rlang::enquo(r)
  f <- rlang::enquo(f)

  # Sum of population in system
  sum_pop <- x %>%
    dplyr::distinct(!!o_id,
                    .keep_all = TRUE) %>%
    dplyr::mutate(sum_pop = !!r*(!!pop)^alpha) %>%
    dplyr::pull(sum_pop) %>%
    sum()

  # Balancing factor size
  f_p <- dplyr::pull(x, !!r) * dplyr::pull(x, !!pop)^alpha / sum_pop

  # Sum of impedance
  sum_impedance <- x %>%
    dplyr::group_by(!!d_id) %>%
    dplyr::summarize(sum_impedance = sum(!!f))

  # Join sum of impedance to table
  x <- x %>%
    dplyr::left_join(sum_impedance,
                     by = rlang::as_name(d_id))

  # Balancing factor impedance
  f_c <- dplyr::pull(x, !!f) / x$sum_impedance

  # Add balancing factors to table
  x$f_c <- f_c
  x$f_p <- f_p

  # Sum of balancing factors for mass and impedance
  sum_pa <- x %>%
    dplyr::group_by(!!d_id) %>%
    dplyr::summarize(sum_pa= sum(f_p * f_c))

  x <- x %>%
    dplyr::left_join(sum_pa,
                     by = rlang::as_name(d_id))

  # Joint balancing factor
  f_t <- (f_p * f_c) / dplyr::pull(x, sum_pa)

  dplyr::pull(x, !!opp) * f_t
}

#detailed
sp_avail_detailed <- function(x, o_id, d_id, pop, opp, r, f, alpha = 1){

  o_id <- rlang::enquo(o_id)
  d_id <- rlang::enquo(d_id)
  pop <- rlang::enquo(pop)
  opp <- rlang::enquo(opp)
  r <- rlang::enquo(r)
  f <- rlang::enquo(f)

  sum_pop <- x %>%
    dplyr::distinct(!!o_id,
                    .keep_all = TRUE) %>%
    dplyr::mutate(sum_pop = !!r*(!!pop)^alpha) %>%
    dplyr::pull(sum_pop) %>%
    sum()

  f_p <- dplyr::pull(x, !!r) * dplyr::pull(x, !!pop)^alpha / sum_pop

  sum_impedance <- x %>%
    dplyr::group_by(!!d_id) %>%
    dplyr::summarize(sum_impedance = sum(!!f))

  x <- x %>%
    dplyr::left_join(sum_impedance,
                     by = rlang::as_name(d_id))

  f_c <- dplyr::pull(x, !!f) / x$sum_impedance

  x$f_c <- f_c
  x$f_p <- f_p

  sum_pa <- x %>%
    dplyr::group_by(!!d_id) %>%
    dplyr::summarize(sum_pa= sum(f_p * f_c))

  x <- x %>%
    dplyr::left_join(sum_pa,
                     by = rlang::as_name(d_id))
  x$f_t <- (f_p * f_c) / dplyr::pull(x, sum_pa)

  x %>%
    dplyr::mutate(V_ij = !!opp * f_t)
}
```

```{r synthetic-data, include=FALSE}
od_tt <- data.frame(i = c("A", "A", "A", "B", "B", "B", "C", "C", "C"), # Three origins
                    j = c("1", "2", "3", "1", "2", "3", "1", "2", "3"), # Three destinations
                    tt_x = c(15, 30, 100, 30, 15, 100, 100, 100, 15), # Travel time, x-mode(original) 
                    tt_z = c(10, 25, 80, 25, 10, 80, 80, 80, 10), # Travel time, z-mode  (the faster mode - like car)
                    pop = c(50000, 50000, 50000, 150000, 150000, 150000, 10000, 10000, 10000), # Population
                    opp = c(100000, 100000, 10000, 100000, 100000, 10000, 100000, 100000, 10000)) # Jobs

#write the dataframe long.
od_tt_m <- data.frame(i = c("A", "A", "A", "A", "A", "A",
                            "B", "B", "B", "B", "B", "B",
                            "C", "C", "C", "C", "C", "C"), # Three origins
                    j = c("1", "2", "3", "1", "2", "3", 
                          "1", "2", "3", "1", "2", "3", 
                          "1", "2", "3", "1", "2", "3"), # Three destinations
                    m = c("z", "z", "z", "x", "x", "x",
                          "z", "z", "z", "x", "x", "x",
                          "z", "z", "z", "x", "x", "x"),
                    tt = c(10, 25, 80,15, 30, 100,  #travel time, z-mode (faster) AND THEN x-mode (the slower mode, original)
                           25,  10,  80, 30, 15, 100, 
                           80, 80, 10, 100, 100, 15), 
                    pop = c(33000, 33000, 33000, 16500, 16500, 16500,
                            90000, 90000, 90000, 60000, 60000,60000,
                            7000,7000, 7000, 3000, 3000, 3000),
                    opp = c(100000, 100000, 10000, 100000, 100000, 10000,
                            100000, 100000, 10000, 100000, 100000, 10000,
                            100000, 100000, 10000, 100000, 100000, 10000)) 
```

```{r data-figure-with-toy-example, include=FALSE}
od <- data.frame(id = c("A", "B", "C", "1", "2", "3"),
                 type = c("Population", "Population", "Population", "Jobs", "Jobs", "Jobs"),
                 size = c(50000, 150000, 10000, 100000, 100000, 10000),
                 x = c(2.5, 2.5, 6.5, 0.5, 0.5, 6.5),
                 y = c(8.0, 2.5, 4.5, 7.5, 2.5, 2.5))

centers <- data.frame(id = c("Urban center", "Suburb", "Satellite town"),
                 radius = rep(1.3, 3),
                 x = c(1.5, 1.5, 6.5),
                 y = c(2.5, 7.5, 3.5))

# od_lines coded as segments
od_lines <- data.frame(x = c(2.5, 2.5, 2.5, 2.5, 2.5, 2.5, 6.5, 6.5, 6.5),
                       y = c(8.0, 8.0, 8.0, 2.5, 2.5, 2.5, 4.5, 4.5, 4.5),
                       xend = c(0.5, 0.5, 6.5, 0.5, 0.5, 6.5, 0.5, 0.5, 6.5),
                       yend = c(7.5, 2.5, 2.5, 7.5, 2.5, 2.5, 7.5, 2.5, 2.5),
                       group = c("A", "A", "A", "B", "B", "B", "C", "C", "C"),
                       time = c("15 min (x)\n10 min (z)", "30 min (x)\n25 min (z)", "100 min (x)\n80 min (z)",
                                "30 min (x)\n25 min (z)", "15 min (x)\n10 min (z)", "100 min (x)\n80 min (z)",
                                "100 min (x)\n80 min (z)", "100 min (x)\n80 min (z)", "15 min (x)\n10 min (z)"))

od_table <- data.frame(Center = c("A", "B", "C", "1", "2", "3"),
  Size = c("50,000 pop (33% x)", "150,000 pop (40% x)", "10,000 pop (30% x)", 
           "100,000 jobs", "100,000 jobs", "10,000 jobs"))

od_table2 <- data.frame(Path = c("A to 1", "A to 2", "A to 3", 
                                "B to 1", "B to 2", "B to 3", 
                                "C to 1", "C to 2", "C to 3"),
  Value = c("15 min (x)\n10 min (z)", "30 min (x)\n25 min (z)", "100 min (x)\n80 min (z)",
                                "30 min (x)\n25 min (z)", "15 min (x)\n10 min (z)", "100 min (x)\n80 min (z)",
                                "100 min (x)\n80 min (z)", "100 min (x)\n80 min (z)", "15 min (x)\n10 min (z)"))
```

```{r, creating-toy-example-plot, eval=FALSE}
toy_example_plot <- ggplot() + 
  # Plot centers
  geom_circle(data = centers,
             aes(x0 = x, 
                 y0 = y,
                 r = radius),
             color = "grey40") +
  annotate(geom = "label", 
           x=c(0.5, 0.6, 7),
           y=c(9, 1, 5.2), 
           label = c("Suburban", "Urban", "Satellite"), 
           size=4, 
           label.size=NA) +
  # Connect origins and destinations with curves
  geom_curve(data = od_lines,
               aes(x = x,
                   y = y,
                   xend = xend,
                   yend = yend,
                   linetype = group,
                   group = group),
             curvature = 0.20,
             linewidth = 1,
             color = "black") +
  # Text on curves
  geom_textcurve(data = od_lines,
               aes(x = x,
                   y = y,
                   xend = xend,
                   yend = yend,
                   linetype = group,
                   group = group,
                   label = time),
              size = 3,
             curvature = 0.20,
             color = "black") +
  # Plot origins and destinations
  geom_point(data = od,
            aes(size = size,
                x = x, 
                y = y, 
                color = type,
                shape = type)) +
  scale_size(range = c(4, 10)) +
  # Label origins and destinations
  geom_text(data = od, 
            aes(x, 
                y, 
                label=id), 
            size=3) +
  # Create table
  annotate(geom = "table",
           x = 8, 
           y = 9, 
           label = list(od_table),
           fill = "white",
           size=3) +
  # Theme the plot
  coord_fixed() + 
  theme_void() + guides(shape = "none", size = "none", color = "none", linetype ="none")

toy_example_plot
ggsave("images/Fig1.png",dpi=600)
```

```{r access-calcs-table, include=FALSE}
#table with all the accessibility calculations
# Impedance parameter
b <- 0.1

# Calculate impedance function
od_tt_m <- od_tt_m %>%
  mutate(f = exp(-b * tt))
```

```{r access-Si-calc, include=FALSE}
S_im <- od_tt_m %>% 
  group_by(i,m) %>% 
  summarize(S_im = sum(opp*f))
            #,pop = mean(pop)

#not needed for this comparison
# S_i <- S_im %>%
#   group_by(i)%>%
#   summarise(S_i = sum(S_im*pop)/sum(pop))
```

```{r access-ai-calc, include=FALSE}
LOS_j <- od_tt_m %>% 
  group_by(j) %>%
  summarize(opp = mean(opp),
            R_j = sum(pop * f)) %>%
  mutate(LOS_j = opp/R_j)

a_im <- od_tt_m %>% 
  merge(LOS_j, by=c("j")) %>%
  group_by(i,m) %>%
  summarize(a_im = sum(LOS_j * f),
            pop = mean(pop))

a_i <- a_im %>%
  group_by(i) %>%
  summarize(a_i = sum(a_im*pop)/sum(pop))

a_i <- a_im %>% merge(a_i, by="i", all.x = T) %>% select(-c("pop"))
```

```{r calculating-beta-0.1-Vi-table-creation}
V_ijm <- od_tt_m %>%
  mutate(catch = 1) %>%
  sp_avail_detailed(o_id = i, #spatial availability
                    d_id = j, 
                    pop = pop,
                    opp = opp,
                    r = catch,
                    f = f, 
                    alpha = 1)

V_im <- V_ijm %>% 
  group_by(i,m) %>%
  summarise(V_im = sum(V_ij),
            pop = first(pop))

V_i <- V_im %>%
  group_by(i) %>%
  summarize(V_i = sum(V_im))

V_i <- V_im %>% merge(V_i, by="i", all.x = T)
```

```{r creating-synth-table}
table_toy_format <- V_i %>% select(-c(pop)) %>%
  left_join(S_im, by = c("i","m")) %>%
  left_join(a_i, by = c("i","m")) %>%
  transmute(i,
            m,
            S_im,
            a_im,
            V_im,
            a_i,
            V_i,
            V_im = ifelse(V_im < 0.001, "< 0.001", round(V_im, 3)),
            S_im = ifelse(S_im < 0.001, "< 0.001", round(S_im, 3)),
            a_im = ifelse(a_im < 0.001, "< 0.001", round(a_im, 3)),
            V_i = ifelse(V_i < 0.001, "< 0.001", round(V_i, 3)),
            a_i = ifelse(a_i < 0.001, "< 0.001", round(a_i, 3))
            )

add_data <- data.frame(i = "TOTALS", m = "" , 
                       S_im = table_toy_format$S_im %>% sum(),
                       a_im = NA,
                       V_im = table_toy_format$V_im %>% sum(),
                       a_i = NA,
                       V_i = 210000
                       )

table_toy_format <- bind_rows(table_toy_format, add_data)
```

```{r creating-flex-table}
table_toy_format_flex <- flextable(table_toy_format) %>% colformat_double(big.mark = ",", digits = 2, na_str = "N/A")

table_toy_format_flex <- compose(table_toy_format_flex,
                                 i = 1, j = c(3), part = "header",
                                 value = as_paragraph("S", as_sub("i"), as_sup("m")))%>%
  compose(
    i = 1, j = c(4), part = "header",
    value = as_paragraph("a", as_sub("i"), as_sup("m"))) %>%
  compose(i = 1, j = c(5), part = "header",
          value = as_paragraph("V", as_sub("i"), as_sup("m"))) %>%
  compose(
    i = 1, j = c(6), part = "header",
    value = as_paragraph("a", as_sub("i"))) %>%
  compose(
    i = 1, j = c(7), part = "header",
    value = as_paragraph("V", as_sub("i"))) %>%
  merge_v(j=c(1)) %>%
  merge_v(j=c(6)) %>%
  merge_v(j=c(7)) %>%
  vline(j = c(2,5)) %>%
  hline(i = c(2,4,6)) %>% 
  autofit()

table_toy_format_flex
```

```{r load-data}
load("data-prep/data/zn1259.RData") 
zn1259 <- zn %>% dplyr::select(c("FID", "CD_ZT12", "ZT1259"))
remove(zn)
load("data-prep/data/zn208.RData")
load("data-prep/data/zn_u.RData") # a key to join the zn1259 to the zn208 + contains municipality names

zn1259 <-  merge(zn1259, zn_u, by="ZT1259")

LEZ_centro <- read_sf("data-prep/data-raw/LEZ.shp") %>% st_transform(4326) %>% mutate(merge_all = 1) %>%
    group_by(merge_all) %>%
    summarise(geometry = sf::st_union(geometry)) %>%
    ungroup()
#add M30 for visualizing
LEZ_M30 <- read_sf("data-prep/data-raw/Zona_Bajas_Emisiones_MADRID.shp") %>% st_transform(4326)

#creating a list of zones that correspond to the zn208, which indicate if they are in a LEZ.
# LEZ_centro_list <- st_intersection(zn208, LEZ_centro) %>% 
#   st_drop_geometry() %>% 
#   dplyr::select(CD_ZT208) %>% 
#   unlist()

# LEZ_M30_list <- st_intersection(zn208, LEZ_M30) %>% 
#   st_drop_geometry() %>% 
#   dplyr::select(CD_ZT208) %>% 
#   unlist()

#import travel times
load("./data-prep/data-raw/intermed/ttresults/calc_bikett.Rdata")
load("./data-prep/data-raw/intermed/ttresults/calc_walktt.Rdata")
load("./data-prep/data-raw/intermed/ttresults/calc_transittt.Rdata")
load("./data-prep/data-raw/intermed/ttresults/calc_cartt.Rdata")
```

```{r travel-data-calculated-filtered}
zn1259_mad <- zn1259 %>% filter(Municipality == "Madrid")
zn1259_mad_list <- zn1259_mad %>% st_drop_geometry() %>% dplyr::select(ZT1259)

calc_walktt <- calc_walktt %>% filter((VORIZT1259 %in% zn1259_mad_list$ZT1259) & (VDESZT1259 %in% zn1259_mad_list$ZT1259))
calc_cartt <- calc_cartt %>% filter((VORIZT1259 %in% zn1259_mad_list$ZT1259) & (VDESZT1259 %in% zn1259_mad_list$ZT1259))
calc_transittt <- calc_transittt %>% filter((VORIZT1259 %in% zn1259_mad_list$ZT1259) & (VDESZT1259 %in% zn1259_mad_list$ZT1259))
calc_bikett <- calc_bikett %>% filter((VORIZT1259 %in% zn1259_mad_list$ZT1259) & (VDESZT1259 %in% zn1259_mad_list$ZT1259))
```

```{r travel-data-from-survey}
#import travel survey 2018
load(file = "data-prep/data/OD_v.RData")

#Keep only the direct home to work trips for the spatial availability calculation (no chained home-to-x-to-work trips, they only make up a small portion of home-to-work trips anyways, see the exploration of this in the 'data-exploration.rmd' file)
OD_v_HW <- OD_v %>% filter(VORI == "1" & VDES == "2" &
                             VORI_Municipality == "Madrid" & VDES_Municipality == "Madrid" &
                             TIEMPO_VIAJES <= 120) #VORG = 1 = casa, VDES = 2 = trabajo

#here we separate all the "car" trips, "transit" trips, "bike" trips, and "walk" trips. We then will create 4 different impedance functions
transit_HW_alltt <- OD_v_HW  %>% 
  filter(MODO_PRIORITARIO <= 9) %>% #(1. Renfe Cercanías ', 2. Autobus interurbano, 3. Autobus urbano otro municipio, 4. Metro, 5. Metro ligero/tranvía, 6. Autobus urbano Madrid EMT, 7. Resto renfe, 8. Autobus discrecional, 9. Autobus de largo recorrido)
  transmute(TIEMPO_VIAJES, 
            ELE_G_POND_ESC2,
            mode = "transit",
            VORIZT1259,
            VDESZT1259)

car_motor_HW_alltt <- OD_v_HW  %>% 
  filter(MODO_PRIORITARIO <= 19 & MODO_PRIORITARIO >= 10) %>% #includes both motorcycles and cars (specifically: 10. Taxi, 11. Coche conductor particular, 12. Coche conductor empresa, 13. Coche conductor alquiler sin conductor, 14. Coche pasajero particular, 15. Coche pasajero empresa, 16. Coche pasajero alquiler con conductor, 17. Moto/ciclomotor particular, 18. Moto/ciclomotor publica, 19. Moto/ciclomotor empresa)
  transmute(TIEMPO_VIAJES, 
            ELE_G_POND_ESC2,
            mode = "car/motor",
            VORIZT1259,
            VDESZT1259)

bike_HW_alltt <- OD_v_HW  %>% 
  filter(MODO_PRIORITARIO <= 23 & MODO_PRIORITARIO >= 20) %>% #includes cycling and 'otros', specificially: 20. Bicicleta particular, 21. Bicicleta publica, 22. Bicicleta empresa, 23. Otros)
  transmute(TIEMPO_VIAJES, 
            ELE_G_POND_ESC2,
            mode = "bike",
            VORIZT1259,
            VDESZT1259)

walk_HW_alltt <- OD_v_HW  %>% filter(MODO_PRIORITARIO == 24) %>% #"24. Andando/pie"
  transmute(TIEMPO_VIAJES,
            ELE_G_POND_ESC2,
            mode = "walk",
            VORIZT1259,
            VDESZT1259)
```

```{r seeing-how-similar-calc-tt-is-to-survey, eval=FALSE}
#Let's see how similar the calc tt is to the survey (only Madrid city travel times). We can see that transit is the most worrisome -- calculating travel times from some arbitrary point in the center of the TAZ is not very accurate. Also travel times for all active modes should also be calculated for trips between 60-120mins (not only 0-60mins). For now let's only work with the survey data.

#This provides justification -- we need to use hexagons! for paper 2.

#walk
compare_walktts <- calc_walktt %>% merge(walk_HW_alltt, by=c("VORIZT1259", "VDESZT1259")) %>% dplyr::select(-c("ELE_G_POND_ESC2", "mode"))
compare_walktts %>% summary()

#bike
compare_biketts <- calc_bikett %>% merge(bike_HW_alltt, by=c("VORIZT1259", "VDESZT1259")) %>% dplyr::select(-c("ELE_G_POND_ESC2", "mode"))
compare_biketts %>% summary()

#car
compare_cartts <- calc_cartt %>% merge(car_motor_HW_alltt, by=c("VORIZT1259", "VDESZT1259")) %>% dplyr::select(-c("ELE_G_POND_ESC2", "mode"))
compare_cartts %>% summary()

#transit 
compare_transittts <- calc_transittt %>% merge(transit_HW_alltt, by=c("VORIZT1259", "VDESZT1259")) %>% dplyr::select(-c("ELE_G_POND_ESC2", "mode"))
compare_transittts %>% summary()
```

```{r fitting-impedance-function1}
#see which curve will fit the best using the descdist() function. This is to calibrate an impedance function for each mode based on the travel survey

#transit
transit_HW_alltt_long <- transit_HW_alltt[rep(seq_len(dim(transit_HW_alltt %>% dplyr::select(-mode))[1]), transit_HW_alltt$ELE_G_POND_ESC2), 1]
#fitdistrplus::descdist(data=transit_HW_alltt_long) #try beta, uniform, and normal

#car
car_motor_HW_alltt_long <- car_motor_HW_alltt[rep(seq_len(dim(car_motor_HW_alltt %>% dplyr::select(-mode))[1]), car_motor_HW_alltt$ELE_G_POND_ESC2), 1]
#fitdistrplus::descdist(data=car_motor_HW_alltt_long) #try beta, gamma

#bike
bike_HW_alltt_long <- bike_HW_alltt[rep(seq_len(dim(bike_HW_alltt %>% dplyr::select(-mode))[1]), bike_HW_alltt$ELE_G_POND_ESC2), 1]
#fitdistrplus::descdist(data=bike_HW_alltt_long) #try beta, gamma

#walk
walk_HW_alltt_long <- walk_HW_alltt[rep(seq_len(dim(walk_HW_alltt %>% dplyr::select(-mode))[1]), walk_HW_alltt$ELE_G_POND_ESC2), 1]
#fitdistrplus::descdist(data=walk_HW_alltt_long) #try beta, gamma, exp

```

```{r fitting-impedance-function2-transit}
# using fitdist function to fit a distribution using the default maximum likelihood estimation method and Nelder-Mead method for direct optimization

#for transit, try: uniform, normal, gamma, lognormal
gamma_ <- fitdistrplus::fitdist(data=transit_HW_alltt$TIEMPO_VIAJES,
                                "gamma", 
                                method="mle",
                                weights = round(transit_HW_alltt$ELE_G_POND_ESC2),
                                optim.method="Nelder-Mead")
unif_ <- fitdistrplus::fitdist(data=transit_HW_alltt$TIEMPO_VIAJES,
                                "unif", 
                                method="mle",
                                weights = round(transit_HW_alltt$ELE_G_POND_ESC2),
                               optim.method="Nelder-Mead")
norm_ <- fitdistrplus::fitdist(data=transit_HW_alltt$TIEMPO_VIAJES,
                                "norm", 
                                method="mle",
                                weights = round(transit_HW_alltt$ELE_G_POND_ESC2),
                                optim.method="Nelder-Mead")
lnorm_ <- fitdistrplus::fitdist(data=transit_HW_alltt$TIEMPO_VIAJES,
                                "lnorm", 
                                method="mle",
                                weights = round(transit_HW_alltt$ELE_G_POND_ESC2),
                                optim.method="Nelder-Mead")
```

```{r fitting-impedance-function2-transit2, eval=FALSE}
gamma_$aic #lowest aic and bic, so we pick this distribution.
unif_$aic 
norm_$aic
lnorm_$aic

gamma_$bic #lowest aic and bic, so we pick this distribution.
unif_$bic 
norm_$bic
lnorm_$bic
```

```{r fitting-impedance-function2-transit3}
HW_transit_gamma <- gamma_
```

```{r fitting-impedance-function2-car}
#for car, try: uniform, lnorm
gamma_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt$TIEMPO_VIAJES,
                                "gamma", 
                                method="mle",
                                weights = round(car_motor_HW_alltt$ELE_G_POND_ESC2),
                                optim.method="Nelder-Mead")

lnorm_ <- fitdistrplus::fitdist(data=car_motor_HW_alltt$TIEMPO_VIAJES,
                                "lnorm", 
                                method="mle",
                                weights = round(car_motor_HW_alltt$ELE_G_POND_ESC2),
                                optim.method="Nelder-Mead")
```

```{r fitting-impedance-function2-car2, eval=FALSE}
gamma_$aic #lowest aic and bic, so we pick this distribution.
lnorm_$aic

gamma_$bic #lowest aic and bic, so we pick this distribution.
lnorm_$bic
```

```{r fitting-impedance-function2-car3}
HW_car_motor_gamma <- gamma_
```

```{r fitting-impedance-function2-bike}
#for bike, try: uniform, lnorm
gamma_ <- fitdistrplus::fitdist(data=bike_HW_alltt$TIEMPO_VIAJES,
                                "gamma", 
                                method="mle",
                                weights = round(bike_HW_alltt$ELE_G_POND_ESC2),
                                optim.method="Nelder-Mead")

lnorm_ <- fitdistrplus::fitdist(data=bike_HW_alltt$TIEMPO_VIAJES,
                                "lnorm", 
                                method="mle",
                                weights = round(bike_HW_alltt$ELE_G_POND_ESC2),
                                optim.method="Nelder-Mead")
```

```{r fitting-impedance-function2-bike2, eval=FALSE}
gamma_$aic 
lnorm_$aic #lowest aic and bic, so we pick this distribution.

gamma_$bic 
lnorm_$bic #lowest aic and bic, so we pick this distribution.
```

```{r fitting-impedance-function2-bike3}
HW_bike_lnorm <- lnorm_
```

```{r fitting-impedance-function2-walk}
#for walk, try: uniform, lnorm, normal
gamma_ <- fitdistrplus::fitdist(data=walk_HW_alltt$TIEMPO_VIAJES,
                                "gamma", 
                                method="mle",
                                weights = round(walk_HW_alltt$ELE_G_POND_ESC2),
                                optim.method="Nelder-Mead")

lnorm_ <- fitdistrplus::fitdist(data=walk_HW_alltt$TIEMPO_VIAJES,
                                "lnorm", 
                                method="mme",
                                weights = round(walk_HW_alltt$ELE_G_POND_ESC2),
                                optim.method="Nelder-Mead")
norm_ <- fitdistrplus::fitdist(data=walk_HW_alltt$TIEMPO_VIAJES,
                                "norm", 
                                method="mle",
                                weights = round(walk_HW_alltt$ELE_G_POND_ESC2),
                                optim.method="Nelder-Mead")
```

```{r fitting-impedance-function2-walk2, eval=FALSE}
gamma_$aic 
lnorm_$aic #lowest aic and bic, so we pick this distribution.
norm_$aic

gamma_$bic 
lnorm_$bic #lowest aic and bic, so we pick this distribution.
norm_$bic
```

```{r fitting-impedance-function2-walk3}
HW_walk_lnorm <- lnorm_
```

```{r}
# To visualize the impedance functions we create a data frame and calculate the density of the distribution given the estimated parameters:

x <- data.frame(t = seq(1, 120, 1))

x_transit <- x %>%
  mutate(f = dgamma(t,
                    HW_transit_gamma$estimate[1],
                    HW_transit_gamma$estimate[2]),
         mode = "transit")

x_car_motor <- x %>%
  mutate(f = dgamma(t,
                    shape = HW_car_motor_gamma$estimate[1],
                    rate = HW_car_motor_gamma$estimate[2]),
         mode = "car/motor")

x_bike <- x %>%
  mutate(f = dlnorm(t,
                    HW_bike_lnorm$estimate[1],
                    HW_bike_lnorm$estimate[2]),
         mode = "bike")

x_walk <- x %>%
  mutate(f = dlnorm(t,
                    HW_walk_lnorm$estimate[1],
                    HW_walk_lnorm$estimate[2]),
         mode = "walk")
```

```{r}
# Bind the empirical and theoretical trip length distributions:
tld_empirical <- rbind(transit_HW_alltt,
      car_motor_HW_alltt,
      bike_HW_alltt,
      walk_HW_alltt) %>%
  mutate(mode = factor(mode,
                       levels = c("bike", "walk", "car/motor", "transit")),
         distribution = "empirical")

tld_theoretical <- rbind(x_transit,
      x_car_motor,
      x_bike,
      x_walk) %>%
  mutate(mode = factor(mode,
                       levels = c("bike", "walk", "car/motor", "transit")))
```

```{r, eval=FALSE, warning=FALSE}
#plot
tlds_curves_m_plot <- ggplot() + 
  geom_histogram(data = tld_empirical,
                 aes(x = TIEMPO_VIAJES, 
                     weights = ELE_G_POND_ESC2, 
                     after_stat(density)),
                 bins = 20) +
  geom_line(data = tld_theoretical,
            aes(x = t, y = f)) + 
  xlab("Travel time from home to work (min)") +
  facet_wrap(~ mode, nrow = 2) +
  theme_classic()

tlds_curves_m_plot
ggsave("images/tlds_curves_m_plot.png")
```

```{r, eval=FALSE}
#here, let's double check the empirical trip length distributions, as we can see from the two generated plots, our input TLD is correct!
rbind(transit_HW_alltt,
      car_motor_HW_alltt,
      bike_HW_alltt,
      walk_HW_alltt) %>%
  mutate(mode = factor(mode,
                       levels = c("bike", "walk", "car/motor", "transit"))) %>%
  ggplot(aes(x = TIEMPO_VIAJES, after_stat(density),
             weights = ELE_G_POND_ESC2
             )) + 
  geom_histogram(bins = 20) + 
  facet_wrap(~ mode, nrow = 2)
```
```{r define-population-balancing-factor}
# We define $P_i$ (for the population balancing factor) for each mode and $O_j$ associated with each zone. To do this, we add the number of trips (by mode) and create objects that contain the number of workers (origin) and jobs (destinations) - by mode:

# Instead of creating various columns like above, create a single column with the mode
OD_v_HW <- OD_v_HW  %>% 
  mutate(mode = case_when(MODO_PRIORITARIO <= 9 ~ "transit",
                          MODO_PRIORITARIO <= 19 & MODO_PRIORITARIO >= 10 ~ "car/motor",
                          MODO_PRIORITARIO <= 23 & MODO_PRIORITARIO >= 20 ~ "bike",
                          MODO_PRIORITARIO == 24 ~ "walk"),
         mode = factor(mode,
                       levels = c("bike", "walk", "car/motor", "transit")))

# The workers are classified by mode and origin. Ex. from Origin 1 five people use bike mode to get to destinations. This is P_i for each m
workers_zn208 <- OD_v_HW %>%
  group_by(VORI_CD_ZT208,
           mode) %>%
  summarize(workers = sum(ELE_G_POND_ESC2),
            total_ttworkers = sum(workers*mean(TIEMPO_VIAJES)),
            .groups = "drop") %>%
  rename(CD_ZT208 = VORI_CD_ZT208)

jobs_zn208 <- OD_v_HW %>%
  group_by(VDES_CD_ZT208) %>%
  summarize(jobs = sum(ELE_G_POND_ESC2),
            .groups = "drop") %>%
  rename(CD_ZT208 = VDES_CD_ZT208)
```

```{r}
# Next join the number of workers summed at each origin to the ZN simple features object:
workers_sf_zn208 <- workers_zn208 %>%
  left_join(zn208,
            by = "CD_ZT208") %>%
  st_as_sf()

jobs_sf_zn208 <- jobs_zn208 %>%
  left_join(zn208,
            by = "CD_ZT208") %>%
  st_as_sf()
```

```{r}
#Now we reshape the data. Each row will represent an any origin to to 1 destination (i.e., destination row is unique). We will average the travel times, by mode, and then use this average travel time to find the 'f' (impedance value for travel for each mode). Hint: this is why we needed to find the impedance function in the first place! The functions represent the cost of travel for each mode based on the travel time for a trip, since we are averaging together the travel time for each mode.

OD <- OD_v_HW %>%
  group_by(VORIZT1259,
           VDESZT1259,
           mode) %>%
  summarize(tt = mean(TIEMPO_VIAJES),
            .groups = "drop")

OD_zn208 <- OD_v_HW %>%
  group_by(VORI_CD_ZT208,
           VDES_CD_ZT208,
           mode) %>%
  summarize(tt = mean(TIEMPO_VIAJES),
            .groups = "drop")
```

```{r, eval=FALSE}
# Check the mean and median travel times by mode, the first is at the level of zn1259 and the second is at zn208
OD_zn208 %>% 
  group_by(mode) %>% 
  summarize(mean_tt = mean(tt),
            median_tt = median(tt),
            min_tt = min(tt),
            max_tt = max(tt))
```

```{r}
#let's also do this at the level of zn208
OD_3 <- OD_zn208 %>%
  left_join(workers_zn208,
            by = c("VORI_CD_ZT208" = "CD_ZT208", "mode"))

OD_3 <- OD_3 %>%
  left_join(jobs_zn208,
            by = c("VDES_CD_ZT208" = "CD_ZT208"))

```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
#Estimate f value using these averaged travel times:
OD_3 <- OD_3 %>%
  mutate(f = case_when(mode == "transit" ~ dgamma(tt,
                                                  HW_transit_gamma$estimate["shape"], 
                                                  HW_transit_gamma$estimate["rate"]),
                       mode == "car/motor" ~ dgamma(tt,
                                                    HW_car_motor_gamma$estimate["shape"],
                                                    HW_car_motor_gamma$estimate["rate"]),
                       mode == "bike" ~ dlnorm(tt,
                                               HW_bike_lnorm$estimate["meanlog"],
                                               HW_bike_lnorm$estimate["sdlog"]),
                       mode == "walk" ~ dlnorm(tt, 
                                               HW_walk_lnorm$estimate["meanlog"], 
                                               HW_walk_lnorm$estimate["sdlog"])),
         jobs_f = jobs*f) #for Hansen-type accessibility (unconstrained)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# ## Calculating hansen-type accessibility (for comparison)
OD_Si_zn208 <- OD_3 %>%
  group_by(VORI_CD_ZT208, mode) %>%
  summarize(S_i = sum(jobs_f)) # Hansen-type accessibility (unconstrained)
```

```{r}
#calculating SA
OD_SA_zn208 <- OD_3 %>%
  mutate(catch = 1) %>%
  sp_avail_detailed(.,
                    o_id = VORI_CD_ZT208,
                    d_id = VDES_CD_ZT208,
                    pop = workers,
                    opp = jobs,
                    r = catch,
                    f = f) %>% 
  mutate(jobs_tt_f_t = jobs*tt*f_t)
```

```{r}
SA_i_zn208 <- OD_SA_zn208 %>%
  group_by(VORI_CD_ZT208) %>%
  summarize(V_i = sum(V_ij))

SA_im_zn208 <- OD_SA_zn208 %>%
  group_by(VORI_CD_ZT208, 
           mode) %>%
  summarize(V_im = sum(V_ij),
            population_m = mean(workers),
            .groups = "drop")

population_zn208 <- workers_zn208 %>%
  group_by(CD_ZT208) %>%
  summarize(population = sum(workers))

jobs_zn208 <- jobs_zn208 %>%
  group_by(CD_ZT208) %>%
  summarize(jobs = sum(jobs))
```

```{r, eval=FALSE}
#Preforming checks. Spatial availability by origin (with check). Here we represent spatial availability per origin across all modes for each i. We know that car mode represents the majority of spatial availability per i overall: 
summary(SA_i_zn208) 
sum(SA_i_zn208$V_i)
```

```{r, eval=FALSE}
#But we can suspect that the spatial distribution of SA is uneven depending on the mode. So here we represent Spatial availability by origin, mode (with check):
summary(SA_im_zn208)
sum(SA_im_zn208$V_im)
```

```{r, eval=FALSE}
#more checks, calc.  Calculate worker population by zone of origin
sum(population_zn208$population)
sum(workers_zn208$workers)
```

```{r}
# Finally, let's join the zonal worker population to the SA_i and SA_im spatial objects
SA_i_zn208 <- SA_i_zn208 %>%
  left_join(population_zn208,
            by = c("VORI_CD_ZT208" = "CD_ZT208"))

SA_i_zn208 <- SA_i_zn208 %>%
  left_join(jobs_zn208,
            by = c("VORI_CD_ZT208" = "CD_ZT208"))

SA_im_zn208 <- SA_im_zn208 %>%
  left_join(population_zn208,
            by = c("VORI_CD_ZT208" = "CD_ZT208"))
```

```{r, eval=FALSE}
# checks
sum(SA_i_zn208$V_i)
sum(SA_im_zn208$V_im)
```

```{r}
# To facilitate the interpretation of spatial availability, here we calculate spatial availability per capita:

SA_i_zn208 <- SA_i_zn208 %>%
  mutate(v_i = V_i/population)

# summary(SA_i_zn208)

SA_im_zn208 <- SA_im_zn208 %>%
  mutate(v_im = V_im/population_m)

# summary(SA_im_zn208)
```

```{r}
# Lastly, join geometry of zones to SA so we can visualize:
SA_i_zn208 <- SA_i_zn208 %>%
  left_join(zn208 %>%
              dplyr::select(CD_ZT208),
            by =c("VORI_CD_ZT208" = "CD_ZT208")) %>%
  st_as_sf()

SA_im_zn208 <- SA_im_zn208 %>%
  left_join(zn208 %>%
              dplyr::select(CD_ZT208),
            by = c("VORI_CD_ZT208" = "CD_ZT208")) %>%
  st_as_sf()

OD_Si_zn208 <- OD_Si_zn208 %>%
  left_join(zn208 %>%
              dplyr::select(CD_ZT208),
            by = c("VORI_CD_ZT208" = "CD_ZT208")) %>%
  st_as_sf()
```
```{r create-bounding-boxes}
#create the bounding box for the Madrid city + a little extra
bbox_new1 <- st_bbox(zn1259_mad) # current bounding box

xrange <- bbox_new1$xmax - bbox_new1$xmin # range of x values
yrange <- bbox_new1$ymax - bbox_new1$ymin # range of y values

bbox_new1[1] <- bbox_new1[1] - (0.0 * xrange) # xmin - left
bbox_new1[3] <- bbox_new1[3] + (0.0 * xrange) # xmax - right
bbox_new1[2] <- bbox_new1[2] - (0.0 * yrange) # ymin - bottom
bbox_new1[4] <- bbox_new1[4] + (0.0 * yrange) # ymax - top

bbox_Madrid_muni <- bbox_new1 %>%  # take the bounding box ...
  st_as_sfc()

sf::sf_use_s2(FALSE) #this terns off the spherical geometry component in the st_union() function - this assumes the sf is just planar. Thats' fine for our purposes
Madrid_muni_border <- zn1259_mad %>%
  st_union()
```
```{r merging-borders-for-plotting}
Madrid_muni_border <- Madrid_muni_border %>% st_as_sf() %>% transmute(Border = "Madrid municipal border") %>% rename( "geometry" = "x")

borders <- rbind(LEZ_centro %>% transmute(Border = "LEZ centro"), 
                 LEZ_M30 %>% transmute(Border = "LEZ M-30"),
                 Madrid_muni_border)
```

```{r, eval=FALSE}
#mapping population P_i^m by mode!
pop_i_m_plot <- ggplot() +
  geom_sf(data = Madrid_muni_border, # border for Madrid muni
          colour=NA,
          fill="grey30") +
  geom_sf(data = SA_im_zn208,
          aes(fill = population_m),
          color = NA) +

  scale_fill_fermenter(palette = "Purples", #legend scale bar
                       na.value = "grey90",
                       direction = 1,
                       name = expression(P['i']^{m}),
                       limits = c(38, 18421),
                       breaks = c(38, 360,2229,3750, 6768, 18421),
                       labels = c("39 (min.)", "360 (20% prob.)", "2,229 (40% prob.)",
                                  "3,750 (60% prob.)", "6,768 (80% prob.)","18,421 (max.)"))+
  geom_sf(data = borders,
          aes(colour=Border),
          size = 0.5,
          fill=NA) +
  scale_color_manual(values = c( "red", "cyan", "black"),
                       name = element_blank())+
  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
                   height = unit(0.15, "cm")) +
  theme_void() +
  theme(legend.key.width = unit(0.2, 'cm')) +
  guides(fill = guide_legend(order = 1), colour = guide_legend(order = 2))+
  facet_wrap(~mode, nrow = 2)
```

```{r, eval=FALSE}
# jobs_i !
jobs_i_plot <- ggplot() +
    geom_sf(data = Madrid_muni_border, # border for Madrid muni
          colour=NA,
          fill="grey30") +
  geom_sf(data = SA_i_zn208,
          aes(fill = jobs),
          color = NA) +
  scale_fill_fermenter(palette = "Oranges", #legend scale bar
                       direction = 1,
                       name = expression(O['j']),
                       limits = c(1201, 39811),
                       breaks = c(1201, 5599,8703,14944,22377, 39811),
                       labels = c("1,201 (min.)", "5,599 (20% prob.)", "8,704 (40% prob.)",
                                  "14,944 (60% prob.)", "22,377 (80% prob.)","39,811 (max.)"))+
  geom_sf(data = borders,
          aes(colour=Border),
          size = 0.5,
          fill=NA) +
  scale_color_manual(values = c( "red", "cyan", "black"),
                       name = element_blank())+
  guides(fill = guide_legend(order = 1), colour = guide_legend(order = 2))+
  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
                   height = unit(0.15, "cm")) +
  theme_void() +

  theme(legend.key.width = unit(0.2, 'cm'))
```

```{r, eval=FALSE}
pop_i_m_plot
ggsave("images/im_populations_zn208_plot.png")
```

```{r, eval=FALSE}
jobs_i_plot
ggsave("images/i_jobs_zn208_plot.png")
```

```{r}
#Plot V (spatial availability) by origin-mode (then zn208). 
SA_im_V_zn208_plot <- ggplot() +
    geom_sf(data = Madrid_muni_border, # border for Madrid muni
          colour=NA,
          fill="grey30") +
  geom_sf(data = SA_im_zn208,
          aes(fill = V_im),
          color = NA) +
  scale_fill_fermenter(palette = "Blues", #legend scale bar
                       name = expression(V["i"]^{m}),
                       direction = 1,
                       limits = c(0.07, 36531),
                       breaks = c(0.07, 81, 670, 6409, 7727, 36531),
                       labels = c("0.07 (min.)", "81 (20% prob.)", "670 (40% prob.)", "2,707 (60% prob.)", "7,727 (80% prob.)", "36,531 (max.)"))+
    geom_sf(data = borders,
          aes(colour=Border),
          size = 0.5,
          fill=NA) +
  scale_color_manual(values = c( "red", "cyan", "black"),
                       name = element_blank())+
  guides(fill = guide_legend(order = 1), colour = guide_legend(order = 2))+
  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
                   height = unit(0.15, "cm")) +
  theme_void() +
  theme(legend.key.width = unit(0.2, 'cm')) +
  facet_wrap(~mode, nrow = 2)

#the accessibility plot (Hansen-type) -- this is for internal comparision.
S_im_plot_zn208_plot <- ggplot() +
    geom_sf(data = Madrid_muni_border, # border for Madrid muni
          colour=NA,
          fill="grey30") +
  geom_sf(data = OD_Si_zn208,
          aes(fill = S_i),
          color = NA) +
  scale_fill_distiller(palette = "YlGn", #legend scale bar
                       name = expression(S["im"]),
                       na.value = "grey80",
                       direction = 1)+
    geom_sf(data = borders,
          aes(colour=Border),
          size = 0.5,
          fill=NA) +
  scale_color_manual(values = c( "red", "cyan", "black"),
                       name = element_blank())+
  guides(fill = guide_legend(order = 1), colour = guide_legend(order = 2))+
  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
                   height = unit(0.15, "cm")) +
  theme_void() +
  theme(legend.key.width = unit(0.2, 'cm')) +
  facet_wrap(~mode, nrow = 2)

```

```{r, eval=FALSE}
S_im_plot_zn208_plot
ggsave("images/Hansen_im_V_zn208_plot.png")
```

```{r, eval=FALSE}
SA_im_V_zn208_plot
ggsave("images/SA_im_V_zn208_plot.png")
```

```{r}
#Plot v (spatial availability per capita) by origin:
SA_im_v_zn208_plot <- ggplot() +
  geom_sf(data = Madrid_muni_border, # border for Madrid muni
          colour=NA,
          fill="grey30") +
  geom_sf(data = SA_im_zn208,
          aes(fill = v_im),
          color = NA) +
  scale_fill_gradient2(low="red", high="green", mid = "white", midpoint = 1,
                       name = expression(v["i"]^{m}),
                       limits = c(0.001106867, 2.086736720),
                       breaks = c( 0.12, 0.31, 0.80, 1.11, 2.086736720),
                       labels = c(">0.12 (20% prob.)", "0.31 (40% prob.)", "0.80 (60% prob.)", "1.11 (80% prob.)", "2.09 (max.)"))+
    geom_sf(data = borders,
          aes(colour=Border),
          size = 0.5,
          fill=NA) +
  scale_color_manual(values = c( "red", "cyan", "black"),
                       name = element_blank())+
  guides(fill = guide_legend(order = 1), colour = guide_legend(order = 2))+
  annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
                   height = unit(0.15, "cm")) +
  theme_void() +
  theme(legend.key.width = unit(0.2, 'cm')) +
  facet_wrap(~mode, nrow = 2)
```
```{r, eval=FALSE}
SA_im_v_zn208_plot
ggsave("images/SA_im_vv_zn208_plot.png")
```
```{r, eval=FALSE}
v_im_per_pop_plot <- ggplot(SA_im_zn208 %>% mutate(area = st_area(SA_im_zn208),
                                                   pop_den = population/area),
       aes(x=population,
         #x = reorder(VORI_CD_ZT208, population),
           y = v_im,
           group=mode,
           color=mode)) +
  geom_line()  + geom_point()+
  xlab("Total population at origin") +
  ylab(expression(v["i"]^{m}))+
  #scale_x_discrete(breaks=c("5","15","30","45","60"),labels=c("100","10000","20000","30000","41000"))
  theme_classic()

v_im_per_pop_plot #this doesn't tell me tooo much.. it just means that levels of SA is similar across population densities for the most part. And that bike/walk are similar while car/transit are similar.
ggsave("images/v_im_per_pop_plot.png")
```

# Abstract
An increasing number of studies within the domain of transport are concerned with the inequities in accessibility to opportunities. A dimension of these inequities arise from differences in access by mode type (e.g., the number of work opportunities that can be reached using a car as opposed to transit in a city). However, methods assessing multimodal accessibility in the literature fall short as aspects of competition for opportunities and the explicit methodological acknowledgement of opportunities being _finite_ are lacking. In this vein, this paper presents an extension of _spatial availability_, a singly-constrained competitive accessibility measure, for the context of multimodal accessibility analysis. We first illustrate the features of spatial availability that lends itself to multimodal analysis. We then demonstrate its use on the case study of Low Emission Zones in Madrid (Spain) and highlight how this policy intervention changes the accessibility of populations using different modes. In summary, spatial availability can be used to create and interpret multimodal policy intervention scenarios unlike previous methods: this creation and interpretation can help regions envision a more sustainable and equitable access-to-opportunity landscape by better identifying differences in accessibility afforded by different modes.

# Introduction

<!-- outlines the use of location-based accessibility measures in the literature: equity, planning -- esp. as it relates to modal competition (brief) -->
Implementing urban policies that re-shape cities through accessibility gains (i.e., the _potential to interact_ with opportunities as a result of land-use mix and transport systems as originally defined by @hansenHowAccessibilityShapes1959) have been widely applied within the transportation literature and is increasingly discussed by planners [@levinsonTransportAccessManual2020; @gowerPlanningInnovationCity2022; @siddiqToolsTradeAssessing2021; @yanAccessibilityBasedPlanningAddressing2021]. An important challenge in the identification of interventions that sustainably and equitably transform cities is the effective evaluation of _trade-offs_: cities are complex and dynamic ecologies, and advantaging one component of the city can disadvantage another area, population, or sub-component In this way, policy evaluation should take a _systems_ approach [@fikselSustainabilityResilienceSystems2006]. One way of considering systems is from the perspective of the _finite_. As an illustration, consider the amount of transport space within a city: the amount is typically finite so re-allocating road space away from one mode directly impacts the performance of the others (see the literature on road space reallocation e.g., @valencaMainChallengesOpportunities2021). Evaluating policy impacts in the context of _finity_ provides a way to contextualize the balance of trade-offs that the citizens of a city should tolerate.

From the perspective of urban transport systems, location-based accessibility measures have been used in the context of policy evaluation. For instance, @leeMeasuringImpactsNew2018 assesses the transit accessibility gains to healthcare and employment opportunities for disadvantaged neighbourhood in Columbus, Ohio, USA after the transit system's re-design and introduction of a rapid bus system. However, a limitation of this study, like others that implement accessibility measures, is they do not calculate results under a _constrained_ framework i.e., one of _finity_. The citizens of Columbus should experience qualitative accessibility gains - but is it at the expense of access to opportunities by other modes? As another example, @mohriClusteringMethodMeasuring2021 implements a modified cumulative opportunity measure to assess differences between private vehicle and transit system accessibility to jobs in Melbourne, but a similar question remains, does the accessibility afforded to the private vehicle using population come at the expense of accessibility losses to transit users?

The two studies discussed in the previous paragraph use _non-competitive_ accessibility measures. There is a branch of location-based accessibility measures that do incorporate the effect of competition for opportunities by the population in the region. However, we argue that these existing methods fall short in acknowledging the _finity_ of opportunities. For instance, @maoMeasuringSpatialAccessibility2013 applies the competitive measure, the two-step floating catchment approach (2SFCA) for the case of access to healthcare services in Florida for both a multimodal network and a single modal network. While the differences in modal access are discussed, the question of how the advantage in access afforded by one mode over another impacts access for different mode users is unanswered.

This question of how much one mode-using population can access at the expense of another mode-using population is pertinent equity question in the evaluation of policy scenarios that are multimodal. For instance, consider the impact of a low emission zone (LEZ). LEZ is a policy of spatial and modal discrimination: the circulation of vehicles that are excessively polluting are restricted in specific areas in a city. In the recognition that opportunities are finite, the implementation of a LEZ explicitly reduces the access that the population using polluting vehicles has to opportunities. This restriction allows the population using other more sustainable modes to potentially have a higher level of access than before the LEZ implementation. This evaluation is especially urgent as LEZ are currently in effect in cities globally; their reception has been mixed [@tarrinoortizPublicAcceptabilityLow2021] and may be having negative impacts on disadvantaged populations who have become mobility restricted [@devrijNooneVisitsMe2022; @verbeekJustManagementUrban2022]. Measures that evaluate the accessibility of modes given both _constrained_ and _competitive_ considerations are lacking in the literature, but are needed, to evaluate such policy interventions to effectively comment on how accessibility changes as a result of the mobility-restricted mode.

In @soukhovIntroducingSpatialAvailability2023, we introduce spatial availability, a type of location-based accessibility measure that is both _constrained_ and _competitive_. In this paper, we extend the spatial availability measure into a multimodal framework and explore its use in answering the question outlined: _"given opportunities are finite, how many are available to a given location depending on the mode used?"._ The answer to this question quantifies how many opportunities can be accessed, considering competition, for different modes. To foreground this exploration, in Section 2, we discuss short falls of a few existing location-based measures in comparison to spatial availability through a synthetic example. In Section 3, the spatial availability of an empirical example of the LEZ in the city of Madrid, Spain is calculated. We demonstrate how the restriction of private vehicles with the LEZ implementation impacts the spatial availability of opportunities for each sub-population using transit, cycling and walking modes. In Section 4, we provide concluding remarks on the strengths of the use of spatial availability as a multi-modal accessibility measure, limitations, and potential future uses in policy planning scenarios.

# A brief review of multimodal accessibility measurement methods

Location-based accessibility indicators are quantitative measures of _potential_ interaction with opportunities for locations within a given region: they are a product of the relationship between land-use and transport systems. Arguably the most commonly used location-based measured are cumulative opportunity measures and weighted cumulative opportunity measures [@levinsonTransportAccessManual2020]. These measures weight the opportunities that can be potentially interacted with from origin $i$ to destination $j$ based on some sort of travel cost function (e.g., travel time, fare, travel distance) otherwise known as a travel impedance function $f^{m}(c^m_{ij})$. Many weighted cumulative opportunities (often refered to as the gravity-based measure) originate from the measure proposed by @hansenHowAccessibilityShapes1959, which can take the following multimodal form: $S_i^m = \sum_j O_j f^m(c_{ij}^m)$ where $m$ is a set of modes which have mode-specific travel costs $(c_{ij}^m)$ and travel impedance functions $f^m(\cdot)$.

The Hansen-type measure does not consider competition between modes nor is it constrained. As an example, the work of @tahmasbiMultimodalAccessibilitybasedEquity2019 uses the Hansen-type measure to measure the potential interaction with retail locations using walking, public transit, and car modes $m$. $S_i^m$ is the sum of retail locations $j$ that can potentially be interacted with under the travel impedance as calculated for each $i$ and $m$. In other words, each $i$ has three $S_i$ values, one per $m$. In this work, they demonstrate that the car mode has the highest $S_i^{m=car}$ values in the majority of $i$, i.e., populations using a car can potentially interact with the most retail opportunities than populations using other modes. However, the higher $S_i^{m=car}$ values are not a result of lower $S_i^{m}$ values for other modes: it is not assumed that car-using populations potentially accessing more opportunities take away potential opportunities for other populations within the measure. Put another way, this measure does not consider competition. This measure is also not constrained: there is no global maximum for $S_i$ or $S_i^m$ values, they are presented as a population normalized accessibility index. This makes the interpretation of the 'potentially interacted opportunities' relative to the region, making comparisons of the results across different regions challenging.

However, opportunities in a region can be considered finite. There are only so many school-seats, hospital capacity, food stores, jobs, etc., in a region and if one person interacts with an opportunity at a given time, it is taken. As such, if one person is advantaged and has the ability to reach more opportunities through a lower travel-cost mode, than they have more opportunities to potentially interact with more opportunities than other people. From the other perspective, their are fewer opportunities left to be potentially interacted with for populations using higher travel-cost modes. In this way, populations using modes with a higher travel impedance are at a higher access disadvantage than populations using lower travel impedance modes. This recognition is the motivation behind integrating _competition_ for opportunities within multimodal accessibility measures. Arguably one of the most popular competitive location-based accessibility measures is the two-step floating catchment area (2SFCA) approach popularized by @luoMeasuresSpatialAccessibility2003 who simplified the approach proposed by @shenLocationCharacteristicsInnercity1998 (with similar considerations for competition in @weibullAxiomaticApproachMeasurement1976 and @josephMeasuringPotentialPhysical1982).

The Shen-type accessibility measure's formulation is: $a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}$ where $D_j^m$ is the potential demand for opportunities equal to travel impedance weighted population $\sum_i P_i^m f^m(c_{ij}^m)$. In this way, the Shen-type measure can be understood as a ratio of the potential opportunity supply over the potential demand for opportunities. The measure considers competition, but it is _non-constrained_. A score of competitive potential accessibility associated is associated with each location $i$ for each mode $m$, but there are no global maximums. In other words, it is difficult to interpret the meaning of differences in Shen-type accessibility scores between modes.

To illustrate, @taoInvestigatingImpactsPublic2020a calculates $a_i^m$ to jobs for different income-group populations in Shenzhen (China) using $m = \text{public transit}$ and $m={car}$. They demonstrate that $i$s with low-income populations have lower $a_i^m$ than $i$s with higher-income populations. Further, they demonstrate that $a_i^{m=\text{public transit}}$ is lower than $a_i^{m=\text{car}}$ at many $i$s, arguing that this may put $i$s with lower-income populations in a further disadvantage. $a_i$ and/or $a_i^m$ are used to compare relative spatial differences in overall competitive accessibility and modal competitive accessibility, but because there is no global maximum, making it is difficult to interpret the significance between differences in $a_i^{m}$ values. Questions such as: What is the impact that competition has on the difference in $a_i^m$ values? How does impact vary spatially? And what is the interpretation of this difference? are left unanswered.

Spatial availability improves on previous multi-modal accessibility approaches as it considers _competition_ in the potential interaction with opportunities in a _constrained_ framework (e.g., finite opportunities). This is done by considering: 1) competition between mass effect (e.g., the advantage of sub-populations residing in relatively low population-density and high opportunity-proximate areas) and 2) competition between travel impedance (e.g., sub-populations with relatively low travel-impedance) through a proportional allocation mechanism. The following sub-section demonstrates how spatial availability compares to the two other measures and the advantages in its interpretation through a synthetic example.

## A synthetic example: calculating accessibility for multiple modes considering finite opportunities

In brief, we define the _spatial availability_ at $i$ ( $V_{i}$ ) as the proportion of all opportunities in the region $O$ that are allocated to location $i$ from all opportunity destinations $j$. $V_{i}$ is a value of how many opportunities are available to each location $i$ out of all the opportunities in the region (e.g., a neighbourhood has 1,000 spatially available jobs out of 100,000 jobs in the total region). The general formulation of spatial availability $V_{i}$ is shown in Equation (\ref{eq:spatial-availability-general}):
```{=tex}
\begin{equation}
\label{eq:spatial-availability-general}
V_i = \sum_{j=1}^J O_jF^t_{ij}
\end{equation}
```

\noindent where:

- $F^t_{ij}$ is a balancing factor that depends on the demand for opportunities $O_j$ and cost of movement in the system $f(c_{ij})$.
- $V_i$ is the number of spatially available opportunities at $i$; the sum of $V_{i}$ is equivalent to the total sum of opportunities in the region (i.e., $\sum_j O_j = \sum_i V_i$)

Spatial availability measure is introduced in @soukhovIntroducingSpatialAvailability2023. The unique feature in the measure is the balancing factor $F^t_{ij}$, a proportional allocation mechanisms, which ensures that the $V_i$ calculated for each $i$ sums to the total number of opportunities. Through $F^t_{ij}$, spatial availability is a  _competitive_ and _constrained_ accessibility measure that handles the number of opportunities in the region in a finite way. $F^t_{ij}$ consists of two components: a population-based balancing factor $F^p_{i} = \frac{P_i}{\sum_i P_i}$ and an impedance-based balancing factor $F^c_{ij} = \frac{F^c_{ij}}{\sum_j F^c_{ij}}$ that, respectively, allocate opportunities to $i$ in proportion to the size of the population at $i$ (the mass effect) and the cost of reaching opportunities at $j$ (the impedance effect). $F^p_{i}$ and $ F^c_{ij}$ are calculated for each $i$ such that they both equal 1 when summed across all $i$ in the region (e.g., $\sum_i F^p_{i} = 1$ and $\sum_i F^c_{ij} = 1$). These balancing factors are combined multiplicatively to yield $F^t_{ij}$ which ensures that a proportion of the opportunities $O_j$ are allocated to each $i$ accordingly. In other words, assuming a finite number of opportunities in the region, $F^t_{ij}$ proportionally allocates $O_j$ to each $i$ such that the resulting $V_i$ value represents the number of opportunities _spatially available_ to the population at $i$. This value can be seen to represent spatial availability as it is a proportion of the opportunities in the region (i.e., $\sum_j O_j = \sum_i V_i$).

The focus of this paper is to extend $V_i$ for multimodal applications. To do so, the balancing factors are reformulated to yield a proportional value for the set of modes $m$ used by populations at each $i$. As these factors are proportional, $F^{pm}_{i}$ and $F^{cm}_{ij}$ can be summed up across each $m$ at each $i$ and across all $i$ to equal to 1. They are also similarly combined multiplicatively to obtain their joint effect, represented as the combined balancing factor $F^{tm}_{ij}$ detailed in Equation (\ref{eq:multimodal-balancing-factors}).

```{=tex}
\begin{equation}
\label{eq:multimodal-balancing-factors}
F^{tm}_{ij} = \frac{F^{pm}_{i} \cdot F^{cm}_{ij}}{\sum_{m=1}^M \sum_{i=1}^N F^{pm}_{i} \cdot F^{cm}_{ij}}
\end{equation}
```

\noindent Where:
- The population balancing factor for each $m$ at each $i$ is $F^{pm}_{i} = \frac{P_{i}^m}{\sum_{m}\sum_{i} P_{i}^m}$
- The cost of travel balancing factor for each $m$ at $i$ is $F_{ij}^{cm} = \frac{f(c_{ij}^m)}{\sum_{m} \sum_{i} f(c_{ij}^m)}$

Implementing $F^{tm}_{ij}$, the following Equation (\ref{eq:spatial-availability-multimodal}) demonstrates the multimodal configuration of spatial availability $V_i^m$:

```{=tex}
\begin{equation}
\label{eq:spatial-availability-multimodal}
V^m_{i} = \sum_{j=1}^J O_j\ F^{tm}_{ij}
\end{equation}
```

\noindent Where:

-   $m$ is a set of modes used by populations in the region $m = 1, \cdots, M$.
-   $F^{tm}_{ij}$ is a balancing factor $F^t_{ij}$ for each $m$ at each $i$.
-   $V^m_{i}$ is the spatial availability $V_{i}$ for mode $m$ at each $i$; the sum of $V^m_{i}$ for all $m$ at each $i$ is equivalent to the total sum of opportunities in the region (i.e., $\sum_j O_j = \sum_i V_i = \sum_{m} \sum_{i} V^m_{i}$)

Consider the following synthetic example: Figure \ref{fig:Fig1} depicts a region with population and jobs at three population centers ($A$, $B$, $C$) and three employment centers ($1$, $2$, $3$). The population at each population center is divided into two sub-groups, one using a faster mode $z$ and another using a slower mode $x$, to travel to employment centers. Population center $A$ is Suburban: it is closest to its own relatively large employment center at $A$, close to the Urban area's equally large employment center $B$, and has a population that is smaller than the Urban center $2$ and larger than the Satellite center $3$. The Urban population center has the largest number of population using slower mode $x$, followed by the Suburban then Satellite area. This synthetic example was inspired by the single-mode example used in @shenLocationCharacteristicsInnercity1998 and reconfigured in @soukhovIntroducingSpatialAvailability2023.

```{r, synthetic-example-plot, fig.cap="\\label{fig:Fig1} Modified synthetic example from Shen (1998) with locations of employment centers (in orange), population centers (in blue), number of jobs and population, and travel times for two modes (slower x mode and faster z mode).", fig.show='hold', fig.align='center'}
knitr::include_graphics("images/Fig1.png")
```

From the perspective of access to a _finite_ amount of opportunities in the region ($210,000$ jobs), the sub-population that is most proximate to jobs, furthest from densely populated centers, and is using the lowest travel cost mode $z$ can potentially access the most job opportunities. This appears to be the sub-population at Suburban center $A$ using the faster mode $z$. From the other perspective, sub-populations located further away from jobs, close to dense populations, and using high cost travel mode $x$ are at a job opportunity access _disadvantage_ relative to the other sub-populations. This could be the sub-populations using the slowest mode $x$ at either Urban $B$ or Satellite $C$ area. From the perspective of inequities, the competition for opportunities between different mode-using populations (i.e., how well the land-use and transport system may serve some and not others), matters.

```{r toy-example-table-all-values}
autonum <- run_autonum(seq_id = "tab", bkm = "Tab1")
table_toy_format_flex <- flextable::set_caption(table_toy_format_flex, 
                                                caption = "Summary description of the synthetic example: accessibility values at each origin per mode m at each origin i and aggregated between modes for each i.",
                                                style = "Table Caption",
                                                autonum = autonum)

table_toy_format_flex
```

The calculated $S_i^m$, $a_i^m$ and $V_i^m$ accessibility values for each $i$ and $m$ are shown in the middle three columns and are aggregated for each $i$ in the final two columns in Table (\ref{tab:Tab1}). We use a negative exponential impedance function $f(c_{ij}) = \exp(-\beta\cdot c_{ij})$ with $\beta=0.1$ for both $x$ and $z$ modes for all accessibility measures calculations.

The Hansen-type measure $S_i^m$ is presented for each origin and mode in third column of Table (\ref{tab:Tab1}). For all $i$, the sub-population using the faster mode $z$ has higher $S_i^m$ values than the $x$ using sub-populations. Additionally, $S_i^m$ is equal for both mode using populations in population centers $A$ and $B$. This is the case because $S_i^m$ does not consider _competition_, it only relies on reflecting the count of opportunities that may be potentially interated with as a product of $f^m(c_{ij}^m)$. Recall, populations in $A$ and $B$ have the same travel impedance to employment centers $1$, $2$ and $3$ (either 15, 30, or 100 minutes for mode $x$ or 10, 25, or 80 minutes for mode $z$). As such, these the calculated $S_i^m$ values are the same for both $A$ and $B$. Furthermore, the total sum of $S_i^m$ in the region is equal to `r S_im$S_im %>% sum() %>% prettyNum(big.mark = ",")`. This value is difficult to interpret: it represents the weighted sum of opportunities that may be interacted with within the region based on travel impedance. It cannot be interpreted as a global maximum or any sort of benchmark as the measure is _non-constrained_. To connect this example to literature, this method of modal accessibility calculation is used in the work of @tahmasbiMultimodalAccessibilitybasedEquity2019; they compare differences in $S_i^m$ values between modes in a relative and comparative sense, but make no further interpretation of the $S_i^m$ values.

In the fourth and sixth column in Table (\ref{tab:Tab1}) the Shen-type measure is calculated: first for both origin and mode $a_i^m$ as well as aggregated by the weighted mean mode-population ( $\sum_m \frac{P_i^m}{P_i}*a_i^m$ ) to represent a value for each origin $a_i$. Unlike $S_i^m$, this measure considers _competition_. For instance, the populations using the same modes in $A$ and $B$ centers do not have the same $a_i^m$ values. In fact, the Suburban $A$ has the highest values since this center has the smallest travel impedance to opportunities (lower than at $C$) and has the lowest relatively proximate to populations (lower than at $B$).

However, the Shen-type measure is _non-constrained_: the total sum of $a_i^m$ or $a_i$ is practically meaningless since it represents a sum of ratios. For instance, the sub-population using the fast mode $z$ at $A$ has a value of `r a_i %>% filter(i=="A" & m=="z") %>% select(a_im) %>% round(digits=2)` potential jobs per potential job-seeking population compared to `r a_i %>% filter(i=="A" & m=="x") %>% select(a_im) %>% round(digits=2)` for the slow mode $x$ population. What is the significance of these values? The difference between these modes is equal to `r ((a_i %>% filter(i=="A" & m=="z") %>% select(a_im)) - (a_i %>% filter(i=="A" & m=="x") %>% select(a_im))) %>% round(digits=2)`, but `r ((a_i %>% filter(i=="A" & m=="z") %>% select(a_im)) - (a_i %>% filter(i=="A" & m=="x") %>% select(a_im))) %>% round(digits=2)` of what? How many more job opportunities are $z$ users interacting with than $x$ users? When $a_i^m$ is aggregated to $a_i$ as shown in the sixth column, the values face similar interpretability issues.

The Shen-type measure is implemented in the previously discussed work of @taoInvestigatingImpactsPublic2020a to calculate modal $a_i^m$ values and the aggregated $a_i$ is implemented in the work of @carpentieriMultimodalAccessibilityPrimary2020. However, similar to the Hansen-type measure, these works discuss relative and spatially comparative differences in values, they do not make further interpretation of the $a_i^m$ or $a_i$ themselves. This may be because the Shen-type measure is _non-constrained_, this is no benchmark or global maximum to which comparisons can be drawn from. Being unable to do so makes the interpretation of competition between modes challenging to tease out.

By contrast, spatial availability $V_i$ considers competition and is constrained such that the total sum of values is equal to the total number of opportunities in the region (i.e., $210,000$ jobs). Seen in fifth column of Table (\ref{tab:Tab1}), $V_i^m$ for the same mode-using populations in $A$ and $B$ are not the same (as this measure considers competition). In fact, at $A$ for instance, the sub-population using faster mode $z$ captures `r ((table_toy_format %>% filter(i == "A" & m == "z") %>% select("V_im")) - (table_toy_format %>% filter(i == "A" & m == "x") %>% select("V_im"))) %>% prettyNum(big.mark = ",")` more spatially available jobs (of the $210,000$ jobs in the region) than the sub-population using mode $x$. The numerical difference have a practical interpretation.

Furthermore, $V_i^m$ values for an $i$ can be aggregated across $m$ and compared across $i$ ( $V_i = \sum_m{\sum_i{V_i^m}}$ ) as a result of the proportional allocation mechanism. This aggregation, $V_i$, is shown in the seventh column in Table (\ref{tab:Tab1}). Again looking at center $A$, $A$ is allocated `r ((V_i %>% filter(i == "A" & m == "z") %>% select("V_i"))) %>% prettyNum(big.mark = ",")` spatially available opportunities for both modes. `r ((V_i %>% filter(i == "A" & m == "z") %>% select("V_im")) / ((V_i %>% filter(i == "A" & m == "z") %>% select("V_i")))) %>% as.numeric() %>% percent() ` of this spatial availability allocated to $A$ is assigned to the $z$-using population despite representing `r 0.66 %>% percent()` of the population at $A$.

Spatial availability can be further aggregated to better interpret competition between modes. Across the entire region,  `r V_ijm %>% filter(m=="z") %>% group_by(i) %>% summarize(pop = first(pop)) %>% select(pop) %>% sum() %>% prettyNum(big.mark = ",")` use the faster mode $z$ (`r ((V_ijm %>% filter(m=="z") %>% group_by(i) %>% summarize(pop = first(pop)) %>% select(pop) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% select("pop") %>% sum())) %>% percent()` of the region population). However, $z$-using populations accounts for `r ((V_ijm %>% filter(m=="z") %>% select(V_ij) %>% sum()) / (V_ijm %>% select(V_ij) %>% sum())) %>% percent()` of the region's total spatial availability - the rest is allocated to the population using $x$ (of which represents `r ((V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% filter(m == "x") %>% dplyr::select(pop) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% dplyr::select(pop) %>% sum())) %>% percent()` of the population). Notably, the $x$ population captures `r (((V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% filter(m == "x") %>% dplyr::select(pop) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% dplyr::select(pop) %>% sum())) - ((V_ijm %>% filter(m=="x") %>% select(V_ij) %>% sum()) / (V_ijm %>% select(V_ij) %>% sum()))) %>% percent()` less spatial availability to opportunities than its population proportion.

Since spatial availability is constrained and has an interpretable meaning of a proportion of the total opportunities in the region, the values have a new significance. Inequality in spatial availability values can be explored through a variety of approaches. For instance, consider travel times. The $z$ population using the faster mode $z$ accounts for `r ((V_ijm %>% group_by(i,m) %>% summarize(V_tt = sum(V_ij*tt), .groups = "drop") %>% filter(m == "z") %>% dplyr::select(V_tt) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(V_tt = sum(V_ij*tt), .groups = "drop") %>% dplyr::select(V_tt) %>% sum())) %>% percent()` of the potential travel time traveled in the region: 6% less travel time than the proportion of spatial available opportunities that the $z$ population has access to. Meaning $z$ population travels less potential minutes overall and has more spatial availability of opportunities than the population using the slower mode $x$.

Alternatively, these inequities in spatial availability between mode-using populations can be explored through proportional benchmarks. A spatial availability per capita as presented in Equation (\ref{eq:SA-per-capita}) is defined at each population center, mode, and/or regionally:

```{=tex}
\begin{equation}
\label{eq:SA-per-capita}
v_{im} = \frac{V_{im}}{P_{im}}
\end{equation}
```

The spatial availability per capita values $v_i^m$ for $A$, $B$, and $C$ for the slower mode $x$ is `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="A" & m =="x") %>% select(v_i) %>% sum() %>% round(digits=2)`, `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="B" & m =="x") %>% select(v_i) %>% sum() %>% round(digits=2)` and `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="C" & m =="x") %>% select(v_i) %>% sum() %>% round(digits=2)` respectively. The benchmarks for $x$ mode are evidently lower than the averages at: `r V_i %>% group_by(i) %>% summarize(v_i = sum(V_im)/sum(pop)) %>% filter(i=="A") %>% select(v_i) %>% sum() %>% round(digits=2)`, `r V_i %>% group_by(i) %>% summarize(v_i = sum(V_im)/sum(pop)) %>% filter(i=="B") %>% select(v_i) %>% sum() %>% round(digits=2)` and `r V_i %>% group_by(i) %>% summarize(v_i = sum(V_im)/sum(pop)) %>% filter(i=="C") %>% select(v_i) %>% sum() %>% round(digits=2)` respectively. Especially for $B$ and $C$.

The average for the faster mode $z$ with values of: `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="A" & m =="z") %>% select(v_i) %>% sum() %>% round(digits=2)`, `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="B" & m =="z") %>% select(v_i) %>% sum() %>% round(digits=2)` and `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="C" & m =="z") %>% select(v_i) %>% sum() %>% round(digits=2)` respectively. Once again, we see the competitive advantage on a per capita bases that populations have to opportunities when commuting using the faster mode.

In what follows, we further explore competition between multimodal accessibility and how competition between different modal-users is captured by spatial availability through an empirical example.

# Empirical multimodal spatial availability

## Data inputs: Madrid's LEZ and travel survey
```{r}
pop_props <- SA_im_zn208 %>% st_drop_geometry() %>% group_by(mode) %>% summarize(population_m = sum(population_m))
```
```{r}
tld_empirical_car <- tld_empirical %>% filter(mode == "car/motor") 
tld_empirical_transit <- tld_empirical %>% filter(mode == "transit") 
tld_empirical_bike <- tld_empirical %>% filter(mode == "bike") 
tld_empirical_walk <- tld_empirical %>% filter(mode == "walk")

car_qs <- weighted_quantile(x=tld_empirical_car$TIEMPO_VIAJES, w=tld_empirical_car$ELE_G_POND_ESC2)
transit_qs <- weighted_quantile(x=tld_empirical_transit$TIEMPO_VIAJES, w=tld_empirical_transit$ELE_G_POND_ESC2)
bike_qs <- weighted_quantile(x=tld_empirical_bike$TIEMPO_VIAJES, w=tld_empirical_bike$ELE_G_POND_ESC2)
walk_qs <- weighted_quantile(x=tld_empirical_walk$TIEMPO_VIAJES, w=tld_empirical_walk$ELE_G_POND_ESC2)
```

<!-- a summary about low emission zones, they are a geographic accessibility restriction. they change the accessibility landscape-->

Low emission zones (LEZ) have been implemented as a climate change policy intervention to reduce GHG emissions, improve air quality, and support sustainable mobility in many countries. Though rules vary depending on legal aspects and cultural norms, LEZ aim to deter or reduce traffic in specially designated zones under the penalty of fines and/or seizure of vehicle. In practice, this limits the volume of traffic by excluding vehicles by license plate, fuel type, or by introducing tolls.

Spain is one of a few countries which have active LEZ and are interested in expanding their implementation. As specified in their recent climate change plan _Plan Nacional Integrado de Energía y Clima 2021-2030_ [@espanaPlanNacionalIntegrado2020] and the _Plan Nacional de Control de la Contaminación Atmosférica_ [@espanaResolucion10Enero2020], more LEZ will soon be implemented throughout the country. Specifically, the national Spanish law 7/2021 (“Ley de Cambio Climático y Transición Energética”) will require all municipalities to implement LEZ by 2023 if they meet at least one of the following requirements: (i) municipalities >50,000 inhab.; (ii) islands; and (iii) municipalities > 20,000 inhab. when air quality exceeds limits specified in “RD 102/2011 de Mejora de Calidad del Aire” [@barcelonaGUIATECNICAPARA2021].

In 2017, a LEZ was implemented in the capital city of Madrid following the goals set out in the national agenda: to fight climate change, cut nitrogen dioxide levels for the benefit of people's health, and prioritize people's movement in the city. In geographic scope, the 2017 boundaries of the LEZ were relatively small (covering 4.72 km^{2}) and within the center (i.e., LEZ Centro). They were expanded in 2023 to just inside of the M-30, a highway in proximity to the city center (i.e., LEZ M-30) and the city has plans to further spatially expand the LEZ in the future. Within the 2017 LEZ Centro implementation, all cars, motorcycles and freight with environmental label A or B (higher polluting classification, associated with older make and model of fossil fuel internal combustion engine vehicles), are banned from driving into the area unless they are used by residents or meet other exemptions [@tarrinoortizAnalyzingImpactLow2022]. This restriction impacts approximately half of all car that made trips into LEZ Centro.

From the perspective of restriction for passenger transport, LEZ are a policy of _geographic discrimination_. LEZ actively change how people access opportunities by making the travel impedance more costly for car-mode users. If seeing opportunities as finite, this discrimination allows populations to access opportunities by other modes more readily than before. In this way, the policy changes the multimodal competitive accessibility landscape of the city. Though the cost of travel for modes do dynamically change as a result of the LEZ implementation (i.e., potentially less car congestion, potentially more transit congestion), the focus of this empirical example is to demonstrate the spatial availability, by mode, of the city of Madrid after the 2017 implementation of the LEZ Centro.

```{r, jobs-plot, fig.cap="\\label{fig:Fig2} Jobs $O_j$ taken by people living and working in Madrid as reported by the 2018 travel survey.", fig.show='hold', fig.align='center'}
knitr::include_graphics("images/i_jobs_zn208_plot.png") 
```

```{r, pop-plot, fig.cap="\\label{fig:Fig3} Population living and working in Madrid, by four summarized modal categories, $P^m_i$ as reported by the 2018 travel survey.", fig.show='hold', fig.align='center'}
knitr::include_graphics("images/im_populations_zn208_plot.png")
```
The 2018 Community of Madrid travel survey is the source of data for the empirical example: it is a representative survey that reflects a snap-shot of the travel patterns for one typical day of the working week (e.g., n=222,744 trips with representative population elevation factors). In this paper, a sample of the travel survey is used, namely the residential home origin to work destination trips of all modes and those that originate and end in Madrid. These totals are displayed in Figure \ref{fig:Fig2} and Figure \ref{fig:Fig3}. Both figures are displayed at the level of traffic analysis zones ($i$ and $j$) that correspond to the 2018 travel survey. The red boundary represents the LEZ Centro in effect in 2017 and thus those travel patterns of car-restriction reflected in the 2018 travel survey data. The cyan boundary represents the LEZ that will be within the boundaries of the M-30 highway in 2023 and is present in the plots as a spatial reference for areas in proximity to the LEZ Centro.

The total sum of jobs $O_j$ that are held are shown in Figure \ref{fig:Fig2} and the populations that go to a work destination by four modal categories $P^m_i$, is reflected in Figure \ref{fig:Fig3}. The modal categories represented in Figure \ref{fig:Fig3} were summarized for the following trip mode types:
   - Car/motor: all cars and operating modes (e.g., cab, private driver, company, rental care, main driver, passenger, etc.) and all public, private or company motorcycle/mopeds.
  - Transit: all bus, trams, and trains
  - Bike: all bicycle trips (e.g., private, public, or company bike trips) and "other" types of micromobility options
  - Walk: walking or by foot

From Figure \ref{fig:Fig2}, it can be seen that the largest concentration of jobs are within, near, and to the north of the LEZ Centro.. The population that is accessing those jobs by mode (Figure \ref{fig:Fig3}), appear spatially distinct. Car and transit trips represent `r (pop_props$population_m[3]/sum(pop_props$population_m)) %>% percent()` and `r (pop_props$population_m[4]/sum(pop_props$population_m)) %>% percent()` of the modal share respectively. The population that travels using transit is more spatially distributed than those using cars - particularly near and within LEZ Centro. This distribution could be a result of a variety of factors including: transit coverage and service within with city,  effective car infrastructure outside of the M-30, and/or the impact of the Central LEZ itself.

From Figure \ref{fig:Fig2}, it can also be seen that biking and walking trips are less common than motorized trips at `r (pop_props$population_m[1]/sum(pop_props$population_m)) %>% percent()` and `r (pop_props$population_m[2]/sum(pop_props$population_m)) %>% percent()` respectively. The distribution of walking and biking trips appear to be similar to that of transit trips. This is to be expected as active transport and a higher diversity of land-use spatially occurs with transit infrastructure. <!--CITE-->

The travel time for each trip is provided within the 2018 survey. These travel times, per modal category, are used to calibrate mode specific travel impedance functions $f^m(c_{ij}^m)$. To illustrate the modal differences in travel lengths, summary descriptive per mode are detailed as follows:

- Car/motor: `r weighted_mean(x=tld_empirical_car$TIEMPO_VIAJES, w=tld_empirical_car$ELE_G_POND_ESC2) %>% round()` min (Min:`r car_qs[1]%>% round()`min, Q2: `r car_qs[2]%>% round()` min, Q3: `r car_qs[4]%>% round()` min, Max: `r car_qs[5]%>% round()` min) 
- Transit: `r weighted_mean(x=tld_empirical_transit$TIEMPO_VIAJES, w=tld_empirical_transit$ELE_G_POND_ESC2)%>% round()` min (Min:`r transit_qs[1]%>% round()` min, Q2: `r transit_qs[2]%>% round()` min, Q3: `r transit_qs[4]%>% round()` min, Max: `r transit_qs[5]%>% round()` min) 
- Bike: `r weighted_mean(x=tld_empirical_bike$TIEMPO_VIAJES, w=tld_empirical_bike$ELE_G_POND_ESC2)%>% round()` min (Min:`r bike_qs[1]%>% round()` min, Q2: `r bike_qs[2]%>% round()` min, Q3: `r bike_qs[4]%>% round()` min, Max: `r bike_qs[5]%>% round()` min) 
- Walk: `r weighted_mean(x=tld_empirical_walk$TIEMPO_VIAJES, w=tld_empirical_walk$ELE_G_POND_ESC2)%>% round()` min (Min:`r walk_qs[1]%>% round()` min, Q2: `r walk_qs[2]%>% round()` min, Q3: `r walk_qs[4]%>% round()` min, Max: `r walk_qs[5]%>% round()` min)

To calculate the mode specific travel impedance functions $f^m(c_{ij}^m)$ from the travel times, a concept known as the trip length distribution (TLD) is used. A TLD represents the proportion of trips that are taken at a specific travel cost such as travel time. This distribution is then used to derive impedance functions as done in previous accessibility research [e.g., works of @lopez_2017_spatial, @horbachov_theoretical_2018, and @batista_estimation_2019]. Maximum likelihood estimation and the Nelder-Mead method for direct optimization available within the R {fitdistrplus} package [@fitdistrplus_2015] is used. As shown as shown in Figure \ref{fig:Fig4}, based on goodness-of-fit criteria and associated diagnostics, the gamma and log-normal probability density function (line curves) are selected as best fitting curves for the motorized and non-motorized modes respectively. The selection of functional form aligns with examples used in the literature (e.g., @reggianiAccessibilityImpedanceForms2011). <!--e.g., car gamma XX and active transport log-normal XX.--> Overall, the plots in Figure \ref{fig:Fig4} display the probability of travel given a trip travel time, based on actual travel behaviour from the 2018 survey. These 'probability of travel' at each travel time for each mode are realized observations reflect the land-use, the transport system, and the population travel preferences/behaviour in Madrid.

<!-- TOO MUCH DETAIL, removed. 
They are used in this study (as similar to other work that uses TLD [e.g., @lopez_2017_spatial, @horbachov_theoretical_2018, @batista_estimation_2019]) to reflect the travel impedance of each mode to work destinations in Madrid. Overall, as seen from the four plots, trips under 5 minutes do not occur frequently for any mode and as such are assigned a lower travel impedance $f^m(c_{ij}^m)$ value. One reason that these under 5 minute trips do not frequently occur as a result of land-use (residential and job mismatch): not everyone lives in immediate proximity to their workplace. Observing the non-motorized modes, shorter trips occurred more frequently overall for walking populations, particularly around 15 minute lengths, so a trip of approximately 15 minutes is assigned the highest $f^m(c_{ij}^m)$. For biking populations, longer travel times are more common so though the highest $f^m(c_{ij}^m)$ value also corresponds to approximately 15 minutes, the curve is more spread out and values decrease less rapidly at longer travel times than for the walk mode curve. A similar trend occurs for the motorized modal options where transit mode is more spread out than car/motor mode.  -->

```{r, tlds-curves-m-plot, fig.cap="\\label{fig:Fig4} Fitted impedance function curve (line) against empirical TLD (bars) per mode.", fig.show='hold', fig.align='center', fig.height=6}
knitr::include_graphics("images/tlds_curves_m_plot.png")
```

## Results: spatial availability $V_i^m$ and $V_i$

Using the data inputs outlined, spatial availability of jobs are calculated for each of the four modal categories at the level of traffic analysis zones in Madrid, $V_i^m$. The spatial distribution of the resulting calculations are demonstrated in this section.

```{r, SA-m-plot, fig.cap="\\label{fig:Fig5} Spatial availability of job opportunities per origin and mode $V_i^m$ in Madrid as reported by the home-to-work origin destination flows from the 2018 travel survey. 2017 central LEZ is shown in blue. 2023 expanded LEZ boundaries shown in red.", fig.show='hold', fig.align='center'}
knitr::include_graphics("images/SA_im_V_zn208_plot.png")
```
```{r, eval=FALSE}
#Check that the totals are maintained, it is!:
sum(OD_SA_zn208$V_ij, na.rm = T) #847574 workers -- 
OD_SA_zn208 %>% group_by(VDES_CD_ZT208) %>% summarize(jobs = mean(jobs)) %>% select(jobs) %>% sum()

sum(OD_SA_zn208$tt)
sum(OD_SA_zn208$total_ttworkers)

#for comparision, the number of travel minutes spent going to jobs. group by jobs zona, and the the number of potential minutes traveling to jobs. This sum is the number of minutes spent commuting to jobs. it is 33680009 minutes, this is approx equal to 43min (mean travel time) *847574 (the total number of workers).

OD_SA_zn208 %>% group_by(VDES_CD_ZT208) %>% 
  summarize(jobs_tt_f_t = sum(jobs_tt_f_t)) %>% 
  select(jobs_tt_f_t) %>% sum() #should equal 33680009 mins total
```

```{r}
# Spatial availability by mode, we can see that car mode captures ~48% of the spatial availability but only accounts for ~36% of the tt_ijm !!:
modal_V_comps <- OD_SA_zn208 %>%
  group_by(mode) %>%
  summarize(V_m = sum(V_ij),
            V_m_perc = sum(V_ij)/847574,
            jobs_tt_f_t = sum(jobs_tt_f_t)/33680009)
#modal_V_comps
```

Figure \ref{fig:Fig5} displays the spatial availability values for the four modal categories at the level of the spatial units used in the 2018 travel survey. These values represent a proportion of the total number of the `r SA_i_zn208$V_i %>% sum()  %>% prettyNum(big.mark = ",")` jobs in the region that are _spatially available_ to the population located at the $i$ based on the travel impedance of the mode (relative to the travel impedance of all modes) and the mode-using population size at the $i$ (relative to the population size of all modes and $i$). Since population, their job locations, and associated travel times are used to calculate $V_i^m$, Figure \ref{fig:Fig5} demonstrates values that represent the _realized_ spatial availability of jobs in Madrid. In other words, if someone decided to move into a neighbourhood at an $i$, the amount of jobs that are spatially available to them based on the population within their $i$ and their travel impedance for their used $m$, both relative to the city, is captured by the value of $V_i^m$.

In Figure \ref{fig:Fig5}, the difference in the magnitudes of $V_i^m$ values between $m$ can be observed. The majority of $V_i^m$ is allocated to the populations using motorized modes. This is to be expected as commuting using motorized modes represents `r (pop_props$population_m[3]/sum(pop_props$population_m) + pop_props$population_m[4]/sum(pop_props$population_m)) %>% percent()` of the population (`r (pop_props$population_m[3]/sum(pop_props$population_m)) %>% percent()` (car/motor) and `r (pop_props$population_m[4]/sum(pop_props$population_m)) %>% percent()` (transit)). However, these modal options capture `r (modal_V_comps$V_m_perc[3] + modal_V_comps$V_m_perc[4]) %>% percent()` of the total spatial availability in Madrid. In particular, the car/motor using population is allocated disproportionately more $V_i^m$ than its modal population (`r (pop_props$population_m[3]/sum(pop_props$population_m)) %>% percent()` of the population vs. `r modal_V_comps$V_m_perc[3] %>% percent()` of the $V_i^m$) compare to the transit using population and its relatively proportional $V_i^m$ value (`r (pop_props$population_m[2]/sum(pop_props$population_m)) %>% percent()` of the population vs. `r modal_V_comps$V_m_perc[4] %>% percent()` $V_i^m$).

How does the $V_i^m$ advantage allocated to car-using population arise? From the perceptive of finite opportunities, $V_i^m$ is allocated to car-using populations from less competitive modal populations. How competitive one mode is compared to other modes varies spatially, but overall car-using populations capture more opportunities per car-using population than other modal populations. Namely, though walking and cycling populations represent `r (pop_props$population_m[2]/sum(pop_props$population_m)) %>% percent(accuracy = 0.01)` and `r (pop_props$population_m[1]/sum(pop_props$population_m)) %>% percent(accuracy = 0.01)` respectively, $V_i^m={walk}$ and $V_i^m={bike}$ is `r modal_V_comps$V_m_perc[2] %>% percent(accuracy = 0.01)` and `r modal_V_comps$V_m_perc[1] %>% percent(accuracy = 0.01)` in the region respectively. These modes are less competitive, especially compared to the car/motor mode, as a result of: 1) their lower travel impedance values at longer travel times (see Figure \ref{fig:Fig4} at travel times beyond ~30 minutes), 2) their low population values values overall, and 3) higher populations present in origins with high motorized mode commuting. These factors all contribute to the the car/motor mode being most advantaged in capturing spatially available job opportunities overall.

```{r, SA-per-capita-m-plot, fig.cap="\\label{fig:Fig6} Spatial availability of job opportunities per capita per origin and mode $v_i^m$ in Madrid as reported by the home-to-work origin destination flows from the 2018 travel survey. 2017 central LEZ is shown in blue. 2023 expanded LEZ boundaries shown in red.", fig.show='hold', fig.align='center'}
knitr::include_graphics("images/SA_im_vv_zn208_plot.png")
```

Furthermore, there are spatial differences in the competitive advantage of spatial availability between modes. Figure \ref{fig:Fig6} visualizes $v_i^m$, the spatial availability divided by the mode population. $v_i^m$ values above 1 are represented in increasing red shades, values below 1 are represented in increasingly green shades, and values equal to 1 are white. These plots illustrates the discussion of the disproportionately high over representation of spatial availability relative to the mode-using population in many of the origins for the car plot (bottom right, denoted with green $v_i^m$ values above 1). These plots also visualize areas that are awarded disproportionately low spatial availability, represented in shades of red. Interestingly, spatial availability for the car mode within the LEZ Centro is below 1 (red). For all other modes, the area with the LEZ is relatively higher than the modal averages or even green in the case transit-using populations. This difference in spatial availability can be seen as a direct result of the LEZ Centro - the observed reduction of opportunities within the LEZ Centro boundaries being accessed by car-using populations allows lesser competitive modes to interact with these opportunities.

```{r, SA-per-capita-m-linear-plot, fig.cap="\\label{fig:Fig7} Spatial availability of job opportunities per capita per origin and mode $v_i^m$ in Madrid as reported by the home-to-work origin destination flows from the 2018 travel survey represented by sorted total population per origin.", fig.show='hold', fig.align='center'}
knitr::include_graphics("images/v_im_per_pop_plot.png")
```

Furthermore, Figure \ref{fig:Fig6} makes it evident that transit-using population's spatial availability to jobs is relatively balanced (i.e., many zones are white). Additionally, values for non-motorized modes are higher in origins that have higher transit accessibility. Transit accessibility and non-motorized modes do not appear to be in direct competition as a result of different travel impedance weighting <!-- (? not sure about this)-->. It is not visually clear in the plots but as demonstrated in a line plot of $v_i^m$ sorted by total population in Figure \ref{fig:Fig7}, it can be observed that transit is only close to or above 1 when car accessibility is relatively low. From Figure \ref{fig:Fig7} we can also observe that overall, motorized modes capture more spatial availability per capita than non-motorized modes. <!-- I'm considering removing Fig 7, the paper is becoming too long for TRB! I don't have enough room to explain it -->

## Discussion and conclusions

Location-based accessibility measures like the Hansen-type measure, Shen-type measure, and spatial availability all have a commonality - they are a weighted sum of opportunities assigned to each spatial unit in a region. In this way, they can be interpreted as a score that represents how many opportunities can be potentially interacted with by the population at each spatial unit. How the weight and sum of the potentially interacted opportunities is what defines the type of accessibility measure. Within this paper, the location-based accessibility measures known as spatial availability, a singly- _constrained_ and _competitive_ measure, is extended for the case of capturing multimodal accessibility to opportunities. A synthetic example and then an empirical case of LEZ in Madrid are detailed to demonstrate the multimodal extension of the spatial availability measure.

The spatial availability measure is capable of capturing a new interpretation of multimodal competition that previous accessibility measures have not yet done. We can hypothesis that populations using modes with lower travel impedance, when competing for a finite set of opportunities, will capture more opportunities. However, with spatial availability, the number of spatially available opportunities that are captured (of the total opportunities in the region) by each mode can be individually calculated. From there, the difference between how many spatially available opportunities one mode captures versus another can be investigated.

The flexibility and need for an accessibility measure such as spatial availability is pertinent in policy scenario evaluation. As showcased in the empirical example of the LEZ in Madrid, competition for job opportunity availability highly varies spatially and between modes. The car and transit modes have the highest spatial availability, with the car mode having highest availability with exception to the areas with LEZ Centro. This finding reflects real conditions: since car travel has been highly restricted within the LEZ Centro, much fewer car-using population and much more people are entering using other modes relative to the areas surrounding the LEZ Centro. This difference in car-using population within and immediately outside the LEZ Centro increases the competitiveness of the transit-using population (the second most competitive mode) as well as the non-motorized modes.

Currently, conventional _non-constrained_ accessibility measures are difficult for planners to operationalize for a variety of reasons. They have also been criticized for being difficult to compute and difficult to interpret as they are a ratio of supply to demand [@levinsonTransportAccessManual2020].  With spatial availability, the magnitude of opportunities that are available as a proportion of all the opportunities in the region is equal to $V_i$. As a result of its proportional allocation mechanism it can be easily extended into multimodal applications, pertinent to model policy scenerios in our cities that are becoming increasingly multimodal. This flexibility and interpretation of the spatial availability measure allows for manipulation of $V_i^m$ values to investigate differences of availability between neighbourhoods, modes, and regions, generate per capita benchmarks, and/or generate average values per population-group.

From a spatial equity perspective, spatial availability measure can provide researchers, policy makers, and citizens a newfound interpretation of accessibility measures. With a plot of spatial availability values, one can begin asking, how much is enough and what level is too much. These interpretations were difficult to be made with accessibility measures in the past.

## Acknowledgements

This research was funded by the Canada Graduate Scholarship - Doctoral Program (CGS D) by the Social Sciences and Humanities Research Council (SSHRC).

#AUTHOR CONTRIBUTIONS

The authors confirm contribution to the paper as follows: study conception and design: AS, JSL, AP.; data collection: AS, JSL, AP.; analysis and interpretation of results: AS, JSL, AP.; draft manuscript preparation: AS, JSL, AP. All authors reviewed the results and approved the final version of the manuscript.

<!-- rejected text is left here -->

<!-- As reviewed in the work of [@levinsonTransportAccessManual2020], cumulative opportunity measures and weighted cumulative opportunity measures are commonly used location-based accessibility measures; they use distance-decay functions (otherwise known as an impedance functions) to weight the opportunities that can be potentially accessed from origin to destination based on some sort of travel cost (e.g., travel time, fare, travel distance). A score is calculated for each location in the region; the score represents the number of opportunities that can be potentially interacted after being discounted by travel impedance.  -->

<!-- Cumulative opportunity and weighted cumulative opportunities measures can be understood to have two properties: no consideration for _constraints_ or _competition_. The cumulative opportunity measure weights using a binary impedance function and the weighted cumulative opportunity measure weights using an impedance function that can take any functional form (e.g, exponential, log-normal, or even any sort of probability density function as discussed in the Trip Length Distribution literature). They are both _non-constrained_ because there is no consistent consideration of regional totals that produce some global maximum; they do not discount the number of opportunities in any constrained manner e.g., no consideration of the total magnitude of opportunities, total population, total inefficiencies in the transport system, etc. These measure are also _non-competitive_ as they do not consider the population demand for opportunities.  -->

<!-- <!-- examples of constrained (competitive) measures, the introduction of spatial availability --> -->
<!-- Competitive measures are are also a weighted sum of opportunities, but do so in different ways. They weight opportunities by the travel impedance in addition to the supply of the opportunities and the demand for those opportunities. They have been criticized for being difficult to compute (relative to unconstrained measures) and difficult to interpret as they are a ratio of supply to demand [@levinsonTransportAccessManual2020].  -->

<!-- In @soukhovIntroducingSpatialAvailability2023, we introduce spatial availability, a type of competitive accessibility measure that is _constrained_ alongside a popular _non-constrained_ and _non-competitive_ accessibility measure (Hansen-type gravity model) and a _non-constrained_ and _competitive_ accessibility measure (two-step floating catchment area (2SFCA)). We title spatial availability _constrained_  and _competitive_ measure as it proportionally allocates all the opportunities (supply) to each origin-location based on competition from the population (demand) and travel impedance in a single step. The output can be simply interpreted as the number of opportunities available in the region, per location, such that if all the values for the locations in the region were summed they would equal the total opportunities in the region. -->

<!-- <!-- policy applications of accessibility broadly, but what's missing is multi-modal accessibility methods --> -->
<!-- The consideration of both supply of opportunities and demand from the population is important for policy applications. Spatial availability can be used to create supply per capita benchmarks for locations, and the delta of supply from this benchmark to existing conditions can be used to motivate policy. Currently, conventional _non-constrained_ accessibility measures are difficult for planners to operationalize for a variety of reasons. Furthermore, existing competitive measures are difficult to interpret. In the following, we demonstrate  -->

<!-- In another study that implemented Shen-type accessibility, @carpentieriMultimodalAccessibilityPrimary2020 calculates an aggregated $a_i$ to primary healthcare services in Naples (Italy) using combined walking and transit modes for older adults. Three older-adult sub-groups and their different walking speeds are considered in using the public transit network. They find that the elderly population in suburban areas have low accessibility - but how much lower is this accessibility than the urban center? What is the impact that competition has on this difference? What is the interpretation of this difference? -->

<!-- If opportunities in a region are considered finite, than one person being able to potentially interact with more opportunities than others in the region means there are fewer opportunities available for others. Conceptualizing potential interaction with opportunities as a region-relative value of what is _available_ to each location $i$ ishelpful from the perspective of comparative interpretation. -->

<!-- It can be seen that the general form of spatial availability is a sum of weighted opportunities. All location-based accessibility measures, including spatial availability, have this commonality - they are a weighted sum of opportunities assigned to each spatial unit in a region. How the weight and sum of the potential interaction opportunities is what defines the type of accessibility measure. Within this paper, location-based accessibility measures are discussed from the classification of being _constrained_ and/or _competitive_. -->

### 2023-07-30 TRB-manuscript/TRB-manuscript.tex
TRB draft - ready for AP edits

% Repository:  https://github.com/chiehrosswang/TRB_LaTeX_tex
%
% Transportation Research Board conference paper template
% version 4.0 Lite (updates made to be compatible in Overleaf and ShareLaTeX)
%
%
% When numbered option is activated, lines are numbered.
\documentclass[numbered]{trbunofficial}
\usepackage{graphicx}
\usepackage{booktabs}

\newread\somefile
\usepackage{xparse}
\usepackage{natbib}
\bibliographystyle{unsrtnat}
\setcitestyle{round}
% \usepackage[colorlinks=true,linkcolor=blue,citecolor=blue]{hyperref}
% For TRB version hide links
\usepackage[hidelinks]{hyperref}

% Put here what will go to headers as author
\AuthorHeaders{Multimodal spatial availability - the case of low
emission zones in Madrid}
\title{Multimodal spatial availability - the case of low emission zones
in Madrid}

% TODO: add macros for easier formatting of \author.
\author{%
    \textbf{Anastasia Soukhov}\\\textit{Corresponding Author}\\
  PhD Candidate\\
  Department of Earth, Environment and Society, McMaster University,
Canada\\
  \href{mailto:soukhoa@mcmaster.ca}{\nolinkurl{soukhoa@mcmaster.ca}}\\
  \hfill\break
    \textbf{Javier Tarriño Ortiz}\\
  PhD Graduate\\
  Universidad Politécnica de Madrid, Spain\\
  \href{mailto:bob@example.com}{\nolinkurl{bob@example.com}}\\
  \hfill\break
    \textbf{Julio A. Soria-Lara}\\
  Associate Professor\\
  Urban and Regional Planning, Universidad Politécnica de Madrid,
Spain\\
  \href{mailto:derek@example.com}{\nolinkurl{derek@example.com}}\\
  \hfill\break
    \textbf{Antonio Páez}\\
  Associate Professor\\
  Department of Earth, Environment and Society, McMaster University,
Canada\\
  \href{mailto:paezha@mcmaster.ca}{\nolinkurl{paezha@mcmaster.ca}}\\
  \hfill\break
  }

% If necessary modify the number of words per table or figure default is set to
% 250 words per table (default defined in cls)

% If words are counted manually, put that number here. This does not include
% figures and tables. This can also be used to avoid problems with texcount
% program i.e. if one does not have it installed.
\TotalWords{7500}

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{multirow}
\usepackage{multicol}
\usepackage{colortbl}
\usepackage{hhline}
\newlength\Oldarrayrulewidth
\newlength\Oldtabcolsep
\usepackage{longtable}
\usepackage{array}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}

\begin{document}
\maketitle

\section{Abstract}
An increasing number of studies within the domain of transportation
planning are concerned with the inequities in accessibility to
opportunities. A dimension of these inequities arises from differences
in access by mode type (e.g., commuting using a car as opposed to
transit). However, methods implemented in current accessibility
literature are lacking within the context of multi-modal analysis. This
paper presents an extension of spatial availability, a
singly-constrained competitive accessibility measure, for the context of
multi-modal accessibility analysis. We first illustrate the features of
spatial availability that lend itself to multi-modal analysis. We then
demonstrate its use on the case study of Low Emission Zones in Madrid
(Spain) and highlight how this policy intervention changes the
accessibility of populations using different modes. In summary, spatial
availability can be used to create and interpret multi-modal policy
intervention scenarios unlike previous methods: this creation and
interpretation can help regions envision a more sustainable and
equitable access-to-opportunity landscape.
\hfill\break%
\hfill\break%
\noindent\textit{Keywords}:  Multimodal, Accessibility, Equity, Policy
scenarios, Low emission zones,  
\newpage

\global\setlength{\Oldarrayrulewidth}{\arrayrulewidth}

\global\setlength{\Oldtabcolsep}{\tabcolsep}

\setlength{\tabcolsep}{0pt}

\renewcommand*{\arraystretch}{1.5}

\providecommand{\ascline}[3]{\noalign{\global\arrayrulewidth #1}\arrayrulecolor[HTML]{#2}\cline{#3}}

\begin{longtable}[c]{|p{0.88in}|p{0.41in}|p{1.05in}|p{0.58in}|p{1.05in}|p{0.58in}|p{1.05in}}

\ascline{1.5pt}{666666}{1-7}

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{i}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{m}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{S}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}} \\

\ascline{1.5pt}{666666}{1-7}\endfirsthead

\ascline{1.5pt}{666666}{1-7}

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{i}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{m}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{S}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}} \\

\ascline{1.5pt}{666666}{1-7}\endhead

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{x}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{27,292.18}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.95}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{18,959.86}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.88in}{\raggedright \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{A}}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{z}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{44,999.80}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.56}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{46,913.04}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.58in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.32}}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{1.05in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{65,872.91}}}}} \\

\ascline{1pt}{666666}{1-7}

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{x}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{27,292.18}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.62}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{30,863.43}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.88in}{\raggedright \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{B}}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{z}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{44,999.80}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.03}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{103,391.95}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.58in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.90}}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{1.05in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{134,255.38}}}}} \\

\ascline{1pt}{666666}{1-7}

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{x}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{2,240.38}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.68}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{2,034.49}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.88in}{\raggedright \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{C}}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{z}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{3,745.89}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.12}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{7,837.22}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.58in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.99}}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{1.05in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{9,871.71}}}}} \\

\ascline{1pt}{666666}{1-7}

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{TOTALS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{150,570.22}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{N/A}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{210,000.00}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{N/A}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{210,000.00}}} \\

\ascline{1.5pt}{666666}{1-7}

\end{longtable}

\arrayrulecolor[HTML]{000000}

\global\setlength{\arrayrulewidth}{\Oldarrayrulewidth}

\global\setlength{\tabcolsep}{\Oldtabcolsep}

\renewcommand*{\arraystretch}{1}

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

An increasing number of studies within the domain of transport are
concerned with the inequities in accessibility to opportunities. A
dimension of these inequities arise from differences in access by mode
type (e.g., the number of work opportunities that can be reached using a
car as opposed to transit in a city). However, methods assessing
multimodal accessibility in the literature fall short as aspects of
competition for opportunities and the explicit methodological
acknowledgement of opportunities being \emph{finite} are lacking. In
this vein, this paper presents an extension of \emph{spatial
availability}, a singly-constrained competitive accessibility measure,
for the context of multimodal accessibility analysis. We first
illustrate the features of spatial availability that lends itself to
multimodal analysis. We then demonstrate its use on the case study of
Low Emission Zones in Madrid (Spain) and highlight how this policy
intervention changes the accessibility of populations using different
modes. In summary, spatial availability can be used to create and
interpret multimodal policy intervention scenarios unlike previous
methods: this creation and interpretation can help regions envision a
more sustainable and equitable access-to-opportunity landscape by better
identifying differences in accessibility afforded by different modes.

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Implementing urban policies that re-shape cities through accessibility
gains (i.e., the \emph{potential to interact} with opportunities as a
result of land-use mix and transport systems as originally defined by
\citet{hansenHowAccessibilityShapes1959}) have been widely applied
within the transportation literature and is increasingly discussed by
planners
\citep{levinsonTransportAccessManual2020, gowerPlanningInnovationCity2022, siddiqToolsTradeAssessing2021, yanAccessibilityBasedPlanningAddressing2021}.
An important challenge in the identification of interventions that
sustainably and equitably transform cities is the effective evaluation
of \emph{trade-offs}: cities are complex and dynamic ecologies, and
advantaging one component of the city can disadvantage another area,
population, or sub-component In this way, policy evaluation should take
a \emph{systems} approach
\citep{fikselSustainabilityResilienceSystems2006}. One way of
considering systems is from the perspective of the \emph{finite}. As an
illustration, consider the amount of transport space within a city: the
amount is typically finite so re-allocating road space away from one
mode directly impacts the performance of the others (see the literature
on road space reallocation e.g.,
\citet{valencaMainChallengesOpportunities2021}). Evaluating policy
impacts in the context of \emph{finity} provides a way to contextualize
the balance of trade-offs that the citizens of a city should tolerate.

From the perspective of urban transport systems, location-based
accessibility measures have been used in the context of policy
evaluation. For instance, \citet{leeMeasuringImpactsNew2018} assesses
the transit accessibility gains to healthcare and employment
opportunities for disadvantaged neighbourhood in Columbus, Ohio, USA
after the transit system's re-design and introduction of a rapid bus
system. However, a limitation of this study, like others that implement
accessibility measures, is they do not calculate results under a
\emph{constrained} framework i.e., one of \emph{finity}. The citizens of
Columbus should experience qualitative accessibility gains - but is it
at the expense of access to opportunities by other modes? As another
example, \citet{mohriClusteringMethodMeasuring2021} implements a
modified cumulative opportunity measure to assess differences between
private vehicle and transit system accessibility to jobs in Melbourne,
but a similar question remains, does the accessibility afforded to the
private vehicle using population come at the expense of accessibility
losses to transit users?

The two studies discussed in the previous paragraph use
\emph{non-competitive} accessibility measures. There is a branch of
location-based accessibility measures that do incorporate the effect of
competition for opportunities by the population in the region. However,
we argue that these existing methods fall short in acknowledging the
\emph{finity} of opportunities. For instance,
\citet{maoMeasuringSpatialAccessibility2013} applies the competitive
measure, the two-step floating catchment approach (2SFCA) for the case
of access to healthcare services in Florida for both a multimodal
network and a single modal network. While the differences in modal
access are discussed, the question of how the advantage in access
afforded by one mode over another impacts access for different mode
users is unanswered.

This question of how much one mode-using population can access at the
expense of another mode-using population is pertinent equity question in
the evaluation of policy scenarios that are multimodal. For instance,
consider the impact of a low emission zone (LEZ). LEZ is a policy of
spatial and modal discrimination: the circulation of vehicles that are
excessively polluting are restricted in specific areas in a city. In the
recognition that opportunities are finite, the implementation of a LEZ
explicitly reduces the access that the population using polluting
vehicles has to opportunities. This restriction allows the population
using other more sustainable modes to potentially have a higher level of
access than before the LEZ implementation. This evaluation is especially
urgent as LEZ are currently in effect in cities globally; their
reception has been mixed \citep{tarrinoortizPublicAcceptabilityLow2021}
and may be having negative impacts on disadvantaged populations who have
become mobility restricted
\citep{devrijNooneVisitsMe2022, verbeekJustManagementUrban2022}.
Measures that evaluate the accessibility of modes given both
\emph{constrained} and \emph{competitive} considerations are lacking in
the literature, but are needed, to evaluate such policy interventions to
effectively comment on how accessibility changes as a result of the
mobility-restricted mode.

In \citet{soukhovIntroducingSpatialAvailability2023}, we introduce
spatial availability, a type of location-based accessibility measure
that is both \emph{constrained} and \emph{competitive}. In this paper,
we extend the spatial availability measure into a multimodal framework
and explore its use in answering the question outlined: \emph{``given
opportunities are finite, how many are available to a given location
depending on the mode used?''.} The answer to this question quantifies
how many opportunities can be accessed, considering competition, for
different modes. To foreground this exploration, in Section 2, we
discuss short falls of a few existing location-based measures in
comparison to spatial availability through a synthetic example. In
Section 3, the spatial availability of an empirical example of the LEZ
in the city of Madrid, Spain is calculated. We demonstrate how the
restriction of private vehicles with the LEZ implementation impacts the
spatial availability of opportunities for each sub-population using
transit, cycling and walking modes. In Section 4, we provide concluding
remarks on the strengths of the use of spatial availability as a
multi-modal accessibility measure, limitations, and potential future
uses in policy planning scenarios.

\hypertarget{a-brief-review-of-multimodal-accessibility-measurement-methods}{%
\section{A brief review of multimodal accessibility measurement
methods}\label{a-brief-review-of-multimodal-accessibility-measurement-methods}}

Location-based accessibility indicators are quantitative measures of
\emph{potential} interaction with opportunities for locations within a
given region: they are a product of the relationship between land-use
and transport systems. Arguably the most commonly used location-based
measured are cumulative opportunity measures and weighted cumulative
opportunity measures \citep{levinsonTransportAccessManual2020}. These
measures weight the opportunities that can be potentially interacted
with from origin \(i\) to destination \(j\) based on some sort of travel
cost function (e.g., travel time, fare, travel distance) otherwise known
as a travel impedance function \(f^{m}(c^m_{ij})\). Many weighted
cumulative opportunities (often refered to as the gravity-based measure)
originate from the measure proposed by
\citet{hansenHowAccessibilityShapes1959}, which can take the following
multimodal form: \(S_i^m = \sum_j O_j f^m(c_{ij}^m)\) where \(m\) is a
set of modes which have mode-specific travel costs \((c_{ij}^m)\) and
travel impedance functions \(f^m(\cdot)\).

The Hansen-type measure does not consider competition between modes nor
is it constrained. As an example, the work of
\citet{tahmasbiMultimodalAccessibilitybasedEquity2019} uses the
Hansen-type measure to measure the potential interaction with retail
locations using walking, public transit, and car modes \(m\). \(S_i^m\)
is the sum of retail locations \(j\) that can potentially be interacted
with under the travel impedance as calculated for each \(i\) and \(m\).
In other words, each \(i\) has three \(S_i\) values, one per \(m\). In
this work, they demonstrate that the car mode has the highest
\(S_i^{m=car}\) values in the majority of \(i\), i.e., populations using
a car can potentially interact with the most retail opportunities than
populations using other modes. However, the higher \(S_i^{m=car}\)
values are not a result of lower \(S_i^{m}\) values for other modes: it
is not assumed that car-using populations potentially accessing more
opportunities take away potential opportunities for other populations
within the measure. Put another way, this measure does not consider
competition. This measure is also not constrained: there is no global
maximum for \(S_i\) or \(S_i^m\) values, they are presented as a
population normalized accessibility index. This makes the interpretation
of the `potentially interacted opportunities' relative to the region,
making comparisons of the results across different regions challenging.

However, opportunities in a region can be considered finite. There are
only so many school-seats, hospital capacity, food stores, jobs, etc.,
in a region and if one person interacts with an opportunity at a given
time, it is taken. As such, if one person is advantaged and has the
ability to reach more opportunities through a lower travel-cost mode,
than they have more opportunities to potentially interact with more
opportunities than other people. From the other perspective, their are
fewer opportunities left to be potentially interacted with for
populations using higher travel-cost modes. In this way, populations
using modes with a higher travel impedance are at a higher access
disadvantage than populations using lower travel impedance modes. This
recognition is the motivation behind integrating \emph{competition} for
opportunities within multimodal accessibility measures. Arguably one of
the most popular competitive location-based accessibility measures is
the two-step floating catchment area (2SFCA) approach popularized by
\citet{luoMeasuresSpatialAccessibility2003} who simplified the approach
proposed by \citet{shenLocationCharacteristicsInnercity1998} (with
similar considerations for competition in
\citet{weibullAxiomaticApproachMeasurement1976} and
\citet{josephMeasuringPotentialPhysical1982}).

The Shen-type accessibility measure's formulation is:
\(a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}\) where \(D_j^m\)
is the potential demand for opportunities equal to travel impedance
weighted population \(\sum_i P_i^m f^m(c_{ij}^m)\). In this way, the
Shen-type measure can be understood as a ratio of the potential
opportunity supply over the potential demand for opportunities. The
measure considers competition, but it is \emph{non-constrained}. A score
of competitive potential accessibility associated is associated with
each location \(i\) for each mode \(m\), but there are no global
maximums. In other words, it is difficult to interpret the meaning of
differences in Shen-type accessibility scores between modes.

To illustrate, \citet{taoInvestigatingImpactsPublic2020a} calculates
\(a_i^m\) to jobs for different income-group populations in Shenzhen
(China) using \(m = \text{public transit}\) and \(m={car}\). They
demonstrate that \(i\)s with low-income populations have lower \(a_i^m\)
than \(i\)s with higher-income populations. Further, they demonstrate
that \(a_i^{m=\text{public transit}}\) is lower than
\(a_i^{m=\text{car}}\) at many \(i\)s, arguing that this may put \(i\)s
with lower-income populations in a further disadvantage. \(a_i\) and/or
\(a_i^m\) are used to compare relative spatial differences in overall
competitive accessibility and modal competitive accessibility, but
because there is no global maximum, making it is difficult to interpret
the significance between differences in \(a_i^{m}\) values. Questions
such as: What is the impact that competition has on the difference in
\(a_i^m\) values? How does impact vary spatially? And what is the
interpretation of this difference? are left unanswered.

Spatial availability improves on previous multi-modal accessibility
approaches as it considers \emph{competition} in the potential
interaction with opportunities in a \emph{constrained} framework (e.g.,
finite opportunities). This is done by considering: 1) competition
between mass effect (e.g., the advantage of sub-populations residing in
relatively low population-density and high opportunity-proximate areas)
and 2) competition between travel impedance (e.g., sub-populations with
relatively low travel-impedance) through a proportional allocation
mechanism. The following sub-section demonstrates how spatial
availability compares to the two other measures and the advantages in
its interpretation through a synthetic example.

\hypertarget{a-synthetic-example-calculating-accessibility-for-multiple-modes-considering-finite-opportunities}{%
\subsection{A synthetic example: calculating accessibility for multiple
modes considering finite
opportunities}\label{a-synthetic-example-calculating-accessibility-for-multiple-modes-considering-finite-opportunities}}

In brief, we define the \emph{spatial availability} at \(i\) ( \(V_{i}\)
) as the proportion of all opportunities in the region \(O\) that are
allocated to location \(i\) from all opportunity destinations \(j\).
\(V_{i}\) is a value of how many opportunities are available to each
location \(i\) out of all the opportunities in the region (e.g., a
neighbourhood has 1,000 spatially available jobs out of 100,000 jobs in
the total region). The general formulation of spatial availability
\(V_{i}\) is shown in Equation (\ref{eq:spatial-availability-general}):

\begin{equation}
\label{eq:spatial-availability-general}
V_i = \sum_{j=1}^J O_jF^t_{ij}
\end{equation}

\noindent where:

\begin{itemize}
\tightlist
\item
  \(F^t_{ij}\) is a balancing factor that depends on the demand for
  opportunities \(O_j\) and cost of movement in the system
  \(f(c_{ij})\).
\item
  \(V_i\) is the number of spatially available opportunities at \(i\);
  the sum of \(V_{i}\) is equivalent to the total sum of opportunities
  in the region (i.e., \(\sum_j O_j = \sum_i V_i\))
\end{itemize}

Spatial availability measure is introduced in
\citet{soukhovIntroducingSpatialAvailability2023}. The unique feature in
the measure is the balancing factor \(F^t_{ij}\), a proportional
allocation mechanisms, which ensures that the \(V_i\) calculated for
each \(i\) sums to the total number of opportunities. Through
\(F^t_{ij}\), spatial availability is a \emph{competitive} and
\emph{constrained} accessibility measure that handles the number of
opportunities in the region in a finite way. \(F^t_{ij}\) consists of
two components: a population-based balancing factor
\(F^p_{i} = \frac{P_i}{\sum_i P_i}\) and an impedance-based balancing
factor \(F^c_{ij} = \frac{F^c_{ij}}{\sum_j F^c_{ij}}\) that,
respectively, allocate opportunities to \(i\) in proportion to the size
of the population at \(i\) (the mass effect) and the cost of reaching
opportunities at \(j\) (the impedance effect). \(F^p_{i}\) and \$
F\^{}c\_\{ij\}\$ are calculated for each \(i\) such that they both equal
1 when summed across all \(i\) in the region (e.g.,
\(\sum_i F^p_{i} = 1\) and \(\sum_i F^c_{ij} = 1\)). These balancing
factors are combined multiplicatively to yield \(F^t_{ij}\) which
ensures that a proportion of the opportunities \(O_j\) are allocated to
each \(i\) accordingly. In other words, assuming a finite number of
opportunities in the region, \(F^t_{ij}\) proportionally allocates
\(O_j\) to each \(i\) such that the resulting \(V_i\) value represents
the number of opportunities \emph{spatially available} to the population
at \(i\). This value can be seen to represent spatial availability as it
is a proportion of the opportunities in the region (i.e.,
\(\sum_j O_j = \sum_i V_i\)).

The focus of this paper is to extend \(V_i\) for multimodal
applications. To do so, the balancing factors are reformulated to yield
a proportional value for the set of modes \(m\) used by populations at
each \(i\). As these factors are proportional, \(F^{pm}_{i}\) and
\(F^{cm}_{ij}\) can be summed up across each \(m\) at each \(i\) and
across all \(i\) to equal to 1. They are also similarly combined
multiplicatively to obtain their joint effect, represented as the
combined balancing factor \(F^{tm}_{ij}\) detailed in Equation
(\ref{eq:multimodal-balancing-factors}).

\begin{equation}
\label{eq:multimodal-balancing-factors}
F^{tm}_{ij} = \frac{F^{pm}_{i} \cdot F^{cm}_{ij}}{\sum_{m=1}^M \sum_{i=1}^N F^{pm}_{i} \cdot F^{cm}_{ij}}
\end{equation}

\noindent Where: - The population balancing factor for each \(m\) at
each \(i\) is \(F^{pm}_{i} = \frac{P_{i}^m}{\sum_{m}\sum_{i} P_{i}^m}\)
- The cost of travel balancing factor for each \(m\) at \(i\) is
\(F_{ij}^{cm} = \frac{f(c_{ij}^m)}{\sum_{m} \sum_{i} f(c_{ij}^m)}\)

Implementing \(F^{tm}_{ij}\), the following Equation
(\ref{eq:spatial-availability-multimodal}) demonstrates the multimodal
configuration of spatial availability \(V_i^m\):

\begin{equation}
\label{eq:spatial-availability-multimodal}
V^m_{i} = \sum_{j=1}^J O_j\ F^{tm}_{ij}
\end{equation}

\noindent Where:

\begin{itemize}
\tightlist
\item
  \(m\) is a set of modes used by populations in the region
  \(m = 1, \cdots, M\).
\item
  \(F^{tm}_{ij}\) is a balancing factor \(F^t_{ij}\) for each \(m\) at
  each \(i\).
\item
  \(V^m_{i}\) is the spatial availability \(V_{i}\) for mode \(m\) at
  each \(i\); the sum of \(V^m_{i}\) for all \(m\) at each \(i\) is
  equivalent to the total sum of opportunities in the region (i.e.,
  \(\sum_j O_j = \sum_i V_i = \sum_{m} \sum_{i} V^m_{i}\))
\end{itemize}

Consider the following synthetic example: Figure \ref{fig:Fig1} depicts
a region with population and jobs at three population centers (\(A\),
\(B\), \(C\)) and three employment centers (\(1\), \(2\), \(3\)). The
population at each population center is divided into two sub-groups, one
using a faster mode \(z\) and another using a slower mode \(x\), to
travel to employment centers. Population center \(A\) is Suburban: it is
closest to its own relatively large employment center at \(A\), close to
the Urban area's equally large employment center \(B\), and has a
population that is smaller than the Urban center \(2\) and larger than
the Satellite center \(3\). The Urban population center has the largest
number of population using slower mode \(x\), followed by the Suburban
then Satellite area. This synthetic example was inspired by the
single-mode example used in
\citet{shenLocationCharacteristicsInnercity1998} and reconfigured in
\citet{soukhovIntroducingSpatialAvailability2023}.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{images/Fig1}

}

\caption{\label{fig:Fig1} Modified synthetic example from Shen (1998) with locations of employment centers (in orange), population centers (in blue), number of jobs and population, and travel times for two modes (slower x mode and faster z mode).}\label{fig:synthetic-example-plot}
\end{figure}

From the perspective of access to a \emph{finite} amount of
opportunities in the region (\(210,000\) jobs), the sub-population that
is most proximate to jobs, furthest from densely populated centers, and
is using the lowest travel cost mode \(z\) can potentially access the
most job opportunities. This appears to be the sub-population at
Suburban center \(A\) using the faster mode \(z\). From the other
perspective, sub-populations located further away from jobs, close to
dense populations, and using high cost travel mode \(x\) are at a job
opportunity access \emph{disadvantage} relative to the other
sub-populations. This could be the sub-populations using the slowest
mode \(x\) at either Urban \(B\) or Satellite \(C\) area. From the
perspective of inequities, the competition for opportunities between
different mode-using populations (i.e., how well the land-use and
transport system may serve some and not others), matters.

\global\setlength{\Oldarrayrulewidth}{\arrayrulewidth}

\global\setlength{\Oldtabcolsep}{\tabcolsep}

\setlength{\tabcolsep}{0pt}

\renewcommand*{\arraystretch}{1.5}

\providecommand{\ascline}[3]{\noalign{\global\arrayrulewidth #1}\arrayrulecolor[HTML]{#2}\cline{#3}}

\begin{longtable}[c]{|p{0.88in}|p{0.41in}|p{1.05in}|p{1.05in}|p{0.58in}|p{1.05in}|p{0.58in}}

\caption{Summary\ description\ of\ the\ synthetic\ example:\ accessibility\ values\ at\ each\ origin\ per\ mode\ m\ at\ each\ origin\ i\ and\ aggregated\ between\ modes\ for\ each\ i.}\\

\ascline{1.5pt}{666666}{1-7}

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{i}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{m}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{S}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}} \\

\ascline{1.5pt}{666666}{1-7}\endfirsthead \caption[]{Summary\ description\ of\ the\ synthetic\ example:\ accessibility\ values\ at\ each\ origin\ per\ mode\ m\ at\ each\ origin\ i\ and\ aggregated\ between\ modes\ for\ each\ i.}\\

\ascline{1.5pt}{666666}{1-7}

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{i}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{m}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{S}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}} \\

\ascline{1.5pt}{666666}{1-7}\endhead

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{x}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{18,959.86}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{27,292.18}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.95}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{65,872.91}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.88in}{\raggedright \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{A}}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{z}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{46,913.04}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{44,999.80}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.56}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{65,872.91}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.58in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.32}}}}} \\

\ascline{1pt}{666666}{1-7}

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{x}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{30,863.43}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{27,292.18}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.62}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{134,255.38}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.88in}{\raggedright \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{B}}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{z}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{103,391.95}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{44,999.80}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.03}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{134,255.38}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.58in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.90}}}}} \\

\ascline{1pt}{666666}{1-7}

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{x}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{2,034.49}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{2,240.38}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.68}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{9,871.71}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{} \\

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.88in}{\raggedright \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{C}}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{z}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{7,837.22}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{3,745.89}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.12}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{9,871.71}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.58in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.99}}}}} \\

\ascline{1pt}{666666}{1-7}

\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{TOTALS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{210,000.00}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{150,570.22}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{N/A}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{210,000.00}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{N/A}}} \\

\ascline{1.5pt}{666666}{1-7}

\end{longtable}

\arrayrulecolor[HTML]{000000}

\global\setlength{\arrayrulewidth}{\Oldarrayrulewidth}

\global\setlength{\tabcolsep}{\Oldtabcolsep}

\renewcommand*{\arraystretch}{1}

The calculated \(S_i^m\), \(a_i^m\) and \(V_i^m\) accessibility values
for each \(i\) and \(m\) are shown in the middle three columns and are
aggregated for each \(i\) in the final two columns in Table
(\ref{tab:Tab1}). We use a negative exponential impedance function
\(f(c_{ij}) = \exp(-\beta\cdot c_{ij})\) with \(\beta=0.1\) for both
\(x\) and \(z\) modes for all accessibility measures calculations.

The Hansen-type measure \(S_i^m\) is presented for each origin and mode
in third column of Table (\ref{tab:Tab1}). For all \(i\), the
sub-population using the faster mode \(z\) has higher \(S_i^m\) values
than the \(x\) using sub-populations. Additionally, \(S_i^m\) is equal
for both mode using populations in population centers \(A\) and \(B\).
This is the case because \(S_i^m\) does not consider \emph{competition},
it only relies on reflecting the count of opportunities that may be
potentially interated with as a product of \(f^m(c_{ij}^m)\). Recall,
populations in \(A\) and \(B\) have the same travel impedance to
employment centers \(1\), \(2\) and \(3\) (either 15, 30, or 100 minutes
for mode \(x\) or 10, 25, or 80 minutes for mode \(z\)). As such, these
the calculated \(S_i^m\) values are the same for both \(A\) and \(B\).
Furthermore, the total sum of \(S_i^m\) in the region is equal to
150,570.2. This value is difficult to interpret: it represents the
weighted sum of opportunities that may be interacted with within the
region based on travel impedance. It cannot be interpreted as a global
maximum or any sort of benchmark as the measure is
\emph{non-constrained}. To connect this example to literature, this
method of modal accessibility calculation is used in the work of
\citet{tahmasbiMultimodalAccessibilitybasedEquity2019}; they compare
differences in \(S_i^m\) values between modes in a relative and
comparative sense, but make no further interpretation of the \(S_i^m\)
values.

In the fourth and sixth column in Table (\ref{tab:Tab1}) the Shen-type
measure is calculated: first for both origin and mode \(a_i^m\) as well
as aggregated by the weighted mean mode-population (
\(\sum_m \frac{P_i^m}{P_i}*a_i^m\) ) to represent a value for each
origin \(a_i\). Unlike \(S_i^m\), this measure considers
\emph{competition}. For instance, the populations using the same modes
in \(A\) and \(B\) centers do not have the same \(a_i^m\) values. In
fact, the Suburban \(A\) has the highest values since this center has
the smallest travel impedance to opportunities (lower than at \(C\)) and
has the lowest relatively proximate to populations (lower than at
\(B\)).

However, the Shen-type measure is \emph{non-constrained}: the total sum
of \(a_i^m\) or \(a_i\) is practically meaningless since it represents a
sum of ratios. For instance, the sub-population using the fast mode
\(z\) at \(A\) has a value of 1.56 potential jobs per potential
job-seeking population compared to 0.95 for the slow mode \(x\)
population. What is the significance of these values? The difference
between these modes is equal to 0.62, but 0.62 of what? How many more
job opportunities are \(z\) users interacting with than \(x\) users?
When \(a_i^m\) is aggregated to \(a_i\) as shown in the sixth column,
the values face similar interpretability issues.

The Shen-type measure is implemented in the previously discussed work of
\citet{taoInvestigatingImpactsPublic2020a} to calculate modal \(a_i^m\)
values and the aggregated \(a_i\) is implemented in the work of
\citet{carpentieriMultimodalAccessibilityPrimary2020}. However, similar
to the Hansen-type measure, these works discuss relative and spatially
comparative differences in values, they do not make further
interpretation of the \(a_i^m\) or \(a_i\) themselves. This may be
because the Shen-type measure is \emph{non-constrained}, this is no
benchmark or global maximum to which comparisons can be drawn from.
Being unable to do so makes the interpretation of competition between
modes challenging to tease out.

By contrast, spatial availability \(V_i\) considers competition and is
constrained such that the total sum of values is equal to the total
number of opportunities in the region (i.e., \(210,000\) jobs). Seen in
fifth column of Table (\ref{tab:Tab1}), \(V_i^m\) for the same
mode-using populations in \(A\) and \(B\) are not the same (as this
measure considers competition). In fact, at \(A\) for instance, the
sub-population using faster mode \(z\) captures 27,953.18 more spatially
available jobs (of the \(210,000\) jobs in the region) than the
sub-population using mode \(x\). The numerical difference have a
practical interpretation.

Furthermore, \(V_i^m\) values for an \(i\) can be aggregated across
\(m\) and compared across \(i\) ( \(V_i = \sum_m{\sum_i{V_i^m}}\) ) as a
result of the proportional allocation mechanism. This aggregation,
\(V_i\), is shown in the seventh column in Table (\ref{tab:Tab1}). Again
looking at center \(A\), \(A\) is allocated 65,872.91 spatially
available opportunities for both modes. 71\% of this spatial
availability allocated to \(A\) is assigned to the \(z\)-using
population despite representing 66\% of the population at \(A\).

Spatial availability can be further aggregated to better interpret
competition between modes. Across the entire region, 137,500 use the
faster mode \(z\) (65\% of the region population). However, \(z\)-using
populations accounts for 75\% of the region's total spatial availability
- the rest is allocated to the population using \(x\) (of which
represents 35\% of the population). Notably, the \(x\) population
captures 10\% less spatial availability to opportunities than its
population proportion.

Since spatial availability is constrained and has an interpretable
meaning of a proportion of the total opportunities in the region, the
values have a new significance. Inequality in spatial availability
values can be explored through a variety of approaches. For instance,
consider travel times. The \(z\) population using the faster mode \(z\)
accounts for 69\% of the potential travel time traveled in the region:
6\% less travel time than the proportion of spatial available
opportunities that the \(z\) population has access to. Meaning \(z\)
population travels less potential minutes overall and has more spatial
availability of opportunities than the population using the slower mode
\(x\).

Alternatively, these inequities in spatial availability between
mode-using populations can be explored through proportional benchmarks.
A spatial availability per capita as presented in Equation
(\ref{eq:SA-per-capita}) is defined at each population center, mode,
and/or regionally:

\begin{equation}
\label{eq:SA-per-capita}
v_{im} = \frac{V_{im}}{P_{im}}
\end{equation}

The spatial availability per capita values \(v_i^m\) for \(A\), \(B\),
and \(C\) for the slower mode \(x\) is 0.95, 0.62 and 0.68 respectively.
The benchmarks for \(x\) mode are evidently lower than the averages at:
1.32, 0.9 and 0.99 respectively. Especially for \(B\) and \(C\).

The average for the faster mode \(z\) with values of: 1.56, 1.03 and
1.12 respectively. Once again, we see the competitive advantage on a per
capita bases that populations have to opportunities when commuting using
the faster mode.

In what follows, we further explore competition between multimodal
accessibility and how competition between different modal-users is
captured by spatial availability through an empirical example.

\hypertarget{empirical-multimodal-spatial-availability}{%
\section{Empirical multimodal spatial
availability}\label{empirical-multimodal-spatial-availability}}

\hypertarget{data-inputs-madrids-lez-and-travel-survey}{%
\subsection{Data inputs: Madrid's LEZ and travel
survey}\label{data-inputs-madrids-lez-and-travel-survey}}

Low emission zones (LEZ) have been implemented as a climate change
policy intervention to reduce GHG emissions, improve air quality, and
support sustainable mobility in many countries. Though rules vary
depending on legal aspects and cultural norms, LEZ aim to deter or
reduce traffic in specially designated zones under the penalty of fines
and/or seizure of vehicle. In practice, this limits the volume of
traffic by excluding vehicles by license plate, fuel type, or by
introducing tolls.

Spain is one of a few countries which have active LEZ and are interested
in expanding their implementation. As specified in their recent climate
change plan \emph{Plan Nacional Integrado de Energía y Clima 2021-2030}
\citep{espanaPlanNacionalIntegrado2020} and the \emph{Plan Nacional de
Control de la Contaminación Atmosférica}
\citep{espanaResolucion10Enero2020}, more LEZ will soon be implemented
throughout the country. Specifically, the national Spanish law 7/2021
(``Ley de Cambio Climático y Transición Energética'') will require all
municipalities to implement LEZ by 2023 if they meet at least one of the
following requirements: (i) municipalities \textgreater50,000 inhab.;
(ii) islands; and (iii) municipalities \textgreater{} 20,000 inhab. when
air quality exceeds limits specified in ``RD 102/2011 de Mejora de
Calidad del Aire'' \citep{barcelonaGUIATECNICAPARA2021}.

In 2017, a LEZ was implemented in the capital city of Madrid following
the goals set out in the national agenda: to fight climate change, cut
nitrogen dioxide levels for the benefit of people's health, and
prioritize people's movement in the city. In geographic scope, the 2017
boundaries of the LEZ were relatively small (covering 4.72 km\^{}\{2\})
and within the center (i.e., LEZ Centro). They were expanded in 2023 to
just inside of the M-30, a highway in proximity to the city center
(i.e., LEZ M-30) and the city has plans to further spatially expand the
LEZ in the future. Within the 2017 LEZ Centro implementation, all cars,
motorcycles and freight with environmental label A or B (higher
polluting classification, associated with older make and model of fossil
fuel internal combustion engine vehicles), are banned from driving into
the area unless they are used by residents or meet other exemptions
\citep{tarrinoortizAnalyzingImpactLow2022}. This restriction impacts
approximately half of all car that made trips into LEZ Centro.

From the perspective of restriction for passenger transport, LEZ are a
policy of \emph{geographic discrimination}. LEZ actively change how
people access opportunities by making the travel impedance more costly
for car-mode users. If seeing opportunities as finite, this
discrimination allows populations to access opportunities by other modes
more readily than before. In this way, the policy changes the multimodal
competitive accessibility landscape of the city. Though the cost of
travel for modes do dynamically change as a result of the LEZ
implementation (i.e., potentially less car congestion, potentially more
transit congestion), the focus of this empirical example is to
demonstrate the spatial availability, by mode, of the city of Madrid
after the 2017 implementation of the LEZ Centro.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{images/i_jobs_zn208_plot}

}

\caption{\label{fig:Fig2} Jobs $O_j$ taken by people living and working in Madrid as reported by the 2018 travel survey.}\label{fig:jobs-plot}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{images/im_populations_zn208_plot}

}

\caption{\label{fig:Fig3} Population living and working in Madrid, by four summarized modal categories, $P^m_i$ as reported by the 2018 travel survey.}\label{fig:pop-plot}
\end{figure}

The 2018 Community of Madrid travel survey is the source of data for the
empirical example: it is a representative survey that reflects a
snap-shot of the travel patterns for one typical day of the working week
(e.g., n=222,744 trips with representative population elevation
factors). In this paper, a sample of the travel survey is used, namely
the residential home origin to work destination trips of all modes and
those that originate and end in Madrid. These totals are displayed in
Figure \ref{fig:Fig2} and Figure \ref{fig:Fig3}. Both figures are
displayed at the level of traffic analysis zones (\(i\) and \(j\)) that
correspond to the 2018 travel survey. The red boundary represents the
LEZ Centro in effect in 2017 and thus those travel patterns of
car-restriction reflected in the 2018 travel survey data. The cyan
boundary represents the LEZ that will be within the boundaries of the
M-30 highway in 2023 and is present in the plots as a spatial reference
for areas in proximity to the LEZ Centro.

The total sum of jobs \(O_j\) that are held are shown in Figure
\ref{fig:Fig2} and the populations that go to a work destination by four
modal categories \(P^m_i\), is reflected in Figure \ref{fig:Fig3}. The
modal categories represented in Figure \ref{fig:Fig3} were summarized
for the following trip mode types: - Car/motor: all cars and operating
modes (e.g., cab, private driver, company, rental care, main driver,
passenger, etc.) and all public, private or company motorcycle/mopeds. -
Transit: all bus, trams, and trains - Bike: all bicycle trips (e.g.,
private, public, or company bike trips) and ``other'' types of
micromobility options - Walk: walking or by foot

From Figure \ref{fig:Fig2}, it can be seen that the largest
concentration of jobs are within, near, and to the north of the LEZ
Centro.. The population that is accessing those jobs by mode (Figure
\ref{fig:Fig3}), appear spatially distinct. Car and transit trips
represent 37\% and 47\% of the modal share respectively. The population
that travels using transit is more spatially distributed than those
using cars - particularly near and within LEZ Centro. This distribution
could be a result of a variety of factors including: transit coverage
and service within with city, effective car infrastructure outside of
the M-30, and/or the impact of the Central LEZ itself.

From Figure \ref{fig:Fig2}, it can also be seen that biking and walking
trips are less common than motorized trips at 1\% and 15\% respectively.
The distribution of walking and biking trips appear to be similar to
that of transit trips. This is to be expected as active transport and a
higher diversity of land-use spatially occurs with transit
infrastructure.

The travel time for each trip is provided within the 2018 survey. These
travel times, per modal category, are used to calibrate mode specific
travel impedance functions \(f^m(c_{ij}^m)\). To illustrate the modal
differences in travel lengths, summary descriptive per mode are detailed
as follows:

\begin{itemize}
\tightlist
\item
  Car/motor: 36 min (Min:0min, Q2: 15 min, Q3: 55 min, Max: 120 min)
\item
  Transit: 55 min (Min:1 min, Q2: 30 min, Q3: 80 min, Max: 120 min)
\item
  Bike: 34 min (Min:5 min, Q2: 15 min, Q3: 40 min, Max: 115 min)
\item
  Walk: 27 min (Min:1 min, Q2: 10 min, Q3: 45 min, Max: 119 min)
\end{itemize}

To calculate the mode specific travel impedance functions
\(f^m(c_{ij}^m)\) from the travel times, a concept known as the trip
length distribution (TLD) is used. A TLD represents the proportion of
trips that are taken at a specific travel cost such as travel time. This
distribution is then used to derive impedance functions as done in
previous accessibility research \citep[e.g., works
of][\citet{horbachov_theoretical_2018}, and
\citet{batista_estimation_2019}]{lopez_2017_spatial}. Maximum likelihood
estimation and the Nelder-Mead method for direct optimization available
within the R \{fitdistrplus\} package \citep{fitdistrplus_2015} is used.
As shown as shown in Figure \ref{fig:Fig4}, based on goodness-of-fit
criteria and associated diagnostics, the gamma and log-normal
probability density function (line curves) are selected as best fitting
curves for the motorized and non-motorized modes respectively. The
selection of functional form aligns with examples used in the literature
(e.g., \citet{reggianiAccessibilityImpedanceForms2011}). Overall, the
plots in Figure \ref{fig:Fig4} display the probability of travel given a
trip travel time, based on actual travel behaviour from the 2018 survey.
These `probability of travel' at each travel time for each mode are
realized observations reflect the land-use, the transport system, and
the population travel preferences/behaviour in Madrid.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{images/tlds_curves_m_plot}

}

\caption{\label{fig:Fig4} Fitted impedance function curve (line) against empirical TLD (bars) per mode.}\label{fig:tlds-curves-m-plot}
\end{figure}

\hypertarget{results-spatial-availability-v_im-and-v_i}{%
\subsection{\texorpdfstring{Results: spatial availability \(V_i^m\) and
\(V_i\)}{Results: spatial availability V\_i\^{}m and V\_i}}\label{results-spatial-availability-v_im-and-v_i}}

Using the data inputs outlined, spatial availability of jobs are
calculated for each of the four modal categories at the level of traffic
analysis zones in Madrid, \(V_i^m\). The spatial distribution of the
resulting calculations are demonstrated in this section.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{images/SA_im_V_zn208_plot}

}

\caption{\label{fig:Fig5} Spatial availability of job opportunities per origin and mode $V_i^m$ in Madrid as reported by the home-to-work origin destination flows from the 2018 travel survey. 2017 central LEZ is shown in blue. 2023 expanded LEZ boundaries shown in red.}\label{fig:SA-m-plot}
\end{figure}

Figure \ref{fig:Fig5} displays the spatial availability values for the
four modal categories at the level of the spatial units used in the 2018
travel survey. These values represent a proportion of the total number
of the 847,574 jobs in the region that are \emph{spatially available} to
the population located at the \(i\) based on the travel impedance of the
mode (relative to the travel impedance of all modes) and the mode-using
population size at the \(i\) (relative to the population size of all
modes and \(i\)). Since population, their job locations, and associated
travel times are used to calculate \(V_i^m\), Figure \ref{fig:Fig5}
demonstrates values that represent the \emph{realized} spatial
availability of jobs in Madrid. In other words, if someone decided to
move into a neighbourhood at an \(i\), the amount of jobs that are
spatially available to them based on the population within their \(i\)
and their travel impedance for their used \(m\), both relative to the
city, is captured by the value of \(V_i^m\).

In Figure \ref{fig:Fig5}, the difference in the magnitudes of \(V_i^m\)
values between \(m\) can be observed. The majority of \(V_i^m\) is
allocated to the populations using motorized modes. This is to be
expected as commuting using motorized modes represents 84\% of the
population (37\% (car/motor) and 47\% (transit)). However, these modal
options capture 95\% of the total spatial availability in Madrid. In
particular, the car/motor using population is allocated
disproportionately more \(V_i^m\) than its modal population (37\% of the
population vs.~48\% of the \(V_i^m\)) compare to the transit using
population and its relatively proportional \(V_i^m\) value (15\% of the
population vs.~47\% \(V_i^m\)).

How does the \(V_i^m\) advantage allocated to car-using population
arise? From the perceptive of finite opportunities, \(V_i^m\) is
allocated to car-using populations from less competitive modal
populations. How competitive one mode is compared to other modes varies
spatially, but overall car-using populations capture more opportunities
per car-using population than other modal populations. Namely, though
walking and cycling populations represent 14.74\% and 1.16\%
respectively, \(V_i^m={walk}\) and \(V_i^m={bike}\) is 4.43\% and 0.23\%
in the region respectively. These modes are less competitive, especially
compared to the car/motor mode, as a result of: 1) their lower travel
impedance values at longer travel times (see Figure \ref{fig:Fig4} at
travel times beyond \textasciitilde30 minutes), 2) their low population
values values overall, and 3) higher populations present in origins with
high motorized mode commuting. These factors all contribute to the the
car/motor mode being most advantaged in capturing spatially available
job opportunities overall.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{images/SA_im_vv_zn208_plot}

}

\caption{\label{fig:Fig6} Spatial availability of job opportunities per capita per origin and mode $v_i^m$ in Madrid as reported by the home-to-work origin destination flows from the 2018 travel survey. 2017 central LEZ is shown in blue. 2023 expanded LEZ boundaries shown in red.}\label{fig:SA-per-capita-m-plot}
\end{figure}

Furthermore, there are spatial differences in the competitive advantage
of spatial availability between modes. Figure \ref{fig:Fig6} visualizes
\(v_i^m\), the spatial availability divided by the mode population.
\(v_i^m\) values above 1 are represented in increasing red shades,
values below 1 are represented in increasingly green shades, and values
equal to 1 are white. These plots illustrates the discussion of the
disproportionately high over representation of spatial availability
relative to the mode-using population in many of the origins for the car
plot (bottom right, denoted with green \(v_i^m\) values above 1). These
plots also visualize areas that are awarded disproportionately low
spatial availability, represented in shades of red. Interestingly,
spatial availability for the car mode within the LEZ Centro is below 1
(red). For all other modes, the area with the LEZ is relatively higher
than the modal averages or even green in the case transit-using
populations. This difference in spatial availability can be seen as a
direct result of the LEZ Centro - the observed reduction of
opportunities within the LEZ Centro boundaries being accessed by
car-using populations allows lesser competitive modes to interact with
these opportunities.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{images/v_im_per_pop_plot}

}

\caption{\label{fig:Fig7} Spatial availability of job opportunities per capita per origin and mode $v_i^m$ in Madrid as reported by the home-to-work origin destination flows from the 2018 travel survey represented by sorted total population per origin.}\label{fig:SA-per-capita-m-linear-plot}
\end{figure}

Furthermore, Figure \ref{fig:Fig6} makes it evident that transit-using
population's spatial availability to jobs is relatively balanced (i.e.,
many zones are white). Additionally, values for non-motorized modes are
higher in origins that have higher transit accessibility. Transit
accessibility and non-motorized modes do not appear to be in direct
competition as a result of different travel impedance weighting . It is
not visually clear in the plots but as demonstrated in a line plot of
\(v_i^m\) sorted by total population in Figure \ref{fig:Fig7}, it can be
observed that transit is only close to or above 1 when car accessibility
is relatively low. From Figure \ref{fig:Fig7} we can also observe that
overall, motorized modes capture more spatial availability per capita
than non-motorized modes.

\hypertarget{discussion-and-conclusions}{%
\subsection{Discussion and
conclusions}\label{discussion-and-conclusions}}

Location-based accessibility measures like the Hansen-type measure,
Shen-type measure, and spatial availability all have a commonality -
they are a weighted sum of opportunities assigned to each spatial unit
in a region. In this way, they can be interpreted as a score that
represents how many opportunities can be potentially interacted with by
the population at each spatial unit. How the weight and sum of the
potentially interacted opportunities is what defines the type of
accessibility measure. Within this paper, the location-based
accessibility measures known as spatial availability, a singly-
\emph{constrained} and \emph{competitive} measure, is extended for the
case of capturing multimodal accessibility to opportunities. A synthetic
example and then an empirical case of LEZ in Madrid are detailed to
demonstrate the multimodal extension of the spatial availability
measure.

The spatial availability measure is capable of capturing a new
interpretation of multimodal competition that previous accessibility
measures have not yet done. We can hypothesis that populations using
modes with lower travel impedance, when competing for a finite set of
opportunities, will capture more opportunities. However, with spatial
availability, the number of spatially available opportunities that are
captured (of the total opportunities in the region) by each mode can be
individually calculated. From there, the difference between how many
spatially available opportunities one mode captures versus another can
be investigated.

The flexibility and need for an accessibility measure such as spatial
availability is pertinent in policy scenario evaluation. As showcased in
the empirical example of the LEZ in Madrid, competition for job
opportunity availability highly varies spatially and between modes. The
car and transit modes have the highest spatial availability, with the
car mode having highest availability with exception to the areas with
LEZ Centro. This finding reflects real conditions: since car travel has
been highly restricted within the LEZ Centro, much fewer car-using
population and much more people are entering using other modes relative
to the areas surrounding the LEZ Centro. This difference in car-using
population within and immediately outside the LEZ Centro increases the
competitiveness of the transit-using population (the second most
competitive mode) as well as the non-motorized modes.

Currently, conventional \emph{non-constrained} accessibility measures
are difficult for planners to operationalize for a variety of reasons.
They have also been criticized for being difficult to compute and
difficult to interpret as they are a ratio of supply to demand
\citep{levinsonTransportAccessManual2020}. With spatial availability,
the magnitude of opportunities that are available as a proportion of all
the opportunities in the region is equal to \(V_i\). As a result of its
proportional allocation mechanism it can be easily extended into
multimodal applications, pertinent to model policy scenerios in our
cities that are becoming increasingly multimodal. This flexibility and
interpretation of the spatial availability measure allows for
manipulation of \(V_i^m\) values to investigate differences of
availability between neighbourhoods, modes, and regions, generate per
capita benchmarks, and/or generate average values per population-group.

From a spatial equity perspective, spatial availability measure can
provide researchers, policy makers, and citizens a newfound
interpretation of accessibility measures. With a plot of spatial
availability values, one can begin asking, how much is enough and what
level is too much. These interpretations were difficult to be made with
accessibility measures in the past.

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

This research was funded by the Canada Graduate Scholarship - Doctoral
Program (CGS D) by the Social Sciences and Humanities Research Council
(SSHRC).

\#AUTHOR CONTRIBUTIONS

The authors confirm contribution to the paper as follows: study
conception and design: AS, JSL, AP.; data collection: AS, JSL, AP.;
analysis and interpretation of results: AS, JSL, AP.; draft manuscript
preparation: AS, JSL, AP. All authors reviewed the results and approved
the final version of the manuscript.

--\textgreater{}

--\textgreater{}

\newpage
\bibliography{mybibfile.bib}

\end{document}

### 2023-07-30 TRB-manuscript/data-prep/01-Travel-time-calculation.Rmd
TRB draft - ready for AP edits

--- 2022-04-08 01-Travel-time-calculation.Rmd
@@
 4) Transit bus
+
+/data-prep/data-raw/intermed/ttresults/output_OD_bike.Rdata
+/data-prep/data-raw/intermed/ttresults/output_OD_car.Rdata
+/data-prep/data-raw/intermed/ttresults/output_OD_transit.Rdata
+/data-prep/data-raw/intermed/ttresults/output_OD_walk.Rdata
 
@@
 ```
+
+Resave but with ZT1259 codes:
+```{r}
+load("data-prep/data/zn1259.RData") 
+zn1259 <- zn %>% dplyr::select(c("FID", "CD_ZT12", "ZT1259"))
+remove(zn)
+
+#associate the ZT11259 ids to the calculated travel times. 
+list_ZT1259_FID <- zn1259 %>% dplyr::select(c("ZT1259", "FID")) %>% st_drop_geometry()
+
+calc_bikett <- list_ZT1259_FID %>% 
+  merge(output_OD_bike, by.x = "FID", by.y="fromId") %>% 
+  rename("VORIZT1259" = "ZT1259") %>% dplyr::select(-FID) %>%
+  merge(list_ZT1259_FID, by.y = "FID", by.x="toId") %>% 
+  rename("VDESZT1259" = "ZT1259") %>% transmute(VORIZT1259, VDESZT1259, travel_time)
+rm("output_OD_bike")
+
+calc_walktt <- list_ZT1259_FID %>% 
+  merge(output_OD_walk, by.x = "FID", by.y="fromId") %>% 
+  rename("VORIZT1259" = "ZT1259") %>% dplyr::select(-FID) %>%
+  merge(list_ZT1259_FID, by.y = "FID", by.x="toId") %>% 
+  rename("VDESZT1259" = "ZT1259") %>% transmute(VORIZT1259, VDESZT1259, travel_time)
+rm("output_OD_walk")
+
+calc_transittt <- list_ZT1259_FID %>% 
+  merge(output_OD_transit, by.x = "FID", by.y="fromId") %>% 
+  rename("VORIZT1259" = "ZT1259") %>% dplyr::select(-FID) %>%
+  merge(list_ZT1259_FID, by.y = "FID", by.x="toId") %>% 
+  rename("VDESZT1259" = "ZT1259") %>% transmute(VORIZT1259, VDESZT1259, travel_time)
+rm("output_OD_transit")
+
+calc_cartt <- list_ZT1259_FID %>% 
+  merge(output_OD_car, by.x = "FID", by.y="fromId") %>% 
+  rename("VORIZT1259" = "ZT1259") %>% dplyr::select(-FID) %>%
+  merge(list_ZT1259_FID, by.y = "FID", by.x="toId") %>% 
+  rename("VDESZT1259" = "ZT1259") %>% transmute(VORIZT1259, VDESZT1259, travel_time)
+rm("output_OD_car")
+
+save("calc_bikett", file = "data-raw/intermed/ttresults/calc_bikett.Rdata")
+save("calc_walktt", file = "data-raw/intermed/ttresults/calc_walktt.Rdata")
+save("calc_transittt", file = "data-raw/intermed/ttresults/calc_transittt.Rdata")
+save("calc_cartt", file = "data-raw/intermed/ttresults/calc_cartt.Rdata")
+```

### 2023-07-31 TRB-manuscript/TRB-manuscript.Rmd
The submitted TRB draft

@@
 ---
-title: Multimodal spatial availability - the case of low emission zones in Madrid
-runningheader: "Multimodal spatial availability - the case of low emission zones in Madrid"
+title: "Competition for Access to Opportunities between Modes: Multimodal Spatial Availability"
+#runningheader: "Capturing Multimodal Access to Opportunities with the Spatial Availability Measure"
 author:
@@
     affiliation: Department of Earth, Environment and Society, McMaster University, Canada
-  - name: Javier Tarriño Ortiz
+  - name: Javier Tarriño-Ortiz
     position: PhD Graduate
-    email: bob@example.com
+    email: javier.tarrino.ortiz@upm.es
     affiliation: Universidad Politécnica de Madrid, Spain
@@
     position: Associate Professor
-    email: derek@example.com
+    email: julio.soria-lara@upm.es
     affiliation: Urban and Regional Planning, Universidad Politécnica de Madrid, Spain
@@
 A dimension of these inequities arises from differences in access by mode type (e.g., commuting using a car as opposed to transit).
-However, methods implemented in current accessibility literature are lacking within the context of multi-modal analysis.
-This paper presents an extension of spatial availability, a singly-constrained competitive accessibility measure, for the context of multi-modal accessibility analysis.
-We first illustrate the features of spatial availability that lend itself to multi-modal analysis.
+However, methods implemented in current accessibility literature are lacking within the context of multimodal analysis.
+This paper presents an extension of spatial availability, a singly-constrained competitive accessibility measure, for the context of multimodal accessibility analysis.
+We first illustrate the features of spatial availability that lend itself to multimodal analysis.
 We then demonstrate its use on the case study of Low Emission Zones in Madrid (Spain) and highlight how this policy intervention changes the accessibility of populations using different modes.
-In summary, spatial availability can be used to create and interpret multi-modal policy intervention scenarios unlike previous methods: this creation and interpretation can help regions envision a more sustainable and equitable access-to-opportunity landscape.
+In summary, spatial availability can be used to create and interpret multimodal policy intervention scenarios unlike previous methods: this creation and interpretation can help regions envision a more sustainable and equitable access-to-opportunity landscape.
 
@@
 classoption:
-  - numbered #  When numbered option is activated, lines are numbered.
-output: rticles::trb_article
-wordcount: 7500
+#- numbered #  When numbered option is activated, lines are numbered.
+output: 
+  #word_document
+  rticles::trb_article
+wordcount: 7195
 ---
@@
 library(ggnewscale)
+library(bookdown)
 ```
@@
   autofit()
-
-table_toy_format_flex
 ```
@@
 ```{r, eval=FALSE}
+all_modes_v_im <- merge(x=car, y=bike, by = "VORI_CD_ZT208", all = T) %>% 
+  merge(transit, by = "VORI_CD_ZT208", all=T)%>%
+  merge(walk, by = "VORI_CD_ZT208", all = T)
+
 v_im_per_pop_plot <- ggplot(SA_im_zn208 %>% mutate(area = st_area(SA_im_zn208),
@@
 
+```{r, eval=FALSE}
+# here we example the relationship between v_im between modes -- I can't tell anything conclusive.
+car <- SA_im_zn208 %>% filter(mode=="car/motor") %>% st_drop_geometry() %>% select(VORI_CD_ZT208,v_im) %>% 
+  rename("v_im_car" = "v_im")
+bike <- SA_im_zn208 %>% filter(mode=="bike") %>% st_drop_geometry() %>% select(VORI_CD_ZT208,v_im) %>% 
+  rename("v_im_bike" = "v_im")
+transit <- SA_im_zn208 %>% filter(mode=="transit") %>% st_drop_geometry() %>% select(VORI_CD_ZT208,v_im) %>% 
+  rename("v_im_transit" = "v_im") 
+walk <- SA_im_zn208 %>% filter(mode=="walk")%>% st_drop_geometry() %>% select(VORI_CD_ZT208,v_im) %>% 
+  rename("v_im_walk" = "v_im")
+
+all_modes_v_im <- merge(x=car, y=bike, by = "VORI_CD_ZT208", all = T) %>% 
+  merge(transit, by = "VORI_CD_ZT208", all=T)%>%
+  merge(walk, by = "VORI_CD_ZT208", all = T)
+
+all_modes_v_im_plot <- ggplot(all_modes_v_im,
+       aes(x=v_im_transit,
+           y = v_im_walk)) +
+  geom_line()  + geom_point()+
+  theme_classic()
+```
+
 # Abstract
@@
 Implementing urban policies that re-shape cities through accessibility gains (i.e., the _potential to interact_ with opportunities as a result of land-use mix and transport systems as originally defined by @hansenHowAccessibilityShapes1959) have been widely applied within the transportation literature and is increasingly discussed by planners [@levinsonTransportAccessManual2020; @gowerPlanningInnovationCity2022; @siddiqToolsTradeAssessing2021; @yanAccessibilityBasedPlanningAddressing2021].
-An important challenge in the identification of interventions that sustainably and equitably transform cities is the effective evaluation of _trade-offs_: cities are complex and dynamic ecologies, and advantaging one component of the city can disadvantage another area, population, or sub-component In this way, policy evaluation should take a _systems_ approach [@fikselSustainabilityResilienceSystems2006].
+An important challenge in the identification of interventions that equitably transform cities is the effective evaluation of _trade-offs_: cities are complex and dynamic ecologies, and advantaging one component of the city can disadvantage another area, population, or sub-component In this way, policy evaluation should take a _systems_ approach [@fikselSustainabilityResilienceSystems2006].
 One way of considering systems is from the perspective of the _finite_.
@@
 However, a limitation of this study, like others that implement accessibility measures, is they do not calculate results under a _constrained_ framework i.e., one of _finity_.
-The citizens of Columbus should experience qualitative accessibility gains - but is it at the expense of access to opportunities by other modes?
-As another example, @mohriClusteringMethodMeasuring2021 implements a modified cumulative opportunity measure to assess differences between private vehicle and transit system accessibility to jobs in Melbourne, but a similar question remains, does the accessibility afforded to the private vehicle using population come at the expense of accessibility losses to transit users?
-
-The two studies discussed in the previous paragraph use _non-competitive_ accessibility measures.
+The citizens of Columbus should experience qualitative accessibility gains - but is it at the expense of access to opportunities to populations using other modes?
+<!-- As another example, @mohriClusteringMethodMeasuring2021 implements a modified cumulative opportunity measure to assess differences between private vehicle and transit system accessibility to jobs in Melbourne, but a similar question remains, does the accessibility afforded to the private vehicle using population come at the expense of accessibility losses to transit users?
+-->
+
+The work of @leeMeasuringImpactsNew2018 uses _non-competitive_ accessibility measures.
 There is a branch of location-based accessibility measures that do incorporate the effect of competition for opportunities by the population in the region.
 However, we argue that these existing methods fall short in acknowledging the _finity_ of opportunities.
-For instance, @maoMeasuringSpatialAccessibility2013 applies the competitive measure, the two-step floating catchment approach (2SFCA) for the case of access to healthcare services in Florida for both a multimodal network and a single modal network.
+For instance, @maoMeasuringSpatialAccessibility2013 applies a competitive measure, two-step floating catchment approach (2SFCA), for the case of access to healthcare services in Florida for both a multimodal network and a single modal network.
 While the differences in modal access are discussed, the question of how the advantage in access afforded by one mode over another impacts access for different mode users is unanswered.
 
-This question of how much one mode-using population can access at the expense of another mode-using population is pertinent equity question in the evaluation of policy scenarios that are multimodal.
+This question of how much one mode-using population can access at the expense of another mode-using population is a pertinent equity question in the evaluation of policy scenarios that are multimodal.
 For instance, consider the impact of a low emission zone (LEZ).
-LEZ is a policy of spatial and modal discrimination: the circulation of vehicles that are excessively polluting are restricted in specific areas in a city.
+LEZ is a policy of spatial and modal discrimination: the circulation of vehicles that are excessively polluting are restricted in specific regions.
 In the recognition that opportunities are finite, the implementation of a LEZ explicitly reduces the access that the population using polluting vehicles has to opportunities.
@@
 This evaluation is especially urgent as LEZ are currently in effect in cities globally; their reception has been mixed [@tarrinoortizPublicAcceptabilityLow2021] and may be having negative impacts on disadvantaged populations who have become mobility restricted [@devrijNooneVisitsMe2022; @verbeekJustManagementUrban2022].
-Measures that evaluate the accessibility of modes given both _constrained_ and _competitive_ considerations are lacking in the literature, but are needed, to evaluate such policy interventions to effectively comment on how accessibility changes as a result of the mobility-restricted mode.
+Measures that evaluate the accessibility of modes given both _constrained_ and _competitive_ considerations are lacking in the literature, but are needed, to evaluate such policy interventions impact on accessibility.
 
@@
 In Section 3, the spatial availability of an empirical example of the LEZ in the city of Madrid, Spain is calculated.
-We demonstrate how the restriction of private vehicles with the LEZ implementation impacts the spatial availability of opportunities for each sub-population using transit, cycling and walking modes.
-In Section 4, we provide concluding remarks on the strengths of the use of spatial availability as a multi-modal accessibility measure, limitations, and potential future uses in policy planning scenarios.
-
-# A brief review of multimodal accessibility measurement methods
-
+We demonstrate how the restriction of car circulation could have impacted the spatial availability of opportunities for each sub-population using transit, cycling and walking modes.
+In Section 4, we provide concluding remarks on the strengths of the use of spatial availability as a multimodal accessibility measure and potential future uses in policy planning scenarios.
+
+# A review of multimodal accessibility measures
 Location-based accessibility indicators are quantitative measures of _potential_ interaction with opportunities for locations within a given region: they are a product of the relationship between land-use and transport systems.
@@
 In this work, they demonstrate that the car mode has the highest $S_i^{m=car}$ values in the majority of $i$, i.e., populations using a car can potentially interact with the most retail opportunities than populations using other modes.
-However, the higher $S_i^{m=car}$ values are not a result of lower $S_i^{m}$ values for other modes: it is not assumed that car-using populations potentially accessing more opportunities take away potential opportunities for other populations within the measure.
-Put another way, this measure does not consider competition.
+However, the higher $S_i^{m=car}$ values are not a result of lower $S_i^{m}$ values for other modes: it is not assumed that car-using populations potentially accessing more opportunities take away potential opportunities for other populations within the measure (no consideration for competition).
 This measure is also not constrained: there is no global maximum for $S_i$ or $S_i^m$ values, they are presented as a population normalized accessibility index.
@@
 However, opportunities in a region can be considered finite.
-There are only so many school-seats, hospital capacity, food stores, jobs, etc., in a region and if one person interacts with an opportunity at a given time, it is taken.
+There are only so many school-seats, hospital capacity, jobs, etc., in a region and if one person interacts with an opportunity at a given time, it is taken.
 As such, if one person is advantaged and has the ability to reach more opportunities through a lower travel-cost mode, than they have more opportunities to potentially interact with more opportunities than other people.
@@
 To illustrate, @taoInvestigatingImpactsPublic2020a calculates $a_i^m$ to jobs for different income-group populations in Shenzhen (China) using $m = \text{public transit}$ and $m={car}$.
-They demonstrate that $i$s with low-income populations have lower $a_i^m$ than $i$s with higher-income populations.
-Further, they demonstrate that $a_i^{m=\text{public transit}}$ is lower than $a_i^{m=\text{car}}$ at many $i$s, arguing that this may put $i$s with lower-income populations in a further disadvantage.
+They demonstrate that $i$ with low-income populations have lower $a_i^m$ than $i$ with higher-income populations.
+Further, they demonstrate that $a_i^{m=\text{public transit}}$ is lower than $a_i^{m=\text{car}}$ at many $i$, arguing that this may put $i$ with lower-income populations in a further disadvantage.
 $a_i$ and/or $a_i^m$ are used to compare relative spatial differences in overall competitive accessibility and modal competitive accessibility, but because there is no global maximum, making it is difficult to interpret the significance between differences in $a_i^{m}$ values.
-Questions such as: What is the impact that competition has on the difference in $a_i^m$ values?
+Questions such as: what is the impact that competition has on the difference in $a_i^m$ values?
 How does impact vary spatially?
@@
 
-Spatial availability improves on previous multi-modal accessibility approaches as it considers _competition_ in the potential interaction with opportunities in a _constrained_ framework (e.g., finite opportunities).
+Spatial availability improves on previous multimodal accessibility approaches as it considers _competition_ in the potential interaction with opportunities in a _constrained_ framework (e.g., finite opportunities).
 This is done by considering: 1) competition between mass effect (e.g., the advantage of sub-populations residing in relatively low population-density and high opportunity-proximate areas) and 2) competition between travel impedance (e.g., sub-populations with relatively low travel-impedance) through a proportional allocation mechanism.
-The following sub-section demonstrates how spatial availability compares to the two other measures and the advantages in its interpretation through a synthetic example.
-
-## A synthetic example: calculating accessibility for multiple modes considering finite opportunities
+The following sub-section demonstrates how spatial availability compares to the Hansen-type and Shen-type measures through a synthetic example.
+
+## Multimodal spatial availability $V_i^m$
 
 In brief, we define the _spatial availability_ at $i$ ( $V_{i}$ ) as the proportion of all opportunities in the region $O$ that are allocated to location $i$ from all opportunity destinations $j$.
-$V_{i}$ is a value of how many opportunities are available to each location $i$ out of all the opportunities in the region (e.g., a neighbourhood has 1,000 spatially available jobs out of 100,000 jobs in the total region).
+$V_{i}$ is a value of how many opportunities are available to each location $i$ out of all the opportunities in the region.
 The general formulation of spatial availability $V_{i}$ is shown in Equation (\ref{eq:spatial-availability-general}):
@@
 
-\noindent where:
+\noindent Where:
 
@@
 Spatial availability measure is introduced in @soukhovIntroducingSpatialAvailability2023.
-The unique feature in the measure is the balancing factor $F^t_{ij}$, a proportional allocation mechanisms, which ensures that the $V_i$ calculated for each $i$ sums to the total number of opportunities.
+The unique feature in the measure is the balancing factor $F^t_{ij}$, a proportional allocation mechanisms, that ensures $V_i$ calculated for each $i$ sums to the total number of opportunities.
 Through $F^t_{ij}$, spatial availability is a  _competitive_ and _constrained_ accessibility measure that handles the number of opportunities in the region in a finite way.
 $F^t_{ij}$ consists of two components: a population-based balancing factor $F^p_{i} = \frac{P_i}{\sum_i P_i}$ and an impedance-based balancing factor $F^c_{ij} = \frac{F^c_{ij}}{\sum_j F^c_{ij}}$ that, respectively, allocate opportunities to $i$ in proportion to the size of the population at $i$ (the mass effect) and the cost of reaching opportunities at $j$ (the impedance effect).
+
 $F^p_{i}$ and $ F^c_{ij}$ are calculated for each $i$ such that they both equal 1 when summed across all $i$ in the region (e.g., $\sum_i F^p_{i} = 1$ and $\sum_i F^c_{ij} = 1$).
@@
 \noindent Where:
+
 - The population balancing factor for each $m$ at each $i$ is $F^{pm}_{i} = \frac{P_{i}^m}{\sum_{m}\sum_{i} P_{i}^m}$
@@
 
--   $m$ is a set of modes used by populations in the region $m = 1, \cdots, M$.
+-   $m$ is a set of modes used by populations in the region.
 -   $F^{tm}_{ij}$ is a balancing factor $F^t_{ij}$ for each $m$ at each $i$.
@@
 
-Consider the following synthetic example: Figure \ref{fig:Fig1} depicts a region with population and jobs at three population centers ($A$, $B$, $C$) and three employment centers ($1$, $2$, $3$).
-The population at each population center is divided into two sub-groups, one using a faster mode $z$ and another using a slower mode $x$, to travel to employment centers.
-Population center $A$ is Suburban: it is closest to its own relatively large employment center at $A$, close to the Urban area's equally large employment center $B$, and has a population that is smaller than the Urban center $2$ and larger than the Satellite center $3$.
-The Urban population center has the largest number of population using slower mode $x$, followed by the Suburban then Satellite area.
+## Synthetic example
+
+Consider the following: Figure \ref{fig:Fig1} depicts a region with population and jobs at three population centers ($A$, $B$, $C$) and three employment centers ($1$, $2$, $3$).
+The population at each population center is divided into two sub-populations, one using a faster mode $z$ and another using a slower mode $x$, to travel to employment centers.
+Population center $A$ is Suburban: it is closest to its own relatively large employment center at $1$, close to the Urban's equally large employment center $2$, and has a population that is smaller than the Urban $B$ and larger than the Satellite $C$.
+$B$ has the largest $x$-using population, followed by then $A$, then $C$.
 This synthetic example was inspired by the single-mode example used in @shenLocationCharacteristicsInnercity1998 and reconfigured in @soukhovIntroducingSpatialAvailability2023.
 
-```{r, synthetic-example-plot, fig.cap="\\label{fig:Fig1} Modified synthetic example from Shen (1998) with locations of employment centers (in orange), population centers (in blue), number of jobs and population, and travel times for two modes (slower x mode and faster z mode).", fig.show='hold', fig.align='center'}
+```{r, synthetic-example-plot, fig.cap="\\label{fig:Fig1} Multimodal synthetic example: locations of employment centers (in orange), population centers (in blue), number of jobs and population, and travel times for two modes (slower mode x and faster mode z).", fig.show='hold', fig.align='center'}
 knitr::include_graphics("images/Fig1.png")
@@
 
-From the perspective of access to a _finite_ amount of opportunities in the region ($210,000$ jobs), the sub-population that is most proximate to jobs, furthest from densely populated centers, and is using the lowest travel cost mode $z$ can potentially access the most job opportunities.
-This appears to be the sub-population at Suburban center $A$ using the faster mode $z$.
-From the other perspective, sub-populations located further away from jobs, close to dense populations, and using high cost travel mode $x$ are at a job opportunity access _disadvantage_ relative to the other sub-populations.
-This could be the sub-populations using the slowest mode $x$ at either Urban $B$ or Satellite $C$ area.
-From the perspective of inequities, the competition for opportunities between different mode-using populations (i.e., how well the land-use and transport system may serve some and not others), matters.
-
-```{r toy-example-table-all-values}
-autonum <- run_autonum(seq_id = "tab", bkm = "Tab1")
-table_toy_format_flex <- flextable::set_caption(table_toy_format_flex, 
-                                                caption = "Summary description of the synthetic example: accessibility values at each origin per mode m at each origin i and aggregated between modes for each i.",
-                                                style = "Table Caption",
-                                                autonum = autonum)
-
-table_toy_format_flex
-```
-
-The calculated $S_i^m$, $a_i^m$ and $V_i^m$ accessibility values for each $i$ and $m$ are shown in the middle three columns and are aggregated for each $i$ in the final two columns in Table (\ref{tab:Tab1}).
+From the perspective of access to a _finite_ amount of opportunities in the region ($210,000$ jobs), the sub-population that is most proximate to jobs, furthest from densely populated centers, and is using the lowest travel-cost mode $z$ can potentially access the most job opportunities.
+This appears to be the sub-population at $A$ using $z$.
+From the other perspective, sub-populations located in opposite conditions (i.e., further away from jobs, close to dense populations, and using $x$) are at a relative job opportunity access _disadvantage_.
+From the perspective of inequities, the competition for opportunities between different mode-using populations matters as it reflects how well the land-use and transport system serves (or doesn't serve) them.
+
+```{r tab1}
+flextable::set_caption(table_toy_format_flex, 
+            caption = "Accessibility values at each origin per mode m at each origin i and aggregated between modes for each i for the synthetic example.")
+```
+
+The calculated $S_i^m$, $a_i^m$ and $V_i^m$ accessibility values for each $i$ and $m$ are shown in the middle three columns and are aggregated for each $i$ in the final two columns in Table 1 <!-- \ref{tab:tab1}-->.
 We use a negative exponential impedance function $f(c_{ij}) = \exp(-\beta\cdot c_{ij})$ with $\beta=0.1$ for both $x$ and $z$ modes for all accessibility measures calculations.
 
-The Hansen-type measure $S_i^m$ is presented for each origin and mode in third column of Table (\ref{tab:Tab1}).
-For all $i$, the sub-population using the faster mode $z$ has higher $S_i^m$ values than the $x$ using sub-populations.
-Additionally, $S_i^m$ is equal for both mode using populations in population centers $A$ and $B$.
-This is the case because $S_i^m$ does not consider _competition_, it only relies on reflecting the count of opportunities that may be potentially interated with as a product of $f^m(c_{ij}^m)$.
-Recall, populations in $A$ and $B$ have the same travel impedance to employment centers $1$, $2$ and $3$ (either 15, 30, or 100 minutes for mode $x$ or 10, 25, or 80 minutes for mode $z$).
+The Hansen-type measure $S_i^m$ is presented for each origin and mode in third column of Table 1 <!-- \ref{tab:tab1}-->.
+For all $i$, the $z$-using sub-population has higher $S_i^m$ values than the $x$-using sub-populations.
+Additionally, $S_i^m$ is equal for both mode-using populations in $A$ and $B$.
+This is the case because $S_i^m$ does not consider _competition_, it only relies on reflecting the count of opportunities that may be interacted with as a product of $f^m(c_{ij}^m)$.
+Recall, populations in $A$ and $B$ have the same travel impedance to employment centers $1$, $2$ and $3$ (either 15, 30, or 100 minutes using $x$ or 10, 25, or 80 minutes using $z$).
 As such, these the calculated $S_i^m$ values are the same for both $A$ and $B$.
@@
 This value is difficult to interpret: it represents the weighted sum of opportunities that may be interacted with within the region based on travel impedance.
-It cannot be interpreted as a global maximum or any sort of benchmark as the measure is _non-constrained_.
-To connect this example to literature, this method of modal accessibility calculation is used in the work of @tahmasbiMultimodalAccessibilitybasedEquity2019; they compare differences in $S_i^m$ values between modes in a relative and comparative sense, but make no further interpretation of the $S_i^m$ values.
-
-In the fourth and sixth column in Table (\ref{tab:Tab1}) the Shen-type measure is calculated: first for both origin and mode $a_i^m$ as well as aggregated by the weighted mean mode-population ( $\sum_m \frac{P_i^m}{P_i}*a_i^m$ ) to represent a value for each origin $a_i$.
+It cannot be interpreted as any sort of benchmark since the measure is _non-constrained_.
+To connect this example to literature, $S_i^m$ is calculated in the work of @tahmasbiMultimodalAccessibilitybasedEquity2019; they compare differences in $S_i^m$ values between modes in a relative and comparative sense, but make no further interpretation of the $S_i^m$ values.
+
+In the fourth and sixth column in Table 1 <!-- \ref{tab:tab1}--> the Shen-type measure is calculated: first for both origin and mode $a_i^m$ as well as aggregated by the weighted mean mode-population ( $\sum_m \frac{P_i^m}{P_i}*a_i^m$ ) to represent a value for each origin $a_i$.
 Unlike $S_i^m$, this measure considers _competition_.
-For instance, the populations using the same modes in $A$ and $B$ centers do not have the same $a_i^m$ values.
-In fact, the Suburban $A$ has the highest values since this center has the smallest travel impedance to opportunities (lower than at $C$) and has the lowest relatively proximate to populations (lower than at $B$).
+For instance, the $x$-using populations in $A$ and $B$ centers do not have the same $a_i^m$ values as the $z$-using.
+In fact, $A$ has the highest values $a_i^m$ and $a_i$ values since this center has the smallest travel impedance to opportunities (lower than at $C$, $A$ and $B$ are equal) and has one of these lowest proximity to a relatively high amount of population (lower than at $B$).
 
 However, the Shen-type measure is _non-constrained_: the total sum of $a_i^m$ or $a_i$ is practically meaningless since it represents a sum of ratios.
-For instance, the sub-population using the fast mode $z$ at $A$ has a value of `r a_i %>% filter(i=="A" & m=="z") %>% select(a_im) %>% round(digits=2)` potential jobs per potential job-seeking population compared to `r a_i %>% filter(i=="A" & m=="x") %>% select(a_im) %>% round(digits=2)` for the slow mode $x$ population.
+For instance, the $z$-using sub-population at $A$ has a value of `r a_i %>% filter(i=="A" & m=="z") %>% select(a_im) %>% pull() %>% round(digits=2)` potential jobs per potential job-seeking population compared to `r a_i %>% filter(i=="A" & m=="x") %>% select(a_im) %>% pull() %>% round(digits=2)` for $x$-using sub-population.
 What is the significance of these values?
-The difference between these modes is equal to `r ((a_i %>% filter(i=="A" & m=="z") %>% select(a_im)) - (a_i %>% filter(i=="A" & m=="x") %>% select(a_im))) %>% round(digits=2)`, but `r ((a_i %>% filter(i=="A" & m=="z") %>% select(a_im)) - (a_i %>% filter(i=="A" & m=="x") %>% select(a_im))) %>% round(digits=2)` of what?
+The difference between these modes is equal to `r ((a_i %>% filter(i=="A" & m=="z") %>% select(a_im)) - (a_i %>% filter(i=="A" & m=="x") %>% select(a_im))) %>% pull() %>% round(digits=2)`, but `r ((a_i %>% filter(i=="A" & m=="z") %>% select(a_im)) - (a_i %>% filter(i=="A" & m=="x") %>% select(a_im))) %>% pull() %>% round(digits=2)` of what?
 How many more job opportunities are $z$ users interacting with than $x$ users?
@@
 This may be because the Shen-type measure is _non-constrained_, this is no benchmark or global maximum to which comparisons can be drawn from.
-Being unable to do so makes the interpretation of competition between modes challenging to tease out.
 
 By contrast, spatial availability $V_i$ considers competition and is constrained such that the total sum of values is equal to the total number of opportunities in the region (i.e., $210,000$ jobs).
-Seen in fifth column of Table (\ref{tab:Tab1}), $V_i^m$ for the same mode-using populations in $A$ and $B$ are not the same (as this measure considers competition).
-In fact, at $A$ for instance, the sub-population using faster mode $z$ captures `r ((table_toy_format %>% filter(i == "A" & m == "z") %>% select("V_im")) - (table_toy_format %>% filter(i == "A" & m == "x") %>% select("V_im"))) %>% prettyNum(big.mark = ",")` more spatially available jobs (of the $210,000$ jobs in the region) than the sub-population using mode $x$.
-The numerical difference have a practical interpretation.
+Seen in fifth column of Table 1 <!-- \ref{tab:tab1}-->, $V_i^m$ for the same mode-using populations in $A$ and $B$ are not the same (as this measure considers competition).
+In fact, at $A$, the $z$-using sub-population captures `r ((table_toy_format %>% filter(i == "A" & m == "z") %>% select("V_im")) - (table_toy_format %>% filter(i == "A" & m == "x") %>% select("V_im"))) %>% pull() %>% prettyNum(big.mark = ",")` more spatially available jobs (of the $210,000$ jobs in the region) than the sub-population using mode $x$.
+The numerical difference has a practical interpretation.
 
 Furthermore, $V_i^m$ values for an $i$ can be aggregated across $m$ and compared across $i$ ( $V_i = \sum_m{\sum_i{V_i^m}}$ ) as a result of the proportional allocation mechanism.
-This aggregation, $V_i$, is shown in the seventh column in Table (\ref{tab:Tab1}).
-Again looking at center $A$, $A$ is allocated `r ((V_i %>% filter(i == "A" & m == "z") %>% select("V_i"))) %>% prettyNum(big.mark = ",")` spatially available opportunities for both modes.
-`r ((V_i %>% filter(i == "A" & m == "z") %>% select("V_im")) / ((V_i %>% filter(i == "A" & m == "z") %>% select("V_i")))) %>% as.numeric() %>% percent() ` of this spatial availability allocated to $A$ is assigned to the $z$-using population despite representing `r 0.66 %>% percent()` of the population at $A$.
+This aggregation, $V_i$, is shown in the seventh column in Table 1 <!-- \ref{tab:tab1}-->.
+Again looking at center $A$, $A$ is allocated `r ((V_i %>% filter(i == "A" & m == "z")  %>% select("V_i"))) %>% pull() %>% prettyNum(big.mark = ",")` spatially available opportunities for both modes.
+`r ((V_i %>% filter(i == "A" & m == "z") %>% select("V_im")) / ((V_i %>% filter(i == "A" & m == "z") %>% select("V_i")))) %>% pull() %>% as.numeric() %>% percent() ` of this spatial availability allocated to $A$ is assigned to the $z$-using population despite representing `r 0.66 %>% percent()` of $A$'s population.
 
 Spatial availability can be further aggregated to better interpret competition between modes.
-Across the entire region,  `r V_ijm %>% filter(m=="z") %>% group_by(i) %>% summarize(pop = first(pop)) %>% select(pop) %>% sum() %>% prettyNum(big.mark = ",")` use the faster mode $z$ (`r ((V_ijm %>% filter(m=="z") %>% group_by(i) %>% summarize(pop = first(pop)) %>% select(pop) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% select("pop") %>% sum())) %>% percent()` of the region population).
-However, $z$-using populations accounts for `r ((V_ijm %>% filter(m=="z") %>% select(V_ij) %>% sum()) / (V_ijm %>% select(V_ij) %>% sum())) %>% percent()` of the region's total spatial availability - the rest is allocated to the population using $x$ (of which represents `r ((V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% filter(m == "x") %>% dplyr::select(pop) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% dplyr::select(pop) %>% sum())) %>% percent()` of the population).
-Notably, the $x$ population captures `r (((V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% filter(m == "x") %>% dplyr::select(pop) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% dplyr::select(pop) %>% sum())) - ((V_ijm %>% filter(m=="x") %>% select(V_ij) %>% sum()) / (V_ijm %>% select(V_ij) %>% sum()))) %>% percent()` less spatial availability to opportunities than its population proportion.
-
-Since spatial availability is constrained and has an interpretable meaning of a proportion of the total opportunities in the region, the values have a new significance.
-Inequality in spatial availability values can be explored through a variety of approaches.
+Across the entire region, `r V_ijm %>% filter(m=="z") %>% group_by(i) %>% summarize(pop = first(pop)) %>% select(pop) %>% sum() %>% prettyNum(big.mark = ",")` people use $z$ (`r ((V_ijm %>% filter(m=="z") %>% group_by(i) %>% summarize(pop = first(pop)) %>% select(pop) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% select("pop") %>% sum())) %>% percent()` of the region population).
+However, the $z$-using population accounts for `r ((V_ijm %>% filter(m=="z") %>% select(V_ij) %>% sum()) / (V_ijm %>% select(V_ij) %>% sum())) %>% percent()` of the region's total spatial availability - the rest is allocated to the $x$-using population (`r ((V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% filter(m == "x") %>% dplyr::select(pop) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% dplyr::select(pop) %>% sum())) %>% percent()`of the total population).
+Notably, the $x$-using population captures `r (((V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% filter(m == "x") %>% dplyr::select(pop) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% dplyr::select(pop) %>% sum())) - ((V_ijm %>% filter(m=="x") %>% select(V_ij) %>% sum()) / (V_ijm %>% select(V_ij) %>% sum()))) %>% percent()` less spatial availability to opportunities than its population proportion.
+This understanding can lead us to ask normative questions such as, how unequal should opportunity access for the two mode-using populations be?
+Can the lower-travel-cost populations spare some spatial availability if a policy of modal-restriction (like a LEZ) was introduced?
+
+<!-- THIS IS GOOD, but cutting to save words for now.
+
+Since spatial availability is constrained and has an interpretable meaning as a proportion of the total opportunities in the region, the values at $i$ have a new significance.
+Inequality in $V_i^m$ values can be explored through a variety of approaches.
 For instance, consider travel times.
-The $z$ population using the faster mode $z$ accounts for `r ((V_ijm %>% group_by(i,m) %>% summarize(V_tt = sum(V_ij*tt), .groups = "drop") %>% filter(m == "z") %>% dplyr::select(V_tt) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(V_tt = sum(V_ij*tt), .groups = "drop") %>% dplyr::select(V_tt) %>% sum())) %>% percent()` of the potential travel time traveled in the region: 6% less travel time than the proportion of spatial available opportunities that the $z$ population has access to.
-Meaning $z$ population travels less potential minutes overall and has more spatial availability of opportunities than the population using the slower mode $x$.
-
-Alternatively, these inequities in spatial availability between mode-using populations can be explored through proportional benchmarks.
-A spatial availability per capita as presented in Equation (\ref{eq:SA-per-capita}) is defined at each population center, mode, and/or regionally:
+The $z$-using population accounts for `r ((V_ijm %>% group_by(i,m) %>% summarize(V_tt = sum(V_ij*tt), .groups = "drop") %>% filter(m == "z") %>% dplyr::select(V_tt) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(V_tt = sum(V_ij*tt), .groups = "drop") %>% dplyr::select(V_tt) %>% sum())) %>% percent()` of the potential travel time traveled in the region: this is `r (((V_ijm %>% filter(m=="z") %>% select(V_ij) %>% sum()) / (V_ijm %>% select(V_ij) %>% sum())) - (V_ijm %>% group_by(i,m) %>% summarize(V_tt = sum(V_ij*tt), .groups = "drop") %>% filter(m == "z") %>% dplyr::select(V_tt) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(V_tt = sum(V_ij*tt), .groups = "drop") %>% dplyr::select(V_tt) %>% sum())) %>% percent()` less travel time than the proportion of spatial available opportunities that is allocated to them.
+In other words, the $z$-using population travels less minutes overall and has more spatial availability of opportunities than the $x$-using population using the slower mode $x$.
+-->
+Alternatively, inequities in spatial availability between mode-using populations can be explored through proportional benchmarks.
+A spatial availability per capita $v_i^m$ as presented in Equation (\ref{eq:SA-per-capita}):
 
@@
 \label{eq:SA-per-capita}
-v_{im} = \frac{V_{im}}{P_{im}}
+v_{i}^m = \frac{V_{i}^m}{P_{i}^m}
 \end{equation}
@@
 
-The spatial availability per capita values $v_i^m$ for $A$, $B$, and $C$ for the slower mode $x$ is `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="A" & m =="x") %>% select(v_i) %>% sum() %>% round(digits=2)`, `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="B" & m =="x") %>% select(v_i) %>% sum() %>% round(digits=2)` and `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="C" & m =="x") %>% select(v_i) %>% sum() %>% round(digits=2)` respectively.
-The benchmarks for $x$ mode are evidently lower than the averages at: `r V_i %>% group_by(i) %>% summarize(v_i = sum(V_im)/sum(pop)) %>% filter(i=="A") %>% select(v_i) %>% sum() %>% round(digits=2)`, `r V_i %>% group_by(i) %>% summarize(v_i = sum(V_im)/sum(pop)) %>% filter(i=="B") %>% select(v_i) %>% sum() %>% round(digits=2)` and `r V_i %>% group_by(i) %>% summarize(v_i = sum(V_im)/sum(pop)) %>% filter(i=="C") %>% select(v_i) %>% sum() %>% round(digits=2)` respectively.
-Especially for $B$ and $C$.
-
-The average for the faster mode $z$ with values of: `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="A" & m =="z") %>% select(v_i) %>% sum() %>% round(digits=2)`, `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="B" & m =="z") %>% select(v_i) %>% sum() %>% round(digits=2)` and `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="C" & m =="z") %>% select(v_i) %>% sum() %>% round(digits=2)` respectively.
-Once again, we see the competitive advantage on a per capita bases that populations have to opportunities when commuting using the faster mode.
-
-In what follows, we further explore competition between multimodal accessibility and how competition between different modal-users is captured by spatial availability through an empirical example.
-
-# Empirical multimodal spatial availability
-
-## Data inputs: Madrid's LEZ and travel survey
+The $v_i^m$ values for $A$, $B$, and $C$ for the $x$-using sub-populations are `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="A" & m =="x") %>% select(v_i) %>% sum() %>% round(digits=2)`, `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="B" & m =="x") %>% select(v_i) %>% sum() %>% round(digits=2)` and `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="C" & m =="x") %>% select(v_i) %>% sum() %>% round(digits=2)` spatially available jobs per capita, respectively.
+The $v_i^m$ for the $z$-using sub-populations are much higher, with values of: `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="A" & m =="z") %>% select(v_i) %>% sum() %>% round(digits=2)`, `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="B" & m =="z") %>% select(v_i) %>% sum() %>% round(digits=2)` and `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="C" & m =="z") %>% select(v_i) %>% sum() %>% round(digits=2)` respectively.
+The $x$-using population, especially at $B$ and $C$, are directly impacted by the jobs that are spatially available to the $z$-using population _in addition to_ the mass effect (occurring at $B$, high population density) and high travel impedance (occurring at the Satellite $C$).
+
+If, lets say, the planning goal is to have one spatially available job per mode-using population, a policy intervention can be put in place, to reduce the $v_i^z$ values and increase $v_i^x$ values.
+This demonstration is to show how simply the $V_i^m$ framework can be manipulated quantify the competitive (dis)advantage in a multimodal application.
+In what follows, we further explore competition between multiple modes through an empirical example.
+
+<!-- The benchmarks for $x$-using populations are evidently lower than the aggregated $A$, $B$, and $C$ $v_i$ values of (`r V_i %>% group_by(i) %>% summarize(v_i = sum(V_im)/sum(pop)) %>% filter(i=="A") %>% select(v_i) %>% sum() %>% round(digits=2)`, `r V_i %>% group_by(i) %>% summarize(v_i = sum(V_im)/sum(pop)) %>% filter(i=="B") %>% select(v_i) %>% sum() %>% round(digits=2)` and `r V_i %>% group_by(i) %>% summarize(v_i = sum(V_im)/sum(pop)) %>% filter(i=="C") %>% select(v_i) %>% sum() %>% round(digits=2)`).-->
+
+# Empirical example: Madrid LEZ
+
+## Multimodal data and methods
 ```{r}
@@
 Low emission zones (LEZ) have been implemented as a climate change policy intervention to reduce GHG emissions, improve air quality, and support sustainable mobility in many countries.
-Though rules vary depending on legal aspects and cultural norms, LEZ aim to deter or reduce traffic in specially designated zones under the penalty of fines and/or seizure of vehicle.
-In practice, this limits the volume of traffic by excluding vehicles by license plate, fuel type, or by introducing tolls.
-
-Spain is one of a few countries which have active LEZ and are interested in expanding their implementation.
-As specified in their recent climate change plan _Plan Nacional Integrado de Energía y Clima 2021-2030_ [@espanaPlanNacionalIntegrado2020] and the _Plan Nacional de Control de la Contaminación Atmosférica_ [@espanaResolucion10Enero2020], more LEZ will soon be implemented throughout the country.
-Specifically, the national Spanish law 7/2021 (“Ley de Cambio Climático y Transición Energética”) will require all municipalities to implement LEZ by 2023 if they meet at least one of the following requirements: (i) municipalities >50,000 inhab.; (ii) islands; and (iii) municipalities > 20,000 inhab.
-when air quality exceeds limits specified in “RD 102/2011 de Mejora de Calidad del Aire” [@barcelonaGUIATECNICAPARA2021].
-
-In 2017, a LEZ was implemented in the capital city of Madrid following the goals set out in the national agenda: to fight climate change, cut nitrogen dioxide levels for the benefit of people's health, and prioritize people's movement in the city.
-In geographic scope, the 2017 boundaries of the LEZ were relatively small (covering 4.72 km^{2}) and within the center (i.e., LEZ Centro).
-They were expanded in 2023 to just inside of the M-30, a highway in proximity to the city center (i.e., LEZ M-30) and the city has plans to further spatially expand the LEZ in the future.
-Within the 2017 LEZ Centro implementation, all cars, motorcycles and freight with environmental label A or B (higher polluting classification, associated with older make and model of fossil fuel internal combustion engine vehicles), are banned from driving into the area unless they are used by residents or meet other exemptions [@tarrinoortizAnalyzingImpactLow2022].
-This restriction impacts approximately half of all car that made trips into LEZ Centro.
-
-From the perspective of restriction for passenger transport, LEZ are a policy of _geographic discrimination_.
+Though rules vary<!-- depending on legal aspects and cultural norms-->, LEZ aim to deter/reduce traffic in designated zones under threat of penalty (e.g., fines, seizure of vehicle).
+<!-- In practice, LEZ limits the volume of traffic by excluding vehicles by license plate, fuel type, or by introducing tolls.--> From the perspective of restriction for passenger transport, LEZ are a policy of _geographic discrimination_.
 LEZ actively change how people access opportunities by making the travel impedance more costly for car-mode users.
 If seeing opportunities as finite, this discrimination allows populations to access opportunities by other modes more readily than before.
-In this way, the policy changes the multimodal competitive accessibility landscape of the city.
-Though the cost of travel for modes do dynamically change as a result of the LEZ implementation (i.e., potentially less car congestion, potentially more transit congestion), the focus of this empirical example is to demonstrate the spatial availability, by mode, of the city of Madrid after the 2017 implementation of the LEZ Centro.
+In this way, LEZ change the multimodal competitive accessibility landscape of a city.
+
+Spain is one of a few countries with active LEZ and plans to expand their implementation as specified in their climate-change-related plans: _Plan Nacional Integrado de Energía y Clima 2021-2030_ [@espanaPlanNacionalIntegrado2020] and _Plan Nacional de Control de la Contaminación Atmosférica_ [@espanaResolucion10Enero2020].
+Specifically, the national Spanish law 7/2021 ( _Ley de Cambio Climático y Transición Energética_) will require all municipalities to implement LEZ by 2023 if they meet at least one of the following requirements: (i) municipalities >50,000 inhab.; (ii) islands; and (iii) municipalities > 20,000 inhab.
+when air quality exceeds limits specified in _RD 102/2011 de Mejora de Calidad del Aire_ [@barcelonaGUIATECNICAPARA2021].
+
+In 2017, a LEZ was implemented in the Spanish capital city of Madrid following the goals set out in the national agenda <!-- : to fight climate change, cut nitrogen dioxide levels for the benefit of people's health, and prioritize people's movement in the city-->.
+In geographic scope, the 2017 boundaries of the LEZ were relatively small (covering 4.72 km^{2}) and within the center (i.e., LEZ Centro).
+These boundaries were expanded in 2023 to inside of the M-30, a highway in proximity to the city center (i.e., LEZ M-30) and the city has plans to further spatially expand the LEZ.
+Within the 2017 LEZ Centro implementation, all cars, motorcycles and freight with environmental label A or B (higher polluting classification, associated with older make and model of fossil fuel internal combustion engine vehicles), are banned from driving into the area (threat of fine) unless they are used by residents or meet other exemptions.
+This restriction impacted approximately half of all car trips that were typically made into the LEZ Centro [@tarrinoortizAnalyzingImpactLow2022].
+
+In this section, we use $V_i^m$ to quantify the competition for spatially available opportunities between modes after the LEZ Centro implementation.
+Particularly, we demonstrate how $V_i^m$ can be used to spectate on how the restriction of car mobility in areas around/within the LEZ Centro allowed the other, more sustainable but often with higher travel impedance modes, to become more competitive.
 
@@
 ```
-The 2018 Community of Madrid travel survey is the source of data for the empirical example: it is a representative survey that reflects a snap-shot of the travel patterns for one typical day of the working week (e.g., n=222,744 trips with representative population elevation factors).
-In this paper, a sample of the travel survey is used, namely the residential home origin to work destination trips of all modes and those that originate and end in Madrid.
+The 2018 Community of Madrid travel survey (@comunidaddemadridResultadosEDM20182020) is the source of data for this empirical example: it is a representative survey that reflects a snap-shot of the travel patterns for one typical day of the working week (e.g., n=222,744 trips with representative population elevation factors).
+In this paper, a sample of the travel survey is used, namely the residential home origin to work destination trips of all modes and those that originate and end in the city of Madrid.
 These totals are displayed in Figure \ref{fig:Fig2} and Figure \ref{fig:Fig3}.
-Both figures are displayed at the level of traffic analysis zones ($i$ and $j$) that correspond to the 2018 travel survey.
-The red boundary represents the LEZ Centro in effect in 2017 and thus those travel patterns of car-restriction reflected in the 2018 travel survey data.
+Both figures are displayed at the level of traffic analysis zones ($i$ and $j$) that correspond to the survey.
+The red boundary represents the LEZ Centro in effect in 2017 and thus those travel patterns of car-restriction reflected in the survey.
 The cyan boundary represents the LEZ that will be within the boundaries of the M-30 highway in 2023 and is present in the plots as a spatial reference for areas in proximity to the LEZ Centro.
@@
 The total sum of jobs $O_j$ that are held are shown in Figure \ref{fig:Fig2} and the populations that go to a work destination by four modal categories $P^m_i$, is reflected in Figure \ref{fig:Fig3}.
-The modal categories represented in Figure \ref{fig:Fig3} were summarized for the following trip mode types:
+The modal categories represented in Figure \ref{fig:Fig3} are summarized for the following trip mode types:
+
    - Car/motor: all cars and operating modes (e.g., cab, private driver, company, rental care, main driver, passenger, etc.) and all public, private or company motorcycle/mopeds.
@@
 
-From Figure \ref{fig:Fig2}, it can be seen that the largest concentration of jobs are within, near, and to the north of the LEZ Centro..
+From Figure \ref{fig:Fig2}, it can be seen that the largest concentration of jobs are within, near, and to the north of the LEZ Centro.
 The population that is accessing those jobs by mode (Figure \ref{fig:Fig3}), appear spatially distinct.
@@
 The distribution of walking and biking trips appear to be similar to that of transit trips.
-This is to be expected as active transport and a higher diversity of land-use spatially occurs with transit infrastructure.
+<!-- This is to be expected as active transport and a higher diversity of land-use spatially occurs with transit infrastructure.
 <!--CITE-->
 
-The travel time for each trip is provided within the 2018 survey.
-These travel times, per modal category, are used to calibrate mode specific travel impedance functions $f^m(c_{ij}^m)$.
-To illustrate the modal differences in travel lengths, summary descriptive per mode are detailed as follows:
-
-- Car/motor: `r weighted_mean(x=tld_empirical_car$TIEMPO_VIAJES, w=tld_empirical_car$ELE_G_POND_ESC2) %>% round()` min (Min:`r car_qs[1]%>% round()`min, Q2: `r car_qs[2]%>% round()` min, Q3: `r car_qs[4]%>% round()` min, Max: `r car_qs[5]%>% round()` min)
-- Transit: `r weighted_mean(x=tld_empirical_transit$TIEMPO_VIAJES, w=tld_empirical_transit$ELE_G_POND_ESC2)%>% round()` min (Min:`r transit_qs[1]%>% round()` min, Q2: `r transit_qs[2]%>% round()` min, Q3: `r transit_qs[4]%>% round()` min, Max: `r transit_qs[5]%>% round()` min)
-- Bike: `r weighted_mean(x=tld_empirical_bike$TIEMPO_VIAJES, w=tld_empirical_bike$ELE_G_POND_ESC2)%>% round()` min (Min:`r bike_qs[1]%>% round()` min, Q2: `r bike_qs[2]%>% round()` min, Q3: `r bike_qs[4]%>% round()` min, Max: `r bike_qs[5]%>% round()` min)
-- Walk: `r weighted_mean(x=tld_empirical_walk$TIEMPO_VIAJES, w=tld_empirical_walk$ELE_G_POND_ESC2)%>% round()` min (Min:`r walk_qs[1]%>% round()` min, Q2: `r walk_qs[2]%>% round()` min, Q3: `r walk_qs[4]%>% round()` min, Max: `r walk_qs[5]%>% round()` min)
-
-To calculate the mode specific travel impedance functions $f^m(c_{ij}^m)$ from the travel times, a concept known as the trip length distribution (TLD) is used.
-A TLD represents the proportion of trips that are taken at a specific travel cost such as travel time.
-This distribution is then used to derive impedance functions as done in previous accessibility research [e.g., works of @lopez_2017_spatial, @horbachov_theoretical_2018, and @batista_estimation_2019].
-Maximum likelihood estimation and the Nelder-Mead method for direct optimization available within the R {fitdistrplus} package [@fitdistrplus_2015] is used.
+The travel time for each trip is provided within the survey.
+These travel times, per modal category, are used to calibrate mode-specific travel impedance functions $f^m(c_{ij}^m)$.
+To illustrate the modal differences in travel lengths, summary descriptive per mode are detailed:
+
+- Car/motor: `r weighted_mean(x=tld_empirical_car$TIEMPO_VIAJES, w=tld_empirical_car$ELE_G_POND_ESC2) %>% round()` min (min:`r car_qs[1]%>% round()` min., Q2: `r car_qs[2]%>% round()` min., Q3: `r car_qs[4]%>% round()` min., max: `r car_qs[5]%>% round()` min.)
+- Transit: `r weighted_mean(x=tld_empirical_transit$TIEMPO_VIAJES, w=tld_empirical_transit$ELE_G_POND_ESC2)%>% round()` min.
+(min:`r transit_qs[1]%>% round()` min., Q2: `r transit_qs[2]%>% round()` min., Q3: `r transit_qs[4]%>% round()` min., max: `r transit_qs[5]%>% round()` min.)
+- Bike: `r weighted_mean(x=tld_empirical_bike$TIEMPO_VIAJES, w=tld_empirical_bike$ELE_G_POND_ESC2)%>% round()` min.
+(min:`r bike_qs[1]%>% round()` min., Q2: `r bike_qs[2]%>% round()` min., Q3: `r bike_qs[4]%>% round()` min., max: `r bike_qs[5]%>% round()` min.)
+- Walk: `r weighted_mean(x=tld_empirical_walk$TIEMPO_VIAJES, w=tld_empirical_walk$ELE_G_POND_ESC2)%>% round()` min.
+(min:`r walk_qs[1]%>% round()` min., Q2: `r walk_qs[2]%>% round()` min., Q3: `r walk_qs[4]%>% round()` min., max: `r walk_qs[5]%>% round()` min.)
+
+To calculate $f^m(c_{ij}^m)$ from the survey travel times, a concept known as the trip length distribution (TLD) was used.
+A TLD represents the proportion of trips that are taken at a specific travel cost such as travel time (i.e., probability density distribution of trips taken by travel cost).
+This distribution is then used to derive impedance functions [e.g., done in the accessibility works of @lopez_2017_spatial, @horbachov_theoretical_2018, and @batista_estimation_2019 for example].
+Maximum likelihood estimation and the Nelder-Mead method for direct optimization available within the R {fitdistrplus} package [@fitdistrplus_2015] is used to fit the impedance functions.
 As shown as shown in Figure \ref{fig:Fig4}, based on goodness-of-fit criteria and associated diagnostics, the gamma and log-normal probability density function (line curves) are selected as best fitting curves for the motorized and non-motorized modes respectively.
 The selection of functional form aligns with examples used in the literature (e.g., @reggianiAccessibilityImpedanceForms2011).
-<!--e.g., car gamma XX and active transport log-normal XX.--> Overall, the plots in Figure \ref{fig:Fig4} display the probability of travel given a trip travel time, based on actual travel behaviour from the 2018 survey.
-These 'probability of travel' at each travel time for each mode are realized observations reflect the land-use, the transport system, and the population travel preferences/behaviour in Madrid.
+<!--e.g., car gamma XX and active transport log-normal XX.--> Overall, the plots in Figure \ref{fig:Fig4} display the probability of travel given a trip travel time, based on trip flows from the survey.
+These 'probability of travel' at each travel time for each mode are realized observations reflect the land-use, the transport system, and the population travel behaviour in Madrid.
 
@@
 
-```{r, tlds-curves-m-plot, fig.cap="\\label{fig:Fig4} Fitted impedance function curve (line) against empirical TLD (bars) per mode.", fig.show='hold', fig.align='center', fig.height=6}
+```{r, tlds-curves-m-plot, fig.cap="\\label{fig:Fig4} Fitted impedance function curve (line) against empirical TLD (bars) corresponding to the home-to-work origin-destination flows from the Madrid 2018 travel survey.", fig.show='hold', fig.align='center', fig.height=6}
 knitr::include_graphics("images/tlds_curves_m_plot.png")
@@
 
-## Results: spatial availability $V_i^m$ and $V_i$
-
-Using the data inputs outlined, spatial availability of jobs are calculated for each of the four modal categories at the level of traffic analysis zones in Madrid, $V_i^m$.
-The spatial distribution of the resulting calculations are demonstrated in this section.
-
-```{r, SA-m-plot, fig.cap="\\label{fig:Fig5} Spatial availability of job opportunities per origin and mode $V_i^m$ in Madrid as reported by the home-to-work origin destination flows from the 2018 travel survey.
-2017 central LEZ is shown in blue.
-2023 expanded LEZ boundaries shown in red.", fig.show='hold', fig.align='center'}
+## Results
+Using the data inputs outlined, $V_i^m$, the spatial availability of jobs, is calculated for each of the four modal categories $m$ at the level of traffic analysis zones $i$ in Madrid and demonstrated in this sub-section.
+
+```{r, SA-m-plot, fig.cap="\\label{fig:Fig5} Spatial availability of job opportunities per origin and mode $V_i^m$ in Madrid.
+Calculated using the home-to-work origin-destination flows from the 2018 travel survey.
+", fig.show='hold', fig.align='center', fig.height=10}
 knitr::include_graphics("images/SA_im_V_zn208_plot.png")
@@
 
-Figure \ref{fig:Fig5} displays the spatial availability values for the four modal categories at the level of the spatial units used in the 2018 travel survey.
-These values represent a proportion of the total number of the `r SA_i_zn208$V_i %>% sum()  %>% prettyNum(big.mark = ",")` jobs in the region that are _spatially available_ to the population located at the $i$ based on the travel impedance of the mode (relative to the travel impedance of all modes) and the mode-using population size at the $i$ (relative to the population size of all modes and $i$).
-Since population, their job locations, and associated travel times are used to calculate $V_i^m$, Figure \ref{fig:Fig5} demonstrates values that represent the _realized_ spatial availability of jobs in Madrid.
-In other words, if someone decided to move into a neighbourhood at an $i$, the amount of jobs that are spatially available to them based on the population within their $i$ and their travel impedance for their used $m$, both relative to the city, is captured by the value of $V_i^m$.
-
-In Figure \ref{fig:Fig5}, the difference in the magnitudes of $V_i^m$ values between $m$ can be observed.
-The majority of $V_i^m$ is allocated to the populations using motorized modes.
-This is to be expected as commuting using motorized modes represents `r (pop_props$population_m[3]/sum(pop_props$population_m) + pop_props$population_m[4]/sum(pop_props$population_m)) %>% percent()` of the population (`r (pop_props$population_m[3]/sum(pop_props$population_m)) %>% percent()` (car/motor) and `r (pop_props$population_m[4]/sum(pop_props$population_m)) %>% percent()` (transit)).
+Figure \ref{fig:Fig5} displays $V_i^m$ for the four modal categories in 2018.
+$V_i^m$ values represent a proportion of the total number of the `r SA_i_zn208$V_i %>% sum()  %>% prettyNum(big.mark = ",")` jobs in the region; put another way, $V_i^m$ is how many jobs are _spatially available_ to the population located at the $i$ based on the travel impedance of the mode (relative to the travel impedance of all modes) and the mode-using population size at the $i$ (relative to the population size of all modes and $i$).
+<!-- Since population, their job locations, and associated travel times are used to calculate $V_i^m$, Figure \ref{fig:Fig5} demonstrates values that represent the _realized_ spatial availability of jobs in Madrid.
+In other words, if someone decided to move into a neighbourhood at an $i$, the amount of jobs that are spatially available to them based on the population within their $i$ and their travel impedance for their used $m$, both relative to the city, is captured by the value of $V_i^m$.--> Note the differences in the magnitudes of $V_i^m$ between modes.
+The majority of $V_i^m$ is allocated to car- and transit- using populations.
+This is to be expected, as commuting using motorized modes represents `r (pop_props$population_m[3]/sum(pop_props$population_m) + pop_props$population_m[4]/sum(pop_props$population_m)) %>% percent()` of the population (`r (pop_props$population_m[3]/sum(pop_props$population_m)) %>% percent()` (car/motor) and `r (pop_props$population_m[4]/sum(pop_props$population_m)) %>% percent()` (transit) in 2018).
 However, these modal options capture `r (modal_V_comps$V_m_perc[3] + modal_V_comps$V_m_perc[4]) %>% percent()` of the total spatial availability in Madrid.
@@
 How does the $V_i^m$ advantage allocated to car-using population arise?
-From the perceptive of finite opportunities, $V_i^m$ is allocated to car-using populations from less competitive modal populations.
+From the perspective of finite opportunities, $V_i^m$ is allocated to car-using populations from less competitive modal populations.
 How competitive one mode is compared to other modes varies spatially, but overall car-using populations capture more opportunities per car-using population than other modal populations.
@@
 
-```{r, SA-per-capita-m-plot, fig.cap="\\label{fig:Fig6} Spatial availability of job opportunities per capita per origin and mode $v_i^m$ in Madrid as reported by the home-to-work origin destination flows from the 2018 travel survey.
-2017 central LEZ is shown in blue.
-2023 expanded LEZ boundaries shown in red.", fig.show='hold', fig.align='center'}
+```{r, SA-per-capita-m-plot, fig.cap="\\label{fig:Fig6} Spatial availability of job opportunities per capita per origin and mode $v_i^m$ in Madrid.
+Calculated using the home-to-work origin destination flows from the 2018 travel survey.", fig.show='hold', fig.align='center', fig.height=10}
 knitr::include_graphics("images/SA_im_vv_zn208_plot.png")
@@
 Furthermore, there are spatial differences in the competitive advantage of spatial availability between modes.
-Figure \ref{fig:Fig6} visualizes $v_i^m$, the spatial availability divided by the mode population.
-$v_i^m$ values above 1 are represented in increasing red shades, values below 1 are represented in increasingly green shades, and values equal to 1 are white.
-These plots illustrates the discussion of the disproportionately high over representation of spatial availability relative to the mode-using population in many of the origins for the car plot (bottom right, denoted with green $v_i^m$ values above 1).
-These plots also visualize areas that are awarded disproportionately low spatial availability, represented in shades of red.
-Interestingly, spatial availability for the car mode within the LEZ Centro is below 1 (red).
-For all other modes, the area with the LEZ is relatively higher than the modal averages or even green in the case transit-using populations.
-This difference in spatial availability can be seen as a direct result of the LEZ Centro - the observed reduction of opportunities within the LEZ Centro boundaries being accessed by car-using populations allows lesser competitive modes to interact with these opportunities.
-
-```{r, SA-per-capita-m-linear-plot, fig.cap="\\label{fig:Fig7} Spatial availability of job opportunities per capita per origin and mode $v_i^m$ in Madrid as reported by the home-to-work origin destination flows from the 2018 travel survey represented by sorted total population per origin.", fig.show='hold', fig.align='center'}
-knitr::include_graphics("images/v_im_per_pop_plot.png")
-```
-
-Furthermore, Figure \ref{fig:Fig6} makes it evident that transit-using population's spatial availability to jobs is relatively balanced (i.e., many zones are white).
-Additionally, values for non-motorized modes are higher in origins that have higher transit accessibility.
-Transit accessibility and non-motorized modes do not appear to be in direct competition as a result of different travel impedance weighting <!-- (?
-not sure about this)-->.
-It is not visually clear in the plots but as demonstrated in a line plot of $v_i^m$ sorted by total population in Figure \ref{fig:Fig7}, it can be observed that transit is only close to or above 1 when car accessibility is relatively low.
-From Figure \ref{fig:Fig7} we can also observe that overall, motorized modes capture more spatial availability per capita than non-motorized modes.
-<!-- I'm considering removing Fig 7, the paper is becoming too long for TRB!
-I don't have enough room to explain it -->
-
-## Discussion and conclusions
-
-Location-based accessibility measures like the Hansen-type measure, Shen-type measure, and spatial availability all have a commonality - they are a weighted sum of opportunities assigned to each spatial unit in a region.
-In this way, they can be interpreted as a score that represents how many opportunities can be potentially interacted with by the population at each spatial unit.
-How the weight and sum of the potentially interacted opportunities is what defines the type of accessibility measure.
-Within this paper, the location-based accessibility measures known as spatial availability, a singly- _constrained_ and _competitive_ measure, is extended for the case of capturing multimodal accessibility to opportunities.
-A synthetic example and then an empirical case of LEZ in Madrid are detailed to demonstrate the multimodal extension of the spatial availability measure.
+Figure \ref{fig:Fig6} visualizes $v_i^m$, the spatial availability divided by the mode-population.
+<!--$v_i^m$ values above 1 are represented in increasing red shades, values below 1 are represented in increasingly green shades, and values equal to 1 are white.--> These plots illustrates the discussion of the disproportionately high over-allocation of spatial availability relative to the mode-using population in many of the origins for the car/motor mode (bottom left plot, areas denoted with green $v_i^m$ values above 1).
+These plots also visualize areas that disproportionately capture lower spatial availability (under 1), represented in shades of red.
+It can be observed that the transit-using population's spatial availability to jobs is relatively balanced (i.e., many zones are white), while the non-motorized modes $v_i^m$ values are low (under 1).
+
+Interestingly, $v_i^m$ for car/motor within and near the LEZ Centro is near or below 1 (white/red): potentially as a consequence, all non-car modes have relatively higher $v_i^m$ values in these areas.
+Though access to travel survey data immediately before the LEZ implementation is unavailable, we know that the LEZ Centro shifted more than half of all car trips into the LEZ to another mode [@tarrinoortizAnalyzingImpactLow2022].
+This restriction decreased the number of car-using population from $i$s going into the LEZ Centro (an area with a large number of jobs - Figure \ref{fig:Fig2}), thus increasing the mass effect for non-car modes and resulting in proportionally higher $v_i^m$ values for non-car modes.
+Though the spatial availability from before the LEZ Centro implementation is unknown, Figure \ref{fig:Fig6} provides a benchmark for quantifying potential LEZ implementations in the future (given 2018 travel conditions).
+Figure \ref{fig:Fig6} also shows that many areas within the M-30 have high (white/green) $v_i^m$ values for car-mode, signaling that the spatial expansion of the LEZ Centro stands to increase the spatial availability of jobs for non-car mode using populations.
+
+<!-- NOTE: SOME COMMENTS FOR FUTURE DISCUSSION POINTS ARE OUTLINED:
+
+These results can be explained in a different/better way..
+not sure how for now but....
+The number of trips going into the LEZ Centro is very small relative to all trips.
+This shift in car trips to non-car, I don't think, is the reason why spatial availability is high in the center for non-car modes.
+I think it is higher for non-car modes because it is highly connected with metro and high walkability/ped congestion that make driving more costly anyways.
+However, it is true that the reduction of population using car-mode DOES increase the competitiveness of other modes, just this impact is very small since the LEZ Centro is very small.
+
+Also, I do not talk about walk and cycling mode that much.
+If LEZ M30 is in place, transit likely captures most of these gains?
+(need to demonstrate this).
+But how much will walk and bike modes capture?
+Does transit competite directly with bike/walk (need to check this out somehow).
+How do we demonstrate this...?
+Maybe we show the change in balancing factors..?
+Perhaps we also show what could happen if we reduce the average travel times for bike/walk (increased walkability/bike infrastructure interventions) in addition to LEZ.
+
+Also, the sample sizes vary drastically between modes, with the cycling sample size being really small. How does this impact inter-modal comparisons?
+
+I conclude that restricting car modes leaves more opportunities for all other modes.
+I think this is because transit accessibility is in direct competition with car travel, and not in direct competition with non-motorized modes (need to demonstrate this - perhaps this can be shown with an x-y correlation between v_im valuesSee v_im_per_pop_plot.png (v_im per population) as well as the all_modes_v_im_plot plot.) -->
+
+# Discussion and conclusions
+Location-based accessibility measures like the Hansen-type $S_i^m$, Shen-type $a_i^m$, and spatial availability $V_i^m$ measures have a commonality - they are a weighted sum of opportunities assigned to each spatial unit $i$ in a region.
+<!-- In this way, they all can be interpreted as a score that represents how many opportunities can be potentially interacted with by the population at $i$.--> How the weight and sum of the potentially-interacted-with opportunities is considered is what defines the type of accessibility measure.
+Within this paper, the location-based singly- _constrained_ and _competitive_ accessibility measure, known as spatial availability $V_i$ (@soukhovIntroducingSpatialAvailability2023), is extended for the case of capturing multimodal accessibility to opportunities $V_i^m$.
+A synthetic example and then an empirical case of LEZ in Madrid are detailed to demonstrate this multimodal extension.
 
 The spatial availability measure is capable of capturing a new interpretation of multimodal competition that previous accessibility measures have not yet done.
-We can hypothesis that populations using modes with lower travel impedance, when competing for a finite set of opportunities, will capture more opportunities.
-However, with spatial availability, the number of spatially available opportunities that are captured (of the total opportunities in the region) by each mode can be individually calculated.
+Competitive measures hypothesis that populations using modes with lower travel impedance, when competing for a finite set of opportunities, will capture more opportunities.
+With spatial availability, the number of opportunities that are captured (of the total opportunities in the region) by each mode can be individually calculated.
 From there, the difference between how many spatially available opportunities one mode captures versus another can be investigated.
+This is the advantage of the spatial availability measure, particularly its multimodal extension.
 
 The flexibility and need for an accessibility measure such as spatial availability is pertinent in policy scenario evaluation.
-As showcased in the empirical example of the LEZ in Madrid, competition for job opportunity availability highly varies spatially and between modes.
-The car and transit modes have the highest spatial availability, with the car mode having highest availability with exception to the areas with LEZ Centro.
-This finding reflects real conditions: since car travel has been highly restricted within the LEZ Centro, much fewer car-using population and much more people are entering using other modes relative to the areas surrounding the LEZ Centro.
-This difference in car-using population within and immediately outside the LEZ Centro increases the competitiveness of the transit-using population (the second most competitive mode) as well as the non-motorized modes.
-
-Currently, conventional _non-constrained_ accessibility measures are difficult for planners to operationalize for a variety of reasons.
-They have also been criticized for being difficult to compute and difficult to interpret as they are a ratio of supply to demand [@levinsonTransportAccessManual2020].
+As showcased in the empirical example of the LEZ in Madrid, competition for job opportunity availability varies spatially _as well as_ between modes.
+The car and transit modes have the highest spatial availability, with the car mode having highest availability with exception to the areas within the LEZ Centro.
+This finding reflects what may suspect are real conditions: since car travel has been highly restricted within the LEZ Centro, fewer car-using people potentially interact with jobs within the LEZ Centro, leaving more _spatially available_ jobs for non-car-using populations.
+This difference in car-using populations in locations accessing jobs within and immediately outside the LEZ Centro increases the competitiveness of the transit-using population (the second most competitive mode) as well as the non-motorized modes.
+
+Spatial availability $V_i^m$ can also be divided by the mode-using population at each $i$ to yield mode-population normalized values.
+These values, reflected in Figure \ref{fig:Fig6}, can be used as a benchmark to investigate existing conditions and plan future LEZ implementation (i.e., target areas with exceptionally high car spatial availability such that more opportunities are available to other mode-users).
+
+In summary, conventional _non-constrained_ accessibility measures are difficult for planners to operationalize for a variety of reasons including issues of computation and interpretability [@levinsonTransportAccessManual2020].
 With spatial availability, the magnitude of opportunities that are available as a proportion of all the opportunities in the region is equal to $V_i$.
-As a result of its proportional allocation mechanism it can be easily extended into multimodal applications, pertinent to model policy scenerios in our cities that are becoming increasingly multimodal.
-This flexibility and interpretation of the spatial availability measure allows for manipulation of $V_i^m$ values to investigate differences of availability between neighbourhoods, modes, and regions, generate per capita benchmarks, and/or generate average values per population-group.
-
-From a spatial equity perspective, spatial availability measure can provide researchers, policy makers, and citizens a newfound interpretation of accessibility measures.
-With a plot of spatial availability values, one can begin asking, how much is enough and what level is too much.
+As a result of its proportional allocation mechanism, $V_i$ can be easily extended into multimodal applications.
+This flexibility is pertinent to model policy scenarios in our cities that are becoming increasingly multimodal.
+The interpretation of $V_i$ allows for manipulation of $V_i^m$ values to investigate differences of availability between neighbourhoods, modes, and regions, generate per capita benchmarks, and/or generate average values per population-group.
+
+From a spatial equity perspective, spatial availability measure can provide researchers, policy makers, and citizens a new-found interpretation of accessibility measures.
+With a plot of spatial availability values, one can begin asking, how much is enough and what level may be too much.
 These interpretations were difficult to be made with accessibility measures in the past.
 
-## Acknowledgements
-
-This research was funded by the Canada Graduate Scholarship - Doctoral Program (CGS D) by the Social Sciences and Humanities Research Council (SSHRC).
-
-#AUTHOR CONTRIBUTIONS
-
-The authors confirm contribution to the paper as follows: study conception and design: AS, JSL, AP.; data collection: AS, JSL, AP.; analysis and interpretation of results: AS, JSL, AP.; draft manuscript preparation: AS, JSL, AP.
+# Acknowledgements
+This research was funded by the Canada Graduate Scholarship - Doctoral Program (CGS D) provided by the Social Sciences and Humanities Research Council (SSHRC).
+
+# Author contributions
+The authors confirm contribution to the paper as follows: study conception and design: AS, JTO, JSL, AP.; data collection: AS, JTO, JSL.; analysis and interpretation of results: AS, JTO, JSL, AP.; draft manuscript preparation: AS, JSL, AP.
 All authors reviewed the results and approved the final version of the manuscript.
@@
 <!-- rejected text is left here -->
-
 <!-- As reviewed in the work of [@levinsonTransportAccessManual2020], cumulative opportunity measures and weighted cumulative opportunity measures are commonly used location-based accessibility measures; they use distance-decay functions (otherwise known as an impedance functions) to weight the opportunities that can be potentially accessed from origin to destination based on some sort of travel cost (e.g., travel time, fare, travel distance).
@@
 
-<!-- <!-- policy applications of accessibility broadly, but what's missing is multi-modal accessibility methods --> -->
+<!-- <!-- policy applications of accessibility broadly, but what's missing is multimodal accessibility methods --> 
 <!-- The consideration of both supply of opportunities and demand from the population is important for policy applications.

### 2023-07-31 TRB-manuscript/TRB-manuscript.tex
The submitted TRB draft

@@
 % When numbered option is activated, lines are numbered.
-\documentclass[numbered]{trbunofficial}
+\documentclass[]{trbunofficial}
 \usepackage{graphicx}
@@
 % Put here what will go to headers as author
-\AuthorHeaders{Multimodal spatial availability - the case of low
-emission zones in Madrid}
-\title{Multimodal spatial availability - the case of low emission zones
-in Madrid}
+\AuthorHeaders{}
+\title{Competition for Access to Opportunities between Modes: Multimodal
+Spatial Availability}
 
@@
   \hfill\break
-    \textbf{Javier Tarriño Ortiz}\\
+    \textbf{Javier Tarriño-Ortiz}\\
   PhD Graduate\\
   Universidad Politécnica de Madrid, Spain\\
-  \href{mailto:bob@example.com}{\nolinkurl{bob@example.com}}\\
+  \href{mailto:javier.tarrino.ortiz@upm.es}{\nolinkurl{javier.tarrino.ortiz@upm.es}}\\
   \hfill\break
@@
 Spain\\
-  \href{mailto:derek@example.com}{\nolinkurl{derek@example.com}}\\
+  \href{mailto:julio.soria-lara@upm.es}{\nolinkurl{julio.soria-lara@upm.es}}\\
   \hfill\break
@@
 % program i.e. if one does not have it installed.
-\TotalWords{7500}
+\TotalWords{7195}
 
@@
 transit). However, methods implemented in current accessibility
-literature are lacking within the context of multi-modal analysis. This
+literature are lacking within the context of multimodal analysis. This
 paper presents an extension of spatial availability, a
 singly-constrained competitive accessibility measure, for the context of
-multi-modal accessibility analysis. We first illustrate the features of
-spatial availability that lend itself to multi-modal analysis. We then
+multimodal accessibility analysis. We first illustrate the features of
+spatial availability that lend itself to multimodal analysis. We then
 demonstrate its use on the case study of Low Emission Zones in Madrid
@@
 accessibility of populations using different modes. In summary, spatial
-availability can be used to create and interpret multi-modal policy
+availability can be used to create and interpret multimodal policy
 intervention scenarios unlike previous methods: this creation and
@@
 \newpage
-
-\global\setlength{\Oldarrayrulewidth}{\arrayrulewidth}
-
-\global\setlength{\Oldtabcolsep}{\tabcolsep}
-
-\setlength{\tabcolsep}{0pt}
-
-\renewcommand*{\arraystretch}{1.5}
-
-\providecommand{\ascline}[3]{\noalign{\global\arrayrulewidth #1}\arrayrulecolor[HTML]{#2}\cline{#3}}
-
-\begin{longtable}[c]{|p{0.88in}|p{0.41in}|p{1.05in}|p{0.58in}|p{1.05in}|p{0.58in}|p{1.05in}}
-
-\ascline{1.5pt}{666666}{1-7}
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{i}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{m}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{S}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}} \\
-
-\ascline{1.5pt}{666666}{1-7}\endfirsthead
-
-\ascline{1.5pt}{666666}{1-7}
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{i}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{m}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{S}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}} \\
-
-\ascline{1.5pt}{666666}{1-7}\endhead
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{x}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{27,292.18}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.95}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{18,959.86}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.88in}{\raggedright \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{A}}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{z}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{44,999.80}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.56}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{46,913.04}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.58in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.32}}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{1.05in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{65,872.91}}}}} \\
-
-\ascline{1pt}{666666}{1-7}
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{x}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{27,292.18}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.62}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{30,863.43}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.88in}{\raggedright \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{B}}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{z}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{44,999.80}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.03}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{103,391.95}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.58in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.90}}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{1.05in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{134,255.38}}}}} \\
-
-\ascline{1pt}{666666}{1-7}
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{x}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{2,240.38}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.68}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{2,034.49}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.88in}{\raggedright \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{C}}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{z}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{3,745.89}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.12}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{7,837.22}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.58in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.99}}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{1.05in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{9,871.71}}}}} \\
-
-\ascline{1pt}{666666}{1-7}
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{TOTALS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{150,570.22}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{N/A}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{210,000.00}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{N/A}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{210,000.00}}} \\
-
-\ascline{1.5pt}{666666}{1-7}
-
-\end{longtable}
-
-\arrayrulecolor[HTML]{000000}
-
-\global\setlength{\arrayrulewidth}{\Oldarrayrulewidth}
-
-\global\setlength{\tabcolsep}{\Oldtabcolsep}
-
-\renewcommand*{\arraystretch}{1}
 
@@
 An important challenge in the identification of interventions that
-sustainably and equitably transform cities is the effective evaluation
-of \emph{trade-offs}: cities are complex and dynamic ecologies, and
+equitably transform cities is the effective evaluation of
+\emph{trade-offs}: cities are complex and dynamic ecologies, and
 advantaging one component of the city can disadvantage another area,
@@
 Columbus should experience qualitative accessibility gains - but is it
-at the expense of access to opportunities by other modes? As another
-example, \citet{mohriClusteringMethodMeasuring2021} implements a
-modified cumulative opportunity measure to assess differences between
-private vehicle and transit system accessibility to jobs in Melbourne,
-but a similar question remains, does the accessibility afforded to the
-private vehicle using population come at the expense of accessibility
-losses to transit users?
-
-The two studies discussed in the previous paragraph use
+at the expense of access to opportunities to populations using other
+modes?
+
+The work of \citet{leeMeasuringImpactsNew2018} uses
 \emph{non-competitive} accessibility measures. There is a branch of
@@
 \emph{finity} of opportunities. For instance,
-\citet{maoMeasuringSpatialAccessibility2013} applies the competitive
-measure, the two-step floating catchment approach (2SFCA) for the case
-of access to healthcare services in Florida for both a multimodal
-network and a single modal network. While the differences in modal
-access are discussed, the question of how the advantage in access
-afforded by one mode over another impacts access for different mode
-users is unanswered.
+\citet{maoMeasuringSpatialAccessibility2013} applies a competitive
+measure, two-step floating catchment approach (2SFCA), for the case of
+access to healthcare services in Florida for both a multimodal network
+and a single modal network. While the differences in modal access are
+discussed, the question of how the advantage in access afforded by one
+mode over another impacts access for different mode users is unanswered.
 
 This question of how much one mode-using population can access at the
-expense of another mode-using population is pertinent equity question in
-the evaluation of policy scenarios that are multimodal. For instance,
+expense of another mode-using population is a pertinent equity question
+in the evaluation of policy scenarios that are multimodal. For instance,
 consider the impact of a low emission zone (LEZ). LEZ is a policy of
 spatial and modal discrimination: the circulation of vehicles that are
-excessively polluting are restricted in specific areas in a city. In the
+excessively polluting are restricted in specific regions. In the
 recognition that opportunities are finite, the implementation of a LEZ
@@
 \emph{constrained} and \emph{competitive} considerations are lacking in
-the literature, but are needed, to evaluate such policy interventions to
-effectively comment on how accessibility changes as a result of the
-mobility-restricted mode.
+the literature, but are needed, to evaluate such policy interventions
+impact on accessibility.
 
@@
 in the city of Madrid, Spain is calculated. We demonstrate how the
-restriction of private vehicles with the LEZ implementation impacts the
-spatial availability of opportunities for each sub-population using
-transit, cycling and walking modes. In Section 4, we provide concluding
-remarks on the strengths of the use of spatial availability as a
-multi-modal accessibility measure, limitations, and potential future
-uses in policy planning scenarios.
-
-\hypertarget{a-brief-review-of-multimodal-accessibility-measurement-methods}{%
-\section{A brief review of multimodal accessibility measurement
-methods}\label{a-brief-review-of-multimodal-accessibility-measurement-methods}}
+restriction of car circulation could have impacted the spatial
+availability of opportunities for each sub-population using transit,
+cycling and walking modes. In Section 4, we provide concluding remarks
+on the strengths of the use of spatial availability as a multimodal
+accessibility measure and potential future uses in policy planning
+scenarios.
+
+\hypertarget{a-review-of-multimodal-accessibility-measures}{%
+\section{A review of multimodal accessibility
+measures}\label{a-review-of-multimodal-accessibility-measures}}
 
@@
 opportunities take away potential opportunities for other populations
-within the measure. Put another way, this measure does not consider
-competition. This measure is also not constrained: there is no global
-maximum for \(S_i\) or \(S_i^m\) values, they are presented as a
-population normalized accessibility index. This makes the interpretation
-of the `potentially interacted opportunities' relative to the region,
-making comparisons of the results across different regions challenging.
+within the measure (no consideration for competition). This measure is
+also not constrained: there is no global maximum for \(S_i\) or
+\(S_i^m\) values, they are presented as a population normalized
+accessibility index. This makes the interpretation of the `potentially
+interacted opportunities' relative to the region, making comparisons of
+the results across different regions challenging.
 
 However, opportunities in a region can be considered finite. There are
-only so many school-seats, hospital capacity, food stores, jobs, etc.,
-in a region and if one person interacts with an opportunity at a given
-time, it is taken. As such, if one person is advantaged and has the
-ability to reach more opportunities through a lower travel-cost mode,
-than they have more opportunities to potentially interact with more
-opportunities than other people. From the other perspective, their are
-fewer opportunities left to be potentially interacted with for
-populations using higher travel-cost modes. In this way, populations
-using modes with a higher travel impedance are at a higher access
-disadvantage than populations using lower travel impedance modes. This
-recognition is the motivation behind integrating \emph{competition} for
-opportunities within multimodal accessibility measures. Arguably one of
-the most popular competitive location-based accessibility measures is
-the two-step floating catchment area (2SFCA) approach popularized by
+only so many school-seats, hospital capacity, jobs, etc., in a region
+and if one person interacts with an opportunity at a given time, it is
+taken. As such, if one person is advantaged and has the ability to reach
+more opportunities through a lower travel-cost mode, than they have more
+opportunities to potentially interact with more opportunities than other
+people. From the other perspective, their are fewer opportunities left
+to be potentially interacted with for populations using higher
+travel-cost modes. In this way, populations using modes with a higher
+travel impedance are at a higher access disadvantage than populations
+using lower travel impedance modes. This recognition is the motivation
+behind integrating \emph{competition} for opportunities within
+multimodal accessibility measures. Arguably one of the most popular
+competitive location-based accessibility measures is the two-step
+floating catchment area (2SFCA) approach popularized by
 \citet{luoMeasuresSpatialAccessibility2003} who simplified the approach
@@
 (China) using \(m = \text{public transit}\) and \(m={car}\). They
-demonstrate that \(i\)s with low-income populations have lower \(a_i^m\)
-than \(i\)s with higher-income populations. Further, they demonstrate
+demonstrate that \(i\) with low-income populations have lower \(a_i^m\)
+than \(i\) with higher-income populations. Further, they demonstrate
 that \(a_i^{m=\text{public transit}}\) is lower than
-\(a_i^{m=\text{car}}\) at many \(i\)s, arguing that this may put \(i\)s
+\(a_i^{m=\text{car}}\) at many \(i\), arguing that this may put \(i\)
 with lower-income populations in a further disadvantage. \(a_i\) and/or
@@
 the significance between differences in \(a_i^{m}\) values. Questions
-such as: What is the impact that competition has on the difference in
+such as: what is the impact that competition has on the difference in
 \(a_i^m\) values? How does impact vary spatially? And what is the
@@
 
-Spatial availability improves on previous multi-modal accessibility
+Spatial availability improves on previous multimodal accessibility
 approaches as it considers \emph{competition} in the potential
@@
 mechanism. The following sub-section demonstrates how spatial
-availability compares to the two other measures and the advantages in
-its interpretation through a synthetic example.
-
-\hypertarget{a-synthetic-example-calculating-accessibility-for-multiple-modes-considering-finite-opportunities}{%
-\subsection{A synthetic example: calculating accessibility for multiple
-modes considering finite
-opportunities}\label{a-synthetic-example-calculating-accessibility-for-multiple-modes-considering-finite-opportunities}}
+availability compares to the Hansen-type and Shen-type measures through
+a synthetic example.
+
+\hypertarget{multimodal-spatial-availability-v_im}{%
+\subsection{\texorpdfstring{Multimodal spatial availability
+\(V_i^m\)}{Multimodal spatial availability V\_i\^{}m}}\label{multimodal-spatial-availability-v_im}}
 
@@
 \(V_{i}\) is a value of how many opportunities are available to each
-location \(i\) out of all the opportunities in the region (e.g., a
-neighbourhood has 1,000 spatially available jobs out of 100,000 jobs in
-the total region). The general formulation of spatial availability
-\(V_{i}\) is shown in Equation (\ref{eq:spatial-availability-general}):
+location \(i\) out of all the opportunities in the region. The general
+formulation of spatial availability \(V_{i}\) is shown in Equation
+(\ref{eq:spatial-availability-general}):
 
@@
 the measure is the balancing factor \(F^t_{ij}\), a proportional
-allocation mechanisms, which ensures that the \(V_i\) calculated for
-each \(i\) sums to the total number of opportunities. Through
-\(F^t_{ij}\), spatial availability is a \emph{competitive} and
-\emph{constrained} accessibility measure that handles the number of
-opportunities in the region in a finite way. \(F^t_{ij}\) consists of
-two components: a population-based balancing factor
-\(F^p_{i} = \frac{P_i}{\sum_i P_i}\) and an impedance-based balancing
-factor \(F^c_{ij} = \frac{F^c_{ij}}{\sum_j F^c_{ij}}\) that,
-respectively, allocate opportunities to \(i\) in proportion to the size
-of the population at \(i\) (the mass effect) and the cost of reaching
-opportunities at \(j\) (the impedance effect). \(F^p_{i}\) and \$
-F\^{}c\_\{ij\}\$ are calculated for each \(i\) such that they both equal
-1 when summed across all \(i\) in the region (e.g.,
+allocation mechanisms, that ensures \(V_i\) calculated for each \(i\)
+sums to the total number of opportunities. Through \(F^t_{ij}\), spatial
+availability is a \emph{competitive} and \emph{constrained}
+accessibility measure that handles the number of opportunities in the
+region in a finite way. \(F^t_{ij}\) consists of two components: a
+population-based balancing factor \(F^p_{i} = \frac{P_i}{\sum_i P_i}\)
+and an impedance-based balancing factor
+\(F^c_{ij} = \frac{F^c_{ij}}{\sum_j F^c_{ij}}\) that, respectively,
+allocate opportunities to \(i\) in proportion to the size of the
+population at \(i\) (the mass effect) and the cost of reaching
+opportunities at \(j\) (the impedance effect).
+
+\(F^p_{i}\) and \$ F\^{}c\_\{ij\}\$ are calculated for each \(i\) such
+that they both equal 1 when summed across all \(i\) in the region (e.g.,
 \(\sum_i F^p_{i} = 1\) and \(\sum_i F^c_{ij} = 1\)). These balancing
@@
 
-\noindent Where: - The population balancing factor for each \(m\) at
-each \(i\) is \(F^{pm}_{i} = \frac{P_{i}^m}{\sum_{m}\sum_{i} P_{i}^m}\)
-- The cost of travel balancing factor for each \(m\) at \(i\) is
+\noindent Where:
+
+\begin{itemize}
+\tightlist
+\item
+  The population balancing factor for each \(m\) at each \(i\) is
+  \(F^{pm}_{i} = \frac{P_{i}^m}{\sum_{m}\sum_{i} P_{i}^m}\)
+\item
+  The cost of travel balancing factor for each \(m\) at \(i\) is
   \(F_{ij}^{cm} = \frac{f(c_{ij}^m)}{\sum_{m} \sum_{i} f(c_{ij}^m)}\)
+\end{itemize}
 
@@
 \item
-  \(m\) is a set of modes used by populations in the region
-  \(m = 1, \cdots, M\).
+  \(m\) is a set of modes used by populations in the region.
 \item
@@
 
-Consider the following synthetic example: Figure \ref{fig:Fig1} depicts
-a region with population and jobs at three population centers (\(A\),
-\(B\), \(C\)) and three employment centers (\(1\), \(2\), \(3\)). The
-population at each population center is divided into two sub-groups, one
-using a faster mode \(z\) and another using a slower mode \(x\), to
-travel to employment centers. Population center \(A\) is Suburban: it is
-closest to its own relatively large employment center at \(A\), close to
-the Urban area's equally large employment center \(B\), and has a
-population that is smaller than the Urban center \(2\) and larger than
-the Satellite center \(3\). The Urban population center has the largest
-number of population using slower mode \(x\), followed by the Suburban
-then Satellite area. This synthetic example was inspired by the
-single-mode example used in
-\citet{shenLocationCharacteristicsInnercity1998} and reconfigured in
-\citet{soukhovIntroducingSpatialAvailability2023}.
+\hypertarget{synthetic-example}{%
+\subsection{Synthetic example}\label{synthetic-example}}
+
+Consider the following: Figure \ref{fig:Fig1} depicts a region with
+population and jobs at three population centers (\(A\), \(B\), \(C\))
+and three employment centers (\(1\), \(2\), \(3\)). The population at
+each population center is divided into two sub-populations, one using a
+faster mode \(z\) and another using a slower mode \(x\), to travel to
+employment centers. Population center \(A\) is Suburban: it is closest
+to its own relatively large employment center at \(1\), close to the
+Urban's equally large employment center \(2\), and has a population that
+is smaller than the Urban \(B\) and larger than the Satellite \(C\).
+\(B\) has the largest \(x\)-using population, followed by then \(A\),
+then \(C\). This synthetic example was inspired by the single-mode
+example used in \citet{shenLocationCharacteristicsInnercity1998} and
+reconfigured in \citet{soukhovIntroducingSpatialAvailability2023}.
 
@@
 
-\caption{\label{fig:Fig1} Modified synthetic example from Shen (1998) with locations of employment centers (in orange), population centers (in blue), number of jobs and population, and travel times for two modes (slower x mode and faster z mode).}\label{fig:synthetic-example-plot}
+\caption{\label{fig:Fig1} Multimodal synthetic example: locations of employment centers (in orange), population centers (in blue), number of jobs and population, and travel times for two modes (slower mode x and faster mode z).}\label{fig:synthetic-example-plot}
 \end{figure}
@@
 is most proximate to jobs, furthest from densely populated centers, and
-is using the lowest travel cost mode \(z\) can potentially access the
-most job opportunities. This appears to be the sub-population at
-Suburban center \(A\) using the faster mode \(z\). From the other
-perspective, sub-populations located further away from jobs, close to
-dense populations, and using high cost travel mode \(x\) are at a job
-opportunity access \emph{disadvantage} relative to the other
-sub-populations. This could be the sub-populations using the slowest
-mode \(x\) at either Urban \(B\) or Satellite \(C\) area. From the
-perspective of inequities, the competition for opportunities between
-different mode-using populations (i.e., how well the land-use and
-transport system may serve some and not others), matters.
+is using the lowest travel-cost mode \(z\) can potentially access the
+most job opportunities. This appears to be the sub-population at \(A\)
+using \(z\). From the other perspective, sub-populations located in
+opposite conditions (i.e., further away from jobs, close to dense
+populations, and using \(x\)) are at a relative job opportunity access
+\emph{disadvantage}. From the perspective of inequities, the competition
+for opportunities between different mode-using populations matters as it
+reflects how well the land-use and transport system serves (or doesn't
+serve) them.
 
@@
 
-\begin{longtable}[c]{|p{0.88in}|p{0.41in}|p{1.05in}|p{1.05in}|p{0.58in}|p{1.05in}|p{0.58in}}
-
-\caption{Summary\ description\ of\ the\ synthetic\ example:\ accessibility\ values\ at\ each\ origin\ per\ mode\ m\ at\ each\ origin\ i\ and\ aggregated\ between\ modes\ for\ each\ i.}\\
+\begin{longtable}[c]{|p{0.88in}|p{0.41in}|p{1.05in}|p{0.58in}|p{1.05in}|p{0.58in}|p{1.05in}}
+
+\caption{Accessibility\ values\ at\ each\ origin\ per\ mode\ m\ at\ each\ origin\ i\ and\ aggregated\ between\ modes\ for\ each\ i\ for\ the\ synthetic\ example.}\\
 
@@
 
-\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{i}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{m}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{S}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}} \\
-
-\ascline{1.5pt}{666666}{1-7}\endfirsthead \caption[]{Summary\ description\ of\ the\ synthetic\ example:\ accessibility\ values\ at\ each\ origin\ per\ mode\ m\ at\ each\ origin\ i\ and\ aggregated\ between\ modes\ for\ each\ i.}\\
+\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{i}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{m}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{S}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}} \\
+
+\ascline{1.5pt}{666666}{1-7}\endfirsthead \caption[]{Accessibility\ values\ at\ each\ origin\ per\ mode\ m\ at\ each\ origin\ i\ and\ aggregated\ between\ modes\ for\ each\ i\ for\ the\ synthetic\ example.}\\
 
@@
 
-\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{i}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{m}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{S}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}} \\
+\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{i}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{m}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{S}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}} \\
 
@@
 
-\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{x}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{18,959.86}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{27,292.18}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.95}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{65,872.91}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.88in}{\raggedright \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{A}}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{z}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{46,913.04}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{44,999.80}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.56}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{65,872.91}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.58in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.32}}}}} \\
+\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{x}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{27,292.18}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.95}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{18,959.86}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{} \\
+
+\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.88in}{\raggedright \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{A}}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{z}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{44,999.80}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.56}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{46,913.04}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.58in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.32}}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{1.05in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{65,872.91}}}}} \\
 
@@
 
-\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{x}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{30,863.43}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{27,292.18}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.62}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{134,255.38}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.88in}{\raggedright \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{B}}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{z}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{103,391.95}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{44,999.80}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.03}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{134,255.38}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.58in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.90}}}}} \\
+\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{x}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{27,292.18}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.62}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{30,863.43}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{} \\
+
+\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.88in}{\raggedright \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{B}}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{z}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{44,999.80}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.03}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{103,391.95}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.58in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.90}}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{1.05in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{134,255.38}}}}} \\
 
@@
 
-\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{x}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{2,034.49}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{2,240.38}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.68}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{9,871.71}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.88in}{\raggedright \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{C}}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{z}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{7,837.22}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{3,745.89}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.12}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{9,871.71}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.58in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.99}}}}} \\
+\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{x}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{2,240.38}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.68}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{2,034.49}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{} \\
+
+\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.88in}{\raggedright \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{C}}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{z}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{3,745.89}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.12}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{7,837.22}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.58in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.99}}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{1.05in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{9,871.71}}}}} \\
 
@@
 
-\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{TOTALS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{210,000.00}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{150,570.22}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{N/A}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{210,000.00}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{N/A}}} \\
+\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{TOTALS}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{150,570.22}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{N/A}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{210,000.00}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{N/A}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{210,000.00}}} \\
 
@@
 for each \(i\) and \(m\) are shown in the middle three columns and are
-aggregated for each \(i\) in the final two columns in Table
-(\ref{tab:Tab1}). We use a negative exponential impedance function
+aggregated for each \(i\) in the final two columns in Table 1 . We use a
+negative exponential impedance function
 \(f(c_{ij}) = \exp(-\beta\cdot c_{ij})\) with \(\beta=0.1\) for both
@@
 The Hansen-type measure \(S_i^m\) is presented for each origin and mode
-in third column of Table (\ref{tab:Tab1}). For all \(i\), the
-sub-population using the faster mode \(z\) has higher \(S_i^m\) values
-than the \(x\) using sub-populations. Additionally, \(S_i^m\) is equal
-for both mode using populations in population centers \(A\) and \(B\).
-This is the case because \(S_i^m\) does not consider \emph{competition},
-it only relies on reflecting the count of opportunities that may be
-potentially interated with as a product of \(f^m(c_{ij}^m)\). Recall,
-populations in \(A\) and \(B\) have the same travel impedance to
-employment centers \(1\), \(2\) and \(3\) (either 15, 30, or 100 minutes
-for mode \(x\) or 10, 25, or 80 minutes for mode \(z\)). As such, these
-the calculated \(S_i^m\) values are the same for both \(A\) and \(B\).
-Furthermore, the total sum of \(S_i^m\) in the region is equal to
-150,570.2. This value is difficult to interpret: it represents the
-weighted sum of opportunities that may be interacted with within the
-region based on travel impedance. It cannot be interpreted as a global
-maximum or any sort of benchmark as the measure is
-\emph{non-constrained}. To connect this example to literature, this
-method of modal accessibility calculation is used in the work of
-\citet{tahmasbiMultimodalAccessibilitybasedEquity2019}; they compare
+in third column of Table 1 . For all \(i\), the \(z\)-using
+sub-population has higher \(S_i^m\) values than the \(x\)-using
+sub-populations. Additionally, \(S_i^m\) is equal for both mode-using
+populations in \(A\) and \(B\). This is the case because \(S_i^m\) does
+not consider \emph{competition}, it only relies on reflecting the count
+of opportunities that may be interacted with as a product of
+\(f^m(c_{ij}^m)\). Recall, populations in \(A\) and \(B\) have the same
+travel impedance to employment centers \(1\), \(2\) and \(3\) (either
+15, 30, or 100 minutes using \(x\) or 10, 25, or 80 minutes using
+\(z\)). As such, these the calculated \(S_i^m\) values are the same for
+both \(A\) and \(B\). Furthermore, the total sum of \(S_i^m\) in the
+region is equal to 150,570.2. This value is difficult to interpret: it
+represents the weighted sum of opportunities that may be interacted with
+within the region based on travel impedance. It cannot be interpreted as
+any sort of benchmark since the measure is \emph{non-constrained}. To
+connect this example to literature, \(S_i^m\) is calculated in the work
+of \citet{tahmasbiMultimodalAccessibilitybasedEquity2019}; they compare
 differences in \(S_i^m\) values between modes in a relative and
@@
 
-In the fourth and sixth column in Table (\ref{tab:Tab1}) the Shen-type
-measure is calculated: first for both origin and mode \(a_i^m\) as well
-as aggregated by the weighted mean mode-population (
+In the fourth and sixth column in Table 1 the Shen-type measure is
+calculated: first for both origin and mode \(a_i^m\) as well as
+aggregated by the weighted mean mode-population (
 \(\sum_m \frac{P_i^m}{P_i}*a_i^m\) ) to represent a value for each
 origin \(a_i\). Unlike \(S_i^m\), this measure considers
-\emph{competition}. For instance, the populations using the same modes
-in \(A\) and \(B\) centers do not have the same \(a_i^m\) values. In
-fact, the Suburban \(A\) has the highest values since this center has
-the smallest travel impedance to opportunities (lower than at \(C\)) and
-has the lowest relatively proximate to populations (lower than at
-\(B\)).
+\emph{competition}. For instance, the \(x\)-using populations in \(A\)
+and \(B\) centers do not have the same \(a_i^m\) values as the
+\(z\)-using. In fact, \(A\) has the highest values \(a_i^m\) and \(a_i\)
+values since this center has the smallest travel impedance to
+opportunities (lower than at \(C\), \(A\) and \(B\) are equal) and has
+one of these lowest proximity to a relatively high amount of population
+(lower than at \(B\)).
 
@@
 of \(a_i^m\) or \(a_i\) is practically meaningless since it represents a
-sum of ratios. For instance, the sub-population using the fast mode
-\(z\) at \(A\) has a value of 1.56 potential jobs per potential
-job-seeking population compared to 0.95 for the slow mode \(x\)
-population. What is the significance of these values? The difference
-between these modes is equal to 0.62, but 0.62 of what? How many more
-job opportunities are \(z\) users interacting with than \(x\) users?
-When \(a_i^m\) is aggregated to \(a_i\) as shown in the sixth column,
-the values face similar interpretability issues.
+sum of ratios. For instance, the \(z\)-using sub-population at \(A\) has
+a value of 1.56 potential jobs per potential job-seeking population
+compared to 0.95 for \(x\)-using sub-population. What is the
+significance of these values? The difference between these modes is
+equal to 0.62, but 0.62 of what? How many more job opportunities are
+\(z\) users interacting with than \(x\) users? When \(a_i^m\) is
+aggregated to \(a_i\) as shown in the sixth column, the values face
+similar interpretability issues.
 
@@
 benchmark or global maximum to which comparisons can be drawn from.
-Being unable to do so makes the interpretation of competition between
-modes challenging to tease out.
 
@@
 number of opportunities in the region (i.e., \(210,000\) jobs). Seen in
-fifth column of Table (\ref{tab:Tab1}), \(V_i^m\) for the same
-mode-using populations in \(A\) and \(B\) are not the same (as this
-measure considers competition). In fact, at \(A\) for instance, the
-sub-population using faster mode \(z\) captures 27,953.18 more spatially
-available jobs (of the \(210,000\) jobs in the region) than the
-sub-population using mode \(x\). The numerical difference have a
-practical interpretation.
+fifth column of Table 1 , \(V_i^m\) for the same mode-using populations
+in \(A\) and \(B\) are not the same (as this measure considers
+competition). In fact, at \(A\), the \(z\)-using sub-population captures
+27,953.18 more spatially available jobs (of the \(210,000\) jobs in the
+region) than the sub-population using mode \(x\). The numerical
+difference has a practical interpretation.
 
@@
 result of the proportional allocation mechanism. This aggregation,
-\(V_i\), is shown in the seventh column in Table (\ref{tab:Tab1}). Again
-looking at center \(A\), \(A\) is allocated 65,872.91 spatially
-available opportunities for both modes. 71\% of this spatial
-availability allocated to \(A\) is assigned to the \(z\)-using
-population despite representing 66\% of the population at \(A\).
+\(V_i\), is shown in the seventh column in Table 1 . Again looking at
+center \(A\), \(A\) is allocated 65,872.91 spatially available
+opportunities for both modes. 71\% of this spatial availability
+allocated to \(A\) is assigned to the \(z\)-using population despite
+representing 66\% of \(A\)'s population.
 
 Spatial availability can be further aggregated to better interpret
-competition between modes. Across the entire region, 137,500 use the
-faster mode \(z\) (65\% of the region population). However, \(z\)-using
-populations accounts for 75\% of the region's total spatial availability
-- the rest is allocated to the population using \(x\) (of which
-represents 35\% of the population). Notably, the \(x\) population
-captures 10\% less spatial availability to opportunities than its
-population proportion.
-
-Since spatial availability is constrained and has an interpretable
-meaning of a proportion of the total opportunities in the region, the
-values have a new significance. Inequality in spatial availability
-values can be explored through a variety of approaches. For instance,
-consider travel times. The \(z\) population using the faster mode \(z\)
-accounts for 69\% of the potential travel time traveled in the region:
-6\% less travel time than the proportion of spatial available
-opportunities that the \(z\) population has access to. Meaning \(z\)
-population travels less potential minutes overall and has more spatial
-availability of opportunities than the population using the slower mode
-\(x\).
-
-Alternatively, these inequities in spatial availability between
-mode-using populations can be explored through proportional benchmarks.
-A spatial availability per capita as presented in Equation
-(\ref{eq:SA-per-capita}) is defined at each population center, mode,
-and/or regionally:
+competition between modes. Across the entire region, 137,500 people use
+\(z\) (65\% of the region population). However, the \(z\)-using
+population accounts for 75\% of the region's total spatial availability
+- the rest is allocated to the \(x\)-using population (35\%of the total
+population). Notably, the \(x\)-using population captures 10\% less
+spatial availability to opportunities than its population proportion.
+This understanding can lead us to ask normative questions such as, how
+unequal should opportunity access for the two mode-using populations be?
+Can the lower-travel-cost populations spare some spatial availability if
+a policy of modal-restriction (like a LEZ) was introduced?
+
+Alternatively, inequities in spatial availability between mode-using
+populations can be explored through proportional benchmarks. A spatial
+availability per capita \(v_i^m\) as presented in Equation
+(\ref{eq:SA-per-capita}):
 
@@
 \label{eq:SA-per-capita}
-v_{im} = \frac{V_{im}}{P_{im}}
+v_{i}^m = \frac{V_{i}^m}{P_{i}^m}
 \end{equation}
 
-The spatial availability per capita values \(v_i^m\) for \(A\), \(B\),
-and \(C\) for the slower mode \(x\) is 0.95, 0.62 and 0.68 respectively.
-The benchmarks for \(x\) mode are evidently lower than the averages at:
-1.32, 0.9 and 0.99 respectively. Especially for \(B\) and \(C\).
-
-The average for the faster mode \(z\) with values of: 1.56, 1.03 and
-1.12 respectively. Once again, we see the competitive advantage on a per
-capita bases that populations have to opportunities when commuting using
-the faster mode.
-
-In what follows, we further explore competition between multimodal
-accessibility and how competition between different modal-users is
-captured by spatial availability through an empirical example.
-
-\hypertarget{empirical-multimodal-spatial-availability}{%
-\section{Empirical multimodal spatial
-availability}\label{empirical-multimodal-spatial-availability}}
-
-\hypertarget{data-inputs-madrids-lez-and-travel-survey}{%
-\subsection{Data inputs: Madrid's LEZ and travel
-survey}\label{data-inputs-madrids-lez-and-travel-survey}}
+The \(v_i^m\) values for \(A\), \(B\), and \(C\) for the \(x\)-using
+sub-populations are 0.95, 0.62 and 0.68 spatially available jobs per
+capita, respectively. The \(v_i^m\) for the \(z\)-using sub-populations
+are much higher, with values of: 1.56, 1.03 and 1.12 respectively. The
+\(x\)-using population, especially at \(B\) and \(C\), are directly
+impacted by the jobs that are spatially available to the \(z\)-using
+population \emph{in addition to} the mass effect (occurring at \(B\),
+high population density) and high travel impedance (occurring at the
+Satellite \(C\)).
+
+If, lets say, the planning goal is to have one spatially available job
+per mode-using population, a policy intervention can be put in place, to
+reduce the \(v_i^z\) values and increase \(v_i^x\) values. This
+demonstration is to show how simply the \(V_i^m\) framework can be
+manipulated quantify the competitive (dis)advantage in a multimodal
+application. In what follows, we further explore competition between
+multiple modes through an empirical example.
+
+\hypertarget{empirical-example-madrid-lez}{%
+\section{Empirical example: Madrid
+LEZ}\label{empirical-example-madrid-lez}}
+
+\hypertarget{multimodal-data-and-methods}{%
+\subsection{Multimodal data and
+methods}\label{multimodal-data-and-methods}}
 
@@
 policy intervention to reduce GHG emissions, improve air quality, and
-support sustainable mobility in many countries. Though rules vary
-depending on legal aspects and cultural norms, LEZ aim to deter or
-reduce traffic in specially designated zones under the penalty of fines
-and/or seizure of vehicle. In practice, this limits the volume of
-traffic by excluding vehicles by license plate, fuel type, or by
-introducing tolls.
-
-Spain is one of a few countries which have active LEZ and are interested
-in expanding their implementation. As specified in their recent climate
-change plan \emph{Plan Nacional Integrado de Energía y Clima 2021-2030}
-\citep{espanaPlanNacionalIntegrado2020} and the \emph{Plan Nacional de
+support sustainable mobility in many countries. Though rules vary, LEZ
+aim to deter/reduce traffic in designated zones under threat of penalty
+(e.g., fines, seizure of vehicle). From the perspective of restriction
+for passenger transport, LEZ are a policy of \emph{geographic
+discrimination}. LEZ actively change how people access opportunities by
+making the travel impedance more costly for car-mode users. If seeing
+opportunities as finite, this discrimination allows populations to
+access opportunities by other modes more readily than before. In this
+way, LEZ change the multimodal competitive accessibility landscape of a
+city.
+
+Spain is one of a few countries with active LEZ and plans to expand
+their implementation as specified in their climate-change-related plans:
+\emph{Plan Nacional Integrado de Energía y Clima 2021-2030}
+\citep{espanaPlanNacionalIntegrado2020} and \emph{Plan Nacional de
 Control de la Contaminación Atmosférica}
-\citep{espanaResolucion10Enero2020}, more LEZ will soon be implemented
-throughout the country. Specifically, the national Spanish law 7/2021
-(``Ley de Cambio Climático y Transición Energética'') will require all
-municipalities to implement LEZ by 2023 if they meet at least one of the
-following requirements: (i) municipalities \textgreater50,000 inhab.;
-(ii) islands; and (iii) municipalities \textgreater{} 20,000 inhab. when
-air quality exceeds limits specified in ``RD 102/2011 de Mejora de
-Calidad del Aire'' \citep{barcelonaGUIATECNICAPARA2021}.
-
-In 2017, a LEZ was implemented in the capital city of Madrid following
-the goals set out in the national agenda: to fight climate change, cut
-nitrogen dioxide levels for the benefit of people's health, and
-prioritize people's movement in the city. In geographic scope, the 2017
-boundaries of the LEZ were relatively small (covering 4.72 km\^{}\{2\})
-and within the center (i.e., LEZ Centro). They were expanded in 2023 to
-just inside of the M-30, a highway in proximity to the city center
-(i.e., LEZ M-30) and the city has plans to further spatially expand the
-LEZ in the future. Within the 2017 LEZ Centro implementation, all cars,
-motorcycles and freight with environmental label A or B (higher
-polluting classification, associated with older make and model of fossil
-fuel internal combustion engine vehicles), are banned from driving into
-the area unless they are used by residents or meet other exemptions
-\citep{tarrinoortizAnalyzingImpactLow2022}. This restriction impacts
-approximately half of all car that made trips into LEZ Centro.
-
-From the perspective of restriction for passenger transport, LEZ are a
-policy of \emph{geographic discrimination}. LEZ actively change how
-people access opportunities by making the travel impedance more costly
-for car-mode users. If seeing opportunities as finite, this
-discrimination allows populations to access opportunities by other modes
-more readily than before. In this way, the policy changes the multimodal
-competitive accessibility landscape of the city. Though the cost of
-travel for modes do dynamically change as a result of the LEZ
-implementation (i.e., potentially less car congestion, potentially more
-transit congestion), the focus of this empirical example is to
-demonstrate the spatial availability, by mode, of the city of Madrid
-after the 2017 implementation of the LEZ Centro.
+\citep{espanaResolucion10Enero2020}. Specifically, the national Spanish
+law 7/2021 ( \emph{Ley de Cambio Climático y Transición Energética})
+will require all municipalities to implement LEZ by 2023 if they meet at
+least one of the following requirements: (i) municipalities
+\textgreater50,000 inhab.; (ii) islands; and (iii) municipalities
+\textgreater{} 20,000 inhab. when air quality exceeds limits specified
+in \emph{RD 102/2011 de Mejora de Calidad del Aire}
+\citep{barcelonaGUIATECNICAPARA2021}.
+
+In 2017, a LEZ was implemented in the Spanish capital city of Madrid
+following the goals set out in the national agenda . In geographic
+scope, the 2017 boundaries of the LEZ were relatively small (covering
+4.72 km\^{}\{2\}) and within the center (i.e., LEZ Centro). These
+boundaries were expanded in 2023 to inside of the M-30, a highway in
+proximity to the city center (i.e., LEZ M-30) and the city has plans to
+further spatially expand the LEZ. Within the 2017 LEZ Centro
+implementation, all cars, motorcycles and freight with environmental
+label A or B (higher polluting classification, associated with older
+make and model of fossil fuel internal combustion engine vehicles), are
+banned from driving into the area (threat of fine) unless they are used
+by residents or meet other exemptions. This restriction impacted
+approximately half of all car trips that were typically made into the
+LEZ Centro \citep{tarrinoortizAnalyzingImpactLow2022}.
+
+In this section, we use \(V_i^m\) to quantify the competition for
+spatially available opportunities between modes after the LEZ Centro
+implementation. Particularly, we demonstrate how \(V_i^m\) can be used
+to spectate on how the restriction of car mobility in areas
+around/within the LEZ Centro allowed the other, more sustainable but
+often with higher travel impedance modes, to become more competitive.
 
@@
 
-The 2018 Community of Madrid travel survey is the source of data for the
-empirical example: it is a representative survey that reflects a
-snap-shot of the travel patterns for one typical day of the working week
-(e.g., n=222,744 trips with representative population elevation
+The 2018 Community of Madrid travel survey
+(\citet{comunidaddemadridResultadosEDM20182020}) is the source of data
+for this empirical example: it is a representative survey that reflects
+a snap-shot of the travel patterns for one typical day of the working
+week (e.g., n=222,744 trips with representative population elevation
 factors). In this paper, a sample of the travel survey is used, namely
 the residential home origin to work destination trips of all modes and
-those that originate and end in Madrid. These totals are displayed in
-Figure \ref{fig:Fig2} and Figure \ref{fig:Fig3}. Both figures are
-displayed at the level of traffic analysis zones (\(i\) and \(j\)) that
-correspond to the 2018 travel survey. The red boundary represents the
+those that originate and end in the city of Madrid. These totals are
+displayed in Figure \ref{fig:Fig2} and Figure \ref{fig:Fig3}. Both
+figures are displayed at the level of traffic analysis zones (\(i\) and
+\(j\)) that correspond to the survey. The red boundary represents the
 LEZ Centro in effect in 2017 and thus those travel patterns of
-car-restriction reflected in the 2018 travel survey data. The cyan
-boundary represents the LEZ that will be within the boundaries of the
-M-30 highway in 2023 and is present in the plots as a spatial reference
-for areas in proximity to the LEZ Centro.
+car-restriction reflected in the survey. The cyan boundary represents
+the LEZ that will be within the boundaries of the M-30 highway in 2023
+and is present in the plots as a spatial reference for areas in
+proximity to the LEZ Centro.
 
@@
 modal categories \(P^m_i\), is reflected in Figure \ref{fig:Fig3}. The
-modal categories represented in Figure \ref{fig:Fig3} were summarized
-for the following trip mode types: - Car/motor: all cars and operating
-modes (e.g., cab, private driver, company, rental care, main driver,
-passenger, etc.) and all public, private or company motorcycle/mopeds. -
-Transit: all bus, trams, and trains - Bike: all bicycle trips (e.g.,
-private, public, or company bike trips) and ``other'' types of
-micromobility options - Walk: walking or by foot
+modal categories represented in Figure \ref{fig:Fig3} are summarized for
+the following trip mode types:
+
+\begin{itemize}
+\tightlist
+\item
+  Car/motor: all cars and operating modes (e.g., cab, private driver,
+  company, rental care, main driver, passenger, etc.) and all public,
+  private or company motorcycle/mopeds.
+\item
+  Transit: all bus, trams, and trains
+\item
+  Bike: all bicycle trips (e.g., private, public, or company bike trips)
+  and ``other'' types of micromobility options
+\item
+  Walk: walking or by foot
+\end{itemize}
 
@@
 concentration of jobs are within, near, and to the north of the LEZ
-Centro.. The population that is accessing those jobs by mode (Figure
+Centro. The population that is accessing those jobs by mode (Figure
 \ref{fig:Fig3}), appear spatially distinct. Car and transit trips
@@
 The distribution of walking and biking trips appear to be similar to
-that of transit trips. This is to be expected as active transport and a
-higher diversity of land-use spatially occurs with transit
-infrastructure.
-
-The travel time for each trip is provided within the 2018 survey. These
-travel times, per modal category, are used to calibrate mode specific
+that of transit trips.
+
+The travel time for each trip is provided within the survey. These
+travel times, per modal category, are used to calibrate mode-specific
 travel impedance functions \(f^m(c_{ij}^m)\). To illustrate the modal
-differences in travel lengths, summary descriptive per mode are detailed
-as follows:
+differences in travel lengths, summary descriptive per mode are
+detailed:
 
@@
 \item
-  Car/motor: 36 min (Min:0min, Q2: 15 min, Q3: 55 min, Max: 120 min)
-\item
-  Transit: 55 min (Min:1 min, Q2: 30 min, Q3: 80 min, Max: 120 min)
-\item
-  Bike: 34 min (Min:5 min, Q2: 15 min, Q3: 40 min, Max: 115 min)
-\item
-  Walk: 27 min (Min:1 min, Q2: 10 min, Q3: 45 min, Max: 119 min)
+  Car/motor: 36 min (min:0 min., Q2: 15 min., Q3: 55 min., max: 120
+  min.)
+\item
+  Transit: 55 min. (min:1 min., Q2: 30 min., Q3: 80 min., max: 120 min.)
+\item
+  Bike: 34 min. (min:5 min., Q2: 15 min., Q3: 40 min., max: 115 min.)
+\item
+  Walk: 27 min. (min:1 min., Q2: 10 min., Q3: 45 min., max: 119 min.)
 \end{itemize}
 
-To calculate the mode specific travel impedance functions
-\(f^m(c_{ij}^m)\) from the travel times, a concept known as the trip
-length distribution (TLD) is used. A TLD represents the proportion of
-trips that are taken at a specific travel cost such as travel time. This
-distribution is then used to derive impedance functions as done in
-previous accessibility research \citep[e.g., works
+To calculate \(f^m(c_{ij}^m)\) from the survey travel times, a concept
+known as the trip length distribution (TLD) was used. A TLD represents
+the proportion of trips that are taken at a specific travel cost such as
+travel time (i.e., probability density distribution of trips taken by
+travel cost). This distribution is then used to derive impedance
+functions \citep[e.g., done in the accessibility works
 of][\citet{horbachov_theoretical_2018}, and
-\citet{batista_estimation_2019}]{lopez_2017_spatial}. Maximum likelihood
-estimation and the Nelder-Mead method for direct optimization available
-within the R \{fitdistrplus\} package \citep{fitdistrplus_2015} is used.
-As shown as shown in Figure \ref{fig:Fig4}, based on goodness-of-fit
+\citet{batista_estimation_2019} for example]{lopez_2017_spatial}.
+Maximum likelihood estimation and the Nelder-Mead method for direct
+optimization available within the R \{fitdistrplus\} package
+\citep{fitdistrplus_2015} is used to fit the impedance functions. As
+shown as shown in Figure \ref{fig:Fig4}, based on goodness-of-fit
 criteria and associated diagnostics, the gamma and log-normal
@@
 plots in Figure \ref{fig:Fig4} display the probability of travel given a
-trip travel time, based on actual travel behaviour from the 2018 survey.
-These `probability of travel' at each travel time for each mode are
-realized observations reflect the land-use, the transport system, and
-the population travel preferences/behaviour in Madrid.
+trip travel time, based on trip flows from the survey. These
+`probability of travel' at each travel time for each mode are realized
+observations reflect the land-use, the transport system, and the
+population travel behaviour in Madrid.
 
@@
 
-\caption{\label{fig:Fig4} Fitted impedance function curve (line) against empirical TLD (bars) per mode.}\label{fig:tlds-curves-m-plot}
+\caption{\label{fig:Fig4} Fitted impedance function curve (line) against empirical TLD (bars) corresponding to the home-to-work origin-destination flows from the Madrid 2018 travel survey.}\label{fig:tlds-curves-m-plot}
 \end{figure}
 
-\hypertarget{results-spatial-availability-v_im-and-v_i}{%
-\subsection{\texorpdfstring{Results: spatial availability \(V_i^m\) and
-\(V_i\)}{Results: spatial availability V\_i\^{}m and V\_i}}\label{results-spatial-availability-v_im-and-v_i}}
-
-Using the data inputs outlined, spatial availability of jobs are
-calculated for each of the four modal categories at the level of traffic
-analysis zones in Madrid, \(V_i^m\). The spatial distribution of the
-resulting calculations are demonstrated in this section.
+\hypertarget{results}{%
+\subsection{Results}\label{results}}
+
+Using the data inputs outlined, \(V_i^m\), the spatial availability of
+jobs, is calculated for each of the four modal categories \(m\) at the
+level of traffic analysis zones \(i\) in Madrid and demonstrated in this
+sub-section.
 
@@
 
-\caption{\label{fig:Fig5} Spatial availability of job opportunities per origin and mode $V_i^m$ in Madrid as reported by the home-to-work origin destination flows from the 2018 travel survey.
-2017 central LEZ is shown in blue.
-2023 expanded LEZ boundaries shown in red.}\label{fig:SA-m-plot}
+\caption{\label{fig:Fig5} Spatial availability of job opportunities per origin and mode $V_i^m$ in Madrid.
+Calculated using the home-to-work origin-destination flows from the 2018 travel survey.
+}\label{fig:SA-m-plot}
 \end{figure}
 
-Figure \ref{fig:Fig5} displays the spatial availability values for the
-four modal categories at the level of the spatial units used in the 2018
-travel survey. These values represent a proportion of the total number
-of the 847,574 jobs in the region that are \emph{spatially available} to
-the population located at the \(i\) based on the travel impedance of the
-mode (relative to the travel impedance of all modes) and the mode-using
-population size at the \(i\) (relative to the population size of all
-modes and \(i\)). Since population, their job locations, and associated
-travel times are used to calculate \(V_i^m\), Figure \ref{fig:Fig5}
-demonstrates values that represent the \emph{realized} spatial
-availability of jobs in Madrid. In other words, if someone decided to
-move into a neighbourhood at an \(i\), the amount of jobs that are
-spatially available to them based on the population within their \(i\)
-and their travel impedance for their used \(m\), both relative to the
-city, is captured by the value of \(V_i^m\).
-
-In Figure \ref{fig:Fig5}, the difference in the magnitudes of \(V_i^m\)
-values between \(m\) can be observed. The majority of \(V_i^m\) is
-allocated to the populations using motorized modes. This is to be
-expected as commuting using motorized modes represents 84\% of the
-population (37\% (car/motor) and 47\% (transit)). However, these modal
-options capture 95\% of the total spatial availability in Madrid. In
-particular, the car/motor using population is allocated
+Figure \ref{fig:Fig5} displays \(V_i^m\) for the four modal categories
+in 2018. \(V_i^m\) values represent a proportion of the total number of
+the 847,574 jobs in the region; put another way, \(V_i^m\) is how many
+jobs are \emph{spatially available} to the population located at the
+\(i\) based on the travel impedance of the mode (relative to the travel
+impedance of all modes) and the mode-using population size at the \(i\)
+(relative to the population size of all modes and \(i\)). Note the
+differences in the magnitudes of \(V_i^m\) between modes. The majority
+of \(V_i^m\) is allocated to car- and transit- using populations. This
+is to be expected, as commuting using motorized modes represents 84\% of
+the population (37\% (car/motor) and 47\% (transit) in 2018). However,
+these modal options capture 95\% of the total spatial availability in
+Madrid. In particular, the car/motor using population is allocated
 disproportionately more \(V_i^m\) than its modal population (37\% of the
@@
 How does the \(V_i^m\) advantage allocated to car-using population
-arise? From the perceptive of finite opportunities, \(V_i^m\) is
+arise? From the perspective of finite opportunities, \(V_i^m\) is
 allocated to car-using populations from less competitive modal
@@
 
-\caption{\label{fig:Fig6} Spatial availability of job opportunities per capita per origin and mode $v_i^m$ in Madrid as reported by the home-to-work origin destination flows from the 2018 travel survey.
-2017 central LEZ is shown in blue.
-2023 expanded LEZ boundaries shown in red.}\label{fig:SA-per-capita-m-plot}
+\caption{\label{fig:Fig6} Spatial availability of job opportunities per capita per origin and mode $v_i^m$ in Madrid.
+Calculated using the home-to-work origin destination flows from the 2018 travel survey.}\label{fig:SA-per-capita-m-plot}
 \end{figure}
@@
 of spatial availability between modes. Figure \ref{fig:Fig6} visualizes
-\(v_i^m\), the spatial availability divided by the mode population.
-\(v_i^m\) values above 1 are represented in increasing red shades,
-values below 1 are represented in increasingly green shades, and values
-equal to 1 are white. These plots illustrates the discussion of the
-disproportionately high over representation of spatial availability
-relative to the mode-using population in many of the origins for the car
-plot (bottom right, denoted with green \(v_i^m\) values above 1). These
-plots also visualize areas that are awarded disproportionately low
-spatial availability, represented in shades of red. Interestingly,
-spatial availability for the car mode within the LEZ Centro is below 1
-(red). For all other modes, the area with the LEZ is relatively higher
-than the modal averages or even green in the case transit-using
-populations. This difference in spatial availability can be seen as a
-direct result of the LEZ Centro - the observed reduction of
-opportunities within the LEZ Centro boundaries being accessed by
-car-using populations allows lesser competitive modes to interact with
-these opportunities.
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{images/v_im_per_pop_plot}
-
-}
-
-\caption{\label{fig:Fig7} Spatial availability of job opportunities per capita per origin and mode $v_i^m$ in Madrid as reported by the home-to-work origin destination flows from the 2018 travel survey represented by sorted total population per origin.}\label{fig:SA-per-capita-m-linear-plot}
-\end{figure}
-
-Furthermore, Figure \ref{fig:Fig6} makes it evident that transit-using
-population's spatial availability to jobs is relatively balanced (i.e.,
-many zones are white). Additionally, values for non-motorized modes are
-higher in origins that have higher transit accessibility. Transit
-accessibility and non-motorized modes do not appear to be in direct
-competition as a result of different travel impedance weighting . It is
-not visually clear in the plots but as demonstrated in a line plot of
-\(v_i^m\) sorted by total population in Figure \ref{fig:Fig7}, it can be
-observed that transit is only close to or above 1 when car accessibility
-is relatively low. From Figure \ref{fig:Fig7} we can also observe that
-overall, motorized modes capture more spatial availability per capita
-than non-motorized modes.
+\(v_i^m\), the spatial availability divided by the mode-population.
+These plots illustrates the discussion of the disproportionately high
+over-allocation of spatial availability relative to the mode-using
+population in many of the origins for the car/motor mode (bottom left
+plot, areas denoted with green \(v_i^m\) values above 1). These plots
+also visualize areas that disproportionately capture lower spatial
+availability (under 1), represented in shades of red. It can be observed
+that the transit-using population's spatial availability to jobs is
+relatively balanced (i.e., many zones are white), while the
+non-motorized modes \(v_i^m\) values are low (under 1).
+
+Interestingly, \(v_i^m\) for car/motor within and near the LEZ Centro is
+near or below 1 (white/red): potentially as a consequence, all non-car
+modes have relatively higher \(v_i^m\) values in these areas. Though
+access to travel survey data immediately before the LEZ implementation
+is unavailable, we know that the LEZ Centro shifted more than half of
+all car trips into the LEZ to another mode
+\citep{tarrinoortizAnalyzingImpactLow2022}. This restriction decreased
+the number of car-using population from \(i\)s going into the LEZ Centro
+(an area with a large number of jobs - Figure \ref{fig:Fig2}), thus
+increasing the mass effect for non-car modes and resulting in
+proportionally higher \(v_i^m\) values for non-car modes. Though the
+spatial availability from before the LEZ Centro implementation is
+unknown, Figure \ref{fig:Fig6} provides a benchmark for quantifying
+potential LEZ implementations in the future (given 2018 travel
+conditions). Figure \ref{fig:Fig6} also shows that many areas within the
+M-30 have high (white/green) \(v_i^m\) values for car-mode, signaling
+that the spatial expansion of the LEZ Centro stands to increase the
+spatial availability of jobs for non-car mode using populations.
 
 \hypertarget{discussion-and-conclusions}{%
-\subsection{Discussion and
-conclusions}\label{discussion-and-conclusions}}
-
-Location-based accessibility measures like the Hansen-type measure,
-Shen-type measure, and spatial availability all have a commonality -
-they are a weighted sum of opportunities assigned to each spatial unit
-in a region. In this way, they can be interpreted as a score that
-represents how many opportunities can be potentially interacted with by
-the population at each spatial unit. How the weight and sum of the
-potentially interacted opportunities is what defines the type of
-accessibility measure. Within this paper, the location-based
-accessibility measures known as spatial availability, a singly-
-\emph{constrained} and \emph{competitive} measure, is extended for the
-case of capturing multimodal accessibility to opportunities. A synthetic
-example and then an empirical case of LEZ in Madrid are detailed to
-demonstrate the multimodal extension of the spatial availability
-measure.
+\section{Discussion and conclusions}\label{discussion-and-conclusions}}
+
+Location-based accessibility measures like the Hansen-type \(S_i^m\),
+Shen-type \(a_i^m\), and spatial availability \(V_i^m\) measures have a
+commonality - they are a weighted sum of opportunities assigned to each
+spatial unit \(i\) in a region. How the weight and sum of the
+potentially-interacted-with opportunities is considered is what defines
+the type of accessibility measure. Within this paper, the location-based
+singly- \emph{constrained} and \emph{competitive} accessibility measure,
+known as spatial availability \(V_i\)
+(\citet{soukhovIntroducingSpatialAvailability2023}), is extended for the
+case of capturing multimodal accessibility to opportunities \(V_i^m\). A
+synthetic example and then an empirical case of LEZ in Madrid are
+detailed to demonstrate this multimodal extension.
 
@@
 interpretation of multimodal competition that previous accessibility
-measures have not yet done. We can hypothesis that populations using
-modes with lower travel impedance, when competing for a finite set of
-opportunities, will capture more opportunities. However, with spatial
-availability, the number of spatially available opportunities that are
-captured (of the total opportunities in the region) by each mode can be
-individually calculated. From there, the difference between how many
-spatially available opportunities one mode captures versus another can
-be investigated.
+measures have not yet done. Competitive measures hypothesis that
+populations using modes with lower travel impedance, when competing for
+a finite set of opportunities, will capture more opportunities. With
+spatial availability, the number of opportunities that are captured (of
+the total opportunities in the region) by each mode can be individually
+calculated. From there, the difference between how many spatially
+available opportunities one mode captures versus another can be
+investigated. This is the advantage of the spatial availability measure,
+particularly its multimodal extension.
 
@@
 the empirical example of the LEZ in Madrid, competition for job
-opportunity availability highly varies spatially and between modes. The
-car and transit modes have the highest spatial availability, with the
-car mode having highest availability with exception to the areas with
-LEZ Centro. This finding reflects real conditions: since car travel has
-been highly restricted within the LEZ Centro, much fewer car-using
-population and much more people are entering using other modes relative
-to the areas surrounding the LEZ Centro. This difference in car-using
-population within and immediately outside the LEZ Centro increases the
-competitiveness of the transit-using population (the second most
-competitive mode) as well as the non-motorized modes.
-
-Currently, conventional \emph{non-constrained} accessibility measures
-are difficult for planners to operationalize for a variety of reasons.
-They have also been criticized for being difficult to compute and
-difficult to interpret as they are a ratio of supply to demand
+opportunity availability varies spatially \emph{as well as} between
+modes. The car and transit modes have the highest spatial availability,
+with the car mode having highest availability with exception to the
+areas within the LEZ Centro. This finding reflects what may suspect are
+real conditions: since car travel has been highly restricted within the
+LEZ Centro, fewer car-using people potentially interact with jobs within
+the LEZ Centro, leaving more \emph{spatially available} jobs for
+non-car-using populations. This difference in car-using populations in
+locations accessing jobs within and immediately outside the LEZ Centro
+increases the competitiveness of the transit-using population (the
+second most competitive mode) as well as the non-motorized modes.
+
+Spatial availability \(V_i^m\) can also be divided by the mode-using
+population at each \(i\) to yield mode-population normalized values.
+These values, reflected in Figure \ref{fig:Fig6}, can be used as a
+benchmark to investigate existing conditions and plan future LEZ
+implementation (i.e., target areas with exceptionally high car spatial
+availability such that more opportunities are available to other
+mode-users).
+
+In summary, conventional \emph{non-constrained} accessibility measures
+are difficult for planners to operationalize for a variety of reasons
+including issues of computation and interpretability
 \citep{levinsonTransportAccessManual2020}. With spatial availability,
@@
 the opportunities in the region is equal to \(V_i\). As a result of its
-proportional allocation mechanism it can be easily extended into
-multimodal applications, pertinent to model policy scenerios in our
-cities that are becoming increasingly multimodal. This flexibility and
-interpretation of the spatial availability measure allows for
-manipulation of \(V_i^m\) values to investigate differences of
-availability between neighbourhoods, modes, and regions, generate per
-capita benchmarks, and/or generate average values per population-group.
+proportional allocation mechanism, \(V_i\) can be easily extended into
+multimodal applications. This flexibility is pertinent to model policy
+scenarios in our cities that are becoming increasingly multimodal. The
+interpretation of \(V_i\) allows for manipulation of \(V_i^m\) values to
+investigate differences of availability between neighbourhoods, modes,
+and regions, generate per capita benchmarks, and/or generate average
+values per population-group.
 
 From a spatial equity perspective, spatial availability measure can
-provide researchers, policy makers, and citizens a newfound
+provide researchers, policy makers, and citizens a new-found
 interpretation of accessibility measures. With a plot of spatial
 availability values, one can begin asking, how much is enough and what
-level is too much. These interpretations were difficult to be made with
-accessibility measures in the past.
+level may be too much. These interpretations were difficult to be made
+with accessibility measures in the past.
 
 \hypertarget{acknowledgements}{%
-\subsection{Acknowledgements}\label{acknowledgements}}
+\section{Acknowledgements}\label{acknowledgements}}
 
 This research was funded by the Canada Graduate Scholarship - Doctoral
-Program (CGS D) by the Social Sciences and Humanities Research Council
-(SSHRC).
-
-\#AUTHOR CONTRIBUTIONS
+Program (CGS D) provided by the Social Sciences and Humanities Research
+Council (SSHRC).
+
+\hypertarget{author-contributions}{%
+\section{Author contributions}\label{author-contributions}}
 
 The authors confirm contribution to the paper as follows: study
-conception and design: AS, JSL, AP.; data collection: AS, JSL, AP.;
-analysis and interpretation of results: AS, JSL, AP.; draft manuscript
-preparation: AS, JSL, AP. All authors reviewed the results and approved
-the final version of the manuscript.
-
---\textgreater{}
+conception and design: AS, JTO, JSL, AP.; data collection: AS, JTO,
+JSL.; analysis and interpretation of results: AS, JTO, JSL, AP.; draft
+manuscript preparation: AS, JSL, AP. All authors reviewed the results
+and approved the final version of the manuscript.
 

### 2023-08-14 TRB-manuscript/TRB-manuscript.Rmd
Suggested title for PLoS ONE paper

@@
 title: "Competition for Access to Opportunities between Modes: Multimodal Spatial Availability"
-#runningheader: "Capturing Multimodal Access to Opportunities with the Spatial Availability Measure"
+#runningheader: "Multimodal Spatial Availability: A Singly-Constrained Measure of Competitive Accessibility by Multiple Modes"
 author:

### 2023-08-14 TRB-manuscript/TRB-manuscript.Rmd
Edited funding info

@@
 # Acknowledgements
-This research was funded by the Canada Graduate Scholarship - Doctoral Program (CGS D) provided by the Social Sciences and Humanities Research Council (SSHRC).
+This research was funded by the Canada Graduate Scholarship - Doctoral Program (CGS D) provided by the Social Sciences and Humanities Research Council (SSHRC) and Project Mobilizing Justice, also supported by SSHRC.
 
@@
 
-<!-- <!-- examples of constrained (competitive) measures, the introduction of spatial availability --> -->
+<!-- <!-- examples of constrained (competitive) measures, the introduction of spatial availability -->
 <!-- Competitive measures are are also a weighted sum of opportunities, but do so in different ways.

### 2023-08-24 Journal-manuscript/Journal-manuscript.Rmd
Submitted this version to journal 1

--- 2023-08-14 TRB-manuscript/TRB-manuscript.Rmd
@@
 ---
-title: "Competition for Access to Opportunities between Modes: Multimodal Spatial Availability"
-#runningheader: "Multimodal Spatial Availability: A Singly-Constrained Measure of Competitive Accessibility by Multiple Modes"
+title: "Multimodal Spatial Availability: A Singly-Constrained Measure of Competitive Accessibility by Multiple Modes"
 author:
@@
 keywords: ["Multimodal", "Accessibility", "Equity", "Policy scenarios", "Low emission zones"]
+
 bibliography: mybibfile.bib
-biblio-style: unsrtnat
-classoption:
-#- numbered #  When numbered option is activated, lines are numbered.
-output: 
-  #word_document
-  rticles::trb_article
-wordcount: 7195
+output: rticles::plos_article
+# plos.csl can be download and used locally
+csl: http://www.zotero.org/styles/plos
 ---
@@
 library(geomtextpath) # Curved Text in 'ggplot2'
+library(sfheaders)
 library(ggpmisc) # Miscellaneous Extensions to 'ggplot2'
@@
 library(scales) # Scale Functions for Visualization
-library(TAM)
+library(TAM) #Test Analysis Modules
 library(officer)
 library(ggnewscale)
-library(bookdown)
+library(cowplot)
 ```
@@
                        y = c(8.0, 8.0, 8.0, 2.5, 2.5, 2.5, 4.5, 4.5, 4.5),
-                       xend = c(0.5, 0.5, 6.5, 0.5, 0.5, 6.5, 0.5, 0.5, 6.5),
-                       yend = c(7.5, 2.5, 2.5, 7.5, 2.5, 2.5, 7.5, 2.5, 2.5),
+                       xend = c(0.8, 0.4, 6.3, 0.4, 0.8, 6.3, 0.8, 0.7, 6.5),
+                       yend = c(7.7, 2.8, 2.7, 7.2, 2.5, 2.5, 7.5, 2.7, 2.7),
                        group = c("A", "A", "A", "B", "B", "B", "C", "C", "C"),
@@
                    group = group),
+             arrow = arrow(length = unit(0.03, "npc")),
              curvature = 0.20,
@@
 ```
-```{r merging-borders-for-plotting}
+```{r merging-borders-for-plotting, warning=FALSE}
 Madrid_muni_border <- Madrid_muni_border %>% st_as_sf() %>% transmute(Border = "Madrid municipal border") %>% rename( "geometry" = "x")
@@
                  Madrid_muni_border)
+
+#to clean up the slivers
+borders <- borders |> st_make_valid() |>
+  group_by(Border) |>
+  summarise() |>
+  sf_remove_holes()
 ```
@@
 ```
+
+```{r, eval=FALSE}
+# here we example the relationship between v_im between modes -- I can't tell anything conclusive.
+car <- SA_im_zn208 %>% filter(mode=="car/motor") %>% st_drop_geometry() %>% select(VORI_CD_ZT208,v_im) %>% 
+  rename("v_im_car" = "v_im")
+bike <- SA_im_zn208 %>% filter(mode=="bike") %>% st_drop_geometry() %>% select(VORI_CD_ZT208,v_im) %>% 
+  rename("v_im_bike" = "v_im")
+transit <- SA_im_zn208 %>% filter(mode=="transit") %>% st_drop_geometry() %>% select(VORI_CD_ZT208,v_im) %>% 
+  rename("v_im_transit" = "v_im") 
+walk <- SA_im_zn208 %>% filter(mode=="walk")%>% st_drop_geometry() %>% select(VORI_CD_ZT208,v_im) %>% 
+  rename("v_im_walk" = "v_im")
+
+all_modes_v_im <- merge(x=car, y=bike, by = "VORI_CD_ZT208", all = T) %>% 
+  merge(transit, by = "VORI_CD_ZT208", all=T)%>%
+  merge(walk, by = "VORI_CD_ZT208", all = T)
+
+all_modes_v_im_plot <- ggplot(all_modes_v_im,
+       aes(x=v_im_transit,
+           y = v_im_walk)) +
+  geom_line()  + geom_point()+
+  theme_classic()
+```
 ```{r, eval=FALSE}
@@
 ```{r, eval=FALSE}
-# here we example the relationship between v_im between modes -- I can't tell anything conclusive.
-car <- SA_im_zn208 %>% filter(mode=="car/motor") %>% st_drop_geometry() %>% select(VORI_CD_ZT208,v_im) %>% 
-  rename("v_im_car" = "v_im")
-bike <- SA_im_zn208 %>% filter(mode=="bike") %>% st_drop_geometry() %>% select(VORI_CD_ZT208,v_im) %>% 
-  rename("v_im_bike" = "v_im")
-transit <- SA_im_zn208 %>% filter(mode=="transit") %>% st_drop_geometry() %>% select(VORI_CD_ZT208,v_im) %>% 
-  rename("v_im_transit" = "v_im") 
-walk <- SA_im_zn208 %>% filter(mode=="walk")%>% st_drop_geometry() %>% select(VORI_CD_ZT208,v_im) %>% 
-  rename("v_im_walk" = "v_im")
-
-all_modes_v_im <- merge(x=car, y=bike, by = "VORI_CD_ZT208", all = T) %>% 
-  merge(transit, by = "VORI_CD_ZT208", all=T)%>%
-  merge(walk, by = "VORI_CD_ZT208", all = T)
-
-all_modes_v_im_plot <- ggplot(all_modes_v_im,
-       aes(x=v_im_transit,
-           y = v_im_walk)) +
-  geom_line()  + geom_point()+
-  theme_classic()
+#Check that the totals are maintained, it is!:
+sum(OD_SA_zn208$V_ij, na.rm = T) #847574 jobs spatially avaialble to the region -- 
+OD_SA_zn208 %>% group_by(VDES_CD_ZT208) %>% summarize(jobs = mean(jobs)) %>% select(jobs) %>% sum()
+
+sum(OD_SA_zn208$tt)
+sum(OD_SA_zn208$total_ttworkers)
+
+#for comparision, the number of travel minutes spent going to jobs.
+group by jobs zona, and the the number of potential minutes traveling to jobs.
+This sum is the number of minutes spent commuting to jobs.
+it is 33680009 minutes, this is approx equal to 43min (mean travel time) *847574 (the total number of workers).
+OD_SA_zn208 %>% group_by(VDES_CD_ZT208) %>% 
+  summarize(jobs_tt_f_t = sum(jobs_tt_f_t)) %>% 
+  select(jobs_tt_f_t) %>% sum() #should equal 33680009 mins total
+```
+
+```{r}
+#Maybe delete this chunk??
+
+# Spatial availability by mode, we can see that car mode captures ~48% of the spatial availability but only accounts for ~36% of the tt_ijm !!:
+modal_V_comps <- OD_SA_zn208 %>%
+  group_by(mode) %>%
+  summarize(workers_m = sum(workers),
+            V_m = sum(V_ij),
+            V_m_perc = sum(V_ij)/847574,
+            jobs_tt_f_t = sum(jobs_tt_f_t)/33680009)
+#modal_V_comps #this issue here is workers should equal V_m in our case, make it do it.
+```
+
+```{r}
+#data for the following plot
+modal_V_comps_4plot <- SA_im_zn208 %>% st_drop_geometry() %>%
+  group_by(mode) %>% 
+  summarize(workers_m = sum(population_m),
+            v_m = mean(v_im),
+            V_m = sum(V_im) ,
+            V_m_perc = sum(V_im)/847574,
+            workers_m_perc = sum(population_m)/847574)
+
+modal_V_comps_4plot <- data.frame(Mode = c("Bike", "Walk", "Car/motor", "Transit",
+                                           "Bike", "Walk", "Car/motor", "Transit"),
+                                  perc = c(modal_V_comps_4plot$workers_m_perc, modal_V_comps_4plot$V_m_perc),
+                                  Measure = c(1,1,1,1,2,2,2,2) ) #for labeling the y-axis
+
+CENTRO_comps <- SA_im_zn208 %>% st_drop_geometry() %>% filter(VORI_CD_ZT208 <= 3) %>%
+  group_by(mode) %>% 
+  summarize(workers_m = sum(population_m),
+            V_m = sum(V_im)) %>%
+  transmute(mode,
+    CENTRO_V_m_perc = V_m/sum(V_m),
+    CENTRO_workers_m_perc = workers_m/sum(workers_m))
+
+CENTRO_comps <- data.frame(Mode = c("Bike", "Walk", "Car/motor", "Transit",
+                                           "Bike", "Walk", "Car/motor", "Transit"),
+                                  perc = c(CENTRO_comps$CENTRO_workers_m_perc, CENTRO_comps$CENTRO_V_m_perc),
+                                  Measure = c(5,5,5,5,6,6,6,6) ) #for labeling the y-axis
+
+M30_comps <- SA_im_zn208 %>% st_drop_geometry() %>% filter(VORI_CD_ZT208 >= 4 & VORI_CD_ZT208 <= 20) %>%
+  group_by(mode) %>% 
+  summarize(workers_m = sum(population_m),
+            V_m = sum(V_im)) %>%
+  transmute(mode,
+    M30_V_m_perc = V_m/sum(V_m),
+    M30_workers_m_perc = workers_m/sum(workers_m))
+
+M30_comps <- data.frame(Mode = c("Bike", "Walk", "Car/motor", "Transit",
+                                           "Bike", "Walk", "Car/motor", "Transit"),
+                                  perc = c(M30_comps$M30_workers_m_perc, M30_comps$M30_V_m_perc),
+                                  Measure = c(3,3,3,3,4,4,4,4) )
+
+modal_V_comps_4plot <- rbind(modal_V_comps_4plot,M30_comps,CENTRO_comps)
+```
+
+```{r}
+#comparision plots of spatial availability vs. population by mode for the full region, just m-30, and just centro
+comps_plot <- ggplot(data = modal_V_comps_4plot,
+       aes(y=perc, x=Measure, fill=Mode, label= perc %>% percent(accuracy = 0.1) )) +
+  scale_x_discrete(limits = c("Population by Mode\n(Total: 847,574)","Spatial Availability\n(Total: 847,574)",
+                              "Population by Mode\n(Total: 243,472)","Spatial Availability\n(Total: 217,219)",
+                              "Population by Mode\n(Total: 44,891)","Spatial Availability\n(Total: 38,942)"),
+                   position = "bottom") +
+  geom_col() +
+  scale_fill_manual(values = c("lightgreen","sienna", "coral", "skyblue"))+
+  geom_text(position = position_stack(vjust=0.5), color="black")+
+  theme_classic()+
+  theme(axis.ticks = element_blank(),
+        axis.title = element_blank(),
+        axis.line = element_blank(),
+        axis.text.y = element_blank(),
+        axis.text = element_text(color="black"),
+        legend.position = "bottom",
+        legend.title = element_blank()) +
+  geom_vline(xintercept=c(0.5,2.5, 4.5, 6.5),
+                color = "black", size=1)
+```
+
+```{r}
+ALL_Plot <- ggplot() +
+    geom_sf(data = borders,
+          size = 0.5,
+          aes(col = Border),
+          fill= NA) +
+  scale_color_manual(values = c( "red", "cyan", "black"),
+                     name = element_blank())+
+  geom_sf(data = borders %>% filter(Border == "Madrid municipal border"),
+          size = 0.5,
+          col ="black",
+          fill="black") +
+  theme_void() +
+  ggtitle("All") +
+  theme(plot.title = element_text(hjust = 0.5),
+        legend.position = "none")
+
+M30_Plot <- ggplot() +
+    geom_sf(data = borders,
+          size = 0.5,
+          aes(col = Border),
+          fill= NA) +
+  scale_color_manual(values = c( "red", "cyan", "black"),
+                     name = element_blank())+
+  geom_sf(data = borders %>% filter(Border == "LEZ M-30"),
+          size = 0.5,
+          col ="cyan",
+          fill="cyan") +
+  theme_void() +
+  ggtitle("M-30") +
+  theme(plot.title = element_text(hjust = 0.5),
+        legend.position = "none")
+
+CENTRO_Plot <- ggplot() +
+    geom_sf(data = borders,
+          size = 0.5,
+          aes(col = Border),
+          fill= NA) +
+  scale_color_manual(values = c( "red", "cyan", "black"),
+                     name = element_blank())+
+  geom_sf(data = borders %>% filter(Border == "LEZ centro"),
+          size = 0.5,
+          col ="red",
+          fill="red") +
+  theme_void() +
+  ggtitle("Centro") +
+  theme(plot.title = element_text(hjust = 0.5),
+        legend.position = "none")
+
+modal_V_comps_4plot_plot <- gridExtra::arrangeGrob(
+                        #labels_4plot,
+  gridExtra::arrangeGrob(ALL_Plot, M30_Plot, CENTRO_Plot, ncol=3),
+                        comps_plot,
+                        nrow=2, heights = c(1,2.5) )
+ggsave("images/modal_V_comps_4plot.png", modal_V_comps_4plot_plot)
 ```
@@
 We first illustrate the features of spatial availability that lends itself to multimodal analysis.
-We then demonstrate its use on the case study of Low Emission Zones in Madrid (Spain) and highlight how this policy intervention changes the accessibility of populations using different modes.
+We then demonstrate its use on the case study of Low Emission Zones in Madrid (Spain) and highlight how this policy intervention changes the accessibility for locations where populations use different modes.
 In summary, spatial availability can be used to create and interpret multimodal policy intervention scenarios unlike previous methods: this creation and interpretation can help regions envision a more sustainable and equitable access-to-opportunity landscape by better identifying differences in accessibility afforded by different modes.
@@
 However, a limitation of this study, like others that implement accessibility measures, is they do not calculate results under a _constrained_ framework i.e., one of _finity_.
-The citizens of Columbus should experience qualitative accessibility gains - but is it at the expense of access to opportunities to populations using other modes?
-<!-- As another example, @mohriClusteringMethodMeasuring2021 implements a modified cumulative opportunity measure to assess differences between private vehicle and transit system accessibility to jobs in Melbourne, but a similar question remains, does the accessibility afforded to the private vehicle using population come at the expense of accessibility losses to transit users?
--->
-
-The work of @leeMeasuringImpactsNew2018 uses _non-competitive_ accessibility measures.
+The citizens of Columbus should experience quantitative accessibility gains - but is it at the expense of access using other modes?
+As another example, @mohriClusteringMethodMeasuring2021 implements a modified cumulative opportunity measure to assess differences between private vehicle and transit system accessibility to jobs in Melbourne, but a similar question remains: does the accessibility afforded to the private vehicle using population come at the expense of accessibility losses to transit users?
+
+@leeMeasuringImpactsNew2018 and @mohriClusteringMethodMeasuring2021 both use _non-competitive_ accessibility measures.
 There is a branch of location-based accessibility measures that do incorporate the effect of competition for opportunities by the population in the region.
@@
 This restriction allows the population using other more sustainable modes to potentially have a higher level of access than before the LEZ implementation.
-This evaluation is especially urgent as LEZ are currently in effect in cities globally; their reception has been mixed [@tarrinoortizPublicAcceptabilityLow2021] and may be having negative impacts on disadvantaged populations who have become mobility restricted [@devrijNooneVisitsMe2022; @verbeekJustManagementUrban2022].
+This evaluation is especially urgent as LEZ are currently in effect in cities globally; their reception has been mixed [@tarrinoortizPublicAcceptabilityLow2021] and may be having negative impacts on disadvantaged populations who have become mobility-restricted [@devrijNooneVisitsMe2022; @verbeekJustManagementUrban2022].
 Measures that evaluate the accessibility of modes given both _constrained_ and _competitive_ considerations are lacking in the literature, but are needed, to evaluate such policy interventions impact on accessibility.
@@
 These measures weight the opportunities that can be potentially interacted with from origin $i$ to destination $j$ based on some sort of travel cost function (e.g., travel time, fare, travel distance) otherwise known as a travel impedance function $f^{m}(c^m_{ij})$.
-Many weighted cumulative opportunities (often refered to as the gravity-based measure) originate from the measure proposed by @hansenHowAccessibilityShapes1959, which can take the following multimodal form: $S_i^m = \sum_j O_j f^m(c_{ij}^m)$ where $m$ is a set of modes which have mode-specific travel costs $(c_{ij}^m)$ and travel impedance functions $f^m(\cdot)$.
+Many weighted cumulative opportunities (often referred to as the gravity-based measure) originate from the measure proposed by @hansenHowAccessibilityShapes1959, which can take the following multimodal form: $S_i^m = \sum_j O_j f^m(c_{ij}^m)$ where $m$ is a set of modes which have mode-specific travel costs $(c_{ij}^m)$ and travel impedance functions $f^m(\cdot)$.
 
@@
 
-In brief, we define the _spatial availability_ at $i$ ( $V_{i}$ ) as the proportion of all opportunities in the region $O$ that are allocated to location $i$ from all opportunity destinations $j$.
+In brief, we define the _spatial availability_ at $i$ ($V_{i}$) as the proportion of all opportunities in the region $O$ that are allocated to location $i$ from all opportunity destinations $j$.
 $V_{i}$ is a value of how many opportunities are available to each location $i$ out of all the opportunities in the region.
@@
 
-Spatial availability measure is introduced in @soukhovIntroducingSpatialAvailability2023.
-The unique feature in the measure is the balancing factor $F^t_{ij}$, a proportional allocation mechanisms, that ensures $V_i$ calculated for each $i$ sums to the total number of opportunities.
-Through $F^t_{ij}$, spatial availability is a  _competitive_ and _constrained_ accessibility measure that handles the number of opportunities in the region in a finite way.
+The spatial availability measure is introduced in @soukhovIntroducingSpatialAvailability2023.
+Spatial availability's unique feature is the balancing factor $F^t_{ij}$, a proportional allocation mechanisms, that ensures the $V_i$ calculated for each $i$ sums, across all $i$ in the region, to equal the total number of opportunities in the region.
+As such, spatial availability is a  _competitive_ and _constrained_ accessibility measure as $F^t_{ij}$ handles the number of opportunities in the region in a finite way (proportional allocation).
 $F^t_{ij}$ consists of two components: a population-based balancing factor $F^p_{i} = \frac{P_i}{\sum_i P_i}$ and an impedance-based balancing factor $F^c_{ij} = \frac{F^c_{ij}}{\sum_j F^c_{ij}}$ that, respectively, allocate opportunities to $i$ in proportion to the size of the population at $i$ (the mass effect) and the cost of reaching opportunities at $j$ (the impedance effect).
@@
 When $a_i^m$ is aggregated to $a_i$ as shown in the sixth column, the values face similar interpretability issues.
-
 The Shen-type measure is implemented in the previously discussed work of @taoInvestigatingImpactsPublic2020a to calculate modal $a_i^m$ values and the aggregated $a_i$ is implemented in the work of @carpentieriMultimodalAccessibilityPrimary2020.
@@
 
-<!-- THIS IS GOOD, but cutting to save words for now.
-
+<!-- THIS IS GOOD, but could be cut to save works-->
 Since spatial availability is constrained and has an interpretable meaning as a proportion of the total opportunities in the region, the values at $i$ have a new significance.
@@
 In other words, the $z$-using population travels less minutes overall and has more spatial availability of opportunities than the $x$-using population using the slower mode $x$.
--->
+
 Alternatively, inequities in spatial availability between mode-using populations can be explored through proportional benchmarks.
@@
 
-<!-- The benchmarks for $x$-using populations are evidently lower than the aggregated $A$, $B$, and $C$ $v_i$ values of (`r V_i %>% group_by(i) %>% summarize(v_i = sum(V_im)/sum(pop)) %>% filter(i=="A") %>% select(v_i) %>% sum() %>% round(digits=2)`, `r V_i %>% group_by(i) %>% summarize(v_i = sum(V_im)/sum(pop)) %>% filter(i=="B") %>% select(v_i) %>% sum() %>% round(digits=2)` and `r V_i %>% group_by(i) %>% summarize(v_i = sum(V_im)/sum(pop)) %>% filter(i=="C") %>% select(v_i) %>% sum() %>% round(digits=2)`).-->
-
 # Empirical example: Madrid LEZ
@@
 
+```{r, eval=FALSE, messages=FALSE}
+#not correlated, but this demonstrates that there's a higher positive linear trend for zones with pops of bike/walk with pops of transit than for pops of car
+
+corr_m_pops <- data.frame(VORI_CD_ZT208 = SA_im_zn208$VORI_CD_ZT208 %>% unique())
+
+corr_m_pops <- merge(corr_m_pops, SA_im_zn208 %>% st_drop_geometry() %>% filter(mode == "bike") %>% select(c(VORI_CD_ZT208, population_m)), by="VORI_CD_ZT208", all = TRUE) %>% rename("population_m_bike" = "population_m")
+
+corr_m_pops <- merge(corr_m_pops, SA_im_zn208 %>% st_drop_geometry() %>% filter(mode == "transit") %>% select(c(VORI_CD_ZT208, population_m)), by="VORI_CD_ZT208", all = TRUE) %>% rename("population_m_transit" = "population_m")
+
+corr_m_pops <- merge(corr_m_pops, SA_im_zn208 %>% st_drop_geometry() %>% filter(mode == "walk") %>% select(c(VORI_CD_ZT208, population_m)), by="VORI_CD_ZT208", all = TRUE) %>% rename("population_m_walk" = "population_m")
+
+corr_m_pops <- merge(corr_m_pops, SA_im_zn208 %>% st_drop_geometry() %>% filter(mode == "car/motor") %>% select(c(VORI_CD_ZT208, population_m)), by="VORI_CD_ZT208", all = TRUE) %>% rename("population_m_car" = "population_m")
+
+#shapiro.test(corr_m_pops$population_m_car) #all pop_ms have p values under 0.05, indicating that they are not statistically different from the null which is that they are something other than normally distirbuted.
+thus wecannot run the correlation.
+#shapiro.test(corr_m_pops$population_m_transit)
+
+ggplot(corr_m_pops, aes(x=population_m_walk, y=population_m_car)) + 
+  geom_point() +
+  geom_smooth(method = "lm")
+ggplot(corr_m_pops, aes(x=population_m_walk, y=population_m_transit)) + 
+  geom_point() +
+  geom_smooth(method = "lm")
+
+ggplot(corr_m_pops, aes(x=population_m_bike, y=population_m_car)) + 
+  geom_point() +
+  geom_smooth(method = "lm")
+ggplot(corr_m_pops, aes(x=population_m_bike, y=population_m_transit)) + 
+  geom_point() +
+  geom_smooth(method = "lm")
+```
+
 <!-- a summary about low emission zones, they are a geographic accessibility restriction. they change the accessibility landscape-->
-
 Low emission zones (LEZ) have been implemented as a climate change policy intervention to reduce GHG emissions, improve air quality, and support sustainable mobility in many countries.
 Though rules vary<!-- depending on legal aspects and cultural norms-->, LEZ aim to deter/reduce traffic in designated zones under threat of penalty (e.g., fines, seizure of vehicle).
-<!-- In practice, LEZ limits the volume of traffic by excluding vehicles by license plate, fuel type, or by introducing tolls.--> From the perspective of restriction for passenger transport, LEZ are a policy of _geographic discrimination_.
-LEZ actively change how people access opportunities by making the travel impedance more costly for car-mode users.
+<!-- In practice, LEZ limits the volume of traffic by excluding vehicles by license plate, fuel type, or by introducing tolls.--> From the perspective of restriction for passenger transport, LEZ are a policy of _geographic discrimination_ as they change how people access opportunities by making the travel impedance more costly for car-mode users.
 If seeing opportunities as finite, this discrimination allows populations to access opportunities by other modes more readily than before.
@@
 
-In 2017, a LEZ was implemented in the Spanish capital city of Madrid following the goals set out in the national agenda <!-- : to fight climate change, cut nitrogen dioxide levels for the benefit of people's health, and prioritize people's movement in the city-->.
+In 2017, LEZs were implemented in the Spanish capital city of Madrid following the goals set out in the national agenda <!-- : to fight climate change, cut nitrogen dioxide levels for the benefit of people's health, and prioritize people's movement in the city-->.
 In geographic scope, the 2017 boundaries of the LEZ were relatively small (covering 4.72 km^{2}) and within the center (i.e., LEZ Centro).
 These boundaries were expanded in 2023 to inside of the M-30, a highway in proximity to the city center (i.e., LEZ M-30) and the city has plans to further spatially expand the LEZ.
-Within the 2017 LEZ Centro implementation, all cars, motorcycles and freight with environmental label A or B (higher polluting classification, associated with older make and model of fossil fuel internal combustion engine vehicles), are banned from driving into the area (threat of fine) unless they are used by residents or meet other exemptions.
+Within the 2017 LEZ Centro implementation, all cars, motorcycles and freight with environmental label A or B (higher polluting classification, associated with older make and model of fossil fuel internal combustion engine vehicles), are not permitted to enter the area unless they are used by residents or meet other exemptions.
 This restriction impacted approximately half of all car trips that were typically made into the LEZ Centro [@tarrinoortizAnalyzingImpactLow2022].
 
-In this section, we use $V_i^m$ to quantify the competition for spatially available opportunities between modes after the LEZ Centro implementation.
+For this case study, we use $V_i^m$ to quantify the competition of spatially available opportunities between modes after the LEZ Centro implementation.
 Particularly, we demonstrate how $V_i^m$ can be used to spectate on how the restriction of car mobility in areas around/within the LEZ Centro allowed the other, more sustainable but often with higher travel impedance modes, to become more competitive.
@@
    - Car/motor: all cars and operating modes (e.g., cab, private driver, company, rental care, main driver, passenger, etc.) and all public, private or company motorcycle/mopeds.
-  - Transit: all bus, trams, and trains
-  - Bike: all bicycle trips (e.g., private, public, or company bike trips) and "other" types of micromobility options
-  - Walk: walking or by foot
+  - Transit: all bus, trams, and trains,
+  - Bike: all bicycle trips (e.g., private, public, or company bike trips) and "other" types of micromobility options,
+  - Walk: walking or by foot,
 
@@
 
-From Figure \ref{fig:Fig2}, it can also be seen that biking and walking trips are less common than motorized trips at `r (pop_props$population_m[1]/sum(pop_props$population_m)) %>% percent()` and `r (pop_props$population_m[2]/sum(pop_props$population_m)) %>% percent()` respectively.
-The distribution of walking and biking trips appear to be similar to that of transit trips.
-<!-- This is to be expected as active transport and a higher diversity of land-use spatially occurs with transit infrastructure.
-<!--CITE-->
+From Figure \ref{fig:Fig3}, it can also be seen that biking and walking trips are less common than motorized trips at `r (pop_props$population_m[1]/sum(pop_props$population_m)) %>% percent()` and `r (pop_props$population_m[2]/sum(pop_props$population_m)) %>% percent()` respectively.
+Noteably, there is a positive trend between the populations of walking and biking trips in zones and populations of transit trips.
+This positive trend is higher than for car trip populations.
+<!-- the populations do not follow a normal distribution, so we can't say there is a correlation, formally.
+see calcs in chunk at the beginning of this subsection-->
 
@@
 A TLD represents the proportion of trips that are taken at a specific travel cost such as travel time (i.e., probability density distribution of trips taken by travel cost).
-This distribution is then used to derive impedance functions [e.g., done in the accessibility works of @lopez_2017_spatial, @horbachov_theoretical_2018, and @batista_estimation_2019 for example].
+This distribution is then used to derive impedance functions (e.g., done in the accessibility works of @lopez_2017_spatial, @horbachov_theoretical_2018, and @batista_estimation_2019).
 Maximum likelihood estimation and the Nelder-Mead method for direct optimization available within the R {fitdistrplus} package [@fitdistrplus_2015] is used to fit the impedance functions.
 As shown as shown in Figure \ref{fig:Fig4}, based on goodness-of-fit criteria and associated diagnostics, the gamma and log-normal probability density function (line curves) are selected as best fitting curves for the motorized and non-motorized modes respectively.
-The selection of functional form aligns with examples used in the literature (e.g., @reggianiAccessibilityImpedanceForms2011).
+The selection of functional form aligns with empirical examples in other regions [@reggianiAccessibilityImpedanceForms2011; @soukhovTTS2016RDataSet2023; @soukhovIntroducingSpatialAvailability2023].
 <!--e.g., car gamma XX and active transport log-normal XX.--> Overall, the plots in Figure \ref{fig:Fig4} display the probability of travel given a trip travel time, based on trip flows from the survey.
-These 'probability of travel' at each travel time for each mode are realized observations reflect the land-use, the transport system, and the population travel behaviour in Madrid.
+These 'probability of travel' at each travel time for each mode are realized observations that reflect the land-use, the transport system, and the population travel behaviour in Madrid.
 
@@
 ## Results
-Using the data inputs outlined, $V_i^m$, the spatial availability of jobs, is calculated for each of the four modal categories $m$ at the level of traffic analysis zones $i$ in Madrid and demonstrated in this sub-section.
+The spatial availability of jobs $V_i^m$ is calculated for each of the four modal categories $m$ at the level of traffic analysis zones $i$ in Madrid and demonstrated in Figure \ref{fig:Fig5}.
 
@@
 ```
-```{r, eval=FALSE}
-#Check that the totals are maintained, it is!:
-sum(OD_SA_zn208$V_ij, na.rm = T) #847574 workers -- 
-OD_SA_zn208 %>% group_by(VDES_CD_ZT208) %>% summarize(jobs = mean(jobs)) %>% select(jobs) %>% sum()
-
-sum(OD_SA_zn208$tt)
-sum(OD_SA_zn208$total_ttworkers)
-
-#for comparision, the number of travel minutes spent going to jobs.
-group by jobs zona, and the the number of potential minutes traveling to jobs.
-This sum is the number of minutes spent commuting to jobs.
-it is 33680009 minutes, this is approx equal to 43min (mean travel time) *847574 (the total number of workers).
-
-OD_SA_zn208 %>% group_by(VDES_CD_ZT208) %>% 
-  summarize(jobs_tt_f_t = sum(jobs_tt_f_t)) %>% 
-  select(jobs_tt_f_t) %>% sum() #should equal 33680009 mins total
-```
-
-```{r}
-# Spatial availability by mode, we can see that car mode captures ~48% of the spatial availability but only accounts for ~36% of the tt_ijm !!:
-modal_V_comps <- OD_SA_zn208 %>%
-  group_by(mode) %>%
-  summarize(V_m = sum(V_ij),
-            V_m_perc = sum(V_ij)/847574,
-            jobs_tt_f_t = sum(jobs_tt_f_t)/33680009)
-#modal_V_comps
-```
-
-Figure \ref{fig:Fig5} displays $V_i^m$ for the four modal categories in 2018.
-$V_i^m$ values represent a proportion of the total number of the `r SA_i_zn208$V_i %>% sum()  %>% prettyNum(big.mark = ",")` jobs in the region; put another way, $V_i^m$ is how many jobs are _spatially available_ to the population located at the $i$ based on the travel impedance of the mode (relative to the travel impedance of all modes) and the mode-using population size at the $i$ (relative to the population size of all modes and $i$).
-<!-- Since population, their job locations, and associated travel times are used to calculate $V_i^m$, Figure \ref{fig:Fig5} demonstrates values that represent the _realized_ spatial availability of jobs in Madrid.
-In other words, if someone decided to move into a neighbourhood at an $i$, the amount of jobs that are spatially available to them based on the population within their $i$ and their travel impedance for their used $m$, both relative to the city, is captured by the value of $V_i^m$.--> Note the differences in the magnitudes of $V_i^m$ between modes.
+
+$V_i^m$ is a proportion of the total number of the `r SA_i_zn208$V_i %>% sum()  %>% prettyNum(big.mark = ",")` jobs in the region and is visualized in Figure \ref{fig:Fig5}.
+Since $V_i^m$ is calculated based on the likelihood of travel from observed home-to-work journeys, the values can be understood as the number of full-time jobs that are spatially available to the full-time working population at that $i$ and their associated $m$, relative to all the jobs in the city.
+$V_i^m$ is the number of jobs that are _spatially available_ to a $m$-using population located at $i$, relative to the travel impedance and size of _all_ populations in the region.
+
+Notable are the differences in the magnitude of $V_i^m$ between modes in Figure \ref{fig:Fig5}.
 The majority of $V_i^m$ is allocated to car- and transit- using populations.
-This is to be expected, as commuting using motorized modes represents `r (pop_props$population_m[3]/sum(pop_props$population_m) + pop_props$population_m[4]/sum(pop_props$population_m)) %>% percent()` of the population (`r (pop_props$population_m[3]/sum(pop_props$population_m)) %>% percent()` (car/motor) and `r (pop_props$population_m[4]/sum(pop_props$population_m)) %>% percent()` (transit) in 2018).
-However, these modal options capture `r (modal_V_comps$V_m_perc[3] + modal_V_comps$V_m_perc[4]) %>% percent()` of the total spatial availability in Madrid.
-In particular, the car/motor using population is allocated disproportionately more $V_i^m$ than its modal population (`r (pop_props$population_m[3]/sum(pop_props$population_m)) %>% percent()` of the population vs.
-`r modal_V_comps$V_m_perc[3] %>% percent()` of the $V_i^m$) compare to the transit using population and its relatively proportional $V_i^m$ value (`r (pop_props$population_m[2]/sum(pop_props$population_m)) %>% percent()` of the population vs.
-`r modal_V_comps$V_m_perc[4] %>% percent()` $V_i^m$).
-
-How does the $V_i^m$ advantage allocated to car-using population arise?
-From the perspective of finite opportunities, $V_i^m$ is allocated to car-using populations from less competitive modal populations.
-How competitive one mode is compared to other modes varies spatially, but overall car-using populations capture more opportunities per car-using population than other modal populations.
-Namely, though walking and cycling populations represent `r (pop_props$population_m[2]/sum(pop_props$population_m)) %>% percent(accuracy = 0.01)` and `r (pop_props$population_m[1]/sum(pop_props$population_m)) %>% percent(accuracy = 0.01)` respectively, $V_i^m={walk}$ and $V_i^m={bike}$ is `r modal_V_comps$V_m_perc[2] %>% percent(accuracy = 0.01)` and `r modal_V_comps$V_m_perc[1] %>% percent(accuracy = 0.01)` in the region respectively.
-These modes are less competitive, especially compared to the car/motor mode, as a result of: 1) their lower travel impedance values at longer travel times (see Figure \ref{fig:Fig4} at travel times beyond ~30 minutes), 2) their low population values values overall, and 3) higher populations present in origins with high motorized mode commuting.
+This is to be expected, as the population that commutes using these modes represents `r (modal_V_comps_4plot$perc[3] + modal_V_comps_4plot$perc[4]) %>% percent(accuracy=0.1)` of the total population.
+Differences in $V_i^m$ values within mode-using populations also exist: car-using populations outside of the M-30 region appear to have greater $V_i^m$ values, while some $i$ areas inside the M-30 appear to have higher $V_i^m$ values for the transit-using populations.
+Overall, the magnitude of $V_i^m$ values for the bikers and walkers are lower than car and transit but the highest $V_i^{bike}$ and $V_i^{walk}$ values tend to be allotted to $i$s within the M-30 and $i$s that have higher $V_i^{transit}$ values.
+
+The differences between the mode-using population and their mode-specific spatial availability highlights the competitive advantage offered to certain modes in certain spatial extents.
+As summarized in the left-most columns in Figure \ref{fig:Fig6}, the 'car/motor' and 'transit' populations represent a combined `r (modal_V_comps_4plot$perc[7] + modal_V_comps_4plot$perc[8]) %>% percent(accuracy=0.1)` of the total spatial availability in the city.
+However, the 'car/motor' using population is allocated disproportionately more $V_i^m$ than its size compared to the transit-using population.
+The car-using and transit-using population is `r modal_V_comps_4plot$perc[3] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[4] %>% percent(accuracy=0.1)` respectively, but is allocated `r modal_V_comps_4plot$perc[7] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[8] %>% percent(accuracy=0.1)` respectively, of the city's spatial availability.
+When treating the number of opportunities that can be reached as a finite value (total: 847,574 opportunities), fewer opportunities are spatial availability to the lesser competitive modes-using populations, in this case walking and cycling.
+These modes are less competitive as a result of: their lower travel impedance values at longer travel times (see Figure \ref{fig:Fig4} at travel times beyond ~30 minutes); their low population values values overall; and higher populations present in origins with high motorized mode commuting.
 These factors all contribute to the the car/motor mode being most advantaged in capturing spatially available job opportunities overall.
-
-```{r, SA-per-capita-m-plot, fig.cap="\\label{fig:Fig6} Spatial availability of job opportunities per capita per origin and mode $v_i^m$ in Madrid.
+```{r, modal-V-comps-plot, fig.cap="\\label{fig:Fig6} Displays the proportion of the working population by mode and spatial availability of job opportunities by mode aggregated for three spatial areas.
+From left to right, the city of Madrid (All), the area within the M-30 highway (M-30),the area within the Centro region (Centro).", fig.show='hold', fig.align='center', fig.height=10}
+knitr::include_graphics("images/modal_V_comps_4plot.png")
+```
+
+There are spatial variations in the competitive advantage of the car-using populations.
+The proportion of car-using population in the Centro is smaller and has higher travel impedance values relative to the inputs in other areas and mode-using populations.
+The LEZ Centro implementation further restricts the car-advantage as it shifted more than half of all car trips into the LEZ to another mode [@tarrinoortizAnalyzingImpactLow2022].
+This restriction decreased the number of car-using population from $i$s going into the LEZ Centro (an area with a large number of jobs overall, see Figure \ref{fig:Fig2}), thus increasing the mass effect for non-car modes and resulting in proportionally higher $v_i^m$ values for non-car modes.
+As such, the lower amount of access to opportunities by car-mode allows more opportunities in the LEZ to be available by populations using other modes.
+
+As summarized in the two right-most columns in Figure \ref{fig:Fig6}, the proportion of spatial availability allocated to the car-using population in the Centro (`r modal_V_comps_4plot$perc[23] %>% percent(accuracy=0.1)` or 5,373 opportunities).
+As a comparative reference, this is less than the proportion of the car-using population in the Centro (`r modal_V_comps_4plot$perc[19] %>% percent(accuracy=0.1)`), evidently less then the proportion of car-using population in the city, and is the opposite of the trend overall (left-most columns) and within the M-30 (middle columns).
+More opportunities are spatial availability to non-car using populations within the Centro, particularly transit-using populations (`r modal_V_comps_4plot$perc[24] %>% percent(accuracy=0.1)` of spatially available jobs in the Centro despite representing `r modal_V_comps_4plot$perc[20] %>% percent(accuracy=0.1)` of the population in the Centro and `r modal_V_comps_4plot$perc[4] %>% percent(accuracy=0.1)` in the city overall).
+
+From Figure \ref{fig:Fig6}, it is also summarized that there is a higher proportion of opportunities spatially available to walking and cycling populations in the Centro than in the City overall and in all areas within the M-30.
+Notably, within the Centro, `r modal_V_comps_4plot$perc[21] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[22] %>% percent(accuracy=0.1)` of opportunities are spatially available to bike and walk modes respectively, while their populations represent smaller proportions of `r modal_V_comps_4plot$perc[1] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[2] %>% percent(accuracy=0.1)` of the population overall.
+Though the proportion of spatial availability for these mode-using populations is still lower than the proportion of mode-using population located in the Centro, these modes are more competitive within the Centro than outside of the Centro.
+By restricting the more competitive car mode through the LEZ, the advantage in the spatial availability of opportunities afforded to the otherwise lesser competitive modes is made apparent.
+
+The spatial differences in the competitive dis/advantage of spatial availability between modes can also be visualized per origin.
+Figure \ref{fig:Fig7} visualizes $v_i^m$, the spatial availability $V_i^m$ divided by the mode-population.
+$v_i^m$ values above 1 are represented in increasing red shades, values below 1 are represented in increasingly green shades, and values equal to 1 are white.
+These plots illustrates the discussion of the disproportionately high over-allocation of spatial availability relative to the mode-using population in many of the origins for the car/motor mode (bottom left plot, areas denoted with green $v_i^m$ values above 1).
+These plots also visualize areas that disproportionately capture lower spatial availability (under 1), represented in shades of red.
+It can be observed that the transit-using population's spatial availability to jobs is relatively balanced (i.e., many zones are white), while the non-motorized modes $v_i^m$ values are low (under 1) overall.
+
+```{r, SA-per-capita-m-plot, fig.cap="\\label{fig:Fig7} Spatial availability of job opportunities per mode-using capita by mode $v_i^m$ per origin in Madrid.
 Calculated using the home-to-work origin destination flows from the 2018 travel survey.", fig.show='hold', fig.align='center', fig.height=10}
@@
 
-Furthermore, there are spatial differences in the competitive advantage of spatial availability between modes.
-Figure \ref{fig:Fig6} visualizes $v_i^m$, the spatial availability divided by the mode-population.
-<!--$v_i^m$ values above 1 are represented in increasing red shades, values below 1 are represented in increasingly green shades, and values equal to 1 are white.--> These plots illustrates the discussion of the disproportionately high over-allocation of spatial availability relative to the mode-using population in many of the origins for the car/motor mode (bottom left plot, areas denoted with green $v_i^m$ values above 1).
-These plots also visualize areas that disproportionately capture lower spatial availability (under 1), represented in shades of red.
-It can be observed that the transit-using population's spatial availability to jobs is relatively balanced (i.e., many zones are white), while the non-motorized modes $v_i^m$ values are low (under 1).
-
-Interestingly, $v_i^m$ for car/motor within and near the LEZ Centro is near or below 1 (white/red): potentially as a consequence, all non-car modes have relatively higher $v_i^m$ values in these areas.
-Though access to travel survey data immediately before the LEZ implementation is unavailable, we know that the LEZ Centro shifted more than half of all car trips into the LEZ to another mode [@tarrinoortizAnalyzingImpactLow2022].
-This restriction decreased the number of car-using population from $i$s going into the LEZ Centro (an area with a large number of jobs - Figure \ref{fig:Fig2}), thus increasing the mass effect for non-car modes and resulting in proportionally higher $v_i^m$ values for non-car modes.
-Though the spatial availability from before the LEZ Centro implementation is unknown, Figure \ref{fig:Fig6} provides a benchmark for quantifying potential LEZ implementations in the future (given 2018 travel conditions).
-Figure \ref{fig:Fig6} also shows that many areas within the M-30 have high (white/green) $v_i^m$ values for car-mode, signaling that the spatial expansion of the LEZ Centro stands to increase the spatial availability of jobs for non-car mode using populations.
+Interestingly, as also represented in Figure \ref{fig:Fig6}, $v_i^m$ for car/motor within and near the LEZ Centro is near or below 1 (white/red) in Figure \ref{fig:Fig7} while all non-car modes have relatively higher $v_i^m$ values.
+Though the spatial availability from before the LEZ Centro implementation is unknown, Figure \ref{fig:Fig7} provides a benchmark for quantifying potential LEZ implementations in the future (given 2018 travel conditions).
+Figure \ref{fig:Fig7} also shows that many areas within the M-30 have high (white/green) $v_i^m$ values for car-mode, signaling that the spatial expansion of the LEZ Centro stands to increase the spatial availability of jobs for non-car mode using populations.
 
@@
 
+ANSWER TO THIS PARAGRAPH, V_i is a location-based measure.
+it tells us that the jobs within the centro can be more easily available to non-motorized modes than motorized modes.
+This is in part because of the LEZ but also because of walkability/land -use, of course.
 These results can be explained in a different/better way..
@@
 # Discussion and conclusions
-Location-based accessibility measures like the Hansen-type $S_i^m$, Shen-type $a_i^m$, and spatial availability $V_i^m$ measures have a commonality - they are a weighted sum of opportunities assigned to each spatial unit $i$ in a region.
-<!-- In this way, they all can be interpreted as a score that represents how many opportunities can be potentially interacted with by the population at $i$.--> How the weight and sum of the potentially-interacted-with opportunities is considered is what defines the type of accessibility measure.
-Within this paper, the location-based singly- _constrained_ and _competitive_ accessibility measure, known as spatial availability $V_i$ (@soukhovIntroducingSpatialAvailability2023), is extended for the case of capturing multimodal accessibility to opportunities $V_i^m$.
+Location-based accessibility measures like the Hansen-type $S_i^m$, Shen-type $a_i^m$, and spatial availability $V_i^m$ measures share a commonality; they are a weighted sum of opportunities assigned to each spatial unit $i$ in a region.
+In this way, they all can be interpreted as a score of how many opportunities can be potentially interacted with by the population at $i$.
+How the weight and sum of the potentially-interacted-with opportunities is considered is what defines the type of accessibility measure.
+
+Within this paper, the location-based singly- _constrained_ and _competitive_ accessibility measure, known as spatial availability $V_i$ [@soukhovIntroducingSpatialAvailability2023], is extended for the case of capturing multimodal accessibility to opportunities $V_i^m$.
 A synthetic example and then an empirical case of LEZ in Madrid are detailed to demonstrate this multimodal extension.
@@
 As showcased in the empirical example of the LEZ in Madrid, competition for job opportunity availability varies spatially _as well as_ between modes.
-The car and transit modes have the highest spatial availability, with the car mode having highest availability with exception to the areas within the LEZ Centro.
-This finding reflects what may suspect are real conditions: since car travel has been highly restricted within the LEZ Centro, fewer car-using people potentially interact with jobs within the LEZ Centro, leaving more _spatially available_ jobs for non-car-using populations.
+The car and transit modes have the highest spatial availability, with the car-mode having highest availability with exception to the areas within the LEZ Centro.
+Since car travel has been highly restricted within the LEZ Centro, fewer car-using people potentially interact with jobs within the LEZ Centro, leaving more _spatially available_ jobs for non-car-using populations.
 This difference in car-using populations in locations accessing jobs within and immediately outside the LEZ Centro increases the competitiveness of the transit-using population (the second most competitive mode) as well as the non-motorized modes.
@@
 Spatial availability $V_i^m$ can also be divided by the mode-using population at each $i$ to yield mode-population normalized values.
-These values, reflected in Figure \ref{fig:Fig6}, can be used as a benchmark to investigate existing conditions and plan future LEZ implementation (i.e., target areas with exceptionally high car spatial availability such that more opportunities are available to other mode-users).
+These values, reflected in Figure \ref{fig:Fig7}, can be used as a benchmark to investigate existing conditions and plan future LEZ implementation (i.e., target areas with exceptionally high car spatial availability such that more opportunities are available to other mode-users).
 
@@
 With spatial availability, the magnitude of opportunities that are available as a proportion of all the opportunities in the region is equal to $V_i$.
-As a result of its proportional allocation mechanism, $V_i$ can be easily extended into multimodal applications.
-This flexibility is pertinent to model policy scenarios in our cities that are becoming increasingly multimodal.
+As a result of its proportional allocation mechanism, $V_i$ can be naturally extended into multimodal applications.
+This flexibility is helpful to modelling policy scenarios in our cities that are increasingly multimodal.
 The interpretation of $V_i$ allows for manipulation of $V_i^m$ values to investigate differences of availability between neighbourhoods, modes, and regions, generate per capita benchmarks, and/or generate average values per population-group.
@@
 This research was funded by the Canada Graduate Scholarship - Doctoral Program (CGS D) provided by the Social Sciences and Humanities Research Council (SSHRC) and Project Mobilizing Justice, also supported by SSHRC.
+All work is fully-reproducible and available within this [GitHub repository](https://github.com/soukhova/Madrid-ZBE-LEZ).
 
@@
 -->
+
+# References {#references .unnumbered}

### 2023-08-24 Journal-manuscript/Journal-manuscript.tex
Submitted this version to journal 1

--- 2023-07-31 TRB-manuscript/TRB-manuscript.tex
@@
-% Repository:  https://github.com/chiehrosswang/TRB_LaTeX_tex
-%
-% Transportation Research Board conference paper template
-% version 4.0 Lite (updates made to be compatible in Overleaf and ShareLaTeX)
-%
-%
-% When numbered option is activated, lines are numbered.
-\documentclass[]{trbunofficial}
-\usepackage{graphicx}
-\usepackage{booktabs}
-
-\newread\somefile
-\usepackage{xparse}
-\usepackage{natbib}
-\bibliographystyle{unsrtnat}
-\setcitestyle{round}
-% \usepackage[colorlinks=true,linkcolor=blue,citecolor=blue]{hyperref}
-% For TRB version hide links
-\usepackage[hidelinks]{hyperref}
-
-% Put here what will go to headers as author
-\AuthorHeaders{}
-\title{Competition for Access to Opportunities between Modes: Multimodal
-Spatial Availability}
-
-% TODO: add macros for easier formatting of \author.
-\author{%
-    \textbf{Anastasia Soukhov}\\\textit{Corresponding Author}\\
-  PhD Candidate\\
-  Department of Earth, Environment and Society, McMaster University,
-Canada\\
-  \href{mailto:soukhoa@mcmaster.ca}{\nolinkurl{soukhoa@mcmaster.ca}}\\
-  \hfill\break
-    \textbf{Javier Tarriño-Ortiz}\\
-  PhD Graduate\\
-  Universidad Politécnica de Madrid, Spain\\
-  \href{mailto:javier.tarrino.ortiz@upm.es}{\nolinkurl{javier.tarrino.ortiz@upm.es}}\\
-  \hfill\break
-    \textbf{Julio A. Soria-Lara}\\
-  Associate Professor\\
-  Urban and Regional Planning, Universidad Politécnica de Madrid,
-Spain\\
-  \href{mailto:julio.soria-lara@upm.es}{\nolinkurl{julio.soria-lara@upm.es}}\\
-  \hfill\break
-    \textbf{Antonio Páez}\\
-  Associate Professor\\
-  Department of Earth, Environment and Society, McMaster University,
-Canada\\
-  \href{mailto:paezha@mcmaster.ca}{\nolinkurl{paezha@mcmaster.ca}}\\
-  \hfill\break
+% Template for PLoS
+% Version 3.5 March 2018
+%
+% % % % % % % % % % % % % % % % % % % % % %
+%
+% -- IMPORTANT NOTE
+%
+% This template contains comments intended
+% to minimize problems and delays during our production
+% process. Please follow the template instructions
+% whenever possible.
+%
+% % % % % % % % % % % % % % % % % % % % % % %
+%
+% Once your paper is accepted for publication,
+% PLEASE REMOVE ALL TRACKED CHANGES in this file
+% and leave only the final text of your manuscript.
+% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
+% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
+%
+%
+% There are no restrictions on package use within the LaTeX files except that
+% no packages listed in the template may be deleted.
+%
+% Please do not include colors or graphics in the text.
+%
+% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
+%
+% % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- FIGURES AND TABLES
+%
+% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
+%
+% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
+% - Figures should be uploaded separately from your manuscript file.
+% - Figures generated using LaTeX should be extracted and removed from the PDF before submission.
+% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
+% For figure citations, please use "Fig" instead of "Figure".
+% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
+%
+% Tables should be cell-based and may not contain:
+% - spacing/line breaks within cells to alter layout or alignment
+% - do not nest tabular environments (no tabular environments within tabular environments)
+% - no graphics or colored text (cell background color/shading OK)
+% See http://journals.plos.org/plosone/s/tables for table guidelines.
+%
+% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
+%
+% IMPORTANT
+% Below are a few tips to help format your equations and other special characters according to our specifications.
+For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
+%
+% For inline equations, please be sure to include all portions of an equation in the math environment.
+%
+% Do not include text that is not math in the math environment.
+%
+% Please add line breaks to long display equations when possible in order to fit size of the column.
+%
+% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
+%
+% When adding superscript or subscripts outside of brackets/braces, please group using {}.
+%
+% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+%
+% Please contact latex@plos.org with any questions.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+
+\documentclass[10pt,letterpaper]{article}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+
+% amsmath and amssymb packages, useful for mathematical formulas and symbols
+\usepackage{amsmath,amssymb}
+
+% Use adjustwidth environment to exceed column width (see example table in text)
+\usepackage{changepage}
+
+% textcomp package and marvosym package for additional characters
+\usepackage{textcomp,marvosym}
+
+% cite package, to clean up citations in the main text. Do not remove.
+% \usepackage{cite}
+
+% Use nameref to cite supporting information files (see Supporting Information section for more info)
+\usepackage{nameref,hyperref}
+
+% line numbers
+\usepackage[right]{lineno}
+
+% ligatures disabled
+\usepackage{microtype}
+\DisableLigatures[f]{encoding = *, family = * }
+
+% color can be used to apply background shading to table cells only
+\usepackage[table]{xcolor}
+
+% array package and thick rules for tables
+\usepackage{array}
+
+% create "+" rule type for thick vertical lines
+\newcolumntype{+}{!{\vrule width 2pt}}
+
+% create \thickcline for thick horizontal lines of variable length
+\newlength\savedwidth
+\newcommand\thickcline[1]{%
+  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+  \cline{#1}%
+  \noalign{\vskip\arrayrulewidth}%
+  \noalign{\global\arrayrulewidth\savedwidth}%
 }
 
-% If necessary modify the number of words per table or figure default is set to
-% 250 words per table (default defined in cls)
-
-% If words are counted manually, put that number here. This does not include
-% figures and tables. This can also be used to avoid problems with texcount
-% program i.e. if one does not have it installed.
-\TotalWords{7195}
+% \thickhline command for thick horizontal lines that span the table
+\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+\hline
+\noalign{\global\arrayrulewidth\savedwidth}}
+
+% Remove comment for double spacing
+%\usepackage{setspace}
+%\doublespacing
+
+% Text layout
+\raggedright
+\setlength{\parindent}{0.5cm}
+\textwidth 5.25in
+\textheight 8.75in
+
+% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
+% Captions will be left justified
+\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
+\renewcommand{\figurename}{Fig}
+
+% Use the PLoS provided BiBTeX style
+% \bibliographystyle{plos2015}
+
+% Remove brackets from numbering in List of References
+\makeatletter
+\renewcommand{\@biblabel}[1]{\quad#1.}
+\makeatother
+
+% Header and Footer with logo
+\usepackage{lastpage,fancyhdr,graphicx}
+\usepackage{epstopdf}
+%\pagestyle{myheadings}
+\pagestyle{fancy}
+\fancyhf{}
+%\setlength{\headheight}{27.023pt}
+%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
+\rfoot{\thepage/\pageref{LastPage}}
+\renewcommand{\headrulewidth}{0pt}
+\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
+\fancyheadoffset[L]{2.25in}
+\fancyfootoffset[L]{2.25in}
+\lfoot{\today}
+
+%% Include all macros below
+
+\newcommand{\lorem}{{\bf LOREM}}
+\newcommand{\ipsum}{{\bf IPSUM}}
 
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+% Pandoc citation processing
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+% for Pandoc 2.8 to 2.10.1
+\newenvironment{cslreferences}%
+  {}%
+  {\par}
+% For Pandoc 2.11+
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 
+\usepackage{forarray}
+\usepackage{xstring}
+\newcommand{\getIndex}[2]{
+  \ForEach{,}{\IfEq{#1}{\thislevelitem}{\number\thislevelcount\ExitForEach}{}}{#2}
+}
+
+\setcounter{secnumdepth}{0}
+
+\newcommand{\getAff}[1]{
+  \getIndex{#1}{}
+}
+
 \begin{document}
-\maketitle
-
-\section{Abstract}
+\vspace*{0.2in}
+
+% Title must be 250 characters or less.
+\begin{flushleft}
+{\Large
+\textbf\newline{Multimodal Spatial Availability: A Singly-Constrained
+Measure of Competitive Accessibility by Multiple
+Modes} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+}
+\newline
+% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
+\\
+Anastasia Soukhov\textsuperscript{\getAff{Department of Earth,
+Environment and Society, McMaster University, Canada}},
+Javier Tarriño-Ortiz\textsuperscript{\getAff{Universidad Politécnica de
+Madrid, Spain}},
+Julio A. Soria-Lara\textsuperscript{\getAff{Urban and Regional Planning,
+Universidad Politécnica de Madrid, Spain}},
+Antonio Páez\textsuperscript{\getAff{Department of Earth, Environment
+and Society, McMaster University, Canada}}\\
+\bigskip
+\bigskip
+\end{flushleft}
+% Please keep the abstract below 300 words
+\section*{Abstract}
 An increasing number of studies within the domain of transportation
@@
 equitable access-to-opportunity landscape.
-\hfill\break%
-\hfill\break%
-\noindent\textit{Keywords}:  Multimodal, Accessibility, Equity, Policy
-scenarios, Low emission zones,  
-\newpage
-
+
+% Please keep the Author Summary between 150 and 200 words
+% Use first person. PLOS ONE authors please skip this step.
+% Author Summary not valid for PLOS ONE submissions.
+
+\linenumbers
+
+% Use "Eq" instead of "Equation" for equation citations.
 \hypertarget{abstract}{%
@@
 Low Emission Zones in Madrid (Spain) and highlight how this policy
-intervention changes the accessibility of populations using different
-modes. In summary, spatial availability can be used to create and
-interpret multimodal policy intervention scenarios unlike previous
-methods: this creation and interpretation can help regions envision a
-more sustainable and equitable access-to-opportunity landscape by better
-identifying differences in accessibility afforded by different modes.
+intervention changes the accessibility for locations where populations
+use different modes. In summary, spatial availability can be used to
+create and interpret multimodal policy intervention scenarios unlike
+previous methods: this creation and interpretation can help regions
+envision a more sustainable and equitable access-to-opportunity
+landscape by better identifying differences in accessibility afforded by
+different modes.
 
@@
 result of land-use mix and transport systems as originally defined by
-\citet{hansenHowAccessibilityShapes1959}) have been widely applied
-within the transportation literature and is increasingly discussed by
-planners
-\citep{levinsonTransportAccessManual2020, gowerPlanningInnovationCity2022, siddiqToolsTradeAssessing2021, yanAccessibilityBasedPlanningAddressing2021}.
-An important challenge in the identification of interventions that
-equitably transform cities is the effective evaluation of
-\emph{trade-offs}: cities are complex and dynamic ecologies, and
-advantaging one component of the city can disadvantage another area,
-population, or sub-component In this way, policy evaluation should take
-a \emph{systems} approach
-\citep{fikselSustainabilityResilienceSystems2006}. One way of
-considering systems is from the perspective of the \emph{finite}. As an
-illustration, consider the amount of transport space within a city: the
-amount is typically finite so re-allocating road space away from one
-mode directly impacts the performance of the others (see the literature
-on road space reallocation e.g.,
-\citet{valencaMainChallengesOpportunities2021}). Evaluating policy
-impacts in the context of \emph{finity} provides a way to contextualize
-the balance of trade-offs that the citizens of a city should tolerate.
+{[}1{]}) have been widely applied within the transportation literature
+and is increasingly discussed by planners {[}2--5{]}. An important
+challenge in the identification of interventions that equitably
+transform cities is the effective evaluation of \emph{trade-offs}:
+cities are complex and dynamic ecologies, and advantaging one component
+of the city can disadvantage another area, population, or sub-component
+In this way, policy evaluation should take a \emph{systems} approach
+{[}6{]}. One way of considering systems is from the perspective of the
+\emph{finite}. As an illustration, consider the amount of transport
+space within a city: the amount is typically finite so re-allocating
+road space away from one mode directly impacts the performance of the
+others (see the literature on road space reallocation e.g., {[}7{]}).
+Evaluating policy impacts in the context of \emph{finity} provides a way
+to contextualize the balance of trade-offs that the citizens of a city
+should tolerate.
 
@@
 accessibility measures have been used in the context of policy
-evaluation. For instance, \citet{leeMeasuringImpactsNew2018} assesses
-the transit accessibility gains to healthcare and employment
-opportunities for disadvantaged neighbourhood in Columbus, Ohio, USA
-after the transit system's re-design and introduction of a rapid bus
-system. However, a limitation of this study, like others that implement
-accessibility measures, is they do not calculate results under a
-\emph{constrained} framework i.e., one of \emph{finity}. The citizens of
-Columbus should experience qualitative accessibility gains - but is it
-at the expense of access to opportunities to populations using other
-modes?
-
-The work of \citet{leeMeasuringImpactsNew2018} uses
-\emph{non-competitive} accessibility measures. There is a branch of
-location-based accessibility measures that do incorporate the effect of
-competition for opportunities by the population in the region. However,
-we argue that these existing methods fall short in acknowledging the
-\emph{finity} of opportunities. For instance,
-\citet{maoMeasuringSpatialAccessibility2013} applies a competitive
-measure, two-step floating catchment approach (2SFCA), for the case of
-access to healthcare services in Florida for both a multimodal network
-and a single modal network. While the differences in modal access are
-discussed, the question of how the advantage in access afforded by one
-mode over another impacts access for different mode users is unanswered.
+evaluation. For instance, {[}8{]} assesses the transit accessibility
+gains to healthcare and employment opportunities for disadvantaged
+neighbourhood in Columbus, Ohio, USA after the transit system's
+re-design and introduction of a rapid bus system. However, a limitation
+of this study, like others that implement accessibility measures, is
+they do not calculate results under a \emph{constrained} framework i.e.,
+one of \emph{finity}. The citizens of Columbus should experience
+quantitative accessibility gains - but is it at the expense of access
+using other modes? As another example, {[}9{]} implements a modified
+cumulative opportunity measure to assess differences between private
+vehicle and transit system accessibility to jobs in Melbourne, but a
+similar question remains: does the accessibility afforded to the private
+vehicle using population come at the expense of accessibility losses to
+transit users?
+
+{[}8{]} and {[}9{]} both use \emph{non-competitive} accessibility
+measures. There is a branch of location-based accessibility measures
+that do incorporate the effect of competition for opportunities by the
+population in the region. However, we argue that these existing methods
+fall short in acknowledging the \emph{finity} of opportunities. For
+instance, {[}10{]} applies a competitive measure, two-step floating
+catchment approach (2SFCA), for the case of access to healthcare
+services in Florida for both a multimodal network and a single modal
+network. While the differences in modal access are discussed, the
+question of how the advantage in access afforded by one mode over
+another impacts access for different mode users is unanswered.
 
@@
 urgent as LEZ are currently in effect in cities globally; their
-reception has been mixed \citep{tarrinoortizPublicAcceptabilityLow2021}
-and may be having negative impacts on disadvantaged populations who have
-become mobility restricted
-\citep{devrijNooneVisitsMe2022, verbeekJustManagementUrban2022}.
-Measures that evaluate the accessibility of modes given both
-\emph{constrained} and \emph{competitive} considerations are lacking in
-the literature, but are needed, to evaluate such policy interventions
-impact on accessibility.
-
-In \citet{soukhovIntroducingSpatialAvailability2023}, we introduce
-spatial availability, a type of location-based accessibility measure
-that is both \emph{constrained} and \emph{competitive}. In this paper,
-we extend the spatial availability measure into a multimodal framework
-and explore its use in answering the question outlined: \emph{``given
-opportunities are finite, how many are available to a given location
-depending on the mode used?''.} The answer to this question quantifies
-how many opportunities can be accessed, considering competition, for
-different modes. To foreground this exploration, in Section 2, we
-discuss short falls of a few existing location-based measures in
-comparison to spatial availability through a synthetic example. In
-Section 3, the spatial availability of an empirical example of the LEZ
-in the city of Madrid, Spain is calculated. We demonstrate how the
-restriction of car circulation could have impacted the spatial
-availability of opportunities for each sub-population using transit,
-cycling and walking modes. In Section 4, we provide concluding remarks
-on the strengths of the use of spatial availability as a multimodal
-accessibility measure and potential future uses in policy planning
-scenarios.
+reception has been mixed {[}11{]} and may be having negative impacts on
+disadvantaged populations who have become mobility-restricted
+{[}12,13{]}. Measures that evaluate the accessibility of modes given
+both \emph{constrained} and \emph{competitive} considerations are
+lacking in the literature, but are needed, to evaluate such policy
+interventions impact on accessibility.
+
+In {[}14{]}, we introduce spatial availability, a type of location-based
+accessibility measure that is both \emph{constrained} and
+\emph{competitive}. In this paper, we extend the spatial availability
+measure into a multimodal framework and explore its use in answering the
+question outlined: \emph{``given opportunities are finite, how many are
+available to a given location depending on the mode used?''.} The answer
+to this question quantifies how many opportunities can be accessed,
+considering competition, for different modes. To foreground this
+exploration, in Section 2, we discuss short falls of a few existing
+location-based measures in comparison to spatial availability through a
+synthetic example. In Section 3, the spatial availability of an
+empirical example of the LEZ in the city of Madrid, Spain is calculated.
+We demonstrate how the restriction of car circulation could have
+impacted the spatial availability of opportunities for each
+sub-population using transit, cycling and walking modes. In Section 4,
+we provide concluding remarks on the strengths of the use of spatial
+availability as a multimodal accessibility measure and potential future
+uses in policy planning scenarios.
 
@@
 measured are cumulative opportunity measures and weighted cumulative
-opportunity measures \citep{levinsonTransportAccessManual2020}. These
-measures weight the opportunities that can be potentially interacted
-with from origin \(i\) to destination \(j\) based on some sort of travel
-cost function (e.g., travel time, fare, travel distance) otherwise known
-as a travel impedance function \(f^{m}(c^m_{ij})\). Many weighted
-cumulative opportunities (often refered to as the gravity-based measure)
-originate from the measure proposed by
-\citet{hansenHowAccessibilityShapes1959}, which can take the following
-multimodal form: \(S_i^m = \sum_j O_j f^m(c_{ij}^m)\) where \(m\) is a
-set of modes which have mode-specific travel costs \((c_{ij}^m)\) and
-travel impedance functions \(f^m(\cdot)\).
+opportunity measures {[}2{]}. These measures weight the opportunities
+that can be potentially interacted with from origin \(i\) to destination
+\(j\) based on some sort of travel cost function (e.g., travel time,
+fare, travel distance) otherwise known as a travel impedance function
+\(f^{m}(c^m_{ij})\). Many weighted cumulative opportunities (often
+referred to as the gravity-based measure) originate from the measure
+proposed by {[}1{]}, which can take the following multimodal form:
+\(S_i^m = \sum_j O_j f^m(c_{ij}^m)\) where \(m\) is a set of modes which
+have mode-specific travel costs \((c_{ij}^m)\) and travel impedance
+functions \(f^m(\cdot)\).
 
 The Hansen-type measure does not consider competition between modes nor
-is it constrained. As an example, the work of
-\citet{tahmasbiMultimodalAccessibilitybasedEquity2019} uses the
+is it constrained. As an example, the work of {[}15{]} uses the
 Hansen-type measure to measure the potential interaction with retail
@@
 competitive location-based accessibility measures is the two-step
-floating catchment area (2SFCA) approach popularized by
-\citet{luoMeasuresSpatialAccessibility2003} who simplified the approach
-proposed by \citet{shenLocationCharacteristicsInnercity1998} (with
-similar considerations for competition in
-\citet{weibullAxiomaticApproachMeasurement1976} and
-\citet{josephMeasuringPotentialPhysical1982}).
+floating catchment area (2SFCA) approach popularized by {[}16{]} who
+simplified the approach proposed by {[}17{]} (with similar
+considerations for competition in {[}18{]} and {[}19{]}).
 
@@
 
-To illustrate, \citet{taoInvestigatingImpactsPublic2020a} calculates
-\(a_i^m\) to jobs for different income-group populations in Shenzhen
-(China) using \(m = \text{public transit}\) and \(m={car}\). They
-demonstrate that \(i\) with low-income populations have lower \(a_i^m\)
-than \(i\) with higher-income populations. Further, they demonstrate
-that \(a_i^{m=\text{public transit}}\) is lower than
-\(a_i^{m=\text{car}}\) at many \(i\), arguing that this may put \(i\)
-with lower-income populations in a further disadvantage. \(a_i\) and/or
-\(a_i^m\) are used to compare relative spatial differences in overall
-competitive accessibility and modal competitive accessibility, but
-because there is no global maximum, making it is difficult to interpret
-the significance between differences in \(a_i^{m}\) values. Questions
-such as: what is the impact that competition has on the difference in
-\(a_i^m\) values? How does impact vary spatially? And what is the
-interpretation of this difference? are left unanswered.
+To illustrate, {[}20{]} calculates \(a_i^m\) to jobs for different
+income-group populations in Shenzhen (China) using
+\(m = \text{public transit}\) and \(m={car}\). They demonstrate that
+\(i\) with low-income populations have lower \(a_i^m\) than \(i\) with
+higher-income populations. Further, they demonstrate that
+\(a_i^{m=\text{public transit}}\) is lower than \(a_i^{m=\text{car}}\)
+at many \(i\), arguing that this may put \(i\) with lower-income
+populations in a further disadvantage. \(a_i\) and/or \(a_i^m\) are used
+to compare relative spatial differences in overall competitive
+accessibility and modal competitive accessibility, but because there is
+no global maximum, making it is difficult to interpret the significance
+between differences in \(a_i^{m}\) values. Questions such as: what is
+the impact that competition has on the difference in \(a_i^m\) values?
+How does impact vary spatially? And what is the interpretation of this
+difference? are left unanswered.
 
@@
 
-In brief, we define the \emph{spatial availability} at \(i\) ( \(V_{i}\)
-) as the proportion of all opportunities in the region \(O\) that are
+In brief, we define the \emph{spatial availability} at \(i\) (\(V_{i}\))
+as the proportion of all opportunities in the region \(O\) that are
 allocated to location \(i\) from all opportunity destinations \(j\).
@@
 
-Spatial availability measure is introduced in
-\citet{soukhovIntroducingSpatialAvailability2023}. The unique feature in
-the measure is the balancing factor \(F^t_{ij}\), a proportional
-allocation mechanisms, that ensures \(V_i\) calculated for each \(i\)
-sums to the total number of opportunities. Through \(F^t_{ij}\), spatial
-availability is a \emph{competitive} and \emph{constrained}
-accessibility measure that handles the number of opportunities in the
-region in a finite way. \(F^t_{ij}\) consists of two components: a
-population-based balancing factor \(F^p_{i} = \frac{P_i}{\sum_i P_i}\)
-and an impedance-based balancing factor
-\(F^c_{ij} = \frac{F^c_{ij}}{\sum_j F^c_{ij}}\) that, respectively,
-allocate opportunities to \(i\) in proportion to the size of the
-population at \(i\) (the mass effect) and the cost of reaching
+The spatial availability measure is introduced in {[}14{]}. Spatial
+availability's unique feature is the balancing factor \(F^t_{ij}\), a
+proportional allocation mechanisms, that ensures the \(V_i\) calculated
+for each \(i\) sums, across all \(i\) in the region, to equal the total
+number of opportunities in the region. As such, spatial availability is
+a \emph{competitive} and \emph{constrained} accessibility measure as
+\(F^t_{ij}\) handles the number of opportunities in the region in a
+finite way (proportional allocation). \(F^t_{ij}\) consists of two
+components: a population-based balancing factor
+\(F^p_{i} = \frac{P_i}{\sum_i P_i}\) and an impedance-based balancing
+factor \(F^c_{ij} = \frac{F^c_{ij}}{\sum_j F^c_{ij}}\) that,
+respectively, allocate opportunities to \(i\) in proportion to the size
+of the population at \(i\) (the mass effect) and the cost of reaching
 opportunities at \(j\) (the impedance effect).
@@
 then \(C\). This synthetic example was inspired by the single-mode
-example used in \citet{shenLocationCharacteristicsInnercity1998} and
-reconfigured in \citet{soukhovIntroducingSpatialAvailability2023}.
+example used in {[}17{]} and reconfigured in {[}14{]}.
 
@@
 
-\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{x}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{27,292.18}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.95}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{18,959.86}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.88in}{\raggedright \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{A}}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{z}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{44,999.80}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.56}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{46,913.04}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.58in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.32}}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{1.05in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{65,872.91}}}}} \\
+\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{x}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{27,292.18}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.95}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{15,696.89}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{} \\
+
+\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.88in}{\raggedright \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{A}}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{z}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{44,999.80}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.57}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{51,785.72}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.58in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.36}}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{1.05in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{67,482.61}}}}} \\
 
@@
 
-\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{x}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{27,292.18}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.62}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{30,863.43}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.88in}{\raggedright \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{B}}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{z}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{44,999.80}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.03}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{103,391.95}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.58in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.90}}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{1.05in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{134,255.38}}}}} \\
+\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{x}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{27,292.18}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.64}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{38,170.03}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{} \\
+
+\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.88in}{\raggedright \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{B}}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{z}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{44,999.80}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.05}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{94,468.91}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.58in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.88}}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{1.05in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{132,638.94}}}}} \\
 
@@
 
-\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{x}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{2,240.38}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.68}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{2,034.49}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{} \\
-
-\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.88in}{\raggedright \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{C}}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{z}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{3,745.89}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.12}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{7,837.22}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.58in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.99}}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{1.05in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{9,871.71}}}}} \\
+\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{x}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{2,240.38}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.68}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{2,035.86}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{} \\
+
+\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.88in}{\raggedright \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{C}}}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{z}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{3,745.89}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{1.12}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{7,842.59}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{0.58in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{0.99}}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\multirow[c]{-2}{*}{\parbox{1.05in}{\raggedleft \textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{9,878.45}}}}} \\
 
@@
 connect this example to literature, \(S_i^m\) is calculated in the work
-of \citet{tahmasbiMultimodalAccessibilitybasedEquity2019}; they compare
-differences in \(S_i^m\) values between modes in a relative and
-comparative sense, but make no further interpretation of the \(S_i^m\)
-values.
+of {[}15{]}; they compare differences in \(S_i^m\) values between modes
+in a relative and comparative sense, but make no further interpretation
+of the \(S_i^m\) values.
 
@@
 sum of ratios. For instance, the \(z\)-using sub-population at \(A\) has
-a value of 1.56 potential jobs per potential job-seeking population
+a value of 1.57 potential jobs per potential job-seeking population
 compared to 0.95 for \(x\)-using sub-population. What is the
@@
 aggregated to \(a_i\) as shown in the sixth column, the values face
-similar interpretability issues.
-
-The Shen-type measure is implemented in the previously discussed work of
-\citet{taoInvestigatingImpactsPublic2020a} to calculate modal \(a_i^m\)
+similar interpretability issues. The Shen-type measure is implemented in
+the previously discussed work of {[}20{]} to calculate modal \(a_i^m\)
 values and the aggregated \(a_i\) is implemented in the work of
-\citet{carpentieriMultimodalAccessibilityPrimary2020}. However, similar
-to the Hansen-type measure, these works discuss relative and spatially
-comparative differences in values, they do not make further
-interpretation of the \(a_i^m\) or \(a_i\) themselves. This may be
-because the Shen-type measure is \emph{non-constrained}, this is no
-benchmark or global maximum to which comparisons can be drawn from.
+{[}21{]}. However, similar to the Hansen-type measure, these works
+discuss relative and spatially comparative differences in values, they
+do not make further interpretation of the \(a_i^m\) or \(a_i\)
+themselves. This may be because the Shen-type measure is
+\emph{non-constrained}, this is no benchmark or global maximum to which
+comparisons can be drawn from.
 
@@
 competition). In fact, at \(A\), the \(z\)-using sub-population captures
-27,953.18 more spatially available jobs (of the \(210,000\) jobs in the
+36,088.84 more spatially available jobs (of the \(210,000\) jobs in the
 region) than the sub-population using mode \(x\). The numerical
@@
 \(V_i\), is shown in the seventh column in Table 1 . Again looking at
-center \(A\), \(A\) is allocated 65,872.91 spatially available
-opportunities for both modes. 71\% of this spatial availability
+center \(A\), \(A\) is allocated 67,482.61 spatially available
+opportunities for both modes. 77\% of this spatial availability
 allocated to \(A\) is assigned to the \(z\)-using population despite
@@
 Spatial availability can be further aggregated to better interpret
-competition between modes. Across the entire region, 137,500 people use
-\(z\) (65\% of the region population). However, the \(z\)-using
-population accounts for 75\% of the region's total spatial availability
-- the rest is allocated to the \(x\)-using population (35\%of the total
-population). Notably, the \(x\)-using population captures 10\% less
+competition between modes. Across the entire region, 130,000 people use
+\(z\) (62\% of the region population). However, the \(z\)-using
+population accounts for 73\% of the region's total spatial availability
+- the rest is allocated to the \(x\)-using population (38\%of the total
+population). Notably, the \(x\)-using population captures 11\% less
 spatial availability to opportunities than its population proportion.
@@
 
+Since spatial availability is constrained and has an interpretable
+meaning as a proportion of the total opportunities in the region, the
+values at \(i\) have a new significance. Inequality in \(V_i^m\) values
+can be explored through a variety of approaches. For instance, consider
+travel times. The \(z\)-using population accounts for 67\% of the
+potential travel time traveled in the region: this is 7\% less travel
+time than the proportion of spatial available opportunities that is
+allocated to them. In other words, the \(z\)-using population travels
+less minutes overall and has more spatial availability of opportunities
+than the \(x\)-using population using the slower mode \(x\).
+
 Alternatively, inequities in spatial availability between mode-using
@@
 The \(v_i^m\) values for \(A\), \(B\), and \(C\) for the \(x\)-using
-sub-populations are 0.95, 0.62 and 0.68 spatially available jobs per
+sub-populations are 0.95, 0.64 and 0.68 spatially available jobs per
 capita, respectively. The \(v_i^m\) for the \(z\)-using sub-populations
-are much higher, with values of: 1.56, 1.03 and 1.12 respectively. The
+are much higher, with values of: 1.57, 1.05 and 1.12 respectively. The
 \(x\)-using population, especially at \(B\) and \(C\), are directly
@@
 for passenger transport, LEZ are a policy of \emph{geographic
-discrimination}. LEZ actively change how people access opportunities by
-making the travel impedance more costly for car-mode users. If seeing
+discrimination} as they change how people access opportunities by making
+the travel impedance more costly for car-mode users. If seeing
 opportunities as finite, this discrimination allows populations to
@@
 their implementation as specified in their climate-change-related plans:
-\emph{Plan Nacional Integrado de Energía y Clima 2021-2030}
-\citep{espanaPlanNacionalIntegrado2020} and \emph{Plan Nacional de
-Control de la Contaminación Atmosférica}
-\citep{espanaResolucion10Enero2020}. Specifically, the national Spanish
-law 7/2021 ( \emph{Ley de Cambio Climático y Transición Energética})
-will require all municipalities to implement LEZ by 2023 if they meet at
-least one of the following requirements: (i) municipalities
-\textgreater50,000 inhab.; (ii) islands; and (iii) municipalities
-\textgreater{} 20,000 inhab. when air quality exceeds limits specified
-in \emph{RD 102/2011 de Mejora de Calidad del Aire}
-\citep{barcelonaGUIATECNICAPARA2021}.
-
-In 2017, a LEZ was implemented in the Spanish capital city of Madrid
+\emph{Plan Nacional Integrado de Energía y Clima 2021-2030} {[}22{]} and
+\emph{Plan Nacional de Control de la Contaminación Atmosférica}
+{[}23{]}. Specifically, the national Spanish law 7/2021 ( \emph{Ley de
+Cambio Climático y Transición Energética}) will require all
+municipalities to implement LEZ by 2023 if they meet at least one of the
+following requirements: (i) municipalities \textgreater50,000 inhab.;
+(ii) islands; and (iii) municipalities \textgreater{} 20,000 inhab. when
+air quality exceeds limits specified in \emph{RD 102/2011 de Mejora de
+Calidad del Aire} {[}24{]}.
+
+In 2017, LEZs were implemented in the Spanish capital city of Madrid
 following the goals set out in the national agenda . In geographic
@@
 make and model of fossil fuel internal combustion engine vehicles), are
-banned from driving into the area (threat of fine) unless they are used
-by residents or meet other exemptions. This restriction impacted
-approximately half of all car trips that were typically made into the
-LEZ Centro \citep{tarrinoortizAnalyzingImpactLow2022}.
-
-In this section, we use \(V_i^m\) to quantify the competition for
+not permitted to enter the area unless they are used by residents or
+meet other exemptions. This restriction impacted approximately half of
+all car trips that were typically made into the LEZ Centro {[}25{]}.
+
+For this case study, we use \(V_i^m\) to quantify the competition of
 spatially available opportunities between modes after the LEZ Centro
@@
 
-The 2018 Community of Madrid travel survey
-(\citet{comunidaddemadridResultadosEDM20182020}) is the source of data
-for this empirical example: it is a representative survey that reflects
-a snap-shot of the travel patterns for one typical day of the working
-week (e.g., n=222,744 trips with representative population elevation
-factors). In this paper, a sample of the travel survey is used, namely
-the residential home origin to work destination trips of all modes and
-those that originate and end in the city of Madrid. These totals are
-displayed in Figure \ref{fig:Fig2} and Figure \ref{fig:Fig3}. Both
-figures are displayed at the level of traffic analysis zones (\(i\) and
-\(j\)) that correspond to the survey. The red boundary represents the
-LEZ Centro in effect in 2017 and thus those travel patterns of
+The 2018 Community of Madrid travel survey ({[}26{]}) is the source of
+data for this empirical example: it is a representative survey that
+reflects a snap-shot of the travel patterns for one typical day of the
+working week (e.g., n=222,744 trips with representative population
+elevation factors). In this paper, a sample of the travel survey is
+used, namely the residential home origin to work destination trips of
+all modes and those that originate and end in the city of Madrid. These
+totals are displayed in Figure \ref{fig:Fig2} and Figure \ref{fig:Fig3}.
+Both figures are displayed at the level of traffic analysis zones (\(i\)
+and \(j\)) that correspond to the survey. The red boundary represents
+the LEZ Centro in effect in 2017 and thus those travel patterns of
 car-restriction reflected in the survey. The cyan boundary represents
@@
 \item
-  Transit: all bus, trams, and trains
+  Transit: all bus, trams, and trains,
 \item
   Bike: all bicycle trips (e.g., private, public, or company bike trips)
-  and ``other'' types of micromobility options
+  and ``other'' types of micromobility options,
 \item
-  Walk: walking or by foot
+  Walk: walking or by foot,
 \end{itemize}
@@
 
-From Figure \ref{fig:Fig2}, it can also be seen that biking and walking
+From Figure \ref{fig:Fig3}, it can also be seen that biking and walking
 trips are less common than motorized trips at 1\% and 15\% respectively.
-The distribution of walking and biking trips appear to be similar to
-that of transit trips.
+Noteably, there is a positive trend between the populations of walking
+and biking trips in zones and populations of transit trips. This
+positive trend is higher than for car trip populations.
 
@@
 travel cost). This distribution is then used to derive impedance
-functions \citep[e.g., done in the accessibility works
-of][\citet{horbachov_theoretical_2018}, and
-\citet{batista_estimation_2019} for example]{lopez_2017_spatial}.
-Maximum likelihood estimation and the Nelder-Mead method for direct
-optimization available within the R \{fitdistrplus\} package
-\citep{fitdistrplus_2015} is used to fit the impedance functions. As
-shown as shown in Figure \ref{fig:Fig4}, based on goodness-of-fit
-criteria and associated diagnostics, the gamma and log-normal
-probability density function (line curves) are selected as best fitting
-curves for the motorized and non-motorized modes respectively. The
-selection of functional form aligns with examples used in the literature
-(e.g., \citet{reggianiAccessibilityImpedanceForms2011}). Overall, the
-plots in Figure \ref{fig:Fig4} display the probability of travel given a
-trip travel time, based on trip flows from the survey. These
+functions (e.g., done in the accessibility works of {[}27{]}, {[}28{]},
+and {[}29{]}). Maximum likelihood estimation and the Nelder-Mead method
+for direct optimization available within the R \{fitdistrplus\} package
+{[}30{]} is used to fit the impedance functions. As shown as shown in
+Figure \ref{fig:Fig4}, based on goodness-of-fit criteria and associated
+diagnostics, the gamma and log-normal probability density function (line
+curves) are selected as best fitting curves for the motorized and
+non-motorized modes respectively. The selection of functional form
+aligns with empirical examples in other regions {[}14,31,32{]}. Overall,
+the plots in Figure \ref{fig:Fig4} display the probability of travel
+given a trip travel time, based on trip flows from the survey. These
 `probability of travel' at each travel time for each mode are realized
-observations reflect the land-use, the transport system, and the
+observations that reflect the land-use, the transport system, and the
 population travel behaviour in Madrid.
@@
 
-Using the data inputs outlined, \(V_i^m\), the spatial availability of
-jobs, is calculated for each of the four modal categories \(m\) at the
-level of traffic analysis zones \(i\) in Madrid and demonstrated in this
-sub-section.
+The spatial availability of jobs \(V_i^m\) is calculated for each of the
+four modal categories \(m\) at the level of traffic analysis zones \(i\)
+in Madrid and demonstrated in Figure \ref{fig:Fig5}.
 
@@
 
-Figure \ref{fig:Fig5} displays \(V_i^m\) for the four modal categories
-in 2018. \(V_i^m\) values represent a proportion of the total number of
-the 847,574 jobs in the region; put another way, \(V_i^m\) is how many
-jobs are \emph{spatially available} to the population located at the
-\(i\) based on the travel impedance of the mode (relative to the travel
-impedance of all modes) and the mode-using population size at the \(i\)
-(relative to the population size of all modes and \(i\)). Note the
-differences in the magnitudes of \(V_i^m\) between modes. The majority
-of \(V_i^m\) is allocated to car- and transit- using populations. This
-is to be expected, as commuting using motorized modes represents 84\% of
-the population (37\% (car/motor) and 47\% (transit) in 2018). However,
-these modal options capture 95\% of the total spatial availability in
-Madrid. In particular, the car/motor using population is allocated
-disproportionately more \(V_i^m\) than its modal population (37\% of the
-population vs.~48\% of the \(V_i^m\)) compare to the transit using
-population and its relatively proportional \(V_i^m\) value (15\% of the
-population vs.~47\% \(V_i^m\)).
-
-How does the \(V_i^m\) advantage allocated to car-using population
-arise? From the perspective of finite opportunities, \(V_i^m\) is
-allocated to car-using populations from less competitive modal
-populations. How competitive one mode is compared to other modes varies
-spatially, but overall car-using populations capture more opportunities
-per car-using population than other modal populations. Namely, though
-walking and cycling populations represent 14.74\% and 1.16\%
-respectively, \(V_i^m={walk}\) and \(V_i^m={bike}\) is 4.43\% and 0.23\%
-in the region respectively. These modes are less competitive, especially
-compared to the car/motor mode, as a result of: 1) their lower travel
-impedance values at longer travel times (see Figure \ref{fig:Fig4} at
-travel times beyond \textasciitilde30 minutes), 2) their low population
-values values overall, and 3) higher populations present in origins with
-high motorized mode commuting. These factors all contribute to the the
-car/motor mode being most advantaged in capturing spatially available
-job opportunities overall.
+\(V_i^m\) is a proportion of the total number of the 847,574 jobs in the
+region and is visualized in Figure \ref{fig:Fig5}. Since \(V_i^m\) is
+calculated based on the likelihood of travel from observed home-to-work
+journeys, the values can be understood as the number of full-time jobs
+that are spatially available to the full-time working population at that
+\(i\) and their associated \(m\), relative to all the jobs in the city.
+\(V_i^m\) is the number of jobs that are \emph{spatially available} to a
+\(m\)-using population located at \(i\), relative to the travel
+impedance and size of \emph{all} populations in the region.
+
+Notable are the differences in the magnitude of \(V_i^m\) between modes
+in Figure \ref{fig:Fig5}. The majority of \(V_i^m\) is allocated to car-
+and transit- using populations. This is to be expected, as the
+population that commutes using these modes represents 84.1\% of the
+total population. Differences in \(V_i^m\) values within mode-using
+populations also exist: car-using populations outside of the M-30 region
+appear to have greater \(V_i^m\) values, while some \(i\) areas inside
+the M-30 appear to have higher \(V_i^m\) values for the transit-using
+populations. Overall, the magnitude of \(V_i^m\) values for the bikers
+and walkers are lower than car and transit but the highest
+\(V_i^{bike}\) and \(V_i^{walk}\) values tend to be allotted to \(i\)s
+within the M-30 and \(i\)s that have higher \(V_i^{transit}\) values.
+
+The differences between the mode-using population and their
+mode-specific spatial availability highlights the competitive advantage
+offered to certain modes in certain spatial extents. As summarized in
+the left-most columns in Figure \ref{fig:Fig6}, the `car/motor' and
+`transit' populations represent a combined 95.3\% of the total spatial
+availability in the city. However, the `car/motor' using population is
+allocated disproportionately more \(V_i^m\) than its size compared to
+the transit-using population. The car-using and transit-using population
+is 36.6\% and 47.5\% respectively, but is allocated 48.0\% and 47.3\%
+respectively, of the city's spatial availability. When treating the
+number of opportunities that can be reached as a finite value (total:
+847,574 opportunities), fewer opportunities are spatial availability to
+the lesser competitive modes-using populations, in this case walking and
+cycling. These modes are less competitive as a result of: their lower
+travel impedance values at longer travel times (see Figure
+\ref{fig:Fig4} at travel times beyond \textasciitilde30 minutes); their
+low population values values overall; and higher populations present in
+origins with high motorized mode commuting. These factors all contribute
+to the the car/motor mode being most advantaged in capturing spatially
+available job opportunities overall.
 
@@
 
+{\centering \includegraphics[width=1\linewidth]{images/modal_V_comps_4plot}
+
+}
+
+\caption{\label{fig:Fig6} Displays the proportion of the working population by mode and spatial availability of job opportunities by mode aggregated for three spatial areas.
+From left to right, the city of Madrid (All), the area within the M-30 highway (M-30),the area within the Centro region (Centro).}\label{fig:modal-V-comps-plot}
+\end{figure}
+
+There are spatial variations in the competitive advantage of the
+car-using populations. The proportion of car-using population in the
+Centro is smaller and has higher travel impedance values relative to the
+inputs in other areas and mode-using populations. The LEZ Centro
+implementation further restricts the car-advantage as it shifted more
+than half of all car trips into the LEZ to another mode {[}25{]}. This
+restriction decreased the number of car-using population from \(i\)s
+going into the LEZ Centro (an area with a large number of jobs overall,
+see Figure \ref{fig:Fig2}), thus increasing the mass effect for non-car
+modes and resulting in proportionally higher \(v_i^m\) values for
+non-car modes. As such, the lower amount of access to opportunities by
+car-mode allows more opportunities in the LEZ to be available by
+populations using other modes.
+
+As summarized in the two right-most columns in Figure \ref{fig:Fig6},
+the proportion of spatial availability allocated to the car-using
+population in the Centro (13.8\% or 5,373 opportunities). As a
+comparative reference, this is less than the proportion of the car-using
+population in the Centro (16.1\%), evidently less then the proportion of
+car-using population in the city, and is the opposite of the trend
+overall (left-most columns) and within the M-30 (middle columns). More
+opportunities are spatial availability to non-car using populations
+within the Centro, particularly transit-using populations (68.5\% of
+spatially available jobs in the Centro despite representing 51.4\% of
+the population in the Centro and 47.5\% in the city overall).
+
+From Figure \ref{fig:Fig6}, it is also summarized that there is a higher
+proportion of opportunities spatially available to walking and cycling
+populations in the Centro than in the City overall and in all areas
+within the M-30. Notably, within the Centro, 1.7\% and 16.1\% of
+opportunities are spatially available to bike and walk modes
+respectively, while their populations represent smaller proportions of
+1.2\% and 14.7\% of the population overall. Though the proportion of
+spatial availability for these mode-using populations is still lower
+than the proportion of mode-using population located in the Centro,
+these modes are more competitive within the Centro than outside of the
+Centro. By restricting the more competitive car mode through the LEZ,
+the advantage in the spatial availability of opportunities afforded to
+the otherwise lesser competitive modes is made apparent.
+
+The spatial differences in the competitive dis/advantage of spatial
+availability between modes can also be visualized per origin. Figure
+\ref{fig:Fig7} visualizes \(v_i^m\), the spatial availability \(V_i^m\)
+divided by the mode-population. \(v_i^m\) values above 1 are represented
+in increasing red shades, values below 1 are represented in increasingly
+green shades, and values equal to 1 are white. These plots illustrates
+the discussion of the disproportionately high over-allocation of spatial
+availability relative to the mode-using population in many of the
+origins for the car/motor mode (bottom left plot, areas denoted with
+green \(v_i^m\) values above 1). These plots also visualize areas that
+disproportionately capture lower spatial availability (under 1),
+represented in shades of red. It can be observed that the transit-using
+population's spatial availability to jobs is relatively balanced (i.e.,
+many zones are white), while the non-motorized modes \(v_i^m\) values
+are low (under 1) overall.
+
+\begin{figure}
+
 {\centering \includegraphics[width=1\linewidth]{images/SA_im_vv_zn208_plot}
@@
 
-\caption{\label{fig:Fig6} Spatial availability of job opportunities per capita per origin and mode $v_i^m$ in Madrid.
+\caption{\label{fig:Fig7} Spatial availability of job opportunities per mode-using capita by mode $v_i^m$ per origin in Madrid.
 Calculated using the home-to-work origin destination flows from the 2018 travel survey.}\label{fig:SA-per-capita-m-plot}
@@
 
-Furthermore, there are spatial differences in the competitive advantage
-of spatial availability between modes. Figure \ref{fig:Fig6} visualizes
-\(v_i^m\), the spatial availability divided by the mode-population.
-These plots illustrates the discussion of the disproportionately high
-over-allocation of spatial availability relative to the mode-using
-population in many of the origins for the car/motor mode (bottom left
-plot, areas denoted with green \(v_i^m\) values above 1). These plots
-also visualize areas that disproportionately capture lower spatial
-availability (under 1), represented in shades of red. It can be observed
-that the transit-using population's spatial availability to jobs is
-relatively balanced (i.e., many zones are white), while the
-non-motorized modes \(v_i^m\) values are low (under 1).
-
-Interestingly, \(v_i^m\) for car/motor within and near the LEZ Centro is
-near or below 1 (white/red): potentially as a consequence, all non-car
-modes have relatively higher \(v_i^m\) values in these areas. Though
-access to travel survey data immediately before the LEZ implementation
-is unavailable, we know that the LEZ Centro shifted more than half of
-all car trips into the LEZ to another mode
-\citep{tarrinoortizAnalyzingImpactLow2022}. This restriction decreased
-the number of car-using population from \(i\)s going into the LEZ Centro
-(an area with a large number of jobs - Figure \ref{fig:Fig2}), thus
-increasing the mass effect for non-car modes and resulting in
-proportionally higher \(v_i^m\) values for non-car modes. Though the
-spatial availability from before the LEZ Centro implementation is
-unknown, Figure \ref{fig:Fig6} provides a benchmark for quantifying
-potential LEZ implementations in the future (given 2018 travel
-conditions). Figure \ref{fig:Fig6} also shows that many areas within the
-M-30 have high (white/green) \(v_i^m\) values for car-mode, signaling
-that the spatial expansion of the LEZ Centro stands to increase the
-spatial availability of jobs for non-car mode using populations.
+Interestingly, as also represented in Figure \ref{fig:Fig6}, \(v_i^m\)
+for car/motor within and near the LEZ Centro is near or below 1
+(white/red) in Figure \ref{fig:Fig7} while all non-car modes have
+relatively higher \(v_i^m\) values. Though the spatial availability from
+before the LEZ Centro implementation is unknown, Figure \ref{fig:Fig7}
+provides a benchmark for quantifying potential LEZ implementations in
+the future (given 2018 travel conditions). Figure \ref{fig:Fig7} also
+shows that many areas within the M-30 have high (white/green) \(v_i^m\)
+values for car-mode, signaling that the spatial expansion of the LEZ
+Centro stands to increase the spatial availability of jobs for non-car
+mode using populations.
 
@@
 Location-based accessibility measures like the Hansen-type \(S_i^m\),
-Shen-type \(a_i^m\), and spatial availability \(V_i^m\) measures have a
-commonality - they are a weighted sum of opportunities assigned to each
-spatial unit \(i\) in a region. How the weight and sum of the
+Shen-type \(a_i^m\), and spatial availability \(V_i^m\) measures share a
+commonality; they are a weighted sum of opportunities assigned to each
+spatial unit \(i\) in a region. In this way, they all can be interpreted
+as a score of how many opportunities can be potentially interacted with
+by the population at \(i\). How the weight and sum of the
 potentially-interacted-with opportunities is considered is what defines
-the type of accessibility measure. Within this paper, the location-based
-singly- \emph{constrained} and \emph{competitive} accessibility measure,
-known as spatial availability \(V_i\)
-(\citet{soukhovIntroducingSpatialAvailability2023}), is extended for the
-case of capturing multimodal accessibility to opportunities \(V_i^m\). A
-synthetic example and then an empirical case of LEZ in Madrid are
-detailed to demonstrate this multimodal extension.
+the type of accessibility measure.
+
+Within this paper, the location-based singly- \emph{constrained} and
+\emph{competitive} accessibility measure, known as spatial availability
+\(V_i\) {[}14{]}, is extended for the case of capturing multimodal
+accessibility to opportunities \(V_i^m\). A synthetic example and then
+an empirical case of LEZ in Madrid are detailed to demonstrate this
+multimodal extension.
 
@@
 modes. The car and transit modes have the highest spatial availability,
-with the car mode having highest availability with exception to the
-areas within the LEZ Centro. This finding reflects what may suspect are
-real conditions: since car travel has been highly restricted within the
-LEZ Centro, fewer car-using people potentially interact with jobs within
-the LEZ Centro, leaving more \emph{spatially available} jobs for
-non-car-using populations. This difference in car-using populations in
-locations accessing jobs within and immediately outside the LEZ Centro
-increases the competitiveness of the transit-using population (the
-second most competitive mode) as well as the non-motorized modes.
+with the car-mode having highest availability with exception to the
+areas within the LEZ Centro. Since car travel has been highly restricted
+within the LEZ Centro, fewer car-using people potentially interact with
+jobs within the LEZ Centro, leaving more \emph{spatially available} jobs
+for non-car-using populations. This difference in car-using populations
+in locations accessing jobs within and immediately outside the LEZ
+Centro increases the competitiveness of the transit-using population
+(the second most competitive mode) as well as the non-motorized modes.
 
@@
 population at each \(i\) to yield mode-population normalized values.
-These values, reflected in Figure \ref{fig:Fig6}, can be used as a
+These values, reflected in Figure \ref{fig:Fig7}, can be used as a
 benchmark to investigate existing conditions and plan future LEZ
@@
 are difficult for planners to operationalize for a variety of reasons
-including issues of computation and interpretability
-\citep{levinsonTransportAccessManual2020}. With spatial availability,
-the magnitude of opportunities that are available as a proportion of all
-the opportunities in the region is equal to \(V_i\). As a result of its
-proportional allocation mechanism, \(V_i\) can be easily extended into
-multimodal applications. This flexibility is pertinent to model policy
-scenarios in our cities that are becoming increasingly multimodal. The
-interpretation of \(V_i\) allows for manipulation of \(V_i^m\) values to
-investigate differences of availability between neighbourhoods, modes,
-and regions, generate per capita benchmarks, and/or generate average
-values per population-group.
+including issues of computation and interpretability {[}2{]}. With
+spatial availability, the magnitude of opportunities that are available
+as a proportion of all the opportunities in the region is equal to
+\(V_i\). As a result of its proportional allocation mechanism, \(V_i\)
+can be naturally extended into multimodal applications. This flexibility
+is helpful to modelling policy scenarios in our cities that are
+increasingly multimodal. The interpretation of \(V_i\) allows for
+manipulation of \(V_i^m\) values to investigate differences of
+availability between neighbourhoods, modes, and regions, generate per
+capita benchmarks, and/or generate average values per population-group.
 
@@
 Program (CGS D) provided by the Social Sciences and Humanities Research
-Council (SSHRC).
+Council (SSHRC) and Project Mobilizing Justice, also supported by SSHRC.
+All work is fully-reproducible and available within this
+\href{https://github.com/soukhova/Madrid-ZBE-LEZ}{GitHub repository}.
 
@@
 
---\textgreater{}
-
-\newpage
-\bibliography{mybibfile.bib}
+\hypertarget{references}{%
+\section*{References}\label{references}}
+\addcontentsline{toc}{section}{References}
+
+\hypertarget{refs}{}
+\begin{CSLReferences}{0}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-hansenHowAccessibilityShapes1959}{}}%
+\CSLLeftMargin{1. }%
+\CSLRightInline{Hansen WG. How accessibility shapes land use. Journal of
+the American Institute of Planners. 1959;25: 73--76.
+doi:\href{https://doi.org/10.1080/01944365908978307}{10.1080/01944365908978307}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-levinsonTransportAccessManual2020}{}}%
+\CSLLeftMargin{2. }%
+\CSLRightInline{Levinson D, King D. Transport access manual: {A} guide
+for measuring connection between people and places. {University of
+Sydney}; 2020. Available:
+\url{https://ses.library.usyd.edu.au/handle/2123/23733}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-gowerPlanningInnovationCity2022}{}}%
+\CSLLeftMargin{3. }%
+\CSLRightInline{Gower A, Grodach C. Planning innovation or city
+branding? Exploring how cities operationalise the 20-minute
+neighbourhood concept. Urban Policy and Research. 2022;40: 36--52.
+doi:\href{https://doi.org/10.1080/08111146.2021.2019701}{10.1080/08111146.2021.2019701}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-siddiqToolsTradeAssessing2021}{}}%
+\CSLLeftMargin{4. }%
+\CSLRightInline{Siddiq F, D. Taylor B. Tools of the trade?: Assessing
+the progress of accessibility measures for planning practice. Journal of
+the American Planning Association. 2021;87: 497--511.
+doi:\href{https://doi.org/10.1080/01944363.2021.1899036}{10.1080/01944363.2021.1899036}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-yanAccessibilityBasedPlanningAddressing2021}{}}%
+\CSLLeftMargin{5. }%
+\CSLRightInline{Yan X. Toward accessibility-based planning addressing
+the myth of travel cost savings. {JOURNAL} {OF} {THE} {AMERICAN}
+{PLANNING} {ASSOCIATION}. 2021;87: 409--423.
+doi:\href{https://doi.org/10.1080/01944363.2020.1850321}{10.1080/01944363.2020.1850321}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-fikselSustainabilityResilienceSystems2006}{}}%
+\CSLLeftMargin{6. }%
+\CSLRightInline{Fiksel J. Sustainability and resilience: Toward a
+systems approach. Sustainability: Science, Practice and Policy. 2006;2:
+14--21.
+doi:\href{https://doi.org/10.1080/15487733.2006.11907980}{10.1080/15487733.2006.11907980}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-valencaMainChallengesOpportunities2021}{}}%
+\CSLLeftMargin{7. }%
+\CSLRightInline{Valença G, Moura F, Morais De Sá A. Main challenges and
+opportunities to dynamic road space allocation: From static to dynamic
+urban designs. Journal of Urban Mobility. 2021;1: 100008.
+doi:\href{https://doi.org/10.1016/j.urbmob.2021.100008}{10.1016/j.urbmob.2021.100008}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-leeMeasuringImpactsNew2018}{}}%
+\CSLLeftMargin{8. }%
+\CSLRightInline{Lee J, Miller HJ. Measuring the impacts of new public
+transit services on space-time accessibility: An analysis of transit
+system redesign and new bus rapid transit in columbus, ohio, {USA}.
+Applied Geography. 2018;93: 47--63.
+doi:\href{https://doi.org/10.1016/j.apgeog.2018.02.012}{10.1016/j.apgeog.2018.02.012}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-mohriClusteringMethodMeasuring2021}{}}%
+\CSLLeftMargin{9. }%
+\CSLRightInline{Mohri S, Mortazavi S, Nassir N. A clustering method for
+measuring accessibility and equity in public transportation service:
+{Case} study of {Melbourne}. SUSTAINABLE CITIES AND SOCIETY. 2021;74.
+doi:\href{https://doi.org/10.1016/j.scs.2021.103241}{10.1016/j.scs.2021.103241}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-maoMeasuringSpatialAccessibility2013}{}}%
+\CSLLeftMargin{10. }%
+\CSLRightInline{Mao L, Nekorchuk D. Measuring spatial accessibility to
+healthcare for populations with multiple transportation modes. Health \&
+Place. 2013;24: 115--122.
+doi:\href{https://doi.org/10.1016/j.healthplace.2013.08.008}{10.1016/j.healthplace.2013.08.008}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-tarrinoortizPublicAcceptabilityLow2021}{}}%
+\CSLLeftMargin{11. }%
+\CSLRightInline{Tarriño-Ortiz J, Soria-Lara JA, Gómez J, Vassallo JM.
+Public acceptability of low emission zones: The case of {``madrid
+central.''} Sustainability. 2021;13: 3251.
+doi:\href{https://doi.org/10.3390/su13063251}{10.3390/su13063251}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-devrijNooneVisitsMe2022}{}}%
+\CSLLeftMargin{12. }%
+\CSLRightInline{De Vrij E, Vanoutrive T. {``No-one visits me anymore''}:
+Low emission zones and social exclusion via sustainable transport
+policy. 2022 {[}cited 27 Jul 2023{]}. Available:
+\url{https://www.tandfonline.com/doi/epdf/10.1080/1523908X.2021.2022465?needAccess=true\&role=button}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-verbeekJustManagementUrban2022}{}}%
+\CSLLeftMargin{13. }%
+\CSLRightInline{Verbeek T, Hincks S. The {``just''} management of urban
+air pollution? A geospatial analysis of low emission zones in brussels
+and london. Applied Geography. 2022;140: 102642.
+doi:\href{https://doi.org/10.1016/j.apgeog.2022.102642}{10.1016/j.apgeog.2022.102642}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-soukhovIntroducingSpatialAvailability2023}{}}%
+\CSLLeftMargin{14. }%
+\CSLRightInline{Soukhov A, Paez A, Higgins CD, Mohamed M. Introducing
+spatial availability, a singly-constrained measure of competitive
+accessibility {\textbar} {PLOS} {ONE}. {PLOS} {ONE}. 2023; 1--30.
+doi:\href{https://\%20doi.org/10.1371/journal.pone.0278468}{https://
+doi.org/10.1371/journal.pone.0278468}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-tahmasbiMultimodalAccessibilitybasedEquity2019}{}}%
+\CSLLeftMargin{15. }%
+\CSLRightInline{Tahmasbi B, Mansourianfar MH, Haghshenas H, Kim I.
+Multimodal accessibility-based equity assessment of urban public
+facilities distribution. Sustainable Cities and Society. 2019;49:
+101633.
+doi:\href{https://doi.org/10.1016/j.scs.2019.101633}{10.1016/j.scs.2019.101633}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-luoMeasuresSpatialAccessibility2003}{}}%
+\CSLLeftMargin{16. }%
+\CSLRightInline{Luo W, Wang F. Measures of spatial accessibility to
+health care in a {GIS} environment: Synthesis and a case study in the
+chicago region. Environ Plann B Plann Des. 2003;30: 865--884.
+doi:\href{https://doi.org/10.1068/b29120}{10.1068/b29120}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-shenLocationCharacteristicsInnercity1998}{}}%
+\CSLLeftMargin{17. }%
+\CSLRightInline{Shen Q. Location characteristics of inner-city
+neighborhoods and employment accessibility of low-wage workers. Environ
+Plann B. 1998;25: 345--365.
+doi:\href{https://doi.org/10.1068/b250345}{10.1068/b250345}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-weibullAxiomaticApproachMeasurement1976}{}}%
+\CSLLeftMargin{18. }%
+\CSLRightInline{Weibull JW. An axiomatic approach to the measurement of
+accessibility. Regional Science and Urban Economics. 1976;6: 357--379.
+doi:\href{https://doi.org/10.1016/0166-0462(76)90031-4}{10.1016/0166-0462(76)90031-4}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-josephMeasuringPotentialPhysical1982}{}}%
+\CSLLeftMargin{19. }%
+\CSLRightInline{Joseph AE, Bantock PR. Measuring potential physical
+accessibility to general practitioners in rural areas: A method and case
+study. Social Science \& Medicine. 1982;16: 85--90.
+doi:\href{https://doi.org/10.1016/0277-9536(82)90428-2}{10.1016/0277-9536(82)90428-2}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-taoInvestigatingImpactsPublic2020a}{}}%
+\CSLLeftMargin{20. }%
+\CSLRightInline{Tao Z, Zhou J, Lin X, Chao H, Li G. Investigating the
+impacts of public transport on job accessibility in {Shenzhen}, {China}:
+A multi-modal approach. Land Use Policy. 2020;99: 105025.
+doi:\href{https://doi.org/10.1016/j.landusepol.2020.105025}{10.1016/j.landusepol.2020.105025}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-carpentieriMultimodalAccessibilityPrimary2020}{}}%
+\CSLLeftMargin{21. }%
+\CSLRightInline{Carpentieri G, Guida C, Masoumi HE. Multimodal
+{Accessibility} to {Primary Health Services} for the {Elderly}: {A Case
+Study} of {Naples}, {Italy}. Sustainability. 2020;12: 781.
+doi:\href{https://doi.org/10.3390/su12030781}{10.3390/su12030781}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-espanaPlanNacionalIntegrado2020}{}}%
+\CSLLeftMargin{22. }%
+\CSLRightInline{España. Plan nacional integrado de energía y clima
+({PNIEC}) 2021-2030. 2020 {[}cited 30 Jul 2023{]}. Available:
+\url{https://www.miteco.gob.es/es/prensa/pniec.html}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-espanaResolucion10Enero2020}{}}%
+\CSLLeftMargin{23. }%
+\CSLRightInline{España. Resolución de 10 de enero de 2020, de la
+dirección general de biodiversidad y calidad ambiental, por la que se
+publica el programa nacional de control de la contaminación atmosférica.
+Resolución Jan 24, 2020 pp. 6947--6947. Available:
+\url{https://www.boe.es/eli/es/res/2020/01/10/(10)}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-barcelonaGUIATECNICAPARA2021}{}}%
+\CSLLeftMargin{24. }%
+\CSLRightInline{Barcelona. {GUÍA} {TÉCNICA} {PARA} {LA} {IMPLEMENTACIÓN}
+{DE} {ZONAS} {DE} {BAJAS} {EMISIONES}. Àrea Metropolitana de Barcelona
+({AMB}): Oficina Técnica de Gerencia del {AMB}; 2021. Available:
+\url{https://revista.dgt.es/images/GUIA-ZBE.pdf}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-tarrinoortizAnalyzingImpactLow2022}{}}%
+\CSLLeftMargin{25. }%
+\CSLRightInline{Tarriño-Ortiz J, Gómez J, Soria-Lara JA, Vassallo JM.
+Analyzing the impact of low emission zones on modal shift. Sustainable
+Cities and Society. 2022;77: 103562.
+doi:\href{https://doi.org/10.1016/j.scs.2021.103562}{10.1016/j.scs.2021.103562}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-comunidaddemadridResultadosEDM20182020}{}}%
+\CSLLeftMargin{26. }%
+\CSLRightInline{Comunidad de Madrid. Resultados de la {EDM} 2018 - Datos
+Abiertos. 2020 {[}cited 31 Jul 2023{]}. Available:
+\url{https://datos.comunidad.madrid/catalogo/dataset/resultados-edm2018}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-lopez_2017_spatial}{}}%
+\CSLLeftMargin{27. }%
+\CSLRightInline{Lopez FA, Paez A. Spatial clustering of high-tech
+manufacturing and knowledge-intensive service firms in the greater
+toronto area. Canadian Geographer-Geographe Canadien. 2017;61: 240--252.
+doi:\href{https://doi.org/10.1111/cag.12326}{10.1111/cag.12326}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-horbachov_theoretical_2018}{}}%
+\CSLLeftMargin{28. }%
+\CSLRightInline{Horbachov P, Svichynskyi S. Theoretical substantiation
+of trip length distribution for home-based work trips in urban transit
+systems. Journal of Transport and Land Use. 2018;11: 593--632.
+Available: \url{https://www.jstor.org/stable/26622420}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-batista_estimation_2019}{}}%
+\CSLLeftMargin{29. }%
+\CSLRightInline{Batista SFA, Leclercq L, Geroliminis N. Estimation of
+regional trip length distributions for the calibration of the aggregated
+network traffic models. Transportation Research Part B: Methodological.
+2019;122: 192--217.
+doi:\href{https://doi.org/10.1016/j.trb.2019.02.009}{10.1016/j.trb.2019.02.009}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-fitdistrplus_2015}{}}%
+\CSLLeftMargin{30. }%
+\CSLRightInline{Delignette-Muller ML, Dutang C. {fitdistrplus}: An {R}
+package for fitting distributions. Journal of Statistical Software.
+2015;64: 1--34. Available:
+\url{https://www.jstatsoft.org/article/view/v064i04}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-reggianiAccessibilityImpedanceForms2011}{}}%
+\CSLLeftMargin{31. }%
+\CSLRightInline{Reggiani A, Bucci P, Russo G. Accessibility and
+impedance forms: Empirical applications to the german commuting network.
+International Regional Science Review. 2011;34: 230--252.
+doi:\href{https://doi.org/10.1177/0160017610387296}{10.1177/0160017610387296}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-soukhovTTS2016RDataSet2023}{}}%
+\CSLLeftMargin{32. }%
+\CSLRightInline{Soukhov A, Páez A. {TTS}2016R: A data set to study
+population and employment patterns from the 2016 transportation tomorrow
+survey in the greater golden horseshoe area, ontario, canada.
+Environment and Planning B: Urban Analytics and City Science. 2023;
+23998083221146781.
+doi:\href{https://doi.org/10.1177/23998083221146781}{10.1177/23998083221146781}}
+
+\end{CSLReferences}
+
+\nolinenumbers
 

### 2023-08-24 Journal-manuscript/data-prep/00-OD-preparation.Rmd
Submitted this version to journal 1

---
title: "00-OD-preparation"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

Here we load the libraries we will be using in this .Rmd file:
```{r}
library(tidyverse) 
library(readxl) #imports excel files
library(geojsonio) #converts geoJSON files to R classes and vice versa
library(sf) # sf= simple features! These are spatial objects in R, read here: https://r-spatial.github.io/sf/index.html
library(lubridate) #for time formating
library(ggspatial)
```

## Import data

```{r}
OD_v <- read_excel("data-raw/EDM2018VIAJES.xlsx") #here we import the trips file, by default this function imports the first sheet in an excel file. This .Rmd file is saved in the same place as the .xlsx file so we just need to write the name of the file we want to read. run ' help(read_excel) ` to see the documentation of this function.

summary(OD_v) #descriptive stats of each row
```

You can also view a few rows of the 'OD' table like this:
```{r}
head(OD_v, 12) #the first 12 rows
```

I'm interested in making the table spatial! It currently does not have coordinates but we can add from the Madrid Region shapefile which can be downloaded from the [Open Data Portal Madrid Regional Transport Consortium](https://datos.crtm.es/datasets/zonificacionzt1259/explore?location=40.420959%2C-3.695073%2C12.93):
```{r}
# # here we download the madrid zona shapefile using the API GeoJSON resource! neat let's download ZT1259 and the lower resolution ZT208
# ZT1259 <- geojson_read("https://opendata.arcgis.com/datasets/97f83bab03664d4e9853acf0e431d893_0.geojson",  what = "sp")
# ZT1259 <- st_as_sf(ZT1259) %>% st_transform(crs = 4326) #transform it into a 'sf' class object and transform coordinates to WGS84
# 
# # downloaded ZT208 from here, the API link doesn't work for our application: https://hub.arcgis.com/datasets/crtm::zonificacionzt208/about
# 
# #let's save both to data-raw just in case this link/resource doesn't work in the future
# st_write(ZT1259, "data-raw/ZT1259.shp")
zn <-st_read("data-raw/ZT1259.shp")
#ZT208 is now in the data-raw, I’ve 
#let's read ZT208 and call it zn208, this will be our main resolution now
zn208 <- st_read("data-raw/ZonificacionZT208.shp")
zn208 <- st_as_sf(zn208) %>% st_transform(crs = 4326) #transform it into a 'sf' class object and transform coordinates to WGS84
```

let's add the CD_ZT208 values to the zn object so that each ZT1259 is associated with a CD_ZT208.
```{r}
sf::sf_use_s2(FALSE)
zn_u <- st_join(zn, zn208,  join = st_intersects, left = TRUE, largest = TRUE) #NOTE THIS IS KEY

zn_u <- zn_u %>% st_drop_geometry() %>% 
  transmute(CD_ZT208,
            ZT1259)
```
Now since that was spatially joined, let's double check the association using the associated CD_ZT208 IDs associated to each household located in a ZT1259 origin! We pulled this data from the household level data ('EMD2018HOGARES').

Note: Why we must  use the st_join dataset is because the associated household level data is not complete, i.e., only ~900 (of the 1259 ZTs) have an associated CD_ZT208. We need ALL ZTs to have an associated CD_ZT208 or the travel survey won't be able to completely be transfered to the CD_ZT208 resolution and we'll miss data.

```{r}
zn208_key <- read_excel("data-raw/ZT1259-CD_ZT208-associated.xlsx")
t <- zn_u %>% merge(zn208_key, by="ZT1259", all = T) %>% 
  mutate(IN = ifelse(CD_ZT208.x == CD_ZT208.y | is.na(CD_ZT208.y), 1, 0)) 
t
sum(t$IN)
#so since the sum is 1259, it means our join equals the associated join + with the NAs populated. 
```
Now importing a list of all the municipalities within the Madrid region and adding them to the zone shape file:
```{r}
t <- read_excel("data-raw/ZonificacionZT1259 complete.xlsx") %>% dplyr::select(ZT1259, Municipality)

zn_u <- zn_u %>% merge(t, by = c("ZT1259"), all.x=TRUE)
```

Let's associate the CD_ZT208 IDs to the trip data, starting with origin trips:
```{r}
OD_v <- OD_v %>% merge(zn_u, by.x = "VORIZT1259", by.y = "ZT1259", all.x=T)

OD_v %>% filter(is.na(CD_ZT208)) #some ZT1259 from the travel survey don't exist in the ZT1259 spatial object! (1,091 trips out of the total 222,744) So within these ZTs we can't associate a CD_ZT208 since they don't have a spatial location... I did not previously notice that these ZT1259 zones existed (e.g., 02-000-000, 01-000-000). These origins don't originated from within Madrid but their destinations are in Madrid.

# And as we can see, only 3 of these trips purposes' are to come home (a destination in madrid)
OD_v %>% filter(is.na(CD_ZT208) & MOTIVO_PRIORITARIO == 1)
# 386 of these trips purposes' are to come to work (a destination in madrid)
OD_v %>% filter(is.na(CD_ZT208) & MOTIVO_PRIORITARIO == 2)
# 205 of these trips purposes' are to come for a personal matter (a destination in madrid)
OD_v %>% filter(is.na(CD_ZT208) & MOTIVO_PRIORITARIO == 10)

#(some examples). Since these trips don't originate in Madrid we don't care about them from the 'supply' side but we do care about them from the 'demand' side (they increase the number of opportunities IN madrid).

#rename column
OD_v <- OD_v %>% rename(VORI_CD_ZT208 = CD_ZT208,
                        VORI_Municipality = Municipality)
```

Now destination trips:
```{r}
OD_v <- OD_v %>% merge(zn_u , by.x = "VDESZT1259", by.y = "ZT1259", all.x=T)

OD_v %>% filter(is.na(CD_ZT208)) #some ZT1259 from the travel survey don't exist in the ZT1259 spatial object! (916 trips out of the total 222,744) So within these ZTs we can't associate a CD_ZT208 since they don't have a spatial location... I did not previously notice that these ZT1259 zones existed (e.g., 02-000-000, 03-000-000). These destinations are not within Madrid but their origins are in Madrid.

# And as we can see,  0 of these trips purposes' are home related ( i.e., home outside of Madrid are not contained in this data set)
OD_v %>% filter(is.na(CD_ZT208) & MOTIVO_PRIORITARIO == 1)
# 397 of these trips purposes' are work related ( i.e., work trips that go outside of Madrid are not contained in this data set)
OD_v %>% filter(is.na(CD_ZT208) & MOTIVO_PRIORITARIO == 2)
# 177 of these trips purposes' are to leave for a personal matter (a destination outside of madrid)
OD_v %>% filter(is.na(CD_ZT208) & MOTIVO_PRIORITARIO == 10)

#(some examples). These trips end outside of Madrid, so these trips increase competition (supply side demand, i.e. more population) but don't add opportunities in Madrid.

#rename column
OD_v <- OD_v %>% rename(VDES_CD_ZT208 = CD_ZT208,
                        VDES_Municipality = Municipality)
```

Add time:
```{r}
OD_v <- OD_v %>%
  mutate(TIEMPO_VIAJES =
           (as.numeric(OD_v$VDESHORAFIN) - 
              as.numeric(OD_v$VORIHORAINI))) %>%
  mutate(TIEMPO_VIAJES = ifelse(TIEMPO_VIAJES == 0, 0.1, TIEMPO_VIAJES)) # set all 0min travel times to 0.1 min
summary(OD_v$TIEMPO_VIAJES)
```

These following chunks will be removed as they apply to ZT1259 and replaced with Javier's manual analysis applied to ZT208. A note though, it contains a beautiful plot so come back for the code.

<!-- ```{r} -->
<!-- ALLLEZ <- st_read("data-raw/Zona_Bajas_Emisiones_MADRID.shp") -->
<!-- ``` -->

<!-- Next, we will add a column which indicates if a zone is in a LEZ or is located within Madrid City Centre (this area contains the LEZ and is contained within the Municipality of Madrid): -->
<!-- ```{r} -->
<!-- #IDS that correspond to CC - city center and LEZ - low emission zone -->
<!-- CC_IDS <- c("079-01", "079-02", "079-03", "079-04", "079-05", "079-06", "079-07") -->
<!-- LEZ_IDS <- c("079-01-001", "079-01-002", "079-01-003", "079-01-004", "079-01-005", "079-01-006", "079-01-007", "079-01-008", "079-01-009", "079-01-010", "079-01-012", "079-01-013", "079-01-014", "079-01-015", "079-01-016", "079-01-017", "079-01-018", "079-01-019","079-01-020", "079-01-021", "079-01-022", "079-01-023") -->

<!-- #manually added column to zn which identifies if an ID is in the city center (1) or not (0). This searches for any instance of "079-01-XXX", "079-02-XXX" etc. -->
<!-- zn$CC <- ifelse(grepl(paste(CC_IDS, collapse = "|"), zn$ZT1259), 1,0) -->

<!-- #more simple to use the mutate function here since the string of IDs is complete for LEZ -->
<!-- zn <- zn %>% mutate(LEZ = ifelse(ZT1259 %in% LEZ_IDS, 1, 0)) -->

<!-- ``` -->

<!-- create a bounding box for the plot: -->
<!-- ```{r} -->
<!-- #create the bounding box for the Toronto region + a little extra -->
<!-- bbox_new1 <- st_bbox(zn %>% filter(CC == 1)) # current bounding box -->

<!-- xrange <- bbox_new1$xmax - bbox_new1$xmin # range of x values -->
<!-- yrange <- bbox_new1$ymax - bbox_new1$ymin # range of y values -->

<!-- bbox_new1[1] <- bbox_new1[1] - (0.2 * xrange) # xmin - left -->
<!-- bbox_new1[3] <- bbox_new1[3] + (0.2 * xrange) # xmax - right -->
<!-- bbox_new1[2] <- bbox_new1[2] - (0.05 * yrange) # ymin - bottom -->
<!-- bbox_new1[4] <- bbox_new1[4] + (0.05 * yrange) # ymax - top -->

<!-- bbox_CityCentre <- bbox_new1 %>%  # take the bounding box ... -->
<!--   st_as_sfc() -->
<!-- ``` -->

<!-- plot: -->
<!-- ```{r} -->
<!-- plot_rmad <- ggplot() + -->
<!--   geom_sf(data = zn, fill = "light grey", col = "dark grey", size=0.5) + #all of madrid region -->
<!--   geom_sf(data = zn %>% filter(Municipality == "Madrid"), aes(fill = "Municipality of Madrid"), col = "yellow3", size=0.5, show.legend = "polygon") + # madrid municiality -->
<!--   geom_sf(data = zn %>% filter(CC == 1), aes(fill =  "City centre"), col = NA, show.legend = "polygon") + # madrid city centre -->
<!--   geom_sf(data = zn %>%  filter(LEZ == 1), aes(fill = "Low emission zone"), col = NA, show.legend = "polygon") + #madrid low emission zone -->
<!--   geom_sf(data = bbox_CityCentre, fill = NA, col = "black", size = 0.5)+ -->
<!--   scale_fill_manual(labels = c("City centre", "Low emission zone", "Municipality of Madrid"), -->
<!--                       values = c("green", "red", "yellow")) +  -->
<!--   annotation_north_arrow(location = "tl", # north arrow for both the main plot and inset -->
<!--                          height = unit(0.8, "cm"),  -->
<!--                          width = unit(0.8, "cm"), -->
<!--                          style = north_arrow_orienteering(line_width = 0.25, -->
<!--                                                           line_col = "dimgrey",  -->
<!--                                                           fill = c("white","dimgrey"))) + -->
<!--   annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset -->
<!--                    height = unit(0.15, "cm")) + -->
<!--   scale_y_continuous(breaks = c(40, 40.4, 40.8, 41.2)) +  -->
<!--   scale_x_continuous(breaks = c(-3.0, -3.8, -4.6)) + -->
<!--   guides(fill=guide_legend(title=NULL)) + -->
<!--   theme_minimal() + theme(legend.position = c(0.25, .16))  -->

<!-- #just the city center and LEZ -->
<!-- plot_ccmad <- ggplot() + -->
<!--   geom_sf(data = zn %>% filter(CC == 1), aes(fill = "City centre"), col = "green3") + # madrid city centre -->
<!--   geom_sf(data = zn %>%  filter(LEZ == 1), aes(fill = "Low emission zone"), col = "red4") + #madrid low emission zone -->
<!--   geom_sf(data = bbox_CityCentre, fill = NA, col = "black", size = 0.5)+ -->
<!--   scale_fill_manual(labels = c("City centre", "Low emission zone"), -->
<!--                     values = c("green", "red")) +  -->
<!--   annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset -->
<!--                    height = unit(0.15, "cm"), -->
<!--                    pad_y = unit(0.05, "cm")) + -->
<!--   guides(fill=guide_legend(title=NULL)) + -->
<!--   theme_void() + theme(legend.position = "none")  -->

<!-- cowplot::plot_grid(plot_rmad, plot_ccmad, ncol = 2, rel_widths = c(2.5,1))  -->
<!-- ``` -->

<!-- ```{r} -->

<!-- ggplot() + -->
<!--   geom_sf(data = zn) + -->
<!--   geom_sf(data = ALLLEZ, fill="red")  -->
<!-- ``` -->

## Exploring the data - filtering OD trips, home to work

```{r}
OD_v_Oc_Dt <- OD_v %>% filter(VORI == "1" & VDES == "2") #only filter in origin casa trips and destination trabajo

count(OD_v_Oc_Dt) # from the 222,744 trips we now have 27,796 trips

summary(OD_v_Oc_Dt$VORI_CD_ZT208)
summary(OD_v_Oc_Dt$VDES_CD_ZT208)
#worrying..
```

What can we say about these work trips? (commented out for now)

<!-- Well, the majority of trip frequency is "A diario, de lunes a viernes" (Daily, from Monday to Friday - 1) -->
<!-- ```{r} -->
<!-- hist(OD_v_Oc_Dt$VFRECUENCIA) -->
<!-- ``` -->
<!-- The average destination arrival time (Destino. Hora llegada) is: -->
<!-- ```{r} -->
<!-- Dtrips <- as.POSIXct(OD_v$VDESHORAFIN, format="%H%M") %>% format("%H:%M") # formatting and removing date so we just how time -->
<!-- seconds_to_period(mean(period_to_seconds(hm(Dtrips)), na.rm=T)) # finding mean -->
<!-- ``` -->
<!-- 9:48am arrival.. not very helpful. -->

<!-- Let's see the distribution: -->
<!-- ```{r} -->
<!-- hist(OD_v_Oc_Dt$VDESHORAFIN%>% as.numeric(), breaks = 24) -->
<!-- ``` -->
<!-- So we see that 8AM, and 9AM arrival times are most frequent. -->

<!-- We also know that the majority do make this trip by personal car "¿Dispone de vehículo particular para hacer este viaje?" (1): -->
<!-- ```{r} -->
<!-- hist(OD_v_Oc_Dt$VVEHICULO) -->

<!-- OD_v_Oc_Dt$VVEHICULO %>% typeof() -->
<!-- ``` -->

<!-- Let's see how long are short-trips are (by distance, assuming in km units): -->
<!-- ```{r} -->
<!-- t <- OD_v_Oc_Dt %>% filter(DISTANCIA_VIAJE <= 60) %>% dplyr::select(DISTANCIA_VIAJE) %>% unlist() %>% as.numeric()  -->
<!-- hist(t) -->
<!-- ``` -->

<!-- As you can see though there are some long distance trip (greater than 60km). There are 101 trips which are between 60km to 600km and 1 trip greater than 600km. For perspective though, these trips make up onl 0.3% of total trips taken ( ~100 trips out of ~28,000 trips). -->
<!-- ```{r} -->
<!-- t <- OD_v_Oc_Dt %>% filter(DISTANCIA_VIAJE >= 60 & DISTANCIA_VIAJE <= 600) %>% dplyr::select(DISTANCIA_VIAJE) %>% unlist() %>% as.numeric()  -->
<!-- hist(t, breaks = 50) -->

<!-- t <- OD_v_Oc_Dt %>% filter(DISTANCIA_VIAJE >= 600) %>% dplyr::select(DISTANCIA_VIAJE) %>% unlist() %>% as.numeric()  -->
<!-- hist(t, breaks = 50) -->
<!-- ``` -->

<!-- Let's check the primary mode:  -->
<!-- ```{r} -->
<!-- hist(OD_v_Oc_Dt$MODO_PRIORITARIO) -->
<!-- ``` -->

<!-- MODO_PRIORITARIO -> 11, 12, 13, 14, 15, 16 correspond to: -->
<!-- 11 '11. Private car - driver -->
<!-- 12 '12. Company car - driver -->
<!-- 13 '13. Rental car without driver -->
<!-- 14 '14. Private car - passenger -->
<!-- 15 '15. Company car- passenger -->
<!-- 16 '16. Car rental with driver - passenger -->

## Making OD home to work trips spatial

Here we add # of origin trips to the zona sf object, so we see where trips are coming from.

First, we count the number of trips coming to each destination:
```{r}
Dtrips_zona <- OD_v_Oc_Dt %>% group_by(VDES_CD_ZT208)  %>% summarise(Dtrips = (n())) # we group by the Destination ZT1259 ID and then we summarize the number of rows that have the same ZT using the summarize() function and the n() function to create a new 'Dtrips' column.

head(Dtrips_zona,12)

sum(Dtrips_zona$Dtrips)
```

Now, we add this new column to the zonas sf object:
```{r}
zn_t <- zn208 %>% merge(Dtrips_zona, by.x = "CD_ZT208", by.y="VDES_CD_ZT208")

sum(zn_t$Dtrips)
```

a simple plot of the number of DESTINATION home-based trips:
```{r}
ggplot() +
  geom_sf(data = zn_t,
          aes(fill = Dtrips),
          color  = NA) +
  scale_fill_distiller(palette = "Spectral")
```

For comparison... lets plot the number of ORIGIN home-based trips:
```{r}
Otrips_zona <- OD_v_Oc_Dt %>% group_by(VORI_CD_ZT208)  %>% summarise(Otrips = (n())) # we group by the origin ZT1259 ID and then we summarize the number of rows that have the same ZT using the summarize() function and the n() function to create a new 'Otrips' column.

zn_t <- zn_t %>% merge(Otrips_zona, by.x = "CD_ZT208", by.y="VORI_CD_ZT208", all.x = TRUE)

ggplot() +
  geom_sf(data = zn_t,
          aes(fill = Otrips),
          color  = NA) +
  scale_fill_distiller(palette = "Spectral")
```
How many trips come into the LEZ from outside of Madrid municipality and how many of those are car trips? Just curious.

First, let's add a column which correspond to if the trip is made by car or not my car.
```{r}
OD_v_Oc_Dt <- OD_v_Oc_Dt %>% mutate(bycar = ifelse(MODO_PRIORITARIO %in% c(11, 12, 13, 14, 15, 16),
                                                                         1, 0),
                                    nocar = ifelse(!(MODO_PRIORITARIO %in% c(11, 12, 13, 14, 15, 16)),
                                                                         1, 0))
```

Next, let's split the data - one subset of all trips that don't go into the LEZ, and another subset of trips that DO go into the LEZ
```{r}
OD_v_Oc_Dt_noDesLEZ <- OD_v_Oc_Dt %>% filter(VDES_Municipality != "Madrid")
OD_v_Oc_Dt_DesLEZ <- OD_v_Oc_Dt  %>% filter(VDES_Municipality == "Madrid")

count(OD_v_Oc_Dt_DesLEZ)/count(OD_v_Oc_Dt)*100
```
we can see that trips into Madrid municipality is the MAJORITY of trips, 60%. Now... how many of these are car trips?

```{r}
sum(OD_v_Oc_Dt_DesLEZ$bycar)/count(OD_v_Oc_Dt_DesLEZ)
```
40% of these LEZ trips are made by car! This could be a huge GHG emission savings... and may have a huge impact on equity.

Let's plot where these car LEZ trips are coming from:
```{r}
D2trips_zona <- OD_v_Oc_Dt_DesLEZ %>% group_by(VORI_CD_ZT208)  %>% summarise(Dtrips = (n()),
                                                                              bycar = sum(bycar),
                                                                              nocar = sum(nocar))

zn_t <- zn208 %>% merge(D2trips_zona, by.x = "CD_ZT208", by.y="VORI_CD_ZT208")
```

plots:
```{r}
# magnitude of trips made by car 
ggplot() +
  geom_sf(data = zn_t,
          aes(fill = (bycar) ),
          color  = NA) +
  scale_fill_distiller(palette = "Spectral") +
  labs(title = "number of work trips by origin (excluding those that originate in LEZ) made by car ")
```

% of car trips which go into the LEZ 
```{r}
# Mag
ggplot() +
  geom_sf(data = zn_t,
          aes(fill = (bycar/Dtrips * 100) ),
          color  = NA) +
  scale_fill_distiller(palette = "Spectral") +
  labs(title = "% work trips (excluding those that originate in LEZ) made by car ")
```

## Saving data
```{r}
save(zn, file = "data/zn1259.RData")
save(zn208, file = "data/zn208.RData")
save(zn_u, file = "data/zn_u.RData")
save(OD_v, file = "data/OD_v.RData")
```

### 2023-08-24 README.Rmd
Submitted this version to journal 1

--- 2022-04-01 README.Rmd
@@
 
-# Madrid zonas de bajas emisiones (low emission zones)
+# Multimodal Spatial Availability: A Singly-Constrained Measure of Competitive Accessibility by Multiple Modes
 
-The repository is home to the analysis of low emissions zones in Madrid using spatial availability measure.
+The repository is home to the analysis of multi-modal extension of spatial availability.
 
-Currently, the following raw data from Madrid is contained here:
-
--  "parque_vehiculos_2021"  includes vehicle fleet characteristics of Madrid City. Of interest: the district and the environmental label associated to each vehicle.
-
-- several excel documents from the Home Mobility Survey of the Community of Madrid.
-These documents include information like citizens' socioeconomic characteristics, origin and destination of the trip, transport mode, distance, time required, activities carried out at origin and destination, etc.
-
-- A figure of Madrid spatial context.
+_An increasing number of studies within the domain of transportation planning are concerned with the inequities in accessibility to opportunities.
+A dimension of these inequities arises from differences in access by mode type (e.g., commuting using a car as opposed to transit).
+However, methods implemented in current accessibility literature are lacking within the context of multimodal analysis.
+This paper presents an extension of spatial availability, a singly-constrained competitive accessibility measure, for the context of multimodal accessibility analysis.
+We first illustrate the features of spatial availability that lend itself to multimodal analysis.
+We then demonstrate its use on the case study of Low Emission Zones in Madrid (Spain) and highlight how this policy intervention changes the accessibility of populations using different modes.
+In summary, spatial availability can be used to create and interpret multimodal policy intervention scenarios unlike previous methods: this creation and interpretation can help regions envision a more sustainable and equitable access-to-opportunity landscape._
 

### 2023-08-25 Journal-manuscript/Journal-manuscript.Rmd
Submitted this version to journal 1
Added title page as requested by editor

@@
 ---
-title: "Multimodal Spatial Availability: A Singly-Constrained Measure of Competitive Accessibility by Multiple Modes"
+title: "Multimodal spatial availability: a singly-constrained measure of competitive accessibility considering multiple modes"
+
 author:
   - name: Anastasia Soukhov
-    note: Corresponding Author
+    corresponding: true
     position: PhD Candidate
-    email: soukhoa@mcmaster.ca
-    affiliation: Department of Earth, Environment and Society, McMaster University, Canada
+    email: soukhoa@mcmaster.ca (AS)
+    affiliation: McMaster
   - name: Javier Tarriño-Ortiz
@@
     email: javier.tarrino.ortiz@upm.es
-    affiliation: Universidad Politécnica de Madrid, Spain
+    affiliation: Madrid
   - name: Julio A. Soria-Lara
@@
     email: julio.soria-lara@upm.es
-    affiliation: Urban and Regional Planning, Universidad Politécnica de Madrid, Spain
+    affiliation: Madrid
   - name: Antonio Páez
@@
     email: paezha@mcmaster.ca
-    affiliation: Department of Earth, Environment and Society, McMaster University, Canada
+    affiliation: McMaster
+address:
+  - code: McMaster
+    address: Department of Earth, Environment and Society, McMaster University, Hamilton, Canada
+  - code: Madrid
+    address: Centro de Investigación del Transporte (TRANSyT), Universidad Politécnica de Madrid, Madrid, Spain
+
 abstract: |
@@
 
-keywords: ["Multimodal", "Accessibility", "Equity", "Policy scenarios", "Low emission zones"]
+keywords: ["Multimodal", "Competitive", "Accessibility", "Equity", "Policy scenarios", "Low emission zones"]
 
@@
 
-# Abstract
-An increasing number of studies within the domain of transport are concerned with the inequities in accessibility to opportunities.
-A dimension of these inequities arise from differences in access by mode type (e.g., the number of work opportunities that can be reached using a car as opposed to transit in a city).
-However, methods assessing multimodal accessibility in the literature fall short as aspects of competition for opportunities and the explicit methodological acknowledgement of opportunities being _finite_ are lacking.
-In this vein, this paper presents an extension of _spatial availability_, a singly-constrained competitive accessibility measure, for the context of multimodal accessibility analysis.
-We first illustrate the features of spatial availability that lends itself to multimodal analysis.
-We then demonstrate its use on the case study of Low Emission Zones in Madrid (Spain) and highlight how this policy intervention changes the accessibility for locations where populations use different modes.
-In summary, spatial availability can be used to create and interpret multimodal policy intervention scenarios unlike previous methods: this creation and interpretation can help regions envision a more sustainable and equitable access-to-opportunity landscape by better identifying differences in accessibility afforded by different modes.
-
 # Introduction
@@
 This research was funded by the Canada Graduate Scholarship - Doctoral Program (CGS D) provided by the Social Sciences and Humanities Research Council (SSHRC) and Project Mobilizing Justice, also supported by SSHRC.
-All work is fully-reproducible and available within this [GitHub repository](https://github.com/soukhova/Madrid-ZBE-LEZ).
+The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.
+
+All work is fully-reproducible and available within this [GitHub repository](https://github.com/soukhova/Multimodal-spatial-availability).
 

### 2023-08-25 Journal-manuscript/Journal-manuscript.tex
Submitted this version to journal 1
Added title page as requested by editor

@@
 \newcommand{\getAff}[1]{
-  \getIndex{#1}{}
+  \getIndex{#1}{McMaster,Madrid}
 }
@@
 {\Large
-\textbf\newline{Multimodal Spatial Availability: A Singly-Constrained
-Measure of Competitive Accessibility by Multiple
-Modes} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+\textbf\newline{Multimodal spatial availability: a singly-constrained
+measure of competitive accessibility considering multiple
+modes} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
 }
@@
 \\
-Anastasia Soukhov\textsuperscript{\getAff{Department of Earth,
-Environment and Society, McMaster University, Canada}},
-Javier Tarriño-Ortiz\textsuperscript{\getAff{Universidad Politécnica de
-Madrid, Spain}},
-Julio A. Soria-Lara\textsuperscript{\getAff{Urban and Regional Planning,
-Universidad Politécnica de Madrid, Spain}},
-Antonio Páez\textsuperscript{\getAff{Department of Earth, Environment
-and Society, McMaster University, Canada}}\\
+Anastasia Soukhov\textsuperscript{\getAff{McMaster}}\textsuperscript{*},
+Javier Tarriño-Ortiz\textsuperscript{\getAff{Madrid}},
+Julio A. Soria-Lara\textsuperscript{\getAff{Madrid}},
+Antonio Páez\textsuperscript{\getAff{McMaster}}\\
 \bigskip
+\textbf{\getAff{McMaster}}Department of Earth, Environment and Society,
+McMaster University, Hamilton, Canada\\
+\textbf{\getAff{Madrid}}Centro de Investigación del Transporte
+(TRANSyT), Universidad Politécnica de Madrid, Madrid, Spain\\
 \bigskip
+* Corresponding author: soukhoa@mcmaster.ca (AS)\\
 \end{flushleft}
@@
 % Use "Eq" instead of "Equation" for equation citations.
-\hypertarget{abstract}{%
-\section{Abstract}\label{abstract}}
-
-An increasing number of studies within the domain of transport are
-concerned with the inequities in accessibility to opportunities. A
-dimension of these inequities arise from differences in access by mode
-type (e.g., the number of work opportunities that can be reached using a
-car as opposed to transit in a city). However, methods assessing
-multimodal accessibility in the literature fall short as aspects of
-competition for opportunities and the explicit methodological
-acknowledgement of opportunities being \emph{finite} are lacking. In
-this vein, this paper presents an extension of \emph{spatial
-availability}, a singly-constrained competitive accessibility measure,
-for the context of multimodal accessibility analysis. We first
-illustrate the features of spatial availability that lends itself to
-multimodal analysis. We then demonstrate its use on the case study of
-Low Emission Zones in Madrid (Spain) and highlight how this policy
-intervention changes the accessibility for locations where populations
-use different modes. In summary, spatial availability can be used to
-create and interpret multimodal policy intervention scenarios unlike
-previous methods: this creation and interpretation can help regions
-envision a more sustainable and equitable access-to-opportunity
-landscape by better identifying differences in accessibility afforded by
-different modes.
-
 \hypertarget{introduction}{%
@@
 Council (SSHRC) and Project Mobilizing Justice, also supported by SSHRC.
+The funders had no role in study design, data collection and analysis,
+decision to publish, or preparation of the manuscript.
+
 All work is fully-reproducible and available within this
-\href{https://github.com/soukhova/Madrid-ZBE-LEZ}{GitHub repository}.
+\href{https://github.com/soukhova/Multimodal-spatial-availability}{GitHub
+repository}.
 

### 2023-08-25 Journal-manuscript/Title-page.Rmd
Added title page as requested by editor

---
title: "Multimodal spatial availability: a singly-constrained measure of competitive accessibility considering multiple modes"
author:
  - name: Anastasia Soukhov
    corresponding: true
    position: PhD Candidate
    email: soukhoa@mcmaster.ca (AS)
    affiliation: McMaster
  - name: Javier Tarriño-Ortiz
    position: PhD Graduate
    email: javier.tarrino.ortiz@upm.es
    affiliation: Madrid
  - name: Julio A. Soria-Lara
    position: Associate Professor
    email: julio.soria-lara@upm.es
    affiliation: Madrid
  - name: Antonio Páez
    position: Associate Professor
    email: paezha@mcmaster.ca
    affiliation: McMaster
address:
  - code: McMaster
    address: Department of Earth, Environment and Society, McMaster University, Hamilton, Canada
  - code: Madrid
    address: Centro de Investigación del Transporte (TRANSyT), Universidad Politécnica de Madrid, Madrid, Spain

output: rticles::plos_article
header-includes:
  - \usepackage{booktabs} 
  - \usepackage{sectsty} \sectionfont{\centering}
  - \renewcommand{\contentsname}{}\vspace{-2cm} 
  - \usepackage{pdfpages}2
# plos.csl can be download and used locally
csl: http://www.zotero.org/styles/plos
---

\includepdf[pages={-}]{Journal-manuscript.pdf}

\newpage

### 2023-10-27 Response-to-reviewers_R1.Rmd
Draft response to reviewers

---
title: "Response-to-reviewers_R1"
output: pdf_document
date: "2023-10-18"
---

Thank you for for the time and care taken by the editorial team and reviewers in providing feedback to this manuscript. No doubt, the revised version submitted is more articulate than the first. We've provided extensive feedback to all comments from each reviewer in \textcolor{blue}{blue} and directly quoted the revised portion of the manuscript as relevant in \textcolor{blue}{\emph{italicized blue}}

## Reviewer #1:

The article extends the concept of spatial accessibility and applies it to a fairness analysis of different modes of transportation. The article uses travel data from a week in Madrid as an example to analyze spatial accessibility in the city. However, the article has the following issues:

1. Lack of Innovation: The article's improvement on spatial accessibility mainly involves incorporating the ratio of travel frequencies as weight in travel costs. This improvement is relatively minor and the method used does not demonstrate superiority over traditional accessibility approaches.

\textcolor{blue}{We disagree. Particuarly, in the original manuscript of the Discussion section, it is made explicit the advantages that spatial availabiltity offers when considering the competitive accessibility to opportunities by different modes. The ratio of travel frequencies are incorporated, but the result is a proportional allocation of opportunities, based on their competitive travel-cost-advantage and relative-mode-using-population, back to each spatial unit. Ultimately, the sum of all spatial availability values for each mode sum up to the total in the region. This is not true for _any_ other type of accessibility measure. It is also not plainly obvious that this can be done using spatial availability; thus this stand alone paper is needed to showcase this. }

\textcolor{blue}{However, your comment demonstrates that the advantages of the method are prehaps not adequately showcased. For this reason, we revised the Introduction and Discussion to make this more clear. The following has been added..........}

2. Insufficient Experimental Data: The article uses questionnaire data from a week in Madrid, but lacks basic descriptions of the questionnaire data. Additionally, the daily travel data of approximately 30,000 trips is significantly limited, and there is no description of the criteria for selecting the dates. The typicality of the experimental data is questionable.

\textcolor{blue}{The experimental data is the best and most complete travel survey conducted in Madrid to date. In filtering for home-to-work travel, of the ~280,000 trips, ~30,000 remain. This is all we can do. Accessibility data isn't typically generalizable anyways, it is very context-specific. The purpose of using the experimental data is do demonstrate what we can presume to be true about LEZ. Specifically: that LEZ restrict car mobility hence you would expect to see higher accessibility for non-car modes. This is intiution, and through spatial availability this can be seen, specifically in using the intution behind spatial availability (which is the proportional allocation of opportunities based on population competition and modal competition)}

3. Lack of Data-Driven Analysis: In the analysis section, a large amount of data is used to analyze people's travel behavior in different areas. The analysis of spatial accessibility only considers the differences in the modes of travel mentioned earlier. The analysis results are somewhat one-sided, and examining the indicator from multiple perspectives would provide a more objective view. It is recommended to expand the dimensions of the analysis.

\textcolor{blue}{Honestly not sure what to say to this... Maybe we make it more clear that the empirical case is a snapshot of what occured (when the travel survey was taken). It reflects potential access. }

4. Disorganized Format: The article's methodology section contains numerous formulas and variables, but the definitions of these variables are unclear and difficult to read. There are also numerous formatting errors, and Table 1 is disorganized and unappealing. It is recommended to revise these issues.

\textcolor{blue}{Thank you! Will fix the table. And will streamline the formulas (but still want to make it clear to readers the steps involved in the processa)}

## Reviewer #2:

This paper has sound mathematical foundations and allows to answer its research question in an elegant way (how to measure competition for e.g. jobs based on spatial accessibility?). I also really appreciated the fact that the paper was very didactic. Still, I have concerns about the relevance of the paper for future research.

More specifically, the new measure proposed by the authors has clear limitations:
i) it focuses on the competition for jobs and is not useful for studying access to non-competitive or semi-competitive resources such as amenities.

\textcolor{blue}{ 1) jobs represents the majority of accessibility research (historically). It can also be used for amentities - with a different weighting system - this will be an application in the future. }

ii) it doesn't allow to study absolute gains or losses in accessibility from public transportation infrastructure improvements or changes.

\textcolor{blue}{ It can be used to do so! Unfortunately, we do not have travel data from before the LEZ implementation, but a before- and after- can certainly be done using spatial availability }

iii) the authors do not allow for modal shifts: they assume that the transport mode choice of households is fixed and cannot evolve due to e.g. transport infrastructure changes.

\textcolor{blue}{ Are you outlining our next paper for us?? Modal shift scenarios can be implemented }

These points limit the relevance of the new accessibility measure. The authors should, at least, specify these limitations early in the paper. The introduction should start by stating the precise research question the new accessibility measure is seeking to answer as well as its limitations). They should also justify, based on the literature, that competition for jobs is a key determinant of job market outcomes, and that there is strong inertia in mode choices.

Finally, the writing of the paper, and particularly of the abstract and introduction, should be improved. The abstract could state the broader relevance of the topic and summarize the results of the case study on LEZs. The introduction should start more directly by introducing the research question and its relevance and describing the new measure and its limitations.

\textcolor{blue}{ Thank you for your thoughtful comments. We have improved the introduction and conclusion to reflect this }

Minor comments:
i) what are the summary statistics p10 (car: 36 min, transit: 55 min,...)? I assume they correspond to the mean.
ii) I also have identified a few formatting issues (e.g. Fcij p10 and 4.72km2 p 8).

## Reviewer #3:

This manuscript extended the authors’ previous work spatial availability measure, which is a type of location-based accessibility measure that is both constrained and competitive compared to Hansen-type measure and Shen-type accessibility measure, into a multimodal framework. The new measure, multimodal spatial availability, strengthened the constrained (or finite) nature of opportunities, and the competitive nature among multimodal accessibility resulting from this constraint through a synthetic example and an empirical example of the LEZ in the city of Madrid. In conclusion, the authors demonstrated one restriction had impacted the spatial availability of opportunities for other modes using and proposed potential future uses in policy planning scenarios.
In general, the manuscript was logical and well-structured. The research problem was well defined. The data were available and quite supported the conclusion. The statistical analysis performed appropriately.

However, there are some issues:

Major issues:
Please demonstrate whether “car/motor & transit” and “bike & walk” are comparable or whether they are in an actual competitive relationship? For example, if I work 3km from where I live, maybe I will never choose to take a transit, I will always walk or ride. But if I work 20km from where I live, walking or riding to work seems impossible for me, I have to drive or take public transportation. Car/motor and transit can be in competitive relationship and people can choose which one they prefer, but not choose between motor and walk. This issue will also have an impact on the results of the research.

\textcolor{blue}{Great comment. We have made this more clear in the text with explaintion of the impedance function for the modes. The impedance functions reflect actual travel using each mode. Average travel times for car/motor and transit are longer than bike and walk. All people don't have access to all options - completely true. But the travel impedances reflect this real travel at an aggregate. And on average, it is assumes that these modes are all in direct competition -- this is a big assumptions. We've made this more clear. However, pending real data that demonstrates who uses what mode on what street / what type of person / who can actually go to what jobs, etc (micro travel data), this is all we have. So under this 'average' assumption, we can see the competitive advantage of one mode to access all opportunities. }

Minor issues:
1. As for Fig 2 and Fig 3, please indicate the meaning of the gray color blocks in illustration.
2. Please change the color scheme of fig 2. The red color scheme makes the LEZ centro area boundary, which is also in red, not visible.

### 2023-10-27 Journal-manuscript/data-prep/01-Travel-time-calculation.Rmd
Cloned GitHub repo and worked with renv to make sure that the project is in a consistent state.

--- 2023-07-30 TRB-manuscript/data-prep/01-Travel-time-calculation.Rmd
@@
 library(tidyr)
-library(tmap)
+#library(tmap)
 library(zoo) # for rollapplyr
@@
 options(future.globals.maxSize = Inf)
-tmap_mode("view")
+#tmap_mode("view")
 ```

### 2023-11-01 Response-to-reviewers_R1.Rmd
Response to reviewers, R1

@@
 
-\textcolor{blue}{We disagree.
-Particuarly, in the original manuscript of the Discussion section, it is made explicit the advantages that spatial availabiltity offers when considering the competitive accessibility to opportunities by different modes.
-The ratio of travel frequencies are incorporated, but the result is a proportional allocation of opportunities, based on their competitive travel-cost-advantage and relative-mode-using-population, back to each spatial unit.
+\textcolor{blue}{Thank you for this comment but we disagree with this characterization of the our manuscript.
+
+The spatial availability measure we extend for multi-modal application does not use _"the ratio of travel frequencies as weight in travel costs"_.
+In our empirical example, we use mode-segmented origin-destination data to estimate travel impedance based on trip lengths.
+The travel impedance can be understood as "travel costs".
+In this way, the travel frequencies are reflected as a weight in the calculation of travel impedance itself as well as reflected in the population factor and job opportunities variable.
+So your comment inaccurately characterizes our method.
+Spatial availability consists of a total balancing factor (based on travel impedance and population) that proportionally allocates opportunities.
+In this manuscript, we demonstrate how the mode-specific travel impedance and populations can be used to calculate _competitive_ accessibility, by mode (considering competition by all modes based on the travel impedance) and _constrained_ to add up to the total sum of opportunities available in the region.
+In this way, each spatial availability value is a proportion of the total sum of opportunities in the region.
+
+Particularly, in the "Discussion and conclusions" section, the advantages that spatial availability offers when considering the competitive accessibility to opportunities by different modes.
+In a way, the ratio of travel frequencies are incorporated, but the result is a proportional allocation of opportunities, based on their competitive travel-cost-advantage and relative-mode-using-population, back to each spatial unit.
 Ultimately, the sum of all spatial availability values for each mode sum up to the total in the region.
 This is not true for _any_ other type of accessibility measure.
-It is also not plainly obvious that this can be done using spatial availability; thus this stand alone paper is needed to showcase this.
+It is also not plainly obvious that this can be done using spatial availability for multiple modes; thus this paper fills this gap.
+
+Further, we believe we do demonstrate superiority of our measure throughout the manuscript.
+For example,we summarize in the Discussion and Conclusions: _"With spatial availability, the number of opportunities that are captured (of the total opportunities in the region) by each mode can be individually calculated.
+From there, the difference between how many spatially available opportunities one mode captures versus another can be investigated.
+This is the advantage of the spatial availability measure, particularly its multimodal extension."_.
+Ultimately, the spatial availability values are a product of proportional allocation: each value is a proportion of the total opportunities in the region.
+Put in another way, the total opportunities in the region are _finite_, and this is how spatial availability is introduced against un/constrained and non/competitive accessibility measures.
+The conceptualization of opportunities as _finite_ within the spatial availability measures gives spatial availability values inherent meaning that is not available for any other type of accessibility measure.
+This meaning allows us to explore, for instance, the difference in spatial availability values between modes (i.e., job opportunities that are spatially available based on the travel-impedance and populations of all modes) and the spatial availability per capita benchmarks as done in the empirical examples.
+
+However, we agree we could make this more advantage of spatial availability more explicitly clear in all sections, as such the "Introduction", "A review of multimodal accessibility measures", and "Discussion and Conclusions" section have been expanded and/or clarified.
+
+Introduction:
+_The question of how much one mode-using population can access at the expense of another mode-using population is a pertinent equity question in the evaluation of policy scenarios, especially those that impact modes differently.
+For instance, consider the impact of a low emission zone (LEZ).
+A LEZ is a policy of spatial and modal discrimination: the circulation of vehicles that are excessively polluting are restricted in specific regions.
+Depending on the level of vehicle restriction, LEZ have been shown to significantly improve air quality [@holmanReviewEfficacyLow2015] and air-quality related health outcomes[@margaryanLowEmissionZones2021].
+From the perspective of accessibility to opportunities, LEZ explicitly lower the access that restricted-vehicle using populations have to opportunities within the LEZ.
+However, if conceptualizing opportunities in a region as finite, one populations' lose is another populations' gain.
+The populations using non-restricted modes (e.g., walking, cycling, transit, and less-polluting vehicles) to travel into LEZ will have more access than before the LEZ implementation._
+
+_This evaluation is especially urgent as LEZ are in effect in cities globally; their reception [@tarrinoortizPublicAcceptabilityLow2021] and impact on certain activity types such as shopping [@tarrinoortizImpactLowEmission2023] have been mixed.
+They may also be having negative impacts on disadvantaged populations who have become mobility-restricted [@devrijNooneVisitsMe2022; @verbeekJustManagementUrban2022; @liottaWhatDrivesInequalities2023].
+Measures that evaluate the accessibility of modes given both constrained and competitive considerations are lacking in the literature, but are needed, to evaluate such an intervention's impact on accessibility gains and losses within a finite conceptualization._
+
+A review of multimodal accessibility measures:
+- extended the explanation of Hasen-type measure and the consideration of competition in accessibility (shen-type measure) with additional examples from the literature. See the revised sections.
+
+Discussion and Conclusions:
+XXXXXXXXXXX
+
 }
-
-\textcolor{blue}{However, your comment demonstrates that the advantages of the method are prehaps not adequately showcased.
-For this reason, we revised the Introduction and Discussion to make this more clear.
-The following has been added..........}
 
@@
 
-\textcolor{blue}{The experimental data is the best and most complete travel survey conducted in Madrid to date.
-In filtering for home-to-work travel, of the ~280,000 trips, ~30,000 remain.
-This is all we can do.
-Accessibility data isn't typically generalizable anyways, it is very context-specific.
+\textcolor{blue}{Thank you for this comment.
+To clarify, the empirical data used is a travel survey conducted by the City of Mardid.
+This is a conventional travel survey that is conducted in cities around the world.
+This experimental data is the best and most complete travel survey conducted in Madrid to date.
+I'm unsure where you got the 30,000 trip number from but I presume you are citing the maximum 'opportunity' or 'population' number in Figure 2 or 3?
+The more intense the colour, the higher the value it represents.
+The most intensely coloured TAZ have ~30,000 jobs while the least intensely coloured have ~1,000 or fewer (figure 2).
+The total number of jobs in the City are 847,574, this is also the sum of the total spatial availability in the City.
+
+Furthermore, results from accessibility data isn't typically generalization anyways, it is highly context-specific.
 The purpose of using the experimental data is do demonstrate what we can presume to be true about LEZ.
 Specifically: that LEZ restrict car mobility hence you would expect to see higher accessibility for non-car modes.
-This is intiution, and through spatial availability this can be seen, specifically in using the intution behind spatial availability (which is the proportional allocation of opportunities based on population competition and modal competition)}
+This is intuition, and through spatial availability this can be seen, specifically in using the intuition behind the multimodal extension of this method (which is the proportional allocation of opportunities based on population competition and modal competition)
+
+We have updated the manuscript to improve clarity for the readers:
+
+Introduction:
+_In Section 3, the spatial availability of four mode categories is calculated for the city of Madrid, Spain after the 2017 LEZ is implemented.
+Using mode-specific home-to-work origin-destination 2018 travel survey as empirical data, we demonstrate inequities of spatial availability within and outside the LEZ for each sub-population using car, transit, cycling and walking modes._
+
+Multimodal data and methods:
+_The 2018 Community of Madrid travel survey [@comunidaddemadridResultadosEDM20182020] is the source of data for this empirical example: it is a representative survey that reflects a snap-shot of the travel patterns for a typical day of the working week in 2018.
+Specifically, it includes 222,744 trips taken from a representative sample of 85,064 households across the traffic analysis zones (TAZ) representing the Community of Madrid (population of 6,507,184 over 3 years old) through population elevation factors._
+
+_In this empirical example, we only use a sub-set of the survey, specifically home-to-work trips, by all modes, to demonstrate multimodal spatial availability of employment opportunities.
+The total population who travels to work from an origin (i.e., the worker), and the destinations (i.e., the employment opportunity), for the City of Madrid (a sub-set of the Community of Madrid) is visualized in Figure \ref{fig:Fig2} and Figure \ref{fig:Fig3}.
+Both figures are displayed at the level of TAZ (the $i$ and $j$ zones) that correspond to the survey._...
+
+Captions to Figure 2 and Figure 3:
+- Jobs $O_j$ taken by people living and working in Madrid as reported by the home-to-work flows in the 2018 travel survey.
+- Figure 3: Population living and working in Madrid by four summarized modal categories $P^m_i$ as reported by the home-to-work flows in the 2018 travel survey.
+}
 
@@
 
-\textcolor{blue}{Honestly not sure what to say to this...
-Maybe we make it more clear that the empirical case is a snapshot of what occured (when the travel survey was taken).
-It reflects potential access.
-}
+\textcolor{blue}{Thanks for this comment. We believe with the additional detail added to the manuscript to address your and other reviewers' comments, overall clarity has been improved.
+
+In this work, we outline issues associated with accessibility measures stemming from the lack of competition considerations and being constrained: namely, neglecting competition can be an oversight when opportunities have a capacity, and neglecting to constraint the global measures makes interpretation in comparisons difficult.
+We elaborate on these issues in a synthetic example of a synthetic region with two populations, one using a faster mode and one using a slower mode to access opportunities.
+
+Having demonstrated the merits in how multimodal spatial availability considers competition and can be more easily interpreted as it is constrained (each value is a score that is a proportion of the total opportunities in the region), we then expand on an empirical example.
+The multimodal spatial availability of employment opportunities in Madrid, a city that has recently put in place a LEZ.
+We demonstrate how the measure can be used to understand differences in spatial availability between modes, and how the restriction on one mode could improve the spatial availability for others.
+The empirical case is used as a snapshot of the estimated spatial availability as informed by the origin-destination flows from the 2018 travel survey.
+It is a proof-of-concept on empirical data that spatially visuals what singly-constrained competitive accessibility to employment, by mode, looks like.
+
+As we believe your comment is referring to the dimensionality of the empirical example, we added the following sentences to the beginning of the Results section to demonstrate this intention clearly:
+_As it is worth re-iterating, the empirical example reflects a snap-shot of the calculated multimodal spatial availability using data from the 2018 travel survey.
+It is intended to visualize the spatial trends in availability of employment opportunities, by mode, and demonstrate how spatial availability can be interpreted to discuss the competitive advantage of lower travel impedance modes within Madrid Centro._}
 
@@
 
-\textcolor{blue}{Thank you! Will fix the table. And will streamline the formulas (but still want to make it clear to readers the steps involved in the processa)}
+\textcolor{blue}{Thank you for this comment, but we are not sure how to address all of these items.
+We have used the PLOS ONE template.
+All definitions of variables are included and variables are correctly included in mathematical notation in the original manuscript.
+As you did not provide us with specific line or page numbers indicating were errors may be, we took a careful read through of the manuscript, particularly focusing on the methodology, and attempted to streamline where we could.
+
+We also made sure Table 1 starts on a new page, when it runs across pages the formatting resulted in some disorganization.}
 
@@
 
-\textcolor{blue}{ 1) jobs represents the majority of accessibility research (historically).
-It can also be used for amentities - with a different weighting system - this will be an application in the future.
-}
+\textcolor{blue}{Thank you for this comment; Indeed, we have grappled with this idea for some time.
+At this point, we are convinced that _all_ opportunities have some sort of capacity.
+For instance, green spaces are often considered non-competitive, however, some standards are provided in the form of unit of green space per capita (e.g., [Natural England](https://redfrogforum.org/wp-content/uploads/2019/11/67-Nature-Nearby%E2%80%99-Accessible-Natural-Greenspace-Guidance.pdf) set an Accessible Natural Greenspace Standard recommending that everyone have at a minimum: a minimum of one ha of statutory Local Nature Reserves per thousand population).
+This is some amount of green space per person, that no one else is using at the same time.
+Though Local Nature Reserves have large capacities - they still have a capacity - and thus spatial availability can be used to measure competitive (and constrained) access to this sort of opportunity.
+As planning literature moves more towards the creation of standards, these standards can be used to directly inform results from spatial availability analysis.
+However, this idea is not discussed in the manuscript itself.
+
+<!--should I discuss this in the conclusion? Perhaps yes, alongside other standards. otherwise not sure what to change in the manuscript to address this comment. -->
+Additionally, even if this idea of all opportunities having a capacity is not conceptually discussed in the paper, this method is still relevant for many opportunity types.
+Historically, competitive opportunities have been extensively studied in accessibility literature.
+For instance, health-care and employment opportunities in the long tradition of competitive accessibility studies, e.g., [@taoInvestigatingImpactsPublic2020a; @maoMeasuringSpatialAccessibility2013; @shenLocationCharacteristicsInnercity1998].
+These opportunities are also more readily available in historically point-of-interest datasets and reported on in long-form censuses.
+Further, places of employment have been seen as proxies for _all_ types of activities (as they correspond to some sort of employment opportunity as activities typically are a service), see work by @hellingChangingIntrametropolitanAccessibility1998.}
 
@@
 
-\textcolor{blue}{ It can be used to do so! Unfortunately, we do not have travel data from before the LEZ implementation, but a before- and after- can certainly be done using spatial availability }
+\textcolor{blue}{Great point - spatial availability can certainly be used to capture absolute gains and losses.
+In fact, logically, the gains and losses produced by using a competitive and constrained measure allows for a clear interpretation.
+
+Unfortunately however, we do not quantify any before- or after- scenarios in this work.
+We only have access to a travel survey from after the LEZ implementation in Madrid for the time being.
+Furthermore, the empirical example is meant as a proof-of-concept of the multimodal spatial availability extension.
+
+We intend to use spatial availability in the scenario analysis of before- and after- LEZ implementation from a modal and socio-economic equity perspective.
+We've added this intuition and limitation to the conclusion:
+_In future work, we intend to use multimodal spatial availability to characterise the equity of modal accessibility losses and gains in varying LEZ policy scenario implementations._}
 
@@
 
-\textcolor{blue}{ Are you outlining our next paper for us?? Modal shift scenarios can be implemented }
+\textcolor{blue}{
+Excellent point.
+We have thought about this potential consideration.
+Spatial availability is certainly compatible with some sort of modal split model.
+It is a flexible framework that can take in different travel impedance curves as well as proportions of populations using those travel impedance (i.e., modal splits changing).
+However, that demonstrate is out of scope in this paper.
+
+The focus of this paper is on the methodological expansion of spatial availability to quantify multiple modal options and characterise their potential competitive advantage for opportunities.
+The empirical example is a demonstrate of this multimodal spatial availability extension.
+We've made it more clear in the conclusion that this is our intention for future work.
+We've also improved the Introduction and Abstraction to make it more clear that this is a methodolgically focused paper with an empirical case study as snap-shot demonstration of conditions in 2018 after the LEZ implementation.}
 
@@
 
+\textcolor{blue}{ 
+<!--Honestly not too sure about this one..
+the point about competition in job markets is sort of tangent to this research.
+And mode choice inertia isn't too relevant either _because_ we are expressing this as a 'snap shot' of conditions in 2018 based on how people _behave_.--> }
+
 Finally, the writing of the paper, and particularly of the abstract and introduction, should be improved.
@@
 
-\textcolor{blue}{ Thank you for your thoughtful comments. We have improved the introduction and conclusion to reflect this }
+\textcolor{blue}{ Thank you for your thoughtful comments. We have streamlined the Introduction based on your comments and other reviewers. We've also updated the Abstract.
+
+Regarding the limitations of the method, the limitations are in the details - specifically the travel impedance selected and the data used.
+The method itself is a framework.
+So we've augmented the Methodology and Results sections to more clearly express the assumptions embedded in the empirical example that limit it's interpretation.
+<!--Not too confident in this response...
+I feel like I should change more of the manuscript but not sure how as the comments aren't entirely relevant.
+I did not update the Abstract or Introduction too drastically.
+I like the flow and I think they accurately reflect what this paper does.-->}
 
@@
 i) what are the summary statistics p10 (car: 36 min, transit: 55 min,...)? I assume they correspond to the mean.
+\textcolor{blue}{
+Updated! Apologies, car: '36 min' corresponds to a mean of 36 minutes and then within the brackets additional descriptive statistics (minumum value, maximum value, etc.). }
 ii) I also have identified a few formatting issues (e.g. Fcij p10 and 4.72km2 p 8).
+\textcolor{blue}{
+Fixed! Thank you. }
 
@@
 
-\textcolor{blue}{Great comment.
-We have made this more clear in the text with explaintion of the impedance function for the modes.
-The impedance functions reflect actual travel using each mode.
-Average travel times for car/motor and transit are longer than bike and walk.
+\textcolor{blue}{
+
+Great comment. Whenever a destination can be reached by more than one mode, users of those mode are in competition for the same opportunities.
+
+We have improve the discussion of impedance function to make this more clear.
+The impedance functions for all four modes are not the same.
+They describe the travel behaviour of home-to-work trips (as informed by the 2018 travel surve).
+If you work 20 km from where you live, and its not possible to walk, you are not competing for jobs at that location with people who walk where you leave (because people don't walk).
+But you are competing with people who can walk (live at other closer destinations) to those jobs.
+
+Futhermore, average travel times for car/motor and transit are longer than bike and walk.
 All people don't have access to all options - completely true.
-But the travel impedances reflect this real travel at an aggregate.
-And on average, it is assumes that these modes are all in direct competition -- this is a big assumptions.
-We've made this more clear.
-However, pending real data that demonstrates who uses what mode on what street / what type of person / who can actually go to what jobs, etc (micro travel data), this is all we have.
-So under this 'average' assumption, we can see the competitive advantage of one mode to access all opportunities.
+But the travel impedances reflect this real travel at an aggregate based on all the trips for a mode.
+And on average, it is assumed that people at each origin that take a mode to a destination are in direct competition for opportunities (as opportunities are finite) -- and a part of the competition is defined by the mode-specific impedance function (the second part is the population balancing factor).
+This assumption, that all populations, no matter their mode, are competing for the same finite set of opportunity, is part of spatial availability.
+We've made this more clear in the text.
+
+Multimodal data and methods:
+
+_Notably, plots in Figure \ref{fig:Fig4} can be observed across modes if curious to know what trips are more likely to occur using what mode depending on trip length (as informed by the observed 2018 survey).
+Trips >5 minutes do not occur frequently for any mode, as such trips with short trip lengths are assigned a lower travel impedance_ $f^m(c_{ij}^m)$ _value.
+One reason that >5 minute trips do not frequently occur as a result of land-use (residential and job mismatch): not everyone lives in immediate proximity to their workplace.
+In terms of the non-motorized modes, shorter trips occurred more frequently overall for walking populations, particularly around 15 minute lengths, so a trip of approximately 15 minutes is assigned the highest_ $f^m(c_{ij}^m)$.
+_For biking populations, longer travel times are more common so though the highest_ $f^m(c_{ij}^m)$ _value also corresponds to approximately 15 minutes, the curve is more spread out and values decrease less rapidly at longer travel times than for the walk mode curve.
+A similar trend can be observed for the motorized modal options where transit mode is more spread out than car/motor mode.
+All in all, these observations demonstrate that, based on a given mode and travel time, a trip is more or less likely to occur and is accordingly represented in the cost of travel balancing factor_ $F_{ij}^m$ _for_ $V_i^m$.
 }
@@
 1. As for Fig 2 and Fig 3, please indicate the meaning of the gray color blocks in illustration.
+\textcolor{blue}{
+Fixed for all Figures, thank you. }
 2. Please change the color scheme of fig 2. The red color scheme makes the LEZ centro area boundary, which is also in red, not visible.
+\textcolor{blue}{
+WILL FIX THANK YOU <!-- and I must change the labels on the legends to be ranges -->}

### 2023-11-01 Journal-manuscript/Journal-manuscript_R1.Rmd
Response to reviewers, R1

--- 2023-08-25 Journal-manuscript/Journal-manuscript.Rmd
@@
 This paper presents an extension of spatial availability, a singly-constrained competitive accessibility measure, for the context of multimodal accessibility analysis.
-We first illustrate the features of spatial availability that lend itself to multimodal analysis.
+We first illustrate the features of spatial availability that lend itself to multimodal analysis using synthetic example.
 We then demonstrate its use on the case study of Low Emission Zones in Madrid (Spain) and highlight how this policy intervention changes the accessibility of populations using different modes.
@@
 ```
-
+<!-- Maybe change title from:
+Multimodal spatial availability: a singly-constrained measure of competitive accessibility considering multiple modes
+to:
+Multimodal spatial availability: a singly-constrained measure of competitive modal accessibility 
+-->
 # Introduction
 
-<!-- outlines the use of location-based accessibility measures in the literature: equity, planning -- esp. as it relates to modal competition (brief) -->
-Implementing urban policies that re-shape cities through accessibility gains (i.e., the _potential to interact_ with opportunities as a result of land-use mix and transport systems as originally defined by @hansenHowAccessibilityShapes1959) have been widely applied within the transportation literature and is increasingly discussed by planners [@levinsonTransportAccessManual2020; @gowerPlanningInnovationCity2022; @siddiqToolsTradeAssessing2021; @yanAccessibilityBasedPlanningAddressing2021].
-An important challenge in the identification of interventions that equitably transform cities is the effective evaluation of _trade-offs_: cities are complex and dynamic ecologies, and advantaging one component of the city can disadvantage another area, population, or sub-component In this way, policy evaluation should take a _systems_ approach [@fikselSustainabilityResilienceSystems2006].
-One way of considering systems is from the perspective of the _finite_.
+Implementing policies that re-shape cities through accessibility gains (i.e., the _potential to interact_ with opportunities as a result of land-use mix and transport systems as defined in @hansenHowAccessibilityShapes1959) have been widely applied within the transportation literature and increasingly discussed by planners [@levinsonTransportAccessManual2020; @gowerPlanningInnovationCity2022; @siddiqToolsTradeAssessing2021; @yanAccessibilityBasedPlanningAddressing2021].
+An important challenge in the identification of interventions that equitably transform cities is the effective evaluation of _trade-offs_: cities are complex and dynamic ecologies, and advantaging one component of the city can disadvantage another area, population, or sub-component.
+In this way, policy evaluation should take a _systems approach_ [@fikselSustainabilityResilienceSystems2006].
+One way of applying a systems approach is from the perspective of the _finite_.
 As an illustration, consider the amount of transport space within a city: the amount is typically finite so re-allocating road space away from one mode directly impacts the performance of the others (see the literature on road space reallocation e.g., @valencaMainChallengesOpportunities2021).
@@
 From the perspective of urban transport systems, location-based accessibility measures have been used in the context of policy evaluation.
-For instance, @leeMeasuringImpactsNew2018 assesses the transit accessibility gains to healthcare and employment opportunities for disadvantaged neighbourhood in Columbus, Ohio, USA after the transit system's re-design and introduction of a rapid bus system.
-However, a limitation of this study, like others that implement accessibility measures, is they do not calculate results under a _constrained_ framework i.e., one of _finity_.
+For instance, @leeMeasuringImpactsNew2018 assesses the transit accessibility gains to healthcare and employment opportunities for disadvantaged neighbourhood in the city of Columbus, USA after the transit system's re-design and introduction of a rapid bus system.
+A limitation of this study is they do not calculate results under a _constrained_ framework i.e., one of _finity_.
 The citizens of Columbus should experience quantitative accessibility gains - but is it at the expense of access using other modes?
+This limitation is similar to others that implement accessibility measures.
 As another example, @mohriClusteringMethodMeasuring2021 implements a modified cumulative opportunity measure to assess differences between private vehicle and transit system accessibility to jobs in Melbourne, but a similar question remains: does the accessibility afforded to the private vehicle using population come at the expense of accessibility losses to transit users?
@@
 For instance, @maoMeasuringSpatialAccessibility2013 applies a competitive measure, two-step floating catchment approach (2SFCA), for the case of access to healthcare services in Florida for both a multimodal network and a single modal network.
-While the differences in modal access are discussed, the question of how the advantage in access afforded by one mode over another impacts access for different mode users is unanswered.
-
-This question of how much one mode-using population can access at the expense of another mode-using population is a pertinent equity question in the evaluation of policy scenarios that are multimodal.
+While the differences in modal access are discussed, the question of how the advantage in access afforded by one mode over another impacts access for different mode users is not addressed.
+
+The question of how much one mode-using population can access at the expense of another mode-using population is a pertinent equity question in the evaluation of policy scenarios, especially those that impact modes differently.
 For instance, consider the impact of a low emission zone (LEZ).
-LEZ is a policy of spatial and modal discrimination: the circulation of vehicles that are excessively polluting are restricted in specific regions.
-In the recognition that opportunities are finite, the implementation of a LEZ explicitly reduces the access that the population using polluting vehicles has to opportunities.
-This restriction allows the population using other more sustainable modes to potentially have a higher level of access than before the LEZ implementation.
-This evaluation is especially urgent as LEZ are currently in effect in cities globally; their reception has been mixed [@tarrinoortizPublicAcceptabilityLow2021] and may be having negative impacts on disadvantaged populations who have become mobility-restricted [@devrijNooneVisitsMe2022; @verbeekJustManagementUrban2022].
-Measures that evaluate the accessibility of modes given both _constrained_ and _competitive_ considerations are lacking in the literature, but are needed, to evaluate such policy interventions impact on accessibility.
+A LEZ is a policy of spatial and modal discrimination: the circulation of vehicles that are excessively polluting are restricted in specific regions.
+Depending on the level of vehicle restriction, LEZ have been shown to significantly improve air quality [@holmanReviewEfficacyLow2015] and air-quality related health outcomes[@margaryanLowEmissionZones2021].
+From the perspective of accessibility to opportunities, LEZ explicitly lower the access that restricted-vehicle using populations have to opportunities within the LEZ.
+However, if conceptualizing opportunities in a region as _finite_, one populations' lose is another populations' gain.
+The populations using non-restricted modes (e.g., walking, cycling, transit, and less-polluting vehicles) to travel into LEZ will have _more_ access than before the LEZ implementation.
+
+This evaluation is especially urgent as LEZ are in effect in cities globally; their reception [@tarrinoortizPublicAcceptabilityLow2021] and impact on certain activity types such as shopping [@tarrinoortizImpactLowEmission2023] have been mixed.
+They may also be having negative impacts on disadvantaged populations who have become mobility-restricted [@devrijNooneVisitsMe2022; @verbeekJustManagementUrban2022; @liottaWhatDrivesInequalities2023].
+Measures that evaluate the accessibility of modes given both _constrained_ and _competitive_ considerations are lacking in the literature, but are needed, to evaluate such an intervention's impact on accessibility gains and losses within a _finite_ conceptualization.
 
 In @soukhovIntroducingSpatialAvailability2023, we introduce spatial availability, a type of location-based accessibility measure that is both _constrained_ and _competitive_.
-In this paper, we extend the spatial availability measure into a multimodal framework and explore its use in answering the question outlined: _"given opportunities are finite, how many are available to a given location depending on the mode used?"._ The answer to this question quantifies how many opportunities can be accessed, considering competition, for different modes.
-To foreground this exploration, in Section 2, we discuss short falls of a few existing location-based measures in comparison to spatial availability through a synthetic example.
-In Section 3, the spatial availability of an empirical example of the LEZ in the city of Madrid, Spain is calculated.
-We demonstrate how the restriction of car circulation could have impacted the spatial availability of opportunities for each sub-population using transit, cycling and walking modes.
+In this paper, we extend the spatial availability measure into a multimodal framework and explore its use in answering the question outlined: _"given opportunities are seen as finite, how many opportunities are available to a mode-using population given a spatial location?"._ The answer to this question quantifies how many opportunities can be accessed, considering competition, for different modes.
+
+To foreground our methodological contribution, in Section 2, we outline shortcomings of a few location-based accessibility measures in comparison to spatial availability using a synthetic example of a multiple mode using population.
+In Section 3, the spatial availability of four mode categories is calculated for the city of Madrid, Spain after the 2017 LEZ is implemented.
+Using mode-specific home-to-work origin-destination 2018 travel survey as empirical data, we demonstrate inequities of spatial availability within and outside the LEZ for each sub-population using car, transit, cycling and walking modes.
 In Section 4, we provide concluding remarks on the strengths of the use of spatial availability as a multimodal accessibility measure and potential future uses in policy planning scenarios.
@@
 Location-based accessibility indicators are quantitative measures of _potential_ interaction with opportunities for locations within a given region: they are a product of the relationship between land-use and transport systems.
-Arguably the most commonly used location-based measured are cumulative opportunity measures and weighted cumulative opportunity measures [@levinsonTransportAccessManual2020].
-These measures weight the opportunities that can be potentially interacted with from origin $i$ to destination $j$ based on some sort of travel cost function (e.g., travel time, fare, travel distance) otherwise known as a travel impedance function $f^{m}(c^m_{ij})$.
-Many weighted cumulative opportunities (often referred to as the gravity-based measure) originate from the measure proposed by @hansenHowAccessibilityShapes1959, which can take the following multimodal form: $S_i^m = \sum_j O_j f^m(c_{ij}^m)$ where $m$ is a set of modes which have mode-specific travel costs $(c_{ij}^m)$ and travel impedance functions $f^m(\cdot)$.
+Arguably, the most commonly used location-based measures is the cumulative opportunity measure and thee weighted cumulative opportunity measures [@levinsonTransportAccessManual2020].
+These measures weight the opportunities that can be potentially interacted with from origin ($i$) to destination ($j$) based on some sort of travel cost function (e.g., travel time, fare, travel distance), otherwise known as a travel impedance function $f^{m}(c^m_{ij})$.
+Many weighted cumulative opportunities (often referred to as the gravity-based measure) originate from the measure proposed by @hansenHowAccessibilityShapes1959, which can take the following multimodal form: $S_i^m = \sum_j O_j f^m(c_{ij}^m)$ where $m$ is a set of modes which have mode-specific travel costs ($(c_{ij}^m)$), travel impedance functions ($f^m(\cdot)$) for the mode-specific Hansen-type value ($S_i^m$).
 
 The Hansen-type measure does not consider competition between modes nor is it constrained.
-As an example, the work of @tahmasbiMultimodalAccessibilitybasedEquity2019 uses the Hansen-type measure to measure the potential interaction with retail locations using walking, public transit, and car modes $m$.
+As an example, the work of @tahmasbiMultimodalAccessibilitybasedEquity2019 uses the Hansen-type measure to measure the potential interaction with retail locations using walking, public transit, and car modes as $m$.
 $S_i^m$ is the sum of retail locations $j$ that can potentially be interacted with under the travel impedance as calculated for each $i$ and $m$.
-In other words, each $i$ has three $S_i$ values, one per $m$.
-In this work, they demonstrate that the car mode has the highest $S_i^{m=car}$ values in the majority of $i$, i.e., populations using a car can potentially interact with the most retail opportunities than populations using other modes.
-However, the higher $S_i^{m=car}$ values are not a result of lower $S_i^{m}$ values for other modes: it is not assumed that car-using populations potentially accessing more opportunities take away potential opportunities for other populations within the measure (no consideration for competition).
-This measure is also not constrained: there is no global maximum for $S_i$ or $S_i^m$ values, they are presented as a population normalized accessibility index.
-This makes the interpretation of the 'potentially interacted opportunities' relative to the region, making comparisons of the results across different regions challenging.
-
-However, opportunities in a region can be considered finite.
-There are only so many school-seats, hospital capacity, jobs, etc., in a region and if one person interacts with an opportunity at a given time, it is taken.
-As such, if one person is advantaged and has the ability to reach more opportunities through a lower travel-cost mode, than they have more opportunities to potentially interact with more opportunities than other people.
-From the other perspective, their are fewer opportunities left to be potentially interacted with for populations using higher travel-cost modes.
-In this way, populations using modes with a higher travel impedance are at a higher access disadvantage than populations using lower travel impedance modes.
-This recognition is the motivation behind integrating _competition_ for opportunities within multimodal accessibility measures.
-Arguably one of the most popular competitive location-based accessibility measures is the two-step floating catchment area (2SFCA) approach popularized by @luoMeasuresSpatialAccessibility2003 who simplified the approach proposed by @shenLocationCharacteristicsInnercity1998 (with similar considerations for competition in @weibullAxiomaticApproachMeasurement1976 and @josephMeasuringPotentialPhysical1982).
-
-The Shen-type accessibility measure's formulation is: $a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}$ where $D_j^m$ is the potential demand for opportunities equal to travel impedance weighted population $\sum_i P_i^m f^m(c_{ij}^m)$.
-In this way, the Shen-type measure can be understood as a ratio of the potential opportunity supply over the potential demand for opportunities.
-The measure considers competition, but it is _non-constrained_.
-A score of competitive potential accessibility associated is associated with each location $i$ for each mode $m$, but there are no global maximums.
-In other words, it is difficult to interpret the meaning of differences in Shen-type accessibility scores between modes.
-
-To illustrate, @taoInvestigatingImpactsPublic2020a calculates $a_i^m$ to jobs for different income-group populations in Shenzhen (China) using $m = \text{public transit}$ and $m={car}$.
+In other words, each $i$ has three $S_i^m$ values, one per $m$.
+In this work, they demonstrate that the car mode has the highest $S_i^{m}$ values in the majority of $i$, i.e., populations using a car can potentially interact with the most retail opportunities than populations using other modes.
+However, the higher car $S_i^{m}$ values are not a result of lower $S_i^{m}$ values for other modes: it is not assumed that car-using populations accessing more opportunities take away potential opportunities from other populations within the measure (i.e., no consideration for competition).
+This measure is also not constrained so there is no global maximum for $S_i$ or $S_i^m$ values.
+As such, $S_i^m$ values are presented as standardized accessibility index in this work, from the lowest value of 0 to the highest value of 1.
+This presentation or discussion of Hansen-type accessibility measures is common in the literature e.g., see the work of @maharjanSpatialEquityModal2022 and @campbell2019accessibility.
+However, this representation of results can make the interpretation of results challenging _because_ values are region-relative and lacking numeric meaning.
+Zones will always have values from between 0 to 1 represented, but is a zone with a low value for pedestrian-modes and a high value for car-modes notable?
+And if notable, what does the difference in these standardized values tangibly mean for planners?
+By how much should transport systems and land-use configurations be changed to improve conditions?
+And in what way can these existing methods be used to track progress over time?
+Or between regions?
+This is challenging to track, since certain values will always be relatively 'low' and 'high'.
+Alternatively, modal accessibility calculated using the Hansen-type measure is also presented without standardization, as in the work by @lunkeModalAccessibilityDisparities2022.
+Often, the meaning of the accessibility value itself is not interpreted, as it is difficult to state what having access to 30,000 employment opportunities by car but having 10,000 employment opportunity by transit tangibly signifies.
+
+However, once opportunities in a region are interpreted as finite, a new meaning is taken on.
+As considered in the long tradition of accessibility work, capacity of opportunities is limited and thus is subject to competition by population [@soukhovIntroducingSpatialAvailability2023; @grengsJobAccessibilityModal2010; @kawabataJobAccessibilityIndicator2006; @kwokUseModalAccessibility2004; @morrisAccessibilityIndicatorsTransport1979; @weibullAxiomaticApproachMeasurement1976; @shenLocationCharacteristicsInnercity1998].
+There are only so many school-seats, hospital capacity, employment opportunities, etc., in a region and if one person interacts with an opportunity at a given time, it is taken: the supply of an opportunity and the demand for that opportunity are two components of accessibility.
+The competition consideration is especially important in the case of multiple modes.
+For instance, people in a zone who are advantaged with a relatively low travel-cost mode have the ability to potentially interact with more opportunities than other people.
+Due to this advantage, through the perspective of finite opportunities, there are fewer opportunities left to be potentially interacted for everyone else, especially zones with populations using higher travel-cost modes.
+This recognition is the motivation behind integrating _competition_ for opportunities within modal accessibility measures.
+
+Arguably, one of the most popular competitive location-based accessibility measures is the two-step floating catchment area (2SFCA) approach popularized by @luoMeasuresSpatialAccessibility2003 who simplified the approach proposed by @shenLocationCharacteristicsInnercity1998 (with similar considerations for competition in @weibullAxiomaticApproachMeasurement1976 and @josephMeasuringPotentialPhysical1982).
+The Shen-type accessibility measure's formulation is: $a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}$ where $D_j^m$ is the potential demand for opportunities equal to travel impedance weighted population $\sum_i P_i^m f^m(c_{ij}^m)$ and the remaining variables are repeated in the Hansen-type measure.
+The Shen-type measure per mode ($a_i^m$) can be understood as a ratio of the travel impedance-weighted supply of opportunities for the $m$-using population in $i$ over the ravel impedance-weighted demand for opportunities.
+In this way, it considers competition but the measure remains _non-constrained_ meaning resulting values are not associated with global maximums.
+In other words, it is difficult to interpret the meaning of differences in Shen-type accessibility scores between modes in similar ways as reflected for Hansen-type measures.
+
+To illustrate, @taoInvestigatingImpactsPublic2020a calculates $a_i^m$ to jobs for different income-group populations in Shenzhen, China using $m = \text{public transit}$ and $m={car}$.
 They demonstrate that $i$ with low-income populations have lower $a_i^m$ than $i$ with higher-income populations.
@@
 And what is the interpretation of this difference?
-are left unanswered.
-
-Spatial availability improves on previous multimodal accessibility approaches as it considers _competition_ in the potential interaction with opportunities in a _constrained_ framework (e.g., finite opportunities).
+are left unaddressed.
+<!--QUESTION: DOES calling this 'modal accessibility measures" instead of "multi-modal" make more sense? -->
+Spatial availability improves on previously discussed modal accessibility approaches using the Hansen-type measure and the Shen-type measure by taking into account _competition_ in the potential interaction with opportunities in a _constrained_ framework (e.g., finite opportunities).
 This is done by considering: 1) competition between mass effect (e.g., the advantage of sub-populations residing in relatively low population-density and high opportunity-proximate areas) and 2) competition between travel impedance (e.g., sub-populations with relatively low travel-impedance) through a proportional allocation mechanism.
@@
 
-In brief, we define the _spatial availability_ at $i$ ($V_{i}$) as the proportion of all opportunities in the region $O$ that are allocated to location $i$ from all opportunity destinations $j$.
+In brief, we define the spatial availability at $i$ ($V_{i}$) as the proportion of all opportunities in the region $O$ that are allocated to location $i$ from all opportunity destinations $j$.
 $V_{i}$ is a value of how many opportunities are available to each location $i$ out of all the opportunities in the region.
@@
 As such, spatial availability is a  _competitive_ and _constrained_ accessibility measure as $F^t_{ij}$ handles the number of opportunities in the region in a finite way (proportional allocation).
-$F^t_{ij}$ consists of two components: a population-based balancing factor $F^p_{i} = \frac{P_i}{\sum_i P_i}$ and an impedance-based balancing factor $F^c_{ij} = \frac{F^c_{ij}}{\sum_j F^c_{ij}}$ that, respectively, allocate opportunities to $i$ in proportion to the size of the population at $i$ (the mass effect) and the cost of reaching opportunities at $j$ (the impedance effect).
+$F^t_{ij}$ consists of two components: a population-based balancing factor $F^p_{i} = \frac{P_i}{\sum_i P_i}$ and an impedance-based balancing factor $F^c_{ij} = \frac{F^c_{ij}}{\sum_j F^c_{ij}}$ that, respectively, allocates opportunities to $i$ in proportion to the size of the population at $i$ (the mass effect) and the cost of reaching opportunities at $j$ (the impedance effect).
 
@@
 In other words, assuming a finite number of opportunities in the region, $F^t_{ij}$ proportionally allocates $O_j$ to each $i$ such that the resulting $V_i$ value represents the number of opportunities _spatially available_ to the population at $i$.
-This value can be seen to represent spatial availability as it is a proportion of the opportunities in the region (i.e., $\sum_j O_j = \sum_i V_i$).
-
-The focus of this paper is to extend $V_i$ for multimodal applications.
+Each zonal value is a proportion of the opportunities in the region (i.e., $\sum_j O_j = \sum_i V_i$).
+
+The focus of this paper is to extend $V_i$ for the measurement of modal accessibility applications.
 To do so, the balancing factors are reformulated to yield a proportional value for the set of modes $m$ used by populations at each $i$.
-As these factors are proportional, $F^{pm}_{i}$ and $F^{cm}_{ij}$ can be summed up across each $m$ at each $i$ and across all $i$ to equal to 1.
+As these factors are proportional, $F^{pm}_{i}$ and $F^{cm}_{ij}$ can be summed across each $m$ at each $i$ and then across all $i$ to equal to 1.
 They are also similarly combined multiplicatively to obtain their joint effect, represented as the combined balancing factor $F^{tm}_{ij}$ detailed in Equation (\ref{eq:multimodal-balancing-factors}).
@@
 - The population balancing factor for each $m$ at each $i$ is $F^{pm}_{i} = \frac{P_{i}^m}{\sum_{m}\sum_{i} P_{i}^m}$
-- The cost of travel balancing factor for each $m$ at $i$ is $F_{ij}^{cm} = \frac{f(c_{ij}^m)}{\sum_{m} \sum_{i} f(c_{ij}^m)}$
+- The cost of travel balancing factor for each $m$ at $i$ is $F_{ij}^{cm} = \frac{f^m(c_{ij}^m)}{\sum_{m} \sum_{i} f^m(c_{ij}^m)}$
 
@@
 
-## Synthetic example
-
-Consider the following: Figure \ref{fig:Fig1} depicts a region with population and jobs at three population centers ($A$, $B$, $C$) and three employment centers ($1$, $2$, $3$).
+## An illustrative synthetic example
+<!--QUESTION: maybe we remove this synthetic example?-->
+Consider the following Figure \ref{fig:Fig1}, it depicts a region with population and jobs at three population centers ($A$, $B$, $C$) and three employment centers ($1$, $2$, $3$).
 The population at each population center is divided into two sub-populations, one using a faster mode $z$ and another using a slower mode $x$, to travel to employment centers.
@@
 From the perspective of access to a _finite_ amount of opportunities in the region ($210,000$ jobs), the sub-population that is most proximate to jobs, furthest from densely populated centers, and is using the lowest travel-cost mode $z$ can potentially access the most job opportunities.
-This appears to be the sub-population at $A$ using $z$.
-From the other perspective, sub-populations located in opposite conditions (i.e., further away from jobs, close to dense populations, and using $x$) are at a relative job opportunity access _disadvantage_.
-From the perspective of inequities, the competition for opportunities between different mode-using populations matters as it reflects how well the land-use and transport system serves (or doesn't serve) them.
+This appears to be the sub-population at $A$ using mode $z$.
+Sub-populations located in opposite conditions (i.e., further away from jobs, close to dense populations, and using $x$) are at a relative job opportunity access _disadvantage_.
+The competition for opportunities between different mode-using populations matters from the perspective of inequities as it reflects how well the land-use and transport system serves (or doesn't serve) certain populations.
 
@@
 The calculated $S_i^m$, $a_i^m$ and $V_i^m$ accessibility values for each $i$ and $m$ are shown in the middle three columns and are aggregated for each $i$ in the final two columns in Table 1 <!-- \ref{tab:tab1}-->.
-We use a negative exponential impedance function $f(c_{ij}) = \exp(-\beta\cdot c_{ij})$ with $\beta=0.1$ for both $x$ and $z$ modes for all accessibility measures calculations.
+We use a negative exponential impedance function $f^m(c_{ij}^m) = \exp(-\beta\cdot c_{ij})$ with $\beta=0.1$ for both $x$ and $z$ modes for all accessibility measures calculations.
 
@@
 Additionally, $S_i^m$ is equal for both mode-using populations in $A$ and $B$.
-This is the case because $S_i^m$ does not consider _competition_, it only relies on reflecting the count of opportunities that may be interacted with as a product of $f^m(c_{ij}^m)$.
+This is the case since $S_i^m$ does not consider _competition_, it only relies on reflecting the count of opportunities that may be interacted with as a product of $f^m(c_{ij}^m)$.
 Recall, populations in $A$ and $B$ have the same travel impedance to employment centers $1$, $2$ and $3$ (either 15, 30, or 100 minutes using $x$ or 10, 25, or 80 minutes using $z$).
@@
 
-<!-- THIS IS GOOD, but could be cut to save works-->
 Since spatial availability is constrained and has an interpretable meaning as a proportion of the total opportunities in the region, the values at $i$ have a new significance.
@@
 <!-- a summary about low emission zones, they are a geographic accessibility restriction. they change the accessibility landscape-->
-Low emission zones (LEZ) have been implemented as a climate change policy intervention to reduce GHG emissions, improve air quality, and support sustainable mobility in many countries.
-Though rules vary<!-- depending on legal aspects and cultural norms-->, LEZ aim to deter/reduce traffic in designated zones under threat of penalty (e.g., fines, seizure of vehicle).
-<!-- In practice, LEZ limits the volume of traffic by excluding vehicles by license plate, fuel type, or by introducing tolls.--> From the perspective of restriction for passenger transport, LEZ are a policy of _geographic discrimination_ as they change how people access opportunities by making the travel impedance more costly for car-mode users.
-If seeing opportunities as finite, this discrimination allows populations to access opportunities by other modes more readily than before.
+Low emission zones (LEZ) have been implemented as a climate change policy intervention to reduce GHG emissions, improve air quality, and support sustainable mobility in many countries [@margaryanLowEmissionZones2021; @verbeekJustManagementUrban2022].
+Though rules of exclusion vary<!-- depending on legal aspects and cultural norms-->, LEZ aim to deter/reduce traffic in designated zones under threat of penalty (e.g., fines, seizure of vehicle).
+<!-- In practice, LEZ limits the volume of traffic by excluding vehicles by license plate, fuel type, or by introducing tolls.--> From the perspective of restriction for passenger transport, LEZ are a policy of _geographic discrimination_ as they change how people access opportunities by making the travel impedance more costly for specific car-mode users.
+When considering opportunities as finite in a region, this discrimination could allow certain populations to access opportunities by other modes more readily than before.
 In this way, LEZ change the multimodal competitive accessibility landscape of a city.
@@
 In 2017, LEZs were implemented in the Spanish capital city of Madrid following the goals set out in the national agenda <!-- : to fight climate change, cut nitrogen dioxide levels for the benefit of people's health, and prioritize people's movement in the city-->.
-In geographic scope, the 2017 boundaries of the LEZ were relatively small (covering 4.72 km^{2}) and within the center (i.e., LEZ Centro).
+In geographic scope, the 2017 boundaries of the LEZ were relatively small (covering 4.72 km$^2$) and within the center (i.e., LEZ Centro).
 These boundaries were expanded in 2023 to inside of the M-30, a highway in proximity to the city center (i.e., LEZ M-30) and the city has plans to further spatially expand the LEZ.
@@
 
-```{r, jobs-plot, fig.cap="\\label{fig:Fig2} Jobs $O_j$ taken by people living and working in Madrid as reported by the 2018 travel survey.", fig.show='hold', fig.align='center'}
+```{r, jobs-plot, fig.cap="\\label{fig:Fig2} Jobs $O_j$ taken by people living and working in Madrid as reported by the home-to-work flows in the 2018 travel survey.
+Gray TAZ has no jobs.", fig.show='hold', fig.align='center'}
 knitr::include_graphics("images/i_jobs_zn208_plot.png") 
@@
 
-```{r, pop-plot, fig.cap="\\label{fig:Fig3} Population living and working in Madrid, by four summarized modal categories, $P^m_i$ as reported by the 2018 travel survey.", fig.show='hold', fig.align='center'}
+```{r, pop-plot, fig.cap="\\label{fig:Fig3} Population living and working in Madrid by four summarized modal categories $P^m_i$ as reported by the home-to-work flows in the 2018 travel survey.
+Gray TAZs have no population.", fig.show='hold', fig.align='center'}
 knitr::include_graphics("images/im_populations_zn208_plot.png")
 ```
-The 2018 Community of Madrid travel survey (@comunidaddemadridResultadosEDM20182020) is the source of data for this empirical example: it is a representative survey that reflects a snap-shot of the travel patterns for one typical day of the working week (e.g., n=222,744 trips with representative population elevation factors).
-In this paper, a sample of the travel survey is used, namely the residential home origin to work destination trips of all modes and those that originate and end in the city of Madrid.
-These totals are displayed in Figure \ref{fig:Fig2} and Figure \ref{fig:Fig3}.
-Both figures are displayed at the level of traffic analysis zones ($i$ and $j$) that correspond to the survey.
+The 2018 Community of Madrid travel survey [@comunidaddemadridResultadosEDM20182020] is the source of data for this empirical example: it is a representative survey that reflects a snap-shot of the travel patterns for a typical day of the working week in 2018.
+Specifically, it includes 222,744 trips taken from a representative sample of 85,064 households across the traffic analysis zones (TAZ) representing the Community of Madrid (population of 6,507,184 over 3 years old) through population elevation factors.
+
+In this empirical example, we only use a sub-set of the survey, specifically home-to-full-time-work trips, by all modes, to demonstrate multimodal spatial availability of employment opportunities.
+The total population who travels to full-time work from an origin (i.e., the worker), and the destinations (i.e., the employment opportunity), for the City of Madrid (a sub-set of the Community of Madrid) is visualized in Figure \ref{fig:Fig2} and Figure \ref{fig:Fig3}.
+Both figures are displayed at the level of TAZ (the $i$ and $j$ zones) that correspond to the survey.
 The red boundary represents the LEZ Centro in effect in 2017 and thus those travel patterns of car-restriction reflected in the survey.
 The cyan boundary represents the LEZ that will be within the boundaries of the M-30 highway in 2023 and is present in the plots as a spatial reference for areas in proximity to the LEZ Centro.
-
+<!--QUESTION: here's an idea, why not use the employment flows to create the modal impedance (like we do) but fit that to all direct home-to-destination O-D flows instead of just work O-D flows?
+All destinations are likely places people could _potentially_ work ??
+this is confusing to justify but an idea.
+-->
 The total sum of jobs $O_j$ that are held are shown in Figure \ref{fig:Fig2} and the populations that go to a work destination by four modal categories $P^m_i$, is reflected in Figure \ref{fig:Fig3}.
-The modal categories represented in Figure \ref{fig:Fig3} are summarized for the following trip mode types:
-
-   - Car/motor: all cars and operating modes (e.g., cab, private driver, company, rental care, main driver, passenger, etc.) and all public, private or company motorcycle/mopeds.
+The four modal categories represented in Figure \ref{fig:Fig3} are created through the grouping of similar mode types as provided in the survey, by the authors discretion.
+The modal categories and the mode types within each category are reported as follows:
+
+   - Car/motor: all cars and operating modes (e.g., cab, private driver, company, rental car, main driver of a private car, passenger in a private car) and all public, private or company motorcycle/mopeds.
   - Transit: all bus, trams, and trains,
   - Bike: all bicycle trips (e.g., private, public, or company bike trips) and "other" types of micromobility options,
-  - Walk: walking or by foot,
+  - Walk: walking or by foot.
 
@@
 The population that travels using transit is more spatially distributed than those using cars - particularly near and within LEZ Centro.
-This distribution could be a result of a variety of factors including: transit coverage and service within with city,  effective car infrastructure outside of the M-30, and/or the impact of the Central LEZ itself.
-
-From Figure \ref{fig:Fig3}, it can also be seen that biking and walking trips are less common than motorized trips at `r (pop_props$population_m[1]/sum(pop_props$population_m)) %>% percent()` and `r (pop_props$population_m[2]/sum(pop_props$population_m)) %>% percent()` respectively.
+This distribution is likely caused by a variety of factors including: transit coverage and service within with city,  effective car infrastructure outside of the M-30, and/or the impact of the LEZ Centro itself.
+From Figure \ref{fig:Fig3}, it can be observed that biking and walking trips are less common than motorized trips at `r (pop_props$population_m[1]/sum(pop_props$population_m)) %>% percent()` and `r (pop_props$population_m[2]/sum(pop_props$population_m)) %>% percent()` respectively.
 Noteably, there is a positive trend between the populations of walking and biking trips in zones and populations of transit trips.
@@
 These travel times, per modal category, are used to calibrate mode-specific travel impedance functions $f^m(c_{ij}^m)$.
-To illustrate the modal differences in travel lengths, summary descriptive per mode are detailed:
-
-- Car/motor: `r weighted_mean(x=tld_empirical_car$TIEMPO_VIAJES, w=tld_empirical_car$ELE_G_POND_ESC2) %>% round()` min (min:`r car_qs[1]%>% round()` min., Q2: `r car_qs[2]%>% round()` min., Q3: `r car_qs[4]%>% round()` min., max: `r car_qs[5]%>% round()` min.)
-- Transit: `r weighted_mean(x=tld_empirical_transit$TIEMPO_VIAJES, w=tld_empirical_transit$ELE_G_POND_ESC2)%>% round()` min.
-(min:`r transit_qs[1]%>% round()` min., Q2: `r transit_qs[2]%>% round()` min., Q3: `r transit_qs[4]%>% round()` min., max: `r transit_qs[5]%>% round()` min.)
-- Bike: `r weighted_mean(x=tld_empirical_bike$TIEMPO_VIAJES, w=tld_empirical_bike$ELE_G_POND_ESC2)%>% round()` min.
-(min:`r bike_qs[1]%>% round()` min., Q2: `r bike_qs[2]%>% round()` min., Q3: `r bike_qs[4]%>% round()` min., max: `r bike_qs[5]%>% round()` min.)
-- Walk: `r weighted_mean(x=tld_empirical_walk$TIEMPO_VIAJES, w=tld_empirical_walk$ELE_G_POND_ESC2)%>% round()` min.
-(min:`r walk_qs[1]%>% round()` min., Q2: `r walk_qs[2]%>% round()` min., Q3: `r walk_qs[4]%>% round()` min., max: `r walk_qs[5]%>% round()` min.)
-
-To calculate $f^m(c_{ij}^m)$ from the survey travel times, a concept known as the trip length distribution (TLD) was used.
+To illustrate the modal differences in travel time lengths, summary descriptive per mode are detailed:
+
+- Car/motor: mean `r weighted_mean(x=tld_empirical_car$TIEMPO_VIAJES, w=tld_empirical_car$ELE_G_POND_ESC2) %>% round()` minutes (min:`r car_qs[1]%>% round()` minutes, Q2: `r car_qs[2]%>% round()` minutes, Q3: `r car_qs[4]%>% round()` minutes, max: `r car_qs[5]%>% round()` minutes)
+- Transit: mean `r weighted_mean(x=tld_empirical_transit$TIEMPO_VIAJES, w=tld_empirical_transit$ELE_G_POND_ESC2)%>% round()` minutes (min:`r transit_qs[1]%>% round()` minutes, Q2: `r transit_qs[2]%>% round()` minutes, Q3: `r transit_qs[4]%>% round()` minutes, max: `r transit_qs[5]%>% round()` minutes)
+- Bike: mean `r weighted_mean(x=tld_empirical_bike$TIEMPO_VIAJES, w=tld_empirical_bike$ELE_G_POND_ESC2)%>% round()` minutes (min:`r bike_qs[1]%>% round()` minutes, Q2: `r bike_qs[2]%>% round()` minutes, Q3: `r bike_qs[4]%>% round()` minutes, max: `r bike_qs[5]%>% round()` minutes)
+- Walk: mean `r weighted_mean(x=tld_empirical_walk$TIEMPO_VIAJES, w=tld_empirical_walk$ELE_G_POND_ESC2)%>% round()` minutes (min:`r walk_qs[1]%>% round()` minutes, Q2: `r walk_qs[2]%>% round()` minutes, Q3: `r walk_qs[4]%>% round()` minutes, max: `r walk_qs[5]%>% round()` minutes)
+
+To calculate $f^m(c_{ij}^m)$ from the survey travel times, a concept known as the trip length distribution (TLD) is used.
 A TLD represents the proportion of trips that are taken at a specific travel cost such as travel time (i.e., probability density distribution of trips taken by travel cost).
-This distribution is then used to derive impedance functions (e.g., done in the accessibility works of @lopez_2017_spatial, @horbachov_theoretical_2018, and @batista_estimation_2019).
-Maximum likelihood estimation and the Nelder-Mead method for direct optimization available within the R {fitdistrplus} package [@fitdistrplus_2015] is used to fit the impedance functions.
-As shown as shown in Figure \ref{fig:Fig4}, based on goodness-of-fit criteria and associated diagnostics, the gamma and log-normal probability density function (line curves) are selected as best fitting curves for the motorized and non-motorized modes respectively.
+This distribution is then used in the derivation of travel impedance functions (e.g., done in the accessibility works of @lopez_2017_spatial, @horbachov_theoretical_2018, and @batista_estimation_2019).
+To fit the impedance functions, we use the Maximum likelihood estimation and the Nelder-Mead method for direct optimization available within the R {fitdistrplus} package [@fitdistrplus_2015].
+Based on goodness-of-fit criteria and associated diagnostics, the gamma and log-normal probability density functions are selected as best fitting curves for the motorized and non-motorized modes respectively.
 The selection of functional form aligns with empirical examples in other regions [@reggianiAccessibilityImpedanceForms2011; @soukhovTTS2016RDataSet2023; @soukhovIntroducingSpatialAvailability2023].
-<!--e.g., car gamma XX and active transport log-normal XX.--> Overall, the plots in Figure \ref{fig:Fig4} display the probability of travel given a trip travel time, based on trip flows from the survey.
-These 'probability of travel' at each travel time for each mode are realized observations that reflect the land-use, the transport system, and the population travel behaviour in Madrid.
-
-<!-- TOO MUCH DETAIL, removed. 
-They are used in this study (as similar to other work that uses TLD [e.g., @lopez_2017_spatial, @horbachov_theoretical_2018, @batista_estimation_2019]) to reflect the travel impedance of each mode to work destinations in Madrid.
-Overall, as seen from the four plots, trips under 5 minutes do not occur frequently for any mode and as such are assigned a lower travel impedance $f^m(c_{ij}^m)$ value.
-One reason that these under 5 minute trips do not frequently occur as a result of land-use (residential and job mismatch): not everyone lives in immediate proximity to their workplace.
-Observing the non-motorized modes, shorter trips occurred more frequently overall for walking populations, particularly around 15 minute lengths, so a trip of approximately 15 minutes is assigned the highest $f^m(c_{ij}^m)$.
+The shape and rate parameters for the gamma functions (motorized modes) are `r HW_car_motor_gamma$estimate[1]` and `r HW_car_motor_gamma$estimate[2]`for car/motor and `r HW_transit_gamma$estimate[1]` and `r HW_transit_gamma$estimate[2]` for transit; for the log-normal functions (non-motorized modes), mean and standard deviation parameters are `r HW_bike_lnorm$estimate[1]` and `r HW_bike_lnorm$estimate[2]` for bike and `r HW_walk_lnorm$estimate[1]` and `r HW_bike_lnorm$estimate[2]` for walk.
+
+Represented in Figure \ref{fig:Fig4} are four plots visualizing the estimated probability density functions (represented as black lines) over top the binned travel time data from the 2018 survey.
+The estimated probability density functions can be interpreted as the probability of travel (y-axis) given a trip travel time (x-axis), based on trip flows from the 2018 travel survey.
+These 'probability of travel' at each travel time for each mode are realized observations that reflect the land-use, the transport system, and the population travel behaviour in Madrid (according to the 2018 survey).
+A trip that leaves $i$ is assigned a probability of travel value (y-axis) based on the mode and travel time of that trip to work: this probability value is the input $f^m(c_{ij}^m)$.
+
+Notably, plots in Figure \ref{fig:Fig4} can be observed across modes if curious to know what trips are more likely to occur using what mode depending on trip length (as informed by the observed 2018 survey).
+Trips >5 minutes do not occur frequently for any mode, as such trips with short trip lengths are assigned a lower travel impedance $f^m(c_{ij}^m)$ value.
+One reason that >5 minute trips do not frequently occur as a result of land-use (residential and job mismatch): not everyone lives in immediate proximity to their workplace.
+In terms of the non-motorized modes, shorter trips occurred more frequently overall for walking populations, particularly around 15 minute lengths, so a trip of approximately 15 minutes is assigned the highest $f^m(c_{ij}^m)$.
 For biking populations, longer travel times are more common so though the highest $f^m(c_{ij}^m)$ value also corresponds to approximately 15 minutes, the curve is more spread out and values decrease less rapidly at longer travel times than for the walk mode curve.
-A similar trend occurs for the motorized modal options where transit mode is more spread out than car/motor mode.
--->
+A similar trend can be observed for the motorized modal options where transit mode is more spread out than car/motor mode.
+All in all, these observations demonstrate that, based on a given mode and travel time, a trip is more or less likely to occur and is accordingly represented in the cost of travel balancing factor $F_{ij}^m$ for  $V_i^m$.
 
@@
 ## Results
+As it is worth re-iterating, the empirical example reflects a snap-shot of the calculated multimodal spatial availability using data from the 2018 travel survey.
+It is intended to visualize the spatial trends in availability of employment opportunities, by mode, and demonstrate how spatial availability can be interpreted to discuss the competitive advantage of lower travel impedance modes within Madrid Centro.
+
 The spatial availability of jobs $V_i^m$ is calculated for each of the four modal categories $m$ at the level of traffic analysis zones $i$ in Madrid and demonstrated in Figure \ref{fig:Fig5}.
@@
 ```{r, SA-m-plot, fig.cap="\\label{fig:Fig5} Spatial availability of job opportunities per origin and mode $V_i^m$ in Madrid.
-Calculated using the home-to-work origin-destination flows from the 2018 travel survey.
-", fig.show='hold', fig.align='center', fig.height=10}
+Calculated using the home-to-work flows from the 2018 travel survey.
+Gray TAZs have no population.", fig.show='hold', fig.align='center', fig.height=10}
 knitr::include_graphics("images/SA_im_V_zn208_plot.png")
@@
 
-$V_i^m$ is a proportion of the total number of the `r SA_i_zn208$V_i %>% sum()  %>% prettyNum(big.mark = ",")` jobs in the region and is visualized in Figure \ref{fig:Fig5}.
+$V_i^m$ is a proportion of the total number of `r SA_i_zn208$V_i %>% sum()  %>% prettyNum(big.mark = ",")` jobs in the region and is visualized in Figure \ref{fig:Fig5}.
 Since $V_i^m$ is calculated based on the likelihood of travel from observed home-to-work journeys, the values can be understood as the number of full-time jobs that are spatially available to the full-time working population at that $i$ and their associated $m$, relative to all the jobs in the city.
@@
 ```{r, SA-per-capita-m-plot, fig.cap="\\label{fig:Fig7} Spatial availability of job opportunities per mode-using capita by mode $v_i^m$ per origin in Madrid.
-Calculated using the home-to-work origin destination flows from the 2018 travel survey.", fig.show='hold', fig.align='center', fig.height=10}
+Calculated using the home-to-work flows from the 2018 travel survey.
+Gray TAZs have no population.", fig.show='hold', fig.align='center', fig.height=10}
 knitr::include_graphics("images/SA_im_vv_zn208_plot.png")
@@
 # Discussion and conclusions
+<!--QUESTION: maybe break up this section and consolidate all limitations/assumptions/potential uses of multimodal spatial availability in the conclusion?-->
 Location-based accessibility measures like the Hansen-type $S_i^m$, Shen-type $a_i^m$, and spatial availability $V_i^m$ measures share a commonality; they are a weighted sum of opportunities assigned to each spatial unit $i$ in a region.
@@
 
+<!-- note: maybe add in a paragraph about how the spatial availabilit ymeasure can take into consideration traveler related characteristics, like mode choice, within the measure.
+This is a current concern with accessibility measures Karner et al.
+2023 ( Emerging perspectives on transportation justice.
+Transportation Research
+1
+Part D: Transport and Environment) -->
+
 The flexibility and need for an accessibility measure such as spatial availability is pertinent in policy scenario evaluation.
@@
 These interpretations were difficult to be made with accessibility measures in the past.
+In future work, we intend to use multimodal spatial availability to characterise the equity of modal accessibility losses and gains in varying LEZ policy scenario implementations.
 

### 2023-11-06 Journal-manuscript/Editorial correspondance/Response-to-reviewers_R1.Rmd
First go at response to reviewers by Antonio after Anastasia

---
title: "Response-to-reviewers_R1"
output: pdf_document
date: "2023-10-18"
---

Thank you for for the time and care taken by the editorial team and reviewers in providing feedback to this manuscript. No doubt, the revised version submitted is more articulate than the first. Our responses to all comments from each reviewer are shown below in \textcolor{blue}{blue} and directly quoted the revised portion of the manuscript as relevant}

## Reviewer #1:

The article extends the concept of spatial accessibility and applies it to a fairness analysis of different modes of transportation. The article uses travel data from a week in Madrid as an example to analyze spatial accessibility in the city. However, the article has the following issues:

\textcolor{blue}{We appreciate your effort reading our submission and respond in detail to your comments below.}

1. Lack of Innovation: The article's improvement on spatial accessibility mainly involves incorporating the ratio of travel frequencies as weight in travel costs. This improvement is relatively minor and the method used does not demonstrate superiority over traditional accessibility approaches.

\textcolor{blue}{Thank you for this comment. First we would like to bring to your attention the editorial philosophy of PLoS ONE, according to which editors "make decisions on submissions based on scientific rigor, regardless of novelty," (see https://journals.plos.org/plosone/s/editorial-and-peer-review-process). Nonetheless we wish to respond to this comment.}

\textcolor{blue}{In this paper we extend our earlier work on [spatial availability](https://doi.org/10.1371/journal.pone.0278468) for the simultaneous analysis of multiple modes of transportation. You appear to be under the misaprehension that the main difference with accessibility measures is to incorporate "the ratio of travel frequencies as weight in travel costs". This is statement is not correct. The main difference is a proportional allocation mechanism that we introduced for spatial availability, and that is extended in this paper to allocate opportunities based on the proportion of travelers by different modes. This is not the same as "weighting the travel cost". In fact, each mode is modelled using its own impedance function, as shown in our empirical example, where we use origin-destination data by mode to estimate travel impedance functions specific to each mode.}

\textcolor{blue}{While this enhancement may appear minor to you, it is not plainly obvious how spatial availability applies to multiple modes, which is why we think this paper is needed. We do not dispute that this is an incremental step in the development of a more general method, but it is a step that considerably expands the range of potential applications. Further, we contend that the method has been rigorously developed and demonstrated using an open, transparent, and reproducible example.}

\textcolor{blue}{In response to this comment we have edited the "Discussion and conclusions" section to more clearly describe the advantages that multimodal spatial availability offers when considering the accessibility to opportunities by different modes. In particular, proportional allocation of opportunities using mode-specific impedance functions means that the travel-cost-advantage of each mode can be analyzed. Further, this mechanism ensures that the sum of all spatial availability values for all modes sum up to the total number of opportunities in the region, which is not true for any other type of accessibility measure.}

\textcolor{blue}{With respect to the "superiority" of the method, a reader of our earlier paper would already be aware of what limitations of accessibility analysis spatial availability aims to address. In this paper we also try to demonstrate throughout the manuscript what our measure does that others can't. For example, in the Discussion and Conclusions we state that: "With spatial availability, the number of opportunities that are captured (of the total opportunities in the region) by each mode can be individually calculated. From there, the difference between how many spatially available opportunities one mode captures versus another can be investigated. This is the advantage of the spatial availability measure, particularly its multimodal extension.". Since the spatial availability values result from proportional allocation, each value is a proportion of the total opportunities. Put another way, we can compare the proportion of opportunities available to users of each mode, to users at each zone, to users of each mode by zone, and so on, and the values relate directly to the total opportunities in the region. This also allows us to calculate values per capita that serve as benchmark values, as shown in the empirical example.}

\textcolor{blue}{In the revised version of the paper, the advantages of spatial availability are stated more explicitly, including in the "Introduction", "A review of multimodal accessibility measures", and "Discussion and Conclusions".}

\textcolor{blue}{Introduction:}

\textcolor{blue}{The question of how much one mode-using population can access at the expense of another mode-using population is a pertinent equity question in the evaluation of policy scenarios, especially those that impact modes differently. For instance, consider the impact of a low emission zone (LEZ). A LEZ is a policy of spatial and modal discrimination: the circulation of vehicles that are excessively polluting are restricted in specific regions. Depending on the level of vehicle restriction, LEZ have been shown to significantly improve air quality [@holmanReviewEfficacyLow2015] and air-quality related health outcomes[@margaryanLowEmissionZones2021]. From the perspective of accessibility to opportunities, LEZ explicitly lower the access that restricted-vehicle using populations have to opportunities within the LEZ. However, if conceptualizing opportunities in a region as finite, one populations' lose is another populations' gain. The populations using non-restricted modes (e.g., walking, cycling, transit, and less-polluting vehicles) to travel into LEZ will have more access than before the LEZ implementation.}

\textcolor{blue}{This evaluation is especially urgent as LEZ are in effect in cities globally; their reception [@tarrinoortizPublicAcceptabilityLow2021] and impact on certain activity types such as shopping [@tarrinoortizImpactLowEmission2023] have been mixed. They may also be having negative impacts on disadvantaged populations who have become mobility-restricted [@devrijNooneVisitsMe2022; @verbeekJustManagementUrban2022; @liottaWhatDrivesInequalities2023]. Measures that evaluate the accessibility of modes given both constrained and competitive considerations are lacking in the literature, but are needed, to evaluate such an intervention's impact on accessibility gains and losses within a finite conceptualization.}

\textcolor{blue}{A review of multimodal accessibility measures:}

\textcolor{blue}{- extended the explanation of Hasen-type measure and the consideration of competition in accessibility (shen-type measure) with additional examples from the literature. See the revised sections.}

\textcolor{blue}{Discussion and Conclusions:}

\textcolor{blue}{XXXXXXXXXXX}

2. Insufficient Experimental Data: The article uses questionnaire data from a week in Madrid, but lacks basic descriptions of the questionnaire data. Additionally, the daily travel data of approximately 30,000 trips is significantly limited, and there is no description of the criteria for selecting the dates. The typicality of the experimental data is questionable.

\textcolor{blue}{Thank you for this comment. To clarify, the data are not experimental. They are observational, since they were collected using a travel survey conducted by the City of Madrid. Travel surveys are a standard instrument in transportation planning and research, and are conducted in cities around the world. The data we work with represent the most recent and most complete travel survey conducted for the region to date. We are not completely sure where you got the figure of "30,000 trips" (which you consider significantly limited); presumably you are citing the maximum 'opportunity' or 'population' numbers in Figure 2 or 3. The most intensely coloured TAZ have ~30,000 jobs while the least intensely coloured have ~1,000 or fewer (figure 2). In fact, there are 847,574 jobs in the city, which is also the sum of the total spatial availability in our analysis, as well as the number of potential trips to work.}

\textcolor{blue}{The "typicallity" of a travel survey is information about travel on a typical day, usually during a period of maximum demand (i.e., not during the summer vacation). This is standard practice for these surveys.}

\textcolor{blue}{We have updated the manuscript to provide further information about the survey for the benefit of readers:}

\textcolor{blue}{Introduction:}

\textcolor{blue}{In Section 3, the spatial availability of four mode categories is calculated for the city of Madrid, Spain after the 2017 LEZ is implemented. Using mode-specific home-to-work origin-destination 2018 travel survey as empirical data, we demonstrate inequities of spatial availability within and outside the LEZ for each sub-population using car, transit, cycling and walking modes.}

\textcolor{blue}{Multimodal data and methods:}

\textcolor{blue}{The 2018 Community of Madrid travel survey [@comunidaddemadridResultadosEDM20182020] is the source of data for this empirical example: it is a representative survey that reflects a snap-shot of the travel patterns for a typical day of the working week in 2018. Specifically, it includes 222,744 trips taken from a representative sample of 85,064 households across the traffic analysis zones (TAZ) representing the Community of Madrid (population of 6,507,184 over 3 years old) through population elevation factors.}

\textcolor{blue}{In this empirical example, we only use a sub-set of the survey, specifically home-to-work trips, by all modes, to demonstrate multimodal spatial availability of employment opportunities. The total population who travels to work from an origin (i.e., the worker), and the destinations (i.e., the employment opportunity), for the City of Madrid (a sub-set of the Community of Madrid) is visualized in Figure \ref{fig:Fig2} and Figure \ref{fig:Fig3}. Both figures are displayed at the level of TAZ (the $i$ and $j$ zones) that correspond to the survey....}

\textcolor{blue}{Captions to Figure 2 and Figure 3:}

\textcolor{blue}{- Jobs $O_j$ taken by people living and working in Madrid as reported by the home-to-work flows in the 2018 travel survey.}
\textcolor{blue}{- Figure 3: Population living and working in Madrid by four summarized modal categories $P^m_i$ as reported by the home-to-work flows in the 2018 travel survey.
}

3. Lack of Data-Driven Analysis: In the analysis section, a large amount of data is used to analyze people's travel behavior in different areas. The analysis of spatial accessibility only considers the differences in the modes of travel mentioned earlier. The analysis results are somewhat one-sided, and examining the indicator from multiple perspectives would provide a more objective view. It is recommended to expand the dimensions of the analysis.

\textcolor{blue}{Thanks for this comment.}

\textcolor{blue}{It is somewhat puzzling that "30,000 trips" would be significantly limited (as per your comment \#2), and at the same time be "a large amount of data". We find that this comment in particular is not actionable due to its vagueness. There are no meaningful responses to "somewhat one-sided" when "somewhat" and "one-sided" do not quantify or refer to a particular side. We used all modes available in the survey, and examined the results from the perspective of each mode. What other perspectives do you suggest? What dimensions should be explored?}

\textcolor{blue}{Nevertheless, with the additional detail added to the manuscript to address your and other reviewers' comments, we believe that overall clarity has been improved.}

4. Disorganized Format: The article's methodology section contains numerous formulas and variables, but the definitions of these variables are unclear and difficult to read. There are also numerous formatting errors, and Table 1 is disorganized and unappealing. It is recommended to revise these issues.

\textcolor{blue}{Reviewer #2 identified some specific formatting issues and we fixed them. Otherwise we proofread the paper and hopefully did not leave a typo behind. We also made sure Table 1 starts on a new page, when it runs across pages the formatting resulted in some disorganization.}

\textcolor{blue}{Thank you again for your efforts trying to review our submission.}

## Reviewer #2:

This paper has sound mathematical foundations and allows to answer its research question in an elegant way (how to measure competition for e.g. jobs based on spatial accessibility?). I also really appreciated the fact that the paper was very didactic. Still, I have concerns about the relevance of the paper for future research.

\textcolor{blue}{Thank you for your thoughtful review of our paper. Your comments were helpful to improve the clarity of the research.}

More specifically, the new measure proposed by the authors has clear limitations:
i) it focuses on the competition for jobs and is not useful for studying access to non-competitive or semi-competitive resources such as amenities.

\textcolor{blue}{Thank you for this comment. We would begin by noting that there are many types of opportunities that are clearly mutually exclusive due to competition. For this paper we focused on jobs, but there are many others, such as beds at hospitals, doctors at clinics, seats at schools, and so on. Thus, even if the measure was applicable only to opportunities subject to competition, there are numerous applications to it.}

\textcolor{blue}{That said, we have grappled for some time with the idea that some opportunities might not be subject to competition. Our current thinking, after much consideration, is that in practice every type of opportunity is subject to congestion or capacity constraint. For instance, green spaces are often considered non-competitive, however, standards for the provision of such amenties are provided in the form of units of amenity per capita. For example, Natural England recommends an Accessible Natural Greenspace Standard such that the minimum supply of space is one ha of statutory Local Nature Reserves per thousand population}^[see  https://redfrogforum.org/wp-content/uploads/2019/11/67-Nature-Nearby%E2%80%99-Accessible-Natural-Greenspace-Guidance.pdf] \textcolor{blue}{. Similarly, the World Health Organization (cited in OECD, 2013) recommends that cities provide a minimum of 9 m2 of green area per inhabitant.}^[see https://doi.org/10.1787/9789264191808-en] \textcolor{blue}{For our purposes, standards of this type translate into "how much of this resource can we give to each individual that is not allocated to anyone else?". For this reason, although Local Nature Reserves have large capacities - they still have a capacity - and thus spatial availability can be used to measure constrained accessibility to this sort of opportunity. As standards are emphasized in the planning literature, in particular for fairness in transportation (see Martens and Golub, 2021), spatial availability analysis can be used to assess standards. We believe that as other researchers discover this new measure, other applications will be found.}

ii) it doesn't allow to study absolute gains or losses in accessibility from public transportation infrastructure improvements or changes.

\textcolor{blue}{Great point. Spatial availability can certainly be used to capture absolute gains and losses. In fact, logically, the gains and losses produced by using a competitive and constrained measure allows for a clear interpretation. This would require the analyst to estimate the accessibility before and after some change to the land use or transportation system. In this paper our empirical application is a single scenario to serve as proof of concept, but in future research we intend to use spatial availability to analyze changes in the size of Madrid's Low Emissions Zone implementation from a modal and socio-economic equity perspective.}

iii) the authors do not allow for modal shifts: they assume that the transport mode choice of households is fixed and cannot evolve due to e.g. transport infrastructure changes.

\textcolor{blue}{This is an excellent point. Although spatial availability is not a model of modal split, it is certainly amenable to changes in the use of various modes. This could be implemented as follows: the results of a modal split model are used to estimate new modal shares, which in turn are used to recalculate spatial availability. The results then can be compared to the baseline scenario. In other words, the framework for spatial availability is sufficiently flexible to take in not only mode-specific travel impedance functions, but also the proportions of the populations using each mode.}

These points limit the relevance of the new accessibility measure. The authors should, at least, specify these limitations early in the paper. The introduction should start by stating the precise research question the new accessibility measure is seeking to answer as well as its limitations). They should also justify, based on the literature, that competition for jobs is a key determinant of job market outcomes, and that there is strong inertia in mode choices.

\textcolor{blue}{Your comments have been very helpful to improve the clarity of the paper, as well as the scope of what we do now, and directions for future research.}

Finally, the writing of the paper, and particularly of the abstract and introduction, should be improved. The abstract could state the broader relevance of the topic and summarize the results of the case study on LEZs. The introduction should start more directly by introducing the research question and its relevance and describing the new measure and its limitations.

\textcolor{blue}{We have done this. Thank you again for your thoughtful comments and suggestions to improve the paper.}

Minor comments:

i) what are the summary statistics p10 (car: 36 min, transit: 55 min,...)? I assume they correspond to the mean.

\textcolor{blue}{Updated! Apologies, car: '36 min' corresponds to a mean of 36 minutes and then within the brackets additional descriptive statistics (minumum value, maximum value, etc.). }

ii) I also have identified a few formatting issues (e.g. Fcij p10 and 4.72km2 p 8).

\textcolor{blue}{Fixed! Thank you.}

## Reviewer #3:

This manuscript extended the authors’ previous work spatial availability measure, which is a type of location-based accessibility measure that is both constrained and competitive compared to Hansen-type measure and Shen-type accessibility measure, into a multimodal framework. The new measure, multimodal spatial availability, strengthened the constrained (or finite) nature of opportunities, and the competitive nature among multimodal accessibility resulting from this constraint through a synthetic example and an empirical example of the LEZ in the city of Madrid. In conclusion, the authors demonstrated one restriction had impacted the spatial availability of opportunities for other modes using and proposed potential future uses in policy planning scenarios.

In general, the manuscript was logical and well-structured. The research problem was well defined. The data were available and quite supported the conclusion. The statistical analysis performed appropriately.

However, there are some issues:

Major issues:

Please demonstrate whether “car/motor & transit” and “bike & walk” are comparable or whether they are in an actual competitive relationship? For example, if I work 3km from where I live, maybe I will never choose to take a transit, I will always walk or ride. But if I work 20km from where I live, walking or riding to work seems impossible for me, I have to drive or take public transportation. Car/motor and transit can be in competitive relationship and people can choose which one they prefer, but not choose between motor and walk. This issue will also have an impact on the results of the research.

\textcolor{blue}{Great comment. The short answer is that whenever a destination can be reached by more than one mode, users of those modes are in competition for the opportunities there.}

\textcolor{blue}{In the paper we have tried to improve the discussion to make this point more clear. The impedance functions for all four modes are not the same. They describe the travel behaviour of commuters as informed by the 2018 travel survey. To follow your example, someone who cannot walk to work because their job is 20 km from where they live, will not compete for that job against people from their same origin who do walk. However, it the place where their work is can be reached by walking by anyone from any other origin (someone who lives closer to that one destination), they would compete for the same opportunity.}

\textcolor{blue}{Futhermore, average travel times for car/motor and transit are longer than bike and walk. All people don't have access to all options - completely true. But the travel impedances reflect this real travel at an aggregate based on all the trips for a mode. And on average, it is assumed that people at each origin that take a mode to a destination are in direct competition for opportunities (as opportunities are finite) -- and a part of the competition is defined by the mode-specific impedance function (the second part is the population balancing factor). This assumption, that all populations, no matter their mode, are competing for the same finite set of opportunity, is part of spatial availability. We've made this more clear in the text.}

\textcolor{blue}{Multimodal data and methods:}

\textcolor{blue}{Notably, plots in Figure \ref{fig:Fig4} can be observed across modes if curious to know what trips are more likely to occur using what mode depending on trip length (as informed by the observed 2018 survey). Trips >5 minutes do not occur frequently for any mode, as such trips with short trip lengths are assigned a lower travel impedance $f^m(c_{ij}^m)$ value. One reason that >5 minute trips do not frequently occur as a result of land-use (residential and job mismatch): not everyone lives in immediate proximity to their workplace. In terms of the non-motorized modes, shorter trips occurred more frequently overall for walking populations, particularly around 15 minute lengths, so a trip of approximately 15 minutes is assigned the highest $f^m(c_{ij}^m)$. For biking populations, longer travel times are more common so though the highest $f^m(c_{ij}^m)$ value also corresponds to approximately 15 minutes, the curve is more spread out and values decrease less rapidly at longer travel times than for the walk mode curve. A similar trend can be observed for the motorized modal options where transit mode is more spread out than car/motor mode. All in all, these observations demonstrate that, based on a given mode and travel time, a trip is more or less likely to occur and is accordingly represented in the cost of travel balancing factor $F_{ij}^m$ for $V_i^m$.}

Minor issues:

1. As for Fig 2 and Fig 3, please indicate the meaning of the gray color blocks in illustration.

\textcolor{blue}{Fixed for all Figures, thank you.}

2. Please change the color scheme of fig 2. The red color scheme makes the LEZ centro area boundary, which is also in red, not visible.

\textcolor{blue}{WILL FIX THANK YOU}

 <!-- and I must change the labels on the legends to be ranges -->

### 2023-11-06 Journal-manuscript/Journal-manuscript.Rmd
First go at response to reviewers by Antonio after Anastasia

@@
 Location-based accessibility indicators are quantitative measures of _potential_ interaction with opportunities for locations within a given region: they are a product of the relationship between land-use and transport systems.
-Arguably the most commonly used location-based measured are cumulative opportunity measures and weighted cumulative opportunity measures [@levinsonTransportAccessManual2020].
-These measures weight the opportunities that can be potentially interacted with from origin $i$ to destination $j$ based on some sort of travel cost function (e.g., travel time, fare, travel distance) otherwise known as a travel impedance function $f^{m}(c^m_{ij})$.
-Many weighted cumulative opportunities (often referred to as the gravity-based measure) originate from the measure proposed by @hansenHowAccessibilityShapes1959, which can take the following multimodal form: $S_i^m = \sum_j O_j f^m(c_{ij}^m)$ where $m$ is a set of modes which have mode-specific travel costs $(c_{ij}^m)$ and travel impedance functions $f^m(\cdot)$.
+Arguably, the most commonly used location-based measures is the cumulative opportunity measure and thee weighted cumulative opportunity measures [@levinsonTransportAccessManual2020].
+These measures weight the opportunities that can be potentially interacted with from origin ($i$) to destination ($j$) based on some sort of travel cost function (e.g., travel time, fare, travel distance), otherwise known as a travel impedance function $f^{m}(c^m_{ij})$.
+Many weighted cumulative opportunities (often referred to as the gravity-based measure) originate from the measure proposed by @hansenHowAccessibilityShapes1959, which can take the following multimodal form: $S_i^m = \sum_j O_j f^m(c_{ij}^m)$ where $m$ is a set of modes which have mode-specific travel costs ($(c_{ij}^m)$), travel impedance functions ($f^m(\cdot)$) for the mode-specific Hansen-type value ($S_i^m$).
 
 The Hansen-type measure does not consider competition between modes nor is it constrained.
-As an example, the work of @tahmasbiMultimodalAccessibilitybasedEquity2019 uses the Hansen-type measure to measure the potential interaction with retail locations using walking, public transit, and car modes $m$.
+As an example, the work of @tahmasbiMultimodalAccessibilitybasedEquity2019 uses the Hansen-type measure to measure the potential interaction with retail locations using walking, public transit, and car modes as $m$.
 $S_i^m$ is the sum of retail locations $j$ that can potentially be interacted with under the travel impedance as calculated for each $i$ and $m$.
-In other words, each $i$ has three $S_i$ values, one per $m$.
-In this work, they demonstrate that the car mode has the highest $S_i^{m=car}$ values in the majority of $i$, i.e., populations using a car can potentially interact with the most retail opportunities than populations using other modes.
-However, the higher $S_i^{m=car}$ values are not a result of lower $S_i^{m}$ values for other modes: it is not assumed that car-using populations potentially accessing more opportunities take away potential opportunities for other populations within the measure (no consideration for competition).
-This measure is also not constrained: there is no global maximum for $S_i$ or $S_i^m$ values, they are presented as a population normalized accessibility index.
-This makes the interpretation of the 'potentially interacted opportunities' relative to the region, making comparisons of the results across different regions challenging.
-
-However, opportunities in a region can be considered finite.
-There are only so many school-seats, hospital capacity, jobs, etc., in a region and if one person interacts with an opportunity at a given time, it is taken.
-As such, if one person is advantaged and has the ability to reach more opportunities through a lower travel-cost mode, than they have more opportunities to potentially interact with more opportunities than other people.
-From the other perspective, their are fewer opportunities left to be potentially interacted with for populations using higher travel-cost modes.
-In this way, populations using modes with a higher travel impedance are at a higher access disadvantage than populations using lower travel impedance modes.
-This recognition is the motivation behind integrating _competition_ for opportunities within multimodal accessibility measures.
-Arguably one of the most popular competitive location-based accessibility measures is the two-step floating catchment area (2SFCA) approach popularized by @luoMeasuresSpatialAccessibility2003 who simplified the approach proposed by @shenLocationCharacteristicsInnercity1998 (with similar considerations for competition in @weibullAxiomaticApproachMeasurement1976 and @josephMeasuringPotentialPhysical1982).
-
-The Shen-type accessibility measure's formulation is: $a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}$ where $D_j^m$ is the potential demand for opportunities equal to travel impedance weighted population $\sum_i P_i^m f^m(c_{ij}^m)$.
-In this way, the Shen-type measure can be understood as a ratio of the potential opportunity supply over the potential demand for opportunities.
-The measure considers competition, but it is _non-constrained_.
-A score of competitive potential accessibility associated is associated with each location $i$ for each mode $m$, but there are no global maximums.
-In other words, it is difficult to interpret the meaning of differences in Shen-type accessibility scores between modes.
-
-To illustrate, @taoInvestigatingImpactsPublic2020a calculates $a_i^m$ to jobs for different income-group populations in Shenzhen (China) using $m = \text{public transit}$ and $m={car}$.
+In other words, each $i$ has three $S_i^m$ values, one per $m$.
+In this work, they demonstrate that the car mode has the highest $S_i^{m}$ values in the majority of $i$, i.e., populations using a car can potentially interact with the most retail opportunities than populations using other modes.
+However, the higher car $S_i^{m}$ values are not a result of lower $S_i^{m}$ values for other modes: it is not assumed that car-using populations accessing more opportunities take away potential opportunities from other populations within the measure (i.e., no consideration for competition).
+This measure is also not constrained so there is no global maximum for $S_i$ or $S_i^m$ values.
+As such, $S_i^m$ values are presented as standardized accessibility index in this work, from the lowest value of 0 to the highest value of 1.
+This presentation or discussion of Hansen-type accessibility measures is common in the literature e.g., see the work of @maharjanSpatialEquityModal2022 and @campbell2019accessibility.
+However, this representation of results can make the interpretation of results challenging _because_ values are region-relative and lacking numeric meaning.
+Zones will always have values from between 0 to 1 represented, but is a zone with a low value for pedestrian-modes and a high value for car-modes notable?
+And if notable, what does the difference in these standardized values tangibly mean for planners?
+By how much should transport systems and land-use configurations be changed to improve conditions?
+And in what way can these existing methods be used to track progress over time?
+Or between regions?
+This is challenging to track, since certain values will always be relatively 'low' and 'high'.
+Alternatively, modal accessibility calculated using the Hansen-type measure is also presented without standardization, as in the work by @lunkeModalAccessibilityDisparities2022.
+Often, the meaning of the accessibility value itself is not interpreted, as it is difficult to state what having access to 30,000 employment opportunities by car but having 10,000 employment opportunity by transit tangibly signifies.
+
+However, once opportunities in a region are interpreted as finite, a new meaning is taken on.
+As considered in the long tradition of accessibility work, capacity of opportunities is limited and thus is subject to competition by population [@soukhovIntroducingSpatialAvailability2023; @grengsJobAccessibilityModal2010; @kawabataJobAccessibilityIndicator2006; @kwokUseModalAccessibility2004; @morrisAccessibilityIndicatorsTransport1979; @weibullAxiomaticApproachMeasurement1976; @shenLocationCharacteristicsInnercity1998].
+There are only so many school-seats, hospital capacity, employment opportunities, etc., in a region and if one person interacts with an opportunity at a given time, it is taken: the supply of an opportunity and the demand for that opportunity are two components of accessibility.
+The competition consideration is especially important in the case of multiple modes.
+For instance, people in a zone who are advantaged with a relatively low travel-cost mode have the ability to potentially interact with more opportunities than other people.
+Due to this advantage, through the perspective of finite opportunities, there are fewer opportunities left to be potentially interacted for everyone else, especially zones with populations using higher travel-cost modes.
+This recognition is the motivation behind integrating _competition_ for opportunities within modal accessibility measures.
+
+Arguably, one of the most popular competitive location-based accessibility measures is the two-step floating catchment area (2SFCA) approach popularized by @luoMeasuresSpatialAccessibility2003 who simplified the approach proposed by @shenLocationCharacteristicsInnercity1998 (with similar considerations for competition in @weibullAxiomaticApproachMeasurement1976 and @josephMeasuringPotentialPhysical1982).
+The Shen-type accessibility measure's formulation is: $a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}$ where $D_j^m$ is the potential demand for opportunities equal to travel impedance weighted population $\sum_i P_i^m f^m(c_{ij}^m)$ and the remaining variables are repeated in the Hansen-type measure.
+The Shen-type measure per mode ($a_i^m$) can be understood as a ratio of the travel impedance-weighted supply of opportunities for the $m$-using population in $i$ over the ravel impedance-weighted demand for opportunities.
+In this way, it considers competition but the measure remains _non-constrained_ meaning resulting values are not associated with global maximums.
+In other words, it is difficult to interpret the meaning of differences in Shen-type accessibility scores between modes in similar ways as reflected for Hansen-type measures.
+
+To illustrate, @taoInvestigatingImpactsPublic2020a calculates $a_i^m$ to jobs for different income-group populations in Shenzhen, China using $m = \text{public transit}$ and $m={car}$.
 They demonstrate that $i$ with low-income populations have lower $a_i^m$ than $i$ with higher-income populations.
 Further, they demonstrate that $a_i^{m=\text{public transit}}$ is lower than $a_i^{m=\text{car}}$ at many $i$, arguing that this may put $i$ with lower-income populations in a further disadvantage.
-$a_i$ and/or $a_i^m$ are used to compare relative spatial differences in overall competitive accessibility and modal competitive accessibility, but because there is no global maximum, making it is difficult to interpret the significance between differences in $a_i^{m}$ values.
+$a_i$ and/or $a_i^m$ are used to compare relative spatial differences in overall competitive accessibility and multimodal competitive accessibility, but because there is no global maximum, making it is difficult to interpret the significance between differences in $a_i^{m}$ values.
 Questions such as: what is the impact that competition has on the difference in $a_i^m$ values?
@@
 And what is the interpretation of this difference?
-are left unanswered.
-
-Spatial availability improves on previous multimodal accessibility approaches as it considers _competition_ in the potential interaction with opportunities in a _constrained_ framework (e.g., finite opportunities).
+are left unaddressed.
+
+Spatial availability improves on previously discussed multimodal accessibility approaches using the Hansen-type measure and the Shen-type measure by taking into account _competition_ in the potential interaction with opportunities in a _constrained_ framework (e.g., finite opportunities).
 This is done by considering: 1) competition between mass effect (e.g., the advantage of sub-populations residing in relatively low population-density and high opportunity-proximate areas) and 2) competition between travel impedance (e.g., sub-populations with relatively low travel-impedance) through a proportional allocation mechanism.
@@
 In other words, assuming a finite number of opportunities in the region, $F^t_{ij}$ proportionally allocates $O_j$ to each $i$ such that the resulting $V_i$ value represents the number of opportunities _spatially available_ to the population at $i$.
-This value can be seen to represent spatial availability as it is a proportion of the opportunities in the region (i.e., $\sum_j O_j = \sum_i V_i$).
-
-The focus of this paper is to extend $V_i$ for multimodal applications.
+Each zonal value is a proportion of the opportunities in the region (i.e., $\sum_j O_j = \sum_i V_i$).
+
+The focus of this paper is to extend $V_i$ for the measurement of multimodal accessibility applications.
 To do so, the balancing factors are reformulated to yield a proportional value for the set of modes $m$ used by populations at each $i$.
-As these factors are proportional, $F^{pm}_{i}$ and $F^{cm}_{ij}$ can be summed up across each $m$ at each $i$ and across all $i$ to equal to 1.
+As these factors are proportional, $F^{pm}_{i}$ and $F^{cm}_{ij}$ can be summed across each $m$ at each $i$ and then across all $i$ to equal to 1.
 They are also similarly combined multiplicatively to obtain their joint effect, represented as the combined balancing factor $F^{tm}_{ij}$ detailed in Equation (\ref{eq:multimodal-balancing-factors}).
@@
 
-## Synthetic example
-
-Consider the following: Figure \ref{fig:Fig1} depicts a region with population and jobs at three population centers ($A$, $B$, $C$) and three employment centers ($1$, $2$, $3$).
+## An illustrative synthetic example
+Consider the following Figure \ref{fig:Fig1}, it depicts a region with population and jobs at three population centers ($A$, $B$, $C$) and three employment centers ($1$, $2$, $3$).
 The population at each population center is divided into two sub-populations, one using a faster mode $z$ and another using a slower mode $x$, to travel to employment centers.
@@
 
-```{r, pop-plot, fig.cap="\\label{fig:Fig3} Population living and working in Madrid, by four summarized modal categories, $P^m_i$ as reported by the 2018 travel survey.", fig.show='hold', fig.align='center'}
+```{r, pop-plot, fig.cap="\\label{fig:Fig3} Population living and working in Madrid by four summarized modal categories $P^m_i$ as reported by the home-to-work flows in the 2018 travel survey.
+Gray TAZs have no population.", fig.show='hold', fig.align='center'}
 knitr::include_graphics("images/im_populations_zn208_plot.png")
 ```
-The 2018 Community of Madrid travel survey (@comunidaddemadridResultadosEDM20182020) is the source of data for this empirical example: it is a representative survey that reflects a snap-shot of the travel patterns for one typical day of the working week (e.g., n=222,744 trips with representative population elevation factors).
-In this paper, a sample of the travel survey is used, namely the residential home origin to work destination trips of all modes and those that originate and end in the city of Madrid.
-These totals are displayed in Figure \ref{fig:Fig2} and Figure \ref{fig:Fig3}.
-Both figures are displayed at the level of traffic analysis zones ($i$ and $j$) that correspond to the survey.
+The 2018 Community of Madrid travel survey [@comunidaddemadridResultadosEDM20182020] is the source of data for this empirical example: it is a representative survey that reflects a snap-shot of the travel patterns for a typical day of the working week in 2018.
+Specifically, it includes 222,744 trips taken from a representative sample of 85,064 households across the traffic analysis zones (TAZ) representing the Community of Madrid (population of 6,507,184 over 3 years old) through population elevation factors.
+
+In this empirical example, we only use a sub-set of the survey, specifically home-to-full-time-work trips, by all modes, to demonstrate multimodal spatial availability of employment opportunities.
+The total population who travels to full-time work from an origin (i.e., the worker), and the destinations (i.e., the employment opportunity), for the City of Madrid (a sub-set of the Community of Madrid) is visualized in Figure \ref{fig:Fig2} and Figure \ref{fig:Fig3}.
+Both figures are displayed at the level of TAZ (the $i$ and $j$ zones) that correspond to the survey.
 The red boundary represents the LEZ Centro in effect in 2017 and thus those travel patterns of car-restriction reflected in the survey.
@@
 The total sum of jobs $O_j$ that are held are shown in Figure \ref{fig:Fig2} and the populations that go to a work destination by four modal categories $P^m_i$, is reflected in Figure \ref{fig:Fig3}.
-The modal categories represented in Figure \ref{fig:Fig3} are summarized for the following trip mode types:
-
-   - Car/motor: all cars and operating modes (e.g., cab, private driver, company, rental care, main driver, passenger, etc.) and all public, private or company motorcycle/mopeds.
+The four modal categories represented in Figure \ref{fig:Fig3} are created through the grouping of similar mode types as provided in the survey, by the authors discretion.
+The modal categories and the mode types within each category are reported as follows:
+
+   - Car/motor: all cars and operating modes (e.g., cab, private driver, company, rental car, main driver of a private car, passenger in a private car) and all public, private or company motorcycle/mopeds.
   - Transit: all bus, trams, and trains,
   - Bike: all bicycle trips (e.g., private, public, or company bike trips) and "other" types of micromobility options,
-  - Walk: walking or by foot,
+  - Walk: walking or by foot.
 
@@
 These interpretations were difficult to be made with accessibility measures in the past.
+In future work, we intend to use multimodal spatial availability to characterise the equity of multimodal accessibility losses and gains in varying LEZ policy scenario implementations.
 

### 2023-11-07 Journal-manuscript/Editorial correspondance/Response-to-reviewers_R1.Rmd
Edits to the introduction

@@
 
-\textcolor{blue}{We have done this. Thank you again for your thoughtful comments and suggestions to improve the paper.}
+\textcolor{blue}{We have done this.
+In the original version of the abstract and introduction we gave the misleading impression that we would analyze changes in the system, when in reality our intent was to demonstrate the application of the measure in an empirical example.
+We do plan to study changes in the system, but this requires more work than can be presented in a single paper, partly for the reasons that you identified above (modal shifts and the calculation of spatial availability for two different scenarios, as well as the analysis of the differences between them.
+We will do this in a future applied paper.}
+
+\textcolor{blue}{Thank you again for your thoughtful comments and suggestions to improve the paper.}
 

### 2023-11-07 Journal-manuscript/Journal-manuscript.Rmd
Edits to the introduction

@@
 ---
-title: "Multimodal spatial availability: a singly-constrained measure of competitive accessibility considering multiple modes"
+title: "Multimodal spatial availability: a singly-constrained measure of accessibility considering multiple modes"
 
@@
 abstract: |
-  An increasing number of studies within the domain of transportation planning are concerned with the inequities in accessibility to opportunities.
-A dimension of these inequities arises from differences in access by mode type (e.g., commuting using a car as opposed to transit).
-However, methods implemented in current accessibility literature are lacking within the context of multimodal analysis.
-This paper presents an extension of spatial availability, a singly-constrained competitive accessibility measure, for the context of multimodal accessibility analysis.
-We first illustrate the features of spatial availability that lend itself to multimodal analysis using synthetic example.
-We then demonstrate its use on the case study of Low Emission Zones in Madrid (Spain) and highlight how this policy intervention changes the accessibility of populations using different modes.
-In summary, spatial availability can be used to create and interpret multimodal policy intervention scenarios unlike previous methods: this creation and interpretation can help regions envision a more sustainable and equitable access-to-opportunity landscape.
-
-keywords: ["Multimodal", "Competitive", "Accessibility", "Equity", "Policy scenarios", "Low emission zones"]
-
-bibliography: mybibfile.bib
+  Recent research has aimed to address the way opportunities are counted in accessibility analysis.
+In conventional accessibility measures, opportunities are often multiply counted, which leads to values of accessibility that are difficult to interpret.
+Constraining the calculations to match a known quantity ensures that the measurements sum up to a predetermined quantity (i.e., the total number of opportunities), and so each value can be meaningfully related to this total.
+A recent such effort is spatial availability, a singly-constrained accessibility measure.
+In this paper we extend spatial availability for use in the case of multiple modes or, more generally, heterogeneous population segments with distinct travel behaviors.
+After deriving a multimodal version of spatial availability We proceed to illustrate its features using a synthetic example.
+Next, we apply it to an empirical example in Madrid.
+We conclude the paper with suggestions for future research.
+
+keywords: ["Spatial availability", "Multimodal", "Singly-constrained", "Accessibility", "Madrid", "Reproducible research"]
+
+bibliography:
+  - "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
 output: rticles::plos_article
@@
 
-Implementing policies that re-shape cities through accessibility gains (i.e., the _potential to interact_ with opportunities as a result of land-use mix and transport systems as defined in @hansenHowAccessibilityShapes1959) have been widely applied within the transportation literature and increasingly discussed by planners [@levinsonTransportAccessManual2020; @gowerPlanningInnovationCity2022; @siddiqToolsTradeAssessing2021; @yanAccessibilityBasedPlanningAddressing2021].
+<!-- WHAT IS ACCESSIBILITY AND WHY IS IT A RELEVANT TOPIC OF RESEARCH -->
+
+Accessibility is a key concept in the analysis of land use and transportation systems [e.g., @hansenHowAccessibilityShapes1959; @geursAccessibility2004; @paezPositive2012], and one that is coming of age from the perspective of planning too [see, _inter alia_, @handyAccessibility2020; @levinsonTransportAccessManual2020; @siddiqToolsTradeAssessing2021; @yanAccessibilityBasedPlanningAddressing2021; @elGeneidyMaking2022].
+Beginning with the work of Hansen [@hansenHowAccessibilityShapes1959], accessibility measures have been widely used to evaluate the efficiency of transportation systems when combined with the distribution of opportunities in space [e.g., ].
+As such, it is a holistic measure of spatial systems that measures the ease of reaching destinations [@handyMeasuring1997; @kwanSpace1998].
+
+<!-- CLEARLY STATE THE PROBLEM -->
+
+In practice, the most common form of accessibility measure is based on the gravity model.
+These measures are sums of weighted opportunities around a focal point (i.e., a potential origin), based on how expensive it is to reach them.
+Recent research in accessibility analysis has paid attention to the way opportunities are counted in the pertinent calculations.
+Conventionally, the sums are not constrained, which means that the same opportunity can enter the sum for different origins.
+Counting the same opportunity multiple times treats it as if it was inexhaustible.
+But opportunities in general are not inexhaustible, and in fact some of them are by definition exclusive: for example, once a job is taken up by someone in the population, the same job is no longer available for any other person to take.
+More generally, opportunities are subject to congestion: for example, multiple people can avail themselves of the services of the same family doctor, but the more people who do, the more congested the service will be.
+
+The issue of congestion was the motivation for the development of floating catchment area approaches in accessibility analysis [@shenLocationCharacteristicsInnercity1998; @luoMeasuresSpatialAccessibility2003].
+While these approaches purport to account for congestions, Paez et al.
+[@paezDemand2019] demonstrate that in general they do not solve the issue of multiple counting of opportunities, thus leading to biases in the calculation of total demand and supply, sometimes inflating them, other times deflating them.
+In response to this, recent research has paid closer attention to the way opportunities are counted in accessibility analysis.
+@paezDemand2019, for example, tackle floating catchment area methods and introduce a normalization of the impedance matrix to allocate the population and then the level of service proportionally.
+More recently, Soukhov et al.
+[@soukhovIntroducingSpatialAvailability2023] introduced a singly-constrained measure of accessibility, called spatial availability, that employs a similar, but more sophisticated proportional allocation mechanism.
+The work of these authors shows that floating catchment area methods can be seen as singly-constrained accessibility measures, and improve on existing approaches by guaranteeing that each opportunity is counted only once - in other words, treating opportunities as _finite_.
+The proportional allocation of spatial availability constrains the calculations to match a known quantity, therefore ensuring that the measurements sum up to a predetermined quantity (i.e., the total number of opportunities), and so each value can be meaningfully related to this total.
+
+A limitation of spatial availability as introduced by Soukhov et al.
+[@soukhovIntroducingSpatialAvailability2023] is that it was developed for the case of a homogeneous population, for example for the case of a single mode of transportation.
+However, the finite nature of opportunities makes the analysis of heterogeneous populations very relevant.
+In the case of multiple modes of transportation, people who travel by slow modes (e.g., active modes) can usually reach fewer opportunities than people who travel by faster modes and whose range is typically far wider (e.g., car).
+This implies that slower travelers will often face increased competition for local opportunities from travelers who can reach said opportunities from farther afield.
+
+<!-- WHAT DOES THIS PAPER DO? -->
+
+The objective of this paper is to address this limitation of spatial availability.
+Our primary motivation is to extend spatial availability for the case of multimodal accessibility, but it is worthwhile noting that this is in fact just one case of heterogeneous populations (i.e., travel by different modes).
+The method itself can easily accommodate other forms of heterogeneity, for example variations in travel behavior between older and younger adults [e.g., @paezHealth2010], the propensity of older adults to use different modes of transportation [@moniruzzamanMode2013], the usually shorter trip lengths of children compared to grown-ups [e.g., @reyesWalking2014], or the more limited travel ranges of single parents [e.g., @paezJobs2013].
+
+The paper rest of this paper is organized as follows.
+In Section we use a synthetic examples with multiple modes to illustrate some relevant issues; this helps us motivate the derivation of the new spatial availability expressions.
+This is followed in Section 3 by an empirical example using data from the city of Madrid after the implementation of its Low Emission Zones.
+Data for this example comes from the city's 2018 travel survey.
+The example shows the differences in spatial availability within and outside the LEZ for travellers using different modes, namely car, transit, cycling and walking.
+In Section 4, we provide concluding remarks on the strengths of the use of spatial availability as a multimodal accessibility measure, and discuss potential future uses in policy planning scenarios as well as directions for future research.
+
+<!--
+In @soukhovIntroducingSpatialAvailability2023, we introduce spatial availability, a type of location-based accessibility measure that is both _constrained_ and _competitive_.
+In this paper, we extend the spatial availability measure into a multimodal framework and explore its use in answering the question outlined: _"given opportunities are seen as finite, how many opportunities are available to a mode-using population given a spatial location?"._ The answer to this question quantifies how many opportunities can be accessed, considering competition, for different modes.
+
+Implementing policies that re-shape cities through accessibility gains (i.e., the _potential to interact_ with opportunities as a result of land-use mix and transport systems as defined in
+
 An important challenge in the identification of interventions that equitably transform cities is the effective evaluation of _trade-offs_: cities are complex and dynamic ecologies, and advantaging one component of the city can disadvantage another area, population, or sub-component.
@@
 Evaluating policy impacts in the context of _finity_ provides a way to contextualize the balance of trade-offs that the citizens of a city should tolerate.
-
-From the perspective of urban transport systems, location-based accessibility measures have been used in the context of policy evaluation.
-For instance, @leeMeasuringImpactsNew2018 assesses the transit accessibility gains to healthcare and employment opportunities for disadvantaged neighbourhood in the city of Columbus, USA after the transit system's re-design and introduction of a rapid bus system.
-A limitation of this study is they do not calculate results under a _constrained_ framework i.e., one of _finity_.
-The citizens of Columbus should experience quantitative accessibility gains - but is it at the expense of access using other modes?
-This limitation is similar to others that implement accessibility measures.
-As another example, @mohriClusteringMethodMeasuring2021 implements a modified cumulative opportunity measure to assess differences between private vehicle and transit system accessibility to jobs in Melbourne, but a similar question remains: does the accessibility afforded to the private vehicle using population come at the expense of accessibility losses to transit users?
-
-@leeMeasuringImpactsNew2018 and @mohriClusteringMethodMeasuring2021 both use _non-competitive_ accessibility measures.
-There is a branch of location-based accessibility measures that do incorporate the effect of competition for opportunities by the population in the region.
-However, we argue that these existing methods fall short in acknowledging the _finity_ of opportunities.
-For instance, @maoMeasuringSpatialAccessibility2013 applies a competitive measure, two-step floating catchment approach (2SFCA), for the case of access to healthcare services in Florida for both a multimodal network and a single modal network.
-While the differences in modal access are discussed, the question of how the advantage in access afforded by one mode over another impacts access for different mode users is not addressed.
 
@@
 Measures that evaluate the accessibility of modes given both _constrained_ and _competitive_ considerations are lacking in the literature, but are needed, to evaluate such an intervention's impact on accessibility gains and losses within a _finite_ conceptualization.
-
-In @soukhovIntroducingSpatialAvailability2023, we introduce spatial availability, a type of location-based accessibility measure that is both _constrained_ and _competitive_.
-In this paper, we extend the spatial availability measure into a multimodal framework and explore its use in answering the question outlined: _"given opportunities are seen as finite, how many opportunities are available to a mode-using population given a spatial location?"._ The answer to this question quantifies how many opportunities can be accessed, considering competition, for different modes.
-
-To foreground our methodological contribution, in Section 2, we outline shortcomings of a few location-based accessibility measures in comparison to spatial availability using a synthetic example of a multiple mode using population.
-In Section 3, the spatial availability of four mode categories is calculated for the city of Madrid, Spain after the 2017 LEZ is implemented.
-Using mode-specific home-to-work origin-destination 2018 travel survey as empirical data, we demonstrate inequities of spatial availability within and outside the LEZ for each sub-population using car, transit, cycling and walking modes.
-In Section 4, we provide concluding remarks on the strengths of the use of spatial availability as a multimodal accessibility measure and potential future uses in policy planning scenarios.
+ -->
 
 # A review of multimodal accessibility measures
-Location-based accessibility indicators are quantitative measures of _potential_ interaction with opportunities for locations within a given region: they are a product of the relationship between land-use and transport systems.
-Arguably, the most commonly used location-based measures is the cumulative opportunity measure and thee weighted cumulative opportunity measures [@levinsonTransportAccessManual2020].
-These measures weight the opportunities that can be potentially interacted with from origin ($i$) to destination ($j$) based on some sort of travel cost function (e.g., travel time, fare, travel distance), otherwise known as a travel impedance function $f^{m}(c^m_{ij})$.
-Many weighted cumulative opportunities (often referred to as the gravity-based measure) originate from the measure proposed by @hansenHowAccessibilityShapes1959, which can take the following multimodal form: $S_i^m = \sum_j O_j f^m(c_{ij}^m)$ where $m$ is a set of modes which have mode-specific travel costs ($(c_{ij}^m)$), travel impedance functions ($f^m(\cdot)$) for the mode-specific Hansen-type value ($S_i^m$).
-
-The Hansen-type measure does not consider competition between modes nor is it constrained.
+
+Location-based accessibility indicators are quantitative measures of _potential_ interaction with opportunities for locations within a given region: they are summary measures of the relationship between land-use and transport systems.
+Arguably, the most commonly used are measures based on the gravity model [@wuUnifyingAccess2020], of which cumulative opportunity measures and weighted cumulative opportunity measures are particular forms [@levinsonTransportAccessManual2020].
+These measures assign a weight to opportunities based on how easy it is to reach them.
+Given an origin ($i$) and a destination ($j$), an impedance function $f^{m}(c^m_{ij})$ converts the cost of travel (e.g., time, money, generalized cost) into a score that represents the propensity for interaction.
+These measures originate from that proposed by @hansenHowAccessibilityShapes1959, which can take the following form in the multimodal case: $S_i^m = \sum_j O_j f^m(c_{ij}^m)$ where $m$ is a set of modes which have mode-specific travel costs ($(c_{ij}^m)$) and/or travel impedance functions ($f^m(\cdot)$).
+
+The Hansen-type measure is not constrained, and as a result does not consider the opportunities as finite.
 As an example, the work of @tahmasbiMultimodalAccessibilitybasedEquity2019 uses the Hansen-type measure to measure the potential interaction with retail locations using walking, public transit, and car modes as $m$.
-$S_i^m$ is the sum of retail locations $j$ that can potentially be interacted with under the travel impedance as calculated for each $i$ and $m$.
+$S_i^m$ is the sum of retail locations $j$ that can potentially be reached under the travel impedance as calculated for each $i$ and $m$.
 In other words, each $i$ has three $S_i^m$ values, one per $m$.
-In this work, they demonstrate that the car mode has the highest $S_i^{m}$ values in the majority of $i$, i.e., populations using a car can potentially interact with the most retail opportunities than populations using other modes.
-However, the higher car $S_i^{m}$ values are not a result of lower $S_i^{m}$ values for other modes: it is not assumed that car-using populations accessing more opportunities take away potential opportunities from other populations within the measure (i.e., no consideration for competition).
-This measure is also not constrained so there is no global maximum for $S_i$ or $S_i^m$ values.
-As such, $S_i^m$ values are presented as standardized accessibility index in this work, from the lowest value of 0 to the highest value of 1.
-This presentation or discussion of Hansen-type accessibility measures is common in the literature e.g., see the work of @maharjanSpatialEquityModal2022 and @campbell2019accessibility.
+In this work, they demonstrate that the car mode has the highest $S_i^{m}$ values in the majority of $i$, i.e., populations using a car can potentially reach more retail opportunities than populations using other modes.
+However, higher $S_i^{m}$ values for car do not affect the values of $S_i^{m}$ for other modes: in effect, each modes is analyzed as if the others did not exist, which would be reasonable for an inexhaustible opportunity, but is more questionable when the opportunity is subject to congestion.
+Since the measure is measure is also not constrained, and each opportunity can be and often is counted multiple times within and between modes, the sum of accessibility is not a meaningful quantity.
+To improve interpretability, $S_i^m$ values are presented as standardized accessibility index in this work, from the lowest value of 0 to the highest value of 1.
+This presentation or discussion of Hansen-type accessibility measures is common in the literature [e.g., in @campbell2019accessibility; and @maharjanSpatialEquityModal2022].
+
 However, this representation of results can make the interpretation of results challenging _because_ values are region-relative and lacking numeric meaning.
@@
 
-## Multimodal data and methods
+## Data and methods
 ```{r}

### 2023-11-07 Journal-manuscript/Journal-manuscript.tex
Edits to the introduction

@@
 \textbf\newline{Multimodal spatial availability: a singly-constrained
-measure of competitive accessibility considering multiple
+measure of accessibility considering multiple
 modes} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
@@
 \section*{Abstract}
-An increasing number of studies within the domain of transportation
-planning are concerned with the inequities in accessibility to
-opportunities. A dimension of these inequities arises from differences
-in access by mode type (e.g., commuting using a car as opposed to
-transit). However, methods implemented in current accessibility
-literature are lacking within the context of multimodal analysis. This
-paper presents an extension of spatial availability, a
-singly-constrained competitive accessibility measure, for the context of
-multimodal accessibility analysis. We first illustrate the features of
-spatial availability that lend itself to multimodal analysis. We then
-demonstrate its use on the case study of Low Emission Zones in Madrid
-(Spain) and highlight how this policy intervention changes the
-accessibility of populations using different modes. In summary, spatial
-availability can be used to create and interpret multimodal policy
-intervention scenarios unlike previous methods: this creation and
-interpretation can help regions envision a more sustainable and
-equitable access-to-opportunity landscape.
+Recent research has tried to address the way opportunities are counted
+in accessibility analysis. In conventional accessibility measures,
+opportunities are often multiply counted, which leads to values of
+accessibility that are difficult to interpret. Constraining the
+calculations to match a known quantity ensures that the measurements sum
+up to a predetermined quantity (i.e., the total number of
+opportunities), and so each value can be meaningfully related to this
+total. A recent such effort is spatial availability, a
+singly-constrained accessibility measure. In this paper we extend
+spatial availability for use in the case of multiple modes or, more
+generally, heterogeneous population segments with distinct travel
+behaviors. After deriving a multimodal version of spatial availability
+We proceed to illustrate its features using a synthetic example. Next,
+we apply it to an empirical example in Madrid. We conclude the paper
+with suggestions for future research.
 
@@
 
-Implementing urban policies that re-shape cities through accessibility
-gains (i.e., the \emph{potential to interact} with opportunities as a
-result of land-use mix and transport systems as originally defined by
-{[}1{]}) have been widely applied within the transportation literature
-and is increasingly discussed by planners {[}2--5{]}. An important
-challenge in the identification of interventions that equitably
-transform cities is the effective evaluation of \emph{trade-offs}:
-cities are complex and dynamic ecologies, and advantaging one component
-of the city can disadvantage another area, population, or sub-component
-In this way, policy evaluation should take a \emph{systems} approach
-{[}6{]}. One way of considering systems is from the perspective of the
-\emph{finite}. As an illustration, consider the amount of transport
-space within a city: the amount is typically finite so re-allocating
-road space away from one mode directly impacts the performance of the
-others (see the literature on road space reallocation e.g., {[}7{]}).
-Evaluating policy impacts in the context of \emph{finity} provides a way
-to contextualize the balance of trade-offs that the citizens of a city
-should tolerate.
-
-From the perspective of urban transport systems, location-based
-accessibility measures have been used in the context of policy
-evaluation. For instance, {[}8{]} assesses the transit accessibility
-gains to healthcare and employment opportunities for disadvantaged
-neighbourhood in Columbus, Ohio, USA after the transit system's
-re-design and introduction of a rapid bus system. However, a limitation
-of this study, like others that implement accessibility measures, is
-they do not calculate results under a \emph{constrained} framework i.e.,
-one of \emph{finity}. The citizens of Columbus should experience
-quantitative accessibility gains - but is it at the expense of access
-using other modes? As another example, {[}9{]} implements a modified
-cumulative opportunity measure to assess differences between private
-vehicle and transit system accessibility to jobs in Melbourne, but a
-similar question remains: does the accessibility afforded to the private
-vehicle using population come at the expense of accessibility losses to
-transit users?
-
-{[}8{]} and {[}9{]} both use \emph{non-competitive} accessibility
-measures. There is a branch of location-based accessibility measures
-that do incorporate the effect of competition for opportunities by the
-population in the region. However, we argue that these existing methods
-fall short in acknowledging the \emph{finity} of opportunities. For
-instance, {[}10{]} applies a competitive measure, two-step floating
-catchment approach (2SFCA), for the case of access to healthcare
-services in Florida for both a multimodal network and a single modal
-network. While the differences in modal access are discussed, the
-question of how the advantage in access afforded by one mode over
-another impacts access for different mode users is unanswered.
-
-This question of how much one mode-using population can access at the
-expense of another mode-using population is a pertinent equity question
-in the evaluation of policy scenarios that are multimodal. For instance,
-consider the impact of a low emission zone (LEZ). LEZ is a policy of
-spatial and modal discrimination: the circulation of vehicles that are
-excessively polluting are restricted in specific regions. In the
-recognition that opportunities are finite, the implementation of a LEZ
-explicitly reduces the access that the population using polluting
-vehicles has to opportunities. This restriction allows the population
-using other more sustainable modes to potentially have a higher level of
-access than before the LEZ implementation. This evaluation is especially
-urgent as LEZ are currently in effect in cities globally; their
-reception has been mixed {[}11{]} and may be having negative impacts on
-disadvantaged populations who have become mobility-restricted
-{[}12,13{]}. Measures that evaluate the accessibility of modes given
-both \emph{constrained} and \emph{competitive} considerations are
-lacking in the literature, but are needed, to evaluate such policy
-interventions impact on accessibility.
-
-In {[}14{]}, we introduce spatial availability, a type of location-based
-accessibility measure that is both \emph{constrained} and
-\emph{competitive}. In this paper, we extend the spatial availability
-measure into a multimodal framework and explore its use in answering the
-question outlined: \emph{``given opportunities are finite, how many are
-available to a given location depending on the mode used?''.} The answer
-to this question quantifies how many opportunities can be accessed,
-considering competition, for different modes. To foreground this
-exploration, in Section 2, we discuss short falls of a few existing
-location-based measures in comparison to spatial availability through a
-synthetic example. In Section 3, the spatial availability of an
-empirical example of the LEZ in the city of Madrid, Spain is calculated.
-We demonstrate how the restriction of car circulation could have
-impacted the spatial availability of opportunities for each
-sub-population using transit, cycling and walking modes. In Section 4,
-we provide concluding remarks on the strengths of the use of spatial
-availability as a multimodal accessibility measure and potential future
-uses in policy planning scenarios.
+Accessibility is a key concept in the analysis of land use and
+transportation systems {[}e.g., 1,2,3{]}, and one that is coming of age
+from the perspective of planning too {[}see, \emph{inter alia},
+4,5--8{]}. Beginning with the work of {[}1{]}, accessibility measures
+have been widely used to evaluate the efficiency of transportation
+systems when combined with the distribution of opportunities in space
+{[}e.g., {]}. As such, it is a holistic measure of spatial systems that
+measures the ease of reaching destinations {[}9,10{]}.
+
+In practice, the most common form of accessibility measure is based on
+the gravity model. These measures are sums of weighted opportunities
+around a focal point (i.e., a potential origin), based on how expensive
+it is to reach them. Recent research in accessibility analysis has paid
+attention to the way opportunities are counted in the pertinent
+calculations. Conventionally, the sums are not constrained, which means
+that the same opportunity can enter the sum for different origins.
+Counting the same opportunity multiple times treats it as if it was
+inexhaustible. But opportunities in general are not inexhaustible, and
+in fact some of them are by definition exclusive: for example, once a
+job is taken up by someone in the population, the same job is no longer
+available for any other person to take. Other types of opportunities are
+at the very least subject to congestion: multiple people can avail
+themselves of the services of the same family doctor, but the more
+people who do, the more congested the service will be.
+
+The issue of congestion was the motivation for the development of
+floating catchment area approaches in accessibility analysis
+{[}11,12{]}. While these approaches purport to account for congestions,
+{[}13{]} demonstrate that in general they do not solve the issue of
+multiple counting of opportunities, thus leading to biases in the
+calculation of total demand and supply, sometimes inflating them, other
+times deflating them. In response to this, recent research has paid
+closer attention to the way opportunities are counted in accessibility
+analysis. {[}13{]}, for example, tackle floating catchment area methods
+and introduce a normalization of the impedance matrix to allocate the
+population and then the level of service proportionally. More recently,
+{[}14{]} introduced a singly-constrained measure of accessibility,
+called spatial availability, that employs a similar, but more
+sophisticated proportional allocation mechanism. The work of these
+authors shows that floating catchment area methods can be seen as
+singly-constrained accessibility measures, and improve on them by
+guaranteeing that each opportunity is counted only once - in other
+words, treating opportunties as \emph{finite}. The proportional
+allocation of spatial availability constrains the calculations to match
+a known quantity, therefore ensuring that the measurements sum up to a
+predetermined quantity (i.e., the total number of opportunities), and so
+each value can be meaningfully related to this total.
+
+A limitation of spatial availability as introduced by {[}14{]} is that
+it was developed for the case of a homogeneous population, for example
+for the case of a single mode of transportation. However, the finity of
+the opportunities makes the analysis of heterogeneous populations very
+relevant. In the case of multiple modes of transportation, people who
+travel by slow modes (e.g., active modes) can usually reach fewer
+opportunities than people who travel by faster modes and whose range is
+typically far wider (e.g., car). This implies that slower travelers will
+often face increased competition for local opportunities from travelers
+who can reach said opportunities from farther afield.
+
+The objective of this paper is to address this limitation of spatial
+availability. Our primary motivation is to extend spatial availability
+for the case of multimodal accessibility, but this is in fact just one
+example of heterogeneous populations (i.e., travel by different modes).
+The method itself can easily accommodate other forms of heterogeneity,
+for example variations in travel behavior between older and younger
+adults {[}e.g., {[}15{]}; {]}, the propensity of older adults to use
+different modes of transportation {[}16{]}, the usually shorter trip
+lengths of children compared to grown-ups {[}e.g., 17{]}, or the more
+limited travel ranges of single parents {[}e.g., 18{]}.
+
+The paper rest of this paper is organized as follows. In Section we use
+a synthetic examples with multiple modes to illustrate some relevant
+issues; this helps us motivate the derivation of the new spatial
+availability expressions. This is followed in Section 3 by an empirical
+example using data from the city of Madrid after the implementation of
+its Low Emission Zones. Data for this example comes from the city's 2018
+travel survey. The example shows the differences in spatial availability
+within and outside the LEZ for travellers using different modes, namely
+car, transit, cycling and walking. In Section 4, we provide concluding
+remarks on the strengths of the use of spatial availability as a
+multimodal accessibility measure, and discuss potential future uses in
+policy planning scenarios as well as directions for future research.
 
@@
 \emph{potential} interaction with opportunities for locations within a
-given region: they are a product of the relationship between land-use
-and transport systems. Arguably the most commonly used location-based
-measured are cumulative opportunity measures and weighted cumulative
-opportunity measures {[}2{]}. These measures weight the opportunities
-that can be potentially interacted with from origin \(i\) to destination
-\(j\) based on some sort of travel cost function (e.g., travel time,
-fare, travel distance) otherwise known as a travel impedance function
-\(f^{m}(c^m_{ij})\). Many weighted cumulative opportunities (often
-referred to as the gravity-based measure) originate from the measure
-proposed by {[}1{]}, which can take the following multimodal form:
+given region: they are summary measures of the relationship between
+land-use and transport systems. Arguably, the most commonly used are
+measures based on the gravity model {[}19{]}, of which cumulative
+opportunity measures and weighted cumulative opportunity measures are
+particular forms {[}5{]}. These measures assign a weight to
+opportunities based on how easy it is to reach them. Given an origin
+(\(i\)) and a destination (\(j\)), an impedance function
+\(f^{m}(c^m_{ij})\) converts the cost of travel (e.g., time, money,
+generalized cost) into a score that represents the propensity for
+interaction. These measures originate from that proposed by {[}1{]},
+which can take the following form in the multimodal case:
 \(S_i^m = \sum_j O_j f^m(c_{ij}^m)\) where \(m\) is a set of modes which
-have mode-specific travel costs \((c_{ij}^m)\) and travel impedance
-functions \(f^m(\cdot)\).
-
-The Hansen-type measure does not consider competition between modes nor
-is it constrained. As an example, the work of {[}15{]} uses the
-Hansen-type measure to measure the potential interaction with retail
-locations using walking, public transit, and car modes \(m\). \(S_i^m\)
-is the sum of retail locations \(j\) that can potentially be interacted
-with under the travel impedance as calculated for each \(i\) and \(m\).
-In other words, each \(i\) has three \(S_i\) values, one per \(m\). In
-this work, they demonstrate that the car mode has the highest
-\(S_i^{m=car}\) values in the majority of \(i\), i.e., populations using
-a car can potentially interact with the most retail opportunities than
-populations using other modes. However, the higher \(S_i^{m=car}\)
-values are not a result of lower \(S_i^{m}\) values for other modes: it
-is not assumed that car-using populations potentially accessing more
-opportunities take away potential opportunities for other populations
-within the measure (no consideration for competition). This measure is
-also not constrained: there is no global maximum for \(S_i\) or
-\(S_i^m\) values, they are presented as a population normalized
-accessibility index. This makes the interpretation of the `potentially
-interacted opportunities' relative to the region, making comparisons of
-the results across different regions challenging.
-
-However, opportunities in a region can be considered finite. There are
-only so many school-seats, hospital capacity, jobs, etc., in a region
-and if one person interacts with an opportunity at a given time, it is
-taken. As such, if one person is advantaged and has the ability to reach
-more opportunities through a lower travel-cost mode, than they have more
-opportunities to potentially interact with more opportunities than other
-people. From the other perspective, their are fewer opportunities left
-to be potentially interacted with for populations using higher
-travel-cost modes. In this way, populations using modes with a higher
-travel impedance are at a higher access disadvantage than populations
-using lower travel impedance modes. This recognition is the motivation
-behind integrating \emph{competition} for opportunities within
-multimodal accessibility measures. Arguably one of the most popular
-competitive location-based accessibility measures is the two-step
-floating catchment area (2SFCA) approach popularized by {[}16{]} who
-simplified the approach proposed by {[}17{]} (with similar
-considerations for competition in {[}18{]} and {[}19{]}).
-
-The Shen-type accessibility measure's formulation is:
+have mode-specific travel costs (\((c_{ij}^m)\)) and/or travel impedance
+functions (\(f^m(\cdot)\)).
+
+The Hansen-type measure is not constrained, and as a result does not
+consider the opportunities as finite. As an example, the work of
+{[}20{]} uses the Hansen-type measure to measure the potential
+interaction with retail locations using walking, public transit, and car
+modes as \(m\). \(S_i^m\) is the sum of retail locations \(j\) that can
+potentially be reached under the travel impedance as calculated for each
+\(i\) and \(m\). In other words, each \(i\) has three \(S_i^m\) values,
+one per \(m\). In this work, they demonstrate that the car mode has the
+highest \(S_i^{m}\) values in the majority of \(i\), i.e., populations
+using a car can potentially reach more retail opportunities than
+populations using other modes. However, higher \(S_i^{m}\) values for
+car do not affect the values of \(S_i^{m}\) for other modes: in effect,
+each modes is analyzed as if the others did not exist, which would be
+reasonable for an inexhaustible opportunity, but is more questionable
+when the opportunity is subject to congestion. Since the measure is
+measure is also not constrained, and each opportunity can be and often
+is counted multiple times within and between modes, the sum of
+accessibility is not a meaningful quantity. To improve interpretability,
+\(S_i^m\) values are presented as standardized accessibility index in
+this work, from the lowest value of 0 to the highest value of 1. This
+presentation or discussion of Hansen-type accessibility measures is
+common in the literature {[}e.g., in 21,and 22{]}.
+
+However, this representation of results can make the interpretation of
+results challenging \emph{because} values are region-relative and
+lacking numeric meaning. Zones will always have values from between 0 to
+1 represented, but is a zone with a low value for pedestrian-modes and a
+high value for car-modes notable? And if notable, what does the
+difference in these standardized values tangibly mean for planners? By
+how much should transport systems and land-use configurations be changed
+to improve conditions? And in what way can these existing methods be
+used to track progress over time? Or between regions? This is
+challenging to track, since certain values will always be relatively
+`low' and `high'. Alternatively, modal accessibility calculated using
+the Hansen-type measure is also presented without standardization, as in
+the work by {[}23{]}. Often, the meaning of the accessibility value
+itself is not interpreted, as it is difficult to state what having
+access to 30,000 employment opportunities by car but having 10,000
+employment opportunity by transit tangibly signifies.
+
+However, once opportunities in a region are interpreted as finite, a new
+meaning is taken on. As considered in the long tradition of
+accessibility work, capacity of opportunities is limited and thus is
+subject to competition by population {[}11,14,24--28{]}. There are only
+so many school-seats, hospital capacity, employment opportunities, etc.,
+in a region and if one person interacts with an opportunity at a given
+time, it is taken: the supply of an opportunity and the demand for that
+opportunity are two components of accessibility. The competition
+consideration is especially important in the case of multiple modes. For
+instance, people in a zone who are advantaged with a relatively low
+travel-cost mode have the ability to potentially interact with more
+opportunities than other people. Due to this advantage, through the
+perspective of finite opportunities, there are fewer opportunities left
+to be potentially interacted for everyone else, especially zones with
+populations using higher travel-cost modes. This recognition is the
+motivation behind integrating \emph{competition} for opportunities
+within modal accessibility measures.
+
+Arguably, one of the most popular competitive location-based
+accessibility measures is the two-step floating catchment area (2SFCA)
+approach popularized by {[}12{]} who simplified the approach proposed by
+{[}11{]} (with similar considerations for competition in {[}28{]} and
+{[}29{]}). The Shen-type accessibility measure's formulation is:
 \(a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}\) where \(D_j^m\)
 is the potential demand for opportunities equal to travel impedance
-weighted population \(\sum_i P_i^m f^m(c_{ij}^m)\). In this way, the
-Shen-type measure can be understood as a ratio of the potential
-opportunity supply over the potential demand for opportunities. The
-measure considers competition, but it is \emph{non-constrained}. A score
-of competitive potential accessibility associated is associated with
-each location \(i\) for each mode \(m\), but there are no global
-maximums. In other words, it is difficult to interpret the meaning of
-differences in Shen-type accessibility scores between modes.
-
-To illustrate, {[}20{]} calculates \(a_i^m\) to jobs for different
-income-group populations in Shenzhen (China) using
+weighted population \(\sum_i P_i^m f^m(c_{ij}^m)\) and the remaining
+variables are repeated in the Hansen-type measure. The Shen-type measure
+per mode (\(a_i^m\)) can be understood as a ratio of the travel
+impedance-weighted supply of opportunities for the \(m\)-using
+population in \(i\) over the ravel impedance-weighted demand for
+opportunities. In this way, it considers competition but the measure
+remains \emph{non-constrained} meaning resulting values are not
+associated with global maximums. In other words, it is difficult to
+interpret the meaning of differences in Shen-type accessibility scores
+between modes in similar ways as reflected for Hansen-type measures.
+
+To illustrate, {[}30{]} calculates \(a_i^m\) to jobs for different
+income-group populations in Shenzhen, China using
 \(m = \text{public transit}\) and \(m={car}\). They demonstrate that
@@
 to compare relative spatial differences in overall competitive
-accessibility and modal competitive accessibility, but because there is
-no global maximum, making it is difficult to interpret the significance
-between differences in \(a_i^{m}\) values. Questions such as: what is
-the impact that competition has on the difference in \(a_i^m\) values?
-How does impact vary spatially? And what is the interpretation of this
-difference? are left unanswered.
-
-Spatial availability improves on previous multimodal accessibility
-approaches as it considers \emph{competition} in the potential
+accessibility and multimodal competitive accessibility, but because
+there is no global maximum, making it is difficult to interpret the
+significance between differences in \(a_i^{m}\) values. Questions such
+as: what is the impact that competition has on the difference in
+\(a_i^m\) values? How does impact vary spatially? And what is the
+interpretation of this difference? are left unaddressed.
+
+Spatial availability improves on previously discussed multimodal
+accessibility approaches using the Hansen-type measure and the Shen-type
+measure by taking into account \emph{competition} in the potential
 interaction with opportunities in a \emph{constrained} framework (e.g.,
@@
 
-In brief, we define the \emph{spatial availability} at \(i\) (\(V_{i}\))
-as the proportion of all opportunities in the region \(O\) that are
-allocated to location \(i\) from all opportunity destinations \(j\).
-\(V_{i}\) is a value of how many opportunities are available to each
-location \(i\) out of all the opportunities in the region. The general
-formulation of spatial availability \(V_{i}\) is shown in Equation
+In brief, we define the spatial availability at \(i\) (\(V_{i}\)) as the
+proportion of all opportunities in the region \(O\) that are allocated
+to location \(i\) from all opportunity destinations \(j\). \(V_{i}\) is
+a value of how many opportunities are available to each location \(i\)
+out of all the opportunities in the region. The general formulation of
+spatial availability \(V_{i}\) is shown in Equation
 (\ref{eq:spatial-availability-general}):
@@
 factor \(F^c_{ij} = \frac{F^c_{ij}}{\sum_j F^c_{ij}}\) that,
-respectively, allocate opportunities to \(i\) in proportion to the size
+respectively, allocates opportunities to \(i\) in proportion to the size
 of the population at \(i\) (the mass effect) and the cost of reaching
@@
 the number of opportunities \emph{spatially available} to the population
-at \(i\). This value can be seen to represent spatial availability as it
-is a proportion of the opportunities in the region (i.e.,
-\(\sum_j O_j = \sum_i V_i\)).
-
-The focus of this paper is to extend \(V_i\) for multimodal
-applications. To do so, the balancing factors are reformulated to yield
-a proportional value for the set of modes \(m\) used by populations at
-each \(i\). As these factors are proportional, \(F^{pm}_{i}\) and
-\(F^{cm}_{ij}\) can be summed up across each \(m\) at each \(i\) and
-across all \(i\) to equal to 1. They are also similarly combined
-multiplicatively to obtain their joint effect, represented as the
-combined balancing factor \(F^{tm}_{ij}\) detailed in Equation
-(\ref{eq:multimodal-balancing-factors}).
+at \(i\). Each zonal value is a proportion of the opportunities in the
+region (i.e., \(\sum_j O_j = \sum_i V_i\)).
+
+The focus of this paper is to extend \(V_i\) for the measurement of
+multimodal accessibility applications. To do so, the balancing factors
+are reformulated to yield a proportional value for the set of modes
+\(m\) used by populations at each \(i\). As these factors are
+proportional, \(F^{pm}_{i}\) and \(F^{cm}_{ij}\) can be summed across
+each \(m\) at each \(i\) and then across all \(i\) to equal to 1. They
+are also similarly combined multiplicatively to obtain their joint
+effect, represented as the combined balancing factor \(F^{tm}_{ij}\)
+detailed in Equation (\ref{eq:multimodal-balancing-factors}).
 
@@
   The cost of travel balancing factor for each \(m\) at \(i\) is
-  \(F_{ij}^{cm} = \frac{f(c_{ij}^m)}{\sum_{m} \sum_{i} f(c_{ij}^m)}\)
+  \(F_{ij}^{cm} = \frac{f^m(c_{ij}^m)}{\sum_{m} \sum_{i} f^m(c_{ij}^m)}\)
 \end{itemize}
@@
 
-\hypertarget{synthetic-example}{%
-\subsection{Synthetic example}\label{synthetic-example}}
-
-Consider the following: Figure \ref{fig:Fig1} depicts a region with
+\hypertarget{an-illustrative-synthetic-example}{%
+\subsection{An illustrative synthetic
+example}\label{an-illustrative-synthetic-example}}
+
+Consider the following Figure \ref{fig:Fig1}, it depicts a region with
 population and jobs at three population centers (\(A\), \(B\), \(C\))
@@
 then \(C\). This synthetic example was inspired by the single-mode
-example used in {[}17{]} and reconfigured in {[}14{]}.
+example used in {[}11{]} and reconfigured in {[}14{]}.
 
@@
 most job opportunities. This appears to be the sub-population at \(A\)
-using \(z\). From the other perspective, sub-populations located in
-opposite conditions (i.e., further away from jobs, close to dense
-populations, and using \(x\)) are at a relative job opportunity access
-\emph{disadvantage}. From the perspective of inequities, the competition
-for opportunities between different mode-using populations matters as it
-reflects how well the land-use and transport system serves (or doesn't
-serve) them.
+using mode \(z\). Sub-populations located in opposite conditions (i.e.,
+further away from jobs, close to dense populations, and using \(x\)) are
+at a relative job opportunity access \emph{disadvantage}. The
+competition for opportunities between different mode-using populations
+matters from the perspective of inequities as it reflects how well the
+land-use and transport system serves (or doesn't serve) certain
+populations.
 
@@
 negative exponential impedance function
-\(f(c_{ij}) = \exp(-\beta\cdot c_{ij})\) with \(\beta=0.1\) for both
+\(f^m(c_{ij}^m) = \exp(-\beta\cdot c_{ij})\) with \(\beta=0.1\) for both
 \(x\) and \(z\) modes for all accessibility measures calculations.
@@
 sub-populations. Additionally, \(S_i^m\) is equal for both mode-using
-populations in \(A\) and \(B\). This is the case because \(S_i^m\) does
+populations in \(A\) and \(B\). This is the case since \(S_i^m\) does
 not consider \emph{competition}, it only relies on reflecting the count
@@
 connect this example to literature, \(S_i^m\) is calculated in the work
-of {[}15{]}; they compare differences in \(S_i^m\) values between modes
+of {[}20{]}; they compare differences in \(S_i^m\) values between modes
 in a relative and comparative sense, but make no further interpretation
@@
 similar interpretability issues. The Shen-type measure is implemented in
-the previously discussed work of {[}20{]} to calculate modal \(a_i^m\)
+the previously discussed work of {[}30{]} to calculate modal \(a_i^m\)
 values and the aggregated \(a_i\) is implemented in the work of
-{[}21{]}. However, similar to the Hansen-type measure, these works
+{[}31{]}. However, similar to the Hansen-type measure, these works
 discuss relative and spatially comparative differences in values, they
@@
 policy intervention to reduce GHG emissions, improve air quality, and
-support sustainable mobility in many countries. Though rules vary, LEZ
-aim to deter/reduce traffic in designated zones under threat of penalty
-(e.g., fines, seizure of vehicle). From the perspective of restriction
-for passenger transport, LEZ are a policy of \emph{geographic
-discrimination} as they change how people access opportunities by making
-the travel impedance more costly for car-mode users. If seeing
-opportunities as finite, this discrimination allows populations to
-access opportunities by other modes more readily than before. In this
-way, LEZ change the multimodal competitive accessibility landscape of a
-city.
+support sustainable mobility in many countries {[}32,33{]}. Though rules
+of exclusion vary, LEZ aim to deter/reduce traffic in designated zones
+under threat of penalty (e.g., fines, seizure of vehicle). From the
+perspective of restriction for passenger transport, LEZ are a policy of
+\emph{geographic discrimination} as they change how people access
+opportunities by making the travel impedance more costly for specific
+car-mode users. When considering opportunities as finite in a region,
+this discrimination could allow certain populations to access
+opportunities by other modes more readily than before. In this way, LEZ
+change the multimodal competitive accessibility landscape of a city.
 
@@
 their implementation as specified in their climate-change-related plans:
-\emph{Plan Nacional Integrado de Energía y Clima 2021-2030} {[}22{]} and
+\emph{Plan Nacional Integrado de Energía y Clima 2021-2030} {[}34{]} and
 \emph{Plan Nacional de Control de la Contaminación Atmosférica}
-{[}23{]}. Specifically, the national Spanish law 7/2021 ( \emph{Ley de
+{[}35{]}. Specifically, the national Spanish law 7/2021 ( \emph{Ley de
 Cambio Climático y Transición Energética}) will require all
@@
 air quality exceeds limits specified in \emph{RD 102/2011 de Mejora de
-Calidad del Aire} {[}24{]}.
+Calidad del Aire} {[}36{]}.
 
@@
 scope, the 2017 boundaries of the LEZ were relatively small (covering
-4.72 km\^{}\{2\}) and within the center (i.e., LEZ Centro). These
+4.72 km\(^2\)) and within the center (i.e., LEZ Centro). These
 boundaries were expanded in 2023 to inside of the M-30, a highway in
@@
 meet other exemptions. This restriction impacted approximately half of
-all car trips that were typically made into the LEZ Centro {[}25{]}.
+all car trips that were typically made into the LEZ Centro {[}37{]}.
 
@@
 
-\caption{\label{fig:Fig2} Jobs $O_j$ taken by people living and working in Madrid as reported by the 2018 travel survey.}\label{fig:jobs-plot}
+\caption{\label{fig:Fig2} Jobs $O_j$ taken by people living and working in Madrid as reported by the home-to-work flows in the 2018 travel survey. Gray TAZ has no jobs.}\label{fig:jobs-plot}
 \end{figure}
@@
 
-\caption{\label{fig:Fig3} Population living and working in Madrid, by four summarized modal categories, $P^m_i$ as reported by the 2018 travel survey.}\label{fig:pop-plot}
+\caption{\label{fig:Fig3} Population living and working in Madrid by four summarized modal categories $P^m_i$ as reported by the home-to-work flows in the 2018 travel survey.
+Gray TAZs have no population.}\label{fig:pop-plot}
 \end{figure}
 
-The 2018 Community of Madrid travel survey ({[}26{]}) is the source of
+The 2018 Community of Madrid travel survey {[}38{]} is the source of
 data for this empirical example: it is a representative survey that
-reflects a snap-shot of the travel patterns for one typical day of the
-working week (e.g., n=222,744 trips with representative population
-elevation factors). In this paper, a sample of the travel survey is
-used, namely the residential home origin to work destination trips of
-all modes and those that originate and end in the city of Madrid. These
-totals are displayed in Figure \ref{fig:Fig2} and Figure \ref{fig:Fig3}.
-Both figures are displayed at the level of traffic analysis zones (\(i\)
-and \(j\)) that correspond to the survey. The red boundary represents
-the LEZ Centro in effect in 2017 and thus those travel patterns of
-car-restriction reflected in the survey. The cyan boundary represents
-the LEZ that will be within the boundaries of the M-30 highway in 2023
-and is present in the plots as a spatial reference for areas in
-proximity to the LEZ Centro.
+reflects a snap-shot of the travel patterns for a typical day of the
+working week in 2018. Specifically, it includes 222,744 trips taken from
+a representative sample of 85,064 households across the traffic analysis
+zones (TAZ) representing the Community of Madrid (population of
+6,507,184 over 3 years old) through population elevation factors.
+
+In this empirical example, we only use a sub-set of the survey,
+specifically home-to-full-time-work trips, by all modes, to demonstrate
+multimodal spatial availability of employment opportunities. The total
+population who travels to full-time work from an origin (i.e., the
+worker), and the destinations (i.e., the employment opportunity), for
+the City of Madrid (a sub-set of the Community of Madrid) is visualized
+in Figure \ref{fig:Fig2} and Figure \ref{fig:Fig3}. Both figures are
+displayed at the level of TAZ (the \(i\) and \(j\) zones) that
+correspond to the survey. The red boundary represents the LEZ Centro in
+effect in 2017 and thus those travel patterns of car-restriction
+reflected in the survey. The cyan boundary represents the LEZ that will
+be within the boundaries of the M-30 highway in 2023 and is present in
+the plots as a spatial reference for areas in proximity to the LEZ
+Centro.
 
@@
 modal categories \(P^m_i\), is reflected in Figure \ref{fig:Fig3}. The
-modal categories represented in Figure \ref{fig:Fig3} are summarized for
-the following trip mode types:
+four modal categories represented in Figure \ref{fig:Fig3} are created
+through the grouping of similar mode types as provided in the survey, by
+the authors discretion. The modal categories and the mode types within
+each category are reported as follows:
 
@@
   Car/motor: all cars and operating modes (e.g., cab, private driver,
-  company, rental care, main driver, passenger, etc.) and all public,
-  private or company motorcycle/mopeds.
+  company, rental car, main driver of a private car, passenger in a
+  private car) and all public, private or company motorcycle/mopeds.
 \item
@@
 \item
-  Walk: walking or by foot,
+  Walk: walking or by foot.
 \end{itemize}
@@
 using cars - particularly near and within LEZ Centro. This distribution
-could be a result of a variety of factors including: transit coverage
-and service within with city, effective car infrastructure outside of
-the M-30, and/or the impact of the Central LEZ itself.
-
-From Figure \ref{fig:Fig3}, it can also be seen that biking and walking
-trips are less common than motorized trips at 1\% and 15\% respectively.
-Noteably, there is a positive trend between the populations of walking
-and biking trips in zones and populations of transit trips. This
-positive trend is higher than for car trip populations.
+is likely caused by a variety of factors including: transit coverage and
+service within with city, effective car infrastructure outside of the
+M-30, and/or the impact of the LEZ Centro itself. From Figure
+\ref{fig:Fig3}, it can be observed that biking and walking trips are
+less common than motorized trips at 1\% and 15\% respectively. Noteably,
+there is a positive trend between the populations of walking and biking
+trips in zones and populations of transit trips. This positive trend is
+higher than for car trip populations.
 
@@
 travel impedance functions \(f^m(c_{ij}^m)\). To illustrate the modal
-differences in travel lengths, summary descriptive per mode are
+differences in travel time lengths, summary descriptive per mode are
 detailed:
@@
 \item
-  Car/motor: 36 min (min:0 min., Q2: 15 min., Q3: 55 min., max: 120
-  min.)
+  Car/motor: mean 36 minutes (min:0 minutes, Q2: 15 minutes, Q3: 55
+  minutes, max: 120 minutes)
 \item
-  Transit: 55 min. (min:1 min., Q2: 30 min., Q3: 80 min., max: 120 min.)
+  Transit: mean 55 minutes (min:1 minutes, Q2: 30 minutes, Q3: 80
+  minutes, max: 120 minutes)
 \item
-  Bike: 34 min. (min:5 min., Q2: 15 min., Q3: 40 min., max: 115 min.)
+  Bike: mean 34 minutes (min:5 minutes, Q2: 15 minutes, Q3: 40 minutes,
+  max: 115 minutes)
 \item
-  Walk: 27 min. (min:1 min., Q2: 10 min., Q3: 45 min., max: 119 min.)
+  Walk: mean 27 minutes (min:1 minutes, Q2: 10 minutes, Q3: 45 minutes,
+  max: 119 minutes)
 \end{itemize}
@@
 To calculate \(f^m(c_{ij}^m)\) from the survey travel times, a concept
-known as the trip length distribution (TLD) was used. A TLD represents
+known as the trip length distribution (TLD) is used. A TLD represents
 the proportion of trips that are taken at a specific travel cost such as
 travel time (i.e., probability density distribution of trips taken by
-travel cost). This distribution is then used to derive impedance
-functions (e.g., done in the accessibility works of {[}27{]}, {[}28{]},
-and {[}29{]}). Maximum likelihood estimation and the Nelder-Mead method
-for direct optimization available within the R \{fitdistrplus\} package
-{[}30{]} is used to fit the impedance functions. As shown as shown in
-Figure \ref{fig:Fig4}, based on goodness-of-fit criteria and associated
-diagnostics, the gamma and log-normal probability density function (line
-curves) are selected as best fitting curves for the motorized and
-non-motorized modes respectively. The selection of functional form
-aligns with empirical examples in other regions {[}14,31,32{]}. Overall,
-the plots in Figure \ref{fig:Fig4} display the probability of travel
-given a trip travel time, based on trip flows from the survey. These
-`probability of travel' at each travel time for each mode are realized
-observations that reflect the land-use, the transport system, and the
-population travel behaviour in Madrid.
+travel cost). This distribution is then used in the derivation of travel
+impedance functions (e.g., done in the accessibility works of {[}39{]},
+{[}40{]}, and {[}41{]}). To fit the impedance functions, we use the
+Maximum likelihood estimation and the Nelder-Mead method for direct
+optimization available within the R \{fitdistrplus\} package {[}42{]}.
+Based on goodness-of-fit criteria and associated diagnostics, the gamma
+and log-normal probability density functions are selected as best
+fitting curves for the motorized and non-motorized modes respectively.
+The selection of functional form aligns with empirical examples in other
+regions {[}14,43,44{]}. The shape and rate parameters for the gamma
+functions (motorized modes) are 1.8651852 and 0.051468for car/motor and
+2.7566235 and 0.0499193 for transit; for the log-normal functions
+(non-motorized modes), mean and standard deviation parameters are
+3.2372212 and 0.7575986 for bike and 2.9918042 and 0.7575986 for walk.
+
+Represented in Figure \ref{fig:Fig4} are four plots visualizing the
+estimated probability density functions (represented as black lines)
+over top the binned travel time data from the 2018 survey. The estimated
+probability density functions can be interpreted as the probability of
+travel (y-axis) given a trip travel time (x-axis), based on trip flows
+from the 2018 travel survey. These `probability of travel' at each
+travel time for each mode are realized observations that reflect the
+land-use, the transport system, and the population travel behaviour in
+Madrid (according to the 2018 survey). A trip that leaves \(i\) is
+assigned a probability of travel value (y-axis) based on the mode and
+travel time of that trip to work: this probability value is the input
+\(f^m(c_{ij}^m)\).
+
+Notably, plots in Figure \ref{fig:Fig4} can be observed across modes if
+curious to know what trips are more likely to occur using what mode
+depending on trip length (as informed by the observed 2018 survey).
+Trips \textgreater5 minutes do not occur frequently for any mode, as
+such trips with short trip lengths are assigned a lower travel impedance
+\(f^m(c_{ij}^m)\) value. One reason that \textgreater5 minute trips do
+not frequently occur as a result of land-use (residential and job
+mismatch): not everyone lives in immediate proximity to their workplace.
+In terms of the non-motorized modes, shorter trips occurred more
+frequently overall for walking populations, particularly around 15
+minute lengths, so a trip of approximately 15 minutes is assigned the
+highest \(f^m(c_{ij}^m)\). For biking populations, longer travel times
+are more common so though the highest \(f^m(c_{ij}^m)\) value also
+corresponds to approximately 15 minutes, the curve is more spread out
+and values decrease less rapidly at longer travel times than for the
+walk mode curve. A similar trend can be observed for the motorized modal
+options where transit mode is more spread out than car/motor mode. All
+in all, these observations demonstrate that, based on a given mode and
+travel time, a trip is more or less likely to occur and is accordingly
+represented in the cost of travel balancing factor \(F_{ij}^m\) for
+\(V_i^m\).
 
@@
 \subsection{Results}\label{results}}
+
+As it is worth re-iterating, the empirical example reflects a snap-shot
+of the calculated multimodal spatial availability using data from the
+2018 travel survey. It is intended to visualize the spatial trends in
+availability of employment opportunities, by mode, and demonstrate how
+spatial availability can be interpreted to discuss the competitive
+advantage of lower travel impedance modes within Madrid Centro.
 
@@
 \caption{\label{fig:Fig5} Spatial availability of job opportunities per origin and mode $V_i^m$ in Madrid.
-Calculated using the home-to-work origin-destination flows from the 2018 travel survey.
-}\label{fig:SA-m-plot}
+Calculated using the home-to-work flows from the 2018 travel survey.
+Gray TAZs have no population.}\label{fig:SA-m-plot}
 \end{figure}
 
-\(V_i^m\) is a proportion of the total number of the 847,574 jobs in the
+\(V_i^m\) is a proportion of the total number of 847,574 jobs in the
 region and is visualized in Figure \ref{fig:Fig5}. Since \(V_i^m\) is
@@
 implementation further restricts the car-advantage as it shifted more
-than half of all car trips into the LEZ to another mode {[}25{]}. This
+than half of all car trips into the LEZ to another mode {[}37{]}. This
 restriction decreased the number of car-using population from \(i\)s
@@
 \caption{\label{fig:Fig7} Spatial availability of job opportunities per mode-using capita by mode $v_i^m$ per origin in Madrid.
-Calculated using the home-to-work origin destination flows from the 2018 travel survey.}\label{fig:SA-per-capita-m-plot}
+Calculated using the home-to-work flows from the 2018 travel survey.
+Gray TAZs have no population.}\label{fig:SA-per-capita-m-plot}
 \end{figure}
@@
 are difficult for planners to operationalize for a variety of reasons
-including issues of computation and interpretability {[}2{]}. With
+including issues of computation and interpretability {[}5{]}. With
 spatial availability, the magnitude of opportunities that are available
@@
 level may be too much. These interpretations were difficult to be made
-with accessibility measures in the past.
+with accessibility measures in the past. In future work, we intend to
+use multimodal spatial availability to characterise the equity of
+multimodal accessibility losses and gains in varying LEZ policy scenario
+implementations.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-geursAccessibility2004}{}}%
+\CSLLeftMargin{2. }%
+\CSLRightInline{Geurs KT, Wee B van. Accessibility evaluation of
+land-use and transport strategies: Review and research directions.
+Journal of Transport Geography. 2004;12: 127--140. Available:
+\href{http://www.sciencedirect.com/science/article/B6VG8-4B28VY7-1/2/61478339c14cab4fa58438ad7b1f4610\%0AC:/Papers/Journal\%20of\%20Transport\%20Geography/Journal\%20of\%20Transport\%20Geography\%20(2004)\%2012\%20(2)\%20127-140.pdf}{http://www.sciencedirect.com/science/article/B6VG8-4B28VY7-1/2/61478339c14cab4fa58438ad7b1f4610
+C:/Papers/Journal of Transport Geography/Journal of Transport Geography
+(2004) 12 (2) 127-140.pdf}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-paezPositive2012}{}}%
+\CSLLeftMargin{3. }%
+\CSLRightInline{Paez A, Scott DM, Morency C. Measuring accessibility:
+Positive and normative implementations of various accessibility
+indicators. Journal of Transport Geography. 2012;25: 141--153.
+doi:\href{https://doi.org/10.1016/j.jtrangeo.2012.03.016}{10.1016/j.jtrangeo.2012.03.016}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-handyAccessibility2020}{}}%
+\CSLLeftMargin{4. }%
+\CSLRightInline{Handy S. Is accessibility an idea whose time has finally
+come? Transportation Research Part D: Transport and Environment.
+2020;83: 102319. doi:\url{https://doi.org/10.1016/j.trd.2020.102319}}
+
 \leavevmode\vadjust pre{\hypertarget{ref-levinsonTransportAccessManual2020}{}}%
-\CSLLeftMargin{2. }%
+\CSLLeftMargin{5. }%
 \CSLRightInline{Levinson D, King D. Transport access manual: {A} guide
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-gowerPlanningInnovationCity2022}{}}%
-\CSLLeftMargin{3. }%
-\CSLRightInline{Gower A, Grodach C. Planning innovation or city
-branding? Exploring how cities operationalise the 20-minute
-neighbourhood concept. Urban Policy and Research. 2022;40: 36--52.
-doi:\href{https://doi.org/10.1080/08111146.2021.2019701}{10.1080/08111146.2021.2019701}}
-
 \leavevmode\vadjust pre{\hypertarget{ref-siddiqToolsTradeAssessing2021}{}}%
-\CSLLeftMargin{4. }%
+\CSLLeftMargin{6. }%
 \CSLRightInline{Siddiq F, D. Taylor B. Tools of the trade?: Assessing
@@
 \leavevmode\vadjust pre{\hypertarget{ref-yanAccessibilityBasedPlanningAddressing2021}{}}%
-\CSLLeftMargin{5. }%
+\CSLLeftMargin{7. }%
 \CSLRightInline{Yan X. Toward accessibility-based planning addressing
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-fikselSustainabilityResilienceSystems2006}{}}%
-\CSLLeftMargin{6. }%
-\CSLRightInline{Fiksel J. Sustainability and resilience: Toward a
-systems approach. Sustainability: Science, Practice and Policy. 2006;2:
-14--21.
-doi:\href{https://doi.org/10.1080/15487733.2006.11907980}{10.1080/15487733.2006.11907980}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-valencaMainChallengesOpportunities2021}{}}%
-\CSLLeftMargin{7. }%
-\CSLRightInline{Valença G, Moura F, Morais De Sá A. Main challenges and
-opportunities to dynamic road space allocation: From static to dynamic
-urban designs. Journal of Urban Mobility. 2021;1: 100008.
-doi:\href{https://doi.org/10.1016/j.urbmob.2021.100008}{10.1016/j.urbmob.2021.100008}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-leeMeasuringImpactsNew2018}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-elGeneidyMaking2022}{}}%
 \CSLLeftMargin{8. }%
-\CSLRightInline{Lee J, Miller HJ. Measuring the impacts of new public
-transit services on space-time accessibility: An analysis of transit
-system redesign and new bus rapid transit in columbus, ohio, {USA}.
-Applied Geography. 2018;93: 47--63.
-doi:\href{https://doi.org/10.1016/j.apgeog.2018.02.012}{10.1016/j.apgeog.2018.02.012}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-mohriClusteringMethodMeasuring2021}{}}%
+\CSLRightInline{El-Geneidy A, Levinson D. Making accessibility work in
+practice. Transport Reviews. 2022;42: 129--133.
+doi:\href{https://doi.org/10.1080/01441647.2021.1975954}{10.1080/01441647.2021.1975954}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-handyMeasuring1997}{}}%
 \CSLLeftMargin{9. }%
-\CSLRightInline{Mohri S, Mortazavi S, Nassir N. A clustering method for
-measuring accessibility and equity in public transportation service:
-{Case} study of {Melbourne}. SUSTAINABLE CITIES AND SOCIETY. 2021;74.
-doi:\href{https://doi.org/10.1016/j.scs.2021.103241}{10.1016/j.scs.2021.103241}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-maoMeasuringSpatialAccessibility2013}{}}%
+\CSLRightInline{Handy S, Niemeier D. Measuring accessibility: An
+exploration of issues and alternatives. Environment and Planning A.
+1997;29: 1175--1194. }
+
+\leavevmode\vadjust pre{\hypertarget{ref-kwanSpace1998}{}}%
 \CSLLeftMargin{10. }%
-\CSLRightInline{Mao L, Nekorchuk D. Measuring spatial accessibility to
-healthcare for populations with multiple transportation modes. Health \&
-Place. 2013;24: 115--122.
-doi:\href{https://doi.org/10.1016/j.healthplace.2013.08.008}{10.1016/j.healthplace.2013.08.008}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-tarrinoortizPublicAcceptabilityLow2021}{}}%
+\CSLRightInline{Kwan MP. Space-time and integral measures of individual
+accessibility: A comparative analysis using a point-based framework.
+Geographical Analysis. 1998;30: 191--216. Available:
+\href{https://ISI:000074579200001\%0AC:/Papers/Geographical\%20Analysis/Geographical\%20Analysis\%20(1998)\%2030\%20(3)\%20191-216.pdf}{ISI:000074579200001
+C:/Papers/Geographical Analysis/Geographical Analysis (1998) 30 (3)
+191-216.pdf}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-shenLocationCharacteristicsInnercity1998}{}}%
 \CSLLeftMargin{11. }%
-\CSLRightInline{Tarriño-Ortiz J, Soria-Lara JA, Gómez J, Vassallo JM.
-Public acceptability of low emission zones: The case of {``madrid
-central.''} Sustainability. 2021;13: 3251.
-doi:\href{https://doi.org/10.3390/su13063251}{10.3390/su13063251}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-devrijNooneVisitsMe2022}{}}%
+\CSLRightInline{Shen Q. Location characteristics of inner-city
+neighborhoods and employment accessibility of low-wage workers. Environ
+Plann B. 1998;25: 345--365.
+doi:\href{https://doi.org/10.1068/b250345}{10.1068/b250345}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-luoMeasuresSpatialAccessibility2003}{}}%
 \CSLLeftMargin{12. }%
-\CSLRightInline{De Vrij E, Vanoutrive T. {``No-one visits me anymore''}:
-Low emission zones and social exclusion via sustainable transport
-policy. 2022 {[}cited 27 Jul 2023{]}. Available:
-\url{https://www.tandfonline.com/doi/epdf/10.1080/1523908X.2021.2022465?needAccess=true\&role=button}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-verbeekJustManagementUrban2022}{}}%
+\CSLRightInline{Luo W, Wang F. Measures of spatial accessibility to
+health care in a {GIS} environment: Synthesis and a case study in the
+chicago region. Environ Plann B Plann Des. 2003;30: 865--884.
+doi:\href{https://doi.org/10.1068/b29120}{10.1068/b29120}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-paezDemand2019}{}}%
 \CSLLeftMargin{13. }%
-\CSLRightInline{Verbeek T, Hincks S. The {``just''} management of urban
-air pollution? A geospatial analysis of low emission zones in brussels
-and london. Applied Geography. 2022;140: 102642.
-doi:\href{https://doi.org/10.1016/j.apgeog.2022.102642}{10.1016/j.apgeog.2022.102642}}
+\CSLRightInline{Paez A, Higgins CD, Vivona SF. Demand and level of
+service inflation in floating catchment area (FCA) methods. PloS one.
+2019;14: e0218773.
+doi:\href{https://doi.org/10.1371/journal.pone.0218773}{10.1371/journal.pone.0218773}}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-paezHealth2010}{}}%
+\CSLLeftMargin{15. }%
+\CSLRightInline{Páez A, Mercado R, Farber S, Morency C, Roorda M.
+Accessibility to health care facilities in montreal island: An
+application of relative accessibility indicators from the perspective of
+senior and non-senior residents. International Journal of Health
+Geographics. 2010;9: 1--9. Available:
+\url{http://www.ij-healthgeographics.com/content/9/1/52}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-moniruzzamanMode2013}{}}%
+\CSLLeftMargin{16. }%
+\CSLRightInline{Moniruzzaman M, Páez A, Nurul Habib KM, Morency C. Mode
+use and trip length of seniors in montreal. Journal of Transport
+Geography. 2013;30: 89--99.
+doi:\url{http://dx.doi.org/10.1016/j.jtrangeo.2013.03.007}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-reyesWalking2014}{}}%
+\CSLLeftMargin{17. }%
+\CSLRightInline{Reyes M, Paez A, Morency C. Walking accessibility to
+urban parks by children: A case study of montreal. Landscape and Urban
+Planning. 2014;125: 38--47.
+doi:\href{https://doi.org/10.1016/j.landurbplan.2014.02.002}{10.1016/j.landurbplan.2014.02.002}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-paezJobs2013}{}}%
+\CSLLeftMargin{18. }%
+\CSLRightInline{Páez A, Farber S, Mercado R, Roorda M, Morency C. Jobs
+and the single parent: An analysis of accessibility to employment in
+toronto. Urban Geography. 2013;34: 815--842.
+doi:\href{https://doi.org/10.1080/02723638.2013.778600}{10.1080/02723638.2013.778600}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-wuUnifyingAccess2020}{}}%
+\CSLLeftMargin{19. }%
+\CSLRightInline{Wu H, Levinson D. Unifying access. Transportation
+Research Part D: Transport and Environment. 2020;83: 102355.
+doi:\href{https://doi.org/10.1016/j.trd.2020.102355}{10.1016/j.trd.2020.102355}}
+
 \leavevmode\vadjust pre{\hypertarget{ref-tahmasbiMultimodalAccessibilitybasedEquity2019}{}}%
-\CSLLeftMargin{15. }%
+\CSLLeftMargin{20. }%
 \CSLRightInline{Tahmasbi B, Mansourianfar MH, Haghshenas H, Kim I.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-luoMeasuresSpatialAccessibility2003}{}}%
-\CSLLeftMargin{16. }%
-\CSLRightInline{Luo W, Wang F. Measures of spatial accessibility to
-health care in a {GIS} environment: Synthesis and a case study in the
-chicago region. Environ Plann B Plann Des. 2003;30: 865--884.
-doi:\href{https://doi.org/10.1068/b29120}{10.1068/b29120}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-shenLocationCharacteristicsInnercity1998}{}}%
-\CSLLeftMargin{17. }%
-\CSLRightInline{Shen Q. Location characteristics of inner-city
-neighborhoods and employment accessibility of low-wage workers. Environ
-Plann B. 1998;25: 345--365.
-doi:\href{https://doi.org/10.1068/b250345}{10.1068/b250345}}
+\leavevmode\vadjust pre{\hypertarget{ref-campbell2019accessibility}{}}%
+\CSLLeftMargin{21. }%
+\CSLRightInline{Campbell KB, Rising JA, Klopp JM, Mbilo JM.
+Accessibility across transport modes and residential developments in
+nairobi. Journal of Transport Geography. 2019;74: 77--90. }
+
+\leavevmode\vadjust pre{\hypertarget{ref-maharjanSpatialEquityModal2022}{}}%
+\CSLLeftMargin{22. }%
+\CSLRightInline{Maharjan S, Tilahun N, Ermagun A. Spatial equity of
+modal access gap to multiple destination types across chicago. Journal
+of Transport Geography. 2022;104: 103437.
+doi:\href{https://doi.org/10.1016/j.jtrangeo.2022.103437}{10.1016/j.jtrangeo.2022.103437}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-lunkeModalAccessibilityDisparities2022}{}}%
+\CSLLeftMargin{23. }%
+\CSLRightInline{Lunke EB. Modal accessibility disparities and transport
+poverty in the oslo region. Transportation Research Part D: Transport
+and Environment. 2022;103: 103171.
+doi:\href{https://doi.org/10.1016/j.trd.2022.103171}{10.1016/j.trd.2022.103171}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-grengsJobAccessibilityModal2010}{}}%
+\CSLLeftMargin{24. }%
+\CSLRightInline{Grengs J. Job accessibility and the modal mismatch in
+detroit. Journal of Transport Geography. 2010;18: 42--54.
+doi:\href{https://doi.org/10.1016/j.jtrangeo.2009.01.012}{10.1016/j.jtrangeo.2009.01.012}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-kawabataJobAccessibilityIndicator2006}{}}%
+\CSLLeftMargin{25. }%
+\CSLRightInline{Kawabata M, Shen Q. Job accessibility as an indicator of
+auto-oriented urban structure: A comparison of boston and los angeles
+with tokyo. Environ Plann B Plann Des. 2006;33: 115--130.
+doi:\href{https://doi.org/10.1068/b31144}{10.1068/b31144}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-kwokUseModalAccessibility2004}{}}%
+\CSLLeftMargin{26. }%
+\CSLRightInline{Kwok RCW, Yeh AGO. The use of modal accessibility gap as
+an indicator for sustainable transport development. Environ Plan A.
+2004;36: 921--936.
+doi:\href{https://doi.org/10.1068/a3673}{10.1068/a3673}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-morrisAccessibilityIndicatorsTransport1979}{}}%
+\CSLLeftMargin{27. }%
+\CSLRightInline{Morris JM, Dumble PL, Wigan MR. Accessibility indicators
+for transport planning. Transportation Research Part A: General.
+1979;13: 91--109.
+doi:\href{https://doi.org/10.1016/0191-2607(79)90012-8}{10.1016/0191-2607(79)90012-8}}
 
 \leavevmode\vadjust pre{\hypertarget{ref-weibullAxiomaticApproachMeasurement1976}{}}%
-\CSLLeftMargin{18. }%
+\CSLLeftMargin{28. }%
 \CSLRightInline{Weibull JW. An axiomatic approach to the measurement of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-josephMeasuringPotentialPhysical1982}{}}%
-\CSLLeftMargin{19. }%
+\CSLLeftMargin{29. }%
 \CSLRightInline{Joseph AE, Bantock PR. Measuring potential physical
@@
 \leavevmode\vadjust pre{\hypertarget{ref-taoInvestigatingImpactsPublic2020a}{}}%
-\CSLLeftMargin{20. }%
+\CSLLeftMargin{30. }%
 \CSLRightInline{Tao Z, Zhou J, Lin X, Chao H, Li G. Investigating the
@@
 \leavevmode\vadjust pre{\hypertarget{ref-carpentieriMultimodalAccessibilityPrimary2020}{}}%
-\CSLLeftMargin{21. }%
+\CSLLeftMargin{31. }%
 \CSLRightInline{Carpentieri G, Guida C, Masoumi HE. Multimodal
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-margaryanLowEmissionZones2021}{}}%
+\CSLLeftMargin{32. }%
+\CSLRightInline{Margaryan S. Low emission zones and population health.
+Journal of Health Economics. 2021;76: 102402.
+doi:\href{https://doi.org/10.1016/j.jhealeco.2020.102402}{10.1016/j.jhealeco.2020.102402}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-verbeekJustManagementUrban2022}{}}%
+\CSLLeftMargin{33. }%
+\CSLRightInline{Verbeek T, Hincks S. The {``just''} management of urban
+air pollution? A geospatial analysis of low emission zones in brussels
+and london. Applied Geography. 2022;140: 102642.
+doi:\href{https://doi.org/10.1016/j.apgeog.2022.102642}{10.1016/j.apgeog.2022.102642}}
+
 \leavevmode\vadjust pre{\hypertarget{ref-espanaPlanNacionalIntegrado2020}{}}%
-\CSLLeftMargin{22. }%
+\CSLLeftMargin{34. }%
 \CSLRightInline{España. Plan nacional integrado de energía y clima
@@
 \leavevmode\vadjust pre{\hypertarget{ref-espanaResolucion10Enero2020}{}}%
-\CSLLeftMargin{23. }%
+\CSLLeftMargin{35. }%
 \CSLRightInline{España. Resolución de 10 de enero de 2020, de la
@@
 \leavevmode\vadjust pre{\hypertarget{ref-barcelonaGUIATECNICAPARA2021}{}}%
-\CSLLeftMargin{24. }%
+\CSLLeftMargin{36. }%
 \CSLRightInline{Barcelona. {GUÍA} {TÉCNICA} {PARA} {LA} {IMPLEMENTACIÓN}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-tarrinoortizAnalyzingImpactLow2022}{}}%
-\CSLLeftMargin{25. }%
+\CSLLeftMargin{37. }%
 \CSLRightInline{Tarriño-Ortiz J, Gómez J, Soria-Lara JA, Vassallo JM.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-comunidaddemadridResultadosEDM20182020}{}}%
-\CSLLeftMargin{26. }%
+\CSLLeftMargin{38. }%
 \CSLRightInline{Comunidad de Madrid. Resultados de la {EDM} 2018 - Datos
@@
 \leavevmode\vadjust pre{\hypertarget{ref-lopez_2017_spatial}{}}%
-\CSLLeftMargin{27. }%
+\CSLLeftMargin{39. }%
 \CSLRightInline{Lopez FA, Paez A. Spatial clustering of high-tech
@@
 \leavevmode\vadjust pre{\hypertarget{ref-horbachov_theoretical_2018}{}}%
-\CSLLeftMargin{28. }%
+\CSLLeftMargin{40. }%
 \CSLRightInline{Horbachov P, Svichynskyi S. Theoretical substantiation
@@
 \leavevmode\vadjust pre{\hypertarget{ref-batista_estimation_2019}{}}%
-\CSLLeftMargin{29. }%
+\CSLLeftMargin{41. }%
 \CSLRightInline{Batista SFA, Leclercq L, Geroliminis N. Estimation of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-fitdistrplus_2015}{}}%
-\CSLLeftMargin{30. }%
+\CSLLeftMargin{42. }%
 \CSLRightInline{Delignette-Muller ML, Dutang C. {fitdistrplus}: An {R}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-reggianiAccessibilityImpedanceForms2011}{}}%
-\CSLLeftMargin{31. }%
+\CSLLeftMargin{43. }%
 \CSLRightInline{Reggiani A, Bucci P, Russo G. Accessibility and
@@
 \leavevmode\vadjust pre{\hypertarget{ref-soukhovTTS2016RDataSet2023}{}}%
-\CSLLeftMargin{32. }%
+\CSLLeftMargin{44. }%
 \CSLRightInline{Soukhov A, Páez A. {TTS}2016R: A data set to study

### 2023-11-07 Journal-manuscript/Journal-manuscript.Rmd
Further edits, close to 80% done

@@
 The paper rest of this paper is organized as follows.
-In Section we use a synthetic examples with multiple modes to illustrate some relevant issues; this helps us motivate the derivation of the new spatial availability expressions.
+In Section we discuss the use a synthetic examples with multiple modes to illustrate some relevant issues; this helps us motivate the derivation of the new spatial availability expressions.
 This is followed in Section 3 by an empirical example using data from the city of Madrid after the implementation of its Low Emission Zones.
 Data for this example comes from the city's 2018 travel survey.
-The example shows the differences in spatial availability within and outside the LEZ for travellers using different modes, namely car, transit, cycling and walking.
+The example shows the differences in spatial availability within and outside the LEZ for travelers using different modes, namely car, transit, cycling and walking.
 In Section 4, we provide concluding remarks on the strengths of the use of spatial availability as a multimodal accessibility measure, and discuss potential future uses in policy planning scenarios as well as directions for future research.
@@
 
-# A review of multimodal accessibility measures
+# A brief review of multimodal accessibility
 
@@
 Given an origin ($i$) and a destination ($j$), an impedance function $f^{m}(c^m_{ij})$ converts the cost of travel (e.g., time, money, generalized cost) into a score that represents the propensity for interaction.
-These measures originate from that proposed by @hansenHowAccessibilityShapes1959, which can take the following form in the multimodal case: $S_i^m = \sum_j O_j f^m(c_{ij}^m)$ where $m$ is a set of modes which have mode-specific travel costs ($(c_{ij}^m)$) and/or travel impedance functions ($f^m(\cdot)$).
-
-The Hansen-type measure is not constrained, and as a result does not consider the opportunities as finite.
-As an example, the work of @tahmasbiMultimodalAccessibilitybasedEquity2019 uses the Hansen-type measure to measure the potential interaction with retail locations using walking, public transit, and car modes as $m$.
+These measures originate from that proposed by @hansenHowAccessibilityShapes1959, which can take the following form in the multimodal case: $S_i^m = \sum_j O_j f^m(c_{ij}^m)$ where $m$ is a set of modes which have mode-specific travel costs ($c_{ij}^m$) and/or travel impedance functions ($f^m(\cdot)$).
+
+Hansen-type accessibility is not constrained, which is to say it does not consider the opportunities as finite.
+To cite an example, Tahmasbi et al.
+[@tahmasbiMultimodalAccessibilitybasedEquity2019] use Hansen-type accessibility to assess the potential interaction with retail locations by three modes: walking, public transit, and car (i.e., $m = w, p, c$).
 $S_i^m$ is the sum of retail locations $j$ that can potentially be reached under the travel impedance as calculated for each $i$ and $m$.
-In other words, each $i$ has three $S_i^m$ values, one per $m$.
-In this work, they demonstrate that the car mode has the highest $S_i^{m}$ values in the majority of $i$, i.e., populations using a car can potentially reach more retail opportunities than populations using other modes.
-However, higher $S_i^{m}$ values for car do not affect the values of $S_i^{m}$ for other modes: in effect, each modes is analyzed as if the others did not exist, which would be reasonable for an inexhaustible opportunity, but is more questionable when the opportunity is subject to congestion.
-Since the measure is measure is also not constrained, and each opportunity can be and often is counted multiple times within and between modes, the sum of accessibility is not a meaningful quantity.
-To improve interpretability, $S_i^m$ values are presented as standardized accessibility index in this work, from the lowest value of 0 to the highest value of 1.
-This presentation or discussion of Hansen-type accessibility measures is common in the literature [e.g., in @campbell2019accessibility; and @maharjanSpatialEquityModal2022].
-
-However, this representation of results can make the interpretation of results challenging _because_ values are region-relative and lacking numeric meaning.
-Zones will always have values from between 0 to 1 represented, but is a zone with a low value for pedestrian-modes and a high value for car-modes notable?
-And if notable, what does the difference in these standardized values tangibly mean for planners?
+In other words, for each origin $i$ three accessibility scores are calculated.
+In this work, Tahmasbi et al.
+[@tahmasbiMultimodalAccessibilitybasedEquity2019] show that car travel affords the highest $S_i^{m}$ values in the majority of $i$, i.e., travelers who use a car can potentially reach more retail opportunities than populations using other modes.
+However, higher $S_i^{m}$ values for car do not affect the values of $S_i^{m}$ for other modes: in effect, each mode is analyzed as if the others did not exist.
+Since the measure is not constrained, each opportunity is typically counted multiple times within and between modes, and as a result the sum of accessibility is not necessarily a meaningful quantity.
+The accessibility scores for the modes are often values that are difficult to interpret beyond making statements about relative size.
+For example, Lunke [@lunkeModalAccessibilityDisparities2022], researching the region of Oslo, reports accessibility scores for car in the order of tens of thousands of employment opportunities.
+The corresponding scores for transit are lower, but still often in the thousands or tens of thousands.
+As reported, the ratio of the transit to the car score can be lower than 0.2 (meaning transit gives access to less than 20% of the opportunities than car).
+But despite the discussion about "sufficient accessibility" (e.g., in the abstract), it is unclear what the unconstrained scores mean: is having access to but 10,000 jobs by transit insufficient?
+After all, 10,000 employment opportunities are still plenty of opportunities.
+These ratios can be found elsewhere in the literature [e.g., @paezRelative2010, Figs.
+7, 8, and 9; @paezHealth2010, Fig.
+5; @paezJobs2013, Figs.
+6, 7, and 8], and they are useful to assess when some members of the public are better or worse off than others, but they do not say much about how bad is "worse".
+
+Besides ratios of accessibility, another way to improve interpretability of scores sometimes seen in the literature is to standardize them to lie in the range [0-1].
+This adjustment is only helpful insofar as it facilitates relative comparisons, but interpretation of the scores remains challenging because the values are specific to a region and convey no meaning about the magnitude of the scores.
+In this approach, zones always have values between 0 and 1, but how remarkable is a zone with a low score for pedestrians and a high value for car?
+And if remarkable, what does the difference in these standardized values mean for planners?
 By how much should transport systems and land-use configurations be changed to improve conditions?
-And in what way can these existing methods be used to track progress over time?
+And in what way can these scores be used to track differences over time?
 Or between regions?
-This is challenging to track, since certain values will always be relatively 'low' and 'high'.
-Alternatively, modal accessibility calculated using the Hansen-type measure is also presented without standardization, as in the work by @lunkeModalAccessibilityDisparities2022.
-Often, the meaning of the accessibility value itself is not interpreted, as it is difficult to state what having access to 30,000 employment opportunities by car but having 10,000 employment opportunity by transit tangibly signifies.
-
-However, once opportunities in a region are interpreted as finite, a new meaning is taken on.
-As considered in the long tradition of accessibility work, capacity of opportunities is limited and thus is subject to competition by population [@soukhovIntroducingSpatialAvailability2023; @grengsJobAccessibilityModal2010; @kawabataJobAccessibilityIndicator2006; @kwokUseModalAccessibility2004; @morrisAccessibilityIndicatorsTransport1979; @weibullAxiomaticApproachMeasurement1976; @shenLocationCharacteristicsInnercity1998].
-There are only so many school-seats, hospital capacity, employment opportunities, etc., in a region and if one person interacts with an opportunity at a given time, it is taken: the supply of an opportunity and the demand for that opportunity are two components of accessibility.
-The competition consideration is especially important in the case of multiple modes.
-For instance, people in a zone who are advantaged with a relatively low travel-cost mode have the ability to potentially interact with more opportunities than other people.
-Due to this advantage, through the perspective of finite opportunities, there are fewer opportunities left to be potentially interacted for everyone else, especially zones with populations using higher travel-cost modes.
-This recognition is the motivation behind integrating _competition_ for opportunities within modal accessibility measures.
-
-Arguably, one of the most popular competitive location-based accessibility measures is the two-step floating catchment area (2SFCA) approach popularized by @luoMeasuresSpatialAccessibility2003 who simplified the approach proposed by @shenLocationCharacteristicsInnercity1998 (with similar considerations for competition in @weibullAxiomaticApproachMeasurement1976 and @josephMeasuringPotentialPhysical1982).
-The Shen-type accessibility measure's formulation is: $a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}$ where $D_j^m$ is the potential demand for opportunities equal to travel impedance weighted population $\sum_i P_i^m f^m(c_{ij}^m)$ and the remaining variables are repeated in the Hansen-type measure.
-The Shen-type measure per mode ($a_i^m$) can be understood as a ratio of the travel impedance-weighted supply of opportunities for the $m$-using population in $i$ over the ravel impedance-weighted demand for opportunities.
-In this way, it considers competition but the measure remains _non-constrained_ meaning resulting values are not associated with global maximums.
-In other words, it is difficult to interpret the meaning of differences in Shen-type accessibility scores between modes in similar ways as reflected for Hansen-type measures.
-
-To illustrate, @taoInvestigatingImpactsPublic2020a calculates $a_i^m$ to jobs for different income-group populations in Shenzhen, China using $m = \text{public transit}$ and $m={car}$.
-They demonstrate that $i$ with low-income populations have lower $a_i^m$ than $i$ with higher-income populations.
-Further, they demonstrate that $a_i^{m=\text{public transit}}$ is lower than $a_i^{m=\text{car}}$ at many $i$, arguing that this may put $i$ with lower-income populations in a further disadvantage.
-$a_i$ and/or $a_i^m$ are used to compare relative spatial differences in overall competitive accessibility and multimodal competitive accessibility, but because there is no global maximum, making it is difficult to interpret the significance between differences in $a_i^{m}$ values.
-Questions such as: what is the impact that competition has on the difference in $a_i^m$ values?
-How does impact vary spatially?
+These questions lack straightforward answers since certain values will always be relatively 'low' or 'high', but do not track to a quantity that can be intuitively understood.
+Presentation or discussion of Hansen-type accessibility that has been standardized in this way is not uncommon in the literature [e.g., @campbell2019accessibility; @maharjanSpatialEquityModal2022].
+
+Once we understand opportunities to be finite, it is possible for an accessibility measure to take a crisper meaning.
+As considered in the long tradition of accessibility research, capacity of opportunities is limited and thus is subject to competition by population [@soukhovIntroducingSpatialAvailability2023; @grengsJobAccessibilityModal2010; @kawabataJobAccessibilityIndicator2006; @kwokUseModalAccessibility2004; @morrisAccessibilityIndicatorsTransport1979; @weibullAxiomaticApproachMeasurement1976; @shenLocationCharacteristicsInnercity1998].
+There are only so many school-seats, hospital capacity, employment opportunities, etc., in a region and if one person reaches an opportunity, it is taken: the supply of an opportunity and the demand for that opportunity are two components of accessibility.
+These are clear examples of opportunities that are clearly competitive.
+But we would go as far as to argue that every type of opportunity is subject to congestion or capacity constraint, even when the opportunities are conventionally seen as non-competitive.
+Amenities are a good example of this.
+For instance, standards for providing green spaces are often stated in the form of _exclusive access_, in units of amenity per capita.
+A case in point is Natural England, an organization that recommends an Accessible Natural Greenspace Standard such that the minimum supply of space is one ha of statutory Local Nature Reserves per thousand population^[see  https://redfrogforum.org/wp-content/uploads/2019/11/67-Nature-Nearby%E2%80%99-Accessible-Natural-Greenspace-Guidance.pdf].
+Similarly, the World Health Organization [cited in @OECDFrameworks2013] recommends that cities provide a minimum of 9$m^2$ of green area per inhabitant.
+For our purposes, standards of this type translate into "how much of a resource can be offered to each individual that is not allocated to anyone else?".
+Green spaces often have large capacities, but they still have a capacity.
+Constraining accessibility is in this way a useful way to evaluate the congested availability of any type of opportunity.
+As development of sound standards is emphasized in the planning literature, in particular in regards to fairness in transportation [see @martensFair2021], spatial availability analysis is a useful way to develop and assess standards.
+
+The relevance of the considerations above is put in sharper relief when we think about the use of multiple modes (or heterogeneous populations).
+If we return to Oslo for a moment [@lunkeModalAccessibilityDisparities2022], we notice that the places that have high accessibility by transit are also the places that have _very high_ accessibility by car (Figure 2).
+Those two populations are going for the same opportunities, and those travelling by transit have fewer to choose from to begin with.
+More generally, people in a zone who are advantaged with relatively cost of travel will have the ability to potentially reach more opportunities than other people.
+Due to this advantage, through the perspective of finite opportunities, there are fewer opportunities left for everyone else, especially for those who use modes that are slower or more expensive.
+
+As noted in the introduction, competitive accessibility was the rationale for developing floating catchment area methods (FCA), popularized by @luoMeasuresSpatialAccessibility2003 [although similar, and earlier, developments are found in @weibullAxiomaticApproachMeasurement1976; @josephMeasuringPotentialPhysical1982; @shenLocationCharacteristicsInnercity1998].
+Shen-type accessibility is formulated as: $a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}$ where $D_j^m$ is the potential demand for opportunities equal to travel impedance weighted population $\sum_i P_i^m f^m(c_{ij}^m)$ and the remaining variables are repeated in the Hansen-type measure.
+Shen-type modal accessibility ($a_i^m$) can be understood as a ratio of the travel impedance-weighted supply of opportunities for $m$-mode in $i$ over the travel impedance-weighted demand for opportunities.
+In this way, it considers competition.
+That said, the measure remains unconstrained, meaning both population _and_ opportunities are multiply counted [see @paezDemand2019].
+In other words, interpretation of differences in Shen-type accessibility scores between modes is fraught as it is for Hansen-type measures.
+
+To illustrate, @taoInvestigatingImpactsPublic2020a calculate $a_i^m$ to jobs for different income-group populations in Shenzhen, China using $m = \text{public transit}$ and $m=\text{car}$.
+Their results indicate that zones with low-income populations have lower $a_i^m$ than zones with higher-income populations.
+Further, they show that $a_i^{m=\text{public transit}}$ is lower than $a_i^{m=\text{car}}$ at many zones, arguing that this may further place those zones with lower-income populations at disadvantage.
+$a_i$ and/or $a_i^m$ are used to compare relative spatial differences in overall competitive accessibility and multimodal competitive accessibility, but because opportunities were doubly counted (entering the sums of both modes), this makes for uneasy interpretations of the differences in $a_i^{m}$ between modes.
+Questions that this approach leaves unaddressed include: what is the impact of competition on the difference in $a_i^m$ values?
+How does the impact vary spatially?
 And what is the interpretation of this difference?
-are left unaddressed.
-
-Spatial availability improves on previously discussed multimodal accessibility approaches using the Hansen-type measure and the Shen-type measure by taking into account _competition_ in the potential interaction with opportunities in a _constrained_ framework (e.g., finite opportunities).
-This is done by considering: 1) competition between mass effect (e.g., the advantage of sub-populations residing in relatively low population-density and high opportunity-proximate areas) and 2) competition between travel impedance (e.g., sub-populations with relatively low travel-impedance) through a proportional allocation mechanism.
-The following sub-section demonstrates how spatial availability compares to the Hansen-type and Shen-type measures through a synthetic example.
-
-## Multimodal spatial availability $V_i^m$
+
+Spatial availability improves on previously discussed accessibility approaches using the Hansen-type measure and the Shen-type measure by constraining the sum of opportunities, that is, by treating opportunities as finite.
+This is done by means of proportional allocation factors that follow well established principles of spatial interaction and the gravity model [see @wilsonFamilty1971].
+In Soukhov et al.
+[@soukhovIntroducingSpatialAvailability2023] these factors consider: 1) the mass effect (e.g., the size of populations at different origins); and 2) the cost of travel from different zones (e.g., some sub-populations face relatively higher or lower costs).
+The following section introduces the multimodal form of spatial availability.
+
+# Multimodal spatial availability $V_i^m$
 
@@
 $V_{i}$ is a value of how many opportunities are available to each location $i$ out of all the opportunities in the region.
-The general formulation of spatial availability $V_{i}$ is shown in Equation (\ref{eq:spatial-availability-general}):
+The general formulation of spatial availability $V_{i}$ is shown in Equation (\ref{eq:spatial-availability-general}) [see @soukhovIntroducingSpatialAvailability2023]:
 ```{=tex}
@@
 
-- $F^t_{ij}$ is a balancing factor that depends on the demand for opportunities $O_j$ and cost of movement in the system $f(c_{ij})$.
-- $V_i$ is the number of spatially available opportunities at $i$; the sum of $V_{i}$ is equivalent to the total sum of opportunities in the region (i.e., $\sum_j O_j = \sum_i V_i$)
-
-The spatial availability measure is introduced in @soukhovIntroducingSpatialAvailability2023.
-Spatial availability's unique feature is the balancing factor $F^t_{ij}$, a proportional allocation mechanisms, that ensures the $V_i$ calculated for each $i$ sums, across all $i$ in the region, to equal the total number of opportunities in the region.
-As such, spatial availability is a  _competitive_ and _constrained_ accessibility measure as $F^t_{ij}$ handles the number of opportunities in the region in a finite way (proportional allocation).
-$F^t_{ij}$ consists of two components: a population-based balancing factor $F^p_{i} = \frac{P_i}{\sum_i P_i}$ and an impedance-based balancing factor $F^c_{ij} = \frac{F^c_{ij}}{\sum_j F^c_{ij}}$ that, respectively, allocates opportunities to $i$ in proportion to the size of the population at $i$ (the mass effect) and the cost of reaching opportunities at $j$ (the impedance effect).
-
-$F^p_{i}$ and $ F^c_{ij}$ are calculated for each $i$ such that they both equal 1 when summed across all $i$ in the region (e.g., $\sum_i F^p_{i} = 1$ and $\sum_i F^c_{ij} = 1$).
-These balancing factors are combined multiplicatively to yield $F^t_{ij}$ which ensures that a proportion of the opportunities $O_j$ are allocated to each $i$ accordingly.
-In other words, assuming a finite number of opportunities in the region, $F^t_{ij}$ proportionally allocates $O_j$ to each $i$ such that the resulting $V_i$ value represents the number of opportunities _spatially available_ to the population at $i$.
+- $F^t_{ij}$ is a balancing factor that depends on the size of the populations at different locations that demand opportunities $O_j$, as well as the cost of movement in the system $f(c_{ij})$.
+
+- $V_i$ is the number of spatially available opportunities at $i$; the sum of $V_{i}$ is identical to the total number of opportunities in the region (i.e., $\sum_j O_j = \sum_i V_i$); in other words, opportunities are dealt with as finite resources.
+
+Compared to Hansen-type accessibility:
+```{=tex}
+\begin{equation}
+\label{eq:hansen-accessibility}
+A_i = \sum_{j=1}^J O_jf(c_{ij})
+\end{equation}
+```
+
+\noindent we see that spatial availability is, like Hansen's measure, a weighted sum of the opportunities.
+What makes spatial availability stand apart from other approaches is the weight used in the sum, balancing factor $F^t_{ij}$, implements a proportional allocation mechanisms to ensure that the sum of $V_i$ is constrained to match the total number of opportunities in the region.
+As such, spatial availability is singly-constrained and naturally implements competition or congestion.
+$F^t_{ij}$ consists of two parts.
+The first part is a population-based proportional allocation factor to model the mass effect of the gravity model:
+$$
+F^p_{i} = \frac{P_i}{\sum_i P_i}
+$$
+
+This factor makes opportunities available based on demand. Secondly, there is an impedance-based proportional allocation factor that model the cost effect:
+$$
+F^c_{ij} = \frac{F^c_{ij}}{\sum_j F^c_{ij}}
+$$
+
+This factor makes opportunities available preferentially to those who can reach them at a lower cost.
+$F^p_{i}$ and $F^c_{ij}$ are designed so that they both equal 1 when summed across all $i$ in the region (e.g., $\sum_i F^p_{i} = 1$ and $\sum_i F^c_{ij} = 1$).
+These factors are combined multiplicatively to yield $F^t_{ij}$ which ensures that a proportion of the opportunities $O_j$ are allocated to each $i$ accordingly.
+In other words, assuming a finite number of opportunities in the region, $F^t_{ij}$ proportionally allocates $O_j$ to each $i$ such that the resulting $V_i$ value represents the number of opportunities _available_ to the population at $i$.
 Each zonal value is a proportion of the opportunities in the region (i.e., $\sum_j O_j = \sum_i V_i$).
 
-The focus of this paper is to extend $V_i$ for the measurement of multimodal accessibility applications.
-To do so, the balancing factors are reformulated to yield a proportional value for the set of modes $m$ used by populations at each $i$.
-As these factors are proportional, $F^{pm}_{i}$ and $F^{cm}_{ij}$ can be summed across each $m$ at each $i$ and then across all $i$ to equal to 1.
-They are also similarly combined multiplicatively to obtain their joint effect, represented as the combined balancing factor $F^{tm}_{ij}$ detailed in Equation (\ref{eq:multimodal-balancing-factors}).
-
+The focus of this paper is to extend $V_i$ for the measurement of multimodal applications (or more generally heterogeneous populations).
+To do so, the balancing factors need to be reformulated so that 1) the mass effect now accounts not only for the size of the population at $i$, but also the size of sub-populations within $i$; and 2) the cost of travel is not only for different zones, but by sub-populations within each zone (e.g., the cost of travel from $i$ by car, transit, walking, etc.) When we introduce modes (or sub-populations) $m$, the proportional allocation factors need to satisfy the condition that $F^{pm}_{i}$ and $F^{cm}_{ij}$ can be summed across each $m$ at each $i$ and then across all $i$ to equal to 1.
+They are also similarly combined multiplicatively to obtain their joint effect, represented as the combined balancing factor $F^{tm}_{ij}$ similar to that detailed in Equation (\ref{eq:multimodal-balancing-factors}).
+This factor is given by:
 ```{=tex}
@@
 
-\noindent Where:
-
-- The population balancing factor for each $m$ at each $i$ is $F^{pm}_{i} = \frac{P_{i}^m}{\sum_{m}\sum_{i} P_{i}^m}$
-- The cost of travel balancing factor for each $m$ at $i$ is $F_{ij}^{cm} = \frac{f^m(c_{ij}^m)}{\sum_{m} \sum_{i} f^m(c_{ij}^m)}$
-
-Implementing $F^{tm}_{ij}$, the following Equation (\ref{eq:spatial-availability-multimodal}) demonstrates the multimodal configuration of spatial availability $V_i^m$:
-
+\noindent where:
+
+- The factor for allocation by population for each $m$ at each $i$ is $F^{pm}_{i} = \frac{P_{i}^m}{\sum_{m}\sum_{i} P_{i}^m}$; and
+- The factor for allocation by cost of travel for each $m$ at $i$ is $F_{ij}^{cm} = \frac{f^m(c_{ij}^m)}{\sum_{m} \sum_{i} f^m(c_{ij}^m)}$
+
+Implementing $F^{tm}_{ij}$, the following Equation (\ref{eq:spatial-availability-multimodal}) gives the multimodal version of spatial availability $V_i^m$:
 ```{=tex}
@@
 
--   $m$ is a set of modes used by populations in the region.
+-   $m = 1, 2,\cdots, M$ is a set of $M$ modes (or sub-populations) of interest.
 -   $F^{tm}_{ij}$ is a balancing factor $F^t_{ij}$ for each $m$ at each $i$.
--   $V^m_{i}$ is the spatial availability $V_{i}$ for mode $m$ at each $i$; the sum of $V^m_{i}$ for all $m$ at each $i$ is equivalent to the total sum of opportunities in the region (i.e., $\sum_j O_j = \sum_i V_i = \sum_{m} \sum_{i} V^m_{i}$)
-
-## An illustrative synthetic example
-Consider the following Figure \ref{fig:Fig1}, it depicts a region with population and jobs at three population centers ($A$, $B$, $C$) and three employment centers ($1$, $2$, $3$).
-The population at each population center is divided into two sub-populations, one using a faster mode $z$ and another using a slower mode $x$, to travel to employment centers.
+-   $V^m_{i}$ is the spatial availability $V_{i}$ for mode $m$ at each $i$; the sum of $V^m_{i}$ for all $m$ at each $i$ is equivalent to the total sum of opportunities in the region (i.e., $\sum_j O_j = \sum_i V_i = \sum_{m} \sum_{i} V^m_{i}$).
+
+Next we use a synthetic example to contrast multimodal accessibility and spatial availability.
+
+# An illustrative synthetic example
+
+Consider the simple system shown in Figure \ref{fig:Fig1}.
+The figure shows a region with population at three population centers ($A$, $B$, $C$) and jobs at three employment centers ($1$, $2$, $3$).
+The population at each origin $i$ is consists of two sub-populations, one using a faster mode $z$ and another using a slower mode $x$, to travel to employment centers.
 Population center $A$ is Suburban: it is closest to its own relatively large employment center at $1$, close to the Urban's equally large employment center $2$, and has a population that is smaller than the Urban $B$ and larger than the Satellite $C$.
 $B$ has the largest $x$-using population, followed by then $A$, then $C$.
-This synthetic example was inspired by the single-mode example used in @shenLocationCharacteristicsInnercity1998 and reconfigured in @soukhovIntroducingSpatialAvailability2023.
+This synthetic example is inspired by the single-mode example used in @shenLocationCharacteristicsInnercity1998 and reconfigured in @soukhovIntroducingSpatialAvailability2023.
 
@@
 
-From the perspective of access to a _finite_ amount of opportunities in the region ($210,000$ jobs), the sub-population that is most proximate to jobs, furthest from densely populated centers, and is using the lowest travel-cost mode $z$ can potentially access the most job opportunities.
+From the perspective of access to a _finite_ amount of opportunities in the region ($210,000$ jobs), the sub-population that is most proximate to jobs (low cost to reach), furthest from large populations (less competition), and uses the fastest mode $z$ (greater range) can potentially reach the largest number of opportunities.
 This appears to be the sub-population at $A$ using mode $z$.
-Sub-populations located in opposite conditions (i.e., further away from jobs, close to dense populations, and using $x$) are at a relative job opportunity access _disadvantage_.
-The competition for opportunities between different mode-using populations matters from the perspective of inequities as it reflects how well the land-use and transport system serves (or doesn't serve) certain populations.
+Sub-populations located in opposite conditions (i.e., more distant from jobs, close to large populations, and using slow mode $x$) are at a relative disadvantage.
+The competition for opportunities between different mode-using populations matters as it reflects how well the land-use and transport system serves (or does not serve) certain populations.
 
@@
 
-The calculated $S_i^m$, $a_i^m$ and $V_i^m$ accessibility values for each $i$ and $m$ are shown in the middle three columns and are aggregated for each $i$ in the final two columns in Table 1 <!-- \ref{tab:tab1}-->.
-We use a negative exponential impedance function $f^m(c_{ij}^m) = \exp(-\beta\cdot c_{ij})$ with $\beta=0.1$ for both $x$ and $z$ modes for all accessibility measures calculations.
-
-The Hansen-type measure $S_i^m$ is presented for each origin and mode in third column of Table 1 <!-- \ref{tab:tab1}-->.
-For all $i$, the $z$-using sub-population has higher $S_i^m$ values than the $x$-using sub-populations.
-Additionally, $S_i^m$ is equal for both mode-using populations in $A$ and $B$.
-This is the case since $S_i^m$ does not consider _competition_, it only relies on reflecting the count of opportunities that may be interacted with as a product of $f^m(c_{ij}^m)$.
-Recall, populations in $A$ and $B$ have the same travel impedance to employment centers $1$, $2$ and $3$ (either 15, 30, or 100 minutes using $x$ or 10, 25, or 80 minutes using $z$).
-As such, these the calculated $S_i^m$ values are the same for both $A$ and $B$.
+The values calculated for $S_i^m$ (Hansen-type accessibility), $a_i^m$ (Shen-type accessibility), and $V_i^m$ (spatial availability) for each $i$ and $m$ are shown in the middle three columns and are aggregated for each $i$ in the final two columns in Table 1 <!-- \ref{tab:tab1}-->.
+As in the example in Shen [@shenLocationCharacteristicsInnercity1998], we use a negative exponential impedance function $f^m(c_{ij}^m) = \exp(-\beta\cdot c_{ij})$ with $\beta=0.1$ for both $x$ and $z$ modes for all accessibility measures calculations.
+Notice that in this example we use the same impedance function but the travel times are different for the two modes.
+More generally, it is possible to use different impedance functions for the modes, as demonstrated in the empirical example below.
+
+Hansen-type accessibility $S_i^m$ is presented for each origin and mode in the third column of Table 1 <!-- \ref{tab:tab1}-->.
+For all $i$, the travel by $z$ results in higher values of $S_i^m$ than travel by $x$.
+Lack of competition, or alternatively the assumption of an inexhaustible resource in the calculation of $S_i^m$, lead to a curious result.
+Since the populations in $A$ and $B$ have the same travel impedance to employment centers $1$, $2$ and $3$ (either 15, 30, or 100 minutes using $x$ or 10, 25, or 80 minutes using $z$), their values of $S_i^m$ are the same for both $A$ and $B$.
 Furthermore, the total sum of $S_i^m$ in the region is equal to `r S_im$S_im %>% sum() %>% prettyNum(big.mark = ",")`.
-This value is difficult to interpret: it represents the weighted sum of opportunities that may be interacted with within the region based on travel impedance.
-It cannot be interpreted as any sort of benchmark since the measure is _non-constrained_.
-To connect this example to literature, $S_i^m$ is calculated in the work of @tahmasbiMultimodalAccessibilitybasedEquity2019; they compare differences in $S_i^m$ values between modes in a relative and comparative sense, but make no further interpretation of the $S_i^m$ values.
-
-In the fourth and sixth column in Table 1 <!-- \ref{tab:tab1}--> the Shen-type measure is calculated: first for both origin and mode $a_i^m$ as well as aggregated by the weighted mean mode-population ( $\sum_m \frac{P_i^m}{P_i}*a_i^m$ ) to represent a value for each origin $a_i$.
-Unlike $S_i^m$, this measure considers _competition_.
-For instance, the $x$-using populations in $A$ and $B$ centers do not have the same $a_i^m$ values as the $z$-using.
-In fact, $A$ has the highest values $a_i^m$ and $a_i$ values since this center has the smallest travel impedance to opportunities (lower than at $C$, $A$ and $B$ are equal) and has one of these lowest proximity to a relatively high amount of population (lower than at $B$).
-
-However, the Shen-type measure is _non-constrained_: the total sum of $a_i^m$ or $a_i$ is practically meaningless since it represents a sum of ratios.
-For instance, the $z$-using sub-population at $A$ has a value of `r a_i %>% filter(i=="A" & m=="z") %>% select(a_im) %>% pull() %>% round(digits=2)` potential jobs per potential job-seeking population compared to `r a_i %>% filter(i=="A" & m=="x") %>% select(a_im) %>% pull() %>% round(digits=2)` for $x$-using sub-population.
-What is the significance of these values?
+This value lacks an intuitive interpretation: it represents the weighted sum of opportunities that may be reached within the region according to the travel impedance (i.e., the travel behavior and the characteristics of the modes) and does not usefully translate into any sort of benchmark.
+To connect this example to the literature, $S_i^m$ is calculated in the work of @tahmasbiMultimodalAccessibilitybasedEquity2019; they compare differences in $S_i^m$ values between modes in a relative and comparative sense, but make no further interpretation of the $S_i^m$ values.
+Large metropolitan regions will have large values and small regions small, but that is simply an artifact of size.
+
+In the fourth and sixth columns in Table 1 <!-- \ref{tab:tab1}--> the results for Shen-type accessibility are reported: first for both origin and mode $a_i^m$ as well as aggregated by the weighted mean mode-population ( $\sum_m \frac{P_i^m}{P_i}*a_i^m$ ) to represent a value for each origin $a_i$.
+Unlike $S_i^m$, this measure does considers _competition_.
+For instance, the population travelling by $x$ from $A$ and $B$ do not have the same values of $a_i^m$ as those travelling by $z$.
+In fact, $A$ has the highest values $a_i^m$ and $a_i$ values since this center has the lowest travel impedance to opportunities (lower than at $C$, $A$ and $B$ are equal) and has faces relatively low competition, not being close to a relatively large population (lower than at $B$).
+
+However, the calculations of $a_i^m$ are not constrained: the total sum of $a_i^m$ or $a_i$ is practically meaningless since it represents a sum of ratios.
+For instance, the population travelling by $z$ from $A$ has a value of `r a_i %>% filter(i=="A" & m=="z") %>% select(a_im) %>% pull() %>% round(digits=2)` jobs per job-seeking population compared to `r a_i %>% filter(i=="A" & m=="x") %>% select(a_im) %>% pull() %>% round(digits=2)` for users of mode $x$.
+What is the meaning of these values?
 The difference between these modes is equal to `r ((a_i %>% filter(i=="A" & m=="z") %>% select(a_im)) - (a_i %>% filter(i=="A" & m=="x") %>% select(a_im))) %>% pull() %>% round(digits=2)`, but `r ((a_i %>% filter(i=="A" & m=="z") %>% select(a_im)) - (a_i %>% filter(i=="A" & m=="x") %>% select(a_im))) %>% pull() %>% round(digits=2)` of what?
-How many more job opportunities are $z$ users interacting with than $x$ users?
+How many more job opportunities can users of $z$ reach compared to user of $x$?
 When $a_i^m$ is aggregated to $a_i$ as shown in the sixth column, the values face similar interpretability issues.
-The Shen-type measure is implemented in the previously discussed work of @taoInvestigatingImpactsPublic2020a to calculate modal $a_i^m$ values and the aggregated $a_i$ is implemented in the work of @carpentieriMultimodalAccessibilityPrimary2020.
-However, similar to the Hansen-type measure, these works discuss relative and spatially comparative differences in values, they do not make further interpretation of the $a_i^m$ or $a_i$ themselves.
-This may be because the Shen-type measure is _non-constrained_, this is no benchmark or global maximum to which comparisons can be drawn from.
-
-By contrast, spatial availability $V_i$ considers competition and is constrained such that the total sum of values is equal to the total number of opportunities in the region (i.e., $210,000$ jobs).
-Seen in fifth column of Table 1 <!-- \ref{tab:tab1}-->, $V_i^m$ for the same mode-using populations in $A$ and $B$ are not the same (as this measure considers competition).
-In fact, at $A$, the $z$-using sub-population captures `r ((table_toy_format %>% filter(i == "A" & m == "z") %>% select("V_im")) - (table_toy_format %>% filter(i == "A" & m == "x") %>% select("V_im"))) %>% pull() %>% prettyNum(big.mark = ",")` more spatially available jobs (of the $210,000$ jobs in the region) than the sub-population using mode $x$.
-The numerical difference has a practical interpretation.
-
-Furthermore, $V_i^m$ values for an $i$ can be aggregated across $m$ and compared across $i$ ( $V_i = \sum_m{\sum_i{V_i^m}}$ ) as a result of the proportional allocation mechanism.
+The Shen-type measure is implemented in aforementioned work of Tao et al.
+[@taoInvestigatingImpactsPublic2020a] to calculate modal $a_i^m$ values and the aggregated $a_i$ is implemented in the work of [@carpentieriMultimodalAccessibilityPrimary2020].
+However, similar to Hansen-type accessibility, these works discuss relative and spatially comparative differences in values, but veer from interpretating the values of $a_i^m$ or $a_i$ themselves.
+In fairness, interpretation is complicated by the multiple counting of opportunities between zones and modes.
+
+In contrast, spatial availability $V_i$ considers competition and is constrained such that the total sum of values is equal to the total number of opportunities in the region (i.e., $210,000$ jobs).
+Seen in fifth column of Table 1 <!-- \ref{tab:tab1}-->, the values of $V_i^m$ in $A$ and $B$ are not the same within mode (as this measure considers competition).
+In fact, at $A$, users of mode $z$ capture `r ((table_toy_format %>% filter(i == "A" & m == "z") %>% select("V_im")) - (table_toy_format %>% filter(i == "A" & m == "x") %>% select("V_im"))) %>% pull() %>% prettyNum(big.mark = ",")` more spatially available jobs (of the $210,000$ jobs in the region) than the sub-population travelling by $x$.
+The numerical difference is clear since it refers to opportunities out of the total.
+
+Furthermore, the proportional allocation mechanism also means that the values of $V_i^m$ for any origin $i$ can be aggregated across $m$ and compared between zones ($V_i = \sum_m{\sum_i{V_i^m}}$).
 This aggregation, $V_i$, is shown in the seventh column in Table 1 <!-- \ref{tab:tab1}-->.
 Again looking at center $A$, $A$ is allocated `r ((V_i %>% filter(i == "A" & m == "z")  %>% select("V_i"))) %>% pull() %>% prettyNum(big.mark = ",")` spatially available opportunities for both modes.
-`r ((V_i %>% filter(i == "A" & m == "z") %>% select("V_im")) / ((V_i %>% filter(i == "A" & m == "z") %>% select("V_i")))) %>% pull() %>% as.numeric() %>% percent() ` of this spatial availability allocated to $A$ is assigned to the $z$-using population despite representing `r 0.66 %>% percent()` of $A$'s population.
+`r ((V_i %>% filter(i == "A" & m == "z") %>% select("V_im")) / ((V_i %>% filter(i == "A" & m == "z") %>% select("V_i")))) %>% pull() %>% as.numeric() %>% percent() ` of this spatial availability allocated to $A$ is assigned to users of mode $z$ despite representing `r 0.66 %>% percent()` of $A$'s population.
 
@@
 Across the entire region, `r V_ijm %>% filter(m=="z") %>% group_by(i) %>% summarize(pop = first(pop)) %>% select(pop) %>% sum() %>% prettyNum(big.mark = ",")` people use $z$ (`r ((V_ijm %>% filter(m=="z") %>% group_by(i) %>% summarize(pop = first(pop)) %>% select(pop) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% select("pop") %>% sum())) %>% percent()` of the region population).
-However, the $z$-using population accounts for `r ((V_ijm %>% filter(m=="z") %>% select(V_ij) %>% sum()) / (V_ijm %>% select(V_ij) %>% sum())) %>% percent()` of the region's total spatial availability - the rest is allocated to the $x$-using population (`r ((V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% filter(m == "x") %>% dplyr::select(pop) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% dplyr::select(pop) %>% sum())) %>% percent()`of the total population).
-Notably, the $x$-using population captures `r (((V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% filter(m == "x") %>% dplyr::select(pop) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% dplyr::select(pop) %>% sum())) - ((V_ijm %>% filter(m=="x") %>% select(V_ij) %>% sum()) / (V_ijm %>% select(V_ij) %>% sum()))) %>% percent()` less spatial availability to opportunities than its population proportion.
-This understanding can lead us to ask normative questions such as, how unequal should opportunity access for the two mode-using populations be?
-Can the lower-travel-cost populations spare some spatial availability if a policy of modal-restriction (like a LEZ) was introduced?
-
-Since spatial availability is constrained and has an interpretable meaning as a proportion of the total opportunities in the region, the values at $i$ have a new significance.
+However, users of $z$ account for `r ((V_ijm %>% filter(m=="z") %>% select(V_ij) %>% sum()) / (V_ijm %>% select(V_ij) %>% sum())) %>% percent()` of the region's total spatial availability - while the remaining `r (1 - (V_ijm %>% filter(m=="z") %>% select(V_ij) %>% sum()) / (V_ijm %>% select(V_ij) %>% sum())) %>% percent()` is allocated to users of mode $x$ who are `r ((V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% filter(m == "x") %>% dplyr::select(pop) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% dplyr::select(pop) %>% sum())) %>% percent()`of the total population.
+Notably, the population who uses $x$ have `r (((V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% filter(m == "x") %>% dplyr::select(pop) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% dplyr::select(pop) %>% sum())) - ((V_ijm %>% filter(m=="x") %>% select(V_ij) %>% sum()) / (V_ijm %>% select(V_ij) %>% sum()))) %>% percent()` fewer spatially available opportunities than its share in the population.
+This realization leads us to ask normative questions such as, how unequal should availability of opportunities be by mode?
+What intervention could help to redistribute spatial availability to sub-populations commensurate with their proportion of the total?
+
+Since spatial availability is constrained and has an interpretable meaning as a proportion of the total opportunities in the region, the values at $i$ have a straighforward interpretation.
 Inequality in $V_i^m$ values can be explored through a variety of approaches.
 For instance, consider travel times.
-The $z$-using population accounts for `r ((V_ijm %>% group_by(i,m) %>% summarize(V_tt = sum(V_ij*tt), .groups = "drop") %>% filter(m == "z") %>% dplyr::select(V_tt) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(V_tt = sum(V_ij*tt), .groups = "drop") %>% dplyr::select(V_tt) %>% sum())) %>% percent()` of the potential travel time traveled in the region: this is `r (((V_ijm %>% filter(m=="z") %>% select(V_ij) %>% sum()) / (V_ijm %>% select(V_ij) %>% sum())) - (V_ijm %>% group_by(i,m) %>% summarize(V_tt = sum(V_ij*tt), .groups = "drop") %>% filter(m == "z") %>% dplyr::select(V_tt) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(V_tt = sum(V_ij*tt), .groups = "drop") %>% dplyr::select(V_tt) %>% sum())) %>% percent()` less travel time than the proportion of spatial available opportunities that is allocated to them.
-In other words, the $z$-using population travels less minutes overall and has more spatial availability of opportunities than the $x$-using population using the slower mode $x$.
-
-Alternatively, inequities in spatial availability between mode-using populations can be explored through proportional benchmarks.
-A spatial availability per capita $v_i^m$ as presented in Equation (\ref{eq:SA-per-capita}):
+The population of travelers who use $z$ accounts for `r ((V_ijm %>% group_by(i,m) %>% summarize(V_tt = sum(V_ij*tt), .groups = "drop") %>% filter(m == "z") %>% dplyr::select(V_tt) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(V_tt = sum(V_ij*tt), .groups = "drop") %>% dplyr::select(V_tt) %>% sum())) %>% percent()` of the potential travel time traveled in the region: this is `r (((V_ijm %>% filter(m=="z") %>% select(V_ij) %>% sum()) / (V_ijm %>% select(V_ij) %>% sum())) - (V_ijm %>% group_by(i,m) %>% summarize(V_tt = sum(V_ij*tt), .groups = "drop") %>% filter(m == "z") %>% dplyr::select(V_tt) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(V_tt = sum(V_ij*tt), .groups = "drop") %>% dplyr::select(V_tt) %>% sum())) %>% percent()` less travel time than the proportion of spatial available opportunities that is allocated to them.
+In other words, the population of users of $z$ travels fewer minutes overall and has more spatial availability of opportunities than users of the slower mode $x$.
+
+Alternatively, inequities in spatial availability between modes can be explored through proportional benchmarks.
+A spatial availability per capita $v_i^m$ is presented in Equation (\ref{eq:SA-per-capita}):
 
@@
 
-The $v_i^m$ values for $A$, $B$, and $C$ for the $x$-using sub-populations are `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="A" & m =="x") %>% select(v_i) %>% sum() %>% round(digits=2)`, `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="B" & m =="x") %>% select(v_i) %>% sum() %>% round(digits=2)` and `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="C" & m =="x") %>% select(v_i) %>% sum() %>% round(digits=2)` spatially available jobs per capita, respectively.
-The $v_i^m$ for the $z$-using sub-populations are much higher, with values of: `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="A" & m =="z") %>% select(v_i) %>% sum() %>% round(digits=2)`, `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="B" & m =="z") %>% select(v_i) %>% sum() %>% round(digits=2)` and `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="C" & m =="z") %>% select(v_i) %>% sum() %>% round(digits=2)` respectively.
-The $x$-using population, especially at $B$ and $C$, are directly impacted by the jobs that are spatially available to the $z$-using population _in addition to_ the mass effect (occurring at $B$, high population density) and high travel impedance (occurring at the Satellite $C$).
-
-If, lets say, the planning goal is to have one spatially available job per mode-using population, a policy intervention can be put in place, to reduce the $v_i^z$ values and increase $v_i^x$ values.
-This demonstration is to show how simply the $V_i^m$ framework can be manipulated quantify the competitive (dis)advantage in a multimodal application.
-In what follows, we further explore competition between multiple modes through an empirical example.
+The values of $v_i^m$ for $A$, $B$, and $C$ for users of $x$ are `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="A" & m =="x") %>% select(v_i) %>% sum() %>% round(digits=2)`, `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="B" & m =="x") %>% select(v_i) %>% sum() %>% round(digits=2)` and `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="C" & m =="x") %>% select(v_i) %>% sum() %>% round(digits=2)` spatially available jobs per capita, respectively.
+The values of $v_i^m$ for users of $z$ are much higher, with values of: `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="A" & m =="z") %>% select(v_i) %>% sum() %>% round(digits=2)`, `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="B" & m =="z") %>% select(v_i) %>% sum() %>% round(digits=2)` and `r V_i %>% group_by(i,m) %>% summarize(v_i = sum(V_im)/sum(pop), .groups="drop") %>% filter(i=="C" & m =="z") %>% select(v_i) %>% sum() %>% round(digits=2)` respectively.
+Users of $x$, especially those at $B$ and $C$, are directly impacted by the jobs that are spatially available to users of $z$ _in addition to_ the mass effect (occurring at $B$, high population density) and high travel impedance (occurring at the Satellite $C$).
+
+If, let us say, the planning goal was to have one spatially available job per mode-using population, a policy intervention could be devised, to reduce the values of $v_i^z$ (making it slower or more expensive) and increase he values of $v_i^x$ (making it faster or les expensive).
+The purpose of this simple demonstration is to show how spatial availability can be used to quantify the competitive (dis)advantage in a multimodal application.
+In what follows, we demonstrate the use of multimodal spatial availability through an empirical example.
 
@@
 ## Data and methods
+
 ```{r}
@@
 <!-- a summary about low emission zones, they are a geographic accessibility restriction. they change the accessibility landscape-->
-Low emission zones (LEZ) have been implemented as a climate change policy intervention to reduce GHG emissions, improve air quality, and support sustainable mobility in many countries [@margaryanLowEmissionZones2021; @verbeekJustManagementUrban2022].
-Though rules of exclusion vary<!-- depending on legal aspects and cultural norms-->, LEZ aim to deter/reduce traffic in designated zones under threat of penalty (e.g., fines, seizure of vehicle).
-<!-- In practice, LEZ limits the volume of traffic by excluding vehicles by license plate, fuel type, or by introducing tolls.--> From the perspective of restriction for passenger transport, LEZ are a policy of _geographic discrimination_ as they change how people access opportunities by making the travel impedance more costly for specific car-mode users.
-When considering opportunities as finite in a region, this discrimination could allow certain populations to access opportunities by other modes more readily than before.
-In this way, LEZ change the multimodal competitive accessibility landscape of a city.
+
+The context for the empirical example is Madrid, in Spain.
+This city implemented a Low Emission Zone in 2017, to attend to goals set out in the national agenda to fight climate change, cut nitrogen dioxide levels, and prioritize people's movement in the city.
+Low emission zones (LEZ) elsewhere have been implemented as a climate change policy intervention to reduce GHG emissions, improve air quality, and support sustainable mobility [@margaryanLowEmissionZones2021; @verbeekJustManagementUrban2022].
+Though the rules of exclusion vary<!-- depending on legal aspects and cultural norms-->, LEZ aim to deter/reduce traffic in designated zones under threat of penalty (e.g., fines, seizure of vehicle).
+<!-- In practice, LEZ limits the volume of traffic by excluding vehicles by license plate, fuel type, or by introducing tolls.--> In other words, LEZ are a policy of _geographic discrimination_ as they change how people can reach opportunities by making the travel impedance more costly for specific forms of travel, typically cars.
+When considering opportunities as finite in a region, this discrimination reduces the competition of one mode and allows other modes to potentially thrive.
+In this way, LEZ aim to change the multimodal competitive accessibility landscape of a city.
 
@@
 
-In 2017, LEZs were implemented in the Spanish capital city of Madrid following the goals set out in the national agenda <!-- : to fight climate change, cut nitrogen dioxide levels for the benefit of people's health, and prioritize people's movement in the city-->.
-In geographic scope, the 2017 boundaries of the LEZ were relatively small (covering 4.72 km$^2$) and within the center (i.e., LEZ Centro).
-These boundaries were expanded in 2023 to inside of the M-30, a highway in proximity to the city center (i.e., LEZ M-30) and the city has plans to further spatially expand the LEZ.
-Within the 2017 LEZ Centro implementation, all cars, motorcycles and freight with environmental label A or B (higher polluting classification, associated with older make and model of fossil fuel internal combustion engine vehicles), are not permitted to enter the area unless they are used by residents or meet other exemptions.
-This restriction impacted approximately half of all car trips that were typically made into the LEZ Centro [@tarrinoortizAnalyzingImpactLow2022].
-
-For this case study, we use $V_i^m$ to quantify the competition of spatially available opportunities between modes after the LEZ Centro implementation.
-Particularly, we demonstrate how $V_i^m$ can be used to spectate on how the restriction of car mobility in areas around/within the LEZ Centro allowed the other, more sustainable but often with higher travel impedance modes, to become more competitive.
+In geographic scope, the 2017 boundaries of the LEZ in Madrid were relatively modest (covering 4.72 km$^2$) and only within the center (i.e., LEZ Centro).
+These boundaries were expanded in 2023 to the area inside the M-30, an orbital highway in proximity to the city center (i.e., LEZ M-30).
+Beyond this, the city has plans to spatially expand the LEZ.
+Within the 2017 LEZ Centro implementation, all cars, motorcycles and freight with environmental labels A or B (older makes and models of fossil fuel internal combustion engine vehicles), are not permitted to enter the area unless they are used by residents or meet other exemptions.
+This restriction impacted approximately half of all car trips that typically travelled into what is now the LEZ Centro [@tarrinoortizAnalyzingImpactLow2022].
+
+For this case study, we use spatial availability to quantify access to opportunities by different modes in Madrid.
+Particularly, we demonstrate how $V_i^m$ can be used to derive insights into how the restriction of car mobility in areas around/within the LEZ Centro may have allowed other, more sustainable but often slower or more costly modes, to become more competitive.
 
@@
 ```
-The 2018 Community of Madrid travel survey [@comunidaddemadridResultadosEDM20182020] is the source of data for this empirical example: it is a representative survey that reflects a snap-shot of the travel patterns for a typical day of the working week in 2018.
-Specifically, it includes 222,744 trips taken from a representative sample of 85,064 households across the traffic analysis zones (TAZ) representing the Community of Madrid (population of 6,507,184 over 3 years old) through population elevation factors.
-
-In this empirical example, we only use a sub-set of the survey, specifically home-to-full-time-work trips, by all modes, to demonstrate multimodal spatial availability of employment opportunities.
-The total population who travels to full-time work from an origin (i.e., the worker), and the destinations (i.e., the employment opportunity), for the City of Madrid (a sub-set of the Community of Madrid) is visualized in Figure \ref{fig:Fig2} and Figure \ref{fig:Fig3}.
-Both figures are displayed at the level of TAZ (the $i$ and $j$ zones) that correspond to the survey.
-The red boundary represents the LEZ Centro in effect in 2017 and thus those travel patterns of car-restriction reflected in the survey.
-The cyan boundary represents the LEZ that will be within the boundaries of the M-30 highway in 2023 and is present in the plots as a spatial reference for areas in proximity to the LEZ Centro.
-
-The total sum of jobs $O_j$ that are held are shown in Figure \ref{fig:Fig2} and the populations that go to a work destination by four modal categories $P^m_i$, is reflected in Figure \ref{fig:Fig3}.
-The four modal categories represented in Figure \ref{fig:Fig3} are created through the grouping of similar mode types as provided in the survey, by the authors discretion.
+
+The source of data for this example is the 2018 Community of Madrid travel survey [@comunidaddemadridResultadosEDM20182020].
+This is a representative survey that gives a snap-shot of the travel patterns for a typical day of the working week in 2018.
+Specifically, it captures 222,744 trips as measured from a representative sample of 85,064 households across the traffic analysis zones (TAZ) in the Community of Madrid (population of 6,507,184 over 3 years old).
+
+In this example, we use all home-to-full-time-work trips, by all modes.
+Figures \ref{fig:Fig2} and \ref{fig:Fig3} show the population of workers and the distribution of jobs in the City of Madrid.
+The zoning system is the trafric analysis zones used in the survey.
+The red boundary represents the LEZ Centro in effect in 2017.
+The survey already captures the travel patterns after the introduction of LEZ Centro.
+The cyan boundary represents the LEZ planned for for the boundaries of the M-30 highway and is present in the plots as a spatial reference for areas in proximity to the LEZ Centro.
+
+The total sum of jobs $O_j$ are shown in Figure \ref{fig:Fig2} and the populations that go to a work destination by four modal categories $P^m_i$, is displayed in Figure \ref{fig:Fig3}.
+The modal shares in Figure \ref{fig:Fig3} are calculated based on those measued in the survey.
 The modal categories and the mode types within each category are reported as follows:
@@
 
-From Figure \ref{fig:Fig2}, it can be seen that the largest concentration of jobs are within, near, and to the north of the LEZ Centro.
-The population that is accessing those jobs by mode (Figure \ref{fig:Fig3}), appear spatially distinct.
-Car and transit trips represent `r (pop_props$population_m[3]/sum(pop_props$population_m)) %>% percent()` and `r (pop_props$population_m[4]/sum(pop_props$population_m)) %>% percent()` of the modal share respectively.
-The population that travels using transit is more spatially distributed than those using cars - particularly near and within LEZ Centro.
+Some aggregation of modes is necessary to calculate the travel impedance functions by mode.
+From Figure \ref{fig:Fig2}, it can be seen that the largest concentration of jobs is within, near, and to the north of LEZ Centro.
+The populations with access to those jobs by mode (Figure \ref{fig:Fig3}) are spatially distinct.
+Travel by car and transit represent `r (pop_props$population_m[3]/sum(pop_props$population_m)) %>% percent()` and `r (pop_props$population_m[4]/sum(pop_props$population_m)) %>% percent()` of the modal share respectively.
+The population that travels by transit is more spatially distributed than those using cars - particularly near and within LEZ Centro.
 This distribution is likely caused by a variety of factors including: transit coverage and service within with city,  effective car infrastructure outside of the M-30, and/or the impact of the LEZ Centro itself.
-From Figure \ref{fig:Fig3}, it can be observed that biking and walking trips are less common than motorized trips at `r (pop_props$population_m[1]/sum(pop_props$population_m)) %>% percent()` and `r (pop_props$population_m[2]/sum(pop_props$population_m)) %>% percent()` respectively.
-Noteably, there is a positive trend between the populations of walking and biking trips in zones and populations of transit trips.
+From Figure \ref{fig:Fig3}, it can be seen that active travel is less common than motorized trips at `r (pop_props$population_m[1]/sum(pop_props$population_m)) %>% percent()` and `r (pop_props$population_m[2]/sum(pop_props$population_m)) %>% percent()` for cycling and walking respectively.
+Noticeably, there is a positive trend between the walking and cycling in zones where transit is also present.
 This positive trend is higher than for car trip populations.
@@
 
-The travel time for each trip is provided within the survey.
-These travel times, per modal category, are used to calibrate mode-specific travel impedance functions $f^m(c_{ij}^m)$.
-To illustrate the modal differences in travel time lengths, summary descriptive per mode are detailed:
+Travel times are provided within the survey by mode.
+This information is used to calibrate mode-specific travel impedance functions $f^m(c_{ij}^m)$.
+To illustrate the modal differences in travel times, the following descriptive statistics per mode are presented:
 
@@
 
-To calculate $f^m(c_{ij}^m)$ from the survey travel times, a concept known as the trip length distribution (TLD) is used.
-A TLD represents the proportion of trips that are taken at a specific travel cost such as travel time (i.e., probability density distribution of trips taken by travel cost).
-This distribution is then used in the derivation of travel impedance functions (e.g., done in the accessibility works of @lopez_2017_spatial, @horbachov_theoretical_2018, and @batista_estimation_2019).
+Impedance functions $f^m(c_{ij}^m)$ are calibrated from the travel times in the survey via the empirical trip length distribution (TLD).
+An empirical TLD is given by the proportion of trips at various travel cost bins.
+This distribution is then used to estimate the parameters of a function for the travel impedance [as done in @lopez_2017_spatial; @horbachov_theoretical_2018; @batista_estimation_2019].
 To fit the impedance functions, we use the Maximum likelihood estimation and the Nelder-Mead method for direct optimization available within the R {fitdistrplus} package [@fitdistrplus_2015].
 Based on goodness-of-fit criteria and associated diagnostics, the gamma and log-normal probability density functions are selected as best fitting curves for the motorized and non-motorized modes respectively.
-The selection of functional form aligns with empirical examples in other regions [@reggianiAccessibilityImpedanceForms2011; @soukhovTTS2016RDataSet2023; @soukhovIntroducingSpatialAvailability2023].
-The shape and rate parameters for the gamma functions (motorized modes) are `r HW_car_motor_gamma$estimate[1]` and `r HW_car_motor_gamma$estimate[2]`for car/motor and `r HW_transit_gamma$estimate[1]` and `r HW_transit_gamma$estimate[2]` for transit; for the log-normal functions (non-motorized modes), mean and standard deviation parameters are `r HW_bike_lnorm$estimate[1]` and `r HW_bike_lnorm$estimate[2]` for bike and `r HW_walk_lnorm$estimate[1]` and `r HW_bike_lnorm$estimate[2]` for walk.
-
-Represented in Figure \ref{fig:Fig4} are four plots visualizing the estimated probability density functions (represented as black lines) over top the binned travel time data from the 2018 survey.
-The estimated probability density functions can be interpreted as the probability of travel (y-axis) given a trip travel time (x-axis), based on trip flows from the 2018 travel survey.
-These 'probability of travel' at each travel time for each mode are realized observations that reflect the land-use, the transport system, and the population travel behaviour in Madrid (according to the 2018 survey).
-A trip that leaves $i$ is assigned a probability of travel value (y-axis) based on the mode and travel time of that trip to work: this probability value is the input $f^m(c_{ij}^m)$.
-
-Notably, plots in Figure \ref{fig:Fig4} can be observed across modes if curious to know what trips are more likely to occur using what mode depending on trip length (as informed by the observed 2018 survey).
-Trips >5 minutes do not occur frequently for any mode, as such trips with short trip lengths are assigned a lower travel impedance $f^m(c_{ij}^m)$ value.
-One reason that >5 minute trips do not frequently occur as a result of land-use (residential and job mismatch): not everyone lives in immediate proximity to their workplace.
-In terms of the non-motorized modes, shorter trips occurred more frequently overall for walking populations, particularly around 15 minute lengths, so a trip of approximately 15 minutes is assigned the highest $f^m(c_{ij}^m)$.
-For biking populations, longer travel times are more common so though the highest $f^m(c_{ij}^m)$ value also corresponds to approximately 15 minutes, the curve is more spread out and values decrease less rapidly at longer travel times than for the walk mode curve.
+The selection of functional forms aligns with empirical examples in other regions [@reggianiAccessibilityImpedanceForms2011; @soukhovTTS2016RDataSet2023; @soukhovIntroducingSpatialAvailability2023].
+The shape and rate parameters for the gamma functions (motorized modes) are `r HW_car_motor_gamma$estimate[1]` and `r HW_car_motor_gamma$estimate[2]`for car/motor and `r HW_transit_gamma$estimate[1]` and `r HW_transit_gamma$estimate[2]` for transit; for the log-normal functions (non-motorized modes), the mean and standard deviation parameters are `r HW_bike_lnorm$estimate[1]` and `r HW_bike_lnorm$estimate[2]` for bike and `r HW_walk_lnorm$estimate[1]` and `r HW_bike_lnorm$estimate[2]` for walk.
+
+Figure \ref{fig:Fig4} includes four plots to visualize the calibrated impedance functions (represented as black lines) superimposed on the empirical TLD.
+The impedance functions can be interpreted as the propensity to travel (y-axis) given a trip travel time (x-axis).
+The functions reflect a combination of possibilities and preferences: the travel behavior given the transportation technologies available.
+For example, trips shorter than 5 minutes do not occur frequently for any mode; this reflects the spatial separation between places of residence and places of work commonly seen in many cities.
+In terms of the non-motorized modes, there is a preference towards walking trips around 15 minutes in duration, as seen from the highest value of $f^{walk}(c_{ij}^{walk})$.
+With respect to travel by bicycle, longer travel times are more common; although the highest value of the impedance also corresponds to approximately 15 minutes, the curve has a longer tail and values decrease less rapidly at longer travel times than is the case of $f^{walk}(c_{ij}^{walk})$.
 A similar trend can be observed for the motorized modal options where transit mode is more spread out than car/motor mode.
-All in all, these observations demonstrate that, based on a given mode and travel time, a trip is more or less likely to occur and is accordingly represented in the cost of travel balancing factor $F_{ij}^m$ for  $V_i^m$.
+All in all, these functions represent the propensity of travel by mode by duration of trip, which are needed to calculate the proportional allocation factors $F_{ij}^m$ for $V_i^m$.
 
@@
 ## Results
-As it is worth re-iterating, the empirical example reflects a snap-shot of the calculated multimodal spatial availability using data from the 2018 travel survey.
+
+At this point it is worthwhile to reiterate that the empirical example is a snap-shot of the calculated multimodal spatial availability using data from the 2018 travel survey.
 It is intended to visualize the spatial trends in availability of employment opportunities, by mode, and demonstrate how spatial availability can be interpreted to discuss the competitive advantage of lower travel impedance modes within Madrid Centro.
 
-The spatial availability of jobs $V_i^m$ is calculated for each of the four modal categories $m$ at the level of traffic analysis zones $i$ in Madrid and demonstrated in Figure \ref{fig:Fig5}.
+The spatial availability of jobs $V_i^m$ is calculated for each of the four modes $m$ at the level of traffic analysis zones $i$ in Madrid and displayed in Figure \ref{fig:Fig5}.
 
@@
 
-$V_i^m$ is a proportion of the total number of `r SA_i_zn208$V_i %>% sum()  %>% prettyNum(big.mark = ",")` jobs in the region and is visualized in Figure \ref{fig:Fig5}.
-Since $V_i^m$ is calculated based on the likelihood of travel from observed home-to-work journeys, the values can be understood as the number of full-time jobs that are spatially available to the full-time working population at that $i$ and their associated $m$, relative to all the jobs in the city.
+$V_i^m$ is a proportion of the total number of `r SA_i_zn208$V_i %>% sum()  %>% prettyNum(big.mark = ",")` jobs in the region and is shown in Figure \ref{fig:Fig5}.
+Since $V_i^m$ is calculated based on the population of workers and the distribution of jobs, the values can be understood as the number of full-time jobs that are spatially available to the full-time working population at that $i$ and their associated $m$, relative to all the jobs in the city.
 $V_i^m$ is the number of jobs that are _spatially available_ to a $m$-using population located at $i$, relative to the travel impedance and size of _all_ populations in the region.
 
-Notable are the differences in the magnitude of $V_i^m$ between modes in Figure \ref{fig:Fig5}.
-The majority of $V_i^m$ is allocated to car- and transit- using populations.
-This is to be expected, as the population that commutes using these modes represents `r (modal_V_comps_4plot$perc[3] + modal_V_comps_4plot$perc[4]) %>% percent(accuracy=0.1)` of the total population.
-Differences in $V_i^m$ values within mode-using populations also exist: car-using populations outside of the M-30 region appear to have greater $V_i^m$ values, while some $i$ areas inside the M-30 appear to have higher $V_i^m$ values for the transit-using populations.
-Overall, the magnitude of $V_i^m$ values for the bikers and walkers are lower than car and transit but the highest $V_i^{bike}$ and $V_i^{walk}$ values tend to be allotted to $i$s within the M-30 and $i$s that have higher $V_i^{transit}$ values.
-
-The differences between the mode-using population and their mode-specific spatial availability highlights the competitive advantage offered to certain modes in certain spatial extents.
-As summarized in the left-most columns in Figure \ref{fig:Fig6}, the 'car/motor' and 'transit' populations represent a combined `r (modal_V_comps_4plot$perc[7] + modal_V_comps_4plot$perc[8]) %>% percent(accuracy=0.1)` of the total spatial availability in the city.
-However, the 'car/motor' using population is allocated disproportionately more $V_i^m$ than its size compared to the transit-using population.
-The car-using and transit-using population is `r modal_V_comps_4plot$perc[3] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[4] %>% percent(accuracy=0.1)` respectively, but is allocated `r modal_V_comps_4plot$perc[7] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[8] %>% percent(accuracy=0.1)` respectively, of the city's spatial availability.
-When treating the number of opportunities that can be reached as a finite value (total: 847,574 opportunities), fewer opportunities are spatial availability to the lesser competitive modes-using populations, in this case walking and cycling.
-These modes are less competitive as a result of: their lower travel impedance values at longer travel times (see Figure \ref{fig:Fig4} at travel times beyond ~30 minutes); their low population values values overall; and higher populations present in origins with high motorized mode commuting.
+Some notable differences in the magnitude of $V_i^m$ occur between modes as seen in Figure \ref{fig:Fig5}.
+The majority of $V_i^m$ (which is to say of spatially available jobs) is allocated to those travelling by car and transit.
+This is to be expected, as the users of these modes represent `r (modal_V_comps_4plot$perc[3] + modal_V_comps_4plot$perc[4]) %>% percent(accuracy=0.1)` of the total population.
+Differences in $V_i^m$ values within modes also exist in space: car users outside of the M-30 region appear to have greater $V_i^m$ values, while some zones inside the M-30 appear to have higher $V_i^m$ values for transit.
+Overall, the magnitude of $V_i^m$ values for cyclists and pedestrians are lower than for car and transit but the highest values of $V_i^{bike}$ and $V_i^{walk}$ tend to be allotted to zones within the M-30 and origins with higher spatial availability by transit.
+
+The differences between the shares of modes and their shares of spatially availabile opportunities highlights the competitive advantage afforded by certain modes, although this effect is not uniform in space.
+As seen in the left-most columns of Figure \ref{fig:Fig6}, 'car/motor' and 'transit' together can avail themselves of `r (modal_V_comps_4plot$perc[7] + modal_V_comps_4plot$perc[8]) %>% percent(accuracy=0.1)` of all jobs in the city.
+However, 'car/motor' has a disproportionate share of $V_i^m$ relative to the population of users of this mode, compared to opportunities that are spatially available to transit users.
+The combined population of car and transit users is `r modal_V_comps_4plot$perc[3] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[4] %>% percent(accuracy=0.1)` respectively, but these populations are allocated `r modal_V_comps_4plot$perc[7] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[8] %>% percent(accuracy=0.1)` respectively, of the city's jobs.
+When treating the number of opportunities that can be reached as a finite value (total: 847,574 opportunities), fewer opportunities are spatially available to slower modes (i.e., walking and cycling), even taking into account that their share is smaller overall.
+These modes are at a disadvantage as a result of: the travel impedance for longer trips (see Figure \ref{fig:Fig4}; their low population values values overall; and larger populations in origins with high shares of travel by motorized modes.
 These factors all contribute to the the car/motor mode being most advantaged in capturing spatially available job opportunities overall.
@@
 
-There are spatial variations in the competitive advantage of the car-using populations.
-The proportion of car-using population in the Centro is smaller and has higher travel impedance values relative to the inputs in other areas and mode-using populations.
-The LEZ Centro implementation further restricts the car-advantage as it shifted more than half of all car trips into the LEZ to another mode [@tarrinoortizAnalyzingImpactLow2022].
-This restriction decreased the number of car-using population from $i$s going into the LEZ Centro (an area with a large number of jobs overall, see Figure \ref{fig:Fig2}), thus increasing the mass effect for non-car modes and resulting in proportionally higher $v_i^m$ values for non-car modes.
-As such, the lower amount of access to opportunities by car-mode allows more opportunities in the LEZ to be available by populations using other modes.
-
-As summarized in the two right-most columns in Figure \ref{fig:Fig6}, the proportion of spatial availability allocated to the car-using population in the Centro (`r modal_V_comps_4plot$perc[23] %>% percent(accuracy=0.1)` or 5,373 opportunities).
-As a comparative reference, this is less than the proportion of the car-using population in the Centro (`r modal_V_comps_4plot$perc[19] %>% percent(accuracy=0.1)`), evidently less then the proportion of car-using population in the city, and is the opposite of the trend overall (left-most columns) and within the M-30 (middle columns).
-More opportunities are spatial availability to non-car using populations within the Centro, particularly transit-using populations (`r modal_V_comps_4plot$perc[24] %>% percent(accuracy=0.1)` of spatially available jobs in the Centro despite representing `r modal_V_comps_4plot$perc[20] %>% percent(accuracy=0.1)` of the population in the Centro and `r modal_V_comps_4plot$perc[4] %>% percent(accuracy=0.1)` in the city overall).
-
-From Figure \ref{fig:Fig6}, it is also summarized that there is a higher proportion of opportunities spatially available to walking and cycling populations in the Centro than in the City overall and in all areas within the M-30.
-Notably, within the Centro, `r modal_V_comps_4plot$perc[21] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[22] %>% percent(accuracy=0.1)` of opportunities are spatially available to bike and walk modes respectively, while their populations represent smaller proportions of `r modal_V_comps_4plot$perc[1] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[2] %>% percent(accuracy=0.1)` of the population overall.
-Though the proportion of spatial availability for these mode-using populations is still lower than the proportion of mode-using population located in the Centro, these modes are more competitive within the Centro than outside of the Centro.
-By restricting the more competitive car mode through the LEZ, the advantage in the spatial availability of opportunities afforded to the otherwise lesser competitive modes is made apparent.
+Despite the big picture, it is important to note that there are spatial variations in the competitive advantage of cars.
+The proportion of travel by car in Centro is smaller and has higher travel impedance values relative to the inputs in other areas and by other modes.
+The LEZ Centro implementation further restricts the advantage of cars by shifting more than half of all car trips into the LEZ to modes [@tarrinoortizAnalyzingImpactLow2022].
+This restriction decreased the number of trips by car from going into the LEZ Centro (an area with a large number of jobs overall, see Figure \ref{fig:Fig2}), thus increasing the mass effect for non-car modes and resulting in proportionally higher $v_i^m$ values for non-car modes.
+As such, the lower amount of access to opportunities by car allows more opportunities in the LEZ to be available to populations using other modes.
+
+As summarized in the two right-most columns in Figure \ref{fig:Fig6}, the proportion of jobs spatially available to car in Centro is (`r modal_V_comps_4plot$perc[23] %>% percent(accuracy=0.1)` or 5,373 opportunities).
+For reference, this is less than the proportion of the car users in Centro (`r modal_V_comps_4plot$perc[19] %>% percent(accuracy=0.1)`), evidently less then the proportion of car users in the city, and is the opposite of the overall trend (left-most columns) and within the M-30 (middle columns).
+More opportunities are spatially available to non-car users within Centro, particularly by transit, a mode with `r modal_V_comps_4plot$perc[24] %>% percent(accuracy=0.1)` of spatially available jobs in Centro, despite representing only `r modal_V_comps_4plot$perc[20] %>% percent(accuracy=0.1)` of the population in Centro and `r modal_V_comps_4plot$perc[4] %>% percent(accuracy=0.1)` in the city overall.
+
+Figure \ref{fig:Fig6} also shows that there is a higher proportion of opportunities that are spatially available to pedestrians and cyclists in Centro than in the City overall and in all areas within the M-30.
+Notably, within Centro, `r modal_V_comps_4plot$perc[21] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[22] %>% percent(accuracy=0.1)` of opportunities are spatially available to bike and walk modes respectively, while their populations represent smaller proportions of `r modal_V_comps_4plot$perc[1] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[2] %>% percent(accuracy=0.1)` of the population.
+Although the proportion of opportunities that are spatially available to these modes is still lower than the respective shares of the population located in the Centro, these modes are more competitive within the Centro than anywhere without.
+By restricting the ability of cars to enter Centro, the LEZ also contributes to level the playing field for slower modes, in particular cycling and walking.
 
 The spatial differences in the competitive dis/advantage of spatial availability between modes can also be visualized per origin.
-Figure \ref{fig:Fig7} visualizes $v_i^m$, the spatial availability $V_i^m$ divided by the mode-population.
-$v_i^m$ values above 1 are represented in increasing red shades, values below 1 are represented in increasingly green shades, and values equal to 1 are white.
-These plots illustrates the discussion of the disproportionately high over-allocation of spatial availability relative to the mode-using population in many of the origins for the car/motor mode (bottom left plot, areas denoted with green $v_i^m$ values above 1).
-These plots also visualize areas that disproportionately capture lower spatial availability (under 1), represented in shades of red.
-It can be observed that the transit-using population's spatial availability to jobs is relatively balanced (i.e., many zones are white), while the non-motorized modes $v_i^m$ values are low (under 1) overall.
+Figure \ref{fig:Fig7} shows $v_i^m$, the spatial availability $V_i^m$ divided by the population that uses $m$.
+Values of $v_i^m$ above 1 are shown in red and represent more than one spatially available opportunity per capita; values below 1 are shown in green, and represent less than one spatially available opportunity per capita; values equal to 1 are shown in white.
+These plots drive home the point of the disproportionately high over-allocation of spatial availability relative to the share of modes in many of the origins for car (bottom left plot, areas denoted with green $v_i^m$ values above 1).
+These plots also show areas with disproportionately few spatially available opportunities (under 1).
+It can be observed that the spatial availability to jobs of transit users is relatively well balanced (i.e., many zones are white), while that for non-motorized modes is low (under 1) overall.
 
@@
 
-Interestingly, as also represented in Figure \ref{fig:Fig6}, $v_i^m$ for car/motor within and near the LEZ Centro is near or below 1 (white/red) in Figure \ref{fig:Fig7} while all non-car modes have relatively higher $v_i^m$ values.
-Though the spatial availability from before the LEZ Centro implementation is unknown, Figure \ref{fig:Fig7} provides a benchmark for quantifying potential LEZ implementations in the future (given 2018 travel conditions).
-Figure \ref{fig:Fig7} also shows that many areas within the M-30 have high (white/green) $v_i^m$ values for car-mode, signaling that the spatial expansion of the LEZ Centro stands to increase the spatial availability of jobs for non-car mode using populations.
+It is also interesting to note that $v_i^m$ for car within and near LEZ Centro is close to or below 1 (white/red) in Figure \ref{fig:Fig7}, while in contrast all non-car modes have relatively higher $v_i^m$ values.
+Since the values are comparable across regions and over time, Figure \ref{fig:Fig7} provides a benchmark for quantifying potential LEZ implementations in the future.
+As Figure \ref{fig:Fig7} also shows, many areas within the M-30 have high (white/green) $v_i^m$ values for car, but the results for LEZ Centro give reasonable grounds to speculate that a spatial expansion of the LEZ to include all areas within the M-30 stands to increase the spatial availability of jobs for more transit, cyclcing and walking.
 
@@
 
+<!--  CONTINUE EDITING FROM HERE  -->
+
 # Discussion and conclusions
+
 <!--QUESTION: maybe break up this section and consolidate all limitations/assumptions/potential uses of multimodal spatial availability in the conclusion?-->
@@
 # Acknowledgements
+
 This research was funded by the Canada Graduate Scholarship - Doctoral Program (CGS D) provided by the Social Sciences and Humanities Research Council (SSHRC) and Project Mobilizing Justice, also supported by SSHRC.

### 2023-11-07 Journal-manuscript/Journal-manuscript.tex
Further edits, close to 80% done

@@
 \section*{Abstract}
-Recent research has tried to address the way opportunities are counted
+Recent research has aimed to address the way opportunities are counted
 in accessibility analysis. In conventional accessibility measures,
@@
 from the perspective of planning too {[}see, \emph{inter alia},
-4,5--8{]}. Beginning with the work of {[}1{]}, accessibility measures
-have been widely used to evaluate the efficiency of transportation
-systems when combined with the distribution of opportunities in space
-{[}e.g., {]}. As such, it is a holistic measure of spatial systems that
-measures the ease of reaching destinations {[}9,10{]}.
+4,5--8{]}. Beginning with the work of Hansen {[}1{]}, accessibility
+measures have been widely used to evaluate the efficiency of
+transportation systems when combined with the distribution of
+opportunities in space {[}e.g., {]}. As such, it is a holistic measure
+of spatial systems that measures the ease of reaching destinations
+{[}9,10{]}.
 
@@
 job is taken up by someone in the population, the same job is no longer
-available for any other person to take. Other types of opportunities are
-at the very least subject to congestion: multiple people can avail
+available for any other person to take. More generally, opportunities
+are subject to congestion: for example, multiple people can avail
 themselves of the services of the same family doctor, but the more
@@
 {[}11,12{]}. While these approaches purport to account for congestions,
-{[}13{]} demonstrate that in general they do not solve the issue of
-multiple counting of opportunities, thus leading to biases in the
-calculation of total demand and supply, sometimes inflating them, other
-times deflating them. In response to this, recent research has paid
-closer attention to the way opportunities are counted in accessibility
-analysis. {[}13{]}, for example, tackle floating catchment area methods
-and introduce a normalization of the impedance matrix to allocate the
-population and then the level of service proportionally. More recently,
-{[}14{]} introduced a singly-constrained measure of accessibility,
-called spatial availability, that employs a similar, but more
-sophisticated proportional allocation mechanism. The work of these
-authors shows that floating catchment area methods can be seen as
-singly-constrained accessibility measures, and improve on them by
-guaranteeing that each opportunity is counted only once - in other
-words, treating opportunties as \emph{finite}. The proportional
-allocation of spatial availability constrains the calculations to match
-a known quantity, therefore ensuring that the measurements sum up to a
-predetermined quantity (i.e., the total number of opportunities), and so
-each value can be meaningfully related to this total.
-
-A limitation of spatial availability as introduced by {[}14{]} is that
-it was developed for the case of a homogeneous population, for example
-for the case of a single mode of transportation. However, the finity of
-the opportunities makes the analysis of heterogeneous populations very
-relevant. In the case of multiple modes of transportation, people who
-travel by slow modes (e.g., active modes) can usually reach fewer
-opportunities than people who travel by faster modes and whose range is
-typically far wider (e.g., car). This implies that slower travelers will
-often face increased competition for local opportunities from travelers
-who can reach said opportunities from farther afield.
+Paez et al. {[}13{]} demonstrate that in general they do not solve the
+issue of multiple counting of opportunities, thus leading to biases in
+the calculation of total demand and supply, sometimes inflating them,
+other times deflating them. In response to this, recent research has
+paid closer attention to the way opportunities are counted in
+accessibility analysis. {[}13{]}, for example, tackle floating catchment
+area methods and introduce a normalization of the impedance matrix to
+allocate the population and then the level of service proportionally.
+More recently, Soukhov et al. {[}14{]} introduced a singly-constrained
+measure of accessibility, called spatial availability, that employs a
+similar, but more sophisticated proportional allocation mechanism. The
+work of these authors shows that floating catchment area methods can be
+seen as singly-constrained accessibility measures, and improve on
+existing approaches by guaranteeing that each opportunity is counted
+only once - in other words, treating opportunities as \emph{finite}. The
+proportional allocation of spatial availability constrains the
+calculations to match a known quantity, therefore ensuring that the
+measurements sum up to a predetermined quantity (i.e., the total number
+of opportunities), and so each value can be meaningfully related to this
+total.
+
+A limitation of spatial availability as introduced by Soukhov et al.
+{[}14{]} is that it was developed for the case of a homogeneous
+population, for example for the case of a single mode of transportation.
+However, the finite nature of opportunities makes the analysis of
+heterogeneous populations very relevant. In the case of multiple modes
+of transportation, people who travel by slow modes (e.g., active modes)
+can usually reach fewer opportunities than people who travel by faster
+modes and whose range is typically far wider (e.g., car). This implies
+that slower travelers will often face increased competition for local
+opportunities from travelers who can reach said opportunities from
+farther afield.
 
@@
 availability. Our primary motivation is to extend spatial availability
-for the case of multimodal accessibility, but this is in fact just one
-example of heterogeneous populations (i.e., travel by different modes).
-The method itself can easily accommodate other forms of heterogeneity,
-for example variations in travel behavior between older and younger
-adults {[}e.g., {[}15{]}; {]}, the propensity of older adults to use
-different modes of transportation {[}16{]}, the usually shorter trip
-lengths of children compared to grown-ups {[}e.g., 17{]}, or the more
-limited travel ranges of single parents {[}e.g., 18{]}.
-
-The paper rest of this paper is organized as follows. In Section we use
-a synthetic examples with multiple modes to illustrate some relevant
-issues; this helps us motivate the derivation of the new spatial
-availability expressions. This is followed in Section 3 by an empirical
-example using data from the city of Madrid after the implementation of
-its Low Emission Zones. Data for this example comes from the city's 2018
-travel survey. The example shows the differences in spatial availability
-within and outside the LEZ for travellers using different modes, namely
-car, transit, cycling and walking. In Section 4, we provide concluding
-remarks on the strengths of the use of spatial availability as a
-multimodal accessibility measure, and discuss potential future uses in
-policy planning scenarios as well as directions for future research.
-
-\hypertarget{a-review-of-multimodal-accessibility-measures}{%
-\section{A review of multimodal accessibility
-measures}\label{a-review-of-multimodal-accessibility-measures}}
+for the case of multimodal accessibility, but it is worthwhile noting
+that this is in fact just one case of heterogeneous populations (i.e.,
+travel by different modes). The method itself can easily accommodate
+other forms of heterogeneity, for example variations in travel behavior
+between older and younger adults {[}e.g., 15{]}, the propensity of older
+adults to use different modes of transportation {[}16{]}, the usually
+shorter trip lengths of children compared to grown-ups {[}e.g., 17{]},
+or the more limited travel ranges of single parents {[}e.g., 18{]}.
+
+The paper rest of this paper is organized as follows. In Section we
+discuss the use a synthetic examples with multiple modes to illustrate
+some relevant issues; this helps us motivate the derivation of the new
+spatial availability expressions. This is followed in Section 3 by an
+empirical example using data from the city of Madrid after the
+implementation of its Low Emission Zones. Data for this example comes
+from the city's 2018 travel survey. The example shows the differences in
+spatial availability within and outside the LEZ for travelers using
+different modes, namely car, transit, cycling and walking. In Section 4,
+we provide concluding remarks on the strengths of the use of spatial
+availability as a multimodal accessibility measure, and discuss
+potential future uses in policy planning scenarios as well as directions
+for future research.
+
+\hypertarget{a-brief-review-of-multimodal-accessibility}{%
+\section{A brief review of multimodal
+accessibility}\label{a-brief-review-of-multimodal-accessibility}}
 
@@
 \(S_i^m = \sum_j O_j f^m(c_{ij}^m)\) where \(m\) is a set of modes which
-have mode-specific travel costs (\((c_{ij}^m)\)) and/or travel impedance
+have mode-specific travel costs (\(c_{ij}^m\)) and/or travel impedance
 functions (\(f^m(\cdot)\)).
 
-The Hansen-type measure is not constrained, and as a result does not
-consider the opportunities as finite. As an example, the work of
-{[}20{]} uses the Hansen-type measure to measure the potential
-interaction with retail locations using walking, public transit, and car
-modes as \(m\). \(S_i^m\) is the sum of retail locations \(j\) that can
-potentially be reached under the travel impedance as calculated for each
-\(i\) and \(m\). In other words, each \(i\) has three \(S_i^m\) values,
-one per \(m\). In this work, they demonstrate that the car mode has the
-highest \(S_i^{m}\) values in the majority of \(i\), i.e., populations
-using a car can potentially reach more retail opportunities than
-populations using other modes. However, higher \(S_i^{m}\) values for
-car do not affect the values of \(S_i^{m}\) for other modes: in effect,
-each modes is analyzed as if the others did not exist, which would be
-reasonable for an inexhaustible opportunity, but is more questionable
-when the opportunity is subject to congestion. Since the measure is
-measure is also not constrained, and each opportunity can be and often
-is counted multiple times within and between modes, the sum of
-accessibility is not a meaningful quantity. To improve interpretability,
-\(S_i^m\) values are presented as standardized accessibility index in
-this work, from the lowest value of 0 to the highest value of 1. This
-presentation or discussion of Hansen-type accessibility measures is
-common in the literature {[}e.g., in 21,and 22{]}.
-
-However, this representation of results can make the interpretation of
-results challenging \emph{because} values are region-relative and
-lacking numeric meaning. Zones will always have values from between 0 to
-1 represented, but is a zone with a low value for pedestrian-modes and a
-high value for car-modes notable? And if notable, what does the
-difference in these standardized values tangibly mean for planners? By
-how much should transport systems and land-use configurations be changed
-to improve conditions? And in what way can these existing methods be
-used to track progress over time? Or between regions? This is
-challenging to track, since certain values will always be relatively
-`low' and `high'. Alternatively, modal accessibility calculated using
-the Hansen-type measure is also presented without standardization, as in
-the work by {[}23{]}. Often, the meaning of the accessibility value
-itself is not interpreted, as it is difficult to state what having
-access to 30,000 employment opportunities by car but having 10,000
-employment opportunity by transit tangibly signifies.
-
-However, once opportunities in a region are interpreted as finite, a new
-meaning is taken on. As considered in the long tradition of
-accessibility work, capacity of opportunities is limited and thus is
-subject to competition by population {[}11,14,24--28{]}. There are only
-so many school-seats, hospital capacity, employment opportunities, etc.,
-in a region and if one person interacts with an opportunity at a given
-time, it is taken: the supply of an opportunity and the demand for that
-opportunity are two components of accessibility. The competition
-consideration is especially important in the case of multiple modes. For
-instance, people in a zone who are advantaged with a relatively low
-travel-cost mode have the ability to potentially interact with more
-opportunities than other people. Due to this advantage, through the
-perspective of finite opportunities, there are fewer opportunities left
-to be potentially interacted for everyone else, especially zones with
-populations using higher travel-cost modes. This recognition is the
-motivation behind integrating \emph{competition} for opportunities
-within modal accessibility measures.
-
-Arguably, one of the most popular competitive location-based
-accessibility measures is the two-step floating catchment area (2SFCA)
-approach popularized by {[}12{]} who simplified the approach proposed by
-{[}11{]} (with similar considerations for competition in {[}28{]} and
-{[}29{]}). The Shen-type accessibility measure's formulation is:
+Hansen-type accessibility is not constrained, which is to say it does
+not consider the opportunities as finite. To cite an example, Tahmasbi
+et al. {[}20{]} use Hansen-type accessibility to assess the potential
+interaction with retail locations by three modes: walking, public
+transit, and car (i.e., \(m = w, p, c\)). \(S_i^m\) is the sum of retail
+locations \(j\) that can potentially be reached under the travel
+impedance as calculated for each \(i\) and \(m\). In other words, for
+each origin \(i\) three accessibility scores are calculated. In this
+work, Tahmasbi et al. {[}20{]} show that car travel affords the highest
+\(S_i^{m}\) values in the majority of \(i\), i.e., travelers who use a
+car can potentially reach more retail opportunities than populations
+using other modes. However, higher \(S_i^{m}\) values for car do not
+affect the values of \(S_i^{m}\) for other modes: in effect, each mode
+is analyzed as if the others did not exist. Since the measure is not
+constrained, each opportunity is typically counted multiple times within
+and between modes, and as a result the sum of accessibility is not
+necessarily a meaningful quantity. The accessibility scores for the
+modes are often values that are difficult to interpret beyond making
+statements about relative size. For example, Lunke {[}21{]}, researching
+the region of Oslo, reports accessibility scores for car in the order of
+tens of thousands of employment opportunities. The corresponding scores
+for transit are lower, but still often in the thousands or tens of
+thousands. As reported, the ratio of the transit to the car score can be
+lower than 0.2 (meaning transit gives access to less than 20\% of the
+opportunities than car). But despite the discussion about ``sufficient
+accessibility'' (e.g., in the abstract), it is unclear what the
+unconstrained scores mean: is having access to but 10,000 jobs by
+transit insufficient? After all, 10,000 employment opportunities are
+still plenty of opportunities. These ratios can be found elsewhere in
+the literature {[}e.g., 22, and 9,15,18, and 8{]}, and they are useful
+to assess when some members of the public are better or worse off than
+others, but they do not say much about how bad is ``worse''.
+
+Besides ratios of accessibility, another way to improve interpretability
+of scores sometimes seen in the literature is to standardize them to lie
+in the range {[}0-1{]}. This adjustment is only helpful insofar as it
+facilitates relative comparisons, but interpretation of the scores
+remains challenging because the values are specific to a region and
+convey no meaning about the magnitude of the scores. In this approach,
+zones always have values between 0 and 1, but how remarkable is a zone
+with a low score for pedestrians and a high value for car? And if
+remarkable, what does the difference in these standardized values mean
+for planners? By how much should transport systems and land-use
+configurations be changed to improve conditions? And in what way can
+these scores be used to track differences over time? Or between regions?
+These questions lack straightforward answers since certain values will
+always be relatively `low' or `high', but do not track to a quantity
+that can be intuitively understood. Presentation or discussion of
+Hansen-type accessibility that has been standardized in this way is not
+uncommon in the literature {[}e.g., 23,24{]}.
+
+Once we understand opportunities to be finite, it is possible for an
+accessibility measure to take a crisper meaning. As considered in the
+long tradition of accessibility research, capacity of opportunities is
+limited and thus is subject to competition by population
+{[}11,14,25--29{]}. There are only so many school-seats, hospital
+capacity, employment opportunities, etc., in a region and if one person
+reaches an opportunity, it is taken: the supply of an opportunity and
+the demand for that opportunity are two components of accessibility.
+These are clear examples of opportunities that are clearly competitive.
+But we would go as far as to argue that every type of opportunity is
+subject to congestion or capacity constraint, even when the
+opportunities are conventionally seen as non-competitive. Amenities are
+a good example of this. For instance, standards for providing green
+spaces are often stated in the form of \emph{exclusive access}, in units
+of amenity per capita. A case in point is Natural England, an
+organization that recommends an Accessible Natural Greenspace Standard
+such that the minimum supply of space is one ha of statutory Local
+Nature Reserves per thousand population\footnote{see
+  https://redfrogforum.org/wp-content/uploads/2019/11/67-Nature-Nearby\%E2\%80\%99-Accessible-Natural-Greenspace-Guidance.pdf}.
+Similarly, the World Health Organization {[}cited in 30{]} recommends
+that cities provide a minimum of 9\(m^2\) of green area per inhabitant.
+For our purposes, standards of this type translate into ``how much of a
+resource can be offered to each individual that is not allocated to
+anyone else?''. Green spaces often have large capacities, but they still
+have a capacity. Constraining accessibility is in this way a useful way
+to evaluate the congested availability of any type of opportunity. As
+development of sound standards is emphasized in the planning literature,
+in particular in regards to fairness in transportation {[}see 31{]},
+spatial availability analysis is a useful way to develop and assess
+standards.
+
+The relevance of the considerations above is put in sharper relief when
+we think about the use of multiple modes (or heterogeneous populations).
+If we return to Oslo for a moment {[}21{]}, we notice that the places
+that have high accessibility by transit are also the places that have
+\emph{very high} accessibility by car (Figure 2). Those two populations
+are going for the same opportunities, and those travelling by transit
+have fewer to choose from to begin with. More generally, people in a
+zone who are advantaged with relatively cost of travel will have the
+ability to potentially reach more opportunities than other people. Due
+to this advantage, through the perspective of finite opportunities,
+there are fewer opportunities left for everyone else, especially for
+those who use modes that are slower or more expensive.
+
+As noted in the introduction, competitive accessibility was the
+rationale for developing floating catchment area methods (FCA),
+popularized by {[}12,11,although similar, and earlier, developments are
+found in 29,32{]}. Shen-type accessibility is formulated as:
 \(a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}\) where \(D_j^m\)
@@
 weighted population \(\sum_i P_i^m f^m(c_{ij}^m)\) and the remaining
-variables are repeated in the Hansen-type measure. The Shen-type measure
-per mode (\(a_i^m\)) can be understood as a ratio of the travel
-impedance-weighted supply of opportunities for the \(m\)-using
-population in \(i\) over the ravel impedance-weighted demand for
-opportunities. In this way, it considers competition but the measure
-remains \emph{non-constrained} meaning resulting values are not
-associated with global maximums. In other words, it is difficult to
-interpret the meaning of differences in Shen-type accessibility scores
-between modes in similar ways as reflected for Hansen-type measures.
-
-To illustrate, {[}30{]} calculates \(a_i^m\) to jobs for different
+variables are repeated in the Hansen-type measure. Shen-type modal
+accessibility (\(a_i^m\)) can be understood as a ratio of the travel
+impedance-weighted supply of opportunities for \(m\)-mode in \(i\) over
+the travel impedance-weighted demand for opportunities. In this way, it
+considers competition. That said, the measure remains unconstrained,
+meaning both population \emph{and} opportunities are multiply counted
+{[}see 13{]}. In other words, interpretation of differences in Shen-type
+accessibility scores between modes is fraught as it is for Hansen-type
+measures.
+
+To illustrate, {[}33{]} calculate \(a_i^m\) to jobs for different
 income-group populations in Shenzhen, China using
-\(m = \text{public transit}\) and \(m={car}\). They demonstrate that
-\(i\) with low-income populations have lower \(a_i^m\) than \(i\) with
-higher-income populations. Further, they demonstrate that
+\(m = \text{public transit}\) and \(m=\text{car}\). Their results
+indicate that zones with low-income populations have lower \(a_i^m\)
+than zones with higher-income populations. Further, they show that
 \(a_i^{m=\text{public transit}}\) is lower than \(a_i^{m=\text{car}}\)
-at many \(i\), arguing that this may put \(i\) with lower-income
-populations in a further disadvantage. \(a_i\) and/or \(a_i^m\) are used
-to compare relative spatial differences in overall competitive
+at many zones, arguing that this may further place those zones with
+lower-income populations at disadvantage. \(a_i\) and/or \(a_i^m\) are
+used to compare relative spatial differences in overall competitive
 accessibility and multimodal competitive accessibility, but because
-there is no global maximum, making it is difficult to interpret the
-significance between differences in \(a_i^{m}\) values. Questions such
-as: what is the impact that competition has on the difference in
-\(a_i^m\) values? How does impact vary spatially? And what is the
-interpretation of this difference? are left unaddressed.
-
-Spatial availability improves on previously discussed multimodal
-accessibility approaches using the Hansen-type measure and the Shen-type
-measure by taking into account \emph{competition} in the potential
-interaction with opportunities in a \emph{constrained} framework (e.g.,
-finite opportunities). This is done by considering: 1) competition
-between mass effect (e.g., the advantage of sub-populations residing in
-relatively low population-density and high opportunity-proximate areas)
-and 2) competition between travel impedance (e.g., sub-populations with
-relatively low travel-impedance) through a proportional allocation
-mechanism. The following sub-section demonstrates how spatial
-availability compares to the Hansen-type and Shen-type measures through
-a synthetic example.
+opportunities were doubly counted (entering the sums of both modes),
+this makes for uneasy interpretations of the differences in \(a_i^{m}\)
+between modes. Questions that this approach leaves unaddressed include:
+what is the impact of competition on the difference in \(a_i^m\) values?
+How does the impact vary spatially? And what is the interpretation of
+this difference?
+
+Spatial availability improves on previously discussed accessibility
+approaches using the Hansen-type measure and the Shen-type measure by
+constraining the sum of opportunities, that is, by treating
+opportunities as finite. This is done by means of proportional
+allocation factors that follow well established principles of spatial
+interaction and the gravity model {[}see 34{]}. In Soukhov et al.
+{[}14{]} these factors consider: 1) the mass effect (e.g., the size of
+populations at different origins); and 2) the cost of travel from
+different zones (e.g., some sub-populations face relatively higher or
+lower costs). The following section introduces the multimodal form of
+spatial availability.
 
 \hypertarget{multimodal-spatial-availability-v_im}{%
-\subsection{\texorpdfstring{Multimodal spatial availability
+\section{\texorpdfstring{Multimodal spatial availability
 \(V_i^m\)}{Multimodal spatial availability V\_i\^{}m}}\label{multimodal-spatial-availability-v_im}}
@@
 spatial availability \(V_{i}\) is shown in Equation
-(\ref{eq:spatial-availability-general}):
+(\ref{eq:spatial-availability-general}) {[}see 14{]}:
 
@@
 \begin{itemize}
-\tightlist
 \item
-  \(F^t_{ij}\) is a balancing factor that depends on the demand for
-  opportunities \(O_j\) and cost of movement in the system
-  \(f(c_{ij})\).
+  \(F^t_{ij}\) is a balancing factor that depends on the size of the
+  populations at different locations that demand opportunities \(O_j\),
+  as well as the cost of movement in the system \(f(c_{ij})\).
 \item
   \(V_i\) is the number of spatially available opportunities at \(i\);
-  the sum of \(V_{i}\) is equivalent to the total sum of opportunities
-  in the region (i.e., \(\sum_j O_j = \sum_i V_i\))
+  the sum of \(V_{i}\) is identical to the total number of opportunities
+  in the region (i.e., \(\sum_j O_j = \sum_i V_i\)); in other words,
+  opportunities are dealt with as finite resources.
 \end{itemize}
 
-The spatial availability measure is introduced in {[}14{]}. Spatial
-availability's unique feature is the balancing factor \(F^t_{ij}\), a
-proportional allocation mechanisms, that ensures the \(V_i\) calculated
-for each \(i\) sums, across all \(i\) in the region, to equal the total
-number of opportunities in the region. As such, spatial availability is
-a \emph{competitive} and \emph{constrained} accessibility measure as
-\(F^t_{ij}\) handles the number of opportunities in the region in a
-finite way (proportional allocation). \(F^t_{ij}\) consists of two
-components: a population-based balancing factor
-\(F^p_{i} = \frac{P_i}{\sum_i P_i}\) and an impedance-based balancing
-factor \(F^c_{ij} = \frac{F^c_{ij}}{\sum_j F^c_{ij}}\) that,
-respectively, allocates opportunities to \(i\) in proportion to the size
-of the population at \(i\) (the mass effect) and the cost of reaching
-opportunities at \(j\) (the impedance effect).
-
-\(F^p_{i}\) and \$ F\^{}c\_\{ij\}\$ are calculated for each \(i\) such
-that they both equal 1 when summed across all \(i\) in the region (e.g.,
-\(\sum_i F^p_{i} = 1\) and \(\sum_i F^c_{ij} = 1\)). These balancing
+Compared to Hansen-type accessibility:
+
+\begin{equation}
+\label{eq:hansen-accessibility}
+A_i = \sum_{j=1}^J O_jf(c_{ij})
+\end{equation}
+
+\noindent we see that spatial availability is, like Hansen's measure, a
+weighted sum of the opportunities. What makes spatial availability stand
+apart from other approaches is the weight used in the sum, balancing
+factor \(F^t_{ij}\), implements a proportional allocation mechanisms to
+ensure that the sum of \(V_i\) is constrained to match the total number
+of opportunities in the region. As such, spatial availability is
+singly-constrained and naturally implements competition or congestion.
+\(F^t_{ij}\) consists of two parts. The first part is a population-based
+proportional allocation factor to model the mass effect of the gravity
+model: \[
+F^p_{i} = \frac{P_i}{\sum_i P_i}
+\]
+
+This factor makes opportunities available based on demand. Secondly,
+there is an impedance-based proportional allocation factor that model
+the cost effect: \[
+F^c_{ij} = \frac{F^c_{ij}}{\sum_j F^c_{ij}}
+\]
+
+This factor makes opportunities available preferentially to those who
+can reach them at a lower cost. \(F^p_{i}\) and \(F^c_{ij}\) are
+designed so that they both equal 1 when summed across all \(i\) in the
+region (e.g., \(\sum_i F^p_{i} = 1\) and \(\sum_i F^c_{ij} = 1\)). These
 factors are combined multiplicatively to yield \(F^t_{ij}\) which
@@
 \(O_j\) to each \(i\) such that the resulting \(V_i\) value represents
-the number of opportunities \emph{spatially available} to the population
-at \(i\). Each zonal value is a proportion of the opportunities in the
-region (i.e., \(\sum_j O_j = \sum_i V_i\)).
+the number of opportunities \emph{available} to the population at \(i\).
+Each zonal value is a proportion of the opportunities in the region
+(i.e., \(\sum_j O_j = \sum_i V_i\)).
 
 The focus of this paper is to extend \(V_i\) for the measurement of
-multimodal accessibility applications. To do so, the balancing factors
-are reformulated to yield a proportional value for the set of modes
-\(m\) used by populations at each \(i\). As these factors are
-proportional, \(F^{pm}_{i}\) and \(F^{cm}_{ij}\) can be summed across
-each \(m\) at each \(i\) and then across all \(i\) to equal to 1. They
-are also similarly combined multiplicatively to obtain their joint
-effect, represented as the combined balancing factor \(F^{tm}_{ij}\)
-detailed in Equation (\ref{eq:multimodal-balancing-factors}).
+multimodal applications (or more generally heterogeneous populations).
+To do so, the balancing factors need to be reformulated so that 1) the
+mass effect now accounts not only for the size of the population at
+\(i\), but also the size of sub-populations within \(i\); and 2) the
+cost of travel is not only for different zones, but by sub-populations
+within each zone (e.g., the cost of travel from \(i\) by car, transit,
+walking, etc.) When we introduce modes (or sub-populations) \(m\), the
+proportional allocation factors need to satisfy the condition that
+\(F^{pm}_{i}\) and \(F^{cm}_{ij}\) can be summed across each \(m\) at
+each \(i\) and then across all \(i\) to equal to 1. They are also
+similarly combined multiplicatively to obtain their joint effect,
+represented as the combined balancing factor \(F^{tm}_{ij}\) similar to
+that detailed in Equation (\ref{eq:multimodal-balancing-factors}). This
+factor is given by:
 
@@
 
-\noindent Where:
+\noindent where:
 
@@
 \item
-  The population balancing factor for each \(m\) at each \(i\) is
-  \(F^{pm}_{i} = \frac{P_{i}^m}{\sum_{m}\sum_{i} P_{i}^m}\)
+  The factor for allocation by population for each \(m\) at each \(i\)
+  is \(F^{pm}_{i} = \frac{P_{i}^m}{\sum_{m}\sum_{i} P_{i}^m}\); and
 \item
-  The cost of travel balancing factor for each \(m\) at \(i\) is
+  The factor for allocation by cost of travel for each \(m\) at \(i\) is
   \(F_{ij}^{cm} = \frac{f^m(c_{ij}^m)}{\sum_{m} \sum_{i} f^m(c_{ij}^m)}\)
@@
 Implementing \(F^{tm}_{ij}\), the following Equation
-(\ref{eq:spatial-availability-multimodal}) demonstrates the multimodal
-configuration of spatial availability \(V_i^m\):
+(\ref{eq:spatial-availability-multimodal}) gives the multimodal version
+of spatial availability \(V_i^m\):
 
@@
 \item
-  \(m\) is a set of modes used by populations in the region.
+  \(m = 1, 2,\cdots, M\) is a set of \(M\) modes (or sub-populations) of
+  interest.
 \item
@@
   equivalent to the total sum of opportunities in the region (i.e.,
-  \(\sum_j O_j = \sum_i V_i = \sum_{m} \sum_{i} V^m_{i}\))
+  \(\sum_j O_j = \sum_i V_i = \sum_{m} \sum_{i} V^m_{i}\)).
 \end{itemize}
 
+Next we use a synthetic example to contrast multimodal accessibility and
+spatial availability.
+
 \hypertarget{an-illustrative-synthetic-example}{%
-\subsection{An illustrative synthetic
+\section{An illustrative synthetic
 example}\label{an-illustrative-synthetic-example}}
 
-Consider the following Figure \ref{fig:Fig1}, it depicts a region with
-population and jobs at three population centers (\(A\), \(B\), \(C\))
-and three employment centers (\(1\), \(2\), \(3\)). The population at
-each population center is divided into two sub-populations, one using a
-faster mode \(z\) and another using a slower mode \(x\), to travel to
-employment centers. Population center \(A\) is Suburban: it is closest
-to its own relatively large employment center at \(1\), close to the
-Urban's equally large employment center \(2\), and has a population that
-is smaller than the Urban \(B\) and larger than the Satellite \(C\).
-\(B\) has the largest \(x\)-using population, followed by then \(A\),
-then \(C\). This synthetic example was inspired by the single-mode
-example used in {[}11{]} and reconfigured in {[}14{]}.
+Consider the simple system shown in Figure \ref{fig:Fig1}. The figure
+shows a region with population at three population centers (\(A\),
+\(B\), \(C\)) and jobs at three employment centers (\(1\), \(2\),
+\(3\)). The population at each origin \(i\) is consists of two
+sub-populations, one using a faster mode \(z\) and another using a
+slower mode \(x\), to travel to employment centers. Population center
+\(A\) is Suburban: it is closest to its own relatively large employment
+center at \(1\), close to the Urban's equally large employment center
+\(2\), and has a population that is smaller than the Urban \(B\) and
+larger than the Satellite \(C\). \(B\) has the largest \(x\)-using
+population, followed by then \(A\), then \(C\). This synthetic example
+is inspired by the single-mode example used in {[}11{]} and reconfigured
+in {[}14{]}.
 
@@
 opportunities in the region (\(210,000\) jobs), the sub-population that
-is most proximate to jobs, furthest from densely populated centers, and
-is using the lowest travel-cost mode \(z\) can potentially access the
-most job opportunities. This appears to be the sub-population at \(A\)
-using mode \(z\). Sub-populations located in opposite conditions (i.e.,
-further away from jobs, close to dense populations, and using \(x\)) are
-at a relative job opportunity access \emph{disadvantage}. The
-competition for opportunities between different mode-using populations
-matters from the perspective of inequities as it reflects how well the
-land-use and transport system serves (or doesn't serve) certain
+is most proximate to jobs (low cost to reach), furthest from large
+populations (less competition), and uses the fastest mode \(z\) (greater
+range) can potentially reach the largest number of opportunities. This
+appears to be the sub-population at \(A\) using mode \(z\).
+Sub-populations located in opposite conditions (i.e., more distant from
+jobs, close to large populations, and using slow mode \(x\)) are at a
+relative disadvantage. The competition for opportunities between
+different mode-using populations matters as it reflects how well the
+land-use and transport system serves (or does not serve) certain
 populations.
@@
 
-The calculated \(S_i^m\), \(a_i^m\) and \(V_i^m\) accessibility values
-for each \(i\) and \(m\) are shown in the middle three columns and are
-aggregated for each \(i\) in the final two columns in Table 1 . We use a
-negative exponential impedance function
+The values calculated for \(S_i^m\) (Hansen-type accessibility),
+\(a_i^m\) (Shen-type accessibility), and \(V_i^m\) (spatial
+availability) for each \(i\) and \(m\) are shown in the middle three
+columns and are aggregated for each \(i\) in the final two columns in
+Table 1 . As in the example in Shen {[}11{]}, we use a negative
+exponential impedance function
 \(f^m(c_{ij}^m) = \exp(-\beta\cdot c_{ij})\) with \(\beta=0.1\) for both
 \(x\) and \(z\) modes for all accessibility measures calculations.
-
-The Hansen-type measure \(S_i^m\) is presented for each origin and mode
-in third column of Table 1 . For all \(i\), the \(z\)-using
-sub-population has higher \(S_i^m\) values than the \(x\)-using
-sub-populations. Additionally, \(S_i^m\) is equal for both mode-using
-populations in \(A\) and \(B\). This is the case since \(S_i^m\) does
-not consider \emph{competition}, it only relies on reflecting the count
-of opportunities that may be interacted with as a product of
-\(f^m(c_{ij}^m)\). Recall, populations in \(A\) and \(B\) have the same
-travel impedance to employment centers \(1\), \(2\) and \(3\) (either
-15, 30, or 100 minutes using \(x\) or 10, 25, or 80 minutes using
-\(z\)). As such, these the calculated \(S_i^m\) values are the same for
-both \(A\) and \(B\). Furthermore, the total sum of \(S_i^m\) in the
-region is equal to 150,570.2. This value is difficult to interpret: it
-represents the weighted sum of opportunities that may be interacted with
-within the region based on travel impedance. It cannot be interpreted as
-any sort of benchmark since the measure is \emph{non-constrained}. To
-connect this example to literature, \(S_i^m\) is calculated in the work
-of {[}20{]}; they compare differences in \(S_i^m\) values between modes
-in a relative and comparative sense, but make no further interpretation
-of the \(S_i^m\) values.
-
-In the fourth and sixth column in Table 1 the Shen-type measure is
-calculated: first for both origin and mode \(a_i^m\) as well as
-aggregated by the weighted mean mode-population (
+Notice that in this example we use the same impedance function but the
+travel times are different for the two modes. More generally, it is
+possible to use different impedance functions for the modes, as
+demonstrated in the empirical example below.
+
+Hansen-type accessibility \(S_i^m\) is presented for each origin and
+mode in the third column of Table 1 . For all \(i\), the travel by \(z\)
+results in higher values of \(S_i^m\) than travel by \(x\). Lack of
+competition, or alternatively the assumption of an inexhaustible
+resource in the calculation of \(S_i^m\), lead to a curious result.
+Since the populations in \(A\) and \(B\) have the same travel impedance
+to employment centers \(1\), \(2\) and \(3\) (either 15, 30, or 100
+minutes using \(x\) or 10, 25, or 80 minutes using \(z\)), their values
+of \(S_i^m\) are the same for both \(A\) and \(B\). Furthermore, the
+total sum of \(S_i^m\) in the region is equal to 150,570.2. This value
+lacks an intuitive interpretation: it represents the weighted sum of
+opportunities that may be reached within the region according to the
+travel impedance (i.e., the travel behavior and the characteristics of
+the modes) and does not usefully translate into any sort of benchmark.
+To connect this example to the literature, \(S_i^m\) is calculated in
+the work of {[}20{]}; they compare differences in \(S_i^m\) values
+between modes in a relative and comparative sense, but make no further
+interpretation of the \(S_i^m\) values. Large metropolitan regions will
+have large values and small regions small, but that is simply an
+artifact of size.
+
+In the fourth and sixth columns in Table 1 the results for Shen-type
+accessibility are reported: first for both origin and mode \(a_i^m\) as
+well as aggregated by the weighted mean mode-population (
 \(\sum_m \frac{P_i^m}{P_i}*a_i^m\) ) to represent a value for each
-origin \(a_i\). Unlike \(S_i^m\), this measure considers
-\emph{competition}. For instance, the \(x\)-using populations in \(A\)
-and \(B\) centers do not have the same \(a_i^m\) values as the
-\(z\)-using. In fact, \(A\) has the highest values \(a_i^m\) and \(a_i\)
-values since this center has the smallest travel impedance to
+origin \(a_i\). Unlike \(S_i^m\), this measure does considers
+\emph{competition}. For instance, the population travelling by \(x\)
+from \(A\) and \(B\) do not have the same values of \(a_i^m\) as those
+travelling by \(z\). In fact, \(A\) has the highest values \(a_i^m\) and
+\(a_i\) values since this center has the lowest travel impedance to
 opportunities (lower than at \(C\), \(A\) and \(B\) are equal) and has
-one of these lowest proximity to a relatively high amount of population
-(lower than at \(B\)).
-
-However, the Shen-type measure is \emph{non-constrained}: the total sum
-of \(a_i^m\) or \(a_i\) is practically meaningless since it represents a
-sum of ratios. For instance, the \(z\)-using sub-population at \(A\) has
-a value of 1.57 potential jobs per potential job-seeking population
-compared to 0.95 for \(x\)-using sub-population. What is the
-significance of these values? The difference between these modes is
-equal to 0.62, but 0.62 of what? How many more job opportunities are
-\(z\) users interacting with than \(x\) users? When \(a_i^m\) is
-aggregated to \(a_i\) as shown in the sixth column, the values face
-similar interpretability issues. The Shen-type measure is implemented in
-the previously discussed work of {[}30{]} to calculate modal \(a_i^m\)
-values and the aggregated \(a_i\) is implemented in the work of
-{[}31{]}. However, similar to the Hansen-type measure, these works
-discuss relative and spatially comparative differences in values, they
-do not make further interpretation of the \(a_i^m\) or \(a_i\)
-themselves. This may be because the Shen-type measure is
-\emph{non-constrained}, this is no benchmark or global maximum to which
-comparisons can be drawn from.
-
-By contrast, spatial availability \(V_i\) considers competition and is
+faces relatively low competition, not being close to a relatively large
+population (lower than at \(B\)).
+
+However, the calculations of \(a_i^m\) are not constrained: the total
+sum of \(a_i^m\) or \(a_i\) is practically meaningless since it
+represents a sum of ratios. For instance, the population travelling by
+\(z\) from \(A\) has a value of 1.57 jobs per job-seeking population
+compared to 0.95 for users of mode \(x\). What is the meaning of these
+values? The difference between these modes is equal to 0.62, but 0.62 of
+what? How many more job opportunities can users of \(z\) reach compared
+to user of \(x\)? When \(a_i^m\) is aggregated to \(a_i\) as shown in
+the sixth column, the values face similar interpretability issues. The
+Shen-type measure is implemented in aforementioned work of Tao et al.
+{[}33{]} to calculate modal \(a_i^m\) values and the aggregated \(a_i\)
+is implemented in the work of {[}35{]}. However, similar to Hansen-type
+accessibility, these works discuss relative and spatially comparative
+differences in values, but veer from interpretating the values of
+\(a_i^m\) or \(a_i\) themselves. In fairness, interpretation is
+complicated by the multiple counting of opportunities between zones and
+modes.
+
+In contrast, spatial availability \(V_i\) considers competition and is
 constrained such that the total sum of values is equal to the total
 number of opportunities in the region (i.e., \(210,000\) jobs). Seen in
-fifth column of Table 1 , \(V_i^m\) for the same mode-using populations
-in \(A\) and \(B\) are not the same (as this measure considers
-competition). In fact, at \(A\), the \(z\)-using sub-population captures
-36,088.84 more spatially available jobs (of the \(210,000\) jobs in the
-region) than the sub-population using mode \(x\). The numerical
-difference has a practical interpretation.
-
-Furthermore, \(V_i^m\) values for an \(i\) can be aggregated across
-\(m\) and compared across \(i\) ( \(V_i = \sum_m{\sum_i{V_i^m}}\) ) as a
-result of the proportional allocation mechanism. This aggregation,
-\(V_i\), is shown in the seventh column in Table 1 . Again looking at
-center \(A\), \(A\) is allocated 67,482.61 spatially available
-opportunities for both modes. 77\% of this spatial availability
-allocated to \(A\) is assigned to the \(z\)-using population despite
-representing 66\% of \(A\)'s population.
+fifth column of Table 1 , the values of \(V_i^m\) in \(A\) and \(B\) are
+not the same within mode (as this measure considers competition). In
+fact, at \(A\), users of mode \(z\) capture 36,088.84 more spatially
+available jobs (of the \(210,000\) jobs in the region) than the
+sub-population travelling by \(x\). The numerical difference is clear
+since it refers to opportunities out of the total.
+
+Furthermore, the proportional allocation mechanism also means that the
+values of \(V_i^m\) for any origin \(i\) can be aggregated across \(m\)
+and compared between zones (\(V_i = \sum_m{\sum_i{V_i^m}}\)). This
+aggregation, \(V_i\), is shown in the seventh column in Table 1 . Again
+looking at center \(A\), \(A\) is allocated 67,482.61 spatially
+available opportunities for both modes. 77\% of this spatial
+availability allocated to \(A\) is assigned to users of mode \(z\)
+despite representing 66\% of \(A\)'s population.
 
@@
 competition between modes. Across the entire region, 130,000 people use
-\(z\) (62\% of the region population). However, the \(z\)-using
-population accounts for 73\% of the region's total spatial availability
-- the rest is allocated to the \(x\)-using population (38\%of the total
-population). Notably, the \(x\)-using population captures 11\% less
-spatial availability to opportunities than its population proportion.
-This understanding can lead us to ask normative questions such as, how
-unequal should opportunity access for the two mode-using populations be?
-Can the lower-travel-cost populations spare some spatial availability if
-a policy of modal-restriction (like a LEZ) was introduced?
+\(z\) (62\% of the region population). However, users of \(z\) account
+for 73\% of the region's total spatial availability - while the
+remaining 27\% is allocated to users of mode \(x\) who are 38\%of the
+total population. Notably, the population who uses \(x\) have 11\% fewer
+spatially available opportunities than its share in the population. This
+realization leads us to ask normative questions such as, how unequal
+should availability of opportunities be by mode? What intervention could
+help to redistribute spatial availability to sub-populations
+commensurate with their proportion of the total?
 
@@
 meaning as a proportion of the total opportunities in the region, the
-values at \(i\) have a new significance. Inequality in \(V_i^m\) values
-can be explored through a variety of approaches. For instance, consider
-travel times. The \(z\)-using population accounts for 67\% of the
-potential travel time traveled in the region: this is 7\% less travel
-time than the proportion of spatial available opportunities that is
-allocated to them. In other words, the \(z\)-using population travels
-less minutes overall and has more spatial availability of opportunities
-than the \(x\)-using population using the slower mode \(x\).
-
-Alternatively, inequities in spatial availability between mode-using
-populations can be explored through proportional benchmarks. A spatial
-availability per capita \(v_i^m\) as presented in Equation
-(\ref{eq:SA-per-capita}):
+values at \(i\) have a straighforward interpretation. Inequality in
+\(V_i^m\) values can be explored through a variety of approaches. For
+instance, consider travel times. The population of travelers who use
+\(z\) accounts for 67\% of the potential travel time traveled in the
+region: this is 7\% less travel time than the proportion of spatial
+available opportunities that is allocated to them. In other words, the
+population of users of \(z\) travels fewer minutes overall and has more
+spatial availability of opportunities than users of the slower mode
+\(x\).
+
+Alternatively, inequities in spatial availability between modes can be
+explored through proportional benchmarks. A spatial availability per
+capita \(v_i^m\) is presented in Equation (\ref{eq:SA-per-capita}):
 
@@
 
-The \(v_i^m\) values for \(A\), \(B\), and \(C\) for the \(x\)-using
-sub-populations are 0.95, 0.64 and 0.68 spatially available jobs per
-capita, respectively. The \(v_i^m\) for the \(z\)-using sub-populations
-are much higher, with values of: 1.57, 1.05 and 1.12 respectively. The
-\(x\)-using population, especially at \(B\) and \(C\), are directly
-impacted by the jobs that are spatially available to the \(z\)-using
-population \emph{in addition to} the mass effect (occurring at \(B\),
-high population density) and high travel impedance (occurring at the
-Satellite \(C\)).
-
-If, lets say, the planning goal is to have one spatially available job
-per mode-using population, a policy intervention can be put in place, to
-reduce the \(v_i^z\) values and increase \(v_i^x\) values. This
-demonstration is to show how simply the \(V_i^m\) framework can be
-manipulated quantify the competitive (dis)advantage in a multimodal
-application. In what follows, we further explore competition between
-multiple modes through an empirical example.
+The values of \(v_i^m\) for \(A\), \(B\), and \(C\) for users of \(x\)
+are 0.95, 0.64 and 0.68 spatially available jobs per capita,
+respectively. The values of \(v_i^m\) for users of \(z\) are much
+higher, with values of: 1.57, 1.05 and 1.12 respectively. Users of
+\(x\), especially those at \(B\) and \(C\), are directly impacted by the
+jobs that are spatially available to users of \(z\) \emph{in addition
+to} the mass effect (occurring at \(B\), high population density) and
+high travel impedance (occurring at the Satellite \(C\)).
+
+If, let us say, the planning goal was to have one spatially available
+job per mode-using population, a policy intervention could be devised,
+to reduce the values of \(v_i^z\) (making it slower or more expensive)
+and increase he values of \(v_i^x\) (making it faster or les expensive).
+The purpose of this simple demonstration is to show how spatial
+availability can be used to quantify the competitive (dis)advantage in a
+multimodal application. In what follows, we demonstrate the use of
+multimodal spatial availability through an empirical example.
 
@@
 
-\hypertarget{multimodal-data-and-methods}{%
-\subsection{Multimodal data and
-methods}\label{multimodal-data-and-methods}}
-
-Low emission zones (LEZ) have been implemented as a climate change
-policy intervention to reduce GHG emissions, improve air quality, and
-support sustainable mobility in many countries {[}32,33{]}. Though rules
-of exclusion vary, LEZ aim to deter/reduce traffic in designated zones
-under threat of penalty (e.g., fines, seizure of vehicle). From the
-perspective of restriction for passenger transport, LEZ are a policy of
-\emph{geographic discrimination} as they change how people access
-opportunities by making the travel impedance more costly for specific
-car-mode users. When considering opportunities as finite in a region,
-this discrimination could allow certain populations to access
-opportunities by other modes more readily than before. In this way, LEZ
-change the multimodal competitive accessibility landscape of a city.
+\hypertarget{data-and-methods}{%
+\subsection{Data and methods}\label{data-and-methods}}
+
+The context for the empirical example is Madrid, in Spain. This city
+implemented a Low Emission Zone in 2017, to attend to goals set out in
+the national agenda to fight climate change, cut nitrogen dioxide
+levels, and prioritize people's movement in the city. Low emission zones
+(LEZ) elsewhere have been implemented as a climate change policy
+intervention to reduce GHG emissions, improve air quality, and support
+sustainable mobility {[}36,37{]}. Though the rules of exclusion vary,
+LEZ aim to deter/reduce traffic in designated zones under threat of
+penalty (e.g., fines, seizure of vehicle). In other words, LEZ are a
+policy of \emph{geographic discrimination} as they change how people can
+reach opportunities by making the travel impedance more costly for
+specific forms of travel, typically cars. When considering opportunities
+as finite in a region, this discrimination reduces the competition of
+one mode and allows other modes to potentially thrive. In this way, LEZ
+aim to change the multimodal competitive accessibility landscape of a
+city.
 
@@
 their implementation as specified in their climate-change-related plans:
-\emph{Plan Nacional Integrado de Energía y Clima 2021-2030} {[}34{]} and
+\emph{Plan Nacional Integrado de Energía y Clima 2021-2030} {[}38{]} and
 \emph{Plan Nacional de Control de la Contaminación Atmosférica}
-{[}35{]}. Specifically, the national Spanish law 7/2021 ( \emph{Ley de
+{[}39{]}. Specifically, the national Spanish law 7/2021 ( \emph{Ley de
 Cambio Climático y Transición Energética}) will require all
@@
 air quality exceeds limits specified in \emph{RD 102/2011 de Mejora de
-Calidad del Aire} {[}36{]}.
-
-In 2017, LEZs were implemented in the Spanish capital city of Madrid
-following the goals set out in the national agenda . In geographic
-scope, the 2017 boundaries of the LEZ were relatively small (covering
-4.72 km\(^2\)) and within the center (i.e., LEZ Centro). These
-boundaries were expanded in 2023 to inside of the M-30, a highway in
-proximity to the city center (i.e., LEZ M-30) and the city has plans to
-further spatially expand the LEZ. Within the 2017 LEZ Centro
-implementation, all cars, motorcycles and freight with environmental
-label A or B (higher polluting classification, associated with older
-make and model of fossil fuel internal combustion engine vehicles), are
-not permitted to enter the area unless they are used by residents or
-meet other exemptions. This restriction impacted approximately half of
-all car trips that were typically made into the LEZ Centro {[}37{]}.
-
-For this case study, we use \(V_i^m\) to quantify the competition of
-spatially available opportunities between modes after the LEZ Centro
-implementation. Particularly, we demonstrate how \(V_i^m\) can be used
-to spectate on how the restriction of car mobility in areas
-around/within the LEZ Centro allowed the other, more sustainable but
-often with higher travel impedance modes, to become more competitive.
+Calidad del Aire} {[}40{]}.
+
+In geographic scope, the 2017 boundaries of the LEZ in Madrid were
+relatively modest (covering 4.72 km\(^2\)) and only within the center
+(i.e., LEZ Centro). These boundaries were expanded in 2023 to the area
+inside the M-30, an orbital highway in proximity to the city center
+(i.e., LEZ M-30). Beyond this, the city has plans to spatially expand
+the LEZ. Within the 2017 LEZ Centro implementation, all cars,
+motorcycles and freight with environmental labels A or B (older makes
+and models of fossil fuel internal combustion engine vehicles), are not
+permitted to enter the area unless they are used by residents or meet
+other exemptions. This restriction impacted approximately half of all
+car trips that typically travelled into what is now the LEZ Centro
+{[}41{]}.
+
+For this case study, we use spatial availability to quantify access to
+opportunities by different modes in Madrid. Particularly, we demonstrate
+how \(V_i^m\) can be used to derive insights into how the restriction of
+car mobility in areas around/within the LEZ Centro may have allowed
+other, more sustainable but often slower or more costly modes, to become
+more competitive.
 
@@
 
-The 2018 Community of Madrid travel survey {[}38{]} is the source of
-data for this empirical example: it is a representative survey that
-reflects a snap-shot of the travel patterns for a typical day of the
-working week in 2018. Specifically, it includes 222,744 trips taken from
-a representative sample of 85,064 households across the traffic analysis
-zones (TAZ) representing the Community of Madrid (population of
-6,507,184 over 3 years old) through population elevation factors.
-
-In this empirical example, we only use a sub-set of the survey,
-specifically home-to-full-time-work trips, by all modes, to demonstrate
-multimodal spatial availability of employment opportunities. The total
-population who travels to full-time work from an origin (i.e., the
-worker), and the destinations (i.e., the employment opportunity), for
-the City of Madrid (a sub-set of the Community of Madrid) is visualized
-in Figure \ref{fig:Fig2} and Figure \ref{fig:Fig3}. Both figures are
-displayed at the level of TAZ (the \(i\) and \(j\) zones) that
-correspond to the survey. The red boundary represents the LEZ Centro in
-effect in 2017 and thus those travel patterns of car-restriction
-reflected in the survey. The cyan boundary represents the LEZ that will
-be within the boundaries of the M-30 highway in 2023 and is present in
-the plots as a spatial reference for areas in proximity to the LEZ
-Centro.
-
-The total sum of jobs \(O_j\) that are held are shown in Figure
-\ref{fig:Fig2} and the populations that go to a work destination by four
-modal categories \(P^m_i\), is reflected in Figure \ref{fig:Fig3}. The
-four modal categories represented in Figure \ref{fig:Fig3} are created
-through the grouping of similar mode types as provided in the survey, by
-the authors discretion. The modal categories and the mode types within
-each category are reported as follows:
+The source of data for this example is the 2018 Community of Madrid
+travel survey {[}42{]}. This is a representative survey that gives a
+snap-shot of the travel patterns for a typical day of the working week
+in 2018. Specifically, it captures 222,744 trips as measured from a
+representative sample of 85,064 households across the traffic analysis
+zones (TAZ) in the Community of Madrid (population of 6,507,184 over 3
+years old).
+
+In this example, we use all home-to-full-time-work trips, by all modes.
+Figures \ref{fig:Fig2} and \ref{fig:Fig3} show the population of workers
+and the distribution of jobs in the City of Madrid. The zoning system is
+the trafric analysis zones used in the survey. The red boundary
+represents the LEZ Centro in effect in 2017. The survey already captures
+the travel patterns after the introduction of LEZ Centro. The cyan
+boundary represents the LEZ planned for for the boundaries of the M-30
+highway and is present in the plots as a spatial reference for areas in
+proximity to the LEZ Centro.
+
+The total sum of jobs \(O_j\) are shown in Figure \ref{fig:Fig2} and the
+populations that go to a work destination by four modal categories
+\(P^m_i\), is displayed in Figure \ref{fig:Fig3}. The modal shares in
+Figure \ref{fig:Fig3} are calculated based on those measued in the
+survey. The modal categories and the mode types within each category are
+reported as follows:
 
@@
 
-From Figure \ref{fig:Fig2}, it can be seen that the largest
-concentration of jobs are within, near, and to the north of the LEZ
-Centro. The population that is accessing those jobs by mode (Figure
-\ref{fig:Fig3}), appear spatially distinct. Car and transit trips
+Some aggregation of modes is necessary to calculate the travel impedance
+functions by mode. From Figure \ref{fig:Fig2}, it can be seen that the
+largest concentration of jobs is within, near, and to the north of LEZ
+Centro. The populations with access to those jobs by mode (Figure
+\ref{fig:Fig3}) are spatially distinct. Travel by car and transit
 represent 37\% and 47\% of the modal share respectively. The population
-that travels using transit is more spatially distributed than those
-using cars - particularly near and within LEZ Centro. This distribution
-is likely caused by a variety of factors including: transit coverage and
+that travels by transit is more spatially distributed than those using
+cars - particularly near and within LEZ Centro. This distribution is
+likely caused by a variety of factors including: transit coverage and
 service within with city, effective car infrastructure outside of the
 M-30, and/or the impact of the LEZ Centro itself. From Figure
-\ref{fig:Fig3}, it can be observed that biking and walking trips are
-less common than motorized trips at 1\% and 15\% respectively. Noteably,
-there is a positive trend between the populations of walking and biking
-trips in zones and populations of transit trips. This positive trend is
-higher than for car trip populations.
-
-The travel time for each trip is provided within the survey. These
-travel times, per modal category, are used to calibrate mode-specific
-travel impedance functions \(f^m(c_{ij}^m)\). To illustrate the modal
-differences in travel time lengths, summary descriptive per mode are
-detailed:
+\ref{fig:Fig3}, it can be seen that active travel is less common than
+motorized trips at 1\% and 15\% for cycling and walking respectively.
+Noticeably, there is a positive trend between the walking and cycling in
+zones where transit is also present. This positive trend is higher than
+for car trip populations.
+
+Travel times are provided within the survey by mode. This information is
+used to calibrate mode-specific travel impedance functions
+\(f^m(c_{ij}^m)\). To illustrate the modal differences in travel times,
+the following descriptive statistics per mode are presented:
 
@@
 
-To calculate \(f^m(c_{ij}^m)\) from the survey travel times, a concept
-known as the trip length distribution (TLD) is used. A TLD represents
-the proportion of trips that are taken at a specific travel cost such as
-travel time (i.e., probability density distribution of trips taken by
-travel cost). This distribution is then used in the derivation of travel
-impedance functions (e.g., done in the accessibility works of {[}39{]},
-{[}40{]}, and {[}41{]}). To fit the impedance functions, we use the
-Maximum likelihood estimation and the Nelder-Mead method for direct
-optimization available within the R \{fitdistrplus\} package {[}42{]}.
-Based on goodness-of-fit criteria and associated diagnostics, the gamma
-and log-normal probability density functions are selected as best
-fitting curves for the motorized and non-motorized modes respectively.
-The selection of functional form aligns with empirical examples in other
-regions {[}14,43,44{]}. The shape and rate parameters for the gamma
-functions (motorized modes) are 1.8651852 and 0.051468for car/motor and
-2.7566235 and 0.0499193 for transit; for the log-normal functions
-(non-motorized modes), mean and standard deviation parameters are
-3.2372212 and 0.7575986 for bike and 2.9918042 and 0.7575986 for walk.
-
-Represented in Figure \ref{fig:Fig4} are four plots visualizing the
-estimated probability density functions (represented as black lines)
-over top the binned travel time data from the 2018 survey. The estimated
-probability density functions can be interpreted as the probability of
-travel (y-axis) given a trip travel time (x-axis), based on trip flows
-from the 2018 travel survey. These `probability of travel' at each
-travel time for each mode are realized observations that reflect the
-land-use, the transport system, and the population travel behaviour in
-Madrid (according to the 2018 survey). A trip that leaves \(i\) is
-assigned a probability of travel value (y-axis) based on the mode and
-travel time of that trip to work: this probability value is the input
-\(f^m(c_{ij}^m)\).
-
-Notably, plots in Figure \ref{fig:Fig4} can be observed across modes if
-curious to know what trips are more likely to occur using what mode
-depending on trip length (as informed by the observed 2018 survey).
-Trips \textgreater5 minutes do not occur frequently for any mode, as
-such trips with short trip lengths are assigned a lower travel impedance
-\(f^m(c_{ij}^m)\) value. One reason that \textgreater5 minute trips do
-not frequently occur as a result of land-use (residential and job
-mismatch): not everyone lives in immediate proximity to their workplace.
-In terms of the non-motorized modes, shorter trips occurred more
-frequently overall for walking populations, particularly around 15
-minute lengths, so a trip of approximately 15 minutes is assigned the
-highest \(f^m(c_{ij}^m)\). For biking populations, longer travel times
-are more common so though the highest \(f^m(c_{ij}^m)\) value also
-corresponds to approximately 15 minutes, the curve is more spread out
-and values decrease less rapidly at longer travel times than for the
-walk mode curve. A similar trend can be observed for the motorized modal
-options where transit mode is more spread out than car/motor mode. All
-in all, these observations demonstrate that, based on a given mode and
-travel time, a trip is more or less likely to occur and is accordingly
-represented in the cost of travel balancing factor \(F_{ij}^m\) for
+Impedance functions \(f^m(c_{ij}^m)\) are calibrated from the travel
+times in the survey via the empirical trip length distribution (TLD). An
+empirical TLD is given by the proportion of trips at various travel cost
+bins. This distribution is then used to estimate the parameters of a
+function for the travel impedance {[}as done in 43,44,45{]}. To fit the
+impedance functions, we use the Maximum likelihood estimation and the
+Nelder-Mead method for direct optimization available within the R
+\{fitdistrplus\} package {[}46{]}. Based on goodness-of-fit criteria and
+associated diagnostics, the gamma and log-normal probability density
+functions are selected as best fitting curves for the motorized and
+non-motorized modes respectively. The selection of functional forms
+aligns with empirical examples in other regions {[}14,47,48{]}. The
+shape and rate parameters for the gamma functions (motorized modes) are
+1.8651852 and 0.051468for car/motor and 2.7566235 and 0.0499193 for
+transit; for the log-normal functions (non-motorized modes), the mean
+and standard deviation parameters are 3.2372212 and 0.7575986 for bike
+and 2.9918042 and 0.7575986 for walk.
+
+Figure \ref{fig:Fig4} includes four plots to visualize the calibrated
+impedance functions (represented as black lines) superimposed on the
+empirical TLD. The impedance functions can be interpreted as the
+propensity to travel (y-axis) given a trip travel time (x-axis). The
+functions reflect a combination of possibilities and preferences: the
+travel behavior given the transportation technologies available. For
+example, trips shorter than 5 minutes do not occur frequently for any
+mode; this reflects the spatial separation between places of residence
+and places of work commonly seen in many cities. In terms of the
+non-motorized modes, there is a preference towards walking trips around
+15 minutes in duration, as seen from the highest value of
+\(f^{walk}(c_{ij}^{walk})\). With respect to travel by bicycle, longer
+travel times are more common; although the highest value of the
+impedance also corresponds to approximately 15 minutes, the curve has a
+longer tail and values decrease less rapidly at longer travel times than
+is the case of \(f^{walk}(c_{ij}^{walk})\). A similar trend can be
+observed for the motorized modal options where transit mode is more
+spread out than car/motor mode. All in all, these functions represent
+the propensity of travel by mode by duration of trip, which are needed
+to calculate the proportional allocation factors \(F_{ij}^m\) for
 \(V_i^m\).
@@
 
-As it is worth re-iterating, the empirical example reflects a snap-shot
-of the calculated multimodal spatial availability using data from the
-2018 travel survey. It is intended to visualize the spatial trends in
-availability of employment opportunities, by mode, and demonstrate how
-spatial availability can be interpreted to discuss the competitive
-advantage of lower travel impedance modes within Madrid Centro.
+At this point it is worthwhile to reiterate that the empirical example
+is a snap-shot of the calculated multimodal spatial availability using
+data from the 2018 travel survey. It is intended to visualize the
+spatial trends in availability of employment opportunities, by mode, and
+demonstrate how spatial availability can be interpreted to discuss the
+competitive advantage of lower travel impedance modes within Madrid
+Centro.
 
 The spatial availability of jobs \(V_i^m\) is calculated for each of the
-four modal categories \(m\) at the level of traffic analysis zones \(i\)
-in Madrid and demonstrated in Figure \ref{fig:Fig5}.
+four modes \(m\) at the level of traffic analysis zones \(i\) in Madrid
+and displayed in Figure \ref{fig:Fig5}.
 
@@
 \(V_i^m\) is a proportion of the total number of 847,574 jobs in the
-region and is visualized in Figure \ref{fig:Fig5}. Since \(V_i^m\) is
-calculated based on the likelihood of travel from observed home-to-work
-journeys, the values can be understood as the number of full-time jobs
-that are spatially available to the full-time working population at that
+region and is shown in Figure \ref{fig:Fig5}. Since \(V_i^m\) is
+calculated based on the population of workers and the distribution of
+jobs, the values can be understood as the number of full-time jobs that
+are spatially available to the full-time working population at that
 \(i\) and their associated \(m\), relative to all the jobs in the city.
@@
 
-Notable are the differences in the magnitude of \(V_i^m\) between modes
-in Figure \ref{fig:Fig5}. The majority of \(V_i^m\) is allocated to car-
-and transit- using populations. This is to be expected, as the
-population that commutes using these modes represents 84.1\% of the
-total population. Differences in \(V_i^m\) values within mode-using
-populations also exist: car-using populations outside of the M-30 region
-appear to have greater \(V_i^m\) values, while some \(i\) areas inside
-the M-30 appear to have higher \(V_i^m\) values for the transit-using
-populations. Overall, the magnitude of \(V_i^m\) values for the bikers
-and walkers are lower than car and transit but the highest
-\(V_i^{bike}\) and \(V_i^{walk}\) values tend to be allotted to \(i\)s
-within the M-30 and \(i\)s that have higher \(V_i^{transit}\) values.
-
-The differences between the mode-using population and their
-mode-specific spatial availability highlights the competitive advantage
-offered to certain modes in certain spatial extents. As summarized in
-the left-most columns in Figure \ref{fig:Fig6}, the `car/motor' and
-`transit' populations represent a combined 95.3\% of the total spatial
-availability in the city. However, the `car/motor' using population is
-allocated disproportionately more \(V_i^m\) than its size compared to
-the transit-using population. The car-using and transit-using population
-is 36.6\% and 47.5\% respectively, but is allocated 48.0\% and 47.3\%
-respectively, of the city's spatial availability. When treating the
-number of opportunities that can be reached as a finite value (total:
-847,574 opportunities), fewer opportunities are spatial availability to
-the lesser competitive modes-using populations, in this case walking and
-cycling. These modes are less competitive as a result of: their lower
-travel impedance values at longer travel times (see Figure
-\ref{fig:Fig4} at travel times beyond \textasciitilde30 minutes); their
-low population values values overall; and higher populations present in
-origins with high motorized mode commuting. These factors all contribute
-to the the car/motor mode being most advantaged in capturing spatially
-available job opportunities overall.
+Some notable differences in the magnitude of \(V_i^m\) occur between
+modes as seen in Figure \ref{fig:Fig5}. The majority of \(V_i^m\) (which
+is to say of spatially available jobs) is allocated to those travelling
+by car and transit. This is to be expected, as the users of these modes
+represent 84.1\% of the total population. Differences in \(V_i^m\)
+values within modes also exist in space: car users outside of the M-30
+region appear to have greater \(V_i^m\) values, while some zones inside
+the M-30 appear to have higher \(V_i^m\) values for transit. Overall,
+the magnitude of \(V_i^m\) values for cyclists and pedestrians are lower
+than for car and transit but the highest values of \(V_i^{bike}\) and
+\(V_i^{walk}\) tend to be allotted to zones within the M-30 and origins
+with higher spatial availability by transit.
+
+The differences between the shares of modes and their shares of
+spatially availabile opportunities highlights the competitive advantage
+afforded by certain modes, although this effect is not uniform in space.
+As seen in the left-most columns of Figure \ref{fig:Fig6}, `car/motor'
+and `transit' together can avail themselves of 95.3\% of all jobs in the
+city. However, `car/motor' has a disproportionate share of \(V_i^m\)
+relative to the population of users of this mode, compared to
+opportunities that are spatially available to transit users. The
+combined population of car and transit users is 36.6\% and 47.5\%
+respectively, but these populations are allocated 48.0\% and 47.3\%
+respectively, of the city's jobs. When treating the number of
+opportunities that can be reached as a finite value (total: 847,574
+opportunities), fewer opportunities are spatially available to slower
+modes (i.e., walking and cycling), even taking into account that their
+share is smaller overall. These modes are at a disadvantage as a result
+of: the travel impedance for longer trips (see Figure \ref{fig:Fig4};
+their low population values values overall; and larger populations in
+origins with high shares of travel by motorized modes. These factors all
+contribute to the the car/motor mode being most advantaged in capturing
+spatially available job opportunities overall.
 
@@
 
-There are spatial variations in the competitive advantage of the
-car-using populations. The proportion of car-using population in the
-Centro is smaller and has higher travel impedance values relative to the
-inputs in other areas and mode-using populations. The LEZ Centro
-implementation further restricts the car-advantage as it shifted more
-than half of all car trips into the LEZ to another mode {[}37{]}. This
-restriction decreased the number of car-using population from \(i\)s
-going into the LEZ Centro (an area with a large number of jobs overall,
-see Figure \ref{fig:Fig2}), thus increasing the mass effect for non-car
-modes and resulting in proportionally higher \(v_i^m\) values for
-non-car modes. As such, the lower amount of access to opportunities by
-car-mode allows more opportunities in the LEZ to be available by
-populations using other modes.
+Despite the big picture, it is important to note that there are spatial
+variations in the competitive advantage of cars. The proportion of
+travel by car in Centro is smaller and has higher travel impedance
+values relative to the inputs in other areas and by other modes. The LEZ
+Centro implementation further restricts the advantage of cars by
+shifting more than half of all car trips into the LEZ to modes {[}41{]}.
+This restriction decreased the number of trips by car from going into
+the LEZ Centro (an area with a large number of jobs overall, see Figure
+\ref{fig:Fig2}), thus increasing the mass effect for non-car modes and
+resulting in proportionally higher \(v_i^m\) values for non-car modes.
+As such, the lower amount of access to opportunities by car allows more
+opportunities in the LEZ to be available to populations using other
+modes.
 
 As summarized in the two right-most columns in Figure \ref{fig:Fig6},
-the proportion of spatial availability allocated to the car-using
-population in the Centro (13.8\% or 5,373 opportunities). As a
-comparative reference, this is less than the proportion of the car-using
-population in the Centro (16.1\%), evidently less then the proportion of
-car-using population in the city, and is the opposite of the trend
-overall (left-most columns) and within the M-30 (middle columns). More
-opportunities are spatial availability to non-car using populations
-within the Centro, particularly transit-using populations (68.5\% of
-spatially available jobs in the Centro despite representing 51.4\% of
-the population in the Centro and 47.5\% in the city overall).
-
-From Figure \ref{fig:Fig6}, it is also summarized that there is a higher
-proportion of opportunities spatially available to walking and cycling
-populations in the Centro than in the City overall and in all areas
-within the M-30. Notably, within the Centro, 1.7\% and 16.1\% of
-opportunities are spatially available to bike and walk modes
-respectively, while their populations represent smaller proportions of
-1.2\% and 14.7\% of the population overall. Though the proportion of
-spatial availability for these mode-using populations is still lower
-than the proportion of mode-using population located in the Centro,
-these modes are more competitive within the Centro than outside of the
-Centro. By restricting the more competitive car mode through the LEZ,
-the advantage in the spatial availability of opportunities afforded to
-the otherwise lesser competitive modes is made apparent.
+the proportion of jobs spatially available to car in Centro is (13.8\%
+or 5,373 opportunities). For reference, this is less than the proportion
+of the car users in Centro (16.1\%), evidently less then the proportion
+of car users in the city, and is the opposite of the overall trend
+(left-most columns) and within the M-30 (middle columns). More
+opportunities are spatially available to non-car users within Centro,
+particularly by transit, a mode with 68.5\% of spatially available jobs
+in Centro, despite representing only 51.4\% of the population in Centro
+and 47.5\% in the city overall.
+
+Figure \ref{fig:Fig6} also shows that there is a higher proportion of
+opportunities that are spatially available to pedestrians and cyclists
+in Centro than in the City overall and in all areas within the M-30.
+Notably, within Centro, 1.7\% and 16.1\% of opportunities are spatially
+available to bike and walk modes respectively, while their populations
+represent smaller proportions of 1.2\% and 14.7\% of the population.
+Although the proportion of opportunities that are spatially available to
+these modes is still lower than the respective shares of the population
+located in the Centro, these modes are more competitive within the
+Centro than anywhere without. By restricting the ability of cars to
+enter Centro, the LEZ also contributes to level the playing field for
+slower modes, in particular cycling and walking.
 
@@
 availability between modes can also be visualized per origin. Figure
-\ref{fig:Fig7} visualizes \(v_i^m\), the spatial availability \(V_i^m\)
-divided by the mode-population. \(v_i^m\) values above 1 are represented
-in increasing red shades, values below 1 are represented in increasingly
-green shades, and values equal to 1 are white. These plots illustrates
-the discussion of the disproportionately high over-allocation of spatial
-availability relative to the mode-using population in many of the
-origins for the car/motor mode (bottom left plot, areas denoted with
-green \(v_i^m\) values above 1). These plots also visualize areas that
-disproportionately capture lower spatial availability (under 1),
-represented in shades of red. It can be observed that the transit-using
-population's spatial availability to jobs is relatively balanced (i.e.,
-many zones are white), while the non-motorized modes \(v_i^m\) values
-are low (under 1) overall.
+\ref{fig:Fig7} shows \(v_i^m\), the spatial availability \(V_i^m\)
+divided by the population that uses \(m\). Values of \(v_i^m\) above 1
+are shown in red and represent more than one spatially available
+opportunity per capita; values below 1 are shown in green, and represent
+less than one spatially available opportunity per capita; values equal
+to 1 are shown in white. These plots drive home the point of the
+disproportionately high over-allocation of spatial availability relative
+to the share of modes in many of the origins for car (bottom left plot,
+areas denoted with green \(v_i^m\) values above 1). These plots also
+show areas with disproportionately few spatially available opportunities
+(under 1). It can be observed that the spatial availability to jobs of
+transit users is relatively well balanced (i.e., many zones are white),
+while that for non-motorized modes is low (under 1) overall.
 
@@
 
-Interestingly, as also represented in Figure \ref{fig:Fig6}, \(v_i^m\)
-for car/motor within and near the LEZ Centro is near or below 1
-(white/red) in Figure \ref{fig:Fig7} while all non-car modes have
-relatively higher \(v_i^m\) values. Though the spatial availability from
-before the LEZ Centro implementation is unknown, Figure \ref{fig:Fig7}
-provides a benchmark for quantifying potential LEZ implementations in
-the future (given 2018 travel conditions). Figure \ref{fig:Fig7} also
-shows that many areas within the M-30 have high (white/green) \(v_i^m\)
-values for car-mode, signaling that the spatial expansion of the LEZ
-Centro stands to increase the spatial availability of jobs for non-car
-mode using populations.
+It is also interesting to note that \(v_i^m\) for car within and near
+LEZ Centro is close to or below 1 (white/red) in Figure \ref{fig:Fig7},
+while in contrast all non-car modes have relatively higher \(v_i^m\)
+values. Since the values are comparable across regions and over time,
+Figure \ref{fig:Fig7} provides a benchmark for quantifying potential LEZ
+implementations in the future. As Figure \ref{fig:Fig7} also shows, many
+areas within the M-30 have high (white/green) \(v_i^m\) values for car,
+but the results for LEZ Centro give reasonable grounds to speculate that
+a spatial expansion of the LEZ to include all areas within the M-30
+stands to increase the spatial availability of jobs for more transit,
+cyclcing and walking.
 
@@
 spatial availability, a singly-constrained measure of competitive
-accessibility {\textbar} {PLOS} {ONE}. {PLOS} {ONE}. 2023; 1--30.
+accessibility \textbar{} {PLOS ONE}. PLOS ONE. 2023; 1--30.
 doi:\href{https://\%20doi.org/10.1371/journal.pone.0278468}{https://
@@
 senior and non-senior residents. International Journal of Health
-Geographics. 2010;9: 1--9. Available:
-\url{http://www.ij-healthgeographics.com/content/9/1/52}}
+Geographics. 2010;9: 1--9. }
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-lunkeModalAccessibilityDisparities2022}{}}%
+\CSLLeftMargin{21. }%
+\CSLRightInline{Lunke EB. Modal accessibility disparities and transport
+poverty in the oslo region. Transportation Research Part D: Transport
+and Environment. 2022;103: 103171.
+doi:\href{https://doi.org/10.1016/j.trd.2022.103171}{10.1016/j.trd.2022.103171}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-paezRelative2010}{}}%
+\CSLLeftMargin{22. }%
+\CSLRightInline{Paez A, Mercado RG, Farber S, Morency C, Roorda M.
+Relative accessibility deprivation indicators for urban settings:
+Definitions and application to food deserts in montreal. Urban Studies.
+2010;47: 1415--1438.
+doi:\href{https://doi.org/10.1177/0042098009353626}{10.1177/0042098009353626}}
+
 \leavevmode\vadjust pre{\hypertarget{ref-campbell2019accessibility}{}}%
-\CSLLeftMargin{21. }%
+\CSLLeftMargin{23. }%
 \CSLRightInline{Campbell KB, Rising JA, Klopp JM, Mbilo JM.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-maharjanSpatialEquityModal2022}{}}%
-\CSLLeftMargin{22. }%
+\CSLLeftMargin{24. }%
 \CSLRightInline{Maharjan S, Tilahun N, Ermagun A. Spatial equity of
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-lunkeModalAccessibilityDisparities2022}{}}%
-\CSLLeftMargin{23. }%
-\CSLRightInline{Lunke EB. Modal accessibility disparities and transport
-poverty in the oslo region. Transportation Research Part D: Transport
-and Environment. 2022;103: 103171.
-doi:\href{https://doi.org/10.1016/j.trd.2022.103171}{10.1016/j.trd.2022.103171}}
-
 \leavevmode\vadjust pre{\hypertarget{ref-grengsJobAccessibilityModal2010}{}}%
-\CSLLeftMargin{24. }%
+\CSLLeftMargin{25. }%
 \CSLRightInline{Grengs J. Job accessibility and the modal mismatch in
@@
 \leavevmode\vadjust pre{\hypertarget{ref-kawabataJobAccessibilityIndicator2006}{}}%
-\CSLLeftMargin{25. }%
+\CSLLeftMargin{26. }%
 \CSLRightInline{Kawabata M, Shen Q. Job accessibility as an indicator of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-kwokUseModalAccessibility2004}{}}%
-\CSLLeftMargin{26. }%
+\CSLLeftMargin{27. }%
 \CSLRightInline{Kwok RCW, Yeh AGO. The use of modal accessibility gap as
@@
 \leavevmode\vadjust pre{\hypertarget{ref-morrisAccessibilityIndicatorsTransport1979}{}}%
-\CSLLeftMargin{27. }%
+\CSLLeftMargin{28. }%
 \CSLRightInline{Morris JM, Dumble PL, Wigan MR. Accessibility indicators
@@
 \leavevmode\vadjust pre{\hypertarget{ref-weibullAxiomaticApproachMeasurement1976}{}}%
-\CSLLeftMargin{28. }%
+\CSLLeftMargin{29. }%
 \CSLRightInline{Weibull JW. An axiomatic approach to the measurement of
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-OECDFrameworks2013}{}}%
+\CSLLeftMargin{30. }%
+\CSLRightInline{OECD. Frameworks and sector policies for urban
+development in chile. OECD urban policy reviews, chile 2013. 2013.
+doi:\url{http://dx.doi.org/10.1787/9789264191808-en}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-martensFair2021}{}}%
+\CSLLeftMargin{31. }%
+\CSLRightInline{Martens K, Golub A. A fair distribution of
+accessibility: Interpreting civil rights regulations for regional
+transportation plans. Journal of Planning Education and Research.
+2021;41: 425--444.
+doi:\href{https://doi.org/10.1177/0739456x18791014}{10.1177/0739456x18791014}}
+
 \leavevmode\vadjust pre{\hypertarget{ref-josephMeasuringPotentialPhysical1982}{}}%
-\CSLLeftMargin{29. }%
+\CSLLeftMargin{32. }%
 \CSLRightInline{Joseph AE, Bantock PR. Measuring potential physical
@@
 \leavevmode\vadjust pre{\hypertarget{ref-taoInvestigatingImpactsPublic2020a}{}}%
-\CSLLeftMargin{30. }%
+\CSLLeftMargin{33. }%
 \CSLRightInline{Tao Z, Zhou J, Lin X, Chao H, Li G. Investigating the
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-wilsonFamilty1971}{}}%
+\CSLLeftMargin{34. }%
+\CSLRightInline{Wilson AG. A family of spatial interaction models, and
+associated developments. Environment and Planning A. 1971;3: 1--32. }
+
 \leavevmode\vadjust pre{\hypertarget{ref-carpentieriMultimodalAccessibilityPrimary2020}{}}%
-\CSLLeftMargin{31. }%
+\CSLLeftMargin{35. }%
 \CSLRightInline{Carpentieri G, Guida C, Masoumi HE. Multimodal
@@
 \leavevmode\vadjust pre{\hypertarget{ref-margaryanLowEmissionZones2021}{}}%
-\CSLLeftMargin{32. }%
+\CSLLeftMargin{36. }%
 \CSLRightInline{Margaryan S. Low emission zones and population health.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-verbeekJustManagementUrban2022}{}}%
-\CSLLeftMargin{33. }%
+\CSLLeftMargin{37. }%
 \CSLRightInline{Verbeek T, Hincks S. The {``just''} management of urban
@@
 \leavevmode\vadjust pre{\hypertarget{ref-espanaPlanNacionalIntegrado2020}{}}%
-\CSLLeftMargin{34. }%
+\CSLLeftMargin{38. }%
 \CSLRightInline{España. Plan nacional integrado de energía y clima
@@
 \leavevmode\vadjust pre{\hypertarget{ref-espanaResolucion10Enero2020}{}}%
-\CSLLeftMargin{35. }%
+\CSLLeftMargin{39. }%
 \CSLRightInline{España. Resolución de 10 de enero de 2020, de la
@@
 \leavevmode\vadjust pre{\hypertarget{ref-barcelonaGUIATECNICAPARA2021}{}}%
-\CSLLeftMargin{36. }%
+\CSLLeftMargin{40. }%
 \CSLRightInline{Barcelona. {GUÍA} {TÉCNICA} {PARA} {LA} {IMPLEMENTACIÓN}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-tarrinoortizAnalyzingImpactLow2022}{}}%
-\CSLLeftMargin{37. }%
+\CSLLeftMargin{41. }%
 \CSLRightInline{Tarriño-Ortiz J, Gómez J, Soria-Lara JA, Vassallo JM.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-comunidaddemadridResultadosEDM20182020}{}}%
-\CSLLeftMargin{38. }%
+\CSLLeftMargin{42. }%
 \CSLRightInline{Comunidad de Madrid. Resultados de la {EDM} 2018 - Datos
@@
 \leavevmode\vadjust pre{\hypertarget{ref-lopez_2017_spatial}{}}%
-\CSLLeftMargin{39. }%
+\CSLLeftMargin{43. }%
 \CSLRightInline{Lopez FA, Paez A. Spatial clustering of high-tech
@@
 \leavevmode\vadjust pre{\hypertarget{ref-horbachov_theoretical_2018}{}}%
-\CSLLeftMargin{40. }%
+\CSLLeftMargin{44. }%
 \CSLRightInline{Horbachov P, Svichynskyi S. Theoretical substantiation
@@
 \leavevmode\vadjust pre{\hypertarget{ref-batista_estimation_2019}{}}%
-\CSLLeftMargin{41. }%
+\CSLLeftMargin{45. }%
 \CSLRightInline{Batista SFA, Leclercq L, Geroliminis N. Estimation of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-fitdistrplus_2015}{}}%
-\CSLLeftMargin{42. }%
+\CSLLeftMargin{46. }%
 \CSLRightInline{Delignette-Muller ML, Dutang C. {fitdistrplus}: An {R}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-reggianiAccessibilityImpedanceForms2011}{}}%
-\CSLLeftMargin{43. }%
+\CSLLeftMargin{47. }%
 \CSLRightInline{Reggiani A, Bucci P, Russo G. Accessibility and
@@
 \leavevmode\vadjust pre{\hypertarget{ref-soukhovTTS2016RDataSet2023}{}}%
-\CSLLeftMargin{44. }%
+\CSLLeftMargin{48. }%
 \CSLRightInline{Soukhov A, Páez A. {TTS}2016R: A data set to study

### 2023-11-09 Journal-manuscript/Editorial correspondance/Response-to-reviewers_R1.Rmd
Antonio completed editions of revised version of paper; now it is going back to Anastasia for proof-reading

@@
 ---
-title: "Response-to-reviewers_R1"
+title: "PLoS ONE submission: Response to reviewers"
 output: pdf_document
 date: "2023-10-18"
+bibliography:
+  - "`r system('kpsewhich ../../bibliography/bibliography.bib', intern=TRUE)`"
 ---
@@
 No doubt, the revised version submitted is more articulate than the first.
-Our responses to all comments from each reviewer are shown below in \textcolor{blue}{blue} and directly quoted the revised portion of the manuscript as relevant}
+Our responses to all comments from each reviewer are shown below in \textcolor{blue}{blue}.
 
@@
 \textcolor{blue}{Thank you for this comment.
-First we would like to bring to your attention the editorial philosophy of PLoS ONE, according to which editors "make decisions on submissions based on scientific rigor, regardless of novelty," (see https://journals.plos.org/plosone/s/editorial-and-peer-review-process).
+We would like to bring to your attention the editorial philosophy of PLoS ONE, according to which editors "make decisions on submissions based on scientific rigor, regardless of novelty," (see https://journals.plos.org/plosone/s/editorial-and-peer-review-process).
 Nonetheless we wish to respond to this comment.}
@@
 You appear to be under the misaprehension that the main difference with accessibility measures is to incorporate "the ratio of travel frequencies as weight in travel costs".
-This is statement is not correct.
-The main difference is a proportional allocation mechanism that we introduced for spatial availability, and that is extended in this paper to allocate opportunities based on the proportion of travelers by different modes.
+This statement is not correct.
+The main difference is the proportional allocation mechanism that we introduced for spatial availability, and that is extended in this paper to allocate opportunities based on the proportion of travelers by different modes.
 This is not the same as "weighting the travel cost".
@@
 
-\textcolor{blue}{In response to this comment we have edited the "Discussion and conclusions" section to more clearly describe the advantages that multimodal spatial availability offers when considering the accessibility to opportunities by different modes.
+\textcolor{blue}{In response to this comment we have edited the paper to more clearly describe the advantages that multimodal spatial availability offers when considering the accessibility to opportunities by different modes.
 In particular, proportional allocation of opportunities using mode-specific impedance functions means that the travel-cost-advantage of each mode can be analyzed.
@@
 In this paper we also try to demonstrate throughout the manuscript what our measure does that others can't.
-For example, in the Discussion and Conclusions we state that: "With spatial availability, the number of opportunities that are captured (of the total opportunities in the region) by each mode can be individually calculated.
-From there, the difference between how many spatially available opportunities one mode captures versus another can be investigated.
-This is the advantage of the spatial availability measure, particularly its multimodal extension.".
-Since the spatial availability values result from proportional allocation, each value is a proportion of the total opportunities.
+Since spatial availability values result from proportional allocation, each value is a proportion of the total number of opportunities.
 Put another way, we can compare the proportion of opportunities available to users of each mode, to users at each zone, to users of each mode by zone, and so on, and the values relate directly to the total opportunities in the region.
 This also allows us to calculate values per capita that serve as benchmark values, as shown in the empirical example.}
-
-\textcolor{blue}{In the revised version of the paper, the advantages of spatial availability are stated more explicitly, including in the "Introduction", "A review of multimodal accessibility measures", and "Discussion and Conclusions".}
-
-\textcolor{blue}{Introduction:}
-
-\textcolor{blue}{The question of how much one mode-using population can access at the expense of another mode-using population is a pertinent equity question in the evaluation of policy scenarios, especially those that impact modes differently.
-For instance, consider the impact of a low emission zone (LEZ).
-A LEZ is a policy of spatial and modal discrimination: the circulation of vehicles that are excessively polluting are restricted in specific regions.
-Depending on the level of vehicle restriction, LEZ have been shown to significantly improve air quality [@holmanReviewEfficacyLow2015] and air-quality related health outcomes[@margaryanLowEmissionZones2021].
-From the perspective of accessibility to opportunities, LEZ explicitly lower the access that restricted-vehicle using populations have to opportunities within the LEZ.
-However, if conceptualizing opportunities in a region as finite, one populations' lose is another populations' gain.
-The populations using non-restricted modes (e.g., walking, cycling, transit, and less-polluting vehicles) to travel into LEZ will have more access than before the LEZ implementation.}
-
-\textcolor{blue}{This evaluation is especially urgent as LEZ are in effect in cities globally; their reception [@tarrinoortizPublicAcceptabilityLow2021] and impact on certain activity types such as shopping [@tarrinoortizImpactLowEmission2023] have been mixed.
-They may also be having negative impacts on disadvantaged populations who have become mobility-restricted [@devrijNooneVisitsMe2022; @verbeekJustManagementUrban2022; @liottaWhatDrivesInequalities2023].
-Measures that evaluate the accessibility of modes given both constrained and competitive considerations are lacking in the literature, but are needed, to evaluate such an intervention's impact on accessibility gains and losses within a finite conceptualization.}
-
-\textcolor{blue}{A review of multimodal accessibility measures:}
-
-\textcolor{blue}{- extended the explanation of Hasen-type measure and the consideration of competition in accessibility (shen-type measure) with additional examples from the literature.
-See the revised sections.}
-
-\textcolor{blue}{Discussion and Conclusions:}
-
-\textcolor{blue}{XXXXXXXXXXX}
 
@@
 The data we work with represent the most recent and most complete travel survey conducted for the region to date.
-We are not completely sure where you got the figure of "30,000 trips" (which you consider significantly limited); presumably you are citing the maximum 'opportunity' or 'population' numbers in Figure 2 or 3.
-The most intensely coloured TAZ have ~30,000 jobs while the least intensely coloured have ~1,000 or fewer (figure 2).
+We are not completely sure where you got the figure of "30,000 trips" (which you consider significantly limited); presumably you are citing the maximum 'opportunity' or 'population' numbers in Figures 2 or 3.
+The most job-rich TAZs have ~30,000 jobs while the least have ~1,000 or fewer (Figure 2).
 In fact, there are 847,574 jobs in the city, which is also the sum of the total spatial availability in our analysis, as well as the number of potential trips to work.}
 
-\textcolor{blue}{The "typicallity" of a travel survey is information about travel on a typical day, usually during a period of maximum demand (i.e., not during the summer vacation).
-This is standard practice for these surveys.}
-
-\textcolor{blue}{We have updated the manuscript to provide further information about the survey for the benefit of readers:}
-
-\textcolor{blue}{Introduction:}
-
-\textcolor{blue}{In Section 3, the spatial availability of four mode categories is calculated for the city of Madrid, Spain after the 2017 LEZ is implemented.
-Using mode-specific home-to-work origin-destination 2018 travel survey as empirical data, we demonstrate inequities of spatial availability within and outside the LEZ for each sub-population using car, transit, cycling and walking modes.}
-
-\textcolor{blue}{Multimodal data and methods:}
-
-\textcolor{blue}{The 2018 Community of Madrid travel survey [@comunidaddemadridResultadosEDM20182020] is the source of data for this empirical example: it is a representative survey that reflects a snap-shot of the travel patterns for a typical day of the working week in 2018.
-Specifically, it includes 222,744 trips taken from a representative sample of 85,064 households across the traffic analysis zones (TAZ) representing the Community of Madrid (population of 6,507,184 over 3 years old) through population elevation factors.}
-
-\textcolor{blue}{In this empirical example, we only use a sub-set of the survey, specifically home-to-work trips, by all modes, to demonstrate multimodal spatial availability of employment opportunities.
-The total population who travels to work from an origin (i.e., the worker), and the destinations (i.e., the employment opportunity), for the City of Madrid (a sub-set of the Community of Madrid) is visualized in Figure \ref{fig:Fig2} and Figure \ref{fig:Fig3}.
-Both figures are displayed at the level of TAZ (the $i$ and $j$ zones) that correspond to the survey....}
-
-\textcolor{blue}{Captions to Figure 2 and Figure 3:}
-
-\textcolor{blue}{- Jobs $O_j$ taken by people living and working in Madrid as reported by the home-to-work flows in the 2018 travel survey.}
-\textcolor{blue}{- Figure 3: Population living and working in Madrid by four summarized modal categories $P^m_i$ as reported by the home-to-work flows in the 2018 travel survey.
-}
+\textcolor{blue}{With respect to The "typicallity" of the data, travel surveys are designed to provide information about personal travel on a typical day, usually during a period of maximum demand (i.e., not during the summer vacation).
+This is standard practice for these surveys, and the one for Madrid is no different in this respect.}
 
@@
 
-\textcolor{blue}{Thanks for this comment.}
-
-\textcolor{blue}{It is somewhat puzzling that "30,000 trips" would be significantly limited (as per your comment \#2), and at the same time be "a large amount of data".
+\textcolor{blue}{Thank you for this comment.
+It is somewhat puzzling that "30,000 trips" would be significantly limited (as per your comment \#2), and at the same time be "a large amount of data".
 We find that this comment in particular is not actionable due to its vagueness.
 There are no meaningful responses to "somewhat one-sided" when "somewhat" and "one-sided" do not quantify or refer to a particular side.
-We used all modes available in the survey, and examined the results from the perspective of each mode.
+We studied the modes available in the region, and examined the results from the perspective of each mode.
 What other perspectives do you suggest?
@@
 
-\textcolor{blue}{Nevertheless, with the additional detail added to the manuscript to address your and other reviewers' comments, we believe that overall clarity has been improved.}
+\textcolor{blue}{Despite this comment being nonactionable, we believe that with the additional detail added to the manuscript to address your and other reviewers' comments, that overall clarity has been improved.}
 
@@
 
+\newpage
+
 ## Reviewer #2:
@@
 
-\textcolor{blue}{Thank you for your thoughtful review of our paper. Your comments were helpful to improve the clarity of the research.}
+\textcolor{blue}{Thank you for your thoughtful review of our paper. Your comments were very helpful to improve the clarity of the research.}
 
@@
 We would begin by noting that there are many types of opportunities that are clearly mutually exclusive due to competition.
-For this paper we focused on jobs, but there are many others, such as beds at hospitals, doctors at clinics, seats at schools, and so on.
-Thus, even if the measure was applicable only to opportunities subject to competition, there are numerous applications to it.}
+For this paper we focused on jobs, but there are many others, such as beds at hospitals, seats at schools, and so on.
+Thus, even if the measure was applicable only to exclusive opportunities, there are numerous applications to choose from.}
 
-\textcolor{blue}{That said, we have grappled for some time with the idea that some opportunities might not be subject to competition.
-Our current thinking, after much consideration, is that in practice every type of opportunity is subject to congestion or capacity constraint.
+\textcolor{blue}{That said, we have grappled for some time with the question of what types of opportunities are best analyzed using spatial availability.
+Our current thinking on this matter, after much consideration, is that in practice every type of opportunity, even when not clearly exclusive, is subject at least to congestion or capacity constraint.
 For instance, green spaces are often considered non-competitive, however, standards for the provision of such amenties are provided in the form of units of amenity per capita.
-For example, Natural England recommends an Accessible Natural Greenspace Standard such that the minimum supply of space is one ha of statutory Local Nature Reserves per thousand population}^[see  https://redfrogforum.org/wp-content/uploads/2019/11/67-Nature-Nearby%E2%80%99-Accessible-Natural-Greenspace-Guidance.pdf] \textcolor{blue}{.
-Similarly, the World Health Organization (cited in OECD, 2013) recommends that cities provide a minimum of 9 m2 of green area per inhabitant.}^[see https://doi.org/10.1787/9789264191808-en] \textcolor{blue}{For our purposes, standards of this type translate into "how much of this resource can we give to each individual that is not allocated to anyone else?".
-For this reason, although Local Nature Reserves have large capacities - they still have a capacity - and thus spatial availability can be used to measure constrained accessibility to this sort of opportunity.
-As standards are emphasized in the planning literature, in particular for fairness in transportation (see Martens and Golub, 2021), spatial availability analysis can be used to assess standards.
-We believe that as other researchers discover this new measure, other applications will be found.}
+For example, A case in point is the Ile-de-France region, a jurisdiction that suggested in a major planning document of 2013 that at the municipal level at least $10m^2$ of public green space should be supplied per inhabitant} [@liottaPlanning2020]\textcolor{blue}{.
+But green spaces are not evenly distributed, which means that who has access to them hinges on where they are and how easy is to reach them.
+Formulating the provision of amenities in these terms in not rare.
+For example, Natural England recommends an Accessible Natural Greenspace Standard such that the minimum supply of space is one ha of statutory Local Nature Reserves \textit{per thousand population}}^[see  https://redfrogforum.org/wp-content/uploads/2019/11/67-Nature-Nearby%E2%80%99-Accessible-Natural-Greenspace-Guidance.pdf] \textcolor{blue}{.
+Similarly, the World Health Organization (cited in OECD, 2013) recommends that cities provide a minimum of 9 $m^2$ of green area per inhabitant.}^[see https://doi.org/10.1787/9789264191808-en] \textcolor{blue}{For our purposes, standards of this type translate into "how much of this resource is available to one individual that has not been claimed by anyone else?".
+Green spaces often have large capacities, but they still have a capacity, and it is not the same for a person to have access to $5m^2$ of \textit{uncongested} green space than to $15m^2$.
+This difference is in fact a matter of justice} [@laraSpace2015; @liottaPlanning2020]\textcolor{blue}{.
+Constraining accessibility is in this way a useful way to evaluate the congested availability of any type of opportunity.
+As standards are emphasized in the planning literature, in particular for fairness in transportation} [see @martensFair2021]\textcolor{blue}{, spatial availability analysis can be used to assess standards.
+We are convinced that as other researchers discover this new approach to measuring accessibility many other applications will be found.}
 
@@
 \textcolor{blue}{This is an excellent point.
-Although spatial availability is not a model of modal split, it is certainly amenable to changes in the use of various modes.
-This could be implemented as follows: the results of a modal split model are used to estimate new modal shares, which in turn are used to recalculate spatial availability.
-The results then can be compared to the baseline scenario.
+Accessibility measures, including spatial availability, are not meant to function as modal split models, but they is certainly amenable to analyze changes in the accessibiity landscape if different modal shares are used as inputs.
+This is similar to the case of destination choice: accessibility measures do not model this, but changes in destination choices can be incorporated via how they affect the impedance function.
+In terms of how this could be implemented, the process can be sketched as follows: the results of a modal split model are used to estimate new modal shares, which in turn are used to recalculate spatial availability.
+The new values then can be compared to the baseline scenario.
 In other words, the framework for spatial availability is sufficiently flexible to take in not only mode-specific travel impedance functions, but also the proportions of the populations using each mode.}
@@
 
-\textcolor{blue}{Your comments have been very helpful to improve the clarity of the paper, as well as the scope of what we do now, and directions for future research.}
+\textcolor{blue}{Your comments have been very helpful to improve the clarity of the paper, as well as the scope of what we do, as well as directions for future research.
+For example, we now note in the "Discussion and conclusions" that "our example dealt with differences in travel by mode only, but it is possible to think of the intersection between mode of travel and different types of travelers.
+This would expand the number of sub-populations in the analysis from, say, $m=M$ (modes) to $m = M\cdot Q$ (modes times population segments), each with their own characteristic impedance function.
+Evaluations of this kind will be especially relevant as LEZ are implemented in cities globally, and the question of their impact on disadvantaged populations who have become mobility-restricted increasingly come to the fore} [@devrijNooneVisitsMe2022; @verbeekJustManagementUrban2022; @liottaWhatDrivesInequalities2023]\textcolor{blue}{."}
 
@@
 In the original version of the abstract and introduction we gave the misleading impression that we would analyze changes in the system, when in reality our intent was to demonstrate the application of the measure in an empirical example.
-We do plan to study changes in the system, but this requires more work than can be presented in a single paper, partly for the reasons that you identified above (modal shifts and the calculation of spatial availability for two different scenarios, as well as the analysis of the differences between them.
-We will do this in a future applied paper.}
+We do plan to study changes in the system, but this requires more work than can be presented in a single paper, partly for the reasons that you identified above (modal shifts and the calculation of spatial availability for two different scenarios, as well as the analysis of the differences between them).
+We plan to do this in a future paper focused on the policy instead of the presentation of a new method.}
 
@@
 \textcolor{blue}{Fixed! Thank you.}
+
+\newpage
 
@@
 
-\textcolor{blue}{Great comment. The short answer is that whenever a destination can be reached by more than one mode, users of those modes are in competition for the opportunities there.}
+\textcolor{blue}{This is an excellent comment.
+The short answer is that whenever a destination can be reached by more than one mode, users of those modes are in competition for the opportunities there.}
 
-\textcolor{blue}{In the paper we have tried to improve the discussion to make this point more clear.
+\textcolor{blue}{In this revision we tried to improve the discussion to make this point more clear.
 The impedance functions for all four modes are not the same.
@@
 To follow your example, someone who cannot walk to work because their job is 20 km from where they live, will not compete for that job against people from their same origin who do walk.
-However, it the place where their work is can be reached by walking by anyone from any other origin (someone who lives closer to that one destination), they would compete for the same opportunity.}
+However, if the place where their work is can be reached by anyone who can walk from other origins (say, someone who lives closer to that destination), they would be in competion for the same opportunity.}
 
@@
 This assumption, that all populations, no matter their mode, are competing for the same finite set of opportunity, is part of spatial availability.
-We've made this more clear in the text.}
-
-\textcolor{blue}{Multimodal data and methods:}
-
-\textcolor{blue}{Notably, plots in Figure \ref{fig:Fig4} can be observed across modes if curious to know what trips are more likely to occur using what mode depending on trip length (as informed by the observed 2018 survey).
-Trips >5 minutes do not occur frequently for any mode, as such trips with short trip lengths are assigned a lower travel impedance $f^m(c_{ij}^m)$ value.
-One reason that >5 minute trips do not frequently occur as a result of land-use (residential and job mismatch): not everyone lives in immediate proximity to their workplace.
-In terms of the non-motorized modes, shorter trips occurred more frequently overall for walking populations, particularly around 15 minute lengths, so a trip of approximately 15 minutes is assigned the highest $f^m(c_{ij}^m)$.
-For biking populations, longer travel times are more common so though the highest $f^m(c_{ij}^m)$ value also corresponds to approximately 15 minutes, the curve is more spread out and values decrease less rapidly at longer travel times than for the walk mode curve.
-A similar trend can be observed for the motorized modal options where transit mode is more spread out than car/motor mode.
-All in all, these observations demonstrate that, based on a given mode and travel time, a trip is more or less likely to occur and is accordingly represented in the cost of travel balancing factor $F_{ij}^m$ for $V_i^m$.}
+We've made an effort to make this more clear in the text.}
 
@@
  <!-- and I must change the labels on the legends to be ranges -->
+
+\newpage
+
+## References

### 2023-11-09 Journal-manuscript/Journal-manuscript.Rmd
Antonio completed editions of revised version of paper; now it is going back to Anastasia for proof-reading

@@
 # plos.csl can be download and used locally
-csl: http://www.zotero.org/styles/plos
+csl: plos.csl
 ---
@@
   comment = '', 
-  out.width = "1\\linewidth")
+  out.width = "0.85\\linewidth")
 
@@
                      after_stat(density)),
-                 bins = 20) +
+                 fill = "gray70",
+                 bins = 15) +
   geom_line(data = tld_theoretical,
@@
 ```
+
 ```{r define-population-balancing-factor}
@@
 ```
+
 ```{r create-bounding-boxes}
@@
 ```
+
 ```{r merging-borders-for-plotting, warning=FALSE}
@@
           colour=NA,
-          fill="grey30") +
+          fill="grey50") +
   geom_sf(data = SA_im_zn208,
@@
 
-  scale_fill_fermenter(palette = "Purples", #legend scale bar
-                       na.value = "grey90",
-                       direction = 1,
+  scale_fill_viridis_c(option = "mako", #legend scale bar
+                       na.value = "gray50",
+                       direction = -1,
                        name = expression(P['i']^{m}),
@@
                        breaks = c(38, 360,2229,3750, 6768, 18421),
-                       labels = c("39 (min.)", "360 (20% prob.)", "2,229 (40% prob.)",
-                                  "3,750 (60% prob.)", "6,768 (80% prob.)","18,421 (max.)"))+
+                       labels = c("< 39", "(39  - 360]", "(360 - 2,229]",
+                                  "(2,229 - 3,750]", "(3,750 - 6,768]","(6,768 - 18,421]"))+
   geom_sf(data = borders,
           aes(colour=Border),
-          size = 0.5,
+          size = 2,
           fill=NA) +
-  scale_color_manual(values = c( "red", "cyan", "black"),
+  scale_color_manual(values = c("gray90", "gray60", "black"),
                        name = element_blank())+
@@
           colour=NA,
-          fill="grey30") +
+          fill="grey50") +
   geom_sf(data = SA_i_zn208,
@@
           color = NA) +
-  scale_fill_fermenter(palette = "Oranges", #legend scale bar
-                       direction = 1,
+  scale_fill_viridis_c(option = "magma", #legend scale bar
+                       direction = -1,
                        name = expression(O['j']),
-                       limits = c(1201, 39811),
-                       breaks = c(1201, 5599,8703,14944,22377, 39811),
-                       labels = c("1,201 (min.)", "5,599 (20% prob.)", "8,704 (40% prob.)",
-                                  "14,944 (60% prob.)", "22,377 (80% prob.)","39,811 (max.)"))+
+                       limits = c(1202, 39811),
+                       breaks = c(1202, 5599,8703,14944,22377, 39811),
+                       labels = c("< 1,202", "(1,202 - 5,599]", "(5,599 - 8,704]",
+                                  "(8,704 - 14,944]", "(14,944 - 22,377]","(22,377 - 39,811]"))+
   geom_sf(data = borders,
           aes(colour=Border),
-          size = 0.5,
+          size = 2,
           fill=NA) +
-  scale_color_manual(values = c( "red", "cyan", "black"),
+  scale_color_manual(values = c("gray90", "gray60", "black"),
                        name = element_blank())+
@@
 pop_i_m_plot
-ggsave("images/im_populations_zn208_plot.png")
+ggsave("images/im_populations_zn208_plot.png",
+       width = 7,
+       units = "in")
 ```
@@
 jobs_i_plot
-ggsave("images/i_jobs_zn208_plot.png")
+ggsave("images/i_jobs_zn208_plot.png",
+       width = 7,
+       units = "in")
 ```
@@
           color = NA) +
-  scale_fill_fermenter(palette = "Blues", #legend scale bar
+  scale_fill_viridis_c(option = "magma", #legend scale bar
                        name = expression(V["i"]^{m}),
-                       direction = 1,
+                       direction = -1,
                        limits = c(0.07, 36531),
-                       breaks = c(0.07, 81, 670, 6409, 7727, 36531),
-                       labels = c("0.07 (min.)", "81 (20% prob.)", "670 (40% prob.)", "2,707 (60% prob.)", "7,727 (80% prob.)", "36,531 (max.)"))+
+                       breaks = c(0.1, 81, 670, 6409, 7727, 36531),
+                       labels = c("< 0.1", "(0.1 - 81]", "(81 - 670]", "(670 - 2,707]", "(2,707 - 7,727]", "(7,727 - 36,531]"))+
     geom_sf(data = borders,
@@
           fill=NA) +
-  scale_color_manual(values = c( "red", "cyan", "black"),
+  scale_color_manual(values = c("gray90", "gray60", "black"),
                        name = element_blank())+
@@
           colour=NA,
-          fill="grey30") +
+          fill="grey50") +
   geom_sf(data = OD_Si_zn208,
@@
                        name = expression(S["im"]),
-                       na.value = "grey80",
+                       na.value = "gray50",
                        direction = 1)+
@@
 S_im_plot_zn208_plot
-ggsave("images/Hansen_im_V_zn208_plot.png")
+ggsave("images/Hansen_im_V_zn208_plot.png",
+       width = 7,
+       units = "in")
 ```
@@
 SA_im_V_zn208_plot
-ggsave("images/SA_im_V_zn208_plot.png")
+ggsave("images/SA_im_V_zn208_plot.png",
+       width = 7,
+       units = "in")
 ```
@@
           color = NA) +
-  scale_fill_gradient2(low="red", high="green", mid = "white", midpoint = 1,
+  scale_fill_gradient2(low="#F05B12", high="#455BCD", mid = "#46F884", midpoint = 1,
                        name = expression(v["i"]^{m}),
                        limits = c(0.001106867, 2.086736720),
-                       breaks = c( 0.12, 0.31, 0.80, 1.11, 2.086736720),
-                       labels = c(">0.12 (20% prob.)", "0.31 (40% prob.)", "0.80 (60% prob.)", "1.11 (80% prob.)", "2.09 (max.)"))+
+                       breaks = c(0.001106867,
+                                  SA_im_zn208 |> filter(v_im <= 1) |> pull(v_im) |> median(),
+                                   1, 
+                                   SA_im_zn208 |> filter(v_im >= 1) |> pull(v_im) |> median(),
+                                   2.086736720),
+                       labels = c("< 0.23", "[0.23 - 1)", "[1 -1.17)", "[1.17 - 2.08)", "2.08 (max)"))+
     geom_sf(data = borders,
@@
 ```
+
 ```{r, eval=FALSE}
 SA_im_v_zn208_plot
-ggsave("images/SA_im_vv_zn208_plot.png")
+ggsave("images/SA_im_vv_zn208_plot.png",
+       width = 7,
+       units = "in")
 ```
@@
 ```
+
 ```{r, eval=FALSE}
@@
 ```{r}
-#comparision plots of spatial availability vs. population by mode for the full region, just m-30, and just centro
+#comparison plots of spatial availability vs. population by mode for the full region, just m-30, and just centro
 comps_plot <- ggplot(data = modal_V_comps_4plot,
        aes(y=perc, x=Measure, fill=Mode, label= perc %>% percent(accuracy = 0.1) )) +
-  scale_x_discrete(limits = c("Population by Mode\n(Total: 847,574)","Spatial Availability\n(Total: 847,574)",
-                              "Population by Mode\n(Total: 243,472)","Spatial Availability\n(Total: 217,219)",
-                              "Population by Mode\n(Total: 44,891)","Spatial Availability\n(Total: 38,942)"),
+  scale_x_discrete(limits = c("Population\n by Mode\n (Total: 847,574)","Spatial Availability\n by Mode\n(Total: 847,574)",
+                              "Population\n by Mode\n (Total: 243,472)","Spatial Availability\n by Mode\n(Total: 217,219)",
+                              "Population\n by Mode\n (Total: 44,891)","Spatial Availability\n by Mode\n(Total: 38,942)"),
                    position = "bottom") +
   geom_col() +
-  scale_fill_manual(values = c("lightgreen","sienna", "coral", "skyblue"))+
+  scale_fill_manual(values = c("#FDE725","#1E9B8A", "#85D54A", "#2D708E"))+
   geom_text(position = position_stack(vjust=0.5), color="black")+
@@
           fill= NA) +
-  scale_color_manual(values = c( "red", "cyan", "black"),
+  scale_color_manual(values = c("gray90", "gray60", "black"),
                      name = element_blank())+
@@
           size = 0.5,
-          col ="cyan",
-          fill="cyan") +
+          col ="black",
+          fill="grey60") +
   theme_void() +
@@
           size = 0.5,
-          col ="red",
-          fill="red") +
+          col ="black",
+          fill="gray90") +
   theme_void() +
@@
                         nrow=2, heights = c(1,2.5) )
-ggsave("images/modal_V_comps_4plot.png", modal_V_comps_4plot_plot)
-```
-<!-- Maybe change title from:
-Multimodal spatial availability: a singly-constrained measure of competitive accessibility considering multiple modes
-to:
-Multimodal spatial availability: a singly-constrained measure of competitive modal accessibility 
--->
+ggsave("images/modal_V_comps_4plot.png", modal_V_comps_4plot_plot,
+        width = 7,
+       units = "in")
+```
+
 # Introduction
@@
 Accessibility is a key concept in the analysis of land use and transportation systems [e.g., @hansenHowAccessibilityShapes1959; @geursAccessibility2004; @paezPositive2012], and one that is coming of age from the perspective of planning too [see, _inter alia_, @handyAccessibility2020; @levinsonTransportAccessManual2020; @siddiqToolsTradeAssessing2021; @yanAccessibilityBasedPlanningAddressing2021; @elGeneidyMaking2022].
-Beginning with the work of Hansen [@hansenHowAccessibilityShapes1959], accessibility measures have been widely used to evaluate the efficiency of transportation systems when combined with the distribution of opportunities in space [e.g., ].
+Beginning with the work of Hansen [@hansenHowAccessibilityShapes1959], accessibility measures have been widely used to evaluate the efficiency of transportation systems when combined with the distribution of opportunities in space [@shiLiterature2020].
 As such, it is a holistic measure of spatial systems that measures the ease of reaching destinations [@handyMeasuring1997; @kwanSpace1998].
@@
 The populations using non-restricted modes (e.g., walking, cycling, transit, and less-polluting vehicles) to travel into LEZ will have _more_ access than before the LEZ implementation.
-
-This evaluation is especially urgent as LEZ are in effect in cities globally; their reception [@tarrinoortizPublicAcceptabilityLow2021] and impact on certain activity types such as shopping [@tarrinoortizImpactLowEmission2023] have been mixed.
-They may also be having negative impacts on disadvantaged populations who have become mobility-restricted [@devrijNooneVisitsMe2022; @verbeekJustManagementUrban2022; @liottaWhatDrivesInequalities2023].
-Measures that evaluate the accessibility of modes given both _constrained_ and _competitive_ considerations are lacking in the literature, but are needed, to evaluate such an intervention's impact on accessibility gains and losses within a _finite_ conceptualization.
  -->
@@
 There are only so many school-seats, hospital capacity, employment opportunities, etc., in a region and if one person reaches an opportunity, it is taken: the supply of an opportunity and the demand for that opportunity are two components of accessibility.
-These are clear examples of opportunities that are clearly competitive.
+These are clear examples of opportunities that are unambiguously competitive.
 But we would go as far as to argue that every type of opportunity is subject to congestion or capacity constraint, even when the opportunities are conventionally seen as non-competitive.
@@
 For instance, standards for providing green spaces are often stated in the form of _exclusive access_, in units of amenity per capita.
-A case in point is Natural England, an organization that recommends an Accessible Natural Greenspace Standard such that the minimum supply of space is one ha of statutory Local Nature Reserves per thousand population^[see  https://redfrogforum.org/wp-content/uploads/2019/11/67-Nature-Nearby%E2%80%99-Accessible-Natural-Greenspace-Guidance.pdf].
+A case in point is the Ile-de-France region, a jurisdiction that suggested in a major planning document of 2013 that at the municipal level at least $10m^2$ of public green space should be supplied _per inhabitant_ [@liottaPlanning2020].
+But green spaces are not evenly distributed, which means that who has access to them hinges on where they are and how easy is to reach them.
+This formulation of provision of amenities is not unusual.
+For example, Natural England, an organization that recommends an Accessible Natural Greenspace Standard such that the minimum supply of space is one ha of statutory Local Nature Reserves per thousand population^[see  https://redfrogforum.org/wp-content/uploads/2019/11/67-Nature-Nearby%E2%80%99-Accessible-Natural-Greenspace-Guidance.pdf].
 Similarly, the World Health Organization [cited in @OECDFrameworks2013] recommends that cities provide a minimum of 9$m^2$ of green area per inhabitant.
-For our purposes, standards of this type translate into "how much of a resource can be offered to each individual that is not allocated to anyone else?".
-Green spaces often have large capacities, but they still have a capacity.
+For our purposes, standards of this type translate into "how much of this resource is available to one individual that has not been claimed by anyone else?".
+Green spaces often have large capacities, but they still have a capacity, and it is not the same a person to have access to $5m^2$ of _uncongested_ green space than to $15m^2$.
+This difference is in fact a matter of justice [@laraSpace2015; @liottaPlanning2020].
 Constraining accessibility is in this way a useful way to evaluate the congested availability of any type of opportunity.
@@
 
-# Multimodal spatial availability $V_i^m$
-
-In brief, we define the spatial availability at $i$ ($V_{i}$) as the proportion of all opportunities in the region $O$ that are allocated to location $i$ from all opportunity destinations $j$.
+# Multimodal spatial availability
+
+In brief, we define the spatial availability at an origin $i$ ($V_{i}$) as the proportion of all opportunities in the region that are allocated to location $i$ from all destinations $j$.
 $V_{i}$ is a value of how many opportunities are available to each location $i$ out of all the opportunities in the region.
@@
 
-# Empirical example: Madrid LEZ
-
-## Data and methods
+# Empirical example
+
+## Context
 
@@
 
-The context for the empirical example is Madrid, in Spain.
-This city implemented a Low Emission Zone in 2017, to attend to goals set out in the national agenda to fight climate change, cut nitrogen dioxide levels, and prioritize people's movement in the city.
-Low emission zones (LEZ) elsewhere have been implemented as a climate change policy intervention to reduce GHG emissions, improve air quality, and support sustainable mobility [@margaryanLowEmissionZones2021; @verbeekJustManagementUrban2022].
-Though the rules of exclusion vary<!-- depending on legal aspects and cultural norms-->, LEZ aim to deter/reduce traffic in designated zones under threat of penalty (e.g., fines, seizure of vehicle).
-<!-- In practice, LEZ limits the volume of traffic by excluding vehicles by license plate, fuel type, or by introducing tolls.--> In other words, LEZ are a policy of _geographic discrimination_ as they change how people can reach opportunities by making the travel impedance more costly for specific forms of travel, typically cars.
-When considering opportunities as finite in a region, this discrimination reduces the competition of one mode and allows other modes to potentially thrive.
-In this way, LEZ aim to change the multimodal competitive accessibility landscape of a city.
-
+The context for the empirical example is Madrid, Spain.
+This city implemented a Low Emission Zone in 2017, to pursue goals set out in the national climate change agenda, to cut nitrogen dioxide levels, and to prioritize people's movement in the city.
+Low Emission Zones (LEZs) elsewhere have been implemented as interventions to reduce GHG emissions, improve air quality, and support sustainable mobility [@margaryanLowEmissionZones2021; @verbeekJustManagementUrban2022].
+Though the rules of exclusion vary by city<!-- depending on legal aspects and cultural norms-->, LEZs aim to deter/reduce traffic in designated zones under threat of penalty (e.g., fines, seizure of vehicle).
+<!-- In practice, LEZ limits the volume of traffic by excluding vehicles by license plate, fuel type, or by introducing tolls.--> In other words, LEZ implement a form of _geographic discrimination_ as they change how people can reach opportunities by making it more costly for some forms of travel, typically cars, to circulate in predetermined zones.
+When considering opportunities as finite in a region, this discrimination reduces the competition of one mode and opens up opportunties for other modes to thrive.
+In this way, at their core LEZs operate by changing the accessibility landscape of a city from the perspective of multiple modes.
+
+<!--
 Spain is one of a few countries with active LEZ and plans to expand their implementation as specified in their climate-change-related plans: _Plan Nacional Integrado de Energía y Clima 2021-2030_ [@espanaPlanNacionalIntegrado2020] and _Plan Nacional de Control de la Contaminación Atmosférica_ [@espanaResolucion10Enero2020].
@@
 when air quality exceeds limits specified in _RD 102/2011 de Mejora de Calidad del Aire_ [@barcelonaGUIATECNICAPARA2021].
-
-In geographic scope, the 2017 boundaries of the LEZ in Madrid were relatively modest (covering 4.72 km$^2$) and only within the center (i.e., LEZ Centro).
-These boundaries were expanded in 2023 to the area inside the M-30, an orbital highway in proximity to the city center (i.e., LEZ M-30).
-Beyond this, the city has plans to spatially expand the LEZ.
-Within the 2017 LEZ Centro implementation, all cars, motorcycles and freight with environmental labels A or B (older makes and models of fossil fuel internal combustion engine vehicles), are not permitted to enter the area unless they are used by residents or meet other exemptions.
-This restriction impacted approximately half of all car trips that typically travelled into what is now the LEZ Centro [@tarrinoortizAnalyzingImpactLow2022].
+-->
+
+In geographic scope, the 2017 boundaries of the LEZ in Madrid were relatively modest, covering only approximately 4.72 $km^2$ of the central business district of the city (the so-called LEZ Centro).
+As of this writing, there are plans to expand these boundaries to the area inside the M-30, an orbital highway in proximity to the city center (i.e., LEZ M-30).
+Within the 2017 LEZ Centro implementation, all cars, motorcycles and freight vehicles with environmental labels A or B (older makes and models of fossil fuel internal combustion engine vehicles), were disallowed from entering the zone unless they are used by residents or meet other exemptions.
+This restriction impacted approximately half of all car trips that used to travel into what is now the LEZ Centro [@tarrinoortizAnalyzingImpactLow2022].
 
 For this case study, we use spatial availability to quantify access to opportunities by different modes in Madrid.
-Particularly, we demonstrate how $V_i^m$ can be used to derive insights into how the restriction of car mobility in areas around/within the LEZ Centro may have allowed other, more sustainable but often slower or more costly modes, to become more competitive.
-
-```{r, jobs-plot, fig.cap="\\label{fig:Fig2} Jobs $O_j$ taken by people living and working in Madrid as reported by the home-to-work flows in the 2018 travel survey.
-Gray TAZ has no jobs.", fig.show='hold', fig.align='center'}
+Particularly, we demonstrate how $V_i^m$ can be used to derive insights into how the restriction of car mobility in areas around/within the LEZ Centro may have allowed more sustainable (but often slower or more costly modes) to become more competitive.
+
+```{r, jobs-plot, fig.cap="\\label{fig:Fig2} Distribution of jobs taken by people living and working in Madrid as reported in the 2018 travel survey.
+Gray TAZ has no jobs.
+Ranges of values in the legend are quintiles.", fig.show='hold', fig.align='center'}
 knitr::include_graphics("images/i_jobs_zn208_plot.png") 
@@
 
-```{r, pop-plot, fig.cap="\\label{fig:Fig3} Population living and working in Madrid by four summarized modal categories $P^m_i$ as reported by the home-to-work flows in the 2018 travel survey.
-Gray TAZs have no population.", fig.show='hold', fig.align='center'}
+```{r, pop-plot, fig.cap="\\label{fig:Fig3} Population living and working in Madrid by mode of transportation as reported in the 2018 travel survey.
+Gray TAZs have no population.
+Ranges of values in the legend are quintiles.", fig.show='hold', fig.align='center'}
 knitr::include_graphics("images/im_populations_zn208_plot.png")
@@
 
-The source of data for this example is the 2018 Community of Madrid travel survey [@comunidaddemadridResultadosEDM20182020].
-This is a representative survey that gives a snap-shot of the travel patterns for a typical day of the working week in 2018.
-Specifically, it captures 222,744 trips as measured from a representative sample of 85,064 households across the traffic analysis zones (TAZ) in the Community of Madrid (population of 6,507,184 over 3 years old).
+## Data
+
+The source of data for our empirical example is the 2018 Travel Survey of the Community of Madrid [@comunidaddemadridResultadosEDM20182020].
+This is a representative survey that offers a snap-shot of travel patterns for a typical weekday in 2018.
+The survey collected 222,744 trips from a representative sample of 85,064 households across the traffic analysis zones (TAZ) in the Community of Madrid.
+For context, the population older than 3 years in the Community is 6,507,184.
 
 In this example, we use all home-to-full-time-work trips, by all modes.
-Figures \ref{fig:Fig2} and \ref{fig:Fig3} show the population of workers and the distribution of jobs in the City of Madrid.
-The zoning system is the trafric analysis zones used in the survey.
-The red boundary represents the LEZ Centro in effect in 2017.
-The survey already captures the travel patterns after the introduction of LEZ Centro.
-The cyan boundary represents the LEZ planned for for the boundaries of the M-30 highway and is present in the plots as a spatial reference for areas in proximity to the LEZ Centro.
+The trips are expanded using population weights.
+Figures \ref{fig:Fig2} and \ref{fig:Fig3} show the number of workers and the distribution of jobs in the City of Madrid by TAZ.
+The light gray boundary represents the LEZ Centro in effect in 2017.
+The survey was conducted after the introduction of LEZ Centro.
+The dark gray boundary represents the LEZ planned for for the boundaries of the M-30 highway and is present in the plots as a spatial reference for areas in proximity to the LEZ Centro.
 
@@
 A similar trend can be observed for the motorized modal options where transit mode is more spread out than car/motor mode.
-All in all, these functions represent the propensity of travel by mode by duration of trip, which are needed to calculate the proportional allocation factors $F_{ij}^m$ for $V_i^m$.
-
-```{r, tlds-curves-m-plot, fig.cap="\\label{fig:Fig4} Fitted impedance function curve (line) against empirical TLD (bars) corresponding to the home-to-work origin-destination flows from the Madrid 2018 travel survey.", fig.show='hold', fig.align='center', fig.height=6}
+All in all, these functions represent the propensity of travel by mode by duration of trip, and are used to calculate the proportional allocation factors $F_{ij}^m$ for $V_i^m$.
+
+```{r, tlds-curves-m-plot, fig.cap="\\label{fig:Fig4} Fitted impedance function  against empirical TLD (bars) corresponding to the home-to-work origin-destination flows from the Madrid 2018 travel survey.", fig.show='hold', fig.align='center', fig.height=6}
 knitr::include_graphics("images/tlds_curves_m_plot.png")
@@
 
-At this point it is worthwhile to reiterate that the empirical example is a snap-shot of the calculated multimodal spatial availability using data from the 2018 travel survey.
-It is intended to visualize the spatial trends in availability of employment opportunities, by mode, and demonstrate how spatial availability can be interpreted to discuss the competitive advantage of lower travel impedance modes within Madrid Centro.
-
+At this point it is worthwhile to reiterate that the empirical example is a snap-shot of spatial availability by mode using data from the 2018 travel survey.
+Our purpose is to investigate the trends in availability of employment opportunities by mode, and illustrate how spatial availability can help in discussions about the competitive advantage of various modes within Madrid Centro.
 The spatial availability of jobs $V_i^m$ is calculated for each of the four modes $m$ at the level of traffic analysis zones $i$ in Madrid and displayed in Figure \ref{fig:Fig5}.
 
-```{r, SA-m-plot, fig.cap="\\label{fig:Fig5} Spatial availability of job opportunities per origin and mode $V_i^m$ in Madrid.
-Calculated using the home-to-work flows from the 2018 travel survey.
-Gray TAZs have no population.", fig.show='hold', fig.align='center', fig.height=10}
+```{r, SA-m-plot, fig.cap="\\label{fig:Fig5} Spatial availability of jobs per origin and mode $V_i^m$ in Madrid.
+Gray TAZs have no population.
+Ranges of values in the legend are quintiles.", fig.show='hold', fig.align='center', fig.height=10}
 knitr::include_graphics("images/SA_im_V_zn208_plot.png")
@@
 
-$V_i^m$ is a proportion of the total number of `r SA_i_zn208$V_i %>% sum()  %>% prettyNum(big.mark = ",")` jobs in the region and is shown in Figure \ref{fig:Fig5}.
-Since $V_i^m$ is calculated based on the population of workers and the distribution of jobs, the values can be understood as the number of full-time jobs that are spatially available to the full-time working population at that $i$ and their associated $m$, relative to all the jobs in the city.
-$V_i^m$ is the number of jobs that are _spatially available_ to a $m$-using population located at $i$, relative to the travel impedance and size of _all_ populations in the region.
-
-Some notable differences in the magnitude of $V_i^m$ occur between modes as seen in Figure \ref{fig:Fig5}.
-The majority of $V_i^m$ (which is to say of spatially available jobs) is allocated to those travelling by car and transit.
-This is to be expected, as the users of these modes represent `r (modal_V_comps_4plot$perc[3] + modal_V_comps_4plot$perc[4]) %>% percent(accuracy=0.1)` of the total population.
-Differences in $V_i^m$ values within modes also exist in space: car users outside of the M-30 region appear to have greater $V_i^m$ values, while some zones inside the M-30 appear to have higher $V_i^m$ values for transit.
-Overall, the magnitude of $V_i^m$ values for cyclists and pedestrians are lower than for car and transit but the highest values of $V_i^{bike}$ and $V_i^{walk}$ tend to be allotted to zones within the M-30 and origins with higher spatial availability by transit.
-
-The differences between the shares of modes and their shares of spatially availabile opportunities highlights the competitive advantage afforded by certain modes, although this effect is not uniform in space.
+In the figure, $V_i^m$ is a proportion of the total number of `r SA_i_zn208$V_i %>% sum()  %>% prettyNum(big.mark = ",")` jobs in the region.
+Since $V_i^m$ is calculated based on the population of workers and the distribution of jobs, the values can be understood as the number of full-time jobs that are spatially available to full-time workers at that $i$ traveling by mode $m$, relative to all the jobs in the city.
+There are noticeable differences in the magnitude of $V_i^m$ between modes as seen in Figure \ref{fig:Fig5}.
+The majority of $V_i^m$ (which is to say of spatially available jobs) are allocated to workers travelling by car and transit.
+In a way, this is to be expected since users of these modes represent `r (modal_V_comps_4plot$perc[3] + modal_V_comps_4plot$perc[4]) %>% percent(accuracy=0.1)` of the total population.
+However, the ability to travel at greater speeds also impacts this results.
+Fuerthermore, differences in $V_i^m$ values within modes also exist in space: car users outside of the M-30 region appear to enjoy greater spatial availability, while some zones inside the M-30 to have greater spatial availability for transit.
+Overall, the magnitude of $V_i^m$ values for cyclists and pedestrians are lower than for car and transit but the highest values of $V_i^{bike}$ and $V_i^{walk}$ tend to be found in zones within the M-30 and origins with higher spatial availability by transit.
+
+The differences between the shares of modes and their shares of spatially available opportunities highlights the competitive advantage of certain modes, although this effect is not geographically uniform.
 As seen in the left-most columns of Figure \ref{fig:Fig6}, 'car/motor' and 'transit' together can avail themselves of `r (modal_V_comps_4plot$perc[7] + modal_V_comps_4plot$perc[8]) %>% percent(accuracy=0.1)` of all jobs in the city.
@@
 These factors all contribute to the the car/motor mode being most advantaged in capturing spatially available job opportunities overall.
-```{r, modal-V-comps-plot, fig.cap="\\label{fig:Fig6} Displays the proportion of the working population by mode and spatial availability of job opportunities by mode aggregated for three spatial areas.
-From left to right, the city of Madrid (All), the area within the M-30 highway (M-30),the area within the Centro region (Centro).", fig.show='hold', fig.align='center', fig.height=10}
+```{r, modal-V-comps-plot, fig.cap="\\label{fig:Fig6} Proportion of population by mode and spatial availability of jobs by mode aggregated for three areas.
+From left to right, the city of Madrid (All), the area within the M-30 highway (M-30), the area within the Centro region (Centro).", fig.show='hold', fig.align='center', fig.height=10}
 knitr::include_graphics("images/modal_V_comps_4plot.png")
@@
 
-Despite the big picture, it is important to note that there are spatial variations in the competitive advantage of cars.
-The proportion of travel by car in Centro is smaller and has higher travel impedance values relative to the inputs in other areas and by other modes.
-The LEZ Centro implementation further restricts the advantage of cars by shifting more than half of all car trips into the LEZ to modes [@tarrinoortizAnalyzingImpactLow2022].
-This restriction decreased the number of trips by car from going into the LEZ Centro (an area with a large number of jobs overall, see Figure \ref{fig:Fig2}), thus increasing the mass effect for non-car modes and resulting in proportionally higher $v_i^m$ values for non-car modes.
-As such, the lower amount of access to opportunities by car allows more opportunities in the LEZ to be available to populations using other modes.
-
+The big picture demonstrates how car is the most advantageous mode, however it is interesting to notice that this advantage appears to be blunted by the LEZ.
+Unlike the results for the whole city and even within the M-30 ring, the proportion of car users in Centro is larger than the proportion of opportunities spatially available to them.
+The restriction on cars in effect reduces competition by this mode, and leads to a relative increment of the mass effect of the modes allowed within the LEZ, which also contains a large number of jobs (see Figure \ref{fig:Fig2}).
 As summarized in the two right-most columns in Figure \ref{fig:Fig6}, the proportion of jobs spatially available to car in Centro is (`r modal_V_comps_4plot$perc[23] %>% percent(accuracy=0.1)` or 5,373 opportunities).
 For reference, this is less than the proportion of the car users in Centro (`r modal_V_comps_4plot$perc[19] %>% percent(accuracy=0.1)`), evidently less then the proportion of car users in the city, and is the opposite of the overall trend (left-most columns) and within the M-30 (middle columns).
-More opportunities are spatially available to non-car users within Centro, particularly by transit, a mode with `r modal_V_comps_4plot$perc[24] %>% percent(accuracy=0.1)` of spatially available jobs in Centro, despite representing only `r modal_V_comps_4plot$perc[20] %>% percent(accuracy=0.1)` of the population in Centro and `r modal_V_comps_4plot$perc[4] %>% percent(accuracy=0.1)` in the city overall.
-
-Figure \ref{fig:Fig6} also shows that there is a higher proportion of opportunities that are spatially available to pedestrians and cyclists in Centro than in the City overall and in all areas within the M-30.
-Notably, within Centro, `r modal_V_comps_4plot$perc[21] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[22] %>% percent(accuracy=0.1)` of opportunities are spatially available to bike and walk modes respectively, while their populations represent smaller proportions of `r modal_V_comps_4plot$perc[1] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[2] %>% percent(accuracy=0.1)` of the population.
-Although the proportion of opportunities that are spatially available to these modes is still lower than the respective shares of the population located in the Centro, these modes are more competitive within the Centro than anywhere without.
-By restricting the ability of cars to enter Centro, the LEZ also contributes to level the playing field for slower modes, in particular cycling and walking.
-
-The spatial differences in the competitive dis/advantage of spatial availability between modes can also be visualized per origin.
-Figure \ref{fig:Fig7} shows $v_i^m$, the spatial availability $V_i^m$ divided by the population that uses $m$.
-Values of $v_i^m$ above 1 are shown in red and represent more than one spatially available opportunity per capita; values below 1 are shown in green, and represent less than one spatially available opportunity per capita; values equal to 1 are shown in white.
-These plots drive home the point of the disproportionately high over-allocation of spatial availability relative to the share of modes in many of the origins for car (bottom left plot, areas denoted with green $v_i^m$ values above 1).
-These plots also show areas with disproportionately few spatially available opportunities (under 1).
-It can be observed that the spatial availability to jobs of transit users is relatively well balanced (i.e., many zones are white), while that for non-motorized modes is low (under 1) overall.
-
-```{r, SA-per-capita-m-plot, fig.cap="\\label{fig:Fig7} Spatial availability of job opportunities per mode-using capita by mode $v_i^m$ per origin in Madrid.
-Calculated using the home-to-work flows from the 2018 travel survey.
-Gray TAZs have no population.", fig.show='hold', fig.align='center', fig.height=10}
+
+It is also clear that more opportunities are spatially available to non-car users within Centro.
+In the case of active travel, the proportions of cyclists and walkers within LEZ Centro still exceed the proportions of jobs spatially available to them --- however the disparity is drastically reduced compared to the rest of the city.
+As seen in Figure \ref{fig:Fig6}, there is a higher proportion of opportunities that are spatially available to pedestrians and cyclists in Centro than in the City overall and in areas within the M-30.
+Notably, within Centro, `r modal_V_comps_4plot$perc[21] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[22] %>% percent(accuracy=0.1)` of opportunities are spatially available to bike and walk modes respectively, while their populations represent  `r modal_V_comps_4plot$perc[1] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[2] %>% percent(accuracy=0.1)` of the population.
+By restricting the ability of cars to enter Centro, the LEZ seems to contributes to level the playing field for slower modes, in particular cycling and walking, but also transit.
+As seen in the figure, transit users are generally close to parity across the region, with nearly as many spatially available jobs as transit users.
+Still, this mode has the greatest advantage in LEZ Centro with `r modal_V_comps_4plot$perc[24] %>% percent(accuracy=0.1)` of spatially available jobs in Centro for `r modal_V_comps_4plot$perc[20] %>% percent(accuracy=0.1)` of transit users in Centro.
+This result makes intuitive sense: after car, it is the mode with the greatest range, and unlike car it is allowed in LEZ Centro.
+
+The spatial differences in the competitive dis/advantage of spatial availability between modes can also be visualized at a finer level of granularity.
+Figure \ref{fig:Fig7} shows $v_i^m$, the spatial availability $V_i^m$ divided by the population of users of $m$.
+Values of $v_i^m$ below 1 are shown in shades of orange, and indicate TAZs with less than one spatially available opportunity per capita for the mode.
+Values above 1 are shown in shades of green, and indicate TAZs with more than one spatially available opportunity per capita for the mode.
+The highest spatial availability per capita (shown in blue) is for car users in a zone northeast just beyond the M-30.
+These plots illustrate in unambiguous fashion, and in a quantity that is comparable over space and time, the advantage in terms of spatial availability of car for most of the city (bottom left plot, areas denoted with green $v_i^m$ values above 1).
+It can also be observed that spatial availability of jobs is relatively well balanced for transit users over most of the regions (i.e., many zones are light orange or light green).
+Spatial availability of jobs for non-motorized modes, in contrast, is low (under 1) overall, although less so within LEZ Centro.
+
+```{r, SA-per-capita-m-plot, fig.cap="\\label{fig:Fig7} Distribution of spatially available jobs per capita by mode of transportatin ($v_i^m$).
+Gray TAZs have no population that use the mode.
+Ranges of values in the legend are quintiles.", fig.show='hold', fig.align='center', fig.height=10}
 knitr::include_graphics("images/SA_im_vv_zn208_plot.png")
@@
 
-It is also interesting to note that $v_i^m$ for car within and near LEZ Centro is close to or below 1 (white/red) in Figure \ref{fig:Fig7}, while in contrast all non-car modes have relatively higher $v_i^m$ values.
-Since the values are comparable across regions and over time, Figure \ref{fig:Fig7} provides a benchmark for quantifying potential LEZ implementations in the future.
-As Figure \ref{fig:Fig7} also shows, many areas within the M-30 have high (white/green) $v_i^m$ values for car, but the results for LEZ Centro give reasonable grounds to speculate that a spatial expansion of the LEZ to include all areas within the M-30 stands to increase the spatial availability of jobs for more transit, cyclcing and walking.
+Incidentally, it is bears noting that $v_i^m$ for car within and near LEZ Centro is close to or below 1 in Figure \ref{fig:Fig7}, while all non-car modes have relatively higher $v_i^m$ values.
+Since these values are comparable across regions and over time, Figure \ref{fig:Fig7} potentially provides a benchmark for quantifying changes in LEZ policies in the future.
+As Figure \ref{fig:Fig7} also shows, many areas within the M-30 have high (white/green) $v_i^m$ values for car, but the results for LEZ Centro give reasonable grounds to speculate that a spatial expansion of the LEZ to include all areas within the M-30 would likely increase the spatial availability of jobs for transit users, cyclists and pedestrians.
 
@@
 
-<!--QUESTION: maybe break up this section and consolidate all limitations/assumptions/potential uses of multimodal spatial availability in the conclusion?-->
-Location-based accessibility measures like the Hansen-type $S_i^m$, Shen-type $a_i^m$, and spatial availability $V_i^m$ measures share a commonality; they are a weighted sum of opportunities assigned to each spatial unit $i$ in a region.
-In this way, they all can be interpreted as a score of how many opportunities can be potentially interacted with by the population at $i$.
-How the weight and sum of the potentially-interacted-with opportunities is considered is what defines the type of accessibility measure.
-
-Within this paper, the location-based singly- _constrained_ and _competitive_ accessibility measure, known as spatial availability $V_i$ [@soukhovIntroducingSpatialAvailability2023], is extended for the case of capturing multimodal accessibility to opportunities $V_i^m$.
-A synthetic example and then an empirical case of LEZ in Madrid are detailed to demonstrate this multimodal extension.
-
-The spatial availability measure is capable of capturing a new interpretation of multimodal competition that previous accessibility measures have not yet done.
-Competitive measures hypothesis that populations using modes with lower travel impedance, when competing for a finite set of opportunities, will capture more opportunities.
-With spatial availability, the number of opportunities that are captured (of the total opportunities in the region) by each mode can be individually calculated.
-From there, the difference between how many spatially available opportunities one mode captures versus another can be investigated.
-This is the advantage of the spatial availability measure, particularly its multimodal extension.
-
-<!-- note: maybe add in a paragraph about how the spatial availabilit ymeasure can take into consideration traveler related characteristics, like mode choice, within the measure.
-This is a current concern with accessibility measures Karner et al.
-2023 ( Emerging perspectives on transportation justice.
-Transportation Research
-1
-Part D: Transport and Environment) -->
-
-The flexibility and need for an accessibility measure such as spatial availability is pertinent in policy scenario evaluation.
-As showcased in the empirical example of the LEZ in Madrid, competition for job opportunity availability varies spatially _as well as_ between modes.
-The car and transit modes have the highest spatial availability, with the car-mode having highest availability with exception to the areas within the LEZ Centro.
-Since car travel has been highly restricted within the LEZ Centro, fewer car-using people potentially interact with jobs within the LEZ Centro, leaving more _spatially available_ jobs for non-car-using populations.
-This difference in car-using populations in locations accessing jobs within and immediately outside the LEZ Centro increases the competitiveness of the transit-using population (the second most competitive mode) as well as the non-motorized modes.
-
-Spatial availability $V_i^m$ can also be divided by the mode-using population at each $i$ to yield mode-population normalized values.
-These values, reflected in Figure \ref{fig:Fig7}, can be used as a benchmark to investigate existing conditions and plan future LEZ implementation (i.e., target areas with exceptionally high car spatial availability such that more opportunities are available to other mode-users).
-
-In summary, conventional _non-constrained_ accessibility measures are difficult for planners to operationalize for a variety of reasons including issues of computation and interpretability [@levinsonTransportAccessManual2020].
-With spatial availability, the magnitude of opportunities that are available as a proportion of all the opportunities in the region is equal to $V_i$.
-As a result of its proportional allocation mechanism, $V_i$ can be naturally extended into multimodal applications.
-This flexibility is helpful to modelling policy scenarios in our cities that are increasingly multimodal.
-The interpretation of $V_i$ allows for manipulation of $V_i^m$ values to investigate differences of availability between neighbourhoods, modes, and regions, generate per capita benchmarks, and/or generate average values per population-group.
-
-From a spatial equity perspective, spatial availability measure can provide researchers, policy makers, and citizens a new-found interpretation of accessibility measures.
-With a plot of spatial availability values, one can begin asking, how much is enough and what level may be too much.
-These interpretations were difficult to be made with accessibility measures in the past.
-In future work, we intend to use multimodal spatial availability to characterise the equity of multimodal accessibility losses and gains in varying LEZ policy scenario implementations.
+Accessibility measures are an important tool in transportation research [@shiLiterature2020] and increasingly also seen as valuable for planning purposes [@handyAccessibility2020; @levinsonTransportAccessManual2020; @siddiqToolsTradeAssessing2021; @yanAccessibilityBasedPlanningAddressing2021; @elGeneidyMaking2022].
+They boast a long history of development, beginning with Hansen-type $S_i^m$ measures, with other developments like Shen's $a_i^m$, to account for competition/congestion.
+The more recent spatial availability measure $V_i^m$ has in common with these accessibility indicators that it is a weighted sum of the opportunities in a region from the perspective of a determined origin $i$.
+Aggregations of opportunities embody principles of gravitational/spatial interaction modelling that date back to at least H.C.
+Carey [@careyPriciples1858], and are part of a line of research that includes the work of Ravenstein [@ravensteinLaws1889], Reilly [@reillyMethods1929], Stewart [@stewartInverse1941; @stewartSuggested1947; @stewartDemographic1948], Zipf [@zipfHypothesis1946; @zipfIntercity1946], Wilson [@wilsonFamilty1971], and many others besides.
+In this way, $S_i^m$, $a_i^m$, and yes, $V_i^m$, can be interpreted as scores of the potential for interaction with opportunities in space.
+
+Different accessibility indicators are characterized by how they weight and aggregate opportunities.
+Spatial availability's contribution to the literature is to incorporate a proportional allocation mechanism that essentially constrains the sums to match the number of opportunities in the region; in this way it is a singly-constrained accessibility measure that naturally accommodates congestion and competition.
+The effort with spatial availability is in line with previous research on proportional allocation by Paez et al.
+[@paezDemand2019].
+As initially introduced by Soukhov et al.
+[@soukhovIntroducingSpatialAvailability2023], spatial availability was designed for a homogeneous population traveling by a single mode of transportation.
+In this paper we extended spatial availability for the case of heterogeneous populations.
+We discussed this in terms of multiple modes of transportation, but the framework can accommodate equally well variations in travel behavior by population segments.
+
+An empirical example using data from Madrid helped to illustrate the potential of multimodal spatial availability analysis, including its ability to account for competition for opportunities within and between modes.
+Particularly relevant is the fact that spatial availability scores relate directly to the total number of opportunities in the region.
+This makes it possible to compare the results to intuitive benchmarks, such as opportunities per population, in ways that other accessibility measures cannot or tend to obfuscate.
+This comparability is preserved between regions and over time.
+The example suggests that once that opportunities are treated as being finite, restrictions to travel by car leave more spatially available opportunities for non-car-users.
+This difference for car travel in locations within and immediately around the LEZ Centro seems to increase the number of opportunities spatially available to transit users (transit being the second most competitive mode), as well as the spatial availability from the perspective of non-motorized modes.
+In effect, a policy such as Low Emission Zones helps to improve the accessibility situation of active travel and transit in the parts of the city where it is implemented.
+
+The purpose of the empirical example was to illustrate the kind of insights that can be derived from the application of multimodal spatial availability.
+But there are some intriguing opportunities for future research.
+Accessibility indicators are not designed to work as modal split models, and yet, in the case of policies that alter the relative cost of various forms of transportation, one can reasonably expect to see some shifts between modes.
+In our empirical example we used data collected _after_ the introduction of LEZ Centro.
+However, given a modal split model to project model shares, accessibility indicators, including spatial availability, can be used to investigate changes to the accessibility landscape.
+Ditto for destination choice.
+Our empirical example in this paper presented but a snapshot of this, and in future research it will be interesting to investigate changes _between_ policy interventions.
+The expansion of Madrid's LEZ to the ring contained by the M-30 orbital presents an excellent opportunity to do so.
+Given the intuitive and straightforward interpretation of spatial availability scores as fractions of opportunities from the total, relative and absolute changes in the accessibility landscape can be assessed, thus helping to evaluate the implications of policy interventions.
+
+Finally, our example dealt with differences in travel by mode only, but it is possible to think of the intersection between mode of travel and different types of travelers.
+This would expand the number of sub-populations in the analysis from, say, $m=M$ (modes) to $m = M\cdot Q$ (modes times population segments), each with their own characteristic impedance function.
+Evaluations of this kind will be especially relevant as LEZ are implemented in cities globally, and the question of their impact on disadvantaged populations who have become mobility-restricted increasingly come to the fore [@devrijNooneVisitsMe2022; @verbeekJustManagementUrban2022; @liottaWhatDrivesInequalities2023].
 

### 2023-11-09 Journal-manuscript/Journal-manuscript.tex
Antonio completed editions of revised version of paper; now it is going back to Anastasia for proof-reading

@@
 transportation systems when combined with the distribution of
-opportunities in space {[}e.g., {]}. As such, it is a holistic measure
-of spatial systems that measures the ease of reaching destinations
-{[}9,10{]}.
+opportunities in space {[}9{]}. As such, it is a holistic measure of
+spatial systems that measures the ease of reaching destinations
+{[}10,11{]}.
 
@@
 floating catchment area approaches in accessibility analysis
-{[}11,12{]}. While these approaches purport to account for congestions,
-Paez et al. {[}13{]} demonstrate that in general they do not solve the
+{[}12,13{]}. While these approaches purport to account for congestions,
+Paez et al. {[}14{]} demonstrate that in general they do not solve the
 issue of multiple counting of opportunities, thus leading to biases in
@@
 paid closer attention to the way opportunities are counted in
-accessibility analysis. {[}13{]}, for example, tackle floating catchment
+accessibility analysis. {[}14{]}, for example, tackle floating catchment
 area methods and introduce a normalization of the impedance matrix to
 allocate the population and then the level of service proportionally.
-More recently, Soukhov et al. {[}14{]} introduced a singly-constrained
+More recently, Soukhov et al. {[}15{]} introduced a singly-constrained
 measure of accessibility, called spatial availability, that employs a
@@
 A limitation of spatial availability as introduced by Soukhov et al.
-{[}14{]} is that it was developed for the case of a homogeneous
+{[}15{]} is that it was developed for the case of a homogeneous
 population, for example for the case of a single mode of transportation.
@@
 other forms of heterogeneity, for example variations in travel behavior
-between older and younger adults {[}e.g., 15{]}, the propensity of older
-adults to use different modes of transportation {[}16{]}, the usually
-shorter trip lengths of children compared to grown-ups {[}e.g., 17{]},
-or the more limited travel ranges of single parents {[}e.g., 18{]}.
+between older and younger adults {[}e.g., 16{]}, the propensity of older
+adults to use different modes of transportation {[}17{]}, the usually
+shorter trip lengths of children compared to grown-ups {[}e.g., 18{]},
+or the more limited travel ranges of single parents {[}e.g., 19{]}.
 
@@
 land-use and transport systems. Arguably, the most commonly used are
-measures based on the gravity model {[}19{]}, of which cumulative
+measures based on the gravity model {[}20{]}, of which cumulative
 opportunity measures and weighted cumulative opportunity measures are
@@
 not consider the opportunities as finite. To cite an example, Tahmasbi
-et al. {[}20{]} use Hansen-type accessibility to assess the potential
+et al. {[}21{]} use Hansen-type accessibility to assess the potential
 interaction with retail locations by three modes: walking, public
@@
 each origin \(i\) three accessibility scores are calculated. In this
-work, Tahmasbi et al. {[}20{]} show that car travel affords the highest
+work, Tahmasbi et al. {[}21{]} show that car travel affords the highest
 \(S_i^{m}\) values in the majority of \(i\), i.e., travelers who use a
@@
 modes are often values that are difficult to interpret beyond making
-statements about relative size. For example, Lunke {[}21{]}, researching
+statements about relative size. For example, Lunke {[}22{]}, researching
 the region of Oslo, reports accessibility scores for car in the order of
@@
 still plenty of opportunities. These ratios can be found elsewhere in
-the literature {[}e.g., 22, and 9,15,18, and 8{]}, and they are useful
+the literature {[}e.g., 23, and 9,16,19, and 8{]}, and they are useful
 to assess when some members of the public are better or worse off than
@@
 Hansen-type accessibility that has been standardized in this way is not
-uncommon in the literature {[}e.g., 23,24{]}.
+uncommon in the literature {[}e.g., 24,25{]}.
 
@@
 limited and thus is subject to competition by population
-{[}11,14,25--29{]}. There are only so many school-seats, hospital
+{[}12,15,26--30{]}. There are only so many school-seats, hospital
 capacity, employment opportunities, etc., in a region and if one person
@@
 the demand for that opportunity are two components of accessibility.
-These are clear examples of opportunities that are clearly competitive.
-But we would go as far as to argue that every type of opportunity is
-subject to congestion or capacity constraint, even when the
-opportunities are conventionally seen as non-competitive. Amenities are
-a good example of this. For instance, standards for providing green
+These are clear examples of opportunities that are unambiguously
+competitive. But we would go as far as to argue that every type of
+opportunity is subject to congestion or capacity constraint, even when
+the opportunities are conventionally seen as non-competitive. Amenities
+are a good example of this. For instance, standards for providing green
 spaces are often stated in the form of \emph{exclusive access}, in units
-of amenity per capita. A case in point is Natural England, an
-organization that recommends an Accessible Natural Greenspace Standard
-such that the minimum supply of space is one ha of statutory Local
-Nature Reserves per thousand population\footnote{see
+of amenity per capita. A case in point is the Ile-de-France region, a
+jurisdiction that suggested in a major planning document of 2013 that at
+the municipal level at least \(10m^2\) of public green space should be
+supplied \emph{per inhabitant} {[}31{]}. But green spaces are not evenly
+distributed, which means that who has access to them hinges on where
+they are and how easy is to reach them. This formulation of provision of
+amenities is not unusual. For example, Natural England, an organization
+that recommends an Accessible Natural Greenspace Standard such that the
+minimum supply of space is one ha of statutory Local Nature Reserves per
+thousand population\footnote{see
   https://redfrogforum.org/wp-content/uploads/2019/11/67-Nature-Nearby\%E2\%80\%99-Accessible-Natural-Greenspace-Guidance.pdf}.
-Similarly, the World Health Organization {[}cited in 30{]} recommends
+Similarly, the World Health Organization {[}cited in 32{]} recommends
 that cities provide a minimum of 9\(m^2\) of green area per inhabitant.
-For our purposes, standards of this type translate into ``how much of a
-resource can be offered to each individual that is not allocated to
-anyone else?''. Green spaces often have large capacities, but they still
-have a capacity. Constraining accessibility is in this way a useful way
-to evaluate the congested availability of any type of opportunity. As
-development of sound standards is emphasized in the planning literature,
-in particular in regards to fairness in transportation {[}see 31{]},
-spatial availability analysis is a useful way to develop and assess
-standards.
+For our purposes, standards of this type translate into ``how much of
+this resource is available to one individual that has not been claimed
+by anyone else?''. Green spaces often have large capacities, but they
+still have a capacity, and it is not the same a person to have access to
+\(5m^2\) of \emph{uncongested} green space than to \(15m^2\). This
+difference is in fact a matter of justice {[}31,33{]}. Constraining
+accessibility is in this way a useful way to evaluate the congested
+availability of any type of opportunity. As development of sound
+standards is emphasized in the planning literature, in particular in
+regards to fairness in transportation {[}see 34{]}, spatial availability
+analysis is a useful way to develop and assess standards.
 
@@
 we think about the use of multiple modes (or heterogeneous populations).
-If we return to Oslo for a moment {[}21{]}, we notice that the places
+If we return to Oslo for a moment {[}22{]}, we notice that the places
 that have high accessibility by transit are also the places that have
@@
 rationale for developing floating catchment area methods (FCA),
-popularized by {[}12,11,although similar, and earlier, developments are
-found in 29,32{]}. Shen-type accessibility is formulated as:
+popularized by {[}13,12,although similar, and earlier, developments are
+found in 30,35{]}. Shen-type accessibility is formulated as:
 \(a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}\) where \(D_j^m\)
@@
 meaning both population \emph{and} opportunities are multiply counted
-{[}see 13{]}. In other words, interpretation of differences in Shen-type
+{[}see 14{]}. In other words, interpretation of differences in Shen-type
 accessibility scores between modes is fraught as it is for Hansen-type
@@
 
-To illustrate, {[}33{]} calculate \(a_i^m\) to jobs for different
+To illustrate, {[}36{]} calculate \(a_i^m\) to jobs for different
 income-group populations in Shenzhen, China using
@@
 allocation factors that follow well established principles of spatial
-interaction and the gravity model {[}see 34{]}. In Soukhov et al.
-{[}14{]} these factors consider: 1) the mass effect (e.g., the size of
+interaction and the gravity model {[}see 37{]}. In Soukhov et al.
+{[}15{]} these factors consider: 1) the mass effect (e.g., the size of
 populations at different origins); and 2) the cost of travel from
@@
 
-\hypertarget{multimodal-spatial-availability-v_im}{%
-\section{\texorpdfstring{Multimodal spatial availability
-\(V_i^m\)}{Multimodal spatial availability V\_i\^{}m}}\label{multimodal-spatial-availability-v_im}}
-
-In brief, we define the spatial availability at \(i\) (\(V_{i}\)) as the
-proportion of all opportunities in the region \(O\) that are allocated
-to location \(i\) from all opportunity destinations \(j\). \(V_{i}\) is
-a value of how many opportunities are available to each location \(i\)
-out of all the opportunities in the region. The general formulation of
-spatial availability \(V_{i}\) is shown in Equation
-(\ref{eq:spatial-availability-general}) {[}see 14{]}:
+\hypertarget{multimodal-spatial-availability}{%
+\section{Multimodal spatial
+availability}\label{multimodal-spatial-availability}}
+
+In brief, we define the spatial availability at an origin \(i\)
+(\(V_{i}\)) as the proportion of all opportunities in the region that
+are allocated to location \(i\) from all destinations \(j\). \(V_{i}\)
+is a value of how many opportunities are available to each location
+\(i\) out of all the opportunities in the region. The general
+formulation of spatial availability \(V_{i}\) is shown in Equation
+(\ref{eq:spatial-availability-general}) {[}see 15{]}:
 
@@
 population, followed by then \(A\), then \(C\). This synthetic example
-is inspired by the single-mode example used in {[}11{]} and reconfigured
-in {[}14{]}.
+is inspired by the single-mode example used in {[}12{]} and reconfigured
+in {[}15{]}.
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{images/Fig1}
+{\centering \includegraphics[width=0.85\linewidth]{images/Fig1}
 
@@
 columns and are aggregated for each \(i\) in the final two columns in
-Table 1 . As in the example in Shen {[}11{]}, we use a negative
+Table 1 . As in the example in Shen {[}12{]}, we use a negative
 exponential impedance function
@@
 To connect this example to the literature, \(S_i^m\) is calculated in
-the work of {[}20{]}; they compare differences in \(S_i^m\) values
+the work of {[}21{]}; they compare differences in \(S_i^m\) values
 between modes in a relative and comparative sense, but make no further
@@
 Shen-type measure is implemented in aforementioned work of Tao et al.
-{[}33{]} to calculate modal \(a_i^m\) values and the aggregated \(a_i\)
-is implemented in the work of {[}35{]}. However, similar to Hansen-type
+{[}36{]} to calculate modal \(a_i^m\) values and the aggregated \(a_i\)
+is implemented in the work of {[}38{]}. However, similar to Hansen-type
 accessibility, these works discuss relative and spatially comparative
@@
 
-\hypertarget{empirical-example-madrid-lez}{%
-\section{Empirical example: Madrid
-LEZ}\label{empirical-example-madrid-lez}}
-
-\hypertarget{data-and-methods}{%
-\subsection{Data and methods}\label{data-and-methods}}
-
-The context for the empirical example is Madrid, in Spain. This city
-implemented a Low Emission Zone in 2017, to attend to goals set out in
-the national agenda to fight climate change, cut nitrogen dioxide
-levels, and prioritize people's movement in the city. Low emission zones
-(LEZ) elsewhere have been implemented as a climate change policy
-intervention to reduce GHG emissions, improve air quality, and support
-sustainable mobility {[}36,37{]}. Though the rules of exclusion vary,
-LEZ aim to deter/reduce traffic in designated zones under threat of
-penalty (e.g., fines, seizure of vehicle). In other words, LEZ are a
-policy of \emph{geographic discrimination} as they change how people can
-reach opportunities by making the travel impedance more costly for
-specific forms of travel, typically cars. When considering opportunities
-as finite in a region, this discrimination reduces the competition of
-one mode and allows other modes to potentially thrive. In this way, LEZ
-aim to change the multimodal competitive accessibility landscape of a
-city.
-
-Spain is one of a few countries with active LEZ and plans to expand
-their implementation as specified in their climate-change-related plans:
-\emph{Plan Nacional Integrado de Energía y Clima 2021-2030} {[}38{]} and
-\emph{Plan Nacional de Control de la Contaminación Atmosférica}
-{[}39{]}. Specifically, the national Spanish law 7/2021 ( \emph{Ley de
-Cambio Climático y Transición Energética}) will require all
-municipalities to implement LEZ by 2023 if they meet at least one of the
-following requirements: (i) municipalities \textgreater50,000 inhab.;
-(ii) islands; and (iii) municipalities \textgreater{} 20,000 inhab. when
-air quality exceeds limits specified in \emph{RD 102/2011 de Mejora de
-Calidad del Aire} {[}40{]}.
+\hypertarget{empirical-example}{%
+\section{Empirical example}\label{empirical-example}}
+
+\hypertarget{context}{%
+\subsection{Context}\label{context}}
+
+The context for the empirical example is Madrid, Spain. This city
+implemented a Low Emission Zone in 2017, to pursue goals set out in the
+national climate change agenda, to cut nitrogen dioxide levels, and to
+prioritize people's movement in the city. Low Emission Zones (LEZs)
+elsewhere have been implemented as interventions to reduce GHG
+emissions, improve air quality, and support sustainable mobility
+{[}39,40{]}. Though the rules of exclusion vary by city, LEZs aim to
+deter/reduce traffic in designated zones under threat of penalty (e.g.,
+fines, seizure of vehicle). In other words, LEZ implement a form of
+\emph{geographic discrimination} as they change how people can reach
+opportunities by making it more costly for some forms of travel,
+typically cars, to circulate in predetermined zones. When considering
+opportunities as finite in a region, this discrimination reduces the
+competition of one mode and opens up opportunties for other modes to
+thrive. In this way, at their core LEZs operate by changing the
+accessibility landscape of a city from the perspective of multiple
+modes.
 
 In geographic scope, the 2017 boundaries of the LEZ in Madrid were
-relatively modest (covering 4.72 km\(^2\)) and only within the center
-(i.e., LEZ Centro). These boundaries were expanded in 2023 to the area
+relatively modest, covering only approximately 4.72 \(km^2\) of the
+central business district of the city (the so-called LEZ Centro). As of
+this writing, there are plans to expand these boundaries to the area
 inside the M-30, an orbital highway in proximity to the city center
-(i.e., LEZ M-30). Beyond this, the city has plans to spatially expand
-the LEZ. Within the 2017 LEZ Centro implementation, all cars,
-motorcycles and freight with environmental labels A or B (older makes
-and models of fossil fuel internal combustion engine vehicles), are not
-permitted to enter the area unless they are used by residents or meet
-other exemptions. This restriction impacted approximately half of all
-car trips that typically travelled into what is now the LEZ Centro
+(i.e., LEZ M-30). Within the 2017 LEZ Centro implementation, all cars,
+motorcycles and freight vehicles with environmental labels A or B (older
+makes and models of fossil fuel internal combustion engine vehicles),
+were disallowed from entering the zone unless they are used by residents
+or meet other exemptions. This restriction impacted approximately half
+of all car trips that used to travel into what is now the LEZ Centro
 {[}41{]}.
@@
 how \(V_i^m\) can be used to derive insights into how the restriction of
-car mobility in areas around/within the LEZ Centro may have allowed
-other, more sustainable but often slower or more costly modes, to become
-more competitive.
+car mobility in areas around/within the LEZ Centro may have allowed more
+sustainable (but often slower or more costly modes) to become more
+competitive.
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{images/i_jobs_zn208_plot}
+{\centering \includegraphics[width=0.85\linewidth]{images/i_jobs_zn208_plot}
 
@@
 
-\caption{\label{fig:Fig2} Jobs $O_j$ taken by people living and working in Madrid as reported by the home-to-work flows in the 2018 travel survey. Gray TAZ has no jobs.}\label{fig:jobs-plot}
+\caption{\label{fig:Fig2} Distribution of jobs taken by people living and working in Madrid as reported in the 2018 travel survey.
+Gray TAZ has no jobs.
+Ranges of values in the legend are quintiles.}\label{fig:jobs-plot}
 \end{figure}
@@
 
-{\centering \includegraphics[width=1\linewidth]{images/im_populations_zn208_plot}
+{\centering \includegraphics[width=0.85\linewidth]{images/im_populations_zn208_plot}
 
@@
 
-\caption{\label{fig:Fig3} Population living and working in Madrid by four summarized modal categories $P^m_i$ as reported by the home-to-work flows in the 2018 travel survey.
-Gray TAZs have no population.}\label{fig:pop-plot}
+\caption{\label{fig:Fig3} Population living and working in Madrid by mode of transportation as reported in the 2018 travel survey.
+Gray TAZs have no population.
+Ranges of values in the legend are quintiles.}\label{fig:pop-plot}
 \end{figure}
 
-The source of data for this example is the 2018 Community of Madrid
-travel survey {[}42{]}. This is a representative survey that gives a
-snap-shot of the travel patterns for a typical day of the working week
-in 2018. Specifically, it captures 222,744 trips as measured from a
-representative sample of 85,064 households across the traffic analysis
-zones (TAZ) in the Community of Madrid (population of 6,507,184 over 3
-years old).
+\hypertarget{data}{%
+\subsection{Data}\label{data}}
+
+The source of data for our empirical example is the 2018 Travel Survey
+of the Community of Madrid {[}42{]}. This is a representative survey
+that offers a snap-shot of travel patterns for a typical weekday in
+2018. The survey collected 222,744 trips from a representative sample of
+85,064 households across the traffic analysis zones (TAZ) in the
+Community of Madrid. For context, the population older than 3 years in
+the Community is 6,507,184.
 
 In this example, we use all home-to-full-time-work trips, by all modes.
-Figures \ref{fig:Fig2} and \ref{fig:Fig3} show the population of workers
-and the distribution of jobs in the City of Madrid. The zoning system is
-the trafric analysis zones used in the survey. The red boundary
-represents the LEZ Centro in effect in 2017. The survey already captures
-the travel patterns after the introduction of LEZ Centro. The cyan
-boundary represents the LEZ planned for for the boundaries of the M-30
-highway and is present in the plots as a spatial reference for areas in
-proximity to the LEZ Centro.
+The trips are expanded using population weights. Figures \ref{fig:Fig2}
+and \ref{fig:Fig3} show the number of workers and the distribution of
+jobs in the City of Madrid by TAZ. The light gray boundary represents
+the LEZ Centro in effect in 2017. The survey was conducted after the
+introduction of LEZ Centro. The dark gray boundary represents the LEZ
+planned for for the boundaries of the M-30 highway and is present in the
+plots as a spatial reference for areas in proximity to the LEZ Centro.
 
@@
 non-motorized modes respectively. The selection of functional forms
-aligns with empirical examples in other regions {[}14,47,48{]}. The
+aligns with empirical examples in other regions {[}15,47,48{]}. The
 shape and rate parameters for the gamma functions (motorized modes) are
@@
 spread out than car/motor mode. All in all, these functions represent
-the propensity of travel by mode by duration of trip, which are needed
-to calculate the proportional allocation factors \(F_{ij}^m\) for
+the propensity of travel by mode by duration of trip, and are used to
+calculate the proportional allocation factors \(F_{ij}^m\) for
 \(V_i^m\).
@@
 
-{\centering \includegraphics[width=1\linewidth]{images/tlds_curves_m_plot}
+{\centering \includegraphics[width=0.85\linewidth]{images/tlds_curves_m_plot}
 
@@
 
-\caption{\label{fig:Fig4} Fitted impedance function curve (line) against empirical TLD (bars) corresponding to the home-to-work origin-destination flows from the Madrid 2018 travel survey.}\label{fig:tlds-curves-m-plot}
+\caption{\label{fig:Fig4} Fitted impedance function  against empirical TLD (bars) corresponding to the home-to-work origin-destination flows from the Madrid 2018 travel survey.}\label{fig:tlds-curves-m-plot}
 \end{figure}
@@
 At this point it is worthwhile to reiterate that the empirical example
-is a snap-shot of the calculated multimodal spatial availability using
-data from the 2018 travel survey. It is intended to visualize the
-spatial trends in availability of employment opportunities, by mode, and
-demonstrate how spatial availability can be interpreted to discuss the
-competitive advantage of lower travel impedance modes within Madrid
-Centro.
-
-The spatial availability of jobs \(V_i^m\) is calculated for each of the
-four modes \(m\) at the level of traffic analysis zones \(i\) in Madrid
-and displayed in Figure \ref{fig:Fig5}.
+is a snap-shot of spatial availability by mode using data from the 2018
+travel survey. Our purpose is to investigate the trends in availability
+of employment opportunities by mode, and illustrate how spatial
+availability can help in discussions about the competitive advantage of
+various modes within Madrid Centro. The spatial availability of jobs
+\(V_i^m\) is calculated for each of the four modes \(m\) at the level of
+traffic analysis zones \(i\) in Madrid and displayed in Figure
+\ref{fig:Fig5}.
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{images/SA_im_V_zn208_plot}
+{\centering \includegraphics[width=0.85\linewidth]{images/SA_im_V_zn208_plot}
 
@@
 
-\caption{\label{fig:Fig5} Spatial availability of job opportunities per origin and mode $V_i^m$ in Madrid.
-Calculated using the home-to-work flows from the 2018 travel survey.
-Gray TAZs have no population.}\label{fig:SA-m-plot}
+\caption{\label{fig:Fig5} Spatial availability of jobs per origin and mode $V_i^m$ in Madrid. Gray TAZs have no population. Ranges of values in the legend are quintiles.}\label{fig:SA-m-plot}
 \end{figure}
 
-\(V_i^m\) is a proportion of the total number of 847,574 jobs in the
-region and is shown in Figure \ref{fig:Fig5}. Since \(V_i^m\) is
-calculated based on the population of workers and the distribution of
-jobs, the values can be understood as the number of full-time jobs that
-are spatially available to the full-time working population at that
-\(i\) and their associated \(m\), relative to all the jobs in the city.
-\(V_i^m\) is the number of jobs that are \emph{spatially available} to a
-\(m\)-using population located at \(i\), relative to the travel
-impedance and size of \emph{all} populations in the region.
-
-Some notable differences in the magnitude of \(V_i^m\) occur between
-modes as seen in Figure \ref{fig:Fig5}. The majority of \(V_i^m\) (which
-is to say of spatially available jobs) is allocated to those travelling
-by car and transit. This is to be expected, as the users of these modes
-represent 84.1\% of the total population. Differences in \(V_i^m\)
-values within modes also exist in space: car users outside of the M-30
-region appear to have greater \(V_i^m\) values, while some zones inside
-the M-30 appear to have higher \(V_i^m\) values for transit. Overall,
-the magnitude of \(V_i^m\) values for cyclists and pedestrians are lower
+In the figure, \(V_i^m\) is a proportion of the total number of 847,574
+jobs in the region. Since \(V_i^m\) is calculated based on the
+population of workers and the distribution of jobs, the values can be
+understood as the number of full-time jobs that are spatially available
+to full-time workers at that \(i\) traveling by mode \(m\), relative to
+all the jobs in the city. There are noticeable differences in the
+magnitude of \(V_i^m\) between modes as seen in Figure \ref{fig:Fig5}.
+The majority of \(V_i^m\) (which is to say of spatially available jobs)
+are allocated to workers travelling by car and transit. In a way, this
+is to be expected since users of these modes represent 84.1\% of the
+total population. However, the ability to travel at greater speeds also
+impacts this results. Fuerthermore, differences in \(V_i^m\) values
+within modes also exist in space: car users outside of the M-30 region
+appear to enjoy greater spatial availability, while some zones inside
+the M-30 to have greater spatial availability for transit. Overall, the
+magnitude of \(V_i^m\) values for cyclists and pedestrians are lower
 than for car and transit but the highest values of \(V_i^{bike}\) and
-\(V_i^{walk}\) tend to be allotted to zones within the M-30 and origins
+\(V_i^{walk}\) tend to be found in zones within the M-30 and origins
 with higher spatial availability by transit.
@@
 The differences between the shares of modes and their shares of
-spatially availabile opportunities highlights the competitive advantage
-afforded by certain modes, although this effect is not uniform in space.
-As seen in the left-most columns of Figure \ref{fig:Fig6}, `car/motor'
-and `transit' together can avail themselves of 95.3\% of all jobs in the
+spatially available opportunities highlights the competitive advantage
+of certain modes, although this effect is not geographically uniform. As
+seen in the left-most columns of Figure \ref{fig:Fig6}, `car/motor' and
+`transit' together can avail themselves of 95.3\% of all jobs in the
 city. However, `car/motor' has a disproportionate share of \(V_i^m\)
@@
 
-{\centering \includegraphics[width=1\linewidth]{images/modal_V_comps_4plot}
+{\centering \includegraphics[width=0.85\linewidth]{images/modal_V_comps_4plot}
 
@@
 
-\caption{\label{fig:Fig6} Displays the proportion of the working population by mode and spatial availability of job opportunities by mode aggregated for three spatial areas.
-From left to right, the city of Madrid (All), the area within the M-30 highway (M-30),the area within the Centro region (Centro).}\label{fig:modal-V-comps-plot}
+\caption{\label{fig:Fig6} Proportion of population by mode and spatial availability of jobs by mode aggregated for three areas.
+From left to right, the city of Madrid (All), the area within the M-30 highway (M-30), the area within the Centro region (Centro).}\label{fig:modal-V-comps-plot}
 \end{figure}
 
-Despite the big picture, it is important to note that there are spatial
-variations in the competitive advantage of cars. The proportion of
-travel by car in Centro is smaller and has higher travel impedance
-values relative to the inputs in other areas and by other modes. The LEZ
-Centro implementation further restricts the advantage of cars by
-shifting more than half of all car trips into the LEZ to modes {[}41{]}.
-This restriction decreased the number of trips by car from going into
-the LEZ Centro (an area with a large number of jobs overall, see Figure
-\ref{fig:Fig2}), thus increasing the mass effect for non-car modes and
-resulting in proportionally higher \(v_i^m\) values for non-car modes.
-As such, the lower amount of access to opportunities by car allows more
-opportunities in the LEZ to be available to populations using other
-modes.
-
-As summarized in the two right-most columns in Figure \ref{fig:Fig6},
-the proportion of jobs spatially available to car in Centro is (13.8\%
-or 5,373 opportunities). For reference, this is less than the proportion
-of the car users in Centro (16.1\%), evidently less then the proportion
-of car users in the city, and is the opposite of the overall trend
-(left-most columns) and within the M-30 (middle columns). More
-opportunities are spatially available to non-car users within Centro,
-particularly by transit, a mode with 68.5\% of spatially available jobs
-in Centro, despite representing only 51.4\% of the population in Centro
-and 47.5\% in the city overall.
-
-Figure \ref{fig:Fig6} also shows that there is a higher proportion of
+The big picture demonstrates how car is the most advantageous mode,
+however it is interesting to notice that this advantage appears to be
+blunted by the LEZ. Unlike the results for the whole city and even
+within the M-30 ring, the proportion of car users in Centro is larger
+than the proportion of opportunities spatially available to them. The
+restriction on cars in effect reduces competition by this mode, and
+leads to a relative increment of the mass effect of the modes allowed
+within the LEZ, which also contains a large number of jobs (see Figure
+\ref{fig:Fig2}). As summarized in the two right-most columns in Figure
+\ref{fig:Fig6}, the proportion of jobs spatially available to car in
+Centro is (13.8\% or 5,373 opportunities). For reference, this is less
+than the proportion of the car users in Centro (16.1\%), evidently less
+then the proportion of car users in the city, and is the opposite of the
+overall trend (left-most columns) and within the M-30 (middle columns).
+
+It is also clear that more opportunities are spatially available to
+non-car users within Centro. In the case of active travel, the
+proportions of cyclists and walkers within LEZ Centro still exceed the
+proportions of jobs spatially available to them --- however the
+disparity is drastically reduced compared to the rest of the city. As
+seen in Figure \ref{fig:Fig6}, there is a higher proportion of
 opportunities that are spatially available to pedestrians and cyclists
-in Centro than in the City overall and in all areas within the M-30.
+in Centro than in the City overall and in areas within the M-30.
 Notably, within Centro, 1.7\% and 16.1\% of opportunities are spatially
 available to bike and walk modes respectively, while their populations
-represent smaller proportions of 1.2\% and 14.7\% of the population.
-Although the proportion of opportunities that are spatially available to
-these modes is still lower than the respective shares of the population
-located in the Centro, these modes are more competitive within the
-Centro than anywhere without. By restricting the ability of cars to
-enter Centro, the LEZ also contributes to level the playing field for
-slower modes, in particular cycling and walking.
+represent 1.2\% and 14.7\% of the population. By restricting the ability
+of cars to enter Centro, the LEZ seems to contributes to level the
+playing field for slower modes, in particular cycling and walking, but
+also transit. As seen in the figure, transit users are generally close
+to parity across the region, with nearly as many spatially available
+jobs as transit users. Still, this mode has the greatest advantage in
+LEZ Centro with 68.5\% of spatially available jobs in Centro for 51.4\%
+of transit users in Centro. This result makes intuitive sense: after
+car, it is the mode with the greatest range, and unlike car it is
+allowed in LEZ Centro.
 
 The spatial differences in the competitive dis/advantage of spatial
-availability between modes can also be visualized per origin. Figure
-\ref{fig:Fig7} shows \(v_i^m\), the spatial availability \(V_i^m\)
-divided by the population that uses \(m\). Values of \(v_i^m\) above 1
-are shown in red and represent more than one spatially available
-opportunity per capita; values below 1 are shown in green, and represent
-less than one spatially available opportunity per capita; values equal
-to 1 are shown in white. These plots drive home the point of the
-disproportionately high over-allocation of spatial availability relative
-to the share of modes in many of the origins for car (bottom left plot,
-areas denoted with green \(v_i^m\) values above 1). These plots also
-show areas with disproportionately few spatially available opportunities
-(under 1). It can be observed that the spatial availability to jobs of
-transit users is relatively well balanced (i.e., many zones are white),
-while that for non-motorized modes is low (under 1) overall.
+availability between modes can also be visualized at a finer level of
+granularity. Figure \ref{fig:Fig7} shows \(v_i^m\), the spatial
+availability \(V_i^m\) divided by the population of users of \(m\).
+Values of \(v_i^m\) below 1 are shown in shades of orange, and indicate
+TAZs with less than one spatially available opportunity per capita for
+the mode. Values above 1 are shown in shades of green, and indicate TAZs
+with more than one spatially available opportunity per capita for the
+mode. The highest spatial availability per capita (shown in blue) is for
+car users in a zone northeast just beyond the M-30. These plots
+illustrate in unambiguous fashion, and in a quantity that is comparable
+over space and time, the advantage in terms of spatial availability of
+car for most of the city (bottom left plot, areas denoted with green
+\(v_i^m\) values above 1). It can also be observed that spatial
+availability of jobs is relatively well balanced for transit users over
+most of the regions (i.e., many zones are light orange or light green).
+Spatial availability of jobs for non-motorized modes, in contrast, is
+low (under 1) overall, although less so within LEZ Centro.
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{images/SA_im_vv_zn208_plot}
+{\centering \includegraphics[width=0.85\linewidth]{images/SA_im_vv_zn208_plot}
 
@@
 
-\caption{\label{fig:Fig7} Spatial availability of job opportunities per mode-using capita by mode $v_i^m$ per origin in Madrid.
-Calculated using the home-to-work flows from the 2018 travel survey.
-Gray TAZs have no population.}\label{fig:SA-per-capita-m-plot}
+\caption{\label{fig:Fig7} Distribution of spatially available jobs per capita by mode of transportatin ($v_i^m$).
+Gray TAZs have no population that use the mode.
+Ranges of values in the legend are quintiles.}\label{fig:SA-per-capita-m-plot}
 \end{figure}
 
-It is also interesting to note that \(v_i^m\) for car within and near
-LEZ Centro is close to or below 1 (white/red) in Figure \ref{fig:Fig7},
-while in contrast all non-car modes have relatively higher \(v_i^m\)
-values. Since the values are comparable across regions and over time,
-Figure \ref{fig:Fig7} provides a benchmark for quantifying potential LEZ
-implementations in the future. As Figure \ref{fig:Fig7} also shows, many
+Incidentally, it is bears noting that \(v_i^m\) for car within and near
+LEZ Centro is close to or below 1 in Figure \ref{fig:Fig7}, while all
+non-car modes have relatively higher \(v_i^m\) values. Since these
+values are comparable across regions and over time, Figure
+\ref{fig:Fig7} potentially provides a benchmark for quantifying changes
+in LEZ policies in the future. As Figure \ref{fig:Fig7} also shows, many
 areas within the M-30 have high (white/green) \(v_i^m\) values for car,
@@
 a spatial expansion of the LEZ to include all areas within the M-30
-stands to increase the spatial availability of jobs for more transit,
-cyclcing and walking.
+would likely increase the spatial availability of jobs for transit
+users, cyclists and pedestrians.
 
@@
 
-Location-based accessibility measures like the Hansen-type \(S_i^m\),
-Shen-type \(a_i^m\), and spatial availability \(V_i^m\) measures share a
-commonality; they are a weighted sum of opportunities assigned to each
-spatial unit \(i\) in a region. In this way, they all can be interpreted
-as a score of how many opportunities can be potentially interacted with
-by the population at \(i\). How the weight and sum of the
-potentially-interacted-with opportunities is considered is what defines
-the type of accessibility measure.
-
-Within this paper, the location-based singly- \emph{constrained} and
-\emph{competitive} accessibility measure, known as spatial availability
-\(V_i\) {[}14{]}, is extended for the case of capturing multimodal
-accessibility to opportunities \(V_i^m\). A synthetic example and then
-an empirical case of LEZ in Madrid are detailed to demonstrate this
-multimodal extension.
-
-The spatial availability measure is capable of capturing a new
-interpretation of multimodal competition that previous accessibility
-measures have not yet done. Competitive measures hypothesis that
-populations using modes with lower travel impedance, when competing for
-a finite set of opportunities, will capture more opportunities. With
-spatial availability, the number of opportunities that are captured (of
-the total opportunities in the region) by each mode can be individually
-calculated. From there, the difference between how many spatially
-available opportunities one mode captures versus another can be
-investigated. This is the advantage of the spatial availability measure,
-particularly its multimodal extension.
-
-The flexibility and need for an accessibility measure such as spatial
-availability is pertinent in policy scenario evaluation. As showcased in
-the empirical example of the LEZ in Madrid, competition for job
-opportunity availability varies spatially \emph{as well as} between
-modes. The car and transit modes have the highest spatial availability,
-with the car-mode having highest availability with exception to the
-areas within the LEZ Centro. Since car travel has been highly restricted
-within the LEZ Centro, fewer car-using people potentially interact with
-jobs within the LEZ Centro, leaving more \emph{spatially available} jobs
-for non-car-using populations. This difference in car-using populations
-in locations accessing jobs within and immediately outside the LEZ
-Centro increases the competitiveness of the transit-using population
-(the second most competitive mode) as well as the non-motorized modes.
-
-Spatial availability \(V_i^m\) can also be divided by the mode-using
-population at each \(i\) to yield mode-population normalized values.
-These values, reflected in Figure \ref{fig:Fig7}, can be used as a
-benchmark to investigate existing conditions and plan future LEZ
-implementation (i.e., target areas with exceptionally high car spatial
-availability such that more opportunities are available to other
-mode-users).
-
-In summary, conventional \emph{non-constrained} accessibility measures
-are difficult for planners to operationalize for a variety of reasons
-including issues of computation and interpretability {[}5{]}. With
-spatial availability, the magnitude of opportunities that are available
-as a proportion of all the opportunities in the region is equal to
-\(V_i\). As a result of its proportional allocation mechanism, \(V_i\)
-can be naturally extended into multimodal applications. This flexibility
-is helpful to modelling policy scenarios in our cities that are
-increasingly multimodal. The interpretation of \(V_i\) allows for
-manipulation of \(V_i^m\) values to investigate differences of
-availability between neighbourhoods, modes, and regions, generate per
-capita benchmarks, and/or generate average values per population-group.
-
-From a spatial equity perspective, spatial availability measure can
-provide researchers, policy makers, and citizens a new-found
-interpretation of accessibility measures. With a plot of spatial
-availability values, one can begin asking, how much is enough and what
-level may be too much. These interpretations were difficult to be made
-with accessibility measures in the past. In future work, we intend to
-use multimodal spatial availability to characterise the equity of
-multimodal accessibility losses and gains in varying LEZ policy scenario
-implementations.
+Accessibility measures are an important tool in transportation research
+{[}9{]} and increasingly also seen as valuable for planning purposes
+{[}4--8{]}. They boast a long history of development, beginning with
+Hansen-type \(S_i^m\) measures, with other developments like Shen's
+\(a_i^m\), to account for competition/congestion. The more recent
+spatial availability measure \(V_i^m\) has in common with these
+accessibility indicators that it is a weighted sum of the opportunities
+in a region from the perspective of a determined origin \(i\).
+Aggregations of opportunities embody principles of gravitational/spatial
+interaction modelling that date back to at least H.C. Carey {[}49{]},
+and are part of a line of research that includes the work of Ravenstein
+{[}50{]}, Reilly {[}51{]}, Stewart {[}52--54{]}, Zipf {[}55,56{]},
+Wilson {[}37{]}, and many others besides. In this way, \(S_i^m\),
+\(a_i^m\), and yes, \(V_i^m\), can be interpreted as scores of the
+potential for interaction with opportunities in space.
+
+Different accessibility indicators are characterized by how they weight
+and aggregate opportunities. Spatial availability's contribution to the
+literature is to incorporate a proportional allocation mechanism that
+essentially constrains the sums to match the number of opportunities in
+the region; in this way it is a singly-constrained accessibility measure
+that naturally accommodates congestion and competition. The effort with
+spatial availability is in line with previous research on proportional
+allocation by Paez et al. {[}14{]}. As initially introduced by Soukhov
+et al. {[}15{]}, spatial availability was designed for a homogeneous
+population traveling by a single mode of transportation. In this paper
+we extended spatial availability for the case of heterogeneous
+populations. We discussed this in terms of multiple modes of
+transportation, but the framework can accommodate equally well
+variations in travel behavior by population segments.
+
+An empirical example using data from Madrid helped to illustrate the
+potential of multimodal spatial availability analysis, including its
+ability to account for competition for opportunities within and between
+modes. Particularly relevant is the fact that spatial availability
+scores relate directly to the total number of opportunities in the
+region. This makes it possible to compare the results to intuitive
+benchmarks, such as opportunities per population, in ways that other
+accessibility measures cannot or tend to obfuscate. This comparability
+is preserved between regions and over time. The example suggests that
+once that opportunities are treated as being finite, restrictions to
+travel by car leave more spatially available opportunities for
+non-car-users. This difference for car travel in locations within and
+immediately around the LEZ Centro seems to increase the number of
+opportunities spatially available to transit users (transit being the
+second most competitive mode), as well as the spatial availability from
+the perspective of non-motorized modes. In effect, a policy such as Low
+Emission Zones helps to improve the accessibility situation of active
+travel and transit in the parts of the city where it is implemented.
+
+The purpose of the empirical example was to illustrate the kind of
+insights that can be derived from the application of multimodal spatial
+availability. But there are some intriguing opportunities for future
+research. Accessibility indicators are not designed to work as modal
+split models, and yet, in the case of policies that alter the relative
+cost of various forms of transportation, one can reasonably expect to
+see some shifts between modes. In our empirical example we used data
+collected \emph{after} the introduction of LEZ Centro. However, given a
+modal split model to project model shares, accessibility indicators,
+including spatial availability, can be used to investigate changes to
+the accessibility landscape. Ditto for destination choice. Our empirical
+example in this paper presented but a snapshot of this, and in future
+research it will be interesting to investigate changes \emph{between}
+policy interventions. The expansion of Madrid's LEZ to the ring
+contained by the M-30 orbital presents an excellent opportunity to do
+so. Given the intuitive and straightforward interpretation of spatial
+availability scores as fractions of opportunities from the total,
+relative and absolute changes in the accessibility landscape can be
+assessed, thus helping to evaluate the implications of policy
+interventions.
+
+Finally, our example dealt with differences in travel by mode only, but
+it is possible to think of the intersection between mode of travel and
+different types of travelers. This would expand the number of
+sub-populations in the analysis from, say, \(m=M\) (modes) to
+\(m = M\cdot Q\) (modes times population segments), each with their own
+characteristic impedance function. Evaluations of this kind will be
+especially relevant as LEZ are implemented in cities globally, and the
+question of their impact on disadvantaged populations who have become
+mobility-restricted increasingly come to the fore {[}40,57,58{]}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-shiLiterature2020}{}}%
+\CSLLeftMargin{9. }%
+\CSLRightInline{Shi Y, Blainey S, Sun C, Jing P. A literature review on
+accessibility using bibliometric analysis techniques. Journal of
+Transport Geography. 2020;87: 102810.
+doi:\href{https://doi.org/10.1016/j.jtrangeo.2020.102810}{10.1016/j.jtrangeo.2020.102810}}
+
 \leavevmode\vadjust pre{\hypertarget{ref-handyMeasuring1997}{}}%
-\CSLLeftMargin{9. }%
+\CSLLeftMargin{10. }%
 \CSLRightInline{Handy S, Niemeier D. Measuring accessibility: An
@@
 \leavevmode\vadjust pre{\hypertarget{ref-kwanSpace1998}{}}%
-\CSLLeftMargin{10. }%
+\CSLLeftMargin{11. }%
 \CSLRightInline{Kwan MP. Space-time and integral measures of individual
@@
 \leavevmode\vadjust pre{\hypertarget{ref-shenLocationCharacteristicsInnercity1998}{}}%
-\CSLLeftMargin{11. }%
+\CSLLeftMargin{12. }%
 \CSLRightInline{Shen Q. Location characteristics of inner-city
@@
 \leavevmode\vadjust pre{\hypertarget{ref-luoMeasuresSpatialAccessibility2003}{}}%
-\CSLLeftMargin{12. }%
+\CSLLeftMargin{13. }%
 \CSLRightInline{Luo W, Wang F. Measures of spatial accessibility to
@@
 \leavevmode\vadjust pre{\hypertarget{ref-paezDemand2019}{}}%
-\CSLLeftMargin{13. }%
+\CSLLeftMargin{14. }%
 \CSLRightInline{Paez A, Higgins CD, Vivona SF. Demand and level of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-soukhovIntroducingSpatialAvailability2023}{}}%
-\CSLLeftMargin{14. }%
+\CSLLeftMargin{15. }%
 \CSLRightInline{Soukhov A, Paez A, Higgins CD, Mohamed M. Introducing
@@
 \leavevmode\vadjust pre{\hypertarget{ref-paezHealth2010}{}}%
-\CSLLeftMargin{15. }%
+\CSLLeftMargin{16. }%
 \CSLRightInline{Páez A, Mercado R, Farber S, Morency C, Roorda M.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-moniruzzamanMode2013}{}}%
-\CSLLeftMargin{16. }%
+\CSLLeftMargin{17. }%
 \CSLRightInline{Moniruzzaman M, Páez A, Nurul Habib KM, Morency C. Mode
@@
 \leavevmode\vadjust pre{\hypertarget{ref-reyesWalking2014}{}}%
-\CSLLeftMargin{17. }%
+\CSLLeftMargin{18. }%
 \CSLRightInline{Reyes M, Paez A, Morency C. Walking accessibility to
@@
 \leavevmode\vadjust pre{\hypertarget{ref-paezJobs2013}{}}%
-\CSLLeftMargin{18. }%
+\CSLLeftMargin{19. }%
 \CSLRightInline{Páez A, Farber S, Mercado R, Roorda M, Morency C. Jobs
@@
 \leavevmode\vadjust pre{\hypertarget{ref-wuUnifyingAccess2020}{}}%
-\CSLLeftMargin{19. }%
+\CSLLeftMargin{20. }%
 \CSLRightInline{Wu H, Levinson D. Unifying access. Transportation
@@
 \leavevmode\vadjust pre{\hypertarget{ref-tahmasbiMultimodalAccessibilitybasedEquity2019}{}}%
-\CSLLeftMargin{20. }%
+\CSLLeftMargin{21. }%
 \CSLRightInline{Tahmasbi B, Mansourianfar MH, Haghshenas H, Kim I.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-lunkeModalAccessibilityDisparities2022}{}}%
-\CSLLeftMargin{21. }%
+\CSLLeftMargin{22. }%
 \CSLRightInline{Lunke EB. Modal accessibility disparities and transport
@@
 \leavevmode\vadjust pre{\hypertarget{ref-paezRelative2010}{}}%
-\CSLLeftMargin{22. }%
+\CSLLeftMargin{23. }%
 \CSLRightInline{Paez A, Mercado RG, Farber S, Morency C, Roorda M.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-campbell2019accessibility}{}}%
-\CSLLeftMargin{23. }%
+\CSLLeftMargin{24. }%
 \CSLRightInline{Campbell KB, Rising JA, Klopp JM, Mbilo JM.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-maharjanSpatialEquityModal2022}{}}%
-\CSLLeftMargin{24. }%
+\CSLLeftMargin{25. }%
 \CSLRightInline{Maharjan S, Tilahun N, Ermagun A. Spatial equity of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-grengsJobAccessibilityModal2010}{}}%
-\CSLLeftMargin{25. }%
+\CSLLeftMargin{26. }%
 \CSLRightInline{Grengs J. Job accessibility and the modal mismatch in
@@
 \leavevmode\vadjust pre{\hypertarget{ref-kawabataJobAccessibilityIndicator2006}{}}%
-\CSLLeftMargin{26. }%
+\CSLLeftMargin{27. }%
 \CSLRightInline{Kawabata M, Shen Q. Job accessibility as an indicator of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-kwokUseModalAccessibility2004}{}}%
-\CSLLeftMargin{27. }%
+\CSLLeftMargin{28. }%
 \CSLRightInline{Kwok RCW, Yeh AGO. The use of modal accessibility gap as
@@
 \leavevmode\vadjust pre{\hypertarget{ref-morrisAccessibilityIndicatorsTransport1979}{}}%
-\CSLLeftMargin{28. }%
+\CSLLeftMargin{29. }%
 \CSLRightInline{Morris JM, Dumble PL, Wigan MR. Accessibility indicators
@@
 \leavevmode\vadjust pre{\hypertarget{ref-weibullAxiomaticApproachMeasurement1976}{}}%
-\CSLLeftMargin{29. }%
+\CSLLeftMargin{30. }%
 \CSLRightInline{Weibull JW. An axiomatic approach to the measurement of
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-liottaPlanning2020}{}}%
+\CSLLeftMargin{31. }%
+\CSLRightInline{Liotta C, Kervinio Y, Levrel H, Tardieu L. Planning for
+environmental justice - reducing well-being inequalities through urban
+greening. Environmental Science \& Policy. 2020;112: 47--60.
+doi:\url{https://doi.org/10.1016/j.envsci.2020.03.017}}
+
 \leavevmode\vadjust pre{\hypertarget{ref-OECDFrameworks2013}{}}%
-\CSLLeftMargin{30. }%
+\CSLLeftMargin{32. }%
 \CSLRightInline{OECD. Frameworks and sector policies for urban
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-laraSpace2015}{}}%
+\CSLLeftMargin{33. }%
+\CSLRightInline{Lara-Valencia F, García-Pérez H. Space for equity:
+Socioeconomic variations in the provision of public parks in hermosillo,
+mexico. Local Environment. 2015;20: 350--368.
+doi:\href{https://doi.org/10.1080/13549839.2013.857647}{10.1080/13549839.2013.857647}}
+
 \leavevmode\vadjust pre{\hypertarget{ref-martensFair2021}{}}%
-\CSLLeftMargin{31. }%
+\CSLLeftMargin{34. }%
 \CSLRightInline{Martens K, Golub A. A fair distribution of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-josephMeasuringPotentialPhysical1982}{}}%
-\CSLLeftMargin{32. }%
+\CSLLeftMargin{35. }%
 \CSLRightInline{Joseph AE, Bantock PR. Measuring potential physical
@@
 \leavevmode\vadjust pre{\hypertarget{ref-taoInvestigatingImpactsPublic2020a}{}}%
-\CSLLeftMargin{33. }%
+\CSLLeftMargin{36. }%
 \CSLRightInline{Tao Z, Zhou J, Lin X, Chao H, Li G. Investigating the
@@
 \leavevmode\vadjust pre{\hypertarget{ref-wilsonFamilty1971}{}}%
-\CSLLeftMargin{34. }%
+\CSLLeftMargin{37. }%
 \CSLRightInline{Wilson AG. A family of spatial interaction models, and
@@
 \leavevmode\vadjust pre{\hypertarget{ref-carpentieriMultimodalAccessibilityPrimary2020}{}}%
-\CSLLeftMargin{35. }%
+\CSLLeftMargin{38. }%
 \CSLRightInline{Carpentieri G, Guida C, Masoumi HE. Multimodal
@@
 \leavevmode\vadjust pre{\hypertarget{ref-margaryanLowEmissionZones2021}{}}%
-\CSLLeftMargin{36. }%
+\CSLLeftMargin{39. }%
 \CSLRightInline{Margaryan S. Low emission zones and population health.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-verbeekJustManagementUrban2022}{}}%
-\CSLLeftMargin{37. }%
+\CSLLeftMargin{40. }%
 \CSLRightInline{Verbeek T, Hincks S. The {``just''} management of urban
@@
 doi:\href{https://doi.org/10.1016/j.apgeog.2022.102642}{10.1016/j.apgeog.2022.102642}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-espanaPlanNacionalIntegrado2020}{}}%
-\CSLLeftMargin{38. }%
-\CSLRightInline{España. Plan nacional integrado de energía y clima
-({PNIEC}) 2021-2030. 2020 {[}cited 30 Jul 2023{]}. Available:
-\url{https://www.miteco.gob.es/es/prensa/pniec.html}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-espanaResolucion10Enero2020}{}}%
-\CSLLeftMargin{39. }%
-\CSLRightInline{España. Resolución de 10 de enero de 2020, de la
-dirección general de biodiversidad y calidad ambiental, por la que se
-publica el programa nacional de control de la contaminación atmosférica.
-Resolución Jan 24, 2020 pp. 6947--6947. Available:
-\url{https://www.boe.es/eli/es/res/2020/01/10/(10)}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-barcelonaGUIATECNICAPARA2021}{}}%
-\CSLLeftMargin{40. }%
-\CSLRightInline{Barcelona. {GUÍA} {TÉCNICA} {PARA} {LA} {IMPLEMENTACIÓN}
-{DE} {ZONAS} {DE} {BAJAS} {EMISIONES}. Àrea Metropolitana de Barcelona
-({AMB}): Oficina Técnica de Gerencia del {AMB}; 2021. Available:
-\url{https://revista.dgt.es/images/GUIA-ZBE.pdf}}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-careyPriciples1858}{}}%
+\CSLLeftMargin{49. }%
+\CSLRightInline{Carey HC. Principles of social science. Philadelphia:
+J.B. Lippincot; Co.; 1858. }
+
+\leavevmode\vadjust pre{\hypertarget{ref-ravensteinLaws1889}{}}%
+\CSLLeftMargin{50. }%
+\CSLRightInline{Ravenstein EG. The laws of migration. Journal of the
+Royal Statistical Society. 1889;52: 241--305.
+doi:\href{https://doi.org/10.2307/2979333}{10.2307/2979333}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-reillyMethods1929}{}}%
+\CSLLeftMargin{51. }%
+\CSLRightInline{Reilly WJ. Methods for the study of retail
+relationships. 1929. }
+
+\leavevmode\vadjust pre{\hypertarget{ref-stewartInverse1941}{}}%
+\CSLLeftMargin{52. }%
+\CSLRightInline{Stewart JQ. An inverse distance variation for certain
+social influences. Science. 1941;93: 89--90. Available:
+\url{http://www.jstor.org/stable/1668130}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-stewartSuggested1947}{}}%
+\CSLLeftMargin{53. }%
+\CSLRightInline{Stewart JQ. Suggested principles of ''social physics''.
+Science. 1947;106: 179--180. Available:
+\url{http://www.jstor.org/stable/1675368}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-stewartDemographic1948}{}}%
+\CSLLeftMargin{54. }%
+\CSLRightInline{Stewart JQ. Demographic gravitation: Evidence and
+applications. Sociometry. 1948;11: 31--58.
+doi:\href{https://doi.org/10.2307/2785468}{10.2307/2785468}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-zipfHypothesis1946}{}}%
+\CSLLeftMargin{55. }%
+\CSLRightInline{Zipf GK. The p 1 p 2/d hypothesis: The case of railway
+express. The Journal of Psychology. 1946;22: 3--8.
+doi:\href{https://doi.org/10.1080/00223980.1946.9917292}{10.1080/00223980.1946.9917292}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-zipfIntercity1946}{}}%
+\CSLLeftMargin{56. }%
+\CSLRightInline{Zipf GK. The P1P2/d hypothesis: On the intercity
+movement of persons. American Sociological Review. 1946;11: 677--686.
+doi:\href{https://doi.org/10.2307/2087063}{10.2307/2087063}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-devrijNooneVisitsMe2022}{}}%
+\CSLLeftMargin{57. }%
+\CSLRightInline{De Vrij E, Vanoutrive T. {``No-one visits me anymore''}:
+Low emission zones and social exclusion via sustainable transport
+policy. 2022 {[}cited 27 Jul 2023{]}. Available:
+\url{https://www.tandfonline.com/doi/epdf/10.1080/1523908X.2021.2022465?needAccess=true\&role=button}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-liottaWhatDrivesInequalities2023}{}}%
+\CSLLeftMargin{58. }%
+\CSLRightInline{Liotta C. What drives inequalities in low emission
+zones'impacts on job accessibility? Virtual; 2023.
+doi:\href{https://doi.org/10.17605/OSF.IO/9VQU7}{10.17605/OSF.IO/9VQU7}}
+
 \end{CSLReferences}

### 2023-11-09 README.Rmd
Updated README

--- 2022-04-01 README.Rmd
@@
 
-# Madrid zonas de bajas emisiones (low emission zones)
-
-The repository is home to the analysis of low emissions zones in Madrid using spatial availability measure.
-
-Currently, the following raw data from Madrid is contained here:
-
--  "parque_vehiculos_2021"  includes vehicle fleet characteristics of Madrid City. Of interest: the district and the environmental label associated to each vehicle.
-
-- several excel documents from the Home Mobility Survey of the Community of Madrid.
-These documents include information like citizens' socioeconomic characteristics, origin and destination of the trip, transport mode, distance, time required, activities carried out at origin and destination, etc.
-
-- A figure of Madrid spatial context.
+# Multimodal spatial availability: a singly-constrained measure of accessibility considering multiple modes
 
@@
 <!-- badges: end -->
+
+Anastasia Soukhov (McMaster University)
+Javier Tarriño-Ortiz (Universidad Politécnica de Madrid)
+Julio Soria-Lara (Universidad Politécnica de Madrid)
+Antonio Páez (McMaster University)
+
+Recent research has aimed to address the way opportunities are counted in accessibility analysis.
+In conventional accessibility measures, opportunities are often multiply counted, which leads to values of accessibility that are difficult to interpret.
+Constraining the calculations to match a known quantity ensures that the measurements sum up to a predetermined quantity (i.e., the total number of opportunities), and so each value can be meaningfully related to this total.
+A recent such effort is spatial availability, a singly-constrained accessibility measure.
+In this paper we extend spatial availability for use in the case of multiple modes or, more generally, heterogeneous population segments with distinct travel behaviors.
+After deriving a multimodal version of spatial availability We proceed to illustrate its features using a synthetic example.
+Next, we apply it to an empirical example in Madrid.
+We conclude the paper with suggestions for future research.
+
+Submitted to PLoS ONE.

### 2023-11-10 Journal-manuscript/Editorial correspondance/Response-to-reviewers_R1.Rmd
Finalized revisions and comments to reviewers (R1)

@@
 output: pdf_document
+urlcolor: blue
 date: "2023-10-18"
@@
 The article extends the concept of spatial accessibility and applies it to a fairness analysis of different modes of transportation.
-The article uses travel data from a week in Madrid as an example to analyze spatial accessibility in the city.
+The article uses travel data from a week in Madrid as an example to analyse spatial accessibility in the city.
 However, the article has the following issues:
@@
 \textcolor{blue}{Thank you for this comment.
-We would like to bring to your attention the editorial philosophy of PLoS ONE, according to which editors "make decisions on submissions based on scientific rigor, regardless of novelty," (see https://journals.plos.org/plosone/s/editorial-and-peer-review-process).
-Nonetheless we wish to respond to this comment.}
+We would like to bring to your attention the editorial philosophy of PLoS ONE, according to which editors "make decisions on submissions based on scientific rigor, regardless of novelty," (see } \textcolor{blue}{\underline{https://journals.plos.org/plosone/s/editorial-and-peer-review-process)}}.
+\textcolor{blue}{Nonetheless we wish to respond to this comment.}
 
-\textcolor{blue}{In this paper we extend our earlier work on [spatial availability](https://doi.org/10.1371/journal.pone.0278468) for the simultaneous analysis of multiple modes of transportation.
+\textcolor{blue}{In this paper we extend our earlier work on spatial availability (}\textcolor{blue}{\underline{https://doi.org/10.1371/journal.pone.0278468}}\textcolor{blue}{) for the simultaneous analysis of multiple modes of transportation.
 You appear to be under the misaprehension that the main difference with accessibility measures is to incorporate "the ratio of travel frequencies as weight in travel costs".
@@
 
-\textcolor{blue}{While this enhancement may appear minor to you, it is not plainly obvious how spatial availability applies to multiple modes, which is why we think this paper is needed.
+\textcolor{blue}{While this enhancement may appear minor in your opinion, it is not plainly obvious how spatial availability applies to multiple modes, which is why we believe this paper is needed.
 We do not dispute that this is an incremental step in the development of a more general method, but it is a step that considerably expands the range of potential applications.
-Further, we contend that the method has been rigorously developed and demonstrated using an open, transparent, and reproducible example.}
+Further, we contend that the method has been rigorously developed and demonstrated this using an open, transparent, and reproducible example.}
 
 \textcolor{blue}{In response to this comment we have edited the paper to more clearly describe the advantages that multimodal spatial availability offers when considering the accessibility to opportunities by different modes.
-In particular, proportional allocation of opportunities using mode-specific impedance functions means that the travel-cost-advantage of each mode can be analyzed.
+In particular, proportional allocation of opportunities using mode-specific impedance functions means that the travel-cost-advantage of each mode can be analysed.
 Further, this mechanism ensures that the sum of all spatial availability values for all modes sum up to the total number of opportunities in the region, which is not true for any other type of accessibility measure.}
@@
 \textcolor{blue}{Thank you for this comment.
-To clarify, the data are not experimental.
-They are observational, since they were collected using a travel survey conducted by the City of Madrid.
+To clarify, the empirical data used is not experimental.
+It is observational, since it is collected using a travel survey conducted by the City of Madrid.
 Travel surveys are a standard instrument in transportation planning and research, and are conducted in cities around the world.
@@
 3.
-Lack of Data-Driven Analysis: In the analysis section, a large amount of data is used to analyze people's travel behavior in different areas.
+Lack of Data-Driven Analysis: In the analysis section, a large amount of data is used to analyse people's travel behavior in different areas.
 The analysis of spatial accessibility only considers the differences in the modes of travel mentioned earlier.
@@
 
-\textcolor{blue}{Reviewer #2 identified some specific formatting issues and we fixed them.
+\textcolor{blue}{Reviewer \#2 identified some specific formatting issues and we fixed them.
 Otherwise we proofread the paper and hopefully did not leave a typo behind.
@@
 \textcolor{blue}{Thank you for this comment.
-We would begin by noting that there are many types of opportunities that are clearly mutually exclusive due to competition.
+We would begin by noting that there are many types of opportunities that are mutually exclusive due to competition.
 For this paper we focused on jobs, but there are many others, such as beds at hospitals, seats at schools, and so on.
@@
 
-\textcolor{blue}{That said, we have grappled for some time with the question of what types of opportunities are best analyzed using spatial availability.
+\textcolor{blue}{That said, we have grappled for some time with the question of what types of opportunities are best analysed using spatial availability.
 Our current thinking on this matter, after much consideration, is that in practice every type of opportunity, even when not clearly exclusive, is subject at least to congestion or capacity constraint.
 For instance, green spaces are often considered non-competitive, however, standards for the provision of such amenties are provided in the form of units of amenity per capita.
-For example, A case in point is the Ile-de-France region, a jurisdiction that suggested in a major planning document of 2013 that at the municipal level at least $10m^2$ of public green space should be supplied per inhabitant} [@liottaPlanning2020]\textcolor{blue}{.
+For example, a case could the Ile-de-France region, a jurisdiction that suggested in a major planning document of 2013 that at the municipal level at least $10m^2$ of public green space should be supplied per inhabitant} \begingroup\color{blue}[@liottaPlanning2020]\endgroup \textcolor{blue}{.
 But green spaces are not evenly distributed, which means that who has access to them hinges on where they are and how easy is to reach them.
 Formulating the provision of amenities in these terms in not rare.
-For example, Natural England recommends an Accessible Natural Greenspace Standard such that the minimum supply of space is one ha of statutory Local Nature Reserves \textit{per thousand population}}^[see  https://redfrogforum.org/wp-content/uploads/2019/11/67-Nature-Nearby%E2%80%99-Accessible-Natural-Greenspace-Guidance.pdf] \textcolor{blue}{.
-Similarly, the World Health Organization (cited in OECD, 2013) recommends that cities provide a minimum of 9 $m^2$ of green area per inhabitant.}^[see https://doi.org/10.1787/9789264191808-en] \textcolor{blue}{For our purposes, standards of this type translate into "how much of this resource is available to one individual that has not been claimed by anyone else?".
+For example, Natural England recommends an Accessible Natural Greenspace Standard such that the minimum supply of space is one ha of statutory Local Nature Reserves\textit{per thousand population}}^[see https://redfrogforum.org/wp-content/uploads/2019/11/67-Nature-Nearby%E2%80%99-Accessible-Natural-Greenspace-Guidance.pdf] \textcolor{blue}{.
+Similarly, the World Health Organization (cited in OECD, 2013) recommends that cities provide a minimum of 9 $m^2$ of green area per inhabitant}^[see https://doi.org/10.1787/9789264191808-en] \textcolor{blue}{.
+For our purposes, standards of this type translate into "how much of this resource is available to one individual that has not been claimed by anyone else?".
 Green spaces often have large capacities, but they still have a capacity, and it is not the same for a person to have access to $5m^2$ of \textit{uncongested} green space than to $15m^2$.
-This difference is in fact a matter of justice} [@laraSpace2015; @liottaPlanning2020]\textcolor{blue}{.
+This difference is in fact a matter of justice }\begingroup\color{blue}[@laraSpace2015; @liottaPlanning2020]\endgroup \textcolor{blue}{.
 Constraining accessibility is in this way a useful way to evaluate the congested availability of any type of opportunity.
-As standards are emphasized in the planning literature, in particular for fairness in transportation} [see @martensFair2021]\textcolor{blue}{, spatial availability analysis can be used to assess standards.
+As standards are emphasized in the planning literature, in particular for fairness in transportation }\begingroup\color{blue}[see @martensFair2021]\endgroup \textcolor{blue}{, spatial availability analysis can be used to assess standards.
 We are convinced that as other researchers discover this new approach to measuring accessibility many other applications will be found.}
@@
 This would require the analyst to estimate the accessibility before and after some change to the land use or transportation system.
-In this paper our empirical application is a single scenario to serve as proof of concept, but in future research we intend to use spatial availability to analyze changes in the size of Madrid's Low Emissions Zone implementation from a modal and socio-economic equity perspective.}
+In this paper our empirical application is a single scenario to serve as proof of concept, but in future research we intend to use spatial availability to analyse changes in the size of Madrid's Low Emissions Zone implementation from a modal and socio-economic equity perspective.}
 
@@
 \textcolor{blue}{This is an excellent point.
-Accessibility measures, including spatial availability, are not meant to function as modal split models, but they is certainly amenable to analyze changes in the accessibiity landscape if different modal shares are used as inputs.
+Accessibility measures, including spatial availability, are not meant to function as modal split models, but they are certainly amenable to analysis of changes in the accessibiity landscape if different modal shares are used as inputs.
 This is similar to the case of destination choice: accessibility measures do not model this, but changes in destination choices can be incorporated via how they affect the impedance function.
@@
 This would expand the number of sub-populations in the analysis from, say, $m=M$ (modes) to $m = M\cdot Q$ (modes times population segments), each with their own characteristic impedance function.
-Evaluations of this kind will be especially relevant as LEZ are implemented in cities globally, and the question of their impact on disadvantaged populations who have become mobility-restricted increasingly come to the fore} [@devrijNooneVisitsMe2022; @verbeekJustManagementUrban2022; @liottaWhatDrivesInequalities2023]\textcolor{blue}{."}
+Evaluations of this kind will be especially relevant as LEZ are implemented in cities globally, and the question of their impact on disadvantaged populations who have become mobility-restricted increasingly come to the fore }\begingroup\color{blue}[@devrijNooneVisitsMe2022; @verbeekJustManagementUrban2022; @liottaWhatDrivesInequalities2023]\endgroup \textcolor{blue}{."}
 
@@
 
-\textcolor{blue}{We have done this.
-In the original version of the abstract and introduction we gave the misleading impression that we would analyze changes in the system, when in reality our intent was to demonstrate the application of the measure in an empirical example.
+\textcolor{blue}{We have now done this.
+In the original version of the abstract and introduction we gave the misleading impression that we would analyse changes in the system, when in reality our intent was to demonstrate the application of the measure in an empirical example.
 We do plan to study changes in the system, but this requires more work than can be presented in a single paper, partly for the reasons that you identified above (modal shifts and the calculation of spatial availability for two different scenarios, as well as the analysis of the differences between them).
@@
 
-\textcolor{blue}{Fixed for all Figures, thank you.}
-
 2. Please change the color scheme of fig 2. The red color scheme makes the LEZ centro area boundary, which is also in red, not visible.
 
-\textcolor{blue}{WILL FIX THANK YOU}
-
- <!-- and I must change the labels on the legends to be ranges -->
+\textcolor{blue}{We have updated the captions of the figures to reflec the meaningof grey colour blocks and the colour scheme of all Figures.
+We've also updated the labels on the legends to make them more interpretable.}
 

### 2023-11-10 Journal-manuscript/Journal-manuscript.Rmd
Finalized revisions and comments to reviewers (R1)

@@
 Constraining the calculations to match a known quantity ensures that the measurements sum up to a predetermined quantity (i.e., the total number of opportunities), and so each value can be meaningfully related to this total.
-A recent such effort is spatial availability, a singly-constrained accessibility measure.
+A recent effort is spatial availability, a singly-constrained accessibility measure.
 In this paper we extend spatial availability for use in the case of multiple modes or, more generally, heterogeneous population segments with distinct travel behaviors.
-After deriving a multimodal version of spatial availability We proceed to illustrate its features using a synthetic example.
-Next, we apply it to an empirical example in Madrid.
+After deriving a multimodal version of spatial availability, we proceed to illustrate its features using a synthetic example.
+Next, we apply it to an empirical example in Madrid, Spain.
 We conclude the paper with suggestions for future research.
@@
 # plos.csl can be download and used locally
-csl: plos.csl
+csl: plos-one.csl
 ---
@@
                      after_stat(density)),
-                 fill = "gray70",
+                 fill = "grey70",
                  bins = 15) +
@@
 tlds_curves_m_plot
-ggsave("images/tlds_curves_m_plot.png")
+ggsave("images/Fig4.png")
 ```
@@
   scale_fill_viridis_c(option = "mako", #legend scale bar
-                       na.value = "gray50",
+                       na.value = "grey50",
                        direction = -1,
@@
           fill=NA) +
-  scale_color_manual(values = c("gray90", "gray60", "black"),
+  scale_color_manual(values = c("grey90", "grey60", "black"),
                        name = element_blank())+
@@
 pop_i_m_plot
-ggsave("images/im_populations_zn208_plot.png",
+ggsave("images/Fig3.png",
        width = 7,
@@
 jobs_i_plot
-ggsave("images/i_jobs_zn208_plot.png",
+ggsave("images/Fig2.png",
        width = 7,
@@
                        name = expression(S["im"]),
-                       na.value = "gray50",
+                       na.value = "grey50",
                        direction = 1)+
@@
 SA_im_V_zn208_plot
-ggsave("images/SA_im_V_zn208_plot.png",
+ggsave("images/Fig5.png",
        width = 7,
@@
 SA_im_v_zn208_plot
-ggsave("images/SA_im_vv_zn208_plot.png",
+ggsave("images/Fig7.png",
        width = 7,
@@
           fill= NA) +
-  scale_color_manual(values = c("gray90", "gray60", "black"),
+  scale_color_manual(values = c("grey90", "grey60", "black"),
                      name = element_blank())+
@@
           col ="black",
-          fill="gray90") +
+          fill="grey90") +
   theme_void() +
@@
                         nrow=2, heights = c(1,2.5) )
-ggsave("images/modal_V_comps_4plot.png", modal_V_comps_4plot_plot,
+ggsave("images/Fig6.png", modal_V_comps_4plot_plot,
         width = 7,
@@
 
-Accessibility is a key concept in the analysis of land use and transportation systems [e.g., @hansenHowAccessibilityShapes1959; @geursAccessibility2004; @paezPositive2012], and one that is coming of age from the perspective of planning too [see, _inter alia_, @handyAccessibility2020; @levinsonTransportAccessManual2020; @siddiqToolsTradeAssessing2021; @yanAccessibilityBasedPlanningAddressing2021; @elGeneidyMaking2022].
+Accessibility is a key concept in the analysis of land use and transportation systems [e.g., @hansenHowAccessibilityShapes1959; @geursAccessibility2004; @paezPositive2012], and too is coming of age from the perspective of planning [see _inter alia_, @handyAccessibility2020; @levinsonTransportAccessManual2020; @siddiqToolsTradeAssessing2021; @yanAccessibilityBasedPlanningAddressing2021; @elGeneidyMaking2022].
 Beginning with the work of Hansen [@hansenHowAccessibilityShapes1959], accessibility measures have been widely used to evaluate the efficiency of transportation systems when combined with the distribution of opportunities in space [@shiLiterature2020].
@@
 But opportunities in general are not inexhaustible, and in fact some of them are by definition exclusive: for example, once a job is taken up by someone in the population, the same job is no longer available for any other person to take.
-More generally, opportunities are subject to congestion: for example, multiple people can avail themselves of the services of the same family doctor, but the more people who do, the more congested the service will be.
-
-The issue of congestion was the motivation for the development of floating catchment area approaches in accessibility analysis [@shenLocationCharacteristicsInnercity1998; @luoMeasuresSpatialAccessibility2003].
-While these approaches purport to account for congestions, Paez et al.
+More generally, opportunities are subject to congestion: for example, multiple people can obtain services from the same family doctor, but the more people who do, the more congested the service will be.
+
+The issue of congestion in accessibility measures was the motivation for the development of floating catchment area approaches [@shenLocationCharacteristicsInnercity1998; @luoMeasuresSpatialAccessibility2003].
+While these approaches purport to account for congestion, Paez et al.
 [@paezDemand2019] demonstrate that in general they do not solve the issue of multiple counting of opportunities, thus leading to biases in the calculation of total demand and supply, sometimes inflating them, other times deflating them.
 In response to this, recent research has paid closer attention to the way opportunities are counted in accessibility analysis.
+Paez et al.
 @paezDemand2019, for example, tackle floating catchment area methods and introduce a normalization of the impedance matrix to allocate the population and then the level of service proportionally.
@@
 [@soukhovIntroducingSpatialAvailability2023] introduced a singly-constrained measure of accessibility, called spatial availability, that employs a similar, but more sophisticated proportional allocation mechanism.
-The work of these authors shows that floating catchment area methods can be seen as singly-constrained accessibility measures, and improve on existing approaches by guaranteeing that each opportunity is counted only once - in other words, treating opportunities as _finite_.
+The work of these authors show that floating catchment area methods can be seen as singly-constrained accessibility measures, and improve on existing approaches by guaranteeing that each opportunity is counted only once - in other words, treating opportunities as _finite_.
 The proportional allocation of spatial availability constrains the calculations to match a known quantity, therefore ensuring that the measurements sum up to a predetermined quantity (i.e., the total number of opportunities), and so each value can be meaningfully related to this total.
@@
 However, the finite nature of opportunities makes the analysis of heterogeneous populations very relevant.
-In the case of multiple modes of transportation, people who travel by slow modes (e.g., active modes) can usually reach fewer opportunities than people who travel by faster modes and whose range is typically far wider (e.g., car).
+In the case of multiple modes of transportation, people who travel by slower modes (e.g., active modes) can usually reach fewer opportunities than people who travel by faster modes and whose range is typically far wider (e.g., car).
 This implies that slower travelers will often face increased competition for local opportunities from travelers who can reach said opportunities from farther afield.
@@
 Our primary motivation is to extend spatial availability for the case of multimodal accessibility, but it is worthwhile noting that this is in fact just one case of heterogeneous populations (i.e., travel by different modes).
-The method itself can easily accommodate other forms of heterogeneity, for example variations in travel behavior between older and younger adults [e.g., @paezHealth2010], the propensity of older adults to use different modes of transportation [@moniruzzamanMode2013], the usually shorter trip lengths of children compared to grown-ups [e.g., @reyesWalking2014], or the more limited travel ranges of single parents [e.g., @paezJobs2013].
+The method itself can easily accommodate other forms of heterogeneity, for example variations in travel behavior between older and younger adults [e.g., @paezHealth2010], the propensity of older adults to use different modes of transportation [e.g., @moniruzzamanMode2013], the usually shorter trip lengths of children compared to grown-ups [e.g., @reyesWalking2014], or the more limited travel ranges of single parents [e.g., @paezJobs2013].
 
 The paper rest of this paper is organized as follows.
-In Section we discuss the use a synthetic examples with multiple modes to illustrate some relevant issues; this helps us motivate the derivation of the new spatial availability expressions.
-This is followed in Section 3 by an empirical example using data from the city of Madrid after the implementation of its Low Emission Zones.
+In Section 2 we provide a brief review of multimodal accessibility.
+In the following Section 3 we demonstrate the derivation of the new spatial availability expression for multiple modes.
+In Section 4 we illustrate relevant issues through a synthetic example.
+This is followed in Section 5 by an empirical example using data from the city of Madrid after the implementation of its Low Emission Zones (LEZ).
 Data for this example comes from the city's 2018 travel survey.
 The example shows the differences in spatial availability within and outside the LEZ for travelers using different modes, namely car, transit, cycling and walking.
-In Section 4, we provide concluding remarks on the strengths of the use of spatial availability as a multimodal accessibility measure, and discuss potential future uses in policy planning scenarios as well as directions for future research.
+In Section 6, we provide concluding remarks on the strengths of the use of spatial availability as a multimodal accessibility measure, and discuss potential future uses in policy planning scenarios as well as directions for future research.
 
@@
 Given an origin ($i$) and a destination ($j$), an impedance function $f^{m}(c^m_{ij})$ converts the cost of travel (e.g., time, money, generalized cost) into a score that represents the propensity for interaction.
-These measures originate from that proposed by @hansenHowAccessibilityShapes1959, which can take the following form in the multimodal case: $S_i^m = \sum_j O_j f^m(c_{ij}^m)$ where $m$ is a set of modes which have mode-specific travel costs ($c_{ij}^m$) and/or travel impedance functions ($f^m(\cdot)$).
+These measures originate from that proposed by Hansen @hansenHowAccessibilityShapes1959, which can take the following form in the multimodal case: $S_i^m = \sum_j O_j f^m(c_{ij}^m)$ where $m$ is a set of modes which have mode-specific travel costs ($c_{ij}^m$) and/or travel impedance functions ($f^m(\cdot)$).
 
@@
 As reported, the ratio of the transit to the car score can be lower than 0.2 (meaning transit gives access to less than 20% of the opportunities than car).
-But despite the discussion about "sufficient accessibility" (e.g., in the abstract), it is unclear what the unconstrained scores mean: is having access to but 10,000 jobs by transit insufficient?
+But despite the discussion about "sufficient accessibility", it is unclear what the unconstrained scores mean: is having access to but 10,000 jobs by transit insufficient?
 After all, 10,000 employment opportunities are still plenty of opportunities.
@@
 5; @paezJobs2013, Figs.
-6, 7, and 8], and they are useful to assess when some members of the public are better or worse off than others, but they do not say much about how bad is "worse".
+6, 7, and 8], and they are useful as relative assessment of when some members of the public are better or worse off than others, but they do not say much about how bad is "worse off".
 
@@
 
-Once we understand opportunities to be finite, it is possible for an accessibility measure to take a crisper meaning.
+Once we understand opportunities to be finite, it is possible for an accessibility measure to take on a crisper meaning.
 As considered in the long tradition of accessibility research, capacity of opportunities is limited and thus is subject to competition by population [@soukhovIntroducingSpatialAvailability2023; @grengsJobAccessibilityModal2010; @kawabataJobAccessibilityIndicator2006; @kwokUseModalAccessibility2004; @morrisAccessibilityIndicatorsTransport1979; @weibullAxiomaticApproachMeasurement1976; @shenLocationCharacteristicsInnercity1998].
@@
 But we would go as far as to argue that every type of opportunity is subject to congestion or capacity constraint, even when the opportunities are conventionally seen as non-competitive.
+
 Amenities are a good example of this.
 For instance, standards for providing green spaces are often stated in the form of _exclusive access_, in units of amenity per capita.
-A case in point is the Ile-de-France region, a jurisdiction that suggested in a major planning document of 2013 that at the municipal level at least $10m^2$ of public green space should be supplied _per inhabitant_ [@liottaPlanning2020].
-But green spaces are not evenly distributed, which means that who has access to them hinges on where they are and how easy is to reach them.
+A case in point is a the Ile-de-France region, a jurisdiction that suggested in a 2013 planning document that at the municipal level at least $10m^2$ of public green space should be supplied _per inhabitant_ [@liottaPlanning2020].
+Green spaces are not evenly distributed, which means that who has access to them hinges on where they are and how easy is to reach them.
 This formulation of provision of amenities is not unusual.
@@
 For our purposes, standards of this type translate into "how much of this resource is available to one individual that has not been claimed by anyone else?".
-Green spaces often have large capacities, but they still have a capacity, and it is not the same a person to have access to $5m^2$ of _uncongested_ green space than to $15m^2$.
+Green spaces often have large capacities, but they still have a capacity, and it is not the same for a person to have access to $5m^2$ of _uncongested_ green space as $15m^2$.
 This difference is in fact a matter of justice [@laraSpace2015; @liottaPlanning2020].
@@
 The relevance of the considerations above is put in sharper relief when we think about the use of multiple modes (or heterogeneous populations).
-If we return to Oslo for a moment [@lunkeModalAccessibilityDisparities2022], we notice that the places that have high accessibility by transit are also the places that have _very high_ accessibility by car (Figure 2).
+If we return to Oslo for a moment [@lunkeModalAccessibilityDisparities2022], we notice that the places that have high accessibility by transit are also the places that have _very high_ accessibility by car (in their Figure 2).
 Those two populations are going for the same opportunities, and those travelling by transit have fewer to choose from to begin with.
-More generally, people in a zone who are advantaged with relatively cost of travel will have the ability to potentially reach more opportunities than other people.
-Due to this advantage, through the perspective of finite opportunities, there are fewer opportunities left for everyone else, especially for those who use modes that are slower or more expensive.
-
-As noted in the introduction, competitive accessibility was the rationale for developing floating catchment area methods (FCA), popularized by @luoMeasuresSpatialAccessibility2003 [although similar, and earlier, developments are found in @weibullAxiomaticApproachMeasurement1976; @josephMeasuringPotentialPhysical1982; @shenLocationCharacteristicsInnercity1998].
+More generally, people in a zone who are advantaged with relatively low cost of travel will have the ability to potentially reach more opportunities than other people.
+Due to this advantage, through the perspective of finite opportunities, there are fewer opportunities left for everyone else, especially for those who use modes that are slower or otherwise more expensive.
+
+As noted in the Introduction, competitive accessibility was the rationale for developing floating catchment area methods (FCA), popularized by Luo et al.
+[@luoMeasuresSpatialAccessibility2003] who reformulated the work of Shen [@shenLocationCharacteristicsInnercity1998] into two steps (although similar, and earlier, developments are found in [@weibullAxiomaticApproachMeasurement1976; @josephMeasuringPotentialPhysical1982]).
 Shen-type accessibility is formulated as: $a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}$ where $D_j^m$ is the potential demand for opportunities equal to travel impedance weighted population $\sum_i P_i^m f^m(c_{ij}^m)$ and the remaining variables are repeated in the Hansen-type measure.
@@
 That said, the measure remains unconstrained, meaning both population _and_ opportunities are multiply counted [see @paezDemand2019].
-In other words, interpretation of differences in Shen-type accessibility scores between modes is fraught as it is for Hansen-type measures.
+In other words, interpretation of the Shen-type accessibility scores between modes is fraught as it is for Hansen-type measures.
 
@@
 Their results indicate that zones with low-income populations have lower $a_i^m$ than zones with higher-income populations.
-Further, they show that $a_i^{m=\text{public transit}}$ is lower than $a_i^{m=\text{car}}$ at many zones, arguing that this may further place those zones with lower-income populations at disadvantage.
+Further, they show that $a_i^{m=\text{public transit}}$ is lower than $a_i^{m=\text{car}}$ at many zones, arguing that this may further place those zones with lower-income populations at a disadvantage.
 $a_i$ and/or $a_i^m$ are used to compare relative spatial differences in overall competitive accessibility and multimodal competitive accessibility, but because opportunities were doubly counted (entering the sums of both modes), this makes for uneasy interpretations of the differences in $a_i^{m}$ between modes.
@@
 In Soukhov et al.
-[@soukhovIntroducingSpatialAvailability2023] these factors consider: 1) the mass effect (e.g., the size of populations at different origins); and 2) the cost of travel from different zones (e.g., some sub-populations face relatively higher or lower costs).
+[@soukhovIntroducingSpatialAvailability2023] these factors consider: the mass effect (e.g., the size of populations at different origins); and the cost of travel from different zones (e.g., some sub-populations face relatively higher or lower costs).
 The following section introduces the multimodal form of spatial availability.
@@
 
-\noindent we see that spatial availability is, like Hansen's measure, a weighted sum of the opportunities.
-What makes spatial availability stand apart from other approaches is the weight used in the sum, balancing factor $F^t_{ij}$, implements a proportional allocation mechanisms to ensure that the sum of $V_i$ is constrained to match the total number of opportunities in the region.
+\noindent we see that spatial availability is, like the Hansen-type measure, a weighted sum of the opportunities.
+What makes spatial availability stand apart from other approaches is how the weight used in the sum, balancing factor $F^t_{ij}$, implements a proportional allocation mechanisms to ensure that the sum of $V_i$ is constrained to match the total number of opportunities in the region.
 As such, spatial availability is singly-constrained and naturally implements competition or congestion.
@@
 
-This factor makes opportunities available based on demand. Secondly, there is an impedance-based proportional allocation factor that model the cost effect:
+This factor makes opportunities available based on demand. Secondly, there is an impedance-based proportional allocation factor that models the cost effect:
 $$
@@
 -   $F^{tm}_{ij}$ is a balancing factor $F^t_{ij}$ for each $m$ at each $i$.
--   $V^m_{i}$ is the spatial availability $V_{i}$ for mode $m$ at each $i$; the sum of $V^m_{i}$ for all $m$ at each $i$ is equivalent to the total sum of opportunities in the region (i.e., $\sum_j O_j = \sum_i V_i = \sum_{m} \sum_{i} V^m_{i}$).
+-   $V^m_{i}$ is the spatial availability $V_{i}$ for each $m$ at each $i$; the sum of $V^m_{i}$ for all $m$ at each $i$ is equivalent to the total sum of opportunities in the region (i.e., $\sum_j O_j = \sum_i V_i = \sum_{m} \sum_{i} V^m_{i}$).
 
@@
 flextable::set_caption(table_toy_format_flex, 
-            caption = "Accessibility values at each origin per mode m at each origin i and aggregated between modes for each i for the synthetic example.")
+            caption = "Accessibility values at each origin (i) per mode (m) (columns three to five) and aggregated per i (columns six and seven) for the synthetic example.")
 ```
@@
 Notice that in this example we use the same impedance function but the travel times are different for the two modes.
-More generally, it is possible to use different impedance functions for the modes, as demonstrated in the empirical example below.
+More generally, it is possible to use different impedance functions for the modes, as demonstrated in the empirical example in the following section.
 
@@
 This value lacks an intuitive interpretation: it represents the weighted sum of opportunities that may be reached within the region according to the travel impedance (i.e., the travel behavior and the characteristics of the modes) and does not usefully translate into any sort of benchmark.
-To connect this example to the literature, $S_i^m$ is calculated in the work of @tahmasbiMultimodalAccessibilitybasedEquity2019; they compare differences in $S_i^m$ values between modes in a relative and comparative sense, but make no further interpretation of the $S_i^m$ values.
-Large metropolitan regions will have large values and small regions small, but that is simply an artifact of size.
+To connect this example to the aforementioned literature, $S_i^m$ is calculated in the work of Tahmasbi et al.
+[@tahmasbiMultimodalAccessibilitybasedEquity2019]; they contrast differences in $S_i^m$ values between modes in a relative and comparative sense, but make no further interpretation of the $S_i^m$ values.
+More densely populated metropolitan regions will tend to have more opportunities and hence large $S_i^m$ values and less densely regions, smaller values; how much of these differences may simply an artifact of region density?
 
@@
 For instance, the population travelling by $x$ from $A$ and $B$ do not have the same values of $a_i^m$ as those travelling by $z$.
-In fact, $A$ has the highest values $a_i^m$ and $a_i$ values since this center has the lowest travel impedance to opportunities (lower than at $C$, $A$ and $B$ are equal) and has faces relatively low competition, not being close to a relatively large population (lower than at $B$).
+In fact, $A$ has the highest values $a_i^m$ and $a_i$ values since this center has the lowest travel impedance to opportunities (lower than at $C$, $A$ and $B$ are equal) and faces relatively low competition, not being close to a relatively large population (lower than at $B$).
 
@@
 The Shen-type measure is implemented in aforementioned work of Tao et al.
-[@taoInvestigatingImpactsPublic2020a] to calculate modal $a_i^m$ values and the aggregated $a_i$ is implemented in the work of [@carpentieriMultimodalAccessibilityPrimary2020].
-However, similar to Hansen-type accessibility, these works discuss relative and spatially comparative differences in values, but veer from interpretating the values of $a_i^m$ or $a_i$ themselves.
+[@taoInvestigatingImpactsPublic2020a] to calculate modal $a_i^m$ values and the aggregated $a_i$ is implemented in the work of Carpentier et al.
+[@carpentieriMultimodalAccessibilityPrimary2020].
+However, similar to Hansen-type accessibility, these works discuss relative and spatially comparative differences in values, but veer from interpreting the values of $a_i^m$ or $a_i$ themselves.
 In fairness, interpretation is complicated by the multiple counting of opportunities between zones and modes.
@@
 In contrast, spatial availability $V_i$ considers competition and is constrained such that the total sum of values is equal to the total number of opportunities in the region (i.e., $210,000$ jobs).
-Seen in fifth column of Table 1 <!-- \ref{tab:tab1}-->, the values of $V_i^m$ in $A$ and $B$ are not the same within mode (as this measure considers competition).
+Seen in fifth column of Table 1 <!-- \ref{tab:tab1}-->, the values of $V_i^m$ in $A$ and $B$ are not the same within each mode (as this measure considers competition).
 In fact, at $A$, users of mode $z$ capture `r ((table_toy_format %>% filter(i == "A" & m == "z") %>% select("V_im")) - (table_toy_format %>% filter(i == "A" & m == "x") %>% select("V_im"))) %>% pull() %>% prettyNum(big.mark = ",")` more spatially available jobs (of the $210,000$ jobs in the region) than the sub-population travelling by $x$.
@@
 The context for the empirical example is Madrid, Spain.
-This city implemented a Low Emission Zone in 2017, to pursue goals set out in the national climate change agenda, to cut nitrogen dioxide levels, and to prioritize people's movement in the city.
-Low Emission Zones (LEZs) elsewhere have been implemented as interventions to reduce GHG emissions, improve air quality, and support sustainable mobility [@margaryanLowEmissionZones2021; @verbeekJustManagementUrban2022].
+This city implemented a Low Emission Zone (LEZ) in 2017 to: pursue goals set out in the national climate change agenda, cut nitrogen dioxide levels, and to prioritize people's movement in the city.
+LEZs elsewhere have similarly  been implemented as interventions to reduce GHG emissions, improve air quality, and support sustainable mobility [@margaryanLowEmissionZones2021; @verbeekJustManagementUrban2022].
 Though the rules of exclusion vary by city<!-- depending on legal aspects and cultural norms-->, LEZs aim to deter/reduce traffic in designated zones under threat of penalty (e.g., fines, seizure of vehicle).
-<!-- In practice, LEZ limits the volume of traffic by excluding vehicles by license plate, fuel type, or by introducing tolls.--> In other words, LEZ implement a form of _geographic discrimination_ as they change how people can reach opportunities by making it more costly for some forms of travel, typically cars, to circulate in predetermined zones.
-When considering opportunities as finite in a region, this discrimination reduces the competition of one mode and opens up opportunties for other modes to thrive.
-In this way, at their core LEZs operate by changing the accessibility landscape of a city from the perspective of multiple modes.
+<!-- In practice, LEZ limits the volume of traffic by excluding vehicles by license plate, fuel type, or by introducing tolls.--> In other words, LEZs implement a form of _geographic discrimination_ as they change how people can reach opportunities by making it more costly for some forms of travel, typically cars, to circulate in predetermined zones.
+When considering opportunities as finite in a region, this discrimination reduces the competition of one mode and opens up opportunities for other modes to better thrive.
+At their core, LEZs operate by changing the accessibility landscape of a city from the perspective of multiple modes.
 
@@
 
-In geographic scope, the 2017 boundaries of the LEZ in Madrid were relatively modest, covering only approximately 4.72 $km^2$ of the central business district of the city (the so-called LEZ Centro).
+In geographic scope, the 2017 boundaries of the LEZ in Madrid were relatively modest, covering only approximately 4.72 km\textsuperscript{2} of the central business district of the city (the so-called LEZ Centro).
 As of this writing, there are plans to expand these boundaries to the area inside the M-30, an orbital highway in proximity to the city center (i.e., LEZ M-30).
@@
 ```{r, jobs-plot, fig.cap="\\label{fig:Fig2} Distribution of jobs taken by people living and working in Madrid as reported in the 2018 travel survey.
-Gray TAZ has no jobs.
+Grey TAZs has no jobs.
 Ranges of values in the legend are quintiles.", fig.show='hold', fig.align='center'}
-knitr::include_graphics("images/i_jobs_zn208_plot.png") 
+knitr::include_graphics("images/Fig2.png") 
 ```
@@
 ```{r, pop-plot, fig.cap="\\label{fig:Fig3} Population living and working in Madrid by mode of transportation as reported in the 2018 travel survey.
-Gray TAZs have no population.
+Grey TAZs have no population.
 Ranges of values in the legend are quintiles.", fig.show='hold', fig.align='center'}
-knitr::include_graphics("images/im_populations_zn208_plot.png")
+knitr::include_graphics("images/Fig3.png")
 ```
@@
 The trips are expanded using population weights.
-Figures \ref{fig:Fig2} and \ref{fig:Fig3} show the number of workers and the distribution of jobs in the City of Madrid by TAZ.
-The light gray boundary represents the LEZ Centro in effect in 2017.
+Figures \ref{fig:Fig2} and \ref{fig:Fig3} show the number of workers and the distribution of full-time jobs in the City of Madrid by TAZ.
+The light grey boundary represents the LEZ Centro in effect in 2017.
 The survey was conducted after the introduction of LEZ Centro.
-The dark gray boundary represents the LEZ planned for for the boundaries of the M-30 highway and is present in the plots as a spatial reference for areas in proximity to the LEZ Centro.
+The dark grey boundary represents the LEZ planned for the boundaries of the M-30 highway and is present in the plots as a spatial reference for areas in proximity to the LEZ Centro.
 
 The total sum of jobs $O_j$ are shown in Figure \ref{fig:Fig2} and the populations that go to a work destination by four modal categories $P^m_i$, is displayed in Figure \ref{fig:Fig3}.
-The modal shares in Figure \ref{fig:Fig3} are calculated based on those measued in the survey.
+The modal shares in Figure \ref{fig:Fig3} are calculated based on those measured in the survey.
 The modal categories and the mode types within each category are reported as follows:
@@
    - Car/motor: all cars and operating modes (e.g., cab, private driver, company, rental car, main driver of a private car, passenger in a private car) and all public, private or company motorcycle/mopeds.
-  - Transit: all bus, trams, and trains,
-  - Bike: all bicycle trips (e.g., private, public, or company bike trips) and "other" types of micromobility options,
+  - Transit: all bus, trams, and trains.
+  - Bike: all bicycle trips (e.g., private, public, or company bike trips) and "other" types of micromobility options.
   - Walk: walking or by foot.
@@
 
-Travel times are provided within the survey by mode.
+Travel times are provided within the travel survey by mode.
 This information is used to calibrate mode-specific travel impedance functions $f^m(c_{ij}^m)$.
@@
 
-- Car/motor: mean `r weighted_mean(x=tld_empirical_car$TIEMPO_VIAJES, w=tld_empirical_car$ELE_G_POND_ESC2) %>% round()` minutes (min:`r car_qs[1]%>% round()` minutes, Q2: `r car_qs[2]%>% round()` minutes, Q3: `r car_qs[4]%>% round()` minutes, max: `r car_qs[5]%>% round()` minutes)
-- Transit: mean `r weighted_mean(x=tld_empirical_transit$TIEMPO_VIAJES, w=tld_empirical_transit$ELE_G_POND_ESC2)%>% round()` minutes (min:`r transit_qs[1]%>% round()` minutes, Q2: `r transit_qs[2]%>% round()` minutes, Q3: `r transit_qs[4]%>% round()` minutes, max: `r transit_qs[5]%>% round()` minutes)
-- Bike: mean `r weighted_mean(x=tld_empirical_bike$TIEMPO_VIAJES, w=tld_empirical_bike$ELE_G_POND_ESC2)%>% round()` minutes (min:`r bike_qs[1]%>% round()` minutes, Q2: `r bike_qs[2]%>% round()` minutes, Q3: `r bike_qs[4]%>% round()` minutes, max: `r bike_qs[5]%>% round()` minutes)
-- Walk: mean `r weighted_mean(x=tld_empirical_walk$TIEMPO_VIAJES, w=tld_empirical_walk$ELE_G_POND_ESC2)%>% round()` minutes (min:`r walk_qs[1]%>% round()` minutes, Q2: `r walk_qs[2]%>% round()` minutes, Q3: `r walk_qs[4]%>% round()` minutes, max: `r walk_qs[5]%>% round()` minutes)
+- Car/motor: mean `r weighted_mean(x=tld_empirical_car$TIEMPO_VIAJES, w=tld_empirical_car$ELE_G_POND_ESC2) %>% round()` minutes (min: `r car_qs[1]%>% round()` minutes, Q2: `r car_qs[2]%>% round()` minutes, Q3: `r car_qs[4]%>% round()` minutes, max: `r car_qs[5]%>% round()` minutes)
+- Transit: mean `r weighted_mean(x=tld_empirical_transit$TIEMPO_VIAJES, w=tld_empirical_transit$ELE_G_POND_ESC2)%>% round()` minutes (min: `r transit_qs[1]%>% round()` minutes, Q2: `r transit_qs[2]%>% round()` minutes, Q3: `r transit_qs[4]%>% round()` minutes, max: `r transit_qs[5]%>% round()` minutes)
+- Bike: mean `r weighted_mean(x=tld_empirical_bike$TIEMPO_VIAJES, w=tld_empirical_bike$ELE_G_POND_ESC2)%>% round()` minutes (min: `r bike_qs[1]%>% round()` minutes, Q2: `r bike_qs[2]%>% round()` minutes, Q3: `r bike_qs[4]%>% round()` minutes, max: `r bike_qs[5]%>% round()` minutes)
+- Walk: mean `r weighted_mean(x=tld_empirical_walk$TIEMPO_VIAJES, w=tld_empirical_walk$ELE_G_POND_ESC2)%>% round()` minutes (min: `r walk_qs[1]%>% round()` minutes, Q2: `r walk_qs[2]%>% round()` minutes, Q3: `r walk_qs[4]%>% round()` minutes, max: `r walk_qs[5]%>% round()` minutes)
 
@@
 The selection of functional forms aligns with empirical examples in other regions [@reggianiAccessibilityImpedanceForms2011; @soukhovTTS2016RDataSet2023; @soukhovIntroducingSpatialAvailability2023].
-The shape and rate parameters for the gamma functions (motorized modes) are `r HW_car_motor_gamma$estimate[1]` and `r HW_car_motor_gamma$estimate[2]`for car/motor and `r HW_transit_gamma$estimate[1]` and `r HW_transit_gamma$estimate[2]` for transit; for the log-normal functions (non-motorized modes), the mean and standard deviation parameters are `r HW_bike_lnorm$estimate[1]` and `r HW_bike_lnorm$estimate[2]` for bike and `r HW_walk_lnorm$estimate[1]` and `r HW_bike_lnorm$estimate[2]` for walk.
+The shape and rate parameters for the gamma functions (motorized modes) are `r HW_car_motor_gamma$estimate[1]` and `r HW_car_motor_gamma$estimate[2]` for car/motor and `r HW_transit_gamma$estimate[1]` and `r HW_transit_gamma$estimate[2]` for transit; for the log-normal functions (non-motorized modes), the mean and standard deviation parameters are `r HW_bike_lnorm$estimate[1]` and `r HW_bike_lnorm$estimate[2]` for bike and `r HW_walk_lnorm$estimate[1]` and `r HW_bike_lnorm$estimate[2]` for walk.
 
@@
 
-```{r, tlds-curves-m-plot, fig.cap="\\label{fig:Fig4} Fitted impedance function  against empirical TLD (bars) corresponding to the home-to-work origin-destination flows from the Madrid 2018 travel survey.", fig.show='hold', fig.align='center', fig.height=6}
-knitr::include_graphics("images/tlds_curves_m_plot.png")
+```{r, tlds-curves-m-plot, fig.cap="\\label{fig:Fig4} Fitted impedance function against empirical TLD (bars) corresponding to the home to full-time work origin destination flows for the City of Madrid from the 2018 travel survey.", fig.show='hold', fig.align='center', fig.height=6}
+knitr::include_graphics("images/Fig4.png")
 ```
@@
 
-At this point it is worthwhile to reiterate that the empirical example is a snap-shot of spatial availability by mode using data from the 2018 travel survey.
-Our purpose is to investigate the trends in availability of employment opportunities by mode, and illustrate how spatial availability can help in discussions about the competitive advantage of various modes within Madrid Centro.
+At this point, it is worthwhile reiterating that the empirical example is a snap-shot of spatial availability by mode using data from the 2018 travel survey.
+Our purpose in this empirical example is to investigate the trends in availability of employment opportunities by mode, and illustrate how spatial availability can be used in discussions about the competitive advantage of various modes within Madrid Centro.
 The spatial availability of jobs $V_i^m$ is calculated for each of the four modes $m$ at the level of traffic analysis zones $i$ in Madrid and displayed in Figure \ref{fig:Fig5}.
@@
 ```{r, SA-m-plot, fig.cap="\\label{fig:Fig5} Spatial availability of jobs per origin and mode $V_i^m$ in Madrid.
-Gray TAZs have no population.
+Grey TAZs have no population.
 Ranges of values in the legend are quintiles.", fig.show='hold', fig.align='center', fig.height=10}
-knitr::include_graphics("images/SA_im_V_zn208_plot.png")
+knitr::include_graphics("images/Fig5.png")
 ```
@@
 In a way, this is to be expected since users of these modes represent `r (modal_V_comps_4plot$perc[3] + modal_V_comps_4plot$perc[4]) %>% percent(accuracy=0.1)` of the total population.
-However, the ability to travel at greater speeds also impacts this results.
-Fuerthermore, differences in $V_i^m$ values within modes also exist in space: car users outside of the M-30 region appear to enjoy greater spatial availability, while some zones inside the M-30 to have greater spatial availability for transit.
+However, the ability to travel at greater speeds also impacts these results.
+Furthermore, differences in $V_i^m$ values within modes also exist in space: car users outside of the M-30 region appear to enjoy greater spatial availability, while some zones inside the M-30 to have greater spatial availability for transit.
 Overall, the magnitude of $V_i^m$ values for cyclists and pedestrians are lower than for car and transit but the highest values of $V_i^{bike}$ and $V_i^{walk}$ tend to be found in zones within the M-30 and origins with higher spatial availability by transit.
@@
 The differences between the shares of modes and their shares of spatially available opportunities highlights the competitive advantage of certain modes, although this effect is not geographically uniform.
-As seen in the left-most columns of Figure \ref{fig:Fig6}, 'car/motor' and 'transit' together can avail themselves of `r (modal_V_comps_4plot$perc[7] + modal_V_comps_4plot$perc[8]) %>% percent(accuracy=0.1)` of all jobs in the city.
-However, 'car/motor' has a disproportionate share of $V_i^m$ relative to the population of users of this mode, compared to opportunities that are spatially available to transit users.
+As seen in the left-most columns of Figure \ref{fig:Fig6}, users of 'car/motor' and 'transit' together can avail `r (modal_V_comps_4plot$perc[7] + modal_V_comps_4plot$perc[8]) %>% percent(accuracy=0.1)` of all jobs in the city (Spatial Availability by Mode).
+However, car/motor users have a disproportionate share of $V_i^m$ relative to the population of users of this mode (Population by Mode), compared to opportunities that are spatially available to transit users.
 The combined population of car and transit users is `r modal_V_comps_4plot$perc[3] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[4] %>% percent(accuracy=0.1)` respectively, but these populations are allocated `r modal_V_comps_4plot$perc[7] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[8] %>% percent(accuracy=0.1)` respectively, of the city's jobs.
@@
 From left to right, the city of Madrid (All), the area within the M-30 highway (M-30), the area within the Centro region (Centro).", fig.show='hold', fig.align='center', fig.height=10}
-knitr::include_graphics("images/modal_V_comps_4plot.png")
+knitr::include_graphics("images/Fig6.png")
 ```
@@
 It is also clear that more opportunities are spatially available to non-car users within Centro.
-In the case of active travel, the proportions of cyclists and walkers within LEZ Centro still exceed the proportions of jobs spatially available to them --- however the disparity is drastically reduced compared to the rest of the city.
+In the case of active travel, the proportions of cyclists and walkers within LEZ Centro still exceeds the proportions of jobs spatially available to them --- however, the disparity is drastically reduced compared to the rest of the city.
 As seen in Figure \ref{fig:Fig6}, there is a higher proportion of opportunities that are spatially available to pedestrians and cyclists in Centro than in the City overall and in areas within the M-30.
 Notably, within Centro, `r modal_V_comps_4plot$perc[21] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[22] %>% percent(accuracy=0.1)` of opportunities are spatially available to bike and walk modes respectively, while their populations represent  `r modal_V_comps_4plot$perc[1] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[2] %>% percent(accuracy=0.1)` of the population.
-By restricting the ability of cars to enter Centro, the LEZ seems to contributes to level the playing field for slower modes, in particular cycling and walking, but also transit.
-As seen in the figure, transit users are generally close to parity across the region, with nearly as many spatially available jobs as transit users.
+By restricting the ability of cars to enter Centro, the LEZ seems to contribute to leveling the playing field for slower modes, in particular cycling and walking, but also transit.
+As seen in the Figure \ref{fig:Fig6}, transit users are generally close to parity across the region, with nearly as many spatially available jobs as transit users.
 Still, this mode has the greatest advantage in LEZ Centro with `r modal_V_comps_4plot$perc[24] %>% percent(accuracy=0.1)` of spatially available jobs in Centro for `r modal_V_comps_4plot$perc[20] %>% percent(accuracy=0.1)` of transit users in Centro.
-This result makes intuitive sense: after car, it is the mode with the greatest range, and unlike car it is allowed in LEZ Centro.
+This result makes intuitive sense: after car, it is the mode with the greatest range, and unlike car it is unrestricted in the LEZ Centro.
 
@@
 Figure \ref{fig:Fig7} shows $v_i^m$, the spatial availability $V_i^m$ divided by the population of users of $m$.
-Values of $v_i^m$ below 1 are shown in shades of orange, and indicate TAZs with less than one spatially available opportunity per capita for the mode.
-Values above 1 are shown in shades of green, and indicate TAZs with more than one spatially available opportunity per capita for the mode.
+Values of $v_i^m$ below one are shown in shades of orange, and indicate TAZs with less than one spatially available opportunity per capita for the mode.
+Values above one are shown in shades of green, and indicate TAZs with more than one spatially available opportunity per capita for the mode.
 The highest spatial availability per capita (shown in blue) is for car users in a zone northeast just beyond the M-30.
-These plots illustrate in unambiguous fashion, and in a quantity that is comparable over space and time, the advantage in terms of spatial availability of car for most of the city (bottom left plot, areas denoted with green $v_i^m$ values above 1).
+These plots illustrate in unambiguous fashion, and in a quantity that is comparable over space and time, the advantage in terms of spatial availability of car for most of the city (bottom left plot, areas denoted with green $v_i^m$ values above one).
 It can also be observed that spatial availability of jobs is relatively well balanced for transit users over most of the regions (i.e., many zones are light orange or light green).
-Spatial availability of jobs for non-motorized modes, in contrast, is low (under 1) overall, although less so within LEZ Centro.
+Spatial availability of jobs for non-motorized modes, in contrast, is low (under one) overall, although less so within LEZ Centro.
 
 ```{r, SA-per-capita-m-plot, fig.cap="\\label{fig:Fig7} Distribution of spatially available jobs per capita by mode of transportatin ($v_i^m$).
-Gray TAZs have no population that use the mode.
+Grey TAZs have no population that use the mode.
 Ranges of values in the legend are quintiles.", fig.show='hold', fig.align='center', fig.height=10}
-knitr::include_graphics("images/SA_im_vv_zn208_plot.png")
-```
-
-Incidentally, it is bears noting that $v_i^m$ for car within and near LEZ Centro is close to or below 1 in Figure \ref{fig:Fig7}, while all non-car modes have relatively higher $v_i^m$ values.
+knitr::include_graphics("images/Fig7.png")
+```
+
+Incidentally, $v_i^m$ values for car within and near LEZ Centro is close to or below one in Figure \ref{fig:Fig7}, while all non-car modes have relatively higher $v_i^m$ values.
 Since these values are comparable across regions and over time, Figure \ref{fig:Fig7} potentially provides a benchmark for quantifying changes in LEZ policies in the future.
@@
 
-Accessibility measures are an important tool in transportation research [@shiLiterature2020] and increasingly also seen as valuable for planning purposes [@handyAccessibility2020; @levinsonTransportAccessManual2020; @siddiqToolsTradeAssessing2021; @yanAccessibilityBasedPlanningAddressing2021; @elGeneidyMaking2022].
+Accessibility measures are an important tool in transportation research [@shiLiterature2020] and are increasingly seen as valuable for planning purposes [@handyAccessibility2020; @levinsonTransportAccessManual2020; @siddiqToolsTradeAssessing2021; @yanAccessibilityBasedPlanningAddressing2021; @elGeneidyMaking2022].
 They boast a long history of development, beginning with Hansen-type $S_i^m$ measures, with other developments like Shen's $a_i^m$, to account for competition/congestion.
@@
 Aggregations of opportunities embody principles of gravitational/spatial interaction modelling that date back to at least H.C.
-Carey [@careyPriciples1858], and are part of a line of research that includes the work of Ravenstein [@ravensteinLaws1889], Reilly [@reillyMethods1929], Stewart [@stewartInverse1941; @stewartSuggested1947; @stewartDemographic1948], Zipf [@zipfHypothesis1946; @zipfIntercity1946], Wilson [@wilsonFamilty1971], and many others besides.
+Carey [@careyPriciples1858], and are part of a line of research that includes the work of Ravenstein [@ravensteinLaws1889], Reilly [@reillyMethods1929], Stewart [@stewartInverse1941; @stewartSuggested1947; @stewartDemographic1948], Zipf [@zipfHypothesis1946; @zipfIntercity1946], Wilson [@wilsonFamilty1971], and many others.
 In this way, $S_i^m$, $a_i^m$, and yes, $V_i^m$, can be interpreted as scores of the potential for interaction with opportunities in space.
@@
 [@soukhovIntroducingSpatialAvailability2023], spatial availability was designed for a homogeneous population traveling by a single mode of transportation.
-In this paper we extended spatial availability for the case of heterogeneous populations.
+In this paper, we extended spatial availability for the case of heterogeneous populations.
 We discussed this in terms of multiple modes of transportation, but the framework can accommodate equally well variations in travel behavior by population segments.
@@
 This difference for car travel in locations within and immediately around the LEZ Centro seems to increase the number of opportunities spatially available to transit users (transit being the second most competitive mode), as well as the spatial availability from the perspective of non-motorized modes.
-In effect, a policy such as Low Emission Zones helps to improve the accessibility situation of active travel and transit in the parts of the city where it is implemented.
-
-The purpose of the empirical example was to illustrate the kind of insights that can be derived from the application of multimodal spatial availability.
+In effect, a policy such as Low Emission Zones help to improve the accessibility situation of active travel and transit in the parts of the city where it is implemented.
+
+The purpose of the empirical example is to illustrate the kind of insights that can be derived from the application of multimodal spatial availability.
 But there are some intriguing opportunities for future research.
@@
 Ditto for destination choice.
-Our empirical example in this paper presented but a snapshot of this, and in future research it will be interesting to investigate changes _between_ policy interventions.
+Our empirical example presented but a snapshot of this, and in future research it will be interesting to investigate changes _between_ policy interventions.
 The expansion of Madrid's LEZ to the ring contained by the M-30 orbital presents an excellent opportunity to do so.

### 2023-11-10 Journal-manuscript/Journal-manuscript.tex
Finalized revisions and comments to reviewers (R1)

@@
 \usepackage{changepage}
+
+% Use Unicode characters when possible
+\usepackage[utf8x]{inputenc}
 
@@
 opportunities), and so each value can be meaningfully related to this
-total. A recent such effort is spatial availability, a
-singly-constrained accessibility measure. In this paper we extend
-spatial availability for use in the case of multiple modes or, more
-generally, heterogeneous population segments with distinct travel
-behaviors. After deriving a multimodal version of spatial availability
-We proceed to illustrate its features using a synthetic example. Next,
-we apply it to an empirical example in Madrid. We conclude the paper
-with suggestions for future research.
+total. A recent effort is spatial availability, a singly-constrained
+accessibility measure. In this paper we extend spatial availability for
+use in the case of multiple modes or, more generally, heterogeneous
+population segments with distinct travel behaviors. After deriving a
+multimodal version of spatial availability, we proceed to illustrate its
+features using a synthetic example. Next, we apply it to an empirical
+example in Madrid, Spain. We conclude the paper with suggestions for
+future research.
 
@@
 Accessibility is a key concept in the analysis of land use and
-transportation systems {[}e.g., 1,2,3{]}, and one that is coming of age
-from the perspective of planning too {[}see, \emph{inter alia},
-4,5--8{]}. Beginning with the work of Hansen {[}1{]}, accessibility
-measures have been widely used to evaluate the efficiency of
-transportation systems when combined with the distribution of
-opportunities in space {[}9{]}. As such, it is a holistic measure of
-spatial systems that measures the ease of reaching destinations
-{[}10,11{]}.
+transportation systems {[}e.g., 1,2,3{]}, and too is coming of age from
+the perspective of planning {[}see \emph{inter alia}, 4,5--8{]}.
+Beginning with the work of Hansen {[}1{]}, accessibility measures have
+been widely used to evaluate the efficiency of transportation systems
+when combined with the distribution of opportunities in space {[}9{]}.
+As such, it is a holistic measure of spatial systems that measures the
+ease of reaching destinations {[}10,11{]}.
 
@@
 available for any other person to take. More generally, opportunities
-are subject to congestion: for example, multiple people can avail
-themselves of the services of the same family doctor, but the more
-people who do, the more congested the service will be.
-
-The issue of congestion was the motivation for the development of
-floating catchment area approaches in accessibility analysis
-{[}12,13{]}. While these approaches purport to account for congestions,
-Paez et al. {[}14{]} demonstrate that in general they do not solve the
-issue of multiple counting of opportunities, thus leading to biases in
-the calculation of total demand and supply, sometimes inflating them,
-other times deflating them. In response to this, recent research has
-paid closer attention to the way opportunities are counted in
-accessibility analysis. {[}14{]}, for example, tackle floating catchment
-area methods and introduce a normalization of the impedance matrix to
-allocate the population and then the level of service proportionally.
-More recently, Soukhov et al. {[}15{]} introduced a singly-constrained
-measure of accessibility, called spatial availability, that employs a
-similar, but more sophisticated proportional allocation mechanism. The
-work of these authors shows that floating catchment area methods can be
-seen as singly-constrained accessibility measures, and improve on
-existing approaches by guaranteeing that each opportunity is counted
-only once - in other words, treating opportunities as \emph{finite}. The
+are subject to congestion: for example, multiple people can obtain
+services from the same family doctor, but the more people who do, the
+more congested the service will be.
+
+The issue of congestion in accessibility measures was the motivation for
+the development of floating catchment area approaches {[}12,13{]}. While
+these approaches purport to account for congestion, Paez et al. {[}14{]}
+demonstrate that in general they do not solve the issue of multiple
+counting of opportunities, thus leading to biases in the calculation of
+total demand and supply, sometimes inflating them, other times deflating
+them. In response to this, recent research has paid closer attention to
+the way opportunities are counted in accessibility analysis. Paez et al.
+{[}14{]}, for example, tackle floating catchment area methods and
+introduce a normalization of the impedance matrix to allocate the
+population and then the level of service proportionally. More recently,
+Soukhov et al. {[}15{]} introduced a singly-constrained measure of
+accessibility, called spatial availability, that employs a similar, but
+more sophisticated proportional allocation mechanism. The work of these
+authors show that floating catchment area methods can be seen as
+singly-constrained accessibility measures, and improve on existing
+approaches by guaranteeing that each opportunity is counted only once -
+in other words, treating opportunities as \emph{finite}. The
 proportional allocation of spatial availability constrains the
@@
 heterogeneous populations very relevant. In the case of multiple modes
-of transportation, people who travel by slow modes (e.g., active modes)
-can usually reach fewer opportunities than people who travel by faster
-modes and whose range is typically far wider (e.g., car). This implies
-that slower travelers will often face increased competition for local
-opportunities from travelers who can reach said opportunities from
+of transportation, people who travel by slower modes (e.g., active
+modes) can usually reach fewer opportunities than people who travel by
+faster modes and whose range is typically far wider (e.g., car). This
+implies that slower travelers will often face increased competition for
+local opportunities from travelers who can reach said opportunities from
 farther afield.
@@
 between older and younger adults {[}e.g., 16{]}, the propensity of older
-adults to use different modes of transportation {[}17{]}, the usually
-shorter trip lengths of children compared to grown-ups {[}e.g., 18{]},
-or the more limited travel ranges of single parents {[}e.g., 19{]}.
-
-The paper rest of this paper is organized as follows. In Section we
-discuss the use a synthetic examples with multiple modes to illustrate
-some relevant issues; this helps us motivate the derivation of the new
-spatial availability expressions. This is followed in Section 3 by an
+adults to use different modes of transportation {[}e.g., 17{]}, the
+usually shorter trip lengths of children compared to grown-ups {[}e.g.,
+18{]}, or the more limited travel ranges of single parents {[}e.g.,
+19{]}.
+
+The paper rest of this paper is organized as follows. In Section 2 we
+provide a brief review of multimodal accessibility. In the following
+Section 3 we demonstrate the derivation of the new spatial availability
+expression for multiple modes. In Section 4 we illustrate relevant
+issues through a synthetic example. This is followed in Section 5 by an
 empirical example using data from the city of Madrid after the
-implementation of its Low Emission Zones. Data for this example comes
-from the city's 2018 travel survey. The example shows the differences in
-spatial availability within and outside the LEZ for travelers using
-different modes, namely car, transit, cycling and walking. In Section 4,
-we provide concluding remarks on the strengths of the use of spatial
-availability as a multimodal accessibility measure, and discuss
-potential future uses in policy planning scenarios as well as directions
-for future research.
+implementation of its Low Emission Zones (LEZ). Data for this example
+comes from the city's 2018 travel survey. The example shows the
+differences in spatial availability within and outside the LEZ for
+travelers using different modes, namely car, transit, cycling and
+walking. In Section 6, we provide concluding remarks on the strengths of
+the use of spatial availability as a multimodal accessibility measure,
+and discuss potential future uses in policy planning scenarios as well
+as directions for future research.
 
@@
 generalized cost) into a score that represents the propensity for
-interaction. These measures originate from that proposed by {[}1{]},
-which can take the following form in the multimodal case:
+interaction. These measures originate from that proposed by Hansen
+{[}1{]}, which can take the following form in the multimodal case:
 \(S_i^m = \sum_j O_j f^m(c_{ij}^m)\) where \(m\) is a set of modes which
@@
 opportunities than car). But despite the discussion about ``sufficient
-accessibility'' (e.g., in the abstract), it is unclear what the
-unconstrained scores mean: is having access to but 10,000 jobs by
-transit insufficient? After all, 10,000 employment opportunities are
-still plenty of opportunities. These ratios can be found elsewhere in
-the literature {[}e.g., 23, and 9,16,19, and 8{]}, and they are useful
-to assess when some members of the public are better or worse off than
-others, but they do not say much about how bad is ``worse''.
+accessibility'', it is unclear what the unconstrained scores mean: is
+having access to but 10,000 jobs by transit insufficient? After all,
+10,000 employment opportunities are still plenty of opportunities. These
+ratios can be found elsewhere in the literature {[}e.g., 23, and
+9,16,19, and 8{]}, and they are useful as relative assessment of when
+some members of the public are better or worse off than others, but they
+do not say much about how bad is ``worse off''.
 
@@
 Once we understand opportunities to be finite, it is possible for an
-accessibility measure to take a crisper meaning. As considered in the
+accessibility measure to take on a crisper meaning. As considered in the
 long tradition of accessibility research, capacity of opportunities is
@@
 opportunity is subject to congestion or capacity constraint, even when
-the opportunities are conventionally seen as non-competitive. Amenities
-are a good example of this. For instance, standards for providing green
-spaces are often stated in the form of \emph{exclusive access}, in units
-of amenity per capita. A case in point is the Ile-de-France region, a
-jurisdiction that suggested in a major planning document of 2013 that at
-the municipal level at least \(10m^2\) of public green space should be
-supplied \emph{per inhabitant} {[}31{]}. But green spaces are not evenly
-distributed, which means that who has access to them hinges on where
-they are and how easy is to reach them. This formulation of provision of
-amenities is not unusual. For example, Natural England, an organization
-that recommends an Accessible Natural Greenspace Standard such that the
-minimum supply of space is one ha of statutory Local Nature Reserves per
-thousand population\footnote{see
+the opportunities are conventionally seen as non-competitive.
+
+Amenities are a good example of this. For instance, standards for
+providing green spaces are often stated in the form of \emph{exclusive
+access}, in units of amenity per capita. A case in point is a the
+Ile-de-France region, a jurisdiction that suggested in a 2013 planning
+document that at the municipal level at least \(10m^2\) of public green
+space should be supplied \emph{per inhabitant} {[}31{]}. Green spaces
+are not evenly distributed, which means that who has access to them
+hinges on where they are and how easy is to reach them. This formulation
+of provision of amenities is not unusual. For example, Natural England,
+an organization that recommends an Accessible Natural Greenspace
+Standard such that the minimum supply of space is one ha of statutory
+Local Nature Reserves per thousand population\footnote{see
   https://redfrogforum.org/wp-content/uploads/2019/11/67-Nature-Nearby\%E2\%80\%99-Accessible-Natural-Greenspace-Guidance.pdf}.
@@
 by anyone else?''. Green spaces often have large capacities, but they
-still have a capacity, and it is not the same a person to have access to
-\(5m^2\) of \emph{uncongested} green space than to \(15m^2\). This
+still have a capacity, and it is not the same for a person to have
+access to \(5m^2\) of \emph{uncongested} green space as \(15m^2\). This
 difference is in fact a matter of justice {[}31,33{]}. Constraining
@@
 that have high accessibility by transit are also the places that have
-\emph{very high} accessibility by car (Figure 2). Those two populations
-are going for the same opportunities, and those travelling by transit
-have fewer to choose from to begin with. More generally, people in a
-zone who are advantaged with relatively cost of travel will have the
-ability to potentially reach more opportunities than other people. Due
-to this advantage, through the perspective of finite opportunities,
-there are fewer opportunities left for everyone else, especially for
-those who use modes that are slower or more expensive.
-
-As noted in the introduction, competitive accessibility was the
+\emph{very high} accessibility by car (in their Figure 2). Those two
+populations are going for the same opportunities, and those travelling
+by transit have fewer to choose from to begin with. More generally,
+people in a zone who are advantaged with relatively low cost of travel
+will have the ability to potentially reach more opportunities than other
+people. Due to this advantage, through the perspective of finite
+opportunities, there are fewer opportunities left for everyone else,
+especially for those who use modes that are slower or otherwise more
+expensive.
+
+As noted in the Introduction, competitive accessibility was the
 rationale for developing floating catchment area methods (FCA),
-popularized by {[}13,12,although similar, and earlier, developments are
-found in 30,35{]}. Shen-type accessibility is formulated as:
+popularized by Luo et al. {[}13{]} who reformulated the work of Shen
+{[}12{]} into two steps (although similar, and earlier, developments are
+found in {[}30,35{]}). Shen-type accessibility is formulated as:
 \(a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}\) where \(D_j^m\)
@@
 meaning both population \emph{and} opportunities are multiply counted
-{[}see 14{]}. In other words, interpretation of differences in Shen-type
+{[}see 14{]}. In other words, interpretation of the Shen-type
 accessibility scores between modes is fraught as it is for Hansen-type
@@
 at many zones, arguing that this may further place those zones with
-lower-income populations at disadvantage. \(a_i\) and/or \(a_i^m\) are
+lower-income populations at a disadvantage. \(a_i\) and/or \(a_i^m\) are
 used to compare relative spatial differences in overall competitive
@@
 interaction and the gravity model {[}see 37{]}. In Soukhov et al.
-{[}15{]} these factors consider: 1) the mass effect (e.g., the size of
-populations at different origins); and 2) the cost of travel from
-different zones (e.g., some sub-populations face relatively higher or
-lower costs). The following section introduces the multimodal form of
-spatial availability.
+{[}15{]} these factors consider: the mass effect (e.g., the size of
+populations at different origins); and the cost of travel from different
+zones (e.g., some sub-populations face relatively higher or lower
+costs). The following section introduces the multimodal form of spatial
+availability.
 
@@
 
-\noindent we see that spatial availability is, like Hansen's measure, a
-weighted sum of the opportunities. What makes spatial availability stand
-apart from other approaches is the weight used in the sum, balancing
-factor \(F^t_{ij}\), implements a proportional allocation mechanisms to
-ensure that the sum of \(V_i\) is constrained to match the total number
-of opportunities in the region. As such, spatial availability is
-singly-constrained and naturally implements competition or congestion.
-\(F^t_{ij}\) consists of two parts. The first part is a population-based
-proportional allocation factor to model the mass effect of the gravity
-model: \[
+\noindent we see that spatial availability is, like the Hansen-type
+measure, a weighted sum of the opportunities. What makes spatial
+availability stand apart from other approaches is how the weight used in
+the sum, balancing factor \(F^t_{ij}\), implements a proportional
+allocation mechanisms to ensure that the sum of \(V_i\) is constrained
+to match the total number of opportunities in the region. As such,
+spatial availability is singly-constrained and naturally implements
+competition or congestion. \(F^t_{ij}\) consists of two parts. The first
+part is a population-based proportional allocation factor to model the
+mass effect of the gravity model: \[
 F^p_{i} = \frac{P_i}{\sum_i P_i}
@@
 This factor makes opportunities available based on demand. Secondly,
-there is an impedance-based proportional allocation factor that model
+there is an impedance-based proportional allocation factor that models
 the cost effect: \[
@@
 \item
-  \(V^m_{i}\) is the spatial availability \(V_{i}\) for mode \(m\) at
+  \(V^m_{i}\) is the spatial availability \(V_{i}\) for each \(m\) at
   each \(i\); the sum of \(V^m_{i}\) for all \(m\) at each \(i\) is
@@
 
-\caption{Accessibility\ values\ at\ each\ origin\ per\ mode\ m\ at\ each\ origin\ i\ and\ aggregated\ between\ modes\ for\ each\ i\ for\ the\ synthetic\ example.}\\
+\caption{Accessibility\ values\ at\ each\ origin\ (i)\ per\ mode\ (m)\ (columns\ three\ to\ five)\ and\ aggregated\ per\ i\ (columns\ six\ and\ seven)\ for\ the\ synthetic\ example.}\\
 
@@
 
-\ascline{1.5pt}{666666}{1-7}\endfirsthead \caption[]{Accessibility\ values\ at\ each\ origin\ per\ mode\ m\ at\ each\ origin\ i\ and\ aggregated\ between\ modes\ for\ each\ i\ for\ the\ synthetic\ example.}\\
+\ascline{1.5pt}{666666}{1-7}\endfirsthead \caption[]{Accessibility\ values\ at\ each\ origin\ (i)\ per\ mode\ (m)\ (columns\ three\ to\ five)\ and\ aggregated\ per\ i\ (columns\ six\ and\ seven)\ for\ the\ synthetic\ example.}\\
 
@@
 possible to use different impedance functions for the modes, as
-demonstrated in the empirical example below.
+demonstrated in the empirical example in the following section.
 
@@
 the modes) and does not usefully translate into any sort of benchmark.
-To connect this example to the literature, \(S_i^m\) is calculated in
-the work of {[}21{]}; they compare differences in \(S_i^m\) values
-between modes in a relative and comparative sense, but make no further
-interpretation of the \(S_i^m\) values. Large metropolitan regions will
-have large values and small regions small, but that is simply an
-artifact of size.
+To connect this example to the aforementioned literature, \(S_i^m\) is
+calculated in the work of Tahmasbi et al. {[}21{]}; they contrast
+differences in \(S_i^m\) values between modes in a relative and
+comparative sense, but make no further interpretation of the \(S_i^m\)
+values. More densely populated metropolitan regions will tend to have
+more opportunities and hence large \(S_i^m\) values and less densely
+regions, smaller values; how much of these differences may simply an
+artifact of region density?
 
@@
 \(a_i\) values since this center has the lowest travel impedance to
-opportunities (lower than at \(C\), \(A\) and \(B\) are equal) and has
-faces relatively low competition, not being close to a relatively large
+opportunities (lower than at \(C\), \(A\) and \(B\) are equal) and faces
+relatively low competition, not being close to a relatively large
 population (lower than at \(B\)).
@@
 {[}36{]} to calculate modal \(a_i^m\) values and the aggregated \(a_i\)
-is implemented in the work of {[}38{]}. However, similar to Hansen-type
-accessibility, these works discuss relative and spatially comparative
-differences in values, but veer from interpretating the values of
-\(a_i^m\) or \(a_i\) themselves. In fairness, interpretation is
-complicated by the multiple counting of opportunities between zones and
-modes.
+is implemented in the work of Carpentier et al. {[}38{]}. However,
+similar to Hansen-type accessibility, these works discuss relative and
+spatially comparative differences in values, but veer from interpreting
+the values of \(a_i^m\) or \(a_i\) themselves. In fairness,
+interpretation is complicated by the multiple counting of opportunities
+between zones and modes.
 
@@
 fifth column of Table 1 , the values of \(V_i^m\) in \(A\) and \(B\) are
-not the same within mode (as this measure considers competition). In
-fact, at \(A\), users of mode \(z\) capture 36,088.84 more spatially
+not the same within each mode (as this measure considers competition).
+In fact, at \(A\), users of mode \(z\) capture 36,088.84 more spatially
 available jobs (of the \(210,000\) jobs in the region) than the
@@
 The context for the empirical example is Madrid, Spain. This city
-implemented a Low Emission Zone in 2017, to pursue goals set out in the
-national climate change agenda, to cut nitrogen dioxide levels, and to
-prioritize people's movement in the city. Low Emission Zones (LEZs)
-elsewhere have been implemented as interventions to reduce GHG
-emissions, improve air quality, and support sustainable mobility
-{[}39,40{]}. Though the rules of exclusion vary by city, LEZs aim to
-deter/reduce traffic in designated zones under threat of penalty (e.g.,
-fines, seizure of vehicle). In other words, LEZ implement a form of
+implemented a Low Emission Zone (LEZ) in 2017 to: pursue goals set out
+in the national climate change agenda, cut nitrogen dioxide levels, and
+to prioritize people's movement in the city. LEZs elsewhere have
+similarly been implemented as interventions to reduce GHG emissions,
+improve air quality, and support sustainable mobility {[}39,40{]}.
+Though the rules of exclusion vary by city, LEZs aim to deter/reduce
+traffic in designated zones under threat of penalty (e.g., fines,
+seizure of vehicle). In other words, LEZs implement a form of
 \emph{geographic discrimination} as they change how people can reach
@@
 opportunities as finite in a region, this discrimination reduces the
-competition of one mode and opens up opportunties for other modes to
-thrive. In this way, at their core LEZs operate by changing the
-accessibility landscape of a city from the perspective of multiple
-modes.
+competition of one mode and opens up opportunities for other modes to
+better thrive. At their core, LEZs operate by changing the accessibility
+landscape of a city from the perspective of multiple modes.
 
 In geographic scope, the 2017 boundaries of the LEZ in Madrid were
-relatively modest, covering only approximately 4.72 \(km^2\) of the
-central business district of the city (the so-called LEZ Centro). As of
-this writing, there are plans to expand these boundaries to the area
-inside the M-30, an orbital highway in proximity to the city center
-(i.e., LEZ M-30). Within the 2017 LEZ Centro implementation, all cars,
-motorcycles and freight vehicles with environmental labels A or B (older
-makes and models of fossil fuel internal combustion engine vehicles),
-were disallowed from entering the zone unless they are used by residents
-or meet other exemptions. This restriction impacted approximately half
-of all car trips that used to travel into what is now the LEZ Centro
-{[}41{]}.
+relatively modest, covering only approximately 4.72
+km\textsuperscript{2} of the central business district of the city (the
+so-called LEZ Centro). As of this writing, there are plans to expand
+these boundaries to the area inside the M-30, an orbital highway in
+proximity to the city center (i.e., LEZ M-30). Within the 2017 LEZ
+Centro implementation, all cars, motorcycles and freight vehicles with
+environmental labels A or B (older makes and models of fossil fuel
+internal combustion engine vehicles), were disallowed from entering the
+zone unless they are used by residents or meet other exemptions. This
+restriction impacted approximately half of all car trips that used to
+travel into what is now the LEZ Centro {[}41{]}.
 
@@
 
-{\centering \includegraphics[width=0.85\linewidth]{images/i_jobs_zn208_plot}
+{\centering \includegraphics[width=0.85\linewidth]{images/Fig2}
 
@@
 \caption{\label{fig:Fig2} Distribution of jobs taken by people living and working in Madrid as reported in the 2018 travel survey.
-Gray TAZ has no jobs.
+Grey TAZs has no jobs.
 Ranges of values in the legend are quintiles.}\label{fig:jobs-plot}
@@
 
-{\centering \includegraphics[width=0.85\linewidth]{images/im_populations_zn208_plot}
+{\centering \includegraphics[width=0.85\linewidth]{images/Fig3}
 
@@
 \caption{\label{fig:Fig3} Population living and working in Madrid by mode of transportation as reported in the 2018 travel survey.
-Gray TAZs have no population.
+Grey TAZs have no population.
 Ranges of values in the legend are quintiles.}\label{fig:pop-plot}
@@
 and \ref{fig:Fig3} show the number of workers and the distribution of
-jobs in the City of Madrid by TAZ. The light gray boundary represents
-the LEZ Centro in effect in 2017. The survey was conducted after the
-introduction of LEZ Centro. The dark gray boundary represents the LEZ
-planned for for the boundaries of the M-30 highway and is present in the
-plots as a spatial reference for areas in proximity to the LEZ Centro.
+full-time jobs in the City of Madrid by TAZ. The light grey boundary
+represents the LEZ Centro in effect in 2017. The survey was conducted
+after the introduction of LEZ Centro. The dark grey boundary represents
+the LEZ planned for the boundaries of the M-30 highway and is present in
+the plots as a spatial reference for areas in proximity to the LEZ
+Centro.
 
@@
 \(P^m_i\), is displayed in Figure \ref{fig:Fig3}. The modal shares in
-Figure \ref{fig:Fig3} are calculated based on those measued in the
+Figure \ref{fig:Fig3} are calculated based on those measured in the
 survey. The modal categories and the mode types within each category are
@@
 \item
-  Transit: all bus, trams, and trains,
+  Transit: all bus, trams, and trains.
 \item
   Bike: all bicycle trips (e.g., private, public, or company bike trips)
-  and ``other'' types of micromobility options,
+  and ``other'' types of micromobility options.
 \item
@@
 
-Travel times are provided within the survey by mode. This information is
-used to calibrate mode-specific travel impedance functions
-\(f^m(c_{ij}^m)\). To illustrate the modal differences in travel times,
-the following descriptive statistics per mode are presented:
+Travel times are provided within the travel survey by mode. This
+information is used to calibrate mode-specific travel impedance
+functions \(f^m(c_{ij}^m)\). To illustrate the modal differences in
+travel times, the following descriptive statistics per mode are
+presented:
 
@@
 \item
-  Car/motor: mean 36 minutes (min:0 minutes, Q2: 15 minutes, Q3: 55
+  Car/motor: mean 36 minutes (min: 0 minutes, Q2: 15 minutes, Q3: 55
   minutes, max: 120 minutes)
 \item
-  Transit: mean 55 minutes (min:1 minutes, Q2: 30 minutes, Q3: 80
+  Transit: mean 55 minutes (min: 1 minutes, Q2: 30 minutes, Q3: 80
   minutes, max: 120 minutes)
 \item
-  Bike: mean 34 minutes (min:5 minutes, Q2: 15 minutes, Q3: 40 minutes,
+  Bike: mean 34 minutes (min: 5 minutes, Q2: 15 minutes, Q3: 40 minutes,
   max: 115 minutes)
 \item
-  Walk: mean 27 minutes (min:1 minutes, Q2: 10 minutes, Q3: 45 minutes,
+  Walk: mean 27 minutes (min: 1 minutes, Q2: 10 minutes, Q3: 45 minutes,
   max: 119 minutes)
@@
 shape and rate parameters for the gamma functions (motorized modes) are
-1.8651852 and 0.051468for car/motor and 2.7566235 and 0.0499193 for
+1.8651852 and 0.051468 for car/motor and 2.7566235 and 0.0499193 for
 transit; for the log-normal functions (non-motorized modes), the mean
@@
 
-{\centering \includegraphics[width=0.85\linewidth]{images/tlds_curves_m_plot}
+{\centering \includegraphics[width=0.85\linewidth]{images/Fig4}
 
@@
 
-\caption{\label{fig:Fig4} Fitted impedance function  against empirical TLD (bars) corresponding to the home-to-work origin-destination flows from the Madrid 2018 travel survey.}\label{fig:tlds-curves-m-plot}
+\caption{\label{fig:Fig4} Fitted impedance function against empirical TLD (bars) corresponding to the home to full-time work origin destination flows for the City of Madrid from the 2018 travel survey.}\label{fig:tlds-curves-m-plot}
 \end{figure}
@@
 
-At this point it is worthwhile to reiterate that the empirical example
+At this point, it is worthwhile reiterating that the empirical example
 is a snap-shot of spatial availability by mode using data from the 2018
-travel survey. Our purpose is to investigate the trends in availability
-of employment opportunities by mode, and illustrate how spatial
-availability can help in discussions about the competitive advantage of
-various modes within Madrid Centro. The spatial availability of jobs
-\(V_i^m\) is calculated for each of the four modes \(m\) at the level of
-traffic analysis zones \(i\) in Madrid and displayed in Figure
-\ref{fig:Fig5}.
+travel survey. Our purpose in this empirical example is to investigate
+the trends in availability of employment opportunities by mode, and
+illustrate how spatial availability can be used in discussions about the
+competitive advantage of various modes within Madrid Centro. The spatial
+availability of jobs \(V_i^m\) is calculated for each of the four modes
+\(m\) at the level of traffic analysis zones \(i\) in Madrid and
+displayed in Figure \ref{fig:Fig5}.
 
@@
 
-{\centering \includegraphics[width=0.85\linewidth]{images/SA_im_V_zn208_plot}
+{\centering \includegraphics[width=0.85\linewidth]{images/Fig5}
 
@@
 
-\caption{\label{fig:Fig5} Spatial availability of jobs per origin and mode $V_i^m$ in Madrid. Gray TAZs have no population. Ranges of values in the legend are quintiles.}\label{fig:SA-m-plot}
+\caption{\label{fig:Fig5} Spatial availability of jobs per origin and mode $V_i^m$ in Madrid. Grey TAZs have no population. Ranges of values in the legend are quintiles.}\label{fig:SA-m-plot}
 \end{figure}
@@
 total population. However, the ability to travel at greater speeds also
-impacts this results. Fuerthermore, differences in \(V_i^m\) values
+impacts these results. Furthermore, differences in \(V_i^m\) values
 within modes also exist in space: car users outside of the M-30 region
@@
 of certain modes, although this effect is not geographically uniform. As
-seen in the left-most columns of Figure \ref{fig:Fig6}, `car/motor' and
-`transit' together can avail themselves of 95.3\% of all jobs in the
-city. However, `car/motor' has a disproportionate share of \(V_i^m\)
-relative to the population of users of this mode, compared to
-opportunities that are spatially available to transit users. The
-combined population of car and transit users is 36.6\% and 47.5\%
-respectively, but these populations are allocated 48.0\% and 47.3\%
-respectively, of the city's jobs. When treating the number of
-opportunities that can be reached as a finite value (total: 847,574
-opportunities), fewer opportunities are spatially available to slower
-modes (i.e., walking and cycling), even taking into account that their
-share is smaller overall. These modes are at a disadvantage as a result
-of: the travel impedance for longer trips (see Figure \ref{fig:Fig4};
-their low population values values overall; and larger populations in
-origins with high shares of travel by motorized modes. These factors all
-contribute to the the car/motor mode being most advantaged in capturing
-spatially available job opportunities overall.
+seen in the left-most columns of Figure \ref{fig:Fig6}, users of
+`car/motor' and `transit' together can avail 95.3\% of all jobs in the
+city (Spatial Availability by Mode). However, car/motor users have a
+disproportionate share of \(V_i^m\) relative to the population of users
+of this mode (Population by Mode), compared to opportunities that are
+spatially available to transit users. The combined population of car and
+transit users is 36.6\% and 47.5\% respectively, but these populations
+are allocated 48.0\% and 47.3\% respectively, of the city's jobs. When
+treating the number of opportunities that can be reached as a finite
+value (total: 847,574 opportunities), fewer opportunities are spatially
+available to slower modes (i.e., walking and cycling), even taking into
+account that their share is smaller overall. These modes are at a
+disadvantage as a result of: the travel impedance for longer trips (see
+Figure \ref{fig:Fig4}; their low population values values overall; and
+larger populations in origins with high shares of travel by motorized
+modes. These factors all contribute to the the car/motor mode being most
+advantaged in capturing spatially available job opportunities overall.
 
@@
 
-{\centering \includegraphics[width=0.85\linewidth]{images/modal_V_comps_4plot}
+{\centering \includegraphics[width=0.85\linewidth]{images/Fig6}
 
@@
 non-car users within Centro. In the case of active travel, the
-proportions of cyclists and walkers within LEZ Centro still exceed the
-proportions of jobs spatially available to them --- however the
+proportions of cyclists and walkers within LEZ Centro still exceeds the
+proportions of jobs spatially available to them --- however, the
 disparity is drastically reduced compared to the rest of the city. As
@@
 represent 1.2\% and 14.7\% of the population. By restricting the ability
-of cars to enter Centro, the LEZ seems to contributes to level the
+of cars to enter Centro, the LEZ seems to contribute to leveling the
 playing field for slower modes, in particular cycling and walking, but
-also transit. As seen in the figure, transit users are generally close
-to parity across the region, with nearly as many spatially available
-jobs as transit users. Still, this mode has the greatest advantage in
-LEZ Centro with 68.5\% of spatially available jobs in Centro for 51.4\%
-of transit users in Centro. This result makes intuitive sense: after
-car, it is the mode with the greatest range, and unlike car it is
-allowed in LEZ Centro.
+also transit. As seen in the Figure \ref{fig:Fig6}, transit users are
+generally close to parity across the region, with nearly as many
+spatially available jobs as transit users. Still, this mode has the
+greatest advantage in LEZ Centro with 68.5\% of spatially available jobs
+in Centro for 51.4\% of transit users in Centro. This result makes
+intuitive sense: after car, it is the mode with the greatest range, and
+unlike car it is unrestricted in the LEZ Centro.
 
@@
 availability \(V_i^m\) divided by the population of users of \(m\).
-Values of \(v_i^m\) below 1 are shown in shades of orange, and indicate
-TAZs with less than one spatially available opportunity per capita for
-the mode. Values above 1 are shown in shades of green, and indicate TAZs
-with more than one spatially available opportunity per capita for the
-mode. The highest spatial availability per capita (shown in blue) is for
-car users in a zone northeast just beyond the M-30. These plots
-illustrate in unambiguous fashion, and in a quantity that is comparable
-over space and time, the advantage in terms of spatial availability of
-car for most of the city (bottom left plot, areas denoted with green
-\(v_i^m\) values above 1). It can also be observed that spatial
-availability of jobs is relatively well balanced for transit users over
-most of the regions (i.e., many zones are light orange or light green).
-Spatial availability of jobs for non-motorized modes, in contrast, is
-low (under 1) overall, although less so within LEZ Centro.
+Values of \(v_i^m\) below one are shown in shades of orange, and
+indicate TAZs with less than one spatially available opportunity per
+capita for the mode. Values above one are shown in shades of green, and
+indicate TAZs with more than one spatially available opportunity per
+capita for the mode. The highest spatial availability per capita (shown
+in blue) is for car users in a zone northeast just beyond the M-30.
+These plots illustrate in unambiguous fashion, and in a quantity that is
+comparable over space and time, the advantage in terms of spatial
+availability of car for most of the city (bottom left plot, areas
+denoted with green \(v_i^m\) values above one). It can also be observed
+that spatial availability of jobs is relatively well balanced for
+transit users over most of the regions (i.e., many zones are light
+orange or light green). Spatial availability of jobs for non-motorized
+modes, in contrast, is low (under one) overall, although less so within
+LEZ Centro.
 
@@
 
-{\centering \includegraphics[width=0.85\linewidth]{images/SA_im_vv_zn208_plot}
+{\centering \includegraphics[width=0.85\linewidth]{images/Fig7}
 
@@
 \caption{\label{fig:Fig7} Distribution of spatially available jobs per capita by mode of transportatin ($v_i^m$).
-Gray TAZs have no population that use the mode.
+Grey TAZs have no population that use the mode.
 Ranges of values in the legend are quintiles.}\label{fig:SA-per-capita-m-plot}
@@
 
-Incidentally, it is bears noting that \(v_i^m\) for car within and near
-LEZ Centro is close to or below 1 in Figure \ref{fig:Fig7}, while all
-non-car modes have relatively higher \(v_i^m\) values. Since these
-values are comparable across regions and over time, Figure
-\ref{fig:Fig7} potentially provides a benchmark for quantifying changes
-in LEZ policies in the future. As Figure \ref{fig:Fig7} also shows, many
-areas within the M-30 have high (white/green) \(v_i^m\) values for car,
-but the results for LEZ Centro give reasonable grounds to speculate that
-a spatial expansion of the LEZ to include all areas within the M-30
-would likely increase the spatial availability of jobs for transit
-users, cyclists and pedestrians.
+Incidentally, \(v_i^m\) values for car within and near LEZ Centro is
+close to or below one in Figure \ref{fig:Fig7}, while all non-car modes
+have relatively higher \(v_i^m\) values. Since these values are
+comparable across regions and over time, Figure \ref{fig:Fig7}
+potentially provides a benchmark for quantifying changes in LEZ policies
+in the future. As Figure \ref{fig:Fig7} also shows, many areas within
+the M-30 have high (white/green) \(v_i^m\) values for car, but the
+results for LEZ Centro give reasonable grounds to speculate that a
+spatial expansion of the LEZ to include all areas within the M-30 would
+likely increase the spatial availability of jobs for transit users,
+cyclists and pedestrians.
 
@@
 Accessibility measures are an important tool in transportation research
-{[}9{]} and increasingly also seen as valuable for planning purposes
+{[}9{]} and are increasingly seen as valuable for planning purposes
 {[}4--8{]}. They boast a long history of development, beginning with
@@
 {[}50{]}, Reilly {[}51{]}, Stewart {[}52--54{]}, Zipf {[}55,56{]},
-Wilson {[}37{]}, and many others besides. In this way, \(S_i^m\),
-\(a_i^m\), and yes, \(V_i^m\), can be interpreted as scores of the
-potential for interaction with opportunities in space.
+Wilson {[}37{]}, and many others. In this way, \(S_i^m\), \(a_i^m\), and
+yes, \(V_i^m\), can be interpreted as scores of the potential for
+interaction with opportunities in space.
 
@@
 et al. {[}15{]}, spatial availability was designed for a homogeneous
-population traveling by a single mode of transportation. In this paper
+population traveling by a single mode of transportation. In this paper,
 we extended spatial availability for the case of heterogeneous
@@
 the perspective of non-motorized modes. In effect, a policy such as Low
-Emission Zones helps to improve the accessibility situation of active
+Emission Zones help to improve the accessibility situation of active
 travel and transit in the parts of the city where it is implemented.
 
-The purpose of the empirical example was to illustrate the kind of
+The purpose of the empirical example is to illustrate the kind of
 insights that can be derived from the application of multimodal spatial
@@
 the accessibility landscape. Ditto for destination choice. Our empirical
-example in this paper presented but a snapshot of this, and in future
-research it will be interesting to investigate changes \emph{between}
-policy interventions. The expansion of Madrid's LEZ to the ring
-contained by the M-30 orbital presents an excellent opportunity to do
-so. Given the intuitive and straightforward interpretation of spatial
-availability scores as fractions of opportunities from the total,
-relative and absolute changes in the accessibility landscape can be
-assessed, thus helping to evaluate the implications of policy
-interventions.
+example presented but a snapshot of this, and in future research it will
+be interesting to investigate changes \emph{between} policy
+interventions. The expansion of Madrid's LEZ to the ring contained by
+the M-30 orbital presents an excellent opportunity to do so. Given the
+intuitive and straightforward interpretation of spatial availability
+scores as fractions of opportunities from the total, relative and
+absolute changes in the accessibility landscape can be assessed, thus
+helping to evaluate the implications of policy interventions.
 

### 2023-11-10 Journal-manuscript/Title-page.Rmd
V0 (Journal-manuscriptV0) .tex, the updated Title-page, and the diff between V0 and R1 (Journal-manuscript) .tex/.pdf created.

@@
 ---
-title: "Multimodal spatial availability: a singly-constrained measure of competitive accessibility considering multiple modes"
+title: "Multimodal spatial availability: a singly-constrained measure of accessibility considering multiple modes"
 author:

### 2023-11-10 Journal-manuscript/Editorial correspondance/Response-to-reviewers_R1.Rmd
The R1 submitted version, and small tweaks to editorial correspondance.

@@
 ---
-title: "PLoS ONE submission: Response to reviewers"
-output: pdf_document
+title: "PLOS ONE submission: Response to reviewers"
+output: pdf
 urlcolor: blue
-date: "2023-10-18"
+date: "2023-11-09"
 bibliography:
@@
 
-\textcolor{blue}{Futhermore, average travel times for car/motor and transit are longer than bike and walk.
+\textcolor{blue}{Furthermore, average travel times for car/motor and transit are longer than bike and walk.
 All people don't have access to all options - completely true.
@@
 
-\textcolor{blue}{We have updated the captions of the figures to reflec the meaningof grey colour blocks and the colour scheme of all Figures.
+\textcolor{blue}{We have updated the captions of the figures to reflect the meaning of grey colour blocks and the colour scheme of all Figures.
 We've also updated the labels on the legends to make them more interpretable.}

### 2023-11-10 Journal-manuscript/Editorial correspondance/Response-to-reviewers_R1.txt
The R1 submitted version, and small tweaks to editorial correspondance.

Please see the 'Response-to-reviewer.pdf' for the formatted letter.

#Reviewer 1

Thank you for this comment regarding "1. Lack of Innovation". We would like to bring to your attention the editorial philosophy of PLoS ONE, according to which editors "make decisions on submissions based on scientific rigor, regardless of novelty," (see https://journals.plos.org/plosone/s/editorial-and-peer-review-process). Nonetheless we wish to respond to this comment.

In this paper we extend our earlier work on spatial availability (https://doi.org/10.1371/journal.pone.0278468) for the simultaneous analysis of multiple modes of transportation. You appear to be under the misaprehension that the main difference with accessibility measures is to incorporate "the ratio of travel frequencies as weight in travel costs". This statement is not correct. The main difference is the proportional allocation mechanism that we introduced for spatial availability, and that is extended in this paper to allocate opportunities based on the proportion of travelers by different modes. This is not the same as "weighting the travel cost". In fact, each mode is modelled using its own impedance function, as shown in our empirical example, where we use origin-destination data by mode to estimate travel impedance functions specific to each mode.

While this enhancement may appear minor in your opinion, it is not plainly obvious how spatial availability applies to multiple modes, which is why we believe this paper is needed. We do not dispute that this is an incremental step in the development of a more general method, but it is a step that considerably expands the
range of potential applications. Further, we contend that the method has been rigorously developed and demonstrated this using an open, transparent, and reproducible example. In response to this comment we have edited the paper to more clearly describe the advantages that multimodal spatial availability offers when considering the accessibility to opportunities by different modes. In particular, proportional allocation of opportunities using mode-specific impedance functions means that the travel- cost-advantage of each mode can be analysed. Further, this mechanism ensures that the sum of all spatial availability values for all modes sum up to the total number of opportunities in the region, which is not true for any other type of accessibility measure. With respect to the "superiority" of the method, a reader of our earlier paper would already be aware of what limitations of accessibility analysis spatial availability aims to address. In this paper we also try to demonstrate throughout the manuscript what our measure does that others can’t. Since spatial availability values result from proportional allocation, each value is a proportion of 
the total number of opportunities. Put another way, we can compare the proportion of opportunities available to users of each mode, to users at each zone, to users of each mode by zone, and so on, and the values relate directly to the total opportunities in the region. This also allows us to calculate values per capita that serve as benchmark values, as shown in the empirical example.

Thank you for this comment regarding "2. Insufficient Experimental Data". To clarify, the empirical data used is not experimental. It is observational, since it is collected using a travel survey conducted by the City of Madrid. Travel surveys are a standard instrument in transportation planning and research, and are conducted in cities around the world. The data we work with represent the most recent and most complete travel survey conducted for the region to date. We are not completely sure where you got the figure of "30,000 trips" (which you consider significantly limited); presumably you are citing the maximum ’opportunity’ or ’population’ numbers in Figures 2 or 3. The most job-rich TAZs have 30,000 jobs while the least have 1,000 or fewer (Figure 2). In fact, there are 847,574 jobs in the city, which is also the sum of the total spatial availability in our analysis, as well as the number of potential trips to work.

With respect to The "typicality" of the data, travel surveys are designed to provide information about personal travel on a typical day, usually during a period of maximum demand (i.e., not during the summer vacation). This is standard practice for these surveys, and the one for Madrid is no different in this respect.

Regarding "3. Lack of Data-Driven Analysis". Thank you for this comment. It is somewhat puzzling that "30,000 trips" would be significantly limited (as per your comment #2), and at the same time be "a large amount of data". We find that this comment in particular is not actionable due to its vagueness. There are no meaningful responses to "somewhat one-sided" when "somewhat" and "one-sided" do not quantify or refer to a particular side. We studied the modes available
in the region, and examined the results from the perspective of each mode. What other perspectives do you suggest? What dimensions should be explored?

Despite this comment being nonactionable, we believe that with the additional detail added to the manuscript to address your and other reviewers’ comments, that overall clarity has been improved.

Regarding "4. Disorganized format". Reviewer #2 identified some specific formatting issues and we fixed them. Otherwise we proofread the paper and hopefully did not leave a typo behind. We also made sure Table 1 starts on a new page, when it runs across pages the formatting resulted in some disorganization. Thank you again for your efforts reviewing our submission.
---

#Reviewer 2

Thank you for your thoughtful review of our paper. Your comments were very helpful to improve the clarity of the research.

Regarding "method limits: focuses on the competition for jobs and is not useful for studying access to non-competitive or semi-competitive resources". Thank you for this comment. We would begin by noting that there are many types of opportunities that are mutually exclusive due to competition. For this paper we focused on jobs, but there are many others, such as beds at hospitals, seats at schools, and so on. Thus, even if the measure was applicable only to exclusive opportunities, there are numerous applications to choose from.

That said, we have grappled for some time with the question of what types of opportunities are best analysed using spatial availability. Our current thinking on this matter, after much consideration, is that in practice every type of opportunity, even when not clearly exclusive, is subject at least to congestion or capacity constraint. For instance, green spaces are often considered non-competitive, however, standards for the provision of such amenties are provided in the form of units of amenity per capita. For example, a case could the Ile-de-France region, a jurisdiction that suggested in a major planning document of 2013 that at the municipal level at least 10m^2 of public green space should be supplied per inhabitant (Liotta et al. 2020). But green spaces are not evenly distributed, which means that who has access to them hinges on where they are and how easy is to reach them. Formulating the provision of amenities in these terms in not rare. For example, Natural England recommends an Accessible Natural Greenspace Standard such that the minimum supply of space is one ha of statutory Local Nature Reserves per thousand population (see https://redfrogforum.org/wp-content/uploads/2019/11/67-Nature-Nearby%E2%80%99-Accessible-Natural-Greenspace-
Guidance.pdf) . Similarly, the World Health Organization (cited in OECD, 2013) recommends that cities provide a minimum of 9m^2 of green area per inhabitant (see https://doi.org/10.1787/9789264191808-en). For our purposes, standards of this type translate into "how much of this resource is available to one individual that has not been claimed by anyone else?". Green spaces often have large capacities, but they still have a capacity, and it is not the same for a person to have access to 5m^2 of uncongested green space than to 15m^2. This difference is in fact a matter of justice (Lara-Valencia and García-Pérez 2015; Liotta et al. 2020). Constraining accessibility is in this way a useful way to evaluate the congested availability of any type of opportunity. As standards are emphasized in the planning literature, in particular for fairness in transportation (see Martens and Golub 2021), spatial availability analysis can be used to assess standards. We are convinced that as other researchers discover this new approach to measuring accessibility many other applications will be found.

Regarding "method limits: does not allow for absolute gains or losses to be studied". Thank you for this comment. Spatial availability can certainly be used to capture absolute gains and losses. In fact, logically, the gains and losses produced by using a competitive and constrained measure allows for a clear interpretation. This would require the analyst to estimate the accessibility before and after some change to the land use or transportation system. In this paper our empirical application is a single scenario to serve as proof of concept, but in future research we intend to use spatial availability to analyse changes in the size of Madrid’s Low Emissions Zone implementation from a modal and socio-economic equity perspective.

Regarding "method limits: modal shift?". This is an excellent point. Accessibility measures, including spatial availability, are not meant to function as modal split models, but they are certainly amenable to analysis of changes in the accessibiity landscape if different modal shares are used as inputs. This is similar to the case of destination choice: accessibility measures do not model this, but changes in destination choices can be incorporated via how they affect the impedance function. In terms of how this could be implemented, the process can be sketched as follows: the results of a modal split model are used to estimate new modal shares, which in turn are used to recalculate spatial availability. The new values then can be compared to the baseline scenario. In other words, the framework for spatial availability is sufficiently flexible to take in not only mode-specific travel impedance functions, but also the proportions of the populations using each mode.

Regarding "suggestions for specifying method limits". Your comments have been very helpful to improve the clarity of the paper, as well as the scope of what we do, as well as directions for future research. For example, we now note in the "Discussion and conclusions" that "our example dealt with differences in travel by mode only, but it is possible to think of the intersection between mode of travel and different types of travelers. This would expand the number of sub-populations in the analysis from, say, m = M (modes) to m = M · Q (modes times population segments), each with their own characteristic impedance function. Evaluations of this kind will be especially relevant as LEZ are implemented in cities globally, and the question of their impact on disadvantaged populations who have become mobility-restricted increasingly come to the fore (De Vrij and Vanoutrive 2022; Verbeek and Hincks 2022; Liotta 2023)."

Regarding "improvements to abstract and introduction". We have now done this. In the original version of the abstract and introduction we gave the misleading impression that we would analyse changes in the system, when in reality our intent was to demonstrate the application of the measure in an empirical example. We do plan to study changes in the system, but this requires more work than can be presented in a single paper, partly for the reasons that you identified above (modal shifts and the calculation of spatial availability for two different scenarios, as well as the analysis of the differences between them). We plan to do this in a future paper focused on the policy instead of the presentation of a new method. Thank you again for your thoughtful comments and suggestions to improve the paper.

Regarding "minor comments". Updated! Apologies, car: ’36 min’ corresponds to a mean of 36 minutes and then within the brackets
additional descriptive statistics (minimum value, maximum value, etc.). We also fixed the formatting issues specified.
---

#Reviewer 3

Regarding "major issues". This is an excellent comment. The short answer is that whenever a destination can be reached by more than one mode, users of those modes are in competition for the opportunities there.

In this revision we tried to improve the discussion to make this point more clear. The impedance functions for all four modes are not the same. They describe the travel behaviour of commuters as informed by the 2018 travel survey. To follow your example, someone who cannot walk to work because their job is 20 km from where they live, will not compete for that job against people from their same origin who do walk. However, if the place where their work is can be reached by anyone who can walk from other origins (say, someone who lives closer to that destination), they would be in competition for the same opportunity.

Furthermore, average travel times for car/motor and transit are longer than bike and walk. All people don’t have access to all options - completely true. But the travel impedances reflect this real travel at an aggregate based on all the trips for a mode. And on average, it is assumed that people at each origin that take a mode to a destination are in direct competition for opportunities (as opportunities are finite) – and a part of the competition is defined by the mode-specific impedance function (the second part is the population balancing factor). This assumption, that all populations, no matter their mode, are competing for the same finite set of opportunity, is part of spatial availability. We’ve made an effort to make this more clear in the text.

Regarding "minor issues". We have updated the captions of the figures to reflect the meaning of grey colour blocks and the colour scheme of all Figures. We’ve also updated the labels on the legends to make them more interpretable.

### 2023-11-10 Journal-manuscript/Journal-manuscript.Rmd
The R1 submitted version, and small tweaks to editorial correspondance.

@@
 After deriving a multimodal version of spatial availability, we proceed to illustrate its features using a synthetic example.
-Next, we apply it to an empirical example in Madrid, Spain.
-We conclude the paper with suggestions for future research.
-
-keywords: ["Spatial availability", "Multimodal", "Singly-constrained", "Accessibility", "Madrid", "Reproducible research"]
+Next, we apply it to an empirical example of low emission zones in Madrid, Spain.
+We conclude the paper with suggestions for future research and its use in evaluating policy interventions.
+
+keywords: ["Spatial availability", "Multimodal", "Singly-constrained", "Accessibility", "Competitive", "Madrid", "Reproducible research"]
 
@@
 [@tahmasbiMultimodalAccessibilitybasedEquity2019] show that car travel affords the highest $S_i^{m}$ values in the majority of $i$, i.e., travelers who use a car can potentially reach more retail opportunities than populations using other modes.
-However, higher $S_i^{m}$ values for car do not affect the values of $S_i^{m}$ for other modes: in effect, each mode is analyzed as if the others did not exist.
+However, higher $S_i^{m}$ values for car do not affect the values of $S_i^{m}$ for other modes: in effect, each mode is analysed as if the others did not exist.
 Since the measure is not constrained, each opportunity is typically counted multiple times within and between modes, and as a result the sum of accessibility is not necessarily a meaningful quantity.
@@
 This distribution is likely caused by a variety of factors including: transit coverage and service within with city,  effective car infrastructure outside of the M-30, and/or the impact of the LEZ Centro itself.
-From Figure \ref{fig:Fig3}, it can be seen that active travel is less common than motorized trips at `r (pop_props$population_m[1]/sum(pop_props$population_m)) %>% percent()` and `r (pop_props$population_m[2]/sum(pop_props$population_m)) %>% percent()` for cycling and walking respectively.
+From Figure \ref{fig:Fig3}, it can be seen that active travel is less common than motorised trips at `r (pop_props$population_m[1]/sum(pop_props$population_m)) %>% percent()` and `r (pop_props$population_m[2]/sum(pop_props$population_m)) %>% percent()` for cycling and walking respectively.
 Noticeably, there is a positive trend between the walking and cycling in zones where transit is also present.
@@
 To fit the impedance functions, we use the Maximum likelihood estimation and the Nelder-Mead method for direct optimization available within the R {fitdistrplus} package [@fitdistrplus_2015].
-Based on goodness-of-fit criteria and associated diagnostics, the gamma and log-normal probability density functions are selected as best fitting curves for the motorized and non-motorized modes respectively.
+Based on goodness-of-fit criteria and associated diagnostics, the gamma and log-normal probability density functions are selected as best fitting curves for the motorised and non-motorised modes respectively.
 The selection of functional forms aligns with empirical examples in other regions [@reggianiAccessibilityImpedanceForms2011; @soukhovTTS2016RDataSet2023; @soukhovIntroducingSpatialAvailability2023].
-The shape and rate parameters for the gamma functions (motorized modes) are `r HW_car_motor_gamma$estimate[1]` and `r HW_car_motor_gamma$estimate[2]` for car/motor and `r HW_transit_gamma$estimate[1]` and `r HW_transit_gamma$estimate[2]` for transit; for the log-normal functions (non-motorized modes), the mean and standard deviation parameters are `r HW_bike_lnorm$estimate[1]` and `r HW_bike_lnorm$estimate[2]` for bike and `r HW_walk_lnorm$estimate[1]` and `r HW_bike_lnorm$estimate[2]` for walk.
+The shape and rate parameters for the gamma functions (motorised modes) are `r HW_car_motor_gamma$estimate[1]` and `r HW_car_motor_gamma$estimate[2]` for car/motor and `r HW_transit_gamma$estimate[1]` and `r HW_transit_gamma$estimate[2]` for transit; for the log-normal functions (non-motorised modes), the mean and standard deviation parameters are `r HW_bike_lnorm$estimate[1]` and `r HW_bike_lnorm$estimate[2]` for bike and `r HW_walk_lnorm$estimate[1]` and `r HW_bike_lnorm$estimate[2]` for walk.
 
@@
 For example, trips shorter than 5 minutes do not occur frequently for any mode; this reflects the spatial separation between places of residence and places of work commonly seen in many cities.
-In terms of the non-motorized modes, there is a preference towards walking trips around 15 minutes in duration, as seen from the highest value of $f^{walk}(c_{ij}^{walk})$.
+In terms of the non-motorised modes, there is a preference towards walking trips around 15 minutes in duration, as seen from the highest value of $f^{walk}(c_{ij}^{walk})$.
 With respect to travel by bicycle, longer travel times are more common; although the highest value of the impedance also corresponds to approximately 15 minutes, the curve has a longer tail and values decrease less rapidly at longer travel times than is the case of $f^{walk}(c_{ij}^{walk})$.
-A similar trend can be observed for the motorized modal options where transit mode is more spread out than car/motor mode.
+A similar trend can be observed for the motorised modal options where transit mode is more spread out than car/motor mode.
 All in all, these functions represent the propensity of travel by mode by duration of trip, and are used to calculate the proportional allocation factors $F_{ij}^m$ for $V_i^m$.
@@
 When treating the number of opportunities that can be reached as a finite value (total: 847,574 opportunities), fewer opportunities are spatially available to slower modes (i.e., walking and cycling), even taking into account that their share is smaller overall.
-These modes are at a disadvantage as a result of: the travel impedance for longer trips (see Figure \ref{fig:Fig4}; their low population values values overall; and larger populations in origins with high shares of travel by motorized modes.
+These modes are at a disadvantage as a result of: the travel impedance for longer trips (see Figure \ref{fig:Fig4}; their low population values values overall; and larger populations in origins with high shares of travel by motorised modes.
 These factors all contribute to the the car/motor mode being most advantaged in capturing spatially available job opportunities overall.
@@
 It can also be observed that spatial availability of jobs is relatively well balanced for transit users over most of the regions (i.e., many zones are light orange or light green).
-Spatial availability of jobs for non-motorized modes, in contrast, is low (under one) overall, although less so within LEZ Centro.
+Spatial availability of jobs for non-motorised modes, in contrast, is low (under one) overall, although less so within LEZ Centro.
 
@@
 ANSWER TO THIS PARAGRAPH, V_i is a location-based measure.
-it tells us that the jobs within the centro can be more easily available to non-motorized modes than motorized modes.
+it tells us that the jobs within the centro can be more easily available to non-motorised modes than motorised modes.
 This is in part because of the LEZ but also because of walkability/land -use, of course.
@@
 I conclude that restricting car modes leaves more opportunities for all other modes.
-I think this is because transit accessibility is in direct competition with car travel, and not in direct competition with non-motorized modes (need to demonstrate this - perhaps this can be shown with an x-y correlation between v_im valuesSee v_im_per_pop_plot.png (v_im per population) as well as the all_modes_v_im_plot plot.) -->
+I think this is because transit accessibility is in direct competition with car travel, and not in direct competition with non-motorised modes (need to demonstrate this - perhaps this can be shown with an x-y correlation between v_im valuesSee v_im_per_pop_plot.png (v_im per population) as well as the all_modes_v_im_plot plot.) -->
 
@@
 The example suggests that once that opportunities are treated as being finite, restrictions to travel by car leave more spatially available opportunities for non-car-users.
-This difference for car travel in locations within and immediately around the LEZ Centro seems to increase the number of opportunities spatially available to transit users (transit being the second most competitive mode), as well as the spatial availability from the perspective of non-motorized modes.
+This difference for car travel in locations within and immediately around the LEZ Centro seems to increase the number of opportunities spatially available to transit users (transit being the second most competitive mode), as well as the spatial availability from the perspective of non-motorised modes.
 In effect, a policy such as Low Emission Zones help to improve the accessibility situation of active travel and transit in the parts of the city where it is implemented.

### 2023-11-10 Journal-manuscript/Journal-manuscript.tex
The R1 submitted version, and small tweaks to editorial correspondance.

@@
 \usepackage{changepage}
-
-% Use Unicode characters when possible
-\usepackage[utf8x]{inputenc}
 
@@
 features using a synthetic example. Next, we apply it to an empirical
-example in Madrid, Spain. We conclude the paper with suggestions for
-future research.
+example of low emission zones in Madrid, Spain. We conclude the paper
+with suggestions for future research and its use in evaluating policy
+interventions.
 
@@
 affect the values of \(S_i^{m}\) for other modes: in effect, each mode
-is analyzed as if the others did not exist. Since the measure is not
+is analysed as if the others did not exist. Since the measure is not
 constrained, each opportunity is typically counted multiple times within
@@
 \ref{fig:Fig3}, it can be seen that active travel is less common than
-motorized trips at 1\% and 15\% for cycling and walking respectively.
+motorised trips at 1\% and 15\% for cycling and walking respectively.
 Noticeably, there is a positive trend between the walking and cycling in
@@
 associated diagnostics, the gamma and log-normal probability density
-functions are selected as best fitting curves for the motorized and
-non-motorized modes respectively. The selection of functional forms
+functions are selected as best fitting curves for the motorised and
+non-motorised modes respectively. The selection of functional forms
 aligns with empirical examples in other regions {[}15,47,48{]}. The
-shape and rate parameters for the gamma functions (motorized modes) are
+shape and rate parameters for the gamma functions (motorised modes) are
 1.8651852 and 0.051468 for car/motor and 2.7566235 and 0.0499193 for
-transit; for the log-normal functions (non-motorized modes), the mean
+transit; for the log-normal functions (non-motorised modes), the mean
 and standard deviation parameters are 3.2372212 and 0.7575986 for bike
@@
 and places of work commonly seen in many cities. In terms of the
-non-motorized modes, there is a preference towards walking trips around
+non-motorised modes, there is a preference towards walking trips around
 15 minutes in duration, as seen from the highest value of
@@
 is the case of \(f^{walk}(c_{ij}^{walk})\). A similar trend can be
-observed for the motorized modal options where transit mode is more
+observed for the motorised modal options where transit mode is more
 spread out than car/motor mode. All in all, these functions represent
@@
 Figure \ref{fig:Fig4}; their low population values values overall; and
-larger populations in origins with high shares of travel by motorized
+larger populations in origins with high shares of travel by motorised
 modes. These factors all contribute to the the car/motor mode being most
@@
 transit users over most of the regions (i.e., many zones are light
-orange or light green). Spatial availability of jobs for non-motorized
+orange or light green). Spatial availability of jobs for non-motorised
 modes, in contrast, is low (under one) overall, although less so within
@@
 second most competitive mode), as well as the spatial availability from
-the perspective of non-motorized modes. In effect, a policy such as Low
+the perspective of non-motorised modes. In effect, a policy such as Low
 Emission Zones help to improve the accessibility situation of active

### 2023-12-02 Journal-manuscript/Journal-manuscript.Rmd
Small tweaks to R1 manuscript as requested by the journal office

@@
 ```{r, jobs-plot, fig.cap="\\label{fig:Fig2} Distribution of jobs taken by people living and working in Madrid as reported in the 2018 travel survey.
-Grey TAZs has no jobs.
-Ranges of values in the legend are quintiles.", fig.show='hold', fig.align='center'}
+Grey TAZs have no jobs.
+Ranges of values in the legend are quintiles.
+The TAZ shapefile is available from the Community of Madrid open data portal.", fig.show='hold', fig.align='center'}
 knitr::include_graphics("images/Fig2.png") 
@@
 
-```{r, pop-plot, fig.cap="\\label{fig:Fig3} Population living and working in Madrid by mode of transportation as reported in the 2018 travel survey.
-Grey TAZs have no population.
-Ranges of values in the legend are quintiles.", fig.show='hold', fig.align='center'}
+```{r, pop-plot, fig.cap="\\label{fig:Fig3} Population living and working in Madrid by mode of transportation as reported in the 2018 travel survey and represented at the level of TAZ.
+Grey TAZ have no population.
+Ranges of values in the legend are quintiles.
+The TAZ shapefile is available from the Community of Madrid open data portal.", fig.show='hold', fig.align='center'}
 knitr::include_graphics("images/Fig3.png")
@@
 Figures \ref{fig:Fig2} and \ref{fig:Fig3} show the number of workers and the distribution of full-time jobs in the City of Madrid by TAZ.
+The TAZ shapefiles are available from the Community of Madrid open data portal [@comunidaddemadridzoniEDM20182020].
 The light grey boundary represents the LEZ Centro in effect in 2017.
@@
 
-```{r, SA-m-plot, fig.cap="\\label{fig:Fig5} Spatial availability of jobs per origin and mode $V_i^m$ in Madrid.
-Grey TAZs have no population.
-Ranges of values in the legend are quintiles.", fig.show='hold', fig.align='center', fig.height=10}
+```{r, SA-m-plot, fig.cap="\\label{fig:Fig5} Spatial availability of jobs per origin and mode $V_i^m$ in Madrid at the level of TAZ.
+Grey TAZ have no population.
+Ranges of values in the legend are quintiles.
+The TAZ shapefile is available from the Community of Madrid open data portal.", fig.show='hold', fig.align='center', fig.height=10}
 knitr::include_graphics("images/Fig5.png")
@@
 
-```{r, SA-per-capita-m-plot, fig.cap="\\label{fig:Fig7} Distribution of spatially available jobs per capita by mode of transportatin ($v_i^m$).
-Grey TAZs have no population that use the mode.
-Ranges of values in the legend are quintiles.", fig.show='hold', fig.align='center', fig.height=10}
+```{r, SA-per-capita-m-plot, fig.cap="\\label{fig:Fig7} Distribution of spatially available jobs per capita by mode of transportatin ($v_i^m$) represented at the level of TAZ.
+Grey TAZ have no population that use the mode.
+Ranges of values in the legend are quintiles.
+The TAZ shapefile is available from the Community of Madrid open data portal.", fig.show='hold', fig.align='center', fig.height=10}
 knitr::include_graphics("images/Fig7.png")
@@
 
-# Acknowledgements
-
-This research was funded by the Canada Graduate Scholarship - Doctoral Program (CGS D) provided by the Social Sciences and Humanities Research Council (SSHRC) and Project Mobilizing Justice, also supported by SSHRC.
+<!-- # Acknowledgements -->
+
+<!-- This research was funded by the Canada Graduate Scholarship - Doctoral Program (CGS D) provided by the Social Sciences and Humanities Research Council (SSHRC) and Project Mobilizing Justice, also supported by SSHRC.
 The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.
-
-All work is fully-reproducible and available within this [GitHub repository](https://github.com/soukhova/Multimodal-spatial-availability).
+-->
+
+<!-- All work is fully-reproducible and available within this [GitHub repository](https://github.com/soukhova/Multimodal-spatial-availability). -->
 

### 2023-12-29 TRB-Poster.Rmd
final draft of TRB poster

---
title: "Multimodal spatial availability: a singly-constrained measure of accessibility considering multiple modes"
poster_height: "121cm"
poster_width: "242cm"
font_family: 'Rasa'
primary_colour: "#7A003C"
secondary_colour: "#d3d3d3"
accent_colour: "#FDBF57"
titlebox_bordercol: "#5E6A71"
titlebox_borderwidth: "0.5cm"
author:
  - name: Anastasia Soukhov
    affil: 1
    email: soukhoa@mcmaster.ca
    twitter: soukhov_a
    main: true
  - name: Javier Tarriño-Ortiz
    affil: 2
    email: javier.tarrino.ortiz@upm.es
  - name: Julio A. Soria-Lara
    affil: 2
    email: julio.soria-lara@upm.es
  - name: Antonio Páez
    email: paezha@mcmaster.ca
    affil: 1
affiliation:
  - num: 1
    address: Department of Earth, Environment and Society, McMaster University, Hamilton, Canada
  - num: 2
    address: Centro de Investigación del Transporte (TRANSyT), Universidad Politécnica de Madrid, Madrid, Spain
column_numbers: 4
logoright_name: Presentation-images/acknowledgements-combo.png
logoleft_name: Presentation-images/soukhova-LinkTree.png
sectitle_textsize: "60pt"
sectitle_borderwidth: "1.5mm"
title_textsize: "110pt"
author_textsize: "80pt"
affiliation_textsize: "80pt"
body_textsize: "53px"
link-citations: true
bibliography: bibliography/bibliography.bib
output: 
  posterdown::posterdown_html:
    self_contained: true
knit: pagedown::chrome_print  #this knits the created html file into a .pdf. comment out if .pdf isn't needed
---
# Abstract
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, eval=FALSE, include=FALSE}
# pagedown::chrome_print("TRB-Poster.Rmd", format = "pdf")
```
```{r load-packages, cache=FALSE, include=FALSE}
library(dplyr)
library(sf)
library(ggplot2)
library(tidyverse)
library(units)
library(patchwork)
library(ggspatial)
library(ggforce)
library(geomtextpath)
library(knitr)
library(gridExtra)
library(grid)
library(huxtable)

options(scipen = 999)
```
```{r sp_avail-function,include=FALSE, include=FALSE}
#defining the spatial availability function, this prints all the intermediates
sp_avail_detailed <- function(x, o_id, d_id, pop, opp, r, f, alpha = 1){

  o_id <- rlang::enquo(o_id)
  d_id <- rlang::enquo(d_id)
  pop <- rlang::enquo(pop)
  opp <- rlang::enquo(opp)
  r <- rlang::enquo(r)
  f <- rlang::enquo(f)

  sum_pop <- x %>%
    dplyr::distinct(!!o_id,
                    .keep_all = TRUE) %>%
    dplyr::mutate(sum_pop = !!r*(!!pop)^alpha) %>%
    dplyr::pull(sum_pop) %>%
    sum()

  f_p <- dplyr::pull(x, !!r) * dplyr::pull(x, !!pop)^alpha / sum_pop

  sum_impedance <- x %>%
    dplyr::group_by(!!d_id) %>%
    dplyr::summarize(sum_impedance = sum(!!f))

  x <- x %>%
    dplyr::left_join(sum_impedance,
                     by = rlang::as_name(d_id))

  f_c <- dplyr::pull(x, !!f) / x$sum_impedance

  x$f_c <- f_c
  x$f_p <- f_p

  sum_pa <- x %>%
    dplyr::group_by(!!d_id) %>%
    dplyr::summarize(sum_pa= sum(f_p * f_c))

  x <- x %>%
    dplyr::left_join(sum_pa,
                     by = rlang::as_name(d_id))
  f_t <- (f_p * f_c) / dplyr::pull(x, sum_pa)

  x %>%
    dplyr::mutate(V_ij = !!opp * !!f_t)
}
```
```{r synthetic-data, include=FALSE}
od_tt <- data.frame(i = c("A", "A", "A", "B", "B", "B", "C", "C", "C"), # Three origins
                    j = c("1", "2", "3", "1", "2", "3", "1", "2", "3"), # Three destinations
                    tt_z = c(15, 30, 100, 30, 15, 100, 100, 100, 15), # Travel time, z-mode (original)
                    tt_x = c(10, 25, 80, 25, 10, 80, 80, 80, 10), # Travel time, x-mode (the faster mode - like car)
                    pop = c(50000, 50000, 50000, 150000, 150000, 150000, 10000, 10000, 10000), # Population
                    opp = c(100000, 100000, 10000, 100000, 100000, 10000, 100000, 100000, 10000)) # Jobs

#write the dataframe long.
od_tt_m <- data.frame(i = c("A", "A", "A", "A", "A", "A",
                            "B", "B", "B", "B", "B", "B",
                            "C", "C", "C", "C", "C", "C"), # Three origins
                    j = c("1", "2", "3", "1", "2", "3", 
                          "1", "2", "3", "1", "2", "3", 
                          "1", "2", "3", "1", "2", "3"), # Three destinations
                    m = c("pt", "pt", "pt", "car", "car", "car",
                          "pt", "pt", "pt", "car", "car", "car",
                          "pt", "pt", "pt", "car", "car", "car"),
                    tt = c(15, 30, 100, 10, 25, 80, #travel time, z-mode (original) AND THEN x-mode (the faster mode)
                           30, 15, 100, 25,  10,  80,
                           100, 100, 15, 80, 80, 10), 
                    pop = c(20000, 20000, 20000, 30000, 30000, 30000,
                            50000, 50000,50000, 100000, 100000, 100000,
                            3000, 3000, 3000, 7000,7000, 7000),
                    opp = c(100000, 100000, 10000, 100000, 100000, 10000,
                            100000, 100000, 10000, 100000, 100000, 10000,
                            100000, 100000, 10000, 100000, 100000, 10000)) 
```
```{r data-figure-with-toy-example, include=FALSE}
od <- data.frame(id = c("A", "B", "C", "1", "2", "3"),
                 type = c("Population", "Population", "Population", "Jobs", "Jobs", "Jobs"),
                 size = c(50000, 150000, 10000, 100000, 100000, 10000),
                 x = c(2.5, 2.5, 8.5, 0.5, 0.5, 8.5),
                 y = c(8.0, 2.5, 4.5, 7.5, 2.5, 2.5))

centers <- data.frame(id = c("Urban center", "Suburb", "Satellite town"),
                 radius = rep(1.3, 3),
                 x = c(1.5, 1.5, 8.5),
                 y = c(2.5, 7.5, 3.5))

# od_lines coded as segments
od_lines <- data.frame(x = c(2.5, 2.5, 2.5, 2.5, 2.5, 2.5, 8.5, 8.5, 8.5),
                       y = c(8.0, 8.0, 8.0, 2.5, 2.5, 2.5, 4.5, 4.5, 4.5),
                       xend = c(0.5, 0.5, 8.5, 0.5, 0.5, 8.5, 0.5, 0.5, 8.5),
                       yend = c(7.5, 2.5, 2.5, 7.5, 2.5, 2.5, 7.5, 2.5, 2.5),
                       group = c("A", "A", "A", "B", "B", "B", "C", "C", "C"),
                       time = c("15 min (x)\n10 min (z)", "30 min (x)\n25 min (z)", "100 min (x)\n80 min (z)",
                                "30 min (x)\n25 min (z)", "15 min (x)\n10 min (z)", "100 min (x)\n80 min (z)",
                                "100 min (x)\n80 min (z)", "100 min (x)\n80 min (z)", "15 min (x)\n10 min (z)"))

od_table <- data.frame(Center = c("A", "B", "C", "1", "2", "3"),
  Size = c("50,000 pop (40% x)", "150,000 pop (33% x)", "10,000 pop (30% x)", 
           "100,000 jobs", "100,000 jobs", "10,000 jobs"))

od_table2 <- data.frame(Path = c("A to 1", "A to 2", "A to 3", 
                                "B to 1", "B to 2", "B to 3", 
                                "C to 1", "C to 2", "C to 3"),
  Value = c("15 min (x)\n10 min (z)", "30 min (x)\n25 min (z)", "100 min (x)\n80 min (z)",
            "30 min (x)\n25 min (z)", "15 min (x)\n10 min (z)", "100 min (x)\n80 min (z)",
            "100 min (x)\n80 min (z)", "100 min (x)\n80 min (z)", "15 min (x)\n10 min (z)"))
```
```{r forming the table of synth value, include=FALSE}
#set b and calculate impedance
b_x <- 0.1 #the faster mode, car
b_z <- 0.1 #the original slower mode, transit

od_tt_m <- od_tt_m |>
  mutate(f = case_when(m=="car" ~ exp(-b_x * tt),
                       m=="pt" ~ exp(-b_z * tt)))

S <- od_tt_m |>
  group_by(i,m) |>
  summarize(S = sum(opp * f),
            .groups = "drop")
S_m <- od_tt_m |>
  group_by(i) |>
  summarize(S = sum(opp * f),
            .groups = "drop")

#set b and calculate impedance
LOS_j <- od_tt_m |>
  group_by(j,m) |>
  summarize(opp = mean(opp),
            D_j = sum(pop * f),
            .groups = "drop") |>
  group_by(j) |>
  summarize(LOS_j = first(opp)/sum(D_j),
            opp = mean(opp),
            D_j = mean(D_j))

a_m <- od_tt_m %>%
  left_join(LOS_j |> 
              dplyr::select(j, LOS_j, opp, D_j),
            by = c("j")) %>%
  group_by(i,m) |>
  summarize(a_i_m = sum(LOS_j * f),
            pop = mean(pop),
            .groups = "drop") |>
  rename("a" = "a_i_m")

a <- a_m %>%
  group_by(i) |>
  summarize(a = sum(pop*a)/sum(pop),
            pop = sum(pop),
            .groups = "drop")

od_tt_m <- od_tt_m %>%
  mutate(catch = 1) %>%
  sp_avail_detailed(o_id = i, #spatial availability
                    d_id = j, 
                    pop = pop,
                    opp = opp,
                    r = catch,
                    f = f, 
                    alpha = 1)

V_m <- od_tt_m %>%
  group_by(i,m) %>%
  summarize(V = sum(V_ij),
            v = V/first(pop))

V <- od_tt_m %>%
  group_by(i) %>%
  summarize(V = sum(V_ij))
```
```{r assembing synth value table, include=FALSE}
access_table <- data.frame(i = c("A", "A", 
                                 "B", "B",
                                 "C", "C", "TOTALS"),
                       Mode = c("z", "x", "z", "x", "z", "x", "N/A"),
                       Population = c(a_m$pop[1], a_m$pop[2], a_m$pop[3], a_m$pop[4], a_m$pop[5], a_m$pop[6], a_m$pop%>%sum()),
                       Population.1 = c(a$pop[1], NA, a$pop[2], a$pop[2], a$pop[3], a$pop[3], a$pop%>%sum()),
                       S = c(S$S[1], S$S[2], S$S[3], S$S[4], S$S[5], S$S[6], S$S%>%sum()),
                       S.1 = c(S_m$S[1], S_m$S[1], S_m$S[2], S_m$S[2], S_m$S[3], S_m$S[3], S$S%>%sum()),
                       a = c(a_m$a[1], a_m$a[2], a_m$a[3], a_m$a[4], a_m$a[5], a_m$a[6], "N/A"),
                       a.1 = c(a$a[1], a$a[1], a$a[2], a$a[2], a$a[3], a$a[3], "N/A"),
                       V = c(V_m$V[1], V_m$V[2], V_m$V[3], V_m$V[4], V_m$V[5], V_m$V[6], V$V%>%sum()),
                       V.1 = c(V$V[1], V$V[1], V$V[2], V$V[2], V$V[3], V$V[3], V$V%>%sum())) 
```
```{r saving synth value table part one, include=FALSE}
ht <- hux(access_table, add_colnames = TRUE)

ht <- ht %>%
  merge_across(1, 3:4) %>%
  merge_across(1, 5:6) %>%
  merge_across(1, 7:8) %>%
  merge_across(1, 9:10) %>%
  merge_across(8, 7:8) %>%
  merge_down(2:3, c(1,4,6,8,10)) %>%
  merge_down(4:5, c(1,4,6,8,10))%>%
  merge_down(6:7, c(1,4,6,8,10)) %>%
  set_background_color(col= 1:4, row=c(6,7), value = "springgreen1") %>%
  set_background_color(col= 1:4, row=2:3, value = "springgreen3") %>%
  set_background_color(col= 1:4, row=4:5, value = "springgreen4") %>%
  set_background_color(col=5:6, value = "#FEFE62") %>%
  set_background_color(col=7:8, value = "#FFC20A") %>%
  set_background_color(col=9:10, value = "#E1BE6A") %>%
  set_background_color(row=8, value = "grey90") %>%
  set_background_color(col= 1:4, row=1, value = "grey90") %>%
  set_all_borders(value = 0.4) %>%
  set_align(value="center") %>%
  set_valign(value="middle") %>%
  set_bold(col=1:10, row=1) %>%
  set_font_size(value = 16) %>%
  set_number_format(fmt_pretty(digits=2)) %>%
  set_col_width(value=c(0.1,0.1,0.1,0.1,0.1,0.1,0.1,0.1,0.1,0.1))

quick_html(ht, file = "Presentation-images/synth_ex_table.html", open = FALSE)
# higher zoom increases resolution and image size
webshot::webshot(url = "Presentation-images/synth_ex_table.html", file = "Presentation-images/synth_ex_table.png",
                 zoom = 5, selector = "table")
```
```{r, create-figure-with-toy-example, echo=FALSE, eval=FALSE, include=FALSE}
toy_example_plot <- ggplot() + 
  # Plot centers
  geom_circle(data = centers,
             aes(x0 = x, 
                 y0 = y,
                 r = radius),
             color = "grey30",
             fill = "grey90") +
  annotate(geom = "label", 
           x=c(1.5, 1.5, 10),
           y=c(9.2, 0.8, 5.2), 
           label = c("Suburban (50,000 jobs)", "Urban (100,000 jobs)", "Satellite (10,000 jobs)"),
           fill = c("turquoise1","turquoise1","turquoise1"),
           size=2.7) +
  # Connect origins and destinations with curves and put the text on curves
  geom_textcurve(data = od_lines,
               aes(x = x,
                   y = y,
                   xend = xend,
                   yend = yend,
                   linetype = group,
                   group = group,
                   label = time),
             color = "black",
             position=position_jitter(width=0.1,height=0.1),
             size = 1.8,
             curvature = 0.20,
             linewidth = 0.5,
             linecolour = "black",
             arrow = arrow(type = "closed", angle = 10, length = unit(0.03, "npc"))) +
  # Plot origins and destinations points
  geom_point(data = od,
            aes(size = size,
                x = x, 
                y = y, 
                color = id,
                shape = type)) +
  scale_size(range = c(3, 7)) +
  scale_shape_manual(values = c("Population" = 17,'Jobs' = 16)) +
  scale_color_manual(values = c("A" = "springgreen3",
                                "B" = "springgreen4",
                                "C" = "springgreen1",
                                '1' = "turquoise1",
                                '2' = "turquoise1",
                                '3' = "turquoise1")) +
  # Label origins and destinations
  geom_text(data = od, 
            aes(x, 
                y, 
                label=id),
            size=2.5) +
  # place the table
    annotation_custom(rasterGrob(png::readPNG(source="Presentation-images/synth_ex_table.png")),
                      xmin= 5.5, xmax=Inf, 
                      ymin=6, ymax=Inf
                      )+
  # Theme the plot
  coord_fixed(xlim = c(0,20), ylim=c(0,12)) + 
  theme_void() + guides(shape = "none", size = "none", color = "none", linetype ="none")

toy_example_plot
ggsave("Presentation-images/toy-example.png",dpi=600)
```

A pitfall in conventional accessibility measures are opportunities are often multiply counted, which leads to values of accessibility that are difficult to interpret.

Our purpose: a method for measuring accessibility for multiple modes that addresses this pitfall.

We extend spatial availability [@soukhovIntroducingSpatialAvailability2023] for the case of multiple modes. Spatial availability constrains calculations to match a known quantity ensures that the measurements sum up to a predetermined quantity (i.e., the total number of opportunities), and so each value can be meaningfully related to this total.

We demonstrate its use in the case of multiple modes or, more generally, heterogeneous population segments with distinct travel behaviors. We proceed to illustrate its features using a synthetic example, an empirical example of low emission zones in Madrid, Spain, and suggestions for future research in evaluating policy interventions.

# Overview of multi-modal accessibility measures

<u>Hansen-type accessibility</u> [@hansenHowAccessibilityShapes1959] - _Non competitive_ and _unconstrained_:

$S_i^m = \sum_j O_j \color{blue}{f^m(c_{ij}^m)}$

<u>Shen-type accessibility</u> [@shenLocationCharacteristicsInnercity1998; @luoMeasuresSpatialAccessibility2003] - **Competitive** and _unconstrained_:

$a_i^m = \sum_j O_j \color{blue}{\frac{f^m(c_{ij}^m)}{\sum_m D_j^m}}  = \sum_j O_j \color{blue}{\frac{f^m(c_{ij}^m)}{{\sum_m \sum_i P_{i}^m f^m(c_{ij}^m)^m}}}$

<u>Spatial availability</u> [@soukhovIntroducingSpatialAvailability2023] - **Competitive** and **constrained**:

$V_i^m = \sum_j O_j \color{blue}{F^{t,m}_{ij}} = \sum_j O_j \color{blue}{\frac{F^{pm}_{i} \cdot F^{cm}_{ij}}{\sum_{m} \sum_{i} F^{pm}_{i} \cdot F^{cm}_{ij}}}$

Where:

-   $m = 1, 2,\cdots, M$ is a set of $M$ modes (or sub-populations) of interest.
-   $c_{ij}^m$ is a measure of the cost of moving between $i$ and $j$ for each $m$.
-   $f^m(\cdot)$ is an impedance function of $c^m_{ij}$ for each $m$, it can take the form of any monotonically decreasing function reflective of travel behaviour.
-   $i$ is a set of origin locations ($i = 1,\cdots,N$).
-   $j$ is a set of destination locations ($j = 1,\cdots,J$).
-   $O_j$ is the number of opportunities at location $j$; $O = \sum_{j=1}^J O_j$ is the total supply of opportunities in the study region.
* $F^{tm}_{ij}$ is the total balancing factor for each $m$ at each $i$; it consists of two factors. 
  + First, the factor for allocation by population for each $m$ at each $i$ is $F^{pm}_{i} = \frac{P_{i}^m}{\sum_{m}\sum_{i} P_{i}^m}$ and 
  + Then the factor for allocation by cost of travel for each $m$ at $i$ is $F_{ij}^{cm} = \frac{f^m(c_{ij}^m)}{\sum_{m} \sum_{i} f^m(c_{ij}^m)}$.

### Spatial availability's proportional allocation mechanism

- $V^m_{i}$ is always singly-constrained: $\sum_{i} V_{i} = \sum_{m}\sum_{i} V_{i}^m = \sum_{j} O_j$

- So, one can calculate spatial availability....
  - **per $i$**: ( $V_{i} = \sum_{m} V^m_{i}$ )
  - **per mode**: ( $v^{m} = \sum_{m} \sum_{i} V^m_{i}$ )
  - **per capita in $i$**: ( $v_{i} = \sum_{m} \frac{V_i^m}{P_{i^m}}$ )
  - Etc.
  - Etc.

# Synthetic example of 3 zones and 2 mode-using populations

The calculated Hansen (**S**), Shen-type (**a**) and Spatial Availability (**V**) assuming an impedance function of $f(c_{ij})^{\text{x}} = f(c_{ij})^{\text{z}} =exp(-0.1 \cdot c_{ij}^m)$ is also shown in the table.

![](Presentation-images/toy-example.png){#id .class width=100%}
When competition is not considered, **S** values for the Suburban A and Urban B are the same. Does this equivalency make sense for the differently sized A and B? Further, if constraints are not incorporated (i.e., **S** and **a**), values are hard to interpret. The regional sums of **S** and **a** are meaningless.

In considering both competition and constraints, **V** is not the same for A and B: Suburban A has more _available_ jobs than mode-using population, the Urban B and the satellite C have fewer _available_ jobs, and the sum of **V** is the total number of jobs in the region. We can interpret that the faster z population captures a higher proportion of availability than population in A, B, and C, unlike x. **Clarity in interpretation is the advantage of using spatial availability.**

# Empirical example: quantifying multimodal access in Madrid

The low emission zone (LEZ) in the Centro of the City of Madrid was established in 2017 to pursue national climate change goals. LEZs implement a form of _geographic discrimination_ as they change how people can reach opportunities by making it more costly for some forms of travel, typically cars, to circulate in predetermined zones. LEZ change the accessibility landscape of a city from the perspective of multiple modes.

We ask: **what is the spatial distribution of availability that can be accessed by different mode-using population, especially for the car-using populations within and outside of the Centro LEZ?**

![](Presentation-images/Fig2.png){#id .class width=100%}

![](Presentation-images/Fig5.png){#id .class width=95%}

- Each $V_i^m$ is a proportion of the 847,574 jobs in the City of Madrid (OD full-time work flows from the 2018 Community of Madrid travel survey [@comunidaddemadridResultadosEDM20182020]). They represent the number jobs spatially available to $m$-using workers at that $i$, relative to all the jobs in the city. 
- Note the differences in magnitudes between modes: the majority of availability is allocated to car- and transit- populations. This is to be expected as these modes represent 84.1% of the total population and reflect greater travel speeds.
- Car users outside of the M-30 broadly have greater spatial availability, while many zones inside the M-30 offer greater spatial availability to transit users. 
- Overall, the magnitude of $V_i^m$ for cyclists and pedestrians are lower than for car and transit, but the highest values of $V_i^{bike}$ and $V_i^{walk}$ are within the M-30 and zones with higher spatial availability by transit.

# Interpreting multimodal spatial availability

![](Presentation-images/Fig6.png){#id .class width=100%}

Note how the differences between these proportions change: car captures more availability than its population proportion overall (black) but this is not the case within the Centro (light grey). As such, non-car modes fair much better at capturing spatial availability within the M-30 (dark grey) and especially the Centro (light grey).

![](Presentation-images/Fig7.png){#id .class width=95%}

Do we want a city where the spatial availability of opportunities is equal for all mode users? $V_i^m$ can be divided by the mode-using population at each $i$ to yield mode-population scaled values and used as a planning benchmark. Zones that are orange should be targets for interventions; and car-using populations can be further targeted.

# Conclusions and future work

Opportunities are finite: spatial availability uses this idea as a constraint to consider competition for opportunities by the population. This consideration, through the proportional allocation factors, adds a new-found interpretation of accessibility values.

With spatial availability, the magnitude of opportunities that are available as a proportion of all the opportunities in the region is equal to $V_i$. Heterogeneous population characteristics, like difference in travel times due to mode used, can be easily incorporated, as done in this multimodal extension. The flexibility of spatial availability can be helpful in identifying zones in need of intervention and highlights the spatial competitive advantage of certain modes.

Future works looks to model policy scenarios around normative equity standards, and consideration of population and opportunities characteristics like income, travel mode used, and quality of opportunity.

# References

### 2024-01-03 TRB-Poster.Rmd
Version of poster to print

@@
 affiliation_textsize: "80pt"
-body_textsize: "53px"
+body_textsize: "50px"
 link-citations: true
@@
 
-A pitfall in conventional accessibility measures are opportunities are often multiply counted, which leads to values of accessibility that are difficult to interpret.
+Accessibility is a key concept in the analysis of land use and transportation systems.
+It is too coming of age from the perspective of planning [see _inter alia_, @handyAccessibility2020; @levinsonTransportAccessManual2020; @elGeneidyMaking2022].
+Beginning with the work of Hansen [@hansenHowAccessibilityShapes1959], accessibility measures have been widely used to evaluate the efficiency of transportation systems when combined with the distribution of opportunities in space.
 
-Our purpose: a method for measuring accessibility for multiple modes that addresses this pitfall.
+The most common form of accessibility measure is based on the gravity model, and recent research has focused on the way opportunities are counted (summed).
+In these calculations, the sums are not constrained, meaning the same opportunity can enter the sum for different origins.
+This unconstrained counting is a pitfall in conventional methods as it leads to values of accessibility that are difficult to interpret, especially when comparing results between multiple modes.
 
-We extend spatial availability [@soukhovIntroducingSpatialAvailability2023] for the case of multiple modes.
-Spatial availability constrains calculations to match a known quantity ensures that the measurements sum up to a predetermined quantity (i.e., the total number of opportunities), and so each value can be meaningfully related to this total.
+In this paper we address this pitfall by extending spatial availability for multiple modes [@soukhovIntroducingSpatialAvailability2023].
+Spatial availability is a singly-constrained competitive accessibility measure that constrains calculations to match a known quantity.
+This ensures the measurements sum up to a predetermined quantity (i.e., the total number of opportunities), and so each value can be meaningfully related to this total.
 
-We demonstrate its use in the case of multiple modes or, more generally, heterogeneous population segments with distinct travel behaviors.
-We proceed to illustrate its features using a synthetic example, an empirical example of low emission zones in Madrid, Spain, and suggestions for future research in evaluating policy interventions.
+We demonstrate spatial availability's use in the case of multiple modes or, more generally, heterogeneous population segments with distinct travel behaviours.
+We illustrate its features using a synthetic example, an empirical example of low emission zones in Madrid, Spain, and suggest future research directions in its use to evaluate policy interventions.
 
@@
 -   $m = 1, 2,\cdots, M$ is a set of $M$ modes (or sub-populations) of interest.
+-   $i$ is a set of origin locations ($i = 1,\cdots,N$) 
+-   $j$ is a set of destination locations ($j = 1,\cdots,J$).
 -   $c_{ij}^m$ is a measure of the cost of moving between $i$ and $j$ for each $m$.
 -   $f^m(\cdot)$ is an impedance function of $c^m_{ij}$ for each $m$, it can take the form of any monotonically decreasing function reflective of travel behaviour.
--   $i$ is a set of origin locations ($i = 1,\cdots,N$).
--   $j$ is a set of destination locations ($j = 1,\cdots,J$).
 -   $O_j$ is the number of opportunities at location $j$; $O = \sum_{j=1}^J O_j$ is the total supply of opportunities in the study region.
@@
 
-### Spatial availability's proportional allocation mechanism
+**Spatial availability's proportional allocation mechanism**
 
@@
 
-- So, one can calculate spatial availability....
-  - **per $i$**: ( $V_{i} = \sum_{m} V^m_{i}$ )
-  - **per mode**: ( $v^{m} = \sum_{m} \sum_{i} V^m_{i}$ )
-  - **per capita in $i$**: ( $v_{i} = \sum_{m} \frac{V_i^m}{P_{i^m}}$ )
-  - Etc.
+- So, one can calculate spatial availability per ...
+  - **$i$**: ( $V_{i} = \sum_{m} V^m_{i}$ )
+  - **mode**: ( $v^{m} = \sum_{m} \sum_{i} V^m_{i}$ )
+  - **capita in $i$**: ( $v_{i} = \sum_{m} \frac{V_i^m}{P_{i^m}}$ )
   - Etc.
@@
 
-The calculated Hansen (**S**), Shen-type (**a**) and Spatial Availability (**V**) assuming an impedance function of $f(c_{ij})^{\text{x}} = f(c_{ij})^{\text{z}} =exp(-0.1 \cdot c_{ij}^m)$ is also shown in the table.
+The calculated Hansen (**S**), Shen-type (**a**) and Spatial Availability (**V**) assuming an impedance function of $f(c_{ij})^{\text{x}} = f(c_{ij})^{\text{z}} =exp(-0.1 \cdot c_{ij}^m)$ is shown in the table.
 
@@
 We can interpret that the faster z population captures a higher proportion of availability than population in A, B, and C, unlike x.
-**Clarity in interpretation is the advantage of using spatial availability.**
+**The advantage of using spatial availability is the clarity in interpretation.**
 
@@
 
-The low emission zone (LEZ) in the Centro of the City of Madrid was established in 2017 to pursue national climate change goals.
+The low emission zone (LEZ) in the Centro of the City of Madrid was established in 2017 to pursue climate change mitigation goals.
 LEZs implement a form of _geographic discrimination_ as they change how people can reach opportunities by making it more costly for some forms of travel, typically cars, to circulate in predetermined zones.
@@
 
-Note how the differences between these proportions change: car captures more availability than its population proportion overall (black) but this is not the case within the Centro (light grey).
-As such, non-car modes fair much better at capturing spatial availability within the M-30 (dark grey) and especially the Centro (light grey).
+Note the differences between proportions between mode depending on the spatial neighbourhood examined: car captures more availability than its population proportion overall (black) but this is not the case within the Centro (light grey).
+As such, non-car modes are better at capturing spatial availability within the M-30 (dark grey) and especially the Centro (light grey).
 
@@
 $V_i^m$ can be divided by the mode-using population at each $i$ to yield mode-population scaled values and used as a planning benchmark.
-Zones that are orange should be targets for interventions; and car-using populations can be further targeted.
+Zones that are orange could be targets for interventions; and car-using populations can be further dissuaded.
 
@@
 Heterogeneous population characteristics, like difference in travel times due to mode used, can be easily incorporated, as done in this multimodal extension.
-The flexibility of spatial availability can be helpful in identifying zones in need of intervention and highlights the spatial competitive advantage of certain modes.
+The flexibility of spatial availability can be helpful in identifying zones and modes in need of intervention and highlights the spatial competitive advantage of certain modes.
 
-Future works looks to model policy scenarios around normative equity standards, and consideration of population and opportunities characteristics like income, travel mode used, and quality of opportunity.
+Future works will look to model policy scenarios considering normative equity standards, and consideration of population and opportunities characteristics like income, travel mode used, and quality of opportunity.
 

### 2024-01-25 TRB-Poster.Rmd
Same changes to TRB poster

@@
 title: "Multimodal spatial availability: a singly-constrained measure of accessibility considering multiple modes"
-poster_height: "121cm"
-poster_width: "242cm"
+poster_height: "36in"
+poster_width: "54in"
 font_family: 'Rasa'
@@
 titlebox_bordercol: "#5E6A71"
-titlebox_borderwidth: "0.5cm"
+titlebox_borderwidth: "0.2cm"
 author:
@@
 logoleft_name: Presentation-images/soukhova-LinkTree.png
-sectitle_textsize: "60pt"
-sectitle_borderwidth: "1.5mm"
-title_textsize: "110pt"
-author_textsize: "80pt"
-affiliation_textsize: "80pt"
-body_textsize: "50px"
+sectitle_textsize: "40pt"
+sectitle_borderwidth: "0.4mm"
+title_textsize: "80pt"
+author_textsize: "50pt"
+affiliation_textsize: "40pt"
+body_textsize: "35px"
 link-citations: true
@@
 ---
-# Abstract
+# Overview
 ```{r setup, include=FALSE}
@@
 Accessibility is a key concept in the analysis of land use and transportation systems.
-It is too coming of age from the perspective of planning [see _inter alia_, @handyAccessibility2020; @levinsonTransportAccessManual2020; @elGeneidyMaking2022].
+It is too coming of age from the perspective of planning [see _inter alia_, @handyAccessibility2020; @elGeneidyMaking2022].
 Beginning with the work of Hansen [@hansenHowAccessibilityShapes1959], accessibility measures have been widely used to evaluate the efficiency of transportation systems when combined with the distribution of opportunities in space.
@@
 The most common form of accessibility measure is based on the gravity model, and recent research has focused on the way opportunities are counted (summed).
-In these calculations, the sums are not constrained, meaning the same opportunity can enter the sum for different origins.
-This unconstrained counting is a pitfall in conventional methods as it leads to values of accessibility that are difficult to interpret, especially when comparing results between multiple modes.
+Notably, the sums are not constrained meaning the same opportunity can enter the sum for different origins.
+This unconstrained counting results in accessibility values that are difficult to interpret, especially when comparing results between multiple modes.
 
@@
 
-We demonstrate spatial availability's use in the case of multiple modes or, more generally, heterogeneous population segments with distinct travel behaviours.
+We demonstrate spatial availability's use in the case of multiple modes (i.e., heterogeneous population segments with distinct travel behaviours).
 We illustrate its features using a synthetic example, an empirical example of low emission zones in Madrid, Spain, and suggest future research directions in its use to evaluate policy interventions.
 
-# Overview of multi-modal accessibility measures
+# General form of multimodal accessibility measures
 
@@
   - **capita in $i$**: ( $v_{i} = \sum_{m} \frac{V_i^m}{P_{i^m}}$ )
-  - Etc.
 
@@
 
-The calculated Hansen (**S**), Shen-type (**a**) and Spatial Availability (**V**) assuming an impedance function of $f(c_{ij})^{\text{x}} = f(c_{ij})^{\text{z}} =exp(-0.1 \cdot c_{ij}^m)$ is shown in the table.
+![_Figure 1: Calculated Hansen $S$, Shen-type $a$ and Spatial Availability $V$ assuming an impedance function of $f(c_{ij})^{\text{x}} = f(c_{ij})^{\text{z}} =exp(-0.1 \cdot c_{ij}^m)$ for the multimodal synthetic population._](Presentation-images/toy-example.png){#id .class width=100%}
 
-![](Presentation-images/toy-example.png){#id .class width=100%}
 When competition is not considered, **S** values for the Suburban A and Urban B are the same.
@@
 
-# Empirical example: quantifying multimodal access in Madrid
+# Empirical example: multimodal access in Madrid
 
@@
 
-![](Presentation-images/Fig2.png){#id .class width=100%}
+![_Figure 2: Spatial distribution of jobs taken by people living and working in Madrid as reported in the 2018 travel survey (left) and trip length distribution (right).
+Grey TAZs have no jobs.
+Ranges of values in the legend are quintiles.
+The TAZ shapefile is available from the Community of Madrid open data portal._](Presentation-images/Fig2.png){#id .class width=100%}
 
-![](Presentation-images/Fig5.png){#id .class width=95%}
+![_Figure 3: Spatial availability of jobs per origin and mode $V_i^m$._](Presentation-images/Fig5.png){#id .class width=98%}
 
@@
 
-![](Presentation-images/Fig6.png){#id .class width=100%}
+![_Figure 4: Proportion of population by mode and spatial availability of jobs by mode ($V_i^m$) aggregated for the city of Madrid (All), the area within the M-30 highway, and within the Centro._](Presentation-images/Fig6.png){#id .class width=100%}
 
@@
 
-![](Presentation-images/Fig7.png){#id .class width=95%}
+![_Figure 5: Distribution of spatially available jobs per capita by mode of transportatin ($v_i^m$)._](Presentation-images/Fig7.png){#id .class width=98%}
 

### 2024-01-25 Journal-manuscript/Editorial correspondance/Response-to-reviewers_R2.Rmd
R2 Journal-manuscript revision. For AP review.

---
title: "PLOS ONE submission: Response to reviewers"
output: pdf_document
urlcolor: blue
date: "2024-01-26"
bibliography:
  - "`r system('kpsewhich ../../bibliography/bibliography.bib', intern=TRUE)`"
---

Thank you for for the time and care taken by the editorial team and reviewers in providing feedback to the second round of revision to the manuscript. No doubt, the further revised version submitted is more articulate than the first. Our responses to comments to Reviewer #2 are shown below in \textcolor{blue}{blue}.

## Reviewer #2:

I appreciate the fact that the authors have largely revised the paper to answer the reviewers’ comments. Still, I feel like some of my comments have been insufficiently addressed in the revised draft, as detailed below.

\textcolor{blue}{Thank you: comments provided by all reviewers were exceptionally helpful in clarifying the manuscript. We address the outstanding three comments and additional minor comments in this response letter.}

The authors have adequately addressed my comment on the fact that their new measure is not useful for studying access to non-competitive resources in the revised introduction.

\textcolor{blue}{We added additional detail to the "Introduction" (second paragraph) to further clarify. Specifically, that opportunity congestion can be seen on continnum and that accessibility methods that consider competition have been increasingly applied to analyse access to opportunity types that are not inexhaustible such as educational facilities [@kelobonyeMeasuringAccessibilitySpatial2020a; @liMeasuringMultiactivitiesAccessibility2024a], greenspace and recreation [@wuWillOpeningCommunity2020; @liMeasuringMultiactivitiesAccessibility2024a], and shopping [@kelobonyeMeasuringAccessibilitySpatial2020a; @chenEnhancingTwoStepFloating2019].}

My comment on the fact that the new measure doesn’t allow to study the absolute gains or losses in accessibility from public transport or land use system changes has inadequately been accounted for. First, what your new measure can or cannot do remains unclear to me: for instance, how would an improvement in the transport sector that benefits all transport modes and locations uniformly, but still reduces all transportation times in the city, be accounted for? Second, I do not see where this point is discussed in the paper.

\textcolor{blue}{Interesting and thoughtful comment. We have added further detail to the text to the "Discussion and Conclusions" (final paragraph) to further shed like on 'competition' and make spatial availability's use case more clear in the closure of the manuscript. 
}
\textcolor{blue}{In the scenario described, a multiple mode (equal travel impedance) in a uniformly distributed population/opportunity landscape, will yield spatial availability before- and after- values that are equal. This is because competition has not changed: spatial availability allocates 'opportunities' in a proportional way based on competition. Competition is a product of the relative travel time (between zones) as well as the supply and demand of population and opportunities, so if those relative differences remained unchanged between interventions then spatial availability remains unchanged as well. Though what would decrease are the system-wide transport costs, and those can be reflected without an accessibility measure. Hopefully the addition of the last paragraph in the discussion clarifies this point about competition considerations within accessibility measures.
}

My comment on modal shift has partially been accounted for. A key question is whether people can shift transport modes easily or not: indeed, it seems to be that your new accessibility measure, based on competition, does only make sense in case people cannot easily shift transport modes. You partially answer that question in the paragraph starting l46 by explaining why some populations, e.g. children, elderly, or single parents, might have different characteristics in terms of transportation choice and mode use. Still, in the case of the Madrid LEZ, you do not explain why private cars and public transport users are in competition. For instance, you state that “restrictions to travel by car leave more spatially available opportunities for non-car-users” (l585). But it seems to me that the LEZ aims at promoting a modal shift toward less polluting cars, public transport, or active modes rather than favoring public transport users? Relatedly, your measure does not make any difference between car users that can easily shift to public transport, and car users that cannot.

\textcolor{blue}{Thank you for your comment, we've further amended the "Results" (first and second paragraph) to make the assumptions within and interpretation of the empirical example more clear. The fourth paragraph in the "Discussion and Conclusions" also touches on your comment.
}
\textcolor{blue}{There is no mode-shifting that occurs in the empirical example. This implementation of spatial availability assumes that all employment opportunities are available to the entire population. What impacts the proportional allocation is the relative travel cost (based on observed modal origin-destination travel behaviour from the travel survey) and the relative amount of the mode-using population residing in the zones. Fig 5 demonstrates the zonal multimodal spatial availability values based on the selected mode, observed home-to-work travel patterns, and competition to work. It is only a snapshot of travel behaviour as informed by the 2018 travel survey (i.e., the majority of cars do not enter the LEZ centro), that we use to demonstrate the multimodal spatial availability landscape. In this context, we discuss the impact that the travel behaviour induced by the LEZ centro has on the spatial availability landscape.
}
\textcolor{blue}{As written in "Discussion and Conclusions", accessibility indicators are not designed to work as modal split models. "In our empirical example we used data collected after the introduction of LEZ Centro. However, given a modal split model to predict model shares, accessibility indicators, including spatial availability, can be used to investigate changes to the accessibility landscape. A similar logic can be applied for destination choice."
}
\textcolor{blue}{I hope this is a comprehensive response (and manuscript amended) to address your comment. The empirical example in the manuscript does not deal with the change from one mode to another, it only looks at the snapshot of travel behaviour as recorded in the 2018 travel survey. Investigating mode-shift would be a different research motivation.
}

My other comments have been accounted for, and the writing of the paper has largely been improved.

Finally, I have a few additional minor comments:

Fig 5: the boundaries of the LEZs, and in particular of the LEZ centro, are really hard to see.

There is a typo l56: “The paper rest of the paper”

\textcolor{blue}{Thank you bringing these comments to our attention. We have changed the colours and slightly increased the size of the spatial borders for LEZ Centro and M-30 on all figures (now the borders are Pink and Purple instead of light grey and medium grey) to improve visibility. We have also corrected the identified typo. Additionally, we have improved phrasing and corrected minor grammatical errors throughout the manuscript. }

\textcolor{blue}{Again, thank you for your thoughtful comments through the revision process so far.}

### 2024-01-25 Journal-manuscript/Journal-manuscript.Rmd
R2 Journal-manuscript revision. For AP review.

@@
 borders <- rbind(LEZ_centro %>% transmute(Border = "LEZ centro"), 
-                 LEZ_M30 %>% transmute(Border = "LEZ M-30"),
+                 LEZ_M30 %>% transmute(Border = "M-30"),
                  Madrid_muni_border)
@@
           fill=NA) +
-  scale_color_manual(values = c("grey90", "grey60", "black"),
+  scale_color_manual(values = c("pink", "purple", "black"),
                        name = element_blank())+
@@
 ```{r}
-#Plot V (spatial availability) by origin-mode (then zn208). 
+#Plot V (spatial availability) by origin-mode
 SA_im_V_zn208_plot <- ggplot() +
@@
           aes(colour=Border),
-          size = 0.5,
+          size = 2,
           fill=NA) +
-  scale_color_manual(values = c("grey90", "grey60", "black"),
+  scale_color_manual(values = c("pink", "purple", "black"),
                        name = element_blank())+
@@
           fill= NA) +
-  scale_color_manual(values = c("grey90", "grey60", "black"),
+  scale_color_manual(values = c("pink", "purple", "black"),
                      name = element_blank())+
@@
           fill= NA) +
-  scale_color_manual(values = c("grey90", "grey60", "black"),
+  scale_color_manual(values = c("pink", "purple", "black"),
                      name = element_blank())+
-  geom_sf(data = borders %>% filter(Border == "LEZ M-30"),
+  geom_sf(data = borders %>% filter(Border == "M-30"),
           size = 0.5,
           col ="black",
-          fill="grey60") +
+          fill="purple") +
   theme_void() +
@@
           col ="black",
-          fill="grey90") +
+          fill="pink") +
   theme_void() +
@@
 
-Accessibility is a key concept in the analysis of land use and transportation systems [e.g., @hansenHowAccessibilityShapes1959; @geursAccessibility2004; @paezPositive2012], and too is coming of age from the perspective of planning [see _inter alia_, @handyAccessibility2020; @levinsonTransportAccessManual2020; @siddiqToolsTradeAssessing2021; @yanAccessibilityBasedPlanningAddressing2021; @elGeneidyMaking2022].
-Beginning with the work of Hansen [@hansenHowAccessibilityShapes1959], accessibility measures have been widely used to evaluate the efficiency of transportation systems when combined with the distribution of opportunities in space [@shiLiterature2020].
-As such, it is a holistic measure of spatial systems that measures the ease of reaching destinations [@handyMeasuring1997; @kwanSpace1998].
+Accessibility is a key concept in the analysis of land use and transportation systems [e.g., @hansenHowAccessibilityShapes1959; @geursAccessibility2004; @paezPositive2012] and is coming of age from the perspective of planning research [see _inter alia_, @handyAccessibility2020; @levinsonTransportAccessManual2020; @siddiqToolsTradeAssessing2021; @yanAccessibilityBasedPlanningAddressing2021; @elGeneidyMaking2022].
+Beginning with the work of Hansen [@hansenHowAccessibilityShapes1959], accessibility measures have been widely used to evaluate the efficiency of transportation systems when combined with the distribution of populations and opportunities in space [@shiLiterature2020].
+In this way, accessibility is a holistic measure of spatial systems' ease of reaching population-relevant destinations [@handyMeasuring1997; @kwanSpace1998].
 
@@
 
-In practice, the most common form of accessibility measure is based on the gravity model.
-These measures are sums of weighted opportunities around a focal point (i.e., a potential origin), based on how expensive it is to reach them.
-Recent research in accessibility analysis has paid attention to the way opportunities are counted in the pertinent calculations.
-Conventionally, the sums are not constrained, which means that the same opportunity can enter the sum for different origins.
-Counting the same opportunity multiple times treats it as if it was inexhaustible.
-But opportunities in general are not inexhaustible, and in fact some of them are by definition exclusive: for example, once a job is taken up by someone in the population, the same job is no longer available for any other person to take.
-More generally, opportunities are subject to congestion: for example, multiple people can obtain services from the same family doctor, but the more people who do, the more congested the service will be.
-
-The issue of congestion in accessibility measures was the motivation for the development of floating catchment area approaches [@shenLocationCharacteristicsInnercity1998; @luoMeasuresSpatialAccessibility2003].
+The most common form of accessibility measure is based the gravity model [e.g.
+@hansenHowAccessibilityShapes1959]: these measures sum "weighted" opportunities around a focal point (i.e., a potential origin), based on how expensive it is to reach them.
+Recent research in accessibility analysis has explored the implications of methods by which opportunities are summed, especially in the context of competitive accessibility [e.g., @paezDemand2019, @soukhovIntroducingSpatialAvailability2023; @wang2SFCAI2SFCAIntegration2021; @merlinDoesCompetitionMatter2017].
+In a typical Hansen-type accessibility measure, the sums around origins are not constrained.
+This means that the same opportunity can enter the sum for different origins; Counting the same opportunity multiple times treats the opportunity as if it was inexhaustible (or non-rival [@linUsingTransportationProblem2023]).
+However, this conceptualisation may not reflect reality as opportunities, in general, are not inexhaustible.
+Some opportunities are by definition exclusive: a typical example is employment, once a job is taken up by someone in the population, the same job is no longer available for any other person to take [@vanweeAccessibilityMeasuresCompetition2001; @merlinDoesCompetitionMatter2017].
+More generally, it could be conceptualised that all opportunities are subject to some amount of congestion or capacity-constraint: multiple people use a specific opportunity at a given time and the more people who do, the more congested the opportunity.
+Congestion can be seen on a continuum, and debate on how congestion within accessibility should be considered is ongoing [@vanweeAccessibilityMeasuresCompetition2001; @merlinDoesCompetitionMatter2017; @kelobonyeMeasuringAccessibilitySpatial2020a].
+Nonetheless, competition has been widely considered for traditionally exclusive opportunities types such as employment and healthcare opportunities [@cuiSpatialAccessPublic2020; @maoMeasuringSpatialAccessibility2013; @josephMeasuringPotentialPhysical1982; @higgsModellingSpatialAccess2017; @vanweeAccessibilityMeasuresCompetition2001; @merlinDoesCompetitionMatter2017; @weibullAxiomaticApproachMeasurement1976; @josephMeasuringPotentialPhysical1982; @shenLocationCharacteristicsInnercity1998] and more recently educational facilities [@kelobonyeMeasuringAccessibilitySpatial2020a; @liMeasuringMultiactivitiesAccessibility2024a], and less exclusive amenities-types such as greenspace and recreation [@wuWillOpeningCommunity2020; @liMeasuringMultiactivitiesAccessibility2024a], and shopping [@kelobonyeMeasuringAccessibilitySpatial2020a; @chenEnhancingTwoStepFloating2019].
+
+The consideration of congestion in accessibility measures was the motivation for the development of approaches that consider competition [@weibullAxiomaticApproachMeasurement1976; @josephMeasuringPotentialPhysical1982; @shenLocationCharacteristicsInnercity1998] and are popularly applied as floating catchment area methods [e.g.
+@luoMeasuresSpatialAccessibility2003].
 While these approaches purport to account for congestion, Paez et al.
-[@paezDemand2019] demonstrate that in general they do not solve the issue of multiple counting of opportunities, thus leading to biases in the calculation of total demand and supply, sometimes inflating them, other times deflating them.
+[@paezDemand2019] demonstrates that they do not solve the issue of multiple counting of opportunities in general, thus leading to biases in the calculation of total demand and supply.
+They sometimes inflate counts and other times deflate them.
 In response to this, recent research has paid closer attention to the way opportunities are counted in accessibility analysis.
 Paez et al.
-@paezDemand2019, for example, tackle floating catchment area methods and introduce a normalization of the impedance matrix to allocate the population and then the level of service proportionally.
+@paezDemand2019, for example, focuses on floating catchment area methods and introduces a normalization of the impedance matrix to allocate the population and then the level of service proportionally.
 More recently, Soukhov et al.
-[@soukhovIntroducingSpatialAvailability2023] introduced a singly-constrained measure of accessibility, called spatial availability, that employs a similar, but more sophisticated proportional allocation mechanism.
-The work of these authors show that floating catchment area methods can be seen as singly-constrained accessibility measures, and improve on existing approaches by guaranteeing that each opportunity is counted only once - in other words, treating opportunities as _finite_.
-The proportional allocation of spatial availability constrains the calculations to match a known quantity, therefore ensuring that the measurements sum up to a predetermined quantity (i.e., the total number of opportunities), and so each value can be meaningfully related to this total.
+[@soukhovIntroducingSpatialAvailability2023] introduced a singly-constrained measure of accessibility, called spatial availability, that employs a similar but more sophisticated proportional allocation mechanism.
+The work of these authors show that floating catchment area methods can be seen as singly-constrained accessibility measures and improve on existing approaches by guaranteeing that each opportunity is counted only once.
+In other words, spatial availability treats the opportunities as _finite_.
+The proportional allocation mechanism of spatial availability constrains the calculations to match a known quantity, therefore ensuring that the measurements sum up to a predetermined quantity (i.e., the total number of opportunities), and so each value can be meaningfully related to this total.
 
@@
 
-The objective of this paper is to address this limitation of spatial availability.
-Our primary motivation is to extend spatial availability for the case of multimodal accessibility, but it is worthwhile noting that this is in fact just one case of heterogeneous populations (i.e., travel by different modes).
-The method itself can easily accommodate other forms of heterogeneity, for example variations in travel behavior between older and younger adults [e.g., @paezHealth2010], the propensity of older adults to use different modes of transportation [e.g., @moniruzzamanMode2013], the usually shorter trip lengths of children compared to grown-ups [e.g., @reyesWalking2014], or the more limited travel ranges of single parents [e.g., @paezJobs2013].
-
-The paper rest of this paper is organized as follows.
+This paper's primary motivation is to extend spatial availability for the case of multimodal accessibility (i.e., travel by different modes).
+It is worth noting though that the consideration for travel costs by different modes is in fact just one case of heterogeneous populations.
+The method itself can easily accommodate other forms of heterogeneity, for example: variations in travel behavior between older and younger adults [e.g., @paezHealth2010], the propensity of older adults to use different modes of transportation [e.g., @moniruzzamanMode2013], the usually shorter trip lengths of children compared to adults [e.g., @reyesWalking2014], or the more limited travel ranges of single parents [e.g., @paezJobs2013].
+
+The rest of this paper is organized as follows.
 In Section 2 we provide a brief review of multimodal accessibility.
@@
 In Section 4 we illustrate relevant issues through a synthetic example.
-This is followed in Section 5 by an empirical example using data from the city of Madrid after the implementation of its Low Emission Zones (LEZ).
-Data for this example comes from the city's 2018 travel survey.
-The example shows the differences in spatial availability within and outside the LEZ for travelers using different modes, namely car, transit, cycling and walking.
+This is followed in Section 5 by an empirical example using home-to-work data from the city of Madrid, Spain after the implementation of its Low Emission Zones (LEZ).
+Data for this example is sourced from the city's 2018 travel survey.
+The empirical example demonstrates the differences in multimodal spatial availability highlighting differences within and outside the LEZ for travelers using different modes, namely car, transit, cycling and walking.
 In Section 6, we provide concluding remarks on the strengths of the use of spatial availability as a multimodal accessibility measure, and discuss potential future uses in policy planning scenarios as well as directions for future research.
@@
 Location-based accessibility indicators are quantitative measures of _potential_ interaction with opportunities for locations within a given region: they are summary measures of the relationship between land-use and transport systems.
-Arguably, the most commonly used are measures based on the gravity model [@wuUnifyingAccess2020], of which cumulative opportunity measures and weighted cumulative opportunity measures are particular forms [@levinsonTransportAccessManual2020].
-These measures assign a weight to opportunities based on how easy it is to reach them.
-Given an origin ($i$) and a destination ($j$), an impedance function $f^{m}(c^m_{ij})$ converts the cost of travel (e.g., time, money, generalized cost) into a score that represents the propensity for interaction.
-These measures originate from that proposed by Hansen @hansenHowAccessibilityShapes1959, which can take the following form in the multimodal case: $S_i^m = \sum_j O_j f^m(c_{ij}^m)$ where $m$ is a set of modes which have mode-specific travel costs ($c_{ij}^m$) and/or travel impedance functions ($f^m(\cdot)$).
+Arguably, the most commonly used are measures based on the gravity model [@wuUnifyingAccess2020]; this includes cumulative opportunity measures which are a special case of the gravity model [@levinsonTransportAccessManual2020; @handyMeasuring1997].
+These gravity model based measures weight opportunities depending on how easy it is to reach them.
+Given an origin ($i$) and a destination ($j$), an impedance function $f^{m}(c^m_{ij})$ converts the cost of travel (e.g., time, money, generalized cost) into a score that represents the propensity for potential interaction.
+These measures originate from the work proposed by Hansen @hansenHowAccessibilityShapes1959, which can take the following form in the multimodal case: $S_i^m = \sum_j O_j f^m(c_{ij}^m)$.
+In this form, $m$ is a set of modes that have mode-specific travel costs ($c_{ij}^m$) and/or travel impedance functions ($f^m(\cdot)$).
 
@@
 5; @paezJobs2013, Figs.
-6, 7, and 8], and they are useful as relative assessment of when some members of the public are better or worse off than others, but they do not say much about how bad is "worse off".
-
-Besides ratios of accessibility, another way to improve interpretability of scores sometimes seen in the literature is to standardize them to lie in the range [0-1].
+6, 7, and 8], and they are useful as relative assessment of when some members of the public are better or worse off than others, but they are silent on how bad is "worse off".
+
+Besides ratios of accessibility, another way seen in the literature to improve interpretability of scores is to standardise them within a [0-1] range.
 This adjustment is only helpful insofar as it facilitates relative comparisons, but interpretation of the scores remains challenging because the values are specific to a region and convey no meaning about the magnitude of the scores.
@@
 These questions lack straightforward answers since certain values will always be relatively 'low' or 'high', but do not track to a quantity that can be intuitively understood.
-Presentation or discussion of Hansen-type accessibility that has been standardized in this way is not uncommon in the literature [e.g., @campbell2019accessibility; @maharjanSpatialEquityModal2022].
+Presentation or discussion of Hansen-type accessibility that has been standardised in this way is not uncommon in the literature [e.g., @campbell2019accessibility; @maharjanSpatialEquityModal2022].
 
 Once we understand opportunities to be finite, it is possible for an accessibility measure to take on a crisper meaning.
-As considered in the long tradition of accessibility research, capacity of opportunities is limited and thus is subject to competition by population [@soukhovIntroducingSpatialAvailability2023; @grengsJobAccessibilityModal2010; @kawabataJobAccessibilityIndicator2006; @kwokUseModalAccessibility2004; @morrisAccessibilityIndicatorsTransport1979; @weibullAxiomaticApproachMeasurement1976; @shenLocationCharacteristicsInnercity1998].
+As considered in the long tradition of accessibility research, capacity of opportunities is limited and thus is subject to competition by population [@soukhovIntroducingSpatialAvailability2023; @grengsJobAccessibilityModal2010; @kawabataJobAccessibilityIndicator2006; @kwokUseModalAccessibility2004; @morrisAccessibilityIndicatorsTransport1979; @weibullAxiomaticApproachMeasurement1976; @shenLocationCharacteristicsInnercity1998; @cuiSpatialAccessPublic2020; @maoMeasuringSpatialAccessibility2013; @josephMeasuringPotentialPhysical1982; @higgsModellingSpatialAccess2017; @vanweeAccessibilityMeasuresCompetition2001; @merlinDoesCompetitionMatter2017;@kelobonyeMeasuringAccessibilitySpatial2020a; @liMeasuringMultiactivitiesAccessibility2024a].
 There are only so many school-seats, hospital capacity, employment opportunities, etc., in a region and if one person reaches an opportunity, it is taken: the supply of an opportunity and the demand for that opportunity are two components of accessibility.
@@
 For instance, standards for providing green spaces are often stated in the form of _exclusive access_, in units of amenity per capita.
-A case in point is a the Ile-de-France region, a jurisdiction that suggested in a 2013 planning document that at the municipal level at least $10m^2$ of public green space should be supplied _per inhabitant_ [@liottaPlanning2020].
+A case in point is a the Ile-de-France region, a jurisdiction that suggested in a 2013 planning document that at the municipal level at least 10 $m^2$ of public green space should be supplied _per inhabitant_ [@liottaPlanning2020].
 Green spaces are not evenly distributed, which means that who has access to them hinges on where they are and how easy is to reach them.
@@
 For example, Natural England, an organization that recommends an Accessible Natural Greenspace Standard such that the minimum supply of space is one ha of statutory Local Nature Reserves per thousand population^[see  https://redfrogforum.org/wp-content/uploads/2019/11/67-Nature-Nearby%E2%80%99-Accessible-Natural-Greenspace-Guidance.pdf].
-Similarly, the World Health Organization [cited in @OECDFrameworks2013] recommends that cities provide a minimum of 9$m^2$ of green area per inhabitant.
+Similarly, the World Health Organization [cited in @OECDFrameworks2013] recommends that cities provide a minimum of 9 $m^2$ of green area per inhabitant.
 For our purposes, standards of this type translate into "how much of this resource is available to one individual that has not been claimed by anyone else?".
-Green spaces often have large capacities, but they still have a capacity, and it is not the same for a person to have access to $5m^2$ of _uncongested_ green space as $15m^2$.
+Green spaces often have large capacities, but they still have a capacity, and it is not the same for a person to have access to 5 $m^2$ of _uncongested_ green space as 15 $m^2$.
 This difference is in fact a matter of justice [@laraSpace2015; @liottaPlanning2020].
@@
 If we return to Oslo for a moment [@lunkeModalAccessibilityDisparities2022], we notice that the places that have high accessibility by transit are also the places that have _very high_ accessibility by car (in their Figure 2).
-Those two populations are going for the same opportunities, and those travelling by transit have fewer to choose from to begin with.
+Those two populations are going for the same opportunities, and those travelling by transit have fewer to choose from the start.
 More generally, people in a zone who are advantaged with relatively low cost of travel will have the ability to potentially reach more opportunities than other people.
@@
 That said, the measure remains unconstrained, meaning both population _and_ opportunities are multiply counted [see @paezDemand2019].
-In other words, interpretation of the Shen-type accessibility scores between modes is fraught as it is for Hansen-type measures.
-
-To illustrate, @taoInvestigatingImpactsPublic2020a calculate $a_i^m$ to jobs for different income-group populations in Shenzhen, China using $m = \text{public transit}$ and $m=\text{car}$.
+In other words, interpretation of the Shen-type accessibility scores between modes is fraught, as it is for Hansen-type measures.
+
+To illustrate, @taoInvestigatingImpactsPublic2020a calculates $a_i^m$ to jobs for different income-group populations in Shenzhen, China using $m = \text{public transit}$ and $m=\text{car}$.
 Their results indicate that zones with low-income populations have lower $a_i^m$ than zones with higher-income populations.
@@
 
-Spatial availability improves on previously discussed accessibility approaches using the Hansen-type measure and the Shen-type measure by constraining the sum of opportunities, that is, by treating opportunities as finite.
+Spatial availability improves on the discussed Hansen-type and Shen-type accessibility approaches by constraining the sum of opportunities, that is, by treating opportunities as finite.
 This is done by means of proportional allocation factors that follow well established principles of spatial interaction and the gravity model [see @wilsonFamilty1971].
@@
 
-Compared to Hansen-type accessibility:
-```{=tex}
-\begin{equation}
-\label{eq:hansen-accessibility}
-A_i = \sum_{j=1}^J O_jf(c_{ij})
-\end{equation}
-```
-
-\noindent we see that spatial availability is, like the Hansen-type measure, a weighted sum of the opportunities.
-What makes spatial availability stand apart from other approaches is how the weight used in the sum, balancing factor $F^t_{ij}$, implements a proportional allocation mechanisms to ensure that the sum of $V_i$ is constrained to match the total number of opportunities in the region.
-As such, spatial availability is singly-constrained and naturally implements competition or congestion.
+Compared to Hansen-type accessibility $S_i = \sum_{j=1}^J O_jf(c_{ij})$, we see that spatial availability is, like the Hansen-type measure, a weighted sum of the opportunities.
+What makes spatial availability stand apart from other approaches is how the weight used in the sum (balancing factor $F^t_{ij}$) implements a proportional allocation mechanisms to ensure that the sum of $V_i$ is constrained to match the total number of opportunities in the region.
+As such, spatial availability is singly-constrained and natively considers competition.
 $F^t_{ij}$ consists of two parts.
@@
 The focus of this paper is to extend $V_i$ for the measurement of multimodal applications (or more generally heterogeneous populations).
-To do so, the balancing factors need to be reformulated so that 1) the mass effect now accounts not only for the size of the population at $i$, but also the size of sub-populations within $i$; and 2) the cost of travel is not only for different zones, but by sub-populations within each zone (e.g., the cost of travel from $i$ by car, transit, walking, etc.) When we introduce modes (or sub-populations) $m$, the proportional allocation factors need to satisfy the condition that $F^{pm}_{i}$ and $F^{cm}_{ij}$ can be summed across each $m$ at each $i$ and then across all $i$ to equal to 1.
+To do so, the balancing factors are reformulated so that 1) the mass effect now accounts not only for the size of the population at $i$, but also the size of sub-populations within $i$; and 2) the cost of travel is not only for different zones, but by sub-populations within each zone (e.g., the cost of travel from $i$ by car, transit, walking, etc.) relative to all zones.
+When we introduce modes (or sub-populations) $m$, the proportional allocation factors need to satisfy the condition that $F^{pm}_{i}$ and $F^{cm}_{ij}$ can be summed across each $m$ at each $i$ and then across all $i$ to equal to 1.
 They are also similarly combined multiplicatively to obtain their joint effect, represented as the combined balancing factor $F^{tm}_{ij}$ similar to that detailed in Equation (\ref{eq:multimodal-balancing-factors}).
@@
 
-Next we use a synthetic example to contrast multimodal accessibility and spatial availability.
+Next we use a synthetic example to contrast multimodal spatial availability with multimodal versions of Hansen-type accessibility (unconstrained) and Shen-type (unconstrained and competitive) accessibility.
 
@@
 This appears to be the sub-population at $A$ using mode $z$.
-Sub-populations located in opposite conditions (i.e., more distant from jobs, close to large populations, and using slow mode $x$) are at a relative disadvantage.
+Sub-populations located in opposite conditions (i.e., more distant from jobs, close to large populations, and using slower mode $x$) are at a relative disadvantage.
 The competition for opportunities between different mode-using populations matters as it reflects how well the land-use and transport system serves (or does not serve) certain populations.
@@
 Notably, the population who uses $x$ have `r (((V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% filter(m == "x") %>% dplyr::select(pop) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% dplyr::select(pop) %>% sum())) - ((V_ijm %>% filter(m=="x") %>% select(V_ij) %>% sum()) / (V_ijm %>% select(V_ij) %>% sum()))) %>% percent()` fewer spatially available opportunities than its share in the population.
-This realization leads us to ask normative questions such as, how unequal should availability of opportunities be by mode?
+This realization could lead us to ask normative questions: how unequal should availability of opportunities be by mode?
 What intervention could help to redistribute spatial availability to sub-populations commensurate with their proportion of the total?
@@
 If, let us say, the planning goal was to have one spatially available job per mode-using population, a policy intervention could be devised, to reduce the values of $v_i^z$ (making it slower or more expensive) and increase he values of $v_i^x$ (making it faster or les expensive).
-The purpose of this simple demonstration is to show how spatial availability can be used to quantify the competitive (dis)advantage in a multimodal application.
+The purpose of this synthetic demonstration is to show how spatial availability can be used to quantify the competitive (dis)advantage in a multimodal application.
 In what follows, we demonstrate the use of multimodal spatial availability through an empirical example.
@@
 The context for the empirical example is Madrid, Spain.
-This city implemented a Low Emission Zone (LEZ) in 2017 to: pursue goals set out in the national climate change agenda, cut nitrogen dioxide levels, and to prioritize people's movement in the city.
+This city implemented a Low Emission Zone (LEZ) in 2017 to pursue goals set out in the national climate change agenda, cut nitrogen dioxide levels, and to prioritize people's movement in the city.
 LEZs elsewhere have similarly been implemented as interventions to reduce GHG emissions, improve air quality, and support sustainable mobility [@margaryanLowEmissionZones2021; @verbeekJustManagementUrban2022].
@@
 
-For this case study, we use spatial availability to quantify access to opportunities by different modes in Madrid.
+For this case study, we quantify spatial availability to opportunities by different modes in Madrid.
 Particularly, we demonstrate how $V_i^m$ can be used to derive insights into how the restriction of car mobility in areas around/within the LEZ Centro may have allowed more sustainable (but often slower or more costly modes) to become more competitive.
@@
 
-The source of data for our empirical example is the 2018 Travel Survey of the Community of Madrid [@comunidaddemadridResultadosEDM20182020].
+The source of origin-destination data for our empirical example is the 2018 Travel Survey of the Community of Madrid [@comunidaddemadridResultadosEDM20182020].
 This is a representative survey that offers a snap-shot of travel patterns for a typical weekday in 2018.
@@
 
-In this example, we use all home-to-full-time-work trips, by all modes.
+In this example, we use all direct home-to-full-time-work trips, by all modes.
 The trips are expanded using population weights.
@@
 The TAZ shapefiles are available from the Community of Madrid open data portal [@comunidaddemadridzoniEDM20182020].
-The light grey boundary represents the LEZ Centro in effect in 2017.
-The survey was conducted after the introduction of LEZ Centro.
-The dark grey boundary represents the LEZ planned for the boundaries of the M-30 highway and is present in the plots as a spatial reference for areas in proximity to the LEZ Centro.
+The pink boundary represents the LEZ Centro in effect in 2017 and reflected in the 2018 travel survey.
+The purple boundary represents the LEZ planned for the boundaries of the M-30 highway and is present in the plots as a spatial reference for areas in proximity to the LEZ Centro.
 
 The total sum of jobs $O_j$ are shown in Figure \ref{fig:Fig2} and the populations that go to a work destination by four modal categories $P^m_i$, is displayed in Figure \ref{fig:Fig3}.
-The modal shares in Figure \ref{fig:Fig3} are calculated based on those measured in the survey.
-The modal categories and the mode types within each category are reported as follows:
+The modal shares in Figure \ref{fig:Fig3} are calculated based on the primary mode specified in the survey and summarized into four categories as follows:
 
@@
 
-At this point, it is worthwhile reiterating that the empirical example is a snap-shot of spatial availability by mode using data from the 2018 travel survey.
-Our purpose in this empirical example is to investigate the trends in availability of employment opportunities by mode, and illustrate how spatial availability can be used in discussions about the competitive advantage of various modes within Madrid Centro.
-The spatial availability of jobs $V_i^m$ is calculated for each of the four modes $m$ at the level of traffic analysis zones $i$ in Madrid and displayed in Figure \ref{fig:Fig5}.
+It is worthwhile reiterating that the empirical example is a snapshot of spatial availability by mode using modal home-work origin-destination data from the 2018 travel survey.
+It reflects the travel behaviour after LEZ implementation, i.e.
+very few motorists commute into the LEZ centro.
+Our purpose in this empirical example is to investigate the trends in spatial availability of employment opportunities by mode, and illustrate how spatial availability can be used to discuss modal competitive advantage.
+We also discuss how multimodal spatial availability landscape displayed in the Figure \ref{fig:Fig5} may have been impacted by the LEZ.
 
@@
 
-In the figure, $V_i^m$ is a proportion of the total number of `r SA_i_zn208$V_i %>% sum()  %>% prettyNum(big.mark = ",")` jobs in the region.
+In Figure \ref{fig:Fig5}, the spatial availability of jobs $V_i^m$ is calculated for each of the four modes $m$ at the level of traffic analysis zones $i$ in Madrid.
+$V_i^m$ is a proportion of the total number of `r SA_i_zn208$V_i %>% sum() %>% prettyNum(big.mark = ",")` jobs in the region.
 Since $V_i^m$ is calculated based on the population of workers and the distribution of jobs, the values can be understood as the number of full-time jobs that are spatially available to full-time workers at that $i$ traveling by mode $m$, relative to all the jobs in the city.
-There are noticeable differences in the magnitude of $V_i^m$ between modes as seen in Figure \ref{fig:Fig5}.
+This implementation of spatial availability assumes that all employment opportunities are of interest to the entire population but are proportional allocated based on relative travel cost (informed by observed modal origin-destination travel behaviour) and the relative amount of the mode-using population residing in the zones.
+
+There are noticeable differences in the magnitude of $V_i^m$ between modes in Figure \ref{fig:Fig5}.
 The majority of $V_i^m$ (which is to say of spatially available jobs) are allocated to workers travelling by car and transit.
@@
 However, the ability to travel at greater speeds also impacts these results.
-Furthermore, differences in $V_i^m$ values within modes also exist in space: car users outside of the M-30 region appear to enjoy greater spatial availability, while some zones inside the M-30 to have greater spatial availability for transit.
+Furthermore, differences in $V_i^m$ values within modes also exist in space: car users outside of the M-30 region appear to enjoy greater spatial availability, while some zones inside the M-30 have greater spatial availability for transit.
 Overall, the magnitude of $V_i^m$ values for cyclists and pedestrians are lower than for car and transit but the highest values of $V_i^{bike}$ and $V_i^{walk}$ tend to be found in zones within the M-30 and origins with higher spatial availability by transit.
@@
 
-The big picture demonstrates how car is the most advantageous mode, however it is interesting to notice that this advantage appears to be blunted by the LEZ.
+The big picture demonstrates how, from the 2018 travel survey, car is the most advantageous mode.
+However, it is interesting to notice that this advantage appears to be blunted by the LEZ.
 Unlike the results for the whole city and even within the M-30 ring, the proportion of car users in Centro is larger than the proportion of opportunities spatially available to them.
@@
 It is also clear that more opportunities are spatially available to non-car users within Centro.
-In the case of active travel, the proportions of cyclists and walkers within LEZ Centro still exceeds the proportions of jobs spatially available to them --- however, the disparity is drastically reduced compared to the rest of the city.
+In the case of active travel, the proportions of cyclists and walkers within LEZ Centro still exceeds the proportions of jobs spatially available to them.
+Notably, this disparity is drastically reduced compared to the rest of the city.
 As seen in Figure \ref{fig:Fig6}, there is a higher proportion of opportunities that are spatially available to pedestrians and cyclists in Centro than in the City overall and in areas within the M-30.
-Notably, within Centro, `r modal_V_comps_4plot$perc[21] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[22] %>% percent(accuracy=0.1)` of opportunities are spatially available to bike and walk modes respectively, while their populations represent  `r modal_V_comps_4plot$perc[1] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[2] %>% percent(accuracy=0.1)` of the population.
+Within Centro, `r modal_V_comps_4plot$perc[21] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[22] %>% percent(accuracy=0.1)` of opportunities are spatially available to bike and walk modes respectively, while their populations represent  `r modal_V_comps_4plot$perc[1] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[2] %>% percent(accuracy=0.1)` of the population.
 By restricting the ability of cars to enter Centro, the LEZ seems to contribute to leveling the playing field for slower modes, in particular cycling and walking, but also transit.
@@
 
-The spatial differences in the competitive dis/advantage of spatial availability between modes can also be visualized at a finer level of granularity.
+The spatial differences in the competitive (dis)advantage of spatial availability between modes can also be visualized at a finer level of granularity.
 Figure \ref{fig:Fig7} shows $v_i^m$, the spatial availability $V_i^m$ divided by the population of users of $m$.
@@
 
-<!--  CONTINUE EDITING FROM HERE  -->
-
 # Discussion and conclusions
@@
 Different accessibility indicators are characterized by how they weight and aggregate opportunities.
-Spatial availability's contribution to the literature is to incorporate a proportional allocation mechanism that essentially constrains the sums to match the number of opportunities in the region; in this way it is a singly-constrained accessibility measure that naturally accommodates congestion and competition.
+Spatial availability's contribution to the literature is to incorporate a proportional allocation mechanism that essentially constrains the sums to match the number of opportunities in the region; in this way it is a singly-constrained accessibility measure that natively accommodates congestion and competition.
 The effort with spatial availability is in line with previous research on proportional allocation by Paez et al.
@@
 
-An empirical example using data from Madrid helped to illustrate the potential of multimodal spatial availability analysis, including its ability to account for competition for opportunities within and between modes.
+An empirical example using data from Madrid helps to illustrate the potential of multimodal spatial availability analysis, including its ability to account for competition for opportunities within and between modes.
 Particularly relevant is the fact that spatial availability scores relate directly to the total number of opportunities in the region.
@@
 This comparability is preserved between regions and over time.
-The example suggests that once that opportunities are treated as being finite, restrictions to travel by car leave more spatially available opportunities for non-car-users.
+The example suggests that once that opportunities are treated as being finite, restrictions to travel by car (due to the LEZ) leave more spatially available opportunities for non-car users.
 This difference for car travel in locations within and immediately around the LEZ Centro seems to increase the number of opportunities spatially available to transit users (transit being the second most competitive mode), as well as the spatial availability from the perspective of non-motorised modes.
-In effect, a policy such as Low Emission Zones help to improve the accessibility situation of active travel and transit in the parts of the city where it is implemented.
+In effect, a policy such as LEZ seems to help improve the spatial availability situation of active travel and transit in the parts of the city where it is implemented.
 
@@
 Accessibility indicators are not designed to work as modal split models, and yet, in the case of policies that alter the relative cost of various forms of transportation, one can reasonably expect to see some shifts between modes.
-In our empirical example we used data collected _after_ the introduction of LEZ Centro.
-However, given a modal split model to project model shares, accessibility indicators, including spatial availability, can be used to investigate changes to the accessibility landscape.
-Ditto for destination choice.
-Our empirical example presented but a snapshot of this, and in future research it will be interesting to investigate changes _between_ policy interventions.
-The expansion of Madrid's LEZ to the ring contained by the M-30 orbital presents an excellent opportunity to do so.
-Given the intuitive and straightforward interpretation of spatial availability scores as fractions of opportunities from the total, relative and absolute changes in the accessibility landscape can be assessed, thus helping to evaluate the implications of policy interventions.
-
-Finally, our example dealt with differences in travel by mode only, but it is possible to think of the intersection between mode of travel and different types of travelers.
+In our empirical example we used data collected after the introduction of LEZ Centro.
+However, given a modal split model to predict model shares, accessibility indicators, including spatial availability, can be used to investigate changes to the accessibility landscape.
+A similar logic can be applied for destination choice.
+Our empirical example presented a snapshot of this, and in future research it may be interesting to investigate changes in spatial availability _between_ policy interventions.
+The expansion of Madrid's LEZ to the ring contained by the M-30 presents an excellent opportunity to do so.
+Given the intuitive interpretation of spatial availability scores as fractions of opportunities from the total, relative and absolute changes in the spatial availability landscape can be assessed, thus helping to evaluate the implications of policy interventions.
+
+Our empirical example dealt with differences in travel by mode only, but it is possible to think of the intersection between mode of travel and different types of travelers.
 This would expand the number of sub-populations in the analysis from, say, $m=M$ (modes) to $m = M\cdot Q$ (modes times population segments), each with their own characteristic impedance function.
 Evaluations of this kind will be especially relevant as LEZ are implemented in cities globally, and the question of their impact on disadvantaged populations who have become mobility-restricted increasingly come to the fore [@devrijNooneVisitsMe2022; @verbeekJustManagementUrban2022; @liottaWhatDrivesInequalities2023].
+
+It is worth reiterating that while the focus of this work was on a multimodal extension of spatial availability, the consideration of multiple modes is just one case of heterogeneous populations (i.e., travel by different modes).
+The method itself can easily accommodate other forms of population or opportunity heterogeneity, for example: variations in rich and poor, young and old, types of opportunities.
+Like accessibility measures before it, the potential application of spatial availability measure are as numerous as potential study contexts.
+
+To close, spatial availability is a singly-constrained accessibility measure that considers competition, a way of accounting for opportunity congestion.
+Spatial availability considers competition by allocating opportunities (the subject of the single-constraint) to zones based on zonal proportions.
+If the demand, supply and travel to opportunities is spatially uniform, then competition is uniform and spatial availability will reflect this.
+It has been suggested that opportunity congestion can be seen on a continuum, but how it is to be considered within accessibility research is an ongoing subject of exploration [@vanweeAccessibilityMeasuresCompetition2001; @merlinDoesCompetitionMatter2017; @kelobonyeMeasuringAccessibilitySpatial2020a].
+This is the context in which we present spatial availability and the multimodal extension.
 

### 2024-01-25 Journal-manuscript/Journal-manuscript.tex
R2 Journal-manuscript revision. For AP review.

@@
 Accessibility is a key concept in the analysis of land use and
-transportation systems {[}e.g., 1,2,3{]}, and too is coming of age from
-the perspective of planning {[}see \emph{inter alia}, 4,5--8{]}.
+transportation systems {[}e.g., 1,2,3{]} and is coming of age from the
+perspective of planning research {[}see \emph{inter alia}, 4,5--8{]}.
 Beginning with the work of Hansen {[}1{]}, accessibility measures have
 been widely used to evaluate the efficiency of transportation systems
-when combined with the distribution of opportunities in space {[}9{]}.
-As such, it is a holistic measure of spatial systems that measures the
-ease of reaching destinations {[}10,11{]}.
-
-In practice, the most common form of accessibility measure is based on
-the gravity model. These measures are sums of weighted opportunities
-around a focal point (i.e., a potential origin), based on how expensive
-it is to reach them. Recent research in accessibility analysis has paid
-attention to the way opportunities are counted in the pertinent
-calculations. Conventionally, the sums are not constrained, which means
-that the same opportunity can enter the sum for different origins.
-Counting the same opportunity multiple times treats it as if it was
-inexhaustible. But opportunities in general are not inexhaustible, and
-in fact some of them are by definition exclusive: for example, once a
-job is taken up by someone in the population, the same job is no longer
-available for any other person to take. More generally, opportunities
-are subject to congestion: for example, multiple people can obtain
-services from the same family doctor, but the more people who do, the
-more congested the service will be.
-
-The issue of congestion in accessibility measures was the motivation for
-the development of floating catchment area approaches {[}12,13{]}. While
-these approaches purport to account for congestion, Paez et al. {[}14{]}
-demonstrate that in general they do not solve the issue of multiple
-counting of opportunities, thus leading to biases in the calculation of
-total demand and supply, sometimes inflating them, other times deflating
-them. In response to this, recent research has paid closer attention to
-the way opportunities are counted in accessibility analysis. Paez et al.
-{[}14{]}, for example, tackle floating catchment area methods and
-introduce a normalization of the impedance matrix to allocate the
-population and then the level of service proportionally. More recently,
-Soukhov et al. {[}15{]} introduced a singly-constrained measure of
-accessibility, called spatial availability, that employs a similar, but
-more sophisticated proportional allocation mechanism. The work of these
-authors show that floating catchment area methods can be seen as
-singly-constrained accessibility measures, and improve on existing
-approaches by guaranteeing that each opportunity is counted only once -
-in other words, treating opportunities as \emph{finite}. The
-proportional allocation of spatial availability constrains the
-calculations to match a known quantity, therefore ensuring that the
-measurements sum up to a predetermined quantity (i.e., the total number
-of opportunities), and so each value can be meaningfully related to this
-total.
+when combined with the distribution of populations and opportunities in
+space {[}9{]}. In this way, accessibility is a holistic measure of
+spatial systems' ease of reaching population-relevant destinations
+{[}10,11{]}.
+
+The most common form of accessibility measure is based the gravity model
+{[}e.g. 1{]}: these measures sum ``weighted'' opportunities around a
+focal point (i.e., a potential origin), based on how expensive it is to
+reach them. Recent research in accessibility analysis has explored the
+implications of methods by which opportunities are summed, especially in
+the context of competitive accessibility {[}15{]}. In a typical
+Hansen-type accessibility measure, the sums around origins are not
+constrained. This means that the same opportunity can enter the sum for
+different origins; Counting the same opportunity multiple times treats
+the opportunity as if it was inexhaustible (or non-rival {[}16{]}).
+However, this conceptualisation may not reflect reality as
+opportunities, in general, are not inexhaustible. Some opportunities are
+by definition exclusive: a typical example is employment, once a job is
+taken up by someone in the population, the same job is no longer
+available for any other person to take {[}14,17{]}. More generally, it
+could be conceptualised that all opportunities are subject to some
+amount of congestion or capacity-constraint: multiple people use a
+specific opportunity at a given time and the more people who do, the
+more congested the opportunity. Congestion can be seen on a continuum,
+and debate on how congestion within accessibility should be considered
+is ongoing {[}14,17,18{]}. Nonetheless, competition has been widely
+considered for traditionally exclusive opportunities types such as
+employment and healthcare opportunities {[}14,17,19--21,21--24{]} and
+more recently educational facilities {[}18,25{]}, and less exclusive
+amenities-types such as greenspace and recreation {[}25,26{]}, and
+shopping {[}18,27{]}.
+
+The consideration of congestion in accessibility measures was the
+motivation for the development of approaches that consider competition
+{[}21,23,24{]} and are popularly applied as floating catchment area
+methods {[}e.g. 28{]}. While these approaches purport to account for
+congestion, Paez et al. {[}12{]} demonstrates that they do not solve the
+issue of multiple counting of opportunities in general, thus leading to
+biases in the calculation of total demand and supply. They sometimes
+inflate counts and other times deflate them. In response to this, recent
+research has paid closer attention to the way opportunities are counted
+in accessibility analysis. Paez et al. {[}12{]}, for example, focuses on
+floating catchment area methods and introduces a normalization of the
+impedance matrix to allocate the population and then the level of
+service proportionally. More recently, Soukhov et al. {[}15{]}
+introduced a singly-constrained measure of accessibility, called spatial
+availability, that employs a similar but more sophisticated proportional
+allocation mechanism. The work of these authors show that floating
+catchment area methods can be seen as singly-constrained accessibility
+measures and improve on existing approaches by guaranteeing that each
+opportunity is counted only once. In other words, spatial availability
+treats the opportunities as \emph{finite}. The proportional allocation
+mechanism of spatial availability constrains the calculations to match a
+known quantity, therefore ensuring that the measurements sum up to a
+predetermined quantity (i.e., the total number of opportunities), and so
+each value can be meaningfully related to this total.
 
@@
 
-The objective of this paper is to address this limitation of spatial
-availability. Our primary motivation is to extend spatial availability
-for the case of multimodal accessibility, but it is worthwhile noting
-that this is in fact just one case of heterogeneous populations (i.e.,
-travel by different modes). The method itself can easily accommodate
-other forms of heterogeneity, for example variations in travel behavior
-between older and younger adults {[}e.g., 16{]}, the propensity of older
-adults to use different modes of transportation {[}e.g., 17{]}, the
-usually shorter trip lengths of children compared to grown-ups {[}e.g.,
-18{]}, or the more limited travel ranges of single parents {[}e.g.,
-19{]}.
-
-The paper rest of this paper is organized as follows. In Section 2 we
-provide a brief review of multimodal accessibility. In the following
-Section 3 we demonstrate the derivation of the new spatial availability
-expression for multiple modes. In Section 4 we illustrate relevant
-issues through a synthetic example. This is followed in Section 5 by an
-empirical example using data from the city of Madrid after the
-implementation of its Low Emission Zones (LEZ). Data for this example
-comes from the city's 2018 travel survey. The example shows the
-differences in spatial availability within and outside the LEZ for
-travelers using different modes, namely car, transit, cycling and
-walking. In Section 6, we provide concluding remarks on the strengths of
-the use of spatial availability as a multimodal accessibility measure,
-and discuss potential future uses in policy planning scenarios as well
-as directions for future research.
+This paper's primary motivation is to extend spatial availability for
+the case of multimodal accessibility (i.e., travel by different modes).
+It is worth noting though that the consideration for travel costs by
+different modes is in fact just one case of heterogeneous populations.
+The method itself can easily accommodate other forms of heterogeneity,
+for example: variations in travel behavior between older and younger
+adults {[}e.g., 29{]}, the propensity of older adults to use different
+modes of transportation {[}e.g., 30{]}, the usually shorter trip lengths
+of children compared to adults {[}e.g., 31{]}, or the more limited
+travel ranges of single parents {[}e.g., 32{]}.
+
+The rest of this paper is organized as follows. In Section 2 we provide
+a brief review of multimodal accessibility. In the following Section 3
+we demonstrate the derivation of the new spatial availability expression
+for multiple modes. In Section 4 we illustrate relevant issues through a
+synthetic example. This is followed in Section 5 by an empirical example
+using home-to-work data from the city of Madrid, Spain after the
+implementation of its Low Emission Zones (LEZ). Data for this example is
+sourced from the city's 2018 travel survey. The empirical example
+demonstrates the differences in multimodal spatial availability
+highlighting differences within and outside the LEZ for travelers using
+different modes, namely car, transit, cycling and walking. In Section 6,
+we provide concluding remarks on the strengths of the use of spatial
+availability as a multimodal accessibility measure, and discuss
+potential future uses in policy planning scenarios as well as directions
+for future research.
 
@@
 land-use and transport systems. Arguably, the most commonly used are
-measures based on the gravity model {[}20{]}, of which cumulative
-opportunity measures and weighted cumulative opportunity measures are
-particular forms {[}5{]}. These measures assign a weight to
-opportunities based on how easy it is to reach them. Given an origin
-(\(i\)) and a destination (\(j\)), an impedance function
-\(f^{m}(c^m_{ij})\) converts the cost of travel (e.g., time, money,
-generalized cost) into a score that represents the propensity for
-interaction. These measures originate from that proposed by Hansen
-{[}1{]}, which can take the following form in the multimodal case:
-\(S_i^m = \sum_j O_j f^m(c_{ij}^m)\) where \(m\) is a set of modes which
-have mode-specific travel costs (\(c_{ij}^m\)) and/or travel impedance
-functions (\(f^m(\cdot)\)).
+measures based on the gravity model {[}33{]}; this includes cumulative
+opportunity measures which are a special case of the gravity model
+{[}5,10{]}. These gravity model based measures weight opportunities
+depending on how easy it is to reach them. Given an origin (\(i\)) and a
+destination (\(j\)), an impedance function \(f^{m}(c^m_{ij})\) converts
+the cost of travel (e.g., time, money, generalized cost) into a score
+that represents the propensity for potential interaction. These measures
+originate from the work proposed by Hansen {[}1{]}, which can take the
+following form in the multimodal case:
+\(S_i^m = \sum_j O_j f^m(c_{ij}^m)\). In this form, \(m\) is a set of
+modes that have mode-specific travel costs (\(c_{ij}^m\)) and/or travel
+impedance functions (\(f^m(\cdot)\)).
 
@@
 not consider the opportunities as finite. To cite an example, Tahmasbi
-et al. {[}21{]} use Hansen-type accessibility to assess the potential
+et al. {[}34{]} use Hansen-type accessibility to assess the potential
 interaction with retail locations by three modes: walking, public
@@
 each origin \(i\) three accessibility scores are calculated. In this
-work, Tahmasbi et al. {[}21{]} show that car travel affords the highest
+work, Tahmasbi et al. {[}34{]} show that car travel affords the highest
 \(S_i^{m}\) values in the majority of \(i\), i.e., travelers who use a
@@
 modes are often values that are difficult to interpret beyond making
-statements about relative size. For example, Lunke {[}22{]}, researching
+statements about relative size. For example, Lunke {[}35{]}, researching
 the region of Oslo, reports accessibility scores for car in the order of
@@
 10,000 employment opportunities are still plenty of opportunities. These
-ratios can be found elsewhere in the literature {[}e.g., 23, and
-9,16,19, and 8{]}, and they are useful as relative assessment of when
+ratios can be found elsewhere in the literature {[}e.g., 36, and
+9,29,32, and 8{]}, and they are useful as relative assessment of when
 some members of the public are better or worse off than others, but they
-do not say much about how bad is ``worse off''.
-
-Besides ratios of accessibility, another way to improve interpretability
-of scores sometimes seen in the literature is to standardize them to lie
-in the range {[}0-1{]}. This adjustment is only helpful insofar as it
+are silent on how bad is ``worse off''.
+
+Besides ratios of accessibility, another way seen in the literature to
+improve interpretability of scores is to standardise them within a
+{[}0-1{]} range. This adjustment is only helpful insofar as it
 facilitates relative comparisons, but interpretation of the scores
@@
 that can be intuitively understood. Presentation or discussion of
-Hansen-type accessibility that has been standardized in this way is not
-uncommon in the literature {[}e.g., 24,25{]}.
+Hansen-type accessibility that has been standardised in this way is not
+uncommon in the literature {[}e.g., 37,38{]}.
 
@@
 limited and thus is subject to competition by population
-{[}12,15,26--30{]}. There are only so many school-seats, hospital
+{[}14,15,17--25,39--42{]}. There are only so many school-seats, hospital
 capacity, employment opportunities, etc., in a region and if one person
@@
 Ile-de-France region, a jurisdiction that suggested in a 2013 planning
-document that at the municipal level at least \(10m^2\) of public green
-space should be supplied \emph{per inhabitant} {[}31{]}. Green spaces
+document that at the municipal level at least 10 \(m^2\) of public green
+space should be supplied \emph{per inhabitant} {[}43{]}. Green spaces
 are not evenly distributed, which means that who has access to them
@@
   https://redfrogforum.org/wp-content/uploads/2019/11/67-Nature-Nearby\%E2\%80\%99-Accessible-Natural-Greenspace-Guidance.pdf}.
-Similarly, the World Health Organization {[}cited in 32{]} recommends
-that cities provide a minimum of 9\(m^2\) of green area per inhabitant.
+Similarly, the World Health Organization {[}cited in 44{]} recommends
+that cities provide a minimum of 9 \(m^2\) of green area per inhabitant.
 For our purposes, standards of this type translate into ``how much of
@@
 still have a capacity, and it is not the same for a person to have
-access to \(5m^2\) of \emph{uncongested} green space as \(15m^2\). This
-difference is in fact a matter of justice {[}31,33{]}. Constraining
+access to 5 \(m^2\) of \emph{uncongested} green space as 15 \(m^2\).
+This difference is in fact a matter of justice {[}43,45{]}. Constraining
 accessibility is in this way a useful way to evaluate the congested
@@
 standards is emphasized in the planning literature, in particular in
-regards to fairness in transportation {[}see 34{]}, spatial availability
+regards to fairness in transportation {[}see 46{]}, spatial availability
 analysis is a useful way to develop and assess standards.
@@
 we think about the use of multiple modes (or heterogeneous populations).
-If we return to Oslo for a moment {[}22{]}, we notice that the places
+If we return to Oslo for a moment {[}35{]}, we notice that the places
 that have high accessibility by transit are also the places that have
@@
 populations are going for the same opportunities, and those travelling
-by transit have fewer to choose from to begin with. More generally,
-people in a zone who are advantaged with relatively low cost of travel
-will have the ability to potentially reach more opportunities than other
+by transit have fewer to choose from the start. More generally, people
+in a zone who are advantaged with relatively low cost of travel will
+have the ability to potentially reach more opportunities than other
 people. Due to this advantage, through the perspective of finite
@@
 rationale for developing floating catchment area methods (FCA),
-popularized by Luo et al. {[}13{]} who reformulated the work of Shen
-{[}12{]} into two steps (although similar, and earlier, developments are
-found in {[}30,35{]}). Shen-type accessibility is formulated as:
+popularized by Luo et al. {[}28{]} who reformulated the work of Shen
+{[}24{]} into two steps (although similar, and earlier, developments are
+found in {[}21,23{]}). Shen-type accessibility is formulated as:
 \(a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}\) where \(D_j^m\)
@@
 meaning both population \emph{and} opportunities are multiply counted
-{[}see 14{]}. In other words, interpretation of the Shen-type
-accessibility scores between modes is fraught as it is for Hansen-type
+{[}see 12{]}. In other words, interpretation of the Shen-type
+accessibility scores between modes is fraught, as it is for Hansen-type
 measures.
 
-To illustrate, {[}36{]} calculate \(a_i^m\) to jobs for different
+To illustrate, {[}47{]} calculates \(a_i^m\) to jobs for different
 income-group populations in Shenzhen, China using
@@
 
-Spatial availability improves on previously discussed accessibility
-approaches using the Hansen-type measure and the Shen-type measure by
-constraining the sum of opportunities, that is, by treating
-opportunities as finite. This is done by means of proportional
-allocation factors that follow well established principles of spatial
-interaction and the gravity model {[}see 37{]}. In Soukhov et al.
-{[}15{]} these factors consider: the mass effect (e.g., the size of
+Spatial availability improves on the discussed Hansen-type and Shen-type
+accessibility approaches by constraining the sum of opportunities, that
+is, by treating opportunities as finite. This is done by means of
+proportional allocation factors that follow well established principles
+of spatial interaction and the gravity model {[}see 48{]}. In Soukhov et
+al. {[}15{]} these factors consider: the mass effect (e.g., the size of
 populations at different origins); and the cost of travel from different
@@
 
-Compared to Hansen-type accessibility:
-
-\begin{equation}
-\label{eq:hansen-accessibility}
-A_i = \sum_{j=1}^J O_jf(c_{ij})
-\end{equation}
-
-\noindent we see that spatial availability is, like the Hansen-type
-measure, a weighted sum of the opportunities. What makes spatial
-availability stand apart from other approaches is how the weight used in
-the sum, balancing factor \(F^t_{ij}\), implements a proportional
-allocation mechanisms to ensure that the sum of \(V_i\) is constrained
-to match the total number of opportunities in the region. As such,
-spatial availability is singly-constrained and naturally implements
-competition or congestion. \(F^t_{ij}\) consists of two parts. The first
-part is a population-based proportional allocation factor to model the
-mass effect of the gravity model: \[
+Compared to Hansen-type accessibility
+\(S_i = \sum_{j=1}^J O_jf(c_{ij})\), we see that spatial availability
+is, like the Hansen-type measure, a weighted sum of the opportunities.
+What makes spatial availability stand apart from other approaches is how
+the weight used in the sum (balancing factor \(F^t_{ij}\)) implements a
+proportional allocation mechanisms to ensure that the sum of \(V_i\) is
+constrained to match the total number of opportunities in the region. As
+such, spatial availability is singly-constrained and natively considers
+competition. \(F^t_{ij}\) consists of two parts. The first part is a
+population-based proportional allocation factor to model the mass effect
+of the gravity model: \[
 F^p_{i} = \frac{P_i}{\sum_i P_i}
@@
 multimodal applications (or more generally heterogeneous populations).
-To do so, the balancing factors need to be reformulated so that 1) the
-mass effect now accounts not only for the size of the population at
-\(i\), but also the size of sub-populations within \(i\); and 2) the
-cost of travel is not only for different zones, but by sub-populations
-within each zone (e.g., the cost of travel from \(i\) by car, transit,
-walking, etc.) When we introduce modes (or sub-populations) \(m\), the
-proportional allocation factors need to satisfy the condition that
-\(F^{pm}_{i}\) and \(F^{cm}_{ij}\) can be summed across each \(m\) at
-each \(i\) and then across all \(i\) to equal to 1. They are also
-similarly combined multiplicatively to obtain their joint effect,
-represented as the combined balancing factor \(F^{tm}_{ij}\) similar to
-that detailed in Equation (\ref{eq:multimodal-balancing-factors}). This
-factor is given by:
+To do so, the balancing factors are reformulated so that 1) the mass
+effect now accounts not only for the size of the population at \(i\),
+but also the size of sub-populations within \(i\); and 2) the cost of
+travel is not only for different zones, but by sub-populations within
+each zone (e.g., the cost of travel from \(i\) by car, transit, walking,
+etc.) relative to all zones. When we introduce modes (or
+sub-populations) \(m\), the proportional allocation factors need to
+satisfy the condition that \(F^{pm}_{i}\) and \(F^{cm}_{ij}\) can be
+summed across each \(m\) at each \(i\) and then across all \(i\) to
+equal to 1. They are also similarly combined multiplicatively to obtain
+their joint effect, represented as the combined balancing factor
+\(F^{tm}_{ij}\) similar to that detailed in Equation
+(\ref{eq:multimodal-balancing-factors}). This factor is given by:
 
@@
 
-Next we use a synthetic example to contrast multimodal accessibility and
-spatial availability.
+Next we use a synthetic example to contrast multimodal spatial
+availability with multimodal versions of Hansen-type accessibility
+(unconstrained) and Shen-type (unconstrained and competitive)
+accessibility.
 
@@
 population, followed by then \(A\), then \(C\). This synthetic example
-is inspired by the single-mode example used in {[}12{]} and reconfigured
+is inspired by the single-mode example used in {[}24{]} and reconfigured
 in {[}15{]}.
@@
 Sub-populations located in opposite conditions (i.e., more distant from
-jobs, close to large populations, and using slow mode \(x\)) are at a
+jobs, close to large populations, and using slower mode \(x\)) are at a
 relative disadvantage. The competition for opportunities between
@@
 columns and are aggregated for each \(i\) in the final two columns in
-Table 1 . As in the example in Shen {[}12{]}, we use a negative
+Table 1 . As in the example in Shen {[}24{]}, we use a negative
 exponential impedance function
@@
 To connect this example to the aforementioned literature, \(S_i^m\) is
-calculated in the work of Tahmasbi et al. {[}21{]}; they contrast
+calculated in the work of Tahmasbi et al. {[}34{]}; they contrast
 differences in \(S_i^m\) values between modes in a relative and
@@
 Shen-type measure is implemented in aforementioned work of Tao et al.
-{[}36{]} to calculate modal \(a_i^m\) values and the aggregated \(a_i\)
-is implemented in the work of Carpentier et al. {[}38{]}. However,
+{[}47{]} to calculate modal \(a_i^m\) values and the aggregated \(a_i\)
+is implemented in the work of Carpentier et al. {[}49{]}. However,
 similar to Hansen-type accessibility, these works discuss relative and
@@
 spatially available opportunities than its share in the population. This
-realization leads us to ask normative questions such as, how unequal
-should availability of opportunities be by mode? What intervention could
-help to redistribute spatial availability to sub-populations
-commensurate with their proportion of the total?
+realization could lead us to ask normative questions: how unequal should
+availability of opportunities be by mode? What intervention could help
+to redistribute spatial availability to sub-populations commensurate
+with their proportion of the total?
 
@@
 and increase he values of \(v_i^x\) (making it faster or les expensive).
-The purpose of this simple demonstration is to show how spatial
+The purpose of this synthetic demonstration is to show how spatial
 availability can be used to quantify the competitive (dis)advantage in a
@@
 The context for the empirical example is Madrid, Spain. This city
-implemented a Low Emission Zone (LEZ) in 2017 to: pursue goals set out
-in the national climate change agenda, cut nitrogen dioxide levels, and
-to prioritize people's movement in the city. LEZs elsewhere have
-similarly been implemented as interventions to reduce GHG emissions,
-improve air quality, and support sustainable mobility {[}39,40{]}.
-Though the rules of exclusion vary by city, LEZs aim to deter/reduce
-traffic in designated zones under threat of penalty (e.g., fines,
-seizure of vehicle). In other words, LEZs implement a form of
-\emph{geographic discrimination} as they change how people can reach
-opportunities by making it more costly for some forms of travel,
-typically cars, to circulate in predetermined zones. When considering
-opportunities as finite in a region, this discrimination reduces the
-competition of one mode and opens up opportunities for other modes to
-better thrive. At their core, LEZs operate by changing the accessibility
-landscape of a city from the perspective of multiple modes.
+implemented a Low Emission Zone (LEZ) in 2017 to pursue goals set out in
+the national climate change agenda, cut nitrogen dioxide levels, and to
+prioritize people's movement in the city. LEZs elsewhere have similarly
+been implemented as interventions to reduce GHG emissions, improve air
+quality, and support sustainable mobility {[}50,51{]}. Though the rules
+of exclusion vary by city, LEZs aim to deter/reduce traffic in
+designated zones under threat of penalty (e.g., fines, seizure of
+vehicle). In other words, LEZs implement a form of \emph{geographic
+discrimination} as they change how people can reach opportunities by
+making it more costly for some forms of travel, typically cars, to
+circulate in predetermined zones. When considering opportunities as
+finite in a region, this discrimination reduces the competition of one
+mode and opens up opportunities for other modes to better thrive. At
+their core, LEZs operate by changing the accessibility landscape of a
+city from the perspective of multiple modes.
 
@@
 restriction impacted approximately half of all car trips that used to
-travel into what is now the LEZ Centro {[}41{]}.
-
-For this case study, we use spatial availability to quantify access to
-opportunities by different modes in Madrid. Particularly, we demonstrate
-how \(V_i^m\) can be used to derive insights into how the restriction of
-car mobility in areas around/within the LEZ Centro may have allowed more
-sustainable (but often slower or more costly modes) to become more
-competitive.
+travel into what is now the LEZ Centro {[}52{]}.
+
+For this case study, we quantify spatial availability to opportunities
+by different modes in Madrid. Particularly, we demonstrate how \(V_i^m\)
+can be used to derive insights into how the restriction of car mobility
+in areas around/within the LEZ Centro may have allowed more sustainable
+(but often slower or more costly modes) to become more competitive.
 
@@
 \caption{\label{fig:Fig2} Distribution of jobs taken by people living and working in Madrid as reported in the 2018 travel survey.
-Grey TAZs has no jobs.
-Ranges of values in the legend are quintiles.}\label{fig:jobs-plot}
+Grey TAZs have no jobs.
+Ranges of values in the legend are quintiles.
+The TAZ shapefile is available from the Community of Madrid open data portal.}\label{fig:jobs-plot}
 \end{figure}
@@
 
-\caption{\label{fig:Fig3} Population living and working in Madrid by mode of transportation as reported in the 2018 travel survey.
-Grey TAZs have no population.
-Ranges of values in the legend are quintiles.}\label{fig:pop-plot}
+\caption{\label{fig:Fig3} Population living and working in Madrid by mode of transportation as reported in the 2018 travel survey and represented at the level of TAZ.
+Grey TAZ have no population.
+Ranges of values in the legend are quintiles.
+The TAZ shapefile is available from the Community of Madrid open data portal.}\label{fig:pop-plot}
 \end{figure}
@@
 
-The source of data for our empirical example is the 2018 Travel Survey
-of the Community of Madrid {[}42{]}. This is a representative survey
-that offers a snap-shot of travel patterns for a typical weekday in
-2018. The survey collected 222,744 trips from a representative sample of
-85,064 households across the traffic analysis zones (TAZ) in the
-Community of Madrid. For context, the population older than 3 years in
-the Community is 6,507,184.
-
-In this example, we use all home-to-full-time-work trips, by all modes.
-The trips are expanded using population weights. Figures \ref{fig:Fig2}
-and \ref{fig:Fig3} show the number of workers and the distribution of
-full-time jobs in the City of Madrid by TAZ. The light grey boundary
-represents the LEZ Centro in effect in 2017. The survey was conducted
-after the introduction of LEZ Centro. The dark grey boundary represents
+The source of origin-destination data for our empirical example is the
+2018 Travel Survey of the Community of Madrid {[}53{]}. This is a
+representative survey that offers a snap-shot of travel patterns for a
+typical weekday in 2018. The survey collected 222,744 trips from a
+representative sample of 85,064 households across the traffic analysis
+zones (TAZ) in the Community of Madrid. For context, the population
+older than 3 years in the Community is 6,507,184.
+
+In this example, we use all direct home-to-full-time-work trips, by all
+modes. The trips are expanded using population weights. Figures
+\ref{fig:Fig2} and \ref{fig:Fig3} show the number of workers and the
+distribution of full-time jobs in the City of Madrid by TAZ. The TAZ
+shapefiles are available from the Community of Madrid open data portal
+{[}54{]}. The pink boundary represents the LEZ Centro in effect in 2017
+and reflected in the 2018 travel survey. The purple boundary represents
 the LEZ planned for the boundaries of the M-30 highway and is present in
@@
 \(P^m_i\), is displayed in Figure \ref{fig:Fig3}. The modal shares in
-Figure \ref{fig:Fig3} are calculated based on those measured in the
-survey. The modal categories and the mode types within each category are
-reported as follows:
+Figure \ref{fig:Fig3} are calculated based on the primary mode specified
+in the survey and summarized into four categories as follows:
 
@@
 bins. This distribution is then used to estimate the parameters of a
-function for the travel impedance {[}as done in 43,44,45{]}. To fit the
+function for the travel impedance {[}as done in 55,56,57{]}. To fit the
 impedance functions, we use the Maximum likelihood estimation and the
 Nelder-Mead method for direct optimization available within the R
-\{fitdistrplus\} package {[}46{]}. Based on goodness-of-fit criteria and
+\{fitdistrplus\} package {[}58{]}. Based on goodness-of-fit criteria and
 associated diagnostics, the gamma and log-normal probability density
@@
 non-motorised modes respectively. The selection of functional forms
-aligns with empirical examples in other regions {[}15,47,48{]}. The
+aligns with empirical examples in other regions {[}15,59,60{]}. The
 shape and rate parameters for the gamma functions (motorised modes) are
@@
 
-At this point, it is worthwhile reiterating that the empirical example
-is a snap-shot of spatial availability by mode using data from the 2018
-travel survey. Our purpose in this empirical example is to investigate
-the trends in availability of employment opportunities by mode, and
-illustrate how spatial availability can be used in discussions about the
-competitive advantage of various modes within Madrid Centro. The spatial
-availability of jobs \(V_i^m\) is calculated for each of the four modes
-\(m\) at the level of traffic analysis zones \(i\) in Madrid and
-displayed in Figure \ref{fig:Fig5}.
+It is worthwhile reiterating that the empirical example is a snapshot of
+spatial availability by mode using modal home-work origin-destination
+data from the 2018 travel survey. It reflects the travel behaviour after
+LEZ implementation, i.e.~very few motorists commute into the LEZ centro.
+Our purpose in this empirical example is to investigate the trends in
+spatial availability of employment opportunities by mode, and illustrate
+how spatial availability can be used to discuss modal competitive
+advantage. We also discuss how multimodal spatial availability landscape
+displayed in the Figure \ref{fig:Fig5} may have been impacted by the
+LEZ.
 
@@
 
-\caption{\label{fig:Fig5} Spatial availability of jobs per origin and mode $V_i^m$ in Madrid. Grey TAZs have no population. Ranges of values in the legend are quintiles.}\label{fig:SA-m-plot}
+\caption{\label{fig:Fig5} Spatial availability of jobs per origin and mode $V_i^m$ in Madrid at the level of TAZ.
+Grey TAZ have no population.
+Ranges of values in the legend are quintiles.
+The TAZ shapefile is available from the Community of Madrid open data portal.}\label{fig:SA-m-plot}
 \end{figure}
 
-In the figure, \(V_i^m\) is a proportion of the total number of 847,574
-jobs in the region. Since \(V_i^m\) is calculated based on the
-population of workers and the distribution of jobs, the values can be
-understood as the number of full-time jobs that are spatially available
-to full-time workers at that \(i\) traveling by mode \(m\), relative to
-all the jobs in the city. There are noticeable differences in the
-magnitude of \(V_i^m\) between modes as seen in Figure \ref{fig:Fig5}.
-The majority of \(V_i^m\) (which is to say of spatially available jobs)
-are allocated to workers travelling by car and transit. In a way, this
-is to be expected since users of these modes represent 84.1\% of the
-total population. However, the ability to travel at greater speeds also
-impacts these results. Furthermore, differences in \(V_i^m\) values
-within modes also exist in space: car users outside of the M-30 region
-appear to enjoy greater spatial availability, while some zones inside
-the M-30 to have greater spatial availability for transit. Overall, the
-magnitude of \(V_i^m\) values for cyclists and pedestrians are lower
-than for car and transit but the highest values of \(V_i^{bike}\) and
-\(V_i^{walk}\) tend to be found in zones within the M-30 and origins
-with higher spatial availability by transit.
+In Figure \ref{fig:Fig5}, the spatial availability of jobs \(V_i^m\) is
+calculated for each of the four modes \(m\) at the level of traffic
+analysis zones \(i\) in Madrid. \(V_i^m\) is a proportion of the total
+number of 847,574 jobs in the region. Since \(V_i^m\) is calculated
+based on the population of workers and the distribution of jobs, the
+values can be understood as the number of full-time jobs that are
+spatially available to full-time workers at that \(i\) traveling by mode
+\(m\), relative to all the jobs in the city. This implementation of
+spatial availability assumes that all employment opportunities are of
+interest to the entire population but are proportional allocated based
+on relative travel cost (informed by observed modal origin-destination
+travel behaviour) and the relative amount of the mode-using population
+residing in the zones.
+
+There are noticeable differences in the magnitude of \(V_i^m\) between
+modes in Figure \ref{fig:Fig5}. The majority of \(V_i^m\) (which is to
+say of spatially available jobs) are allocated to workers travelling by
+car and transit. In a way, this is to be expected since users of these
+modes represent 84.1\% of the total population. However, the ability to
+travel at greater speeds also impacts these results. Furthermore,
+differences in \(V_i^m\) values within modes also exist in space: car
+users outside of the M-30 region appear to enjoy greater spatial
+availability, while some zones inside the M-30 have greater spatial
+availability for transit. Overall, the magnitude of \(V_i^m\) values for
+cyclists and pedestrians are lower than for car and transit but the
+highest values of \(V_i^{bike}\) and \(V_i^{walk}\) tend to be found in
+zones within the M-30 and origins with higher spatial availability by
+transit.
 
@@
 
-The big picture demonstrates how car is the most advantageous mode,
-however it is interesting to notice that this advantage appears to be
-blunted by the LEZ. Unlike the results for the whole city and even
-within the M-30 ring, the proportion of car users in Centro is larger
-than the proportion of opportunities spatially available to them. The
-restriction on cars in effect reduces competition by this mode, and
-leads to a relative increment of the mass effect of the modes allowed
-within the LEZ, which also contains a large number of jobs (see Figure
-\ref{fig:Fig2}). As summarized in the two right-most columns in Figure
-\ref{fig:Fig6}, the proportion of jobs spatially available to car in
-Centro is (13.8\% or 5,373 opportunities). For reference, this is less
-than the proportion of the car users in Centro (16.1\%), evidently less
-then the proportion of car users in the city, and is the opposite of the
-overall trend (left-most columns) and within the M-30 (middle columns).
+The big picture demonstrates how, from the 2018 travel survey, car is
+the most advantageous mode. However, it is interesting to notice that
+this advantage appears to be blunted by the LEZ. Unlike the results for
+the whole city and even within the M-30 ring, the proportion of car
+users in Centro is larger than the proportion of opportunities spatially
+available to them. The restriction on cars in effect reduces competition
+by this mode, and leads to a relative increment of the mass effect of
+the modes allowed within the LEZ, which also contains a large number of
+jobs (see Figure \ref{fig:Fig2}). As summarized in the two right-most
+columns in Figure \ref{fig:Fig6}, the proportion of jobs spatially
+available to car in Centro is (13.8\% or 5,373 opportunities). For
+reference, this is less than the proportion of the car users in Centro
+(16.1\%), evidently less then the proportion of car users in the city,
+and is the opposite of the overall trend (left-most columns) and within
+the M-30 (middle columns).
 
@@
 proportions of cyclists and walkers within LEZ Centro still exceeds the
-proportions of jobs spatially available to them --- however, the
-disparity is drastically reduced compared to the rest of the city. As
-seen in Figure \ref{fig:Fig6}, there is a higher proportion of
-opportunities that are spatially available to pedestrians and cyclists
-in Centro than in the City overall and in areas within the M-30.
-Notably, within Centro, 1.7\% and 16.1\% of opportunities are spatially
-available to bike and walk modes respectively, while their populations
-represent 1.2\% and 14.7\% of the population. By restricting the ability
-of cars to enter Centro, the LEZ seems to contribute to leveling the
-playing field for slower modes, in particular cycling and walking, but
-also transit. As seen in the Figure \ref{fig:Fig6}, transit users are
-generally close to parity across the region, with nearly as many
-spatially available jobs as transit users. Still, this mode has the
-greatest advantage in LEZ Centro with 68.5\% of spatially available jobs
-in Centro for 51.4\% of transit users in Centro. This result makes
-intuitive sense: after car, it is the mode with the greatest range, and
-unlike car it is unrestricted in the LEZ Centro.
-
-The spatial differences in the competitive dis/advantage of spatial
+proportions of jobs spatially available to them. Notably, this disparity
+is drastically reduced compared to the rest of the city. As seen in
+Figure \ref{fig:Fig6}, there is a higher proportion of opportunities
+that are spatially available to pedestrians and cyclists in Centro than
+in the City overall and in areas within the M-30. Within Centro, 1.7\%
+and 16.1\% of opportunities are spatially available to bike and walk
+modes respectively, while their populations represent 1.2\% and 14.7\%
+of the population. By restricting the ability of cars to enter Centro,
+the LEZ seems to contribute to leveling the playing field for slower
+modes, in particular cycling and walking, but also transit. As seen in
+the Figure \ref{fig:Fig6}, transit users are generally close to parity
+across the region, with nearly as many spatially available jobs as
+transit users. Still, this mode has the greatest advantage in LEZ Centro
+with 68.5\% of spatially available jobs in Centro for 51.4\% of transit
+users in Centro. This result makes intuitive sense: after car, it is the
+mode with the greatest range, and unlike car it is unrestricted in the
+LEZ Centro.
+
+The spatial differences in the competitive (dis)advantage of spatial
 availability between modes can also be visualized at a finer level of
@@
 
-\caption{\label{fig:Fig7} Distribution of spatially available jobs per capita by mode of transportatin ($v_i^m$).
-Grey TAZs have no population that use the mode.
-Ranges of values in the legend are quintiles.}\label{fig:SA-per-capita-m-plot}
+\caption{\label{fig:Fig7} Distribution of spatially available jobs per capita by mode of transportatin ($v_i^m$) represented at the level of TAZ.
+Grey TAZ have no population that use the mode.
+Ranges of values in the legend are quintiles.
+The TAZ shapefile is available from the Community of Madrid open data portal.}\label{fig:SA-per-capita-m-plot}
 \end{figure}
@@
 Aggregations of opportunities embody principles of gravitational/spatial
-interaction modelling that date back to at least H.C. Carey {[}49{]},
+interaction modelling that date back to at least H.C. Carey {[}61{]},
 and are part of a line of research that includes the work of Ravenstein
-{[}50{]}, Reilly {[}51{]}, Stewart {[}52--54{]}, Zipf {[}55,56{]},
-Wilson {[}37{]}, and many others. In this way, \(S_i^m\), \(a_i^m\), and
+{[}62{]}, Reilly {[}63{]}, Stewart {[}64--66{]}, Zipf {[}67,68{]},
+Wilson {[}48{]}, and many others. In this way, \(S_i^m\), \(a_i^m\), and
 yes, \(V_i^m\), can be interpreted as scores of the potential for
@@
 the region; in this way it is a singly-constrained accessibility measure
-that naturally accommodates congestion and competition. The effort with
+that natively accommodates congestion and competition. The effort with
 spatial availability is in line with previous research on proportional
-allocation by Paez et al. {[}14{]}. As initially introduced by Soukhov
+allocation by Paez et al. {[}12{]}. As initially introduced by Soukhov
 et al. {[}15{]}, spatial availability was designed for a homogeneous
@@
 
-An empirical example using data from Madrid helped to illustrate the
+An empirical example using data from Madrid helps to illustrate the
 potential of multimodal spatial availability analysis, including its
@@
 once that opportunities are treated as being finite, restrictions to
-travel by car leave more spatially available opportunities for
-non-car-users. This difference for car travel in locations within and
-immediately around the LEZ Centro seems to increase the number of
-opportunities spatially available to transit users (transit being the
-second most competitive mode), as well as the spatial availability from
-the perspective of non-motorised modes. In effect, a policy such as Low
-Emission Zones help to improve the accessibility situation of active
-travel and transit in the parts of the city where it is implemented.
+travel by car (due to the LEZ) leave more spatially available
+opportunities for non-car users. This difference for car travel in
+locations within and immediately around the LEZ Centro seems to increase
+the number of opportunities spatially available to transit users
+(transit being the second most competitive mode), as well as the spatial
+availability from the perspective of non-motorised modes. In effect, a
+policy such as LEZ seems to help improve the spatial availability
+situation of active travel and transit in the parts of the city where it
+is implemented.
 
@@
 see some shifts between modes. In our empirical example we used data
-collected \emph{after} the introduction of LEZ Centro. However, given a
-modal split model to project model shares, accessibility indicators,
-including spatial availability, can be used to investigate changes to
-the accessibility landscape. Ditto for destination choice. Our empirical
-example presented but a snapshot of this, and in future research it will
-be interesting to investigate changes \emph{between} policy
-interventions. The expansion of Madrid's LEZ to the ring contained by
-the M-30 orbital presents an excellent opportunity to do so. Given the
-intuitive and straightforward interpretation of spatial availability
-scores as fractions of opportunities from the total, relative and
-absolute changes in the accessibility landscape can be assessed, thus
-helping to evaluate the implications of policy interventions.
-
-Finally, our example dealt with differences in travel by mode only, but
+collected after the introduction of LEZ Centro. However, given a modal
+split model to predict model shares, accessibility indicators, including
+spatial availability, can be used to investigate changes to the
+accessibility landscape. A similar logic can be applied for destination
+choice. Our empirical example presented a snapshot of this, and in
+future research it may be interesting to investigate changes in spatial
+availability \emph{between} policy interventions. The expansion of
+Madrid's LEZ to the ring contained by the M-30 presents an excellent
+opportunity to do so. Given the intuitive interpretation of spatial
+availability scores as fractions of opportunities from the total,
+relative and absolute changes in the spatial availability landscape can
+be assessed, thus helping to evaluate the implications of policy
+interventions.
+
+Our empirical example dealt with differences in travel by mode only, but
 it is possible to think of the intersection between mode of travel and
@@
 question of their impact on disadvantaged populations who have become
-mobility-restricted increasingly come to the fore {[}40,57,58{]}.
-
-\hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
-
-This research was funded by the Canada Graduate Scholarship - Doctoral
-Program (CGS D) provided by the Social Sciences and Humanities Research
-Council (SSHRC) and Project Mobilizing Justice, also supported by SSHRC.
-The funders had no role in study design, data collection and analysis,
-decision to publish, or preparation of the manuscript.
-
-All work is fully-reproducible and available within this
-\href{https://github.com/soukhova/Multimodal-spatial-availability}{GitHub
-repository}.
+mobility-restricted increasingly come to the fore {[}51,69,70{]}.
+
+It is worth reiterating that while the focus of this work was on a
+multimodal extension of spatial availability, the consideration of
+multiple modes is just one case of heterogeneous populations (i.e.,
+travel by different modes). The method itself can easily accommodate
+other forms of population or opportunity heterogeneity, for example:
+variations in rich and poor, young and old, types of opportunities. Like
+accessibility measures before it, the potential application of spatial
+availability measure are as numerous as potential study contexts.
+
+To close, spatial availability is a singly-constrained accessibility
+measure that considers competition, a way of accounting for opportunity
+congestion. Spatial availability considers competition by allocating
+opportunities (the subject of the single-constraint) to zones based on
+zonal proportions. If the demand, supply and travel to opportunities is
+spatially uniform, then competition is uniform and spatial availability
+will reflect this. It has been suggested that opportunity congestion can
+be seen on a continuum, but how it is to be considered within
+accessibility research is an ongoing subject of exploration
+{[}14,17,18{]}. This is the context in which we present spatial
+availability and the multimodal extension.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-shenLocationCharacteristicsInnercity1998}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-paezDemand2019}{}}%
 \CSLLeftMargin{12. }%
-\CSLRightInline{Shen Q. Location characteristics of inner-city
-neighborhoods and employment accessibility of low-wage workers. Environ
-Plann B. 1998;25: 345--365.
-doi:\href{https://doi.org/10.1068/b250345}{10.1068/b250345}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-luoMeasuresSpatialAccessibility2003}{}}%
-\CSLLeftMargin{13. }%
-\CSLRightInline{Luo W, Wang F. Measures of spatial accessibility to
-health care in a {GIS} environment: Synthesis and a case study in the
-chicago region. Environ Plann B Plann Des. 2003;30: 865--884.
-doi:\href{https://doi.org/10.1068/b29120}{10.1068/b29120}}
-
-\leavevmode\vadjust pre{\hypertarget{ref-paezDemand2019}{}}%
-\CSLLeftMargin{14. }%
 \CSLRightInline{Paez A, Higgins CD, Vivona SF. Demand and level of
@@
 doi:\href{https://doi.org/10.1371/journal.pone.0218773}{10.1371/journal.pone.0218773}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-wang2SFCAI2SFCAIntegration2021}{}}%
+\CSLLeftMargin{13. }%
+\CSLRightInline{Wang F. From 2SFCA to i2SFCA: Integration, derivation
+and validation. International Journal of Geographical Information
+Science. 2021;35: 628--638.
+doi:\href{https://doi.org/10.1080/13658816.2020.1811868}{10.1080/13658816.2020.1811868}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-merlinDoesCompetitionMatter2017}{}}%
+\CSLLeftMargin{14. }%
+\CSLRightInline{Merlin LA, Hu L. Does competition matter in measures of
+job accessibility? Explaining employment in los angeles. Journal of
+Transport Geography. 2017;64: 77--88.
+doi:\href{https://doi.org/10.1016/j.jtrangeo.2017.08.009}{10.1016/j.jtrangeo.2017.08.009}}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-linUsingTransportationProblem2023}{}}%
+\CSLLeftMargin{16. }%
+\CSLRightInline{Lin J, Cromley G. Using the transportation problem to
+build a congestion/threshold constrained spatial accessibility model.
+Journal of Transport Geography. 2023;112: 103691.
+doi:\href{https://doi.org/10.1016/j.jtrangeo.2023.103691}{10.1016/j.jtrangeo.2023.103691}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-vanweeAccessibilityMeasuresCompetition2001}{}}%
+\CSLLeftMargin{17. }%
+\CSLRightInline{Van Wee B, Hagoort M, Annema JA. Accessibility measures
+with competition. Journal of Transport Geography. 2001;9: 199--208.
+doi:\href{https://doi.org/10.1016/S0966-6923(01)00010-2}{10.1016/S0966-6923(01)00010-2}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-kelobonyeMeasuringAccessibilitySpatial2020a}{}}%
+\CSLLeftMargin{18. }%
+\CSLRightInline{Kelobonye K, Zhou H, McCarney G, Xia J(Cecilia).
+Measuring the accessibility and spatial equity of urban services under
+competition using the cumulative opportunities measure. Journal of
+Transport Geography. 2020;85: 102706.
+doi:\href{https://doi.org/10.1016/j.jtrangeo.2020.102706}{10.1016/j.jtrangeo.2020.102706}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-cuiSpatialAccessPublic2020}{}}%
+\CSLLeftMargin{19. }%
+\CSLRightInline{Cui B, Boisjoly G, Wasfi R, Orpana H, Manaugh K, Buliung
+R, et al. Spatial access by public transport and likelihood of
+healthcare consultations at hospitals. {TRANSPORTATION} {RESEARCH}
+{RECORD}. 2020;2674: 188--198.
+doi:\href{https://doi.org/10.1177/0361198120952793}{10.1177/0361198120952793}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-maoMeasuringSpatialAccessibility2013}{}}%
+\CSLLeftMargin{20. }%
+\CSLRightInline{Mao L, Nekorchuk D. Measuring spatial accessibility to
+healthcare for populations with multiple transportation modes. Health \&
+Place. 2013;24: 115--122.
+doi:\href{https://doi.org/10.1016/j.healthplace.2013.08.008}{10.1016/j.healthplace.2013.08.008}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-josephMeasuringPotentialPhysical1982}{}}%
+\CSLLeftMargin{21. }%
+\CSLRightInline{Joseph AE, Bantock PR. Measuring potential physical
+accessibility to general practitioners in rural areas: A method and case
+study. Social Science \& Medicine. 1982;16: 85--90.
+doi:\href{https://doi.org/10.1016/0277-9536(82)90428-2}{10.1016/0277-9536(82)90428-2}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-higgsModellingSpatialAccess2017}{}}%
+\CSLLeftMargin{22. }%
+\CSLRightInline{Higgs G, Zahnow R, Corcoran J, Langford M, Fry R.
+Modelling spatial access to general practitioner surgeries: Does public
+transport availability matter? {JOURNAL} {OF} {TRANSPORT} \& {HEALTH}.
+2017;6: 143--154.
+doi:\href{https://doi.org/10.1016/j.jth.2017.05.361}{10.1016/j.jth.2017.05.361}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-weibullAxiomaticApproachMeasurement1976}{}}%
+\CSLLeftMargin{23. }%
+\CSLRightInline{Weibull JW. An axiomatic approach to the measurement of
+accessibility. Regional Science and Urban Economics. 1976;6: 357--379.
+doi:\href{https://doi.org/10.1016/0166-0462(76)90031-4}{10.1016/0166-0462(76)90031-4}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-shenLocationCharacteristicsInnercity1998}{}}%
+\CSLLeftMargin{24. }%
+\CSLRightInline{Shen Q. Location characteristics of inner-city
+neighborhoods and employment accessibility of low-wage workers. Environ
+Plann B. 1998;25: 345--365.
+doi:\href{https://doi.org/10.1068/b250345}{10.1068/b250345}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-liMeasuringMultiactivitiesAccessibility2024a}{}}%
+\CSLLeftMargin{25. }%
+\CSLRightInline{Li C, Wang J. Measuring multi-activities accessibility
+and equity with accessibility-oriented development strategies.
+Transportation Research Part D: Transport and Environment. 2024;126:
+104035.
+doi:\href{https://doi.org/10.1016/j.trd.2023.104035}{10.1016/j.trd.2023.104035}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-wuWillOpeningCommunity2020}{}}%
+\CSLLeftMargin{26. }%
+\CSLRightInline{Wu J, Chen H, Wang H, He Q, Zhou K. Will the opening
+community policy improve the equity of green accessibility and in what
+ways? --- response based on a 2-step floating catchment area method and
+genetic algorithm. Journal of Cleaner Production. 2020;263: 121454.
+doi:\href{https://doi.org/10.1016/j.jclepro.2020.121454}{10.1016/j.jclepro.2020.121454}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-chenEnhancingTwoStepFloating2019}{}}%
+\CSLLeftMargin{27. }%
+\CSLRightInline{Chen X. Enhancing the two-step floating catchment area
+model for community food access mapping: Case of the supplemental
+nutrition assistance program. The Professional Geographer. 2019;71:
+668--680.
+doi:\href{https://doi.org/10.1080/00330124.2019.1578978}{10.1080/00330124.2019.1578978}}
+
+\leavevmode\vadjust pre{\hypertarget{ref-luoMeasuresSpatialAccessibility2003}{}}%
+\CSLLeftMargin{28. }%
+\CSLRightInline{Luo W, Wang F. Measures of spatial accessibility to
+health care in a {GIS} environment: Synthesis and a case study in the
+chicago region. Environ Plann B Plann Des. 2003;30: 865--884.
+doi:\href{https://doi.org/10.1068/b29120}{10.1068/b29120}}
+
 \leavevmode\vadjust pre{\hypertarget{ref-paezHealth2010}{}}%
-\CSLLeftMargin{16. }%
+\CSLLeftMargin{29. }%
 \CSLRightInline{Páez A, Mercado R, Farber S, Morency C, Roorda M.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-moniruzzamanMode2013}{}}%
-\CSLLeftMargin{17. }%
+\CSLLeftMargin{30. }%
 \CSLRightInline{Moniruzzaman M, Páez A, Nurul Habib KM, Morency C. Mode
@@
 \leavevmode\vadjust pre{\hypertarget{ref-reyesWalking2014}{}}%
-\CSLLeftMargin{18. }%
+\CSLLeftMargin{31. }%
 \CSLRightInline{Reyes M, Paez A, Morency C. Walking accessibility to
@@
 \leavevmode\vadjust pre{\hypertarget{ref-paezJobs2013}{}}%
-\CSLLeftMargin{19. }%
+\CSLLeftMargin{32. }%
 \CSLRightInline{Páez A, Farber S, Mercado R, Roorda M, Morency C. Jobs
@@
 \leavevmode\vadjust pre{\hypertarget{ref-wuUnifyingAccess2020}{}}%
-\CSLLeftMargin{20. }%
+\CSLLeftMargin{33. }%
 \CSLRightInline{Wu H, Levinson D. Unifying access. Transportation
@@
 \leavevmode\vadjust pre{\hypertarget{ref-tahmasbiMultimodalAccessibilitybasedEquity2019}{}}%
-\CSLLeftMargin{21. }%
+\CSLLeftMargin{34. }%
 \CSLRightInline{Tahmasbi B, Mansourianfar MH, Haghshenas H, Kim I.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-lunkeModalAccessibilityDisparities2022}{}}%
-\CSLLeftMargin{22. }%
+\CSLLeftMargin{35. }%
 \CSLRightInline{Lunke EB. Modal accessibility disparities and transport
@@
 \leavevmode\vadjust pre{\hypertarget{ref-paezRelative2010}{}}%
-\CSLLeftMargin{23. }%
+\CSLLeftMargin{36. }%
 \CSLRightInline{Paez A, Mercado RG, Farber S, Morency C, Roorda M.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-campbell2019accessibility}{}}%
-\CSLLeftMargin{24. }%
+\CSLLeftMargin{37. }%
 \CSLRightInline{Campbell KB, Rising JA, Klopp JM, Mbilo JM.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-maharjanSpatialEquityModal2022}{}}%
-\CSLLeftMargin{25. }%
+\CSLLeftMargin{38. }%
 \CSLRightInline{Maharjan S, Tilahun N, Ermagun A. Spatial equity of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-grengsJobAccessibilityModal2010}{}}%
-\CSLLeftMargin{26. }%
+\CSLLeftMargin{39. }%
 \CSLRightInline{Grengs J. Job accessibility and the modal mismatch in
@@
 \leavevmode\vadjust pre{\hypertarget{ref-kawabataJobAccessibilityIndicator2006}{}}%
-\CSLLeftMargin{27. }%
+\CSLLeftMargin{40. }%
 \CSLRightInline{Kawabata M, Shen Q. Job accessibility as an indicator of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-kwokUseModalAccessibility2004}{}}%
-\CSLLeftMargin{28. }%
+\CSLLeftMargin{41. }%
 \CSLRightInline{Kwok RCW, Yeh AGO. The use of modal accessibility gap as
@@
 \leavevmode\vadjust pre{\hypertarget{ref-morrisAccessibilityIndicatorsTransport1979}{}}%
-\CSLLeftMargin{29. }%
+\CSLLeftMargin{42. }%
 \CSLRightInline{Morris JM, Dumble PL, Wigan MR. Accessibility indicators
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-weibullAxiomaticApproachMeasurement1976}{}}%
-\CSLLeftMargin{30. }%
-\CSLRightInline{Weibull JW. An axiomatic approach to the measurement of
-accessibility. Regional Science and Urban Economics. 1976;6: 357--379.
-doi:\href{https://doi.org/10.1016/0166-0462(76)90031-4}{10.1016/0166-0462(76)90031-4}}
-
 \leavevmode\vadjust pre{\hypertarget{ref-liottaPlanning2020}{}}%
-\CSLLeftMargin{31. }%
+\CSLLeftMargin{43. }%
 \CSLRightInline{Liotta C, Kervinio Y, Levrel H, Tardieu L. Planning for
@@
 \leavevmode\vadjust pre{\hypertarget{ref-OECDFrameworks2013}{}}%
-\CSLLeftMargin{32. }%
+\CSLLeftMargin{44. }%
 \CSLRightInline{OECD. Frameworks and sector policies for urban
@@
 \leavevmode\vadjust pre{\hypertarget{ref-laraSpace2015}{}}%
-\CSLLeftMargin{33. }%
+\CSLLeftMargin{45. }%
 \CSLRightInline{Lara-Valencia F, García-Pérez H. Space for equity:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-martensFair2021}{}}%
-\CSLLeftMargin{34. }%
+\CSLLeftMargin{46. }%
 \CSLRightInline{Martens K, Golub A. A fair distribution of
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-josephMeasuringPotentialPhysical1982}{}}%
-\CSLLeftMargin{35. }%
-\CSLRightInline{Joseph AE, Bantock PR. Measuring potential physical
-accessibility to general practitioners in rural areas: A method and case
-study. Social Science \& Medicine. 1982;16: 85--90.
-doi:\href{https://doi.org/10.1016/0277-9536(82)90428-2}{10.1016/0277-9536(82)90428-2}}
-
 \leavevmode\vadjust pre{\hypertarget{ref-taoInvestigatingImpactsPublic2020a}{}}%
-\CSLLeftMargin{36. }%
+\CSLLeftMargin{47. }%
 \CSLRightInline{Tao Z, Zhou J, Lin X, Chao H, Li G. Investigating the
@@
 \leavevmode\vadjust pre{\hypertarget{ref-wilsonFamilty1971}{}}%
-\CSLLeftMargin{37. }%
+\CSLLeftMargin{48. }%
 \CSLRightInline{Wilson AG. A family of spatial interaction models, and
@@
 \leavevmode\vadjust pre{\hypertarget{ref-carpentieriMultimodalAccessibilityPrimary2020}{}}%
-\CSLLeftMargin{38. }%
+\CSLLeftMargin{49. }%
 \CSLRightInline{Carpentieri G, Guida C, Masoumi HE. Multimodal
@@
 \leavevmode\vadjust pre{\hypertarget{ref-margaryanLowEmissionZones2021}{}}%
-\CSLLeftMargin{39. }%
+\CSLLeftMargin{50. }%
 \CSLRightInline{Margaryan S. Low emission zones and population health.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-verbeekJustManagementUrban2022}{}}%
-\CSLLeftMargin{40. }%
+\CSLLeftMargin{51. }%
 \CSLRightInline{Verbeek T, Hincks S. The {``just''} management of urban
@@
 \leavevmode\vadjust pre{\hypertarget{ref-tarrinoortizAnalyzingImpactLow2022}{}}%
-\CSLLeftMargin{41. }%
+\CSLLeftMargin{52. }%
 \CSLRightInline{Tarriño-Ortiz J, Gómez J, Soria-Lara JA, Vassallo JM.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-comunidaddemadridResultadosEDM20182020}{}}%
-\CSLLeftMargin{42. }%
+\CSLLeftMargin{53. }%
 \CSLRightInline{Comunidad de Madrid. Resultados de la {EDM} 2018 - Datos
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-comunidaddemadridzoniEDM20182020}{}}%
+\CSLLeftMargin{54. }%
+\CSLRightInline{Comunidad de Madrid. Zonificación de transporte ZT208 de
+la EDM2018 - Datos Abiertos. 2020 {[}cited 1 Dec 2023{]}. Available:
+\url{https://datos.crtm.es/datasets/zonificacionzt208}}
+
 \leavevmode\vadjust pre{\hypertarget{ref-lopez_2017_spatial}{}}%
-\CSLLeftMargin{43. }%
+\CSLLeftMargin{55. }%
 \CSLRightInline{Lopez FA, Paez A. Spatial clustering of high-tech
@@
 \leavevmode\vadjust pre{\hypertarget{ref-horbachov_theoretical_2018}{}}%
-\CSLLeftMargin{44. }%
+\CSLLeftMargin{56. }%
 \CSLRightInline{Horbachov P, Svichynskyi S. Theoretical substantiation
@@
 \leavevmode\vadjust pre{\hypertarget{ref-batista_estimation_2019}{}}%
-\CSLLeftMargin{45. }%
+\CSLLeftMargin{57. }%
 \CSLRightInline{Batista SFA, Leclercq L, Geroliminis N. Estimation of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-fitdistrplus_2015}{}}%
-\CSLLeftMargin{46. }%
+\CSLLeftMargin{58. }%
 \CSLRightInline{Delignette-Muller ML, Dutang C. {fitdistrplus}: An {R}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-reggianiAccessibilityImpedanceForms2011}{}}%
-\CSLLeftMargin{47. }%
+\CSLLeftMargin{59. }%
 \CSLRightInline{Reggiani A, Bucci P, Russo G. Accessibility and
@@
 \leavevmode\vadjust pre{\hypertarget{ref-soukhovTTS2016RDataSet2023}{}}%
-\CSLLeftMargin{48. }%
+\CSLLeftMargin{60. }%
 \CSLRightInline{Soukhov A, Páez A. {TTS}2016R: A data set to study
@@
 \leavevmode\vadjust pre{\hypertarget{ref-careyPriciples1858}{}}%
-\CSLLeftMargin{49. }%
+\CSLLeftMargin{61. }%
 \CSLRightInline{Carey HC. Principles of social science. Philadelphia:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-ravensteinLaws1889}{}}%
-\CSLLeftMargin{50. }%
+\CSLLeftMargin{62. }%
 \CSLRightInline{Ravenstein EG. The laws of migration. Journal of the
@@
 \leavevmode\vadjust pre{\hypertarget{ref-reillyMethods1929}{}}%
-\CSLLeftMargin{51. }%
+\CSLLeftMargin{63. }%
 \CSLRightInline{Reilly WJ. Methods for the study of retail
@@
 \leavevmode\vadjust pre{\hypertarget{ref-stewartInverse1941}{}}%
-\CSLLeftMargin{52. }%
+\CSLLeftMargin{64. }%
 \CSLRightInline{Stewart JQ. An inverse distance variation for certain
@@
 \leavevmode\vadjust pre{\hypertarget{ref-stewartSuggested1947}{}}%
-\CSLLeftMargin{53. }%
+\CSLLeftMargin{65. }%
 \CSLRightInline{Stewart JQ. Suggested principles of ''social physics''.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-stewartDemographic1948}{}}%
-\CSLLeftMargin{54. }%
+\CSLLeftMargin{66. }%
 \CSLRightInline{Stewart JQ. Demographic gravitation: Evidence and
@@
 \leavevmode\vadjust pre{\hypertarget{ref-zipfHypothesis1946}{}}%
-\CSLLeftMargin{55. }%
+\CSLLeftMargin{67. }%
 \CSLRightInline{Zipf GK. The p 1 p 2/d hypothesis: The case of railway
@@
 \leavevmode\vadjust pre{\hypertarget{ref-zipfIntercity1946}{}}%
-\CSLLeftMargin{56. }%
+\CSLLeftMargin{68. }%
 \CSLRightInline{Zipf GK. The P1P2/d hypothesis: On the intercity
@@
 \leavevmode\vadjust pre{\hypertarget{ref-devrijNooneVisitsMe2022}{}}%
-\CSLLeftMargin{57. }%
+\CSLLeftMargin{69. }%
 \CSLRightInline{De Vrij E, Vanoutrive T. {``No-one visits me anymore''}:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-liottaWhatDrivesInequalities2023}{}}%
-\CSLLeftMargin{58. }%
+\CSLLeftMargin{70. }%
 \CSLRightInline{Liotta C. What drives inequalities in low emission

### 2024-01-25 Journal-manuscript/Journal-manuscriptV0.tex
R2 Journal-manuscript revision. For AP review.

--- 2023-11-10 Journal-manuscript/Journal-manuscript.tex
@@
 and \ref{fig:Fig3} show the number of workers and the distribution of
-full-time jobs in the City of Madrid by TAZ. The light grey boundary
-represents the LEZ Centro in effect in 2017. The survey was conducted
-after the introduction of LEZ Centro. The dark grey boundary represents
-the LEZ planned for the boundaries of the M-30 highway and is present in
-the plots as a spatial reference for areas in proximity to the LEZ
-Centro.
+full-time jobs in the City of Madrid by TAZ. The TAZ shapefiles are
+available from the Community of Madrid open data portal {[}43{]}. The
+light grey boundary represents the LEZ Centro in effect in 2017. The
+survey was conducted after the introduction of LEZ Centro. The dark grey
+boundary represents the LEZ planned for the boundaries of the M-30
+highway and is present in the plots as a spatial reference for areas in
+proximity to the LEZ Centro.
 
@@
 bins. This distribution is then used to estimate the parameters of a
-function for the travel impedance {[}as done in 43,44,45{]}. To fit the
+function for the travel impedance {[}as done in 44,45,46{]}. To fit the
 impedance functions, we use the Maximum likelihood estimation and the
 Nelder-Mead method for direct optimization available within the R
-\{fitdistrplus\} package {[}46{]}. Based on goodness-of-fit criteria and
+\{fitdistrplus\} package {[}47{]}. Based on goodness-of-fit criteria and
 associated diagnostics, the gamma and log-normal probability density
@@
 non-motorised modes respectively. The selection of functional forms
-aligns with empirical examples in other regions {[}15,47,48{]}. The
+aligns with empirical examples in other regions {[}15,48,49{]}. The
 shape and rate parameters for the gamma functions (motorised modes) are
@@
 
-\caption{\label{fig:Fig5} Spatial availability of jobs per origin and mode $V_i^m$ in Madrid. Grey TAZs have no population. Ranges of values in the legend are quintiles.}\label{fig:SA-m-plot}
+\caption{\label{fig:Fig5} Spatial availability of jobs per origin and mode $V_i^m$ in Madrid at the level of TAZ.
+Grey TAZ have no population.
+Ranges of values in the legend are quintiles.
+The TAZ shapefile is available from the Community of Madrid open data portal.}\label{fig:SA-m-plot}
 \end{figure}
@@
 Aggregations of opportunities embody principles of gravitational/spatial
-interaction modelling that date back to at least H.C. Carey {[}49{]},
+interaction modelling that date back to at least H.C. Carey {[}50{]},
 and are part of a line of research that includes the work of Ravenstein
-{[}50{]}, Reilly {[}51{]}, Stewart {[}52--54{]}, Zipf {[}55,56{]},
+{[}51{]}, Reilly {[}52{]}, Stewart {[}53--55{]}, Zipf {[}56,57{]},
 Wilson {[}37{]}, and many others. In this way, \(S_i^m\), \(a_i^m\), and
@@
 question of their impact on disadvantaged populations who have become
-mobility-restricted increasingly come to the fore {[}40,57,58{]}.
-
-\hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
-
-This research was funded by the Canada Graduate Scholarship - Doctoral
-Program (CGS D) provided by the Social Sciences and Humanities Research
-Council (SSHRC) and Project Mobilizing Justice, also supported by SSHRC.
-The funders had no role in study design, data collection and analysis,
-decision to publish, or preparation of the manuscript.
-
-All work is fully-reproducible and available within this
-\href{https://github.com/soukhova/Multimodal-spatial-availability}{GitHub
-repository}.
+mobility-restricted increasingly come to the fore {[}40,58,59{]}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-comunidaddemadridzoniEDM20182020}{}}%
+\CSLLeftMargin{43. }%
+\CSLRightInline{Comunidad de Madrid. Zonificación de transporte ZT208 de
+la EDM2018 - Datos Abiertos. 2020 {[}cited 1 Dec 2023{]}. Available:
+\url{https://datos.crtm.es/datasets/zonificacionzt208}}
+
 \leavevmode\vadjust pre{\hypertarget{ref-lopez_2017_spatial}{}}%
-\CSLLeftMargin{43. }%
+\CSLLeftMargin{44. }%
 \CSLRightInline{Lopez FA, Paez A. Spatial clustering of high-tech
@@
 \leavevmode\vadjust pre{\hypertarget{ref-horbachov_theoretical_2018}{}}%
-\CSLLeftMargin{44. }%
+\CSLLeftMargin{45. }%
 \CSLRightInline{Horbachov P, Svichynskyi S. Theoretical substantiation
@@
 \leavevmode\vadjust pre{\hypertarget{ref-batista_estimation_2019}{}}%
-\CSLLeftMargin{45. }%
+\CSLLeftMargin{46. }%
 \CSLRightInline{Batista SFA, Leclercq L, Geroliminis N. Estimation of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-fitdistrplus_2015}{}}%
-\CSLLeftMargin{46. }%
+\CSLLeftMargin{47. }%
 \CSLRightInline{Delignette-Muller ML, Dutang C. {fitdistrplus}: An {R}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-reggianiAccessibilityImpedanceForms2011}{}}%
-\CSLLeftMargin{47. }%
+\CSLLeftMargin{48. }%
 \CSLRightInline{Reggiani A, Bucci P, Russo G. Accessibility and
@@
 \leavevmode\vadjust pre{\hypertarget{ref-soukhovTTS2016RDataSet2023}{}}%
-\CSLLeftMargin{48. }%
+\CSLLeftMargin{49. }%
 \CSLRightInline{Soukhov A, Páez A. {TTS}2016R: A data set to study
@@
 \leavevmode\vadjust pre{\hypertarget{ref-careyPriciples1858}{}}%
-\CSLLeftMargin{49. }%
+\CSLLeftMargin{50. }%
 \CSLRightInline{Carey HC. Principles of social science. Philadelphia:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-ravensteinLaws1889}{}}%
-\CSLLeftMargin{50. }%
+\CSLLeftMargin{51. }%
 \CSLRightInline{Ravenstein EG. The laws of migration. Journal of the
@@
 \leavevmode\vadjust pre{\hypertarget{ref-reillyMethods1929}{}}%
-\CSLLeftMargin{51. }%
+\CSLLeftMargin{52. }%
 \CSLRightInline{Reilly WJ. Methods for the study of retail
@@
 \leavevmode\vadjust pre{\hypertarget{ref-stewartInverse1941}{}}%
-\CSLLeftMargin{52. }%
+\CSLLeftMargin{53. }%
 \CSLRightInline{Stewart JQ. An inverse distance variation for certain
@@
 \leavevmode\vadjust pre{\hypertarget{ref-stewartSuggested1947}{}}%
-\CSLLeftMargin{53. }%
+\CSLLeftMargin{54. }%
 \CSLRightInline{Stewart JQ. Suggested principles of ''social physics''.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-stewartDemographic1948}{}}%
-\CSLLeftMargin{54. }%
+\CSLLeftMargin{55. }%
 \CSLRightInline{Stewart JQ. Demographic gravitation: Evidence and
@@
 \leavevmode\vadjust pre{\hypertarget{ref-zipfHypothesis1946}{}}%
-\CSLLeftMargin{55. }%
+\CSLLeftMargin{56. }%
 \CSLRightInline{Zipf GK. The p 1 p 2/d hypothesis: The case of railway
@@
 \leavevmode\vadjust pre{\hypertarget{ref-zipfIntercity1946}{}}%
-\CSLLeftMargin{56. }%
+\CSLLeftMargin{57. }%
 \CSLRightInline{Zipf GK. The P1P2/d hypothesis: On the intercity
@@
 \leavevmode\vadjust pre{\hypertarget{ref-devrijNooneVisitsMe2022}{}}%
-\CSLLeftMargin{57. }%
+\CSLLeftMargin{58. }%
 \CSLRightInline{De Vrij E, Vanoutrive T. {``No-one visits me anymore''}:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-liottaWhatDrivesInequalities2023}{}}%
-\CSLLeftMargin{58. }%
+\CSLLeftMargin{59. }%
 \CSLRightInline{Liotta C. What drives inequalities in low emission

### 2024-01-29 Journal-manuscript/Editorial correspondance/Response-to-reviewers_R2.Rmd
Edits to response to reviewer

@@
 
-\textcolor{blue}{We added additional detail to the "Introduction" (second paragraph) to further clarify.
-Specifically, that opportunity congestion can be seen on continnum and that accessibility methods that consider competition have been increasingly applied to analyse access to opportunity types that are not inexhaustible such as educational facilities [@kelobonyeMeasuringAccessibilitySpatial2020a; @liMeasuringMultiactivitiesAccessibility2024a], greenspace and recreation [@wuWillOpeningCommunity2020; @liMeasuringMultiactivitiesAccessibility2024a], and shopping [@kelobonyeMeasuringAccessibilitySpatial2020a; @chenEnhancingTwoStepFloating2019].}
+\textcolor{blue}{Thank you, we appreciate your understanding.}
 
@@
 
-\textcolor{blue}{Interesting and thoughtful comment.
+\textcolor{blue}{This is a great question, and we think we know the answer.
+We interpreted your comment initially as to mean changes in accessibility, and we responded accordingly.
+Your example makes it clear that you had something else in mind.
+The answer to your question can be found in point 2 in the conclusion of Soukhov et al.} \begingroup\color{blue}[-@soukhovIntroducingSpatialAvailability2023]\endgroup \hspace{1em}\textcolor{blue}{ (p.
+25)  where it says "Shen-type accessibility can be used to compute the availability of jobs (the rate and the absolute values if the original definition is corrected), however, if the analyst is interested in internal values and secondary analysis of the results, spatial availability should be considered".
+The internal values referred to there are the estimates of trip length/duration/cost associated with the spatial availability of opportunities; the secondary analysis, in turn, referred to estimating the system-wide cost of competitive accessibility.
+To use your example, suppose that there is a change in the transportation system that is uniform: accessibility does not change but travel times are shorter for every mode.
+As Soukhov et al.} \begingroup\color{blue}[-@soukhovIntroducingSpatialAvailability2023]\endgroup \hspace{1em}\textcolor{blue}{demonstrate in the section \textit{Why does proportional allocation matter?} (p.
+15), a difference between spatial availability and unconstrained accessibility measures, is that spatial availability allocates the actual number of opportunities to each origin.
+As a result, the estimates of travel time can be used to quantify the system-wide impacts of spatial availability.
+In your hypothetical example, spatial availability would remain constant, but the system-wide cost of travel, as calculated from the intermediate values (i.e., the detailed table of opportunities allocated to origins by origin-destination pair), would decrease.}
+<!--
+Interesting and thoughtful comment.
 We have added further detail to the text to the "Discussion and Conclusions" (final paragraph) to further shed like on 'competition' and make spatial availability's use case more clear in the closure of the manuscript.
@@
 }
+-->
 
@@
 A key question is whether people can shift transport modes easily or not: indeed, it seems to be that your new accessibility measure, based on competition, does only make sense in case people cannot easily shift transport modes.
+
+\textcolor{blue}{Thank you for your comment.
+For the empirical example we use data collected at least one year \textit{after} the introduction of Madrid LEZ.
+We then make the (hopefully reasonable) assumption that all modal shifts caused by LEZ have already taken place.
+We could (if we had the data) calculate spatial availability using the mode shares \textit{before} the introduction of Madrid LEZ.
+Here, however, we are not conducting a before/after comparison of the policy, but only of the spatial availability already in the context of LEZ.
+In other words, our discussion is based not on a comparison over time, but a comparison over space.}
+
 You partially answer that question in the paragraph starting l46 by explaining why some populations, e.g.
@@
 For instance, you state that “restrictions to travel by car leave more spatially available opportunities for non-car-users” (l585).
+
+\textcolor{blue}{Our discussion of competition is not based on hypothetical modal shifts caused by LEZ, but rather on the comparison of spatial availability after LEZ over space.
+Fig.
+6 in the paper shows that bike and walk consistently have fewer jobs available than their share of users in the population.
+But, they receive a larger share \textit{within} the boundaries of Centro, where the number of cars that can enter is limited by the emissions policy.
+In other words, the restriction, which is what distinguishes Centro from the boundary of the M-30 or the whole region, seems to open up a window of opportunity for jobs to become available to cyclists and pedestrians.}
+
 But it seems to me that the LEZ aims at promoting a modal shift toward less polluting cars, public transport, or active modes rather than favoring public transport users?
+
+\textcolor{blue}{No question that promoting a modal shift is part of the policy.
+However, as we noted above, we assume that the data already incorporates said shift.
+Given the modal shares, our analysis indicates that the winning mode within the boundaries of LEZ is transit (Fig.
+6).}
+
 Relatedly, your measure does not make any difference between car users that can easily shift to public transport, and car users that cannot.
 
-\textcolor{blue}{Thank you for your comment, we've further amended the "Results" (first and second paragraph) to make the assumptions within and interpretation of the empirical example more clear.
-The fourth paragraph in the "Discussion and Conclusions" also touches on your comment.
-}
-\textcolor{blue}{There is no mode-shifting that occurs in the empirical example.
-This implementation of spatial availability assumes that all employment opportunities are available to the entire population.
-What impacts the proportional allocation is the relative travel cost (based on observed modal origin-destination travel behaviour from the travel survey) and the relative amount of the mode-using population residing in the zones.
-Fig 5 demonstrates the zonal multimodal spatial availability values based on the selected mode, observed home-to-work travel patterns, and competition to work.
-It is only a snapshot of travel behaviour as informed by the 2018 travel survey (i.e., the majority of cars do not enter the LEZ centro), that we use to demonstrate the multimodal spatial availability landscape.
-In this context, we discuss the impact that the travel behaviour induced by the LEZ centro has on the spatial availability landscape.
-}
-\textcolor{blue}{As written in "Discussion and Conclusions", accessibility indicators are not designed to work as modal split models.
-"In our empirical example we used data collected after the introduction of LEZ Centro.
-However, given a modal split model to predict model shares, accessibility indicators, including spatial availability, can be used to investigate changes to the accessibility landscape.
-A similar logic can be applied for destination choice."
-}
-\textcolor{blue}{I hope this is a comprehensive response (and manuscript amended) to address your comment.
-The empirical example in the manuscript does not deal with the change from one mode to another, it only looks at the snapshot of travel behaviour as recorded in the 2018 travel survey.
-Investigating mode-shift would be a different research motivation.
-}
+\textcolor{blue}{You are absolutely right.
+That said, the job of allocating users to various modes is not the job of our measure, but the job of a mode split or mode choice model.
+The scope of our measure is to allocate opportunities to users by mode.
+Given changes in the shares of different modes, the measure can be calculated to examine the spatial availability under those conditions.}
 
@@
 
-\textcolor{blue}{Again, thank you for your thoughtful comments through the revision process so far.}
+\textcolor{blue}{We wish to thank you for your thoughtful and constructive comments throughout the revision process so far.
+A sticking point is the shift in modes.
+In this respect, we would like to reiterate that our measure is not intended to do the things that modal split or mode choice models are supposed to do.
+Our empirical example is cross-sectional, and we believe bypasses the question of modal shifts while still giving useful information about the spatial distribution of opportunities and the ability of different modes to reach them.
+We would argue that the materials presented here are already an advance over existing methods--even in the case when accessibility remains constant but the cost of travel declines uniformly.
+Still, we wish to acknowledge that the use of multimodal spatial availability in over-time comparisons would benefit from being paired with a modal split or mode choice model, something that we are prevented from trying in our empirical example due to data limitations.}
+
+## References

### 2024-02-01 Journal-manuscript/Editorial correspondance/Response-to-reviewers_R2.Rmd
Revised manu draft 2- ready to submit back to reviewers

@@
 Here, however, we are not conducting a before/after comparison of the policy, but only of the spatial availability already in the context of LEZ.
-In other words, our discussion is based not on a comparison over time, but a comparison over space.}
+In other words, our discussion is based not on a comparison over time, but a comparison over space.
+We have further enhanced the results to make this interpretation more clear.
+}
 
@@
 6 in the paper shows that bike and walk consistently have fewer jobs available than their share of users in the population.
-But, they receive a larger share \textit{within} the boundaries of Centro, where the number of cars that can enter is limited by the emissions policy.
+But, they receive a larger share \textit{within} the boundaries of Centro, where the number of cars that can enter is limited by the LEZ policy.
 In other words, the restriction, which is what distinguishes Centro from the boundary of the M-30 or the whole region, seems to open up a window of opportunity for jobs to become available to cyclists and pedestrians.}
@@
 \textcolor{blue}{Thank you bringing these comments to our attention.
-We have changed the colours and slightly increased the size of the spatial borders for LEZ Centro and M-30 on all figures (now the borders are Pink and Purple instead of light grey and medium grey) to improve visibility.
+We have changed the colours and slightly increased the size of the spatial borders for LEZ Centro and M-30 on all figures (now the borders are Pink and Purple instead of Light and Medium Grey) to improve visibility.
 We have also corrected the identified typo.
-Additionally, we have improved phrasing and corrected minor grammatical errors throughout the manuscript.
+Additionally, we have improved phrasing and corrected grammatical errors throughout the manuscript.
 }

### 2024-02-01 Journal-manuscript/Journal-manuscript.Rmd
Revised manu draft 2- ready to submit back to reviewers

@@
 abstract: |
-  Recent research has aimed to address the way opportunities are counted in accessibility analysis.
-In conventional accessibility measures, opportunities are often multiply counted, which leads to values of accessibility that are difficult to interpret.
-Constraining the calculations to match a known quantity ensures that the measurements sum up to a predetermined quantity (i.e., the total number of opportunities), and so each value can be meaningfully related to this total.
-A recent effort is spatial availability, a singly-constrained accessibility measure.
-In this paper we extend spatial availability for use in the case of multiple modes or, more generally, heterogeneous population segments with distinct travel behaviors.
+  Place-based accessibility measures communicate the potential interaction with opportunities at a zone that populations can access.
+Recent research has explored the implications of how opportunities are counted by different accessibility methods.
+In conventional measures, opportunities are multiply counted if more than one zone offers access to the same opportunity.
+This multi-count of opportunities leads to values of accessibility that are difficult to interpret.
+A possible solution to enhance the meaning-making of accessibility results is by constraining the calculations to match a known quantity.
+This ensures all zonal values sum up to a predetermined quantity (i.e., the total number of opportunities).
+In this way, each value can be meaningfully related to this total.
+A recent effort that implements this solution is spatial availability, a singly-constrained accessibility measure.
+In this paper, we extend spatial availability for use in the case of multiple modes or more generally, heterogeneous population segments with distinct travel behaviors.
 After deriving a multimodal version of spatial availability, we proceed to illustrate its features using a synthetic example.
-Next, we apply it to an empirical example of low emission zones in Madrid, Spain.
-We conclude the paper with suggestions for future research and its use in evaluating policy interventions.
+We then apply it to an empirical example of low emission zones in Madrid, Spain.
+We conclude with suggestions for future research and its use in evaluating policy interventions.
 
@@
 
-Accessibility is a key concept in the analysis of land use and transportation systems [e.g., @hansenHowAccessibilityShapes1959; @geursAccessibility2004; @paezPositive2012] and is coming of age from the perspective of planning research [see _inter alia_, @handyAccessibility2020; @levinsonTransportAccessManual2020; @siddiqToolsTradeAssessing2021; @yanAccessibilityBasedPlanningAddressing2021; @elGeneidyMaking2022].
-Beginning with the work of Hansen [@hansenHowAccessibilityShapes1959], accessibility measures have been widely used to evaluate the efficiency of transportation systems when combined with the distribution of populations and opportunities in space [@shiLiterature2020].
+Accessibility is a key concept in the analysis of land use and transportation systems [@hansenHowAccessibilityShapes1959; @geursAccessibility2004; @paezPositive2012] and is coming of age from the perspective of planning research [see _inter alia_ @handyAccessibility2020; @levinsonTransportAccessManual2020; @siddiqToolsTradeAssessing2021; @yanAccessibilityBasedPlanningAddressing2021; @elGeneidyMaking2022].
+Beginning with the work of Hansen (1959) [@hansenHowAccessibilityShapes1959], accessibility measures have been widely used to evaluate the efficiency of transportation systems when combined with the distribution of populations and opportunities in space [@shiLiterature2020].
 In this way, accessibility is a holistic measure of spatial systems' ease of reaching population-relevant destinations [@handyMeasuring1997; @kwanSpace1998].
@@
 
-The most common form of accessibility measure is based the gravity model [e.g.
-@hansenHowAccessibilityShapes1959]: these measures sum "weighted" opportunities around a focal point (i.e., a potential origin), based on how expensive it is to reach them.
-Recent research in accessibility analysis has explored the implications of methods by which opportunities are summed, especially in the context of competitive accessibility [e.g., @paezDemand2019, @soukhovIntroducingSpatialAvailability2023; @wang2SFCAI2SFCAIntegration2021; @merlinDoesCompetitionMatter2017].
-In a typical Hansen-type accessibility measure, the sums around origins are not constrained.
-This means that the same opportunity can enter the sum for different origins; Counting the same opportunity multiple times treats the opportunity as if it was inexhaustible (or non-rival [@linUsingTransportationProblem2023]).
-However, this conceptualisation may not reflect reality as opportunities, in general, are not inexhaustible.
+The most common form of accessibility measure is based on the gravity model e.g.
+the Hansen-type measure [@hansenHowAccessibilityShapes1959]: these measures sum "weighted" opportunities around a focal point (i.e., a potential origin), based on how expensive it is to reach them.
+Recent research in accessibility analysis has explored the implications of methods by which opportunities are summed, especially in the context of competitive accessibility [@paezDemand2019, @soukhovIntroducingSpatialAvailability2023; @wang2SFCAI2SFCAIntegration2021; @merlinDoesCompetitionMatter2017].
+In a typical Hansen-type accessibility measure, the sums around origins are not constrained: the same opportunity can enter the sum for different origins.
+Counting the same opportunity multiple times treats it as it was inexhaustible (or non-rival [@linUsingTransportationProblem2023]).
+However, this conceptualisation may not reflect reality, as opportunities are often more exclusive than not.
 Some opportunities are by definition exclusive: a typical example is employment, once a job is taken up by someone in the population, the same job is no longer available for any other person to take [@vanweeAccessibilityMeasuresCompetition2001; @merlinDoesCompetitionMatter2017].
-More generally, it could be conceptualised that all opportunities are subject to some amount of congestion or capacity-constraint: multiple people use a specific opportunity at a given time and the more people who do, the more congested the opportunity.
-Congestion can be seen on a continuum, and debate on how congestion within accessibility should be considered is ongoing [@vanweeAccessibilityMeasuresCompetition2001; @merlinDoesCompetitionMatter2017; @kelobonyeMeasuringAccessibilitySpatial2020a].
-Nonetheless, competition has been widely considered for traditionally exclusive opportunities types such as employment and healthcare opportunities [@cuiSpatialAccessPublic2020; @maoMeasuringSpatialAccessibility2013; @josephMeasuringPotentialPhysical1982; @higgsModellingSpatialAccess2017; @vanweeAccessibilityMeasuresCompetition2001; @merlinDoesCompetitionMatter2017; @weibullAxiomaticApproachMeasurement1976; @josephMeasuringPotentialPhysical1982; @shenLocationCharacteristicsInnercity1998] and more recently educational facilities [@kelobonyeMeasuringAccessibilitySpatial2020a; @liMeasuringMultiactivitiesAccessibility2024a], and less exclusive amenities-types such as greenspace and recreation [@wuWillOpeningCommunity2020; @liMeasuringMultiactivitiesAccessibility2024a], and shopping [@kelobonyeMeasuringAccessibilitySpatial2020a; @chenEnhancingTwoStepFloating2019].
-
-The consideration of congestion in accessibility measures was the motivation for the development of approaches that consider competition [@weibullAxiomaticApproachMeasurement1976; @josephMeasuringPotentialPhysical1982; @shenLocationCharacteristicsInnercity1998] and are popularly applied as floating catchment area methods [e.g.
-@luoMeasuresSpatialAccessibility2003].
-While these approaches purport to account for congestion, Paez et al.
-[@paezDemand2019] demonstrates that they do not solve the issue of multiple counting of opportunities in general, thus leading to biases in the calculation of total demand and supply.
+More generally, it could be conceptualised that all opportunities are somewhat exclusive so they are subject to some amount of congestion or capacity-constraint: multiple people use a specific opportunity at a given time and the more people who do, the more congested the opportunity.
+Congestion can be seen on a continuum, and debate on how congestion within accessibility measures can be considered is ongoing [@vanweeAccessibilityMeasuresCompetition2001; @merlinDoesCompetitionMatter2017; @kelobonyeMeasuringAccessibilitySpatial2020a].
+Nonetheless, competition has been widely considered for traditionally exclusive opportunities types such as employment and healthcare opportunities [@cuiSpatialAccessPublic2020; @maoMeasuringSpatialAccessibility2013; @josephMeasuringPotentialPhysical1982; @higgsModellingSpatialAccess2017; @vanweeAccessibilityMeasuresCompetition2001; @merlinDoesCompetitionMatter2017; @weibullAxiomaticApproachMeasurement1976; @josephMeasuringPotentialPhysical1982; @shenLocationCharacteristicsInnercity1998] and more recently for educational facilities [@kelobonyeMeasuringAccessibilitySpatial2020a; @liMeasuringMultiactivitiesAccessibility2024a] as well as less exclusive amenities types such as greenspace and recreation amenities [@wuWillOpeningCommunity2020; @liMeasuringMultiactivitiesAccessibility2024a], and shopping destinations [@kelobonyeMeasuringAccessibilitySpatial2020a; @chenEnhancingTwoStepFloating2019].
+
+The consideration of congestion in accessibility measures was the motivation for the development of approaches that consider competition [@weibullAxiomaticApproachMeasurement1976; @josephMeasuringPotentialPhysical1982; @shenLocationCharacteristicsInnercity1998] and are popularly applied in the literature as floating catchment area methods [@luoMeasuresSpatialAccessibility2003].
+While these approaches purport to account for congestion, Páez et al.
+(2019) [@paezDemand2019] demonstrates that they do not solve the issue of multiple counting of opportunities in general, thus leading to biases in the calculation of total demand (population) and supply (of opportunities).
 They sometimes inflate counts and other times deflate them.
-In response to this, recent research has paid closer attention to the way opportunities are counted in accessibility analysis.
-Paez et al.
-@paezDemand2019, for example, focuses on floating catchment area methods and introduces a normalization of the impedance matrix to allocate the population and then the level of service proportionally.
+In this line, recent research has paid closer attention to the way opportunities are counted in accessibility analysis.
+Páez et al.
+(2019) [@paezDemand2019], for example, focuses on floating catchment area methods and introduces a normalization of the impedance matrix to allocate the population and then the level of service (i.e., the supply to demand) proportionally.
 More recently, Soukhov et al.
-[@soukhovIntroducingSpatialAvailability2023] introduced a singly-constrained measure of accessibility, called spatial availability, that employs a similar but more sophisticated proportional allocation mechanism.
+(2023) [@soukhovIntroducingSpatialAvailability2023] introduced a singly-constrained measure of accessibility, called spatial availability, that employs a similar but more sophisticated proportional allocation mechanism.
 The work of these authors show that floating catchment area methods can be seen as singly-constrained accessibility measures and improve on existing approaches by guaranteeing that each opportunity is counted only once.
@@
 A limitation of spatial availability as introduced by Soukhov et al.
-[@soukhovIntroducingSpatialAvailability2023] is that it was developed for the case of a homogeneous population, for example for the case of a single mode of transportation.
+(2023) [@soukhovIntroducingSpatialAvailability2023] is that it was developed for the case of a homogeneous population, for example for the case of a single mode of transportation.
 However, the finite nature of opportunities makes the analysis of heterogeneous populations very relevant.
@@
 It is worth noting though that the consideration for travel costs by different modes is in fact just one case of heterogeneous populations.
-The method itself can easily accommodate other forms of heterogeneity, for example: variations in travel behavior between older and younger adults [e.g., @paezHealth2010], the propensity of older adults to use different modes of transportation [e.g., @moniruzzamanMode2013], the usually shorter trip lengths of children compared to adults [e.g., @reyesWalking2014], or the more limited travel ranges of single parents [e.g., @paezJobs2013].
+The method itself can easily accommodate other forms of heterogeneity, for example: variations in travel behavior between older and younger adults (e.g., Páez et al.
+(2010) [@paezHealth2010]), the propensity of older adults to use different modes of transportation (e.g., Moniruzzaman et al.
+(2013) [@moniruzzamanMode2013]), the usually shorter trip lengths of children compared to adults (e.g., Reyes et al.
+(2014) [@reyesWalking2014]), or the more limited travel ranges of single parents (e.g., Páez et al.
+(2010) [@paezJobs2013]).
 
 The rest of this paper is organized as follows.
-In Section 2 we provide a brief review of multimodal accessibility.
-In the following Section 3 we demonstrate the derivation of the new spatial availability expression for multiple modes.
-In Section 4 we illustrate relevant issues through a synthetic example.
+In Section 2, we provide a brief review of multimodal accessibility.
+In the following Section 3, we demonstrate the derivation of the spatial availability expression for multiple modes.
+In Section 4, we illustrate relevant issues through a synthetic example.
 This is followed in Section 5 by an empirical example using home-to-work data from the city of Madrid, Spain after the implementation of its Low Emission Zones (LEZ).
 Data for this example is sourced from the city's 2018 travel survey.
-The empirical example demonstrates the differences in multimodal spatial availability highlighting differences within and outside the LEZ for travelers using different modes, namely car, transit, cycling and walking.
+The empirical example demonstrates the multimodal spatial availability landscape in 2018.
+It highlights the differences within and outside the LEZ for travelers using different modes, namely, car, transit, cycling and walking.
 In Section 6, we provide concluding remarks on the strengths of the use of spatial availability as a multimodal accessibility measure, and discuss potential future uses in policy planning scenarios as well as directions for future research.
@@
 
-Location-based accessibility indicators are quantitative measures of _potential_ interaction with opportunities for locations within a given region: they are summary measures of the relationship between land-use and transport systems.
-Arguably, the most commonly used are measures based on the gravity model [@wuUnifyingAccess2020]; this includes cumulative opportunity measures which are a special case of the gravity model [@levinsonTransportAccessManual2020; @handyMeasuring1997].
-These gravity model based measures weight opportunities depending on how easy it is to reach them.
-Given an origin ($i$) and a destination ($j$), an impedance function $f^{m}(c^m_{ij})$ converts the cost of travel (e.g., time, money, generalized cost) into a score that represents the propensity for potential interaction.
-These measures originate from the work proposed by Hansen @hansenHowAccessibilityShapes1959, which can take the following form in the multimodal case: $S_i^m = \sum_j O_j f^m(c_{ij}^m)$.
-In this form, $m$ is a set of modes that have mode-specific travel costs ($c_{ij}^m$) and/or travel impedance functions ($f^m(\cdot)$).
+Place-based accessibility indicators are quantitative measures of _potential_ interaction with opportunities for locations within a given region: they are summary measures of the relationship between land-use and transport systems.
+Arguably, the most commonly used measures are based on the gravity model [@wuUnifyingAccess2020]; this includes cumulative opportunity measures that are a special case of the gravity model [@levinsonTransportAccessManual2020; @handyMeasuring1997].
+These gravity model based measures weight opportunities depending on the ease of reaching them.
+Given an origin $i$ and a destination $j$, an impedance function $f^{m}(c^m_{ij})$ converts the cost of travel (e.g., time, money, generalized cost) into a score that represents the propensity for potential interaction.
+These measures originate from the work proposed by Hansen (1959) [@hansenHowAccessibilityShapes1959], which can take the following form in the multimodal case: $S_i^m = \sum_j O_j f^m(c_{ij}^m)$.
+In this form, $m$ is a set of modes that have mode-specific travel costs ($c_{ij}^m$) and/or travel impedance functions $f^m(\cdot)$.
 
@@
 To cite an example, Tahmasbi et al.
-[@tahmasbiMultimodalAccessibilitybasedEquity2019] use Hansen-type accessibility to assess the potential interaction with retail locations by three modes: walking, public transit, and car (i.e., $m = w, p, c$).
+(2019) [@tahmasbiMultimodalAccessibilitybasedEquity2019] use Hansen-type accessibility to assess the potential interaction with retail locations by three modes: walking, public transit, and car (i.e., $m = w, p, c$).
 $S_i^m$ is the sum of retail locations $j$ that can potentially be reached under the travel impedance as calculated for each $i$ and $m$.
@@
 In this work, Tahmasbi et al.
-[@tahmasbiMultimodalAccessibilitybasedEquity2019] show that car travel affords the highest $S_i^{m}$ values in the majority of $i$, i.e., travelers who use a car can potentially reach more retail opportunities than populations using other modes.
+[@tahmasbiMultimodalAccessibilitybasedEquity2019] show that car travel affords the highest $S_i^{m}$ values in the majority of $i$ i.e., travelers who use a car can potentially reach more retail opportunities than populations using other modes.
 However, higher $S_i^{m}$ values for car do not affect the values of $S_i^{m}$ for other modes: in effect, each mode is analysed as if the others did not exist.
@@
 The accessibility scores for the modes are often values that are difficult to interpret beyond making statements about relative size.
-For example, Lunke [@lunkeModalAccessibilityDisparities2022], researching the region of Oslo, reports accessibility scores for car in the order of tens of thousands of employment opportunities.
+For example, Lunke (2022) [@lunkeModalAccessibilityDisparities2022] reports accessibility scores for car in the order of tens of thousands of employment opportunities in the Oslo region.
 The corresponding scores for transit are lower, but still often in the thousands or tens of thousands.
 As reported, the ratio of the transit to the car score can be lower than 0.2 (meaning transit gives access to less than 20% of the opportunities than car).
-But despite the discussion about "sufficient accessibility", it is unclear what the unconstrained scores mean: is having access to but 10,000 jobs by transit insufficient?
+But despite the discussion about "sufficient accessibility", it is unclear what the unconstrained scores mean: is having access to 10,000 jobs by transit insufficient?
 After all, 10,000 employment opportunities are still plenty of opportunities.
-These ratios can be found elsewhere in the literature [e.g., @paezRelative2010, Figs.
-7, 8, and 9; @paezHealth2010, Fig.
-5; @paezJobs2013, Figs.
-6, 7, and 8], and they are useful as relative assessment of when some members of the public are better or worse off than others, but they are silent on how bad is "worse off".
+These ratios can be found elsewhere in the literature e.g., Figs.
+7, 8, and 9 in Páez et al.
+(2010) [@paezRelative2010], Fig.
+5 in Páez et al.
+(2010) [@paezHealth2010], Páez et al.
+(2013) Figs.
+6, 7, and 8 in [@paezJobs2013].
+They are useful as relative assessment of when some members of the public are better or worse off than others, but they are silent on how bad is "worse off".
 
@@
 These questions lack straightforward answers since certain values will always be relatively 'low' or 'high', but do not track to a quantity that can be intuitively understood.
-Presentation or discussion of Hansen-type accessibility that has been standardised in this way is not uncommon in the literature [e.g., @campbell2019accessibility; @maharjanSpatialEquityModal2022].
-
-Once we understand opportunities to be finite, it is possible for an accessibility measure to take on a crisper meaning.
-As considered in the long tradition of accessibility research, capacity of opportunities is limited and thus is subject to competition by population [@soukhovIntroducingSpatialAvailability2023; @grengsJobAccessibilityModal2010; @kawabataJobAccessibilityIndicator2006; @kwokUseModalAccessibility2004; @morrisAccessibilityIndicatorsTransport1979; @weibullAxiomaticApproachMeasurement1976; @shenLocationCharacteristicsInnercity1998; @cuiSpatialAccessPublic2020; @maoMeasuringSpatialAccessibility2013; @josephMeasuringPotentialPhysical1982; @higgsModellingSpatialAccess2017; @vanweeAccessibilityMeasuresCompetition2001; @merlinDoesCompetitionMatter2017;@kelobonyeMeasuringAccessibilitySpatial2020a; @liMeasuringMultiactivitiesAccessibility2024a].
-There are only so many school-seats, hospital capacity, employment opportunities, etc., in a region and if one person reaches an opportunity, it is taken: the supply of an opportunity and the demand for that opportunity are two components of accessibility.
-These are clear examples of opportunities that are unambiguously competitive.
-But we would go as far as to argue that every type of opportunity is subject to congestion or capacity constraint, even when the opportunities are conventionally seen as non-competitive.
+Presentation or discussion of Hansen-type accessibility that has been standardised in this way is not uncommon in the literature [@campbell2019accessibility; @maharjanSpatialEquityModal2022].
+
+If we understand opportunities to be finite and/or subject to some levels of congestion, it is possible for an accessibility measure to take on a crisper meaning.
+Accessibility research has a history of considering opportunity competition, especially regarding school-seats, hospital capacity, and employment opportunities [@soukhovIntroducingSpatialAvailability2023; @grengsJobAccessibilityModal2010; @kawabataJobAccessibilityIndicator2006; @kwokUseModalAccessibility2004; @morrisAccessibilityIndicatorsTransport1979; @weibullAxiomaticApproachMeasurement1976; @shenLocationCharacteristicsInnercity1998; @cuiSpatialAccessPublic2020; @maoMeasuringSpatialAccessibility2013; @josephMeasuringPotentialPhysical1982; @higgsModellingSpatialAccess2017; @vanweeAccessibilityMeasuresCompetition2001; @merlinDoesCompetitionMatter2017;@kelobonyeMeasuringAccessibilitySpatial2020a; @liMeasuringMultiactivitiesAccessibility2024a].
+If one person reaches an opportunity - it is taken: the supply of an opportunity and the demand for that opportunity are the nodes in accessibility analysis.
+These types of opportunities are unambiguously exclusive.
+But we would go as far as to argue that every type of opportunity is subject to congestion or capacity constraint, even when the opportunities are conventionally seen as inexhaustible.
 
@@
 For instance, standards for providing green spaces are often stated in the form of _exclusive access_, in units of amenity per capita.
-A case in point is a the Ile-de-France region, a jurisdiction that suggested in a 2013 planning document that at the municipal level at least 10 $m^2$ of public green space should be supplied _per inhabitant_ [@liottaPlanning2020].
-Green spaces are not evenly distributed, which means that who has access to them hinges on where they are and how easy is to reach them.
-This formulation of provision of amenities is not unusual.
-For example, Natural England, an organization that recommends an Accessible Natural Greenspace Standard such that the minimum supply of space is one ha of statutory Local Nature Reserves per thousand population^[see  https://redfrogforum.org/wp-content/uploads/2019/11/67-Nature-Nearby%E2%80%99-Accessible-Natural-Greenspace-Guidance.pdf].
-Similarly, the World Health Organization [cited in @OECDFrameworks2013] recommends that cities provide a minimum of 9 $m^2$ of green area per inhabitant.
+For example, a 2013 planning document for the Ile-de-France region suggested a public green space municipal standard of 10 $m^2$ _per inhabitant_ [@liottaPlanning2020].
+Green spaces are not evenly distributed, meaning those who have access to them depends on where they are and how easy they are to reach.
+This formulation of amenity provision is not unusual.
+As another example, Natural England recommends a national "accessible natural greenspace standard" such that the minimum supply of space is 1 $ha$ of statutory local nature reserves per thousand population [@naturalenglandNatureNearbyAccessible2010].
+Similarly, the World Health Organization [@OECDFrameworks2013] recommends that cities provide a minimum of 9 $m^2$ of green area per inhabitant.
 For our purposes, standards of this type translate into "how much of this resource is available to one individual that has not been claimed by anyone else?".
-Green spaces often have large capacities, but they still have a capacity, and it is not the same for a person to have access to 5 $m^2$ of _uncongested_ green space as 15 $m^2$.
+Green spaces often have large capacities, but they still have a capacity and it is not the same for a person to have access to 5 $m^2$ of _uncongested_ green space as 15 $m^2$.
 This difference is in fact a matter of justice [@laraSpace2015; @liottaPlanning2020].
-Constraining accessibility is in this way a useful way to evaluate the congested availability of any type of opportunity.
-As development of sound standards is emphasized in the planning literature, in particular in regards to fairness in transportation [see @martensFair2021], spatial availability analysis is a useful way to develop and assess standards.
+Constraining accessibility is a useful way to evaluate the congested availability of any type of opportunity.
+As development of sound standards is emphasized in the planning literature, in particular in regards to fairness in transportation [@martensFair2021], spatial availability analysis can be used to develop and assess standards.
 
@@
 As noted in the Introduction, competitive accessibility was the rationale for developing floating catchment area methods (FCA), popularized by Luo et al.
-[@luoMeasuresSpatialAccessibility2003] who reformulated the work of Shen [@shenLocationCharacteristicsInnercity1998] into two steps (although similar, and earlier, developments are found in [@weibullAxiomaticApproachMeasurement1976; @josephMeasuringPotentialPhysical1982]).
+(2003) [@luoMeasuresSpatialAccessibility2003] who reformulated the work of Shen (1998) [@shenLocationCharacteristicsInnercity1998] into two steps (although similar, and earlier, developments are found in [@weibullAxiomaticApproachMeasurement1976; @josephMeasuringPotentialPhysical1982]).
 Shen-type accessibility is formulated as: $a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}$ where $D_j^m$ is the potential demand for opportunities equal to travel impedance weighted population $\sum_i P_i^m f^m(c_{ij}^m)$ and the remaining variables are repeated in the Hansen-type measure.
@@
 In this way, it considers competition.
-That said, the measure remains unconstrained, meaning both population _and_ opportunities are multiply counted [see @paezDemand2019].
+That said, the measure remains unconstrained, meaning both population _and_ opportunities are multiply counted [@paezDemand2019].
 In other words, interpretation of the Shen-type accessibility scores between modes is fraught, as it is for Hansen-type measures.
 
-To illustrate, @taoInvestigatingImpactsPublic2020a calculates $a_i^m$ to jobs for different income-group populations in Shenzhen, China using $m = \text{public transit}$ and $m=\text{car}$.
+To illustrate, Tao et al.
+(2020) @taoInvestigatingImpactsPublic2020a calculates $a_i^m$ to jobs for different income-group populations in Shenzhen, China for those using transit and car.
 Their results indicate that zones with low-income populations have lower $a_i^m$ than zones with higher-income populations.
-Further, they show that $a_i^{m=\text{public transit}}$ is lower than $a_i^{m=\text{car}}$ at many zones, arguing that this may further place those zones with lower-income populations at a disadvantage.
+Further, they show that $a_i^{\text{transit}}$ is lower than $a_i^{\text{car}}$ in many zones, arguing that this may further place those zones with lower-income populations at a disadvantage.
 $a_i$ and/or $a_i^m$ are used to compare relative spatial differences in overall competitive accessibility and multimodal competitive accessibility, but because opportunities were doubly counted (entering the sums of both modes), this makes for uneasy interpretations of the differences in $a_i^{m}$ between modes.
@@
 Spatial availability improves on the discussed Hansen-type and Shen-type accessibility approaches by constraining the sum of opportunities, that is, by treating opportunities as finite.
-This is done by means of proportional allocation factors that follow well established principles of spatial interaction and the gravity model [see @wilsonFamilty1971].
+This is done by means of proportional allocation factors that follow well established principles of spatial interaction and the gravity model [@wilsonFamilty1971].
 In Soukhov et al.
-[@soukhovIntroducingSpatialAvailability2023] these factors consider: the mass effect (e.g., the size of populations at different origins); and the cost of travel from different zones (e.g., some sub-populations face relatively higher or lower costs).
+(2023) [@soukhovIntroducingSpatialAvailability2023] these factors consider: the mass effect (e.g., the size of populations at different origins); and the cost of travel from different zones (e.g., some sub-populations face relatively higher or lower costs).
 The following section introduces the multimodal form of spatial availability.
@@
 
-In brief, we define the spatial availability at an origin $i$ ($V_{i}$) as the proportion of all opportunities in the region that are allocated to location $i$ from all destinations $j$.
-$V_{i}$ is a value of how many opportunities are available to each location $i$ out of all the opportunities in the region.
-The general formulation of spatial availability $V_{i}$ is shown in Equation (\ref{eq:spatial-availability-general}) [see @soukhovIntroducingSpatialAvailability2023]:
+In brief, we define the spatial availability $V_{i}$ at an origin as the proportion of all opportunities in the region that are allocated to origin $i$ from all destinations $j$.
+The general formulation of spatial availability is shown in Equation (\ref{eq:spatial-availability-general}) [@soukhovIntroducingSpatialAvailability2023]:
 ```{=tex}
@@
 
-- $V_i$ is the number of spatially available opportunities at $i$; the sum of $V_{i}$ is identical to the total number of opportunities in the region (i.e., $\sum_j O_j = \sum_i V_i$); in other words, opportunities are dealt with as finite resources.
+- $V_i$ is the number of spatially available opportunities at $i$; the sum of $V_{i}$ is identical to the total number of opportunities in the region (i.e., $\sum_j O_j = \sum_i V_i$).
 
 Compared to Hansen-type accessibility $S_i = \sum_{j=1}^J O_jf(c_{ij})$, we see that spatial availability is, like the Hansen-type measure, a weighted sum of the opportunities.
-What makes spatial availability stand apart from other approaches is how the weight used in the sum (balancing factor $F^t_{ij}$) implements a proportional allocation mechanisms to ensure that the sum of $V_i$ is constrained to match the total number of opportunities in the region.
+What makes spatial availability stand apart from other approaches is how the weight used in the sum (balancing factor $F^t_{ij}$) implements a proportional allocation mechanism to ensure that the sum of $V_i$ is constrained to match the total number of opportunities in the region.
 As such, spatial availability is singly-constrained and natively considers competition.
@@
 
-Implementing $F^{tm}_{ij}$, the following Equation (\ref{eq:spatial-availability-multimodal}) gives the multimodal version of spatial availability $V_i^m$:
+Implementing $F^{tm}_{ij}$, Equation (\ref{eq:spatial-availability-multimodal}) gives the multimodal version of spatial availability $V_i^m$:
 ```{=tex}
@@
 Population center $A$ is Suburban: it is closest to its own relatively large employment center at $1$, close to the Urban's equally large employment center $2$, and has a population that is smaller than the Urban $B$ and larger than the Satellite $C$.
-$B$ has the largest $x$-using population, followed by then $A$, then $C$.
-This synthetic example is inspired by the single-mode example used in @shenLocationCharacteristicsInnercity1998 and reconfigured in @soukhovIntroducingSpatialAvailability2023.
+$B$ has the largest $x$-using population (40%), followed by $A$ (33%) then $C$ (30%).
+This synthetic example is inspired by the single-mode example used in Shen (1998) @shenLocationCharacteristicsInnercity1998 and reconfigured in Soukhov et al.
+(2023) @soukhovIntroducingSpatialAvailability2023.
 
@@
 
-From the perspective of access to a _finite_ amount of opportunities in the region ($210,000$ jobs), the sub-population that is most proximate to jobs (low cost to reach), furthest from large populations (less competition), and uses the fastest mode $z$ (greater range) can potentially reach the largest number of opportunities.
+From the perspective of access to a _finite_ amount of opportunities in the region ($210,000$ jobs), the sub-population that is most proximate to jobs (lowest cost to reach), furthest from large populations (least competition), and uses the fastest mode $z$ (greatest range) can potentially reach the largest number of opportunities.
 This appears to be the sub-population at $A$ using mode $z$.
@@
 The values calculated for $S_i^m$ (Hansen-type accessibility), $a_i^m$ (Shen-type accessibility), and $V_i^m$ (spatial availability) for each $i$ and $m$ are shown in the middle three columns and are aggregated for each $i$ in the final two columns in Table 1 <!-- \ref{tab:tab1}-->.
-As in the example in Shen [@shenLocationCharacteristicsInnercity1998], we use a negative exponential impedance function $f^m(c_{ij}^m) = \exp(-\beta\cdot c_{ij})$ with $\beta=0.1$ for both $x$ and $z$ modes for all accessibility measures calculations.
+As in the example in Shen (1998) [@shenLocationCharacteristicsInnercity1998], we use a negative exponential impedance function $f^m(c_{ij}^m) = \exp(-\beta\cdot c_{ij})$ with $\beta=0.1$ for both $x$ and $z$ modes for all accessibility measures calculations.
 Notice that in this example we use the same impedance function but the travel times are different for the two modes.
-More generally, it is possible to use different impedance functions for the modes, as demonstrated in the empirical example in the following section.
+More generally, it is possible to use different impedance functions for the modes, as the empirical example in the following section demonstrates.
 
@@
 For all $i$, the travel by $z$ results in higher values of $S_i^m$ than travel by $x$.
-Lack of competition, or alternatively the assumption of an inexhaustible resource in the calculation of $S_i^m$, lead to a curious result.
+Lack of competition, or alternatively the assumption of an inexhaustible resource in the calculation of $S_i^m$, leds to a curious result.
 Since the populations in $A$ and $B$ have the same travel impedance to employment centers $1$, $2$ and $3$ (either 15, 30, or 100 minutes using $x$ or 10, 25, or 80 minutes using $z$), their values of $S_i^m$ are the same for both $A$ and $B$.
@@
 To connect this example to the aforementioned literature, $S_i^m$ is calculated in the work of Tahmasbi et al.
-[@tahmasbiMultimodalAccessibilitybasedEquity2019]; they contrast differences in $S_i^m$ values between modes in a relative and comparative sense, but make no further interpretation of the $S_i^m$ values.
+(2019) [@tahmasbiMultimodalAccessibilitybasedEquity2019]; they contrast differences in $S_i^m$ values between modes in a relative and comparative sense, but make no further interpretation of the $S_i^m$ values.
 More densely populated metropolitan regions will tend to have more opportunities and hence large $S_i^m$ values and less densely regions, smaller values; how much of these differences may simply an artifact of region density?
@@
 The Shen-type measure is implemented in aforementioned work of Tao et al.
-[@taoInvestigatingImpactsPublic2020a] to calculate modal $a_i^m$ values and the aggregated $a_i$ is implemented in the work of Carpentier et al.
-[@carpentieriMultimodalAccessibilityPrimary2020].
+(2020) [@taoInvestigatingImpactsPublic2020a] to calculate modal $a_i^m$ values and the aggregated $a_i$ is implemented in the work of Carpentier et al.
+(2020) [@carpentieriMultimodalAccessibilityPrimary2020].
 However, similar to Hansen-type accessibility, these works discuss relative and spatially comparative differences in values, but veer from interpreting the values of $a_i^m$ or $a_i$ themselves.
@@
 In contrast, spatial availability $V_i$ considers competition and is constrained such that the total sum of values is equal to the total number of opportunities in the region (i.e., $210,000$ jobs).
-Seen in fifth column of Table 1 <!-- \ref{tab:tab1}-->, the values of $V_i^m$ in $A$ and $B$ are not the same within each mode (as this measure considers competition).
-In fact, at $A$, users of mode $z$ capture `r ((table_toy_format %>% filter(i == "A" & m == "z") %>% select("V_im")) - (table_toy_format %>% filter(i == "A" & m == "x") %>% select("V_im"))) %>% pull() %>% prettyNum(big.mark = ",")` more spatially available jobs (of the $210,000$ jobs in the region) than the sub-population travelling by $x$.
+Seen in fifth column of Table 1 <!-- \ref{tab:tab1}-->, the values of $V_i^m$ for $A$ and $B$ are not the same within each mode (as this measure considers competition).
+In fact, at $A$, users of mode $z$ capture `r ((table_toy_format %>% filter(i == "A" & m == "z") %>% select("V_im")) - (table_toy_format %>% filter(i == "A" & m == "x") %>% select("V_im"))) %>% pull() %>% prettyNum(big.mark = ",")` more spatially available jobs (of the $210,000$ jobs in the region) than the sub-population travelling by the slower-mode $x$.
 The numerical difference is clear since it refers to opportunities out of the total.
@@
 This aggregation, $V_i$, is shown in the seventh column in Table 1 <!-- \ref{tab:tab1}-->.
-Again looking at center $A$, $A$ is allocated `r ((V_i %>% filter(i == "A" & m == "z")  %>% select("V_i"))) %>% pull() %>% prettyNum(big.mark = ",")` spatially available opportunities for both modes.
-`r ((V_i %>% filter(i == "A" & m == "z") %>% select("V_im")) / ((V_i %>% filter(i == "A" & m == "z") %>% select("V_i")))) %>% pull() %>% as.numeric() %>% percent() ` of this spatial availability allocated to $A$ is assigned to users of mode $z$ despite representing `r 0.66 %>% percent()` of $A$'s population.
+$A$ is allocated `r ((V_i %>% filter(i == "A" & m == "z")  %>% select("V_i"))) %>% pull() %>% prettyNum(big.mark = ",")` spatially available opportunities for both modes.
+`r ((V_i %>% filter(i == "A" & m == "z") %>% select("V_im")) / ((V_i %>% filter(i == "A" & m == "z") %>% select("V_i")))) %>% pull() %>% as.numeric() %>% percent()` of this spatial availability allocated to $A$ is assigned to users of mode $z$ despite representing `r 0.66 %>% percent()` of $A$'s population.
 
@@
 Across the entire region, `r V_ijm %>% filter(m=="z") %>% group_by(i) %>% summarize(pop = first(pop)) %>% select(pop) %>% sum() %>% prettyNum(big.mark = ",")` people use $z$ (`r ((V_ijm %>% filter(m=="z") %>% group_by(i) %>% summarize(pop = first(pop)) %>% select(pop) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% select("pop") %>% sum())) %>% percent()` of the region population).
-However, users of $z$ account for `r ((V_ijm %>% filter(m=="z") %>% select(V_ij) %>% sum()) / (V_ijm %>% select(V_ij) %>% sum())) %>% percent()` of the region's total spatial availability - while the remaining `r (1 - (V_ijm %>% filter(m=="z") %>% select(V_ij) %>% sum()) / (V_ijm %>% select(V_ij) %>% sum())) %>% percent()` is allocated to users of mode $x$ who are `r ((V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% filter(m == "x") %>% dplyr::select(pop) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% dplyr::select(pop) %>% sum())) %>% percent()`of the total population.
+However, users of $z$ account for `r ((V_ijm %>% filter(m=="z") %>% select(V_ij) %>% sum()) / (V_ijm %>% select(V_ij) %>% sum())) %>% percent()` of the region's total spatial availability - while the remaining `r (1 - (V_ijm %>% filter(m=="z") %>% select(V_ij) %>% sum()) / (V_ijm %>% select(V_ij) %>% sum())) %>% percent()` is allocated to users of mode $x$ who are `r ((V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% filter(m == "x") %>% dplyr::select(pop) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% dplyr::select(pop) %>% sum())) %>% percent()` of the total population.
 Notably, the population who uses $x$ have `r (((V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% filter(m == "x") %>% dplyr::select(pop) %>% sum()) / (V_ijm %>% group_by(i,m) %>% summarize(pop = first(pop), .groups = "drop") %>% dplyr::select(pop) %>% sum())) - ((V_ijm %>% filter(m=="x") %>% select(V_ij) %>% sum()) / (V_ijm %>% select(V_ij) %>% sum()))) %>% percent()` fewer spatially available opportunities than its share in the population.
@@
 
-Since spatial availability is constrained and has an interpretable meaning as a proportion of the total opportunities in the region, the values at $i$ have a straighforward interpretation.
+Since spatial availability is constrained to the total opportunities in the region, the values at $i$ have a straighforward interpretation.
 Inequality in $V_i^m$ values can be explored through a variety of approaches.
@@
 Users of $x$, especially those at $B$ and $C$, are directly impacted by the jobs that are spatially available to users of $z$ _in addition to_ the mass effect (occurring at $B$, high population density) and high travel impedance (occurring at the Satellite $C$).
-
-If, let us say, the planning goal was to have one spatially available job per mode-using population, a policy intervention could be devised, to reduce the values of $v_i^z$ (making it slower or more expensive) and increase he values of $v_i^x$ (making it faster or les expensive).
+Notably, $v_i^m$ values are equal to the ratios of the Shen-type measure $a_i^m$ in this case, however they take on a different interpretation.
+
+If, let us say, the planning goal was to have one spatially available job per mode-using population, a policy intervention could be devised, to reduce the values of $v_i^z$ (making it slower or more expensive) and increase the values of $v_i^x$ (making it faster or less expensive).
 The purpose of this synthetic demonstration is to show how spatial availability can be used to quantify the competitive (dis)advantage in a multimodal application.
@@
 
-The context for the empirical example is Madrid, Spain.
-This city implemented a Low Emission Zone (LEZ) in 2017 to pursue goals set out in the national climate change agenda, cut nitrogen dioxide levels, and to prioritize people's movement in the city.
+The context for the empirical example is the city of Madrid, Spain.
+This city implemented a Low Emission Zone (LEZ) in 2017 to pursue goals set out in the national climate change agenda such as cutting nitrogen dioxide levels and prioritizing people's movement in the city.
 LEZs elsewhere have similarly been implemented as interventions to reduce GHG emissions, improve air quality, and support sustainable mobility [@margaryanLowEmissionZones2021; @verbeekJustManagementUrban2022].
@@
 In geographic scope, the 2017 boundaries of the LEZ in Madrid were relatively modest, covering only approximately 4.72 km\textsuperscript{2} of the central business district of the city (the so-called LEZ Centro).
-As of this writing, there are plans to expand these boundaries to the area inside the M-30, an orbital highway in proximity to the city center (i.e., LEZ M-30).
+As of this writing, there are plans to expand these boundaries to the area inside the M-30, an orbital highway in proximity to the city center.
 Within the 2017 LEZ Centro implementation, all cars, motorcycles and freight vehicles with environmental labels A or B (older makes and models of fossil fuel internal combustion engine vehicles), were disallowed from entering the zone unless they are used by residents or meet other exemptions.
@@
 
-For this case study, we quantify spatial availability to opportunities by different modes in Madrid.
-Particularly, we demonstrate how $V_i^m$ can be used to derive insights into how the restriction of car mobility in areas around/within the LEZ Centro may have allowed more sustainable (but often slower or more costly modes) to become more competitive.
+The purpose of this empirical example is to quantify spatial availability to employment opportunities by different modes in Madrid after LEZ Centro implementation.
+Particularly, we demonstrate how $V_i^m$ can be used to illustrate the spatial availability advantage that more sustainable modes (that are often slower than car) gain within/around the LEZ Centro.
+We speculate on how this competitive advantage is heightened as a result of the LEZ implementation that restricts car mobility.
 
@@
   - Bike: all bicycle trips (e.g., private, public, or company bike trips) and "other" types of micromobility options.
-  - Walk: walking or by foot.
+  - Walk: pedestrian mode.
 
 Some aggregation of modes is necessary to calculate the travel impedance functions by mode.
-From Figure \ref{fig:Fig2}, it can be seen that the largest concentration of jobs is within, near, and to the north of LEZ Centro.
+From Figure \ref{fig:Fig2}, the largest concentration of jobs is within, near, and to the north of LEZ Centro.
 The populations with access to those jobs by mode (Figure \ref{fig:Fig3}) are spatially distinct.
@@
 This distribution is likely caused by a variety of factors including: transit coverage and service within with city, effective car infrastructure outside of the M-30, and/or the impact of the LEZ Centro itself.
-From Figure \ref{fig:Fig3}, it can be seen that active travel is less common than motorised trips at `r (pop_props$population_m[1]/sum(pop_props$population_m)) %>% percent()` and `r (pop_props$population_m[2]/sum(pop_props$population_m)) %>% percent()` for cycling and walking respectively.
+From Figure \ref{fig:Fig3}, active travel is less common than motorised trips at `r (pop_props$population_m[1]/sum(pop_props$population_m)) %>% percent()` and `r (pop_props$population_m[2]/sum(pop_props$population_m)) %>% percent()` for cycling and walking respectively.
 Noticeably, there is a positive trend between the walking and cycling in zones where transit is also present.
-This positive trend is higher than for car trip populations.
-<!-- the populations do not follow a normal distribution, so we can't say there is a correlation, formally.
-see calcs in chunk at the beginning of this subsection-->
+This positive trend is higher than for car-using populations.
 
@@
 This distribution is then used to estimate the parameters of a function for the travel impedance [as done in @lopez_2017_spatial; @horbachov_theoretical_2018; @batista_estimation_2019].
-To fit the impedance functions, we use the Maximum likelihood estimation and the Nelder-Mead method for direct optimization available within the R {fitdistrplus} package [@fitdistrplus_2015].
+To fit the impedance functions, we use the Maximum likelihood estimation and the Nelder-Mead method for direct optimization available within the {fitdistrplus} R package [@fitdistrplus_2015].
 Based on goodness-of-fit criteria and associated diagnostics, the gamma and log-normal probability density functions are selected as best fitting curves for the motorised and non-motorised modes respectively.
@@
 
-It is worthwhile reiterating that the empirical example is a snapshot of spatial availability by mode using modal home-work origin-destination data from the 2018 travel survey.
-It reflects the travel behaviour after LEZ implementation, i.e.
-very few motorists commute into the LEZ centro.
-Our purpose in this empirical example is to investigate the trends in spatial availability of employment opportunities by mode, and illustrate how spatial availability can be used to discuss modal competitive advantage.
-We also discuss how multimodal spatial availability landscape displayed in the Figure \ref{fig:Fig5} may have been impacted by the LEZ.
-
+Figure \ref{fig:Fig5} illustrates the multimodal spatial availability landscape for each of the four modes $m$ at the level of traffic analysis zones $i$.
+$V_i^m$ is a proportion of the total number of `r SA_i_zn208$V_i %>% sum() %>% prettyNum(big.mark = ",")` jobs in the city.
+Since $V_i^m$ is calculated based on the population of workers and the distribution of jobs, the values can be understood as the number of full-time jobs that are spatially available to full-time workers at that $i$ traveling by mode $m$, relative to all the jobs in the city.
 ```{r, SA-m-plot, fig.cap="\\label{fig:Fig5} Spatial availability of jobs per origin and mode $V_i^m$ in Madrid at the level of TAZ.
@@
 
-In Figure \ref{fig:Fig5}, the spatial availability of jobs $V_i^m$ is calculated for each of the four modes $m$ at the level of traffic analysis zones $i$ in Madrid.
-$V_i^m$ is a proportion of the total number of `r SA_i_zn208$V_i %>% sum() %>% prettyNum(big.mark = ",")` jobs in the region.
-Since $V_i^m$ is calculated based on the population of workers and the distribution of jobs, the values can be understood as the number of full-time jobs that are spatially available to full-time workers at that $i$ traveling by mode $m$, relative to all the jobs in the city.
-This implementation of spatial availability assumes that all employment opportunities are of interest to the entire population but are proportional allocated based on relative travel cost (informed by observed modal origin-destination travel behaviour) and the relative amount of the mode-using population residing in the zones.
-
-There are noticeable differences in the magnitude of $V_i^m$ between modes in Figure \ref{fig:Fig5}.
-The majority of $V_i^m$ (which is to say of spatially available jobs) are allocated to workers travelling by car and transit.
-In a way, this is to be expected since users of these modes represent `r (modal_V_comps_4plot$perc[3] + modal_V_comps_4plot$perc[4]) %>% percent(accuracy=0.1)` of the total population.
-However, the ability to travel at greater speeds also impacts these results.
-Furthermore, differences in $V_i^m$ values within modes also exist in space: car users outside of the M-30 region appear to enjoy greater spatial availability, while some zones inside the M-30 have greater spatial availability for transit.
-Overall, the magnitude of $V_i^m$ values for cyclists and pedestrians are lower than for car and transit but the highest values of $V_i^{bike}$ and $V_i^{walk}$ tend to be found in zones within the M-30 and origins with higher spatial availability by transit.
-
-The differences between the shares of modes and their shares of spatially available opportunities highlights the competitive advantage of certain modes, although this effect is not geographically uniform.
-As seen in the left-most columns of Figure \ref{fig:Fig6}, users of 'car/motor' and 'transit' together can avail `r (modal_V_comps_4plot$perc[7] + modal_V_comps_4plot$perc[8]) %>% percent(accuracy=0.1)` of all jobs in the city (Spatial Availability by Mode).
-However, car/motor users have a disproportionate share of $V_i^m$ relative to the population of users of this mode (Population by Mode), compared to opportunities that are spatially available to transit users.
-The combined population of car and transit users is `r modal_V_comps_4plot$perc[3] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[4] %>% percent(accuracy=0.1)` respectively, but these populations are allocated `r modal_V_comps_4plot$perc[7] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[8] %>% percent(accuracy=0.1)` respectively, of the city's jobs.
-When treating the number of opportunities that can be reached as a finite value (total: 847,574 opportunities), fewer opportunities are spatially available to slower modes (i.e., walking and cycling), even taking into account that their share is smaller overall.
-These modes are at a disadvantage as a result of: the travel impedance for longer trips (see Figure \ref{fig:Fig4}; their low population values values overall; and larger populations in origins with high shares of travel by motorised modes.
-These factors all contribute to the the car/motor mode being most advantaged in capturing spatially available job opportunities overall.
+Figure \ref{fig:Fig5} is a snapshot of the spatial availability as reflected by the multimodal origin-destination flows and travel times from the 2018 travel survey: it incorporates the travel behaviour after LEZ implementation i.e., reduced car trips into the LEZ Centro.
+Furthermore, this implementation of spatial availability assumes that all employment opportunities are of interest to the entire population.
+Also, opportunities are proportionally allocated to mode-using populations based on travel time and mode-using population residing in the zones relative to the total travel time and population.
+
+There are noticeable differences in the magnitude of $V_i^m$ between modes.
+In Figure \ref{fig:Fig5}, the majority of $V_i^m$ are allocated to workers travelling by car and transit.
+This can be expected since users of these modes represent `r (modal_V_comps_4plot$perc[3] + modal_V_comps_4plot$perc[4]) %>% percent(accuracy=0.1)` of the total population.
+The ability to travel at greater speeds also advantages these modes compared to the non-motorized modes.
+However, differences in $V_i^\text{car}$ and $V_i^\text{transit}$ values exist in space: car users outside of the M-30 are allocated greater spatial availability, while some zones inside the M-30 have greater spatial availability for transit.
+Overall, the magnitude of $V_i^m$ values for non-motorized mode-users are lower than for car and transit, but the highest values of $V_i^{bike}$ and $V_i^{walk}$ tend to be found in zones within the M-30 and origins with higher $V_i^{transit}$.
+
+To highlight the spatial differences in modal competitive advantage, Figure \ref{fig:Fig6} displays the spatial availability and population per mode aggregated for three areas of the city.
+Shifting focus to the left-most bars of Figure \ref{fig:Fig6}, motorised mode users can avail `r (modal_V_comps_4plot$perc[7] + modal_V_comps_4plot$perc[8]) %>% percent(accuracy=0.1)` of all jobs in the city (Spatial Availability by Mode).
+However, car users are allocated a disproportionate share of spatial availability relative to their city-wide population.
+The combined population of car and transit users is `r modal_V_comps_4plot$perc[3] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[4] %>% percent(accuracy=0.1)` respectively, and these populations are respectively allocated `r modal_V_comps_4plot$perc[7] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[8] %>% percent(accuracy=0.1)` of the city's job availability.
+When conceptualising the number of opportunities accessed as a finite value (total: 847,574 jobs), fewer opportunities are spatially available to lesser competitive modes.
+Modes are at a disadvantage when relative travel cost is high (see Figure \ref{fig:Fig4}) and the mode-using population is relatively small compared to all populations (especially populations with lower travel costs).
 ```{r, modal-V-comps-plot, fig.cap="\\label{fig:Fig6} Proportion of population by mode and spatial availability of jobs by mode aggregated for three areas.
@@
 
-The big picture demonstrates how, from the 2018 travel survey, car is the most advantageous mode.
-However, it is interesting to notice that this advantage appears to be blunted by the LEZ.
-Unlike the results for the whole city and even within the M-30 ring, the proportion of car users in Centro is larger than the proportion of opportunities spatially available to them.
-The restriction on cars in effect reduces competition by this mode, and leads to a relative increment of the mass effect of the modes allowed within the LEZ, which also contains a large number of jobs (see Figure \ref{fig:Fig2}).
-As summarized in the two right-most columns in Figure \ref{fig:Fig6}, the proportion of jobs spatially available to car in Centro is (`r modal_V_comps_4plot$perc[23] %>% percent(accuracy=0.1)` or 5,373 opportunities).
-For reference, this is less than the proportion of the car users in Centro (`r modal_V_comps_4plot$perc[19] %>% percent(accuracy=0.1)`), evidently less then the proportion of car users in the city, and is the opposite of the overall trend (left-most columns) and within the M-30 (middle columns).
-
-It is also clear that more opportunities are spatially available to non-car users within Centro.
-In the case of active travel, the proportions of cyclists and walkers within LEZ Centro still exceeds the proportions of jobs spatially available to them.
-Notably, this disparity is drastically reduced compared to the rest of the city.
-As seen in Figure \ref{fig:Fig6}, there is a higher proportion of opportunities that are spatially available to pedestrians and cyclists in Centro than in the City overall and in areas within the M-30.
-Within Centro, `r modal_V_comps_4plot$perc[21] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[22] %>% percent(accuracy=0.1)` of opportunities are spatially available to bike and walk modes respectively, while their populations represent  `r modal_V_comps_4plot$perc[1] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[2] %>% percent(accuracy=0.1)` of the population.
-By restricting the ability of cars to enter Centro, the LEZ seems to contribute to leveling the playing field for slower modes, in particular cycling and walking, but also transit.
-As seen in the Figure \ref{fig:Fig6}, transit users are generally close to parity across the region, with nearly as many spatially available jobs as transit users.
-Still, this mode has the greatest advantage in LEZ Centro with `r modal_V_comps_4plot$perc[24] %>% percent(accuracy=0.1)` of spatially available jobs in Centro for `r modal_V_comps_4plot$perc[20] %>% percent(accuracy=0.1)` of transit users in Centro.
-This result makes intuitive sense: after car, it is the mode with the greatest range, and unlike car it is unrestricted in the LEZ Centro.
-
-The spatial differences in the competitive (dis)advantage of spatial availability between modes can also be visualized at a finer level of granularity.
+Though car mode offers the most spatial availability overall, this is not the case within the Centro.
+As summarized in the two right-most bars in Figure \ref{fig:Fig6}, the proportion of jobs spatially available to car users in Centro is `r modal_V_comps_4plot$perc[23] %>% percent(accuracy=0.1)` (5,373 opportunities), less than the proportion of the car users in the Centro (`r modal_V_comps_4plot$perc[19] %>% percent(accuracy=0.1)`).
+The trend in the Centro for car-users is opposite to that of the city overall (left-most bars) and the areas inside the M-30 (middle bars).
+We suspect that car-mode's competitive advantage is blunted by the LEZ: the number of car trips are relatively reduced in the area making non-car modes more competitive.
+Car mode's advantage in the Centro is also diminished by the relative increment of the mass effect and concentration of jobs (Figure \ref{fig:Fig2}) within/around the Centro.
+
+Since car mode is less competitive within the Centro, other modes are relatively _more_ competitive.
+Referring to the active travel modes in the right-most bars in Figure \ref{fig:Fig6}, `r modal_V_comps_4plot$perc[21] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[22] %>% percent(accuracy=0.1)` of opportunities are spatially available to bike and walk modes respectively, while their populations represent  `r modal_V_comps_4plot$perc[1] %>% percent(accuracy=0.1)` and `r modal_V_comps_4plot$perc[2] %>% percent(accuracy=0.1)` of the population.
+The disparity between the proportions of cyclists and walkers and the proportions of jobs spatially available to them is _smaller_ than displayed in the other two aggregations (All and M-30).
+We suspect that by restricting the ability of cars to enter Centro, the LEZ contributes to leveling the playing field for slower modes, in particular cycling and walking but also transit.
+Transit users are generally close to parity city-wide (left-most bars), with nearly as many spatially available jobs as transit users.
+Still, transit mode has the greatest advantage in the Centro with `r modal_V_comps_4plot$perc[24] %>% percent(accuracy=0.1)` of spatially available jobs for `r modal_V_comps_4plot$perc[20] %>% percent(accuracy=0.1)` of transit users in Centro.
+This result makes intuitive sense: after car, transit is the mode with the greatest range and, unlike car, it faced no restrictions by the LEZ.
+
+The spatial differences in the competitive (dis)advantage of spatial availability between modes can also be visualized at a finer level of spatial granularity.
 Figure \ref{fig:Fig7} shows $v_i^m$, the spatial availability $V_i^m$ divided by the population of users of $m$.
@@
 These plots illustrate in unambiguous fashion, and in a quantity that is comparable over space and time, the advantage in terms of spatial availability of car for most of the city (bottom left plot, areas denoted with green $v_i^m$ values above one).
-It can also be observed that spatial availability of jobs is relatively well balanced for transit users over most of the regions (i.e., many zones are light orange or light green).
-Spatial availability of jobs for non-motorised modes, in contrast, is low (under one) overall, although less so within LEZ Centro.
+It can also be observed that spatial availability of jobs is relatively well balanced for transit users over most of the regions (i.e., many zones are light green).
+In contrast, spatial availability of jobs for non-motorised modes is low (under one) overall, although less so within/around the LEZ Centro.
 
@@
 
-Incidentally, $v_i^m$ values for car within and near LEZ Centro is close to or below one in Figure \ref{fig:Fig7}, while all non-car modes have relatively higher $v_i^m$ values.
-Since these values are comparable across regions and over time, Figure \ref{fig:Fig7} potentially provides a benchmark for quantifying changes in LEZ policies in the future.
-As Figure \ref{fig:Fig7} also shows, many areas within the M-30 have high (white/green) $v_i^m$ values for car, but the results for LEZ Centro give reasonable grounds to speculate that a spatial expansion of the LEZ to include all areas within the M-30 would likely increase the spatial availability of jobs for transit users, cyclists and pedestrians.
-
-<!-- NOTE: SOME COMMENTS FOR FUTURE DISCUSSION POINTS ARE OUTLINED:
-
-ANSWER TO THIS PARAGRAPH, V_i is a location-based measure.
-it tells us that the jobs within the centro can be more easily available to non-motorised modes than motorised modes.
-This is in part because of the LEZ but also because of walkability/land -use, of course.
-These results can be explained in a different/better way..
-not sure how for now but....
-The number of trips going into the LEZ Centro is very small relative to all trips.
-This shift in car trips to non-car, I don't think, is the reason why spatial availability is high in the center for non-car modes.
-I think it is higher for non-car modes because it is highly connected with metro and high walkability/ped congestion that make driving more costly anyways.
-However, it is true that the reduction of population using car-mode DOES increase the competitiveness of other modes, just this impact is very small since the LEZ Centro is very small.
-
-Also, I do not talk about walk and cycling mode that much.
-If LEZ M30 is in place, transit likely captures most of these gains?
-(need to demonstrate this).
-But how much will walk and bike modes capture?
-Does transit competite directly with bike/walk (need to check this out somehow).
-How do we demonstrate this...?
-Maybe we show the change in balancing factors..?
-Perhaps we also show what could happen if we reduce the average travel times for bike/walk (increased walkability/bike infrastructure interventions) in addition to LEZ.
-
-Also, the sample sizes vary drastically between modes, with the cycling sample size being really small. How does this impact inter-modal comparisons?
-
-I conclude that restricting car modes leaves more opportunities for all other modes.
-I think this is because transit accessibility is in direct competition with car travel, and not in direct competition with non-motorised modes (need to demonstrate this - perhaps this can be shown with an x-y correlation between v_im valuesSee v_im_per_pop_plot.png (v_im per population) as well as the all_modes_v_im_plot plot.) -->
+Since $v_i^m$ values are comparable across regions and over time, Figure \ref{fig:Fig7} potentially provides a benchmark for quantifying changes in LEZ policies in the future.
+Even within the M-30, $v_i^car$ values are still high (over 1) for most zones.
+These results may give reasonable grounds to speculate that a spatial expansion of the LEZ to include all areas within the M-30 would increase the spatial availability of jobs for transit users, cyclists and pedestrians.
+However, further investigation is needed.
 
@@
 Carey [@careyPriciples1858], and are part of a line of research that includes the work of Ravenstein [@ravensteinLaws1889], Reilly [@reillyMethods1929], Stewart [@stewartInverse1941; @stewartSuggested1947; @stewartDemographic1948], Zipf [@zipfHypothesis1946; @zipfIntercity1946], Wilson [@wilsonFamilty1971], and many others.
-In this way, $S_i^m$, $a_i^m$, and yes, $V_i^m$, can be interpreted as scores of the potential for interaction with opportunities in space.
+In this way, $S_i^m$, $a_i^m$, and $V_i^m$ can be interpreted as scores of the potential for interaction with opportunities in space.
 
 Different accessibility indicators are characterized by how they weight and aggregate opportunities.
-Spatial availability's contribution to the literature is to incorporate a proportional allocation mechanism that essentially constrains the sums to match the number of opportunities in the region; in this way it is a singly-constrained accessibility measure that natively accommodates congestion and competition.
-The effort with spatial availability is in line with previous research on proportional allocation by Paez et al.
-[@paezDemand2019].
+Spatial availability's contribution to the literature is to incorporate a proportional allocation mechanism that essentially constrains the sums to match the number of opportunities in the region: it is a singly-constrained accessibility measure that natively accommodates congestion and competition.
+The effort with spatial availability is in line with previous research on proportional allocation by Páez et al.
+(2019) [@paezDemand2019].
 As initially introduced by Soukhov et al.
-[@soukhovIntroducingSpatialAvailability2023], spatial availability was designed for a homogeneous population traveling by a single mode of transportation.
+(2023) [@soukhovIntroducingSpatialAvailability2023], spatial availability was designed for a homogeneous population traveling by a single mode of transportation.
 In this paper, we extended spatial availability for the case of heterogeneous populations.
@@
 This comparability is preserved between regions and over time.
-The example suggests that once that opportunities are treated as being finite, restrictions to travel by car (due to the LEZ) leave more spatially available opportunities for non-car users.
-This difference for car travel in locations within and immediately around the LEZ Centro seems to increase the number of opportunities spatially available to transit users (transit being the second most competitive mode), as well as the spatial availability from the perspective of non-motorised modes.
-In effect, a policy such as LEZ seems to help improve the spatial availability situation of active travel and transit in the parts of the city where it is implemented.
-
+The example suggests that once opportunities are treated as finite, it could be suggested that restrictions to travel by car (due to the LEZ) leave more spatially available opportunities for non-car users.
+This difference for car travel in locations within/around the LEZ Centro seem to increase the number of opportunities spatially available to transit users (transit being the second most competitive mode) as well as non-motorised modes.
+In effect, a policy such as LEZ appears to help improve the spatial availability situation of active travel and transit mode users in the parts of the city where it is implemented, though additional research is needed.
+To further speculate, the spatial availability allocated to car-users near but outside the LEZ Centro is still relatively high, potentially supporting the case for LEZ expansion from this perspective.
 The purpose of the empirical example is to illustrate the kind of insights that can be derived from the application of multimodal spatial availability.
-But there are some intriguing opportunities for future research.
+
+There are some intriguing opportunities for future research.
 Accessibility indicators are not designed to work as modal split models, and yet, in the case of policies that alter the relative cost of various forms of transportation, one can reasonably expect to see some shifts between modes.
-In our empirical example we used data collected after the introduction of LEZ Centro.
+In our empirical example, we used data collected after the introduction of LEZ Centro.
 However, given a modal split model to predict model shares, accessibility indicators, including spatial availability, can be used to investigate changes to the accessibility landscape.
@@
 Our empirical example presented a snapshot of this, and in future research it may be interesting to investigate changes in spatial availability _between_ policy interventions.
-The expansion of Madrid's LEZ to the ring contained by the M-30 presents an excellent opportunity to do so.
+The plan to expand Madrid's LEZ to the ring contained by the M-30 presents an excellent opportunity.
 Given the intuitive interpretation of spatial availability scores as fractions of opportunities from the total, relative and absolute changes in the spatial availability landscape can be assessed, thus helping to evaluate the implications of policy interventions.
@@
 
-It is worth reiterating that while the focus of this work was on a multimodal extension of spatial availability, the consideration of multiple modes is just one case of heterogeneous populations (i.e., travel by different modes).
-The method itself can easily accommodate other forms of population or opportunity heterogeneity, for example: variations in rich and poor, young and old, types of opportunities.
-Like accessibility measures before it, the potential application of spatial availability measure are as numerous as potential study contexts.
-
-To close, spatial availability is a singly-constrained accessibility measure that considers competition, a way of accounting for opportunity congestion.
-Spatial availability considers competition by allocating opportunities (the subject of the single-constraint) to zones based on zonal proportions.
-If the demand, supply and travel to opportunities is spatially uniform, then competition is uniform and spatial availability will reflect this.
-It has been suggested that opportunity congestion can be seen on a continuum, but how it is to be considered within accessibility research is an ongoing subject of exploration [@vanweeAccessibilityMeasuresCompetition2001; @merlinDoesCompetitionMatter2017; @kelobonyeMeasuringAccessibilitySpatial2020a].
-This is the context in which we present spatial availability and the multimodal extension.
+To close, in this work spatial availability considers competition by allocating opportunities (the subject of the single-constraint) to modal populations in zones based on zonal proportions.
+Opportunity congestion can be seen on a continuum, but how it is to be considered within accessibility research is an ongoing subject of exploration [@vanweeAccessibilityMeasuresCompetition2001; @merlinDoesCompetitionMatter2017; @kelobonyeMeasuringAccessibilitySpatial2020a].
+It is in this context we present spatial availability and its multimodal extension.
+While our focus was on a multiple modes, this consideration is just one case of heterogeneous populations (i.e., travel by different modes).
+The multimodal method itself can easily accommodate other forms of population or opportunity heterogeneity, for example: variations in rich and poor, young and old, types of opportunities.
+Like preceding accessibility measures, the potential applications of spatial availability are as numerous as the potential study contexts.
 

### 2024-02-01 Journal-manuscript/Journal-manuscript.tex
Revised manu draft 2- ready to submit back to reviewers

@@
 \section*{Abstract}
-Recent research has aimed to address the way opportunities are counted
-in accessibility analysis. In conventional accessibility measures,
-opportunities are often multiply counted, which leads to values of
-accessibility that are difficult to interpret. Constraining the
-calculations to match a known quantity ensures that the measurements sum
-up to a predetermined quantity (i.e., the total number of
-opportunities), and so each value can be meaningfully related to this
-total. A recent effort is spatial availability, a singly-constrained
-accessibility measure. In this paper we extend spatial availability for
-use in the case of multiple modes or, more generally, heterogeneous
-population segments with distinct travel behaviors. After deriving a
-multimodal version of spatial availability, we proceed to illustrate its
-features using a synthetic example. Next, we apply it to an empirical
-example of low emission zones in Madrid, Spain. We conclude the paper
-with suggestions for future research and its use in evaluating policy
-interventions.
+Place-based accessibility measures communicate the potential interaction
+with opportunities at a zone that populations can access. Recent
+research has explored the implications of how opportunities are counted
+by different accessibility methods. In conventional measures,
+opportunities are multiply counted if more than one zone offers access
+to the same opportunity. This multi-count of opportunities leads to
+values of accessibility that are difficult to interpret. A possible
+solution to enhance the meaning-making of accessibility results is by
+constraining the calculations to match a known quantity. This ensures
+all zonal values sum up to a predetermined quantity (i.e., the total
+number of opportunities). In this way, each value can be meaningfully
+related to this total. A recent effort that implements this solution is
+spatial availability, a singly-constrained accessibility measure. In
+this paper, we extend spatial availability for use in the case of
+multiple modes or more generally, heterogeneous population segments with
+distinct travel behaviors. After deriving a multimodal version of
+spatial availability, we proceed to illustrate its features using a
+synthetic example. We then apply it to an empirical example of low
+emission zones in Madrid, Spain. We conclude with suggestions for future
+research and its use in evaluating policy interventions.
 
@@
 Accessibility is a key concept in the analysis of land use and
-transportation systems {[}e.g., 1,2,3{]} and is coming of age from the
-perspective of planning research {[}see \emph{inter alia}, 4,5--8{]}.
-Beginning with the work of Hansen {[}1{]}, accessibility measures have
-been widely used to evaluate the efficiency of transportation systems
-when combined with the distribution of populations and opportunities in
-space {[}9{]}. In this way, accessibility is a holistic measure of
-spatial systems' ease of reaching population-relevant destinations
-{[}10,11{]}.
-
-The most common form of accessibility measure is based the gravity model
-{[}e.g. 1{]}: these measures sum ``weighted'' opportunities around a
-focal point (i.e., a potential origin), based on how expensive it is to
-reach them. Recent research in accessibility analysis has explored the
-implications of methods by which opportunities are summed, especially in
-the context of competitive accessibility {[}15{]}. In a typical
-Hansen-type accessibility measure, the sums around origins are not
-constrained. This means that the same opportunity can enter the sum for
-different origins; Counting the same opportunity multiple times treats
-the opportunity as if it was inexhaustible (or non-rival {[}16{]}).
-However, this conceptualisation may not reflect reality as
-opportunities, in general, are not inexhaustible. Some opportunities are
+transportation systems {[}1--3{]} and is coming of age from the
+perspective of planning research {[}see \emph{inter alia} 4,5--8{]}.
+Beginning with the work of Hansen (1959) {[}1{]}, accessibility measures
+have been widely used to evaluate the efficiency of transportation
+systems when combined with the distribution of populations and
+opportunities in space {[}9{]}. In this way, accessibility is a holistic
+measure of spatial systems' ease of reaching population-relevant
+destinations {[}10,11{]}.
+
+The most common form of accessibility measure is based on the gravity
+model e.g.~the Hansen-type measure {[}1{]}: these measures sum
+``weighted'' opportunities around a focal point (i.e., a potential
+origin), based on how expensive it is to reach them. Recent research in
+accessibility analysis has explored the implications of methods by which
+opportunities are summed, especially in the context of competitive
+accessibility {[}15{]}. In a typical Hansen-type accessibility measure,
+the sums around origins are not constrained: the same opportunity can
+enter the sum for different origins. Counting the same opportunity
+multiple times treats it as it was inexhaustible (or non-rival
+{[}16{]}). However, this conceptualisation may not reflect reality, as
+opportunities are often more exclusive than not. Some opportunities are
 by definition exclusive: a typical example is employment, once a job is
@@
 available for any other person to take {[}14,17{]}. More generally, it
-could be conceptualised that all opportunities are subject to some
-amount of congestion or capacity-constraint: multiple people use a
-specific opportunity at a given time and the more people who do, the
-more congested the opportunity. Congestion can be seen on a continuum,
-and debate on how congestion within accessibility should be considered
-is ongoing {[}14,17,18{]}. Nonetheless, competition has been widely
-considered for traditionally exclusive opportunities types such as
-employment and healthcare opportunities {[}14,17,19--21,21--24{]} and
-more recently educational facilities {[}18,25{]}, and less exclusive
-amenities-types such as greenspace and recreation {[}25,26{]}, and
-shopping {[}18,27{]}.
+could be conceptualised that all opportunities are somewhat exclusive so
+they are subject to some amount of congestion or capacity-constraint:
+multiple people use a specific opportunity at a given time and the more
+people who do, the more congested the opportunity. Congestion can be
+seen on a continuum, and debate on how congestion within accessibility
+measures can be considered is ongoing {[}14,17,18{]}. Nonetheless,
+competition has been widely considered for traditionally exclusive
+opportunities types such as employment and healthcare opportunities
+{[}14,17,19--21,21--24{]} and more recently for educational facilities
+{[}18,25{]} as well as less exclusive amenities types such as greenspace
+and recreation amenities {[}25,26{]}, and shopping destinations
+{[}18,27{]}.
 
@@
 motivation for the development of approaches that consider competition
-{[}21,23,24{]} and are popularly applied as floating catchment area
-methods {[}e.g. 28{]}. While these approaches purport to account for
-congestion, Paez et al. {[}12{]} demonstrates that they do not solve the
-issue of multiple counting of opportunities in general, thus leading to
-biases in the calculation of total demand and supply. They sometimes
-inflate counts and other times deflate them. In response to this, recent
-research has paid closer attention to the way opportunities are counted
-in accessibility analysis. Paez et al. {[}12{]}, for example, focuses on
-floating catchment area methods and introduces a normalization of the
-impedance matrix to allocate the population and then the level of
-service proportionally. More recently, Soukhov et al. {[}15{]}
-introduced a singly-constrained measure of accessibility, called spatial
+{[}21,23,24{]} and are popularly applied in the literature as floating
+catchment area methods {[}28{]}. While these approaches purport to
+account for congestion, Páez et al.~(2019) {[}12{]} demonstrates that
+they do not solve the issue of multiple counting of opportunities in
+general, thus leading to biases in the calculation of total demand
+(population) and supply (of opportunities). They sometimes inflate
+counts and other times deflate them. In this line, recent research has
+paid closer attention to the way opportunities are counted in
+accessibility analysis. Páez et al.~(2019) {[}12{]}, for example,
+focuses on floating catchment area methods and introduces a
+normalization of the impedance matrix to allocate the population and
+then the level of service (i.e., the supply to demand) proportionally.
+More recently, Soukhov et al.~(2023) {[}15{]} introduced a
+singly-constrained measure of accessibility, called spatial
 availability, that employs a similar but more sophisticated proportional
@@
 
-A limitation of spatial availability as introduced by Soukhov et al.
-{[}15{]} is that it was developed for the case of a homogeneous
-population, for example for the case of a single mode of transportation.
-However, the finite nature of opportunities makes the analysis of
-heterogeneous populations very relevant. In the case of multiple modes
-of transportation, people who travel by slower modes (e.g., active
-modes) can usually reach fewer opportunities than people who travel by
-faster modes and whose range is typically far wider (e.g., car). This
-implies that slower travelers will often face increased competition for
-local opportunities from travelers who can reach said opportunities from
-farther afield.
+A limitation of spatial availability as introduced by Soukhov et
+al.~(2023) {[}15{]} is that it was developed for the case of a
+homogeneous population, for example for the case of a single mode of
+transportation. However, the finite nature of opportunities makes the
+analysis of heterogeneous populations very relevant. In the case of
+multiple modes of transportation, people who travel by slower modes
+(e.g., active modes) can usually reach fewer opportunities than people
+who travel by faster modes and whose range is typically far wider (e.g.,
+car). This implies that slower travelers will often face increased
+competition for local opportunities from travelers who can reach said
+opportunities from farther afield.
 
@@
 for example: variations in travel behavior between older and younger
-adults {[}e.g., 29{]}, the propensity of older adults to use different
-modes of transportation {[}e.g., 30{]}, the usually shorter trip lengths
-of children compared to adults {[}e.g., 31{]}, or the more limited
-travel ranges of single parents {[}e.g., 32{]}.
-
-The rest of this paper is organized as follows. In Section 2 we provide
-a brief review of multimodal accessibility. In the following Section 3
-we demonstrate the derivation of the new spatial availability expression
-for multiple modes. In Section 4 we illustrate relevant issues through a
+adults (e.g., Páez et al.~(2010) {[}29{]}), the propensity of older
+adults to use different modes of transportation (e.g., Moniruzzaman et
+al.~(2013) {[}30{]}), the usually shorter trip lengths of children
+compared to adults (e.g., Reyes et al.~(2014) {[}31{]}), or the more
+limited travel ranges of single parents (e.g., Páez et al.~(2010)
+{[}32{]}).
+
+The rest of this paper is organized as follows. In Section 2, we provide
+a brief review of multimodal accessibility. In the following Section 3,
+we demonstrate the derivation of the spatial availability expression for
+multiple modes. In Section 4, we illustrate relevant issues through a
 synthetic example. This is followed in Section 5 by an empirical example
@@
 sourced from the city's 2018 travel survey. The empirical example
-demonstrates the differences in multimodal spatial availability
-highlighting differences within and outside the LEZ for travelers using
-different modes, namely car, transit, cycling and walking. In Section 6,
-we provide concluding remarks on the strengths of the use of spatial
-availability as a multimodal accessibility measure, and discuss
+demonstrates the multimodal spatial availability landscape in 2018. It
+highlights the differences within and outside the LEZ for travelers
+using different modes, namely, car, transit, cycling and walking. In
+Section 6, we provide concluding remarks on the strengths of the use of
+spatial availability as a multimodal accessibility measure, and discuss
 potential future uses in policy planning scenarios as well as directions
@@
 
-Location-based accessibility indicators are quantitative measures of
+Place-based accessibility indicators are quantitative measures of
 \emph{potential} interaction with opportunities for locations within a
 given region: they are summary measures of the relationship between
-land-use and transport systems. Arguably, the most commonly used are
-measures based on the gravity model {[}33{]}; this includes cumulative
-opportunity measures which are a special case of the gravity model
-{[}5,10{]}. These gravity model based measures weight opportunities
-depending on how easy it is to reach them. Given an origin (\(i\)) and a
-destination (\(j\)), an impedance function \(f^{m}(c^m_{ij})\) converts
-the cost of travel (e.g., time, money, generalized cost) into a score
-that represents the propensity for potential interaction. These measures
-originate from the work proposed by Hansen {[}1{]}, which can take the
-following form in the multimodal case:
+land-use and transport systems. Arguably, the most commonly used
+measures are based on the gravity model {[}33{]}; this includes
+cumulative opportunity measures that are a special case of the gravity
+model {[}5,10{]}. These gravity model based measures weight
+opportunities depending on the ease of reaching them. Given an origin
+\(i\) and a destination \(j\), an impedance function \(f^{m}(c^m_{ij})\)
+converts the cost of travel (e.g., time, money, generalized cost) into a
+score that represents the propensity for potential interaction. These
+measures originate from the work proposed by Hansen (1959) {[}1{]},
+which can take the following form in the multimodal case:
 \(S_i^m = \sum_j O_j f^m(c_{ij}^m)\). In this form, \(m\) is a set of
 modes that have mode-specific travel costs (\(c_{ij}^m\)) and/or travel
-impedance functions (\(f^m(\cdot)\)).
+impedance functions \(f^m(\cdot)\).
 
@@
 not consider the opportunities as finite. To cite an example, Tahmasbi
-et al. {[}34{]} use Hansen-type accessibility to assess the potential
-interaction with retail locations by three modes: walking, public
-transit, and car (i.e., \(m = w, p, c\)). \(S_i^m\) is the sum of retail
-locations \(j\) that can potentially be reached under the travel
+et al.~(2019) {[}34{]} use Hansen-type accessibility to assess the
+potential interaction with retail locations by three modes: walking,
+public transit, and car (i.e., \(m = w, p, c\)). \(S_i^m\) is the sum of
+retail locations \(j\) that can potentially be reached under the travel
 impedance as calculated for each \(i\) and \(m\). In other words, for
@@
 work, Tahmasbi et al. {[}34{]} show that car travel affords the highest
-\(S_i^{m}\) values in the majority of \(i\), i.e., travelers who use a
+\(S_i^{m}\) values in the majority of \(i\) i.e., travelers who use a
 car can potentially reach more retail opportunities than populations
@@
 modes are often values that are difficult to interpret beyond making
-statements about relative size. For example, Lunke {[}35{]}, researching
-the region of Oslo, reports accessibility scores for car in the order of
-tens of thousands of employment opportunities. The corresponding scores
+statements about relative size. For example, Lunke (2022) {[}35{]}
+reports accessibility scores for car in the order of tens of thousands
+of employment opportunities in the Oslo region. The corresponding scores
 for transit are lower, but still often in the thousands or tens of
@@
 accessibility'', it is unclear what the unconstrained scores mean: is
-having access to but 10,000 jobs by transit insufficient? After all,
-10,000 employment opportunities are still plenty of opportunities. These
-ratios can be found elsewhere in the literature {[}e.g., 36, and
-9,29,32, and 8{]}, and they are useful as relative assessment of when
-some members of the public are better or worse off than others, but they
-are silent on how bad is ``worse off''.
+having access to 10,000 jobs by transit insufficient? After all, 10,000
+employment opportunities are still plenty of opportunities. These ratios
+can be found elsewhere in the literature e.g., Figs. 7, 8, and 9 in Páez
+et al.~(2010) {[}36{]}, Fig. 5 in Páez et al.~(2010) {[}29{]}, Páez et
+al.~(2013) Figs. 6, 7, and 8 in {[}32{]}. They are useful as relative
+assessment of when some members of the public are better or worse off
+than others, but they are silent on how bad is ``worse off''.
 
@@
 Hansen-type accessibility that has been standardised in this way is not
-uncommon in the literature {[}e.g., 37,38{]}.
-
-Once we understand opportunities to be finite, it is possible for an
-accessibility measure to take on a crisper meaning. As considered in the
-long tradition of accessibility research, capacity of opportunities is
-limited and thus is subject to competition by population
-{[}14,15,17--25,39--42{]}. There are only so many school-seats, hospital
-capacity, employment opportunities, etc., in a region and if one person
-reaches an opportunity, it is taken: the supply of an opportunity and
-the demand for that opportunity are two components of accessibility.
-These are clear examples of opportunities that are unambiguously
-competitive. But we would go as far as to argue that every type of
-opportunity is subject to congestion or capacity constraint, even when
-the opportunities are conventionally seen as non-competitive.
+uncommon in the literature {[}37,38{]}.
+
+If we understand opportunities to be finite and/or subject to some
+levels of congestion, it is possible for an accessibility measure to
+take on a crisper meaning. Accessibility research has a history of
+considering opportunity competition, especially regarding school-seats,
+hospital capacity, and employment opportunities
+{[}14,15,17--25,39--42{]}. If one person reaches an opportunity - it is
+taken: the supply of an opportunity and the demand for that opportunity
+are the nodes in accessibility analysis. These types of opportunities
+are unambiguously exclusive. But we would go as far as to argue that
+every type of opportunity is subject to congestion or capacity
+constraint, even when the opportunities are conventionally seen as
+inexhaustible.
 
@@
 providing green spaces are often stated in the form of \emph{exclusive
-access}, in units of amenity per capita. A case in point is a the
-Ile-de-France region, a jurisdiction that suggested in a 2013 planning
-document that at the municipal level at least 10 \(m^2\) of public green
-space should be supplied \emph{per inhabitant} {[}43{]}. Green spaces
-are not evenly distributed, which means that who has access to them
-hinges on where they are and how easy is to reach them. This formulation
-of provision of amenities is not unusual. For example, Natural England,
-an organization that recommends an Accessible Natural Greenspace
-Standard such that the minimum supply of space is one ha of statutory
-Local Nature Reserves per thousand population\footnote{see
-  https://redfrogforum.org/wp-content/uploads/2019/11/67-Nature-Nearby\%E2\%80\%99-Accessible-Natural-Greenspace-Guidance.pdf}.
-Similarly, the World Health Organization {[}cited in 44{]} recommends
-that cities provide a minimum of 9 \(m^2\) of green area per inhabitant.
-For our purposes, standards of this type translate into ``how much of
-this resource is available to one individual that has not been claimed
-by anyone else?''. Green spaces often have large capacities, but they
-still have a capacity, and it is not the same for a person to have
-access to 5 \(m^2\) of \emph{uncongested} green space as 15 \(m^2\).
-This difference is in fact a matter of justice {[}43,45{]}. Constraining
-accessibility is in this way a useful way to evaluate the congested
-availability of any type of opportunity. As development of sound
-standards is emphasized in the planning literature, in particular in
-regards to fairness in transportation {[}see 46{]}, spatial availability
-analysis is a useful way to develop and assess standards.
+access}, in units of amenity per capita. For example, a 2013 planning
+document for the Ile-de-France region suggested a public green space
+municipal standard of 10 \(m^2\) \emph{per inhabitant} {[}43{]}. Green
+spaces are not evenly distributed, meaning those who have access to them
+depends on where they are and how easy they are to reach. This
+formulation of amenity provision is not unusual. As another example,
+Natural England recommends a national ``accessible natural greenspace
+standard'' such that the minimum supply of space is 1 \(ha\) of
+statutory local nature reserves per thousand population {[}44{]}.
+Similarly, the World Health Organization {[}45{]} recommends that cities
+provide a minimum of 9 \(m^2\) of green area per inhabitant. For our
+purposes, standards of this type translate into ``how much of this
+resource is available to one individual that has not been claimed by
+anyone else?''. Green spaces often have large capacities, but they still
+have a capacity and it is not the same for a person to have access to 5
+\(m^2\) of \emph{uncongested} green space as 15 \(m^2\). This difference
+is in fact a matter of justice {[}43,46{]}. Constraining accessibility
+is a useful way to evaluate the congested availability of any type of
+opportunity. As development of sound standards is emphasized in the
+planning literature, in particular in regards to fairness in
+transportation {[}47{]}, spatial availability analysis can be used to
+develop and assess standards.
 
@@
 rationale for developing floating catchment area methods (FCA),
-popularized by Luo et al. {[}28{]} who reformulated the work of Shen
-{[}24{]} into two steps (although similar, and earlier, developments are
-found in {[}21,23{]}). Shen-type accessibility is formulated as:
-\(a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}\) where \(D_j^m\)
-is the potential demand for opportunities equal to travel impedance
-weighted population \(\sum_i P_i^m f^m(c_{ij}^m)\) and the remaining
-variables are repeated in the Hansen-type measure. Shen-type modal
-accessibility (\(a_i^m\)) can be understood as a ratio of the travel
-impedance-weighted supply of opportunities for \(m\)-mode in \(i\) over
-the travel impedance-weighted demand for opportunities. In this way, it
-considers competition. That said, the measure remains unconstrained,
-meaning both population \emph{and} opportunities are multiply counted
-{[}see 12{]}. In other words, interpretation of the Shen-type
-accessibility scores between modes is fraught, as it is for Hansen-type
-measures.
-
-To illustrate, {[}47{]} calculates \(a_i^m\) to jobs for different
-income-group populations in Shenzhen, China using
-\(m = \text{public transit}\) and \(m=\text{car}\). Their results
-indicate that zones with low-income populations have lower \(a_i^m\)
-than zones with higher-income populations. Further, they show that
-\(a_i^{m=\text{public transit}}\) is lower than \(a_i^{m=\text{car}}\)
-at many zones, arguing that this may further place those zones with
-lower-income populations at a disadvantage. \(a_i\) and/or \(a_i^m\) are
-used to compare relative spatial differences in overall competitive
-accessibility and multimodal competitive accessibility, but because
-opportunities were doubly counted (entering the sums of both modes),
-this makes for uneasy interpretations of the differences in \(a_i^{m}\)
-between modes. Questions that this approach leaves unaddressed include:
-what is the impact of competition on the difference in \(a_i^m\) values?
-How does the impact vary spatially? And what is the interpretation of
-this difference?
+popularized by Luo et al.~(2003) {[}28{]} who reformulated the work of
+Shen (1998) {[}24{]} into two steps (although similar, and earlier,
+developments are found in {[}21,23{]}). Shen-type accessibility is
+formulated as: \(a_i^m = \sum_j \frac{O_jf^m(c_{ij}^m)}{\sum_m D_j^m}\)
+where \(D_j^m\) is the potential demand for opportunities equal to
+travel impedance weighted population \(\sum_i P_i^m f^m(c_{ij}^m)\) and
+the remaining variables are repeated in the Hansen-type measure.
+Shen-type modal accessibility (\(a_i^m\)) can be understood as a ratio
+of the travel impedance-weighted supply of opportunities for \(m\)-mode
+in \(i\) over the travel impedance-weighted demand for opportunities. In
+this way, it considers competition. That said, the measure remains
+unconstrained, meaning both population \emph{and} opportunities are
+multiply counted {[}12{]}. In other words, interpretation of the
+Shen-type accessibility scores between modes is fraught, as it is for
+Hansen-type measures.
+
+To illustrate, Tao et al.~(2020) {[}48{]} calculates \(a_i^m\) to jobs
+for different income-group populations in Shenzhen, China for those
+using transit and car. Their results indicate that zones with low-income
+populations have lower \(a_i^m\) than zones with higher-income
+populations. Further, they show that \(a_i^{\text{transit}}\) is lower
+than \(a_i^{\text{car}}\) in many zones, arguing that this may further
+place those zones with lower-income populations at a disadvantage.
+\(a_i\) and/or \(a_i^m\) are used to compare relative spatial
+differences in overall competitive accessibility and multimodal
+competitive accessibility, but because opportunities were doubly counted
+(entering the sums of both modes), this makes for uneasy interpretations
+of the differences in \(a_i^{m}\) between modes. Questions that this
+approach leaves unaddressed include: what is the impact of competition
+on the difference in \(a_i^m\) values? How does the impact vary
+spatially? And what is the interpretation of this difference?
 
@@
 proportional allocation factors that follow well established principles
-of spatial interaction and the gravity model {[}see 48{]}. In Soukhov et
-al. {[}15{]} these factors consider: the mass effect (e.g., the size of
-populations at different origins); and the cost of travel from different
-zones (e.g., some sub-populations face relatively higher or lower
-costs). The following section introduces the multimodal form of spatial
-availability.
+of spatial interaction and the gravity model {[}49{]}. In Soukhov et
+al.~(2023) {[}15{]} these factors consider: the mass effect (e.g., the
+size of populations at different origins); and the cost of travel from
+different zones (e.g., some sub-populations face relatively higher or
+lower costs). The following section introduces the multimodal form of
+spatial availability.
 
@@
 
-In brief, we define the spatial availability at an origin \(i\)
-(\(V_{i}\)) as the proportion of all opportunities in the region that
-are allocated to location \(i\) from all destinations \(j\). \(V_{i}\)
-is a value of how many opportunities are available to each location
-\(i\) out of all the opportunities in the region. The general
-formulation of spatial availability \(V_{i}\) is shown in Equation
-(\ref{eq:spatial-availability-general}) {[}see 15{]}:
+In brief, we define the spatial availability \(V_{i}\) at an origin as
+the proportion of all opportunities in the region that are allocated to
+origin \(i\) from all destinations \(j\). The general formulation of
+spatial availability is shown in Equation
+(\ref{eq:spatial-availability-general}) {[}15{]}:
 
@@
   the sum of \(V_{i}\) is identical to the total number of opportunities
-  in the region (i.e., \(\sum_j O_j = \sum_i V_i\)); in other words,
-  opportunities are dealt with as finite resources.
+  in the region (i.e., \(\sum_j O_j = \sum_i V_i\)).
 \end{itemize}
@@
 the weight used in the sum (balancing factor \(F^t_{ij}\)) implements a
-proportional allocation mechanisms to ensure that the sum of \(V_i\) is
+proportional allocation mechanism to ensure that the sum of \(V_i\) is
 constrained to match the total number of opportunities in the region. As
@@
 
-Implementing \(F^{tm}_{ij}\), the following Equation
+Implementing \(F^{tm}_{ij}\), Equation
 (\ref{eq:spatial-availability-multimodal}) gives the multimodal version
@@
 larger than the Satellite \(C\). \(B\) has the largest \(x\)-using
-population, followed by then \(A\), then \(C\). This synthetic example
-is inspired by the single-mode example used in {[}24{]} and reconfigured
-in {[}15{]}.
+population (40\%), followed by \(A\) (33\%) then \(C\) (30\%). This
+synthetic example is inspired by the single-mode example used in Shen
+(1998) {[}24{]} and reconfigured in Soukhov et al.~(2023) {[}15{]}.
 
@@
 opportunities in the region (\(210,000\) jobs), the sub-population that
-is most proximate to jobs (low cost to reach), furthest from large
-populations (less competition), and uses the fastest mode \(z\) (greater
-range) can potentially reach the largest number of opportunities. This
-appears to be the sub-population at \(A\) using mode \(z\).
-Sub-populations located in opposite conditions (i.e., more distant from
-jobs, close to large populations, and using slower mode \(x\)) are at a
-relative disadvantage. The competition for opportunities between
-different mode-using populations matters as it reflects how well the
-land-use and transport system serves (or does not serve) certain
+is most proximate to jobs (lowest cost to reach), furthest from large
+populations (least competition), and uses the fastest mode \(z\)
+(greatest range) can potentially reach the largest number of
+opportunities. This appears to be the sub-population at \(A\) using mode
+\(z\). Sub-populations located in opposite conditions (i.e., more
+distant from jobs, close to large populations, and using slower mode
+\(x\)) are at a relative disadvantage. The competition for opportunities
+between different mode-using populations matters as it reflects how well
+the land-use and transport system serves (or does not serve) certain
 populations.
@@
 columns and are aggregated for each \(i\) in the final two columns in
-Table 1 . As in the example in Shen {[}24{]}, we use a negative
+Table 1 . As in the example in Shen (1998) {[}24{]}, we use a negative
 exponential impedance function
@@
 travel times are different for the two modes. More generally, it is
-possible to use different impedance functions for the modes, as
-demonstrated in the empirical example in the following section.
+possible to use different impedance functions for the modes, as the
+empirical example in the following section demonstrates.
 
@@
 competition, or alternatively the assumption of an inexhaustible
-resource in the calculation of \(S_i^m\), lead to a curious result.
+resource in the calculation of \(S_i^m\), leds to a curious result.
 Since the populations in \(A\) and \(B\) have the same travel impedance
@@
 To connect this example to the aforementioned literature, \(S_i^m\) is
-calculated in the work of Tahmasbi et al. {[}34{]}; they contrast
+calculated in the work of Tahmasbi et al.~(2019) {[}34{]}; they contrast
 differences in \(S_i^m\) values between modes in a relative and
@@
 the sixth column, the values face similar interpretability issues. The
-Shen-type measure is implemented in aforementioned work of Tao et al.
-{[}47{]} to calculate modal \(a_i^m\) values and the aggregated \(a_i\)
-is implemented in the work of Carpentier et al. {[}49{]}. However,
-similar to Hansen-type accessibility, these works discuss relative and
-spatially comparative differences in values, but veer from interpreting
-the values of \(a_i^m\) or \(a_i\) themselves. In fairness,
-interpretation is complicated by the multiple counting of opportunities
-between zones and modes.
+Shen-type measure is implemented in aforementioned work of Tao et
+al.~(2020) {[}48{]} to calculate modal \(a_i^m\) values and the
+aggregated \(a_i\) is implemented in the work of Carpentier et
+al.~(2020) {[}50{]}. However, similar to Hansen-type accessibility,
+these works discuss relative and spatially comparative differences in
+values, but veer from interpreting the values of \(a_i^m\) or \(a_i\)
+themselves. In fairness, interpretation is complicated by the multiple
+counting of opportunities between zones and modes.
 
@@
 number of opportunities in the region (i.e., \(210,000\) jobs). Seen in
-fifth column of Table 1 , the values of \(V_i^m\) in \(A\) and \(B\) are
-not the same within each mode (as this measure considers competition).
-In fact, at \(A\), users of mode \(z\) capture 36,088.84 more spatially
-available jobs (of the \(210,000\) jobs in the region) than the
-sub-population travelling by \(x\). The numerical difference is clear
-since it refers to opportunities out of the total.
+fifth column of Table 1 , the values of \(V_i^m\) for \(A\) and \(B\)
+are not the same within each mode (as this measure considers
+competition). In fact, at \(A\), users of mode \(z\) capture 36,088.84
+more spatially available jobs (of the \(210,000\) jobs in the region)
+than the sub-population travelling by the slower-mode \(x\). The
+numerical difference is clear since it refers to opportunities out of
+the total.
 
@@
 and compared between zones (\(V_i = \sum_m{\sum_i{V_i^m}}\)). This
-aggregation, \(V_i\), is shown in the seventh column in Table 1 . Again
-looking at center \(A\), \(A\) is allocated 67,482.61 spatially
-available opportunities for both modes. 77\% of this spatial
-availability allocated to \(A\) is assigned to users of mode \(z\)
-despite representing 66\% of \(A\)'s population.
+aggregation, \(V_i\), is shown in the seventh column in Table 1 . \(A\)
+is allocated 67,482.61 spatially available opportunities for both modes.
+77\% of this spatial availability allocated to \(A\) is assigned to
+users of mode \(z\) despite representing 66\% of \(A\)'s population.
 
@@
 for 73\% of the region's total spatial availability - while the
-remaining 27\% is allocated to users of mode \(x\) who are 38\%of the
+remaining 27\% is allocated to users of mode \(x\) who are 38\% of the
 total population. Notably, the population who uses \(x\) have 11\% fewer
@@
 
-Since spatial availability is constrained and has an interpretable
-meaning as a proportion of the total opportunities in the region, the
-values at \(i\) have a straighforward interpretation. Inequality in
-\(V_i^m\) values can be explored through a variety of approaches. For
-instance, consider travel times. The population of travelers who use
-\(z\) accounts for 67\% of the potential travel time traveled in the
-region: this is 7\% less travel time than the proportion of spatial
-available opportunities that is allocated to them. In other words, the
-population of users of \(z\) travels fewer minutes overall and has more
-spatial availability of opportunities than users of the slower mode
-\(x\).
+Since spatial availability is constrained to the total opportunities in
+the region, the values at \(i\) have a straighforward interpretation.
+Inequality in \(V_i^m\) values can be explored through a variety of
+approaches. For instance, consider travel times. The population of
+travelers who use \(z\) accounts for 67\% of the potential travel time
+traveled in the region: this is 7\% less travel time than the proportion
+of spatial available opportunities that is allocated to them. In other
+words, the population of users of \(z\) travels fewer minutes overall
+and has more spatial availability of opportunities than users of the
+slower mode \(x\).
 
@@
 to} the mass effect (occurring at \(B\), high population density) and
-high travel impedance (occurring at the Satellite \(C\)).
+high travel impedance (occurring at the Satellite \(C\)). Notably,
+\(v_i^m\) values are equal to the ratios of the Shen-type measure
+\(a_i^m\) in this case, however they take on a different interpretation.
 
@@
 to reduce the values of \(v_i^z\) (making it slower or more expensive)
-and increase he values of \(v_i^x\) (making it faster or les expensive).
-The purpose of this synthetic demonstration is to show how spatial
-availability can be used to quantify the competitive (dis)advantage in a
-multimodal application. In what follows, we demonstrate the use of
-multimodal spatial availability through an empirical example.
+and increase the values of \(v_i^x\) (making it faster or less
+expensive). The purpose of this synthetic demonstration is to show how
+spatial availability can be used to quantify the competitive
+(dis)advantage in a multimodal application. In what follows, we
+demonstrate the use of multimodal spatial availability through an
+empirical example.
 
@@
 
-The context for the empirical example is Madrid, Spain. This city
-implemented a Low Emission Zone (LEZ) in 2017 to pursue goals set out in
-the national climate change agenda, cut nitrogen dioxide levels, and to
-prioritize people's movement in the city. LEZs elsewhere have similarly
-been implemented as interventions to reduce GHG emissions, improve air
-quality, and support sustainable mobility {[}50,51{]}. Though the rules
-of exclusion vary by city, LEZs aim to deter/reduce traffic in
-designated zones under threat of penalty (e.g., fines, seizure of
-vehicle). In other words, LEZs implement a form of \emph{geographic
-discrimination} as they change how people can reach opportunities by
-making it more costly for some forms of travel, typically cars, to
-circulate in predetermined zones. When considering opportunities as
-finite in a region, this discrimination reduces the competition of one
-mode and opens up opportunities for other modes to better thrive. At
-their core, LEZs operate by changing the accessibility landscape of a
-city from the perspective of multiple modes.
+The context for the empirical example is the city of Madrid, Spain. This
+city implemented a Low Emission Zone (LEZ) in 2017 to pursue goals set
+out in the national climate change agenda such as cutting nitrogen
+dioxide levels and prioritizing people's movement in the city. LEZs
+elsewhere have similarly been implemented as interventions to reduce GHG
+emissions, improve air quality, and support sustainable mobility
+{[}51,52{]}. Though the rules of exclusion vary by city, LEZs aim to
+deter/reduce traffic in designated zones under threat of penalty (e.g.,
+fines, seizure of vehicle). In other words, LEZs implement a form of
+\emph{geographic discrimination} as they change how people can reach
+opportunities by making it more costly for some forms of travel,
+typically cars, to circulate in predetermined zones. When considering
+opportunities as finite in a region, this discrimination reduces the
+competition of one mode and opens up opportunities for other modes to
+better thrive. At their core, LEZs operate by changing the accessibility
+landscape of a city from the perspective of multiple modes.
 
@@
 these boundaries to the area inside the M-30, an orbital highway in
-proximity to the city center (i.e., LEZ M-30). Within the 2017 LEZ
-Centro implementation, all cars, motorcycles and freight vehicles with
-environmental labels A or B (older makes and models of fossil fuel
-internal combustion engine vehicles), were disallowed from entering the
-zone unless they are used by residents or meet other exemptions. This
-restriction impacted approximately half of all car trips that used to
-travel into what is now the LEZ Centro {[}52{]}.
-
-For this case study, we quantify spatial availability to opportunities
-by different modes in Madrid. Particularly, we demonstrate how \(V_i^m\)
-can be used to derive insights into how the restriction of car mobility
-in areas around/within the LEZ Centro may have allowed more sustainable
-(but often slower or more costly modes) to become more competitive.
+proximity to the city center. Within the 2017 LEZ Centro implementation,
+all cars, motorcycles and freight vehicles with environmental labels A
+or B (older makes and models of fossil fuel internal combustion engine
+vehicles), were disallowed from entering the zone unless they are used
+by residents or meet other exemptions. This restriction impacted
+approximately half of all car trips that used to travel into what is now
+the LEZ Centro {[}53{]}.
+
+The purpose of this empirical example is to quantify spatial
+availability to employment opportunities by different modes in Madrid
+after LEZ Centro implementation. Particularly, we demonstrate how
+\(V_i^m\) can be used to illustrate the spatial availability advantage
+that more sustainable modes (that are often slower than car) gain
+within/around the LEZ Centro. We speculate on how this competitive
+advantage is heightened as a result of the LEZ implementation that
+restricts car mobility.
 
@@
 The source of origin-destination data for our empirical example is the
-2018 Travel Survey of the Community of Madrid {[}53{]}. This is a
+2018 Travel Survey of the Community of Madrid {[}54{]}. This is a
 representative survey that offers a snap-shot of travel patterns for a
@@
 shapefiles are available from the Community of Madrid open data portal
-{[}54{]}. The pink boundary represents the LEZ Centro in effect in 2017
+{[}55{]}. The pink boundary represents the LEZ Centro in effect in 2017
 and reflected in the 2018 travel survey. The purple boundary represents
@@
 \item
-  Walk: walking or by foot.
+  Walk: pedestrian mode.
 \end{itemize}
@@
 Some aggregation of modes is necessary to calculate the travel impedance
-functions by mode. From Figure \ref{fig:Fig2}, it can be seen that the
-largest concentration of jobs is within, near, and to the north of LEZ
-Centro. The populations with access to those jobs by mode (Figure
-\ref{fig:Fig3}) are spatially distinct. Travel by car and transit
-represent 37\% and 47\% of the modal share respectively. The population
-that travels by transit is more spatially distributed than those using
-cars - particularly near and within LEZ Centro. This distribution is
-likely caused by a variety of factors including: transit coverage and
-service within with city, effective car infrastructure outside of the
-M-30, and/or the impact of the LEZ Centro itself. From Figure
-\ref{fig:Fig3}, it can be seen that active travel is less common than
-motorised trips at 1\% and 15\% for cycling and walking respectively.
-Noticeably, there is a positive trend between the walking and cycling in
-zones where transit is also present. This positive trend is higher than
-for car trip populations.
+functions by mode. From Figure \ref{fig:Fig2}, the largest concentration
+of jobs is within, near, and to the north of LEZ Centro. The populations
+with access to those jobs by mode (Figure \ref{fig:Fig3}) are spatially
+distinct. Travel by car and transit represent 37\% and 47\% of the modal
+share respectively. The population that travels by transit is more
+spatially distributed than those using cars - particularly near and
+within LEZ Centro. This distribution is likely caused by a variety of
+factors including: transit coverage and service within with city,
+effective car infrastructure outside of the M-30, and/or the impact of
+the LEZ Centro itself. From Figure \ref{fig:Fig3}, active travel is less
+common than motorised trips at 1\% and 15\% for cycling and walking
+respectively. Noticeably, there is a positive trend between the walking
+and cycling in zones where transit is also present. This positive trend
+is higher than for car-using populations.
 
@@
 bins. This distribution is then used to estimate the parameters of a
-function for the travel impedance {[}as done in 55,56,57{]}. To fit the
+function for the travel impedance {[}as done in 56,57,58{]}. To fit the
 impedance functions, we use the Maximum likelihood estimation and the
-Nelder-Mead method for direct optimization available within the R
-\{fitdistrplus\} package {[}58{]}. Based on goodness-of-fit criteria and
-associated diagnostics, the gamma and log-normal probability density
+Nelder-Mead method for direct optimization available within the
+\{fitdistrplus\} R package {[}59{]}. Based on goodness-of-fit criteria
+and associated diagnostics, the gamma and log-normal probability density
 functions are selected as best fitting curves for the motorised and
 non-motorised modes respectively. The selection of functional forms
-aligns with empirical examples in other regions {[}15,59,60{]}. The
+aligns with empirical examples in other regions {[}15,60,61{]}. The
 shape and rate parameters for the gamma functions (motorised modes) are
@@
 
-It is worthwhile reiterating that the empirical example is a snapshot of
-spatial availability by mode using modal home-work origin-destination
-data from the 2018 travel survey. It reflects the travel behaviour after
-LEZ implementation, i.e.~very few motorists commute into the LEZ centro.
-Our purpose in this empirical example is to investigate the trends in
-spatial availability of employment opportunities by mode, and illustrate
-how spatial availability can be used to discuss modal competitive
-advantage. We also discuss how multimodal spatial availability landscape
-displayed in the Figure \ref{fig:Fig5} may have been impacted by the
-LEZ.
+Figure \ref{fig:Fig5} illustrates the multimodal spatial availability
+landscape for each of the four modes \(m\) at the level of traffic
+analysis zones \(i\). \(V_i^m\) is a proportion of the total number of
+847,574 jobs in the city. Since \(V_i^m\) is calculated based on the
+population of workers and the distribution of jobs, the values can be
+understood as the number of full-time jobs that are spatially available
+to full-time workers at that \(i\) traveling by mode \(m\), relative to
+all the jobs in the city.
 
@@
 
-In Figure \ref{fig:Fig5}, the spatial availability of jobs \(V_i^m\) is
-calculated for each of the four modes \(m\) at the level of traffic
-analysis zones \(i\) in Madrid. \(V_i^m\) is a proportion of the total
-number of 847,574 jobs in the region. Since \(V_i^m\) is calculated
-based on the population of workers and the distribution of jobs, the
-values can be understood as the number of full-time jobs that are
-spatially available to full-time workers at that \(i\) traveling by mode
-\(m\), relative to all the jobs in the city. This implementation of
-spatial availability assumes that all employment opportunities are of
-interest to the entire population but are proportional allocated based
-on relative travel cost (informed by observed modal origin-destination
-travel behaviour) and the relative amount of the mode-using population
-residing in the zones.
+Figure \ref{fig:Fig5} is a snapshot of the spatial availability as
+reflected by the multimodal origin-destination flows and travel times
+from the 2018 travel survey: it incorporates the travel behaviour after
+LEZ implementation i.e., reduced car trips into the LEZ Centro.
+Furthermore, this implementation of spatial availability assumes that
+all employment opportunities are of interest to the entire population.
+Also, opportunities are proportionally allocated to mode-using
+populations based on travel time and mode-using population residing in
+the zones relative to the total travel time and population.
 
 There are noticeable differences in the magnitude of \(V_i^m\) between
-modes in Figure \ref{fig:Fig5}. The majority of \(V_i^m\) (which is to
-say of spatially available jobs) are allocated to workers travelling by
-car and transit. In a way, this is to be expected since users of these
-modes represent 84.1\% of the total population. However, the ability to
-travel at greater speeds also impacts these results. Furthermore,
-differences in \(V_i^m\) values within modes also exist in space: car
-users outside of the M-30 region appear to enjoy greater spatial
-availability, while some zones inside the M-30 have greater spatial
-availability for transit. Overall, the magnitude of \(V_i^m\) values for
-cyclists and pedestrians are lower than for car and transit but the
-highest values of \(V_i^{bike}\) and \(V_i^{walk}\) tend to be found in
-zones within the M-30 and origins with higher spatial availability by
-transit.
-
-The differences between the shares of modes and their shares of
-spatially available opportunities highlights the competitive advantage
-of certain modes, although this effect is not geographically uniform. As
-seen in the left-most columns of Figure \ref{fig:Fig6}, users of
-`car/motor' and `transit' together can avail 95.3\% of all jobs in the
-city (Spatial Availability by Mode). However, car/motor users have a
-disproportionate share of \(V_i^m\) relative to the population of users
-of this mode (Population by Mode), compared to opportunities that are
-spatially available to transit users. The combined population of car and
-transit users is 36.6\% and 47.5\% respectively, but these populations
-are allocated 48.0\% and 47.3\% respectively, of the city's jobs. When
-treating the number of opportunities that can be reached as a finite
-value (total: 847,574 opportunities), fewer opportunities are spatially
-available to slower modes (i.e., walking and cycling), even taking into
-account that their share is smaller overall. These modes are at a
-disadvantage as a result of: the travel impedance for longer trips (see
-Figure \ref{fig:Fig4}; their low population values values overall; and
-larger populations in origins with high shares of travel by motorised
-modes. These factors all contribute to the the car/motor mode being most
-advantaged in capturing spatially available job opportunities overall.
+modes. In Figure \ref{fig:Fig5}, the majority of \(V_i^m\) are allocated
+to workers travelling by car and transit. This can be expected since
+users of these modes represent 84.1\% of the total population. The
+ability to travel at greater speeds also advantages these modes compared
+to the non-motorized modes. However, differences in \(V_i^\text{car}\)
+and \(V_i^\text{transit}\) values exist in space: car users outside of
+the M-30 are allocated greater spatial availability, while some zones
+inside the M-30 have greater spatial availability for transit. Overall,
+the magnitude of \(V_i^m\) values for non-motorized mode-users are lower
+than for car and transit, but the highest values of \(V_i^{bike}\) and
+\(V_i^{walk}\) tend to be found in zones within the M-30 and origins
+with higher \(V_i^{transit}\).
+
+To highlight the spatial differences in modal competitive advantage,
+Figure \ref{fig:Fig6} displays the spatial availability and population
+per mode aggregated for three areas of the city. Shifting focus to the
+left-most bars of Figure \ref{fig:Fig6}, motorised mode users can avail
+95.3\% of all jobs in the city (Spatial Availability by Mode). However,
+car users are allocated a disproportionate share of spatial availability
+relative to their city-wide population. The combined population of car
+and transit users is 36.6\% and 47.5\% respectively, and these
+populations are respectively allocated 48.0\% and 47.3\% of the city's
+job availability. When conceptualising the number of opportunities
+accessed as a finite value (total: 847,574 jobs), fewer opportunities
+are spatially available to lesser competitive modes. Modes are at a
+disadvantage when relative travel cost is high (see Figure
+\ref{fig:Fig4}) and the mode-using population is relatively small
+compared to all populations (especially populations with lower travel
+costs).
 
@@
 
-The big picture demonstrates how, from the 2018 travel survey, car is
-the most advantageous mode. However, it is interesting to notice that
-this advantage appears to be blunted by the LEZ. Unlike the results for
-the whole city and even within the M-30 ring, the proportion of car
-users in Centro is larger than the proportion of opportunities spatially
-available to them. The restriction on cars in effect reduces competition
-by this mode, and leads to a relative increment of the mass effect of
-the modes allowed within the LEZ, which also contains a large number of
-jobs (see Figure \ref{fig:Fig2}). As summarized in the two right-most
-columns in Figure \ref{fig:Fig6}, the proportion of jobs spatially
-available to car in Centro is (13.8\% or 5,373 opportunities). For
-reference, this is less than the proportion of the car users in Centro
-(16.1\%), evidently less then the proportion of car users in the city,
-and is the opposite of the overall trend (left-most columns) and within
-the M-30 (middle columns).
-
-It is also clear that more opportunities are spatially available to
-non-car users within Centro. In the case of active travel, the
-proportions of cyclists and walkers within LEZ Centro still exceeds the
-proportions of jobs spatially available to them. Notably, this disparity
-is drastically reduced compared to the rest of the city. As seen in
-Figure \ref{fig:Fig6}, there is a higher proportion of opportunities
-that are spatially available to pedestrians and cyclists in Centro than
-in the City overall and in areas within the M-30. Within Centro, 1.7\%
-and 16.1\% of opportunities are spatially available to bike and walk
-modes respectively, while their populations represent 1.2\% and 14.7\%
-of the population. By restricting the ability of cars to enter Centro,
-the LEZ seems to contribute to leveling the playing field for slower
-modes, in particular cycling and walking, but also transit. As seen in
-the Figure \ref{fig:Fig6}, transit users are generally close to parity
-across the region, with nearly as many spatially available jobs as
-transit users. Still, this mode has the greatest advantage in LEZ Centro
-with 68.5\% of spatially available jobs in Centro for 51.4\% of transit
-users in Centro. This result makes intuitive sense: after car, it is the
-mode with the greatest range, and unlike car it is unrestricted in the
-LEZ Centro.
+Though car mode offers the most spatial availability overall, this is
+not the case within the Centro. As summarized in the two right-most bars
+in Figure \ref{fig:Fig6}, the proportion of jobs spatially available to
+car users in Centro is 13.8\% (5,373 opportunities), less than the
+proportion of the car users in the Centro (16.1\%). The trend in the
+Centro for car-users is opposite to that of the city overall (left-most
+bars) and the areas inside the M-30 (middle bars). We suspect that
+car-mode's competitive advantage is blunted by the LEZ: the number of
+car trips are relatively reduced in the area making non-car modes more
+competitive. Car mode's advantage in the Centro is also diminished by
+the relative increment of the mass effect and concentration of jobs
+(Figure \ref{fig:Fig2}) within/around the Centro.
+
+Since car mode is less competitive within the Centro, other modes are
+relatively \emph{more} competitive. Referring to the active travel modes
+in the right-most bars in Figure \ref{fig:Fig6}, 1.7\% and 16.1\% of
+opportunities are spatially available to bike and walk modes
+respectively, while their populations represent 1.2\% and 14.7\% of the
+population. The disparity between the proportions of cyclists and
+walkers and the proportions of jobs spatially available to them is
+\emph{smaller} than displayed in the other two aggregations (All and
+M-30). We suspect that by restricting the ability of cars to enter
+Centro, the LEZ contributes to leveling the playing field for slower
+modes, in particular cycling and walking but also transit. Transit users
+are generally close to parity city-wide (left-most bars), with nearly as
+many spatially available jobs as transit users. Still, transit mode has
+the greatest advantage in the Centro with 68.5\% of spatially available
+jobs for 51.4\% of transit users in Centro. This result makes intuitive
+sense: after car, transit is the mode with the greatest range and,
+unlike car, it faced no restrictions by the LEZ.
 
@@
 availability between modes can also be visualized at a finer level of
-granularity. Figure \ref{fig:Fig7} shows \(v_i^m\), the spatial
+spatial granularity. Figure \ref{fig:Fig7} shows \(v_i^m\), the spatial
 availability \(V_i^m\) divided by the population of users of \(m\).
@@
 transit users over most of the regions (i.e., many zones are light
-orange or light green). Spatial availability of jobs for non-motorised
-modes, in contrast, is low (under one) overall, although less so within
-LEZ Centro.
+green). In contrast, spatial availability of jobs for non-motorised
+modes is low (under one) overall, although less so within/around the LEZ
+Centro.
 
@@
 
-Incidentally, \(v_i^m\) values for car within and near LEZ Centro is
-close to or below one in Figure \ref{fig:Fig7}, while all non-car modes
-have relatively higher \(v_i^m\) values. Since these values are
-comparable across regions and over time, Figure \ref{fig:Fig7}
-potentially provides a benchmark for quantifying changes in LEZ policies
-in the future. As Figure \ref{fig:Fig7} also shows, many areas within
-the M-30 have high (white/green) \(v_i^m\) values for car, but the
-results for LEZ Centro give reasonable grounds to speculate that a
-spatial expansion of the LEZ to include all areas within the M-30 would
-likely increase the spatial availability of jobs for transit users,
-cyclists and pedestrians.
+Since \(v_i^m\) values are comparable across regions and over time,
+Figure \ref{fig:Fig7} potentially provides a benchmark for quantifying
+changes in LEZ policies in the future. Even within the M-30, \(v_i^car\)
+values are still high (over 1) for most zones. These results may give
+reasonable grounds to speculate that a spatial expansion of the LEZ to
+include all areas within the M-30 would increase the spatial
+availability of jobs for transit users, cyclists and pedestrians.
+However, further investigation is needed.
 
@@
 Aggregations of opportunities embody principles of gravitational/spatial
-interaction modelling that date back to at least H.C. Carey {[}61{]},
+interaction modelling that date back to at least H.C. Carey {[}62{]},
 and are part of a line of research that includes the work of Ravenstein
-{[}62{]}, Reilly {[}63{]}, Stewart {[}64--66{]}, Zipf {[}67,68{]},
-Wilson {[}48{]}, and many others. In this way, \(S_i^m\), \(a_i^m\), and
-yes, \(V_i^m\), can be interpreted as scores of the potential for
-interaction with opportunities in space.
+{[}63{]}, Reilly {[}64{]}, Stewart {[}65--67{]}, Zipf {[}68,69{]},
+Wilson {[}49{]}, and many others. In this way, \(S_i^m\), \(a_i^m\), and
+\(V_i^m\) can be interpreted as scores of the potential for interaction
+with opportunities in space.
 
@@
 essentially constrains the sums to match the number of opportunities in
-the region; in this way it is a singly-constrained accessibility measure
-that natively accommodates congestion and competition. The effort with
+the region: it is a singly-constrained accessibility measure that
+natively accommodates congestion and competition. The effort with
 spatial availability is in line with previous research on proportional
-allocation by Paez et al. {[}12{]}. As initially introduced by Soukhov
-et al. {[}15{]}, spatial availability was designed for a homogeneous
-population traveling by a single mode of transportation. In this paper,
-we extended spatial availability for the case of heterogeneous
-populations. We discussed this in terms of multiple modes of
-transportation, but the framework can accommodate equally well
+allocation by Páez et al.~(2019) {[}12{]}. As initially introduced by
+Soukhov et al.~(2023) {[}15{]}, spatial availability was designed for a
+homogeneous population traveling by a single mode of transportation. In
+this paper, we extended spatial availability for the case of
+heterogeneous populations. We discussed this in terms of multiple modes
+of transportation, but the framework can accommodate equally well
 variations in travel behavior by population segments.
@@
 is preserved between regions and over time. The example suggests that
-once that opportunities are treated as being finite, restrictions to
-travel by car (due to the LEZ) leave more spatially available
-opportunities for non-car users. This difference for car travel in
-locations within and immediately around the LEZ Centro seems to increase
-the number of opportunities spatially available to transit users
-(transit being the second most competitive mode), as well as the spatial
-availability from the perspective of non-motorised modes. In effect, a
-policy such as LEZ seems to help improve the spatial availability
-situation of active travel and transit in the parts of the city where it
-is implemented.
-
+once opportunities are treated as finite, it could be suggested that
+restrictions to travel by car (due to the LEZ) leave more spatially
+available opportunities for non-car users. This difference for car
+travel in locations within/around the LEZ Centro seem to increase the
+number of opportunities spatially available to transit users (transit
+being the second most competitive mode) as well as non-motorised modes.
+In effect, a policy such as LEZ appears to help improve the spatial
+availability situation of active travel and transit mode users in the
+parts of the city where it is implemented, though additional research is
+needed. To further speculate, the spatial availability allocated to
+car-users near but outside the LEZ Centro is still relatively high,
+potentially supporting the case for LEZ expansion from this perspective.
 The purpose of the empirical example is to illustrate the kind of
 insights that can be derived from the application of multimodal spatial
-availability. But there are some intriguing opportunities for future
-research. Accessibility indicators are not designed to work as modal
-split models, and yet, in the case of policies that alter the relative
-cost of various forms of transportation, one can reasonably expect to
-see some shifts between modes. In our empirical example we used data
-collected after the introduction of LEZ Centro. However, given a modal
-split model to predict model shares, accessibility indicators, including
-spatial availability, can be used to investigate changes to the
-accessibility landscape. A similar logic can be applied for destination
-choice. Our empirical example presented a snapshot of this, and in
-future research it may be interesting to investigate changes in spatial
-availability \emph{between} policy interventions. The expansion of
-Madrid's LEZ to the ring contained by the M-30 presents an excellent
-opportunity to do so. Given the intuitive interpretation of spatial
-availability scores as fractions of opportunities from the total,
-relative and absolute changes in the spatial availability landscape can
-be assessed, thus helping to evaluate the implications of policy
-interventions.
+availability.
+
+There are some intriguing opportunities for future research.
+Accessibility indicators are not designed to work as modal split models,
+and yet, in the case of policies that alter the relative cost of various
+forms of transportation, one can reasonably expect to see some shifts
+between modes. In our empirical example, we used data collected after
+the introduction of LEZ Centro. However, given a modal split model to
+predict model shares, accessibility indicators, including spatial
+availability, can be used to investigate changes to the accessibility
+landscape. A similar logic can be applied for destination choice. Our
+empirical example presented a snapshot of this, and in future research
+it may be interesting to investigate changes in spatial availability
+\emph{between} policy interventions. The plan to expand Madrid's LEZ to
+the ring contained by the M-30 presents an excellent opportunity. Given
+the intuitive interpretation of spatial availability scores as fractions
+of opportunities from the total, relative and absolute changes in the
+spatial availability landscape can be assessed, thus helping to evaluate
+the implications of policy interventions.
 
@@
 question of their impact on disadvantaged populations who have become
-mobility-restricted increasingly come to the fore {[}51,69,70{]}.
-
-It is worth reiterating that while the focus of this work was on a
-multimodal extension of spatial availability, the consideration of
-multiple modes is just one case of heterogeneous populations (i.e.,
-travel by different modes). The method itself can easily accommodate
-other forms of population or opportunity heterogeneity, for example:
-variations in rich and poor, young and old, types of opportunities. Like
-accessibility measures before it, the potential application of spatial
-availability measure are as numerous as potential study contexts.
-
-To close, spatial availability is a singly-constrained accessibility
-measure that considers competition, a way of accounting for opportunity
-congestion. Spatial availability considers competition by allocating
-opportunities (the subject of the single-constraint) to zones based on
-zonal proportions. If the demand, supply and travel to opportunities is
-spatially uniform, then competition is uniform and spatial availability
-will reflect this. It has been suggested that opportunity congestion can
-be seen on a continuum, but how it is to be considered within
+mobility-restricted increasingly come to the fore {[}52,70,71{]}.
+
+To close, in this work spatial availability considers competition by
+allocating opportunities (the subject of the single-constraint) to modal
+populations in zones based on zonal proportions. Opportunity congestion
+can be seen on a continuum, but how it is to be considered within
 accessibility research is an ongoing subject of exploration
-{[}14,17,18{]}. This is the context in which we present spatial
-availability and the multimodal extension.
+{[}14,17,18{]}. It is in this context we present spatial availability
+and its multimodal extension. While our focus was on a multiple modes,
+this consideration is just one case of heterogeneous populations (i.e.,
+travel by different modes). The multimodal method itself can easily
+accommodate other forms of population or opportunity heterogeneity, for
+example: variations in rich and poor, young and old, types of
+opportunities. Like preceding accessibility measures, the potential
+applications of spatial availability are as numerous as the potential
+study contexts.
 
@@
 \CSLLeftMargin{3. }%
-\CSLRightInline{Paez A, Scott DM, Morency C. Measuring accessibility:
+\CSLRightInline{Páez A, Scott DM, Morency C. Measuring accessibility:
 Positive and normative implementations of various accessibility
@@
 \CSLLeftMargin{12. }%
-\CSLRightInline{Paez A, Higgins CD, Vivona SF. Demand and level of
+\CSLRightInline{Páez A, Higgins CD, Vivona SF. Demand and level of
 service inflation in floating catchment area (FCA) methods. PloS one.
@@
 \CSLLeftMargin{15. }%
-\CSLRightInline{Soukhov A, Paez A, Higgins CD, Mohamed M. Introducing
+\CSLRightInline{Soukhov A, Páez A, Higgins CD, Mohamed M. Introducing
 spatial availability, a singly-constrained measure of competitive
@@
 \CSLLeftMargin{31. }%
-\CSLRightInline{Reyes M, Paez A, Morency C. Walking accessibility to
+\CSLRightInline{Reyes M, Páez A, Morency C. Walking accessibility to
 urban parks by children: A case study of montreal. Landscape and Urban
@@
 \CSLLeftMargin{36. }%
-\CSLRightInline{Paez A, Mercado RG, Farber S, Morency C, Roorda M.
+\CSLRightInline{Páez A, Mercado RG, Farber S, Morency C, Roorda M.
 Relative accessibility deprivation indicators for urban settings:
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-naturalenglandNatureNearbyAccessible2010}{}}%
+\CSLLeftMargin{44. }%
+\CSLRightInline{Natural England. Nature nearby: Accessible natural
+greenspace guidance. http://www.naturalengland.org.uk/; 2010 Mar.
+Available:
+\url{https://redfrogforum.org/wp-content/uploads/2019/11/67-Nature-Nearby\%E2\%80\%99-Accessible-Natural-Greenspace-Guidance.pdf}}
+
 \leavevmode\vadjust pre{\hypertarget{ref-OECDFrameworks2013}{}}%
-\CSLLeftMargin{44. }%
+\CSLLeftMargin{45. }%
 \CSLRightInline{OECD. Frameworks and sector policies for urban
@@
 \leavevmode\vadjust pre{\hypertarget{ref-laraSpace2015}{}}%
-\CSLLeftMargin{45. }%
+\CSLLeftMargin{46. }%
 \CSLRightInline{Lara-Valencia F, García-Pérez H. Space for equity:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-martensFair2021}{}}%
-\CSLLeftMargin{46. }%
+\CSLLeftMargin{47. }%
 \CSLRightInline{Martens K, Golub A. A fair distribution of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-taoInvestigatingImpactsPublic2020a}{}}%
-\CSLLeftMargin{47. }%
+\CSLLeftMargin{48. }%
 \CSLRightInline{Tao Z, Zhou J, Lin X, Chao H, Li G. Investigating the
@@
 \leavevmode\vadjust pre{\hypertarget{ref-wilsonFamilty1971}{}}%
-\CSLLeftMargin{48. }%
+\CSLLeftMargin{49. }%
 \CSLRightInline{Wilson AG. A family of spatial interaction models, and
@@
 \leavevmode\vadjust pre{\hypertarget{ref-carpentieriMultimodalAccessibilityPrimary2020}{}}%
-\CSLLeftMargin{49. }%
+\CSLLeftMargin{50. }%
 \CSLRightInline{Carpentieri G, Guida C, Masoumi HE. Multimodal
@@
 \leavevmode\vadjust pre{\hypertarget{ref-margaryanLowEmissionZones2021}{}}%
-\CSLLeftMargin{50. }%
+\CSLLeftMargin{51. }%
 \CSLRightInline{Margaryan S. Low emission zones and population health.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-verbeekJustManagementUrban2022}{}}%
-\CSLLeftMargin{51. }%
+\CSLLeftMargin{52. }%
 \CSLRightInline{Verbeek T, Hincks S. The {``just''} management of urban
@@
 \leavevmode\vadjust pre{\hypertarget{ref-tarrinoortizAnalyzingImpactLow2022}{}}%
-\CSLLeftMargin{52. }%
+\CSLLeftMargin{53. }%
 \CSLRightInline{Tarriño-Ortiz J, Gómez J, Soria-Lara JA, Vassallo JM.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-comunidaddemadridResultadosEDM20182020}{}}%
-\CSLLeftMargin{53. }%
+\CSLLeftMargin{54. }%
 \CSLRightInline{Comunidad de Madrid. Resultados de la {EDM} 2018 - Datos
@@
 \leavevmode\vadjust pre{\hypertarget{ref-comunidaddemadridzoniEDM20182020}{}}%
-\CSLLeftMargin{54. }%
+\CSLLeftMargin{55. }%
 \CSLRightInline{Comunidad de Madrid. Zonificación de transporte ZT208 de
@@
 \leavevmode\vadjust pre{\hypertarget{ref-lopez_2017_spatial}{}}%
-\CSLLeftMargin{55. }%
-\CSLRightInline{Lopez FA, Paez A. Spatial clustering of high-tech
+\CSLLeftMargin{56. }%
+\CSLRightInline{Lopez FA, Páez A. Spatial clustering of high-tech
 manufacturing and knowledge-intensive service firms in the greater
@@
 \leavevmode\vadjust pre{\hypertarget{ref-horbachov_theoretical_2018}{}}%
-\CSLLeftMargin{56. }%
+\CSLLeftMargin{57. }%
 \CSLRightInline{Horbachov P, Svichynskyi S. Theoretical substantiation
@@
 \leavevmode\vadjust pre{\hypertarget{ref-batista_estimation_2019}{}}%
-\CSLLeftMargin{57. }%
+\CSLLeftMargin{58. }%
 \CSLRightInline{Batista SFA, Leclercq L, Geroliminis N. Estimation of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-fitdistrplus_2015}{}}%
-\CSLLeftMargin{58. }%
+\CSLLeftMargin{59. }%
 \CSLRightInline{Delignette-Muller ML, Dutang C. {fitdistrplus}: An {R}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-reggianiAccessibilityImpedanceForms2011}{}}%
-\CSLLeftMargin{59. }%
+\CSLLeftMargin{60. }%
 \CSLRightInline{Reggiani A, Bucci P, Russo G. Accessibility and
@@
 \leavevmode\vadjust pre{\hypertarget{ref-soukhovTTS2016RDataSet2023}{}}%
-\CSLLeftMargin{60. }%
+\CSLLeftMargin{61. }%
 \CSLRightInline{Soukhov A, Páez A. {TTS}2016R: A data set to study
@@
 \leavevmode\vadjust pre{\hypertarget{ref-careyPriciples1858}{}}%
-\CSLLeftMargin{61. }%
+\CSLLeftMargin{62. }%
 \CSLRightInline{Carey HC. Principles of social science. Philadelphia:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-ravensteinLaws1889}{}}%
-\CSLLeftMargin{62. }%
+\CSLLeftMargin{63. }%
 \CSLRightInline{Ravenstein EG. The laws of migration. Journal of the
@@
 \leavevmode\vadjust pre{\hypertarget{ref-reillyMethods1929}{}}%
-\CSLLeftMargin{63. }%
+\CSLLeftMargin{64. }%
 \CSLRightInline{Reilly WJ. Methods for the study of retail
@@
 \leavevmode\vadjust pre{\hypertarget{ref-stewartInverse1941}{}}%
-\CSLLeftMargin{64. }%
+\CSLLeftMargin{65. }%
 \CSLRightInline{Stewart JQ. An inverse distance variation for certain
@@
 \leavevmode\vadjust pre{\hypertarget{ref-stewartSuggested1947}{}}%
-\CSLLeftMargin{65. }%
+\CSLLeftMargin{66. }%
 \CSLRightInline{Stewart JQ. Suggested principles of ''social physics''.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-stewartDemographic1948}{}}%
-\CSLLeftMargin{66. }%
+\CSLLeftMargin{67. }%
 \CSLRightInline{Stewart JQ. Demographic gravitation: Evidence and
@@
 \leavevmode\vadjust pre{\hypertarget{ref-zipfHypothesis1946}{}}%
-\CSLLeftMargin{67. }%
+\CSLLeftMargin{68. }%
 \CSLRightInline{Zipf GK. The p 1 p 2/d hypothesis: The case of railway
@@
 \leavevmode\vadjust pre{\hypertarget{ref-zipfIntercity1946}{}}%
-\CSLLeftMargin{68. }%
+\CSLLeftMargin{69. }%
 \CSLRightInline{Zipf GK. The P1P2/d hypothesis: On the intercity
@@
 \leavevmode\vadjust pre{\hypertarget{ref-devrijNooneVisitsMe2022}{}}%
-\CSLLeftMargin{69. }%
+\CSLLeftMargin{70. }%
 \CSLRightInline{De Vrij E, Vanoutrive T. {``No-one visits me anymore''}:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-liottaWhatDrivesInequalities2023}{}}%
-\CSLLeftMargin{70. }%
+\CSLLeftMargin{71. }%
 \CSLRightInline{Liotta C. What drives inequalities in low emission

### 2024-02-01 Journal-manuscript/Editorial correspondance/Response-to-reviewers_R2.Rmd
Revision 2 submission package

@@
 Our responses to comments to Reviewer #2 are shown below in \textcolor{blue}{blue}.
+Reviewer #3 and #1 did not have any additional comments.
 
@@
 In your hypothetical example, spatial availability would remain constant, but the system-wide cost of travel, as calculated from the intermediate values (i.e., the detailed table of opportunities allocated to origins by origin-destination pair), would decrease.}
-<!--
-Interesting and thoughtful comment.
-We have added further detail to the text to the "Discussion and Conclusions" (final paragraph) to further shed like on 'competition' and make spatial availability's use case more clear in the closure of the manuscript.
-}
-\textcolor{blue}{In the scenario described, a multiple mode (equal travel impedance) in a uniformly distributed population/opportunity landscape, will yield spatial availability before- and after- values that are equal.
-This is because competition has not changed: spatial availability allocates 'opportunities' in a proportional way based on competition.
-Competition is a product of the relative travel time (between zones) as well as the supply and demand of population and opportunities, so if those relative differences remained unchanged between interventions then spatial availability remains unchanged as well.
-Though what would decrease are the system-wide transport costs, and those can be reflected without an accessibility measure.
-Hopefully the addition of the last paragraph in the discussion clarifies this point about competition considerations within accessibility measures.
-}
--->
 

### 2024-02-07 Journal-manuscript/Journal-manuscript.Rmd
Version submitted to (journal) production

@@
 
-keywords: ["Spatial availability", "Multimodal", "Singly-constrained", "Accessibility", "Competitive", "Madrid", "Reproducible research"]
+keywords: ["Spatial availability", "Multimodal", "Singly-constrained", "Accessibility", "Competition", "Madrid", "Reproducible research"]
 
@@
   echo = FALSE,
-  cache = TRUE,
+  cache = FALSE,
   warning = FALSE,
@@
             ELE_G_POND_ESC2,
-            mode = "transit",
+            mode = "Transit",
             VORIZT1259,
@@
             ELE_G_POND_ESC2,
-            mode = "car/motor",
+            mode = "Car/motor",
             VORIZT1259,
@@
             ELE_G_POND_ESC2,
-            mode = "bike",
+            mode = "Bike",
             VORIZT1259,
@@
             ELE_G_POND_ESC2,
-            mode = "walk",
+            mode = "Walk",
             VORIZT1259,
@@
                     HW_transit_gamma$estimate[2]),
-         mode = "transit")
+         mode = "Transit")
 
@@
                     rate = HW_car_motor_gamma$estimate[2]),
-         mode = "car/motor")
+         mode = "Car/motor")
 
@@
                     HW_bike_lnorm$estimate[2]),
-         mode = "bike")
+         mode = "Bike")
 
@@
                     HW_walk_lnorm$estimate[2]),
-         mode = "walk")
+         mode = "Walk")
 ```
@@
   mutate(mode = factor(mode,
-                       levels = c("bike", "walk", "car/motor", "transit")),
+                       levels = c("Bike", "Walk", "Car/motor", "Transit")),
          distribution = "empirical")
@@
   mutate(mode = factor(mode,
-                       levels = c("bike", "walk", "car/motor", "transit")))
+                       levels = c("Bike", "Walk", "Car/motor", "Transit")))
 ```
@@
   mutate(mode = factor(mode,
-                       levels = c("bike", "walk", "car/motor", "transit"))) %>%
+                       levels = c("Bike", "Walk", "Car/motor", "Transit"))) %>%
   ggplot(aes(x = TIEMPO_VIAJES, after_stat(density),
@@
 OD_v_HW <- OD_v_HW  %>% 
-  mutate(mode = case_when(MODO_PRIORITARIO <= 9 ~ "transit",
-                          MODO_PRIORITARIO <= 19 & MODO_PRIORITARIO >= 10 ~ "car/motor",
-                          MODO_PRIORITARIO <= 23 & MODO_PRIORITARIO >= 20 ~ "bike",
-                          MODO_PRIORITARIO == 24 ~ "walk"),
+  mutate(mode = case_when(MODO_PRIORITARIO <= 9 ~ "Transit",
+                          MODO_PRIORITARIO <= 19 & MODO_PRIORITARIO >= 10 ~ "Car/motor",
+                          MODO_PRIORITARIO <= 23 & MODO_PRIORITARIO >= 20 ~ "Bike",
+                          MODO_PRIORITARIO == 24 ~ "Walk"),
          mode = factor(mode,
-                       levels = c("bike", "walk", "car/motor", "transit")))
+                       levels = c("Bike", "Walk", "Car/motor", "Transit")))
 
@@
 OD_3 <- OD_3 %>%
-  mutate(f = case_when(mode == "transit" ~ dgamma(tt,
+  mutate(f = case_when(mode == "Transit" ~ dgamma(tt,
                                                   HW_transit_gamma$estimate["shape"], 
                                                   HW_transit_gamma$estimate["rate"]),
-                       mode == "car/motor" ~ dgamma(tt,
+                       mode == "Car/motor" ~ dgamma(tt,
                                                     HW_car_motor_gamma$estimate["shape"],
                                                     HW_car_motor_gamma$estimate["rate"]),
-                       mode == "bike" ~ dlnorm(tt,
+                       mode == "Bike" ~ dlnorm(tt,
                                                HW_bike_lnorm$estimate["meanlog"],
                                                HW_bike_lnorm$estimate["sdlog"]),
-                       mode == "walk" ~ dlnorm(tt, 
+                       mode == "Walk" ~ dlnorm(tt, 
                                                HW_walk_lnorm$estimate["meanlog"], 
@@
           color = NA) +
-
   scale_fill_viridis_c(option = "mako", #legend scale bar
@@
   annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
-                   height = unit(0.15, "cm")) +
+                   height = unit(0.10, "cm")) +
+  annotation_north_arrow(location = "tr", which_north = "true",  height = unit(0.75, "cm"), width = unit(0.75, "cm"), pad_y = unit(0, "cm"), style = north_arrow_fancy_orienteering(text_size=7)) +
   theme_void() +
@@
   annotation_scale(bar_cols = c("dimgrey", "white"), # scale bar for both the main plot and inset
-                   height = unit(0.15, "cm")) +
+                   height = unit(0.2, "cm")) +
+  annotation_north_arrow(location = "tr", which_north = "true",  height = unit(1, "cm"), width = unit(1, "cm"), pad_y = unit(0.4, "cm"), pad_x = unit(1, "cm"), style = north_arrow_fancy_orienteering(text_size=7)) +
   theme_void() +
@@
 # here we example the relationship between v_im between modes -- I can't tell anything conclusive.
-car <- SA_im_zn208 %>% filter(mode=="car/motor") %>% st_drop_geometry() %>% select(VORI_CD_ZT208,v_im) %>% 
+car <- SA_im_zn208 %>% filter(mode=="Car/motor") %>% st_drop_geometry() %>% select(VORI_CD_ZT208,v_im) %>% 
   rename("v_im_car" = "v_im")
-bike <- SA_im_zn208 %>% filter(mode=="bike") %>% st_drop_geometry() %>% select(VORI_CD_ZT208,v_im) %>% 
+bike <- SA_im_zn208 %>% filter(mode=="Bike") %>% st_drop_geometry() %>% select(VORI_CD_ZT208,v_im) %>% 
   rename("v_im_bike" = "v_im")
-transit <- SA_im_zn208 %>% filter(mode=="transit") %>% st_drop_geometry() %>% select(VORI_CD_ZT208,v_im) %>% 
+transit <- SA_im_zn208 %>% filter(mode=="Transit") %>% st_drop_geometry() %>% select(VORI_CD_ZT208,v_im) %>% 
   rename("v_im_transit" = "v_im") 
-walk <- SA_im_zn208 %>% filter(mode=="walk")%>% st_drop_geometry() %>% select(VORI_CD_ZT208,v_im) %>% 
+walk <- SA_im_zn208 %>% filter(mode=="Walk")%>% st_drop_geometry() %>% select(VORI_CD_ZT208,v_im) %>% 
   rename("v_im_walk" = "v_im")
@@
 flextable::set_caption(table_toy_format_flex, 
-            caption = "Accessibility values at each origin (i) per mode (m) (columns three to five) and aggregated per i (columns six and seven) for the synthetic example.")
+            caption = "Accessibility values (S, a, V) at each origin (i) for the synthetic example. Displayed per mode (m) (columns three to five) and aggregated per i (columns six and seven).")
 ```
@@
 ```{r}
-tld_empirical_car <- tld_empirical %>% filter(mode == "car/motor") 
-tld_empirical_transit <- tld_empirical %>% filter(mode == "transit") 
-tld_empirical_bike <- tld_empirical %>% filter(mode == "bike") 
-tld_empirical_walk <- tld_empirical %>% filter(mode == "walk")
+tld_empirical_car <- tld_empirical %>% filter(mode == "Car/motor") 
+tld_empirical_transit <- tld_empirical %>% filter(mode == "Transit") 
+tld_empirical_bike <- tld_empirical %>% filter(mode == "Bike") 
+tld_empirical_walk <- tld_empirical %>% filter(mode == "Walk")
 
@@
 
-corr_m_pops <- merge(corr_m_pops, SA_im_zn208 %>% st_drop_geometry() %>% filter(mode == "bike") %>% select(c(VORI_CD_ZT208, population_m)), by="VORI_CD_ZT208", all = TRUE) %>% rename("population_m_bike" = "population_m")
-
-corr_m_pops <- merge(corr_m_pops, SA_im_zn208 %>% st_drop_geometry() %>% filter(mode == "transit") %>% select(c(VORI_CD_ZT208, population_m)), by="VORI_CD_ZT208", all = TRUE) %>% rename("population_m_transit" = "population_m")
-
-corr_m_pops <- merge(corr_m_pops, SA_im_zn208 %>% st_drop_geometry() %>% filter(mode == "walk") %>% select(c(VORI_CD_ZT208, population_m)), by="VORI_CD_ZT208", all = TRUE) %>% rename("population_m_walk" = "population_m")
-
-corr_m_pops <- merge(corr_m_pops, SA_im_zn208 %>% st_drop_geometry() %>% filter(mode == "car/motor") %>% select(c(VORI_CD_ZT208, population_m)), by="VORI_CD_ZT208", all = TRUE) %>% rename("population_m_car" = "population_m")
+corr_m_pops <- merge(corr_m_pops, SA_im_zn208 %>% st_drop_geometry() %>% filter(mode == "Bike") %>% select(c(VORI_CD_ZT208, population_m)), by="VORI_CD_ZT208", all = TRUE) %>% rename("population_m_bike" = "population_m")
+
+corr_m_pops <- merge(corr_m_pops, SA_im_zn208 %>% st_drop_geometry() %>% filter(mode == "Transit") %>% select(c(VORI_CD_ZT208, population_m)), by="VORI_CD_ZT208", all = TRUE) %>% rename("population_m_transit" = "population_m")
+
+corr_m_pops <- merge(corr_m_pops, SA_im_zn208 %>% st_drop_geometry() %>% filter(mode == "Walk") %>% select(c(VORI_CD_ZT208, population_m)), by="VORI_CD_ZT208", all = TRUE) %>% rename("population_m_walk" = "population_m")
+
+corr_m_pops <- merge(corr_m_pops, SA_im_zn208 %>% st_drop_geometry() %>% filter(mode == "Car/motor") %>% select(c(VORI_CD_ZT208, population_m)), by="VORI_CD_ZT208", all = TRUE) %>% rename("population_m_car" = "population_m")
 
@@
 The multimodal method itself can easily accommodate other forms of population or opportunity heterogeneity, for example: variations in rich and poor, young and old, types of opportunities.
-Like preceding accessibility measures, the potential applications of spatial availability are as numerous as the potential study contexts.
+Similar to preceding accessibility measures, the applications of spatial availability are as numerous as potential study contexts.
+To encourage open and reproducible science, the manuscript and all associated analysis is available within the lead author's [GitHub repository](https://github.com/soukhova/Multimodal-spatial-availability).
 
@@
 <!-- All work is fully-reproducible and available within this [GitHub repository](https://github.com/soukhova/Multimodal-spatial-availability). -->
-
-# Author contributions
-The authors confirm contribution to the paper as follows: study conception and design: AS, JTO, JSL, AP.; data collection: AS, JTO, JSL.; analysis and interpretation of results: AS, JTO, JSL, AP.; draft manuscript preparation: AS, JSL, AP.
-All authors reviewed the results and approved the final version of the manuscript.
 

### 2024-02-07 Journal-manuscript/Journal-manuscript.tex
Version submitted to (journal) production

@@
 
-\caption{Accessibility\ values\ at\ each\ origin\ (i)\ per\ mode\ (m)\ (columns\ three\ to\ five)\ and\ aggregated\ per\ i\ (columns\ six\ and\ seven)\ for\ the\ synthetic\ example.}\\
+\caption{Accessibility\ values\ (S,\ a,\ V)\ at\ each\ origin\ (i)\ for\ the\ synthetic\ example.\ Displayed\ per\ mode\ (m)\ (columns\ three\ to\ five)\ and\ aggregated\ per\ i\ (columns\ six\ and\ seven).}\\
 
@@
 
-\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{i}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{m}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{S}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}} \\
-
-\ascline{1.5pt}{666666}{1-7}\endfirsthead \caption[]{Accessibility\ values\ at\ each\ origin\ (i)\ per\ mode\ (m)\ (columns\ three\ to\ five)\ and\ aggregated\ per\ i\ (columns\ six\ and\ seven)\ for\ the\ synthetic\ example.}\\
+\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{i}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{m}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{S}}\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textsubscript{i}}}\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textsuperscript{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textsubscript{i}}}\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textsuperscript{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textsubscript{i}}}\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textsuperscript{m}}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textsubscript{i}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textsubscript{i}}}} \\
+
+\ascline{1.5pt}{666666}{1-7}\endfirsthead \caption[]{Accessibility\ values\ (S,\ a,\ V)\ at\ each\ origin\ (i)\ for\ the\ synthetic\ example.\ Displayed\ per\ mode\ (m)\ (columns\ three\ to\ five)\ and\ aggregated\ per\ i\ (columns\ six\ and\ seven).}\\
 
@@
 
-\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{i}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{m}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{S}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}\textcolor[HTML]{000000}{\textsuperscript{\fontsize{11}{11}\selectfont{m}}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\textsubscript{\fontsize{11}{11}\selectfont{i}}}} \\
+\multicolumn{1}{>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{i}}} & \multicolumn{1}{>{\raggedright}m{\dimexpr 0.41in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{m}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{S}}\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textsubscript{i}}}\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textsuperscript{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textsubscript{i}}}\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textsuperscript{m}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textsubscript{i}}}\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textsuperscript{m}}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedleft}m{\dimexpr 0.58in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{a}}\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textsubscript{i}}}} & \multicolumn{1}{>{\raggedleft}m{\dimexpr 1.05in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{V}}\textcolor[HTML]{000000}{\fontsize{11}{11}\selectfont{\textsubscript{i}}}} \\
 
@@
 example: variations in rich and poor, young and old, types of
-opportunities. Like preceding accessibility measures, the potential
-applications of spatial availability are as numerous as the potential
-study contexts.
-
-\hypertarget{author-contributions}{%
-\section{Author contributions}\label{author-contributions}}
-
-The authors confirm contribution to the paper as follows: study
-conception and design: AS, JTO, JSL, AP.; data collection: AS, JTO,
-JSL.; analysis and interpretation of results: AS, JTO, JSL, AP.; draft
-manuscript preparation: AS, JSL, AP. All authors reviewed the results
-and approved the final version of the manuscript.
+opportunities. Similar to preceding accessibility measures, the
+applications of spatial availability are as numerous as potential study
+contexts. To encourage open and reproducible science, the manuscript and
+all associated analysis is available within the lead author's
+\href{https://github.com/soukhova/Multimodal-spatial-availability}{GitHub
+repository}.
 

### 2024-02-07 Presentations/CIT-Presentation.Rmd
Re-organized repo, checked to make sure all .rmds run. Updated lockfile to reflect current state of package dependencies.

--- 2023-07-30 NECTAR-Coimbra-Presentation.Rmd
@@
 ---
-title: "Quantifying multi-modal competition using spatial availability"
-subtitle: "NECTAR - Sustainable Neighbourhoods"
+title: "Multi-modal spatial availability: a research proposal"
+subtitle: "CIT - Tenerife 2023"
 author: "Anastasia Soukhov, J. Tarriño Ortiz, J. A. Soria-Lara, A. Páez"
-date: "21-06-2023"
+date: "15-06-2023"
 output:
@@
     nature:
-      slideNumberFormat: "%current% / 17" 
+      slideNumberFormat: "%current% / 12" 
       highlightStyle: github
@@
 library(kableExtra)
-library("ggpmisc")
 options(scipen = 999)
@@
                            100, 100, 15, 80, 80, 10), 
-                    pop = c(16666.66666, 16666.66666, 16666.66666, 33333.33333, 33333.33333, 33333.33333,
-                            60000, 60000,60000, 90000, 90000, 90000,
+                    pop = c(20000, 20000, 20000, 30000, 30000, 30000,
+                            50000, 50000,50000, 100000, 100000, 100000,
                             3000, 3000, 3000, 7000,7000, 7000),
@@
 od_table <- data.frame(Center = c("A", "B", "C", "1", "2", "3"),
-  Size = c("50,000 pop.", "150,000 pop.", "10,000 pop.", 
+  Size = c("50,000 pop (40% pt)", "150,000 pop (33% pt)", "10,000 pop (30% pt)", 
            "100,000 jobs", "100,000 jobs", "10,000 jobs"))
@@
   annotate(geom = "label", 
-           x=c(1.6, 1.7, 7.5),
-           y=c(9.1, 0.9, 5.2), 
+           x=c(0.5, 0.6, 7),
+           y=c(9, 1, 5.2), 
+           label = c("Suburban", "Urban", "Satellite"), 
            size=4, 
@@
              color = "black") +
+  # Text on curves
+  geom_textcurve(data = od_lines,
+               aes(x = x,
+                   y = y,
+                   xend = xend,
+                   yend = yend,
+                   linetype = group,
+                   group = group,
+                   label = time),
+               size = 3.5,
+             curvature = 0.20,
+             color = "black") +
   # Plot origins and destinations
@@
                 shape = type)) +
-  scale_size(range = c(6, 16)) +
+  scale_size(range = c(4, 10)) +
   # Label origins and destinations
@@
   # Create table
-  annotate(geom = 'table',
+  annotate(geom = "table",
            x = 9, 
@@
            label = list(od_table),
-           size=4) +
+           fill = "white",
+           size=3) +
   # Theme the plot
@@
 toy_example_plot
-ggsave("Presentations/toy-example1.png",dpi=600)
+ggsave("Presentation-images/toy-example.png",dpi=600)
 ```
-
-## Accessibility competition between modes
-
-**Motivation**: interpreting accessibility measures (location-based _potential interaction_) for policy interventions
-
-```{r, toy-example0, out.width='40%', fig.align='center', echo=FALSE}
-knitr::include_graphics("Presentations/access-demo.png")
-```
-
----
-## Accessibility competition between modes
-
-**Motivation**: interpreting accessibility measures (location-based _potential interaction_) for policy interventions
-
-```{r, toy-example00, out.width='40%', fig.align='center', echo=FALSE}
-knitr::include_graphics("Presentations/access-demo.png")
-```
-
-### Overview
-
-**1.** Types of current accessibility measures and their limitations
-
-**2.** Multi-modal **spatial availability**
-
-**3.** Planning accessibility policy interventions considering multi-modal competition
-
----
 class: center, middle
-## Classification of multi-modal accessibility measures
+## Overview of multi-modal accessibility measures
 
@@
 ???
+<!-- for the title slide --> 
+In this presentation I will speak on how current accessibility measures _miss_ considering competition
+**1.** Limitations of current accessibility measures
+**2.** Explaining **spatial availability** with a multi-modal extension
+**3.** Policy scenario planning applications - equity
 
@@
 ---
+class: center
+### Simulated example: uncompetitive and unconstrained
+
+```{r, toy-example0, out.width='80%', echo=FALSE}
+knitr::include_graphics("Presentation-images/toy-example.png")
+```
+
+---
+count: false
 
@@
 ```{r, toy-example1, out.width='120%', echo=FALSE}
-knitr::include_graphics("Presentations/toy-example1.png")
+knitr::include_graphics("Presentation-images/toy-example.png")
 ```
@@
 ```{r, out.width='120%',echo=FALSE}
-knitr::include_graphics("Presentations/toy-example1.png")
+knitr::include_graphics("Presentation-images/toy-example.png")
 ```
@@
 ##### Shen-type accessibility
-$a_i^m = \sum_j \frac{O_j f^m(c_{ij}^m)}{\sum_m \sum_i P_{i}^m f^m(c_{ij}^m)}$
+$a_i^m = \sum_j \frac{O_j f^m(c_{ij}^m)}{\sum_m \sum_i P_{i}^m f^m(c_{ij}^m)^m}$
 
@@
 ---
+count: false
 ### Simulated example: competitive and unconstrained
@@
 
-##### Spatial availability per origin and mode
+##### Spatial availability
 $V^m_{i} = \sum_{j} O_j\ F^{t,m}_{ij}$
@@
 ---
-
+count: false
 ### Simulated example: competitive _and_ constrained
@@
 ```{r,out.width='120%', echo=FALSE}
-knitr::include_graphics("Presentations/toy-example1.png")
+knitr::include_graphics("Presentation-images/toy-example.png")
 ```
@@
 
-##### Spatial availability per origin, per mode
+##### Spatial availability
 $V_{i} = \sum_{m} V^m_{i}$
@@
 ```{r, echo=FALSE, message=FALSE}
-knitr::kable(V,
+knitr::kable(V %>% dplyr::select(-v),
              format = "html",
@@
 ```{r, echo=FALSE, out.width='80%', fig.align='center'}
-knitr::include_graphics("Presentations/madrid-region.png")
+knitr::include_graphics("Presentation-images/madrid-region.png")
 ```
@@
 ```{r, echo=FALSE, out.width='50%', fig.align='center'}
-knitr::include_graphics("Presentations/city-centre.png")
+knitr::include_graphics("Presentation-images/city-centre.png")
 ```
@@
 ```{r, echo=FALSE, out.width='70%',fig.align='center'}
-knitr::include_graphics("Presentations/im_populations_zn208_plot.png")
-```
-Worker population, by mode, at resolution Zona 208
-
----
-class: center
-### Multi-modal home-to-work accessibility in Madrid: 2018 travel survey
-```{r, echo=FALSE, out.width='70%',fig.align='center'}
-knitr::include_graphics("Presentations/Hansen_im_V_zn208_plot.png")
+knitr::include_graphics("Presentation-images/Hansen_im_V_zn208_plot.png")
 ```
@@
 ```{r, echo=FALSE, out.width='70%',fig.align='center'}
-knitr::include_graphics("Presentations/SA_im_V_zn208_plot.png")
+knitr::include_graphics("Presentation-images/SA_im_V_zn208_plot.png")
 ```
@@
 ```{r, echo=FALSE, out.width='70%', fig.align='center'}
-knitr::include_graphics("Presentations/SA_i_V_zn208_plot.png")
+knitr::include_graphics("Presentation-images/SA_i_V_zn208_plot.png")
 ```
@@
 
-- **Scenario evaluation**: accessibility when...
+- Scenario evaluation: accessibility when...
   - LEZ boundaries change
@@
   - Travel behaviour changes
-  - Spatial unit of analysis changes
 
---
-
-- **Questions of equity**: which neighbourhoods, what people, what modes?
+- Questions of equity: which neighbourhoods, what people, what modes?
   - GHG emission reductions
   - Advantages to non-car users
-  - Who are impacted disproportionally
+  - Who is impacted disproportionally
 
@@
 ```{r, echo=FALSE, out.width="20%"}
-knitr::include_graphics("Presentations/MJ_logo.png")
+knitr::include_graphics("Presentation-images/MJ_logo.png")
 ```
@@
 ```{r, echo=FALSE, out.width="30%"}
-knitr::include_graphics("Presentations/mcm-sci-sees_left-col.png")
+knitr::include_graphics("Presentation-images/mcm-sci-sees_left-col.png")
 ```
@@
 ```{r, echo=FALSE, out.width="10%"}
-knitr::include_graphics("Presentations/spatial-availability.png")
+knitr::include_graphics("Presentation-images/spatial-availability.png")
 ```

### 2024-02-07 Presentations/NECTAR-Coimbra-Presentation.Rmd
Re-organized repo, checked to make sure all .rmds run. Updated lockfile to reflect current state of package dependencies.

--- 2023-07-30 NECTAR-Coimbra-Presentation.Rmd
@@
 toy_example_plot
-ggsave("Presentations/toy-example1.png",dpi=600)
+ggsave("Presentation-images/toy-example1.png",dpi=600)
 ```
@@
 ```{r, toy-example0, out.width='40%', fig.align='center', echo=FALSE}
-knitr::include_graphics("Presentations/access-demo.png")
+knitr::include_graphics("Presentation-images/access-demo.png")
 ```
@@
 ```{r, toy-example00, out.width='40%', fig.align='center', echo=FALSE}
-knitr::include_graphics("Presentations/access-demo.png")
+knitr::include_graphics("Presentation-images/access-demo.png")
 ```
@@
 ```{r, toy-example1, out.width='120%', echo=FALSE}
-knitr::include_graphics("Presentations/toy-example1.png")
+knitr::include_graphics("Presentation-images/toy-example1.png")
 ```
@@
 ```{r, out.width='120%',echo=FALSE}
-knitr::include_graphics("Presentations/toy-example1.png")
+knitr::include_graphics("Presentation-images/toy-example1.png")
 ```
@@
 ```{r,out.width='120%', echo=FALSE}
-knitr::include_graphics("Presentations/toy-example1.png")
+knitr::include_graphics("Presentation-images/toy-example1.png")
 ```
@@
 ```{r, echo=FALSE, out.width='70%',fig.align='center'}
-knitr::include_graphics("Presentations/im_populations_zn208_plot.png")
+knitr::include_graphics("Presentation-images/im_populations_zn208_plot.png")
 ```
@@
 ```{r, echo=FALSE, out.width='70%',fig.align='center'}
-knitr::include_graphics("Presentations/Hansen_im_V_zn208_plot.png")
+knitr::include_graphics("Presentation-images/Hansen_im_V_zn208_plot.png")
 ```

### 2024-02-07 TRB/TRB-Poster.Rmd
Re-organized repo, checked to make sure all .rmds run. Updated lockfile to reflect current state of package dependencies.

--- 2024-01-25 TRB-Poster.Rmd
@@
 column_numbers: 4
-logoright_name: Presentation-images/acknowledgements-combo.png
-logoleft_name: Presentation-images/soukhova-LinkTree.png
+logoright_name: images/acknowledgements-combo.png
+logoleft_name: images/soukhova-LinkTree.png
 sectitle_textsize: "40pt"
@@
 link-citations: true
-bibliography: bibliography/bibliography.bib
+bibliography:
+    - "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
 output: 
@@
 
-quick_html(ht, file = "Presentation-images/synth_ex_table.html", open = FALSE)
+quick_html(ht, file = "images/synth_ex_table.html", open = FALSE)
 # higher zoom increases resolution and image size
-webshot::webshot(url = "Presentation-images/synth_ex_table.html", file = "Presentation-images/synth_ex_table.png",
+webshot::webshot(url = "images/synth_ex_table.html", file = "images/synth_ex_table.png",
                  zoom = 5, selector = "table")
@@
   # place the table
-    annotation_custom(rasterGrob(png::readPNG(source="Presentation-images/synth_ex_table.png")),
+    annotation_custom(rasterGrob(png::readPNG(source="images/synth_ex_table.png")),
                       xmin= 5.5, xmax=Inf, 
@@
 toy_example_plot
-ggsave("Presentation-images/toy-example.png",dpi=600)
+ggsave("images/toy-example.png",dpi=600)
 ```
@@
 
-![_Figure 1: Calculated Hansen $S$, Shen-type $a$ and Spatial Availability $V$ assuming an impedance function of $f(c_{ij})^{\text{x}} = f(c_{ij})^{\text{z}} =exp(-0.1 \cdot c_{ij}^m)$ for the multimodal synthetic population._](Presentation-images/toy-example.png){#id .class width=100%}
+![_Figure 1: Calculated Hansen $S$, Shen-type $a$ and Spatial Availability $V$ assuming an impedance function of $f(c_{ij})^{\text{x}} = f(c_{ij})^{\text{z}} =exp(-0.1 \cdot c_{ij}^m)$ for the multimodal synthetic population._](images/toy-example.png){#id .class width=100%}
 
@@
 Ranges of values in the legend are quintiles.
-The TAZ shapefile is available from the Community of Madrid open data portal._](Presentation-images/Fig2.png){#id .class width=100%}
+The TAZ shapefile is available from the Community of Madrid open data portal._](images/Fig2.png){#id .class width=100%}
 
-![_Figure 3: Spatial availability of jobs per origin and mode $V_i^m$._](Presentation-images/Fig5.png){#id .class width=98%}
+![_Figure 3: Spatial availability of jobs per origin and mode $V_i^m$._](images/Fig5.png){#id .class width=98%}
 
@@
 
-![_Figure 4: Proportion of population by mode and spatial availability of jobs by mode ($V_i^m$) aggregated for the city of Madrid (All), the area within the M-30 highway, and within the Centro._](Presentation-images/Fig6.png){#id .class width=100%}
+![_Figure 4: Proportion of population by mode and spatial availability of jobs by mode ($V_i^m$) aggregated for the city of Madrid (All), the area within the M-30 highway, and within the Centro._](images/Fig6.png){#id .class width=100%}
 
@@
 
-![_Figure 5: Distribution of spatially available jobs per capita by mode of transportatin ($v_i^m$)._](Presentation-images/Fig7.png){#id .class width=98%}
+![_Figure 5: Distribution of spatially available jobs per capita by mode of transportatin ($v_i^m$)._](images/Fig7.png){#id .class width=98%}
 
